---
title: "The Quantitative Research Journey"
author: "Jinsong Chen"
bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom
references:
  - id: hernan2020causal
    type: book
    title: "Causal inference: What if"
    author:
      - family: Hernán
        given: Miguel A.
      - family: Robins
        given: James M.
    issued:
      date-parts: [[2020]]
    publisher: Chapman & Hall/CRC
    URL: https://miguelhernan.org/whatifbook
  - id: angrist1996iv
    type: article-journal
    title: "Identification of causal effects using instrumental variables"
    author:
      - family: Angrist
        given: Joshua D.
      - family: Imbens
        given: Guido W.
      - family: Rubin
        given: Donald B.
    issued:
      date-parts: [[1996]]
    container-title: Journal of the American Statistical Association
    volume: 91
    issue: 434
    page: 444-455
    DOI: 10.1080/01621459.1996.10476902
  - id: box1976
    type: article-journal
    title: "Science and statistics"
    author:
      - family: Box
        given: George E. P.
    issued:
      date-parts: [[1976]]
    container-title: Journal of the American Statistical Association
    volume: 71
    issue: 356
    page: 791-799
    DOI: 10.1080/01621459.1976.10480949
  - id: box1979
    type: chapter
    title: "Robustness in the strategy of scientific model building"
    author:
      - family: Box
        given: George E. P.
    editor:
      - family: Launer
        given: Robert L.
      - family: Wilkinson
        given: Graham N.
    issued:
      date-parts: [[1979]]
    container-title: Robustness in Statistics
    publisher: Academic Press
    publisher-place: New York
    page: 201-236
  - id: bloom2015wfh
    type: article-journal
    title: "Does working from home work? Evidence from a Chinese experiment"
    author:
      - family: Bloom
        given: Nicholas
      - family: Liang
        given: James
      - family: Roberts
        given: John
      - family: Ying
        given: Zhichun Jenny
    issued:
      date-parts: [[2015]]
    container-title: The Quarterly Journal of Economics
    volume: 130
    issue: 1
    page: 165-218
    DOI: 10.1093/qje/qju032
  - id: realinho2022
    type: article-journal
    title: "Predicting student dropout and academic success"
    author:
      - family: Realinho
        given: Valentim
      - family: Machado
        given: Jorge
      - family: Baptista
        given: Luís
      - family: Martins
        given: Mónica V.
    issued:
      date-parts: [[2022]]
    container-title: Data
    volume: 7
    issue: 11
    page: 146
    DOI: 10.3390/data7110146
  - id: uci697
    type: dataset
    title: "Predict students' dropout and academic success"
    author:
      - family: Realinho
        given: Valentim
      - family: Vieira Martins
        given: Mónica
      - family: Machado
        given: Jorge
      - family: Baptista
        given: Luís
    issued:
      date-parts: [[2021]]
    publisher: UCI Machine Learning Repository
    DOI: 10.24432/C5MC89
  - id: bates2015lmm
    type: article-journal
    title: "Fitting linear mixed-effects models using lme4"
    author:
      - family: Bates
        given: Douglas
      - family: Mächler
        given: Martin
      - family: Bolker
        given: Ben
      - family: Walker
        given: Steve
    issued:
      date-parts: [[2015]]
    container-title: Journal of Statistical Software
    volume: 67
    issue: 1
    page: 1-48
    DOI: 10.18637/jss.v067.i01
execute:
  echo: false
  warning: false
  message: false
  cache: false
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    documentclass: scrartcl
    fontsize: 11pt
    linestretch: 1.06
    geometry: margin=1in
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    keep-tex: false
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          % Preserve the imported PDF's transparency group in an outer Form XObject.
          % Exact-token checks fail on an unexpected driver so the patch can be reviewed.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup\clearpage}
          \usepackage{caption}
          \captionsetup[table]{position=above,format=plain,justification=raggedright,singlelinecheck=false,labelsep=newline,labelfont=bf,textfont=it}
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \renewcommand{\arraystretch}{1.1}
          \usepackage{float}
          \usepackage{needspace}
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
  html:
    toc: true
    toc-depth: 3
    number-sections: true
editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup-journey
#| include: false
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")

# Student outcomes (Realinho et al., 2022; UCI dataset 697, CC BY 4.0).
# realinho2022_students.csv keeps three source fields in the original row
# order; the refresh chunk below documents its derivation from the release.
students <- read.csv("realinho2022_students.csv", stringsAsFactors = FALSE)
stopifnot(identical(names(students), c("recorded_outcome", "admission_grade", "age")),
          nrow(students) == 4424L, !anyNA(students),
          identical(as.integer(table(students$recorded_outcome)), c(1421L, 794L, 2209L)),
          abs(sum(students$admission_grade) - 561751.2) < 1e-6,
          sum(students$age) == 102925L)
students$dropout <- as.integer(students$recorded_outcome == "Dropout")
# Simplified analysis: stratified 70% training split, one logistic rule,
# and a held-out comparison with the training-prevalence rule. The RNG
# kind is pinned so the split does not depend on session settings.
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
set.seed(12345)
train_rows <- sort(c(
  sample(which(students$dropout == 0), floor(.70 * sum(students$dropout == 0))),
  sample(which(students$dropout == 1), floor(.70 * sum(students$dropout == 1)))
))
train <- students[train_rows, ]
test <- students[-train_rows, ]
record_counts <- function(x)
  table(factor(x$recorded_outcome, levels = c("Dropout", "Enrolled", "Graduate")))
student_counts <- rbind(full = record_counts(students),
                        training = record_counts(train), test = record_counts(test))
stopifnot(identical(as.integer(student_counts["training", ]), c(994L, 562L, 1540L)),
          identical(as.integer(student_counts["test", ]), c(427L, 232L, 669L)))
dropout_fit <- glm(dropout ~ admission_grade + age, data = train, family = binomial())
stopifnot(dropout_fit$converged)
dropout_probability <- predict(dropout_fit, newdata = test, type = "response")
brier <- c(logistic = mean((dropout_probability - test$dropout)^2),
           training_prevalence = mean((mean(train$dropout) - test$dropout)^2))
dropout_identified <- mean(dropout_probability[test$dropout == 1] >= .5)
stopifnot(abs(brier["logistic"] - .200011325828) < 1e-9,
          abs(brier["training_prevalence"] - .218150879563) < 1e-9,
          brier["logistic"] < brier["training_prevalence"],
          abs(dropout_identified - 73 / 427) < 1e-12, dropout_identified < .5)


```

```{r}
#| label: refresh-student-record
#| eval: false
#| include: false
# Maintenance only: rebuild the canonical instructor/chapter_01 CSV from
# the public release. Run interactively from this folder with a network
# connection, then use instructor/tools/sync_data.py to refresh delivery copies.
# Source ZIP MD5 287bbb2a21cf094a13d7f53c805b61c2;
# SHA-256 e90e55fd65ec462ae283ebeb2cca409319e3460ed898d8754f62fb35cc83a65d.
zip_file <- tempfile(fileext = ".zip")
download.file(paste0("https://archive.ics.uci.edu/static/public/697/",
                     "predict%2Bstudents%2Bdropout%2Band%2Bacademic%2Bsuccess.zip"),
              zip_file, mode = "wb")
stopifnot(unname(tools::md5sum(zip_file)) == "287bbb2a21cf094a13d7f53c805b61c2")
raw <- read.csv(unz(zip_file, "data.csv"), sep = ";",
                check.names = FALSE, encoding = "UTF-8")
unlink(zip_file)
stopifnot(nrow(raw) == 4424L, ncol(raw) == 37L)
rebuilt <- data.frame(recorded_outcome = raw$Target,
                      admission_grade = raw[["Admission grade"]],
                      age = raw[["Age at enrollment"]])
stopifnot(!anyNA(rebuilt),
          identical(sort(unique(rebuilt$recorded_outcome)),
                    c("Dropout", "Enrolled", "Graduate")),
          identical(as.integer(table(rebuilt$recorded_outcome)),
                    c(1421L, 794L, 2209L)),
          abs(sum(rebuilt$admission_grade) - 561751.2) < 1e-6,
          sum(rebuilt$age) == 102925L)
# Write with LF line endings on every platform, verify, then replace.
candidate <- tempfile(fileext = ".csv")
con <- file(candidate, open = "wb")
write.csv(rebuilt, con, row.names = FALSE, eol = "\n")
close(con)
check <- read.csv(candidate, stringsAsFactors = FALSE)
stopifnot(identical(names(check), names(rebuilt)),
          identical(check$recorded_outcome, rebuilt$recorded_outcome),
          identical(check$age, rebuilt$age),
          max(abs(check$admission_grade - rebuilt$admission_grade)) < 1e-9,
          unname(tools::md5sum(candidate)) == "38e601f248c7b90acc68d39f5a9638a3")
canonical_dir <- file.path("..", "..", "instructor", "chapter_01")
if (!dir.exists(canonical_dir)) stop("Canonical chapter_01 directory is required")
stopifnot(file.copy(candidate, file.path(canonical_dir, "realinho2022_students.csv"),
                    overwrite = TRUE))
unlink(candidate)
```

::: {.callout-note title="Chapter overview"}
This chapter follows the research journey from questions to data, research
design, methods, and interpretation. Published studies of student outcomes and working
from home, together with invented feedback, measurement, and language
examples, make the ideas concrete. The feedback study then shows how
these contributions form an argument. Reading and comprehending a study
is the approach's primary use; evaluating its quality is a separate
secondary use that depends on that reading. A brief construction
workflow offers auxiliary guidance for conducting research. Fuller
documentary readings and coordinated studies are supplementary depth.
By the end of the chapter you will be able to frame a concern as a
question with an estimand, identify the data and how they were produced,
recognize a method family and the reasons for its use, distinguish an
estimate from its uncertainty and interpretation, and follow the support
to a bounded answer.
:::

Five concerns give the methods in this chapter a purpose. A college
wants to identify students who may need support. Teachers wonder whether
a different feedback format would help students learn. An employer is
considering working from home. A researcher wants to understand
students' confidence and sense of belonging. A language program wants
to follow how its students progress over a year.

How can a concern become a quantitative question, and how do measurement,
design, and statistical methods connect that question to an answer with
stated bounds?

## From a Substantive Problem to a Quantitative Question

Why do some students thrive while others struggle? What shapes a
person's decision to seek help? Does a change in working conditions
improve performance? Curiosity about social and behavioral life often
begins with questions like these. They are familiar enough to invite an
immediate explanation, yet complex enough that several explanations may
seem plausible. Personal experience can suggest a useful idea. Research
gives us ways to examine that idea beyond the experiences that first
brought it to mind.

**Empirical research** grounds knowledge in systematically obtained
observations and experience. Researchers connect what they propose to
what they can observe, then use those observations to develop, examine,
or revise their account. An explanation about motivation, for example,
needs some way of connecting that idea to students' responses or
behavior. Theory helps decide what would be informative; observation
helps reveal where an expectation holds, where it changes, and what
remains unresolved.

Quantitative research is one way of being empirical. It records
observations numerically and uses statistical methods to study
**variation**: differences across people, places, conditions, and time.
The numbers might be amounts, scores, or counts of people in categories.
Their value comes from what they represent and the comparisons they
make possible. Qualitative research also uses empirical evidence;
mixed-methods research brings quantitative and qualitative inquiry
together. The methods developed here are quantitative.

A social or behavioral outcome usually varies for many reasons at once.
Any statistical model selects some of those features and leaves others
in the background. **Parsimony** is a useful modeling principle: seek a
reasonably simple account that retains what matters for the question.
A more elaborate model earns its complexity when it represents
important differences or improves the intended inference. When we try
to model the complexity of the world with parsimony, a familiar caution
applies: as the statistician George Box put it, all models are wrong,
but some are useful [@box1976; @box1979]. A model earns its place by
what it helps us learn, rather than by reproducing every feature of
the world.

Some questions go further than describing variation. They ask whether
changing one thing would produce a change in another. **Causal
questions** require support for that proposed comparison, beyond
observing that two things vary together. They are one important kind of
quantitative question, alongside questions about what is present, how
variables are associated, and what can be predicted. The kind of answer
wanted determines what the research needs to establish.

The five opening concerns each present a different research difficulty.

**Student support: who might need help?** A college wants to help
students before difficulties accumulate. Prior grades, age, finances,
course choices, and family circumstances may all matter. Students share
courses and enter in different years, while their eventual outcomes
become known much later. At enrollment, the college can act only on
information available then. Which of those details could help it
recognize a need for support?

**Feedback: would a different format help?** Teachers spend considerable
time commenting on assignments, yet students repeat similar mistakes.
Learning also depends on prior achievement, the teacher, and the class.
Classmates share conditions, and students may respond differently to the
same comments. Teachers choose formats for reasons that may themselves
affect learning. How could a study distinguish what the feedback
contributes?

**Working from home: what changes?** An employer wonders whether working
from home helps or harms performance. Job responsibilities, tenure, home
conditions, and commuting may matter, and performance changes from week
to week. Employees who volunteer to work from home may differ from
others. The arrangement could affect performance through several
intervening changes. What comparison would help the employer understand
its consequences?

**Confidence and belonging: what do students feel?** A college wants to
know whether students feel able to handle their coursework and whether
they feel part of the academic community. Neither feeling appears on a
transcript, and students may describe similar experiences in different
words. A student who is confident about the work can still feel out of
place. How could a college find out what its students feel?

**Language progress: what changes over a year, and for whom?** A
language program wants to know how students improve across a year.
Progress may come quickly at first and then slow, and two students who
finish at the same level may have arrived there very differently.
Classmates also share a teacher and a timetable. What would show how
learning unfolds and how students differ?

These complexities give statistical methods a purpose. Several
simultaneous influences motivate comparisons that take more than one
variable into account. Ideas such as confidence need observable
indicators. Shared classrooms and repeated occasions call for attention
to dependence. Self-selection makes design important, and intervening
changes invite questions about pathways.

### Questions, aims, theory, and hypotheses

Take the feedback concern: "Does a different kind of feedback improve
students' learning?" It is a useful starting point, but *learning* could mean
performance on an immediate quiz, a delayed assessment, course
completion, or confidence. *Students* could refer to one class or to a
wider educational system. *Improve* could refer to a change observed
over time or to an effect caused by the feedback. An answerable
quantitative question names whom or what it concerns, the outcome,
the comparison or relationship, and the kind of answer sought.

A **topic** identifies an area of interest, such as student support. An
**aim** states what the study intends to accomplish: for example, to
assess how well information available at enrollment predicts later
recorded student outcomes. A **research question** expresses the uncertainty
more directly: how well does a stated prediction rule perform for new
students represented by these data? An aim can be sufficiently
specific to supply a question even when a paper has no sentence ending
with a question mark.

Substantive theory helps explain why the question matters and what might
be expected. A theory of feedback might propose that explanations help
students recognize errors and choose more effective practice. A
**substantive hypothesis** could therefore predict better achievement
under explanatory feedback than under the usual format. In the invented
feedback study, classes in both conditions follow the same six-week
program; the feedback format differs.
That expectation also suggests possible measurements,
such as practice time and the kinds of errors students make.

A **statistical hypothesis** states a proposition about a specified
statistical quantity or model. For the feedback question, one
proposition, the one a test usually assumes, is that the average
assignment effect on the follow-up score is zero. That proposition is
the **null hypothesis**. It is more precise, and narrower in scope, than
the theory that feedback helps learning, and Section 4.4 shows how the
evidence is examined for compatibility with it. A statistical test
examines a proposition under stated conditions; the broader educational
explanation needs additional support.

A **result** is a quantity produced by analysis, such as a mean,
difference, interval, or prediction. A **claim** is a statement the
researchers make after interpreting the result. These terms are related,
but they are not interchangeable. The order of responsibility is:

1. The question states the answer obligation.
2. A hypothesis may record an expectation before the focal evidence is
   examined.
3. Analysis produces one or more results.
4. Reasoning connects those results to an eventual bounded answer, which
   may itself be inconclusive.

Descriptive and exploratory questions can be valuable without a
directional hypothesis. When a hypothesis is present, its content and its
timing are separate pieces of information. A hypothesis about a mean
difference has the same statistical meaning whether formulated before
data collection or suggested by a plot; its status as prospectively
specified or exploratory differs. A transparent study preserves both.

### Description, association, prediction, and causal effects

Consider four questions a college might ask. How are recorded student
outcomes distributed? How is admission grade associated with those
outcomes? How well does an enrollment-time prediction rule perform for
other students? What would a feedback intervention change in achievement?
Administrative records contribute to the first three; an intervention
study can supply a designed comparison for the fourth. A common interest
in student success gives these questions context, while their intended
answers require different support.

These are **descriptive**, **associational**, **predictive**, and **causal**
questions. @tbl-journey-question-modes summarizes their different
tasks. Its third column introduces the **estimand**: the quantity the study
seeks to learn. The next subsection explains how to specify it.

For the feedback question, the **causal effect** is the contrast between
the achievement outcomes that would occur under assignment to explanatory
feedback and to the usual format for the same students. That contrast could be zero.
Learning about it requires a comparison supported by the study design and
its assumptions; Section 3.1 develops this reasoning.

```{=latex}
\Needspace{10\baselineskip}
```

| Mode | Beginner question form | Example estimand | Typical support question |
|:--|:--|:--|:--|
| **Descriptive** | What is present, how much is there, or how is it distributed? | A mean or proportion for a named group, place, and period | Are the cases and measurements suitable for what is being described? |
| **Associational** | How do measured variables differ or vary together? | A difference or relationship for named variables and cases | Does the comparison describe an association without implying causation? |
| **Predictive** | How well can an outcome be predicted for new cases, settings, or times? | A specified performance measure for a stated prediction rule in a defined population of new cases, settings, or times | Was prediction checked on suitable observations not used to fit the procedure? |
| **Causal** | What would differ under one intervention or exposure rather than another? | An average outcome contrast under two named alternatives | Do the design and assumptions permit the intended causal comparison? |

: Four question modes, example estimands, and support questions. {#tbl-journey-question-modes tbl-colwidths="[18,27,25,30]"}

A descriptive answer may be exactly what a decision requires. Knowing
which students are missing from a service can be more useful than fitting
a complex model to those already attending. Likewise, a prediction can
be useful without explaining why an outcome occurs, provided its
performance and intended use are understood. The four modes describe
different tasks rather than a ranking of research quality.

The word *explain* deserves attention. Explaining an observed admission-grade
association may mean describing how recorded outcomes vary with that grade. Explaining
a causal mechanism asks how changing something produces a change in an
outcome. A regression equation can organize the first task without
establishing the second.

Measurement and growth questions fit within this map. Describing the
relationships among confidence indicators is an associational task;
predicting later achievement from them is predictive. Describing an
average achievement trajectory differs from asking whether a policy
causes that trajectory to change. The subject matter and the intended
claim work together to define the question.

### The estimand and the eventual answer

An **estimand** is the target quantity: what the study seeks to learn for
the named units, conditions, outcome, and scope. An **estimate** is the
numerical value calculated from the available data. Put simply, the
estimand is what you want to learn; the estimate is what the analysis
produces as its numerical approximation [@lundberg2021].

For the student-outcomes question, an estimand could be a probability of
recorded Dropout status at a specified admission grade and age. For
feedback, it could be the average difference in follow-up achievement if
the same eligible students' classes were assigned explanatory feedback
rather than the usual format. For prediction, it could be a stated
rule's average probability prediction error for new students represented
by the source dataset: on average, how far the probabilities the rule
assigns fall from what actually happened for students it has not seen.
A predicted probability for one student is an output of that rule; its
performance across new students is a different quantity. A later question
about next year's entrants adds a future-cohort scope that also needs
evidence about stability over time.

These specifications carry information that an everyday question can
leave implicit. A feedback-effect estimand needs the students of
interest, the two assignments being compared, the assessment and its
timing, and the way individual effects are averaged. An average that gives
each student equal weight can differ from one that gives each school
equal weight when school sizes vary. Both can be useful, but they answer
different questions.

For a purely descriptive question about a complete, fixed dataset, the
estimand may be its exact mean or proportion. Once those data are
available, calculating that quantity leaves no sampling uncertainty about
that particular dataset. Questions about recording errors, missing people,
or a wider population can still remain. The word *estimand* identifies
the quantity sought; it leaves the source and extent of uncertainty to be
specified.

An estimand usually supplies the numerical focus of a **question-answer
pair**. The question states what needs answering, and the reported claim
supplies the study's eventual answer. A readable question can remain short,
with its full specification beside it. For the feedback study:

> **Question:** What is the average effect of assigning
> explanatory feedback rather than the usual format on the
> eligible students' follow-up achievement?
>
> **Specification:** The students belong to 24 participating classes in
> 12 schools. Achievement is the score on a common 0-40 assessment two
> weeks after a six-week program. The comparison gives each of the 480
> eligible students equal weight and concerns assignment of their class
> to a format, regardless of their individual attendance.

## Producing Evidence: From People and Concepts to Data

Data are recorded information, and a dataset organizes that information
for a study. It is already the product of several decisions. Someone
defined eligibility, selected a
source, administered an instrument, recorded a response, or applied a
coding rule. Understanding those decisions makes the numbers informative.

Begin with information about one student in each case. The student-outcomes
dataset combines admission information with an eventual administrative
outcome, so its variables concern different times. The feedback dataset
combines the student's assigned class, format, and delayed assessment
score. Understanding whom and what the data describe gives meaning to
a count, average, or comparison.

This section follows how a study produces evidence: what it concerns,
how measurements are made, how observations connect, and what becomes
available for analysis. Chapter 2 takes the resulting dataset as its
starting point. It develops the reading of records, layouts, variable
forms, and codes, then establishes which observations support each
calculation. Understanding the study gives meaning to the file; examining
the file shows how those meanings are represented in the analysis.

### Units, variables, and study scope

For the student-support concern, a real study provides a useful dataset.
Realinho and colleagues studied dropout and academic success at a
Portuguese polytechnic [@realinho2022]. The released file contains 4,424
students: 1,421 with recorded **Dropout** status, 794 **Enrolled**, and
2,209 **Graduate** at the normal duration of their degree [@uci697].
The study built machine-learning models to predict these outcomes. A
simplified analysis here uses logistic regression, which we will
describe in words before following its implications.

A **case** is the entity or measurement occasion described by the data;
here each case is a student. A **unit** is the entity selected, assigned,
measured, or analyzed in a particular operation. These operations may
concern different entities. In the feedback study, whole classes receive
an assignment, achievement is measured on students, and the analysis
compares the class means under the two formats. Identifying the operation makes the
unit clear. Measurement units such as points or minutes have a different
role: they express the amount recorded, rather than identifying the
entity studied. Section 2.3 uses these entities to explain the connections
among their observations.

A **population** is the collection of units a question concerns. A
**sample** is a set of units selected or observed from that collection. The scope
depends on the question: the participating students may themselves be the
population of interest for a narrowly specified assignment effect. A
wider question about all students in the region would need an account of
how the participants relate to that wider population.

A **variable** is a characteristic that can take different values across
units or occasions. Admission grade, age at enrollment, course, and
recorded student status are variables in the administrative file.
Quantitative variables express amounts: time on task is a continuous
measurement, while an approved-unit count or a whole-number score takes
discrete values. Categorical variables identify groups. Course is
nominal, with no intrinsic order; a low-to-high confidence rating is
ordinal, with an order but no established equal spacing. A Dropout
indicator is binary, with two categories. A numerical course code remains
categorical because its number names a course rather than an amount.
Chapter 2 develops these forms and their implications for calculations.

In the released student data, $1{,}421/4{,}424\approx0.3212$, or 32.1203%,
have recorded Dropout status. A Dropout-versus-other-status question
combines the 794 Enrolled and 2,209 Graduate students as the other
category, while keeping their distinct meanings. Recorded Dropout can
include changing course or institution; it need not mean leaving higher
education altogether [@realinho2022; @uci697].

Variables also have **roles** in a question. The outcome is what an
analysis seeks to describe, compare, or predict; predictors help express
its relationship with other information. In the simplified student-outcome
analysis, recorded Dropout is the outcome to be predicted, while admission
grade and age at enrollment are predictors. Achievement is an outcome when
asking about feedback, but prior achievement can be a predictor when
forecasting a later recorded outcome. The same kind of score can therefore serve
different roles at different times or in different questions.

The prediction time also determines which information is available.
First-semester results arrive too late for a prediction made at enrollment,
however strongly they relate to the eventual outcome. They may be useful
predictors when the question concerns a prediction made after that
semester. The question specifies both what is to be predicted and when
the prediction will be made.

The familiar labels **response variable** and **dependent variable**
often refer to the outcome; **explanatory variable** and **independent
variable** often refer to a predictor. In an experiment, a predictor
may be deliberately assigned; in an observational study it may simply
be measured. These role labels alone establish neither causation nor
statistical independence. More specific labels, such as treatment or
confounder, describe the variable's role in the study's design.

### Measurement instruments, procedures, and meaning

In an invented measurement study, researchers want to study
*learning confidence*: students' confidence in understanding course
material and dealing with unfamiliar problems. They collect three
continuously scored indicators, each from 0 to 100: confidence in explaining
an idea, applying it to a new problem, and checking a solution. They also
collect three indicators of *belonging*: feeling accepted, included,
and able to participate in the academic community. Confidence concerns
one's ability to handle learning tasks; belonging concerns one's place
in that community. Each has its own proposed meaning and indicators.

The substantive ideas are **constructs**: concepts defined through
knowledge of the domain. The recorded responses are **observed
variables**, and the observed variables chosen to represent a construct
are its **indicators**.
An instrument links them through questions, tasks, instructions, response
scales, and scoring procedures. The data are more informative when these
details are available: a response given immediately after a difficult test
may have a different meaning from one given at the beginning of a semester.

A researcher might average the three confidence indicators to form a
**composite score**. For example, one student's responses of 60, 80, and
70 give an average of $(60+80+70)/3=70.0000$. The three responses are observed
indicator values; their average is a calculated score. Another researcher
might use a factor model to represent their
relationships through a **latent variable**, a variable specified within
the model rather than directly observed. The construct guides the choice
and interpretation of indicators; the latent variable supplies a
particular statistical representation of their shared variation. A
score estimated from this factor model depends on the model and scoring
method. The construct, the latent variable, and a calculated score are
related, but each has a different role. Some studies use observed
measures or composites without specifying a latent variable.

To see why the distinction matters, imagine that all three indicators
respond strongly to test anxiety. Their scores could move together even
if they cover only a narrow part of learning confidence. Factor analysis
can describe their statistical relationship. Theory, item content,
response processes, and other evidence help explain what that
relationship means. A suitable statistical model and a convincing
substantive interpretation contribute different kinds of support.

Measurement also involves **reliability** and **validity**. Reliability
concerns the consistency or precision of scores under specified
conditions. Validity concerns the support for proposed interpretations
and uses of those scores. Consistent scoring is valuable, while the
meaning of a score still depends on what the instrument measures and how
it is used [@aeraApaNcme2014].

For the feedback study, a common scoring guide helps make scores
comparable across formats. An assessment blueprint connecting tasks
to the taught material gives reasons for interpreting those scores as
course-topic achievement. Extending them to creativity or lifelong
learning would require additional evidence. The instrument and procedure
tell us how observations were produced; the reasons for their interpretation
become important when explaining why an interpretation is supported.
Chapter 2 uses this account of measurement when interpreting recorded
scales and codes and choosing suitable numerical descriptions.

### Data structure: how observations are connected

Once we have identified the units and measurements, we can ask how their
observations connect. A **cross-sectional** view describes units for a
particular occasion or period. **Longitudinal** data follow the same units
over time; **panel data** commonly organize that follow-up into successive
waves. Following the same students supports description of individual
change. **Repeated cross-sectional** observations instead describe a
population at successive times without necessarily following the same
people, so changes in the population can involve different individuals.

An invented language-learning study follows students across four equally
spaced assessments. Student A scores 58, 65, 68, and 76; Student B scores
62, 63, 67, and 68. Their successive gains are 7, 3, and 8 points, and
1, 4, and 1 point. Across three intervals, average changes are
$(76-58)/3=6.0000$ and $(68-62)/3=2.0000$ points. Equal time spacing thus accompanies
uneven progress. Keeping the sequences reveals differences that a single
average for each student would hide.

Repeated measurements can also concern conditions within one session,
such as the same students completing tasks under two sets of
instructions. The observations share a person even when the question
concerns conditions rather than change over a year. Administrative fields
collected at different times can describe a student's history; studying
a growth trajectory additionally requires repeated measurements of the
outcome on a common scale and their occasions.

A **long** arrangement records one student at one occasion per row,
with the score and relevant identifiers. Student and occasion identifiers
connect that student's repeated observations; classroom and school
identifiers show group membership. Chapter 2 develops record counts and
equivalent long and wide layouts.

Students sharing a classroom experience some common teaching conditions;
classrooms can in turn share a school. This is a **hierarchical**
arrangement: occasions within students, students within classrooms, and
classrooms within schools. Sharing people or groups creates
**clustering**. These features can coexist: the language study is
longitudinal and clustered. Section 3.3 considers what those connections
mean for comparisons, information, and uncertainty.

### Implementation, exclusions, and missing observations

A record of feedback assignment tells us which format a class was offered.
Attendance and delivery records tell us what students and teachers
actually did. Both can matter. An assignment-effect question follows
students according to the assigned format; a question about attending
every session asks something different, because attendance may depend on
motivation, health, or difficulty with the material.

Similarly, a student-support study needs to establish who appears in its
file. Consider an invented alternative to the released student data:
an administrative extract contains only students still enrolled at year
end. That file can describe year-end enrollees. Studying withdrawal
requires the entering cohort and information about who left, including
students absent from the year-end file. A missing score for an included
student and a person omitted from the dataset are different problems
for understanding the data.

The **analysis sample** is the set of cases or observations used in a
particular analysis. Eligibility, participation, available measurements,
and exclusions can make this set differ from the observations originally
sought or collected. It may also differ across outcomes or methods. A
student with a recorded baseline score but no follow-up score supplies
baseline information, for example, but cannot supply an observed
before–after change. Following the study from entry through assessment
to analysis explains how the available evidence took shape.

Reading the collection and preparation account helps establish who
entered the study and analysis and which observations became available.
A **data codebook** records variable meanings, units, and codes.
Chapter 2 develops its practical use alongside missing values,
preparation decisions, and the denominator for each calculation.

## Research Design and Assumptions: What Supports an Inference?

An achievement difference of 2.5 points can arise in several kinds of
study. Classes might have been randomly assigned to two formats.
Teachers might have chosen whether to adopt explanatory feedback. Or one
class might have improved by 2.5 points between a pretest and a
posttest. The numerical
difference alone leaves the interpretation open.

**Research design** connects the question to a plan for producing an
informative comparison. Reading that plan means identifying both what
was done and why the study treats it as support for its inference.

### Causal comparisons and study designs

A causal question compares what would happen under different conditions.
For one student, imagine the follow-up score if the class were assigned
explanatory feedback and the score if it were assigned the usual format. Only
one is observed. The other is the **counterfactual** outcome: the outcome
under the alternative condition. The causal effect compares the outcomes
under these two specified conditions for the same student. An average
causal effect makes that comparison for the named units or target
population; it can be zero. The counterfactual is an outcome, whereas the
effect is a contrast between outcomes. Because each student's two
condition-specific outcomes cannot both be observed, the design and its
assumptions must supply a defensible comparison.

An **experiment** deliberately assigns conditions; a randomized experiment
uses a specified probability mechanism for that allocation. The mechanism
gives the comparison a known basis, such as the random draw allocating
classes to the two formats in the feedback study. Chance imbalances can
still occur in the realized groups. Section 3.2 explains how the actual
allocation design supports the causal comparison and how this differs
from selecting participants.

In an **observational study**, the investigator records conditions or
exposures without assigning the focal intervention or exposure. Suppose
better-resourced teachers are more likely to adopt explanatory feedback,
and their resources also improve achievement. The observed difference
then mixes the feedback format with other relevant differences.
**Confounding** occurs when a comparison mixes the exposure's effect with
effects of other factors that influence both exposure and outcome.
A mediator instead lies along a proposed pathway through which the exposure
affects the outcome. Measuring and adjusting for common causes can help,
under conditions that need to be stated and supported.

A **quasi-experimental study** uses a design-based comparison around an
intervention or exposure without randomized allocation. For example, a
clearly enforced eligibility threshold might supply a useful comparison
between otherwise similar units close to that threshold. Its causal
interpretation depends on how that rule operates and the assumptions
around the comparison. Voluntary adoption alone describes a selection
process; a credible quasi-experimental argument needs more than that
label [@shadish2002].

Three considerations help organize causal reasoning: an association or
contrast in outcomes, the appropriate temporal order, and support for
addressing alternative explanations. Confounding is one such explanation.
**Reverse direction** is another: an outcome or its earlier development
may influence the exposure, reversing the proposed direction of
influence. For example, earlier achievement difficulties might prompt a
teacher to adopt a new feedback format, so a relationship between format
and achievement would need an account of that sequence.
Maturation, concurrent curriculum changes, changes in measurement,
selective follow-up, and influence between classes can also matter. These
considerations work together rather than acting as three boxes whose
completion automatically establishes causality.

### Sampling, assignment, and self-selection {#sampling-and-assignment-answer-different-questions}

**Sampling or recruitment** concerns how units enter a study, through
probability selection, volunteering, or availability. **Assignment**
concerns the conditions allocated to participating units. The condition
assigned and the treatment actually received can differ: a class assigned
explanatory feedback might receive only part of the planned program.
Study entry, allocation, and receipt therefore identify different parts
of how the evidence was produced.

**Random sampling**, also called **probability sampling**, selects units
through a probability mechanism that gives every eligible unit in a
defined population or sampling frame a known, nonzero chance of selection.
Those chances need not be equal. Here a frame identifies the units from
which selection is made; Chapter 2 develops this idea and compares
sampling schemes. The phrase *random selection* alone can describe other
chance-based decisions, so identifying what is selected and for what
purpose makes its meaning clear. Probability sampling supplies a basis
for population inference under the selection design, contributing to
**generalizability or external validity**: support for extending findings
to other people or settings.

**Random assignment** uses a specified probability mechanism to allocate
conditions among participating units.
The assigned units may be individuals or groups. This mechanism
contributes to **internal validity**: support for a causal interpretation
of the comparison within the study. A study may use either, both, or
neither form of randomization. Each contribution depends on implementation
and analysis; population selection and causal allocation answer different
questions [@shadish2002].

The causal justification follows the actual assignment design. Chance
allocation supplies a known basis for comparison; the realized groups
can still differ in prior characteristics. The analysis must respect the
allocation mechanism and consider any later selection into the available
outcomes.

In the invented feedback study, 12 schools volunteer, each contributing
two classes with 20 eligible students per class. A random allocation
assigns 12 of the 24 classes to explanatory feedback and the remaining
12 to the usual format. Assignment therefore concerns classes, while the
outcome is measured on students. Comparing equally weighted class means
also gives students equal weight because class sizes are equal. The
comparison concerns these participating students. A claim about other
schools needs further support concerning recruitment, curriculum,
resources, and implementation.

The working-from-home concern was examined in a real experiment by
Bloom and colleagues at Ctrip, a Chinese travel company. Employees first
volunteered and met eligibility conditions; 249 were then randomly
assigned, 131 to home work and 118 to office work. Home assignment meant
four days at home and one in the office each week over nine months
[@bloom2015wfh, p. 169 and Appendix Table A.2]. These employees chose to
enter the experiment; the allocation procedure then determined their
assigned working arrangement.

**Self-selection** occurs when people's choices influence participation,
exposure, or continued involvement. At the sampling or recruitment stage,
volunteering chiefly raises a question about generalizability. For example,
volunteers' home circumstances might make home work more beneficial for
them than for other employees. Random allocation among volunteers can
still support their treatment comparison; extending that effect requires
additional evidence. Self-selection names a process, rather than proving
that every estimate is biased.

At the assignment or exposure stage, imagine instead that employees
choose home or office work. If prior managerial support influences both
their choice and performance, it is a **confounder**. Random assignment
addresses this source of systematic difference; a voluntary-choice
comparison needs other support for its causal interpretation.

**Selection bias** concerns systematic distortion relative to the
declared estimand arising from which people or observations enter the
comparison or analysis. Terminology varies: treatment-choice confounding
is sometimes called selection bias; identifying the mechanism is more
informative than the label alone. Recruitment chiefly raises external
validity concerns, while allocation chiefly raises internal validity
concerns. The distinction is a guide, not an exclusive division: selective
loss of outcomes after allocation can also undermine a randomized
comparison [@hernan2020causal].

Responses should match that mechanism. Suitable sampling, recruitment,
and follow-up, together with information about nonparticipants or losses,
help establish whom the findings can describe. Adjustment or weighting
can support a broader inference under explicit assumptions about
selection and relevant measured differences. For treatment comparisons,
regression adjustment, matching, or weighting may address measured common
causes; unmeasured confounding requires additional support. More elaborate
designs include **regression discontinuity**, comparing units near an
assignment threshold, and **difference-in-differences**, comparing changes
across groups under assumptions about their untreated trends
[@hernan2020causal; @shadish2002].

An **instrumental variable** changes treatment take-up, as a lottery
offer might do in an invented program. Its use requires support for
independence from other outcome determinants and an effect on the outcome
only through treatment. Further assumptions determine the identified
effect, often for people whose take-up the offer changes. This supports
a particular causal comparison without establishing volunteer
representativeness [@angrist1996iv].

Other variable roles become visible in invented extensions of the
working-from-home setting. If home work reduced commuting stress, which
then affected performance, stress would be a proposed **mediator**, or
intervening process. A baseline characteristic would be a **moderator**
if the effect varied across its values. If both job satisfaction and
health affected participation in a follow-up survey, participation would
be a **collider**, a common consequence of the two. Restricting the analysis
to survey participants can create an association
between satisfaction and health even if none existed beforehand. The
restriction is a form of **conditioning**: considering the relationship
among people who share a value of another variable, here participation.
The pattern can arise from who participated rather than a relationship in
the full employee group [@hernan2020causal]. These examples illustrate roles;
they are not pathways established by the original working-from-home study.

A variable's timing and role help determine whether adjusting for it
clarifies the intended comparison or changes or biases it.

### Timing, comparisons, and dependence

The units identified in Section 2.1 and the connections among observations
in Section 2.3 help us understand what a comparison can establish. A
**between-subjects** comparison contrasts different people or groups;
a **within-subjects** comparison uses repeated observations of the same
people across occasions or conditions. Both kinds of comparison can occur
in experimental, quasi-experimental, and observational studies.

Return to the language-learning scores measured four times during a year.
Measurements from the same student share a history. Students within a
school also share some conditions. These connections can create
dependence, affecting the information supplied by the data. Four scores from one
student answer questions about that student's trajectory, while four
students measured once provide information about differences between
students. The same number of measurements can support quite different questions.

Timing helps establish the order of an exposure and an outcome.
Repeated measurements can also contribute to a causal question. A
feedback rollout observed in intervention and comparison classes before
and after implementation helps reveal pre-existing differences and
changes over time. Its causal interpretation still needs reasons that
the comparison describes what would have happened without the rollout.
A new teaching policy introduced only in the intervention classes, for
example, could provide another explanation for their change.

Combining groups and time phases creates a richer design. Read its
allocation, comparison conditions, timing, and dependence together.
Such combinations make experimental design a substantial field of
study in its own right.

The level of a relationship matters too. A change within a language
learner differs from a difference between learners' average scores.
Comparing classroom averages introduces another level. Each can answer a
useful question, provided the units and comparison remain clear.

Design also affects the uncertainty of an estimate. Many measurements
of the same students or classes cannot automatically replace information
from more units. Reading a study's account of allocation, measurement,
and dependence helps explain how informative its comparison can be.
Chapter 3 develops precision, power, and their roles in planning.

### Reasons for analytical choices and the assumptions they invoke

A method is usually selected through several connected considerations:
the question and estimand, the outcome's scale, the structure of the
data, the design, and the intended use of the result. A numerical score
alone points to no unique procedure. Comparing two independently assigned
groups, describing repeated scores, and predicting a new student's score
give that score different analytical roles.

An **assumption** is a condition invoked in the reasoning. A straight-line
baseline-to-delayed-achievement model represents a constant slope over its
stated range. A procedure for independent observations requires an account
of what makes that independence plausible. A growth model instead
represents dependence among repeated measurements. A causal
interpretation of observational adoption of explanatory feedback needs
reasons for addressing confounding and for comparing sufficiently
similar classes.

Several familiar names identify statistical conditions. **Linearity**
describes the form of a modeled relationship; **independence** concerns
connections among observations or errors. **Normality** may concern a
model's conditional errors, the parts of an outcome the model does not
account for, or an approximation used by the procedure that produces the
estimate, rather than every recorded variable. **I.i.d.** means independent and
identically distributed. Repeated observations or shared classroom
conditions often call for a representation of their dependence instead.
These conditions belong to particular methods and inferences; the task
is to recognize which ones the study actually invokes.

**Causality is an interpretation of a comparison**, supported by design
and causal identification assumptions, such as a defensible account of
confounding. Assuming normally distributed errors or specifying a straight
line supplies no causal interpretation by itself. Statistical assumptions
and causal support answer related but distinct questions.

Substantive reasons accompany these statistical ones. The choice of a
follow-up assessment depends on what it represents; a proposed mediation
path depends on a plausible sequence of events; interpretation of a
confidence factor depends on the indicators and the setting. Statistical
and substantive knowledge therefore contribute from the beginning.

Notice the distinction between identifying a reason and judging its
adequacy. A report might state that it used a straight-line model because
the baseline-to-delayed-score plot appeared approximately linear. That sentence
documents the offered rationale. Whether the plot and other checks
sufficiently support the choice is a further question. We return to this
distinction when reading and evaluating the argument.

@tbl-journey-feedback-designs compares the randomized feedback study
with two invented redesigns. Each arrangement names its own causal
question, students, outcome period, and intended estimand.

```{=latex}
\Needspace{10\baselineskip}
```

| Research design | Research question | Intended estimand | Possible confounding and other threats |
|:--|:--|:--|:--|
| **Experimental:** 24 classes randomly allocated between two formats | What would assigning explanatory feedback rather than the usual format change in delayed achievement for the eligible students? | Average assignment effect on the common delayed score for eligible students in participating classes | Random assignment addresses systematic allocation by prior achievement or teacher resources. Chance imbalance remains possible; loss of students and influence between classes are additional threats. |
| **Quasi-experimental:** designated rollout classes and comparable not-yet-treated classes, with baseline information | What would delayed achievement in rollout classes have been under explanatory rather than usual feedback? | Average effect of replacing usual with explanatory feedback for eligible students in rollout classes at the stated follow-up | Prior achievement, teacher resources, and reasons for rollout timing may affect adoption and outcomes. The rollout rule and comparison need to support the counterfactual. |
| **Repeated measurement within a quasi-experimental design:** intervention and comparison classes at several pre- and post-rollout occasions | What effect did explanatory feedback have on average achievement during the named post-rollout period for eligible students in rollout classes? | Average effect on scores in that period for those students, relative to continuing usual feedback | Different pre-existing trends, group-specific maturation, or concurrent teaching changes can distort the comparison. Repeated measurements help examine these possibilities. |

: Causal questions and intended estimands under three feedback-study arrangements. {#tbl-journey-feedback-designs tbl-colwidths="[25,24,24,27]"}

Repeated measurement is a feature that can accompany different assignment
designs, which is why the last design also names a quasi-experimental
arrangement. A one-group before-and-after dataset can describe change;
attributing that change to feedback requires further support for the
unobserved alternative. An adjusted association or an average pre-post
change can be an estimand in its own right, under its own question.
Keeping the intended estimand distinct from an available comparison helps
the reader understand what the analysis is trying to establish.

## Statistical Methods, Models, and Analysis

Statistical analysis connects data to questions through summaries,
relationships, estimates, and uncertainty. Method names help readers
recognize what an analysis learns.

### A question-led map of method families

Descriptive statistics summarize observed data through means,
proportions, measures of variation, and plots. A mean can describe the
delayed achievement of a class; a proportion can describe what share of
students have recorded Dropout status. Variation shows how much the
observations differ around such summaries. Plots make patterns visible,
including features a single number might conceal.

Read @tbl-journey-method-families by connecting each question, method
family, and estimated quantity. Each entry maps the possible questions
within a case rather than a single question-answer pair. Held-out data
are set aside while a rule is developed. The surrounding explanation
identifies the data and comparison; the second table
develops their connection to design.

```{=latex}
\Needspace{10\baselineskip}
```

| Familiar case and question | Related method families | What the analysis estimates |
|:--|:--|:--|
| **Student outcomes:** How does the chance of recorded dropout relate to admission grade and age, and how well can enrollment-time information predict the outcome for other students? | Proportions and descriptive summaries; binary regression within generalized linear models (GLMs); prediction checked on held-out data | A proportion in each recorded status; a conditional probability of recorded dropout; the average error of a fixed prediction rule for new students represented by the dataset |
| **Feedback:** What is the average effect of assigning explanatory feedback rather than the usual format on delayed achievement? | Mean comparisons, analysis of variance (ANOVA), and regression with a format indicator | The student-average achievement contrast under the two assignments in the named classes and period |
| **Working from home:** What did assignment to home work change in performance, and might the effect differ across employees or work through an intervening change? | Group comparison and regression with repeated outcomes; moderation and mediation as previews | The average assignment effect on a named performance outcome over the experiment; a difference in effects across a baseline characteristic; a proposed indirect effect, which needs its own support |
| **Confidence and belonging:** What do several indicators have in common, and how are the two constructs related? | Exploratory and confirmatory factor analysis (EFA, CFA) and structural equation modeling (SEM) | Indicator-factor relationships (loadings); correlations or modeled relationships between factors (factor correlations or structural coefficients) |
| **Language progress:** How does achievement change over four occasions, and how do trajectories differ among students who share classrooms? | Linear mixed-effects models (LMM) and longitudinal models; latent growth as a related representation | An average rate of change and between-student variation in trajectories for a stated model and time scale |

: Questions, method families, and what each analysis estimates in the five cases. {#tbl-journey-method-families tbl-colwidths="[36,30,34]"}

For student outcomes, first describe the proportions recorded as
Dropout, Enrolled, and Graduate. A two-way frequency table could count
these statuses within each course; dividing by the number of students
in that course gives its outcome proportions. Binary regression can
then express the probability of dropout given admission grade and age.
This is a conditional probability because it refers to specified
predictor values. A prediction assessment checks those probabilities
on held-out students. Look for the outcome definition and the distinction
between fitting a rule and assessing it on other students' data.

For feedback, mean comparisons estimate the average difference under
the two assigned formats. Analysis of variance (ANOVA) relates variation
between group means to variation within groups. A **grouping factor**
is a categorical variable defining groups; the two formats are levels
of one factor. One-way and two-way ANOVA use one and two grouping
factors, respectively, each of which can have several levels.

In the feedback study, class means represent outcomes under the two
assigned formats. The analysis accounts for class allocation when
estimating the difference and its uncertainty. The 480 student scores
supply measurements from 24 assigned units. Regression can express the
format comparison while retaining that design and the intended student
weighting.

For working from home, regression can compare performance under the
two assignments while accounting for repeated measurements from each
employee. Many weekly measurements represent repeated observations of
the same people, rather than that many independent employees. A report
may present an average performance difference and a plot of how the
groups compare over time. Further analyses could ask whether the effect
varies across employees or works through an intervening change.
Identify the performance outcome, period, and employees each result
concerns, and whether a proposed pathway belongs to the original study
or an invented extension.

For confidence and belonging, the measurements are several indicators
intended to represent two substantive ideas. EFA explores how their
shared variation might be represented
by fewer latent variables; CFA examines
a specified representation. Here a factor is a latent variable, distinct
from the categorical grouping factor in ANOVA. A loading describes an indicator's relationship with
a factor, and a factor correlation describes how two factors vary
together. SEM can also express
relationships between factors through structural coefficients, each a
modeled relationship between one factor and another [@bollen1989].
@fig-journey-measurement-map in the appendix pictures the six indicators
and two proposed factors. These outputs become interpretable alongside
the item content, scoring, and intended construct definitions. A
report should let you recognize both what the indicators measure and
what the proposed model says about their relationships.

For language progress, repeated assessments reveal starting levels and
change within students. An LMM can represent
an average trajectory while allowing students to differ around it,
including when they share classrooms or schools [@bates2015lmm]. The
mixed-effects results might describe average improvement per assessment
interval and
variation in students' starting scores or rates of improvement. Latent
growth is an alternative representation of the same trajectories within
a factor model of the kind just described. In a report,
identify the time scale and distinguish an average trajectory from
the individual patterns contributing to it. A rising group average
can coexist with students who improve quickly, slowly, or very little.

The second table connects each case's data and design to the method
families.

```{=latex}
\Needspace{10\baselineskip}
```

| Familiar case and data | Research design and structure | Related method families | What connects them |
|:--|:--|:--|:--|
| Student outcomes: admission information and three recorded outcome categories | Observational administrative data; each case is a student, with variables measured at different times | Binary regression within GLMs; predictive modeling | Define the outcome and the prediction time; use predictors available then; assess performance on held-out data |
| Feedback: delayed achievement and assigned format | Random allocation of 24 classes to two formats | Group comparisons, ANOVA, and regression | Match the mean contrast, uncertainty, and weighting to class assignment |
| Working from home: assigned location and repeated employee outcomes | Individual randomization among eligible volunteers; repeated measurements | Comparisons and regression; longitudinal or LMM extensions | Represent assignment and dependence; added pathway questions need their own support |
| Confidence and belonging: multiple indicator scores | Observational measurement study; a separate dataset for examining a proposed representation | EFA, CFA, and SEM | Represent shared variation; connect measurement and structural relationships to construct meaning |
| Language progress: repeated scores from students sharing classrooms or schools | Longitudinal and hierarchical data; experimental or observational depending on allocation | LMM and multilevel models; latent growth | Represent within-person dependence, grouping, and variation in trajectories |

: How data, design, and methods connect in the five cases. {#tbl-journey-data-design-methods tbl-colwidths="[23,26,22,29]"}

These connections are many-to-many. Regression can serve association,
prediction, or a design-supported causal comparison. Repeated binary
outcomes combine an outcome-scale question with a dependence question.
Method choice therefore connects the estimand, measurements, design,
and assumptions.
“Mixed effects” in an LMM names statistical effects, distinct from
quantitative-qualitative mixed-methods research.

### Regression as a connecting language

Consider delayed achievement plotted against baseline achievement in
a small invented feedback dataset. Each point represents two scores
for one student: the baseline score gives its horizontal position, and
the delayed score gives its vertical position. A fitted line summarizes
how delayed scores tend to differ across baseline scores. Students with
the same baseline score can still have different delayed outcomes;
the line describes a pattern around which individual scores vary.

The line's **intercept** is the fitted delayed score at baseline zero.
Its **slope** is the fitted difference in delayed score for each
additional baseline point. A **residual** is an observed delayed score
minus its fitted value. A point above the line has a positive residual;
a point below it has a negative residual. Residuals make the gaps
between the summary and the individual observations visible.

The data in @fig-journey-feedback-pattern are separate
from the 480-student experimental dataset. Both panels have the same
average delayed score and the same fitted straight line:

$$
\widehat{Y}=8+0.70x.
$$

Here $x$ is baseline achievement and $\widehat{Y}$, read “Y-hat,” is
fitted delayed achievement, both in assessment points. The intercept
is 8 and the slope is 0.70. At baseline 20, the fitted score is
$8+0.70(20)=22.0000$. The left-panel student with an observed score of 21
is one point below that fitted value, so its residual is $-1.0000$
point. The slope describes an association between the two scores;
it does not by itself describe what an intervention raising baseline
achievement would do.

```{r}
#| label: fig-journey-feedback-pattern
#| echo: false
#| fig-cap: "Feedback-score patterns with the same fitted straight line. The left panel follows that line closely; on the right the same line, drawn dashed, misses a pattern that flattens at higher baseline scores, which the solid curve follows."
#| fig-alt: "Two panels plot delayed achievement against baseline achievement. The left scores follow a straight pattern and the right scores follow a flattening curve. Both have the same fitted straight line, delayed score equals eight plus zero point seven times baseline score, drawn solid on the left and dashed on the right beside a solid curved fit."
#| fig-width: 6.25
#| fig-height: 3.2
#| fig-pos: H
par(mfrow = c(1, 2), mar = c(4.1, 4.1, 2.0, .6),
    mgp = c(2.5, .7, 0), las = 1, cex = .88)
baseline_demo <- seq(0, 30, by = 2)
residual_demo <- rep(c(1, -1, -1, 1), 4)
straight_demo <- 8 + .7 * baseline_demo + residual_demo
curved_demo <- straight_demo - .02 *
  ((baseline_demo - 15)^2 - mean((baseline_demo - 15)^2))
straight_fit <- lm(straight_demo ~ baseline_demo)
curved_line <- lm(curved_demo ~ baseline_demo)
curve_fit <- lm(curved_demo ~ baseline_demo + I(baseline_demo^2))
stopifnot(max(abs(coef(straight_fit) - c(8, .7))) < 1e-10,
          max(abs(coef(curved_line) - c(8, .7))) < 1e-10,
          abs(mean(straight_demo) - mean(curved_demo)) < 1e-10)
plot(baseline_demo, straight_demo, pch = 21, bg = "#AECAD5",
     col = "#315269", xlab = "Baseline achievement",
     ylab = "Delayed achievement", main = "A straight pattern",
     xlim = c(0, 30), ylim = c(0, 34))
abline(straight_fit, col = "#315269", lwd = 2)
plot(baseline_demo, curved_demo, pch = 21, bg = "#B9D6CB",
     col = "#267C78", xlab = "Baseline achievement",
     ylab = "Delayed achievement", main = "A curved pattern",
     xlim = c(0, 30), ylim = c(0, 34))
abline(curved_line, col = "#71818C", lty = 2, lwd = 2)
baseline_grid <- data.frame(baseline_demo = seq(0, 30, length.out = 150))
lines(baseline_grid$baseline_demo, predict(curve_fit, baseline_grid),
      col = "#267C78", lwd = 2)
legend("topleft", c("Straight fit", "Curved fit"),
       lty = c(2, 1), lwd = 2, col = c("#71818C", "#267C78"),
       bty = "n", cex = .85)
```

The right-hand pattern flattens as baseline achievement increases,
although its mean and fitted line match the left-hand example. Reading
the graph alongside the summary reveals the difference. Section 4.5
returns to this curve as an example of model modification.

Regression also connects to the feedback group comparison. Code usual
feedback as 0 and explanatory feedback as 1, and a model can express
the difference between their means. Adding baseline achievement allows
the analysis to describe how delayed scores relate to both starting
achievement and format. An interaction allows the baseline relationship
itself to differ by format, rather than requiring the same slope in
both groups. In the class-randomized experiment, the analysis still
needs to respect class assignment and the intended student weighting [@gelman2020].

In an invented working-from-home extension, moderation asks whether the
assignment effect varies with baseline commuting time, while mediation
asks whether fewer interruptions help account for an effect on
performance. The proposed pathway needs additional evidence and
assumptions because randomizing work location also leaves other
influences on interruptions and performance to be considered.

The language-learning data add change over time. Student A's scores
are 58, 65, 68, and 76 across four equally spaced occasions; Student B's
are 62, 63, 67, and 68. Their mean rises from 60.0000 at the first assessment
to 72.0000 at the last, but their starting levels and patterns differ.
Student A's successive gains are 7, 3, and 8 points; Student B's are
1, 4, and 1. The total gains of 18 and 6 points correspond to average
changes of 6.0000 and 2.0000 points per interval. Those averages summarize the
whole period, while the observed changes vary from one interval to the
next.

```{r}
#| label: fig-journey-growth
#| echo: false
#| fig-cap: "Two invented language-learning trajectories. Equally spaced assessment times allow uneven score changes; the lines join observations rather than represent a fitted growth model."
#| fig-alt: "At four equally spaced occasions, Student A scores 58, 65, 68, and 76, while Student B scores 62, 63, 67, and 68. Student A starts lower and gains more overall. Both trajectories have uneven interval gains. Points mark observed scores, joined by lines."
#| fig-width: 6.25
#| fig-height: 2.8
#| fig-pos: H
par(mar = c(4, 4.2, .6, .7), mgp = c(2.6, .7, 0), las = 1)
occasion <- 0:3
student_a <- c(58, 65, 68, 76)
student_b <- c(62, 63, 67, 68)
plot(occasion, student_a, type = "o", pch = 16, col = "#315269",
     lwd = 1.8, xaxt = "n", xlab = "Measurement occasion",
     ylab = "Language achievement", ylim = c(56, 79))
axis(1, at = occasion, labels = 1:4)
lines(occasion, student_b, type = "o", pch = 17,
      col = "#267C78", lwd = 1.8)
legend("topleft", c("Student A", "Student B"),
       col = c("#315269", "#267C78"), pch = c(16, 17),
       lty = 1, lwd = 1.8, bty = "n", cex = .88)
```

An LMM can represent an average trajectory together with differences
among students and occasion-to-occasion variation. Even a model with a
straight average trajectory allows observed scores to depart from that
line. If students share classrooms, the model can also represent relevant
grouping. An average rate of improvement and differences among students
answer related but distinct questions. The two trajectories in
@fig-journey-growth show why both matter; data from more students are needed to
estimate the proposed student and classroom variation.

### Model building and estimation

**Model specification** means deciding what statistical relationship
to represent: the outcome, its predictors, the form of their relationship,
and any dependence among observations. For the fitted line above,
delayed achievement is the outcome, baseline achievement is the
predictor, and the chosen form is straight. This choice represents
the same fitted increase per baseline point throughout the range
being described. A curve would allow that increase to change.

Adding feedback format changes what the model represents. A common
baseline slope describes the same baseline-to-delayed relationship in
both formats while allowing their fitted levels to differ. Allowing
format-specific slopes gives a more flexible description. These choices
should follow the question and the patterns the analysis seeks to
represent. The average assignment effect remains a different quantity
from the association between baseline and delayed scores, even when
both are examined using the same students' data.

A model and an estimation method do different work. The model states
the proposed relationship and its conditions. The estimation method
uses the observations to find numerical values for its unknown
quantities. **Least squares**, for example, chooses the fitted line
that minimizes the sum of squared residuals. Squaring prevents positive
and negative residuals from canceling each other and gives larger
gaps more weight. The resulting line makes those squared gaps as
small as possible overall; it need not pass through every observation.

Software performs the specified calculations. R, the computational
tool used throughout this series, can fit the line, calculate fitted
values, and produce the graph. “Analyzed in R” identifies the software.
The equation, data description, and analytical explanation identify
the model, estimation method, and uncertainty procedure it implemented.

The **generalized linear model (GLM)** family extends regression to
describe different kinds of outcomes, including a probability or an
expected count [@nelder1972]. The recorded-dropout question concerns
a probability, which stays between zero and one. **Logistic
regression** is a commonly used member of this family for a two-category
outcome; it is the rule used here for recorded dropout and reported in
the appendix. A question about calls handled during a defined work
period concerns a count. The model should represent that outcome
appropriately, and the reported result should name its scale. A count
model is one possible analysis of call counts, rather than a
description of every analysis in the working-from-home paper.

For the student-outcome analysis, keep three quantities distinguishable.
A student's recorded status is an observation. The proportion with
dropout status summarizes a collection of observations. A fitted
probability describes the model's estimated chance of recorded dropout
given its predictors. Students recorded as Enrolled remain distinct
from graduates, even though the binary question combines both as
outcomes other than dropout. That coding answers the defined dropout
question without predicting eventual graduation for enrolled students.

### Statistical uncertainty, hypothesis testing, and model comparison

An estimate summarizes the available observations, while the quantity
it seeks to describe can remain uncertain. **Standard deviation (SD)**
summarizes spread around a mean. Applied to students' scores, it describes
variation among those scores. Repeating a stated sampling or assignment
process can instead produce different estimates. Their distribution is
a **sampling distribution**, and its SD is the **standard error (SE)**.
SE therefore describes the variation of a statistic used as an estimate,
under the specified repetition and conditions.

For feedback, different allowed assignments of classes could produce
different estimates of the effect. This is assignment variation within
the participating cohort, rather than selection of new students.
The important distinction is whose variation is summarized: individual
measurements or estimates across the relevant repetitions.

A **confidence interval** combines an estimate with an uncertainty
procedure. Under the procedure's conditions, its confidence level
describes how often intervals constructed across the relevant
repetitions would cover the estimand. The reported interval indicates
the range left compatible with the data by that procedure. It concerns
uncertainty about the estimated quantity, rather than a range containing
95% of individual students' outcomes.

For this reading example, the invented feedback report supplies an
estimated effect of 2.50 points and a 95% confidence interval from 1.00 to
4.00 points. The single-number estimate is the **point estimate**. The
report's uncertainty procedure reflects assignment of classes, while the
result concerns the eligible students in those classes. Complete follow-up
still leaves uncertainty about the average assignment effect because
each student was observed under only one assigned format. The appendix
provides the study description and supporting extracts; the reported
quantities are supplied for interpretation here.

A statistical hypothesis states a proposition about a statistical
quantity or model. One null hypothesis for feedback is that the
average assignment effect is zero. A test asks how incompatible the
observed results are with that proposition under stated conditions.
The test expresses discrepancy through a test statistic. Its
**p-value** is the probability, assuming the null hypothesis and the
stated conditions, of a test statistic as extreme as or more extreme
than the one observed, with “extreme” defined by that test. A result is
called **statistically significant** when its p-value falls below a
threshold stated in advance. For a two-sided test at the 5% level built
from the same procedure, the 95% confidence interval excludes the null
value when the result is statistically significant.
A smaller p-value indicates greater incompatibility by that measure;
it gives neither the probability that the null is true nor the
practical importance of the effect [@wasserstein2016]. The estimate
and its interval remain needed to describe the size and uncertainty
of the difference.

**Model comparison** asks a further question about competing descriptions.
For the same feedback data, a model with a common baseline slope
can be compared with one allowing a different slope in each format.
The comparison examines whether the extra flexibility helps describe
the relationship under the chosen comparison procedure. The outcome,
data, and purpose of the comparison need to remain clear. In
prediction, the relevant comparison instead asks which fixed rule
does better on suitable data excluded from its development.
Fitting the available observations more closely and predicting other
observations better are distinct achievements.

### Model diagnostics, modification, sensitivity, and validation

After fitting a model, examine what it captures. **Diagnostics** check
particular features of the fitted analysis. In the feedback example,
residuals that trace a curved pattern suggest that the straight
baseline relationship misses something systematic. The points are
not simply scattered above and below the line: their positions change
in a recognizable way across baseline scores. The diagnostic directs
attention to the shape of the relationship.

**Model modification** changes a specification in response to
substantive reasoning or analytical evidence. Allowing a justified
curve is one response to that residual pattern. The report should
explain what changed, why it changed, and whether the choice was
planned or informed by the observed results. The revised model then
needs examination too. Keeping the change visible helps readers
understand which description was originally proposed and what the
analysis contributed to its development.

**Sensitivity analysis** examines how a result depends on choices
or assumptions. For feedback, one possible check compares the declared average assignment
contrast with an analysis that adjusts for baseline achievement while
retaining the design and intended student weighting. The comparison
shows whether the estimated answer changes appreciably when that
additional information is used. Interpreting the comparison requires
the same question to remain visible; an analysis of a different
outcome or population may be answering a different question.

**Validation** checks a prediction rule on suitable observations
outside its development data. The simplified student-outcome
analysis builds a dropout rule from admission grade and age for
one group of students and checks it on held-out students. A comparison
rule gives every held-out student the same probability: the dropout
proportion calculated from the training data. Under a simple
measure of probability error, the average squared gap between each
predicted probability and the recorded outcome coded 0 or 1, the
fitted rule performs somewhat better. This is an internal check within
the released dataset; a later cohort would provide a different test of
its performance [@james2021].

The held-out observations are useful because their outcomes did not
guide the original fit. Repeatedly consulting them to choose model
changes turns them into development information, so a final assessment
then needs its own protected evidence. **Data leakage** occurs when
information unavailable at the intended prediction time, or information
reserved for assessment, improperly enters development. Using later
semester results to make an enrollment-time prediction is one example
that can make the rule appear more useful than it would be then.

## Interpreting Results and Formulating Bounded Answers

Numerical results become answers through interpretation. An estimated
difference is an analytical output; a statistical conclusion states
what it indicates about the declared comparison; a substantive
interpretation explains what that comparison means in its setting.
Following this movement keeps the participants, outcome, conditions,
and uncertainty attached to the answer.

### From numerical output to a statistical conclusion

The feedback result is an estimated average assignment effect of
2.50 points, with an approximate 95% interval of 1.00 to 4.00 points
for the eligible cohort. A statistical conclusion can state that
the comparison indicates a positive average effect under the
reported analysis, while retaining uncertainty about its size.
The estimate concerns assignment to a format, rather than the
effect of attending an additional session. It is an average, rather
than a claim that every student benefits by the same amount.

For the dropout rule, the statistical conclusion concerns prediction
on held-out students in the released dataset. The simplified rule
performed somewhat better than the comparison rule using the
dropout proportion from the training data for everyone. This supports a
conclusion about that internal assessment, with performance potentially
varying across other partitions or groups of students. A claim about
future cohorts is a further question, requiring evidence about
prediction in those cohorts and at the intended decision time.

### Substantive meaning, importance, and scope

Units are part of meaning. A difference of 2.50 achievement points
refers to the named delayed assessment and the course topics it
covers. Interpreting it as improved learning depends on what that
assessment measures and when it was administered. A reader should
be able to carry the result back to recognizable tasks students
were expected to perform, rather than treating “achievement” as
an unlimited description of educational benefit.

For an explicitly illustrative comparison, suppose two groups have
recorded dropout proportions of 20% and 30%. Their difference is
ten percentage points. These figures illustrate the unit; they are
not outputs from the fitted student-outcome rule. Whether a support
intervention would reduce dropout remains a separate question.

Bloom and colleagues reported about 13% more calls per week under
home assignment among the 134 order takers included in that analysis
over the nine-month experiment [@bloom2015wfh, pp. 187--188]. This
outcome and subgroup matter: the result concerns weekly calls among
those order takers. “Calls per week” gives the observation period
that turns a count into a rate. Calls per minute of working time
would describe a different aspect of performance.

**Practical importance** connects the size of a result to a
substantive concern. The feedback **protocol**, the document recording
the study's planned decisions before the data were examined, identifies
a two-point improvement as a locally meaningful benchmark. Its
scoring guide awards two points for a fully correct application task,
and the curriculum team judged an average gain of that size worth
considering relative to modest additional preparation. The benchmark
therefore has a stated educational rationale; it is not a universal
threshold for useful feedback.

The estimate exceeds two points, while its interval spans values
below and above that benchmark. The estimated improvement is
positive, with its practical size still uncertain.

In the student case, using probabilities to flag students requires choosing
a cut-off. At 0.5, the rule identifies 73 of the 427 held-out students
recorded as Dropout: $73/427\approx17.0960\%$. Better average probability
predictions can therefore coexist with missing many recorded dropouts at
a selected cut-off. Choosing a cut-off requires considering missed cases
and unnecessary alerts. Recorded dropout is also only one possible sign
of support need. The effectiveness of any subsequent support remains a
separate question; prediction and substantive judgment inform different
parts of that decision.

### Reading numerical results accurately

**Read an estimate with its estimand.** “The difference is 2.50”
gives a number. “The estimated student-average assignment effect
is 2.50 points on the delayed assessment in the eligible cohort”
identifies the quantity it is intended to describe. Keep that
comparison and scope when moving from a table to a written answer.

**Read significance alongside magnitude and uncertainty.** A
small p-value describes incompatibility with a stated null under
stated conditions. A large benefit is a statement about magnitude.
The estimate, its units, and the interval are needed to see how
large and how precisely estimated a difference is; substantive
knowledge helps explain why its size matters.

**Read a nonsignificant finding through its interval and question.**
Using the two-sided test and interval procedure described in Section 4.4,
an illustrative 95% interval from -8 to 8 points contains zero and gives
a nonsignificant result. It leaves
substantial effects in both directions unresolved. An interval from
-0.3 to 0.3 would be much more informative if the study had justified
one point as the smallest important difference. Evidence that an effect
is small needs an appropriate comparison and sufficient precision;
“nonsignificant” alone establishes neither equality nor absence
of an effect.

**Read a fitted model as a specified account of relationships.**
A well-fitting measurement model represents relationships among
the indicators under its conditions. Item content and substantive
knowledge help explain what the factors mean. Similarly, interpreting
a fitted pathway as a causal mechanism needs temporal, measurement,
and design support beyond the existence of a fitted arrow.

**Read results as contributions to an answer.** Retain the outcome,
timing, assessment, and intended use when paraphrasing a reported
conclusion.

### Following the reported answer and its qualifications

Consider these excerpts from the feedback report:

> **Results:** The estimated student-average assignment effect was
> 2.50 points on the 0-40 assessment administered two weeks after the
> six-week program. The approximate 95% confidence interval was
> 1.00 to 4.00 points, with uncertainty reflecting class assignment.
>
> **Discussion:** Assignment to explanatory feedback improved average
> performance on the assessed course topics for eligible students in
> these participating classes, under the reported implementation and
> inferential conditions. Whether the improvement reached the locally
> selected two-point benchmark remained uncertain. Broader learning,
> other settings, and individual effects require additional evidence.

The Results excerpt states the statistical comparison. The Discussion
connects it to course performance while preserving participants,
assessment, conditions, and unresolved magnitude. Reading identifies
what the authors claim. If an exercise asks you to formulate an answer
from supplied results, that answer is your own synthesis. Keeping these
activities recognizable lets a subsequent evaluation examine the
reported claim without silently substituting a different one.

## Connecting Evidence to Answers: The Argument-Based Approach

So far, you have followed a question through its data, design,
analysis, and interpretation. An **argument-based approach** connects
these contributions to the study's answer. An argument here means the
account of how evidence and reasons support an inference and a claim.

One question-answer pair has **one argument with linked layers**. Count
pairs by separately owed substantive answers, rather than by estimates,
tests, or coefficients. Several results can contribute to one answer.
The statistical layer (STAT) concerns recorded variables, statistical
support, analysis, and the statistical conclusion. The substantive layer
(SUB) connects these contributions to the people, practices, meanings,
and setting named by the question. Substantive reasoning enters when a
question and its measures are chosen, as well as when results are
interpreted.

### Recognizing the four functions in Sections 2–5

The four functions already have familiar work to do:

- **Evidence (E):** observations and their production, developed in
  Section 2.
- **Justification (J):** reasons and conditions supporting the comparison,
  developed in Section 3.
- **Inference (I):** analysis and interpretation connecting that support
  to a conclusion, developed in Sections 4 and 5.
- **Bounded claim (C):** the reported answer with its conditions and
  limits, developed in Section 5.

Their logical connection is **E + J → I → C**. Each layer contributes
all four functions, as @tbl-journey-component-definitions explains.
The component names identify what information does in an argument.
A design paragraph, for example, may describe an allocation procedure
and explain why that procedure supports a comparison. These are
different functions within the same paragraph.

```{=latex}
\Needspace{10\baselineskip}
```

| Function | Statistical component | Substantive component | Cross-layer action |
|:--|:--|:--|:--|
| **Evidence (E)** | **STAT-E:** information about the observations, variables, and procedures through which the analysis data were produced or acquired | **SUB-E:** information about the people, setting, activities, and substantive meaning represented by those data | SUB-E **situates** STAT-E |
| **Justification (J)** | **STAT-J:** reasons and conditions invoked to support the intended statistical comparison and inference | **SUB-J:** domain-based reasons supporting measurement meaning, implementation, explanation, and applicability | SUB-J **augments and contextualizes** STAT-J |
| **Inference (I)** | **STAT-I:** the analytical reasoning and work connecting evidence and justification to statistical conclusions, including estimates, uncertainty, comparisons, and checks | **SUB-I:** interpretation of that statistical inference in substantive terms, retaining the relevant uncertainty and conditions | SUB-I **interprets** STAT-I |
| **Bounded claim (C)** | **STAT-C:** the reported statistical conclusion about the estimand, with uncertainty, conditions, and scope | **SUB-C:** the reported question-facing answer, with its substantive scope, qualifications, and unresolved matters | SUB-C **bounds** the substantive reach of STAT-C |

: The component functions in the linked statistical and substantive layers. {#tbl-journey-component-definitions tbl-colwidths="[17,30,32,21]"}

The cross-layer arrows in @fig-journey-linked-argument run from the
substantive components to the statistical components they situate,
augment and contextualize, interpret, or bound. These contributions give
statistical work its meaning for the question. They operate alongside
the within-layer support relation, rather than marking separate stages
of research.

```{r}
#| label: fig-journey-linked-argument
#| fig-cap: "One argument with linked layers: logical support and substantive meaning."
#| fig-alt: "Four boxes in the statistical layer are Evidence, Justification, Inference, and bounded Claim. Evidence plus Justification support Inference, which supports Claim. The substantive layer has the same structure. Upward links from the substantive to the statistical components are labeled situates, augments and contextualizes, interprets, and bounds. These links describe logical contributions, not a sequence of research stages."
#| fig-width: 6.25
#| fig-height: 3.5
#| fig-pos: H
par(mar = c(.1, .1, .1, .1), family = "sans")
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 5), xaxs = "i", yaxs = "i")
centres <- c(1.2, 3.65, 6.25, 8.8)
left <- centres - 1.05
right <- centres + 1.05
labels <- c("Evidence", "Justification", "Inference", "Bounded claim")
text(.15, 4.7, "STATISTICAL LAYER", adj = 0, cex = .9,
     font = 2, col = "#315269")
text(.15, .27, "SUBSTANTIVE LAYER", adj = 0, cex = .9,
     font = 2, col = "#267C78")
for (j in 1:4) {
  rect(left[j], 3.55, right[j], 4.3, col = "#E9F0F5", border = "#688AA0")
  rect(left[j], .8, right[j], 1.55, col = "#E9F3F0", border = "#6D9D91")
  text(centres[j], 3.925, paste0("STAT-", c("E", "J", "I", "C")[j],
       "\n", labels[j]), cex = .91)
  text(centres[j], 1.175, paste0("SUB-", c("E", "J", "I", "C")[j],
       "\n", labels[j]), cex = .91)
  arrows(centres[j], 2.85, centres[j], 3.4, length = .07,
         col = "#71818C", lwd = 1.2)
  text(centres[j], 2.5, c("Situates", "Augments and\ncontextualizes",
       "Interprets", "Bounds")[j], cex = .83, col = "#315269")
  arrows(centres[j], 1.72, centres[j], 2.15, length = .07,
         col = "#71818C", lwd = 1.2)
}
for (height in c(3.925, 1.175)) {
  text((right[1] + left[2]) / 2, height, "+", cex = 1.15)
  arrows(right[2] + .06, height, left[3] - .06, height,
         length = .07, col = "#465B66", lwd = 1.3)
  arrows(right[3] + .06, height, left[4] - .06, height,
         length = .07, col = "#465B66", lwd = 1.3)
}
```

Follow the feedback study's Methods, protocol, assessment and
implementation supplement, and analytical supplement through their
extracts in the appendix. @tbl-journey-argument-functions identifies
how these materials contribute to the study's average assignment-effect
argument. Focus on what each contribution does; quality evaluation
considers its adequacy separately.

```{=latex}
\Needspace{10\baselineskip}
```

| Function | Statistical contribution in the feedback study | Substantive contribution in the same argument |
|:--|:--|:--|
| Evidence (E) | **STAT-E:** eligible students and classes; class allocation; assignment, assessment, follow-up, and analysis-sample information | **SUB-E:** who the students are; what explanatory and usual feedback involved; what the achievement assessment represents |
| Justification (J) | **STAT-J:** reasons class randomization, the class-level comparison, declared weighting, and the uncertainty procedure support the assignment-effect inference | **SUB-J:** reasons the assessment meaning, comparable teaching conditions, and delivery records support the educational interpretation |
| Inference (I) | **STAT-I:** the specified analysis; the estimated student-average contrast and its interval; coding, follow-up, and sensitivity checks | **SUB-I:** interpretation of the contrast and uncertainty on the achievement scale at the stated follow-up, against the local benchmark |
| Bounded claim (C) | **STAT-C:** the reported assignment-effect conclusion for the eligible cohort, with uncertainty and inferential conditions | **SUB-C:** the reported answer about assessed performance in these classes and conditions, with its qualifications and unresolved matters |

: Four functions connecting statistical work to a substantive answer. {#tbl-journey-argument-functions tbl-colwidths="[18,41,41]"}

Method families and argument components organize different things.
Regression is a family of statistical tools; its study-specific use
contributes across components. The variables and data contribute
evidence, reasons for the comparison contribute justification, modeling
and estimation contribute inference, and the reported conclusion
contributes a claim. STAT-I usually holds the most technically demanding
statistical work. It traces the central statistical information of the
argument: how evidence, analyzed under the study's stated conditions and
justification, supports a statistical conclusion. Learning methods
develops that reasoning together with an understanding of the evidence
and support it requires.

### Traceable information and where to find it

**Traceable information** is specific study content through which you
can identify an argument component's function at a documentary location.
It may describe an action or state an assumption, interpretation, or
claim. It bridges a general function and the passage or material doing
that work:

**Question-answer pair → argument component → traceable information →
documentary location.**

You can begin with a component and look for its contribution, or begin
with an informative passage and ask what function it performs. For
example, "12 of the 24 classes were randomly allocated to each format" describes
evidence production in STAT-E. The protocol's explanation that this
allocation supports the assignment comparison gives a reason in STAT-J.
The topic is shared, while the particular information performs
different work.

Similarly, "the common assessment was scored without knowledge of
format" describes a measurement procedure in STAT-E. The assessment
blueprint's reasons for treating its tasks as evidence of taught-topic
achievement support SUB-J. Describing which topics the tasks cover
situates the data in SUB-E. Professional knowledge helps you identify
these functions: knowing what random assignment or an assessment
blueprint contributes makes the corresponding information recognizable.
Reading records the offered reasons faithfully; their adequacy is a
further question.

@tbl-journey-study-locations shows where each part of this chain
usually sits in a study's materials.

```{=latex}
\Needspace{10\baselineskip}
```

| Part of the argument | Possible study location | Information a reader may seek |
|:--|:--|:--|
| **Question and candidate answer** | Abstract, introduction, or protocol | Purpose, quantitative question, the kind of answer sought (descriptive, associational, predictive, or causal), hypothesis when present, and important planned choices |
| **Evidence and its production** | Methods, data-collection and preparation documentation, instrument documentation, data codebook | Population and unit, recruitment, assignment or exposure, measurement, timing, implementation, exclusions, and completeness |
| **Design and justification** | Protocol, Methods, analysis plan, supplement | Why the evidence, comparison, measurement interpretation, and uncertainty procedure can support the intended move |
| **Analysis and inference** | Results, code, or analytical supplement | Estimates such as differences, slopes, probabilities, loadings, or trajectories; uncertainty; and relevant checks |
| **Bounded answer and reservations** | Discussion and conclusion | Substantive meaning, stated answer, scope, alternatives, and unresolved reservations |

: Where support for a question-answer pair may appear. {#tbl-journey-study-locations tbl-colwidths="[25,31,44]"}

The information's function determines its component. Methods may
contain both production
details and supporting reasons. Results may contain calculations and
claims. A reason may also be distributed across the paper, protocol,
and supplement. Keep a useful page, paragraph, table, or code reference
so another reader can follow your interpretation.

When something consequential remains unclear, describe it in ordinary
language. If a paper and supplement give different follow-up counts,
identify both accounts and the unresolved difference. If you cannot find
how allocation was performed, say which materials you examined and what
their account leaves unclear. This describes the accessible support,
while leaving open what may have occurred beyond the available record.

### Reading first, with quality evaluation available separately

Reading asks what the study's argument contains and how its contributions
fit together. **Quality evaluation** asks whether that support is
adequate for the question and answer. It is a separate, secondary use
that depends on understanding the study. You may notice a concern while
reading and record it alongside your interpretation; identification and
judgment remain distinguishable without requiring two rigid passes.

The six evaluation questions in @tbl-journey-evaluation fall into three
groups: alignment, support, and translation.

```{=latex}
\Needspace{10\baselineskip}
```

| Group | Evaluation question | What to consider |
|:--|:--|:--|
| **Alignment** | **Question alignment** | Whether the question, estimand, design, evidence, analysis, and answer remain directed at the same aim |
| **Support** | **Evidence integrity** | Whether the observations and contextual records, their production, and their representation are trustworthy, traceable, and appropriate |
| **Support** | **Justification adequacy** | Whether the statistical, design, measurement, and contextual reasons warrant the proposed moves |
| **Support** | **Inferential adequacy** | Whether reasoning within both layers is adequate, including estimation, uncertainty, checks, sensitivity, and substantive interpretation |
| **Translation** | **Transition fidelity** | Whether the links between layers retain meaning, uncertainty, design status, and scope |
| **Translation** | **Claim calibration** | Whether the reported answer's strength and breadth match its support |

: Six questions for evaluating one question-level argument. {#tbl-journey-evaluation tbl-colwidths="[17,28,55]"}

For an individual evaluation question, use one of four judgments:
**adequate for this purpose**, **needs revision**, **insufficient
information to judge**, or **not applicable**. Add a reason explaining
the particular function or connection being judged.

Consider a variant of the feedback Discussion: "The program produces
important gains in every form of learning for all students." The
numerical results stay the same. Reading identifies the claim;
evaluation identifies why it **needs revision** under Transition
fidelity and Claim calibration. A course-topic assessment has become
every form of learning, and a cohort-average effect has become a benefit
for every student. The claim also settles practical importance more
firmly than the interval permits.

The whole question-answer pair also receives one **whole-pair
conclusion**:

- **Supported as stated:** the available support warrants the answer as
  reported, including a carefully bounded answer that remains
  inconclusive.
- **Supported only after narrowing or revision:** a defensible answer
  remains after correcting its wording, scope, or analytical support.
- **Not supported:** the available record gives sufficient grounds to
  conclude that the proposed answer is unwarranted.
- **Not assessable from the accessible record:** essential information
  for judging the whole answer remains unavailable or unresolved.

The first two conclusions identify an answer warranted by the record,
either as reported or after revision. The last two distinguish an
unwarranted answer from one whose support cannot be assessed from the
accessible record. An argument can fail to support a causal answer
without showing that the effect is absent, and an appropriately
inconclusive answer can still be supported as stated.

The overextended feedback variant is **supported only after narrowing
or revision**, provided the remaining support warrants the narrower
answer. A revised answer keeps the positive average estimate, its
interval, the eligible cohort, the assessed topics and conditions, and
uncertainty about the two-point benchmark. This preserves the original
question while correcting what is claimed for its answer. An unclear
allocation account could instead leave Justification adequacy with
**insufficient information to judge**; a local judgment and a whole-pair
conclusion answer different questions.

A fuller supplementary reading tutorial develops practice in tracing
these contributions, and an advanced supplementary tutorial develops
coordinated questions.

## Conducting Research and Making Its Support Visible

Researchers face the complementary task of producing an informative
study and making its support accessible. Ethics, documentation, and
reporting accompany this work throughout.

### Ethics and responsibility across the study

In the feedback study, participation arrangements, consent where
required,
confidentiality, and fair access to instruction deserve attention before
allocation. Implementation needs to respect those arrangements, and
reporting should describe relevant departures. Protecting students and
preserving an interpretable comparison are connected responsibilities.

In student support, the use of a prediction also matters. A rule with
reasonable average performance may still systematically miss some
students who need help or falsely label others. What follows a prediction
can change the consequences: offering an additional conversation differs
from withholding a service. Understanding those consequences calls for
substantive and ethical judgment as well as statistical evaluation.

In the working-from-home case, linking employee surveys and performance
records also calls for confidentiality and care about employment
consequences. A question about performance can be legitimate while the
handling and use of identifiable records remain consequential.

Responsibilities extend to access and communication. Researchers can
make aggregate results and analysis procedures inspectable while
protecting identifiable student records. An account of exclusions and
missing outcomes lets readers understand whose experience contributes
to the findings. Ethical responsibility belongs throughout the study,
rather than in a final checkbox or a single argument component.

### Reproducibility, reporting, and a visible decision history

A protocol records intended decisions. A data codebook explains the
dataset. An analytical script connects the designated data to outputs.
An assessment guide and implementation account explain their substantive
meaning. Together these materials let a reader follow the support
identified in the location table.

**Computational reproducibility** means that the same data and specified
computations can produce the reported results [@peng2011]. This makes
calculations checkable; design and substantive reasoning determine what
their interpretation supports.

Reporting also preserves the history of decisions. A study can identify
what was planned, what was implemented, what changed, when it changed,
and what evidence was available at the time. Exploratory work has a
legitimate place in that history. Its value is clearer when the reader
can distinguish an analysis suggested by the results from one specified
before those results were examined.

Larger studies require deliberate coordination among analysts, subject
specialists, and authors so that results remain connected to their
questions and support. Interpreting and evaluating that support still
requires their expertise.

### A compact construction-workflow overview

The **construction workflow** organizes the work of developing one
question-level argument in six stages and two phases. It covers much
of the research process, while project administration, collaboration,
and ongoing ethical responsibilities extend across and beyond its
stages. Within this approach, it is an auxiliary use for conducting
research.

**Phase 1: Understand and plan**

1. **Review the problem, setting, and relevant literature.** Develop the
   statistical and substantive reasons for the study.
2. **Frame the question and, where useful, a candidate answer.** State
   what needs answering and its intended scope.
3. **Align design, measurement, estimand, evidence-production or
   acquisition plan, and inferential plan.** Specify how the proposed
   data and analysis will address that question.

```{=latex}
\Needspace{8\baselineskip}
```

**Phase 2: Produce and complete**

4. **Produce or acquire, and document, empirical and contextual
   evidence.** Preserve the observations and the information needed to
   understand their production and meaning.
5. **Estimate, check, and analyse.** Carry out and examine the analytical
   work while recording relevant results and changes.
6. **Contextualize, interpret, and complete the bounded answer.** Connect
   the inference to the substantive question with its uncertainty and
   scope.

Iteration stays within each phase of a cycle: planning can refine the
question and design, while completion can revisit calculations, checks,
and interpretation. The Stage 3-to-4 handoff is one-way within that
cycle. This convention makes prior commitments and their revision
visible; it leaves room for practical research to reuse data and
procedures. For an existing dataset, producing evidence can mean
acquiring or designating the question-specific analysis data.

There are three ways a cycle can end. **Normal completion** preserves
the question, estimand, and prior commitments, even when the answer is
inconclusive or contradicts an expectation. **Phase 2 repair** corrects
work within those commitments, such as a scoring error, and repeats the
affected analysis. Once Phase 2 has begun, changing the question,
estimand, or a prior commitment starts an **additional or revised
cycle**. When the planned evidence cannot answer the original question,
record that cycle as **not answerable as planned**. That label describes
the cycle rather than adding a fourth ending. A new outcome chosen after
examining relevant results receives an explicitly exploratory question and
cycle for those data. A documented revision made before access to those
results can remain prospectively specified; preserve its timing and reasons.
The appendix gives the fuller Phase 1 anchor and
shows how the feedback records keep these commitments visible.

## Reading Bridge

Across the five cases, you can now frame a question and its estimand,
identify the data and how they were produced, and recognize a method
family and the reasons for its use. You can also distinguish an estimate
from its uncertainty and interpretation, then follow the support to a
bounded answer. The argument-based approach connects these contributions
for reading a study. Quality evaluation remains a separate secondary
use, and the construction workflow an auxiliary use for conducting
research.

Chapter 2 builds on this account of how observations arise by examining
recorded data: what the entries
represent, how they are organized, which cases support a calculation,
and how those cases were selected. It then develops distributions, group
summaries, and relationships, applying the design reasoning when
interpreting them.
The same ideas of center, spread, and shape then describe statistics
across repeated samples, giving meaning to sampling distributions and
SE. Statistical inference develops the
calculations and conditions for estimates, intervals, and tests.
Group comparisons bring these ideas together through one-way and two-way
ANOVA, contrasts, and multiple comparisons. Computational practice in R
and regression develop their application further.
Generalized linear models extend the kinds of outcomes a
regression can represent; factor analysis and SEM develop measurement
and linked relationships; mixed-effects models address grouped and
repeated observations.

For further reading, *Introduction to Modern Statistics*
[@cetinkaya2024] connects questions, data, and statistical reasoning,
and *Regression and Other Stories* [@gelman2020] connects regression,
design, and interpretation. Work on estimands [@lundberg2021],
educational measurement standards [@aeraApaNcme2014], and the ASA
statement on p-values [@wasserstein2016] develops specific distinctions.

The argument-based approach is a tutorial synthesis informed by
Toulmin's account of argument [@toulmin2003], LeMire's pedagogical link
between testing and subject-matter claims [@lemire2010], and Kane's
argument-based treatment of interpretation and use [@kane2013]. The
linked components, evaluation questions, and workflow convention bring
those influences together for this tutorial, rather than reproducing
one source's arrangement.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

### Key terms {.unnumbered}

Use this short glossary to revisit the chapter's own vocabulary. The
concept reference in the Part One introduction gives common
alternative names and the main explanation of the statistical ideas
the later chapters develop.

```{=latex}
\Needspace{10\baselineskip}
```

| Term and common alternatives | Meaning |
|:--|:--|
| Question-answer pair | What a study asks and the answer it eventually offers for that question. Each separately owed substantive answer has its own argument. |
| Traceable information | Specific study content through which an argument component's function can be identified at a documentary location. Identifying that contribution precedes judging its adequacy. |
| Estimand | The target quantity sought for the named units, conditions, outcome, and scope. It can concern a fixed dataset, a population, a comparison, or prediction performance. |
| Outcome variable (response variable; dependent variable) | The variable an analysis seeks to describe, compare, or predict. These alternative names share that role in many analyses; a variable's role depends on the question. |
| Predictor variable (explanatory variable; independent variable) | A variable used to represent an outcome's relationship with other information. These analytical-role labels alone establish neither a causal effect nor statistical independence. |
| Sampling (recruitment; selection into a study) | How units come to enter a study, by probability selection, volunteering, or availability. It is contrasted with assignment, which concerns the conditions allocated to participating units. |
| Assignment (allocation) | The allocation of study conditions to participating units, by any mechanism. Actual treatment received may differ from the assigned condition. |
| Random sampling (probability sampling) | Selecting units from a defined population or sampling frame through a probability mechanism that gives every eligible unit a known, nonzero chance of selection, not necessarily the same chance. "Random selection" alone can describe other chance-based decisions. |
| Random assignment (random allocation) | Allocating study conditions to participating units through a specified probability mechanism. The assigned units may be individuals or groups; realized groups need not be identical. |
| Causal effect (counterfactual contrast) | The difference between the outcomes that would occur for the same units or target population under two specified conditions, only one of which is observed for each unit. The contrast may be zero; interpreting a comparison causally needs design and identification support. |
| Confounding | Mixing an exposure's effect with effects of other factors that influence both exposure and outcome. Such a factor is a confounder, a specific causal role. |
| External validity (generalizability) | Support for extending findings to other people, settings, or conditions. The intended extension determines which similarities and differences matter. |
| Unit | The entity selected, assigned, measured, or analyzed in a particular operation. The units of assignment, measurement, and analysis can differ, and all are distinct from measurement units such as points or minutes. |
| Evidence | Observations and contextual records contributing to an answer, including information about how they were produced. |
| Justification | Reasons and conditions invoked to support a comparison, analytical move, or interpretation. |
| Inference | Reasoning and analysis connecting evidence and justification to a conclusion. |
| Bounded claim | The reported conclusion with its uncertainty, conditions, and scope: in the statistical layer, the conclusion about the estimand; in the substantive layer, the question-facing answer with its qualifications. |

: Selected terms for the quantitative research journey. {#tbl-journey-glossary tbl-colwidths="[28,72]"}

```{=latex}
\Needspace{10\baselineskip}
\phantomsection
```

## Exercises {.unnumbered}

Each exercise stands on its own and can be completed by hand, with no
software required. **Foundational** exercises isolate one reading move.
**Integrative** exercises combine the moves. Use only the evidence
provided unless a prompt explicitly invites outside material.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

<!-- prompt-id: CH01-E01 -->
### Exercise 1: Read a research question and answer (Foundational) {.unnumbered #ex-ch01-01}

Consider this short passage from an invented feedback study:

> We asked whether assigning explanatory feedback rather than the usual
> format would improve delayed achievement for eligible students in
> 24 participating classes. We expected explanatory feedback to help
> students apply the taught material. Classes were randomly allocated to
> the two formats. The estimated average assignment effect was 2.50 points
> on a common assessment two weeks after the six-week program, with a
> 95% confidence interval from 1.00 to 4.00 points. We conclude
> that assignment improved average performance on the assessed topics in
> these classes under the study conditions.

a. Identify the research question, substantive expectation, numerical
   result, and reported answer. Explain how the result contributes to the
   answer and what the answer adds.
b. State the estimand in words, including the comparison, outcome,
   students, and timing. Explain how it differs from the estimate of
   2.50 points.
c. Write a descriptive question that the same assessment records could
   address. Explain how the answer sought differs from the assignment
   effect in the passage.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

<!-- prompt-id: CH01-E02 -->
### Exercise 2: Make sense of variables and records (Foundational) {.unnumbered #ex-ch01-02}

The student-outcomes codebook includes admission grade, first-semester
approved-unit count, and course and outcome categories. An invented
extension adds a low-to-high ordered rating of confidence. A separate
invented language-learning record stores student ID, classroom ID,
school ID, assessment occasion, and achievement score in long form,
with four equally spaced occasions for each student. Student A's scores
are 58, 65, 68, and 76; Student B's are 62, 63, 67, and 68.

a. Identify quantitative measurements or scores, discrete counts,
   nominal categories, and ordinal categories. Explain why a course
   code remains categorical even when its stored value is a number.
b. For enrollment-time prediction, identify the outcome and one eligible
   predictor from the codebook. Explain why a first-semester result
   would become available too late, even if it predicts the outcome
   well. What would change if the prediction were made after that
   semester?
c. Describe what one row represents in the language-learning record and
   where repeated and hierarchical structure appear. Name one item a
   codebook or data-record description should clarify. Use the two
   score sequences to explain how equally spaced assessment times can
   accompany uneven changes in achievement.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

<!-- prompt-id: CH01-E03 -->
### Exercise 3: Read design before interpreting a difference (Integrative) {.unnumbered #ex-ch01-03}

Two invented studies concern peer study and course achievement.

- **Study A:** A probability sample of 1,000 students from a university
  reports its weekly participation in peer study. Researchers obtain
  semester grades and compare participants with nonparticipants.
- **Study B:** Among 160 eligible volunteers at the same university,
  researchers randomly assign 80 to an offer of a peer-study program and
  80 to usual study arrangements. All supply semester grades. Some
  students offered the program attend few sessions.

a. Identify how students entered each study and whether the researchers
   assigned a condition. Explain what random sampling and random
   assignment contribute to the respective comparisons.
b. Study A finds higher average grades among peer-study participants.
   Explain how prior motivation could contribute to that difference.
   What causal question can Study B's assigned-group comparison address,
   and why does it differ from the effect of attending every session?
c. For each study, state a population or group its result could concern
   under the stated design. Identify one piece of further information
   needed before extending Study B's finding to all university students.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

<!-- prompt-id: CH01-E04 -->
### Exercise 4: Connect methods to the work they perform (Foundational) {.unnumbered #ex-ch01-04}

Consider enrollment-time prediction of recorded Dropout status,
comparing delayed achievement after class-randomized feedback, and
examining language scores recorded repeatedly for students who share
classrooms.

a. Suggest a method family for each question and explain one reason
   involving the outcome, design, or data structure. More than one
   family may provide a reasonable representation.
b. In the feedback analysis, distinguish specifying a straight-line
   relationship between baseline and delayed scores, estimating its
   intercept and slope, comparing a common slope with format-specific
   slopes, and modifying the specification after seeing a curved
   residual pattern.
c. Explain why a GLM could be useful for the student prediction
   question. What would held-out records tell you after model
   development? What additional evidence would strengthen a claim about
   performance for students enrolling in a later cohort?

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

<!-- prompt-id: CH01-E05 -->
### Exercise 5: Turn a result into a bounded answer (Integrative) {.unnumbered #ex-ch01-05}

In the feedback study, the estimated assignment effect is 2.50 points on
the common 0 to 40 delayed achievement assessment, with an approximate
95% interval from 1.00 to 4.00 points. The local educational benchmark
is two points and has a stated assessment-based rationale. The study
concerns eligible students in participating classes, assessed two weeks
after a six-week program.

a. Distinguish the estimand, point estimate, and interval. What does the
   interval leave unresolved about the size of the effect relative to
   the benchmark?
b. Explain why this result alone would leave a claim about all learning,
   in all schools, over a full year unresolved.
c. Write a short answer to the original feedback question that retains
   magnitude, uncertainty, and the stated outcome and setting.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

<!-- prompt-id: CH01-E06 -->
### Exercise 6: Read an argument, then evaluate its quality (Integrative) {.unnumbered #ex-ch01-06}

Use the feedback study's report and protocol extracts in the appendix.

a. Identify the question-answer pair. Find information contributing to
   each statistical component and explain its function, indicating the
   passage or material that helped you identify it.
b. Identify substantive information that situates the scores, augments
   the statistical justification, interprets the result, and bounds the
   question-facing answer. Explain one of these connections in your own
   words.
c. Now evaluate the variant conclusion, "This format produces
   substantial, lasting improvements in learning in any school." Apply
   Question alignment and Claim calibration, and any other question
   needed for your reasoning. Give a whole-pair conclusion and a
   justified revised answer, clearly separating this evaluation from
   your reading account.

```{=latex}
\Needspace{10\baselineskip}
\phantomsection
```

## Appendix: Records for the Shared Cases {.unnumbered}

These records supply the material behind the examples. Use them to
follow a reported number, revisit a study's conditions, or locate
information for the worked reading and exercises.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

### Student outcomes {.unnumbered}

The source follows entrants at the Polytechnic Institute of Portalegre,
Portugal, from 2008/09 to 2018/19. Outcomes were recorded at the normal
duration of their degree: three years, or four for nursing
[@realinho2022]. Recorded Dropout can include a change of course or
institution; Enrolled means still enrolled at that endpoint. Graduate
identifies completion. The released data cover 4,424 students
[@uci697].

The simplified analysis asks how well admission grade and age at
enrollment predict recorded Dropout versus the other two categories
combined. It retains the original categories in the dataset and assigns
about 70% of each binary outcome group to training, leaving the rest
for an internal assessment.

```{=latex}
\Needspace{10\baselineskip}
```

| Dataset | Dropout | Enrolled | Graduate | Total |
|:--|--:|--:|--:|--:|
| Full release | 1,421 | 794 | 2,209 | 4,424 |
| Training | 994 | 562 | 1,540 | 3,096 |
| Held-out test | 427 | 232 | 669 | 1,328 |

: Students in the simplified prediction analysis. {#tbl-journey-student-record tbl-colwidths="[30,17,17,18,18]"}

On the held-out data, average squared prediction error was 0.2000 for
the logistic rule using admission grade and age, against 0.2182 for
predicting the training-data dropout proportion for everyone:

$$
\frac{1}{1328}\sum_i(\widehat p_i-y_i)^2=0.2000\ \text{(logistic rule)}
\qquad\text{versus}\qquad 0.2182\ \text{(training-prevalence rule)},
$$

where $\widehat p_i$ is the predicted probability and $y_i$ the recorded
Dropout indicator, coded 1 for Dropout and 0 otherwise, for each of the
1,328 held-out students.
At a probability threshold of 0.5, the logistic rule
identified 73 of the 427 students with recorded Dropout status, about
17.0960%. Its use for later cohorts and the value of acting on an alert
remain to be established. Semester results
were excluded because they become available after enrollment; the
released file also has no cohort-year field for a later-cohort check.

The data are provided by Realinho, Vieira Martins, Machado, and
Baptista under [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/)
[@uci697]. The three fields used here, the binary coding, the split, and
the logistic rule are adaptations for this chapter; the source authors
built other machine-learning classifiers, and the analysis can be
repeated from the public release. Reading the chapter requires no
programming.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

### Feedback {.unnumbered}

The invented report supplies an estimated assignment effect of 2.50
points and a 95% interval from 1.00 to 4.00 points. These quantities support
the chapter's reading exercises; a numerical derivation is outside this
example.

The report and supporting materials describe the study as follows.

> **Paper, Methods:** Twelve participating schools contributed two
> classes each, with 20 eligible students per class. Before the six-week
> program, 12 of the 24 classes were randomly assigned to each format
> across the full class pool. A common 0-40 assessment was administered two weeks after the
> program. All 480 eligible students supplied scores and were analyzed
> according to their class's assignment.
>
> **Protocol, design and analysis:** Random allocation supplies the
> basis for the assignment comparison. Comparing equally weighted class
> means gives students equal weight because class sizes are equal. The
> uncertainty procedure reflects assignment of classes. The inference assumes implementation of the allocation,
> a common outcome definition, complete follow-up, and no influence of
> one class's assignment on another class's outcomes.
>
> **Assessment and implementation supplement:** Both formats used the
> same taught topics and scheduled time. Explanatory feedback added
> explanations of errors; the usual format supplied correctness
> feedback. Delivery logs documented these activities and separate
> materials. The assessment blueprint linked tasks to the topics,
> and scorers worked without knowledge of format. The curriculum team
> judged a two-point gain, equivalent to full credit on an application
> task, worth considering against the additional preparation.
>
> **Analytical supplement:** Checks confirmed the score range, class
> assignments, denominators, and complete outcomes. Recalculating the
> mean contrast with each school omitted in turn left the point estimate
> positive.
> Those reduced-record checks describe influence on the estimate;
> the full eligible cohort remains the basis of the primary answer.

The report concludes that assignment improved average assessed
course-topic performance in these classes and conditions, with the
practical size uncertain and broader learning and individual effects
unresolved. Its Results and Discussion excerpts appear in Section 5.
The separate straight-line and curved plots illustrate relationships
with baseline achievement; their values are separate from the supplied
assignment-effect result.

The **Phase 1 anchor** records the question and mode, estimand, design,
measurement, evidence-production or acquisition plan, inferential plan,
and, where relevant, a stopping rule stating when data collection can
end. Here the protocol commits to the
eligible-cohort assignment effect, formats, delayed assessment, student
weighting, and analysis. The codebook records school, class, assigned
format, pre-program baseline achievement score, delayed assessment score,
and analysis membership. The analysis record connects
those commitments to the contrast, interval, checks, and reported
answer, while dated changes remain visible.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

### Short supporting records {.unnumbered}

**Working from home.** Group counts are 131 assigned to home work and
118 to office work (Appendix Table A.2, p. 214); eligibility and the
nine-month, four-home-day schedule appear on p. 169. Table II, column 4
(p. 187) reports the result for 134 order takers, interpreted on p. 188
as about 13% more weekly calls [@bloom2015wfh]. Other outcomes use
different measures or employee groups. The
[author's paper, appendix, and replication materials](https://nbloom.people.stanford.edu/research)
also include nonexperimental employees for Table IV; record analysis
membership when reusing these files.

**Confidence and belonging.** Confidence concerns handling learning
tasks; belonging concerns acceptance and inclusion in the academic
community. Three 0-100 confidence indicators concern explaining,
applying, and checking; three belonging indicators concern acceptance,
inclusion, and participation. @fig-journey-measurement-map shows a
candidate representation. Each indicator also has variation beyond
its factor; item content, scoring, and occasion help establish what the
representation means.

```{r}
#| label: fig-journey-measurement-map
#| fig-cap: "A candidate measurement map for confidence and belonging. Arrows show proposed measurement relationships; the two-headed link represents a possible factor correlation. Indicator-specific variation is omitted from this sketch."
#| fig-alt: "A confidence factor points to explaining, applying, and checking indicators. A belonging factor points to acceptance, inclusion, and participation indicators. A two-headed link connects the factors. This is a proposed measurement map, not an estimated model."
#| fig-width: 6.25
#| fig-height: 1.25
#| fig-pos: H
par(mar = c(.15, .1, .15, .1), family = "sans")
plot.new()
plot.window(xlim = c(0, 6.5), ylim = c(0, 2.2), xaxs = "i", yaxs = "i")
angle <- seq(0, 2 * pi, length.out = 150)
for (k in 1:2) {
  centre <- c(1.55, 4.95)[k]
  polygon(centre + .72 * cos(angle), 1.7 + .32 * sin(angle),
          col = c("#E9F0F5", "#E9F3F0")[k], border = "#71818C")
  text(centre, 1.7, c("Confidence", "Belonging")[k], cex = .88)
  xs <- centre + c(-1.0, 0, 1.0)
  words <- list(c("Explaining", "Applying", "Checking"),
                c("Acceptance", "Inclusion", "Participation"))[[k]]
  for (j in 1:3) {
    arrows(centre + c(-.35, 0, .35)[j], 1.4, xs[j], .78,
           length = .065, col = "#71818C")
    rect(xs[j] - .48, .25, xs[j] + .48, .7,
         col = "#F7F8F8", border = "#71818C")
    text(xs[j], .475, words[j], cex = .80)
  }
}
arrows(2.35, 1.7, 4.15, 1.7, code = 3, length = .075, col = "#71818C")
```

**Language-learning progress.** The four equally spaced assessments
use a common 0-100 scale. Student A scores 58, 65, 68, and 76; Student B
scores 62, 63, 67, and 68. Their successive gains are 7, 3, and 8 points
and 1, 4, and 1 point, respectively. Student, classroom, school, and
occasion identifiers preserve the grouping and timing. These trajectories
show different starting levels, overall gains, and uneven changes
between assessments. A full growth study would also document other
students, scoring comparability, missed occasions, and changing learning
conditions.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

## References {.unnumbered}

::: {#refs}
:::
