---
title: "Group Comparisons with Analysis of Variance"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \widowpenalty=10000
          \clubpenalty=10000
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{12\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \usepackage{xcolor}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.pos = "H", collapse = TRUE, comment = "#>")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
feedback_data <- data.frame(
  student = 1:24,
  feedback = factor(rep(c("Rereading", "Answer feedback",
    "Explanatory feedback"), each = 8),
    levels = c("Rereading", "Answer feedback", "Explanatory feedback")),
  practice = factor(rep(rep(c("Massed", "Spaced"), each = 4), 3),
    levels = c("Massed", "Spaced")),
  score = c(4, 5, 7, 8, 4, 5, 7, 8,
    6, 7, 9, 10, 10, 11, 13, 14,
    8, 9, 11, 12, 16, 17, 19, 20))
massed <- subset(feedback_data, practice == "Massed")
one_way <- aov(score ~ feedback, data = massed)
full_fit <- aov(score ~ feedback * practice, data = feedback_data)
```

\clearpage

::: {.callout-note title="Chapter overview"}
This chapter compares means by examining variation. One invented study
provides the observations throughout: we first compare three groups,
then use the full two-factor layout to ask how a comparison changes
across practice schedules. We build the one-way ANOVA calculation from
individual scores, distinguish its omnibus question from a particular
contrast, and explain how the comparison family determines an interval's
coverage. Cell means and simple effects then reveal what main effects
can conceal. By the end, you will be able to reproduce an ANOVA table,
interpret a contrast and its uncertainty, and connect a factorial pattern
to the design and model conditions that support its interpretation.
:::

Twenty-four students study the same fractions material and take a common
20-item quiz. In this **invented study**, random allocation places four
different students in each combination of three feedback conditions and
two practice schedules. *Rereading* offers the worked material again;
*Answer feedback* supplies the correct answer after an attempt;
*Explanatory feedback* supplies the answer with a worked explanation.
Massed practice occurs in one session, and Spaced practice in two shorter
sessions with the same total practice time. Everyone takes the quiz at
the same specified follow-up point after their final practice. All design
details and scores are invented for teaching.

Our first question concerns the students assigned to Massed practice:
how do the three feedback-condition means compare? Chapter 3 developed
an estimate, SE, interval and test for one comparison. We extend that
reasoning to several groups, keeping the difference between an overall
pattern and particular comparisons visible. Later, the full table lets
us ask how the feedback comparison depends on practice schedule.

## Comparison Questions and Designs {#sec-ch04-design}

### Outcome, factors, cells and units

A quiz score is the number of correctly answered items, from 0 to 20.
Higher values represent more correct answers on this assessment. A
four-point difference therefore means four more correct answers on
average; it does not by itself describe every aspect of learning
fractions. The outcome and its scale give a numerical comparison its
substantive meaning.

A **factor** in analysis of variance is a categorical variable defining
groups or conditions. Its categories are **levels**. Feedback has three
levels and practice schedule has two. This use of *factor* differs from
a latent factor used to represent shared indicator variation in a
measurement model.

A **cell** contains cases sharing one level of each factor. Answer
feedback with Spaced practice is one cell. Crossing three feedback
levels with two practice levels gives six cells; four students per cell
makes the layout **balanced**. One-way and two-way refer to the number
of factors, not to the number of levels. A one-way analysis can compare
three, four or more groups.

\Needspace{10\baselineskip}

| Feedback condition | Massed practice | Spaced practice |
|:--|:--|:--|
| Rereading | 4, 5, 7, 8 | 4, 5, 7, 8 |
| Answer feedback | 6, 7, 9, 10 | 10, 11, 13, 14 |
| Explanatory feedback | 8, 9, 11, 12 | 16, 17, 19, 20 |

: Invented fractions-quiz scores, four different students per cell, on a 0--20 scale. {#tbl-ch04-feedback-scores tbl-colwidths="[38,31,31]"}

Each score in @tbl-ch04-feedback-scores belongs to a different student.
The student is both the stipulated assignment unit and the measurement
unit. The four scores in a cell are not repeated measurements on one
student; nor are the Massed and Spaced scores paired. The repeated
within-cell pattern deliberately simplifies the arithmetic. It is a
teaching construction, not evidence of Normal errors, equal variances
or an educational effect.

The complete course data file, `feedback_practice.csv`, represents the
same table in one row per student, with columns `student`, `feedback`,
`practice` and `score`. Students are numbered 1--24, ordered by feedback,
then practice, then student within each cell. The printed scores are
sufficient for every calculation in this chapter.

### A one-way slice and the full factorial question

For the first analysis, restrict attention to the twelve Massed-practice
students. There are three groups, defined by feedback alone. Their
model means, written $\mu_R$, $\mu_A$ and $\mu_E$, refer to Rereading,
Answer feedback and Explanatory feedback within this practice schedule.
For example, $\mu_A-\mu_R$ states a direction and an outcome scale.
The observed difference estimates this particular contrast under the
model introduced below.

This one-way analysis answers a prespecified question within a factorial
study. Its conclusion concerns the Massed slice. It does not silently
average over the Spaced students or establish that the same contrast
holds in both schedules. Returning to all six cells will allow the
second question: does the feedback comparison change with practice?

\Needspace{16\baselineskip}

| Analysis | Records used | Question addressed |
|:--|:--|:--|
| One-way | 12 Massed students; three feedback groups | Do the feedback means differ within Massed practice? |
| Two-way | All 24 students; six feedback-by-practice cells | What are the average factor comparisons, and how do comparisons change across the other factor? |

: Two questions using stated parts of the same invented study. {#tbl-ch04-layouts tbl-colwidths="[19,38,43]"}

A two-factor ANOVA and Chapter 3's two-way frequency table have different
outcomes. ANOVA compares a numerical score across factor combinations.
A frequency table counts cases in combinations of categories. The word
*two-way* describes a layout; it does not identify a shared statistical
test.

### Design determines the relevant comparison

Random allocation is stipulated in the invented story. If implemented
in a real study, that design could support a causal comparison among
the participating students, subject to adherence, relevant interference
and missing outcomes. It would not make the participants a random sample
of all students. The data table alone establishes neither implementation
of assignment nor representativeness.

Different designs can produce superficially similar tables. If whole
classes receive conditions, students within the same class may share
influences and the class may be the assignment unit. If each student is
measured twice, the two outcomes belong to a pair. An uncertainty formula
must reflect these connections rather than treating every row as an
independent assignment.

**Blocking** makes condition comparisons within groups of similar units.
Chapter 3's [paired-school reference](chapter_03.qmd#app-ch03-school-pairs)
gives the complete illustration: each school contributes two classes,
with one class assigned to each format. The school-specific differences
preserve that allocation structure. The six cells in our present quiz
example instead contain distinct individually assigned students.
Sampling strata, which organize selection into a sample, and allocation
blocks, which organize treatment comparisons, have different roles.

\Needspace{12\baselineskip}

## One-Way ANOVA {#sec-mean-comparison}

### Inspect grouped observations before reducing them

The three Massed groups have means 6.0000, 8.0000 and 10.0000 points.
Each has four scores and the same sample SD, 1.8257. Their ranges overlap:
for example, some Rereading scores exceed some Answer-feedback scores.
A difference between means is a summary of group distributions; it does
not say that every student in one group outperformed every student in
another.

\Needspace{12\baselineskip}

| Feedback | n | Mean | SD | Range |
|:--|--:|--:|--:|:--|
| Rereading | 4 | 6.0000 | 1.8257 | 4--8 |
| Answer feedback | 4 | 8.0000 | 1.8257 | 6--10 |
| Explanatory feedback | 4 | 10.0000 | 1.8257 | 8--12 |

: Descriptions of the twelve Massed-practice scores. {#tbl-ch04-group-summary tbl-colwidths="[36,10,18,18,18]"}

```{r}
#| label: fig-ch04-feedback-groups
#| fig-cap: "Massed-practice quiz scores by feedback condition. Every point is a different student; crosses mark group means."
#| fig-alt: "Each of three groups has four scores. Rereading scores 4, 5, 7, 8 center on 6; Answer feedback scores 6, 7, 9, 10 center on 8; Explanatory feedback scores 8, 9, 11, 12 center on 10. The ranges overlap."
#| fig-width: 7
#| fig-height: 3.5
#| out-width: 100%
par(mar = c(4, 9, 1, 1))
stripchart(score ~ feedback, data = massed,
  method = "stack", pch = 16, las = 1,
  xlab = "Quiz score (0-20)", ylab = "", xlim = c(0, 20))
points(tapply(massed$score, massed$feedback, mean), 1:3,
  pch = 4, cex = 1.7, lwd = 2)
```

With only four observations per group, individual dots communicate the
record more directly than a boxplot. A bar chart of means would hide the
within-group variation entirely. That variation matters to inference:
the same distance between group means can be more or less convincing
depending on how widely individual scores vary within groups.

```{r}
#| label: fig-ch04-separation-spread
#| fig-cap: "The same group means can provide different evidence when within-group spread changes. Counts and mean positions are held fixed in these constructed panels."
#| fig-alt: "Both panels have group means 6, 8 and 10 with four observations each. The second doubles each observation's distance from its group mean. Between-group variation stays fixed while within-group variation quadruples."
#| fig-width: 7
#| fig-height: 3.4
#| out-width: 100%
group_mean <- ave(massed$score, massed$feedback, FUN = mean)
wider <- group_mean + 2 * (massed$score - group_mean)
par(mfrow = c(1, 2), mar = c(4, 3, 3, 1))
for (panel in 1:2) {
  v <- list(massed$score, wider)[[panel]]
  stripchart(v ~ massed$feedback, vertical = TRUE,
    method = "stack", pch = 16, ylim = c(0, 16),
    group.names = c("Reread", "Answer", "Explain"),
    main = c("Original spread", "Doubled deviations")[panel],
    xlab = "Feedback", ylab = "Score")
  points(1:3, c(6, 8, 10), pch = 4, cex = 1.5, lwd = 2)
}
par(mfrow = c(1, 1))
```

In @fig-ch04-separation-spread the mean positions and counts stay fixed.
Only each score's distance from its group mean changes. Doubling those
distances multiplies their squares by four. The second panel therefore
has the same between-group separation but four times the within-group
squared variation. ANOVA compares these two kinds of variation after
putting them on comparable scales [@cetinkaya2024; @kutner2005].

### Mean model and deviation decomposition

Write $Y_{ij}$ for the random score of student $i$ in feedback group $j$.
Here $j$ indexes a group, rather than a data-file column. The one-way
model is

$$
Y_{ij}=\mu_j+\varepsilon_{ij}.
$$

The unknown $\mu_j$ is the group mean in the model, and
$\varepsilon_{ij}$ is that student's unknown departure from it. After
fitting the group means, the **fitted value** is
$\widehat y_{ij}=\bar y_j$, and the **residual** is
$e_{ij}=y_{ij}-\bar y_j$. A residual is calculated from observed scores
and fitted means; a model error is a departure from an unknown model
mean. This distinction will recur in regression.

The observed grand mean is $\bar y=(6+8+10)/3=8.0000$ because the
three groups are equally sized. For the Rereading student who scored 4,
the distance from the grand mean separates into two parts:

$$
4-8=(6-8)+(4-6)=-2+(-2).
$$

The group mean is two points below the grand mean, and this student is
another two points below the group mean. The general identity is

$$
y_{ij}-\bar y=(\bar y_j-\bar y)+(y_{ij}-\bar y_j).
$$

Squaring every deviation and adding over students gives a corresponding
partition of total squared variation:

$$
SS_T=SS_B+SS_W,
$$

$$
\begin{aligned}
SS_T&=\sum_{j=1}^J\sum_{i=1}^{n_j}(y_{ij}-\bar y)^2,\\
SS_B&=\sum_{j=1}^Jn_j(\bar y_j-\bar y)^2,\\
SS_W&=\sum_{j=1}^J\sum_{i=1}^{n_j}(y_{ij}-\bar y_j)^2.
\end{aligned}
$$

$SS_B$ is the **between-group sum of squares**: each group-mean
departure is counted once for each student in that group. $SS_W$ is
the **within-group sum of squares**: the remaining squared residuals
around those group means. The total $SS_T$ describes departures around
one grand mean. All three quantities are in squared quiz points.

Here the between-group calculation is

$$
SS_B=4(6-8)^2+4(8-8)^2+4(10-8)^2=32.0000.
$$

Each group's residuals are $-2,-1,1,2$, whose squared sum is
$4+1+1+4=10$. Thus

$$
SS_W=3(10)=30.0000,\qquad SS_T=32+30=62.0000.
$$

Squaring an individual deviation split introduces a cross-product.
Those cross-products cancel only after summing within each group,
because its residuals sum to zero. [Appendix A](#app-ch04-calculations)
shows that cancellation explicitly. The decomposition is arithmetic;
it needs no Normality or independence assumption. Those conditions
enter when the observed decomposition is used for inference.

### Degrees of freedom, mean squares and the F test

A sum of squares increases with both variation and the number of
contributing observations. **Mean squares** divide by the appropriate
degrees of freedom before the two sources are compared.

There are $J=3$ group-mean departures from the grand mean, but their
weighted sum is zero. Only $J-1=2$ departures can vary freely. Within
each group, four residuals sum to zero, leaving three free residuals.
Across three groups this gives $N-J=12-3=9$ within-group df. The total
has $N-1=11$ df, since all deviations from the grand mean sum to zero.
The component df add: $2+9=11$.

$$
MS_B=\frac{SS_B}{J-1}=\frac{32}{2}=16.0000,
\qquad
MS_W=\frac{SS_W}{N-J}=\frac{30}{9}=3.3333.
$$

The **F statistic** compares these mean squares:

$$
F=\frac{MS_B}{MS_W}=\frac{16}{30/9}=4.8000.
$$

\Needspace{16\baselineskip}

| Source | SS | df | MS | F | p |
|:--|--:|--:|--:|--:|--:|
| Feedback | 32.0000 | 2 | 16.0000 | 4.8000 | .0381 |
| Within groups | 30.0000 | 9 | 3.3333 | | |
| Total | 62.0000 | 11 | | | |

: One-way ANOVA for the 12 Massed-practice students. {#tbl-ch04-one-way-anova}

The omnibus null hypothesis is

$$
H_0:\mu_R=\mu_A=\mu_E.
$$

The alternative says that at least two of these means differ. Under
independent observations with Normal errors and a common error
variance $\sigma^2$, the null statistic has an $F(2,9)$ reference
distribution. Its two df describe the numerator and denominator.
Normality concerns the departures around the group means, not the
pooled mixture of all scores.

Under the null, both mean squares estimate the same error variance.
The numerator also responds to separation of the model means under
alternatives. A relatively large ratio is therefore evidence against
equality. The null does not force an observed ratio to equal 1, and
the expectation of the ratio is not exactly 1. Sampling variation
produces a distribution of possible positive ratios.

```{r}
#| label: fig-ch04-f-reference
#| fig-cap: "The one-way statistic 4.8000 and its right-tail probability under an F(2,9) null reference."
#| fig-alt: "A right-skewed F density begins at zero and declines across positive values. The area to the right of F=4.8 is shaded and labelled p=.0381."
#| fig-width: 6.6
#| fig-height: 3.1
xx <- seq(0, 12, length.out = 600)
par(mar = c(4, 4, 1, 1))
plot(xx, df(xx, 2, 9), type = "l", lwd = 2,
  xlab = "F statistic", ylab = "Density")
tail_x <- seq(4.8, 12, length.out = 200)
polygon(c(4.8, tail_x, 12), c(0, df(tail_x, 2, 9), 0),
  col = "grey80", border = NA)
abline(v = 4.8, lty = 2)
text(7.5, .22, "F = 4.8000\np = .0381")
```

The right-tail probability in @fig-ch04-f-reference is .0381. Under
the null and model conditions, a ratio at least as large as 4.8000
occurs with that probability. At a prespecified .05 significance
level the equality null is rejected. This is evidence about the
three-mean pattern, not a probability that the null is true and not
an identification of which particular comparisons differ.

Statistical software fits this group-mean model from the twelve Massed
records and reports the same table; the displayed arithmetic explains
what that table means. The Computer Lab supplies the software routes,
and Chapter 6 develops them further. A different software interface does not
change the need to identify the records, factor levels and error model.

### Read an ANOVA table {#sec-ch04-anova-table}

Every ANOVA table expresses one sentence about the observations:

$$
\text{Data}=\text{fit}+\text{residual}.
$$

The *fit* is what the estimated means account for, and the *residual*
is what is left around them. Squaring and adding turns that sentence
into the one-way partition already calculated, with a matching
partition of degrees of freedom:

$$
SS_T=SS_B+SS_W,\qquad df_T=df_B+df_W,
$$

$$
62.0000=32.0000+30.0000,\qquad 11=2+9.
$$

A balanced two-factor layout keeps the same sentence and divides the
fit part further, into two factor components and their interaction:

$$
SS_T=SS_A+SS_B+SS_{AB}+SS_E,\qquad
df_T=df_A+df_B+df_{AB}+df_E.
$$

For the complete study in @tbl-ch04-two-way-anova, this reads
$476.0000=256.0000+96.0000+64.0000+60.0000$ and $23=2+1+2+18$. The
fit part is therefore $256+96+64=416.0000$ on $2+1+2=5$ df. Both
additivity checks are useful arithmetic checks on any printed table.

The compact six-column table used here keeps one row for each source
of interest. Statistical software often prints a longer version of the
same decomposition, adding rows that serve the fitting algorithm rather
than the comparison question. @tbl-ch04-table-map maps one layout onto
the other.

\Needspace{20\baselineskip}

| Extended output row | One-way value here | Read or ignore |
|:--|:--|:--|
| Corrected Model | 32.0000 on 2 df | The whole fit part. With one factor it repeats the factor row; in the two-way layout it is 416.0000 on 5 df |
| Intercept | A grand-mean term | Ignore. It tests whether the overall mean is zero, which is not a comparison question for quiz scores |
| Factor rows | Feedback 32.0000 on 2 df | Read. In the two-way layout these are 256.0000, 96.0000 and 64.0000 on 2, 1 and 2 df |
| Error | 30.0000 on 9 df | Read. It supplies $MS_W$, or $MS_E$, and the denominator df of every F ratio |
| Total | Includes the grand-mean term | Ignore. This uncorrected total does not equal the sum of the rows above |
| Corrected Total | 62.0000 on 11 df | Read. It equals the fit plus error rows, 476.0000 on 23 df in the two-way layout |

: How the compact table rows correspond to an extended software layout. {#tbl-ch04-table-map tbl-colwidths="[24,29,47]"}

Read the factor rows, the error row and the corrected total; the
intercept and uncorrected total answer no question asked here. The
same reading order applies to the factorial table later in this
chapter, where the interaction row is examined before the two main
effects.

### Two groups, direction and magnitude

For exactly two independent groups under the common-variance model,
the classical pooled two-sample t test and one-way F test give

$$
F(1,N-2)=t_{N-2}^2.
$$

The tests have the same p-value when the t alternative is two-sided.
Squaring removes direction: positive and negative t values of the
same magnitude give the same F. Keep the signed mean difference
when explaining which group has the higher mean.

For a matched numerical example, use only the Massed Rereading and
Answer groups. Their mean difference, Rereading minus Answer, is
$6-8=-2.0000$. Each sample variance is $10/3$, so the pooled
variance is also $10/3$. These eight students give six residual df:

$$
t(6)=\frac{-2}{\sqrt{(10/3)(1/4+1/4)}}=-1.5492,
\qquad F(1,6)=2.4000,
$$

with the same two-sided $p=.1723$. This reduced fit differs from the
three-group analysis, which had nine residual df. A selected contrast
using that three-group error estimate is another calculation. The
identity also does not equate a Welch t procedure, the default in
some software, with a classical pooled ANOVA. [Appendix A](#app-ch04-pooled-identity) derives
the matched identity and Chapter 3 explains the different calibrations.

Magnitude begins on the quiz scale. The observed Answer-minus-Rereading
difference is 2.0000 points, and the Explanatory-minus-Rereading difference
is 4.0000 points within Massed practice. Whether a difference of this size
matters educationally needs a substantive criterion, not only a p-value.

One descriptive summary of the overall separation is **eta squared**:

$$
\eta^2=\frac{SS_B}{SS_T}=\frac{32}{62}=.5161.
$$

Group means account for .5161 of the observed squared variation in
these twelve records. This is a fraction of a sample decomposition,
not the fraction of learning caused by feedback. It does not replace
the actual group means or their pairwise differences. A standardized
mean difference similarly needs its denominator stated: dividing a
difference by an individual-score SD produces a different quantity
from dividing it by the SE of that difference. A bias-adjusted
omnibus summary, omega squared, is described briefly in
[Appendix B](#app-ch04-weighting).

\Needspace{12\baselineskip}

## Contrasts and Multiple Comparisons {#sec-ch04-contrasts}

### State the comparison through its weights

The omnibus question asks whether all three model means agree. A
**contrast** states a particular comparison as a weighted combination
of means whose weights sum to zero:

$$
C=\sum_{j=1}^Jc_j\mu_j,\qquad \sum_{j=1}^Jc_j=0,
\qquad \widehat C=\sum_{j=1}^Jc_j\bar y_j.
$$

The symbol $C$ here names a contrast, and $\widehat C$ its estimate.
The weights specify both the comparison and its direction. In the order
Rereading, Answer feedback, Explanatory feedback:

\Needspace{16\baselineskip}

| Comparison | Weights | Estimate |
|:--|:--|--:|
| Answer minus Rereading | $(-1,1,0)$ | $-6+8=2.0000$ |
| Explanatory minus Rereading | $(-1,0,1)$ | $-6+10=4.0000$ |
| Equal average of the two feedback approaches minus Rereading | $(-1,1/2,1/2)$ | $-6+4+5=3.0000$ |

: Three contrast meanings within Massed practice. {#tbl-ch04-weights tbl-colwidths="[49,23,28]"}

The third comparison gives the two feedback approaches equal weight.
It is not a substitute for both separate comparisons when both are
scientifically required. In unequal groups, equal method weights and
weights proportional to student counts answer different questions.
The present groups have equal counts, so the two averaging rules
coincide; [Appendix B](#app-ch04-weighting) develops the distinction.

A grand mean with weights $(1/3,1/3,1/3)$ is a useful linear combination
but not a contrast, because those weights sum to 1. Multiplying every
contrast weight by two doubles the estimate and SE. It preserves a
zero-null t statistic, but changes the quantity being reported. Naming
the comparison in ordinary words prevents a correct calculation from
answering the wrong question.

### Calculate the uncertainty for that comparison

Under the independent common-variance model, each estimated group mean
has variance $\sigma^2/n_j$. The variance of a weighted sum uses the
**squared weights**. Opposite signs define a difference; they do not
make independent sources of uncertainty cancel. Replacing $\sigma^2$
by the full three-group estimate $MS_W$ gives

$$
\widehat{SE}(\widehat C)=
\sqrt{MS_W\sum_{j=1}^J\frac{c_j^2}{n_j}}.
$$

For each Rereading-referenced difference,

$$
\widehat{SE}=
\sqrt{\frac{10}{3}\left(\frac14+\frac14\right)}
=1.2910\text{ points}.
$$

The three-group fit supplies nine error df even though each contrast
uses two of its means. The error variance was estimated from all three
groups under the common-variance assumption. This differs from refitting
only two groups, as in the t/F identity above.

For the equal-average contrast, the squared weights give

$$
\widehat{SE}=
\sqrt{\frac{10}{3}\left(\frac14+\frac{(1/2)^2}{4}
+\frac{(1/2)^2}{4}\right)}=1.1180.
$$

This smaller SE belongs to a different estimand, an average of two
means compared with the reference. It does not establish that either
separate comparison is more precise than before. The
[NIST contrast explanation](https://www.itl.nist.gov/div898/handbook/prc/section4/prc436.htm)
develops the same weighted-mean variance calculation.

### Declare the family before choosing the critical value

Suppose the planned family contains exactly two two-sided comparisons:
Answer minus Rereading and Explanatory minus Rereading. We want an
interval procedure with at least 95% **simultaneous coverage** for
both target contrasts under the model.

Bonferroni allocates the .05 error allowance across these two
intervals, giving .025 to each interval and .0125 to each tail. Its
critical value is therefore

$$
t^*=t_{1-.05/(2\times2);9}=t_{.9875;9}=2.6850.
$$

Using unrounded values, each margin is
$2.685010847(1.290994449)=3.4663$. The intervals are
$\widehat C\pm t^*\widehat{SE}$:

\Needspace{10\baselineskip}

| Comparison | Estimate | SE | Family lower | Family upper |
|:--|--:|--:|--:|--:|
| Answer minus Rereading | 2.0000 | 1.2910 | -1.4663 | 5.4663 |
| Explanatory minus Rereading | 4.0000 | 1.2910 | 0.5337 | 7.4663 |

: Bonferroni 95% simultaneous intervals for the two stated comparisons in Massed practice. {#tbl-ch04-contrasts tbl-colwidths="[36,16,16,16,16]"}

The two estimates share the Rereading mean and are therefore related.
Bonferroni's bound does not require their independence: the probability
of at least one coverage failure is no larger than the sum of the two
individual failure probabilities. The guarantee concerns repeated use
of the whole stated procedure, not a 95% probability assigned to the
two realized intervals. The [NIST Bonferroni reference](https://www.itl.nist.gov/div898/handbook/prc/section4/prc473.htm)
explains this joint-coverage bound.

**Tukey's method** answers a different family question: simultaneous
coverage for all three pairwise mean differences. Its intervals and
adjusted p-values for these data are:

\Needspace{12\baselineskip}

| Pairwise comparison | Estimate | Lower | Upper | Adjusted p |
|:--|--:|--:|--:|--:|
| Answer minus Rereading | 2.0000 | -1.6045 | 5.6045 | .3150 |
| Explanatory minus Rereading | 4.0000 | 0.3955 | 7.6045 | .0310 |
| Explanatory minus Answer | 2.0000 | -1.6045 | 5.6045 | .3150 |

: Tukey 95% simultaneous intervals for all three pairs. {#tbl-ch04-tukey tbl-colwidths="[36,16,16,16,16]"}

The intervals differ from Bonferroni's because both the protected family
and its calibration differ. Statistical software returns these all-pairs comparisons from the
fitted one-way model; the Computer Lab supplies the software routes. Dunnett's method
addresses comparisons with a common reference; Scheffé's method can
protect all contrasts in a specified mean-comparison space. The choice
comes from the comparisons for which coverage is owed, not from which
method happens to exclude zero [@tukey1953; @kutner2005].

A contrast is a mathematical form. It can be defined before relevant
results are seen or selected after examining them. Selection matters
because looking across possibilities can favor unusually large observed
differences. A reported selected comparison needs uncertainty appropriate
to that selection. Nor is rejection of an omnibus test a universal
prerequisite for a prespecified comparison or a simultaneous interval.
State the question, the family, the selection process and the calibration.

### Read the omnibus and particular results together

The three-group F test concerns equality of all three model means.
The two Bonferroni intervals concern the two narrower reference
comparisons. The Explanatory-minus-Rereading interval excludes zero;
the Answer-minus-Rereading interval includes zero. The latter remains
compatible with both negative and positive differences and does not
establish equivalence. No practical equivalence margin was declared.

A connected report would say that, in the Massed slice, observed means
were 6.0000, 8.0000 and 10.0000 points. The omnibus common-variance
Normal-model test gave $F(2,9)=4.8000$, $p=.0381$. For the two
declared reference comparisons, the estimates were 2.0000 and 4.0000
points, with simultaneous intervals $[-1.4663,5.4663]$ and
$[0.5337,7.4663]$. These statements preserve magnitude, uncertainty,
family and the restriction to Massed practice. They describe calculations
for invented scores, not an empirical educational finding.

## Two-Way ANOVA {#sec-ch04-two-way}

### What interaction means, and why study two factors at once {#sec-ch04-interaction-idea}

Two factors **interact** when the effect of one factor differs across
the levels of the other. Nothing more technical is required to state
the idea. A combination of levels then produces a result that the two
separate comparisons would not have led us to expect.

Here is one invented illustration, outside the quiz study. Suppose a
short vocabulary preview is given before a reading passage, and that
students are also grouped by whether they began with weak or strong
vocabulary. The preview raises comprehension scores substantially for
the students who began weak, and barely changes them for the students
who began strong. The effect of the preview *depends on* prior
vocabulary: that dependence is the interaction. Describing only an
average preview effect would tell a reader that the preview helps
somewhat, and would conceal that it helps one group a great deal and
the other hardly at all.

Studying two factors together in one crossed design has three
advantages over running two separate one-factor studies.

1. **It is more efficient.** Every student contributes to both factor
   comparisons at once, because the levels of one factor are spread
   across the levels of the other. A crossed design therefore needs
   fewer cases per condition than a series of separate one-factor
   studies aimed at the same comparisons.
2. **It reduces residual variation.** A second factor that influences
   the outcome moves variation out of the error term and into the fit.
   If all 24 students had been analyzed with feedback as the only
   factor, the practice and interaction components, 96.0000 and
   64.0000, would have stayed inside the residual term instead of
   leaving the 60.0000 within-cell error used below.
3. **It makes interaction examinable.** Only a crossed design supports
   the question of whether a comparison changes across the other
   factor's levels. A series of one-factor studies supplies no
   estimate of that change, however carefully each one is run.

Adding practice schedule to the quiz study therefore gives each
comparison a context. We can compare feedback conditions within Massed
practice, within Spaced practice, or after averaging over the two
schedules. These are distinct quantities. A **main effect** compares
one factor's levels after averaging across the other factor with
stated weights. A **simple effect** makes the comparison at a
specified level of the other factor. **Interaction** means that a
comparison changes across those levels.

### Main effects and interaction through graphs

In a two-factor design either main effect, the interaction, or any
combination of the three can be present. Six schematic two-by-two
patterns make the whole range visible before we return to the full
three-by-two data. Each line in @fig-ch04-interaction-patterns
represents a feedback level, A or B; the horizontal positions
represent Massed and Spaced practice. These are mean-only drawings,
not additional observed studies, and they carry no significance claim.

```{r}
#| label: fig-ch04-interaction-patterns
#| fig-cap: "Six schematic patterns of cell means, covering the whole range of main-effect and interaction combinations. Lines identify feedback levels A and B; the drawings illustrate averages and changing differences, without significance claims."
#| fig-alt: "Six panels. No effects: A and B coincide as one flat line at 10. One main effect: two separated flat lines at 8 and 12. Two main effects: parallel rising lines. Interaction only: crossing lines with equal marginal means. Interaction qualifies: both lines rise from 8, B more steeply. Interaction reverses: A rises from 8 to 10 while B falls from 16 to 10."
#| fig-width: 7
#| fig-height: 4.2
#| out-width: 100%
patterns <- list(
  "No effects" = matrix(c(10, 10, 10, 10), 2, byrow = TRUE),
  "One main effect" = matrix(c(8, 8, 12, 12), 2, byrow = TRUE),
  "Two main effects" = matrix(c(8, 10, 12, 14), 2, byrow = TRUE),
  "Interaction only" = matrix(c(8, 14, 14, 8), 2, byrow = TRUE),
  "Interaction qualifies" = matrix(c(8, 10, 8, 14), 2, byrow = TRUE),
  "Interaction reverses" = matrix(c(8, 10, 16, 10), 2, byrow = TRUE))
par(mfrow = c(2, 3), mar = c(3.6, 3.6, 2.6, .8), mgp = c(2.2, .7, 0))
for (nm in names(patterns)) {
  matplot(1:2, t(patterns[[nm]]), type = "b", pch = c(16, 1),
    lty = c(1, 2), col = c("black", "grey40"),
    ylim = c(6, 19), xlim = c(.85, 2.15), xaxt = "n",
    xlab = "Practice", ylab = "Mean score", main = nm)
  axis(1, at = 1:2, labels = c("Massed", "Spaced"))
  legend("topleft", c("A", "B"), pch = c(16, 1),
    lty = c(1, 2), col = c("black", "grey40"), bty = "n",
    cex = .8, horiz = TRUE)
}
par(mfrow = c(1, 1))
```

Read each pattern by taking differences, not only by recognizing a
shape. Define a schedule difference as Spaced minus Massed and a
difference of differences as B's schedule difference minus A's.

\Needspace{16\baselineskip}

| Pattern | A: Massed, Spaced | B: Massed, Spaced | Schedule differences A, B | Difference of differences |
|:--|:--|:--|:--|--:|
| No effects | 10, 10 | 10, 10 | 0, 0 | 0 |
| One main effect | 8, 8 | 12, 12 | 0, 0 | 0 |
| Two main effects | 8, 10 | 12, 14 | 2, 2 | 0 |
| Interaction only | 8, 14 | 14, 8 | 6, -6 | -12 |
| Interaction qualifies | 8, 10 | 8, 14 | 2, 6 | 4 |
| Interaction reverses | 8, 10 | 16, 10 | 2, -6 | -8 |

: Arithmetic behind the six schematic patterns. {#tbl-ch04-patterns tbl-colwidths="[27,17,17,21,18]"}

In the first panel nothing varies: both feedback levels and both
schedules have mean 10, and the two lines coincide. In the second,
B exceeds A by 4 points under either schedule while the schedule
comparison is 0 under either feedback level; there is one main
effect and no interaction. In the third, B still exceeds A by 4
points everywhere, and Spaced now exceeds Massed by 2 points
everywhere: two main effects, again with a zero difference of
differences. Parallel lines express those constant comparisons.

The last three panels have a nonzero difference of differences. In
the crossing panel, both feedback marginal means are 11 and both
practice marginal means are 11 when cells receive equal weight. Both
averaged comparisons are zero despite large conditional differences,
which is why a small main effect need not mean that a factor is
irrelevant at every level of another factor.

In the qualifying panel the schedule comparison is 2 for A and 6 for
B. Both are positive, so the averaged 4-point schedule comparison
points the right way for both feedback levels but describes neither
exactly. The lines need not cross for interaction to be present. In
the reversing panel the schedule comparison is 2 for A and $-6$ for
B, so the averaged comparison of $-2$ points the wrong way for A.
Reporting only the two main effects would then mislead a reader
about every conditional comparison. Examine the interaction first,
and read a main effect as an average whose parts are still visible.

These drawings establish only the numerical patterns: significance
also needs information about variation, sample size and the
reference distribution.

### Cells, marginal means and the full data

Return to all 24 quiz scores. Each of the six cells contains four
different students; no person contributes a score under the other
practice schedule. The first one-way analysis used one three-cell
slice, whereas the full analysis uses all six groups.

Write $\bar y_{jk}$ for a cell mean, with $j=1,2,3$ indexing feedback
and $k=1,2$ indexing practice. A dot replaces an index that has been
averaged over: $\bar y_{j\cdot}$ averages a feedback level over
practice, and $\bar y_{\cdot k}$ averages a practice level over
feedback.

\Needspace{13\baselineskip}

| Feedback | Massed | Spaced | Feedback mean |
|:--|--:|--:|--:|
| Rereading | 6.0000 | 6.0000 | 6.0000 |
| Answer feedback | 8.0000 | 12.0000 | 10.0000 |
| Explanatory feedback | 10.0000 | 18.0000 | 14.0000 |
| Practice mean | 8.0000 | 12.0000 | Grand mean 10.0000 |

: Cell and equally weighted marginal means for the complete invented study. {#tbl-ch04-cell-means tbl-colwidths="[37,19,19,25]"}

Answer feedback averages $(8+12)/2=10.0000$ over schedules. Since
cell sizes are equal, this equals the mean of its eight student
scores. The Massed marginal mean is $(6+8+10)/3=8.0000$, and the
Spaced marginal mean is $(6+12+18)/3=12.0000$. The grand mean is
10.0000.

The practice comparison averages $12-8=4.0000$ points across the
three feedback conditions with equal weights. Its component schedule
differences are 0.0000, 4.0000 and 8.0000. A four-point main effect
is therefore a meaningful stated average, but it describes none of
the conditional comparisons except Answer feedback's.

```{r}
#| label: fig-ch04-cell-means
#| fig-cap: "Quiz cell means across practice schedules in the full 3 × 2 example. Each line joins means of different students; the schedule differences are 0, 4 and 8 points."
#| fig-alt: "Rereading stays at 6 from Massed to Spaced practice. Answer feedback rises from 8 to 12. Explanatory feedback rises from 10 to 18. The nonparallel lines show that the schedule comparison changes across feedback conditions."
#| fig-width: 7
#| fig-height: 3.8
#| out-width: 100%
par(mar = c(4, 4, 1, 1), cex = .9)
with(feedback_data,
  interaction.plot(practice, feedback, score, type = "b",
    pch = c(16, 1, 17), lty = 1:3,
    col = c("#202020", "#2B6CA3", "#9B4922"),
    xlab = "Practice schedule", ylab = "Mean quiz score",
    trace.label = "Feedback", ylim = c(0, 20)))
```

The connected lines in @fig-ch04-cell-means organize cell means. They
do not show individual students moving from one condition to another.
The nonparallel pattern describes interaction in the observed means;
the full model supplies the uncertainty calculation for an inferential
claim.

### Additivity and the full model

An additive prediction combines the grand mean with the two marginal
departures. For Answer-Spaced, it is

$$
10+(10-10)+(12-10)=12,
$$

which agrees with that observed cell mean. For Explanatory-Spaced,
the additive prediction is

$$
10+(14-10)+(12-10)=16,
$$

while the cell mean is 18. The extra 2 points are an **interaction
departure** from that additive prediction. Agreement for one cell
does not imply that the entire layout is additive.

The full mean model includes such departures:

$$
Y_{ijk}=\mu+\alpha_j+\beta_k+(\alpha\beta)_{jk}
        +\varepsilon_{ijk}.
$$

Here $i=1,\ldots,4$ identifies a student within a cell, $j=1,2,3$
identifies feedback and $k=1,2$ identifies practice. The grand model
mean is $\mu$; $\alpha_j$ and $\beta_k$ are the factor effects;
$(\alpha\beta)_{jk}$ is the interaction departure; and
$\varepsilon_{ijk}$ is the individual error around the cell mean.
These indexed effects are distinct from the unindexed error
probabilities $\alpha$ and $\beta$ used in hypothesis testing.

In the balanced decomposition, main-effect departures sum to zero.
Interaction departures sum to zero along each row and each column,
so they contain only the remaining pattern after the two main effects.
The observed feedback departures are $-4,0,4$, and practice departures
are $-2,2$. The interaction departures are:

\Needspace{12\baselineskip}

| Feedback | Massed | Spaced |
|:--|--:|--:|
| Rereading | 2 | -2 |
| Answer feedback | 0 | 0 |
| Explanatory feedback | -2 | 2 |

: Observed interaction departures from the additive prediction. {#tbl-ch04-interaction-departures tbl-colwidths="[50,25,25]"}

With two columns, choosing a departure in the first column fixes
the opposite departure in the second. Of the three first-column
departures, only two are free because they sum to zero. Thus the
interaction has $(3-1)(2-1)=2$ df. It describes a two-dimensional
pattern, rather than one particular four-cell difference.

For a common cell size $n=4$, the total squared variation splits as

$$
SS_T=SS_A+SS_B+SS_{AB}+SS_E.
$$

In this **two-way** table, A denotes feedback and B denotes practice.
Thus $SS_B$ now means the practice component, not the entire
between-group component of the one-way table. $SS_E$ is the
within-cell error sum of squares. Stating the local meanings keeps
the two familiar uses of B distinct.

$$
\begin{aligned}
SS_A&=2n\sum_j(\bar y_{j\cdot}-\bar y)^2=256.0000,\\
SS_B&=3n\sum_k(\bar y_{\cdot k}-\bar y)^2=96.0000,\\
SS_{AB}&=n\sum_j\sum_k
 (\bar y_{jk}-\bar y_{j\cdot}-\bar y_{\cdot k}+\bar y)^2
 =64.0000,\\
SS_E&=\sum_j\sum_k\sum_i(y_{ijk}-\bar y_{jk})^2=60.0000.
\end{aligned}
$$

For example, the feedback component counts each marginal departure
eight times: $8\{(-4)^2+0^2+4^2\}=256$. The practice component
counts each departure twelve times: $12\{(-2)^2+2^2\}=96$. Four
students share each interaction departure, giving
$4(4+4+0+0+4+4)=64$. Each cell contributes 10 residual squared
points, giving $6(10)=60$. Their sum is $SS_T=476$.

Each of six cells has an estimated mean, so error df are
$24-6=18$. Feedback has two df, practice has one, and interaction
has two. Together they give $2+1+2+18=23=N-1$.

\Needspace{14\baselineskip}

| Source | SS | df | MS | F | p |
|:--|--:|--:|--:|--:|--:|
| Feedback (A) | 256.0000 | 2 | 128.0000 | 38.4000 | <.0001 |
| Practice (B) | 96.0000 | 1 | 96.0000 | 28.8000 | <.0001 |
| A × B | 64.0000 | 2 | 32.0000 | 9.6000 | .0015 |
| Error | 60.0000 | 18 | 3.3333 | | |
| Total | 476.0000 | 23 | | | |

: Full two-way ANOVA using all 24 students. {#tbl-ch04-two-way-anova}

Read this table with the reading order set out in
[Read an ANOVA table](#sec-ch04-anova-table): the three source rows,
the error row and the corrected total. An extended software layout
prints the same components with a Corrected Model row of 416.0000 on
5 df, plus an intercept row and an uncorrected total that answer no
question asked here.

Every displayed F divides its source mean square by the same
$MS_E=60/18=10/3$. The corresponding null hypotheses address
different features of the model:

\Needspace{16\baselineskip}

| Source | Null hypothesis | Comparison represented |
|:--|:--|:--|
| Feedback | $\mu_{R\cdot}=\mu_{A\cdot}=\mu_{E\cdot}$ | Equal feedback means, each averaging the two schedules equally |
| Practice | $\mu_{\cdot M}=\mu_{\cdot S}$ | Equal schedule means, each averaging the three feedback levels equally |
| Interaction | $(\alpha\beta)_{jk}=0$ for every cell | Each feedback comparison is constant across schedules, equivalently each schedule difference is the same across feedback levels |

: Three null hypotheses with their averaging rules. {#tbl-ch04-factorial-nulls tbl-colwidths="[17,37,46]"}

Independent common-variance Normal errors supply the classical F
references. Under these conditions, the interaction result
$F(2,18)=9.6000$, $p=.0015$ gives evidence against an additive
pattern. Both main effects also have small p-values, but their
meanings remain averaged comparisons. Reading them as “each factor
works equally well everywhere” would discard the interaction.

The clean component decomposition uses the balanced layout. With
unequal cell counts, the averaging target and the model comparison
need more care; different software sums of squares can address
different hypotheses. The [weighting reference](#app-ch04-weighting)
explains the first distinction without changing these data.

### Simple effects and a selected interaction contrast

The Answer-feedback schedule comparison is a **simple effect**:
$12-8=4.0000$ points. Its SE under the full model is

$$
\widehat{SE}=\sqrt{\frac{10}{3}
 \left(\frac14+\frac14\right)}=1.2910.
$$

With 18 residual df, a pointwise 95% interval is

$$
4\pm t_{.975,18}\sqrt{5/3}
\approx[1.2877,6.7123].
$$

This interval is calibrated for the one stated simple comparison.
It is not protected for every simple effect that a reader might
inspect after seeing the graph. The two means come from different
students, so this is an independent-cell calculation, not a paired
calculation.

Now suppose a reader selects the Answer-versus-Rereading change in
the schedule comparison after inspecting the pattern. In model means,
the target is

$$
C=(\mu_{A,S}-\mu_{A,M})-(\mu_{R,S}-\mu_{R,M}).
$$

Its estimate is one **difference of differences**:

$$
\widehat C=(12-8)-(6-6)=4.0000.
$$

The four independent cell means have weights $1,-1,-1,1$ in the
displayed expression. Each has variance estimated by $MS_E/4$;
squaring the weights and adding gives

$$
\widehat{SE}(\widehat C)=
\sqrt{\frac{10}{3}\left(\frac14+\frac14+\frac14+\frac14\right)}
=1.8257.
$$

To account for selection among interaction contrasts, choose
Scheffé coverage for **all contrasts in the two-dimensional
interaction space**. The multiplier is
$\sqrt{2F_{.95;2,18}}=2.6663$. The complete simultaneous calculation is

$$
\begin{aligned}
\widehat C\pm\sqrt{2F_{.95;2,18}}\widehat{SE}(\widehat C)
 &=4\pm2.666292237\sqrt{10/3}\\
 &\approx[-0.8680,8.8680].
\end{aligned}
$$

The interval includes zero even though the omnibus interaction
p-value is .0015. There is no contradiction. The joint test assesses
the entire two-dimensional departure from additivity; the selected
contrast concerns one direction, with protection for choosing among
all interaction contrasts. Evidence that some interaction direction
is nonzero does not establish every selected direction.

The simple Answer schedule effect includes the practice main-effect
component as well as an interaction component. It therefore lies
outside this interaction-only Scheffé family. A claim covering all
simple effects would need a corresponding family and calibration.
[Appendix B](#app-ch04-families) compares the selected contrast's
pointwise and simultaneous intervals using the same estimate and SE.

### Give a connected interpretation

The observed schedule differences are 0.0000, 4.0000 and 8.0000
quiz points for Rereading, Answer feedback and Explanatory feedback.
The full-model interaction test gives $F(2,18)=9.6000$, $p=.0015$
under the stated conditions. The selected Answer-versus-Rereading
change in schedule difference is 4.0000 points, with a 95% Scheffé
interval $[-0.8680,8.8680]$ over all interaction contrasts. The
overall result does not establish that particular selected comparison.

Feedback and practice marginal means summarize explicit equal
weights. The cell means show what those averages can conceal. The
one-way conclusion concerns only the Massed students, whereas the
full factorial conclusion uses all 24 students and 18 error df.
These are analyses of invented data, not empirical evidence that a
teaching method succeeds in a population of students.

## Conditions and Interpretation {#sec-ch04-conditions}

### Explain the conditions in familiar terms

Three conditions support the classical ANOVA references used here:
**independent observations**, **similar variances within groups or
cells**, and **approximately Normal residuals**. More precisely,
the exact finite-sample F and t results assume independent Normal
errors with a common variance around the relevant model means.
Residuals provide observable clues about those errors, together with
the design and substantive knowledge [@kutner2005].

Independence is principally a design question. It concerns how
outcomes can share influences, including repeated observations,
classroom clustering or interference between students. A residual
plot cannot establish independent assignment or independent errors.
Similar within-cell spread is a different question: does a common
error variance reasonably describe each cell? The pooled spread of
all outcomes need not be small or unimodal, because the groups can
have different means.

```{r}
#| label: fig-ch04-residuals
#| fig-cap: "Residuals around the fitted cell means. The repeated vertical pattern was deliberately built into the invented scores; it is not independent evidence that the model conditions hold."
#| fig-alt: "At each distinct fitted cell mean, residuals are minus two, minus one, one and two. Some points overlap because two cells share a fitted mean. The repeated pattern is a feature of the teaching construction."
#| fig-width: 6.6
#| fig-height: 3.2
par(mar = c(4, 4, 1, 1))
plot(fitted(full_fit), residuals(full_fit),
  xlab = "Fitted cell mean", ylab = "Residual", pch = 16)
abline(h = 0, lty = 2)
```

A residual-versus-fitted display can reveal changing spread or an
unusual residual. Here each cell has residuals $-2,-1,1,2$, because
we deliberately constructed the scores that way. Four students per
cell provide very little information for judging a distribution's
tails. Identical plotted spreads in this example do not validate a
Normal model for real quiz scores.

A Normal quantile plot compares ordered residuals with quantiles of
a Normal distribution. It can make pronounced tail departures more
visible, but a small apparently straight display offers limited
reassurance. Treat such graphs as information to interpret; a
sequence of Normality or variance tests should not become an
automatic permission rule for ANOVA.

\Needspace{25\baselineskip}

| Question | Information to examine | Implication for the calculation |
|:--|:--|:--|
| Are observations independent? | Assignment unit, repeated measurements, classes and schools | Dependence changes the covariance and relevant error structure |
| Is a common variance reasonable? | Group/cell spreads, residual displays and subject-matter knowledge | Unequal-variance procedures use a different uncertainty calibration |
| Is the error shape adequate for the reference? | Residual distribution, unusual observations and cell sizes | Small samples can be sensitive to departures from the stated reference model |
| Does the comparison fit the question? | Outcome, factor order, selected records and averaging weights | A correct calculation can otherwise describe a different target |

: Conditions connect design and observed patterns to interpretation. {#tbl-ch04-conditions tbl-colwidths="[28,35,37]"}

Welch's one-way method allows unequal group variances and changes
the uncertainty calibration. It is not obtained by swapping one
number in the classical sum-of-squares table. It does not solve
clustering, and its one-way form does not supply the full factorial
interaction analysis. Chapter 6 shows the alternative alongside
the common-variance calculation, with its conditions kept visible.

### Missing outcomes and the scope of the answer

A report should compare assigned and analyzed counts and explain
known reasons for losses. Missing outcomes can alter who contributes
to a comparison and can require assumptions beyond the displayed
ANOVA model. Our invented dataset has all 24 scores, so it supplies
no missingness experiment or evidence about a deletion rule. In a
real study, silently analyzing only available records would need
justification at the intended target.

Keep three claims distinct. The observed means and sums of squares
describe these records. Intervals and tests express uncertainty
under a specified model or design-based approximation. Causal and
population interpretations require support from allocation,
measurement, participation and the intended scope. A small p-value
does not complete that support, and a large p-value does not prove
that an effect is absent or unimportant.

## Reading Bridge {#sec-ch04-bridge}

Comparing several means extends the estimate-and-SE reasoning of
Chapter 3. ANOVA separates sources of variation to ask an omnibus
question; a contrast states a narrower question and its uncertainty.
Main effects average across factor levels, while cell means and
simple effects show the conditional pattern. A joint interaction
test and one selected interaction contrast can therefore lead to
different, compatible conclusions.

The same group-mean model can be written using indicator variables.
A measured baseline can also enter an adjusted model, called
analysis of covariance (ANCOVA). Chapter 6 gives their first complete
applications, with stated comparison targets and conditions.
Chapters 7--8 develop the regression mathematics, several controls
and standardized predictions. Baseline adjustment accounts for a measured
source of score variation and compares fitted outcomes at a stated common
baseline value, or averages them over the same stated baseline distribution.
The groups need adequate overlap in those baseline values, often called
common support, for those comparisons to be informed by observations.
Adjustment is not a general cure for confounding or missing outcomes.

Blocking organizes comparisons within allocation groups, as in
Chapter 3's paired-school reference. Repeated measurements and
classroom clustering change the covariance structure and the
relevant error term. The shape of a data table cannot settle these
design questions. Chapters 5--6 now make the statistical foundations
executable while keeping the question, records and interpretation
beside each calculation.

### Key terms {.unnumbered}

Use this reference to revisit the concepts principally explained here.
The Part One introduction collects earlier terms, including estimand,
SE, confidence interval and familywise error rate.

```{=latex}
\small
```

| Term and common alternatives | Meaning |
|:--|:--|
| Factor (grouping factor) | In ANOVA, a categorical variable defining groups or conditions; its categories are levels. This meaning differs from a latent factor used to model shared indicator variation. |
| Analysis of variance (ANOVA) | A method for studying mean differences through sources of variation. One-way and two-way ANOVA use one and two grouping factors, respectively; either factor can have several levels. |
| Fitted value | The value a fitted model assigns to an observed case. In a one-way group-mean model, it is that case's group mean. |
| Residual | An observed outcome minus its fitted value. A residual is calculated from the fitted model; an error is the corresponding unknown departure in the population model. |
| Mean square (MS) | A sum of squares divided by its degrees of freedom. In classical ANOVA, the within-group mean square estimates the common error variance under the model's conditions. |
| F statistic (F ratio) | In classical ANOVA, a ratio of the mean square for a tested source to the appropriate error mean square. Its null F reference depends on the model and both degrees of freedom. |
| Omnibus test | A test of one joint null hypothesis about a whole pattern, such as equality of all group means. Rejecting it does not identify which comparisons differ or how large they are. |
| Standardized mean difference | A difference in means divided by a specified SD. The chosen denominator determines the standardized quantity; the raw difference retains the outcome's original units. |
| Eta squared | The proportion of total observed variation attributed to a specified source in an ANOVA decomposition. In one-way ANOVA, it is between-group divided by total sum of squares. |
| Contrast (linear contrast) | A weighted combination of means whose weights sum to zero, expressing a specific comparison. The means, weights, and target population determine the quantity compared. |
| Simultaneous confidence intervals | A set of intervals calibrated so that, under their conditions, all of them cover their estimands together at the stated familywise rate; Bonferroni, Tukey, and Dunnett are different calibrations for different families. |
| Cell mean and marginal mean | In a two-factor layout, a cell mean is the mean of the cases sharing one level of each factor; a marginal mean averages the cell means across the levels of the other factor, with stated weights. |
| Main effect | A comparison across one factor's levels averaged over the other factor's levels, using stated weights. It remains an average when the conditional pattern differs across levels. |
| Interaction | A pattern in which a comparison across one factor's levels changes with the other factor's level. A difference of differences expresses one such comparison. |
| Simple effect | A comparison across one factor's levels at a specified level of the other factor. It describes a conditional comparison within the factorial design. |

: Selected concepts for group comparisons and ANOVA. {#tbl-ch04-key-terms tbl-colwidths="[28,72]"}

```{=latex}
\normalsize
```

### Further reading {.unnumbered}

- *Introduction to Modern Statistics* develops the transition from
  grouped observations to [inference for many means](https://openintro-ims.netlify.app/inference-many-means)
  [@cetinkaya2024].
- [Penn State STAT 200, Lesson 10](https://online.stat.psu.edu/stat200/Lesson10)
  provides an introductory ANOVA treatment; its
  [factorial-design lesson](https://online.stat.psu.edu/stat503/Lesson05)
  develops main effects and interaction.
- The NIST handbook explains [contrasts and their SEs](https://www.itl.nist.gov/div898/handbook/prc/section4/prc436.htm)
  and [Bonferroni simultaneous intervals](https://www.itl.nist.gov/div898/handbook/prc/section4/prc473.htm).

## Exercises {.unnumbered}

The supplied scores and constants support calculator-based work.
Retain full precision until reporting calculated results to four
decimal places. State the relevant records, conditions and comparison
direction in each interpretation.

<!-- prompt-id: CH04-E01 -->
### Exercise 1: Read grouped observations (Foundational) {.unnumbered #ex-ch04-01}

Use the twelve Massed-practice scores from the invented study:

- Rereading: 4, 5, 7, 8.
- Answer feedback: 6, 7, 9, 10.
- Explanatory feedback: 8, 9, 11, 12.

1. Calculate each mean, range and sample SD. Compare the centers,
   spread and overlap. Use the sample-variance denominator $n_j-1$.
2. Draw or describe a dotplot displaying every score. Explain what
   it reveals that a bar chart showing only the means would conceal.
3. Explain why the identical within-group spreads do not establish
   that a common-variance Normal model will hold in a real study.
4. Identify the assignment and measurement units stipulated in this
   invented study. Explain what the scores alone cannot establish
   about causal interpretation or generalization.

<!-- prompt-id: CH04-E02 -->
### Exercise 2: Reproduce the one-way calculation (Foundational) {.unnumbered #ex-ch04-02}

Use Exercise 1's scores. There are $J=3$ groups and $N=12$ students.
The group means are 6, 8 and 10 and the grand mean is 8. For the
classical test, assume independent Normal errors with a common variance.

1. For the Rereading score of 4, split its deviation from the grand
   mean into a group-mean departure and a residual. Calculate the
   squared total deviation, the two squared components and their
   cross-product. Explain why the sum-of-squares identity applies
   after summing, rather than to those two squares alone for every
   individual.
2. Calculate $SS_B$, $SS_W$ and $SS_T$ from the scores. Explain
   why the within-group contribution is the same for all three groups.
3. Find all three df, both mean squares and the F statistic. Explain
   the constraint behind each df and what the ratio compares.
4. State the omnibus null. Interpret the supplied right-tail
   $p=.0381$ at $\alpha=.05$, including the model conditions and
   what the conclusion leaves unanswered about particular comparisons.

<!-- prompt-id: CH04-E03 -->
### Exercise 3: Define and estimate contrasts (Foundational) {.unnumbered #ex-ch04-03}

The Massed group means are 6, 8 and 10 in the order Rereading, Answer
feedback, Explanatory feedback. Each group has four students and the
common-variance estimate is $MS_W=10/3$, with nine residual df.

1. Identify which weight vectors define contrasts:
   $(-1,1,0)$, $(-1,0,1)$, $(-1,1/2,1/2)$ and
   $(1/3,1/3,1/3)$. State the ordinary-language question represented
   by each contrast.
2. Calculate each contrast estimate and its SE from
   $\sqrt{MS_W\sum_j c_j^2/n_j}$. Explain the squared weights
   and the use of the full three-group error estimate.
3. Explain why the equal-average contrast cannot replace both
   separate feedback-versus-Rereading comparisons when both are required.
4. For a weighting illustration only, suppose Answer and Explanatory
   means stay at 8 and 10 but their counts are 10 and 30. The
   Rereading mean stays at 6. Compare the equally weighted
   method-average difference with the student-weighted difference.
   State the question answered by each. These illustrative counts
   do not change the chapter's actual balanced dataset.

<!-- prompt-id: CH04-E04 -->
### Exercise 4: Interpret comparison families (Integrative) {.unnumbered #ex-ch04-04}

The declared simultaneous family contains Answer minus Rereading
and Explanatory minus Rereading in Massed practice. Their estimates
are 2 and 4, each with $SE=\sqrt{5/3}$ and nine error df. Use
the Bonferroni critical value 2.685010847 under the independent
common-variance Normal model.

1. Divide a .05 familywise error allowance between the two
   intervals and their tails. Explain the resulting quantile .9875.
2. Calculate the two simultaneous intervals and interpret them
   together. Explain why an interval containing zero does not
   establish equivalence.
3. Tukey's all-pairs intervals are $[-1.6045,5.6045]$ for Answer
   minus Rereading, $[0.3955,7.6045]$ for Explanatory minus
   Rereading, and $[-1.6045,5.6045]$ for Explanatory minus Answer.
   Identify its family and explain why its limits differ from those
   in part 2. Give a reason for choosing a family before inspecting
   which intervals exclude zero.
4. Both reference comparisons use the same Rereading mean. Explain
   why this creates dependence and why it does not invalidate the
   Bonferroni bound.
5. Distinguish an omnibus rejection, a prespecified contrast and a
   contrast selected after inspecting results. State what should
   accompany uncertainty reported for the selected contrast.

<!-- prompt-id: CH04-E05 -->
### Exercise 5: Read the factorial pattern (Foundational) {.unnumbered #ex-ch04-05}

Use the complete invented study below. There are four different
students per cell, individually allocated to the six combinations.

| Feedback | Massed scores | Spaced scores |
|:--|:--|:--|
| Rereading | 4, 5, 7, 8 | 4, 5, 7, 8 |
| Answer feedback | 6, 7, 9, 10 | 10, 11, 13, 14 |
| Explanatory feedback | 8, 9, 11, 12 | 16, 17, 19, 20 |

1. Identify the outcome, the two factors and their levels, the
   assignment unit, the number of cells and the meaning of balance.
2. Calculate the Answer-Spaced cell mean, the Answer marginal mean,
   both practice marginal means and the grand mean. Explain why
   equal-cell and student-weighted averages coincide here.
3. Calculate the additive prediction for Explanatory-Spaced from
   the grand mean and the two marginal departures. Compare it
   with the observed cell mean and identify the interaction departure.
4. Calculate Spaced minus Massed for each feedback condition.
   Then calculate Answer's schedule difference minus Rereading's.
   Explain how these values appear in the cell-mean plot.
5. Explain why the four-cell difference in part 4 is one interaction
   contrast, while the full three-by-two interaction has two df.
   Explain why the connected lines do not represent paired students.

<!-- prompt-id: CH04-E06 -->
### Exercise 6: Distinguish joint and selected results (Integrative) {.unnumbered #ex-ch04-06}

For the full feedback-by-practice model, the feedback, practice,
interaction and error sums of squares are 256, 96, 64 and 60.
Their df are 2, 1, 2 and 18. Assume independent common-variance
Normal errors. Use $t_{.975,18}=2.100922040$ and the Scheffé
multiplier $\sqrt{2F_{.95;2,18}}=2.666292237$.

1. Reconstruct the full ANOVA table, including total SS and df,
   all mean squares and the three F ratios. The interaction
   p-value is .0015. State and interpret its joint null.
2. The schedule marginal means are 8 and 12, but the conditional
   schedule differences are 0, 4 and 8. Explain what the practice
   main effect averages and why it does not describe every condition.
3. Form the pointwise interval for Answer's 4-point schedule
   difference using $SE=\sqrt{5/3}$. State the family limitation
   of this single-comparison interval.
4. After inspecting the plot, a reader selects Answer's schedule
   difference minus Rereading's, estimated as 4. Calculate its
   SE from four independent cell means of size 4 and $MS_E=10/3$.
   Form its Scheffé interval and identify the protected family.
5. Explain why the interval in part 4 can contain zero when the
   omnibus interaction test is significant, and why that
   interaction-only family does not also protect part 3's simple effect.

<!-- prompt-id: CH04-E07 -->
### Exercise 7: Repair a claim about teaching (Integrative) {.unnumbered #ex-ch04-07}

Repair this claim about the invented study:

> Spaced practice improves every student's learning by four points,
> whatever feedback is provided. The significant interaction proves
> that Answer feedback benefits more from spacing than Rereading,
> and the results apply to all students.

All 24 quiz scores are observed. The stipulated design allocates
four different students to each cell. The equal-weight practice
main effect is 4.0000 points; conditional schedule differences
are 0.0000, 4.0000 and 8.0000. The interaction test gives
$F(2,18)=9.6000$, $p=.0015$. The selected
Answer-minus-Rereading difference of schedule differences is
4.0000, with a 95% interaction-family Scheffé interval
$[-0.8680,8.8680]$.

1. Identify at least four words or claims that exceed the
   supplied statistical and design information.
2. Write a bounded replacement retaining the conditional pattern,
   the joint interaction result, the selected estimate and interval,
   the 0--20 quiz scale, and the invented nature of the example.
3. Explain what random allocation could support in a real study
   and what it would not establish about population sampling or
   individual treatment effects.
4. Explain how class-level assignment, repeated observations or
   missing outcomes would affect the information needed for the
   analysis. Do not invent losses in the present dataset. State
   why four observations per cell limit diagnostic reassurance.

<!-- prompt-id: CH04-E08 -->
### Exercise 8: Match the analysis to the design (Integrative) {.unnumbered #ex-ch04-08}

The Massed Rereading scores are 4, 5, 7, 8 and the Massed Answer
scores are 6, 7, 9, 10. They belong to eight different students.
The pooled within-group variance is $10/3$. For these matched
classical calculations, the two-sided p-value is .1723.

1. Calculate Rereading minus Answer, its SE, the pooled t statistic
   and the corresponding one-way F statistic. State both tests'
   df and explain why their p-values agree although F loses direction.
2. Suppose instead that two measurements came from each of four
   students. Explain why a paired analysis would use within-student
   differences and why the independent-group calculation in part 1
   would no longer be the appropriate starting point. Name the
   dependence information needed to calculate its SE.
3. Explain how assigning one class to each condition within each
   school creates a blocked comparison. Identify the comparison
   units in Chapter 3's paired-school illustration and distinguish
   an allocation block from a sampling stratum.
4. A researcher proposes measuring a fractions pretest before
   assigning feedback. Explain what baseline adjustment adds,
   why an adjusted comparison needs a specified common value or
   averaging distribution, and why adjustment alone does not
   establish a causal interpretation.

## Appendix A: Reproducing the ANOVA Arithmetic {.unnumbered #app-ch04-calculations}

The calculations here use the same scores as the main text. They
show why the decompositions work and how the two-group identity
follows. The numerical identities describe observed data; the model
conditions are added when a reference distribution is used.

### A.1 Reconstruct the one-way decomposition {.unnumbered #app-ch04-small-data}

For the Massed slice, the grand mean is 8. The table separates
each group's contribution to the total.

\Needspace{16\baselineskip}

| Feedback | Mean | $n_j(\bar y_j-8)^2$ | Within-group SS | Total deviations from 8 |
|:--|--:|--:|--:|--:|
| Rereading | 6.0000 | 16.0000 | 10.0000 | 26.0000 |
| Answer feedback | 8.0000 | 0.0000 | 10.0000 | 10.0000 |
| Explanatory feedback | 10.0000 | 16.0000 | 10.0000 | 26.0000 |
| Total | | 32.0000 | 30.0000 | 62.0000 |

: Contributions to the one-way sums of squares. {#tbl-ch04-ss-contributions tbl-colwidths="[29,15,20,18,18]"}

For Rereading, the within-group calculation is

$$
(4-6)^2+(5-6)^2+(7-6)^2+(8-6)^2
=4+1+1+4=10.
$$

Its total squared deviation from the grand mean is

$$
(4-8)^2+(5-8)^2+(7-8)^2+(8-8)^2
=16+9+1+0=26.
$$

The between-group contribution is $4(6-8)^2=16$, and
$26=16+10$. The same calculation for Explanatory feedback gives
26 because its group mean is equally far above 8. The Answer
mean equals the grand mean, so its between-group contribution is
zero although its individual scores still vary.

The general proof begins by squaring the deviation split:

$$
\begin{aligned}
(y_{ij}-\bar y)^2
={}&(\bar y_j-\bar y)^2+(y_{ij}-\bar y_j)^2\\
&+2(\bar y_j-\bar y)(y_{ij}-\bar y_j).
\end{aligned}
$$

Within group $j$, its mean departure is constant. Summing the
cross-products therefore gives

$$
2(\bar y_j-\bar y)\sum_i(y_{ij}-\bar y_j)=0.
$$

The last sum is zero by the definition of a group mean. Summing
the remaining terms over groups establishes $SS_T=SS_B+SS_W$.
For the individual score 4, however, the squared total deviation
is 16 while the two squared components sum to $4+4=8$; its
cross-product contributes another 8. The cancellation is a
property of the full sum, not of each record separately.

Dividing by $N-1$ yields the total sample variance
$62/11=5.6364$. By contrast, the common within-group variance
estimate is $30/9=3.3333$. These denominators and targets differ:
the first describes variation around one grand mean, the second
estimates error variation around the three fitted group means.

### A.2 Reconstruct the balanced two-way table {.unnumbered #app-ch04-two-way-arithmetic}

The six cell means are $(6,6)$, $(8,12)$ and $(10,18)$ in feedback
rows and practice columns. For each observed score,

$$
\begin{aligned}
y_{ijk}-\bar y={}&
(\bar y_{j\cdot}-\bar y)+(\bar y_{\cdot k}-\bar y)\\
&+(\bar y_{jk}-\bar y_{j\cdot}-\bar y_{\cdot k}+\bar y)\\
&+(y_{ijk}-\bar y_{jk}).
\end{aligned}
$$

The four parts are the feedback departure, practice departure,
interaction departure and individual residual. Consider the
Explanatory-Spaced student scoring 16. The grand mean is 10,
the two marginal means are 14 and 12, and the cell mean is 18:

$$
16-10=(14-10)+(12-10)+(18-14-12+10)+(16-18)
=4+2+2-2.
$$

The observed total departure is 6. Balance and the zero-sum
constraints make the cross-products cancel when the full set
of squared decompositions is summed. This yields the four
components in the ANOVA table.

\Needspace{16\baselineskip}

| Feedback | Additive Massed | Additive Spaced | Interaction departures | Observed means |
|:--|--:|--:|:--|:--|
| Rereading | 4 | 8 | $2,-2$ | $6,6$ |
| Answer feedback | 8 | 12 | $0,0$ | $8,12$ |
| Explanatory feedback | 12 | 16 | $-2,2$ | $10,18$ |

: Add the interaction departures to reconstruct every cell mean. {#tbl-ch04-cell-reconstruction tbl-colwidths="[29,17,17,20,17]"}

The feedback term uses two free marginal departures, the practice
term one, and the interaction two. The full model therefore fits
one grand mean plus $2+1+2=5$ effects, equivalent to six separate
cell means. Within each of six cells, the four residuals have
one sum constraint, leaving $6(4-1)=18$ error df.

The total can also be checked directly by centering every score
at 10. The six cells contribute $74,74,26,26,10,266$ in table
order, which sum to 476. Equivalently, their mean departures
from 10 are $-4,-4,-2,2,0,8$, so

$$
SS_T=4(16+16+4+4+0+64)+60=416+60=476.
$$

The between-cell quantity 416 divides into feedback 256,
practice 96 and interaction 64. This gives the independent
checks $256+96+64+60=476$ and $2+1+2+18=23$.

### A.3 Derive the matched pooled-t identity {.unnumbered #app-ch04-pooled-identity}

For two independent groups, let
$\widehat{\Delta\mu}=\bar y_1-\bar y_2$ and $N=n_1+n_2$. Their grand mean is

$$
\bar y=\frac{n_1\bar y_1+n_2\bar y_2}{N}.
$$

Subtracting it from each group mean gives

$$
\bar y_1-\bar y=\frac{n_2}{N}\widehat{\Delta\mu},
\qquad
\bar y_2-\bar y=-\frac{n_1}{N}\widehat{\Delta\mu}.
$$

Consequently,

$$
\begin{aligned}
SS_B
&=n_1\left(\frac{n_2}{N}\widehat{\Delta\mu}\right)^2
  +n_2\left(-\frac{n_1}{N}\widehat{\Delta\mu}\right)^2\\
&=\frac{n_1n_2}{N}\widehat{\Delta\mu}^2.
\end{aligned}
$$

There is one numerator df, so $MS_B=SS_B$. Define the pooled
variance $s_p^2=SS_W/(N-2)$. Then

$$
F=\frac{SS_B}{s_p^2}
=\frac{\widehat{\Delta\mu}^2}{s_p^2(1/n_1+1/n_2)}=t^2.
$$

For the Rereading/Answer subset, $n_1=n_2=4$, $\widehat{\Delta\mu}=-2$,
$SS_B=8$, $SS_W=20$ and $s_p^2=20/6=10/3$. Thus
$F=8/(10/3)=2.4000$, equal to the square of
$t=-2/\sqrt{5/3}=-1.5492$. Under the matched Normal
common-variance model, the upper-tail $F(1,6)$ probability
equals the two-sided $t_6$ probability, .1723.

This algebra does not supply a paired-data SE: paired inference
works on within-pair differences and their variation. A Welch
test also has its own variance estimate and reference df. The
identity requires matching the data, pooled error estimate,
degrees of freedom and two-sided alternative.

## Appendix B: Families, Weighting and Magnitude {.unnumbered #app-ch04-comparisons}

### B.1 What each interval family covers {.unnumbered #app-ch04-families}

A coverage statement needs both a target and a family. The
following procedures used in the chapter make different promises.

\Needspace{16\baselineskip}

| Procedure | Family specified here | Interpretation |
|:--|:--|:--|
| Pointwise t interval | One comparison specified without selection over relevant results | Covers that target at the stated rate under the model |
| Bonferroni | The two declared Rereading-referenced comparisons | Joint coverage of both targets is at least .95 under the model |
| Tukey | All three pairwise feedback differences | Simultaneous coverage for the all-pairs family |
| Scheffé | All contrasts in the two-dimensional interaction space | Simultaneous coverage across this entire space permits a selected interaction direction |

: The family is part of the inferential statement. {#tbl-ch04-family-reference tbl-colwidths="[20,39,41]"}

For the selected difference of differences, retain the same
estimate 4.0000, $SE=\sqrt{10/3}$ and 18 error df throughout.
Only the calibration changes:

$$
\begin{aligned}
\text{Pointwise: }&4\pm2.100922040\sqrt{10/3}
  \approx[0.1643,7.8357],\\
\text{Interaction family: }&4\pm2.666292237\sqrt{10/3}
  \approx[-0.8680,8.8680].
\end{aligned}
$$

The pointwise interval excludes zero and the simultaneous interval
does not. Selecting the narrower one because it produces a
preferred conclusion would discard the selection protection. The
width difference reflects the broader coverage obligation, not a
change in the observed contrast.

The interaction-only family has two dimensions here. It is narrower
than all possible contrasts among six arbitrary cell means, which
would have five dimensions. An interaction contrast cancels both
main-effect components through its row and column weight sums.
The difference of differences does this; a single schedule effect
within one feedback condition does not. This explains why one
multiplier cannot be applied indiscriminately to every comparison.

The two one-way reference estimates also illustrate dependence.
Writing them as $\bar Y_A-\bar Y_R$ and $\bar Y_E-\bar Y_R$,
their covariance under independent group means is
$\operatorname{Var}(\bar Y_R)=\sigma^2/4$. A low reference mean
increases both differences. Bonferroni remains valid because its
joint-coverage bound uses the sum of individual failure probabilities
without assuming independence.

### B.2 Unequal-cell weights and effect-size denominators {.unnumbered #app-ch04-weighting}

In the chapter's balanced data, equal-cell and student-weighted
marginal means coincide. For unequal cell counts, define them
separately:

$$
\bar y_{j\cdot}^{\text{students}}
=\frac{\sum_kn_{jk}\bar y_{jk}}{\sum_kn_{jk}},
\qquad
\bar y_{j\cdot}^{\text{equal cells}}
=\frac1K\sum_k\bar y_{jk}.
$$

The first describes the mixture of students observed in that
feedback group. The second gives each of the $K$ practice conditions
the same weight. Neither is automatically the right target for
every question.

For an illustration, retain Answer's cell means 8 and 12 but
imagine counts of 2 Massed and 6 Spaced students. The equally
weighted mean is $(8+12)/2=10.0000$; the student-weighted mean is
$(2\times8+6\times12)/8=11.0000$. These are hypothetical weights,
not changes to the actual four-per-cell dataset. A comparison of
feedback groups using different schedule mixtures can combine a
feedback comparison with a weighting difference. Adjusted means in
later chapters make their common profile or averaging distribution
explicit for this reason.

The main one-way effect-size summary was $\eta^2=SS_B/SS_T=.5161$.
A common reference correction, **omega squared**, subtracts an
estimated noise contribution from the between-group sum of squares:

$$
\omega^2=
\frac{SS_B-(J-1)MS_W}{SS_T+MS_W}
=\frac{32-2(10/3)}{62+10/3}=.3878.
$$

The subtraction recognizes that group means vary even under equal
model means. This correction is motivated by the common-variance
model; it is not the same descriptive fraction as eta squared,
and it does not establish causation. In some samples its numerator
can be negative, illustrating its role as an estimator rather
than a literal nonnegative share of observed variation. The raw
mean differences remain necessary for understanding the outcome
scale.

In factorial software output, **partial eta squared** often uses
a term-specific denominator:

$$
\eta^2_{\text{partial, term}}=
\frac{SS_{\text{term}}}{SS_{\text{term}}+SS_E},
\qquad
\eta^2_{\text{term}}=\frac{SS_{\text{term}}}{SS_T}.
$$

They agree in one-way ANOVA because the total contains only
that factor and error. In our factorial model, interaction has
$64/(64+60)=.5161$ partial eta squared but $64/476=.1345$
ordinary eta squared. The former excludes the other factor
components from its denominator. Always name the summary and
its denominator before interpreting a software effect-size column.

## References {.unnumbered}

::: {#refs}
:::
