---
title: "Data, Distributions, and Variation"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-cap: true
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  out.width = "100%"
)
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(dev = "cairo_pdf")
}

options(digits = 3, width = 72, scipen = 4)
library(ggplot2)
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      plot.title.position = "plot"
    )
)

## The curated UCI secondary-school record of Cortez and Silva (2008),
## reduced to the columns this chapter analyses and embedded so that the
## chapter renders without an external data file. Values are unchanged;
## pass reproduces the documented cut point of 10 on the 0-20 scale.
sp_absences <- c(
 6,4,10,2,4,10,0,6,0,0,0,4,2,2,0,4,6,4,16,4,0,0,2,0,2,14,2,4,4,16,0,0,0,
 0,0,0,2,7,2,8,25,8,2,0,14,8,12,4,2,2,2,2,6,0,6,8,0,4,2,2,6,6,4,2,0,2,4,
 4,2,12,0,0,2,2,54,6,8,0,2,12,2,4,10,4,2,6,4,4,12,18,0,4,4,0,6,2,2,2,6,
 0,14,0,4,26,0,10,8,2,6,4,6,0,6,10,8,2,2,0,20,6,2,6,2,18,0,0,0,2,0,8,0,
 0,12,16,0,0,0,0,0,0,0,8,2,2,0,0,0,2,0,0,0,6,8,0,0,2,8,6,2,4,0,6,0,2,0,
 16,4,0,0,0,0,2,0,0,4,4,2,4,10,4,10,2,0,56,14,12,2,0,6,4,10,0,12,8,0,0,
 4,8,24,0,2,6,4,18,6,28,5,10,6,6,10,13,0,15,12,2,22,13,3,4,2,0,2,0,0,16,
 10,2,14,10,14,4,14,2,18,10,4,20,2,0,14,2,0,0,0,6,4,16,8,0,0,6,4,0,0,2,
 6,12,8,0,21,2,1,4,0,13,2,8,10,0,15,4,2,2,2,6,75,22,15,8,30,19,1,4,4,2,
 5,6,6,9,11,0,12,6,8,4,0,10,0,5,14,0,0,0,20,8,0,38,0,18,0,20,3,22,14,40,
 0,9,0,2,23,12,3,1,0,3,3,8,7,4,2,7,0,0,0,16,12,0,7,4,4,0,11,0,4,7,9,0,0,
 10,8,2,7,4,4,0,4,2,4,0,0,2,0,0,0,4,0,0,0,10,4,3,8,14,0,2,4,4,0,17,4,5,
 2,0,14,2,7,0,0,0,11,3,3,0,5
)

sp_grade_period_2 <- c(
 6,5,8,14,10,15,12,5,18,15,8,12,14,10,16,14,14,10,5,10,14,15,15,13,9,9,
 12,16,11,12,11,16,16,10,14,7,16,16,12,13,10,12,18,8,10,8,12,19,15,7,13,
 13,11,10,13,9,15,15,10,16,11,8,10,9,10,15,13,7,9,16,15,10,6,12,12,9,11,
 11,8,5,12,10,6,15,10,9,7,14,10,6,7,17,6,10,13,10,15,9,14,9,7,17,13,6,
 18,11,8,18,13,15,19,10,13,19,9,15,13,14,7,13,15,14,13,11,7,13,10,8,4,
 18,0,0,13,11,0,0,0,0,12,16,9,9,11,14,0,11,7,11,6,9,5,13,10,0,11,8,12,8,
 15,12,6,9,0,10,8,11,10,15,7,14,5,15,11,7,11,9,13,5,8,10,8,13,17,9,13,
 12,12,15,7,9,12,8,8,9,14,15,15,9,18,9,16,10,9,6,10,9,7,12,9,7,8,12,13,
 7,10,15,6,6,7,10,6,5,16,13,13,8,15,11,8,10,13,11,9,13,7,9,13,12,11,7,
 12,11,0,12,0,18,12,8,5,15,8,10,9,9,12,9,12,11,14,9,18,8,12,9,10,17,9,
 10,9,0,9,14,11,14,10,12,9,9,8,11,8,9,12,9,9,10,18,12,14,13,11,15,12,18,
 13,12,9,8,13,15,10,11,12,17,14,12,18,9,12,10,9,12,11,10,13,11,8,10,11,
 11,13,9,11,14,15,12,15,10,9,14,8,14,0,8,9,15,13,8,15,10,12,10,15,8,10,
 13,15,10,15,13,7,13,7,8,11,9,13,12,10,16,13,12,11,15,11,10,13,6,10,12,
 7,12,11,5,18,8,14,9,15,10,14,6,11,5,5,9,5,5,9,5,9,16,8,12,9
)

sp_final_grade <- c(
 6,6,10,15,10,15,11,6,19,15,9,12,14,11,16,14,14,10,5,10,15,15,16,12,8,8,
 11,15,11,11,12,17,16,12,15,6,18,15,11,13,11,12,18,11,9,6,11,20,14,7,13,
 13,10,11,13,10,15,15,9,16,11,11,9,9,10,15,12,6,8,16,15,10,5,14,11,10,
 10,11,10,5,12,11,6,15,10,8,6,14,10,7,8,18,6,10,14,10,15,10,14,8,5,17,
 14,6,18,11,8,18,13,16,19,10,13,19,9,16,14,13,8,13,15,15,13,13,8,12,11,
 9,0,18,0,0,12,11,0,0,0,0,12,15,0,9,11,13,0,11,0,11,0,10,0,14,10,0,12,8,
 13,10,15,12,0,7,0,10,7,12,10,16,0,14,0,16,10,0,9,9,11,6,9,11,8,12,17,8,
 12,11,11,15,9,10,13,9,8,10,14,15,16,10,18,10,16,10,10,6,11,9,7,13,10,7,
 8,13,14,8,10,15,4,8,8,10,6,0,17,13,14,7,15,12,9,12,14,11,9,13,6,10,13,
 12,11,0,12,12,0,12,0,18,13,8,5,15,8,10,8,8,12,8,13,11,14,0,18,8,12,9,0,
 17,10,11,10,0,9,14,11,14,10,12,9,9,8,10,8,10,12,10,11,11,19,12,14,15,
 11,15,13,18,14,11,0,8,14,16,11,10,14,18,13,12,18,8,12,10,0,13,11,11,13,
 11,0,9,10,11,13,9,11,15,15,11,16,10,9,14,8,14,0,0,0,15,13,0,17,10,11,0,
 15,0,10,14,16,9,15,13,8,13,8,8,11,9,13,11,10,16,13,12,10,15,12,10,13,0,
 10,11,9,12,11,5,19,10,15,10,15,10,14,7,10,0,5,10,6,0,8,0,9,16,7,10,9
)

sp_higher_no <- c(
 79,128,149,151,154,161,164,165,166,167,240,248,251,253,270,277,312,353,
 362,372
)

sp_school_ms <- c(
 350,351,352,353,354,355,356,357,358,359,360,361,362,363,364,365,366,
 367,368,369,370,371,372,373,374,375,376,377,378,379,380,381,382,383,
 384,385,386,387,388,389,390,391,392,393,394,395
)

students <- data.frame(
  student_id = sprintf("U%03d", seq_along(sp_absences)),
  school = "GP",
  absences = as.integer(sp_absences),
  grade_period_2 = as.integer(sp_grade_period_2),
  final_grade = as.integer(sp_final_grade),
  higher_education = "yes",
  stringsAsFactors = FALSE
)
students$school[sp_school_ms] <- "MS"
students$higher_education[sp_higher_no] <- "no"
students$pass <- ifelse(students$final_grade >= 10, "Yes", "No")

grade_zero <- students$final_grade == 0
students_no_grade_zero <- students[!grade_zero, ]
stopifnot(
  sum(grade_zero) == 38L,
  all(students$absences[grade_zero] == 0),
  sum(students$grade_period_2[grade_zero] > 0) == 25L
)
grade_zero_sensitivity <- c(
  n = nrow(students_no_grade_zero),
  mean = mean(students_no_grade_zero$final_grade),
  median = median(students_no_grade_zero$final_grade),
  variance = var(students_no_grade_zero$final_grade),
  sd = sd(students_no_grade_zero$final_grade),
  IQR = IQR(students_no_grade_zero$final_grade),
  covariance = cov(
    students_no_grade_zero$absences,
    students_no_grade_zero$final_grade
  ),
  correlation = cor(
    students_no_grade_zero$absences,
    students_no_grade_zero$final_grade
  )
)

grade_summary <- data.frame(
  Summary = c("Mean", "Median", "Sample variance", "Sample standard deviation", "IQR"),
  Value = c(
    mean(students$final_grade),
    median(students$final_grade),
    var(students$final_grade),
    sd(students$final_grade),
    IQR(students$final_grade)
  ),
  Units = c("grade points", "grade points", "squared grade points",
            "grade points", "grade points")
)

same_mean <- data.frame(
  Pattern = factor(
    rep(c("Concentrated", "Spread", "Right-skewed"), each = 9),
    levels = c("Concentrated", "Spread", "Right-skewed")
  ),
  Value = c(
    8, 9, 9, 10, 10, 10, 11, 11, 12,
    2, 4, 6, 8, 10, 12, 14, 16, 18,
    5, 6, 7, 8, 9, 10, 11, 13, 21
  ),
  Index = rep(seq_len(9), 3)
)

set.seed(60202)
pattern_x <- seq(-2, 2, length.out = 80)
relationship_patterns <- rbind(
  data.frame(
    X = pattern_x,
    Y = 1.2 + 1.1 * pattern_x + rnorm(80, 0, .45),
    Pattern = "Positive linear"
  ),
  data.frame(
    X = pattern_x,
    Y = 1.2 - 1.1 * pattern_x + rnorm(80, 0, .45),
    Pattern = "Negative linear"
  ),
  data.frame(
    X = pattern_x,
    Y = pattern_x^2 + rnorm(80, 0, .25),
    Pattern = "Curvilinear"
  ),
  data.frame(
    X = runif(80, -2, 2),
    Y = rnorm(80, 1.2, 1.1),
    Pattern = "No clear pattern"
  )
)
relationship_patterns$Pattern <- factor(
  relationship_patterns$Pattern,
  levels = c("Positive linear", "Negative linear", "Curvilinear", "No clear pattern")
)

normal_grid <- data.frame(Z = seq(-3.4, 3.4, length.out = 500))
normal_grid$Density <- dnorm(normal_grid$Z)

set.seed(60203)
repetitions <- 5000L
sample_sizes <- c(5L, 30L, 100L)
sampling_means <- do.call(rbind, lapply(sample_sizes, function(n) {
  data.frame(
    Sample_size = factor(n, levels = sample_sizes),
    Sample_mean = replicate(repetitions, mean(rexp(n, rate = 1 / 10)))
  )
}))
sampling_split <- split(sampling_means$Sample_mean, sampling_means$Sample_size)
sampling_summary <- data.frame(
  n = sample_sizes,
  Mean_of_means = vapply(sampling_split, mean, numeric(1)),
  Empirical_SE = vapply(sampling_split, sd, numeric(1)),
  Model_SE = 10 / sqrt(sample_sizes),
  row.names = NULL
)

group_times <- data.frame(
  Group = factor(c("A", "A", "B", "B", "B", "B")),
  Minutes = c(1, 3, 3, 3, 5, 9)
)
intention_counts <- table(
  factor(students$higher_education, levels = c("no", "yes")),
  factor(students$pass, levels = c("Yes", "No"))
)
stopifnot(
  identical(as.integer(intention_counts), c(7L, 258L, 13L, 117L)),
  abs(cor(students$absences, students$final_grade) - 0.03424731615) < 1e-8,
  abs(mean(group_times$Minutes) - 4) < 1e-12
)
```

{{< pagebreak >}}

::: {.callout-note title="Chapter overview"}

Chapter 1 explained what study observations represent and how they
arise. This chapter develops their descriptive use: how to read the
recorded data and turn measurements into useful displays and summaries.
Its small invented examples make the arithmetic visible before we
return to a larger school dataset. A final simulation extends the same
ideas of center, spread, and shape to summaries from repeated samples,
preparing us to understand statistical uncertainty.

By the end of the chapter you will be able to read a dataset as cases,
variables, and values, recognize long and wide layouts, and identify
the denominator and sample-selection procedure; see and summarize one
distribution by shape, center, and spread; compare groups while keeping
their variation visible; connect variable forms to probability models,
calculate transformations and Normal areas, and describe a relationship
with a graph, covariance, correlation and linear adjustment; interpret
each description within its study setting;
explain what a sampling distribution and standard error show about
repeated samples; and distinguish one sample's sampling error from
that distribution and its standard error.
:::

## Understanding the Dataset

Chapter 1 connected study units and measurement procedures to the
observations available for analysis. Here we examine the resulting
records: how observations are arranged, what scales and codes mean,
and which cases and denominator support each summary. These details
turn the study's account of its evidence into data we can describe.

Schools often want to know whether students with more absences tend to
receive lower grades. Before choosing a statistical method, it helps to
see what information is actually available.

Our main example uses the mathematics dataset from Cortez and Silva's
student-performance study, based on school reports and questionnaires
from two Portuguese secondary schools [@cortez2008using; @cortez2008data].
It contains data for **395 students**, including their absence counts and
final mathematics grades. The repository also provides a separate
Portuguese-language dataset; the calculations here use the mathematics
data.

We will ask: **Among these 395 students, how do recorded absences and final
mathematics grades vary together?** To answer, we first describe grades
on their own, then compare groups and examine relationships. We put that
description back into its school setting before asking a related question:
how could a summary change if we observed a different sample?

### Cases, variables, values, and data layout

A **dataset** brings together information about people, objects, or
measurement occasions. A **case** is the entity or occasion being
described. In this dataset each case is one student. A **variable** is a
feature that can differ across students, such as absences or final grade;
a **value** is the measurement or category for a particular student.

@tbl-student-preview shows six selected students. The student IDs are
labels added for this tutorial. School codes GP and MS identify the two
schools. Reading across U001's row gives six absences and a final grade
of six: the same numeral for two different measurements, a count and a
score. Reading down the Absences column compares that count across
students. The identifier keeps the two measurements linked to the
student they describe.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

```{r}
#| label: tbl-student-preview
#| tbl-cap: "Selected students from the mathematics dataset"
#| tbl-colwidths: [23, 19, 25, 33]
preview <- students[c(1, 2, 3, 350, 351, 352),
                    c("student_id", "school", "absences", "final_grade")]
names(preview) <- c("Student ID", "School", "Absences", "Final grade")
knitr::kable(preview, booktabs = TRUE, row.names = FALSE,
             align = c("l", "l", "r", "r"))
```

The two numerical columns can also be written as a **data matrix**,
with one row for each displayed student and one column for each variable.
Let $x_{ij}$ denote the entry in row $i$ and column $j$: here $j=1$
means absences and $j=2$ final grade. The first index numbers these six
displayed rows, rather than their positions in the original dataset.
Thus $x_{31}=10$ absences and $x_{42}=13$ grade points.

$$
\mathbf X=(x_{ij})=
\begin{pmatrix}6&6\\4&6\\10&10\\10&13\\8&8\\2&13\end{pmatrix},
\qquad i=1,\ldots,6,\quad j=1,2.
$$

Each column can be summarized separately. The symbol $\sum$ means to
add the indicated entries. For example, the sum for the first column is
$\sum_{i=1}^6 x_{i1}=6+4+10+10+8+2=40$, so its mean is
$40/6=6.6667$ absences. The grade column has mean $56/6=9.3333$
grade points. These summarize the six displayed cases; the later
summaries use all 395 students. In general, for $n$ rows,

$$
\bar x_j=\frac1n\sum_{i=1}^n x_{ij},\qquad
s_j^2=\frac1{n-1}\sum_{i=1}^n(x_{ij}-\bar x_j)^2.
$$

The first formula gives a column's mean. The second gives its sample
variance, developed through a small calculation in Section 2.2. This
notation organizes numerical entries; identifiers and school labels
still have their own meanings and are not quantities to average.
For two named variables we can write $x_i=x_{i1}$ and $y_i=x_{i2}$.
Subscripts have locally stated meanings: an index for a column here
can have a different declared use in a later grouped-data calculation.

Chapter 1 distinguished the units a study observes from the occasions
on which they are measured. We can recognize that connection in a
dataset's arrangement. Suppose a separate, invented study records a
score for each of 120 students at the beginning and end of a term,
with both scores observed for every student. There are 120 students
and 240 measurement occasions. In a **long layout**, each row describes
one student on one occasion; an occasion column identifies when the
score was recorded. In a **wide layout**, each row describes one
student, and separate columns hold the beginning and end scores.
@tbl-ch02-layout-long and @tbl-ch02-layout-wide show the same six
invented scores for three of the students.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

| Student ID | Occasion | Score |
|:--|:--|--:|
| S001 | Beginning | 52 |
| S001 | End | 58 |
| S002 | Beginning | 61 |
| S002 | End | 64 |
| S003 | Beginning | 47 |
| S003 | End | 53 |

: Invented repeated scores in long form {#tbl-ch02-layout-long tbl-colwidths="[30,40,30]"}

```{=latex}
\Needspace{10\baselineskip}
\phantomsection
```

| Student ID | Beginning score | End score |
|:--|--:|--:|
| S001 | 52 | 58 |
| S002 | 61 | 64 |
| S003 | 47 | 53 |

: The same invented scores in wide form {#tbl-ch02-layout-wide tbl-colwidths="[30,35,35]"}

For all 120 students, the long table contains 240 rows and the wide
table contains 120 rows. Both contain 240 scores from the same 120
people. The identifier links each student's two scores, whether they
occupy two rows or two columns. Reshaping changes the arrangement while
preserving the observations and their connections. The 240 scores
describe repeated observations of 120 students. Treating every score
as independent would ignore the pairing used to describe change. In
longitudinal or panel data, these identifiers let us follow the same people. Repeated
cross-sectional data instead describe different samples on successive
occasions, so they cannot by themselves show each person's change.
Group identifiers, such as school, can accompany either layout. In our
main dataset, each student's absence count and final grade contribute
one pair to the relationship we examine.

### Variable forms and roles

Numerical, or **quantitative**, variables describe amounts. Counts such
as absences take discrete values: a student can have two or three counted
absences. An absence count also depends on the number of scheduled
sessions in its measurement period. Elapsed time can vary continuously,
even if a device rounds it to whole seconds. Equal differences in
minutes represent equal amounts of elapsed time: from 20 to 25 minutes
and from 40 to 45 minutes both add five minutes. Final grades here are
discrete numerical scores on a 0--20 scale, recorded in whole points.
A one-point difference is a difference on that grading scale; its
meaning for learning also depends on what the assessment measures.

**Categorical** variables place cases in groups. A nominal category,
such as school, names a group without ranking it. An ordinal category
also has an order, as with low, moderate, and high satisfaction.
Numerical codes can represent these categories, but an average of school
codes would have no useful meaning. For ordered categories, the spacing
between codes need not represent equal differences in the underlying
experience.

Variable form helps us choose a description: counts and proportions
show how cases fall into categories, while a graph or numerical
summary can show the distribution of amounts. Roles come from the question.
For the absence--grade question, absences is the **predictor** and final
grade is the **outcome**. These are the explanatory and response roles
developed in Chapter 1, Section 2.1. The roles organize the comparison; the study
design determines whether a causal interpretation is justified. School
could instead define groups for a comparison of grade distributions.

A **codebook** explains the variables, their units, coding rules, and
measurement periods. Here it identifies the final grade as G3,
recorded from 0 to 20, and the
second-period grade as G2. We use readable variable names in the
calculations while preserving the source values. A binary pass indicator
will later distinguish grades of 10 or above from grades below 10.

### Coding, missing values, and denominators

A codebook distinguishes a recorded value from its meaning. Zero can
be an observed count; a missing value means the intended measurement
is unavailable. The UCI documentation reports no missing values, but
the joint pattern of zero grades and zero absences needs contextual
checking, as Section 3.4 explains.

The **analysis sample** is the set of cases contributing to a particular
calculation. Its denominator can change with the measurements that
calculation needs. In a separate invented clinic dataset of 1,240
visits, 36 lack waiting times, 81 lack return statuses, and 10 lack both.
The available counts are

$$
n_{\text{wait}}=1240-36=1204,\qquad
n_{\text{return}}=1240-81=1159,
$$

$$
n_{\text{both}}=1240-(36+81-10)=1133.
$$

Adding 36 and 81 counts the 10 visits missing both measurements twice;
subtracting that overlap once counts each incomplete visit once.
These denominators describe visits, not necessarily distinct patients.
Likewise, 20% of sessions and 20% of students answer different questions.
Name the contributing cases and measurement period with each summary.

### The students observed and the population of interest

Chapter 1 distinguished whom a study's question concerns from the
units that supply its observations. Consider an invented school system
that wants to describe all 20,000 students enrolled on a particular
date and selects 100 of them. The 20,000 students form the **population**,
the collection about which the system wants to learn; the 100 selected
students form its **sample**. A sample may come from a probability
selection process, volunteers, or an available source. How the students
enter the sample matters for extending a description to the population.

The next subsection compares selection procedures in this invented
system, and Section 4 returns to it to examine repeated samples. Our
395-student dataset comes instead from two schools. Its source supports
a description of these students, while extending the description to all
secondary students would need an additional basis.

The distinction is practical. The mean calculated from all 395 supplied
grades describes that collection directly. A statement about students
elsewhere asks more of the study. For now, we will develop the
description first and revisit its scope at the end.

The Student Performance resource is distributed under
[Creative Commons Attribution 4.0](https://creativecommons.org/licenses/by/4.0/)
[@cortez2008data]. This tutorial selects and renames variables, adds
identifiers and a pass indicator, and retains the original selected
values.

### How samples are selected

Chapter 1, Section 3.2 explained how study entry differs from the
allocation of conditions among participants. We now make the selection
procedure visible in the invented school system. This example lets us
compare ways of selecting students; the source of the separate
Portuguese dataset does not establish a probability sample of all
secondary students.

The school system's population contains 20,000 eligible students. A
**sampling frame** is the list or other operational record used to
identify units for selection: here it could be an up-to-date student
register. The **sampling design**, also called the **sampling scheme**,
specifies how selection proceeds. It can include several stages and
use information about the students or their schools.

Chapter 1 introduced **random sampling**, also called **probability
sampling**. Under such a design, every eligible unit in the population
covered by the frame has a known, nonzero chance of inclusion. That
unit's chance of entering the sample is its **inclusion probability**;
the probabilities need not be equal. Coverage matters: if eligible
students are absent from the register, random selection from that
register cannot select them.

Designs can use groups in different ways. **Strata** partition the
population into defined groups, such as school types, with a sample
selected within every stratum. **Clusters** are groups that can be
selected as units, such as schools whose students will be observed.
@tbl-ch02-sampling-schemes compares five probability designs
[@statcan2021probability]; its illustrations use our invented school
system.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

| Scheme | Selection rule | Invented school-system illustration |
|:--|:--|:--|
| Simple random sampling | Every possible sample of the specified size has the same probability. | Select 100 distinct students from the full register, giving every possible set of 100 the same chance. |
| Systematic sampling | Choose a random start, then every $k$th unit. | Select a start at random from 1 to 200, then every 200th student on the 20,000-student list. |
| Stratified random sampling | Select a probability sample within every defined stratum. | Divide the register by school type, then randomly select students within each type. |
| Cluster sampling | Select groups through a probability mechanism; in the simplest version, include all eligible units in them. | Randomly select schools from the school list and include all their eligible students. |
| Multistage sampling | Use probability selection at each stage. | Randomly select schools, then randomly select students within each selected school. |

: Probability sampling schemes in one invented school system {#tbl-ch02-sampling-schemes tbl-colwidths="[22,37,41]"}

The distinction between simple random and systematic sampling concerns
the possible samples. In the systematic design, a random start from 1
to 200 followed by every 200th student gives each student an inclusion
probability of $1/200$. There are only 200 possible samples, one for each
starting position. Simple random sampling permits every possible set
of 100 students. Equal individual chances alone therefore do not define
simple random sampling. The ordering of a systematic-sampling list
also matters: a repeating pattern that matches the interval can make
the selected students unusually alike.

Stratification samples within every stratum, while cluster sampling
selects some clusters. A design can combine these features, such as
stratifying schools before selecting schools and then their students.

**Convenience sampling** might use students available in a nearby
classroom. Their inclusion probabilities are not supplied by a
probability design [@statcan2021nonprobability]. Similarly, an open
invitation describes how volunteers enter a dataset, rather than a
probability selection mechanism. After random selection, some selected
students may decline to participate. Frame coverage, selection and
subsequent nonresponse are distinct parts of the account.

Groups also enter three different parts of a study. Clustered data
describe relationships among observations, such as students sharing a
school. Cluster sampling selects groups into the sample. Cluster
random assignment allocates participating groups to conditions. The
same study may use more than one of these features, so identifying
grouped observations alone does not establish how groups were selected
or assigned.

Selection and dependence matter when a sample supports a population
summary. If students have unequal inclusion probabilities, the
calculation can require weights to account for how they were selected.
If observations are related, their contribution to precision differs
from that of independent observations. Chapter 3 develops uncertainty
for paired and clustered observations and for finite-population
sampling without replacement. Section 4.1 below repeats the simple
random sampling procedure from this overview. Its later waiting-time
illustration uses a separate independent-observation model; an SE from
that model does not apply unchanged to every sampling scheme.

## Describing Distributions and Groups

A **distribution** describes how a variable's values are arranged across cases.
When it describes observed data, we call it an **empirical distribution**.
We begin with the distribution of final mathematics grades. A graph shows
where students' values cluster and how far they spread; the numerical
summaries then give concise descriptions of those features.

### See shape before reducing it


A display is useful when its features answer the descriptive question.
@tbl-ch02-graph-choices gives a small guide. A **histogram** counts
quantitative values in adjacent intervals; changing the intervals can
change the visible shape. A **boxplot** marks the median inside a box
spanning the middle half of the values. In the convention used here,
whiskers reach the most extreme observations within 1.5 IQRs of the
box ends, and more distant observations appear separately. These are
potential outliers to investigate, not automatic errors. The plotting
quartiles use R's convention, which can differ from a hand-calculation
convention in a small sample.

| Data and question | Useful display | What to read |
|:--|:--|:--|
| Category frequencies | Bar chart; pie chart for a simple part-to-whole display | Counts or proportions with their denominator; bars ease comparisons |
| Quantitative values | Dot plot or histogram | Individual values or counts in intervals; inspect gaps, tails and bin choices |
| Center and spread across groups | Side-by-side boxplots with points where feasible | Medians, IQRs, overlap and potential outliers |
| Smoothed shape | Density curve | Area across a range; height is not the probability of one exact value |
| Two quantitative variables | Scatterplot | Paired positions; developed in Section 3.2 |

: Graphs answer different descriptive questions. {#tbl-ch02-graph-choices tbl-colwidths="[27,31,42]"}

@fig-ch02-graph-gallery shows how those choices work with invented
learning records. Bars ease comparisons between category counts;
a pie chart emphasizes shares of a whole. The four displays of
completion times use the same sixteen values. A **density curve**
smooths their shape, with the degree of smoothing controlled by a
bandwidth. Its height is not the probability of one exact time.
Here the smoother is shown only over feasible positive times; it is
an exploratory summary rather than a fitted probability law. Individual
points keep the observations visible in a small sample.

```{r}
#| label: fig-ch02-graph-gallery
#| fig-cap: "Several views of invented learning records: category proportions and the distribution of completion times."
#| fig-alt: "Bar and pie panels show counts four, seven and five for three learning formats. A histogram, density curve and boxplot show the same sixteen completion times, with a long right tail. The final panel shows individual times."
#| fig-height: 5.5
#| fig-pos: "H"

times <- c(2, 3, 3, 4, 4, 5, 5, 5, 6, 6, 7, 8, 9, 11, 14, 20)
formats <- c(Reading = 4, Practice = 7, Discussion = 5)
op <- par(mfrow = c(2, 3), mar = c(4, 4, 2, 1),
          mgp = c(2.2, .7, 0))
barplot(formats, names.arg = c("Read", "Practice", "Discuss"),
        cex.names = .9, ylab = "Number", main = "Categories: bars")
pie(formats, main = "Categories: shares")
hist(times, breaks = seq(0, 24, 4), xlab = "Minutes",
     main = "Quantitative: histogram")
plot(density(times, from = 0), xlab = "Minutes", main = "Density curve")
boxplot(times, ylab = "Minutes", main = "Boxplot")
stripchart(times, method = "stack", xlab = "Minutes",
           main = "Individual observations", pch = 19)
par(op)
```

@fig-grade-distribution displays all 395 final grades. The graph makes
several features visible: the score is bounded between 0 and 20, centered near
11, discrete rather than continuous, and has a visible group at zero.

```{r}
#| label: fig-grade-distribution
#| fig-cap: "Distribution of final mathematics grades for the 395 students. Dashed and dotted lines mark the mean and median, respectively."
#| fig-alt: "A histogram of grades from zero to twenty is centered near eleven, bounded at both ends, and shows a visible group at zero."
#| fig-height: 3
#| fig-pos: "H"

ggplot(students, aes(x = final_grade)) +
  geom_histogram(
    binwidth = 1,
    boundary = -.5,
    fill = "#B9D5E5",
    color = "white"
  ) +
  geom_vline(xintercept = mean(students$final_grade),
             color = "#9B3D45", linetype = 2, linewidth = .7) +
  geom_vline(xintercept = median(students$final_grade),
             color = "#315B45", linetype = 3, linewidth = .7) +
  scale_x_continuous(breaks = seq(0, 20, 2)) +
  labs(
    x = "Final mathematics grade (0--20)",
    y = "Number of students"
  )
```

Describe a distribution with four kinds of language:

- **Shape:** symmetry or skewness, clusters, gaps, and discreteness.
- **Center:** where a typical or balancing value lies.
- **Spread:** how widely observations vary.
- **Striking patterns and potential outliers:** values, clusters or gaps
  needing contextual checking.

An unusual value is not automatically a mistake. Its recorded meaning and
provenance should be investigated before it is changed or removed.

### Five summaries with distinct jobs

We use five summaries to describe center and spread: mean, median, variance,
standard deviation, and interquartile range. A small example shows what each
one contributes before we write the general formulas.

Consider four invented completion times, already ordered: **1, 3, 3, and 5
minutes**. Their sum is $1+3+3+5=12$. The **mean** shares that total equally
across the four observations, giving $\bar x=12/4=3.0000$ minutes. The **median**
is the middle of the ordered values: here the two middle observations are both
3, so their average is also 3.0000 minutes.

**Quartiles** locate the lower and upper quarters of an ordered distribution.
The first quartile $Q_1$ is approximately the 25th percentile and the third
quartile $Q_3$ approximately the 75th. For these small examples, use the median
of the lower half and the median of the upper half, excluding the overall
middle observation when the number of values is odd. Here,

$$
Q_1=(1+3)/2=2.0000,\qquad Q_3=(3+5)/2=4.0000,\qquad
\operatorname{IQR}=4-2=2.0000\text{ minutes}.
$$

The **interquartile range (IQR)** is the width of the middle half of the
distribution. Software may interpolate between ordered values instead; the
large-data summaries in this chapter use R's default quartile convention.
Different conventions can give slightly different answers in small datasets.

To describe spread around the mean, consider each observation's **deviation**:
its value minus the mean. The deviations sum to zero because positive and
negative distances cancel. Squaring them preserves the contribution of each
distance to spread.

| Observation $x_i$ | Deviation $x_i-3$ | Squared deviation $(x_i-3)^2$ |
|-----------------:|-----------------:|----------------------------:|
| 1 | $-2.0000$ | 4.0000 |
| 3 | 0.0000 | 0.0000 |
| 3 | 0.0000 | 0.0000 |
| 5 | 2.0000 | 4.0000 |
| **Sum** | **0.0000** | **8.0000** |

: *Working from observations to squared-deviation spread.* {#tbl-worked-variance tbl-colwidths="[25,30,45]"}

The conventional **sample variance** divides this sum of squared deviations
by $n-1$, where $n$ is the number of observations. The **sample standard
deviation (SD)** is its square root:

$$
s^2=\frac{8}{4-1}=\frac83=2.6667\text{ squared minutes},\qquad
s=\sqrt{\frac83}=1.6330\text{ minutes}.
$$

Keep the unrounded variance while taking its square root. Variance expresses
spread in squared units; SD returns to the units of the observations and gives
a scale for their deviations from the mean.

```{=latex}
\Needspace{10\baselineskip}
\phantomsection
```

Why does $n-1$ appear? Deviations from their own mean must sum to zero. Once
the first three deviations $-2,0,0$ are known, the last must be 2. Only three
can vary freely. More generally, estimating the center uses one **degree of
freedom**, leaving $n-1$. This explains how many deviations can vary freely
under the zero-sum constraint.
The formula can summarize a fixed set of observed values. Its further role
in estimating a population variance, including the sampling reason for this
divisor, will be developed with estimation.

We can now write the general formulas. In $x_1,\ldots,x_n$, the subscript
$i$ identifies an observation. The symbol $\sum$ means to add the indicated
terms: $\sum_{i=1}^n x_i=x_1+x_2+\cdots+x_n$.

$$
\bar{x}=\frac{1}{n}\sum_{i=1}^{n}x_i,\qquad
s^2=\frac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar{x})^2,
$$

$$
s=\sqrt{s^2},\qquad \operatorname{IQR}=Q_3-Q_1.
$$

For any ordered dataset, the median is the middle observation when $n$ is
odd and the average of the two middle observations when $n$ is even.
@tbl-five-summaries brings the five summaries together.

| Summary | Interpretive job | Units | Main caution |
|:--------|:-----------------|:------|:-------------|
| Mean $\bar{x}$ | Arithmetic balance point | Original units | Can move strongly with skewness or extreme values |
| Median | Ordered midpoint | Original units | Does not describe spread or tail shape |
| Sample variance $s^2$ | Squared-deviation spread, using divisor $n-1$ | Squared units | Harder to interpret directly on the original scale |
| Sample standard deviation $s$ | Scale of deviation around the mean | Original units | Can be influenced by extreme values |
| IQR | Width of the middle half | Original units | Leaves the tails undescribed |

: *Five selected summaries and the questions they answer.* {#tbl-five-summaries tbl-colwidths="[19,33,17,31]"}

These five summaries differ from the **five-number summary**, which lists
the minimum, $Q_1$, median, $Q_3$, and maximum.

A fixed set of values also has a mean squared deviation obtained by dividing
by $n$: $8/4=2.0000$ squared minutes here. Both conventions describe spread when
clearly named. We use the conventional sample variance throughout, including
when describing the 395 grades. The choice of divisor alone says
nothing about how the observations were selected.

```{r}
#| label: tbl-grade-summary
#| tbl-cap: "Five numerical summaries of final mathematics grade."
#| tbl-colwidths: [32, 24, 28]

grade_print <- grade_summary
grade_print$Value <- sprintf("%.4f", grade_summary$Value)
knitr::kable(
  grade_print,
  booktabs = TRUE,
  align = c("l", "r", "l")
)
```

The mean grade is 10.4152 and the median is 11.0000. The variance is 20.9896 squared
grade points; SD is 4.5814 grade points, and IQR is 6.0000 grade points. The graph
helps us interpret these summaries, particularly the visible group at zero.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

### Equal means can hide different distributions

@fig-same-mean shows three invented distributions with the same mean
of 10. Their shapes and spreads are plainly different.

```{r}
#| label: fig-same-mean
#| fig-cap: "Three invented distributions share a mean of 10.0000 but differ in shape and spread."
#| fig-alt: "Faceted dot plots show a concentrated distribution around ten, a broad evenly spaced distribution, and a right-skewed distribution. Each has mean ten."
#| fig-height: 2.1
#| fig-pos: "H"

ggplot(same_mean, aes(x = Value, y = 0)) +
  geom_dotplot(binaxis = "x", stackdir = "up", binwidth = .55,
               fill = "#6C9EAE", color = "#28424F") +
  geom_vline(xintercept = 10, linetype = 2, color = "#9B3D45") +
  facet_wrap(~ Pattern, nrow = 1) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(x = "Observed value")
```

The mean gives a balance point, while the graph shows how the values are
arranged around it. Mean and SD are useful when the center and spread around
that center serve the question. Median and IQR are **resistant** summaries:
they change relatively little when a small number of extreme values changes.
They often describe a skewed distribution more clearly. In either case,
the graph supplies information about tails, gaps, and clusters.


@fig-ch02-skew-centers compares three further invented shapes without
holding their means fixed. The symmetric values have mean and median
6.0000. For the right-skewed values, the longer upper tail pulls the
mean to 5.1111 while the median is 4.0000. Reflecting those values gives
the left-skewed panel, with mean 4.8889 and median 6.0000. The dashed
and dotted lines mark the mean and median. This ordering is a common
pattern to interpret alongside the graph, not a theorem for every
distribution described as skewed.

```{r}
#| label: fig-ch02-skew-centers
#| fig-cap: "Three invented shapes: the mean moves toward the longer tail."
#| fig-alt: "The symmetric values have mean and median six. Right-skewed values have mean 5.1111 and median four; reflected left-skewed values have mean 4.8889 and median six. Dashed and dotted lines mark mean and median."
#| fig-height: 2.5
#| fig-pos: "H"

shapes <- list(Symmetric = c(2, 3, 4, 5, 6, 7, 8, 9, 10),
               Right = c(1, 2, 2, 3, 4, 5, 6, 8, 15))
shapes$Left <- 10 - shapes$Right
op <- par(mfrow = c(1, 3), mar = c(4, 2, 2, 1))
for (nm in names(shapes)) {
  v <- shapes[[nm]]
  stripchart(v, method = "stack", pch = 19, xlim = c(-5, 15),
             xlab = "Value", main = nm)
  abline(v = mean(v), lty = 2, col = "firebrick")
  abline(v = median(v), lty = 3, col = "navy")
}
par(op)
```

### Comparing groups while keeping their variation visible

A single distribution can combine groups with different centers or
spreads. In the school data, grades might be displayed separately by
school. Before doing that with many students, consider six invented
task-completion times from two training groups:

| Group | Completion times (minutes) | Number of learners | Mean (minutes) |
|:------|:---------------------------|-------------------:|---------------:|
| A | 1, 3 | 2 | 2.0000 |
| B | 3, 3, 5, 9 | 4 | 5.0000 |

: *A small grouped dataset for describing center and spread* {#tbl-ch02-group-times tbl-colwidths="[15,39,24,22]"}

A **grouping variable** identifies the groups being compared. The
two-category variable here is **binary**, or **dichotomous**. We can
represent it by an **indicator**, also called a **dummy variable**:
let $D_i=1$ for Group B and $D_i=0$ for Group A. In row order,

$$
(D_1,\ldots,D_6)=(0,0,1,1,1,1),\qquad
\bar D=\frac46=.6667.
$$

The mean of a 0/1 indicator is the proportion in the coded-one category.
Reversing the coding changes that category and the direction used for
a difference between the group means; the observed groups remain the
same. With three or more categories, codes 1, 2 and 3 do not establish
quantitative spacing. Later regression uses a declared set of indicators
to represent such group membership.

Group B takes three minutes longer on average, but its learners are
also more spread out. The dotplot in @fig-ch02-groups keeps both
features visible. Some learners in different groups have the same time,
and the group means summarize rather than replace those observations.

```{r}
#| label: fig-ch02-groups
#| fig-cap: "Completion times and group means in the invented six-learner example"
#| fig-alt: "Group A has times one and three, with mean two. Group B has times three, three, five, and nine, with mean five. The two three-minute learners in group B are shown separately. Diamonds mark group means."
#| fig-height: 2.7
#| fig-pos: "H"
group_plot <- group_times
group_plot$Position <- c(.96, 1.04, 1.94, 2.06, 2.14, 2)
ggplot(group_plot, aes(x = Minutes, y = Position)) +
  geom_point(size = 2.6, color = "#315D70") +
  geom_point(data = data.frame(Minutes = c(2, 5), Position = c(1, 2)),
             shape = 18, size = 3.5, color = "#A24B2A") +
  scale_x_continuous(breaks = 0:10, limits = c(0, 10)) +
  scale_y_continuous(breaks = c(1, 2), labels = c("A", "B"),
                     limits = c(.6, 2.4)) +
  labs(x = "Completion time (minutes); diamonds mark means", y = "Group")
```

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

The **grand mean** is the mean across all six learners:

$$
\bar x=\frac{1+3+3+3+5+9}{6}=\frac{24}{6}=4.0000\text{ minutes}.
$$

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

We can obtain it from the group means too, provided each group receives
its learner count as a weight:

$$
\bar x=\frac{2(2)+4(5)}{2+4}=4.0000.
$$

The unweighted average $(2+5)/2=3.5000$ would give the two groups equal
influence despite their different sizes. That answers a different
averaging question. When the aim is to describe all learners equally,
the larger group contributes more to the grand mean.

There is also variation **within** each group. For Group A, the squared
deviations from 2 are 1 and 1, so its sample variance is
$2/(2-1)=2.0000$ and its SD is 1.4142 minutes. For Group B, deviations from
5 are $-2,-2,0,4$. Thus,

$$
s_B^2=\frac{4+4+0+16}{4-1}=8.0000,\qquad
s_B=\sqrt{8}=2.8284\text{ minutes}.
$$

Keeping each group's spread makes the comparison more informative than
reporting two means alone. These calculations describe the groups; they
do not assume a common variance.

Finally, one learner's distance from the grand mean can be understood
in two steps. The nine-minute learner belongs to Group B, whose mean
is five minutes. That learner is four minutes above the group mean,
and the group mean is one minute above the grand mean:

$$
9-4=(9-5)+(5-4)=4+1=5.0000\text{ minutes}.
$$

In words, the total deviation combines the learner's position
**within the group** and the group's position relative to the
**overall mean**. This elementary identity prepares the later
analysis of between-group and within-group variation. For now,
the displays, means, and separate SDs give the descriptive comparison.
Appendix A develops weighted means and squared deviations a little
further.

### Common probability models

An empirical distribution describes recorded observations. A
**probability model** instead describes possible observations and their
probabilities. A **random variable** gives a numerical description of a
possible observation: uppercase $X$ denotes the variable, and lowercase
$x$ an observed or specified value. A categorical observation can be
recorded by a label or represented by indicators; arbitrary category
codes do not become quantitative measurements.

The model mean, or **expectation**, $E(X)$ is a probability-weighted
average. With finitely many possible values,

$$
E(X)=\sum_x xP(X=x).
$$

For a Bernoulli response, $X=1$ means correct and $X=0$ incorrect. If
$P(X=1)=p$, its mean is $0(1-p)+1p=p$. A correct-response probability
of .7 therefore gives a model mean of .7, even though each individual
response is either zero or one. The model variance,
$\operatorname{Var}(X)$, describes squared spread around that mean.
@tbl-ch02-models connects several familiar variable forms to useful
models [@nistDistributionGallery]. The mathematical properties in the
last column describe those models; an observed shape alone does not
establish that a model is appropriate.

| Model and support | Meaning and educational example | Parameters and useful properties |
|:--|:--|:--|
| Bernoulli: $X=0,1$ | One correct/incorrect response | $P(X=1)=p$; $E(X)=p$, $\operatorname{Var}(X)=p(1-p)$ |
| Binomial: $X=0,\ldots,m$ | Number correct in $m$ independent trials with common success probability; item heterogeneity can violate the model | $X\sim\operatorname{Bin}(m,p)$; mean $mp$, variance $mp(1-p)$ |
| Categorical: one of $K$ labels | A student's selected learning format | Probabilities $\pi_1,\ldots,\pi_K$ sum to one; each category has an indicator; numerical means of arbitrary labels are not meaningful |
| Poisson: $X=0,1,\ldots$ | Number of help requests during a specified observation period | $X\sim\operatorname{Pois}(\lambda)$; mean and variance $\lambda$; different exposures require explicit treatment |
| Normal: real values | A continuous score model when its symmetric shape is adequate | $X\sim N(\mu,\sigma^2)$; mean $\mu$, SD $\sigma$ |
| Lognormal: positive values | A positive completion time under a multiplicative model | $\log X\sim N(\mu_{\log},\sigma_{\log}^2)$; median $\exp(\mu_{\log})$, mean $\exp(\mu_{\log}+\sigma_{\log}^2/2)$ |

: Models for different kinds of possible observations. {#tbl-ch02-models tbl-colwidths="[23,40,37]"}

The notation $\sim$ means "is distributed as." For example,
$X\sim\operatorname{Bin}(m,p)$ states a Binomial model with $m$ trials
and common success probability $p$. Here $m$ counts trials within one
modeled observation, while $n$ counts recorded cases. One Bernoulli
trial is a Binomial with $m=1$ and also a two-category categorical model.
A Binomial count has an upper bound $m$; a Poisson count has no fixed
upper bound. Observation periods matter for counts: a help-request
rate for one lesson cannot be applied unchanged to a whole term.
The Lognormal describes positive values whose logarithms follow a
Normal model; its log-scale SD differs from its original-scale SD.

For **discrete** outcomes, probabilities at the possible values add to
one. For a **continuous** variable, a density curve has total area one,
and the probability of a range is its area. A curve height is not a
probability at one exact value. @fig-ch02-model-gallery shows masses
and densities separately. The displayed axes show informative ranges;
they do not make a Normal, Lognormal or Poisson tail end at the plot
boundary. Positive skew alone establishes none of these models.

```{r}
#| label: fig-ch02-model-gallery
#| fig-cap: "Six probability models with declared parameters: masses for discrete outcomes and density curves for continuous outcomes."
#| fig-alt: "Bernoulli probabilities at zero and one are .3 and .7. A Binomial with ten trials and success probability .7 peaks near seven. Three category probabilities are .2, .5 and .3. A Poisson with mean three has a right tail. A Normal with mean fifty and SD ten is symmetric; a Lognormal with log mean two and log SD .5 is positive and right-skewed."
#| fig-height: 5.5
#| fig-pos: "H"

op <- par(mfrow = c(2, 3), mar = c(4, 4, 2, 1),
          mgp = c(2.2, .7, 0))
plot(0:1, dbinom(0:1, size = 1, prob = .7), type = "h", lwd = 4,
     ylim = c(0, 1), xlab = "Outcome", ylab = "Probability",
     main = "Bernoulli: p = .7", xaxt = "n")
axis(1, at = 0:1)
plot(0:10, dbinom(0:10, size = 10, prob = .7), type = "h", lwd = 3,
     xlab = "Number correct", ylab = "Probability", main = "Binomial")
barplot(c(Reading = .2, Practice = .5, Discussion = .3),
        names.arg = c("Read", "Practice", "Discuss"), cex.names = .9,
        ylim = c(0, 1), ylab = "Probability", main = "Categorical")
plot(0:12, dpois(0:12, lambda = 3), type = "h", lwd = 3,
     xlab = "Requests per period", ylab = "Probability", main = "Poisson")
curve(dnorm(x, mean = 50, sd = 10), from = 10, to = 90,
      xlab = "Score", ylab = "Density", main = "Normal")
curve(dlnorm(x, meanlog = 2, sdlog = .5), from = 0, to = 35,
      xlab = "Minutes", ylab = "Density", main = "Lognormal")
par(op)
```

The **cumulative distribution function (CDF)**,
$F(x)=P(X\le x)$, gives the probability up to a specified value.
A **quantile** is a value reaching a stated cumulative probability.
These ideas let us ask both "What proportion is below this score?"
and "What score marks the upper 5%?" The next subsection works both
questions with a Normal model.

A brief preview connects this catalogue to later regression. The
**logistic distribution** is continuous and symmetric, with location
$\mu_L$ and positive scale $s_L$ [@rCoreLogistic]:

$$
F(x)=\frac{1}{1+\exp[-(x-\mu_L)/s_L]},\qquad
SD(X)=\frac{\pi s_L}{\sqrt3}.
$$

Its scale is therefore different from its SD. With location zero and
scale one, its S-shaped CDF is the inverse-logit mapping used in
Chapter 11 to turn a linear predictor into a probability. The binary
outcome in logistic regression follows a Bernoulli model; it is not
itself a continuous logistic variable.

### Linear transformations, standardized values, and Normal areas

Changing measurement units connects numerical summaries to the scale
on which we read them. Suppose $w_i=a+bx_i$: every value is multiplied
by the same constant $b$, then shifted by $a$. Substitution into the mean
gives

$$
\bar w=\frac1n\sum_{i=1}^n(a+bx_i)
=a+b\bar x.
$$

Each deviation becomes $w_i-\bar w=b(x_i-\bar x)$. Squaring the
deviations therefore multiplies their sum by $b^2$, so

$$
s_w^2=b^2s_x^2,\qquad s_w=|b|s_x.
$$

The four completion times 1, 3, 3 and 5 minutes have mean 3.0000
minutes and SD 1.6330 minutes. Converting to seconds multiplies the
mean by 60 to give 180.0000 seconds and the unrounded SD by 60 to give
97.9796 seconds. Adding a constant moves the center without changing
the spread. A negative multiplier reverses the order, while its absolute
value scales SD. A zero multiplier produces a constant with zero spread.
The corresponding model properties are

$$
E(a+bX)=a+bE(X),\qquad
\operatorname{Var}(a+bX)=b^2\operatorname{Var}(X).
$$

A **standardized score**, or **z score**, expresses distance from a
specified mean in SD units. In an invented commute sample with mean
26 minutes and SD 6 minutes, a 38-minute commute gives

$$
z=\frac{38-26}{6}=2.0000.
$$

For observed values, $z_i=(x_i-\bar x)/s$. Subtracting the mean
changes the origin and dividing by a positive SD changes the scale.
The ordering and shape remain the same; standardizing skewed values
leaves them skewed. Standardization is undefined when the SD is zero.

A Normal model supplies the additional information needed for Normal
probability calculations. If $X\sim N(\mu,\sigma^2)$, then

$$
Z=\frac{X-\mu}{\sigma}\sim N(0,1).
$$

The **standard Normal distribution** has mean zero and SD one. Its
symmetry and density let us read central probabilities in
@fig-standard-normal:

$$
P(|Z|\le1)=.6827,\quad P(|Z|\le2)=.9545,\quad P(|Z|\le3)=.9973.
$$

```{r}
#| label: fig-standard-normal
#| fig-cap: "Central Normal probabilities within one, two and three SDs."
#| fig-alt: "Three identical standard Normal curves shade the areas between minus and plus one, two, and three. Their central probabilities are .6827, .9545 and .9973."
#| fig-height: 2.6
#| fig-pos: "H"

op <- par(mfrow = c(1, 3), mar = c(4, 4, 2, 1),
          mgp = c(2.2, .7, 0))
for (k in 1:3) {
  curve(dnorm(x), from = -4, to = 4, ylim = c(0, .42),
        xlab = "z", ylab = "Density",
        main = sprintf("Within %d SD: %.4f", k, 2 * pnorm(k) - 1))
  zz <- seq(-k, k, length.out = 300)
  polygon(c(-k, zz, k), c(0, dnorm(zz), 0),
          col = "lightblue", border = NA)
  curve(dnorm(x), add = TRUE)
}
par(op)
```

These give the familiar approximate 68%, 95%, 99.7% rule. The exact
central 95% range uses approximately $\pm1.9600$ rather than exactly
$\pm2$. Let $\Phi$ denote the standard Normal CDF and $\Phi^{-1}$ its
quantile function. Under a score model $X\sim N(50,10^2)$, a score of
65 is $(65-50)/10=1.5000$ SDs above the mean. Therefore,

$$
P(X\le65)=P(Z\le1.5)=\Phi(1.5)=.9332.
$$

About 93.3193% of values lie at or below 65 under this model. To find the
95th percentile, begin with the standard Normal value
$\Phi^{-1}(.95)=1.6449$ and transform back, retaining its unrounded
value during calculation:

$$
x_{.95}=50+10\Phi^{-1}(.95)=66.4485.
$$

The subscript on $x_{.95}$ marks a percentile, rather than a case
number. About 95% of the model's values are at or below 66.4485.
The calculations use a supplied probability table or quantile value.
Applying them to recorded scores requires a suitable Normal model;
computing sample z scores alone does not establish that suitability.

## Describing and Interpreting Relationships

Describing a relationship means keeping two measurements together.
The useful summary depends on what those measurements represent. We
first use two categorical variables from the school dataset to show how
counts become meaningful percentages. We then return to the chapter's
absence--grade question, which pairs two quantitative variables. The
last two subsections interpret the resulting description in the study
setting that produced it.

### Counts and proportions for two categorical variables

The dataset includes whether each student wanted to pursue higher
education. This is a reported intention, not a record of eventual
enrollment. We can compare it with **pass status**, defined here as a
final mathematics grade of 10 or higher. The following table counts
every student once.

```{r}
#| label: tbl-ch02-intention-pass
#| tbl-cap: "Higher-education intention and mathematics pass status for the 395 students"
#| tbl-colwidths: [40, 20, 22, 18]
counts_print <- data.frame(
  "Intends higher education" = c("No", "Yes", "Total"),
  "Passed" = c(intention_counts[, 1], sum(intention_counts[, 1])),
  "Did not pass" = c(intention_counts[, 2], sum(intention_counts[, 2])),
  "Total" = c(rowSums(intention_counts), sum(intention_counts)),
  check.names = FALSE
)
knitr::kable(counts_print, booktabs = TRUE, row.names = FALSE,
             align = c("l", "r", "r", "r"))
```

A **two-way frequency table** groups cases by two categorical variables.
Here the counts tell us that 20 students reported no intention to pursue
higher education and 375 reported that intention. Because the groups
are so different in size, their raw passing counts are an incomplete
comparison.

To compare passing within the two groups, divide each passing count by
the size of its own group:

$$
\frac{7}{20}=.3500=35.0000\%,\qquad
\frac{258}{375}=.6880=68.8000\%.
$$

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

These are **conditional proportions**: each describes passing among
students meeting a stated condition. The difference is

$$
68.8000\%-35.0000\%=33.8000\text{ percentage points}.
$$

Thus the passing proportion is higher among the students reporting a
higher-education intention. Percentage points express the subtraction
of two percentages; the calculation above is not a relative percentage
increase.

Changing the question changes the denominator. Among the 265 passing
students, the proportion reporting higher-education intention is
$258/265=97.3585\%$, rather than the 68.8000% passing within the
intention group. For reference, $258/395=65.3165\%$ of all students both
report that intention and pass, while overall passing is
$265/395=67.0886\%$. The word *among* identifies the conditioning group.
These are descriptions of observed counts. Intention was not randomly
assigned, so their difference alone does not establish a causal effect.

### Form before coefficient

For two quantitative variables, a scatterplot shows one point for each
paired observation. The four invented examples in @fig-relationship-forms
show how different relationships can look before we summarize them with
a coefficient.

```{r}
#| label: fig-relationship-forms
#| fig-cap: "Linear, curvilinear, and patternless relationships require different descriptions."
#| fig-alt: "Four panels show a positive line, a negative line, a U-shaped curve, and no clear pattern."
#| fig-height: 3.4
#| fig-pos: "H"

ggplot(relationship_patterns, aes(x = X, y = Y)) +
  geom_point(alpha = .55, size = 1.2, color = "#2F6073") +
  geom_smooth(method = "loess", se = FALSE, linewidth = .8, color = "#A24B2A") +
  facet_wrap(~ Pattern, nrow = 2, scales = "free_y") +
  labs(x = "X", y = "Y")
```

Describe a scatterplot by:

1. **form**: linear, curved, clustered, or otherwise structured;
2. **direction**: increasing, decreasing, or mixed;
3. **strength**: how tightly observations follow the form;
4. **support**: where observations actually occur; and
5. **unusual observations**: points or groups that may change the summary.

*Nonlinear* or *curvilinear* describes form. It does not mean random,
unimportant, or causal. A strong U-shaped relationship can have Pearson's
correlation near zero because its negative and positive linear tendencies
cancel.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

### Covariance and Pearson's correlation

**Covariance** describes how paired deviations vary together. Deviations
on the same side of their respective means give a positive product:
both values can be above their means or both can be below them.
Deviations on opposite sides give a negative product. A value exactly
at its mean contributes a zero deviation and hence a zero product.

Consider measurements on four invented learners with $x=(1,3,3,5)$ practice hours
and $y=(2,2,4,4)$ score points. Both means are 3.0000. The deviations and their
products are

$$
x_i-\bar x=(-2,0,0,2),\qquad y_i-\bar y=(-1,-1,1,1),
$$

$$
(x_i-\bar x)(y_i-\bar y)=(2,0,0,2).
$$

The first pair shows why two below-mean values contribute positively:
$(-2)(-1)=2$. The products sum to 4. Dividing by $n-1=3$ gives the sample covariance
$s_{xy}=4/3=1.3333$ hour-points. The positive value agrees with the pattern: the
larger practice values tend to accompany the larger scores.

For any paired observations, the calculation is

$$
s_{xy}=\frac{1}{n-1}\sum_{i=1}^{n}
(x_i-\bar{x})(y_i-\bar{y}).
$$

Covariance carries the product of the two measurement units. To express the
linear relationship on a common, unit-free scale, **Pearson's correlation**
divides covariance by the two SDs:

$$
r=\frac{s_{xy}}{s_xs_y}.
$$

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

For the four pairs, $s_x=\sqrt{8/3}$ and $s_y=\sqrt{4/3}$. Substitution gives

$$
r=\frac{4/3}{\sqrt{8/3}\sqrt{4/3}}
=\frac{4}{\sqrt{8\times4}}=0.7071.
$$

The units cancel as well: hour-points divided by hours times points is
unit-free. If hours become minutes, both covariance and $s_x$ are multiplied
by 60. Covariance becomes $60(4/3)=80.0000$ minute-points, while $r$ remains .7071.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

The formula also has a useful squared-deviation form:

$$
r=\frac{\sum_i(x_i-\bar{x})(y_i-\bar{y})}
{\sqrt{\sum_i(x_i-\bar{x})^2\sum_i(y_i-\bar{y})^2}}.
$$

The $n-1$ factors cancel between covariance and the two SDs. This connection
between paired variation and squared deviations will also be useful in
regression.

A second connection uses standardized values. With
$z_{xi}=(x_i-\bar x)/s_x$ and $z_{yi}=(y_i-\bar y)/s_y$,

$$
r=\frac{1}{n-1}\sum_{i=1}^n z_{xi}z_{yi}.
$$

Positive standardized products contribute to a positive linear association;
negative products contribute to a negative one. The divisor is $n-1$
because these z scores use sample SDs. Correlation ranges from $-1$ to $1$,
keeps its value under positive rescaling, and reverses sign when one scale
is reversed. If either SD is zero, the denominator is zero and correlation
is undefined. Appendix A explains the mathematical bound and why zero correlation can
coexist with a relationship.

| Feature | Sample covariance $s_{xy}$ | Pearson correlation $r$ |
|:--------|:-------------------------------|:--------------------------|
| What it summarizes | Sum of paired-deviation products divided by $n-1$ | Standardized linear direction and strength |
| Units | Product of the two variables' units | None |
| Range | Unbounded; scale dependent | From $-1$ to $1$ |
| Effect of changing measurement units | Magnitude changes | Unchanged by positive linear rescaling |
| What it cannot reveal alone | Curvature, clusters, restricted support, or influential observations | The same omissions; near zero is not proof of no relationship |

: *Covariance and Pearson's correlation have connected but different jobs.* {#tbl-covariance-correlation tbl-colwidths="[23,38,39]"}


The graph remains essential: covariance and correlation summarize a linear
pattern, while a display shows curvature, clusters, sparse regions, and
unusual observations.

```{=latex}
\Needspace{8\baselineskip}
\phantomsection
```


@fig-ch02-r-gallery makes direction and strength visible on common
axes. Each panel contains 80 simulated pairs, and each label reports
the correlation actually computed from its displayed points. Moving
from a strong negative to a strong positive coefficient changes the
linear pattern. The earlier U-shaped example remains a reminder that
one linear summary cannot describe every form of relationship.

```{r}
#| label: fig-ch02-r-gallery
#| fig-cap: "Six simulated scatterplots show different directions and strengths of linear association; labels report the sample correlation of each panel."
#| fig-alt: "Six panels use common horizontal and vertical scales and equal sample sizes. They range from strongly negative through weak association to strongly positive. Each label gives its computed Pearson correlation."
#| fig-height: 4.8
#| fig-pos: "H"

set.seed(60206)
targets <- c(-.9, -.5, 0, .3, .6, .9)
panels <- lapply(targets, function(rho) {
  x <- rnorm(80)
  y <- rho * x + sqrt(1 - rho^2) * rnorm(80)
  data.frame(x = x, y = y)
})
lim <- range(unlist(panels))
op <- par(mfrow = c(2, 3), mar = c(4, 4, 2, 1),
          mgp = c(2.2, .7, 0))
for (d in panels) {
  plot(d$x, d$y, pch = 19, cex = .6, xlim = lim, ylim = lim,
       xlab = "x", ylab = "y",
       main = sprintf("Sample r = %.4f", cor(d$x, d$y)))
}
par(op)
```

Different questions can motivate different correlations
[@rCoreCorrelation]. **Ranks** replace each value by its ordered
position. Spearman correlation applies Pearson's calculation to those
ranks, so it can summarize a monotonic pattern that consistently rises
or falls without being straight. For Kendall's coefficient, consider
two cases at a time: their ordering is concordant if the case with the
higher value of one variable also has the higher value of the other,
and discordant if the order reverses. A tie-aware version handles equal
values. @tbl-ch02-correlation-types places these summaries beside
Pearson's coefficient.

| Correlation | Useful question | Key qualification |
|:--|:--|:--|
| Pearson $r$ | How strongly do two quantitative variables follow a linear pattern? | Always inspect their scatterplot and influential values |
| Spearman rank correlation $r_s$ | Do larger ranks on one variable tend to accompany larger ranks on the other? | Pearson correlation of ranks; describes monotonic rather than specifically linear association |
| Kendall $\tau_K$ | Do pairs of cases tend to have concordant rather than discordant ordering? | Use a version accounting for ties when needed; its scale is not numerically interchangeable with Pearson's |
| Point-biserial correlation | How does a quantitative outcome relate to a binary group indicator? | Pearson correlation with a 0/1 variable; reversing the coding reverses its sign |

: Different correlations summarize different patterns. {#tbl-ch02-correlation-types tbl-colwidths="[23,39,38]"}

We use $\tau_K$ for Kendall's coefficient to distinguish it from a
later mean-difference estimand. Point-biserial correlation is the
ordinary Pearson correlation when one variable is coded zero/one;
reversing the group coding reverses its sign. A non-Normal histogram
by itself does not determine which coefficient answers the relationship
question. Inspect the paired pattern, variable meaning and unusual
observations before choosing a summary.

### Returning to absences and grades

We can now specify the **estimand** for the school comparison: the Pearson
correlation between recorded absences and final mathematics grade among the
395 students. An estimand is the precisely specified quantity
sought by the question. Here it describes the linear relationship in this
complete paired dataset.

@fig-absence-grade preserves all 395 paired observations. Jitter makes
overlapping integer-valued points visible; it does not change the values used
for the coefficient.

```{r}
#| label: fig-absence-grade
#| fig-cap: "Recorded absences and final mathematics grade in the observational two-school dataset."
#| fig-alt: "A jittered scatterplot shows most students below twenty absences, a concentrated set at zero absences and zero final grade, a few much larger absence counts up to seventy-five, and little overall linear pattern. The smooth rises from zero absences to low positive counts, then declines, with fewer students in the high-absence tail."
#| fig-height: 3.1
#| fig-pos: "H"

set.seed(60205)
ggplot(students, aes(x = absences, y = final_grade)) +
  geom_jitter(width = .35, height = .18, alpha = .48,
              color = "#28536B", size = 1.35) +
  geom_smooth(method = "loess", se = FALSE, color = "#A24B2A",
              linewidth = .9) +
  scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60, 75)) +
  scale_y_continuous(breaks = seq(0, 20, 5), limits = c(-.5, 20.5)) +
  labs(
    x = "Recorded absences (count)",
    y = "Final mathematics grade\n(0--20)"
  )
```

Most students have relatively few recorded absences; a smaller number have
much larger counts. The sample covariance is 1.2557 absence-count-by-grade-point
units, and Pearson's $r=.0342$. The coefficient is positive but small: the overall linear
association among these students is close to zero. A **smooth** is a curve
summarizing nearby observations, helping us inspect a relationship's form.
Here it rises from zero absences to the low positive counts, then declines
gradually. The high-absence tail contains fewer students and supplies less
information about the curve's shape. The overall linear coefficient and
the local curve therefore describe different features of the same data.

Because all 395 pairs defining this estimand are observed, the
calculated correlation gives its value for this dataset directly. Coding and measurement questions remain relevant to its meaning.
Extending the result to students outside this dataset would require an
appropriate sampling design or other explicit basis for that inference.

All 38 zero-grade students also have zero absences; 25 have positive
second-period grades from 4 to 10. The accessible codebook leaves this
joint pattern unexplained. A **sensitivity analysis** examines how a
result changes when an analytical choice changes. Omitting these 38
records gives $r=-.2131$ among 357 students, compared with $.0342$
among all 395. Relative to the original means, the $(0,0)$ observations
contribute positive products of deviations; removing them also changes
the means used for every remaining deviation.

The omission is a diagnostic of a different record set, rather than
evidence that the zeros are errors. Its 357-student correlation describes
a different set from the original estimand. We keep the full set primary
and use the contrast to show why the unexplained coding pattern matters.

### Relationships after linear adjustment

A different descriptive question keeps the same cases and accounts for
a third variable. Consider separate, invented summaries of practice
time $x$, final score $y$, and baseline score $z$. Suppose the Pearson
correlations are $r_{xy}=.60$, $r_{xz}=.50$, and $r_{yz}=.60$, all
calculated from the same complete set of students. These are stipulated
summary inputs, not correlations from the Portuguese school dataset.

The **partial correlation** of practice time and final score after
linear adjustment for baseline is

$$
r_{xy\cdot z}=\frac{r_{xy}-r_{xz}r_{yz}}
 {\sqrt{(1-r_{xz}^2)(1-r_{yz}^2)}}
=\frac{.60-(.50)(.60)}{\sqrt{(1-.50^2)(1-.60^2)}}
=.4330.
$$

To understand the operation, imagine removing each variable's fitted
straight-line relationship with baseline score, then correlating the
two remaining parts. Those remaining parts are **residuals**. Chapter 7
develops fitted lines and their residual calculations, while Chapter 8
extends the adjustment to several variables. The formula above gives
the descriptive calculation directly. In this example, adjustment
reduces the correlation from .6000 to .4330.

The result describes association after this specified linear adjustment.
It does not establish a direct causal effect, remove every influence of
baseline, repair an ambiguous code or adjust for omitted variables.
Choosing what to adjust for depends on the study question and design.
A variable along the process from practice to the outcome, or one
affected by both quantities, can change or distort the intended
comparison when included as a control.

The two descriptions therefore answer different questions: the
zero-code check changes the cases being described; partial correlation
keeps the same cases and removes their linear relationships with a
stated third variable. Neither operation replaces the other.

### Interpret and report the relationship

The calculations now give us a description of the school dataset. To explain
what it means, we connect the patterns to the way the students were observed
and the measurements were made. This develops Chapter 1's design discussion
through the data we have just examined.

Before reporting the relationship, ask what could have produced it.
Prior academic difficulty, motivation, or illness could influence both
attendance and final performance, providing common-cause explanations
of the observed relationship. Earlier low grades might instead lead
to disengagement and later absence, illustrating reverse direction.
These are possibilities to investigate. The recorded pairs alone do
not establish which explanation applies; we need the timing of the
absences and assessments and a design that supports the intended
comparison. Chapter 1, Section 3.2 explains how sampling and assignment
contribute to that design.

Association in a causal comparison need not appear as a nonzero
Pearson correlation or a statistically significant result in the
sense Chapter 1 defines. A near-zero overall correlation can conceal
curvature or different subgroup patterns. Its value summarizes the
linear relationship among the recorded measurements; it does not
settle what would happen if attendance were changed.

We can now answer the chapter's question about how absences and grades
vary together in this set of students. A near-zero
linear correlation cannot settle the effect of changing attendance, and
the omission check shows why the coding context matters.

A results paragraph can state both the primary description and the
sensitivity:

> Among the 395 students recorded in two secondary schools, recorded absences
> and final mathematics grades showed almost no overall linear association
> ($r=.0342$). Omitting the 38 unexplained zero-grade students as a diagnostic
> gave a modest negative correlation ($r=-.2131$). The interpretation is
> therefore sensitive to those values. These observational results do not
> establish the effect of changing attendance or its importance for students
> in other settings.

Each qualification has a concrete role. The named students identify who was
described; *linear* identifies the feature summarized; the omission check
makes the sensitivity visible; and *observational* connects the
interpretation to the design. Rather than repeat a generic caution, explain the particular
limitation that matters for the result.

Further investigation would first seek an explanation of the joint zero
pattern in the codebook or source records. A newly discovered coding error
could be corrected and the calculations repeated with the change documented.
A causal question about improving attendance would instead need a design
that supports that comparison. These are different research tasks, even
when some data can be reused.

### From describing these students to learning about others

In the school dataset, no attendance intervention was assigned, and
the accessible source does not establish random sampling from all
secondary students. These features bound what the results can tell us
about changing attendance or describing students elsewhere.

This still leaves a useful task: describing these students accurately.
The analysis identifies the grade distribution, a difference in passing
proportions, and an absence--grade correlation that changes when a
particular set of values is omitted. Each result names what it
describes and supplies a starting point for further investigation.

If we instead selected another sample from a defined population, its
grades and summaries could differ. We can begin to understand that
repeat-to-repeat variation using the same tools we have already used:
graphs, means, and standard deviations. The next section makes the
repetition explicit, first through an invented school system and then
through a simulation whose behavior we can inspect.

## From One Sample to Repeated Samples

Until now, our distributions have described measurements from the
students or other cases in a dataset. A second kind of distribution
describes what happens to a summary when the sample changes. Seeing
this distinction is an important step toward understanding uncertainty:
a mean calculated from one sample is useful, but another sample could
give a different mean.

### From one sample to several sample means

Return to the invented school system introduced in Section 1.4 and the
selection procedure described in Section 1.5. The system contains
**20,000 enrolled students** on a specified date. We want their mean grade and select
100 students by simple random sampling without replacement. Each
possible set of 100 students has the same chance of selection, and a
student can appear only once in that sample. This setting supplies a
stated sampling design; it makes no additional claim about the
selection of the 395 students in the Portuguese dataset.

The mean grade of all 20,000 students is a **parameter**: a numerical
characteristic of the defined population. It is the estimand for this
question. The mean calculated from the 100 selected students is a
**statistic**: a numerical summary calculated from sample data. It
provides an estimate of the population mean. The contrast is simple:
the population mean is what we want to know, while the sample mean is
what we can calculate from the selected students.

Suppose the first sample mean is 10.6 grade points. If we could repeat
the same selection process, a second sample might give 11.2 and a third
10.9. These are illustrative values, not results calculated from the
Portuguese dataset. The population and its mean stay the same across
these imagined selections; the selected students, and therefore the
sample means, change. Students selected in one sample can appear
again in another, because each repetition starts from the full
population.

There are now two different sets of numbers to keep in view. Within
the first sample, the 100 individual grades have a distribution. Across
repetitions, the sample means 10.6, 11.2, 10.9, and so on have another
distribution. A dot in the first distribution represents a student's
grade; a dot in the second represents the mean from an entire sample
of 100 students.

```{r}
#| label: fig-ch02-sampling-process
#| fig-cap: "Repeating the same sampling process produces a distribution of sample means. The grade values are invented illustrations."
#| fig-alt: "The same population of twenty thousand students supplies three repeated samples of one hundred. Their sample means are ten point six, eleven point two, and ten point nine. The individual grades within a sample and the sample means across repetitions form different distributions."
#| fig-height: 3.1
#| fig-pos: "H"
par(mar = c(0, 0, 0, 0), family = "sans")
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 5), xaxs = "i", yaxs = "i")
rect(0.1, 1.5, 3.0, 3.5, col = "#E9F0F5", border = "#44657D")
text(1.55, 2.5, "Same population\n20,000 students", cex = 1.05)
sample_y <- c(4.2, 2.5, 0.8)
for (j in seq_along(sample_y)) {
  arrows(3.1, 2.5, 4.0, sample_y[j], length = .07, col = "#44657D")
  rect(4.1, sample_y[j] - .55, 7.0, sample_y[j] + .55,
       col = "#F3F5F5", border = "#647780")
  text(5.55, sample_y[j], paste0("Sample ", j, "\n100 students"), cex = 1.0)
  arrows(7.1, sample_y[j], 8.0, sample_y[j], length = .07,
         col = "#44657D")
}
text(9, 4.8, "Sample mean", cex = 1.0)
text(rep(9, 3), sample_y, c("10.6", "11.2", "10.9"), cex = 1.15)
```

The **sampling distribution** of a statistic describes its possible
values under a specified repetition of the sampling or data-generating
process. Here the statistic is the mean, the sample size is 100, and
the process is simple random selection from the school system. Naming
these features matters: changing the sample size or selecting whole
classes instead of individual students would change the repetition we
are considering.

In practice, researchers usually have one sample. Repeated sampling
helps us reason about the variation behind that one result. Simulation
makes this idea visible by creating many samples under conditions we
have explicitly chosen.

### See repeated sampling in a waiting-time model

We now change from bounded grades to waiting times in an invented
service where people wait for assistance. The chosen model allows
nonnegative waiting times without a fixed upper limit, making a
right-skewed distribution and its repeated means easy to inspect.
We use an **exponential probability model** for
individual waiting times. This model allows many short waits and
occasional long waits, giving a right-skewed distribution. We choose a
population mean of 10 minutes and a population SD of 10 minutes. The
two values are equal for this particular model; means and SDs generally
describe different features and need not be equal.

Each simulated waiting time is generated independently from that same
model. This simplifies the example: actual service data might show
dependence because several people arrive during the same busy period.
The simulation asks what happens under the specified independent
process, rather than claiming that all waiting-time data behave this
way.

The procedure is straightforward. First generate five waiting times,
then calculate their mean. Start afresh and do this 5,000 times,
saving one mean from each repetition. Next repeat the procedure with
30 waiting times per sample, and finally with 100. Each panel in
@fig-ch02-sampling-means therefore summarizes **5,000 sample means**,
not 5,000 individual waiting times.

```{r}
#| label: fig-ch02-sampling-means
#| fig-cap: "Simulated sampling distributions of the mean waiting time. Each panel contains 5,000 sample means from independent exponential waiting times with population mean and SD both 10 minutes. The dashed line marks the population mean."
#| fig-alt: "Three histograms share an axis from zero to thirty-five minutes. For samples of five, sample means are broad and right-skewed. For samples of thirty they are narrower and more symmetric. For samples of one hundred they are tightly concentrated around ten and approximately bell-shaped."
#| fig-height: 3.4
#| fig-pos: "H"
ggplot(sampling_means, aes(x = Sample_mean)) +
  geom_histogram(binwidth = .5, boundary = 0,
                 fill = "#527D98", color = "white", linewidth = .15) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "#9D4938") +
  facet_wrap(~ Sample_size, nrow = 1,
             labeller = labeller(Sample_size = function(x) paste("n =", x))) +
  scale_x_continuous(breaks = c(0, 10, 20, 30)) +
  coord_cartesian(xlim = c(0, 35)) +
  labs(x = "Sample mean waiting time (minutes)", y = "Number of sample means") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        strip.text = element_text(size = 12))
```

All three distributions are centered near 10 minutes, the mean built
into the model. With only five people per sample, the means vary
considerably and the distribution still has a noticeable right tail.
One long wait can pull a small sample's mean upward. With 100 people
per sample, individual long waits have less influence on the average,
and sample means tend to be closer to 10.

The simulation also connects with the Normal model introduced earlier.
The individual waiting times are right-skewed, yet the sample means
become more nearly bell-shaped as sample size increases. The variable
being graphed has changed: it is now a mean calculated from several
waiting times. Neither collecting more cases nor standardizing their
values makes the distribution of the individual waiting times Normal.

### Standard error as the spread of sample means

We can summarize each histogram just as we summarized grades: find its
center and its spread. The mean of the 5,000 sample means describes
the center. Their SD describes how much those means vary across the
simulated repetitions.

This introduces **standard error (SE)**: the standard deviation of a
statistic's sampling distribution. For a sample mean, SE describes
variation among sample means under the specified repetition. In this
simulation, the SD of the 5,000 means is an empirical approximation
to that SE. Statistical procedures can also calculate or estimate SE
from a model and the information in one sample; those calculations
build on the meaning we can see here.

```{r}
#| label: tbl-ch02-sampling-summary
#| tbl-cap: "Center and spread of the simulated sample means, in minutes"
#| tbl-colwidths: [18, 27, 28, 27]
sampling_print <- sampling_summary
sampling_print[-1] <- lapply(sampling_print[-1], function(x) sprintf("%.4f", x))
knitr::kable(
  sampling_print,
  col.names = c("Sample size", "Mean of sample means", "Empirical SE", "Model SE"),
  align = "rrrr", booktabs = TRUE, row.names = FALSE
)
```

The empirical SE column in @tbl-ch02-sampling-summary comes from the
simulation. The model SE column gives the corresponding reference
values for the independent waiting-time model; the general calculation
is developed with statistical inference. Comparing the two columns
checks whether the simulation is behaving as expected. Their small
differences reflect the finite number of simulation repetitions.

For samples of five, a mean varies on a scale of about 4.4721 minutes.
For samples of 100, that scale is about 1.0000 minute. Both are much smaller
than the **10-minute SD of individual waiting times**, which stays the
same in all three conditions. SD and SE are therefore answering
different questions: how much do people's waiting times differ, and
how much would the mean differ across samples of the stated size?

It also helps to distinguish SE from the discrepancy in one particular
sample. Suppose one sample of 100 has a mean of 11.3 minutes. Because
the model's population mean is known to be 10, this sample's estimation
error is $11.3-10=1.3000$ minutes. **Sampling error** is the part of an
estimate's departure from the population quantity that arises from
observing a sample rather than the whole population. Under this
illustration's stated model, the 1.3000-minute departure is attributable
to sampling. Other samples can have positive or
negative errors. SE describes the spread across those possible
results; it is neither the signed error of this one result nor a bound
that every error must fall within. In an actual study the population
mean is usually unknown, so the exact error of the observed estimate
is unknown too.

Coverage, nonresponse, and measurement error are separate sources of
error. A missing group in the sampling frame, a selected student's
nonresponse, or an inaccurate measurement can change what an estimate
describes. SE characterizes sampling spread under the stated process;
it does not encompass every source of error in a study.

### Sample size and number of repetitions

The simulation has two counts with different jobs. **Sample size** is
the number of people whose waits enter one mean: 5, 30, or 100.
**Number of repetitions** is the number of independently generated
samples used to display how that mean varies: 5,000 for each condition.

Increasing sample size changes the statistic's sampling distribution.
Each mean combines more observations, and in this independent model
its spread becomes smaller. Increasing the number of repetitions gives
a clearer picture of the sampling distribution for the existing
sample size. For example, generating 20,000 samples of five would make
the first histogram smoother and its empirical SE more stable. It
would still describe means based on five people, with an SE close to
4.4721 minutes.

The model SE column shows a useful square-root regularity. Moving
from 5 to 100 people multiplies sample size by 20, while SE falls
from about 4.4721 to 1.0000, a reduction by about $\sqrt{20}=4.4721$ rather
than by 20. Under the same independent model, quadrupling sample
size from 100 to 400 would halve SE from 1.0000 to 0.5000 minute. This
explains why substantial gains in precision can require substantial
increases in sample size.

The conditions remain important. Sampling students in the same class
or observing patients during the same busy period can supply less
independent information than the same number of unrelated observations.
More cases can improve precision while leaving a systematic selection
problem in place. A very large volunteer sample may describe its
volunteers precisely and still differ from the population of interest.

### What stabilizes and what becomes more nearly Normal

Two changes are visible as sample size increases, and it helps to
describe them separately. First, the means concentrate more closely
around 10. Larger samples are less likely to give a mean far from the
model's population mean. This illustrates the stabilization of an
average: short and long waits tend to balance more effectively across
many observations.

Second, the distribution of sample means becomes more nearly symmetric
and bell-shaped. Shape and spread are different features. A narrow
distribution need not be Normal, and a Normal distribution can be wide
or narrow. Here the three panels show both concentration and a change
in shape, under one explicitly stated model.

The law of large numbers and the central limit theorem provide formal
accounts of these ideas under their respective conditions. The
simulation gives us something concrete to attach to those later
accounts: averages settling near a population value, and the shape of
their variation across samples becoming more nearly Normal. A sample
size such as 30 is not a universal threshold; the behavior depends on
the distribution and dependence of the observations.

The progression from grades to sample means is now complete. We began
by describing variation among cases, then used the same descriptive
tools to examine variation among summaries. The next step is to use a
stated sampling or data-generating process to quantify uncertainty
about an estimand from the one sample actually available.

## Reading Bridge

This chapter began by reading cases, variables, layouts, denominators,
and sample-selection procedures. We then turned measurements into
descriptions: the shape, center, and spread of one distribution;
group means with separate spreads and a count-weighted grand mean;
conditional proportions for two categorical variables; and a
scatterplot with covariance and correlation for two quantitative ones.
Grouping indicators made category membership explicit. Probability models
describe possible values; transformations and Normal areas connect
measurement scales with probabilities. Partial correlation adds a
specified linear adjustment while retaining the same cases.

We then interpreted a description in its study setting and watched the
same tools describe sample means under a stated repetition. A sampling
distribution shows how a statistic varies, and its standard error
describes that spread. One sample's sampling error is its particular
departure attributable to sampling. The selection design matters to
the repetition and its spread.
Larger samples and more simulation repetitions change different things.

Chapter 3 uses the stated repetition and its conditions to calculate
uncertainty for an estimate from one sample, through standard errors,
confidence intervals, and tests. Chapter 4 uses group means, deviations,
and spread to compare several groups with one-way and two-way analysis
of variance.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

### Key terms {.unnumbered #sec-ch02-key-terms}

Use this short reference to revisit the main concepts as needed. The
Part One introduction gives the broader reference and the principal
home of each concept; familiar alternative labels appear there when
helpful.

| Term and common alternatives | Meaning |
|:--|:--|
| Case | The entity or measurement occasion described in a dataset. Depending on the study, a case may be a student, a class, or one measurement occasion for a student. |
| Variable | A characteristic that can take different values across cases or occasions. A dataset records values of one or more variables for its cases. |
| Quantitative variable (numerical variable) | A variable expressing an amount, including continuous measurements and discrete counts or scores. Its units and scale determine what numerical differences mean. |
| Categorical variable | A variable whose values identify categories. Nominal categories have no intrinsic order; ordinal categories have an order; a binary variable has two categories. |
| Population | The collection of units a question concerns. It may be the participating students themselves or a wider collection to which the findings might extend. |
| Sample | A selected set of units from a larger collection. Calling a set a sample does not by itself explain how it was selected or which population it represents. |
| Analysis sample | The units whose data contribute to a particular analysis. Eligibility, missing measurements, and exclusions can make this set and its denominator differ across analyses. |
| Sampling design (sampling scheme) | The specified procedure for selecting a sample, including selection stages, grouping, and, for probability sampling, the selection probabilities. |
| Simple random sampling | Selection in which every possible sample of the specified size has the same probability of being chosen. |
| Empirical distribution (observed distribution; data distribution) | How the observed values are distributed across values or intervals, including their frequencies, center, spread, and shape. A probability model is a separate description of possible values. |
| Mean (arithmetic mean) | The sum of the numerical values divided by their count. Every included value contributes, so unusual values can shift the mean. |
| Median | The middle value after ordering the data, or the average of the two middle values when the count is even. It describes the center by position. |
| Quartile | A cut point dividing ordered data into quarters. The first and third quartiles mark approximately the lower and upper edges of the middle half; calculation conventions can differ. |
| Interquartile range (IQR) | The third quartile minus the first quartile: the width of the middle half of the ordered data, expressed in the variable's original units. |
| Variance | A measure of spread based on squared deviations from the mean. The usual sample variance divides their sum by $n - 1$ and is expressed in squared units. |
| Standard deviation (SD) | The square root of the variance, expressing spread around the mean in the variable's original units. Applied to individual observations, it describes variation among their values; applied to a statistic's sampling distribution, it is the SE. |
| Grand mean | The mean across all included cases. It equals the average of the group means weighted by their group counts, so larger groups contribute more. |
| Degrees of freedom (df) | A count reflecting freely varying information after constraints or fitted quantities are taken into account. The $n$ deviations from their own mean sum to zero, leaving $n - 1$ free. |
| Standardized value (z score) | A value minus a specified mean, divided by a specified standard deviation. It expresses distance in SD units; standardizing alone preserves the distribution's shape. |
| Normal distribution (Gaussian distribution) | A symmetric, bell-shaped probability model described by a mean and standard deviation. The standard Normal has mean zero and SD one; observed data need not follow this model. |
| Conditional proportion | The proportion in one category among cases satisfying a stated condition. Its denominator is the count in that conditioning group, such as students in one education-intention category. |
| Covariance | A summary based on products of paired deviations from the two means. Its sign shows the direction of joint variation; its magnitude depends on both variables' units and spreads. |
| Pearson's correlation (Pearson's r; product-moment correlation) | Covariance divided by the product of the two SDs. For variables with nonzero spread, it describes direction and strength of linear association on a scale from -1 to 1. |
| Sampling distribution | The distribution of a statistic, such as a sample mean, across repeated samples generated by a stated sampling procedure or model. Repeated sample summaries in a simulation approximate this distribution. |
| Sampling error | The part of an estimate's departure from the population quantity that arises from observing a sample rather than the whole population. It varies across repetitions, and its spread is the SE; coverage, nonresponse, and measurement error are separate. |
| Standard error (SE) | The standard deviation of a statistic's sampling distribution under a stated sampling procedure or model. It describes spread across repeated sample summaries, distinct from individual-value SD and one sample's difference from the population quantity. |
| Data matrix | A rectangular arrangement with cases in rows and variables in columns; $x_{ij}$ denotes case $i$ on variable $j$. |
| Binary variable (0/1 coding) | A binary variable has two possible categories; a 0/1 coding marks membership or occurrence. Its sample mean is the proportion coded 1. |
| Linear transformation | $Y=a+bX$ changes the center to $a+bE(X)$ and the SD to $\lvert b\rvert SD(X)$; a negative $b$ also reverses order. |
| Bernoulli; binomial; categorical; Poisson; lognormal; logistic | Respectively a binary outcome, a count of successes in fixed independent equal-probability trials, one category, an event count under a specified rate model, a positive variable with Normal log, and a continuous symmetric distribution used here as a preview. Logistic regression later models Bernoulli outcomes. |
| Density; cumulative distribution function; quantile | Density determines probability by area for a continuous variable; the CDF gives $P(X\le x)$; a quantile is a value reaching the stated cumulative probability. |
| Rank correlation; point-biserial correlation | Spearman uses ranks; Kendall summarizes concordance; point-biserial is Pearson correlation with one binary variable coded 0/1. Each addresses a specified relationship. |
| Partial correlation | Correlation between two variables after their linear relationships with stated controls are removed by regression; it describes residual association and does not establish causal adjustment. |
| Parameter | A numerical characteristic of a population or probability model, such as a population mean. A parameter may be the estimand; the question determines which quantity is sought. |
| Statistic | A quantity calculated from sample data, such as a sample mean or proportion. The same statistic can summarize the sample or serve as an estimate for a wider target. |

: Selected concepts for describing data and sampling variation {#tbl-ch02-key-terms tbl-colwidths="[28,72]"}

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

### Further reading {.unnumbered #sec-ch02-further-reading}

- OpenIntro connects graphical descriptions, numerical summaries, and
  foundations for inference [@cetinkaya2024].
- The ASA GAISE College Report emphasizes learning statistics through
  questions, data, analysis, and interpretation [@asa2016gaise].
- Moore, McCabe, and Craig develop the connections among descriptive
  summaries, sampling, and inference [@moore2014].
- NIST's distribution gallery and the R reference explain distribution
  properties and correlation definitions [@nistDistributionGallery;
  @rCoreCorrelation; @rCoreLogistic].
- The UCI source supplies the student-performance data and codebook
  [@cortez2008data].
- Statistics Canada's introductory sampling overview compares probability
  schemes and nonprobability selection
  [@statcan2021probability; @statcan2021nonprobability].

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

## Exercises {.unnumbered #sec-ch02-exercises}

These exercises can be completed by hand with a standard calculator.
Each supplies its own data or study description.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

<!-- prompt-id: CH02-E01 -->
### Exercise 1: Understand the cases and variables (Foundational) {.unnumbered #ex-ch02-01}

An invented clinic dataset contains information for individual visits. Some
patients have several visits. The variables are patient ID, clinic, visit
date, waiting time in minutes,
and whether the patient returned within 30 days. Of 1,240 visits, waiting time
is missing for 36, return status is missing for 81, and 10 visits are missing
both.

Identify:

1. the case represented by each observation;
2. the observational unit or occasion;
3. one identifier, one quantitative variable, and one categorical variable;
4. why 1,240 visits should not be described as 1,240 independent patients;
5. the complete-case denominator for an analysis using waiting time and return
   status; and
6. one question that must be answered before generalizing beyond these visits.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

<!-- prompt-id: CH02-E02 -->
### Exercise 2: See and summarize one distribution (Foundational) {.unnumbered #ex-ch02-02}

Eight invented completion times, in minutes, are:

> 2, 4, 4, 4, 5, 5, 7, 9

1. Sketch a dotplot or ordered display.
2. Calculate the mean, median, sample variance, sample standard deviation, and
   IQR. For quartiles, use the medians of the lower and upper halves.
3. State the units of the variance and standard deviation.
4. If the question asks for a resistant description of typical time and spread,
   choose a pair of summaries and explain why.
5. Name one feature that the chosen pair does not show.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

<!-- prompt-id: CH02-E03 -->
### Exercise 3: Interpret a Normal model and standardized value (Foundational) {.unnumbered #ex-ch02-03}

A reference model for a score is Normal with mean 50 and standard deviation 10.
A learner receives 65.

1. Calculate and interpret the learner's z score.
2. Distinguish this model-based standardization from using a sample mean and
   sample standard deviation.
3. Explain why transforming every observed score to z scores would not make a
   skewed observed distribution Normal.
4. State one question that cannot be answered without additional probability
   information or a Normal table.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

<!-- prompt-id: CH02-E04 -->
### Exercise 4: Compare a graph, covariance, and correlation (Foundational) {.unnumbered #ex-ch02-04}

In an invented study, a scatterplot of weekly practice hours and assessment score follows a clear
U-shaped curve. The sample covariance is 0.80 hour-points and Pearson's
correlation is .03.

1. Why can the graph show a strong relationship while $r$ is near zero?
2. What does the positive sign of the covariance summarize?
3. Why is the covariance magnitude not directly comparable across measurement
   units?
4. If hours are converted to minutes, what happens to the covariance and to
   Pearson's correlation?
5. Write one descriptive sentence that reflects the pattern in the graph.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

<!-- prompt-id: CH02-E05 -->
### Exercise 5: Choose the denominator for a categorical comparison (Foundational) {.unnumbered #ex-ch02-05}

In the mathematics dataset, 375 students reported an intention to pursue
higher education: 258 passed and 117 did not. Of the 20 students without
that intention, 7 passed and 13 did not.

1. Organize the counts in a two-way table with totals.
2. Calculate the passing proportion within each intention group and
   their difference in percentage points.
3. Among students who passed, what proportion reported a
   higher-education intention? Explain why its denominator differs
   from the denominators in part 2.
4. Explain why comparing 258 passes with 7 passes alone would be
   misleading.
5. Give one plausible reason for the association other than a causal
   effect of the reported intention.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

<!-- prompt-id: CH02-E06 -->
### Exercise 6: Describe groups and their grand mean (Integrative) {.unnumbered #ex-ch02-06}

An invented task produces the following completion times, in minutes:

- Group A: 2, 4.
- Group B: 4, 4, 6, 10.

1. Display the observations by group and calculate each group mean.
2. Calculate the grand mean from all six times, then reproduce it
   using the group means and group sizes.
3. Explain why an unweighted average of the two group means gives a
   different answer.
4. Calculate each group's sample variance and SD. Describe the
   difference in both center and spread.
5. Split the ten-minute learner's deviation from the grand mean into
   a within-group deviation and the group mean's deviation from the
   grand mean.
6. Does the difference in group means establish that membership in
   Group B caused a longer completion time? State the design
   information you would need.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

<!-- prompt-id: CH02-E07 -->
### Exercise 7: Interpret an incomplete observational dataset (Integrative) {.unnumbered #ex-ch02-07}

A report asks: "Among 260 volunteers in an online training course, how strong
is the linear relationship between logged study minutes and final quiz score?"
It documents measurements for each volunteer, 18 missing quiz scores, a Pearson
correlation of .41 among 242 complete cases, and a scatterplot showing a
moderate positive linear pattern and two volunteers whose study times are
much higher than those of the other volunteers.
The report concludes: "More study time raises scores for online learners
everywhere." The course platform's codebook is available, but the recruitment
notice and analysis script are not supplied.

1. Identify the predictor and outcome, their observational unit,
   and the group named by the research question. What precisely is the
   estimand for that group?
2. Explain why the calculated correlation uses 242 rather than 260 volunteers.
   Does it describe the relationship among all 260 without further information
   or assumptions about the missing scores?
3. Interpret $r=.41$ alongside the scatterplot. Explain what the coefficient
   summarizes and what inspecting the two unusually high study times could add.
4. Identify two things you would seek in the codebook, methods, or analysis
   details before interpreting the result more broadly. Explain why each
   would matter.
5. Explain the difficulty with "raises scores" and "online learners everywhere."
   Which features of the design and participant selection would be needed to
   support those claims?
6. Rewrite the conclusion for the 242 complete cases, preserving the
   distinction from the original 260-volunteer question and the uncertainty
   created by the missing scores.
7. Suppose checking those two volunteers' measurements reveals a coding error.
   Explain how you would correct and report the analysis. How would this
   differ from asking a new causal question about assigning additional
   study time?


```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

<!-- prompt-id: CH02-E08 -->
### Exercise 8: Distinguish individual variation from sampling variation (Integrative) {.unnumbered #ex-ch02-08}

Use the same kind of independent waiting-time model as in Section 4,
with population mean and SD both 10 minutes. For this exercise, suppose
a simulation generates 5,000 samples of 100 waiting times each and
supplies the following illustrative results: the mean of the 5,000
sample means is 10.02 minutes, their SD is 1.01 minutes, and one sample
has a mean of 11.3 minutes. These supplied results serve the exercise;
the chapter's particular seeded run has its own reported summaries.
The model SE for samples of 100 is 1.00 minute.

1. Identify the population parameter being estimated and the statistic
   calculated in each sample. What does one value in the simulated
   sampling distribution represent?
2. Explain the different meanings of the individual SD of 10 minutes
   and the empirical SE of 1.01 minutes. What does comparing 1.01 with
   the model SE of 1.00 help you check?
3. Calculate the estimation error of the sample mean 11.3. Explain how
   this one error differs from SE, and why its exact value is known in
   this simulation but usually unknown in a study.
4. Under the same independent model, what approximate SE would you
   expect for samples of 400? Explain using the square-root regularity.
   What would change instead if you kept sample size at 100 but
   generated 20,000 simulation repetitions?
5. A student concludes that individual waiting times must become
   Normally distributed when the sample size is large. Use the
   simulation to distinguish the distribution of individual waits
   from the distribution of their sample mean.

```{=latex}
\Needspace{12\baselineskip}
\phantomsection
```

## Appendix A: Weighted Means and Descriptive Identities {.unnumbered #sec-ch02-appendix}

These short developments explain relationships used in the main text.
They concern observed values and can be followed without a probability
model.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

### Why the grand mean uses group sizes {.unnumbered #sec-ch02-weighted-mean}

Let group $j$ contain $n_j$ observations with mean $\bar x_j$. The
sum of its observations is $n_j\bar x_j$. Adding those sums over $J$
groups and dividing by the total number of observations gives

$$
\bar x=\frac{\sum_{j=1}^{J}n_j\bar x_j}{\sum_{j=1}^{J}n_j}.
$$

For the two training groups, this is $(2\times2+4\times5)/6=4.0000$.
Equal group sizes make the simple average of group means agree with
the grand mean. Unequal sizes require the learner counts when each
learner is to contribute equally. Equal weighting of groups is also
possible, but it describes an average across groups rather than across
all learners.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

### Why the mean is a useful center for squared deviations {.unnumbered #sec-ch02-squared-deviations}

Take any proposed center $a$. Write each deviation as

$$
x_i-a=(x_i-\bar x)+(\bar x-a).
$$

After squaring and adding, the cross-products cancel because
$\sum_i(x_i-\bar x)=0$. Therefore,

$$
\sum_i(x_i-a)^2
=\sum_i(x_i-\bar x)^2+n(\bar x-a)^2.
$$

The last term is nonnegative and equals zero when $a=\bar x$.
Thus the mean gives the smallest sum of squared deviations.

For the four completion times $1,3,3,5$, the squared deviations around
their mean of 3 sum to 8. Around a proposed center of 4 they sum to
$9+1+1+1=12$. The identity gives the same answer:
$8+4(3-4)^2=12$. This is the descriptive basis for the least-squares
idea developed with statistical models.

The conventional sample variance is the sum around the observed mean
divided by $n-1$. Dividing by $n$ instead gives the mean squared
deviation of that fixed set. Which convention is reported should be
clear; the additional reason for using $n-1$ to estimate a population
variance belongs with sampling and estimation.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

### Why correlation is bounded {.unnumbered #sec-ch02-correlation-bound}

When both sample SDs are positive, the standardized values satisfy
$\sum_i z_{xi}^2=\sum_i z_{yi}^2=n-1$.
From the main-text correlation formula,

$$
\sum_i(z_{xi}-z_{yi})^2=2(n-1)(1-r).
$$

The left side is a sum of squares and cannot be negative, so $r\le1$.
Similarly,

$$
\sum_i(z_{xi}+z_{yi})^2=2(n-1)(1+r)\ge0,
$$

which gives $r\ge-1$. The endpoints require standardized values that
align perfectly, either in the same or the opposite direction.

Zero correlation can still accompany a fully patterned relationship.
For invented values $x=(-2,-1,0,1,2)$ and $y=x^2=(4,1,0,1,4)$,
the means are 0 and 2. The products of deviations are
$(-4,1,0,-1,4)$ and sum to zero. The covariance and correlation
therefore equal zero, although each $y$ value is determined by its
paired $x$ value. The U-shaped graph supplies the relationship that
the linear coefficient leaves out.

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

## References {.unnumbered #sec-ch02-references}
