---
title: "Description and Basic Inference with R"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: true
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \widowpenalty=10000
          \clubpenalty=10000
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{12\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                     collapse = TRUE, comment = "#>")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 10, width = 65)
```

\clearpage

::: {.callout-note title="Chapter overview"}
This chapter makes the descriptions of Chapter 2 and the mean inference
of Chapter 3 executable in R. We read and select observations, describe
distributions and groups, and simulate the variation of an estimator.
We then connect an estimate, its standard error, a reference distribution
and a critical value to an interval and a test. Relationships receive
the same treatment: a graph and covariance lead to correlation matrices,
selected correlation tests and one-control linear adjustment.
Throughout, the calculation stays beside the observations, conditions
and question that give it meaning. A short Quarto report connects the
final answer to the code that produced it.
:::

Three students scored 50, 60 and 70 points. Their mean is
$(50+60+70)/3=60.0000$ points. R gives the same result by applying a
function to a vector:

```{r}
#| label: ch05-scores
scores <- c(50, 60, 70)
mean(scores)
```

`c()` collects the numbers into a **vector**, an ordered collection of
values. The assignment symbol `<-` gives the vector the name `scores`.
`mean(scores)` supplies it to the function `mean()`, which returns 60.
Naming the observations lets later commands use the same values without
retyping them. This answer describes the three recorded scores directly.
Using it to estimate another group's mean would add a target and an
account of how these observations inform that target.

The chapter assumes no previous R course. Familiar browser examples from
Chapters 3–4 provide a starting point; a saved script now lets us run a
connected analysis again. The Computer Lab gives short setup checks and
practice using the same calculations. 

## Read and Prepare Observations {#sec-ch5-inspect}

[]{#sec-ch5-start}
[]{#ch05-data}

### R, RStudio, a script, and Quarto

**R** performs statistical calculations and draws graphs [@rcore2025].
An **object** is a named value, vector, table or result in its session;
a **function** performs an operation on inputs. **RStudio** provides an
editor and Console for writing and running R commands. Save the opening
commands in a `.R` script: the script records how an answer was obtained,
and running it in order recreates the objects in a fresh session.

**Quarto** combines prose, mathematics, executable code, tables, figures
and references in a `.qmd` source and renders a report [@quarto2026].
Section 6 shows what that connection accomplishes. The Computer Lab's
setup reference covers installation, launch and Help; its final round
uses a prepared report rather than requiring a new document from scratch.

All statistical calculations and plots here use **base R**, including
the stats, graphics, utils and datasets packages supplied with R.
A **package** collects functions and data. Additional report support,
such as knitr and rmarkdown, lets Quarto execute R; its installation
belongs with the reporting example. Chapter 6 introduces a small
additional plotting vocabulary after this base-R route.

Square brackets select positions or values. R positions start at 1;
a comparison produces a logical vector of `TRUE` and `FALSE` values:

```{r}
#| label: ch05-vector-operations
options(digits = 10, width = 65)
length(scores)
scores[2]
scores >= 60
scores[scores >= 60]
mean(x = scores[scores >= 60])
```

The selected scores are 60 and 70, whose mean is 65.0000. Selection
leaves `scores` unchanged. In the last call, `x` names an **argument**,
an input or option supplied to a function. Inside a call, `=` supplies
an argument; `<-` assigns an object. This computing meaning of argument
differs from a study argument. R is case-sensitive, so `scores` and
`Scores` name different objects. The Console marker `[1]` indicates
the first displayed position, rather than being part of the answer.
The `digits` setting changes printed precision, not stored values.

### A data frame is a table of cases and variables

A **data frame** is a table whose columns can hold different types of
values. In Chapter 2’s data-matrix notation, $x_{ij}$ is case $i$’s value
on variable $j$; the arrangement also accommodates text and factors. Its layout should make the observational unit clear. In the next
table, a row represents one student, while the columns give a student
identifier, a programme code, a baseline score, and a follow-up score.
Student B's follow-up score is unavailable. Chapter 2 explained how IDs,
long and wide layouts, and calculation denominators represent the same
observations. Here R carries out those operations. A row can instead record
one person on one occasion or under one condition; count the represented
units as well as the rows.

```{r}
small <- data.frame(
  student = c("A", "B", "C"), program_code = c(1, 2, 2),
  baseline = c(50, 60, 70), followup = c(55, NA, 78))
small
```

The `$` operator selects a named column. Thus `small$followup` is the
vector containing 55, `NA`, and 78. An ordinary `mean()` of that vector
returns `NA`, signalling that a value needed for the calculation is
unavailable. Setting `na.rm = TRUE` calculates the mean of the recorded
values. We count those same values beside the mean:

```{r}
mean(small$followup, na.rm = TRUE)
sum(!is.na(small$followup))
```

`is.na()` asks whether each entry is missing; `!` reverses the logical
answers. `sum()` counts the `TRUE` values as ones, so the second line
gives 2. The available follow-up mean is $(55+78)/2=66.5000$ points.
Treating the missing score as zero would answer a different question
using an invented measurement. Excluding it from the calculation makes
the denominator explicit, without recovering the score or establishing
that the included students represent anyone else.

For change, subtract the baseline score from the follow-up score within
each student:

```{r}
paired_rows <- complete.cases(small[c("baseline", "followup")])
small$change <- small$followup - small$baseline
mean(small$change[paired_rows])
```

`complete.cases()` identifies the students with both measurements.
The changes are 5, `NA`, and 8 points, giving a complete-pair mean of
$(5+8)/2=6.5000$. Subtracting the two separate available means happens to
give $66.5000-60=6.5000$ too. However, the baseline mean uses three students
and the follow-up mean uses two. Their numerical agreement here does
not establish agreement of the student sets. A paired calculation keeps
the same students in both measurements, a principle we return to in
Section 3.4.

### Inspect, select, and derive a variable

The [airquality documentation](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/airquality.html)
describes daily New York measurements from May to September 1973.
A row is a day. `Ozone` is ozone in parts per billion (ppb), `Wind`
wind speed in miles per hour, `Temp` maximum temperature in degrees
Fahrenheit, and `Solar.R` solar radiation in Langleys. `Month` and
`Day` identify dates within that period. Inspect the table before
choosing a calculation:

```{r}
#| label: ch05-air-inspection
air <- airquality
head(air, 2)
dim(air)
c(rows = nrow(air), recorded = sum(!is.na(air$Ozone)),
  missing = sum(is.na(air$Ozone)))
ozone <- air$Ozone[!is.na(air$Ozone)]
```

`head()` shows the first rows; `dim()` returns 153 rows and six columns.
`is.na()` identifies unavailable values and `!` reverses those logical
answers. `sum()` counts `TRUE` as one, giving 116 recorded ozone values
and 37 missing values. The mean of `ozone` will therefore describe 116
days. A percentage of days with measurements would instead use all 153
as its denominator: $100(116/153)=75.8170\%$.

Data-frame indexing uses `[rows, columns]`. Leaving the column position
empty keeps every column from the chosen rows; `&` requires both rules:

```{r}
#| label: ch05-selection-recoding
july <- air[air$Month == 7 & !is.na(air$Ozone), ]
c(n_July = nrow(july), mean_July = mean(july$Ozone))
air$temp_C <- (air$Temp - 32) * 5 / 9
small$program <- factor(small$program_code, levels = c(1, 2),
                        labels = c("Standard", "Practice"))
small[c("student", "program_code", "program")]
```

July contributes 26 recorded days with mean 59.1154 ppb. This restriction
selects a different set of days. Converting temperature to Celsius keeps
the observations and changes their units. The programme codes 1 and 2
name categories: their numerical storage does not make an average code
a meaningful average programme. A **factor** stores these categories
with named levels and an explicit order.

A relationship requires jointly recorded measurements. For ozone and
solar radiation, only the rows that record both values enter the
calculation.

```{r}
#| label: ch05-complete-pairs
#| echo: false
solar_pairs <- air[complete.cases(air[c("Ozone", "Solar.R")]), ]
nrow(solar_pairs)
```

There are 111 complete pairs. A missing value in an unrelated column
does not remove a pair under this rule. Complete records define the
calculation's cases; they do not establish random sampling. Likewise,
`NA` represents an unavailable value, whereas zero is a recorded value.
Chapter 2's zero-grade example shows why a measurement definition and
codebook must support any decision to convert zeros to missing values.

### Recorded cases and a proposed population

The 153 days form a fixed historical record for a particular city and
period. An observed mean can describe its recorded measurements fully.
A claim about the 37 days with unavailable ozone, other years, or
other cities needs an additional target and a justification connecting
the available days to it.

For example, a question about the mean recorded ozone in this dataset
can be answered directly. A question about the mean ozone of all New
York summer days treats these observations as evidence about a wider
population. The arithmetic might begin with the same mean, but the
second use adds assumptions about selection, time, and measurement.
Section 3 makes those assumptions visible when moving from an observed
SD to the modeled uncertainty of a mean.


Chapter 2 also distinguished selection schemes. They describe how units
enter a study; filtering recorded values in R describes which observations
enter a particular calculation. A complete-case filter supplies no random
sampling mechanism. The invented school system gives a compact reminder:

\Needspace{16\baselineskip}

| Scheme | Selection from the school system |
|:--|:--|
| Simple random sampling | Every set of 100 students from the register has the same chance. |
| Systematic sampling | Choose a random start, then every 200th student; the ordering of the list matters. |
| Stratified random sampling | Select a probability sample within every school-type stratum. |
| Cluster sampling | Randomly select schools and include their eligible students. |
| Multistage sampling | Randomly select schools, then students within those schools. |

: Recognize the sampling procedure before calculating. {#tbl-ch05-sampling-schemes tbl-colwidths="[30,70]"}

Probability sampling gives eligible units known nonzero inclusion
probabilities, which need not be equal. An open invitation, convenience
recruitment, or quotas alone does not supply those probabilities. Frame
coverage and nonresponse still matter after probability selection.
Sampling strata concern selection; Chapter 3’s school blocks concern allocation
of conditions. Cluster sampling and assignment of clusters are different
design decisions, even when both produce records grouped within schools.

We can carry out the simple random selection with a register of IDs.
For the invented population of 20,000 students, select 100 without replacement:

```{r}
set.seed(60508)
selected_id <- sample.int(20000, 100, replace = FALSE)
head(selected_id)
c(selected = length(selected_id), distinct = length(unique(selected_id)))
```

`sample.int()` selects from the integers 1 through 20,000. Setting
`replace = FALSE` prevents selecting an ID twice in this sample.
The selected count is 100; each student has inclusion probability
$100/20{,}000=.0050$. A new repetition starts from the full register,
so students may appear again in a different sample. The IDs are identifiers,
not grades: without a population score vector this code cannot reproduce
Chapter 2's illustrative means of 10.6, 11.2, and 10.9. Section 3 changes
to generating measurements from a stated probability model. General
unequal-probability estimation and the SEs for all survey designs extend
beyond the methods developed here.

## Describe Distributions and Groups {#sec-ch5-summaries}

[]{#ch05-descriptions}

### Shape, centre, and spread

A distribution shows how the values are arranged. Before reducing
the 116 ozone measurements to numbers, inspect their shape. A histogram
groups nearby values into intervals and counts the observations in each
interval; a boxplot compresses their location and spread.

```{r}
#| label: fig-ch05-ozone
#| fig-cap: "Recorded ozone values on 116 days, displayed as a histogram and boxplot."
#| fig-alt: "The histogram has a long right tail reaching 168 ppb. The horizontal boxplot places the median near the lower values and marks several high observations beyond its upper whisker."
#| fig-width: 6.2
#| fig-height: 3.4
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 1, .5))
hist(ozone, breaks = seq(0, 180, by = 20), col = "grey85",
     xlab = "Ozone (ppb)", ylab = "Recorded days", main = "")
boxplot(ozone, horizontal = TRUE, xlab = "Ozone (ppb)")
par(oldpar)
```

The `breaks` vector specifies the histogram boundaries. Changing those
boundaries can reveal or conceal detail while leaving the observations
unchanged. Most days have lower values, with a right tail reaching 168
ppb. The boxplot marks the median and a central box between its hinges.
Each whisker reaches the most distant value within 1.5 box lengths of
the corresponding hinge. Points beyond whiskers invite examination;
the display alone does not justify removing them. Boxplot hinges can
differ from the numerical quartile convention used below.
Chapter 2 used the mean and median to describe location, and variance,
SD, and IQR to describe spread. Their calculations preserve different
information. For observations $x_1,\ldots,x_n$,

$$
\bar x=\frac{1}{n}\sum_{i=1}^{n}x_i,\qquad
s^2=\frac{\sum_{i=1}^{n}(x_i-\bar x)^2}{n-1},
$$

$$
s=\sqrt{s^2},\qquad IQR=Q_3-Q_1.
$$

The summation sign $\sum$ means to add the indicated terms. The variance
first measures each value's deviation from the mean, squares those
deviations, adds them, and divides by $n-1$. Squaring prevents positive
and negative deviations from cancelling, but gives variance squared
measurement units. Taking the square root returns the SD to the original
units.

For the three quiz scores, the deviations from 60 are $-10$, 0, and 10.
Their squared deviations sum to 200. Thus $s^2=200/(3-1)=100.0000$ squared
points and $s=10.0000$ points. R lets us inspect the sum as well as the
final summaries:

```{r}
c(mean = mean(scores), ss = sum((scores - mean(scores))^2),
  var = var(scores), sd = sd(scores))
changed_scores <- c(50, 60, 100)
c(mean = mean(changed_scores), median = median(changed_scores),
  sd = sd(changed_scores))
```

Replacing 70 by 100 moves the mean from 60.0000 to 70.0000 and raises the SD
to 26.4575. The median remains 60.0000 because it is still the middle ordered
value. This illustrates the median's resistance to one extreme value;
it does not make either center universally preferable.

Why divide by $n-1$? The deviations around a sample mean sum to zero.
Once $n-1$ are known, the last deviation is determined. Under independent
sampling from a common distribution with finite variance, this denominator
also makes the sample variance unbiased for the process variance.
Section 3.3 shows that property through repeated samples. It is a claim
about the estimator's average across repetitions, not a guarantee that
one sample variance equals the process variance.

For ozone, use the same 116-value vector throughout:

\Needspace{12\baselineskip}

```{r}
c(mean = mean(ozone), median = median(ozone), sd = sd(ozone))
quantile(ozone, probs = c(0, .25, .5, .75, 1), type = 7)
IQR(ozone, type = 7)
```

The mean is 42.1293 ppb, median 31.5000 ppb, and SD 32.9879 ppb. The mean
exceeds the median, consistent with the long right tail. The first and
third quartiles are 18.0000 and 63.2500 ppb, giving IQR 45.2500 ppb; the range
is 1--168 ppb. The SD describes spread around the mean, while the IQR
describes the middle half of the ordered observations.

Small datasets make quartile conventions especially visible. R's
default `type = 7` interpolates between ordered values. For a proportion
$p$, it uses the position $h=1+(n-1)p$. Applied to 1, 3, 3, and 5,

$$
\begin{aligned}
Q_1&=1+.75(3-1)=2.5000,\\
Q_3&=3+.25(5-3)=3.5000.
\end{aligned}
$$

The positions are 1.75 and 3.25, respectively, so each calculation moves
part of the distance between neighboring values. This gives IQR 1.0000,
whereas Chapter 2's medians-of-halves rule gives quartiles 2.0000 and 4.0000,
and IQR 2.0000. R's own `quantile(x, c(.25, .75), type = 7)` returns
2.5000 and 3.5000 for these four values, and `IQR(x, type = 7)` returns
the matching 1.0000.

```{r}
#| label: ch05-quartile-convention
#| echo: false
quantile(c(1, 3, 3, 5), c(.25, .75), type = 7)
IQR(c(1, 3, 3, 5), type = 7)
```

A reproducible report names the convention when it affects the answer.
The different IQRs here come from different rules applied to the same
observations.

Consider two invented groups with scores 58, 60, 62 and 40, 60, 80.
Both balance at 60, but their distances from that center differ.

```{r}
#| label: ch05-equal-means-spreads
#| echo: false
group_1 <- c(58, 60, 62)
group_2 <- c(40, 60, 80)
c(mean(group_1), mean(group_2))
c(sd(group_1), sd(group_2))
```

Their means are both 60.0000, while their SDs are 2.0000 and 20.0000. A report of
equal means would conceal a tenfold difference in SD. This is why a
group comparison keeps counts, centers, and spreads together, and why
a display often adds information that the means alone cannot provide.

### Group summaries and weighted means

For the ozone records, the month codes identify groups of days. Turn
the codes into a factor whose labels appear in calendar order:

```{r}
air_observed <- air[!is.na(air$Ozone), ]
air_observed$month <- factor(air_observed$Month, levels = 5:9,
  labels = c("May", "June", "July", "August", "September"))
cbind(days = table(air_observed$month),
      mean = tapply(ozone, air_observed$month, mean),
      sd = tapply(ozone, air_observed$month, sd))
```

`5:9` generates the integers from 5 through 9. The `levels` argument
states their order, and `labels` gives the corresponding month names.
`table()` counts the observations in each group; `tapply()` applies
a supplied function, here `mean` or `sd`, to the values within each
factor level; `cbind()` places the three results side by side.

The summaries keep each mean beside the number and spread of its
recorded values:

| Month | Recorded days | Mean ozone | SD ozone |
|:--|--:|--:|--:|
| May | 26 | 23.6154 | 22.2244 |
| June | 9 | 29.4444 | 18.2079 |
| July | 26 | 59.1154 | 31.6358 |
| August | 26 | 59.9615 | 39.6812 |
| September | 29 | 31.4483 | 24.1418 |

: Recorded ozone by month; means and SDs are in ppb. {#tbl-ch05-monthly tbl-colwidths="[28,24,24,24]"}

June's mean describes only nine recorded days. Its count matters when
interpreting its position among the other monthly means. The Computer Lab varies the month selection and weighting within this
same analysis.

```{r}
#| label: fig-ch05-months
#| fig-pos: H
#| echo: false
#| fig-cap: "Recorded ozone distributions by month; each box uses that month's available measurements."
#| fig-alt: "July and August have higher median ozone than May, June, and September, with different spreads and only nine observed days in June."
#| fig-width: 6
#| fig-height: 3.4
boxplot(Ozone ~ month, data = air_observed,
        xlab = "Month", ylab = "Ozone (ppb)", col = "grey90")
```

The boxes show the higher centers in July and August alongside the
within-month variation. The observed monthly differences summarize this
record. They do not by themselves isolate an effect of season from
other changes during the period.

The **grand mean** over individual observations weights each group mean
by its group count:

$$
\bar x=\frac{\sum_g n_g\bar x_g}{\sum_g n_g}.
$$

For two observations with mean 20 and eight with mean 50, the grand
mean is $(2\times20+8\times50)/10=44.0000$. The unweighted average
$(20+50)/2=35.0000$ gives each group equal weight instead. Chapter 2's
invented waiting groups make this visible from the observations:
A contains 1 and 3 minutes, while B contains 3, 3, 5, and 9. Their
means are 2.0000 and 5.0000, their SDs are 1.4142 and 2.8284, and the overall mean
is $(2\times2+4\times5)/6=4.0000$ minutes.

\Needspace{12\baselineskip}

```{r}
#| label: ch05-wait-groups
#| echo: false
wait_a <- c(1, 3)
wait_b <- c(3, 3, 5, 9)
c(mean_A = mean(wait_a), mean_B = mean(wait_b),
  sd_A = sd(wait_a), sd_B = sd(wait_b))
mean(c(wait_a, wait_b))
```

Combining the two groups gives 4.0000 minutes over the six waits. The
same count-weighting rule applied to the five monthly ozone means
returns 42.1293 ppb. Giving each month equal weight instead would
describe the average month, rather than the average recorded day.
For June and July, the two recorded-day counts are 9 and 26:

```{r}
#| label: ch05-month-weighting
month_n <- tapply(air$Ozone, air$Month, function(x) sum(!is.na(x)))
month_mean <- tapply(air$Ozone, air$Month, mean, na.rm = TRUE)
c(recorded_day_mean = weighted.mean(month_mean[c("6", "7")],
                                     month_n[c("6", "7")]),
  equal_month_mean = mean(month_mean[c("6", "7")]))
```

The recorded-day mean is 51.4857 ppb; the equal-month mean is 44.2799
ppb. Neither weighting rule can be read from the two means alone:
the question determines whether days or months receive equal weight.

### Counts, proportions, and a binary indicator

The [mtcars documentation](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html)
describes 32 car-model records from the 1974 *Motor Trend* magazine,
covering 1973--1974 models. Each row is a car model. The variable `mpg`
gives fuel economy in miles per US gallon, while `am` codes automatic
transmission as 0 and manual transmission as 1.

Use 20 mpg as a teaching threshold, and label both categories before
forming their two-way table:

```{r}
cars <- mtcars
cars$transmission <- factor(cars$am, levels = c(0, 1),
  labels = c("Automatic", "Manual"))
cars$mpg20 <- factor(cars$mpg >= 20, levels = c(FALSE, TRUE),
  labels = c("Below 20", "At least 20"))
tab <- table(cars$transmission, cars$mpg20)
tab
100 * prop.table(tab, margin = 1)
100 * prop.table(tab, margin = 2)
```

The table contains 15 automatic models below the threshold and 4 at
or above it, compared with 3 and 10 manual models. `prop.table()`
turns counts into proportions. With `margin = 1`, each row supplies
its own denominator; multiplying by 100 expresses percentages.
Thus $100(4/19)=21.0526\%$ of automatic models and
$100(10/13)=76.9231\%$ of manual models meet the threshold, a
manual-minus-automatic difference of 55.8704 percentage points.

With `margin = 2`, the denominator is instead a column total.
Among the 14 threshold-meeting models, $100(10/14)=71.4286\%$ are
manual. “Meeting the threshold among manual models” and “manual
among threshold-meeting models” reverse the condition and answer
different questions. Naming the denominator keeps the difference clear.

These percentages describe an association among the 32 recorded
models. Transmission groups also differ in weight and engine
characteristics. Chapter 6 applies proportion and count inference to a declared count model;
it cannot by itself identify what changing a transmission would do
to fuel economy.

A binary indicator has an additional useful identity: its mean is the
proportion coded 1. Thus `mean(cars$mpg >= 20)` is 14/32=.4375.
Chapter 2's indicator idea also connects a binary variable to a
point-biserial correlation in Section 5.3. Counts and proportions here
remain descriptions of the recorded car models.

### Transformations and the Normal model

A linear transformation changes location and units predictably. If
$y_i=a+bx_i$, its mean is $a+b\bar x$, variance $b^2s^2$ and SD
$|b|s$. The Celsius conversion has $a=-32(5/9)$ and $b=5/9$.

```{r}
#| label: ch05-transformation
#| echo: false
c(mean_F = mean(air$Temp), SD_F = sd(air$Temp),
  mean_C = mean(air$temp_C), SD_C = sd(air$temp_C))
```

The mean is 77.8824 degrees Fahrenheit or 25.4902 degrees Celsius;
the SDs are 9.4653 and 5.2585. Subtracting a constant affects the mean
but leaves the spread unchanged. Multiplying by $5/9$ rescales both.

A standardized observation measures distance from the mean in SD units:

$$
z_i=\frac{x_i-\bar x}{s},\qquad
z_{100\text{ ppb}}=\frac{100-42.1293}{32.9879}=1.7543.
$$

`scale()` performs this transformation on a vector. Standardizing
changes location and scale, while preserving distributional shape:
right-skewed ozone remains right-skewed. This individual z score uses
the observations' SD. Section 4's mean-test statistic instead divides
by the mean's SE.

R's Normal functions let us compare the observed proportions with a
model. `pnorm()` gives a cumulative probability; `lower.tail = FALSE`
requests the probability above the supplied value:

```{r}
#| label: ch05-normal-empirical
z_100 <- (100 - mean(ozone)) / sd(ozone)
c(z = z_100, model_above = pnorm(z_100, lower.tail = FALSE),
  observed_above = mean(ozone >= 100),
  model_within_1SD = pnorm(1) - pnorm(-1),
  observed_within_1SD = mean(abs(scale(ozone)) <= 1))
```

The upper probability is .0397 under a Normal model with the ozone
mean and SD. The observed fraction is $7/116=.0603$. The proportions
within one SD are .6827 for a Normal model and .7241 for these data.
The model calculation evaluates its supplied assumptions; it does not
establish that the distribution describes ozone well. The familiar
68–95–99.7 rule belongs to the Normal model, not to every standardized
dataset.

For a separate stipulated score distribution $X\sim N(100,15^2)$,
subtract cumulative areas to obtain an interval probability:

```{r}
#| label: ch05-normal-score-areas
c(below_100 = pnorm(100, 100, 15),
  above_140 = pnorm(140, 100, 15, lower.tail = FALSE),
  between_100_120 = pnorm(120, 100, 15) - pnorm(100, 100, 15))
qnorm(c(.025, .975), mean = 100, sd = 15)
```

The probabilities are .5000, .0038 and .4088. `qnorm()` reverses the
question: it returns a value at a stated cumulative probability. Here
the central 95% endpoints are 70.6005 and 129.3995. They bound individual
scores under the stipulated model, rather than forming a confidence
interval for an unknown mean.

```{r}
#| label: fig-ch05-normal-area
#| echo: false
#| fig-cap: "Probability is an area under a stipulated Normal density."
#| fig-alt: "The score range from 100 to 120 is shaded under a Normal curve with mean 100 and SD 15. Its area is .4088."
#| fig-width: 6.2
#| fig-height: 3.3
curve(dnorm(x, mean = 100, sd = 15), 40, 160,
      xlab = "Score", ylab = "Density")
shade_x <- seq(100, 120, length.out = 200)
polygon(c(100, shade_x, 120),
        c(0, dnorm(shade_x, 100, 15), 0), col = "grey80")
```

The curve's height is a density. Probability belongs to an area,
such as the shaded region, rather than to the height at one exact score.

### A compact distribution catalogue in R {#ch05-distributions}

Chapter 2 connected variable forms to probability models. R uses a
common naming pattern across these families. A `d` function gives a
mass at a discrete value or a continuous density height; `p` gives
the cumulative probability $P(X\le x)$; `q` gives a cumulative quantile;
and `r` generates observations from the stated model. The parameters
identify which member of a family is being used.

\Needspace{24\baselineskip}

| Model | Object and parameters | Base-R family |
|:--|:--|:--|
| Bernoulli | One 0/1 response, probability $p$ of 1 | `binom`, with `size=1` |
| Binomial | Success count in $m$ independent trials with common $p$ | `dbinom`, `pbinom`, `qbinom`, `rbinom` |
| Categorical | One label among categories with stated probabilities | `sample()` with `prob`; a table of masses |
| Poisson | Nonnegative count with model mean $\lambda$ | `dpois`, `ppois`, `qpois`, `rpois` |
| Normal | Continuous value with mean $\mu$ and SD $\sigma$ | `dnorm`, `pnorm`, `qnorm`, `rnorm` |
| Lognormal | Positive value whose log is Normal; parameters on the log scale | `dlnorm`, `plnorm`, `qlnorm`, `rlnorm` |
| Logistic | Continuous symmetric value with location $\mu_L$ and positive scale $s_L$ | `dlogis`, `plogis`, `qlogis`, `rlogis` |
| Student t | Symmetric reference distribution with specified df | `dt`, `pt`, `qt`, `rt` |

: Distribution functions respect the model's support and parameters. {#tbl-ch05-models tbl-colwidths="[16,49,35]"}

For example, $X\sim\operatorname{Binomial}(20,.3)$ counts successes
in 20 independent trials with success probability .3. Compare the
probability of exactly four with the probability of at most four:

```{r}
#| label: ch05-binomial-functions
c(mass_4 = dbinom(4, size = 20, prob = .3),
  at_most_4 = pbinom(4, size = 20, prob = .3),
  quantile_95 = qbinom(.95, size = 20, prob = .3))
set.seed(20517)
rbinom(6, size = 20, prob = .3)
```

The mass is .1304 and the cumulative probability .2375. The .95
quantile is 9, the smallest integer whose cumulative probability reaches
.95; discreteness means the attained probability need not be exactly
.95. The six draws are 6, 8, 6, 3, 5 and 6. Each draw is a count
from one set of 20 trials. The first argument, 6, is the number of
counts generated, whereas `size=20` is the number of trials per count.
A seed makes these draws reproducible without changing model probabilities.

The same vocabulary works across other supports:

\Needspace{19\baselineskip}

```{r}
#| label: ch05-other-model-functions
c(Bernoulli_1 = dbinom(1, size = 1, prob = .3),
  Poisson_3 = dpois(3, lambda = 2),
  lognormal_below_2 = plnorm(2, meanlog = 0, sdlog = .8),
  logistic_quantile_95 = qlogis(.95, location = 0, scale = 1))
labels <- c("A", "B", "C")
probabilities <- c(.2, .3, .5)
set.seed(20517)
sample(labels, size = 6, replace = TRUE, prob = probabilities)
```

The four numerical answers are .3000, .1804, .8069 and 2.9444;
the category draws are B, B, C, C, C and B. Categorical probabilities
refer to labels. A cumulative probability requires an explicit ordering
of those labels; arbitrary category numbers need not have one.
Lognormal `meanlog` and `sdlog` describe the logarithm, not the mean
and SD of the positive observations. The logistic distribution is
continuous and differs from both lognormal and binary outcomes.
Its cumulative function later supplies a probability curve in Chapter 11.

```{r}
#| label: fig-ch05-models
#| fig-pos: H
#| echo: false
#| fig-cap: "Examples of discrete masses and continuous densities under specified models."
#| fig-alt: "Four panels show Bernoulli, binomial, categorical and Poisson masses. Four others show Normal, lognormal, logistic and t densities on their respective supports."
#| fig-width: 7
#| fig-height: 5.5
oldpar <- par(mfrow = c(2, 4), mar = c(3.5, 3.5, 2, .5),
              mgp = c(2, .6, 0))
plot(0:1, dbinom(0:1, 1, .3), type = "h", lwd = 3,
     ylim = c(0, 1), xaxt = "n", xlab = "Outcome",
     ylab = "Mass", main = "Bernoulli")
axis(1, at = 0:1)
plot(0:20, dbinom(0:20, 20, .3), type = "h",
     xlab = "Count", ylab = "Mass", main = "Binomial")
barplot(probabilities, names.arg = labels,
        ylab = "Mass", main = "Categorical")
plot(0:9, dpois(0:9, 2), type = "h",
     xlab = "Count", ylab = "Mass", main = "Poisson")
curve(dnorm(x), -4, 4, xlab = "Value", ylab = "Density",
      main = "Normal")
curve(dlnorm(x, 0, .8), .01, 6, xlab = "Value",
      ylab = "Density", main = "Lognormal")
curve(dlogis(x), -6, 6, xlab = "Value", ylab = "Density",
      main = "Logistic")
curve(dt(x, df = 5), -4, 4, xlab = "Value",
      ylab = "Density", main = "t, df = 5")
par(oldpar)
```

The upper panels assign probability to discrete outcomes, so their
masses sum to one over the complete support. The lower panels use area:
their densities integrate to one, although any one height need not be
less than one. The t curve has heavier tails than the Normal at small
df and approaches it as df grows. A family and its parameters together
determine these shapes; a function call evaluates that specified model.
The Computer Lab provides the figure construction and one-parameter
variations as reference examples.

## Sampling Variation and Standard Errors {#sec-ch5-sampling}

[]{#ch05-sampling}

### A target, an estimator, and a realized estimate

Chapter 2's invented school system contained 20,000 students. Repeated
simple random samples of 100 could give different means while the
population mean stayed fixed. Section 1.4 selected IDs without replacement
from that finite register. We now change the generating process: instead
of selecting IDs from a fixed population, generate independent observations
from a probability distribution whose parameters we choose.

For a separate numerical illustration, let the population distribution
be Normal with mean 50 and SD 10. Generate 30 independent draws:

```{r}
set.seed(60506)
one_sample <- rnorm(30, mean = 50, sd = 10)
c(model_mean = 50, model_sd = 10,
  sample_mean = mean(one_sample), sample_sd = sd(one_sample))
```

`rnorm()` takes the number of observations, the model mean, and the
model SD as arguments. The resulting vector is one sample from that
distribution. Its mean is 49.5691 and its SD 7.0051, neither of which has
to equal the corresponding model value.

\Needspace{10\baselineskip}

| Quantity | Kind | Value |
|:--|:--|--:|
| Model mean $\mu$ | Parameter | 50 |
| Model SD $\sigma$ | Parameter | 10 |
| Sample mean $\bar x$ | Statistic | 49.5691 |
| Sample SD $s$ | Statistic | 7.0051 |

: One sample of 30 draws from a declared Normal model (seed 60506). {#tbl-ch05-parameter-statistic tbl-colwidths="[36,32,32]"}

The **parameters** are fixed features of the model; the **statistics**
are calculated from the generated observations. Running `rnorm()` again
without resetting the seed gives another sample and generally different
statistics. Resetting the seed recreates the same draws. A seed controls
repeatability of the numerical experiment; it supplies no evidence that
the Normal model describes a real study.

A simulation gives us a known target, but a study usually has a target
we cannot observe directly. Recall Chapter 3's invented special-practice
study. It assigned 120 participants, 60 to Special Practice and 60 to
Usual Practice, then measured delayed-assessment scores. The group means
were 27.1 and 24.0 points, with SDs 6.0 and 6.4.

The **estimand** is the mean effect of assignment to Special rather than
Usual Practice among these 120 participants. Write $Y_i(1)$ for
participant $i$'s potential score under Special assignment and $Y_i(0)$
for the score under Usual assignment. The target and estimator are

$$
\Delta\mu=\frac{1}{120}\sum_{i=1}^{120}\{Y_i(1)-Y_i(0)\},
$$

$$
\widehat{\Delta\mu}=\bar y_1-\bar y_0,\qquad\text{observed estimate}=27.1-24.0=3.1000\text{ points}.
$$

Each participant reveals only the outcome under the assigned condition.
We therefore cannot calculate every individual effect and average them.
The **estimator**, the difference in assigned-group means, uses the
observed outcomes. **Estimation** applies that rule to the data and
checks its inputs. The **estimate** is the resulting 3.1000 points.

We can reproduce the arithmetic directly from the supplied summaries:

```{r}
n_special <- 60
n_usual <- 60
mean_special <- 27.1
mean_usual <- 24.0
sd_special <- 6.0
sd_usual <- 6.4
estimate <- mean_special - mean_usual
estimate
```

The repetition behind uncertainty reassigns the same 120 participants,
with exactly 60 in each condition. The participants and their potential
outcomes stay fixed; the assigned groups and observed outcomes can vary.
This differs from repeatedly sampling new people. Section 4 applies
the familiar Welch procedure as an approximate working calibration,
retaining Chapter 3’s assignment interpretation and conditions.

| Object | Working definition | In the special-practice study | In R |
|:--|:--|:--|:--|
| Estimand | The precisely specified quantity the question seeks | Mean delayed-score difference caused by Special rather than Usual assignment among the 120 participants | Not an object; stated in words before any code |
| Estimator | The rule applied in each admissible repetition | Special-group mean minus Usual-group mean | The expression `mean_special - mean_usual` |
| Estimate | The number the estimator produced | $27.1-24.0=3.1000$ points | The value of `estimate` |
| Statistic | A numerical function of observed data | A group mean, the 3.1000-point estimate | Any computed summary |
| Parameter | A fixed feature named by the estimand under the chosen interpretation | The participant-specific mean assignment contrast | Unknown here; Section 3.3 changes the target to a known waiting-time variance |
| Standard error | The modeled spread of the estimator across the declared repetitions | 1.1325 points | `sqrt(sd_special^2 / n_special + sd_usual^2 / n_usual)` |

: Inferential objects with their R counterparts. {#tbl-ch05-inferential-objects tbl-colwidths="[13,27,34,26]"}

The table distinguishes the scientific target from its computational
representation. Storing a result in an object named `estimate` does
not define its estimand; that definition comes from the question,
participant set, comparison, and outcome.

### The first repeated-sample calculation

Return to Chapter 2's waiting-time model: independent exponential waits
with mean and SD both 10 minutes. The exponential distribution permits
many short waits and occasional long ones. R's `rexp()` uses a
**rate**, the reciprocal of the mean, so `rate = 1 / 10` gives a
ten-minute mean.

Begin with one sample of five waits and its mean:

```{r}
set.seed(60203)
one_waiting_sample <- rexp(5, rate = 1 / 10)
one_waiting_sample
mean(one_waiting_sample)
```

The first printed vector contains individual waiting times. Its mean
compresses the entire sample into one statistic. To see how that
statistic varies, repeat the draw-and-average operation:

```{r}
set.seed(60203)
B <- 5000
n_middle <- 30
means_5 <- replicate(B, mean(rexp(5, rate = 1 / 10)))
mean(means_5)
sd(means_5)
```

`replicate()` evaluates its second argument repeatedly and collects the
results. Here it returns a vector of 5,000 sample means, not 5,000
individual waits. Resetting the seed before the
block lets the first repeated sample reproduce the one-sample example.

The mean of `means_5` is about 10.0141 minutes and its SD about
4.5395 minutes. Next repeat the same operation for sample sizes 30 and
100. Keep the draws in the same order as Chapter 2 to reproduce its
numerical results:

```{r}
means_middle <- replicate(
  B, mean(rexp(n_middle, rate = 1 / 10)))
means_100 <- replicate(B, mean(rexp(100, rate = 1 / 10)))
sample_sizes <- c(5, n_middle, 100)
c(mean(means_middle), sd(means_middle))
c(mean(means_100), sd(means_100))
10 / sqrt(sample_sizes)
```

Each histogram below collects 5,000 sample means from one sample size.
The display and table use these three vectors; no further random draws
are needed. The Computer Lab supplies the plotting construction.

```{r}
#| label: fig-ch05-sampling-means
#| echo: false
#| fig-cap: "Sampling distributions from 5,000 means at each sample size; the individual waiting-time model is unchanged."
#| fig-alt: "Three histograms share both axes. Means from samples of five are broad and right-skewed; means from thirty and one hundred observations concentrate more tightly around ten and become more nearly symmetric."
#| fig-width: 6.5
#| fig-height: 3.3
oldpar <- par(mfrow = c(1, 3), mar = c(4, 3, 2, .5))
means_list <- list(means_5, means_middle, means_100)
breaks_wait <- seq(0, ceiling(max(unlist(means_list))), by = .5)
for (j in seq_along(means_list)) {
  hist(means_list[[j]], breaks = breaks_wait, probability = TRUE,
       xlim = range(breaks_wait), ylim = c(0, .45), col = "grey85",
       xlab = "Sample mean (minutes)", main = sample_sizes[j])
  abline(v = 10, lty = 2)
}
par(oldpar)
```

One histogram bar groups sample means in its interval. The vertical
density scale divides its count by $B$ times the bin width, so the
bar’s area is the fraction of repeated means in that interval. In the first
panel, each mean comes from five waits; in the last, each comes from
100. The individual-wait model is unchanged. What changes is the
distribution of the mean under each sample size.

| Sample size | Mean of sample means | Empirical SE | Model SE |
|--:|--:|--:|--:|
| 5 | 10.0141 | 4.5395 | 4.4721 |
| 30 | 9.9626 | 1.7878 | 1.8257 |
| 100 | 10.0027 | 1.0076 | 1.0000 |

: Centre and spread of the simulated sample means, in minutes. {#tbl-ch05-sampling-summary}

The means of sample means are 10.0141, 9.9626, and 10.0027 minutes, close to the fixed model
mean. Their empirical SDs are 4.5395, 1.7878, and 1.0076 minutes.
The SE label connects these empirical spreads to the model calculation.

A statistic's **standard error** is its SD across the stated
repetitions. The SD of the 5,000 generated means approximates that
sampling-distribution SD, so it is an *empirical SE*. For independent
observations from a common distribution with SD $\sigma$, the mean's
model SE is

$$
SE(\bar Y)=\frac{\sigma}{\sqrt n}.
$$

For these waits,

$$
\frac{10}{\sqrt5}=4.4721,\qquad
\frac{10}{\sqrt{30}}=1.8257,\qquad
\frac{10}{\sqrt{100}}=1.0000\text{ minutes}.
$$

The empirical SEs in @tbl-ch05-sampling-summary approximate these values.
With finitely many repetitions, they need not agree exactly.

Keep three quantities distinct. The 10-minute SD describes individual
waits. The SE describes how sample means vary at a stated sample size.
A particular sample's **error**, such as $11.3-\nobreak10=\nobreak1.3000$ minutes for an
illustrative mean of 11.3, is one realized departure from the target.
An SE summarizes a distribution of possible estimates; it does not tell
us the unknown error of a particular estimate.

Increasing $n$ from 5 to 100 multiplies sample size by 20 and divides
the SE by $\sqrt{20}\approx4.4721$. To halve the SE from 1.0000 to
0.5000 minute under this model requires four times as many observations:

$$
\frac{10}{\sqrt{400}}=0.5000\text{ minute}.
$$

Increasing the number of simulation repetitions has a different effect:
it shows the same sampling distribution more steadily. It does not
change the study sample size or its model SE. To explore the distinction,
change `n_middle` or `B` and rerun the simulation from its
seed. Change one setting at a time so the cause of the changed display
remains clear.

### Bias and precision

An estimator's **expectation** is its probability-weighted average
across the declared repetitions. Its **bias** is that expectation
minus the target. Chapter 3 illustrated this with an estimator averaging
3.4 when the target was 3.0, a bias of 0.4000. **Precision** concerns how
tightly its possible values cluster. An estimator can be centered
correctly and still vary widely from one sample to another.

To compare the rules, change the target to the variance of the
waiting-time process: 100 squared minutes. For each of 5,000 independent
samples of five waits, compute one variance with divisor $n-1$ and one
with divisor $n$. The Computer Lab reference supplies the complete simulation; its
results show the distinction here:

| Variance rule | Mean over 5,000 samples | Expectation under the model |
|:--|--:|--:|
| Divide by n - 1 | 101.2241 | 100.0000 |
| Divide by n | 80.9793 | 80.0000 |

: Variance estimates from independent exponential samples of five waits. {#tbl-ch05-bias-evidence tbl-colwidths="[34,33,33]"}

Their theoretical expectations are

$$
\operatorname{Bias}(\widehat\theta)
=E(\widehat\theta)-\theta,\qquad
E(s^2)=\sigma^2,
$$

$$
E\left\{\frac{1}{n}\sum_{i=1}^{n}(X_i-\bar X)^2\right\}
=\frac{n-1}{n}\sigma^2
=\frac45(100)=80.0000.
$$

The $n-1$ rule is unbiased under the independent common-distribution
model. The $n$ rule has downward bias $\sigma^2/n=20.0000$ here.
The finite-simulation average 101.2241 can differ from 100 even for an
unbiased rule: 5,000 samples approximate the expectation rather than
enumerating every possible sample.

Both rules can give an estimate far from 100 in a single sample. The
$n$ estimate is always four-fifths of the $n-1$ estimate here, so its
spread across repetitions is also smaller, even though its center is
wrong. This gives a concrete reason to distinguish precision from bias.
An apparently stable estimate can systematically miss the target.

### SE follows the unit and the comparison {#sec-ch5-reshape}

An SE follows the observational units that can vary independently.
For repeated measurements, alignment comes before subtraction. The
[sleep documentation](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/sleep.html)
describes extra hours of sleep relative to control under two drugs for
ten people. `ID` identifies a person, `group` a drug condition and
`extra` the extra hours. Twenty measurement rows represent ten people;
these are two drug measurements relative to control, not before-and-after
scores. Negative extra sleep means less sleep than under control.

```{r}
#| label: ch05-align-sleep
all(table(sleep$ID, sleep$group) == 1)
sleep_wide <- reshape(sleep, idvar = "ID", timevar = "group",
                       direction = "wide")
sleep_wide <- sleep_wide[order(sleep_wide$ID), ]
sleep_d <- sleep_wide$extra.2 - sleep_wide$extra.1
head(sleep_wide, 3)
c(n = length(sleep_d), mean = mean(sleep_d), SD = sd(sleep_d))
```

The `all()` check verifies one record per ID and condition. `reshape()` uses
the ID to align measurements, creates `extra.1` and `extra.2`, and
places each person on one row. Duplicate ID–condition keys need review
before reshaping an unfamiliar file. The subtraction is **drug 2 minus
drug 1**. For the first person it gives $1.9-0.7=1.2000$ hours.
The ten differences have mean 1.5800 and SD 1.2300 hours. With the same
pairs in each calculation, their mean equals the difference of means:
$2.3300-0.7500=1.5800$. Their variance also retains the pairing:

$$
s_d^2=s_2^2+s_1^2-2s_{12}.
$$

Here $s_{12}$ is the sample covariance of the two measurements.
People with high values in both conditions contribute shared variation
that subtraction removes. Verify both identities on the same people:

```{r}
#| label: ch05-paired-identities
mean(sleep_wide$extra.2) - mean(sleep_wide$extra.1)
c(variance_difference = var(sleep_d),
  identity = var(sleep_wide$extra.2) + var(sleep_wide$extra.1) -
    2 * cov(sleep_wide$extra.2, sleep_wide$extra.1))
```

The variance is 1.5129 squared hours. The covariance is positive,
2.8483, so it reduces the difference variance relative to the sum of
the two variances. Separately available means and variances could use
different people when a condition is missing; the identities require
the same pairs. A high correlation between conditions would describe
shared variation, not equality of their means.


For independent observations from the same finite-variance distribution,
the sample SD estimates the unknown process SD. The estimated SE is

$$
\widehat{SE}(\bar Y)=\frac{s}{\sqrt n}.
$$

For the 116 recorded ozone values, the formula gives
$32.9879/\sqrt{116}=3.0628$ ppb.

```{r}
#| label: ch05-ozone-se
#| echo: false
sd(ozone) / sqrt(length(ozone))
```

This is an SE **under a working assumption of independent observations
from a common process**. The historical daily record does not establish
that assumption; nearby days could be related, and the process could
change over the period. The calculation demonstrates what the formula
returns. Its inferential use needs a justification for the model and
target, in addition to the recorded values.

For two independent groups, each mean contributes uncertainty, and their
variances add. The special-practice summaries give the same familiar
numerical calculation:

$$
\begin{aligned}
\widehat{SE}(\bar y_1-\bar y_0)
&=\sqrt{\frac{6.0^2}{60}+\frac{6.4^2}{60}}\\
&=\sqrt{0.6000+0.682666\ldots}
=1.1325\text{ points}.
\end{aligned}
$$

```{r}
standard_error <- sqrt(
  sd_special^2 / n_special + sd_usual^2 / n_usual)
standard_error
```

Its interpretation in this fixed-participant assignment study is
different from independent sampling. Exactly 60 of the 120 participants
are assigned to each group, so the two assigned-group means are dependent.
The exact assignment variance includes a subtraction for variation in
individual assignment effects, which cannot be recovered from the
observed outcomes. The squared SE above omits that term and estimates
an upper bound on assignment variance *on average*. It need not exceed
the true variance in every realized study. Chapter 3 develops this
distinction and the approximate Welch calibration used in Section 4.

Pairing changes which SD belongs in the numerator. If independent
people each supply two measurements, calculate the within-person
differences and use their SD:

$$
\widehat{SE}(\bar d)=\frac{s_d}{\sqrt n}
=\frac{1.2300}{\sqrt{10}}=0.3890\text{ hours}.
$$

```{r}
#| label: ch05-paired-se
#| echo: false
sd(sleep_d) / sqrt(length(sleep_d))
sqrt((var(sleep_wide$extra.2) + var(sleep_wide$extra.1)) / 10)
```

The second value, 0.8491 hours, imposes an independence structure that is
inappropriate for these paired measurements. It omits their
positive within-person covariance. The number of people has stayed at ten;
it is the variance calculation, not the count, that changed.

Clustered assignment requires the same attention to the units that can
vary independently. Chapter 3’s Appendix B, “Paired schools: a comparison
of class means,” gives the complete school-blocked feedback example.
It has 480 students in 24 classes within 12 schools, with one class per
format assigned within each school. It differs from Chapter 1’s
assignment of twelve of twenty-four classes across the whole class pool.
Here each value is one explanatory-minus-usual class-mean contrast, and
the twelve contrasts are $-2$, 0, 1, 1, 2, 2, 3, 3, 4, 4, 5 and 7:

```{r}
#| label: ch05-school-contrasts
#| echo: false
contrasts_12 <- c(-2, 0, 1, 1, 2, 2, 3, 3, 4, 4, 5, 7)
c(mean = mean(contrasts_12), sd = sd(contrasts_12),
  se = sd(contrasts_12) / sqrt(12))
```

Their mean is 2.5000 points, SD 2.3932 points, and estimated SE
$s_d/\sqrt{12}\approx.6908$ points, using the unrounded SD. This calculation follows the
school-block contrasts. Counting 480 student identifiers would not
establish 480 independent assignments.

Sampling without replacement also changes uncertainty. If $n$ members
are selected by simple random sampling from a finite population of
size $N$, whose variance $\sigma^2$ uses denominator $N$, then

$$
SE_{\text{finite}}
=\frac{\sigma}{\sqrt n}\sqrt{\frac{N-n}{N-1}}.
$$

The correction reflects information gained by observing a larger fraction
of the fixed population; observing everyone leaves no sampling uncertainty
about its mean. Chapter 3's Appendix A explains the variance conventions
and the estimated-SE calculation. With sample variance $s^2$ using
divisor $n-1$, the estimated SE is $(s/\sqrt n)\sqrt{1-n/N}$.
The formula applies specifically to
simple random sampling without replacement, rather than to every method
that selects observations from a finite population. Across these settings, effective
information depends on the estimand, dependence, weighting, and design.
Row counts alone cannot determine an uncertainty calculation.

### Concentration and sampling shape

Two results help explain the changing distributions of sample means.
The **law of large numbers** concerns concentration: under its conditions,
larger samples make the mean more likely to be close to the population
mean. The **central limit theorem** concerns shape: under conditions such
as independent, identically distributed observations with finite variance,
the centered and appropriately scaled sample mean approaches a Normal
distribution.

A running mean displays the first idea along one sequence of draws.
After draw $k$, average the first $k$ observations:

```{r}
#| label: fig-ch05-running-mean
#| echo: false
#| fig-cap: "One running mean from independent exponential waits with model mean ten minutes."
#| fig-alt: "The running mean fluctuates irregularly around ten; it is 9.7207 after 400 observations. Adding a draw can move it closer to or farther from the target."
#| fig-width: 6.2
#| fig-height: 3.3
set.seed(60204)
running_waits <- rexp(400, rate = 1 / 10)
running_mean <- cumsum(running_waits) / seq_along(running_waits)
plot(running_mean, type = "l", xlab = "Number of observations",
     ylab = "Running mean (minutes)", ylim = range(c(running_mean, 10)))
abline(h = 10, lty = 2)
```

The running means are 9.0830 after
10 draws, 7.9852 after 50, 8.8373 after 100, and 9.7207 after 400.
This path is irregular: adding observations can temporarily move the
mean farther from 10. Concentration is a probabilistic statement as
sample size grows, not a promise of monotonic improvement along one
sequence or a guarantee at a particular sample size.

The waiting-time histograms in @fig-ch05-sampling-means also change shape.
Means of larger samples become more nearly symmetric even though
individual waits retain their exponential distribution. Concentration
describes closeness to the target; the CLT describes the sampling shape
after centering and scaling. Neither result makes individual observations
Normal. Strong dependence, extreme tails, or a dominating observation
can make a familiar approximation poor. There is no universal sample-size
cutoff. Chapter 3’s Appendix A gives the formal statements and conditions.
The Computer Lab reference applies the reasoning to a second skewed
process and supplies the running-mean construction code.

## Intervals and Tests for Means {#ch05-mean-inference}

### Estimate, SE, critical value, and interval {#ch05-z-t}

An estimate answers a numerical question; an interval describes uncertainty
about its target under a stated procedure. Before asking R for either,
name the outcome and its units, the included observations, the parameter,
and the repetition that makes an SE meaningful. The model and design
determine the calculation. A function's default cannot supply them.

The common construction from Chapter 3 is

$$
\text{interval}=\widehat\theta\pm c^*
\widehat{\operatorname{SE}}(\widehat\theta),\qquad
T=\frac{\widehat\theta-\theta_0}
{\widehat{\operatorname{SE}}(\widehat\theta)}.
$$

The estimate and SE have outcome units. Multiplying the SE by the
**critical value** $c^*$ gives a margin in those same units. The test
statistic instead expresses departure from a null value $\theta_0$ in
standard-error units. The confidence level, tail convention and reference
distribution determine the critical value and the test's p-value.

For a mean of independent Normal observations whose population SD
$\sigma$ is stipulated independently, the SE is $\sigma/\sqrt n$ and
the standardized statistic has the standard Normal reference. When the
SD is estimated from those observations, use $s/\sqrt n$ and a t
reference with $n-1$ degrees of freedom. Estimating the scale introduces
uncertainty, reflected in the heavier tails of t [@student1908].

The following quantiles leave .025 in the upper tail, as required for a
central 95% interval. At low df the t multiplier is noticeably larger;
it approaches the Normal multiplier as df increases.

```{r}
#| label: ch05-mean-critical
#| echo: true
c(z_975 = qnorm(.975), t_df5 = qt(.975, 5),
  t_df9 = qt(.975, 9), t_df30 = qt(.975, 30))
```

These are 1.9600, 2.5706, 2.2622 and 2.0423. A large sample does not
make an estimated SD a known population SD; it makes the two reference
distributions more similar.

The 32 recorded car models have mean fuel economy 20.0906 mpg. We can
use the same observations to compare two explicitly different working
assumptions. First, **suppose independently that the process SD is known
to be 5 mpg**. This teaching value is not an estimate from `mtcars`.
The null process mean is 20 mpg. `abs()` uses the magnitude of the
observed z statistic, and `lower.tail = FALSE` requests the upper-tail
probability. Doubling it accounts for departures in either direction.

\Needspace{18\baselineskip}

```{r}
#| label: ch05-one-z
#| echo: true
mpg <- cars$mpg
n_mpg <- length(mpg)
mean_mpg <- mean(mpg)
sigma_known <- 5
SE_z <- sigma_known / sqrt(n_mpg)
z_mpg <- (mean_mpg - 20) / SE_z
c(estimate = mean_mpg, SE = SE_z, z = z_mpg,
  p = 2 * pnorm(abs(z_mpg), lower.tail = FALSE))
mean_mpg + c(-1, 1) * qnorm(.975) * SE_z
```

The vector `c(-1, 1)` subtracts and adds the same margin, returning
the lower and upper endpoint. Now estimate the SD from the records:

```{r}
#| label: ch05-one-t
#| echo: true
one_t <- t.test(mpg, mu = 20)
one_t
c(SD = sd(mpg), SE = one_t$stderr,
  critical = qt(.975, n_mpg - 1))
```

`t.test()` defaults to a two-sided alternative and 95% confidence level.
Printing `one_t` gives the method, statistic, df, p-value, stated
alternative, interval and estimated mean. The returned object also
stores those components: `$estimate`, `$stderr`, `$statistic`,
`$parameter`, `$p.value` and `$conf.int`. Here `$parameter` is the
reference df, not the estimated process mean.

| Procedure | SE | Statistic | df | Two-sided p | 95% interval |
|:--|--:|--:|--:|--:|:--|
| Known-SD z, $\sigma=5$ | 0.8839 | 0.1025 | -- | .9183 | [18.3582, 21.8230] |
| Estimated-SD t | 1.0654 | 0.0851 | 31 | .9328 | [17.9177, 22.2636] |

: One mean, two explicitly different uncertainty assumptions. {#tbl-ch05-one-mean}

The t calculation uses observed SD 6.0269 and critical value 2.0395.
Both its SE and multiplier exceed those of the stipulated known-SD
calculation, so its interval is wider. These calculations illustrate
working process models. The historical collection of automobile models
does not become a probability sample of a broader population when we
call `t.test()`. A broader inference needs an explicitly supported target
process and appropriate dependence and distribution conditions.

Both intervals contain 20 mpg, matching their two-sided tests' failure
to reject that null at .05. This matching holds when the interval and
test use the same method, assumptions and calibration. A p-value is a
tail probability under the null and those conditions, not the
probability that the null is true. Including 20 establishes neither
exact equality nor practical equivalence. An interval for a process
mean also differs from a range containing 95% of individual car values.

### Supplied summaries and a planned direction

Chapter 3's separate illustration supplies $n=25$, mean 52 and sample
SD 5 under independent Normal sampling. These summaries are enough to
calculate its mean interval and test. We should use the summaries
directly rather than inventing observations to pass into `t.test()`.
For the null mean 50, the estimated SE is $5/\sqrt{25}=1$ and the
observed t statistic is $(52-50)/1=2$.

```{r}
#| label: ch05-summary-mean
#| echo: true
n_sum <- 25
mean_sum <- 52
sd_sum <- 5
SE_sum <- sd_sum / sqrt(n_sum)
t_sum <- (mean_sum - 50) / SE_sum
c(SE = SE_sum, critical = qt(.975, n_sum - 1), t = t_sum,
  two_sided_p = 2 * pt(abs(t_sum), n_sum - 1, lower.tail = FALSE),
  greater_p = pt(t_sum, n_sum - 1, lower.tail = FALSE))
mean_sum + c(-1, 1) * qt(.975, n_sum - 1) * SE_sum
```

The SE is 1.0000, critical t is 2.0639, and the interval is
[49.9361, 54.0639]. The two-sided p-value is .0569; the upper-tail
p-value for the alternative $\mu>50$ is .0285. These differ because
they answer different prespecified questions. A directional alternative
must be chosen from the question before inspecting the result. For a
negative statistic, the upper-tail probability would exceed .5;
blindly halving a two-sided p-value would then give the wrong tail.

The two-sided interval contains 50, consistent with the two-sided test
at .05. The upper-tail test does not match that same two-sided 95%
interval. If a task stipulates the rounded critical value 2.064,
its endpoints are [49.9360, 54.0640]. Use the supplied value for that
task and retain the software quantile when reproducing the software
procedure; this is a rounding difference, not a changed conclusion.

### Paired observations and complete pairs {#ch05-paired}

The sleep measurements return us to a comparison whose unit is a
person. Section 3 aligned measurements by ID into `sleep_wide` and
defined `sleep_d` as drug 2 minus drug 1. There are ten people with
complete pairs, not twenty independent people. Their condition means
are 0.7500 and 2.3300 hours, and the mean of their aligned differences
is 1.5800 hours. Nine differences are positive and one is zero;
they range from 0 to 4.6 hours.

```{r}
#| label: fig-ch05-pairs
#| echo: false
#| fig-cap: "Two measurements for each of ten people."
#| fig-alt: "Lines connect drug 1 and drug 2 for each person; nine rise and one is flat."
#| fig-width: 6.2
#| fig-height: 3.4
matplot(1:2, t(as.matrix(sleep_wide[c("extra.1", "extra.2")])),
  type = "b", lty = 1, pch = 16, col = "grey40",
  xaxt = "n", xlab = "Drug", ylab = "Extra sleep (hours)")
axis(1, at = 1:2, labels = c("1", "2"))
```

The inferential target is a process mean difference $\mu_D$. Under a
working model of independent Normal differences between people,
$s_d/\sqrt n$ estimates the SE of their mean and a t reference has
$n-1$ df. Normality concerns the differences. Dependence between the
two measurements on one person is preserved in their difference.

The observed difference SD is 1.2300 hours and its mean SE is 0.3890
hours. Thus the interval and zero-null statistic are

$$
\bar d\pm t_{.975,9}\frac{s_d}{\sqrt{10}}
=[0.7001,\ 2.4599]\ \text{hours},\qquad
t=\frac{\bar d}{s_d/\sqrt{10}}=4.0621.
$$

`paired = TRUE` tells R to subtract measurements at corresponding
positions. The ID-based alignment performed earlier is what makes
those positions correct. `complete.cases()` checks that both required
measurements are present on the same row. Here every pair is complete.

```{r}
#| label: ch05-paired-test
#| echo: true
complete_sleep <- complete.cases(sleep_wide[c("extra.1", "extra.2")])
table(complete_sleep)
paired_t <- t.test(sleep_wide$extra.2[complete_sleep],
  sleep_wide$extra.1[complete_sleep], paired = TRUE)
c(t = unname(paired_t$statistic), df = unname(paired_t$parameter),
  p = paired_t$p.value)
paired_t$conf.int
c(estimate = mean(sleep_d), SE = sd(sleep_d) / sqrt(length(sleep_d)),
  critical = qt(.975, length(sleep_d) - 1))
t.test(sleep_d, conf.level = .99)$conf.int
```

The extracted components give $t(9)=4.0621$, $p=.0028$ and the 95%
interval [0.7001, 2.4599] hours. The critical t is 2.2622. A paired t calculation
is a one-sample t calculation on the within-person differences, so the
last call uses `sleep_d` directly. Increasing confidence to 99% widens
its interval to [0.3159, 2.8441]; the observations and estimate stay
unchanged. The Computer Lab checks the two interfaces on the same pairs.

With missing measurements, retain the rows having both values before
forming differences. Removing missing entries independently from each
column could align different people. Complete-pair analysis also changes
who contributes and may need a missingness justification; a software
exclusion is not that justification.

The model interval concerns a process mean, not a guarantee of benefit
for every person. Ten differences give limited protection against
non-Normal tails or influential cases. Neither the paired call nor the
dataset establishes random sampling, treatment order, absence of
carryover, or other conditions needed for a causal interpretation.

The same arithmetic reproduces Chapter 3's invented eight-student
illustration. Keep after minus before so that improvement is positive:

```{r}
#| label: ch05-student-pairs
#| echo: true
before <- c(50, 54, 48, 60, 55, 58, 62, 53)
after <- c(49, 54, 49, 62, 57, 61, 66, 58)
change <- after - before
student_t <- t.test(after, before, paired = TRUE)
c(t = unname(student_t$statistic), df = unname(student_t$parameter),
  p = student_t$p.value)
student_t$conf.int
c(mean = mean(change), SD = sd(change),
  SE = sd(change) / sqrt(length(change)))
```

The mean change is 2.0000 points, SD 2.0000 and SE 0.7071; the test
gives $t(7)=2.8284$, $p=.0255$, and interval [0.3280, 3.6720]. Pairing
preserves the comparison; design and timing evidence still determine
whether the change can be attributed to an intervention. Chapter 3's
paired-school appendix applies the same form to school-specific
class-mean contrasts. Its twelve school blocks, assignment interpretation
and complete data remain there; the Computer Lab reference reproduces
that calculation without treating 480 student scores as independent
assignment contrasts.

### Two independent groups and Welch's method {#ch05-welch}

Now compare separate automatic and manual car models. Let
$\Delta\mu=\mu_A-\mu_M$ be the automatic-minus-manual process mean
mpg difference. Under an independent-groups working model, the two
means contribute separate variance terms. Welch's method estimates
both and allows their group variances to differ [@welch1947].

$$
\widehat{\Delta\mu}=\bar y_A-\bar y_M,\qquad
a=\frac{s_A^2}{n_A},\quad b=\frac{s_M^2}{n_M},\qquad
\widehat{\operatorname{SE}}=\sqrt{a+b},
$$
$$
\nu=\frac{(a+b)^2}{a^2/(n_A-1)+b^2/(n_M-1)}.
$$

The Welch--Satterthwaite df $\nu$ account approximately for the two
estimated variance terms. They need not be an integer. R codes automatic
transmission as `am == 0` and manual as `am == 1`; extracting the two
vectors in a stated order makes the subtraction unambiguous.

```{r}
#| label: ch05-independent-welch
#| echo: true
auto <- cars$mpg[cars$am == 0]
manual <- cars$mpg[cars$am == 1]
a <- var(auto) / length(auto)
b <- var(manual) / length(manual)
df_welch <- (a + b)^2 /
  (a^2 / (length(auto) - 1) + b^2 / (length(manual) - 1))
c(n_auto = length(auto), n_manual = length(manual),
  mean_auto = mean(auto), mean_manual = mean(manual),
  difference = mean(auto) - mean(manual), SE = sqrt(a + b),
  df = df_welch)
welch <- t.test(auto, manual)
c(t = unname(welch$statistic), df = unname(welch$parameter),
  p = welch$p.value)
welch$conf.int
```

There are 19 automatic and 13 manual models, with means 17.1474 and
24.3923 mpg. The difference is $-7.2449$ mpg, SE 1.9232 and
approximate df 18.3323. The two-sided p-value is .0014 and the 95%
interval is [$-11.2802$, $-3.2097$]. The negative sign describes
the declared order. Reversing it would reverse the estimate and
negate and exchange the endpoints, while preserving the two-sided p-value.

The two-sample `t.test()` default is Welch's method. Pooled-variance
inference, requested with `var.equal = TRUE`, gives SE 1.7644, df 30,
p=.0003 and interval [$-10.8484$, $-3.6415$].

```{r}
#| label: ch05-independent-pooled
#| echo: false
pooled <- t.test(auto, manual, var.equal = TRUE)
c(SE = pooled$stderr, df = pooled$parameter, p = pooled$p.value)
pooled$conf.int
```

The pooled t reference is exact for independent Normal groups with a
common variance. Welch's t reference remains
approximate even for independent unequal-variance Normal groups.
The estimate and observations are unchanged; the variance assumptions
and calibration differ. A smaller p-value is not a reason to select
the common-variance assumption. Neither method resolves differences
in weight, engine and other features associated with transmission.
These observational comparisons do not isolate a transmission effect.

We can also complete Chapter 3's special-practice calculation from its
supplied summaries. The invented study assigned the same 120
participants, 60 to each format, and observed every delayed assessment.
Its estimand is the mean effect of Special rather than Usual assignment
among those participants. The repetition reassigns the fixed people;
it does not independently sample two new groups.

Section 3 explained why the familiar squared SE estimates an upper
bound on assignment variance on average. We retain Chapter 3's
approximate Welch working reference for that assignment analysis.
The shared arithmetic does not replace its design-based interpretation.
The earlier objects hold means 27.1 and 24.0 and SDs 6.0 and 6.4:

\Needspace{20\baselineskip}

```{r}
#| label: ch05-special-practice-inference
#| echo: true
a_special <- sd_special^2 / n_special
a_usual <- sd_usual^2 / n_usual
df_special <- (a_special + a_usual)^2 /
  (a_special^2 / (n_special - 1) + a_usual^2 / (n_usual - 1))
t_special <- estimate / standard_error
c(estimate = estimate, SE = standard_error, df = df_special,
  critical = qt(.975, df_special), t = t_special,
  p = 2 * pt(abs(t_special), df_special, lower.tail = FALSE))
estimate + c(-1, 1) * qt(.975, df_special) * standard_error
```

The estimate is 3.1000 points, SE 1.1325, df 117.5119 and critical
t 1.9804. The statistic is 2.7372, p=.0072, with 95% interval
[0.8571, 5.3429]. This interval includes effects below a stated
two-point importance benchmark as well as above it. Evidence against
a zero effect therefore supports less than a claim that the mean
improvement exceeds two points. Generalization beyond the participating
students needs further support. The supplied rounded summaries are
also distinct from the companion CSV's slightly different SDs; they
should not silently be substituted for one another.

### Coverage belongs to the procedure {#ch05-coverage}

What makes an interval procedure a 95% procedure? We can inspect the
answer by generating data from a known process. Let independent
differences be Normal with mean 1.5 hours and SD 1.2 hours. These are
teaching constants, not a reconstruction of the historical sleep study.
Draw ten differences, estimate their mean and SE, and form a t interval.

```{r}
#| label: ch05-one-coverage-study
#| echo: false
set.seed(60610)
one_sample <- rnorm(10, mean = 1.5, sd = 1.2)
one_mean <- mean(one_sample)
one_se <- sd(one_sample) / sqrt(10)
one_interval <- one_mean + c(-1, 1) * qt(.975, 9) * one_se
one_interval
1.5 >= one_interval[1] & 1.5 <= one_interval[2]
```

This first interval is [1.0213, 2.2119] hours and contains 1.5, so
the logical check returns `TRUE`: both endpoint conditions hold. A
repeated experiment generates fresh observations and repeats these
steps, keeping the true mean fixed. Recalculating the interval on the
same observations would produce the same interval, not a coverage
experiment. The Computer Lab reference supplies the full loop and
changes in sample size or repetitions.

\Needspace{11\baselineskip}

| n | Mean of estimates | Empirical SE | Model SE | Coverage |
|--:|--:|--:|--:|--:|
| 10 | 1.5015 | 0.3766 | 0.3795 | .9515 |
| 20 | 1.5006 | 0.2591 | 0.2683 | .9510 |
| 40 | 1.5014 | 0.1886 | 0.1897 | .9460 |

: Coverage of mean intervals from the Normal process, using 2,000 repetitions at each sample size and resetting seed 60610 for each row. {#tbl-ch05-coverage tbl-colwidths="[10,25,25,20,20]"}

The empirical SE is the SD of the generated means; the model SE is
$1.2/\sqrt n$. Coverage is the fraction of generated intervals
containing 1.5. Under the declared Normal model, the t procedure has
exact 95% coverage. A finite simulation approximates that value, so
its observed fraction need not equal .9500.

The fraction has its own **Monte Carlo SE**. Treat each independent
generated study's coverage indicator as a Bernoulli outcome. At
coverage probability .95 and $B=2000$ repetitions,

$$
\operatorname{MCSE}(\widehat p_{\mathrm{cover}})
=\sqrt{\frac{.95(.05)}{2000}}=0.0049.
$$

Increasing $B$ reduces simulation noise; quadrupling it halves this
MCSE. Increasing $n$ supplies more information in each study and
narrows intervals: changing 10 to 40 halves the model SE. It does
not systematically raise the nominal 95% coverage above 95%. The
individual-difference SD, the mean's SE and the simulation's MCSE
describe different sources of variation.

Chapter 3's coverage display shows the same fixed-parameter reasoning.
For one observed interval the parameter is either inside or outside;
the 95% describes repeated use of the procedure. A seed makes the
simulation repeatable, and suitable model and design information make
an interval interpretation defensible in an actual study.

## Correlations and Their Inference {#sec-ch5-relationships}

[]{#ch05-correlations}

### Begin with the plotted observations

Mean comparisons group observations by a category. A relationship between
two quantitative variables instead begins with pairs of values. For the
32 car models, `wt` measures weight in thousands of pounds and `mpg`
fuel economy in miles per US gallon. Keep the same complete rows for
the graph and every numerical description of that pair:

```{r}
#| label: ch05-car-relationship
car_pairs <- mtcars[complete.cases(mtcars[c("wt", "mpg")]), ]
c(n = nrow(car_pairs), covariance = cov(car_pairs$wt, car_pairs$mpg),
  correlation = cor(car_pairs$wt, car_pairs$mpg))
```

```{r}
#| label: fig-ch05-car-pairs
#| fig-cap: "Mileage and weight for 32 recorded car models."
#| fig-alt: "Heavier models generally have lower mpg, with substantial scatter around the decreasing pattern and possible curvature."
#| fig-width: 6.2
#| fig-height: 3.5
plot(car_pairs$wt, car_pairs$mpg, pch = 16,
     xlab = "Weight (thousand pounds)", ylab = "Miles per gallon")
```

Read direction, strength and form before reducing the display to a
coefficient. Here heavier models generally have lower mpg. Inspect
unusual observations, the range of weights, possible subgroups and
changes in spread. A coefficient compresses these features, so the
plot helps determine what that number can usefully describe.

Covariance averages products of paired deviations. On the same $n$
complete observations, sample covariance and Pearson correlation are

$$
s_{xy}=\frac{\sum_i(x_i-\bar x)(y_i-\bar y)}{n-1},\qquad
r_{xy}=\frac{s_{xy}}{s_xs_y}.
$$

A pair contributes positively when both values are above their means
or both are below; opposite deviations contribute negatively.
Covariance retains the product of the two units. Dividing by the SDs
gives a unitless coefficient between $-1$ and 1 when both variables
have nonzero spread. For $x=(1,2,3)$ and $y=(2,4,6)$, the deviation
products sum to four, covariance is $4/(3-1)=2.0000$, and the SDs
are one and two. Thus $r=1.0000$: all points lie on one increasing line.

The car covariance is $-5.1167$ thousand-pounds times mpg and
$r=-.8677$. Changing the weight unit changes covariance predictably.

```{r}
#| label: ch05-covariance-rescaling
#| echo: false
c(covariance_pounds = cov(1000 * car_pairs$wt, car_pairs$mpg),
  correlation_pounds = cor(1000 * car_pairs$wt, car_pairs$mpg))
```

Covariance multiplies by 1,000; the weight SD does too, leaving the
correlation unchanged. Adding a constant changes neither covariance
nor correlation. Multiplying one variable by a negative constant
reverses both signs.

A small correlation need not mean little relationship. The five points
$x=(-2,-1,0,1,2)$ and $y=x^2=(4,1,0,1,4)$ follow an exact curve while
having zero covariance and correlation.

```{r}
#| label: ch05-curve-example
#| echo: false
x_curve <- -2:2
y_curve <- x_curve^2
c(covariance = cov(x_curve, y_curve), r = cor(x_curve, y_curve))
```

```{r}
#| label: fig-ch05-curve
#| echo: false
#| fig-cap: "An exact curved relationship with zero Pearson correlation."
#| fig-alt: "Five points at x equal to minus two, minus one, zero, one and two have y values four, one, zero, one and four, forming a U."
#| fig-width: 5.5
#| fig-height: 2.8
plot(x_curve, y_curve, type = "b", pch = 16,
     xlab = "x", ylab = expression(y == x^2))
```

Positive and negative cross-products cancel by symmetry, although
knowing $x$ determines $y$. Pearson correlation describes linear
association, so “the variables are unrelated” would overstate this result.

Chapter 2's gallery gives a compact visual reference. Each panel below
contains 80 simulated pairs and uses the same axes. The generating
correlations are $-.9,-.5,0,.3,.6,.9$, but each label reports the
correlation of the displayed observations. Sampling variation separates
a model parameter from the realized sample coefficient.

```{r}
#| label: fig-ch05-correlation-gallery
#| echo: false
#| fig-cap: "Realized samples with different directions and strengths of linear association."
#| fig-alt: "Six equal-size samples on common axes range from strongly negative through weak association to strongly positive. Their observed correlations are minus .8916, minus .6252, minus .1328, .2808, .4803 and .8800."
#| fig-width: 6.5
#| fig-height: 4.5
set.seed(60206)
target_r <- c(-.9, -.5, 0, .3, .6, .9)
gallery <- lapply(target_r, function(rho) {
  x <- rnorm(80)
  y <- rho * x + sqrt(1 - rho^2) * rnorm(80)
  data.frame(x, y)
})
gallery_limits <- range(unlist(gallery))
oldpar <- par(mfrow = c(2, 3), mar = c(4, 4, 2, 1),
              mgp = c(2.2, .7, 0))
for (j in seq_along(gallery)) {
  g <- gallery[[j]]
  plot(g$x, g$y, pch = 16, cex = .55,
       xlim = gallery_limits, ylim = gallery_limits,
       xlab = "x", ylab = "y",
       main = sprintf("Observed r = %.4f", cor(g$x, g$y)))
}
par(oldpar)
```

The sign gives direction and the absolute value summarizes linear
strength. The model with generating correlation zero happened to
produce a slightly negative sample coefficient. A generated display
illustrates the model; it provides no evidence that actual study
observations follow that model. The lab reference supplies the complete
generating and plotting code.

### Covariance and correlation matrices {#ch05-matrices}

A covariance matrix extends the pairwise calculation to several
variables. Its diagonal contains variances, and its off-diagonal
entries contain covariances. A correlation matrix standardizes each
entry; diagonal entries are one for variables with positive variance.

Select one common complete-case set before forming either matrix:

```{r}
#| label: ch05-matrix-calculation
variables <- c("mpg", "hp", "wt")
X <- mtcars[complete.cases(mtcars[variables]), variables]
S <- cov(X)
R <- cov2cor(S)
s <- sqrt(diag(S))
S_rebuilt <- R * outer(s, s)
c(n = nrow(X), largest_reconstruction_error =
    max(abs(S_rebuilt - S)))
```

All 32 rows are complete for fuel economy, gross horsepower (`hp`)
and weight. `diag(S)` extracts the variances; taking their square roots
gives SDs. `outer(s,s)` forms every pairwise SD product. Multiplying it
entry by entry by `R` restores the covariance matrix:

| Variable | mpg | hp | wt |
|:--|--:|--:|--:|
| mpg | 36.3241 | $-320.7321$ | $-5.1167$ |
| hp | $-320.7321$ | 4700.8669 | 44.1927 |
| wt | $-5.1167$ | 44.1927 | 0.9574 |

: Sample covariance matrix for the same 32 car models. {#tbl-ch05-covariance}

| Variable | mpg | hp | wt |
|:--|--:|--:|--:|
| mpg | 1.0000 | $-.7762$ | $-.8677$ |
| hp | $-.7762$ | 1.0000 | .6587 |
| wt | $-.8677$ | .6587 | 1.0000 |

: Corresponding Pearson correlation matrix. {#tbl-ch05-correlation}

For example, the mpg variance has squared-mpg units, while its
covariance with horsepower has mpg-times-horsepower units. Correlations
remove those units. If $D$ is the diagonal matrix of positive sample
SDs, the two operations are

$$
R=D^{-1}SD^{-1},\qquad S=DRD.
$$

The reconstructed values agree up to floating-point precision.
Correlations alone cannot recover the original covariance matrix:
the SDs and their units are also needed. A constant column has zero SD
and an undefined correlation; detect it before dividing or converting.

Using common rows matters when measurements are missing. Computing
each entry from a different pairwise subset can produce a matrix that
does not describe one joint sample and is not positive semidefinite,
so it can imply an impossible negative variance for a linear combination.
For airquality's ozone, solar radiation and wind, the individual
recorded counts are 116, 146 and 153, but the common set contains 111
days. The Computer Lab reference computes that common-case matrix and
checks that standardizing `X` leaves its correlation matrix unchanged.

### Pearson, rank, and point-biserial coefficients

Pearson correlation uses the measured scales and describes linear
association. **Spearman correlation** applies correlation to ranks and
describes monotone association: values tend to move consistently in
one direction, even when the pattern is curved. **Kendall correlation**
compares concordant and discordant ordering of pairs, with an adjustment
for ties. The coefficients address related but distinct features.

```{r}
#| label: ch05-correlation-types
#| echo: false
c(Pearson = cor(X$mpg, X$hp),
  Spearman = cor(X$mpg, X$hp, method = "spearman"),
  Kendall = cor(X$mpg, X$hp, method = "kendall"))
point_biserial <- cor(mtcars$am, mtcars$mpg)
c(point_biserial = point_biserial,
  reversed_coding = cor(1 - mtcars$am, mtcars$mpg))
```

For mpg and horsepower, the coefficients are $-.7762,-.8947,-.7428$.
Tied horsepower values retain their tied ranks; their presence also
affects test calibration. The different magnitudes do not make the
largest coefficient the universally best description. The question and
plot determine which association is relevant.

A **point-biserial correlation** is Pearson correlation with one binary
variable coded 0/1. Here `am=1` means manual and `am=0` automatic;
$r=.5998$ corresponds to means 24.3923 and 17.1474 mpg, respectively.
Reversing the coding reverses the coefficient's sign while leaving
the same two groups. For more than two categories, arbitrary numerical
labels generally do not define a meaningful quantitative scale for
correlation. Coding must follow the variable's role.

### A selected correlation test {#ch05-correlation-tests}

The descriptive weight–mpg correlation concerns the recorded models.
Inference adds a target process correlation $\rho$ and a supporting
model. Suppose we selected the two-sided null $H_0:\rho=0$ before
examining the result, with level $\alpha=.05$. Under independent
bivariate-Normal sampling, the zero-correlation statistic has an exact
$t_{n-2}$ reference:

$$
t=r\sqrt{\frac{n-2}{1-r^2}},\qquad df=n-2.
$$

Use the same 32 complete pairs as in the graph:

```{r}
#| label: ch05-pearson-test
pearson_test <- cor.test(car_pairs$wt, car_pairs$mpg,
                         method = "pearson")
c(r = unname(pearson_test$estimate), t = unname(pearson_test$statistic),
  df = unname(pearson_test$parameter), p = pearson_test$p.value)
pearson_test$conf.int
r_wt <- cor(car_pairs$wt, car_pairs$mpg)
t_wt <- r_wt * sqrt((nrow(car_pairs) - 2) / (1 - r_wt^2))
c(t = t_wt, df = nrow(car_pairs) - 2,
  p = 2 * pt(abs(t_wt), nrow(car_pairs) - 2, lower.tail = FALSE))
```

The coefficient is $r=-.8677$, with $t(30)=-9.5590$ and $p<.0001$.
R prints the very small p-value in scientific notation.
The result supplies evidence against zero process correlation under
the stated reference model; a small p-value does not establish that
model or turn the historical car records into a probability sample.
Normality is unnecessary to calculate descriptive $r$. Independence,
shape, influential cases and the observed range matter for inference.

The printed 95% interval is $[-.9338,-.7441]$. It uses Fisher's
transformation, a different approximation from the exact zero-null test.
Transform the bounded correlation, form an approximately symmetric
interval on that scale, then transform back:

$$
CI_\rho=\tanh\left\{\operatorname{atanh}(r)
\pm\frac{z_{.975}}{\sqrt{n-3}}\right\}.
$$

```{r}
#| label: ch05-fisher-interval
fisher_interval <- tanh(atanh(r_wt) +
  c(-1, 1) * qnorm(.975) / sqrt(nrow(car_pairs) - 3))
fisher_interval
```

This interval requires at least four complete pairs and remains
approximate even under bivariate Normality. Substantial nonlinearity
can make linear correlation a poor answer to the substantive question,
regardless of the numerical width.

For a preselected rank-association question, keep the coefficient
beside the test and retain the statistic's native name. The mpg and
horsepower data contain ties, so the approximate calibrations are
requested explicitly with `exact = FALSE`.

```{r}
#| label: ch05-rank-tests
#| echo: false
spearman_test <- cor.test(X$mpg, X$hp, method = "spearman",
                          exact = FALSE)
c(S = unname(spearman_test$statistic), p = spearman_test$p.value)
kendall_test <- cor.test(X$mpg, X$hp, method = "kendall",
                         exact = FALSE)
c(z = unname(kendall_test$statistic), p = kendall_test$p.value)
```

Spearman's output reports $S=10337.2903$, while Kendall's reports
$z=-5.8710$; both give $p<.0001$. These are not Pearson t statistics.
The rank tests still require an appropriate independent-observation
model and a question about the relevant association. They do not supply
the same Fisher interval as the Pearson call. Choosing a method because
its p-value is smallest changes the testing procedure; a matrix of
unplanned tests also creates a comparison family, developed in Chapter 6.

For a binary predictor, the point-biserial t statistic equals a pooled
two-group mean-test statistic. Keep manual first, matching the positive
indicator coding:

```{r}
#| label: ch05-point-biserial-equivalence
#| echo: false
r_pb <- cor(mtcars$am, mtcars$mpg)
t_pb <- r_pb * sqrt((nrow(mtcars) - 2) / (1 - r_pb^2))
pooled_manual_first <- t.test(manual, auto, var.equal = TRUE)
c(point_biserial_t = t_pb,
  pooled_t = unname(pooled_manual_first$statistic),
  df = nrow(mtcars) - 2,
  p = 2 * pt(abs(t_pb), nrow(mtcars) - 2, lower.tail = FALSE))
```

Both give $t(30)=4.1061$ and two-sided $p=.0003$. The supporting model
is independent, conditionally Normal outcomes with equal variance
in the two groups; a binary predictor itself cannot have a bivariate
Normal distribution with the outcome. This equivalence is to the
pooled procedure and generally does not hold for Welch's test.
Chapter 7 connects it further to a regression slope.

### One-control partial correlation {#ch05-partial}

Chapter 2 distinguished changing the cases from adjusting a relationship
on the same cases. For variables $X,Y,Z$, the **partial correlation**
of $X$ and $Y$ controlling linearly for $Z$ is

$$
r_{XY\cdot Z}=\frac{r_{XY}-r_{XZ}r_{YZ}}
{\sqrt{(1-r_{XZ}^2)(1-r_{YZ}^2)}}.
$$

Its supplied practice-time, score and baseline correlations were
$r_{XY}=.60,r_{XZ}=.50,r_{YZ}=.60$. Compute that descriptive adjustment
first, then apply the same formula to mpg and horsepower controlling
for weight:

```{r}
#| label: ch05-partial-formula
r_xy <- .60; r_xz <- .50; r_yz <- .60
(r_xy - r_xz * r_yz) / sqrt((1 - r_xz^2) * (1 - r_yz^2))
r_partial <- (R["mpg", "hp"] - R["mpg", "wt"] * R["hp", "wt"]) /
  sqrt((1 - R["mpg", "wt"]^2) * (1 - R["hp", "wt"]^2))
r_partial
```

The supplied correlations give .4330. For the car models, the mpg–hp
coefficient changes from $-.7762$ to $-.5470$ after linear adjustment
for weight. The adjustment uses the same 32 rows; it does not exclude
heavy cars or replace their observations.

There is a second way to see the calculation. Fit a straight line from
weight to each variable and retain its **residuals**, the observed
values minus their fitted values. `lm()` fits the line specified by
its formula; `mpg ~ wt` places the outcome left of `~` and the predictor
right of it. Each fit below includes an intercept. `residuals()` returns
what remains after that fitted linear relationship:

```{r}
#| label: ch05-partial-residuals
resid_mpg <- residuals(lm(mpg ~ wt, data = X))
resid_hp <- residuals(lm(hp ~ wt, data = X))
c(formula = r_partial, residual_correlation = cor(resid_mpg, resid_hp))
```

```{r}
#| label: fig-ch05-partial
#| echo: false
#| fig-cap: "Mileage and horsepower after separately fitting their linear relationships with weight."
#| fig-alt: "The two residual vectors from the same 32 car models retain a negative association, with partial correlation minus .5470."
#| fig-width: 6.2
#| fig-height: 3.5
plot(resid_hp, resid_mpg, pch = 16,
     xlab = "Horsepower residual after weight adjustment",
     ylab = "Mpg residual after weight adjustment")
```

Both calculations give $-.5470$. A model with several predictors
develops this interpretation further in Chapter 8. Here the important
distinction is that specified linear relationships have been removed.
This operation does not remove all confounding, omitted nonlinear
relationships or measurement problems, and it does not establish a
causal effect of horsepower.

If the partial coefficient is the preselected inferential question,
the reference must account for fitting the control. With $k$ controls
and $n$ common observations, the classical statistic uses $n-k-2$ df:

$$
t=r_{XY\cdot Z}\sqrt{\frac{n-k-2}{1-r_{XY\cdot Z}^{\,2}}}.
$$

```{r}
#| label: ch05-partial-test
k_controls <- 1
df_partial <- nrow(X) - k_controls - 2
t_partial <- r_partial * sqrt(df_partial / (1 - r_partial^2))
p_partial <- 2 * pt(abs(t_partial), df_partial, lower.tail = FALSE)
c(r = r_partial, t = t_partial, df = df_partial, p = p_partial)
```

The result is $t=-3.5187$, $df=29$ and $p=.0015$. This exact classical
calibration requires independent observations, full-rank linear
adjustment (without redundant fitted terms) and the relevant Normal
conditional-error model, or a
multivariate-Normal model. Applying ordinary `cor.test()` to the fitted
residual vectors would use $n-2=30$ df and ignore the estimated control.
The correctly adjusted reference makes the fitting step visible.

A sensitivity analysis asks a different question. Restricting the car
record to weight at most 5,000 pounds changes its cases and gives
$r=-.8792$ between weight and mpg. Removing three ozone days above
120 ppb changes its mean to 39.4867, SD to 28.8942 and median to 30.0000.
These are consequences of stated restrictions, not reasons by themselves
to delete observations. The lab reference compares them with the full
recorded sets. Linear adjustment and case selection remain distinct,
even when both change a reported association.

## Report the Answer {#sec-ch5-report}

[]{#ch05-report}

### A result connects numbers to a question

A statistical report returns from the function output to the question.
Name the people or records, outcome and units, comparison, estimate and
uncertainty. Then state the conditions and scope that support the
interpretation. The sleep result can be reported as follows:

> For the ten recorded people, extra sleep under drug 2 averaged
> 1.5800 hours more than under drug 1. The SD of the ten within-person
> differences was 1.2300 hours. Under an independent Normal-differences
> working model, the mean difference has SE 0.3890 hours and a 95% t
> interval of [0.7001, 2.4599] hours ($t(9)=4.0621$, two-sided $p=.0028$).
> This uncertainty concerns the stated process mean. A causal account
> also requires information about treatment assignment, order and carryover.

The first two sentences describe the recorded people. The next adds
model-based uncertainty for a named target, rather than a range of
individual responses. The final sentence identifies specific information
needed for a causal claim. A successful computation supplies the result;
its study setting supplies the argument for using it [@peng2011].

### Connect a report to its source

A saved script records calculations. A Quarto source places them beside
the prose that explains them, so a render executes the code and inserts
its output. The prepared two-file example uses a short `.qmd` source and
an accompanying bibliography. Its paired analysis uses these same data,
subtraction direction, model and base-R commands. One chunk computes
the result; an inline value inserts a number into a sentence; a table
or graph helps readers see what it describes.

During the final Computer Lab round, render the prepared source, change
confidence from .95 to .99 and revise the corresponding sentence. The
interval becomes [0.3159, 2.8441] while the estimate stays 1.5800 hours.
That small change reveals the connection between an input, a calculation,
output and interpretation. The lab reference explains the header,
chunks, cross-references, citations and report-support installation.
Its first HTML render requires no PDF typesetting setup.

When a data-selection rule or a method changes, computed values can
update automatically; the interpretation still needs to be reread.
A reproducible report keeps question, data, code, output and explanation
consistent. Rendering demonstrates that a document can be produced,
while the evidence and reasoning determine what its claims establish.

## Reading Bridge {#sec-ch5-summary}

When a study reports a mean, interval or association, identify what was
recorded, which observations entered the calculation, the statistic
or comparison and its units. Next locate the repetition or model that
supports the SE, reference distribution, interval or test. Finally
separate the numerical result from the assumptions needed for population,
practical or causal claims. R reproduces a computation; the study's
evidence and design support its interpretation.

This chapter connected base-R descriptions to sampling variation,
mean inference and selected correlation inference. Chapter 6 applies
the same reading pattern to counts and group models, introduces a small
ggplot2 plotting vocabulary, and examines fitting criteria and procedures
across repetitions. Chapters 7–8 extend correlations, comparisons and
linear adjustment into regression arguments.

### Key terms {.unnumbered}

These entries use the shared Part One definitions. The first-use
explanations above connect them to specific observations and commands;
Chapters 2–3 provide the broader descriptive and inferential reference.

| Term and common alternatives | Meaning |
|:--|:--|
| Object (R object) | A named value, vector, table, model result, or other item stored in an R session. Assignment with `<-` names an object; the Environment pane lists the current objects. |
| Vector | A sequence of values of one basic type: numeric, character, or logical. `NA` marks an unavailable value; it is not zero. A data frame is a table whose columns are vectors of equal length, one row per case. |
| Function and argument (R function) | A function performs an operation on supplied inputs; an argument names an input or an option, such as `na.rm = TRUE`. An R argument is distinct from a study argument in the argument-based approach. |
| Package (library) | An installed collection of functions and data, such as stats or graphics, which are supplied with R. Installing an additional package makes it available on the computer; `library()` attaches it for the session; `package::function()` names one exported function explicitly. |
| Script and Quarto source | A saved `.R` script records commands so an analysis can be rerun; a `.qmd` source connects prose, mathematics, executable code, tables, figures, and references, and renders to PDF, HTML, or Word. |
| Simulation and seed (Monte Carlo simulation) | Generating data from a declared process to see what a statistic or procedure does across repetitions. A seed makes the random draws repeatable; it does not make the process true for an actual study. |
| Monte Carlo standard error | The numerical uncertainty of a simulated probability or mean that comes from using finitely many repetitions; more repetitions shrink it. It is distinct from a study's SE, which depends on the study's sample size. |

: Computing and simulation terms. {#tbl-ch05-computing-terms tbl-colwidths="[28,72]"}

| Term and common alternatives | Meaning |
|:--|:--|
| Covariance matrix | A square table of variances on the diagonal and covariances off the diagonal for a stated set of variables and observations. |
| Correlation matrix | A square table of pairwise correlation coefficients for a stated set of variables and observations; its diagonal is 1 for variables with positive variance. |

: Data-summary terms. {#tbl-ch05-matrix-terms tbl-colwidths="[28,72]"}

### Further reading {.unnumbered}

R's reference system [@rcore2025] and local Help for `mean`, `sd`,
`quantile`, `t.test`, `cor`, `cov2cor` and `cor.test` describe their
arguments, defaults and computational conventions. The original papers
by Student [-@student1908] and Welch [-@welch1947] explain the classical
mean procedures developed here. Quarto's documentation [@quarto2026]
gives examples of executable reports, and Peng [-@peng2011] connects
reproducible computation with broader scientific reporting.

## Exercises {.unnumbered}

Use the stated records and conventions, including each prompt's supplied
rounded summaries. Explain the statistical meaning as well as the calculation.
Report numerical results to the requested precision.

<!-- prompt-id: CH05-E01 -->

### Exercise 1: Count the right observations (Foundational) {.unnumbered #ex-ch05-01}

A table contains 40 students with one baseline and one follow-up row each,
plus one duplicate follow-up row. There are 81 rows. Explain the difference
between a row count and a student count. If four students have missing
follow-up scores, give the denominator for the mean observed follow-up score
after the documented duplicate is resolved. State what the mean describes.

<!-- END EX05-01 PROMPT -->

<!-- prompt-id: CH05-E02 -->

### Exercise 2: Center, spread and a changed observation (Foundational) {.unnumbered #ex-ch05-02}

Three recorded scores are 50, 60 and 70 points. Calculate the mean, median,
sample variance and sample SD. Replace 70 by 100 and calculate the new mean,
median and SD. Explain what changed statistically. Use the sample-variance
denominator $n-1$ and report SDs to two decimal places.

<!-- END EX05-02 PROMPT -->

<!-- prompt-id: CH05-E03 -->

### Exercise 3: A missing score is not zero (Foundational) {.unnumbered #ex-ch05-03}

Students A, B and C have baseline scores 50, 60 and 70, and corresponding
follow-up scores 55, missing and 78. Calculate the available follow-up mean,
the complete-pair mean change and the difference of the separate available
means. State each denominator. Explain why equality of two numerical answers
would not establish that they used the same student sets.

<!-- END EX05-03 PROMPT -->

<!-- prompt-id: CH05-E04 -->

### Exercise 4: Choose the group denominator (Foundational) {.unnumbered #ex-ch05-04}

A table classifies 19 automatic car models, of which 4 meet a fuel-economy
threshold, and 13 manual models, of which 10 meet it. Calculate the percentage
meeting the threshold within each transmission group and their manual-minus-
automatic percentage-point difference. Then calculate the percentage of
threshold-meeting models that are manual. Explain the different denominators.

<!-- END EX05-04 PROMPT -->

<!-- prompt-id: CH05-E05 -->

### Exercise 5: Read a relationship before evaluating it (Integrative) {.unnumbered #ex-ch05-05}

For five paired observations, $x=(-2,-1,0,1,2)$ and $y=(4,1,0,1,4)$.
Calculate covariance and Pearson correlation and describe the scatterplot.
A report says, "The correlation is zero, so the variables are unrelated."
First identify its numerical result and reported conclusion, then evaluate
whether that conclusion follows. What happens to covariance and correlation
if $y$ is multiplied by 10?

<!-- END EX05-05 PROMPT -->

<!-- prompt-id: CH05-E06 -->

### Exercise 6: Report paired differences (Integrative) {.unnumbered #ex-ch05-06}

Four people have condition-1 values 1, 2, 4 and 5 and corresponding
condition-2 values 2, 4, 4 and 8 hours. Calculate the mean and sample SD
of condition 2 minus condition 1, and verify the difference-of-means identity.
Write a descriptive result that names the people, direction and units.
Explain why it does not say that every person gained the mean difference.

<!-- END EX05-06 PROMPT -->

<!-- prompt-id: CH05-E07 -->

### Exercise 7: Simulate a sampling distribution and read its SE (Integrative) {.unnumbered #ex-ch05-07}

The waiting-time simulation in Section 3.2 reports, for $n=30$, a mean
of sample means of 9.96 and an empirical SE of 1.79 against a model SE
of 1.83.

1. Explain what one histogram bar counts and why the empirical and model
   SEs differ.
2. Without running R, give the model SE for $n=120$ and for $n=480$ and
   explain the pattern.
3. Say what changes, and what does not, if the number of repetitions is
   raised from 5,000 to 20,000.
4. In R, change `n_middle <- 30` in the supplied code to 120, rerun
   from the seed, and report the empirical SE to three decimals.
5. The running mean and the histograms of repeated sample means show
   different changes as sample size increases. Explain concentration
   toward a target and increasingly Normal standardized sampling shape,
   and say whether either makes individual waiting times Normal.

<!-- END EX05-07 PROMPT -->

<!-- prompt-id: CH05-E08 -->

### Exercise 8: Name the estimand, estimator, and estimate, then calculate the SE (Integrative) {.unnumbered #ex-ch05-08}

The special-practice study assigned 60 students to each of two formats;
the delayed-assessment means were 27.1 and 24.0 with SDs 6.0 and 6.4.

1. State the estimand, the estimator, and the estimate in one sentence
   each, and name the repetition behind the uncertainty.
2. Calculate the SE of the mean difference in R from these summaries
   and report it to four decimals.
3. For the ten sleep differences (SD 1.23 hours), calculate the SE of
   the mean difference and explain why using the two condition SDs,
   1.79 and 2.00, as if the two sets of condition measurements were
   independent would be a different and larger calculation.
4. A separate simulation estimates a waiting-time variance of 100.
   Across 5,000 samples of five waits, variances using the $n-1$
   denominator average 101.22 and those using $n$ average 80.98. Explain
   which rule is unbiased, why 101.22 need not equal 100 in this finite
   run, and why unbiasedness does not ensure that one estimate is close
   to the target.

<!-- END EX05-08 PROMPT -->

<!-- prompt-id: CH05-E09 -->

### Exercise 9: An interval and its question (Foundational) {.unnumbered #ex-ch05-09}

Twenty-five independent paired differences have mean 3.0 and SD 5.0.
Under the Normal-difference model use $t_{.975,24}=2.064$. Calculate
the estimated SE, 95% interval and statistic for a zero process mean.
Then name the appropriate procedures for (a) the same 20 people measured
twice and (b) independent groups of 20 and 25 without equal variances.
State why the interval is not a range for individual differences.

<!-- END EX05-09 PROMPT -->

\Needspace{14\baselineskip}

<!-- prompt-id: CH05-E10 -->

### Exercise 10: A selected relationship (Integrative) {.unnumbered #ex-ch05-10}

Use the common 32 rows of mpg, hp, and wt. (a) Convert their covariance
matrix to a correlation matrix and reconstruct the covariance matrix
from the correlations and SDs. (b) Report Pearson and Spearman correlations
for mpg/hp and state which association each describes. (c) Compute the
Pearson partial correlation controlling linearly for wt, its classical
t statistic and correct df, and explain why adjustment alone does not
establish a causal effect. (d) For the preselected wt/mpg Pearson test,
report the estimate, 95% interval and the conditions supporting the reference.

<!-- END EX05-10 PROMPT -->

## References {.unnumbered}
