---
title: "Estimates, Uncertainty, and Statistical Tests"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - ../koma_heading_tags.tex
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{12\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \usepackage{xcolor}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false

if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
n_special <- 60; n_usual <- 60
mean_special <- 27.1; mean_usual <- 24.0
sd_special <- 6.0; sd_usual <- 6.4
estimate <- mean_special - mean_usual
pooled_var <- ((n_special - 1) * sd_special^2 +
  (n_usual - 1) * sd_usual^2) / (n_special + n_usual - 2)
pooled_df <- n_special + n_usual - 2
standard_error <- sqrt(pooled_var * (1 / n_special + 1 / n_usual))
test_statistic <- estimate / standard_error
p_value <- 2 * pt(-abs(test_statistic), df = pooled_df)
confidence_interval <- estimate +
  c(-1, 1) * qt(.975, df = pooled_df) * standard_error
interval_90 <- estimate +
  c(-1, 1) * qt(.95, df = pooled_df) * standard_error
welch_df <- (sd_special^2 / n_special + sd_usual^2 / n_usual)^2 /
  ((sd_special^2 / n_special)^2 / (n_special - 1) +
   (sd_usual^2 / n_usual)^2 / (n_usual - 1))
records <- read.csv("chapter_03_special_practice.csv")
special_scores <- records$score[records$group == "Special Practice"]
usual_scores <- records$score[records$group == "Usual Practice"]
ink <- "#243447"; blue <- "#2B6F8A"; blue_light <- "#DDECF2"
gold <- "#B7791F"; red <- "#A23B3B"; grey <- "#6B7280"
## Display helper: round once to four decimals and drop trailing zeros; a
## vector is one set and shares the fewest decimals that show every member;
## lead = FALSE drops the leading zero of a p-value or proportion; minus
## signs are U+2212 (figures pass minus = "-" for a cleanly embedded font).
fmt4 <- function(x, lead = TRUE, minus = "\u2212") {
  r <- round(x, 4)
  out <- format(r, digits = 15, scientific = FALSE, trim = TRUE)
  tiny <- r == 0 & x != 0
  out[tiny] <- ifelse(x[tiny] > 0, "< 0.0001", "> -0.0001")
  if (!lead) out <- sub("^([<>] )?(-?)0[.]", "\\1\\2.", out)
  gsub("-", minus, out)
}
```

{{< pagebreak >}}

::: {.callout-note title="Chapter overview"}
This chapter builds on the ideas Chapter 2 developed: population and
sample, parameter and statistic, sampling distribution and standard
error. We first say precisely what quantity a study seeks and which rule
estimates it, and name the repetition behind its uncertainty. We then
calculate standard errors for a mean and for a difference between two
group means, build confidence intervals and read their confidence level,
and carry out hypothesis tests, reading p-values and statistical
significance. Each idea appears first with one mean and then with the
continuing two-group comparison. The t tests for one sample, paired
samples and two independent samples follow as one family: the
equal-variance test used for the continuing study, then Welch's test for
groups whose spreads may differ. The same test logic then carries over
to the association between two categorical variables, with the
chi-square distribution as its reference, and a three-way table is read
as two-way tables within layers. We close by weighing the size of a
difference, the errors a
decision rule can make and the choices made after seeing results, so
that one qualified conclusion can be written.
:::

A first-year course wanted to know whether a new practice format helps
students remember what they have learned. Its instructors recruited 120
consenting students and randomly assigned 60 to Special Practice and 60
to Usual Practice. Both formats covered the same content over four
sessions with equal total study time. Two weeks later, all 120 completed
the same assessment, scored from 0 to 40. Special Practice averaged 27.1
points and Usual Practice 24.0, a difference of 3.1 points. This study is
invented for the chapter.

Is 3.1 points a difference we should trust, and is it large enough
to matter?

Section 1 names the target and the repetition behind its uncertainty;
Sections 2, 3 and 4 then calculate a standard error, build a confidence
interval and carry out a hypothesis test, each meeting its idea first
with one mean and then with the practice study. Section 5 gathers the
t tests into one family, Section 6 carries the same logic to the
association between two categorical variables, and Section 7 returns to
whether 3.1 points is large enough to matter.

## From a Sample to an Estimate {#sec-ch03-estimation}

### The study and its estimated difference

The study asks about the **mean effect of assignment to Special rather
than Usual Practice on the delayed assessment among these 120
participants**. Each part matters. Assignment is the intervention being
compared; attending every session would be a different comparison. A mean
effect summarizes the participants and allows individual responses to
differ. The two-week assessment measures a specified outcome at a
specified time. An effect on durable learning or among students elsewhere
would require additional information.

The course team also chose a practical benchmark before seeing the
outcomes. It treated two points as the smallest average difference that
would change its recommendation: the marking scheme awards two points
for one completed application task, and the new format adds little
preparation time. The benchmark is course-specific, and Section 7
returns to it after uncertainty has been calculated.

| Item | Special Practice | Usual Practice |
|:--|--:|--:|
| Assigned students | 60 | 60 |
| Delayed assessments observed | 60 | 60 |
| Delayed-assessment mean | 27.1 | 24.0 |
| Delayed-assessment SD | 6.0 | 6.4 |

: Observed summaries for the invented special-practice study. {#tbl-ch03-observed tbl-colwidths="[50,25,25]"}

The standard deviations describe differences among students' scores
within each group. @fig-ch03-scores draws the study's own 120 records,
so that the 3.1-point shift between the means can be set against the
variation among students that makes an uncertainty calculation
necessary. It shows why the difference needs more than arithmetic.
The two groups overlap heavily: 17 of the 60 Usual Practice students
scored at or above the Special Practice median of 27, and scores within
each group spread over almost 30 points, with SDs of about six. Against
that spread the means differ by 3.1 points. A different assignment
of the same students would have placed different students in each
group and produced a different difference, so the next sections ask
how far such differences would vary. The figure also lets us check the
records directly: no single score stands far from the rest, and the
three scores beyond the whiskers lie within two points of the whisker
ends. Measuring that variation begins by naming the target and the
rule used to estimate it.

```{r}
#| label: fig-ch03-scores
#| fig-cap: "Delayed scores of the 120 students by assigned format. Each dot is one student; boxes span the quartiles, whiskers reach the most extreme scores within 1.5 IQR of the box, open circles mark the three scores beyond them, and diamonds mark the group means, 27.1 and 24.0 points. The figure draws the chapter's 120 invented teaching records, whose SDs, 5.9992 and 6.3992, round to the 6.0 and 6.4 of @tbl-ch03-observed; the chapter's calculations use those rounded summaries."
#| fig-alt: "Two horizontal boxplots with stacked dots on a score axis from 10 to 40, Special Practice above Usual Practice. Special Practice has quartiles 24, 27 and 31, whiskers from 14 to 40 and one low score of 13 beyond them; Usual Practice has quartiles 20, 24 and 27, whiskers from 11 to 37 and scores of 38 and 39 beyond them. A red bracket marks the 3.1-point gap between the group means, small beside the spread of each group's scores."
#| fig-width: 7.2
#| fig-height: 2.8
#| fig-pos: "!ht"

local({
  stack_dots <- function(x, base, step = 0.07) {
    x <- sort(x)
    height <- ave(x, x, FUN = seq_along)
    points(x, base + step * height, pch = 16, cex = 0.62, col = blue)
  }
  saved_par <- par(mar = c(4.1, 7.2, 0.6, 0.8))
  on.exit(par(saved_par))
  plot(NA, xlim = c(10, 41), ylim = c(0.45, 3.2), yaxt = "n",
       xlab = "Delayed assessment score (points, 0-40 scale)", ylab = "",
       col.axis = ink, col.lab = ink)
  boxplot(list(usual_scores, special_scores), horizontal = TRUE,
          at = c(0.85, 2.05), boxwex = 0.42, add = TRUE, axes = FALSE,
          col = blue_light, border = ink, outcol = ink, outcex = 0.8)
  stack_dots(usual_scores, 1.10)
  stack_dots(special_scores, 2.30)
  points(c(mean(usual_scores), mean(special_scores)), c(0.85, 2.05),
         pch = 23, bg = gold, col = ink, cex = 1.3)
  axis(2, at = c(1.2, 2.4), labels = c("Usual\nPractice", "Special\nPractice"),
       las = 1, tick = FALSE, col.axis = ink)
  arrows(mean(usual_scores), 0.52, mean(special_scores), 0.52,
         code = 3, angle = 90, length = 0.04, col = red, lwd = 1.6)
  text(mean(special_scores), 0.52, "means 3.1 points apart", pos = 4,
       col = red, cex = 0.75)
})
```

### Estimand, estimator, estimation, and estimate

Chapter 2 distinguished the population and sample, introduced parameter
and statistic, and explained sampling distributions and standard error.
We now use those ideas to distinguish the quantity sought, the
estimation rule, and the estimate obtained.

An **estimand** is the precisely specified quantity the question seeks.
For this study, it is the mean assignment effect just described. To make
the target explicit, let $Y_{i1}$ be student $i$'s delayed score if
assigned to Special Practice and $Y_{i0}$ the score if assigned to Usual
Practice. These are the student's two **potential outcomes**. Only the
one corresponding to the actual assignment is observed; Chapter 1 called
the unobserved alternative the counterfactual outcome. The target is

$$
\Delta\mu=\frac{1}{n}\sum_{i=1}^{n}\{Y_{i1}-Y_{i0}\}.
$$

Inside the braces is one student's assignment effect. The sum adds all
effects, and division by $n=120$ takes their mean. Written this way, $\Delta\mu$ is a mean
difference over these 120 participants, not a population quantity. The notation allows different students to
have different effects. We assume the two formats have a consistent
meaning and one student's assignment does not change another student's
outcome in a relevant way.

An **estimator** is a rule for using observed data to approximate this
target. Here the rule subtracts the Usual assigned-group mean from the
Special assigned-group mean. **Estimation** is the process of applying
and checking that rule and characterizing its uncertainty. The
**estimate** is the numerical value obtained:

$$
\widehat{\Delta\mu}=\bar y_1-\bar y_0=27.1-24.0=3.1\text{ points}.
$$

The bar denotes a group mean, and the hat marks an estimate. The rule
would stay the same under another admissible assignment, although the
observed outcomes and the two group means could change. Reversing the
subtraction gives $-3.1$; it expresses the reverse comparison using the
same information.

The observed difference exactly describes the two realized means. That
description is known. The participant mean assignment effect remains
unknown because no student reveals both potential outcomes. This is why
an exact piece of arithmetic can still be an uncertain estimate.

| Object | Working definition | In this study | Common reading error |
|:--|:--|:--|:--|
| Estimand | The precisely specified quantity the question seeks | Mean delayed-score difference caused by Special rather than Usual assignment among the 120 randomized participants | Naming only "the treatment effect" |
| Estimator | The rule that would be applied in each admissible repetition | Special-group mean minus Usual-group mean | Treating the rule as the observed result |
| Estimation | Applying and checking the estimator for the specified estimand and characterizing the estimate and its uncertainty | Apply the mean-difference rule, verify its inputs and conditions, and quantify uncertainty | Treating estimation as another number |
| Estimate | The number the estimator produced in this study | $27.1-24.0=3.1$ points | Reporting a number without units or direction |
| Parameter | The fixed feature named by the estimand under the chosen population or process interpretation | The participant-specific mean assignment contrast | Assuming it is the sample estimate or a universal truth |
| Statistic | A numerical function of observed data | A group mean, the 3.1-point estimate, or the 2.7372 test statistic | Assuming every statistic estimates the same thing |
| Standard error | The modeled spread of an estimator across the declared repetitions | 1.1325 points for the mean-difference estimator | Calling it variation among students |

: Inferential objects in the special-practice study. {#tbl-ch03-inferential-objects tbl-colwidths="[13,29,33,25]"}

A parameter is fixed within the specified population or process
interpretation, but fixed does not mean universal: changing the
participant set, outcome, time, or condition comparison may change the
estimand and its parameter.

### Name the repetition and working model

Uncertainty needs a description of what could have varied. In a sampling
question, different people might have been selected from a population.
In the special-practice study, the primary target concerns these same
120 participants. The relevant repetition assigns exactly 60 of them to
each format again. Their potential outcomes remain fixed; the assigned
groups and observed outcomes can change.

Random assignment makes the difference of assigned-group means useful
for estimating the participant effect. It does not turn these consenting
students into a representative sample of all first-year students.
Inference about a wider population would need a stated model or sampling
justification and an explicit extension of the target.

The numerical analysis uses a familiar working model: two independent
groups whose scores share one population spread. Under that model the
two observed variances estimate the same quantity, so Section 2 combines
them into one pooled estimate, and the interval and test use a t
reference with $60+60-2=118$ degrees of freedom. The observed SDs, 6.0
and 6.4 points, are similar, and the two formats differ only in the
practice activity, so a common spread is a reasonable starting
assumption. Section 5 shows what changes when the groups' spreads or
sizes differ.

The model's formulas come from independent sampling, but complete
assignment of 60 of the 120 students does not make the two
assigned-group means independent. The model therefore serves as an
approximate calibration for this design rather than an exact
randomization distribution. Section 2 explains what the familiar
formula estimates here, and [Appendix A](#app-ch03-assignment) gives the
design-based variance identity and the distinction between an
average-effect null and a sharp null.

### Expectation, bias, and precision

Imagine applying the estimator over all repetitions allowed by the
design or model. Its expectation, in Chapter 2's sense, is the probability-weighted
average of the values it would produce across those repetitions. If an estimator averages 3.4 points when
the target is 3.0, its bias is $3.4-3.0=0.4$ points. This describes the
rule's systematic displacement, not the error of one particular result.

Under complete random assignment and the stated potential-outcome
conditions, each assigned-group mean is centered on the corresponding
participant mean potential outcome. Their difference is therefore
unbiased for the participant mean assignment effect:

$$
\operatorname{Bias}(\bar Y_1-\bar Y_0)
=E_{\mathrm{assign}}(\bar Y_1-\bar Y_0)-\Delta\mu=0.
$$

The equality does not say that this estimate, 3.1, equals the unknown
effect. **Precision** concerns how tightly the estimator's values cluster
across assignments:

$$
SE(\bar Y_1-\bar Y_0)=SD(\bar Y_1-\bar Y_0\text{ across possible assignments}).
$$

An unbiased rule can be imprecise; a tightly concentrated rule can be
biased. A third useful consideration is how its behavior changes under
unusual observations or departures from its conditions. A small SE under
an unsuitable model can give unwarranted confidence.

[]{#app-ch03-mse}

Accuracy combines both qualities. The mean squared error, the average
squared distance of an estimator from its estimand over the
repetitions, adds the variance and the squared bias,
$\operatorname{MSE}(\widehat\theta)=\operatorname{Var}(\widehat\theta)+\operatorname{Bias}(\widehat\theta)^2$.
An unbiased estimator with SD 2 has MSE $2^2=4.00$, while one with SD
1 and bias 1.5 has MSE $1+1.5^2=3.25$: under this criterion the
biased but more precise rule lands closer on average, a verdict that
depends on the criterion chosen.

## Standard Errors {#sec-ch03-se}

### From the visible sampling pattern to the SE of a mean

Chapter 2, Sections 4.2--4.4, drew the pattern this section explains.
Its waiting-time model had mean and SD both equal to 10 minutes, and its
sampling-means figure set the individual waits beside 5,000 sample
means at each of three sample sizes, 5, 30 and 100. The individual
waits kept their 10-minute spread at every size, while the sample means
varied less when each sample contained more observations. Increasing
the number of simulated repetitions made their distribution easier to
see; increasing the size of each sample changed its spread.

We can now explain that spread algebraically. Suppose $Y_1,\ldots,Y_n$
are independent observations from the same distribution, with finite
variance $\sigma^2$. The sum has variance $n\sigma^2$, because
independent variances add. The mean divides that sum by $n$; multiplying
a variable by $1/n$ multiplies its variance by $1/n^2$. Thus

$$
\operatorname{Var}(\bar Y)=\frac{n\sigma^2}{n^2}=\frac{\sigma^2}{n},
\qquad SE(\bar Y)=\frac{\sigma}{\sqrt n},
\qquad \widehat{SE}(\bar Y)=\frac{s}{\sqrt n}.
$$

The last expression estimates the unknown population SD with the sample
SD. The estimated SE is calculated from one sample; the SE it estimates
describes the mean across repetitions. Both are standard deviations,
but of different distributions from the individual waiting times.

Applied to the waiting-time model, $\sigma/\sqrt n$ gives the model
SEs of Chapter 2's sampling table, which the simulated sample means
approximate, and its square root explains why halving an SE takes four
times as many independent observations: at $n=120$ the model SE is
0.9129 minutes, half its value at $n=30$.

The estimated SE uses the sample SD $s$, whose squared deviations from
the sample mean are divided by $n-1$. Chapter 2 noted that the sample
mean is the value that makes the sum of squared deviations smallest;
the deviations from it sum to zero, so once $n-1$ of them are known the
last is fixed. [Appendix A](#app-ch03-variance) derives both facts and
shows why the divisor $n-1$ makes $s^2$ unbiased for $\sigma^2$.

### Uncertainty for a difference in means {#sec-ch03-pooled}

For two independent samples, uncertainty comes from both means. Their
variances add when one mean is subtracted from the other:

$$
\operatorname{Var}(\bar Y_1-\bar Y_0)
=\frac{\sigma_1^2}{n_1}+\frac{\sigma_0^2}{n_0}.
$$

The minus sign changes the direction of a difference, not the sign of a
variance contribution. Independence removes the covariance term; the
next subsection shows what happens when that condition fails.

Section 1.3 stated the working model for the practice study: two
independent groups whose scores share one population spread. If both
populations have the same variance $\sigma^2$, the two contributions
share that factor:

$$
\sigma_1^2=\sigma_0^2=\sigma^2:\quad
\operatorname{Var}(\bar Y_1-\bar Y_0)
=\sigma^2\left(\frac{1}{n_1}+\frac{1}{n_0}\right).
$$

Each group's sample variance then estimates the same $\sigma^2$. The
**pooled variance** $s_p^2$ combines them, weighting each by its degrees
of freedom, $n-1$; its square root $s_p$ is the pooled SD:

$$
s_p^2=\frac{(n_1-1)s_1^2+(n_0-1)s_0^2}{n_1+n_0-2},
\qquad
\widehat{SE}(\bar y_1-\bar y_0)=s_p\sqrt{\frac{1}{n_1}+\frac{1}{n_0}}.
$$

A larger group estimates the common spread from more observations, so
its variance receives more weight. The denominator $n_1+n_0-2$ counts
the degrees of freedom that remain after estimating the two group means.
For the practice study the pooled estimate therefore uses $60+60-2=118$
df, which is why the interval in Section 3 and the test in Section 4 use
a t reference with 118 df. Substitution gives

$$
\begin{aligned}
s_p^2&=\frac{59(6.0^2)+59(6.4^2)}{118}=38.48,
\qquad s_p=6.2032\text{ points},\\
\widehat{SE}(\bar y_1-\bar y_0)&=6.2032\sqrt{\frac{1}{60}+\frac{1}{60}}
=1.1325\text{ points}.
\end{aligned}
$$

With 60 students in each group the pooled variance is simply the
average of the two variances, in squared points, and the pooled SD lies
between the two group SDs. The pooled SE is then algebraically the same
as the square root of the two estimated mean variances added,
$\sqrt{s_1^2/n_1+s_0^2/n_0}$, so both routes give 1.1325 points. With
unequal group sizes the two forms generally differ, and Section 5.4
returns to that case. Equal group sizes make the two SEs coincide; they
do not by themselves establish equal population variances.

For the fixed 120 participants, the SE describes how the estimate would
vary across the possible assignments of 60 students to each format.
Re-assigning these records under the null of no effect, as Section 4.4
describes and the R support page runs, gives differences whose spread,
1.1671 points, is close to this 1.1325. The exact
assignment variance subtracts a term for the variability of individual
effects, which no record reveals, so the familiar estimated
variance, and with equal group sizes the pooled squared SE, estimates
an upper bound on that variance on average.
[Appendix A](#app-ch03-assignment) gives the identity [@bai2024]. The
working t reference also needs adequate information in each group and
no dominating observations; random assignment alone does not make it
exact.

### Pairing, clustering, and sampling without replacement

The independence calculation fails in several common designs, and each
failure is a covariance story. For any two quantities with finite
variances,

$$
\operatorname{Var}(U-V)=\operatorname{Var}(U)+\operatorname{Var}(V)
-2\operatorname{Cov}(U,V).
$$

If paired measurements tend to be high or low together, subtracting
within a pair removes some shared variation. In a separate illustration,
suppose both measurements have SD 4, their covariance is 12, and there
are 25 independent pairs. Let $d$ be the within-pair difference. Then

$$
\begin{aligned}
\operatorname{Var}(d)&=4^2+4^2-2(12)=8,\\
\widehat{SE}(\bar d)&=\sqrt{\frac{8}{25}}=0.5657,\\
\text{ignoring the pairing: }&\sqrt{\frac{16+16}{25}}=1.1314.
\end{aligned}
$$

The reduction comes from the positive covariance, not from a change in
the number of observations. Actual paired-data analysis computes each
difference and uses the SD of those differences. Negative covariance
would change the result in the opposite direction.

[]{#app-ch03-dependence}

Sampling without replacement from a fixed list also changes the
variance, even when selection is random. For a simple random sample of
$n$ units from $N$ fixed values whose variance $\sigma^2$ uses the
divisor $N$,

$$
SE(\bar Y)=\frac{\sigma}{\sqrt n}\sqrt{\frac{N-n}{N-1}}.
$$

The **finite-population correction** $\sqrt{(N-n)/(N-1)}$ shrinks the
SE as the sample covers more of the list, reaching zero at a census. In
Chapter 2's invented system of $N=20{,}000$ students with
$\sigma=4.4259$ grade points, a simple random sample of 100 has SE
0.4426 grade points under independent sampling and 0.4415 with the
correction factor .9975; with 400 students the factor is .9900. Other
selection schemes need their own design-based calculation.

Clustering works in the opposite direction. Chapter 2, Section 4.4,
showed that a sample of 4 schools with 25 students each spreads its
means about twice as widely as a simple random sample of 100 students,
and introduced the design effect,

$$
\mathrm{deff}=1+(m-1)\,\mathrm{ICC},
$$

for a mean from equal clusters of $m$ units. The covariance terms
explain it: each student has $m-1$ school-mates whose scores covary
with theirs by $\mathrm{ICC}\times\sigma^2$, and those terms multiply
the independent-sampling variance of the mean by $\mathrm{deff}$. With
$m=25$ and Chapter 2's intraclass correlation .1304, the design effect
is 4.1299. The SE of the multistage mean, the simple-random-sample SE
multiplied by the square root of the design effect, is then about
0.8972 grade points, close to the 0.8730 that Chapter 2's simulation
showed. The 100 students came from only four sampled schools; the SE
follows the design, not the count of students.

[]{#sec-ch03-school-pairs}

An invented school-blocked extension of Chapter 1's feedback study
shows pairing and clustering together. Each of twelve participating
schools contributed two classes of 20 students, and within every school
one class was randomly assigned to explanatory feedback and the other
to usual feedback; all 480 students took the same delayed 0--40
assessment. The class is the unit of assignment, and each school
supplies one explanatory-minus-usual difference of its two class means.
@fig-ch03-school-pairs shows the twelve differences, $-2$, 0, 1, 1, 2,
2, 3, 3, 4, 4, 5 and 7 points, with mean 2.5 and SD 2.3932. Their
uncertainty comes from these twelve independent within-school
comparisons: $\widehat{SE}(\bar d)=2.3932/\sqrt{12}=0.6908$ points.
Dividing the same SD by $\sqrt{480}$, as if the 480 students rather
than the twelve school comparisons were the independent units, would
give 0.1092 points, an SE $\sqrt{40}=6.3246$ times too small. That
calculation mixes levels: the SD describes class-mean differences,
whereas 480 counts students, who were assigned in whole classes.
Section 5.2 completes the paired analysis of the twelve differences.

```{r}
#| label: fig-ch03-school-pairs
#| fig-cap: "Twelve school differences in class means under within-school assignment (invented data). Each dot is one school's explanatory-minus-usual difference; the vertical line marks their mean, 2.5 points. The upper bar spans one SE either side of the mean from the twelve school comparisons, 0.6908 points; the lower bar spans the 0.1092 that wrongly counting the 480 students as the independent units would give."
#| fig-alt: "A dot plot of twelve differences from minus 2 to 7 points, stacked where values repeat, with a vertical line at 2.5 and a dashed line at zero. Below the dots a blue bar from about 1.8 to 3.2 marks one SE either side of the mean from the twelve school comparisons, and a much shorter red bar from about 2.4 to 2.6 marks the SE that wrongly treating the 480 students as the independent units would give."
#| fig-width: 7.2
#| fig-height: 2.0
#| fig-pos: "!ht"

local({
  school_difference <- c(-2, 0, 1, 1, 2, 2, 3, 3, 4, 4, 5, 7)
  se_schools <- sd(school_difference) / sqrt(12)
  se_students <- sd(school_difference) / sqrt(480)
  saved_par <- par(mar = c(4.0, 1.0, 0.6, 1.0))
  on.exit(par(saved_par))
  plot(NA, xlim = c(-3, 8), ylim = c(0.3, 2.6), yaxt = "n",
       xlab = "Explanatory-minus-usual difference in class means (points)",
       ylab = "", col.axis = ink, col.lab = ink)
  abline(v = 0, col = grey, lty = 2)
  height <- ave(school_difference, school_difference, FUN = seq_along)
  points(school_difference, 1.55 + 0.28 * height, pch = 16, cex = 1.3,
         col = blue)
  abline(v = mean(school_difference), col = gold, lwd = 1.6)
  segments(2.5 - se_schools, 1.05, 2.5 + se_schools, 1.05, col = blue, lwd = 4)
  segments(2.5 - se_students, 0.55, 2.5 + se_students, 0.55, col = red, lwd = 4)
  text(2.5 + se_schools, 1.05,
       paste("+/- 1 SE from 12 schools:", fmt4(se_schools)),
       pos = 4, cex = 0.75, col = blue)
  text(2.5 + se_schools, 0.55,
       paste("+/- 1 SE if 480 students were counted:", fmt4(se_students)),
       pos = 4, cex = 0.75, col = red)
})
```

This is a **blocked assignment**: each school forms a block, and the
two conditions are allocated within every block, so each comparison is
made inside a school and school-level variation drops out of it.
**Stratified sampling** instead selects units within population
strata; it decides who enters a sample, whereas blocking allocates
conditions after entry, and grouping records by school does not make
either mechanism imply the other.

### Concentration, shape, and the conditions behind them

[]{#app-ch03-limits}

Chapter 2's sampling-means figure shows two different things happening
as the sample size grows. The **law of large numbers** describes
concentration: for independent, identically distributed observations
with a finite mean, the chance that $\bar Y_n$ lies farther than any
fixed distance from $\mu$ shrinks toward zero. The **central limit
theorem** describes shape: with finite variance $\sigma^2$, the
standardized mean approaches a standard Normal distribution,

$$
\frac{\bar Y_n-\mu}{\sigma/\sqrt n}\ \text{approaches}\ N(0,1)
\quad\text{as } n \text{ grows}.
$$

The individual waiting times in the figure's first panel stay skewed
at every sample size; only the means concentrate and become nearly
Normal. Both results concern increasing sample size, not repeatedly
simulating the same small sample, and both need conditions: strong
dependence, extreme tails, or an observation that dominates the mean
can make a Normal approximation poor, and no universal sample-size
cutoff settles the issue for every distribution and design. The SE
formula answers how wide a sampling distribution is; its approximate
shape then helps choose an interval or test.

## Confidence Intervals and Confidence Level {#sec-ch03-intervals}

### Estimate, SE, critical value, and interval

A confidence interval combines an estimate with a multiple of its SE.
For the symmetric procedures used here,

$$
\text{estimate}\pm\text{critical value}\times\text{standard error}.
$$

The SE supplies a scale. A **critical value** says how far to extend in
SE units to obtain the intended coverage under a reference distribution.
Chapter 2 introduced the standard Normal distribution, centered at zero
with SD one. Its .975 quantile is about 1.96: .025 of the probability
lies beyond that point, and another .025 lies below $-1.96$.

The Normal critical value suits a mean whose population SD is known
independently of the sample. Usually the SD must be estimated from the
same observations, and that estimate is itself uncertain. A **t
distribution** accounts for this extra uncertainty. It is symmetric
around zero, like the standard Normal, but has heavier tails. Its
**degrees of freedom** (df) index how much information the estimated SD
carries; for one mean they equal $n-1$, the divisor of the sample
variance ([Appendix A](#app-ch03-variance)).
With few df the tails are heavy and the critical value is larger; as df
increase, the t reference approaches the standard Normal. An estimated
spread does not make every statistic exactly t-distributed: the exact
one-mean result requires independent Normal sampling, and the two-group
reference for the practice study is an approximate working calibration
(Section 1.3).

For a separate, invented illustration, suppose a simple random sample
of 25 students from a large district completes a reading scale whose
published reference mean is 50 points. The sample mean is $\bar x=52$
points and the SD is $s=5$ points. The estimand is the district's
population mean score, $\mu$. Treating the scores as independent
observations from an approximately Normal population, Section 2.1
gives $\widehat{SE}(\bar X)=5/\sqrt{25}=1$ point. Because the SD
is estimated from the same 25 scores, the reference is a t
distribution with $25-1=24$ degrees of freedom. Its .975 quantile,
2.064 in a t table, leaves .025 in each tail:

$$
52\pm2.064(1)=52\pm2.064=[49.936,\ 54.064]\text{ points}.
$$

The interval estimates the district mean in points under the sampling
model. Its estimand and repetition differ from the participant
assignment effect of the practice study.

The 95% reference values below leave .025 in each tail. The table gives
full-precision quantiles, so the t-table value 2.064 appears as 2.0639,
and its third row previews the Special Practice reference of Section
3.4.

| Reference | 95% critical value | Role in the examples |
|:--|--:|:--|
| Standard Normal | 1.9600 | Mean with a population SD known independently of the sample, under a suitable Normal model |
| t with 24 df | 2.0639 | The one-mean illustration: SD estimated from 25 independent Normal observations |
| t with 118 df | 1.9803 | The Special Practice comparison under the equal-variance working model (Section 3.4) |

: Reference choice and critical values. {#tbl-ch03-critical-values tbl-colwidths="[25,20,55]"}

The t curves in @fig-ch03-z-t spread more probability into their tails
than the standard Normal, most visibly with few degrees of freedom,
which is why their critical values are larger.

```{r}
#| label: fig-ch03-z-t
#| fig-pos: "!ht"
#| results: hide
#| fig-cap: "Normal and t references at the same location and scale of the test statistic. The heavier t tails are most visible with few degrees of freedom."
#| fig-alt: "Overlaid standard Normal and t curves with 9 and 99 degrees of freedom. Vertical lines mark minus and plus 2; the t curve with 9 degrees of freedom has the largest probability beyond them."
#| fig-width: 6.8
#| fig-height: 2.9
z_grid <- seq(-4, 4, length.out = 801)
old <- par(mar = c(4.1, 4.1, 0.6, 0.8))
plot(z_grid, dnorm(z_grid), type = "l", lwd = 2,
     xlab = "Standardized test statistic", ylab = "Density")
lines(z_grid, dt(z_grid, 9), col = "#A23B3B", lwd = 2)
lines(z_grid, dt(z_grid, 99), col = "#2B6F8A", lwd = 2, lty = 2)
abline(v = c(-2, 2), lty = 3)
legend("topright", c("Normal", "t: df = 9", "t: df = 99"),
       col = c("black", "#A23B3B", "#2B6F8A"),
       lty = c(1, 1, 2), lwd = 2, bty = "n")
par(old)
c(Normal = 2 * pnorm(-2), t9 = 2 * pt(-2, 9),
  t99 = 2 * pt(-2, 99))
```

The procedure and its conditions determine the reference. Choosing the
smaller critical value merely to narrow an interval would change the
procedure without justification.

### Confidence level and coverage

The **confidence level** names the procedure's target coverage: a 95%
procedure aims to cover the fixed estimand in 95% of its repeated
applications under its conditions. The level enters the calculation
through the critical value. For the district sample, a 90% interval
uses the t quantile that leaves .05 in each tail, 1.711 in a t table,
instead of 2.064.

Predict the change before calculating. The estimate and its SE stay the
same and only the multiplier becomes smaller, so the 90% interval must
be narrower, as @tbl-ch03-confidence-levels in Section 3.4 confirms.
The 95% interval contains the published reference value 50, whereas the
90% interval, [50.289, 53.711], excludes it; Section 4.3 uses this
pair to connect intervals with tests.

What does the target coverage describe? @fig-interval-coverage
simulates 40 district samples under a stipulated population mean of 52
points and a known SE of 1, and gives each estimate an interval of
$\pm1.9600$ SEs at 95% and $\pm1.6449$ SEs at 90%.

```{r}
#| label: fig-interval-coverage
#| fig-cap: "The same 40 simulated district samples under two confidence levels. The population mean, 52 points, is fixed for display, and each sample's estimate carries its interval with it. At 95%, 38 of the 40 intervals cover the mean; at 90%, with the smaller multiplier 1.6449 in place of 1.9600, 36 do."
#| fig-alt: "Two panels of 40 horizontal intervals around a vertical line at 52 points, with the same estimates in both panels. In the 95 percent panel, studies 11 and 22 miss the line and are drawn in red; in the shorter 90 percent intervals, studies 2, 11, 14 and 22 miss."
#| fig-width: 7.2
#| fig-height: 3.1
#| fig-pos: "!ht"

local({
  set.seed(30306)
  one_mean <- 52; one_mean_se <- 1
  coverage_estimate <- rnorm(40, mean = one_mean, sd = one_mean_se)
  saved_par <- par(mfrow = c(1, 2), mar = c(4.4, 3.6, 2.0, 0.6),
                   mgp = c(2.5, 0.7, 0))
  on.exit(par(saved_par))
  for (level in c(.95, .90)) {
    multiplier <- qnorm(1 - (1 - level) / 2)
    lower <- coverage_estimate - multiplier * one_mean_se
    upper <- coverage_estimate + multiplier * one_mean_se
    hit <- lower <= one_mean & one_mean <= upper
    plot(NA, xlim = c(48.6, 55.4), ylim = c(0.5, 40.5),
         xlab = "Interval for the population mean (points)",
         ylab = "Repeated study", yaxt = "n", col.axis = ink, col.lab = ink,
         main = sprintf("%d%% intervals: %d of 40 cover",
                        round(100 * level), sum(hit)),
         cex.main = 0.9, font.main = 1, col.main = ink)
    abline(v = one_mean, col = gold, lwd = 2)
    segments(lower, seq_len(40), upper, seq_len(40),
             col = ifelse(hit, blue, red), lwd = 1.4)
    points(coverage_estimate, seq_len(40), pch = 16, cex = 0.5,
           col = ifelse(hit, blue, red))
    axis(2, at = c(1, 10, 20, 30, 40), las = 1, col.axis = ink)
  }
})
```

Across repetitions the samples, estimates and endpoints vary while the
target stays fixed. A miss involves no calculation error: an estimate
that happens to fall far from the population mean carries its interval
with it, so even a valid 95% procedure misses occasionally. The two
panels of @fig-interval-coverage use the same 40 estimates, and only
the multiplier changes. Studies 2 and 14, whose estimates fell between
1.6449 and 1.9600 SEs from 52, are covered at 95% but missed at 90%: a
lower level buys a narrower interval at the price of more misses. A
batch of 40 need not show exactly 38 or 36 successes; the target
coverage describes the procedure over many repetitions. Because the
display uses a simplified Normal model with known SE, its coverage is a
property of the illustration, and fixing the mean at 52 is a display
choice rather than a claim about the unknown district mean.

Once an interval has been calculated, the coverage statement describes
the success rate of the *interval-producing procedure*; it does not
assign a 95% probability to the fixed estimand, and it does not say that
95% of the district's students score between 49.936 and 54.064. Read
the endpoints in points as values compatible with the observations under
the procedure: the 95% interval allows district means slightly below
the reference value 50 as well as means near 54 points, with no natural
cliff at its rounded endpoints.

### What a narrower interval can and cannot resolve

Holding the model and data fixed, a higher confidence level needs a
wider interval. More outcome variability generally widens it, while
more relevant information narrows it, and changes in the design or
variance estimator can also change its width. The 90% interval of
Section 3.2 is narrower because it asks for less coverage, not because
the sample holds more information: changing the level adds no
observations and changes neither the estimate nor its SE. An interval
also quantifies only the uncertainty its procedure represents. A narrow
interval can coexist with systematic selection, poor measurement,
unmodeled dependence, unavailable outcomes, or an unsupported
generalization. Measuring the wrong outcome very consistently can yield
a small SE for the wrong target, and adding many pupils from a few
jointly assigned classes does not reproduce the information from the same number of individually assigned pupils. Precision helps only when the
procedure remains aligned with the question and design.

### The Special Practice interval

The same construction applies to the continuing study. Its estimate is
3.1 points, Section 2.2 gave the pooled SE 1.1325 points, and the
equal-variance t reference has 118 df, whose 95% critical value is
1.9803. Under the equal-variance working model, whose calibration is
approximate for the assignment design (Section 1.3), the interval is

$$
(\bar y_1-\bar y_0)\pm t_{.975,\,118}\,\widehat{SE}(\bar y_1-\bar y_0)
\approx3.1\pm1.9803(1.1325)
\approx[0.8572,\ 5.3428]\text{ points}.
$$

The endpoints retain the Special-minus-Usual direction and
assessment-point units. They allow positive effects near one point and
effects above five points. The data therefore give a useful estimate
without pinpointing its magnitude.

Changing only the confidence level gives the intervals below; in each
example the estimate, the SE and the observations stay the same.

| Example and confidence level | Critical value | Margin | Interval in points | Full width |
|:--|--:|--:|:--|--:|
| District mean, 90% | 1.711 | 1.7110 | [50.2890, 53.7110] | 3.4220 |
| District mean, 95% | 2.064 | 2.0640 | [49.9360, 54.0640] | 4.1280 |
| Special Practice difference, 90% | 1.6579 | 1.8776 | [1.2224, 4.9776] | 3.7552 |
| Special Practice difference, 95% | 1.9803 | 2.2428 | [0.8572, 5.3428] | 4.4855 |

: Changing only the confidence level: the estimate and SE stay fixed, and the critical value alone moves each interval. {#tbl-ch03-confidence-levels tbl-colwidths="[40,13,11,22,14]"}

The district critical values are the supplied t-table values; the
Special Practice values are full-precision quantiles of t with 118 df.
In both examples the smaller critical value narrows the 90% interval
without adding observations or reducing the SE; the narrower interval
has a lower coverage target. Both Special Practice intervals lie above
zero and span the two-point benchmark: they can support a positive mean
assignment effect under the procedure while leaving its practical
importance to Section 7.

## Hypothesis Tests, p-Values and Significance {#sec-ch03-tests}

### Hypotheses, test statistics, and reference distributions

A statistical hypothesis makes a proposition about a target. Return to
the district sample. A natural question asks whether the district's
mean differs from the published reference value. A **null hypothesis**
states the reference proposition, $H_0:\mu=50$; the two-sided
**alternative hypothesis** $H_A:\mu\ne50$ allows a difference in either
direction. A **test statistic** measures the estimate's departure from
the null value in SE units:

$$
t=\frac{\bar x-\mu_0}{s/\sqrt n}=\frac{52-50}{5/\sqrt{25}}=\frac{2}{1}=2,
\qquad df=24.
$$

The sample mean lies two estimated SEs above 50. If the null and the
sampling model hold, the statistic follows a t distribution with 24
degrees of freedom, its **reference distribution**.

Many test statistics share this common form:

$$
\text{test statistic}=
\frac{\text{estimate}-\text{null value}}
{\text{standard error under the test procedure}}.
$$

The numerator measures departure in outcome units. The denominator
expresses how large that departure is relative to the procedure's
uncertainty. The reference distribution then shows which values of the
statistic would be common, and which unusual, if the null and the
procedure's conditions held. Equivalently, subtracting 50 from every
score creates a centered variable whose null mean is zero; the question
stays the same. A **test** is the procedure for assessing the null
proposition, while a test result is the numerical output obtained from
the data.

### Tail areas, p-values, and statistical significance

The p-value comes from a tail area of the reference distribution: the
area beyond the observed statistic, in the direction or directions the
alternative names. The district alternative counts departures in both
directions, so the two tails beyond $\pm2$ are added:

$$
p=P(T_{24}\le-2)+P(T_{24}\ge2)=2P(T_{24}\ge2)=.0569,
\qquad
P(T_{24}\ge2)=.0285.
$$

The symbol $T_{24}$ denotes a variable with the reference t distribution
on 24 df. Software or a reference table supplies the area; each tail
holds .0285.

A **p-value** is a conditional tail probability. If the null and the
procedure's conditions hold, a result at least as far from the null
value as the one observed would occur with probability .0569 here. A
**significance level**, $\alpha$, is a threshold chosen before the
result is inspected; a level-.05 rule rejects the null when the p-value
is below .05. Because .0569 is not below .05, the district result is
not **statistically significant** at the .05 level. The test does not
reject 50, and that decision does not show the mean equals 50.

The label *statistically significant* records a comparison with the
threshold. The p-value calculation starts by assuming the null for its
reference; it does not calculate the probability that the null itself
is true [@wasserstein2016]. Other values, including means near 54, remain
compatible with the district data. The size of a difference, its
importance, and the probability that a claim is true are different
questions.

The district test is two-sided because either direction counts. A
one-sided test needs an alternative whose direction was chosen before
the result was inspected. Had the study specified $H_A:\mu>50$ in
advance, its p-value would be the single upper tail, .0285. Choosing
that tail after seeing a positive estimate, so that .0285 falls below
.05, changes the testing procedure and its error calibration. A large
departure in the opposite direction does not support a one-sided
alternative either.

### A test and an interval built from one procedure

A two-sided test at level $\alpha$ and a $100(1-\alpha)\%$ interval
built from the same procedure agree about any null value. The district
sample shows the agreement at two levels. With the null value 50, the
statistic is the 2 of Section 4.1. Against the supplied two-sided
95% critical value
2.064 the statistic falls short, and the interval [49.936, 54.064]
contains 50: the level-.05 test does not reject. Against the 90%
critical value 1.711 the statistic exceeds the cutoff, and the interval
[50.289, 53.711] excludes 50: a level-.1 test chosen in advance would
reject. In both cases the interval contains exactly the null values the
matching two-sided test would not reject.

Algebra explains the agreement. A null value $\theta_0$ lies outside the
interval $\widehat\theta\pm t^*\widehat{SE}$ exactly when its distance
from the estimate exceeds the margin, $|\widehat\theta-\theta_0|>
t^*\widehat{SE}$. Dividing both sides by the SE gives the test's
comparison of its statistic with the critical value:

$$
\text{the null value }\theta_0\text{ lies outside }\widehat\theta\pm t^*\widehat{SE}
\quad\Longleftrightarrow\quad
\frac{|\widehat\theta-\theta_0|}{\widehat{SE}}>t^*.
$$

The interval is therefore the set of null values the test would not
reject. The agreement requires the same estimator, SE, reference
distribution, sidedness, and comparison, and two calculations aligned in
this way can still share an inadequate model. Nor does the correspondence at each level license choosing
the level after seeing where 50 falls: the 90% rejection counts only if
that level was chosen before the data were examined.

### The Special Practice test

Return to the continuing study. For two populations, the null
$H_0:\Delta\mu=0$ concerns $\Delta\mu=\mu_1-\mu_0$, the difference
between their means; the subscripts 1 and 0 identify the two groups,
here Special and Usual, in that order, so $\mu_0$ now names a group
mean rather than the one-mean null value of Section 4.1. For Special
Practice, $\Delta\mu$ remains the fixed participants' mean assignment
effect defined in Section 1.2, and its null preserves that target:

$$
H_0:\Delta\mu=0\qquad\text{versus}\qquad H_A:\Delta\mu\ne0.
$$

The null says the participant mean assignment effect is zero; the
alternative allows either a positive or a negative effect. Under the
equal-variance working model, the statistic divides the estimate by the
pooled SE of Section 2.2 and uses the t reference with 118 df:

$$
t=\frac{\widehat{\Delta\mu}-0}{\widehat{SE}(\bar y_1-\bar y_0)}
\approx2.7372,\qquad df=118,\qquad
p=2P(T_{118}\ge|t_{\mathrm{obs}}|)\approx.0072.
$$

The estimate is about 2.7372 estimated SEs above zero. Its upper-tail
probability is .0036, and the two-sided test adds the equally extreme
negative tail, giving .0072. The null and the procedure's conditions
define the reference; the observed statistic selects the tail area. At
$\alpha=.05$ the result is statistically significant, and the 95%
interval of Section 3.4, [0.8572, 5.3428], excludes zero, as Section
4.3 predicts.

```{r}
#| label: fig-test-interval
#| fig-cap: "The Special Practice test read two ways; panel (a) is adapted from @biau2010. (a) The t reference with 118 df on the scale of the test statistic: the dashed lines mark the two-sided .05 critical values $\\pm1.9803$, beyond which the light regions hold .025 each, and the red marks show the observed statistic 2.7372 and its mirror image $-2.7372$, which lie beyond them. (b) The upper tail, enlarged on its own vertical scale: the dark area beyond 2.7372 holds .0036, and with the matching lower tail it gives the two-sided p-value .0072. (c) The 95% and 90% intervals from the same equal-variance procedure, read against zero and the two-point benchmark."
#| fig-alt: "Three panels. The first shows a bell-shaped t density with 118 degrees of freedom on a test-statistic axis from about minus 5 to 5; dashed grey lines at minus and plus 1.9803, labelled at the top, mark the .05 critical values, the regions beyond them are shaded light red, and red marks labelled minus and plus 2.7372 stand farther out, where the tails are too thin to see. The second enlarges the upper tail from 1.6 to 4.6 on a density axis up to 0.12: the light region beyond the critical value is labelled critical 1.9803, .025 beyond, and the dark red region beyond the observed statistic is labelled observed 2.7372, .0036 beyond. The third shows the estimate 3.1 with a 95 percent interval from 0.8572 to 5.3428 and a 90 percent interval from 1.2224 to 4.9776, beside a dashed line at zero, labelled no difference, and a dotted line at the two-point benchmark; both intervals exclude zero and cross the benchmark."
#| fig-width: 7.2
#| fig-height: 3.4
#| fig-pos: "!ht"

local({
  p4 <- function(p) fmt4(p, lead = FALSE)
  critical <- qt(.975, pooled_df)
  tail_p <- pt(-test_statistic, pooled_df)
  pink <- "#F4C7C7"
  shade <- function(from, to, col) {
    g <- seq(from, to, length.out = 200)
    polygon(c(from, g, to), c(0, dt(g, pooled_df), 0), col = col,
            border = NA)
  }
  saved_par <- par(mar = c(2.9, 2.9, 1.5, 0.4), mgp = c(1.75, 0.45, 0),
                   tcl = -0.3, cex.axis = 0.8, cex.lab = 0.84,
                   cex.main = 0.88, font.main = 1, col.main = ink,
                   col.axis = ink, col.lab = ink)
  on.exit(par(saved_par))
  layout(matrix(1:3, nrow = 1), widths = c(1.3, 0.95, 0.9))
  par(cex = 1)
  x <- seq(-5.2, 5.2, length.out = 1041)
  plot(NA, xlim = c(-5.2, 5.2), ylim = c(0, 0.5), xaxs = "i", yaxs = "i",
       xlab = "Test statistic t (estimate / SE)", ylab = "Density",
       main = "(a) The t reference with 118 df")
  shade(-5.2, -critical, pink); shade(critical, 5.2, pink)
  shade(-5.2, -test_statistic, red); shade(test_statistic, 5.2, red)
  lines(x, dt(x, pooled_df), col = ink, lwd = 1.8)
  segments(c(-critical, critical), 0, c(-critical, critical), 0.41,
           col = grey, lty = 2, lwd = 1.2)
  text(c(-critical, critical), 0.445, fmt4(c(-critical, critical), minus = "-"),
       col = grey, cex = 0.83)
  segments(c(-test_statistic, test_statistic), 0,
           c(-test_statistic, test_statistic), 0.13, col = red, lwd = 1.6)
  text(c(-test_statistic, test_statistic), 0.13,
       fmt4(c(-test_statistic, test_statistic), minus = "-"),
       pos = c(2, 4), offset = 0.2, col = red, cex = 0.83)
  xt <- seq(1.6, 4.6, length.out = 400)
  plot(NA, xlim = c(1.6, 4.6), ylim = c(0, 0.12), xaxs = "i", yaxs = "i",
       xlab = "Test statistic t", ylab = "Density",
       main = "(b) Upper tail, enlarged")
  shade(critical, 4.6, pink); shade(test_statistic, 4.6, red)
  lines(xt, dt(xt, pooled_df), col = ink, lwd = 1.8)
  segments(critical, 0, critical, 0.095, col = grey, lty = 2, lwd = 1.2)
  text(critical, 0.095, paste0("critical\n", fmt4(critical), "\n.025 beyond"),
       pos = 4, offset = 0.2, col = grey, cex = 0.83)
  segments(test_statistic, 0, test_statistic, 0.05, col = red, lwd = 1.6)
  text(test_statistic, 0.05,
       paste0("observed\n", fmt4(test_statistic), "\n", p4(tail_p), " beyond"),
       pos = 4, offset = 0.2, col = red, cex = 0.83)
  par(mar = c(2.9, 1.0, 1.5, 0.4))
  plot(NA, xlim = c(-1, 7), ylim = c(0.35, 2.75), yaxt = "n",
       xlab = "Special minus Usual (points)", ylab = "",
       main = "(c) Estimate and intervals")
  abline(v = 0, col = grey, lty = 2)
  abline(v = 2, col = gold, lty = 3, lwd = 1.6)
  segments(c(confidence_interval[1], interval_90[1]), c(2, 1),
           c(confidence_interval[2], interval_90[2]), c(2, 1),
           col = blue, lwd = 4)
  points(c(estimate, estimate), c(2, 1), pch = 23, bg = gold, col = ink,
         cex = 1.3)
  text(4.4, c(2.35, 1.35), c("95% interval", "90% interval"),
       cex = 0.83, col = ink)
  text(-0.55, 1.5, "no difference", srt = 90, cex = 0.83, col = grey)
  text(2, 0.62, "two-point\nbenchmark", pos = 4, offset = 0.3, cex = 0.83,
       col = gold)
})
```

Panel (a) of @fig-test-interval draws the reference the test uses, as
Section 4.2 did in words for the district mean. The dashed lines at
$\pm1.9803$ mark the two-sided .05 critical values, the same multiplier
that built the 95% interval in Section 3.4, and the light regions
beyond them hold .025 each. The observed statistic lies beyond them, so
the level-.05 test rejects. The tails beyond $\pm2.7372$ are too thin
to see at this scale, so panel (b) enlarges the upper tail: the dark
area beyond the observed statistic holds .0036, the matching lower tail
holds the same, and together they give the two-sided p-value .0072.
Panel (c) reads the same result as intervals: the 95% and the 90%
interval both exclude zero, and both span the two-point benchmark.

The same p-value can be checked against the design itself. If no
student's score depended on format, re-assigning the 120 records at
random to two groups of 60 would reproduce the design's repetition
under that null, and in 10,000 such re-assignments a share of .0077
gave a difference at least 3.1 points from zero, close to the t test's
.0072; the chapter's R support page runs them. The two need not agree
exactly, because the re-assignments test the stronger null that no
student's score depends on format, using the actual records, whereas
the t test models the average effect
([Appendix A](#app-ch03-assignment)). Such re-assignments form a **null
distribution**, built by imposing the null on the design; a bootstrap
distribution ([Appendix B](#app-ch03-resampling)) instead resamples the
data without imposing a null and describes the estimate's variability,
so it is not a reference for testing zero.

What moves the test statistic? @tbl-ch03-what-moves-t starts from the
Special Practice summaries and changes one input at a time, the last
row two.

| Change from the observed study | SE (points) | t (df) | Two-sided p | 95% interval (points) |
|:--|--:|--:|--:|:--|
| As observed: difference 3.1, SDs 6.0 and 6.4, 60 per group | 1.1325 | 2.7372 (118) | .0072 | [0.8572, 5.3428] |
| Spread 1.5 times larger: SDs 9.0 and 9.6 | 1.6988 | 1.8248 (118) | .0706 | $[-0.2641,\ 6.4641]$ |
| Smaller groups: 15 per group | 2.2651 | 1.3686 (28) | .1820 | $[-1.5398,\ 7.7398]$ |
| Separation halved: difference 1.55 | 1.1325 | 1.3686 (118) | .1737 | $[-0.6928,\ 3.7928]$ |
| Separation halved, 240 per group | 0.5663 | 2.7372 (478) | .0064 | [0.4373, 2.6627] |

: What moves the test statistic: stipulated variations of the Special Practice summaries under the equal-variance procedure. {#tbl-ch03-what-moves-t tbl-colwidths="[40,13,16,12,19]"}

The statistic divides the separation between the means by an SE that
grows with the spread of the scores and shrinks with the group sizes,
$\widehat{SE}=s_p\sqrt{1/n_1+1/n_0}$. Making both SDs 1.5 times larger
multiplies the SE by 1.5, and the same 3.1-point difference no longer
excludes zero, $p=.0706$; a quarter of the students doubles the SE,
$p=.1820$; halving the separation halves the statistic while the SE
stays the same, $p=.1737$. The last row halves the separation and
quadruples the groups. Its statistic returns exactly to 2.7372, so a
study with half the effect and four times the students is just as
significant, $p=.0064$ on 478 df, yet its interval, [0.4373, 2.6627],
is centered at 1.55 points, below the two-point benchmark. A test
statistic measures separation relative to its SE, not the size of an
effect; the estimate, its interval and the benchmark carry the
magnitude.

Calling the result statistically significant records that .0072 is
below the advance threshold .05. It supplies neither the size of the
effect nor a study-specific probability of having made an error. The
table below collects six common mistranslations of the displayed
p-value and their repairs; reporting the estimate and interval alongside
the p-value keeps these distinctions visible.

\Needspace{22\baselineskip}

| Mistranslation | Why it fails | Repair |
|:--|:--|:--|
| "There is a 0.72% chance the null is true." | The p-value conditions on the null; it does not assign probability to it | Under the zero-contrast null and procedure, results at least this incompatible have probability .0072 |
| "There is a 0.72% chance the result happened by chance." | Chance is not a complete model; the result already happened | Name the null, design, statistic, and reference process |
| "The study has a 99.28% chance of replicating." | Replication depends on the effect, design, sample, and analysis | Report the current estimate and uncertainty; discuss replication separately |
| "The effect is large or important because $p=.0072$." | A p-value is not an effect-size or importance measure | Interpret 3.1 points and its interval against a defensible benchmark |
| "The study's false-positive probability is 0.72%." | A long-run error rule and a data-dependent p-value differ | State any prespecified decision rule; avoid a posterior error claim |
| "The p-value keeps its usual meaning after this outcome or model was selected from inspected results." | The selection rule is part of the procedure and can alter calibration | Report the family of analyses and whether the comparison was chosen in advance; interpret the selected result accordingly |

: Six mistranslations of the displayed $p=.0072$ and their repairs. {#tbl-ch03-p-mistranslations tbl-colwidths="[30,32,38]"}

## t Tests for One, Paired and Two Independent Samples {#sec-ch03-t-tests}

Sections 3 and 4 used two members of a family of named t tests, one for
a single mean and one for the difference between two group means. The
members differ in which observations carry the information, how the SE
is calculated, and which df the reference uses; their shared name does
not make paired and independent data interchangeable. The table lists
them in the order this section teaches them.

\Needspace{10\baselineskip}

| Procedure | Data and null | SE and reference |
|:--|:--|:--|
| One-sample t | One quantitative sample; $H_0:\mu=\mu_{\mathrm{null}}$ | $s/\sqrt n$; $df=n-1$ |
| Paired-samples t | Correctly matched observations, independent pairs; $H_0:\mu_D=0$ | $s_d/\sqrt n$ on pair differences; $df=n-1$ |
| Equal-variance two-sample t | Two independent samples with a common population variance; $H_0:\mu_1-\mu_0=0$ | $s_p\sqrt{1/n_1+1/n_0}$; $df=n_1+n_0-2$ |
| Welch's unequal-variance t | Two independent samples whose variances may differ; same mean null | $\sqrt{s_1^2/n_1+s_0^2/n_0}$; approximate df estimated from the data |

: Named t tests and the observations that carry information. {#tbl-ch03-t-procedures tbl-colwidths="[24,40,36]"}

Two of these procedures, the one-sample t test (Section 5.1) and the
equal-variance two-sample t test (Section 5.3), are followed after
their worked result by a box titled *Justifying the analysis*, and
Section 6 gives one to the chi-square test of association. Chapter 1,
Section 6.1, used the word *justification* for the reasons and
conditions invoked to support a comparison, analytical move or
interpretation, and its Section 3.4 introduced the assumptions such
reasons invoke. The box makes that justification concrete for one
analysis in four labelled lines. **Target** states what the number
estimates, in its units and for which units. **Conditions** sorts what
the analysis relies on into three kinds: conditions *secured by
design*, such as random assignment, random sampling, pairing and
independent units; *model assumptions* the procedure adds, such as a
common spread or the shape of a sampling distribution; and *choices and
defaults* the analyst sets, such as the sidedness, the confidence level
and the variance model. **Check** gives the evidence the data offer
about each model assumption and the result that would change the
decision, or says plainly that the data cannot check it. **Bound**
states what the result supports and what it does not. The paired and
Welch t tests are variants of the procedures with boxes, so each states
in a sentence or two what changes, and a later use of any procedure
refers to its box and adds only what changes.
[Appendix B](#app-ch03-procedure-guide) collects the five procedures in
one table with their conditions and the usual response when a condition
fails.

### One-sample t test and its z counterpart

The district calculation of Sections 3 and 4 is a **one-sample t test**
with its matching interval: SE 1 point, a t reference with 24 df,
the statistic 2, the two-sided p-value .0569 and the 95% interval
[49.936, 54.064]. A close relative uses a known SD instead of an
estimated one.

Chapter 2 used a **z score**, $z_i=(x_i-\mu)/\sigma$, to locate
one observation in population SD units. Sample standardization,
$(x_i-\bar x)/s$, describes an observation relative to its sample.
Neither operation makes the original distribution Normal. For a mean,
the quantity being standardized is instead its estimator, so the
denominator is an SE:

$$
z=\frac{\bar x-\mu_0}{\sigma/\sqrt n},\qquad
t=\frac{\bar x-\mu_0}{s/\sqrt n},\qquad df=n-1.
$$

The **one-sample z test** uses a population SD known independently of
the sample. The **one-sample t test** estimates that SD with $s$.
Its exact t reference requires independent observations from a Normal
population; suitable distributional conditions and enough information
can support an approximation more broadly. A large sample does not turn
an unknown SD into a known one or remove dependence.

For the same statistic 2, the two-sided probabilities beyond the
lines of @fig-ch03-z-t in Section 3.1 are .0455 under the Normal, .0766
under t with 9 df and .0482 under t with 99 df. If the district's
population SD were known independently to be 5 points, the z test would
give .0455 rather than the t test's .0569. The change comes from the
reference tails, not from the arithmetic of the statistic, and only
genuine knowledge of the SD justifies the z reference; falling on the
other side of .05 does not. Changing a study's sample size can also
change its SE and observed statistic, which is a separate comparison.

\Needspace{6\baselineskip}

[]{#ch03-justify-one-sample}

::: {.callout-tip title="Justifying the analysis"}
- **Target:** The mean reading score $\mu$ of the district's students,
  in scale points, compared with the published reference value 50.
- **Conditions:**
  - *Secured by design:* A simple random sample of 25 students from a
    large district, each measured once, so the scores are independent
    observations.
  - *Model assumptions:* An approximately Normal population of scores,
    so that the t reference with 24 df applies to a mean whose SD is
    estimated from the same 25 scores.
  - *Choices and defaults:* A two-sided alternative, $H_A:\mu\ne50$,
    chosen in advance; 95% confidence and level .05; one planned test;
    all 25 scores observed.
- **Check:** The summaries, mean 52 and SD 5, cannot show the shape of
  the scores. With the 25 scores in hand, strong skew or one score far
  from the rest would weaken the t approximation and call for a
  bootstrap comparison ([Appendix B](#app-ch03-resampling)). The
  decision is close, $p=.0569$ against .05, so the reference must follow from the conditions: the
  z reference's .0455 would need an SD known independently of the
  sample.
- **Bound:** The interval [49.936, 54.064] describes the district mean
  under the sampling model: means from just below 50 to about 54 points
  remain compatible, and the level-.05 test does not reject 50. The
  result does not show that the mean equals 50, and it concerns this
  district only.
:::

### Paired-samples t test {#sec-ch03-paired}

In a paired design each observation in one set is matched with one in
the other, most often because the same person is measured twice. The
pair, not the single score, is then the unit that carries the
information. Eight invented students, separate from the Special
Practice records, completed the same assessment before and after a
short practice session (@tbl-ch03-student-pairs). Define each student's
difference as $d_i=\mathrm{after}_i-\mathrm{before}_i$, keeping the
person and the subtraction direction before any summary is calculated.

\Needspace{13\baselineskip}

| Student | Before | After | Difference |
|:--|--:|--:|--:|
| 1 | 50 | 49 | $-1$ |
| 2 | 54 | 54 | 0 |
| 3 | 48 | 49 | 1 |
| 4 | 60 | 62 | 2 |
| 5 | 55 | 57 | 2 |
| 6 | 58 | 61 | 3 |
| 7 | 62 | 66 | 4 |
| 8 | 53 | 58 | 5 |

: Invented paired student scores; these are not Special Practice records. {#tbl-ch03-student-pairs tbl-colwidths="[25,25,25,25]"}

Once the differences are formed, the **paired-samples t test** is the
one-sample t test of Section 5.1 applied to them: the null is a zero
mean change, $H_0:\mu_D=0$, the SE is $s_d/\sqrt n$ for the $n$ pairs,
and the reference is t with $n-1$ df. For the eight students the mean
change is 2 points with SD 2, so
$\widehat{SE}(\bar d)=0.7071$, $t(7)=2.8284$, $p=.0255$, and the 95%
interval is [0.328, 3.672] points.

[]{#ch03-justify-paired}

The justification of the one-sample t test in
[Section 5.1](#ch03-justify-one-sample) carries over with the pair as
the unit: the design forms the pairs, the eight pairs must be
independent of one another, which rests on how the students were
chosen, and the near-Normal shape concerns the differences, so they,
not the two sets of scores, are what to inspect. The bound changes
too: without an untreated comparison, a mean change of 2 points
does not separate the practice session from time trends, repeated
testing or other changes between the two measurements.

Why must the pairing be kept? The variance identity of Section 2.3
gives
$\operatorname{Var}(Y_{\mathrm{after}}-Y_{\mathrm{before}})
=\operatorname{Var}(Y_{\mathrm{after}})+\operatorname{Var}(Y_{\mathrm{before}})
-2\operatorname{Cov}(Y_{\mathrm{after}},Y_{\mathrm{before}})$:
when a student's two scores rise and fall together, the covariance term
removes the variation they share.

```{r}
#| label: fig-ch03-paired
#| fig-cap: "The eight invented students' scores three ways. (a) Before and after as two separate sets, with their means; (b) each student's two scores joined; (c) the eight after-minus-before differences, with the paired 95% interval, [0.3280, 3.6720] (blue), and the interval from analysing the scores as two independent groups, $[-3.8877,\\ 7.8877]$ (red)."
#| fig-alt: "Three panels. In the first, eight before scores from 48 to 62 and eight after scores from 49 to 66 overlap heavily, with means 55 and 57 marked. In the second, lines join each student's two scores and run nearly parallel, most rising slightly. In the third, the eight differences from minus 1 to 5 sit above a short blue interval from 0.3280 to 3.6720 and a long red interval from minus 3.8877 to 7.8877 that crosses zero."
#| fig-width: 7.2
#| fig-height: 2.6
#| fig-pos: "!ht"

local({
  before <- c(50, 54, 48, 60, 55, 58, 62, 53)
  after <- c(49, 54, 49, 62, 57, 61, 66, 58)
  difference <- after - before
  saved_par <- par(mar = c(2.9, 2.9, 1.5, 0.4), mgp = c(1.75, 0.45, 0),
                   tcl = -0.3, cex.axis = 0.8, cex.lab = 0.84,
                   cex.main = 0.88, font.main = 1, col.main = ink,
                   col.axis = ink, col.lab = ink)
  on.exit(par(saved_par))
  layout(matrix(1:3, nrow = 1), widths = c(1, 1, 1.2))
  par(cex = 1)
  for (panel in 1:2) {
    plot(NA, xlim = c(0.6, 2.4), ylim = c(46, 67), xaxt = "n",
         xlab = "", ylab = "Score (points)",
         main = if (panel == 1) "(a) Two sets of scores"
                else "(b) Each student's pair")
    axis(1, at = 1:2, labels = c("Before", "After"))
    if (panel == 2) segments(1, before, 2, after, col = grey, lwd = 1.2)
    points(rep(1, 8), before, pch = 16, col = blue)
    points(rep(2, 8), after, pch = 16, col = blue)
    if (panel == 1) {
      segments(c(0.8, 1.8), c(mean(before), mean(after)),
               c(1.2, 2.2), c(mean(before), mean(after)), col = gold, lwd = 2.4)
    }
  }
  paired_ci <- t.test(after, before, paired = TRUE)$conf.int
  ignored_ci <- t.test(after, before, var.equal = TRUE)$conf.int
  par(mar = c(2.9, 1.0, 1.5, 0.4))
  plot(NA, xlim = c(-5, 9), ylim = c(0.4, 3.4), yaxt = "n",
       xlab = "After minus before (points)", ylab = "",
       main = "(c) The eight differences")
  abline(v = 0, col = grey, lty = 2)
  height <- ave(difference, difference, FUN = seq_along)
  points(difference, 2.45 + 0.18 * height, pch = 16, col = blue)
  segments(paired_ci[1], 1.75, paired_ci[2], 1.75, col = blue, lwd = 4)
  segments(ignored_ci[1], 0.85, ignored_ci[2], 0.85, col = red, lwd = 4)
  points(c(2, 2), c(1.75, 0.85), pch = 23, bg = gold, col = ink, cex = 1.2)
  text(paired_ci[2], 1.75, "paired", pos = 4, cex = 0.83, col = blue)
  text(0.1, 1.25, "pairing ignored", pos = 4, cex = 0.83, col = red)
})
```

@fig-ch03-paired shows what the covariance term does. Panel (a) is
what an analysis of two independent groups sees: two overlapping sets
of eight scores whose spread, SDs of about 5 and 6 points, swamps a
2-point shift. Panel (b) restores the pairing: students who scored high
before also scored high after, so the lines run nearly parallel and the
scores correlate at .9599. The differences in panel (c) keep the change
and drop the shared between-student variation. Their interval is short
and excludes zero, whereas analysing the same scores as two independent
groups gives an SE of 2.7451 rather than 0.7071 and the interval
$[-3.8877,\ 7.8877]$, which spans zero. The data did not change; the
relevant variation did. Because the pairing comes from how the scores
were obtained, only the paired analysis describes their uncertainty.
Chapter 4's term *pairwise comparisons* refers to several group
contrasts, a different use of the word.

The school-blocked design of [Section 2.3](#sec-ch03-school-pairs) is
a paired design too: each school supplies one explanatory-minus-usual
difference of class means, and the paired t test applied to the twelve
differences gives $t(11)=3.6187$, $p=.004$ and the 95% interval
[0.9795, 4.0205] points. Here each pair is a school, and random
assignment of the two classes within each school supports a causal
reading for these participating schools, which the before-and-after
scores could not; the schools were not sampled, so the result does not
by itself reach other schools.

### Equal-variance two-sample t test {#sec-ch03-equal-variance}

The procedure used for Special Practice throughout Sections 3 and 4 is
the **equal-variance two-sample t test**, also called the pooled or
Student's two-sample t test. Its model has two independent groups whose
populations share one variance. With Normal populations its t reference
is exact; with adequate group sizes and moderate departures from
Normality it is a good approximation. Its SE is the pooled SE of
Section 2.2, $s_p\sqrt{1/n_1+1/n_0}$, and its reference has $n_1+n_0-2$
df. For Special Practice, Sections 2.2, 3.4 and 4.4 have already
produced the complete result: SE 1.1325 points on 118 df, $t=2.7372$,
$p=.0072$ and the 95% interval [0.8572, 5.3428] points. The new step is
to read the procedure's conditions against the design, which the box
below does.

\Needspace{6\baselineskip}

[]{#ch03-justify-equal-variance}

::: {.callout-tip title="Justifying the analysis"}
- **Target:** The mean effect $\Delta\mu$ of assignment to Special rather
  than Usual Practice on the delayed 0--40 assessment, in points, for
  these 120 participants.
- **Conditions:**
  - *Secured by design:* Random assignment of 60 students to each
    format, which supports a causal comparison among the participants;
    different students in the two groups.
  - *Model assumptions:* One population spread shared by the two
    groups; a near-Normal sampling distribution for the difference in
    means; no relevant interference, so one student's format does not
    change another's score. The t reference with 118 df is an
    approximate calibration for the assignment design (Section 1.3).
  - *Choices and defaults:* Special minus Usual; a two-sided test of
    zero; 95% confidence and level .05; the equal-variance model, chosen
    from the design before the analysis; one comparison and the
    two-point benchmark, both fixed before the outcomes were seen;
    complete follow-up of all 120, so no missing scores to handle.
- **Check:** The SDs, 6.0 and 6.4 points, are similar, and with 60
  students in each group the pooled and unpooled SEs are identical
  whatever the spreads; Welch's procedure gives the same statistic with
  117.5119 df and the same p-value to four decimals (Section 5.4). The
  records in @fig-ch03-scores (Section 1.1) show no score dominating
  either mean: the three scores beyond the whiskers, 13 in Special
  Practice and 38 and 39 in Usual Practice, lie within two points of
  the whisker ends, and each mean averages 60 scores. Interference
  cannot be checked from the scores; it rests on how the sessions were
  run.
- **Bound:** A positive mean assignment effect among these participants
  is supported, 3.1 points, 95% CI [0.8572, 5.3428]. Its practical
  importance against the two-point benchmark remains open (Section 7.1),
  and the result does not speak to attendance, mechanisms, longer-term
  learning or students elsewhere.
:::

A results sentence names the procedure and reports the df beside the
statistic:

> Under the equal-variance working model, Special Practice scored 3.1
> points higher on average, 95% CI [0.8572, 5.3428], *t*(118) = 2.7372,
> *p* = .0072.

### Welch's unequal-variance t test {#sec-ch03-welch}

Pooling assumes that both groups estimate one population variance.
**Welch's unequal-variance t test** drops that assumption and keeps
each group's own variance in the SE [@welch1947]:

$$
\widehat{SE}_{\mathrm{Welch}}=\sqrt{\frac{s_1^2}{n_1}+\frac{s_0^2}{n_0}},
\qquad
t=\frac{\bar y_1-\bar y_0}{\widehat{SE}_{\mathrm{Welch}}}.
$$

Its reference is an approximate t distribution whose df, the
**Welch--Satterthwaite degrees of freedom**, are estimated from the two
variance contributions $s_1^2/n_1$ and $s_0^2/n_0$. They lie between
$\min(n_1,n_0)-1$ and $n_1+n_0-2$. When one group's mean carries almost
all the uncertainty, they fall toward that group's own $n-1$, which is
the lower bound when it is the smaller group. They reach the upper
bound, the equal-variance test's df, only when each contribution is
proportional to its group's df, $n_1-1$ or $n_0-1$, as with equal sizes
and equal SDs. Software computes them, and fractional df describe the
approximation rather than counting people.

Welch's test is therefore not simply a more cautious version of the
equal-variance test: it asks the same question under a weaker
assumption, and how much the choice matters depends on the design.
With equal group sizes the two SEs and t statistics coincide
(Section 2.2), but the df differ unless the two SDs are equal too. With
groups of moderate size that difference barely moves the p-value, and
Welch's test gives up almost no power; with small groups and very
unequal spreads it can change the conclusion. With five students per
group, SDs 10 and 1, and $t=2.5$, the equal-variance test gives
$p=.0369$ on 8 df and Welch's test $p=.0655$ on 4.08 df. With equal
spreads but unequal sizes Welch's test gives up a little power. With
unequal group sizes and unequal spreads, however, the pooled variance, a
df-weighted average dominated by the larger group, misstates the
variance of the smaller group's mean, and the equal-variance test's
error rate drifts in one direction or the other.

For a separate, invented comparison, suppose a district draws two
independent random samples of students who took the same 0--40
assessment: 10 from a small intensive-support programme and 40 from its
usual programme. The intensive group averaged 29 points with SD 8;
the usual group averaged 24 with SD 4. The estimand is the
difference between the two programmes' population mean scores, a
comparison of populations rather than an effect of assignment. Before
seeing these scores, staff expected a wider spread in the intensive
programme, which enrols students with a broad range of prior attainment
and may help some of them much more than others.
@tbl-ch03-welch-example sets the two analyses side by side.

\Needspace{9\baselineskip}

| Result for the 5-point difference | Equal-variance t | Welch's t |
|:--|--:|--:|
| SE of the difference (points) | 1.7678 | 2.6077 |
| t statistic | 2.8284 | 1.9174 |
| Degrees of freedom | 48 | 10.151 |
| Two-sided p-value | .0068 | .0837 |
| 95% interval (points) | [1.4457, 8.5543] | $[-0.7986,\ 10.7986]$ |

: The same summaries under two variance models. {#tbl-ch03-welch-example tbl-colwidths="[44,28,28]"}

The equal-variance SE is much smaller than Welch's because the pooled
variance lies near the usual group's small spread, which supplies 39 of
its 48 df, and pooling lends that small spread to the intensive group's
mean. The equal-variance interval therefore excludes zero while Welch's
includes it. Welch's df, 10.151, lie close to the lower bound of 9
because the intensive group's mean carries almost all of the
uncertainty. Because the staff expected unequal spreads before seeing
the scores, Welch's procedure is the one this design calls for.
@fig-ch03-welch-dots draws one invented set of 50 scores with exactly
these summaries and shows what pooling does to each group's mean.

```{r}
#| label: fig-ch03-welch-dots
#| fig-cap: "One invented set of scores with the two programmes' summaries: 10 intensive-programme scores (mean 29, SD 8) and 40 usual-programme scores (mean 24, SD 4). Below each group's dots, the blue bar spans one SE either side of the mean from the group's own SD, and the red bar one SE from the pooled SD, 5."
#| fig-alt: "Two stacked dot plots on an assessment scale from 14 to 42. The ten intensive scores spread from 18 to 40 in two loose clusters, 18 to 24 and 29 to 40; the forty usual scores run from 16 to 33, most between 19 and 30. For the intensive mean of 29 the blue own-SD bar, plus or minus 2.5298, is much wider than the red pooled bar, plus or minus 1.5811; for the usual mean of 24 the blue bar, plus or minus 0.6325, is slightly shorter than the red one, plus or minus 0.7906."
#| fig-width: 7.2
#| fig-height: 2.6
#| fig-pos: "!ht"

local({
  intensive <- c(18, 20, 20, 24, 29, 33, 34, 34, 38, 40)
  usual <- c(16, 18, 19, 19, 20, 20, 20, 20, 20, 21, 21, 21, 22, 22, 22, 22,
             23, 23, 23, 23, 24, 24, 25, 25, 25, 25, 25, 26, 26, 26, 26, 27,
             29, 29, 29, 29, 30, 30, 32, 33)
  pooled_sd <- sqrt((9 * var(intensive) + 39 * var(usual)) / 48)
  saved_par <- par(mar = c(4.1, 7.0, 0.6, 0.8))
  on.exit(par(saved_par))
  plot(NA, xlim = c(14, 42), ylim = c(0.2, 3.3), yaxt = "n",
       xlab = "Assessment score (points, 0-40 scale)", ylab = "",
       col.axis = ink, col.lab = ink)
  axis(2, at = c(0.9, 2.4),
       labels = c("Usual\n(n = 40)", "Intensive\n(n = 10)"),
       las = 1, tick = FALSE, col.axis = ink)
  groups <- list(usual, intensive); base <- c(1.05, 2.55)
  for (g in 1:2) {
    x <- groups[[g]]
    height <- ave(x, x, FUN = seq_along)
    points(x, base[g] + 0.09 * height, pch = 16, cex = 0.8, col = grey)
    own <- sd(x) / sqrt(length(x)); pooled <- pooled_sd / sqrt(length(x))
    segments(mean(x) - c(own, pooled), base[g] - c(0.18, 0.42),
             mean(x) + c(own, pooled), base[g] - c(0.18, 0.42),
             col = c(blue, red), lwd = 4)
    points(rep(mean(x), 2), base[g] - c(0.18, 0.42), pch = 23, bg = gold,
           col = ink, cex = 1.1)
  }
  legend("bottomright", c("mean +/- 1 SE from the group's own SD",
                          "mean +/- 1 SE from the pooled SD"),
         col = c(blue, red), lwd = 4, bty = "n", cex = 0.75)
})
```

Pooling lends the usual group's tighter spread to the intensive group's
mean: the SE of that mean shrinks from its own 2.5298 points to 1.5811,
while the usual group's grows only from 0.6325 to 0.7906. The pooled SE
of the difference therefore understates the uncertainty that the ten
widely spread intensive scores carry.

[]{#ch03-justify-welch}

The justification of the equal-variance test in
[Section 5.3](#ch03-justify-equal-variance) carries over with three
changes: the target is a difference between two populations' means,
because the programmes were sampled rather than assigned; no common
spread is assumed, so the model assumptions concern each group mean,
above all the ten intensive scores, whose dot plot shows no single
dominating score; and Welch's variance model was chosen before the
scores were seen. Its bound is correspondingly modest: the interval
$[-0.7986,\ 10.7986]$, $p=.0837$, does not establish which programme
has the higher population mean.

How much does the variance model matter? @tbl-ch03-welch-simulation
answers by simulation. Each row generates 50,000 studies of two
independent samples from Normal populations with equal means and the
stated SDs, so every rejection of a zero difference rejects a true
null, and records how often each procedure's two-sided test at level
.05 rejects.

\Needspace{8\baselineskip}

| Group sizes | Population SDs | Equal-variance t | Welch's t |
|:--|:--|--:|--:|
| 10 and 40 | 8 and 4, as in the two-programme comparison | .1848 | .0507 |
| 10 and 40 | 4 and 8 | .0055 | .0513 |
| 25 and 25 | 8 and 4 | .0514 | .0493 |

: Rejection rates of a true zero difference at nominal level .05, from 50,000 simulated studies per design. {#tbl-ch03-welch-simulation tbl-colwidths="[16,48,18,18]"}

When the smaller group is the more variable, the equal-variance test
rejects the true null in .1848 of the studies instead of .05, so only
81.52% of its nominal 95% intervals cover the true difference; when the
smaller group is the less variable it rejects in only .0055, a test
that has become needlessly conservative and lost power. With equal
group sizes both tests stay near .05, and Welch's test stays near .05
in all three designs. This is why many texts recommend Welch's test
whenever the spreads may differ, and statistical software such as R
uses it by default [@delacre2017].

For Special Practice the two procedures nearly coincide: with 60
students in each group both SEs are 1.1325 points, so both give
$t=2.7372$, and Welch's df, 117.5119, fall just below 118 because the
SDs 6.0 and 6.4 are close but not equal, so both p-values display as
.0072. The agreement follows from equal group sizes and similar SDs; it
does not by itself establish equal population variances.

Choose between the procedures before the analysis, from the design and
subject knowledge: whether a common spread is plausible given how the
groups were formed and what the conditions might do to individual
responses, and whether the group sizes are equal. A preliminary test of
equal variances is not a good basis for the choice: the two-stage
procedure alters the error rates it is meant to protect, such tests
have least power in the small samples where the choice matters most,
and a nonsignificant variance test does not establish equal variances
[@zimmerman2004; @delacre2017]. Choosing whichever procedure excludes
zero is selecting an analysis after inspection, a problem Section 7.3
takes up.

## Association Between Categorical Variables: The Chi-Square Test {#sec-categorical-comparison}

The test logic of Section 4 is not limited to means. Chapter 2,
Section 3.1, described the relationship between two categorical
variables with conditional proportions, comparing passing among
students with and without a higher-education intention. Such a
relationship is an **association** between categorical variables: the
distribution of one variable changes across the categories of the
other. When the outcome is a category, such as completed or not, each
record falls in one cell of a table and the data are counts. This
section builds a test of association with the same three parts as the
tests of Section 4: a null hypothesis, a statistic that measures
departure from it, and a reference distribution for that statistic.
Its last subsection reads tables of three or more categorical
variables.

### Counts, proportions and expected counts

Consider an invented teaching model with 100 independently sampled
learners. For each learner the participation format, Online or In
person, and whether the learner completed the course were recorded. The
format was observed rather than randomly assigned, so the question
concerns association: is completion related to format?
@tbl-ch03-two-way crosses the two variables in a **two-way table**,
with one cell for each combination and the totals in the margins. The
number of ways counts the variables crossed, not their categories: a
table of three formats by completion would still be a two-way table.

\Needspace{8\baselineskip}

| Participation format | Completed | Did not complete | Total | Completion proportion |
|:--|--:|--:|--:|--:|
| Online | 32 | 18 | 50 | .64 |
| In person | 18 | 32 | 50 | .36 |
| Total | 50 | 50 | 100 | .50 |

: Invented counts of completion by participation format for 100 learners. {#tbl-ch03-two-way tbl-colwidths="[28,16,20,12,24]"}

Among the 50 Online learners 32 completed, a **conditional proportion**
of $32/50=.64$; among the 50 In-person learners 18 completed,
$18/50=.36$. Each proportion divides by its own format's total, so
the two formats can be compared even when their sizes differ. The
Online-minus-In-person difference is .28, or 28 percentage
points. @fig-ch03-completion draws the two proportions beside the
.5 of all 100 learners.

```{r}
#| label: fig-ch03-completion
#| fig-cap: "Completion by participation format in the invented table of 100 learners. Each bar divides one format's 50 learners into those who completed (dark) and those who did not (light); the dashed line marks the completion proportion of all 100 learners, .5, which both formats would share if format and completion were independent."
#| fig-alt: "Two horizontal bars, Online above In person, each running from 0 to 1. The dark completed part reaches .64 for Online and .36 for In person, with the remainders .36 and .64 in a light shade. A dashed red line at .5 is labelled as the proportion of all 100 learners; the Online bar's completed part extends past it and the In-person bar's stops short of it."
#| fig-width: 7.2
#| fig-height: 1.9
#| fig-pos: "!ht"

local({
  completed <- c("In person" = 18 / 50, Online = 32 / 50)
  p2 <- function(p) fmt4(p, lead = FALSE)
  saved_par <- par(mar = c(3.6, 6.2, 1.2, 1.0), mgp = c(2.2, 0.6, 0))
  on.exit(par(saved_par))
  at <- barplot(rbind(completed, 1 - completed), horiz = TRUE,
                col = c(blue, blue_light), border = ink, xlim = c(0, 1),
                las = 1, xlab = "Proportion of the format's 50 learners",
                space = 0.5, col.axis = ink, col.lab = ink, axes = FALSE)
  axis(1, at = seq(0, 1, .25), labels = c("0", ".25", ".50", ".75", "1"),
       col.axis = ink)
  text(completed / 2, at, paste("completed", p2(completed)),
       col = "white", cex = 0.83)
  text((1 + completed) / 2, at, paste("did not", p2(1 - completed)),
       col = ink, cex = 0.83)
  abline(v = .5, lty = 2, col = red, lwd = 1.4)
  text(.5, max(at) + 0.75, "all 100 learners: .5", pos = 4,
       offset = 0.3, cex = 0.83, col = red, xpd = TRUE)
})
```

\Needspace{8\baselineskip}

Are the two rates different enough, beyond sampling variation, to
indicate an association in the population? The test sets two
hypotheses about the stipulated population or process:

$$
\begin{aligned}
H_0&:\ \text{format and completion are not associated},\\
H_A&:\ \text{format and completion are associated}.
\end{aligned}
$$

No association means that format and completion are **independent**:
the completion probability is the same in both formats. Estimated from
all the records, that common probability is
$50/100=.5$, so independence predicts 25 completions among the 50
Online learners and 25 among the 50 In-person learners. In general a
cell's **expected count** is

$$
E_{rc}=\frac{(\text{row }r\text{ total})(\text{column }c\text{ total})}{n},
$$

which for the Online/Completed cell gives $50\times50/100=25$. Expected
counts keep the observed margins while imposing the null relationship,
and in other tables they need not be whole numbers.
@tbl-ch03-counts sets the observed counts beside the expected ones.

\Needspace{8\baselineskip}

| Participation format | Observed: completed | Observed: did not | Expected: completed | Expected: did not |
|:--|--:|--:|--:|--:|
| Online | 32 | 18 | 25 | 25 |
| In person | 18 | 32 | 25 | 25 |
| Total | 50 | 50 | 50 | 50 |

: Observed counts and the counts expected under independence, with the same margins. {#tbl-ch03-counts tbl-colwidths="[24,19,19,19,19]"}

Every observed count lies 7 from its expected count: more Online and
fewer In-person learners completed than independence predicts.

### The chi-square statistic and its distribution

Pearson's **chi-square statistic** gathers the departures into one
number. Each observed-minus-expected difference is squared, so that
departures in either direction count, and divided by its expected
count, so that a departure is judged against the size of its cell:

$$
X^2=\sum_{\text{cells}}\frac{(O_{rc}-E_{rc})^2}{E_{rc}}.
$$

Here each of the four cells contributes $(32-25)^2/25=1.96$, so
$X^2=7.84$. A table that matched independence exactly would give 0,
and larger values mean larger departures.

How large is large? If format and completion are independent, $X^2$
follows approximately a **chi-square distribution**. Panel (a) of
@fig-ch03-chisq shows that this is a family of right-skewed
distributions on positive values whose shape is set by the degrees of
freedom: the mean of each curve equals its df, so the curves move right
and spread out as the df grow. For a two-way table with $r$ rows and
$c$ columns the df are $(r-1)(c-1)$. In a two-by-two table, once the
row and column totals are fixed, the count in one cell determines the
other three, so only one cell is free to depart from independence and
$df=(2-1)(2-1)=1$.

```{r}
#| label: fig-ch03-chisq
#| fig-cap: "The chi-square reference. (a) Chi-square distributions with 1, 4 and 8 degrees of freedom: right-skewed curves on positive values whose centre and spread grow with the df. (b) The upper tail of the distribution with 1 df, the reference for a two-by-two table, enlarged on its own vertical scale: the shaded area beyond the .05 critical value 3.8415 holds .05, and the dark part beyond the observed statistic 7.84 is the p-value, $p=.0051$."
#| fig-alt: "Two panels. The first shows three density curves on an axis from 0 to 20: with 1 degree of freedom the curve falls steeply from a high value near zero; with 4 it peaks near 2; with 8 it peaks near 6 and spreads further right. The second enlarges the upper tail of the curve with 1 degree of freedom, from 3 to 12 on a density axis up to 0.04: a dashed grey line at 3.8415 is labelled critical 3.8415, .05 beyond, and the region beyond it is shaded light red; a red mark at 7.84 is labelled observed 7.84, p = .0051 beyond, and the smaller region beyond it is dark red."
#| fig-width: 7.2
#| fig-height: 2.9
#| fig-pos: "!ht"

local({
  saved_par <- par(mfrow = c(1, 2), mar = c(4.0, 3.6, 1.8, 0.6),
                   mgp = c(2.3, 0.6, 0))
  on.exit(par(saved_par))
  x <- seq(0.02, 20, length.out = 1000)
  plot(NA, xlim = c(0, 20), ylim = c(0, 0.5), yaxs = "i",
       xlab = "Value of the chi-square statistic", ylab = "Density",
       main = "(a) Chi-square distributions", cex.main = 0.85,
       font.main = 1, col.main = ink, col.axis = ink, col.lab = ink)
  dfs <- c(1, 4, 8); colours <- c(red, blue, gold)
  for (k in 1:3) lines(x, dchisq(x, dfs[k]), col = colours[k], lwd = 1.8)
  text(c(1.6, 3.4, 9.2), c(0.33, 0.2, 0.125), paste("df =", dfs),
       pos = 4, col = colours, cex = 0.83)
  critical <- qchisq(.95, 1); observed <- 7.84
  shade <- function(from, to, col) {
    g <- seq(from, to, length.out = 200)
    polygon(c(from, g, to), c(0, dchisq(g, 1), 0), col = col, border = NA)
  }
  x1 <- seq(3, 12, length.out = 600)
  plot(NA, xlim = c(3, 12), ylim = c(0, 0.04), yaxs = "i",
       xlab = "Value of the chi-square statistic", ylab = "Density",
       main = "(b) Upper tail of df = 1, enlarged", cex.main = 0.85,
       font.main = 1, col.main = ink, col.axis = ink, col.lab = ink)
  shade(critical, 12, "#F4C7C7"); shade(observed, 12, red)
  lines(x1, dchisq(x1, 1), col = ink, lwd = 1.8)
  segments(critical, 0, critical, 0.038, col = grey, lty = 2, lwd = 1.2)
  text(critical, 0.0368, paste0("critical ", fmt4(critical), ": .05 beyond"),
       pos = 4, offset = 0.15, col = grey, cex = 0.83)
  segments(observed, 0, observed, 0.016, col = red, lwd = 1.6)
  text(observed, 0.016, paste0("observed ", fmt4(observed), "\np = ",
       fmt4(pchisq(observed, 1, lower.tail = FALSE), lead = FALSE), " beyond"),
       pos = 4, offset = 0.15, col = red, cex = 0.83)
})
```

Only the upper tail counts. Because every contribution is squared,
departures from independence in either direction make $X^2$ larger, so
the p-value is the area beyond the observed statistic in the upper
tail. Panel (b) enlarges the upper tail of the chi-square distribution
with 1 df: 5% of it lies beyond the critical value 3.8415, and the
observed 7.84 lies well beyond that, cutting off the dark upper-tail
area $p=.0051$. If format and completion were
independent, a table departing from independence at least this much
would arise in about 5 of every 1,000 samples of 100 learners.

### Reading the result and its conditions

Under the stipulated model, the table gives evidence against
independence. The proportions give the direction and size that the
squared statistic discards: completion is 28 percentage points
more frequent in the Online group. This observed association does not
establish that assigning a format would change completion, because
learners chose or were placed in their formats by processes the table
does not record.

\Needspace{6\baselineskip}

[]{#ch03-justify-counts}

::: {.callout-tip title="Justifying the analysis"}
- **Target:** Whether participation format and completion are associated
  in the stipulated population, with direction and size given by the
  completion proportions, .64 Online and .36 In person.
- **Conditions:**
  - *Secured by design:* 100 independently sampled learners, each
    counted once in exactly one of the four cells. The format was
    observed, not assigned.
  - *Model assumptions:* Expected counts large enough for the chi-square
    reference with 1 df to approximate the statistic's distribution
    under independence.
  - *Choices and defaults:* Pearson's statistic without a continuity
    correction, as calculated above; the upper tail, because only large
    discrepancies count against independence; level .05; Online minus
    In person for the proportion difference.
- **Check:** Every expected count is 25, far above the small counts at
  which the approximation weakens; a common guideline flags expected
  counts below 5, where an exact test (Chapter 6) would take its place.
  Independence of the records cannot be checked from the table: learners
  counted twice or sharing a class would make the counts overstate the
  information.
- **Bound:** Evidence against independence in the stipulated
  population, $X^2(1)=7.84$, $p=.0051$, with a completion proportion
  .28 higher Online. Because the format was observed rather than
  assigned, the table does not show that assigning a format would change
  completion.
:::

Statistical software often applies a continuity correction to a
two-by-two table by default, a related but different calculation;
Chapter 6 reproduces the uncorrected statistic taught here in R and
develops further categorical analyses. Exercise 7 supplies a separate
200-record table for applying the same reasoning.

### Three or more categorical variables {#sec-ch03-multiway}

A two-way table holds two categorical variables. Crossing a third gives
a **three-way table**, and a fourth a four-way table; any table of
three or more is a **multiway table**. Suppose the invented record had
also noted whether each learner had studied the subject before, with
25 learners of each format in each prior-study group.
@tbl-ch03-three-way arranges the same 100 learners by prior study,
format and completion. It is read as two-way tables within the
**layers** of the third variable: one format-by-completion table for
the learners who had studied the subject before, and one for those new
to it. Adding the two layers cell by cell returns the two-way table of
Section 6.1.

\Needspace{12\baselineskip}

| Prior study | Participation format | Completed | Did not complete | Total | Completion proportion |
|:--|:--|--:|--:|--:|--:|
| Studied before | Online | 19 | 6 | 25 | .76 |
|  | In person | 15 | 10 | 25 | .60 |
| New to the subject | Online | 13 | 12 | 25 | .52 |
|  | In person | 3 | 22 | 25 | .12 |
| Both layers added | Online | 32 | 18 | 50 | .64 |
|  | In person | 18 | 32 | 50 | .36 |

: Invented counts of completion by participation format within two layers of prior study, for the same 100 learners. {#tbl-ch03-three-way tbl-colwidths="[25,19,15,16,9,16]"}

Within each layer, Online learners completed more often: .76 against
.60 among those who had studied the subject before, a difference of
.16, and .52 against .12 among those new to it, a difference of .40.
Prior study also goes with completion in both formats. Because each
layer holds the same number of learners of each format, the added
table's difference, .28, lies midway between the layer differences.
When the layers hold different mixes of the groups compared, the added
table can differ from every layer; Chapter 14 reads a real admissions
table in which the added comparison points the other way from most of
its layers.

The chi-square test of Sections 6.1--6.3 applies to any one two-way
table in the display. Applied to one layer, it asks whether format and
completion are associated among that layer's 50 learners; applied to
the added table, it asks the question of Section 6.1, with prior study
set aside. No test of one two-way table answers a question about the
three variables jointly: whether the format difference is the same in
both layers, or whether format and completion are associated once
prior study is held fixed, with all 100 learners used at once. Such
questions need a method that uses all three variables at once, such as
the models for multiway tables that begin with logistic regression in
Chapter 11.

Comparing the two layers' tests does not settle whether their
associations differ. Each layer's p-value depends on the size of its
association and on the number of learners in the layer, so a p-value
below .05 in one layer beside one above .05 in the other shows that the
evidence differs, not that the associations do [@gelman2006]. The
difference between the layer differences, .40 against .16, is a
quantity with its own uncertainty, which a joint model estimates.

A fourth variable, such as the term in which a learner enrolled, adds a
second layer variable. The four-way table is read as two-way tables
within each combination of the two layer variables, four layers when
each has two categories, and every added variable divides the same
learners among more layers, leaving fewer records in each.

## Magnitude, Uncertainty, and Several Results {#sec-ch03-interpretation}

### Magnitude and practical importance

The special-practice estimate, interval, test, and benchmark answer
different parts of the substantive question. The raw difference is
3.1 points. Its approximate interval [0.8572, 5.3428] lies above zero and
agrees with the two-sided $p=.0072$ result. It also spans the
course-specific two-point importance criterion.

Consequently, a positive mean assignment effect is supported under
the procedure, while its practical importance remains unresolved.
Effects below the criterion and effects clearly above it are both
compatible with the data. The estimate exceeding two points does not
settle this issue. Nor would an interval crossing zero automatically
mean that all effects of practical interest had been ruled out.

A **standardized mean difference** expresses a raw difference relative
to a named SD. One familiar denominator is the pooled SD of Section 2.2,
$s_p=6.2032$ points, which gives

$$
d=\frac{\bar y_1-\bar y_0}{s_p}=\frac{3.1}{6.2032}=.4997.
$$

The mean difference is about half of this particular within-group SD.
Hedges' correction adjusts $d$ for small-sample bias under its
reference model; its usual approximation multiplies $d$ by a factor
just below one [@lakens2013]:

$$
g\approx d\left(1-\frac{3}{4(n_1+n_0)-9}\right)=.4966.
$$

Changing the denominator changes the comparison's scale. It does not
broaden the participant scope or supply a universal importance threshold.
For the current course decision, assessment points and the two-point
benchmark remain the most direct terms. Chapter 4 develops effect-size
interpretation across group comparisons.

### Errors, power, and precision planning

A test ends in a decision, to reject the null or not to reject it, and
the decision can be wrong in two ways. A **Type I error** rejects a
null hypothesis that is true. Its probability is the significance level
$\alpha$, chosen in advance: a level-.05 rule applied when the null and
the procedure's conditions hold rejects in 5% of repetitions. A **Type
II error** fails to reject a null that is false. Its probability,
$\beta$, depends on how far the truth lies from the null, so it is
defined only under a specified alternative, such as a true difference
of three points. The **power** of the test at that alternative is the
probability that it rejects there, $1-\beta$: the chance that the study
detects an effect of that size. @tbl-ch03-errors sets out the four
combinations of decision and truth; it says *do not reject* because
the decision does not reveal that the null is true.

\Needspace{9\baselineskip}

| Decision | Null is true | Specified alternative is true |
|:--|:--|:--|
| Do not reject $H_0$ | Correct nonrejection; probability $1-\alpha$ for an exactly calibrated continuous test | Type II error; probability $\beta$ at this alternative |
| Reject $H_0$ | Type I error; probability $\alpha$ for that calibrated test | Correct rejection; power $1-\beta$ at this alternative |

: Decisions and error probabilities under specified states of the model. {#tbl-ch03-errors tbl-colwidths="[22,39,39]"}

Both error probabilities condition on a state of the model; neither is
the probability that the decision in one completed study is wrong. For
an approximately calibrated procedure, nominal $\alpha$ need not equal
the actual rejection probability, as the Welch demonstration of Section
5.4 showed.

These probabilities matter most when a study is planned, and planning
can serve two purposes. **Precision planning** asks for an interval
narrow enough for the intended use. For a separate independent
two-group study, suppose each group has anticipated SD 6.2 points and
we want a 95% interval with half-width 2 points. With $n$ per group the
half-width is about $h=z^*\sigma\sqrt{2/n}$, and solving for $n$ gives

$$
n\approx2\left(\frac{z^*\sigma}{h}\right)^2
=2\left(\frac{1.96\times6.2}{2}\right)^2=73.8356,
$$

so 74 observed outcomes per group. The calculation uses a stipulated SD
and an independent-sampling approximation; it does not redesign the
fixed 120-participant study.

**Power planning** asks how likely the study is to reject a zero
difference when a specified difference is real. At 74 per group the SE
of the estimated difference is $6.2\sqrt{2/74}=1.0193$ points. Treating
the planning SD as known, so that the Normal reference applies, a
two-sided .05 test rejects when the estimate lies beyond $\pm1.9977$
points. Panel (a) of @fig-ch03-errors-power draws two sampling
distributions of the estimate on one axis, under the null, centred at
zero, and under an alternative with a true difference of 3 points,
together with that one rejection rule. The null's two outer tails hold
$\alpha/2=.025$ each: a rejection there is a Type I error. Under the
alternative, the hatched area between the boundaries, $\beta=.1627$, is
the probability of a Type II error, and the rest of its curve, .8373,
is the power at 3 points. Moving the alternative further from zero, or
narrowing both curves with a larger sample, shrinks $\beta$ and raises
the power.

```{r}
#| label: fig-ch03-errors-power
#| fig-cap: "Errors and power for a planned study with 74 per group, outcome SD 6.2 and a two-sided .05 test, adapted from @biau2010. (a) Sampling distributions of the estimated difference under the null (true difference 0) and under an alternative (true difference 3), with the rejection boundaries $\\pm1.9977$: the null's dark outer tails hold $\\alpha/2=.025$ each; under the alternative, the hatched area between the boundaries is $\\beta=.1627$ and the light area beyond them is the power, .8373. (b) Power of the same rule across true differences for 37, 74 and 148 per group."
#| fig-alt: "Two panels. The first shows two bell-shaped curves of equal height on an estimated-difference axis, a solid null curve centred at 0 and a dashed alternative curve centred at 3, with dashed boundary lines at minus and plus 1.9977 marked on the axis. The null curve's two small outer tails are dark blue; under the alternative curve the region between the boundaries is hatched red and labelled beta = .1627, and the larger region beyond 1.9977 is light blue and labelled power = .8373. The second plots power against true differences from 0 to 5 points for 37, 74 and 148 per group; every curve starts at .05, the 74-per-group curve passes .5009 at 2 and .8373 at 3, and the 148-per-group curve reaches .7925 at 2, just below a dotted line at .8."
#| fig-width: 7.2
#| fig-height: 3.1
#| fig-pos: "!ht"

local({
  plan_se <- 6.2 * sqrt(2 / 74)
  cutoff <- qnorm(.975) * plan_se
  power_at <- function(delta, n, sigma = 6.2) {
    se <- sigma * sqrt(2 / n); cut <- qnorm(.975) * se
    pnorm(-cut, delta, se) + 1 - pnorm(cut, delta, se)
  }
  saved_par <- par(mar = c(2.9, 2.9, 1.5, 0.4), mgp = c(1.75, 0.45, 0),
                   tcl = -0.3, cex.axis = 0.8, cex.lab = 0.84,
                   cex.main = 0.88, font.main = 1, col.main = ink,
                   col.axis = ink, col.lab = ink)
  on.exit(par(saved_par))
  layout(matrix(1:2, nrow = 1), widths = c(1.4, 1))
  x <- seq(-4.5, 7.5, length.out = 1201)
  null_d <- dnorm(x, 0, plan_se); alt_d <- dnorm(x, 3, plan_se)
  area <- function(keep, dens, ...) {
    polygon(c(x[keep], rev(x[keep])), c(dens[keep], rep(0, sum(keep))), ...)
  }
  plot(NA, xlim = c(-4.5, 7.5), ylim = c(0, 0.78), yaxs = "i", xaxt = "n",
       xlab = "Estimated difference (points)", ylab = "Density",
       main = "(a) One rejection rule, two possible truths")
  axis(1, at = c(-4, -cutoff, 0, cutoff, 4, 6),
       labels = c("-4", "-1.9977", "0", "1.9977", "4", "6"))
  area(x >= cutoff, alt_d, col = "#CFE3EC", border = NA)
  area(abs(x) < cutoff, alt_d, col = red, border = NA, density = 24,
       angle = 45)
  area(x <= -cutoff, null_d, col = blue, border = NA)
  area(x >= cutoff, null_d, col = blue, border = NA)
  lines(x, null_d, col = ink, lwd = 1.8)
  lines(x, alt_d, col = ink, lwd = 1.8, lty = 5)
  segments(c(-cutoff, cutoff), 0, c(-cutoff, cutoff), 0.3, col = grey,
           lty = 2, lwd = 1.2)
  curves <- legend("topleft", c("null: true difference 0",
                                "alternative: true difference 3"),
                   lty = c(1, 5), lwd = 1.8, col = ink, bty = "n",
                   cex = 0.83)
  legend(curves$rect$left, curves$rect$top - curves$rect$h + 0.03,
         legend = c(expression(paste(alpha, "/2 = .025 each tail")),
                    expression(paste(beta, " = .1627")), "power = .8373"),
         fill = c(blue, red, "#CFE3EC"), border = c(ink, red, ink),
         density = c(NA, 30, NA), angle = 45, bty = "n", cex = 0.83)
  plot(NA, xlim = c(0, 5), ylim = c(0, 1), xlab = "True difference (points)",
       ylab = "Power", main = "(b) Power across true differences")
  abline(h = c(.05, .80), lty = 3, col = grey)
  difference <- seq(0, 5, length.out = 201)
  sizes <- c(37, 74, 148); colours <- c(grey, blue, red); types <- c(2, 1, 4)
  for (k in 1:3) lines(difference, power_at(difference, sizes[k]),
                       col = colours[k], lty = types[k], lwd = 2)
  marked <- c(power_at(c(2, 3), 74), power_at(2, 148))
  points(c(2, 3, 2), marked, pch = 21, bg = gold, col = ink, cex = 1.1)
  text(c(2, 3, 2), marked + c(0, 0, 0.08),
       fmt4(marked, lead = FALSE), pos = c(4, 4, 2), cex = 0.83)
  legend("bottomright", paste(sizes, "per group"), col = colours,
         lty = types, lwd = 2, bty = "n", cex = 0.83, inset = c(0, 0.08))
})
```

Panel (b) repeats the calculation across true differences for three
group sizes. Every curve starts at .05 at a true difference of zero,
where a rejection is a Type I error, and rises with the true difference
and the group size; a smaller outcome SD would raise it too. At 74 per
group, the size chosen for a two-point half-width, power is .5009 at a
true difference of 2 points, about an even chance of rejecting at the
benchmark itself, and .8373 at 3 points; doubling to 148 per group
raises power at 2 points to .7925. Reaching power .8 at 2 points takes
about 151 per group (152 with the t reference), roughly twice the
number that precision planning gave. A study has a power curve, not one
power value, and a sample size chosen for precision is not
automatically a power target.

Planning also needs a plausible spread, allocation, dependence
structure, procedure, sidedness, threshold, measurement quality and
feasible follow-up. For a class-assigned study, additional randomly assigned classes supply information that more students
within existing classes cannot replace, as the design effect of Section
2.3 quantifies. If 10% of outcomes are expected to be unavailable
independently of relevant outcomes, recruiting at least 83 per group
($74/.9=82.2222$) gives 74 on average, which neither guarantees
complete data nor repairs bias from selective attrition.

[]{#app-ch03-thousand-studies}

Error probabilities describe a rule applied over many studies, not a
single result. Suppose, as stipulated expected counts rather than data,
that 900 of 1,000 studies test a true null and 100 test a specified
alternative, each at $\alpha=.05$ with power .8 at its alternative
(@tbl-ch03-thousand).

\Needspace{8\baselineskip}

| Model state | Reject | Do not reject | Total |
|:--|--:|--:|--:|
| True null | 45 | 855 | 900 |
| Specified alternative | 80 | 20 | 100 |
| Total | 125 | 875 | 1,000 |

: Expected decisions under stipulated prevalence, alpha and power. {#tbl-ch03-thousand tbl-colwidths="[40,20,20,20]"}

The Type I error probability is $45/900=.05$ and the Type II error
probability $20/100=.2$, but among the 125 expected rejections,
$45/125=.36$ come from true nulls. This false-positive fraction has a
different denominator: it changes with the share of true nulls and with
the power, and it is neither an observed p-value nor a probability for
one completed study.

### Choices made after seeing the results {#sec-ch03-selection}

One analysis can be part of a larger search. A researcher might inspect
several outcomes, subgroups, models, exclusion rules or stopping
points, then report the most favorable result. The reported tail area
can still be calculated, but a calibration for one advance-specified
test does not cover that selection process [@simmons2011]. A report
should therefore say which decisions were made before and which after
the relevant results were examined. Choosing in advance makes the
intended comparison and the scope for selection visible; a comparison
noticed later can still be informative when its exploratory origin is
reported with it.

Selection also changes what the retained estimates look like. Consider
a Normal sampling model with true difference 1 point and known SE 2
points. A two-sided .05 test of zero using the critical value 1.96
retains only estimates above 3.92 or below $-3.92$
(@fig-ch03-selection).

```{r}
#| label: fig-ch03-selection
#| fig-pos: "!ht"
#| fig-cap: "A significance filter retains both tails of a Normal estimate distribution centered at the true difference of one point, with known SE two points."
#| fig-alt: "Normal density centered at one, with a gold line at the true difference labelled above the peak and dashed red lines at minus and plus 3.92. The shaded tails beyond them are labelled: selected estimates below minus 3.92 have the wrong sign, and selected estimates above 3.92 substantially exceed the true effect."
#| fig-width: 6.8
#| fig-height: 2.3
local({
  saved_par <- par(mar = c(4.2, 4.2, 0.6, .7))
  on.exit(par(saved_par))
  x <- seq(-7, 9, length.out = 1601)
  plot(x, dnorm(x, 1, 2), type = "l", lwd = 2, col = ink, ylim = c(0, .235),
       xlab = "Estimated difference (points)", ylab = "Density")
  for (selected in list(x[x < -3.92], x[x > 3.92])) {
    polygon(c(selected, rev(selected)),
            c(dnorm(selected, 1, 2), rep(0, length(selected))),
            col = "#F4C7C7", border = NA)
  }
  lines(x, dnorm(x, 1, 2), lwd = 2, col = ink)
  abline(v = 1, col = gold, lwd = 2)
  abline(v = c(-3.92, 3.92), col = red, lty = 2)
  text(1, .222, "true difference = 1", pos = 2, offset = 0.3, cex = .8,
       col = gold)
  text(-3.92, .105, "-3.9200", pos = 2, cex = .8, col = red)
  text(3.92, .105, "3.9200", pos = 4, cex = .8, col = red)
  text(-5.6, .055, "selected:\nwrong sign", cex = .72, col = red)
  text(6.9, .065, "selected:\nexaggerated", cex = .72, col = red)
})
```

The selected positive estimates are much larger than the truth; the
selected negative estimates have the wrong sign. The underlying estimand
stays at one point, while selection changes which estimates are retained.
The estimator's original unbiasedness and its interval's unconditional
coverage need not hold within this selected set.

[]{#app-ch03-planning}

Running several tests raises a related problem. Each test at level .05
carries its own 5% chance of a false rejection, so five independent
tests of true nulls have probability $1-.95^5=.2262$ of at least one.
Chapter 4 develops how a declared family of comparisons controls that
chance.

\Needspace{16\baselineskip}

### Report the result with its conditions and scope

| Item | Result |
|:--|:--|
| Design and follow-up | 120 consenting students, 60 randomly assigned to each format; all 120 delayed scores observed |
| Mean difference, Special minus Usual | 3.1 points (27.1 versus 24.0) |
| Standard error of the difference | 1.1325 points |
| 95% confidence interval | [0.8572, 5.3428] points |
| Two-sided test from the same procedure | $t(118)=2.7372$, $p=.0072$ |
| Benchmark for potential importance | 2 points; fixed before outcome access, course-specific |

: The invented special-practice study and its results. {#tbl-ch03-study-record tbl-colwidths="[44,56]"}

The table assembles results already developed; the group summaries stay
in @tbl-ch03-observed. A useful results paragraph explains how they
relate, rather than substituting the phrase “Special Practice worked”
for the comparison:

> Among the 120 consenting first-year students randomly assigned to
> Special Practice or Usual Practice, the mean delayed-assessment
> score was 3.1 points higher under Special Practice (27.1 versus
> 24.0 on the 0--40 assessment), with an approximate 95% interval
> from 0.8572 to 5.3428 points and $p=.0072$ for the two-sided test of
> no difference. The comparison and the two-point benchmark for
> practical importance were chosen before the outcomes were seen. The
> interval excludes zero but spans the benchmark, so the direction is
> well supported while the practical importance of the difference
> remains unresolved. The result concerns these participants, this
> assessment two weeks after four sessions, and the equal-variance
> working procedure; it does not by itself speak to attendance,
> mechanisms, durable learning, or students elsewhere.

Random assignment supports the participant comparison, and complete
follow-up avoids missing outcomes for this assessment, as
[the box in Section 5.3](#ch03-justify-equal-variance) records. This
combination of estimates, conditions, and scope makes the result
interpretable [@appelbaum2018].

Suppose the researchers now replace the primary assessment because a
different one gives a larger difference. They have changed the outcome
and selected it after inspection. The report should identify the
original result, the new comparison, the timing and reason for the
change, and the analyses considered. The new result can motivate
further work; its ordinary p-value cannot be presented as if it arose
from the original advance-specified procedure.

## Reading Bridge {#sec-ch03-bridge}

This chapter used a stated repetition to quantify uncertainty about one
estimate. We specified the estimand and the estimator, named the
repetition behind the uncertainty, and calculated standard errors,
pooling two group variances under a common spread. We built confidence
intervals, read their confidence level, and carried out tests from the
same procedure, first for one mean and then for the practice study. The
t tests formed one family, from one sample through paired scores to two
independent groups with equal or unequal spreads, and the same test
logic carried over to the association between two categorical
variables through expected counts and the chi-square distribution,
with a three-way table read as two-way tables within layers. Pairing,
clustering and the school-blocked
design showed that the units which vary independently, not the count of
records, carry the uncertainty. We closed by reading magnitude, the
errors a decision rule can make and the choices made after seeing
results together in one qualified conclusion.

Chapter 4 compares several groups at once. It extends the pooled
variance to several groups as the within-group mean square, splits
variation into between-group and within-group parts, forms the F ratio,
estimates specific contrasts with a calibration for the family of
comparisons, and reads main effects and interaction in a two-factor
design. The SE, interval, testing and selection ideas developed here
will be used at each step.

### Key terms {.unnumbered}

Use this short reference to revisit the main concepts as needed. The
Part One Concept Reference gives the broader reference and the
principal home of each concept, including the sampling ideas developed
in Chapter 2.

```{=latex}
\begingroup\small
```

| Term and common alternatives | Meaning |
|:--|:--|
| Estimate | A numerical approximation to an estimand, calculated from the available data. A point estimate expresses the approximation as one value. |
| Estimator | The rule used to calculate an estimate from data. Its bias and precision concern how the rule behaves across the repetitions specified by the study's procedure or model. |
| Bias (of an estimator) | The difference between an estimator's expected value over the specified repetitions and the estimand. An unbiased estimator has bias zero; precision describes its spread, a separate quality. |
| Pooled variance ($s_p^2$; pooled within-group variance) | A weighted average of within-group sample variances, each weighted by its degrees of freedom: $s_p^2=\{(n_1-1)s_1^2+(n_0-1)s_0^2\}/(n_1+n_0-2)$ for two groups. It estimates a variance the groups are assumed to share; its square root is the pooled SD. |
| Confidence interval (CI) | An interval expressing uncertainty about an estimand. Under its conditions, the procedure covers that estimand at its stated rate across repetitions; the rate describes the procedure. |
| Critical value | The cut-point of a reference distribution, such as the Normal or t, selected by a confidence or significance level. It multiplies the SE in an interval and marks the rejection region of a test. |
| Coverage | The proportion of repetitions in which an interval procedure contains the estimand under its conditions. The confidence level is the procedure's target coverage, not a probability for one realized interval. |
| Null hypothesis | The statistical proposition used as the reference for a test, such as equality of specified population means. The alternative hypothesis states the departure being considered. |
| Test statistic | A quantity calculated from data to assess departure from a null hypothesis. The test compares it with a reference distribution justified under that null and stated conditions. |
| p-value | Under a specified null hypothesis and conditions, the probability of a test result at least as extreme as the observed one, using the test's stated tail rule. |
| Significance level (alpha) | The threshold used for rejecting a null hypothesis, chosen to control the probability of rejecting a true null under the test's conditions. This error is called a Type I error. |
| One-sample z test; one-sample t test | Both compare a sample mean with a stated null mean using its SE; z uses a known process SD under its model, whereas t estimates that SD and uses the appropriate t reference. Neither choice follows an automatic sample-size cutoff. |
| Paired-samples t test | A one-sample t test of within-pair differences against a specified mean difference, with independent pairs under the stated model. |
| Equal-variance two-sample t test (pooled t test; Student's two-sample t test) | One of the two independent-samples t tests. It compares two independent group means under the condition that the populations share one variance: its SE is $s_p\sqrt{1/n_1+1/n_0}$, and its reference is t with $n_1+n_0-2$ df. |
| Welch's unequal-variance t test (Welch test) | The other independent-samples t test. It compares two independent group means without assuming a common variance: its SE adds the two estimated mean variances, and its approximate t reference uses Welch--Satterthwaite df. With equal group sizes its statistic equals the equal-variance statistic, and only the df can differ. |
| Expected count | The count a cell of a two-way table would have under a stated null pattern such as independence, given the observed margins: row total times column total, divided by the overall total. |
| Pearson's chi-square statistic ($X^2$) | The sum over cells of squared differences between observed and expected counts, each divided by its expected count, compared with a chi-square reference with $(R-1)(C-1)$ degrees of freedom under the stated model. |
| Multiway table (three-way table; layered two-way tables) | A table that cross-classifies three or more categorical variables: a two-way table holds two variables, a three-way table three, and a four-way table four. A three-way table is read as two-way tables within the layers of the third variable; a chi-square test of one two-way table cannot test the variables jointly, which needs a method that uses all three variables at once, such as a model for multiway tables. |
| Standardized mean difference | A difference in means divided by a specified SD. The chosen denominator determines the standardized quantity; the raw difference retains the outcome's original units. |
| Power (statistical power) | The probability of rejecting the null under a specified alternative, design, and test. Failure to reject under that alternative is a Type II error; power is one minus its probability. |

: Selected concepts for estimates, uncertainty, and tests. {#tbl-ch03-key-terms tbl-colwidths="[28,72]"}

```{=latex}
\endgroup
```

### Further reading {.unnumbered}

The GAISE College Report emphasizes statistical investigation,
conceptual understanding, and interpretation [@asa2016gaise].
*Introduction to Modern Statistics* develops inference through
repetition, with separate treatments of paired means and two-way
tables [@cetinkaya2024]. [Penn State STAT 200, Lesson 4](https://online.stat.psu.edu/stat200/Lesson04)
connects interval construction with its interpretation. The ASA
p-value statement [@wasserstein2016] and [Moving to a World Beyond
$p<.05$](https://doi.org/10.1080/00031305.2019.1583913) explain why
uncertainty and context deserve more attention than a threshold label.


## Exercises {.unnumbered}

These exercises use the information supplied and a standard calculator. They practise interpretation, the reading of the chapter's figures and tables, and the conditions behind a conclusion, with a short calculation where the reasoning needs it.

<!-- prompt-id: CH03-E01 -->
### Exercise 1: Name the inferential objects (Foundational) {.unnumbered #ex-ch03-01}

A report states:

> The estimand was the mean Special-minus-Usual assignment contrast in
> delayed-assessment score among the 120 randomized participants. In every
> admissible repetition, the analyst would subtract the Usual assigned-group
> mean from the Special assigned-group mean. The observed difference was
> 3.1 points, with an estimated standard error of 1.1325 points.

Identify:

1. the estimand;
2. the estimator;
3. the estimation process;
4. the estimate and one statistic used in obtaining it;
5. the standard error and what varies in its repeated-process interpretation;
6. the parameter; and
7. two bounds that must remain in the answer.

Explain why the estimate and the parameter should not be given the same label.

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH03-E02 -->
### Exercise 2: Match the SE to the unit of assignment (Foundational) {.unnumbered #ex-ch03-02}

An invented first-year course compared Special and Usual Practice on a common
0--40 assessment taken two weeks after four practice sessions. Its 120
students belonged to six tutorial classes of 20. The course randomly
assigned whole classes, three to each format, so every student
practised in the format given to their class. Each format therefore
has 60 students. The Special and Usual Practice means are 27.1 and
24.0 points, a difference of 3.1 points, and their delayed-score
SDs are 6.0 and 6.4 points. The three Special Practice classes had
means of 29.4, 26.6 and 25.3 points; the three Usual Practice classes
had means of 26.3, 24.2 and 21.5 points.

An analyst proposes this SE for the Special-minus-Usual difference,

$$
\widehat{SE}=\sqrt{\frac{s_1^2}{n_1}+\frac{s_0^2}{n_0}},
$$

with $n_1=n_0=60$ students.

1. What changed from assigning students one at a time? Name the
   independent units of assignment, and explain whether the proposed
   SE describes the uncertainty of the Special-minus-Usual difference
   under this design. Does the change affect the estimate, its SE, or
   both?
2. With these inputs the proposed SE is 1.1325 points. How many units
   of assignment does this calculation assume, and what would have to
   be true of the students' scores for it to describe the uncertainty
   under class assignment?
3. Treat each class mean as one observation. The SDs of the three
   class means are 2.0952 points for Special Practice and 2.4062 points
   for Usual Practice. Calculate the SE from the same formula with
   $n_1=n_0=3$ classes and compare it with 1.1325 points. On a t
   reference with 4 df, the 95% interval for the difference is
   $[-2.0145,\ 8.2145]$. Which conclusion about Special Practice
   remains supported?
4. Which quantities describe variation among students' scores, which
   describe variation among class means, and which estimates variation
   of the mean difference across possible assignments of the classes?

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH03-E03 -->
### Exercise 3: Interpret interval coverage (Foundational) {.unnumbered #ex-ch03-03}

A student writes:

> The 95% confidence interval [0.8572, 5.3428] means there is a 95%
> probability that the true mean assignment effect lies between
> 0.8572 and 5.3428 points. If the method is valid, every 95%
> interval should contain the true effect.

Rewrite the statement using repeated-procedure language. Then:

1. The coverage figure in Section 3.2 draws the same 40 district
   samples twice: at 95%, 38 of the 40 intervals cover the fixed
   population mean; at 90%, 36 do. Explain why more intervals miss
   at 90% although no sample changed, and why one study cannot tell
   whether its own interval is one of the misses.
2. Explain what the endpoints contribute to reading this study.
3. Name three sources of uncertainty or bias that the conventional
   interval may omit.

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH03-E04 -->
### Exercise 4: Translate a p-value (Foundational) {.unnumbered #ex-ch03-04}

For the matching two-sided zero-contrast test, *t*(118) = 2.7372 and
*p* = .0072. Consider four interpretations:

A. There is a 0.72% probability that the null hypothesis is true.

B. There is a 99.28% probability that Special Practice will work again.

C. Under the zero-contrast null model and the stated procedure, results at
least as incompatible as the observed one have probability about .0072.

D. The 3.1-point effect is educationally important because *p* = .0072.

Choose the defensible interpretation and diagnose each rejected one.
Then:

1. Section 4.4 re-assigns the 120 students' scores at random to two
   groups of 60, 10,000 times, supposing that no student's score
   depends on format; 77 re-assignments give a difference at least 3.1
   points from zero.
   State what the re-assignments held fixed and what they re-drew,
   and what the share 77/10,000 = .0077 estimates.
2. Explain why .0077 is close to, but need not equal, the t test's
   *p* = .0072.
3. The 95% interval from the same procedure as the t test is
   [0.8572, 5.3428]. Explain why its exclusion of zero agrees with
   the two-sided *p* < .05 result, without implying exact
   randomization calibration.
4. Write one sentence reporting estimate, interval, and p-value
   without using **prove**, **accept**, or **worked**.

No calculation is needed; the tail areas and the interval are
supplied.

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH03-E05 -->
### Exercise 5: Interpret one published active-learning result (Integrative) {.unnumbered #ex-ch03-05}

Deslauriers and colleagues (2019) used a randomized crossover study in introductory
college physics: students were randomly assigned to two groups, and each
group experienced both instructional conditions in successive class meetings,
in opposite orders. In a regression model, performance on a test of learning was
0.46 SD higher under active instruction than under passive instruction,
*p* < .001. The brief result supplied here does not report a confidence
interval [@deslauriers2019].

Use the supplied description to:

1. mark the participant set, condition contrast, outcome, estimate, units,
   p-value, and design;
2. explain what the estimate and p-value each tell you;
3. identify the information supplied and what else would help you understand
   the result; and
4. state a bounded one-sentence version of the reported answer.

Then explain which design and analysis details would help judge the uncertainty
calculation and scope. State what further information you would seek in the
Methods or supplement and name one broader claim this excerpt cannot support.



\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH03-E06 -->
### Exercise 6: Repair the special-practice conclusion (Integrative) {.unnumbered #ex-ch03-06}

Rewrite:

> Special Practice worked, *p* < .05.

Use this study record (the same one summarized in the chapter's study-and-results table):
120 consenting students in one course were randomly assigned, 60 per group,
to four sessions of Special or Usual Practice with equal total study time.
All completed a common 0--40 assessment two weeks later. Special had mean
27.1 (SD 6.0), and Usual had mean 24.0 (SD 6.4). The prespecified primary
comparison gave 3.1 points, 95% CI [0.8572, 5.3428], approximate equal-variance
*t*(118) = 2.7372, *p* = .0072. Before outcome access, the protocol specified
the two-point benchmark as potentially important: the course team treated two
points as the smallest average difference that would change its recommendation,
because the marking scheme awards two points to one completed application task
and the new format would add little preparation time. Your revision must
retain:

1. the participant set and assignment comparison;
2. the outcome, time, direction, and raw units;
3. the estimate and 95% interval;
4. the reported numerical p-value only after the estimate and interval;
5. the choice of the primary comparison before outcome access;
6. what random assignment and complete follow-up support;
7. the interval's relation to the prespecified, course-specific two-point
   benchmark for potential importance;
   and
8. at least three scope boundaries.

After the rewrite, choose three phrases and explain what each repairs in the
original sentence.

After viewing the outcomes, the researchers decide that a different assessment
should replace the planned assessment as primary because it gives a larger
difference. Explain how this changes the comparison and its interpretation.
State what information should be reported about the original and new analyses,
what advance-planning claim can no longer be made, and why the new assessment's
p-value cannot be read as if that assessment had been chosen before inspection.

<!-- prompt-id: CH03-E07 -->
### Exercise 7: Check expected counts and a chi-square result (Foundational) {.unnumbered #ex-ch03-07}

An invented study records participation format and completion. It is
distinct from the UCI student data.

| Participation format | Completed | Did not complete | Total |
|:--|--:|--:|--:|
| Online | 84 | 36 | 120 |
| In person | 63 | 17 | 80 |
| Total | 147 | 53 | 200 |

: Participation-format outcomes for Exercise 7. {#tbl-ch03-ex07-counts tbl-colwidths="[34,22,28,16]"}

The expected counts under independence are 88.2, 31.8, 58.8,
and 21.2.
The Online/Completed contribution is $(84-88.2)^2/88.2=0.2$; the
four contributions (0.2000, 0.5547, 0.3000, 0.8321) sum to
$X^2=1.8868$ on $(2-1)(2-1)=1$ df, giving $p=.1696$.
Report calculated results to four decimal places. The completion proportions are
$84/120=.7000$ and $63/80=.7875$.

1. Explain what the Online/Completed expected count, 88.2,
   represents, using the overall completion proportion
   $147/200=.735$ and the 120 online learners, and why the table has
   one degree of freedom.
2. Identify the cells in which the observed count exceeds the expected count.
3. State the omnibus null in words.
4. Write one conclusion that distinguishes the observed proportion
   difference, evidence against independence, association magnitude,
   and causation, and explain why $p=.1696$ is not proof that format
   and completion are independent.
5. Suppose every count were doubled, so that the completion
   proportions stay .7000 and .7875. Predict what happens to the
   expected counts, to each cell's contribution, to $X^2$ and to the
   proportion difference, then check your prediction against
   $X^2(1)=3.7736$, $p=.0521$ for the doubled table. What does this
   say about reading $X^2$ or its p-value as the strength of an
   association?
6. Suppose the same 200 learners had also been recorded as working or
   not working while they studied (@tbl-ch03-ex07-layers, invented).
   Compare the two formats' completion proportions within each layer
   with the added table's .7000 and .7875, and explain why the added
   In-person-minus-Online difference, .0875, is larger than both layer
   differences. Then name one question about format, completion and
   work that no chi-square test of a single two-way table in the
   display can answer.

\Needspace{12\baselineskip}

| Work while studying | Participation format | Completed | Did not complete | Total | Completion proportion |
|:--|:--|--:|--:|--:|--:|
| Working | Online | 52 | 28 | 80 | .6500 |
|  | In person | 14 | 6 | 20 | .7000 |
| Not working | Online | 32 | 8 | 40 | .8000 |
|  | In person | 49 | 11 | 60 | .8167 |

: Invented counts of completion by participation format within two layers of work while studying, for Exercise 7's 200 learners. {#tbl-ch03-ex07-layers tbl-colwidths="[25,19,15,16,9,16]"}


## Appendix A: Probability and Repetition Foundations {.unnumbered #app-probability}

This appendix derives two results behind the chapter's uncertainty
calculations, each for its stated repetition: drawing observations from
a model, or reassigning the same participants.

### Variance, sample variance, and least squares {.unnumbered #app-ch03-variance}

The variance of a random variable is its expected squared distance from
its expectation: $\operatorname{Var}(Y)=E\{(Y-E(Y))^2\}$. For fixed
weights $a_i$ and observations with finite second moments,

$$
\operatorname{Var}\left(\sum_i a_iY_i\right)
=\sum_i a_i^2\operatorname{Var}(Y_i)
+2\sum_{i<j}a_ia_j\operatorname{Cov}(Y_i,Y_j).
$$

Here covariance is the expected product of two centered observations,
and $i<j$ includes each distinct pair once. Expanding the square of the
weighted sum gives the squared terms and the cross-products in this
identity. For an average, every weight is $1/n$; independent
observations have zero covariance, so with common variance $\sigma^2$
the expression becomes $n\sigma^2/n^2=\sigma^2/n$. For a difference,
weights $1$ and $-1$ instead give
$\operatorname{Var}(U)+\operatorname{Var}(V)-2\operatorname{Cov}(U,V)$.
These two applications explain the mean, independent-group, and
paired-difference calculations, and the covariance terms are those
Section 2.3 uses for pairing, clustering and the design effect.

Estimating $\sigma^2$ uses the sample variance
$s^2=\sum_i(Y_i-\bar Y)^2/(n-1)$. For independent observations with
common mean $\mu$ and variance $\sigma^2$,

$$
\sum_i(Y_i-\bar Y)^2=\sum_i(Y_i-\mu)^2-n(\bar Y-\mu)^2 .
$$

The first term on the right has expectation $n\sigma^2$ and the second
$n(\sigma^2/n)=\sigma^2$, so $E(s^2)=(n\sigma^2-\sigma^2)/(n-1)=\sigma^2$.
For the values 2, 4 and 6 the mean is 4 and the squared deviations sum
to 8, so $s^2=8/2=4.0000$, whereas dividing by 3 gives 2.6667.
Normality is unnecessary for this result, and an unbiased $s^2$ does not
make $s$ unbiased for $\sigma$. The same algebra with observed values
and any proposed center $a$ gives

$$
\sum_i(y_i-a)^2=\sum_i(y_i-\bar y)^2+n(\bar y-a)^2 ,
$$

whose last term is nonnegative and vanishes at $a=\bar y$: the mean
minimizes the sum of squared deviations, the **least-squares** property
Chapter 2 states.

### Random assignment and the familiar SE {.unnumbered #app-ch03-assignment}

For the fixed $N=120$ Special Practice participants, let $S_1^2$ and
$S_0^2$ be the potential-outcome variances under the two assignments,
with divisor $N-1$. Let $S_{\mathrm{eff}}^2$ be the variance of the
individual effects $Y_{i1}-Y_{i0}$, using the same divisor. Under
complete random assignment of $n_1$ and $n_0$ participants, consistent
conditions and no relevant interference [@bai2024],

$$
\operatorname{Var}_{\mathrm{assign}}(\bar Y_1-\bar Y_0)
=\frac{S_1^2}{n_1}+\frac{S_0^2}{n_0}-\frac{S_{\mathrm{eff}}^2}{N},
\qquad
E_{\mathrm{assign}}\left(\frac{s_1^2}{n_1}+\frac{s_0^2}{n_0}\right)
=\operatorname{Var}_{\mathrm{assign}}(\bar Y_1-\bar Y_0)+\frac{S_{\mathrm{eff}}^2}{N}.
$$

Only one potential outcome is observed per participant, so the
individual-effect variance is unidentified. The familiar summed
variance is therefore conservative in expectation under this
complete-assignment design, with equality when individual effects are
constant; with equal group sizes it equals the pooled squared SE of
Section 2.2. A particular estimate need not be too large, and the
property does not by itself make the equal-variance or Welch t reference
exact or ensure conservative finite-sample coverage.

A zero average effect allows positive and negative individual effects
to cancel, whereas the stronger **sharp null** says that every
participant's effect is zero. The re-assignments of Section 4.4 test
that sharp null by reproducing the actual assignment mechanism, so
their reference owes nothing to a Normal or t model.

## Appendix B: Inferential Extensions {.unnumbered #app-procedures}

This appendix gathers two references for the chapter's procedures: a
table of their conditions and the steps of a bootstrap, which the
checks of the procedures call on.

### Procedures and their conditions {.unnumbered #app-ch03-procedure-guide}

@tbl-ch03-procedures collects the five procedures of Sections 5 and 6
with the conditions each relies on, the evidence that checks them and the
response when a condition fails. Each procedure's choices and defaults
stay with its *Justifying the analysis* box, because they vary with the
study rather than the procedure.

```{=latex}
\begingroup\footnotesize
```

| Procedure and target | Secured by design | Model assumptions and how to check them | If a condition fails | Main explanation |
|:--|:--|:--|:--|:--|
| One-sample t test: the mean $\mu$ of the sampled population, in outcome units, compared with a stated null value | Random sampling from the named population; independent units, each measured once | A Normal population gives an exact t reference; otherwise the sample mean needs a near-Normal sampling distribution. Look for strong skew, heavy tails or one dominating value in the scores; no sample-size cutoff settles the question | Dependent units: an SE that represents the pairs or clusters (Ch. 3, Sec. 2.3). Strong skew with few units: disclose the weaker approximation and compare a bootstrap interval (Ch. 3, App. B). Without random sampling, a population reading needs a stated model | Ch. 3, Sec. 5.1 |
| Paired-samples t test: the mean within-pair difference $\mu_D$, in outcome units, with the subtraction direction stated | Pairs formed by the design, such as one student measured twice or two classes in one school; independent pairs. Random assignment within pairs supports a causal reading | Near-Normal pair differences give the t reference: inspect the differences themselves rather than the two sets of scores; a few pairs say little about shape | Ignoring the pairing: an independent-groups analysis omits the within-pair covariance and misstates the SE (Ch. 3, Sec. 5.2). No untreated comparison: a mean change alone cannot be separated from time trends or repeated testing. Skewed differences: resample whole pairs (Ch. 3, App. B) | Ch. 3, Sec. 5.2 |
| Equal-variance two-sample t test: a difference between two means, in outcome units, either a mean assignment effect over the participants or a difference between two populations' means | Random assignment of individuals, supporting a causal comparison among the participants, or two independent random samples; different units in the two groups | One common population variance and a near-Normal sampling distribution of the difference: compare the group SDs with the group sizes and look for skew or dominating values in each group. After random assignment the t reference is approximate. No interference between units is argued from the setting; the data cannot check it | Unequal spreads with unequal group sizes: the pooled SE is too small when the smaller group varies more, so choose Welch's t in advance from the design (Ch. 3, Sec. 5.4), not from a preliminary variance test. With equal group sizes the two SEs coincide | Ch. 3, Sec. 5.3 |
| Welch's unequal-variance t test: a difference between two means, in outcome units, when the two groups' variances may differ | Two independent random samples or randomly assigned groups, with different units in each group | No common variance is assumed. Each group mean needs a near-Normal sampling distribution, and the Welch--Satterthwaite df approximate the reference: inspect each group's distribution, especially the smaller group's | Skew or a dominating value in a small group: disclose the weaker approximation and compare a bootstrap that resamples within each group (Ch. 3, App. B). Paired or clustered units: an SE that represents them (Ch. 3, Sec. 2.3) | Ch. 3, Sec. 5.4 |
| Pearson chi-square test of association: whether two categorical variables are associated in the stipulated population or process, against the null of no association (independence); conditional proportions give the direction and size | Independent sampling of units, each counted once in one of mutually exclusive categories | Adequate expected counts for the chi-square approximation: calculate every expected count, row total times column total divided by $n$, before reading the p-value | Small expected counts: an exact conditional test such as Fisher's (Ch. 6, Sec. 4). Repeated or clustered records: the counts overstate the independent information. An observed rather than assigned grouping supports association only | Ch. 3, Sec. 6 |

: Procedures and their conditions. {#tbl-ch03-procedures tbl-colwidths="[17,15,26,27,15]"}

```{=latex}
\endgroup
```

### Resampling and uncertainty {.unnumbered #app-ch03-resampling}

A **bootstrap distribution** approximates an estimator's variability by
repeatedly resampling from observed data. Its center is usually near the
observed estimate. In a simple case bootstrap for independent observations:

1. Identify the independent observational unit and the original number of units.
2. Draw that many units with replacement from the observed units.
3. Apply the same estimator to the resampled data.
4. Repeat the resampling and estimation many times.
5. Use the distribution of estimates to examine variability or construct an appropriate interval.

The bootstrap SE is the sample SD of the resulting estimates. It is
distinct from the SD of individual observations and from the uncertainty
caused by using only finitely many simulation repetitions. Preserve the
study structure: resample pairs together, or clusters when independent
clusters are the appropriate units. Resampling individual students from
clustered classes can misrepresent uncertainty, and the number
resampled is the observed unit count. As Section 4.4 explains, a
bootstrap distribution centered near the estimate is not a reference
for testing zero, and resampling cannot recover unobserved cases or
repair biased selection, invalid measurement, or an unsuitable estimand.

## References {.unnumbered}

\small

::: {#refs}
:::

\normalsize
