---
title: "Group Comparisons and Statistical Modelling with R"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: true
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \widowpenalty=10000
          \clubpenalty=10000
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\linespread{0.95}\selectfont\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{10\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AtBeginDocument{\renewcommand{\boldsymbol}[1]{\symbf{#1}}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---


```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                     collapse = TRUE, comment = "#>")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 10, width = 65, show.signif.stars = FALSE)
```

::: {.callout-note title="Chapter overview"}
Chapter 5 made estimates, intervals and tests executable in base R. Here we
return to Chapter 4's group comparisons and use R to connect a fitted model
with the question it answers. We begin with three condition means from
Chapter 4's background-sound experiment and compare selected differences
inside stated families. We then examine a six-cell pattern from its
feedback-by-practice study and test how one comparison changes across a
second factor by comparing two named models on the same records.

Counts require a different response model, and the size of a difference
between two groups' proportions can be stated on several scales, which we
set side by side. The same counts provide a small example of likelihood,
which makes a fitting criterion explicit. Finally, simulation examines
procedures under a stated generating process, and planning relates sample
size to either power or interval width. Throughout, a useful result keeps
its data, target, uncertainty calculation and conditions together.
:::

Studies rarely stop at one mean. Chapter 4's invented background-sound
experiment compares mean reading comprehension under three conditions,
silence, instrumental music and music with lyrics, and its second
invented study asks how the difference between spaced and massed
practice changes across three kinds of feedback. Means that differ in a
record differ for more than one reason, so the question is what a stated
model and its conditions let us say about several means at once, and
what a single answer to that question leaves unsettled. Both studies are
invented teaching inputs, so each analysis shows what a procedure
reports, not what background sound or feedback does.

## Fit and read group means {#ch06-group-models}

### Begin with the question and records

Does mean reading comprehension differ across three background-sound
conditions? Chapter 4 developed this question with an invented experiment.
Each of thirty first-year university students read the same passage for
20 minutes and then answered a 20-item reading-comprehension quiz, scored
from 0 to 20. Ten students were randomly assigned to each sound condition:
Silence, Instrumental music or Music with lyrics. Each student supplies
one quiz score, so @tbl-ch06-sound-data holds thirty scores from thirty
different students.

| Sound condition | Scores (out of 20) |
|:--|:--|
| Silence | 8, 9, 9, 11, 12, 13, 13, 14, 15, 16 |
| Instrumental music | 6, 8, 8, 9, 9, 11, 11, 12, 12, 14 |
| Music with lyrics | 4, 5, 6, 7, 8, 8, 9, 10, 11, 12 |

: The same 30 invented comprehension scores as Chapter 4. {#tbl-ch06-sound-data tbl-colwidths="[30,70]"}

These values were constructed to keep the calculations transparent. They
are teaching inputs, not empirical findings about music and reading. The
student is both the unit of random assignment and the unit of
measurement: the ten scores in a condition come from ten different
students, not from repeated measurements of one student. Silence is the
natural reference condition, so it comes first.

The following code creates the complete data. The factor levels fix the
display order, make Silence the reference level for later model
coefficients and fix the direction of each subtraction. The course data
file `background_sound.csv` holds the same thirty rows with a student
identifier; reading it with `read.csv()` and applying the same `factor()`
levels gives identical results.

```{r}
#| label: ch06-sound-setup
sound_data <- data.frame(
  sound = factor(rep(c("Silence", "Instrumental music",
    "Music with lyrics"), each = 10),
    levels = c("Silence", "Instrumental music",
      "Music with lyrics")),
  score = c(8, 9, 9, 11, 12, 13, 13, 14, 15, 16,
    6, 8, 8, 9, 9, 11, 11, 12, 12, 14,
    4, 5, 6, 7, 8, 8, 9, 10, 11, 12))
table(sound_data$sound)
with(sound_data, tapply(score, sound, mean))
with(sound_data, tapply(score, sound, sd))
```

Each condition holds ten students. The condition means are 12, 10 and 8
correct answers, and the grand mean of all thirty scores is 10. The SDs,
2.7080, 2.4037 and 2.5820, are similar without being identical. Before
asking whether a difference is large relative to its uncertainty, we can
see each difference on the quiz's own scale: Instrumental music lies
2 points and Music with lyrics 4 points below Silence.

@fig-ch06-sound-groups shows every student beside the three condition
means. Each point is one student's score, moved a small random distance
sideways so that equal scores in a condition stay visible, and each
diamond marks a condition mean. Two kinds of variation appear together.
The diamonds step down by 2 points from Silence to Instrumental music and
by another 2 points to Music with lyrics, while the ten scores in each
condition spread across 8 points around their own diamond. Whether a
2-point step is large depends on that spread within conditions, and the
model below turns this visual comparison into numbers.

```{r}
#| label: fig-ch06-sound-groups
#| echo: false
#| fig-width: 6.4
#| fig-height: 3.5
#| fig-cap: "Individual comprehension scores and their three condition means."
#| fig-alt: "Ten scores in each of Silence, Instrumental music and Music with lyrics, spread slightly sideways so that tied scores stay visible, lie around means 12, 10 and 8. Larger diamonds mark the means; the three spreads are similar."
library(ggplot2)
set.seed(12345)
group_means <- aggregate(score ~ sound, sound_data, mean)
p <- ggplot(sound_data, aes(sound, score))
p + geom_jitter(width = 0.15, height = 0, size = 2) +
  geom_point(data = group_means, shape = 18, size = 5,
    colour = "black") +
  labs(x = "Background sound",
    y = "Comprehension score (out of 20)") +
  theme_minimal(base_size = 11)
```

The figure is drawn with the `ggplot2` package, which describes a graph
as a data frame, a mapping of variables to position or colour, and layers
of marks; Computer Lab 6 builds it one layer at a time and sets this way
of working beside base R and the wider tidyverse.

### Fit the group model and read its decomposition

For student $i$ in sound condition $j$, the one-way model is

$$
Y_{ij}=\mu_j+\varepsilon_{ij}.
$$

Here $\mu_j$ is the condition's process mean and $\varepsilon_{ij}$ is its
student's deviation from that mean. Fitting by least squares estimates
each $\mu_j$ with the corresponding sample mean, so
$\widehat Y_{ij}=\bar Y_j$. The residual is $e_{ij}=y_{ij}-\bar y_j$: an
observed score minus its fitted condition mean. For Silence the fitted
values are all 12 and the residuals are $-4,-3,-3,-1,0,1,1,2,3,4$.

The model separates differences between conditions from differences
among students within a condition. Chapter 4's decomposition becomes

$$
\begin{aligned}
SS_T&=\sum_{j}\sum_i(y_{ij}-\bar y)^2=SS_B+SS_W,\\
SS_B&=\sum_j n_j(\bar y_j-\bar y)^2,\\
SS_W&=\sum_j\sum_i(y_{ij}-\bar y_j)^2.
\end{aligned}
$$

The total measures deviation from the grand mean. The between-group term
measures deviation of the three condition means from that grand mean,
weighted by their group counts. The within-group term measures deviation
of students from their own fitted condition means. In this example,
$SS_B=10(12-10)^2+10(10-10)^2+10(8-10)^2=80$. The squared residuals
within Silence, Instrumental music and Music with lyrics sum to 66, 52
and 60, giving $SS_W=178$ and $SS_T=80+178=258$. In
@fig-ch06-sound-groups these are the steps between the diamonds and the
spread of the points around them.

The formula `score ~ sound` tells `aov()` to explain score using the
sound factor, and `anova()` extracts the decomposition from the fitted
object.

\Needspace{10\baselineskip}

```{r}
#| label: ch06-oneway-fit
one_fit <- aov(score ~ sound, data = sound_data)
anova(one_fit)
```

Three fitted condition means use three parameters. Consequently the
within-group degrees of freedom are $30-3=27$. The between-group component
has $3-1=2$ degrees of freedom, and the total has $30-1=29$. Dividing each
sum of squares by its degrees of freedom gives a mean square. The
within-group mean square $MS_W=178/27=6.5926$ estimates the common error
variance.

| Source | SS | df | MS | F | p |
|:--|--:|--:|--:|--:|--:|
| Sound condition | 80 | 2 | 40.0000 | 6.0674 | .0067 |
| Within groups | 178 | 27 | 6.5926 | | |
| Total | 258 | 29 | | | |

: One-way ANOVA on the thirty background-sound scores. {#tbl-ch06-oneway}

The null hypothesis is $H_0:\mu_S=\mu_I=\mu_L$ for Silence, Instrumental
music and Music with lyrics. Under independent Normal errors with a common
variance, the ratio

$$
F=\frac{MS_B}{MS_W}=\frac{40}{178/27}=6.0674
$$

has an $F_{2,27}$ reference distribution when that null holds. The p-value
.0067 is the probability, under this null model, of an F statistic at least
this large. At a .05 threshold it provides evidence against equal process
means. It does not say that every pair of conditions differs, that every
kind of background sound changes comprehension, or that the invented
result establishes a real effect of music on reading. The next section
asks the individual comparison questions.

The observed share of score variation associated with these condition
means is

$$
\eta^2=\frac{SS_B}{SS_T}=\frac{80}{258}=.3101.
$$

This is an in-sample variation summary, accompanying differences expressed
in score points. It is neither the probability that background sound
caused a result nor a replacement for a confidence interval. The
bias-adjusted omega-squared of Chapter 4, Appendix B.2, is .2525 for
these scores, and the Computer Lab rebuilds both summaries from the scores.

### Express the same fit with indicators

A factor with three levels can also enter a linear model as two indicator
variables. Let $D_I=1$ for Instrumental music and 0 otherwise, and let
$D_L=1$ for Music with lyrics and 0 otherwise. Silence has both indicators
0. The fitted equation is

$$
\widehat y=12-2D_I-4D_L.
$$

Substituting the indicators gives 12 for Silence, $12-2=10$ for
Instrumental music and $12-4=8$ for Music with lyrics. The intercept is
the reference mean; each other coefficient is a difference from that mean,
negative here because both music conditions have lower means than Silence.
The factor order in the original setup makes Silence the reference level.

```{r}
#| label: ch06-indicator-fit
indicator_fit <- lm(score ~ sound, data = sound_data)
coef(indicator_fit)
```

`lm()` and `aov()` describe the same set of fitted means, so they give the
same residuals and the same ANOVA table. The formula names the outcome and
the explanatory variable; the study's design supplies the independence
assumption. A different reference level changes which comparisons the
coefficients name, but not the fitted means, the residuals or the table:
with Instrumental music as the reference, the coefficients become 10, 2
and $-2$. This indicator representation prepares the named models of
Section 3.2. Chapter 7 develops a continuous predictor's slope, and
Chapter 8 extends the conditional interpretation to several predictors.

## Ask comparisons within stated families {#ch06-contrast-families}

### Specify the comparison and its standard error

Suppose the two questions compare each music condition with Silence:
Instrumental music versus Silence, and Music with lyrics versus Silence.
In the order Silence, Instrumental music, Music with lyrics, their
coefficient vectors are $(-1,1,0)$ and $(-1,0,1)$. Each vector sums to
zero. A quantity

$$
C=\sum_jc_j\mu_j,\qquad \sum_jc_j=0,
$$

is a contrast of the process means. Its estimate replaces process means with
the observed group means:
$\widehat C=\sum_jc_j\bar y_j$. Thus our estimates are $-2$ and $-4$
points: both music conditions lie below Silence. Calling a comparison a
contrast specifies this mathematical form; it does not say whether the
comparison was planned before seeing the results or selected afterward.
That is additional information about the analysis.

For independent groups with a common error variance, Chapter 4 extends
Chapter 3's pooled SE, $s_p\sqrt{1/n_1+1/n_0}$, from two means to a
contrast by replacing $s_p^2$ with $MS_W$:

$$
SE(\widehat C)=\sqrt{MS_W\sum_j\frac{c_j^2}{n_j}}.
$$

Each comparison uses two of the means, so both have estimated SE
$\sqrt{(178/27)(1/10+1/10)}=1.1483$ points. The third condition still
contributes to the common error estimate and its 27 residual degrees of
freedom. This matters when comparing these results with an analysis
restricted to only two conditions.

### Calibrate the two-comparison family

An individual 95% interval uses a .05 error probability for that one
comparison. The declared family here is *each music condition against
Silence*. To cover both of its comparisons with at least 95% familywise
confidence, Bonferroni allocates .025 to each two-sided comparison. Each
tail then receives $.05/(2\times2)=.0125$. The calibration uses the $t$
reference distribution with the model's 27 residual degrees of freedom
[@student1908], giving the critical value

$$
t_{1-.05/(2\times2),27}=2.3734.
$$

The calculations retain the same estimates and SEs but multiply each SE by
this larger critical value. The following code extracts the common variance
estimate from the fitted object and calculates both results from unrounded
inputs.

\Needspace{8\baselineskip}

```{r}
#| label: ch06-bonferroni
mse <- deviance(one_fit) / df.residual(one_fit)
means <- with(sound_data, tapply(score, sound, mean))
estimate <- c(Instrumental = unname(means[2] - means[1]),
  Lyrics = unname(means[3] - means[1]))
se <- sqrt(mse * (1/10 + 1/10))
critical <- qt(1 - .05/(2*2), df.residual(one_fit))
raw_p <- 2 * pt(-abs(estimate/se), df.residual(one_fit))
cbind(estimate, SE = se, lower = estimate-critical*se,
  upper = estimate+critical*se,
  adjusted_p = p.adjust(raw_p, method = "bonferroni"))
```

| Comparison | Estimate | SE | Simultaneous interval | Adjusted p |
|:--|--:|--:|:--|--:|
| Instrumental music minus Silence | $-2$ | 1.1483 | [$-4.7253$, 0.7253] | .1859 |
| Music with lyrics minus Silence | $-4$ | 1.1483 | [$-6.7253$, $-1.2747$] | .0034 |

: Bonferroni results for the declared family of two comparisons. {#tbl-ch06-bonferroni tbl-colwidths="[34,13,11,28,14]"}

Multiplying each raw p-value by two and capping at one gives the
corresponding Bonferroni-adjusted p-value. Music with lyrics minus Silence
is distinguishable from zero with this familywise calibration: its whole
interval lies below zero. The Instrumental-music interval includes zero
although the omnibus ANOVA rejected equality. It allows a difference from
4.7253 points below Silence to 0.7253 points above it. The two results are
compatible, because the omnibus test and each contrast address different
statements.

The family calibration has a purpose and a price. Chapter 4, Section 3.3,
defined the familywise error rate as the probability that a declared
family produces at least one false rejection when every null in it holds.
Two independent comparisons, each calibrated at .05, would leave that rate
at $1-.95^2=.0975$. These two comparisons share the Silence mean, and
dividing .05 across the family bounds the rate at .05 under any
dependence between them. The guarantee is paid for in width, which
Section 2.3 compares, and in power. Suppose a replication of this study
used three balanced groups and took this study's error SD,
$\sqrt{178/27}=2.5676$, as its planning value. A 2-point difference would
first reach .8 power at 27 students per group when the comparison stands
alone, and at 33 per group when it belongs to the two-comparison family.
Section 6 shows how such targets are calculated, there with the smaller
planning SD of the feedback study. Declare the family from the question
before seeing the results; enlarging it afterwards spends that power on
questions nobody asked.

### Compare with an all-pairs family

If the question includes every pair, there are three comparisons. Tukey's
procedure uses the joint distribution of pairwise mean differences under the
common-variance Normal-error model. `TukeyHSD()` supplies those simultaneous
intervals and their adjusted p-values [@tukey1953].

\Needspace{8\baselineskip}

```{r}
#| label: ch06-tukey
TukeyHSD(one_fit, "sound", conf.level = .95)
```

The output labels each subtraction direction. Music with lyrics minus
Silence is $-4$, with interval [$-6.847$, $-1.153$] and adjusted p
.0047. Instrumental music minus Silence and Music with lyrics minus
Instrumental music both equal $-2$, with interval [$-4.847$, 0.847] and
adjusted p .2085. @tbl-ch06-calibrations sets one comparison, Instrumental
music minus Silence, under the three calibrations met so far.

\Needspace{8\baselineskip}

| Calibration | Family protected | Multiplier | 95% interval |
|:--|:--|--:|:--|
| Pointwise t | This comparison alone | 2.0518 | [$-4.3561$, 0.3561] |
| Bonferroni | The two comparisons with Silence | 2.3734 | [$-4.7253$, 0.7253] |
| Tukey | All three pairs | 2.4794 | [$-4.8470$, 0.8470] |

: Instrumental music minus Silence, $-2$ points with SE 1.1483, under three calibrations. {#tbl-ch06-calibrations tbl-colwidths="[17,39,16,28]"}

The estimate, $-2$, and its SE, 1.1483, are the same in every row; only
the multiplier changes. A larger protected family needs a larger
multiplier to hold its familywise error rate at .05, so its intervals are
wider. An interval should therefore be read together with the comparison
and the family it represents.

The all-pairs family also covers a comparison chosen after seeing the
means. Suppose Music with lyrics minus Silence were reported because it is
the largest observed difference. That choice looked at all three pairs, so
its calibration would come from the all-pairs family, [$-6.8470$,
$-1.1530$], rather than from the declared two-comparison family,
[$-6.7253$, $-1.2747$]. An omnibus rejection does not by itself protect
any comparison selected after examining the means. Conversely, a
separately specified contrast can have its own valid uncertainty procedure
without requiring omnibus rejection as an automatic preliminary gate. The
relevant information is the question, selection process and calibration,
rather than a universal sequence of significance decisions.

### Check what changes when the variance model changes

The common-variance model assumes that within-group variability estimates
one shared error variance. In a real study, design information and observed
patterns help assess that working model. A small normality or variance test
cannot certify it, and ten students per condition give limited information
about the error distribution. In these invented data the three SDs, 2.7080,
2.4037 and 2.5820, are similar by construction, and every condition's
residuals lie between $-4$ and 4. In @fig-ch06-sound-groups each residual
is a point's vertical distance from its diamond, and the three conditions
spread alike.

For independent groups whose error variances may differ, Welch's one-way
procedure provides an alternative [@welch1947]. It is the several-group
counterpart of Chapter 3's Welch t test, as Chapter 4's conditions section
explains. R selects it by default in `oneway.test()` [@rcore2025]:

\Needspace{8\baselineskip}

```{r}
#| label: ch06-welch-oneway
oneway.test(score ~ sound, data = sound_data)
```

It gives $F(2, 17.9561)=5.5144$ and p .0136, beside the pooled
$F(2,27)=6.0674$ and p .0067. Welch weights each condition mean by
$n_j/s_j^2$, the inverse of its own estimated sampling variance, and
refers the statistic to fewer denominator degrees of freedom because each
condition now estimates its own variance. With similar observed SDs the
two procedures reach the same conclusion at .05, while their p-values
differ. Each p-value belongs to its own variance model and small-sample
calibration, so the two are results to explain through those assumptions
rather than votes on which numerical conclusion to prefer. With small
groups or clearly unequal spreads, the model choice can matter
appreciably.

A useful exact identity connects Chapter 3's equal-variance (pooled)
two-sample t test, requested with `var.equal = TRUE`, with a two-group
ANOVA: $F=t^2$, with the same two-sided p-value. It requires the same records,
equal-variance denominator and group comparison. R's default `t.test()` is
Welch's test, Chapter 3's last t procedure, and does not meet that
specification. Nor is our three-group contrast the same calculation,
because it uses residual information from the third condition. The
Computer Lab reproduces the matched identity for Silence and Instrumental
music.

## Compare a two-way pattern {#ch06-factorial-comparisons}

### Read cells and margins before fitting

Chapter 4's second study asks how one comparison changes across a second
factor. In this invented study, each of 24 pupils was individually
assigned to one combination of feedback (Rereading, Answer feedback or
Explanatory feedback) and practice (Massed or Spaced). The outcome is the
number correct on a 20-item assessment. A cell is one combination of
feedback and practice, and four pupils contribute scores to each cell in
@tbl-ch06-feedback-data. The question is how the Spaced-minus-Massed
comparison changes across feedback conditions.

| Feedback | Massed scores | Spaced scores |
|:--|:--|:--|
| Rereading | 4, 5, 7, 8 | 4, 5, 7, 8 |
| Answer feedback | 6, 7, 9, 10 | 10, 11, 13, 14 |
| Explanatory feedback | 8, 9, 11, 12 | 16, 17, 19, 20 |

: The same 24 authored scores as Chapter 4. {#tbl-ch06-feedback-data tbl-colwidths="[35,32,33]"}

These values were constructed to make the calculations transparent. They are
teaching inputs, not empirical evidence that one feedback condition works
better. Each score comes from a different pupil; the four scores within a cell
are not repeated measurements of one pupil. Repeated patterns in their
deviations were deliberately chosen and cannot establish assumptions about a
population.

The following code creates the complete data. The factor levels specify
the display order and the reference levels for later model coefficients.

```{r}
#| label: ch06-feedback-setup
feedback_data <- data.frame(
  student = 1:24,
  feedback = factor(rep(c("Rereading", "Answer feedback",
    "Explanatory feedback"), each = 8),
    levels = c("Rereading", "Answer feedback",
      "Explanatory feedback")),
  practice = factor(rep(rep(c("Massed", "Spaced"),
    each = 4), 3), levels = c("Massed", "Spaced")),
  score = c(4,5,7,8, 4,5,7,8, 6,7,9,10, 10,11,13,14,
    8,9,11,12, 16,17,19,20))
```

Computing cell counts and means before fitting any model makes the six
cells and their denominators visible.

\Needspace{15\baselineskip}

```{r}
#| label: ch06-cell-summaries
with(feedback_data, table(feedback, practice))
cell_mean <- with(feedback_data,
  tapply(score, list(feedback, practice), mean))
cell_mean
```

\Needspace{12\baselineskip}

@tbl-ch06-cell-means adds the margins and the three practice changes to
these cell means.

| Feedback | Massed | Spaced | Equally weighted margin | Spaced minus Massed |
|:--|--:|--:|--:|--:|
| Rereading | 6 | 6 | 6 | 0 |
| Answer feedback | 8 | 12 | 10 | 4 |
| Explanatory feedback | 10 | 18 | 14 | 8 |
| Practice margin | 8 | 12 | 10 | 4 |

: Cell means and equally weighted margins of the balanced design. {#tbl-ch06-cell-means tbl-colwidths="[31,12,12,23,22]"}

Each cell has four pupils, so averaging the two practice cells equally gives
the feedback group's overall mean. Similarly, each practice margin averages
equally across the three feedback conditions. In an unbalanced design, a
margin must specify its weights: the observed mix of pupils and an equally
weighted target can answer different questions.

@fig-ch06-feedback-interaction draws the six cell means with one line for
each feedback condition. Each line joins two separate cell means; it does
not trace individual pupils across repeated measurements.

```{r}
#| label: fig-ch06-feedback-interaction
#| echo: false
#| fig-pos: "!ht"
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: "The practice difference depends on feedback."
#| fig-alt: "Rereading joins 6 to 6, Answer feedback joins 8 to 12, and Explanatory feedback joins 10 to 18 from Massed to Spaced. The lines have different slopes."
cell_plot <- aggregate(score ~ feedback + practice,
  feedback_data, mean)
ggplot(cell_plot, aes(practice, score, colour = feedback,
  group = feedback)) + geom_line() + geom_point(size = 2) +
  labs(x = "Practice", y = "Mean number correct",
    colour = "Feedback") +
  theme_minimal(base_size = 11)
```

The observed changes are 0, 4 and 8 points, so the three lines have
different slopes. This describes interaction in the observed means.
Inference about a process interaction additionally requires a model and a
measure of residual variation.

### Fit the full model, identify its tests and compare nested models

Chapter 4 wrote the two-way model as

$$
Y_{ijk}=\mu+\alpha_j+\beta_k+(\alpha\beta)_{jk}
  +\varepsilon_{ijk},
$$

where $j$ indexes feedback, $k$ indexes practice and $i$ indexes pupils
within a cell. With the usual sum-to-zero constraints, $\mu$ is the grand
mean, $\alpha_j$ and $\beta_k$ describe departures of the margins from it,
and $(\alpha\beta)_{jk}$ describes the remaining cell departure. The
constraints make this representation identifiable: the effects cannot all
shift freely while leaving the same fitted means.

There is one grand-mean parameter, $3-1=2$ feedback degrees of freedom,
$2-1=1$ practice degree of freedom, and $(3-1)(2-1)=2$ interaction degrees
of freedom. Together these describe six fitted cell means. The model
therefore has $24-6=18$ residual degrees of freedom. In an R formula,
`feedback * practice` includes both main effects and their interaction.
It expands to `feedback + practice + feedback:practice`.

\Needspace{12\baselineskip}

```{r}
#| label: ch06-factorial-fit
two_fit <- aov(score ~ feedback * practice,
  data = feedback_data)
anova(two_fit)
```

\Needspace{9\baselineskip}

| Source | SS | df | MS | F | p |
|:--|--:|--:|--:|--:|--:|
| Feedback | 256 | 2 | 128.0000 | 38.4 | < .0001 |
| Practice | 96 | 1 | 96.0000 | 28.8 | < .0001 |
| Interaction | 64 | 2 | 32.0000 | 9.6 | .0015 |
| Residual | 60 | 18 | 3.3333 | | |
| Total | 476 | 23 | | | |

: Full balanced factorial decomposition. {#tbl-ch06-factorial}

The feedback test concerns equality of feedback margins averaged equally
over practice. The practice test concerns equality of practice margins
averaged equally over feedback. The interaction test asks whether the
practice difference is the same across all feedback conditions. With
independent Normal errors of common variance, its null reference is
$F_{2,18}$; the observed value 9.6 and p .0015 provide evidence against
that common-difference null.

The balanced layout makes the source components orthogonal: the variation
attributed to one component does not compete with another component for
the same fitted direction. A single-model `anova()` table ordinarily uses
sequential sums of squares, entering terms in formula order. This convenient
balanced decomposition should not be extended automatically to unbalanced
data.

To see why, suppose the two Explanatory–Spaced pupils who scored 19 and 20
were unavailable. The cell counts become 4, 4, 4, 4, 4 and 2, and the full
model leaves $22-6=16$ residual degrees of freedom. The equally weighted
Spaced margin is then $(6+12+16.5)/3=11.5$, whereas the mean of the ten
remaining Spaced pupils is 10.5: two answers to two different questions.
@tbl-ch06-unbalanced refits the full model to the 22 remaining pupils with
the formula written in each order.

\Needspace{10\baselineskip}

| Term | Formula `feedback * practice` | Formula `practice * feedback` |
|:--|--:|--:|
| Feedback | 139.7576 | 157.7500 |
| Practice | 52.0833 | 34.0909 |
| Interaction | 36.2500 | 36.2500 |
| Residual | 50.5000 | 50.5000 |

: Sequential sums of squares for the 22 remaining pupils, by the order of the formula. {#tbl-ch06-unbalanced tbl-colwidths="[28,36,36]"}

Sequential sums of squares give each term only what the terms entered
before it have not taken. The feedback and practice rows therefore move
with the order, while the interaction and the residual stay where they
are. With unequal cells, one sequential table does not supply order-free
tests of the main effects. Naming the two models to be compared does.

The interaction row can itself be read as a comparison of two fits to the
same 24 pupils. The **additive model**, `score ~ feedback + practice`,
gives every feedback condition the same Spaced-minus-Massed difference;
the **full model**, `score ~ feedback * practice`, lets that difference
vary. For nested models fitted to the same response and records, with
residual sum of squares $RSS_0$ on $df_0$ degrees of freedom for the
smaller model and $RSS_1$ on $df_1$ for the larger,

$$
F=\frac{(RSS_0-RSS_1)/(df_0-df_1)}{RSS_1/df_1}
=\frac{(124-60)/(20-18)}{60/18}=9.6,
$$

with the denominator taken from the larger model. The following code fits
both models with `lm()` and compares them.

\Needspace{8\baselineskip}

```{r}
#| label: ch06-nested-f
additive_fit <- lm(score ~ feedback + practice, data = feedback_data)
full_fit <- lm(score ~ feedback * practice, data = feedback_data)
coef(additive_fit)
anova(additive_fit, full_fit)
```

The additive coefficients, 4 for Rereading under Massed practice, 4 and 8
for Answer and Explanatory feedback minus Rereading, and 4 for Spaced
minus Massed, show the single practice difference that the smaller model
imposes. The comparison returns $F(2,18)=9.6$ and $p=.0015$: the
interaction row of @tbl-ch06-factorial, now read as a question about two
named models. Because it names both models, no term's position in a
formula enters it. For the 22 remaining pupils, the additive model (RSS
86.75 on 18 df) against the full model (RSS 50.50 on 16 df) gives
$F(2,16)=5.7426$ and $p=.0132$, whichever order the formula uses. Keep
the response and the records explicitly the same, because default removal
of missing values can otherwise change the comparison. Chapter 9,
Section 2, uses the same comparison of named models for the analysis of
covariance and its parallel-slopes check.

The feedback and practice margins remain meaningful descriptions, but the
interaction clarifies their limits. For example, the 4-point overall practice
difference averages a 0-point, a 4-point and an 8-point difference. A result
paragraph should convey this varying pattern when the scientific question
concerns how practice works across feedback conditions.

### Work the selected difference of differences completely

Consider Answer feedback's practice change minus Rereading's practice change:

$$
\widehat{DD}=(12-8)-(6-6)=4.
$$

The contrast has coefficients $(1,-1,-1,1)$ in the cell order
Rereading–Massed, Rereading–Spaced, Answer–Massed, Answer–Spaced.
Its four cell means come from different pupils. Under the common-variance
model, their variances therefore add after multiplying by the squared
coefficients:

$$
SE(\widehat{DD})=
\sqrt{MS_E\left(\frac14+\frac14+\frac14+\frac14\right)}
=1.8257.
$$

The omnibus interaction spans two independent directions. For simultaneous
confidence over all contrasts in that two-dimensional interaction space,
Scheffe's method uses the multiplier

$$
\sqrt{2F_{.95;2,18}}=2.6663.
$$

The complete calculation keeps the cell subtraction, residual variance and
family calibration visible:

\Needspace{13\baselineskip}

```{r}
#| label: ch06-selected-interaction
dd <- (cell_mean["Answer feedback", "Spaced"] -
  cell_mean["Answer feedback", "Massed"]) -
  (cell_mean["Rereading", "Spaced"] -
  cell_mean["Rereading", "Massed"])
mse_two <- deviance(two_fit) / df.residual(two_fit)
se_dd <- sqrt(mse_two * 4/4)
scheffe <- sqrt(2 * qf(.95, 2, 18))
c(estimate = dd, SE = se_dd, multiplier = scheffe)
dd + c(-1, 1) * scheffe * se_dd
```

The resulting interval is $4\pm2.6663(1.8257)=[-0.868,8.868]$.
It includes zero even though the interaction F test rejects its null.
The F test asks whether both interaction directions are zero jointly;
this interval addresses one selected direction while allowing for the
entire family. Other directions can carry substantial evidence, and a
simultaneous interval for this direction can remain broad. Neither result
cancels the other.

@tbl-ch06-dd-quantities sets the selected DD beside Answer feedback's
simple practice effect, which happens to share its estimate, 4.

\Needspace{9\baselineskip}

| Quantity | Estimate | SE | 95% interval | Calibration |
|:--|--:|--:|:--|:--|
| Answer's simple practice effect | 4 | 1.2910 | [1.2877, 6.7123] | Pointwise |
| Answer-minus-Rereading DD | 4 | 1.8257 | [0.1643, 7.8357] | Pointwise |
| Answer-minus-Rereading DD | 4 | 1.8257 | [$-0.8680$, 8.8680] | Scheffé |

: Three quantities from the interaction section; the Scheffé interval covers the two-dimensional interaction family. {#tbl-ch06-dd-quantities tbl-colwidths="[38,11,10,25,16]"}

The simple effect compares only two cell means, so its SE is smaller. It
also contains the overall practice component as well as an interaction
component, so it lies outside the interaction-only family, and the
Scheffé guarantee does not transfer to it. The DD's two intervals differ
only in their multiplier: the pointwise interval excludes zero and the
simultaneous one includes it. Choosing the narrower interval after seeing
which conclusion it gives does not keep the family's guarantee. State
which quantity was selected, which family was considered and which
interval was used before interpreting a result.

Report the interaction's magnitude in score points first. The practice
difference is 0 points under Rereading, 4 under Answer feedback and 8 under
Explanatory feedback, and the selected DD is 4 points with SE 1.8257. A
variation share may follow once its denominator is named. The interaction
sum of squares is $64/476=.1345$ of the total sum of squares, but
$64/124=.5161$ of the interaction and residual sums of squares together.
Both shares are in-sample summaries of an authored design, and neither
replaces the pattern of practice differences.

## Analyse proportions and counts {#ch06-counts}

### Keep the event and denominator visible

A score model is not the appropriate starting point for every outcome.
Chapter 3 introduced the invented completion table in
@tbl-ch06-completion-counts. It contains aggregate counts for 100 records,
classified by learning format and completion. These records have no
linkage to either score dataset above. The counts themselves establish
neither random allocation nor representative sampling.

\Needspace{9\baselineskip}

| Format | Completed | Not completed | Total |
|:--|--:|--:|--:|
| Online | 32 | 18 | 50 |
| In person | 18 | 32 | 50 |
| Total | 50 | 50 | 100 |

: The same authored completion counts as Chapter 3. {#tbl-ch06-completion-counts}

For the Online group, the event is completion and the denominator is its
50 records. Under an independent Bernoulli model with a common completion
probability $p$, the count of completions has a binomial distribution. The
estimated probability is $\widehat p=32/50=.64$. If records were clustered
within dependent classes or came from repeated observations of the same
people, this ordinary independent-unit uncertainty model would require
modification.

Suppose the benchmark is $p_0=.5$. The score test standardizes the difference
using the variance under that null:

$$
z=\frac{\widehat p-p_0}{\sqrt{p_0(1-p_0)/50}}
=\frac{.64-.5}{\sqrt{.5(.5)/50}}=1.9799.
$$

Squaring gives $X^2=z^2=3.92$, compared with a chi-squared distribution
on one degree of freedom. The procedure uses a large-sample reference;
the expected numbers of successes and failures under this null are both 25.

\Needspace{8\baselineskip}

```{r}
#| label: ch06-one-proportion
one_prop <- prop.test(32, 50, p = .5, correct = FALSE)
one_prop$estimate
one_prop$statistic
one_prop$p.value
one_prop$conf.int
```

The two-sided p-value is .0477. The estimated completion probability is
.64, with 95% Wilson score interval [.5014, .7586]. `correct = FALSE`
explicitly selects the uncorrected score procedure; accepting R's default
continuity correction would specify a different calculation. The call and
the reported result should identify which procedure was used.

The score interval contains the candidate probabilities that their own
two-sided score tests would retain [@wilson1927]. Each candidate supplies
the variance for its null calculation. This inversion explains why the
interval is linked to the score test and is not simply
$\widehat p\pm1.96\sqrt{\widehat p(1-\widehat p)/n}$, the Wald interval,
and why its center lies slightly closer to .5 than .64 does. The Computer
Lab computes the Wilson endpoints directly. The p-value assesses the
stipulated benchmark; it is not the probability that $p=.5$ is true or a
measure of the completion rate's practical importance.

### Compare the two groups through their full table

Now ask whether completion probability differs by format. This uses both
rows and a different null: format and completion are independent in the
working population model, equivalently equal completion probabilities in
this two-group setting. Constructing labelled counts prevents an accidental
change in the event or group direction.

\Needspace{12\baselineskip}

```{r}
#| label: ch06-two-way-counts
counts <- matrix(c(32, 18, 18, 32), nrow = 2,
  byrow = TRUE,
  dimnames = list(Format = c("Online", "In person"),
    Result = c("Completed", "Not completed")))
prop.table(counts, margin = 1)
count_test <- chisq.test(counts, correct = FALSE)
count_test$expected
count_test$statistic
count_test$p.value
count_test$residuals
```

`margin = 1` divides each row by its own total. Completion rates are .64
for Online and .36 for In person, a difference of .28, or 28 percentage
points, in the Online-minus-In-person direction. Under independence the
fitted expected count in row $j$, column $k$ is

$$
E_{jk}=\frac{(\text{row }j\text{ total})
(\text{column }k\text{ total})}{N}.
$$

All four expected counts here equal $50\times50/100=25$. Pearson's
statistic accumulates the squared discrepancies relative to those expected
counts:

$$
X^2=\sum_{j,k}\frac{(O_{jk}-E_{jk})^2}{E_{jk}}
=4\left(\frac{7^2}{25}\right)=7.84.
$$

For a $J$-by-$K$ table the usual independence reference has
$(J-1)(K-1)$ degrees of freedom. Here $df=1$ and p .0051. The observed
Pearson statistic $X^2$ is a number calculated from the table; $\chi^2_1$
names its approximate reference distribution under the null model.

The Pearson residuals are $(O-E)/\sqrt E$: 1.4, $-1.4$, $-1.4$, 1.4
in row order. Their signs locate the pattern: more Online completions and
fewer Online non-completions than the independence fit predicts, with the
opposite departures for In person. Squaring and adding these residuals
recovers $X^2$. They are related through the table margins, so their signs
do not constitute four independent cell tests.

The approximation concerns expected counts under the fitted null model,
along with independent observational units and a sampling model appropriate
to the table. Looking only at observed cell counts misses part of this
check. Sparse expected cells may require an alternative reference or a more
careful model. In the present example all expected counts are 25, supporting
the ordinary chi-squared approximation under the stated unit model.

The one-proportion test asked whether Online completion probability equals
a fixed .5 benchmark. The complete-table test asks whether the two format
probabilities are equal; it estimates their common probability under the
null from the full table. Their different p-values answer different
questions. Neither test establishes that changing format would cause
completion to change without relevant design information.

### Choose an effect scale and keep alternatives within the analysis {#ch06-categorical-reference}

A chi-square test says whether the table departs from no association, not
how large the association is or on which scale. @tbl-ch06-effect-scales
sets four common scales side by side for the completion table, with
$p_1$ the Online and $p_0$ the In-person completion proportion, and
gives the question that each one answers [@agresti2018].

\Needspace{17\baselineskip}

| Scale | Definition | Value | What it answers |
|:--|:--|--:|:--|
| Difference in proportions | $p_1-p_0$ | .2800 | How many percentage points more often the event occurs in one group: an absolute gap |
| Risk ratio | $p_1/p_0$ | 1.7778 | How many times as probable the event is in one group: a relative gap, which changes when the other outcome is called the event |
| Odds ratio | $\dfrac{p_1/(1-p_1)}{p_0/(1-p_0)}$ | 3.1605 | How many times as large the odds of the event are; it stays the same when rows and columns exchange roles, so a design that samples on the outcome can still estimate it |
| Cramér's V | $\sqrt{\dfrac{X^2}{N\min(J-1,K-1)}}$ | .2800 | How strong the association is, from 0 to 1, for a table of any size; it carries no direction and names no cell comparison |

: Four effect scales for the completion table, Online versus In person. {#tbl-ch06-effect-scales tbl-colwidths="[15,25,9,51]"}

The same counts give a gap of 28 percentage points, a risk ratio of
1.7778 and an odds ratio of 3.1605. The odds ratio lies farther from 1
than the risk ratio because completion is common: Online's completion
odds are $32/18=1.7778$ and In person's are $18/32=.5625$, so an odds
ratio is not read as a risk ratio. Calling non-completion the event
changes the risk ratio to $.36/.64=.5625$ and the odds ratio to its
reciprocal, .3164, so the event and the direction are named beside every
scale. Cramér's V [@cramer1946] equals the difference in proportions here
only because every row and column total is 50. Chapter 11 reads the odds
ratio again as an exponentiated logistic coefficient.

Other procedures change one part of the analysis while keeping the same
table, and each answers its own question. The two-proportion interface
gives a large-sample interval for the difference in proportions, a
different quantity from the one-probability Wilson interval of
Section 4.1. A continuity correction changes the calibration of the
Pearson statistic. Fisher's exact test [@fisher1935] conditions on the
table margins and uses an exact reference, which matters when expected
counts are small. The Computer Lab runs each alternative on this table and
then continues the observational student intention-by-pass example of
Chapter 2, Section 3.1, where these scales, the choice of procedure and an
unresolved source-coding question meet in an existing dataset.

## Fit by an objective function and likelihood {#ch06-likelihood}

### Fix the data and vary a candidate parameter

The fitted means in Section 1 were chosen by a rule: minimize the sum
of squared deviations from the observations. We can inspect this rule
with the ten Silence scores, 8, 9, 9, 11, 12, 13, 13, 14, 15 and 16.
These observations stay fixed. What changes is the proposed center,
denoted by $c$.

An **objective function** assigns a numerical criterion to each candidate
parameter value for fixed data. For a center fitted by least squares,
the objective is

$$
Q(c)=\sum_{i=1}^{10}(y_i-c)^2
     =(8-c)^2+(9-c)^2+\cdots+(16-c)^2.
$$

At $c=12$, the deviations are the Silence residuals of Section 1, and
their squares sum to 66 squared score points. At either 11 or 13, they
sum to 76. Chapter 2, Section 2.2, states that the sample mean minimizes
this criterion, and Chapter 3's Appendix A derives the identity behind
it. Here the mean is 12 points. A numerical search lets us reproduce the
minimizing value without writing it into the fitting instruction.

\Needspace{6\baselineskip}

```{r}
#| label: ch06-scalar-squared-loss
silence <- subset(sound_data, sound == "Silence")$score
loss <- function(center) sum((silence - center)^2)
mean_fit <- optimize(loss, interval = range(silence))
c(center = mean_fit$minimum, squared_loss = mean_fit$objective)
```

`subset()` keeps the Section 1 rows whose condition is Silence, and
`loss()` returns the sum of squared deviations for any candidate center.
`optimize()` searches for a minimum of that function over the interval
supplied, here `range(silence)`, or $[8,16]$, which contains the sample
mean. The objective is a single smooth bowl over this interval, so the
scalar search has one minimum to locate.

The two returned quantities have different meanings. `minimum` is
the **parameter location**, 12 points; `objective` is the function
value there, 66 squared points. The latter is a sum of squared
deviations, Silence's contribution to $SS_W$ in Section 1, rather than an
SE or an estimate of the mean. The search interval
also belongs to the fitting instruction: excluding the best center
would restrict which answer the search could return. Numerical stopping
tolerance concerns the accuracy of that search; statistical uncertainty
concerns what an estimate would do across the repetitions justified by
the study.

These least-squares identities are arithmetic. An uncertainty statement
for the fitted mean adds a model or design and its conditions. Chapter 7
extends this criterion from one center to a line with an intercept and
slope. The Computer Lab evaluates the criterion at several centers and
narrows the search interval while keeping the observations fixed.

### Change the fitting criterion for a binomial outcome

For the completion example, the observations are 32 completions in 50
Online trials. Let $p$ denote a candidate probability of completion.
Under the stipulated model of independent trials with a common
probability, the number completed follows a binomial distribution.
The model assigns probability

$$
P_p(X=32)={50\choose32}p^{32}(1-p)^{18}.
$$

In a probability calculation, we choose a value of $p$ and consider
which counts might occur. For fitting, we have already observed the
count: 32 stays fixed, as does the total of 50. We now evaluate the
same expression at different candidate values of $p$. Viewed in this
direction, it is the **likelihood function**,

$$
L(p)={50\choose32}p^{32}(1-p)^{18},\qquad 0\le p\le1.
$$

The binomial coefficient counts the ways 32 completions can occur
among 50 trials. It is constant as $p$ varies. In this binomial model, likelihood is the
probability of the observed data under each candidate model parameter;
it is not a probability distribution over that parameter. In
particular, a high likelihood at a candidate $p$ does not give the
probability that $p$ is true.

**Maximum likelihood estimation** chooses the parameter value that
makes the observed data most likely under the declared model. The
logarithm is increasing, so maximizing the log-likelihood gives the
same parameter value. For $0<p<1$,

$$
\ell(p)=\log {50\choose32}+32\log p+18\log(1-p).
$$

Logs turn a product of powers into a sum and make calculation easier.
The binomial coefficient affects the reported function value, although
it cannot change the maximizing $p$. Here the interior maximum can also
be found analytically. Differentiating with respect to the candidate
probability gives

$$
\ell'(p)=\frac{32}{p}-\frac{18}{1-p}
         =\frac{32-50p}{p(1-p)}.
$$

Setting this slope to zero gives $32-50p=0$, and therefore
$\widehat p=32/50=.64$. The denominator is positive inside $(0,1)$.
The slope is positive below .64 and negative above .64: the curve rises
to this point and then falls. This verifies a maximum rather than
merely finding a flat point.

To use the default minimization operation in `optimize()`, supply the
**negative** log-likelihood. A smaller negative log-likelihood means a
larger log-likelihood, so its minimum locates the same fitted probability.

\Needspace{11\baselineskip}

```{r}
#| label: ch06-scalar-binomial-mle
negative_loglik <- function(p) {
  -dbinom(32, size = 50, prob = p, log = TRUE)
}
likelihood_fit <- optimize(negative_loglik, interval = c(0, 1))
c(numerical = likelihood_fit$minimum, analytic = 32/50,
  negative_loglik = likelihood_fit$objective)
negative_loglik(.5)
```

`dbinom()` evaluates the binomial probability for the fixed count 32.
With `log = TRUE`, it computes the log probability directly. The
numerical and analytic estimates both round to .64. The small
numerical difference from exactly .64 reflects the search tolerance.
At the fitted probability the negative log-likelihood is 2.1465,
compared with 4.1330 at $p=.5$. The smaller value is the better fit
according to this criterion. The parameter estimate .64 is a
completion probability; the criterion value 2.1465 is not.

```{r}
#| label: fig-ch06-likelihood
#| echo: false
#| fig-pos: "!ht"
#| fig-width: 6.1
#| fig-height: 3.4
#| fig-cap: "The binomial log-likelihood is largest at the observed proportion."
#| fig-alt: "For 32 successes in 50 fixed trials, the log-likelihood curve peaks at candidate probability .64. A dashed line at .5 intersects the curve at a lower value."
likelihood_grid <- data.frame(p = seq(.05, .95, length.out = 200))
likelihood_grid$loglik <- dbinom(32, size = 50,
  prob = likelihood_grid$p, log = TRUE)
ggplot(likelihood_grid, aes(p, loglik)) + geom_line() +
  geom_vline(xintercept = .64) +
  geom_vline(xintercept = .5, linetype = 2) +
  labs(x = "Candidate completion probability", y = "Log-likelihood") +
  theme_minimal(base_size = 11)
```

The horizontal axis contains candidate parameter values, and the
vertical axis gives the log-likelihood of the same observed count at
each candidate. Thus the curve displays a fitting criterion, not a
sampling distribution or a distribution of probabilities over $p$.
The peak identifies the fitted value. The score test and interval in
Section 4 answer additional questions about a proposed null and the
uncertainty of estimation under their stated calibration.

Finally, check the boundaries of the parameter space. With both
completions and non-completions observed, $p=0$ and $p=1$ each give
likelihood zero, so the maximum is interior. If all 50 trials instead
had zero completions, the maximum would be at $p=0$; if all completed,
it would be at $p=1$. An interior numerical search can approach such
an endpoint without returning it exactly. The endpoint values must
therefore be examined explicitly in those cases. Chapter 11 develops
these boundary questions further when likelihood is used to fit
logistic regression.

## Examine procedures by repetition and plan precision {#ch06-simulation}

The preceding analyses started with observed data and produced estimates,
intervals and tests. We can also begin with a declared generating model,
create data from it, and examine what an analysis does over repeated studies.
Chapter 5 used this approach to examine sampling variation and interval
coverage. Here the questions concern an omnibus ANOVA procedure and the
sample size for a specified comparison.

### Follow one generated study before repeating it

Imagine three independent groups whose outcomes follow Normal distributions,
each with standard deviation $\sqrt{10/3}=1.8257$ points. First set all three
population means to 8, so that the equal-means null hypothesis is true;
later the unequal means 6, 8 and 10 describe an alternative. These values
borrow the scale of the feedback study in Section 3, whose Massed cells
have means 6, 8 and 10 with four pupils each and whose factorial fit has
residual mean square $10/3$. They define a hypothetical world in which we
know the truth; the constructed scores do not establish that this
generating model describes real pupils.

Begin with four observations per group and analyse one generated study.

\Needspace{13\baselineskip}

```{r}
#| label: ch06-one-generated-anova
set.seed(60630)
n_per_group <- 4
sim_group <- factor(rep(1:3, each = n_per_group))
sim_score <- rnorm(3*n_per_group,
  mean = rep(c(8, 8, 8), each = n_per_group), sd = sqrt(10/3))
anova(lm(sim_score ~ sim_group))
```

`rep()` and `factor()` label four observations in each of three groups,
`rnorm()` draws the twelve outcomes from the declared means and common
SD, and `set.seed()` makes this particular draw reproducible. The formula
`sim_score ~ sim_group` fits a separate mean for each group; `anova()`
compares that fit with a common mean. Its table has 2 group degrees of
freedom and $12-3=9$ error degrees of freedom. For these generated data,
$SS_B=7.5977$ and $SS_W=32.9851$, giving $F=1.0365$ and $p=.3935$.
Applying the rule $p<.05$ therefore does not reject the equal-means null
in this study. Although the three generating means are identical, the
three sample means vary, so the between-group sum of squares is
positive. The estimated error variance, $MS_W=3.665$, also varies around
the generating variance $10/3$.

One generated study cannot describe a procedure's long-run behaviour.
Repeating the same calculation many times gives a collection of p-values
to compare with .05. @tbl-ch06-simulation reports 2,000 studies for each
of four conditions. The null conditions use means $(8,8,8)$; the
alternative conditions use $(6,8,10)$. Each condition retains independent
Normal observations and the same common SD, and the Computer Lab
reproduces the table from seed 60630.

| Pupils per group | Null rejection rate | MCSE | Power at means 6/8/10 | MCSE |
|--:|--:|--:|--:|--:|
| 4 | .056 | 0.0051 | .6545 | 0.0106 |
| 12 | .050 | 0.0049 | .9990 | 0.0007 |

: Operating characteristics in the declared generating models, $B=2000$ studies per condition. {#tbl-ch06-simulation}

The **null rejection rate** estimates the Type I error probability: how
often this .05 omnibus procedure rejects when all three population means
are equal. Both estimated rates are close to .05. The **power** column
estimates the probability of rejecting that same null under the specified
unequal means. It rises from .6545 with four pupils per group to .9990 with
12. This is power against the particular means 6, 8 and 10, with the stated
spread and independence; a smaller separation would give different power.

Each rate is itself an estimate based on finitely many repetitions. If
$\widehat q$ is a simulated rejection proportion from $B$ independent studies,
its approximate Monte Carlo standard error is

$$
MCSE(\widehat q)=\sqrt{\frac{\widehat q(1-\widehat q)}{B}}.
$$

For the estimated power .6545, it is $\sqrt{.6545(1-.6545)/2000}=0.0106$,
and the table's MCSE columns apply the same formula to each rate. This
value describes the numerical uncertainty from using 2,000 simulations:
a fresh set of 2,000 simulated studies can give a somewhat different
power estimate. The MCSE is distinct from a pupil outcome's SD or the SE
of a group mean or contrast within one study. A rate very close to one,
such as .999, also deserves care: only two of the 2,000 studies failed
to reject, so its tiny plug-in MCSE rests on very few non-rejections.

Increasing the number of pupils $n$ changes the information within a study,
its error degrees of freedom and its ability to detect the specified
difference. Increasing the number of repetitions $B$ makes the simulated
rate more precise; it leaves the study being simulated unchanged.
Simulation describes a procedure under the declared model; agreement with
its expected behaviour does not establish that the model fits a
particular real study.

### Plan for a specified comparison and precision target

The omnibus simulation asks whether some population group means differ.
A future study may instead be planned to estimate or detect one particular
difference. Consider a future balanced one-factor study of the three
feedback conditions, with a declared **two-comparison Bonferroni family**:
Answer feedback versus Rereading, and Explanatory feedback versus
Rereading. The family has the same form as Section 2's two comparisons
with Silence. Keep it, with family significance level .05, even when
calculating power for just one of its comparisons.

Suppose the comparison of interest has a population difference of
$\delta=2$ points. Plan equal group sizes $n$, independent Normal outcomes
and common SD $\sigma=\sqrt{10/3}$, the square root of the factorial fit's
residual mean square in Section 3. The planning SE for a difference of two
group means is $\sigma\sqrt{2/n}$. All three groups contribute to the
pooled error variance, so its degrees of freedom are $3(n-1)$, even though
the contrast compares only two of the means.

Bonferroni assigns two-sided significance level $.05/2=.025$ to each
comparison. Dividing that amount between its two tails puts .0125 in each
tail. The rejection cutoffs are therefore $-c$ and $c$, where

$$
c=t_{1-.05/(2\times2),\,3(n-1)}=t_{.9875,\,3(n-1)}.
$$

Under the zero-difference null, the test statistic has the central
$t$ distribution already used for the comparison. Under a specified
nonzero difference, its distribution moves towards that difference.
The **noncentral $t$ distribution** describes this statistic when its
numerator has a nonzero mean and its denominator still estimates the
common spread. Its noncentrality parameter expresses the population
difference in planning SE units:

$$
\lambda=\frac{\delta}{\sigma\sqrt{2/n}}.
$$

For this alternative, power is the area beyond the same two rejection
cutoffs: $P(T<-c)+P(T>c)$ for a noncentral $t$ variable with $3(n-1)$
degrees of freedom and noncentrality $\lambda$. These are probabilities
over future studies, rather than a p-value from observed data. Both tails
belong to the two-sided procedure, even when the planning difference is
positive.

\Needspace{10\baselineskip}

```{r}
#| label: ch06-contrast-power
contrast_power <- function(n, delta = 2, sigma = sqrt(10/3)) {
  df <- 3*(n-1)
  critical <- qt(1 - .05/(2*2), df)
  ncp <- delta / (sigma * sqrt(2/n))
  pt(-critical, df, ncp = ncp) +
    pt(critical, df, ncp = ncp, lower.tail = FALSE)
}
round(c(n16 = contrast_power(16), n17 = contrast_power(17)), 4)
```

The function accepts a proposed group size and returns power. Its first
`pt()` call obtains the lower-tail probability; the second obtains the
upper-tail probability. At $n=17$, the planning SE is 0.6262, the error
degrees of freedom are 48, the critical value is 2.3139, and
$\lambda=3.1937$. Summing the two tail areas gives power
.8073. Checking successive integer sizes gives 17 pupils per group,
51 in total, as the first size reaching .8 power; 16 gives .7793.

This is the power of **one specified comparison within the declared
family**. It is neither the probability that both comparisons reject nor
the power of the omnibus F test. For example, with 12 pupils per group
the specified two-point comparison has power .6328, whereas the omnibus
simulation in @tbl-ch06-simulation gives estimated power .9990 against
the complete pattern $(6,8,10)$. The hypotheses and rejection rules differ,
so these numbers answer different questions.

A precision target concerns the interval's width rather than its
probability of excluding zero. Suppose the aim is a simultaneous interval
whose planned half-width is no more than one point. The family and
degrees of freedom stay the same, giving

$$
\text{planned half-width}
=t_{.9875,\,3(n-1)}\,\sigma\sqrt{\frac{2}{n}}.
$$

This calculation substitutes the planning SD $\sigma$ for the error SD
that the future study will estimate. The planned half-width is 1.0078
points at 34 per group and 0.9929 at 35, so 35 per group, or 105 pupils
in total, is the first size meeting this target. The realised interval
uses the observed $\sqrt{MS_W}$ and therefore has a random width.
Choosing 35 does not guarantee that its half-width will be at most one in
every study. @fig-ch06-planning shows why a power target and an
interval-width target can select quite different sample sizes.

```{r}
#| label: fig-ch06-planning
#| echo: false
#| fig-width: 6.4
#| fig-height: 2.7
#| fig-cap: "Power and planned interval width impose different sample-size targets."
#| fig-alt: "For the same two-comparison three-group procedure, power rises above .8 at 17 pupils per group. Planned simultaneous interval half-width falls below one point at 35 pupils per group."
n_grid <- 4:50
powers <- sapply(n_grid, contrast_power)
half_width <- qt(1-.05/(2*2), 3*(n_grid-1)) *
  sqrt(2*(10/3)/n_grid)
panel <- c("Power", "Planned half-width (points)")
planning <- data.frame(n = rep(n_grid, 2),
  value = c(powers, half_width),
  target = factor(rep(panel, each = length(n_grid)), levels = panel))
targets <- data.frame(target = factor(panel, levels = panel),
  value = c(.8, 1), n = c(17, 35))
ggplot(planning, aes(n, value)) + geom_line() +
  geom_hline(aes(yintercept = value), data = targets, linetype = 2) +
  geom_vline(aes(xintercept = n), data = targets, linetype = 2) +
  facet_wrap(~ target, scales = "free_y") +
  labs(x = "Pupils per group", y = NULL) +
  theme_minimal(base_size = 11)
```

Finally, distinguish the desired number of available outcomes from the
number recruited. If outcomes are expected to be available for 90% of
recruits, the 17-per-group power target needs
$\lceil17/.9\rceil=19$ recruits per group, 57 in total, with an expected
17.1 available outcomes per group under that retention assumption. It is
an average allowance, not a guarantee of 17 outcomes in every group or a
correction for bias from systematically missing outcomes.

The Computer Lab searches the full grid of sample sizes and changes the
target difference and the retention rate. A two-group power calculation
declares a different study, with different error degrees of freedom, and
cannot silently replace this three-group calculation. Choose a planning
difference from the substantive question and relevant prior evidence. As
Chapter 3's significance-filter example showed, selecting statistically
significant results can favour exaggerated estimates; an observed
difference from such a selection is not automatically a suitable planning
alternative.

## Reading Bridge {#ch06-reading-bridge}

The chapter compared group means, separated omnibus questions from
calibrated contrasts, and read a comparison that changes across another
factor, testing that change by comparing two named models on the same
records. Counts required a different response model and their own effect
scales; likelihood and an objective function made the fitting criterion
explicit. Repetition examined procedures under a declared generating
model, while power and precision described different planning aims.

When reading a study, identify the question, analysed records, model,
estimate and uncertainty procedure before evaluating their adequacy.
For a group comparison, locate its subtraction direction and the records
it uses. For a test or interval, identify the null, reference distribution
and comparison family. For an effect in a table, identify the event and
the scale. For a simulation or planning result, locate the generating
assumptions and the particular procedure being evaluated. These details
connect the reported numbers to the claims they can support.

[]{#ch06-appendix-b}Later chapters build on each of these steps. Chapter 7
extends squared loss from a fitted center to a line: it derives the
line's normal equations, relates fitted values and residuals through
projection, and develops uncertainty for the slope and the fitted mean.
Chapter 8 conditions on several controls and distinguishes a coefficient
conditional on specified predictors from a fitted mean at a profile, a
prediction for a new case and an average of fitted values over a named
population. Chapter 9, Section 2, develops the analysis of covariance as
regression and checks its parallel slopes with the nested-model
comparison of Section 3.2. Chapter 11 develops likelihood for binary
regression, including exact boundary estimates, a likelihood with more
than one parameter, and the starting values and convergence checks that
numerical fitting then requires.

\Needspace{10\baselineskip}

### Key terms and notation {.unnumbered}

ANOVA, contrasts, interactions and comparison families keep their main
explanations in Chapter 4, and the Part One Concept Reference gives the
broader reference and the principal home of each concept. The following
concepts were introduced or developed through their R applications here.

| Term and common alternatives | Meaning |
|:--|:--|
| Model formula (R formula) | Notation such as `score ~ group` naming an outcome and the variables that explain it for a fitting function. It does not state whether observations are independent or paired; that comes from the design. |
| Categorical effect scales (difference in proportions; risk ratio; odds ratio; Cramér's V) | Four ways to state the size of an association in a table, each answering its own question: the difference in proportions gives an absolute gap, the risk ratio a relative gap, the odds ratio a ratio of odds that stays the same when rows and columns exchange roles, and Cramér's V a unit-free strength, from 0 to 1, for a table of any size. |
| Objective function | A numerical criterion evaluated at candidate parameter values for fixed data; fitting seeks a stated minimum or maximum. |
| Likelihood and log-likelihood | The probability or density of the observed data under a declared model, viewed as a function of its parameters; the log-likelihood is its logarithm. |
| Maximum likelihood estimation (MLE) | Choosing parameter values that maximize the likelihood, equivalently the log-likelihood, over the stated parameter space. |

: Selected concepts introduced or developed here. {#tbl-ch06-terms tbl-colwidths="[30,70]"}

The notation follows the local question. $C=\sum_jc_j\mu_j$ is a contrast,
whereas $L(\theta)$ is a likelihood function of a parameter. Similarly, Type II error probability
$\beta$ differs from a model coefficient such as $\beta_x$.
Observed Pearson $X^2$ differs from the reference distribution $\chi^2$,
and $\Delta p=p_1-p_0$ is a difference in proportions.
In simulations, $n$ counts observations per stated study or group and $B$
counts repeated studies. An SE describes an estimator's sampling variation;
an MCSE describes finite-simulation uncertainty in a computed summary.

### Further reading {.unnumbered}

For the linear-model basis of ANOVA, contrasts and nested-model
comparisons, see @kutner2005. @tukey1953 develops simultaneous comparison
ideas, and @wilson1927 provides the score-interval construction for a
proportion. @agresti2018 develops the effect scales for two-way tables.
@cohen1988 discusses power and the dependence of a planning result on its
chosen effect and procedure, and @cetinkaya2024 develops the
simulation-based treatment of operating characteristics used in Section 6.
R's documentation gives the arguments and returned components for `aov`,
`lm`, `TukeyHSD`, `oneway.test`, `prop.test`, `chisq.test` and `optimize`
[@rcore2025]. The plotting chapters of @wickham2023 develop the data,
mapping and layer vocabulary behind the chapter's figures, which the
Computer Lab builds.

## Exercises {.unnumbered}

The chapter's printed output, tables and figures supply the values each
exercise reads, with a calculator for the few calculations that the
reasoning needs. State the records, working conditions and comparison
direction in each interpretation.

<!-- prompt-id: CH06-E01 -->

### Exercise 1: Read the one-way fit from its output {.unnumbered #ex-ch06-01}

For the thirty background-sound scores, `anova(one_fit)` reports sums of
squares 80 and 178 on 2 and 27 df, $F=6.0674$ and $p=.0067$, and
`coef(indicator_fit)` returns 12, $-2$ and $-4$ with Silence as the
reference level.

1. Say what each sum of squares measures in these scores, and why
   Instrumental music, whose mean equals the grand mean of 10, adds
   nothing to the between-group sum of squares although its ten scores
   vary.
2. Explain what the within-group df, 27, count, and why $F$ divides by
   the residual mean square, 6.5926, rather than by the variance of all
   thirty scores, $258/29=8.8966$.
3. State the equal-means null, and interpret $p=.0067$ under the working
   model, including what the result leaves unanswered about particular
   pairs of conditions.
4. Predict the three coefficients if Instrumental music were made the
   reference level, and say which parts of the ANOVA output would change.

<!-- END EX06-01 PROMPT -->

<!-- prompt-id: CH06-E02 -->

### Exercise 2: Keep the comparison and its family together {.unnumbered #ex-ch06-02}

Instrumental music minus Silence is $-2$ points with SE 1.1483 on 27 df.
Its 95% interval is $[-4.3561, 0.3561]$ when the comparison stands alone,
$[-4.7253, 0.7253]$ inside the declared family of the two comparisons
with Silence (Bonferroni) and $[-4.8470, 0.8470]$ inside the family of
all three pairs (Tukey).

1. The estimate and SE are the same in all three intervals. Explain why
   the limits differ, naming the family that each interval protects.
2. Explain why the SE uses 27 residual df, although the comparison
   involves only the twenty students in two conditions.
3. An analyst reports Music with lyrics minus Silence, $-4$ points,
   because it is the largest of the three observed differences. Say which
   family its interval should come from, and why.
4. The omnibus F rejected equal means, yet all three intervals above
   include zero. Explain why these results are compatible, and why an
   interval that includes zero does not show that the two means are
   equal.

<!-- END EX06-02 PROMPT -->

<!-- prompt-id: CH06-E03 -->

### Exercise 3: Read interaction as a change in comparison {.unnumbered #ex-ch06-03}

In the 24-pupil feedback-by-practice study, the Massed and Spaced cell
means are 6 and 6 for Rereading, 8 and 12 for Answer feedback and 10 and
18 for Explanatory feedback, and the full two-way model gives the
interaction test $F(2,18)=9.6$, $p=.0015$.

1. Give the three Spaced-minus-Massed changes, and say what the
   nonparallel lines of the interaction plot show and what they cannot
   show without a model.
2. The practice margins are 8 and 12. Explain why the 4-point practice
   main effect needs the interaction beside it when a reader asks how
   spaced practice works.
3. The selected difference of differences, Answer feedback's change
   minus Rereading's change, is 4 points with SE 1.8257. Its pointwise
   interval is $[0.1643, 7.8357]$, and its Scheffé interval over the
   two-dimensional interaction family is $[-0.8680, 8.8680]$. Explain how
   the Scheffé interval can include zero while the interaction F test
   rejects, and say which interval is warranted when this direction was
   chosen after the plot was inspected.
4. Answer feedback's simple practice effect is also 4 points, with SE
   1.291. Explain why its SE is smaller and why the interaction family's
   guarantee does not cover it.

<!-- END EX06-03 PROMPT -->

<!-- prompt-id: CH06-E04 -->

### Exercise 4: Name the two models behind an interaction test {.unnumbered #ex-ch06-04}

The interaction test for the 24 feedback/practice scores compares two
models fitted to the same pupils: the additive model, with residual sum
of squares 124 on 20 df, and the full factorial model, with 60 on 18 df.

1. Name the two models and say what the larger one adds.
2. The comparison gives $F=9.6$ on 2 and 18 df. Explain why its
   denominator comes from the full model rather than the additive model.
3. After the two Explanatory-Spaced pupils scoring 19 and 20 are
   removed, the sequential sums of squares for feedback and practice
   change with the order of the formula, yet the comparison of the
   additive and full models gives $F(2,16)=5.7426$ in either order.
   Explain why.
4. Say what a nonsignificant result of this comparison would and would
   not show about additivity.

<!-- END EX06-04 PROMPT -->

<!-- prompt-id: CH06-E05 -->

### Exercise 5: Preserve the event, denominator and expected counts {.unnumbered #ex-ch06-05}

In the invented table, Online has 32 completions and 18 non-completions;
In person has 18 completions and 32 non-completions. For Online, the
uncorrected score test of $p=.5$ gives $X^2=3.92$ and $p=.0477$, with the
95% Wilson interval $[.5014, .7586]$. For the complete table, Pearson's
test gives $X^2=7.84$ on 1 df and $p=.0051$.

1. State the null and the denominator of each test, and explain why the
   two p-values answer different questions.
2. Calculate the expected count of Online completions under
   independence, and explain why the chi-squared approximation is checked
   against expected rather than observed counts.
3. The difference in proportions is .28, the risk ratio 1.7778 and the
   odds ratio 3.1605. Explain why the odds ratio lies farther from 1 than
   the risk ratio here, and why Cramér's V, .28, equals the difference in
   proportions only in a table like this one.
4. Say what design information a claim that the online format causes
   completion would additionally need.

<!-- END EX06-05 PROMPT -->

<!-- prompt-id: CH06-E06 -->

### Exercise 6: Distinguish study size, repetitions and planning targets {.unnumbered #ex-ch06-06}

In the chapter's simulation, each condition repeats 2,000 generated
studies of three Normal groups with common SD $\sqrt{10/3}$ and applies
the .05 omnibus ANOVA. With equal means $(8,8,8)$ the rejection rates are
.056 at four pupils per group and .050 at twelve; with means 6, 8 and
10 the estimated power is .6545 at four and .9990 at twelve.

1. Say what the rejection rates under equal means estimate, and explain
   why the power rises from four to twelve pupils per group while those
   rates stay near .05.
2. The Monte Carlo standard error (MCSE) of the power estimate .6545
   with $B=2000$ is 0.0106. Say what it measures, and why it differs
   from the SE of a contrast within one simulated study.
3. Predict what quadrupling $B$ to 8,000 would do to this MCSE and to the
   power that the simulation estimates.
4. For the declared two-comparison family, 17 pupils per group first
   reach power .8 for a 2-point comparison, and 35 per group first reach
   a planned simultaneous half-width of at most one point. Say what each
   target controls, and why a realised interval can have a different
   width from its planning value.

<!-- END EX06-06 PROMPT -->

<!-- prompt-id: CH06-E07 -->

### Exercise 7: Distinguish an estimate from its fitting criterion {.unnumbered #ex-ch06-07}

Keep 32 successes in 50 trials fixed. Minimizing the negative binomial
log-likelihood with `optimize()` over $[0,1]$ returns `minimum` .64 and
`objective` 2.1465; at $p=.5$ the negative log-likelihood is 4.1330.

1. Say which returned value is the parameter estimate and which is the
   criterion value, and why 2.1465 is neither a probability that $p=.64$
   nor an SE.
2. The derivative of the log-likelihood is
   $\ell'(p)=(32-50p)/\{p(1-p)\}$. Use its sign to explain why the
   maximum lies at $32/50$.
3. Before calculating, predict how the fitted probability and the fall in
   log-likelihood from $p=.64$ to $p=.5$ would change with 64 successes in
   100 trials. Then compare your prediction with the supplied falls,
   1.9864 for 32 of 50 and 3.9729 for 64 of 100.
4. Explain why the likelihood curve is not a probability distribution
   over $p$, and what an interior numerical search would return if all
   50 trials had been successes.

<!-- END EX06-07 PROMPT -->

<!-- prompt-id: CH06-E08 -->

### Exercise 8: Write a qualified result paragraph {.unnumbered #ex-ch06-08}

Choose the one-way comparisons, the selected interaction contrast or the
two-group count comparison from this chapter. Write a short paragraph giving the
question, data and denominator, estimate, interval and its family or
pointwise status, working conditions and scope. Identify one tempting claim
that the numerical result does not establish, and explain the missing support.

<!-- END EX06-08 PROMPT -->

## References {.unnumbered}
