---
title: "Multiple Regression I: Conditioning, Checks, and Coordinated Questions"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AtBeginDocument{\renewcommand{\boldsymbol}[1]{\symbf{#1}}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.pos = "!ht")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 7, width = 68)
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 11))
```


\clearpage

::: {.callout-note title="Chapter overview"}
One set of records can owe several answers, and multiple regression is
the model behind most of them. This chapter first writes the multiple
regression model, says what a partial slope compares, and draws a
simulated process with two predictors. It then fits the model to the
98 occupations of Chapter 7 that have a recorded type, attaches
uncertainty to the education slope and checks the model, computes a
fitted profile and an average over named profiles, and compares two
prediction rules on held-out occupations. It closes with a short report
that owes three answers, a marginal description, a conditional
association and a prediction comparison, and reads and evaluates them.
By the end you will be able to write, fit and read an additive multiple
regression in R, keep changes of records, comparison, coefficient
coordinates and uncertainty calculation apart, judge a held-out
comparison, and evaluate coordinated answers without merging them.
:::

A report on occupational income may ask whether education and income move
together, whether that relationship remains when other characteristics of
the occupations are held constant, and whether those characteristics help
predict income. The questions inform one another, but answering one does
not answer the other two.

We use a **retrospective teaching study written for this chapter**. Its
records are the *Prestige* data distributed with the carData package
[@carData2022; @fox2019], whose documentation describes the data; the
analyses and the short report are ours.

A simple regression models an outcome with one predictor,
$Y_i=\beta_0+\beta_1x_i+\varepsilon_i$. Multiple regression keeps that
form and adds one term for each further predictor.

## Model Formulation and Interpretation {#sec-ch8-comparisons}

### The population and observation forms

With several predictors, the model for an outcome is

$$
Y=\beta_0+\beta_1x_1+\cdots+\beta_{p-1}x_{p-1}+\varepsilon,
$$

$$
E(\varepsilon\mid x_1,\ldots,x_{p-1})=0,\qquad
\operatorname{Var}(\varepsilon\mid x_1,\ldots,x_{p-1})=\sigma^2;
$$

$$
Y_i=\beta_0+\beta_1x_{i1}+\cdots+\beta_{p-1}x_{i,p-1}+\varepsilon_i,
\qquad i=1,\ldots,n.
$$

The part $\beta_0+\beta_1x_1+\cdots+\beta_{p-1}x_{p-1}$ is the mean of
$Y$ among units that share those predictor values, and the error
$\varepsilon$ collects everything else. The model has $p$
coefficients, counting the intercept. $\beta_0$, the intercept, is the
mean when every predictor is zero; it anchors the equation and
describes real units only if that combination is plausible within the
data. $\beta_j$, a partial slope, is the difference in the mean of $Y$
for a one-unit difference in $x_j$, holding the other predictors
fixed. $\sigma$ is the typical spread of outcomes around the mean at
fixed predictor values. The conditions apply at every combination of
predictor values: the errors average zero and
share one variance $\sigma^2$; they are independent of one another,
which comes from how the data arose; and for the exact $t$ reference
they are Normal. The intervals used here require independent errors, the classical SE
also needs the common variance, and the exact $t$ also needs Normal
errors; the HC3 covariance of Section 3.3 relaxes the common variance
under its own large-sample conditions.

Under the Normal-error model, $Y_i$ at its fixed predictor values is
Normal with mean $\beta_0+\sum_{j=1}^{p-1}\beta_jx_{ij}$ and SD
$\sigma$. Estimating $p$ coefficients leaves $n-p$ residual degrees of
freedom. A factor with $J$ categories contributes $J-1$ indicator
columns, as Section 2.3 explains using the occupation records.

### What a partial slope compares {#ch08-several-controls}

Take two predictors, $x$ and $z$. The model mean is
$\mu(x,z)=\beta_0+\beta_Xx+\beta_Zz$. Comparing two means a unit apart
in $x$ at the same $z$ gives

$$
\mu(x+1,z)-\mu(x,z)=\beta_X(x+1)-\beta_Xx=\beta_X.
$$

All the shared terms cancel. Holding constant names this comparison of
model means; it does not mean that $x$ was changed experimentally, that
two observed units match on everything else, or that unmeasured
differences have been controlled. A partial slope's meaning belongs to
its named conditioning set. The same predictor can have a different
coefficient when a different set is held fixed. The **additive** form
also makes its comparison the same at every value of the other
predictors; Chapter 10 asks when that comparison itself varies.

Partial correlation holds variables fixed another way, by
residualizing: it removes their fitted linear parts from two variables
before correlating what remains. A fitted equation with
several predictors holds the others fixed directly, because two of its
fitted means that differ in one predictor share every other term;
Section 3.1 shows that the two routes give the same slope. Chapter 9,
Section 2, applies the same idea to adjusted group means in the
analysis of covariance.

### Simulating and graphing a two-predictor model

A known process makes the comparison concrete. Suppose

$$
Y=50+2X+5Z+\varepsilon,
\qquad E(\varepsilon\mid X,Z)=0,
$$

with finite variances and
$\operatorname{Cov}(X,Z)/\operatorname{Var}(X)=0.65$.
The conditional $X$ coefficient is 2, and the marginal linear slope is

$$
\frac{\operatorname{Cov}(X,Y)}{\operatorname{Var}(X)}
=2+5\frac{\operatorname{Cov}(X,Z)}{\operatorname{Var}(X)}
=2+5(0.65)=5.25.
$$

Generate independent standard Normal values $X$ and $U$, let
$Z=.65X+\sqrt{1-.65^2}U$, and add independent Normal errors with SD 6.
Both predictors then have variance one and correlation .65.
@fig-ch8-known-process draws one generated sample of 250 cases from this
process, with the process's own mean lines rather than fitted ones. Each
solid line holds $Z$ at one value and rises 2 for each unit of $X$. The
dashed marginal line rises 5.25, because higher $X$ comes with higher $Z$
(the darker points sit toward the upper right) and the marginal line
carries that association along.

```{r}
#| label: fig-ch8-known-process
#| fig-cap: "One generated sample from the known process, with its marginal line and three lines at fixed $Z$."
#| fig-alt: "Scatterplot of 250 generated points of Y against standardized X, shaded by Z, with darker points higher and to the right. Three parallel solid lines at Z equal to minus one, zero and one rise with slope 2. A steeper dashed line, the marginal line, rises with slope 5.25 and crosses the three solid lines."
#| fig-width: 6.1
#| fig-height: 3.5
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
set.seed(82108)
ch8a_n <- 250
ch8a_x <- rnorm(ch8a_n)
ch8a_z <- .65 * ch8a_x + sqrt(1 - .65^2) * rnorm(ch8a_n)
ch8a_y <- 50 + 2 * ch8a_x + 5 * ch8a_z + rnorm(ch8a_n, sd = 6)
ch8a_sim <- data.frame(strategy = ch8a_x, prior = ch8a_z,
                       score = ch8a_y)
sim <- data.frame(x = ch8a_x, z = ch8a_z, y = ch8a_y)
ch8_lines <- data.frame(prior = c(-1, 0, 1),
                        intercept = 50 + 5 * c(-1, 0, 1))
ggplot2::ggplot(ch8a_sim, ggplot2::aes(strategy, score)) +
  ggplot2::geom_point(ggplot2::aes(colour = prior), size = 1.2,
                      alpha = .8) +
  ggplot2::geom_abline(data = ch8_lines,
                       ggplot2::aes(intercept = intercept, slope = 2),
                       colour = "#245C78", linewidth = .8) +
  ggplot2::geom_abline(intercept = 50, slope = 5.25, linetype = 2,
                       linewidth = .9) +
  ggplot2::annotate("text", x = 3.1, y = 50 + 2 * 3.1 + 5 * c(-1, 0, 1),
                    label = c("Z = -1", "Z = 0", "Z = 1"),
                    hjust = 0, vjust = -0.35, size = 3.1,
                    colour = "#245C78") +
  ggplot2::annotate("text", x = -2.6, y = 74, hjust = 0, vjust = 1,
                    size = 3.1,
                    label = "Dashed: marginal line, slope 5.25\nSolid: fixed Z, slope 2") +
  ggplot2::scale_colour_gradient(low = "grey85", high = "black") +
  ggplot2::scale_x_continuous(limits = c(-2.6, 3.6), breaks = -2:3) +
  ggplot2::labs(x = "X (standardized)", y = "Y", colour = "Z")
```

With these generated columns named `x`, `z` and `y` in `sim`, the
following calls estimate the marginal and conditional equations. The
short examples throughout this chapter show the statistical operation
on the data just described.

```{r}
#| label: ch8-known-demo
#| echo: true
#| eval: false
coef(lm(y ~ x, data = sim))
coef(lm(y ~ x + z, data = sim))
```

In this sample the fitted $X$ coefficient is 3.0259 (SE 0.5325), the
$Z$ coefficient 4.3991 (SE 0.5282), and the marginal slope 5.9971
(SE 0.4466). Each lies within two SEs of its programmed value, 2, 5,
and 5.25 respectively. A single sample estimates the process rather
than reproducing its parameters exactly.

Mapping `x` and `y` to position and `z` to colour draws the points in
@fig-ch8-known-process. The fitted slopes and the known process lines
answer related but distinct questions.

```{r}
#| label: ch8-known-plot-demo
#| echo: true
#| eval: false
library(ggplot2)
ggplot(sim, aes(x, y, colour = z)) + geom_point()
```

The two slopes answer different questions: one holds $Z$ fixed, the
other includes how $Z$ moves with $X$, and neither alone says what would
happen if someone intervened on $X$. Calling 5.25 "biased" requires
naming the target it is used for; as the marginal linear slope it is
correct. Appendix A gives the derivation and the repeated samples.

## Estimating the Model on the Occupation Records {#sec-ch8-questions}

### The records and the working model

A row is an **occupation**, not a worker. Income is the occupation's
average annual income in 1971 Canadian dollars, divided here by 1,000;
education is average years of education; *women* is the percentage of
women; prestige is a rating from a mid-1960s survey, earlier than the
1971 income and education; and type is blue-collar (*bc*),
professional/managerial/technical (*prof*), or white-collar (*wc*). These
historical occupational averages are not current individual earnings.

Chapter 7 used all 102 occupations. The full variable set leaves 98
complete records, because athletes, newsboys, babysitters, and farmers
have no recorded type; the three types keep 44, 31, and 23 records. We
keep the source row order, which also makes the validation split of
Section 6 reproducible. The original 102 records are `d0`; `dat` holds
the 98 complete records, with `income_k` in thousand dollars and type
levels ordered `bc`, `prof`, `wc`.

```{=latex}
\Needspace{10\baselineskip}
```

```{r}
#| label: ch8-records-prepare
#| include: false
d0 <- carData::Prestige
all_occ <- d0
all_occ$occupation <- rownames(all_occ)
all_occ$income_k <- all_occ$income / 1000
all_occ$type <- factor(all_occ$type, levels = c("bc", "prof", "wc"))
vars <- c("income_k", "education", "prestige", "women", "type")
dat <- all_occ[complete.cases(all_occ[, vars]), ]
```

```{r}
#| label: ch8-records
#| echo: true
#| eval: false
c(original = nrow(d0), retained = nrow(dat))
table(dat$type)
```

We refit the education-only model on these **same 98 records** before any
adjusted comparison, so that the move from 102 to 98, a change in which
records are described, is never presented as an adjustment effect.
Complete records make the computations comparable; they do not make the
omitted occupations irrelevant or the 98 a probability sample of Canadian
occupations.

The working occupation model is

$$
\mathrm{income}_k=\beta_0+\beta_Ee+\beta_Pp+\beta_Ww
+\delta_{\mathrm{prof}}D_{\mathrm{prof}}+\delta_{\mathrm{wc}}D_{\mathrm{wc}}
+\varepsilon,
$$

where $e$ is education, $p$ prestige, $w$ percentage of women, and each
indicator $D$ is 1 for its named type. Blue-collar, with both indicators
zero, is the reference type. There are six coefficients. The
conditional question concerns $\beta_E$, comparing education at the same
prestige, women and type. Least squares describes the recorded
occupations whether or not that stochastic model holds; Section 3
applies its conditions to attach uncertainty.

A **best linear projection**, summarizing a curved relationship over a
stated predictor distribution, is a different declared target, not a
rescue adopted after curvature appears. Chapter 9, Appendix A,
develops it.

### Least squares and the fits on the same rows

Least squares chooses the coefficients that make the residual sum of
squares smallest. Its residual SD,
$s_e=\sqrt{RSS/(n-p)}$, describes the typical residual size; the full
model below gives 2.6187 thousand dollars on 92 df. $R^2$ is the share
of centered outcome variation accounted for in these records; Appendix B
introduces its adjusted version.

```{r}
#| label: ch8-fits
#| echo: true
m0 <- lm(income_k ~ education, data = dat)
m_w <- lm(income_k ~ education + women, data = dat)
m1 <- lm(income_k ~ education + prestige, data = dat)
m2 <- lm(income_k ~ education + prestige + women + type, data = dat)
coef(m_w)
coef(m2)
```

| Records and predictors | Education slope | In-sample $R^2$ | Role here |
|:--|--:|--:|:--|
| Earlier 102, education only | 0.8988 | .3336 | Chapter 7's line on different records |
| Same 98, education only | 0.8830 | .3296 | The marginal description |
| Same 98, education and women | 0.9292 | .5641 | Holds women fixed; the slope barely moves |
| Same 98, education and prestige | $-0.2179$ | .4996 | Holds prestige fixed; the slope changes sign |
| Same 98, all declared predictors | 0.1164 | .6362 | The conditional answer about $\beta_E$ |

: The model sequence separates sample composition from conditioning. {#tbl-ch8-sequence tbl-colwidths="[36,18,16,30]"}

The education-and-women fit gives a concrete reading: two occupations
one year apart in education, with the same percentage of women, differ
in fitted income by 0.9292 thousand dollars (education SE 0.1044).
At the same education, one percentage point more women corresponds to
0.0654 thousand dollars lower fitted income (SE 0.0091). Its $R^2$ is
.5641. These are comparisons of occupations, not effects of changing an
individual's education or occupation.

Why does adding prestige reverse the education slope while adding
women barely changes it? Education correlates .8665 with prestige and
.0619 with women. Correlation alone is only part of the explanation:
how far the education slope moves depends on the added predictor's
coefficient *and* its slope on education. For least-squares fits on the
same rows, the sample omitted-variable identity is

$$
\begin{aligned}
\text{marginal slope}&=\text{slope with }Z\\
&\quad+(\text{coefficient of }Z)(\text{slope of }Z\text{ on education}),\\
0.8830&=-0.2179+0.2043(5.3884)\qquad(Z=\text{prestige}),\\
0.8830&=0.9292+(-0.0654)(0.7069)\qquad(Z=\text{women}).
\end{aligned}
$$

Appendix A derives the identity; the displayed numbers are rounded.
The carried parts are 1.1009 for prestige and $-0.0462$ for women.
@fig-ch8-grid shows the corresponding fitted lines beside the marginal
line, extending the known-process display to real records.

```{r}
#| label: fig-ch8-grid
#| fig-width: 6.4
#| fig-height: 3.5
#| fig-cap: "Fitted income against education at three held values of prestige (a) or of the percentage of women (b), each line drawn over the education observed near its held value, with the dashed marginal education-only line."
#| fig-alt: "Both panels show the 98 occupation records and a rising dashed marginal line. In panel a, gently falling solid lines at prestige 30, 50 and 70 sit over successively higher education ranges. In panel b, rising solid lines at 10, 50 and 90 percent women are almost parallel to the marginal line, with the line for more women lower."
ch8_lines <- function(fit, held, values, panel) {
  do.call(rbind, lapply(values, function(v) {
    nearby <- dat[abs(dat[[held]] - v) <= 7.5, ]
    grid <- data.frame(education = seq(min(nearby$education),
      max(nearby$education), length.out = 40))
    grid[[held]] <- v
    data.frame(Panel = panel, Held = v, education = grid$education,
      fit = predict(fit, newdata = grid))
  }))
}
ch8_panels <- c("(a) Prestige held at 30, 50, 70",
  "(b) Women held at 10, 50, 90 percent")
ch8_grid <- rbind(ch8_lines(m1, "prestige", c(30, 50, 70), ch8_panels[1]),
  ch8_lines(m_w, "women", c(10, 50, 90), ch8_panels[2]))
ch8_points <- data.frame(Panel = rep(ch8_panels, each = nrow(dat)),
  education = dat$education, income_k = dat$income_k)
ch8_ends <- ch8_grid[!duplicated(ch8_grid[c("Panel", "Held")],
  fromLast = TRUE), ]
ggplot2::ggplot(ch8_grid, ggplot2::aes(education, fit)) +
  ggplot2::geom_point(data = ch8_points, ggplot2::aes(y = income_k),
    colour = "grey70", size = .9) +
  ggplot2::geom_abline(intercept = coef(m0)[1], slope = coef(m0)[2],
    linetype = 2, linewidth = .7) +
  ggplot2::geom_line(ggplot2::aes(group = Held), colour = "#245C78",
    linewidth = .9) +
  ggplot2::geom_text(data = ch8_ends, ggplot2::aes(label = Held),
    hjust = -.25, size = 3, colour = "#245C78") +
  ggplot2::facet_wrap(~Panel) +
  ggplot2::scale_x_continuous(limits = c(6, 17.2)) +
  ggplot2::labs(x = "Education (years)",
    y = "Income (thousand 1971 dollars)")
```

In panel (a), each solid line falls 0.2179 per education year at fixed
prestige. The higher-prestige lines sit over higher education, so the
rising marginal line crosses them. In panel (b), the solid lines rise
0.9292 per year, nearly the marginal slope, because women hardly tracks
education. Each line uses the education range among records within 7.5
points of its held value, a display choice that respects nearby
observations; Section 4.2 checks support more fully.

For example, the first panel's line at prestige 50 combines the fitted
intercept with $50\widehat\beta_P$; its slope is $\widehat\beta_E$.
The nearby points keep the comparison close to observed education
values.

```{r}
#| label: ch8-adjusted-line-demo
#| echo: true
#| eval: false
near <- dat[abs(dat$prestige - 50) <= 7.5, ]
ggplot(near, aes(education, income_k)) + geom_point() +
  geom_abline(intercept = coef(m1)[1] + 50*coef(m1)["prestige"],
              slope = coef(m1)["education"])
```

The sequence of slopes names different comparisons; it is not a true
effect being explained away. Adding predictors on the same rows cannot
increase the smallest achievable RSS, so training $R^2$ cannot fall.
The separate gains need not add up: women adds .2345, prestige adds
.1701, and the full model adds .3066 over education alone. These gains
concern training fit, not held-out prediction.

### Indicators and a fitted profile

A three-level factor needs two indicators. With blue-collar as reference,
$D_{\mathrm{prof}}$ and $D_{\mathrm{wc}}$ identify the other types;
the full occupation fit is

$$
\begin{aligned}
\widehat{\mathrm{income}}_k
={}&0.4205+0.1164e+0.1406p-0.0535w\\
 &+0.3242D_{\mathrm{prof}}+0.2353D_{\mathrm{wc}}.
\end{aligned}
$$

Subtract two fitted means one education year apart at the same prestige,
percentage of women and type. Every term except education cancels:

$$
\widehat m(e+1,p,w,t)-\widehat m(e,p,w,t)=\widehat\beta_E=0.1164.
$$

Thus two such occupations differ by about 116 dollars in fitted income.
At the same values of all the other predictors, a prestige point compares
about 141 dollars more, and a percentage point more women compares about
54 dollars less. Each sentence names what is held fixed.

For a white-collar profile with education 11, prestige 45, and women 50,
$D_{\mathrm{prof}}=0$ and $D_{\mathrm{wc}}=1$, and the unrounded fitted
coefficients give

$$
\begin{aligned}
\widehat m_*
={}&\widehat\beta_0+11\widehat\beta_E+45\widehat\beta_P
+50\widehat\beta_W+\widehat\delta_{\mathrm{wc}}\\
\approx{}&5.5886,
\end{aligned}
$$

a fitted mean of about 5,589 dollars on the historical income scale, not
a guarantee for any occupation or worker.

The intercept refers to zero education, prestige, and women in the
reference type; it anchors the fitted surface and need not describe a
plausible occupation. Nor is one type coefficient "the effect of
occupational type". When a factor stands alone, each coefficient is a
difference between group means; here each type
coefficient compares fitted means at the same education, prestige, and
women. Appendix A gives the group-mean and releveling calculations, and
Chapter 9 develops contrasts, block tests, and the analysis-of-variance
identity.

The substitution uses six multipliers, one for each coefficient.
@tbl-ch8-design-rows puts them in coefficient order for three actual
occupations and for the declared profile. The first multiplier is 1
because every fitted value includes the intercept. A professional
record selects the professional coefficient, a white-collar record the
white-collar coefficient, and a blue-collar record selects neither.

| Record or profile | Intercept | Education | Prestige | Women | Professional | White-collar |
|:--|--:|--:|--:|--:|--:|--:|
| Nursing aides | 1 | 9.45 | 34.9 | 76.14 | 0 | 0 |
| Government administrators | 1 | 13.11 | 68.8 | 11.16 | 1 | 0 |
| Medical technicians | 1 | 12.79 | 67.5 | 76.04 | 0 | 1 |
| Declared white-collar profile | 1 | 11.00 | 45.0 | 50.00 | 0 | 1 |

: Each row lists the multipliers of the six coefficients. {#tbl-ch8-design-rows tbl-colwidths="[30,9,12,12,11,13,13]"}

This gives a concrete meaning to the vector notation. Write the
multipliers as the column vector
$\mathbf x_i=(1,e_i,p_i,w_i,D_{\mathrm{prof},i},D_{\mathrm{wc},i})^{\mathsf T}$
and the corresponding coefficients as
$\boldsymbol\beta=(\beta_0,\beta_E,\beta_P,\beta_W,
\delta_{\mathrm{prof}},\delta_{\mathrm{wc}})^{\mathsf T}$.
A **vector** is an ordered collection of numbers; the superscript
$\mathsf T$ denotes the transpose, which turns a column into a row.
The product $\mathbf x_i^{\mathsf T}\boldsymbol\beta$ adds the six
multiplier--coefficient products, exactly as the scalar equation does.

Stacking the 98 observed rows gives the $98\times6$ **design matrix**
$\mathbf X$. Stacking the outcomes and errors gives the compact model
$\mathbf Y=\mathbf X\boldsymbol\beta+\boldsymbol\varepsilon$.
The design matrix stores predictors and indicators, not income. Distinct
coefficients require that no column be an exact linear combination of
the others, the **full-column-rank condition**. Near collinearity affects
precision, as Section 4.2 explains. The
profile is a new row for prediction, not an extra record used to fit
the model. `model.matrix()` recovers the observed multiplier rows;
multiplying `row` by the fitted coefficients recovers the profile mean.

```{r}
#| label: ch8-design-demo
#| echo: true
#| eval: false
rows <- c("nursing.aides", "gov.administrators", "medical.technicians")
model.matrix(m2)[rows, ]
row <- c(1, 11, 45, 50, 0, 1)
sum(row * coef(m2))
```

Section 5 uses this notation to carry coefficient uncertainty to the
profile; Chapter 9, Appendix A, develops the matrix mathematics. Section 4.2 checks whether the three continuous values of that white-collar
profile occur together, using a weighted combination of observed records.

## Inference for the Partial Slope {#sec-ch8-estimate}

Apply Section 1.1's model to the occupation records: incomes vary
independently around the linear mean of Section 2.1 at the recorded
predictor rows, with a common error SD $\sigma$. The residual SD 2.6187
estimates it on 92 df. The repetition imagined for the classical
intervals is new incomes at the same 98 predictor rows; the historical
records do not establish that model or a probability sample.

### The partial slope from residuals {#ch08-partial-regression}

The partial slope can be recovered by removing the fitted parts of
the other predictors from both variables. Regress education on
prestige, percentage of women, and type and keep the residuals
$r_{E,i}$; do the same for income to obtain $r_{Y,i}$. Both auxiliary
fits include an intercept and use exactly the 98 rows of the full fit.
Then

$$
b_E=\frac{\sum_i r_{E,i}r_{Y,i}}{\sum_i r_{E,i}^2}
    =r_{EY\cdot Z}\frac{s_{r_Y}}{s_{r_E}},
$$

where $Z$ denotes the control columns and $r_{EY\cdot Z}$, the
correlation of the two residual vectors, is the partial correlation. The
numerator is the linear co-variation left after the controls and the
denominator the education variation left; the ratio of residual SDs turns
the unit-free correlation into thousand dollars per education year.

```{r}
#| label: ch8-partial-demo
#| echo: true
#| eval: false
r_E <- residuals(lm(education ~ prestige + women + type, data = dat))
r_Y <- residuals(lm(income_k ~ prestige + women + type, data = dat))
sum(r_E * r_Y) / sum(r_E^2)
```

The uncertainty must count the fitted controls. Prestige and women supply
one column each and type two, so with education and the intercept the
model estimates $p=6$ coefficients from $n=98$ rows, leaving $n-p=92$
residual degrees of freedom. Using the residual variance $s_e^2=RSS/(n-p)$ of Section 2.2,

$$
SE(b_E)=\sqrt{\frac{s_e^2}
                         {\sum_i r_{E,i}^2}},
\qquad
t_E=r_{EY\cdot Z}\sqrt{\frac{n-p}{1-r_{EY\cdot Z}^2}}.
$$

The slope identity is algebraic; the SE and the $t_{92}$ reference for
$H_0:\beta_E=0$ rest on the classical model of Section 1.1. In
@tbl-ch8-partial-df the first row is the full model's education row,
which these formulas reproduce, and the second is a fresh fit of residual
income on residual education.

\Needspace{10\baselineskip}

| Calculation on the same 98 rows | Slope | SE | df | $t$ | $p$ |
|:--|--:|--:|--:|--:|--:|
| Full model `m2`, education row; the residual formulas with $n-p=92$ | 0.1164 | 0.2752 | 92 | 0.4230 | .6732 |
| A fresh fit of residual income on residual education | 0.1164 | 0.2694 | 96 | 0.4321 | .6666 |

: The residual slope reproduces the full-model coefficient; only the full model's df gives its uncertainty. {#tbl-ch8-partial-df tbl-colwidths="[44,11,11,8,12,14]"}

Both rows share the slope 0.1164 and the partial correlation .0441. The
fresh fit counts $98-2=96$ df, as if fitting the four control columns had
cost nothing, so its SE is smaller by the factor $\sqrt{92/96}$. Use
`m2`, or the formulas with $n-p=92$, for uncertainty.

Only the education variation left after the controls informs this
coefficient: its residual SD is 0.9661 years, against 2.7489 before
residualizing. The added-variable display below shows what remains.

```{r}
#| label: fig-ch8-partial
#| fig-cap: "Marginal and residualized income--education comparisons in the same 98 occupations."
#| fig-alt: "The marginal scatterplot has a positive fitted line. After linearly removing prestige, women, and type from both variables, the fitted line is nearly flat and substantial residual variation remains."
#| fig-width: 6.4
#| fig-height: 3.5
r_E <- residuals(lm(education ~ prestige + women + type, data = dat))
r_Y <- residuals(lm(income_k ~ prestige + women + type, data = dat))
oldpar <- par(mfrow = c(1, 2), mar = c(4.3, 4.2, 1, .5), mgp = c(2.7, .8, 0))
plot(dat$education, dat$income_k, pch = 1, xlab = "Education (years)",
     ylab = "Income (thousand dollars)")
abline(m0, col = "#245C78", lwd = 2)
plot(r_E, r_Y, pch = 1, xlab = "Residual education (years)",
     ylab = "Residual income (thousand dollars)")
abline(a = 0, b = coef(m2)["education"], col = "#245C78", lwd = 2)
par(oldpar)
```

@fig-ch8-partial sets the two comparisons side by side. The marginal
line rises 0.8830 per year; after the fitted linear parts of prestige,
women, and type are removed from both variables, the residual line rises
only 0.1164, with wide scatter around it. Residualizing removes those
**fitted linear components**, not every substantive difference between
occupations. The residualized panel can be drawn directly from these
two residual vectors and the full model's slope.

```{r}
#| label: ch8-partial-plot-demo
#| echo: true
#| eval: false
ggplot(data.frame(r_E, r_Y), aes(r_E, r_Y)) + geom_point() +
  geom_abline(intercept = 0, slope = coef(m2)["education"])
```

### Two covariance matrices

Covariance now has two meanings. `cov()` describes how
recorded variables spread and move together; `vcov()` estimates the
covariance of the fitted coefficients. @tbl-ch8-covariance-objects
compares them on the same 98 rows.

\Needspace{16\baselineskip}

| | Recorded-data covariance | Coefficient covariance |
|:--|:--|:--|
| R call | `cov()` of income, education, prestige and women | `vcov(m2)` |
| Rows and columns | 4 numeric variables; the type factor is left out | 6 fitted coefficients, including the intercept and two type indicators |
| Diagonal entries | Sample variances | Squared coefficient SEs |
| Off-diagonal entries | Sample covariances between variables | Joint uncertainty of two coefficients |
| Education entry | 7.5567 years squared | 0.0757; its square root is the SE 0.2752 |
| Question it answers | How the recorded variables spread and move together across occupations | How uncertain the fitted coefficients are, and how their errors move together |

: Two covariance matrices from the same 98 records answer different questions. {#tbl-ch8-covariance-objects tbl-colwidths="[22,39,39]"}

```{r}
#| label: ch8-covariance-demo
#| echo: true
#| eval: false
cov(dat[c("income_k", "education", "prestige", "women")])
vcov(m2)
```

The education entries differ by a factor of about 100 because they
measure different things: the spread of education across occupations
and the squared uncertainty of one fitted coefficient. The matrices are
not interchangeable; Section 5 uses the coefficient covariance in
$\mathbf x_*^{\mathsf T}\widehat{\mathbf V}\mathbf x_*$.

### Interval and tests

With 92 residual degrees of freedom, the classical 95% interval for
education is

$$
\begin{aligned}
b_E\pm t_{.975,92}SE(b_E)
 &=0.1164\pm1.9861(0.2752)\\
 &\approx[-0.4302,\ 0.6631].
\end{aligned}
$$

Under the classical assumptions the interval is compatible with negative
and positive conditional slopes, including differences of several
hundred dollars per year of education. With no margin of practical
equivalence declared, the direction and the practically relevant size
remain unresolved; the interval does not show that the slope is zero or
that adjustment has removed an effect of education.

```{r}
#| label: ch8-uncertainty
#| echo: true
confint(m2, "education")
hc3_covariance <- sandwich::vcovHC(m2, type = "HC3")
hc3_se <- sqrt(hc3_covariance["education", "education"])
coef(m2)["education"] + c(-1, 1) * qt(.975, 92) * hc3_se
```

HC3 keeps the coefficient and replaces its covariance estimate, giving
SE 0.3823 and, with the same $t_{92}$ convention, the approximate
interval $[-0.6429,0.8758]$. It allows unequal error variances among
independent observations under large-sample conditions and is not an
exact small-sample guarantee. Neither interval allows for unmodeled
dependence, selection, measurement error, or a wrong mean form.

The matching test addresses $H_0:\beta_E=0$ in the stated full model,
not "education has no effect". Its classical $t=0.423$ and $p=.6732$ do
not prove the null, and the test does not decide which of the three
questions deserves an answer.

The **overall $F$ test** compares the full regression with a one-mean
model on the same rows, setting all $p-1$ non-intercept coefficients to
zero. This extends Chapter 4's comparison of named mean models. Under
the same Normal-error conditions as the coefficient tests,
its statistic is

$$
F_{\mathrm{overall}}=
\frac{RegSS/(p-1)}{RSS/(n-p)},
\qquad RegSS=\sum_i(\widehat y_i-\bar y)^2.
$$

\Needspace{10\baselineskip}

```{r}
#| label: ch8-overall-f
#| echo: true
summary(m2)$fstatistic
```

The full model gives $F(5,92)=32.1747$, $p<.0001$; the education-only model on the
same 98 records gives $F(1,96)=47.1956$. For education alone the statistic is the square of its slope's
$t=6.8699$. Each tests its stated joint restriction; the full-model test
does not identify which coefficient differs from zero. Neither verifies the
mean function or shows how well the fitted rule will predict a held-out
outcome. Chapter 9 develops the general nested comparison behind this
statistic, its matrix form, and the contrasts, block tests, and adjusted
group means that a factor with several levels invites.

Not every change to a model changes the scientific estimand. In a
randomized study, well-specified covariate adjustment can estimate the
same declared average assignment effect more precisely; releveling a
factor changes the coefficient coordinates but not the fitted values;
and a different uncertainty estimator changes the interval but not the
coefficient.

## Checking the Model {#sec-ch8-check}

What would these plots look like if the working model described the
data well? A fit produces numbers even when its mean or uncertainty
model is unsuitable. Checking asks which part of the interpretation the
evidence supports; no single graph passes or fails the model.

### Residuals, leverage, and influence

A residual is an observed deviation from a fitted value, not the
unobserved error. Leverage measures how unusual a record's combination
of predictor values is; a high-leverage record pulls the fit toward
itself, so its residual tends to be smaller. A standardized residual
therefore divides each residual by its own estimated SD, which combines
the common residual scale with the record's leverage, and an externally
studentized residual does the same with the scale estimated with that
record left out. Influence, measured by Cook's distance, combines
leverage with the size of the residual [@cook1977]. This call lists
the five occupations with the largest Cook's distance.

\Needspace{8\baselineskip}

```{r}
#| label: ch8-diagnostics-prepare
#| include: false
ch8_diag <- data.frame(
  occupation = rownames(dat), fitted = fitted(m2),
  residual = residuals(m2), standardized = rstandard(m2),
  studentized = rstudent(m2), leverage = hatvalues(m2),
  cooks = cooks.distance(m2)
)
ch8_top <- order(ch8_diag$cooks, decreasing = TRUE)[1:5]
ch8_diag[ch8_top, c("residual", "studentized", "leverage", "cooks")]
```

```{r}
#| label: ch8-diagnostics
#| echo: true
#| eval: false
sort(cooks.distance(m2), decreasing = TRUE)[1:5]
```

@fig-ch8-diagnostics shows four views. If the linear mean described the
records well, the residuals in panel (a) would scatter evenly around zero
with a flat smooth; here the smooth drifts downward and the scatter
widens at high fitted income. With constant spread, the smooth in panel
(b) would stay level; it rises. Panel (c) follows the Normal reference
line in the middle but departs in both tails, most strongly on the
right, where a few occupations earn far more than the model predicts;
on the left, ministers earn far less. The two largest positive
residuals, general managers and physicians in the printed list, also
have the largest Cook's distances, drawn as the biggest circles in
panel (d).

```{r}
#| label: fig-ch8-diagnostics
#| echo: false
#| fig-cap: "Four complementary diagnostic views for the full additive model."
#| fig-alt: >
#|   Residual and scale plots show structure and changing spread.
#|   A normal quantile plot departs in both tails, most strongly on
#|   the right. The final panel relates leverage and studentized
#|   residuals, with point sizes indicating Cook's distance.
#| fig-width: 6.7
#| fig-height: 6.2
local({
  old <- par(no.readonly = TRUE)
  on.exit(par(old))
  par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
  z <- ch8_diag
  plot(z$fitted, z$residual, pch = 1,
       xlab = "Fitted income", ylab = "Residual",
       main = "(a) Mean pattern")
  abline(h = 0, lty = 2)
  lines(lowess(z$fitted, z$residual), lwd = 2)
  plot(z$fitted, sqrt(abs(z$standardized)), pch = 1,
       xlab = "Fitted income", ylab = "Sqrt |standardized|",
       main = "(b) Scale pattern")
  lines(lowess(z$fitted, sqrt(abs(z$standardized))), lwd = 2)
  qqnorm(z$studentized, pch = 1, main = "(c) Normal reference")
  qqline(z$studentized, lwd = 2)
  plot(z$leverage, z$studentized, pch = 1,
       cex = .8 + 2 * sqrt(z$cooks / max(z$cooks)),
       xlab = "Leverage", ylab = "Studentized residual",
       main = "(d) Influence")
  abline(h = 0, lty = 2)
})
```

The first panel compares fitted means with observed deviations. A
horizontal reference at zero makes systematic departures visible.

```{r}
#| label: ch8-residual-plot-demo
#| echo: true
#| eval: false
check <- data.frame(fit = fitted(m2), resid = residuals(m2))
ggplot(check, aes(fit, resid)) + geom_point() +
  geom_hline(yintercept = 0)
```

These features challenge the linear mean form and the constant-variance
description, and the heavy tails matter for exact small-sample
Normal-theory inference, although OLS can still be calculated. A large
residual, high leverage, and high influence are different features, and
none alone shows that a record is invalid. No residual graph can
establish that occupations are independent draws from a wider
population.

### Joint support and collinearity

Joint support concerns combinations of predictor values. First,
@tbl-ch8-support asks a coarser question: how far does each variable
extend within each recorded type?

```{=latex}
\Needspace{8\baselineskip}
```

```{r}
#| label: tbl-ch8-support
#| echo: false
#| tbl-cap: "Predictor ranges within each recorded occupation type."
ch8_ranges <- do.call(rbind, lapply(split(dat, dat$type), function(g) {
  data.frame(n = nrow(g),
    Education = sprintf("%.2f--%.2f",
                        min(g$education), max(g$education)),
    Prestige = sprintf("%.1f--%.1f",
                       min(g$prestige), max(g$prestige)),
    Women = sprintf("%.2f--%.2f", min(g$women), max(g$women)))
}))
knitr::kable(cbind(Type = rownames(ch8_ranges), ch8_ranges),
  row.names = FALSE)
```

Blue-collar education ends at 10.93 years, below the professional minimum
of 11.09, so comparing those types at the same education crosses a gap
with no records. The model still returns predictions there, but software
output does not create observations: a lack of support limits what the
data can say about a comparison without making the comparison undefined.

Do the three values of Section 2.3's white-collar profile, education
11, prestige 45 and women 50, occur together in the records?
Separate ranges cannot say. A **convex combination** of records is a
weighted average whose weights are nonnegative and sum to one, and a
**joint-support witness** is a convex combination of observed records
that returns the profile. The four white-collar records in
@tbl-ch8-profile-support, weighted as listed, return education 11,
prestige 45, and women 50 at once.

\Needspace{11\baselineskip}

```{r}
#| label: tbl-ch8-profile-support
#| echo: false
#| tbl-cap: "Four observed white-collar records recover the declared profile."
support_names <- c("radio.tv.announcers", "tellers.cashiers",
                   "receptionsts", "sales.supervisors")
support_rows <- dat[support_names, c("education", "prestige", "women")]
support_weights <- c(0.245243736035, 0.230972611567,
                     0.226143656937, 0.297639995461)
support_display <- data.frame(
  Occupation = c("Radio/TV announcers", "Tellers/cashiers",
                 "Receptionists", "Sales supervisors"),
  support_rows,
  Weight = sub("^0", "", format(round(unname(support_weights), 4),
    digits = 15, scientific = FALSE, trim = TRUE)),
  row.names = NULL
)
knitr::kable(support_display,
  col.names = c("Occupation", "Education", "Prestige", "Women", "Weight"),
  digits = c(0, 2, 1, 2, 4), align = c("l", "r", "r", "r", "r"))
```

The four weights are positive and sum to one. With the records in
`support_rows` and the unrounded weights in `support_weights`, the
weighted column sums recover all three profile values.

```{r}
#| label: ch8-support-demo
#| echo: true
#| eval: false
colSums(support_rows * support_weights)
```

The witness places the profile
inside the joint covariate hull of the white-collar records on all three
continuous predictors. It does not show that many occupations lie near
the profile, that one matches it exactly, or that a causal comparison is
supported.

Section 2.2's adjusted-versus-marginal figure restricts its lines to
nearby records. Its education ranges are plotting neighborhoods, not a
formal guarantee of support for the full-model profile.

**Collinearity** means that a predictor is strongly linearly related to
the other predictors, so they largely predict it, and it costs
precision. The auxiliary $R^2$ from regressing education on the other
predictors, .8765, measures how strongly: they account for most of
education's variation, and education's **variance inflation factor** is

$$
VIF_E=\frac{1}{1-R^2_{E\mid P,W,T}}\approx8.0968,
\qquad \sqrt{VIF_E}\approx2.8455.
$$

@fig-ch8-vif draws the SE multiplier $\sqrt{VIF}$ against the auxiliary
$R^2$ for the three continuous predictors. The curve rises slowly at
first, so women (auxiliary $R^2$ .2312, multiplier 1.1405) loses little
precision, and it climbs steeply beyond about .8, where prestige (.8066,
multiplier 2.2738) and education sit.

```{r}
#| label: fig-ch8-vif
#| fig-cap: "The SE multiplier grows steeply as the other predictors account for more of a predictor's variation."
#| fig-alt: "A curve of the square root of the VIF against the auxiliary R-squared rises slowly from 1 at zero and steeply beyond about 0.8. Points mark women near 0.23 with multiplier 1.14, prestige near 0.81 with 2.27, and education near 0.88 with 2.85."
#| fig-width: 6.1
#| fig-height: 3.1
ch8_vif <- car::vif(m2)[c("women", "prestige", "education"), 1]
ch8_vif_points <- data.frame(name = c("Women", "Prestige", "Education"),
                             r2 = 1 - 1 / ch8_vif,
                             multiplier = sqrt(ch8_vif))
ch8_vif_curve <- data.frame(r2 = seq(0, .95, length.out = 200))
ch8_vif_curve$multiplier <- sqrt(1 / (1 - ch8_vif_curve$r2))
ggplot2::ggplot(ch8_vif_curve, ggplot2::aes(r2, multiplier)) +
  ggplot2::geom_line(linewidth = .8) +
  ggplot2::geom_point(data = ch8_vif_points, size = 2.3,
                      colour = "#245C78") +
  ggplot2::geom_text(data = ch8_vif_points,
                     ggplot2::aes(label = name),
                     hjust = 1.15, vjust = -.4, size = 3.2) +
  ggplot2::labs(x = "Auxiliary R-squared of the predictor on the others",
                y = "SE multiplier, square root of VIF")
```

```{r}
#| label: ch8-vif-demo
#| echo: true
#| eval: false
r2 <- summary(lm(education ~ prestige + women + type, data = dat))$r.squared
c(VIF = 1/(1 - r2), SE_multiplier = sqrt(1/(1 - r2)))
```

Under the classical model, holding the residual variance and education's
total variation fixed, education's SE is 2.8455 times what it would be in
an orthogonal design, where education is uncorrelated with the other
predictors. This compares two variance formulas: deleting a predictor
would change the comparison and can change the residual variance, so a
VIF is not a deletion rule, and near collinearity does not by itself
bias the coefficient under the maintained zero-conditional-mean model.

### Sensitivity to curvature and to one record

A regression is linear in its coefficients: a squared predictor can
enter a linear model even though its mean curve bends. Its form is a
changed specification to assess, not a transformation selected
automatically by a significant test.

A component-plus-residual plot adds the fitted prestige term back to
each residual, $e_i+\widehat\beta_P P_i$. In @fig-ch8-partial-residual
the smooth bends upward at high prestige, away from the straight fitted
component. That motivates a declared form sensitivity; it does not show
that any one curve is the true process.

```{r}
#| label: fig-ch8-partial-residual
#| echo: false
#| fig-cap: "The prestige component plus residual in the full additive model."
#| fig-alt: >
#|   A component-plus-residual scatterplot has a positive trend.
#|   A curved smooth bends upward at high prestige relative to the
#|   straight fitted component.
#| fig-width: 6.1
#| fig-height: 3.5
ch8_partial <- data.frame(
  prestige = dat$prestige,
  component = residuals(m2) + coef(m2)["prestige"] * dat$prestige
)
ggplot2::ggplot(ch8_partial, ggplot2::aes(prestige, component)) +
  ggplot2::geom_point(shape = 1) +
  ggplot2::geom_smooth(method = "lm", se = FALSE,
                      colour = "black", linetype = 2) +
  ggplot2::geom_smooth(method = "loess", se = FALSE,
                      colour = "black") +
  ggplot2::labs(x = "Prestige", y = "Component plus residual")
```

The refit adds a squared prestige term to *m2*. The two fits are
**nested**: they share the outcome and the 98 rows, and the smaller
fit's predictors are a subset of the larger fit's. The comparison of
named mean models applies here with continuous predictors. Its null hypothesis is that the squared-prestige
coefficient is zero, so that *m2*'s mean function suffices; the
statistic sets the residual sum of squares that the added term removes
against the larger fit's residual variance, on 1 and $98-7=91$ degrees
of freedom. A small $p$-value says the term removes more variation than
the error scale would ordinarily produce under that null. It does not
make the enlarged equation the correct mean function, and it is not a
rule for choosing a model; Chapter 9 develops the general comparison.

\Needspace{26\baselineskip}

```{r}
#| label: ch8-form-sensitivity
#| echo: true
ch8_curve <- update(m2, . ~ . + I(prestige^2))
anova(m2, ch8_curve)
c(full_education = unname(coef(m2)["education"]),
  curve_education = unname(coef(ch8_curve)["education"]))
```

The comparison gives $F(1,91)=6.736$ and $p=.011$, and education moves
from 0.1164 to $-0.0298$; in the enlarged equation the prestige slope,
$\beta_P+2\beta_{P^2}P$, depends on prestige. This is a focused
sensitivity for Pair 2, not a replacement for the prediction rule that
Pair 3 fixed in advance.

General managers have the largest Cook's distance, about 0.4467; a high
income relative to the model is not proof of an error. Refitting the
formula to the other 97 records moves
education from 0.1164 to 0.4372. We keep the full-data estimate and
disclose the sensitivity: excluding a record needs evidence about its
coding, measurement, or distinctiveness, never a more attractive
$p$-value. @tbl-ch8-checks collects what each check shows and what
follows for the answer.

```{=latex}
\Needspace{30\baselineskip}
```

| Check | What is visible here | Consequence for the answer |
|:--|:--|:--|
| Functional form | Adding a prestige-squared term changes the education slope to about $-0.0298$ | Treat the additive form as a working approximation; do not silently reinterpret it as an exact conditional law |
| Spread | HC3 raises the education SE from 0.2752 to 0.3823 | Show sensitivity; a robust SE does not repair all model problems |
| Influence | Removing the record for general managers, which has the largest Cook's distance, gives an education slope about 0.4372 | Inspect the record and disclose sensitivity; do not delete it merely to obtain a preferred result |
| Collinearity | Education's auxiliary $R^2$ is .8765 | Little education variation remains after the other predictors; separate conditional coefficients can be imprecise |
| Joint support | Professional and blue-collar education ranges do not overlap | Cross-type comparisons at common covariates may depend heavily on extrapolation |
| Dependence and scope | Occupational aggregates have shared economic and classification contexts | Independence and same-source transport are assumptions, not verified by a residual plot |

: Checks change the bounded answer rather than producing one omnibus pass/fail score. {#tbl-ch8-checks tbl-colwidths="[19,38,43]"}

Appendix B gives the formulas for Cook's distance, the HC3 covariance,
the generalized VIF, and the model-comparison criteria; the results
needed to interpret the answer are all in this section.

## Fitted and Standardized Quantities {#sec-ch8-quantities}

The fitted model also gives quantities for particular profiles: a fitted
mean with its two intervals, and an average over a named set of
profiles. These are teaching extensions of the model, not further
answers the study owes; Pair 2 remains about $\beta_E$.

### A fitted mean and its two intervals

For a complete input row, `predict()` returns two intervals:
`interval = "confidence"` concerns the model's mean at that profile, and
`interval = "prediction"` concerns one new outcome at the same profile.
Both use the full fitted model and its classical assumptions.

\Needspace{17\baselineskip}

```{r}
#| label: ch8-profile-predict-check
#| echo: true
profile <- data.frame(education = 11, prestige = 45, women = 50,
                      type = factor("wc", levels = levels(dat$type)))
predict(m2, newdata = profile, interval = "confidence")
predict(m2, newdata = profile, interval = "prediction")
```

Where do these intervals come from? Use the Section 2.3 profile vector
$\mathbf x_*=(1,11,45,50,0,1)^{\mathsf T}$. With
$\widehat{\mathbf V}$, the coefficient
covariance of Section 3.2, the fitted mean and its estimated classical
variance are

$$
\widehat m_*=\mathbf x_*^{\mathsf T}\widehat{\boldsymbol\beta},\qquad
\widehat{\operatorname{Var}}(\widehat m_*)=\mathbf x_*^{\mathsf T}\widehat{\mathbf V}\mathbf x_*,
\quad \widehat{\mathbf V}=\operatorname{vcov}(m2).
$$

The variance includes the coefficient variances **and their
covariances**: two terms $a\widehat\beta_j+b\widehat\beta_k$ contribute
$a^2\widehat{\operatorname{Var}}(\widehat\beta_j)+
b^2\widehat{\operatorname{Var}}(\widehat\beta_k)+
2ab\widehat{\operatorname{Cov}}(\widehat\beta_j,\widehat\beta_k)$.
Adding coefficient SEs, or ignoring the covariances, is a different
calculation, and the square root comes only after the variance is
complete. Here the fitted mean is 5.5886, its variance 0.308, and its SE
0.5549 thousand dollars, giving the fitted-mean interval
$[4.4865,6.6908]$ that `predict()` printed.

```{r}
#| label: ch8-profile-se-demo
#| echo: true
#| eval: false
sqrt(drop(row %*% vcov(m2) %*% row))
```

**One new outcome** adds its own error variance, $s^2=6.8575$, so its SE
is $\sqrt{s^2+\mathbf x_*^{\mathsf T}\widehat{\mathbf V}\mathbf x_*}\approx2.6768$
and its interval $[0.2722,10.9051]$. @fig-ch8-bands draws both intervals
across the white-collar education range at prestige 45 and women 50. The
dark fitted-mean band is narrowest near education 11.2, about 2.2
thousand dollars wide, and widens toward the edges of the records, to
about 3.1 at 9.17 years. The light band for one new occupation stays
between about 10.6 and 10.9 wide along the whole range, because the new
occupation's error variance, 6.8575, dominates a fitted-mean variance of
about 0.3.

```{r}
#| label: fig-ch8-bands
#| fig-cap: "Two 95% intervals for a white-collar profile at prestige 45 and women 50, across the white-collar education range."
#| fig-alt: "A nearly flat fitted line runs from about 5.4 to 5.8 thousand dollars as education rises from 9.17 to 12.79 years. A narrow dark band around it, the fitted-mean interval, is narrowest near 11 years and widens toward both ends. A much wider light band, the interval for one new occupation, runs from about 0 to 11 thousand dollars along the whole range. A dotted vertical line marks education 11."
#| fig-width: 6.1
#| fig-height: 3.3
wc_range <- range(dat$education[dat$type == "wc"])
band <- data.frame(education = seq(wc_range[1], wc_range[2],
                                   length.out = 60),
                   prestige = 45, women = 50,
                   type = factor("wc", levels = levels(dat$type)))
mean_band <- predict(m2, newdata = band, interval = "confidence")
new_band <- predict(m2, newdata = band, interval = "prediction")
band <- cbind(band, fit = mean_band[, "fit"],
              mean_lwr = mean_band[, "lwr"], mean_upr = mean_band[, "upr"],
              new_lwr = new_band[, "lwr"], new_upr = new_band[, "upr"])
ggplot2::ggplot(band, ggplot2::aes(education)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = new_lwr, ymax = new_upr),
                       fill = "grey88") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = mean_lwr, ymax = mean_upr),
                       fill = "#9DB9CC") +
  ggplot2::geom_line(ggplot2::aes(y = fit), linewidth = .8) +
  ggplot2::geom_vline(xintercept = 11, linetype = 3) +
  ggplot2::annotate("text", x = 12.7, y = c(10.2, 6.75), hjust = 1,
                    size = 3.1,
                    label = c("One new occupation", "Fitted mean")) +
  ggplot2::labs(x = "Education (years), over the white-collar range",
                y = "Income (thousand 1971 dollars)")
```

Both intervals rest on the independent, Normal, common-variance model,
which the checks of Section 4 qualify rather than confirm; replacing
$\widehat{\mathbf V}$ with HC3 alone does not estimate a new outcome's
conditional error distribution. Neither interval says how accurately a
learning rule predicts held-out records; that is Pair 3's question.

### Standardizing over the observed profiles

A standardized fitted value averages over a named set of profiles rather
than using one. Copy the 98 observed profiles, change every type to
white-collar, predict each fitted value, and average them with equal
weights.

\Needspace{11\baselineskip}

```{r}
#| label: ch8-standardization-first
#| echo: true
white_profiles <- dat
white_profiles$type <- factor("wc", levels = levels(dat$type))
mean(predict(m2, newdata = white_profiles))
```

The result, 7.0164, is an average of model predictions after a specified
change of profile, not the observed white-collar group mean. Every one
of the 98 profiles receives weight $1/98$. For its uncertainty, use
Section 5.1's coefficient-covariance formula with the average of these
98 design rows replacing the single profile row $\mathbf x_*$. This
gives SE 0.6177 and classical interval $[5.7896,8.2432]$, treating the
profiles and weights as fixed. The predictions share fitted coefficients;
averaging their interval endpoints would ignore that joint uncertainty.
Repeating the operation for each type gives @tbl-ch8-standardized,
which sets the results beside the observed
mean income of each type.

\Needspace{9\baselineskip}

| Type | Observed mean | Standardized estimate | SE | Lower | Upper |
|:--|--:|--:|--:|--:|--:|
| bc | 5.3741 | 6.7811 | 0.6942 | 5.4023 | 8.1598 |
| prof | 10.5595 | 7.1053 | 0.9011 | 5.3156 | 8.8950 |
| wc | 5.0523 | 7.0164 | 0.6177 | 5.7896 | 8.2432 |

: Observed type means and empirical-profile standardization under the full additive model. {#tbl-ch8-standardized tbl-colwidths="[12,18,22,14,17,17]"}

The observed means differ sharply: professional occupations average
10.5595 thousand dollars, about twice the other two types, but they also
have more education, higher prestige, and different percentages of
women. The standardized estimates give every type the same 98 profiles,
so they differ only by the type coefficients, 0.3242 for professional
and 0.2353 for white-collar relative to blue-collar.

Because *m2* is linear and additive in the continuous predictors, its
average fitted value equals its fitted value at their means (education
averages 10.7951 years); with squared terms, interactions, or a
nonlinear response function, the prediction at the means need not equal
the mean prediction. Any other fixed weighting used for all three
types, whether another named set of profiles or unequal weights on these
98, would move the three standardized levels but leave their differences
at the type coefficients.

Relabelling every profile professional or white-collar creates some
combinations the records do not support, and averaging does not remove
that problem; the white-collar profile of Section 5.1 is a different,
locally supported quantity. Standardization alone neither identifies a
causal effect nor validates out-of-sample performance.

\Needspace{12\baselineskip}

## The Held-Out Prediction Question {#sec-ch8-prediction}

### Learning rules and the loss

A learning rule says how training records become a prediction equation.
Our two rules are fixed: education-only OLS, and the full additive OLS
equation with education, prestige, women, and type. OLS **fits** by
minimizing squared training errors; we separately choose the **mean
absolute error (MAE)** to evaluate the predictions in income units.

A prediction opportunity here is a hypothetical same-source occupation
whose predictors are all available before its income is used, not a
forecast from information that arrives later. The records cannot show
that this exchangeability holds for other countries, periods, individual
workers, or types absent from the dictionary.

For held-out outcomes $y_i$ and predictions $\widehat y_i^{(-)}$ fitted
without those outcomes,

$$
MAE=\frac{1}{n_{\mathrm{test}}}
       \sum_{i\in\mathrm{test}}\left|y_i-\widehat y_i^{(-)}\right|.
$$

A smaller MAE means a smaller average absolute error on the evaluated
cases, not a better prediction for every case.

### Shared ten-fold validation

In shared ten-fold cross-validation the 98 records are divided into ten
folds, and each fold takes a turn as the held-out set: both rules are
fitted to the other records and predict the held-out cases
(@fig-ch8-folds), and the 98 case-level errors are then pooled.
@tbl-ch8-validation records the protocol.

```{r}
#| label: fig-ch8-folds
#| fig-cap: "Shared held-out cases let two fixed learning rules face the same prediction task."
#| fig-alt: "A training-record box branches to education-only OLS and full OLS. Both arrows lead to predictions for the same held-out records, then to paired absolute errors. The diagram is repeated over ten folds."
#| fig-width: 6.3
#| fig-height: 3.3
par(mar = c(.2, .2, .2, .2))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 5))
boxtext <- function(x1, y1, x2, y2, label, fill = "#EFF4F7") {
  rect(x1, y1, x2, y2, col = fill, border = "#426278")
  text((x1 + x2)/2, (y1 + y2)/2, label, cex = .9)
}
boxtext(.1, 1.8, 2.3, 3.2, "Training records\n88 or 89")
boxtext(3.1, 3.1, 6.2, 4.4, "Education-only OLS")
boxtext(3.1, .6, 6.2, 1.9, "Full additive OLS")
boxtext(7, 1.8, 9.9, 3.2, "Same held-out cases\nPaired absolute errors")
arrows(2.3, 2.8, 3.1, 3.6, length = .08)
arrows(2.3, 2.2, 3.1, 1.4, length = .08)
arrows(6.2, 3.6, 7, 2.8, length = .08)
arrows(6.2, 1.4, 7, 2.2, length = .08)
text(5, .12, "Repeat across ten folds; pool the 98 case-level errors.", cex = .9)
```

\Needspace{17\baselineskip}

| Decision | Frozen teaching protocol | Why it matters |
|:--|:--|:--|
| Candidates | Education-only and the declared full additive OLS rule | No search for the best-looking model after validation |
| Folds | Shared balanced random ten-fold assignment, seed 80806 | Both rules face the same held-out cases |
| Summary | Pool absolute errors over all 98 held-out cases | Each case receives equal weight; folds of different size do not |
| Sensitivity | 200 new resplits of the same records, seed 80807 | Describes sensitivity to splitting, not new independent studies |

: The internal-validation protocol was fixed before this teaching comparison. {#tbl-ch8-validation tbl-colwidths="[20,42,38]"}

This is a retrospective reanalysis of familiar data, so "fixed before
this comparison" is not a claim of historical preregistration. It means
that the candidates were not revised after the primary errors were seen.
The sensitivity refits of Section 4 are not added to the candidate set.
The 98 complete rows, unit conversion and type dictionary stay fixed.
Every training set must contain all types and have full column rank.
One fold illustrates the protocol: fold 9 holds the first source record,
government administrators, with eight other occupations. `train` contains
the other 89 records and `held` these nine cases. Both rules use the
same training records and predict the same held-out records.

\Needspace{10\baselineskip}

```{r}
#| label: ch8-fold-prepare
#| include: false
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
set.seed(80806)
fold <- sample(rep(seq_len(10), length.out = nrow(dat)))
held_fold <- fold[1]
train <- dat[fold != held_fold, ]
held <- dat[fold == held_fold, ]
c(fold = held_fold, training = nrow(train), held_out = nrow(held))
fold_simple <- lm(income_k ~ education, data = train)
fold_full <- lm(income_k ~ education + prestige + women + type,
                data = train)
first_case <- dat[1, ]
first_predictions <- c(
  Education = unname(predict(fold_simple, first_case)),
  Full = unname(predict(fold_full, first_case)))
first_predictions
abs(first_predictions - first_case$income_k)
```

```{r}
#| label: ch8-one-fold
#| echo: true
#| eval: false
fold_simple <- lm(income_k ~ education, data = train)
fold_full <- lm(income_k ~ education + prestige + women + type,
                data = train)
pred <- predict(fold_full, newdata = held)
```

Both fits exclude government administrators' outcome. Every training
set in this comparison contains all three types and has full column
rank, so both specified rules are estimable. Chapter 13 develops the
validation protocol further.

No held-out outcome is used to estimate a coefficient. Learned
preprocessing, such as scaling, imputation, or variable selection, would
also have to be fitted on training records only; a fixed unit conversion
such as dollars to thousands is not learned.

### Held-out errors and resplits

For government administrators, observed income is 12.351 thousand
dollars. The full-rule prediction is 11.5948, giving

$$
|12.351-11.5948|=0.7562,
$$

and the education-only prediction is 8.9461, giving
$|12.351-8.9461|=3.4049$.
This record favors the full rule, but one record is not the comparison;
@tbl-ch8-mae pools all 98 errors.

For the nine cases in the illustrated fold, the MAE formula becomes
one average of absolute prediction errors:

```{r}
#| label: ch8-mae-demo
#| echo: true
#| eval: false
mean(abs(held$income_k - pred))
```

The table uses the errors from **all ten folds**, so every one of the
98 records contributes once.

| Learning rule | Held-out MAE (thousand dollars) |
|:--|--:|
| Education-only OLS | 2.4031 |
| Full additive OLS | 1.5412 |
| Difference: full minus education-only | $-0.8619$ |

: Case-weighted absolute errors from the frozen shared-fold comparison. {#tbl-ch8-mae tbl-colwidths="[60,40]"}

The full rule's average held-out absolute error is about 862 dollars
lower. Because the two errors are subtracted for **each same case**
before averaging, differences in difficulty between cases cannot favor
either rule through different test sets. The same predictions give the
root mean squared error, which weights large errors more heavily: 3.5051
for education-only and 2.7111 for the full rule, the same preference
here, although the two losses need not agree in other data.

Eight fits use 88 training records and two use 89; the pooled errors
describe the learning rules at those training sizes, not the one
equation fitted to all 98. Chapter 13, Section 2, develops the full
protocol and the training-size mixture.

The 200 resplits give differences from $-0.914$ to $-0.741$, all
favoring the full rule; this is sensitivity to splitting the same
records, not a confidence interval from independent studies.
New-sample uncertainty is not quantified, so the advantage remains
provisional.

The comparison supports lower internal held-out error for the full rule
in these records under this protocol. It does not isolate the predictive
contribution of education, because the rules differ by prestige, women,
and type together, and it does not establish external transport,
individual earnings prediction, or a causal education effect.

\Needspace{14\baselineskip}

## Reading and Evaluating the Three Answers {#sec-ch8-coordinate}

### The study report

```{=latex}
\Needspace{20\baselineskip}
```

| Pair and role | Question | Estimand and owed answer |
|:--|:--|:--|
| **1. Contextual description** | How does income vary linearly with education across the 98 retained occupations? | The finite-record marginal least-squares slope; a descriptive answer for these records |
| **2. Primary conditional association** | How does income differ with education at the same prestige, percentage of women, and type, within support? | The linear conditional-mean coefficient $\beta_E$ under the declared working model; fitted $b_E$ is its estimate from the 98 retained records |
| **3. Secondary prediction analysis** | Does the full additive learning rule predict eligible same-source occupation records with lower expected absolute error than the education-only rule? | Difference in expected absolute error of the two learning procedures at the declared training sizes; an internal validation answer |

: Three separately owed answers in the occupation teaching study. {#tbl-ch8-pairs tbl-colwidths="[22,36,42]"}

Each row is a question--answer pair with **one argument with linked
layers**. The three are not statistical strands of one argument: the
report owes separate conclusions about marginal association, conditional
association, and prediction. Estimands become strands only when one
integrated conclusion needs them all, and printing several coefficients
creates neither extra questions nor extra strands.

The report below gathers the analyses of Sections 1--6. It is our
teaching record, not a statement by the source investigators.

> **R0 -- Common record and declared scope.** We analyzed the 98 complete
> occupation records described in @sec-ch8-questions. Income was
> scaled to thousand
> 1971 Canadian dollars. Education and income were historical occupational
> averages; prestige came from an earlier rating survey. We retained the
> same rows for all three questions and make no claim of verified probability
> sampling or randomized education.
>
> **R1 -- Marginal answer.** Across these records the education-only
> least-squares slope was 0.883 thousand dollars per year. This is a
> descriptive linear summary: occupations with more education tended to
> have higher average income. It is not an individual or causal effect.
>
> **R2 -- Conditional answer.** The intended estimand was the linear
> conditional-mean education coefficient $\beta_E$, comparing education at the
> same prestige, percentage of women, and occupational type. Its fitted
> estimate was $b_E=0.1164$, with a classical 95% interval [$-0.4302$, 0.6631]
> under the independent linear conditional-mean working model with common
> variance. HC3 gave [$-0.6429$, 0.8758]. Influence, functional-form, and support
> checks limit that model's interpretation. The direction and practically
> relevant magnitude remain unresolved under these calculations; the result
> does not establish no conditional relationship.
>
> **R3 -- Prediction answer.** The two fixed OLS learning rules were compared
> with the shared ten-fold protocol in @sec-ch8-prediction. The full
> rule had MAE
> 1.5412 versus 2.4031 for education-only, difference $-0.8619$. The examined
> resplits all favored the full rule. This is internal evidence about
> learning procedures trained on 88/89 eligible records under the stated
> same-source exchangeability assumptions. Lower expected error is a
> provisional conclusion because new-sample uncertainty is not quantified;
> this is not external validation of a final 98-record equation.
>
> **R4 -- Synthesis.** Education has a positive marginal linear association
> with income in these historical records. Its conditional association is
> imprecisely resolved by the specified additive model, while the full
> learning rule predicts the held-out records more accurately than the
> education-only rule. These conclusions answer different questions and
> do not establish that changing education would change income.

### Functions and traces in the report

Reading starts by locating what each passage does. In Chapter 1's
terms, a component is a logical function, such as STAT-E or SUB-J, and
the report's study-specific content serves it. One passage can serve
several functions. R0, for example, identifies the records for STAT-E
and supplies their historical context for SUB-E. It would become
justification (J) content only if it offered a reason connecting those
records to an inference; "we used R" offers none. @tbl-ch8-traces lists
each function once, with the content shared by the three pairs and the
content specific to each.

| Function | Shared or pair-specific content | Trace in this teaching record |
|:--|:--|:--|
| **STAT-E** | Named 98 rows, variables, exclusions, and computations' inputs | R0; the row audit and import code in @sec-ch8-questions |
| **SUB-E** | Occupations rather than workers; historical dollars, education, ratings, and types | R0; variable/source descriptions |
| **STAT-J** | Pair 1: fixed-record least-squares definition. Pair 2: declared stochastic model and uncertainty conditions. Pair 3: held-out fitting, fixed rules, loss, and exchangeability | R1--R3; model declaration; validation protocol and code |
| **SUB-J** | Reasons connecting those measurements and comparisons to historical occupational income, with limits on individual, current, and causal interpretations | R0--R3; unit/time explanations and prediction-opportunity definition |
| **STAT-I** | Pair 1's 0.8830 slope; Pair 2's 0.1164 estimate, intervals, coefficient test and checks; Pair 3's case-level errors, MAEs, and split sensitivity | @sec-ch8-estimate through @sec-ch8-prediction; R1--R3 |
| **STAT-C** | Descriptive positive marginal slope; unresolved conditional slope under the model; smaller internal error for the full rule under the protocol | R1, R2, and R3 respectively |
| **SUB-I** | Interpret years, income units, conditioning, and prediction errors in the named historical occupation setting | Worked substitutions and error calculation; R1--R3 |
| **SUB-C** | Each question-facing answer with its own limits; none supplies a causal or individual-earnings answer | R1--R3; their distinctions preserved in R4 |

: Reading the linked-layer functions without three duplicate maps. {#tbl-ch8-traces tbl-colwidths="[15,47,38]"}

SUB-E situates the evidence, SUB-J **augments and contextualizes** the
statistical justification, SUB-I interprets the inference, and SUB-C
bounds the substantive answer. Sharing a dataset does not make Pair 2's
independent-error model into Pair 3's validation procedure.

For each supporting statement, record what the report says and where
the reader finds it: in the report, the data documentation, or the
executable calculations. Gaps and conflicts are stated in ordinary
prose. R2 states the independence assumption, for example, but nothing in
the record shows that independence holds under the original selection of
occupations; a stated assumption and evidence for it are different
information. The unit conversion, by contrast, makes reading 0.883
thousand dollars as about 883 dollars traceable. This record contains no
conflicting source, though a real study may contain incompatible
descriptions to record and investigate. A reason's documentary status is
simply what has been stated or located and which support is missing; it
is not a quality score or a set of categories to learn.

Professional knowledge helps identify a trace's function but does not
license adding an unreported design feature. An invalid reason such as
"all rows are in the same dataset, therefore they are independent" is
still **stated J**, whose adequacy is judged separately: reading records
the reason actually offered, not a repaired one.

### Evaluating each answer

Evaluation now judges the quality of the reasoning just identified.
Chapter 1's six questions apply to each argument (@tbl-ch8-quality); they
are not six scores to total.

| Quality-evaluation question | Application to these arguments |
|:--|:--|
| **Question alignment** | R1 matches its finite-record slope. R2 names $\beta_E$, its fitted estimate, the conditioning set, and the model extension. R3 names the learning procedure, loss, and training sizes rather than substituting a coefficient test. |
| **Evidence integrity** | The 102-to-98 accounting and common rows are explicit. Missing type still limits what is represented. CV excludes held-out outcomes from fitting; source selection is not repaired by that procedure. |
| **Justification adequacy** | Least squares is sufficient for the declared finite description. Pair 2's model conditions and Pair 3's same-source exchangeability remain assumptions with incomplete empirical support. Statistical computation alone cannot establish them. |
| **Inferential adequacy** | The slope, intervals, error arithmetic, and comparisons are reproducible. Neither zero-crossing nor a large p-value proves absence; repeated folds are not independent replication. Substantive reasoning must retain those distinctions. |
| **Transition fidelity** | Occupation-level average income remains occupation-level average income; partial adjustment is not a causal intervention, and model fit is not held-out performance. |
| **Claim calibration** | R1 remains finite and historical; R2 reports unresolved magnitude/direction rather than no relationship; R3 remains internal and procedure-specific. No claim extends to individual workers or modern populations. |

: Apply the six questions before evaluating the study-level synthesis. {#tbl-ch8-quality tbl-colwidths="[28,72]"}

Chapter 1 distinguishes judgments about a reason from the conclusion
about a whole question--answer pair. Here the finite description is
**supported as stated**; the explicitly unresolved conditional answer is
also **supported as stated**, under its disclosed model limits. A
correctly reported inconclusive answer need not establish a stronger
claim to be supported. The prediction comparison is supported as a
qualified internal result, while external superiority of the final
all-record equation is **not assessable from the accessible record**.
Argument Chapter B, Section 4, develops the full judgment account and
its alternatives.

### Is the synthesis coherent?

Coordinated questions add one evaluation question, **Cross-question
coherence**: does the synthesis preserve the distinct questions,
estimands, methods, analytic statuses, uncertainty, and limits of the
separate answers? We introduce it here; Argument Chapter B develops it
further but is not needed to apply it.

R4 reports a positive marginal slope, an unresolved conditional slope,
and better internal prediction. **Predict and explain:** can an interval
that includes zero for the education coefficient coexist with a lower
held-out MAE for the full rule? Identify the question each result
answers before reading this alternative synthesis:

> Once the other predictors were controlled, education had no effect;
> nevertheless, the significant full regression proves it will predict
> workers' incomes accurately.

The results can coexist: the interval concerns one conditional
coefficient, whereas the rules differ by prestige, women and type
together. Better predictions from the full rule do not isolate
education's contribution. R4 retains these distinct answers.

Reading first identifies the inference and claim the alternative offers. Evaluation
then finds three failures: "no effect" substitutes a causal null for the
conditional-association answer, an omnibus significance test is not
held-out validation, and occupations have become workers. The corrected
synthesis is R4, not a fourth overarching argument that erases the
separate conclusions.

These records were collected for earlier purposes, not prospectively
for our three teaching questions. Chapter 1's repair-versus-new-cycle
rule applies: correcting an error under the same target is a repair;
choosing a causal target after seeing an association is a revised or
new cycle whose timing must be stated. Argument Chapter B, Section 5,
develops the full coordinated construction workflow.

## Reading Bridge {#sec-ch8-summary}

Start with a model naming its outcome, predictors, coefficients and
error variation. The known process showed partial slopes at fixed other
predictors before the occupation records supplied estimates. Multiple
regression adds conditional comparisons without conferring automatic
causal control. Subtracting fitted means showed what a coefficient compares;
residualizing connected it to covariance; residual, influence, support,
and precision checks each bounded the answer differently; a fitted
profile and a standardized average extended the conditional model; and
held-out errors answered a separately owed question. The occupation study
owed three answers, a finite-record marginal description, a working-model
conditional interval with unresolved magnitude, and a qualified internal
prediction comparison, and its shared records made the questions
coordinated without making their justifications identical. Chapter 9 develops extensions of this model (factors with several levels and their
contrasts, nested model comparison and information criteria, changed
predictor and outcome forms, and robust, clustered, and bootstrap
uncertainty) and transfers the reading to Project STAR, where a
randomized design supplies a different warrant. Chapter 10 then extends
the interaction reasoning of Chapters 4 and 6 to regression.

\Needspace{22\baselineskip}

### Key terms {.unnumbered}

Use this short reference to revisit the concepts introduced here. The
Part One Concept Reference gives the broader reference and the principal home
of each concept; the simple-regression, interval, ANOVA, and reading
vocabulary used in this chapter is explained in Chapters 1--4 and 7.

| Term and common alternatives | Meaning |
|:--|:--|
| Multiple regression (multiple regression model; additive model) | A model in which an outcome's mean is an intercept plus one term per predictor, $E(Y\mid x_1,\ldots,x_{p-1})=\beta_0+\beta_1x_1+\cdots+\beta_{p-1}x_{p-1}$, with errors of one common SD $\sigma$; its $p$ coefficients leave $n-p$ residual df, and least squares estimates them. Each coefficient compares means at the same values of the other predictors; the additive form assumes each comparison is the same across those values. |
| Partial slope (conditional coefficient; adjusted coefficient) | The difference in fitted outcome per one-unit difference in one predictor, holding the model's other predictors fixed. Its meaning depends on the conditioning set named with it; it is not an intervention effect or a claim about unmeasured differences. |
| Indicator variable (dummy variable) and reference category | A 0/1 variable marking membership in one category of a factor; a factor with $J$ levels needs $J-1$ indicators, and the omitted level is the reference. The intercept refers to the reference category; each indicator coefficient is a difference from it at the same other predictors. Releveling changes coordinates, not fitted values. |
| Partial regression (residualization; added-variable form) | Regressing both the outcome and one predictor on the other predictors, then relating their residuals. The residual slope equals the partial slope; the residual correlation is the partial correlation, unit-free. |
| Coefficient covariance matrix | The estimated variances and covariances of the fitted coefficients (`vcov()` in R). Any linear combination of coefficients, such as a fitted mean or a contrast, has variance $\mathbf c^{\mathsf T}\widehat{\mathbf V}\mathbf c$; adding SEs or ignoring covariances is a different, incorrect calculation. |
| Variance inflation factor (VIF) | One divided by one minus the $R^2$ from regressing a predictor on the other predictors. Its square root multiplies that coefficient's SE relative to an orthogonal design, holding residual variance fixed; it describes precision, not bias, and is not a deletion rule. |
| Joint support (covariate hull) | Whether a combination of predictor values lies within the region where observations actually occur, not merely within each variable's separate range. A fitted value outside joint support is extrapolation even if each coordinate is observed somewhere. |
| Standardization over a named distribution (model-averaged fitted value) | Averaging the model's fitted values after setting one predictor to a chosen value for every profile in a declared set with fixed weights. Its uncertainty uses the coefficient covariance; it is not an observed group mean, a causal effect, or held-out performance. |
| Learning rule (learning procedure) | A fixed recipe that turns training records into a prediction equation, such as fitting a declared OLS formula. Prediction performance concerns the rule at stated training sizes, not one final fitted equation. |
| Cross-validation (held-out validation) | Dividing records into folds, fitting a rule on the other folds, predicting the held-out fold, and pooling the errors so that no outcome contributes to its own prediction. Resplits describe sensitivity to the partition, not new independent studies. |
| Mean absolute error (MAE) | The average absolute difference between held-out outcomes and their predictions, in outcome units. It is an evaluation loss chosen separately from the squared-error criterion that OLS minimizes. |
| Cross-question coherence | The additional evaluation question for coordinated answers: does the synthesis preserve the distinct questions, estimands, methods, uncertainty, and limits of the separate answers rather than merging them into one omnibus claim? |

: Selected concepts for multiple regression across coordinated questions. {#tbl-ch08-key-terms tbl-colwidths="[28,72]"}


### Further reading {.unnumbered}

- The [OpenIntro multiple-predictor tutorial](https://openintro-ims.netlify.app/model-mlr)
  offers concrete equations and indicator interpretations; this chapter
  borrows its equation-first device but not its stepwise selection.
- The [ISLR regression lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch3-linreg-lab.html)
  connects fitting, coefficients, checks, and predictions, and the
  [ISLR cross-validation lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch5-resample-lab.html)
  develops the train, predict, and check sequence [@james2021].
- The [carData documentation](https://cran.r-project.org/web/packages/carData/carData.pdf)
  [@carData2022; @fox2019] supplies the Prestige variable meanings and
  provenance; the three questions and the R0--R4 report are this
  chapter's own.
- Fox [@fox2016] and Kutner and colleagues [@kutner2005] develop the
  linear-model algebra behind partial slopes, indicators, and
  diagnostics, and Argument Chapter B develops coordination and its
  auxiliary construction workflow in greater detail.

## Exercises {.unnumbered}

Use the supplied numbers; rounding within 0.02 thousand dollars is
acceptable unless a different precision is requested. These prompts do not
require Argument Chapter B or the appendices.

\Needspace{18\baselineskip}

<!-- prompt-id: CH08-E01 -->
### Exercise 1: Keep the three answers distinct (Foundational) {.unnumbered #ex-ch08-01}

A report promises: (a) a description of education and income across 98
occupations, (b) an education comparison at the same prestige, women, and
type, and (c) a comparison of two prediction rules on held-out records.

1. Explain why these are three question--answer pairs rather than three
   statistical strands serving one integrated answer.
2. For each, name its estimand in plain language and one kind of output
   that could contribute to its answer.
3. Does sharing the same 98 records establish the same justification for
   all three? Give one concrete reason.

\Needspace{18\baselineskip}

<!-- prompt-id: CH08-E02 -->
### Exercise 2: Name what changes (Foundational) {.unnumbered #ex-ch08-02}

Classify each change as primarily changing the records, the comparison,
the coefficient coordinates, or the uncertainty calculation. State whether
it necessarily creates a new scientific question.

1. Refit education-only regression after excluding four missing-type records.
2. Add prestige, women, and type to the education-only association equation.
3. Change the reference category from blue-collar to white-collar.
4. Keep the fitted equation but replace classical SEs with HC3 SEs.

Finally, explain why appropriately specified adjustment in a randomized
study need not change its declared average assignment-effect estimand.

<!-- prompt-id: CH08-E03 -->
### Exercise 3: Interpret a fitted comparison (Foundational) {.unnumbered #ex-ch08-03}

Use this rounded hypothetical equation, with income in thousand dollars:

$$
\widehat y=1+0.20e+0.10p-0.04w+0.30D_{\mathrm{prof}}
           +0.15D_{\mathrm{wc}}.
$$

1. Two white-collar profiles share prestige $p=40$ and women $w=50$;
   one has education $e=12$ and the other $e=10$. Which terms cancel
   when their fitted means are subtracted? Give the modeled difference
   and interpret it in dollars.
2. What does the coefficient 0.15 compare, and at which values of the
   other predictors?
3. What must be checked before calling the education comparison in
   part 1 supported by the observed data?
4. Why is the modeled difference not an estimate of the effect of
   giving a worker two more years of education?

\Needspace{18\baselineskip}

<!-- prompt-id: CH08-E04 -->
### Exercise 4: Read held-out errors (Foundational) {.unnumbered #ex-ch08-04}

Four held-out occupations have the observed incomes and predictions
below, in thousand dollars. Each prediction was fitted without the
corresponding outcome.

| Occupation | Observed | Education-only prediction | Full-rule prediction |
|:--|--:|--:|--:|
| 1 | 5 | 4 | 6 |
| 2 | 8 | 10 | 8 |
| 3 | 10 | 7 | 9 |
| 4 | 13 | 12 | 11 |

1. Calculate the MAE of each rule.
2. Explain why the average of the four paired differences, full minus
   education-only, equals the difference between the two MAEs, and why
   it matters that both rules predict the same four occupations.
3. Does the better average rule have to be better for every
   occupation? Point to the table.
4. Explain why this comparison is neither a coefficient interval nor
   validation of a model subsequently fitted to all available records.

\Needspace{22\baselineskip}

<!-- prompt-id: CH08-E05 -->
### Exercise 5: Locate a reason before judging it (Integrative) {.unnumbered #ex-ch08-05}

An accessible report states:

> We retained the same 98 occupations for both equations. This makes the
> observations independent. The full equation has higher training
> $R^2$, so we conclude it predicts new occupations more accurately.

The report provides import code and fitted-model output, but no validation
split, held-out predictions, or study-specific evidence about dependence.

1. Identify the reasons offered, the inference, and the claim.
2. Record the documentary status of the independence reason and the
   held-out prediction evidence separately. Do not rewrite “stated” as
   “not located” merely because a reason is wrong.
3. Apply Justification adequacy, Inferential adequacy, and Claim calibration.
4. Give a whole-pair evaluation conclusion for the prediction answer and
   name the additional work it would require.

\Needspace{25\baselineskip}

<!-- prompt-id: CH08-E06 -->
### Exercise 6: Evaluate the study-level synthesis (Integrative) {.unnumbered #ex-ch08-06}

Use the occupation study's reported results: marginal slope 0.883;
fitted conditional estimate $b_E=0.1164$ of $\beta_E$, with classical
interval [$-0.4302$, 0.6631];
and held-out MAEs 2.4031 for education-only versus 1.5412 for the full rule.

A summary says:

> Education does not matter after adjustment, but the model proves that
> increasing workers' education raises income and predicts earnings well
> in other countries.

1. Predict whether the conditional interval's including zero conflicts
   with the full rule's lower held-out MAE. Explain which question each
   result answers, then identify the claims actually offered in the
   summary before evaluating them.
2. Identify which separate answer obligation is being replaced or exceeded
   by each clause.
3. Apply Cross-question coherence and write a corrected three-sentence
   synthesis that retains the three reported answers and their limits.
4. If the authors now want a causal effect of education, is that a repair
   of their existing uncertainty calculation or a new/revised question-level
   cycle? Explain.

## Appendix A: Omitted Variables and Group Means {.unnumbered #app-ch8-a}

This appendix develops the mathematics behind @sec-ch8-comparisons. Its
occupation calculations use the same 98-row object *dat*, the models *m0*,
*m1*, and *m2* of the main text, and the type codes *bc*, *prof*, and
*wc*; they introduce no other sample. Chapter 9 develops contrasts, block
tests, adjusted means, the matrix form, and the Project STAR transfer.

### Why the marginal slope includes another association {.unnumbered}

Suppose a population process is

$$
Y=\beta_0+\beta_X X+\beta_Z Z+\varepsilon,
\qquad E(\varepsilon\mid X,Z)=0,
$$

with finite variances and $\operatorname{Var}(X)>0$. The marginal
least-squares slope is a covariance divided by a variance. Substituting
the outcome equation gives

$$
\begin{aligned}
\frac{\operatorname{Cov}(X,Y)}{\operatorname{Var}(X)}
&=\frac{\beta_X\operatorname{Var}(X)
+\beta_Z\operatorname{Cov}(X,Z)
+\operatorname{Cov}(X,\varepsilon)}{\operatorname{Var}(X)}\\
&=\beta_X+\beta_Z
\frac{\operatorname{Cov}(X,Z)}{\operatorname{Var}(X)}.
\end{aligned}
$$ {#eq-ch8a-omission}

The zero-error-covariance condition removes the final term. The sample
counterpart holds exactly for least-squares fits on the same rows, as
Section 2.2's prestige and women calculations show. The
discrepancy therefore involves both the association between $X$ and $Z$
and $Z$'s coefficient in the outcome model; neither ingredient alone
determines its size. If the target is the conditional coefficient,
omitting $Z$ creates bias relative to that target. If the target is the
marginal slope, that marginal quantity is not a biased version of an
unstated conditional target. The two coefficients name different
comparisons, and neither supplies a causal warrant for observational
variables.

### Repeated samples from the known process {.unnumbered}

Let independent standard Normal variables $X$ and $U$ generate
$Z=.65X+\sqrt{1-.65^2}U$, so that both predictors have variance one and
correlation .65. Independently generate a Normal error with SD 6, and set

$$
Y=50+2X+5Z+\varepsilon.
$$ {#eq-ch8a-sim-process}

This is the process of @sec-ch8-comparisons: the marginal slope is
$2+5(.65)=5.25$ and the conditional $X$ slope is 2. The sample in
@fig-ch8-known-process has 250 observations and uses seed 82108. Its
marginal slope is 5.9971 and its conditional slope 3.0259: within two SEs of their
targets, as Section 1.3 reads from the generated sample.

A separate repeated-study design, with its own stated sample size, shows
the sampling variation: 1,000 samples of 150 observations each, with seed
82109. Each sample fits the marginal and conditional equations to new
observations from the same process. The density curves summarize their
estimated slopes.

```{r}
#| label: ch8a-repetition
#| echo: false
#| results: hide
set.seed(82109)
ch8a_one_rep <- function() {
  x <- rnorm(150)
  z <- .65 * x + sqrt(1 - .65^2) * rnorm(150)
  y <- 50 + 2 * x + 5 * z + rnorm(150, sd = 6)
  c(Marginal = unname(coef(lm(y ~ x))["x"]),
    Conditional = unname(coef(lm(y ~ x + z))["x"]))
}
ch8a_slopes <- t(replicate(1000, ch8a_one_rep()))
```

```{r}
#| label: tbl-ch8a-repeated
#| echo: false
#| tbl-cap: "Targets and results across 1,000 generated samples."
knitr::kable(data.frame(
  Model = colnames(ch8a_slopes),
  Target = format(round(c(5.25, 2), 4), digits = 15, trim = TRUE),
  Mean = format(round(colMeans(ch8a_slopes), 4), digits = 15, trim = TRUE),
  SD = format(round(apply(ch8a_slopes, 2, sd), 4), digits = 15,
              trim = TRUE), row.names = NULL
), align = c("l", "r", "r", "r"))
```

```{r}
#| label: fig-ch8a-repeated
#| echo: false
#| fig-cap: "Sampling distributions around two different targets."
#| fig-alt: >
#|   Separate panels show marginal slopes centered near 5.25 and
#|   conditional slopes centered near 2. Dashed lines mark their
#|   respective population targets.
#| fig-width: 6.1
#| fig-height: 3.8
ch8a_long <- data.frame(
  Slope = c(ch8a_slopes[, 1], ch8a_slopes[, 2]),
  Model = rep(c("Marginal", "Conditional"), each = 1000)
)
ch8a_targets <- data.frame(
  Model = c("Marginal", "Conditional"), Target = c(5.25, 2)
)
ggplot2::ggplot(ch8a_long, ggplot2::aes(Slope)) +
  ggplot2::geom_density(fill = "grey80") +
  ggplot2::geom_vline(
    data = ch8a_targets, ggplot2::aes(xintercept = Target),
    linetype = 2
  ) +
  ggplot2::facet_wrap(~Model, scales = "free_x") +
  ggplot2::labs(x = "Estimated slope of X", y = "Density")
```

Both procedures center near their own targets. The conditional estimates
are more variable here, because separating correlated predictors leaves
less residual variation in $X$. That is a result of this specified
process, not a general promise that adjustment always lowers precision:
prognostic covariates and their relations to the focal predictor both
matter. A simulation shows the consequences of its assumptions; it cannot
certify the observational occupation study.

\Needspace{12\baselineskip}

### Fitted group means from indicators {.unnumbered}

Start with a model containing only the three-level type factor. With
blue-collar as reference,

$$
\widehat y_g=\widehat\alpha+
\widehat\delta_{\mathrm{prof}}D_{\mathrm{prof}}+
\widehat\delta_{\mathrm{wc}}D_{\mathrm{wc}}.
$$

The fitted means are $\widehat\mu_{\mathrm{bc}}=\widehat\alpha$,
$\widehat\mu_{\mathrm{prof}}=\widehat\alpha+\widehat\delta_{\mathrm{prof}}$,
and $\widehat\mu_{\mathrm{wc}}=\widehat\alpha+\widehat\delta_{\mathrm{wc}}$.
OLS minimizes squared deviations separately within each group, so these
equal the observed group means; the hats distinguish fitted summaries
from hypothetical population means.

@tbl-ch8a-group-means reads the fit on the 98 records. The group-only
coefficients are $\widehat\alpha=5.3741$, $\widehat\delta_{\mathrm{prof}}=5.1853$
and $\widehat\delta_{\mathrm{wc}}=-0.3218$; each fitted mean combines them
as the indicator columns direct and equals the observed mean income of
its type. Sums of the rounded coefficients can differ from the fitted
means in the last digit.

\Needspace{9\baselineskip}

| Type | $D_{\mathrm{prof}}$ | $D_{\mathrm{wc}}$ | Coefficients combined | Fitted mean | Observed mean |
|:--|--:|--:|:--|--:|--:|
| bc | 0 | 0 | $\widehat\alpha$ | 5.3741 | 5.3741 |
| prof | 1 | 0 | $\widehat\alpha+\widehat\delta_{\mathrm{prof}}$ | 10.5595 | 10.5595 |
| wc | 0 | 1 | $\widehat\alpha+\widehat\delta_{\mathrm{wc}}$ | 5.0523 | 5.0523 |

: The type-only regression reproduces the three observed group means. {#tbl-ch8a-group-means tbl-colwidths="[10,10,10,34,18,18]"}

This group-only description is not the adjusted type comparison in *m2*,
where the indicators compare fitted values at the same education,
prestige, and percentage of women. The factor is one term represented by
two columns; it is neither an ordered numeric score nor one selected
reference-category coefficient.

\Needspace{8\baselineskip}

Releveling changes the intercept and indicator coefficients, not the
fitted values, the residuals, or a fixed substantive contrast;
releveling leaves the fitted values equal up to floating-point noise. No new question
appears merely because the coordinates changed. Centering the continuous
predictors likewise changes the intercept's reference profile without
changing the fitted values, and it cannot create cross-type support.

## Appendix B: Formulas Behind the Checks and Criteria {.unnumbered #app-ch8-b}

This appendix gives the formulas and fuller tables behind Sections 3
and 4, on the same *dat*, *m0*, *m1*, and *m2*. None of these refinements supplies a missing
design, measurement, independence, or transport warrant.

\Needspace{12\baselineskip}

### Cook's distance {.unnumbered}

With $p$ coefficient columns including the intercept, leverage is
$h_{ii}=\mathbf x_i^{\mathsf T}(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf x_i$. Cook's distance can be calculated as

$$
D_i=\frac{e_i^2}{p s^2}\frac{h_{ii}}{(1-h_{ii})^2}
=\frac{(\widehat{\boldsymbol\beta}-\widehat{\boldsymbol\beta}_{(-i)})^{\mathsf T}\mathbf X^{\mathsf T}\mathbf X
(\widehat{\boldsymbol\beta}-\widehat{\boldsymbol\beta}_{(-i)})}{p s^2}.
$$

The second expression measures the change in all coefficients when
record $i$ is deleted, in the design matrix's metric rather than as an
unweighted sum of squared coefficients with incompatible units.

```{=latex}
\Needspace{12\baselineskip}
```

```{r}
#| label: tbl-ch8b-influence
#| echo: false
#| tbl-cap: "The five largest Cook's distances in the full fit."
ch8_minus4 <- function(x) {
  out <- format(round(x, 4), digits = 15, scientific = FALSE, trim = TRUE)
  ifelse(x < 0, paste0("$", out, "$"), out)
}
ch8_influence <- ch8_diag[ch8_top,
  c("occupation", "residual", "leverage", "cooks")]
ch8_influence[-1] <- lapply(ch8_influence[-1], ch8_minus4)
knitr::kable(ch8_influence, row.names = FALSE, escape = FALSE,
  align = c("l", "r", "r", "r"),
  col.names = c("Occupation", "Residual", "Leverage", "Cook's distance"))
```

General managers have the largest Cook's distance, approximately 0.4467,
and Section 4.3 explains why the record is inspected and retained.
Refitting without it moves education from 0.1164 to 0.4372, and the type
coefficients move as well. The comparison records a diagnostic perturbation of the fit,
not six new research questions or evidence that the reduced sample is
preferable.

\Needspace{9\baselineskip}

### The HC3 covariance {.unnumbered}

For independent observations, the HC3 covariance estimator is

$$
\widehat{\mathbf V}_{\mathrm{HC3}}=(\mathbf X^{\mathsf T}\mathbf X)^{-1}
\left\{\sum_i \mathbf x_i\mathbf x_i^{\mathsf T}\frac{e_i^2}{(1-h_{ii})^2}\right\}
(\mathbf X^{\mathsf T}\mathbf X)^{-1}.
$$

The outer factors are the familiar design component. The middle
component lets each record carry its own residual scale, inflated for
leverage [@long2000; @zeileis2004]. The OLS coefficients do not change;
only their covariance does, as @tbl-ch8b-hc3 shows for all six
coefficients; `vcovHC()` evaluates this covariance calculation.

```{r}
#| label: tbl-ch8b-hc3
#| echo: false
#| tbl-cap: "Coefficient estimates and two covariance summaries."
knitr::kable(data.frame(
  Term = names(coef(m2)), Estimate = ch8_minus4(coef(m2)),
  Classical_SE = ch8_minus4(sqrt(diag(vcov(m2)))),
  HC3_SE = ch8_minus4(sqrt(diag(hc3_covariance))), row.names = NULL
), row.names = FALSE, escape = FALSE, align = c("l", "r", "r", "r"),
  col.names = c("Term", "Estimate", "Classical SE", "HC3 SE"))
```

Education's HC3 SE is 0.3823 rather than 0.2752, and with $t_{92}$ as an
approximate reference the interval is $[-0.6429,0.8758]$; this is not an
exact finite-sample HC3 coverage result. Women's HC3 SE is smaller than
its classical counterpart: robust covariance need not widen every
interval.

For a maintained linear conditional mean, HC3 addresses unequal variance
among independent observations. For a random-design best-linear-projection
target, a suitable sandwich covariance can support large-sample
projection inference without making the mean exactly linear. Neither use
repairs measurement error, dependence, selection, causal identification,
or absent joint support.

\Needspace{13\baselineskip}

### VIF and generalized VIF {.unnumbered}

For a single predictor column, the classical conditional variance of its
coefficient can be written

$$
\operatorname{Var}(\widehat\beta_j\mid\mathbf X)=
\frac{\sigma^2}{\sum_i(X_{ij}-\bar X_j)^2}
\frac{1}{1-R_j^2}.
$$

Here $R_j^2$ comes from regressing that predictor on the remaining
columns, including the intercept. The variance inflation factor is
$VIF_j=1/(1-R_j^2)$, and its square root is an SE multiplier, holding the
residual variance and the focal predictor's variation fixed
[@kutner2005]. Education gives $R_j^2=.8765$, VIF 8.0968, and SE
multiplier 2.8455 (Section 4.2), from the auxiliary regression or
`car::vif(m2)`. The education--prestige
correlation is .8665, but the VIF uses all the other columns, not just
that pair.

The two-column type term has a generalized VIF. Its raw GVIF is 7.7001;
the dimension-adjusted value $GVIF^{1/(2\,df)}=1.6658$ is on a scale
comparable to the square root of a one-column VIF [@fox2019], whereas the
raw two-column GVIF is not comparable with an ordinary one-column VIF.

Exact collinearity makes the design rank-deficient. Near collinearity
leaves the coefficients estimable but possibly imprecise, and under the
maintained zero-conditional-mean model it does not by itself bias OLS. A
cutoff of 5 or 10 neither establishes the model's assumptions nor
justifies removing a variable the question requires. Measurement problems
need their own evidence: the familiar attenuation story requires
classical error assumptions and is not a rule for every coefficient in a
multiple-predictor model.

### The Normal likelihood, AIC, and BIC {.unnumbered}

Chapter 6 varied the candidate probability for a fixed binomial count of
32 successes in 50 trials. Here the data again stay fixed, but the
response is continuous and the parameters are several coefficients plus
an error scale. For a declared independent Normal linear model, the
**likelihood** is the joint density of the observed outcomes at their
fixed predictor values, viewed as a function of those parameters; it is
not a probability assigned to a parameter, nor the probability of the
exact observed vector. The **log-likelihood** adds the log densities of
the independent records.

Write the predictor row for record $i$ as $\mathbf x_i^{\mathsf T}$ and let $\sigma>0$ be the
error SD. Before fitting, the log-likelihood is

$$
\ell(\boldsymbol\beta,\sigma)=-n\log\sigma-\frac n2\log(2\pi)
-\frac{1}{2\sigma^2}\sum_i(y_i-\mathbf x_i^{\mathsf T}\boldsymbol\beta)^2.
$$

At any fixed positive scale only the last term changes with the
coefficients, so maximizing over $\boldsymbol\beta$ minimizes the residual
sum of squares: the Normal maximum-likelihood coefficients equal the OLS
coefficients. This connects Chapter 7's squared-loss criterion to this
particular distributional model, although least squares did not need the
model to calculate a minimizing line.

Maximizing over the scale gives
$\widehat\sigma_{\mathrm{ML}}=\sqrt{RSS/n}$, whereas the residual SD used
for classical coefficient uncertainty is $s=\sqrt{RSS/(n-p)}$, with $p$
counting the regression coefficients. The two serve different purposes,
and the ML scale, not the residual SD, belongs in the fitted likelihood.
At the fitted coefficients and ML scale,

$$
\ell(\widehat{\boldsymbol\beta},\widehat\sigma_{\mathrm{ML}})
=-\frac n2\left\{\log(2\pi)+1+\log(RSS/n)\right\}.
$$

R's `logLik()` returns this maximized value for an `lm()` fit: the Normal
log densities of the 98 outcomes at their fitted means and the ML scale
add to the same number. Two
information criteria combine it with a penalty for the number of
parameters:

$$
AIC=-2\log\widehat L+2k,\qquad
BIC=-2\log\widehat L+k\log n.
$$

Here $k$ counts the regression coefficients **and** the estimated
residual variance, so the full occupation model has $k=7$. Smaller values
are preferred under each criterion; neither is a held-out MAE
[@akaike1974; @schwarz1978]. For *m2* the ML scale is 2.5373 and the
residual SD 2.6187; the maximized log-likelihood is $-230.302$, so its
AIC is 474.6040 and its BIC 492.6988.

@tbl-ch8b-criteria compares the three declared models on the same 98
rows, response scale, and likelihood convention. Its adjusted $R^2$, $1-(1-R^2)(n-1)/(n-p)$,
charges each added coefficient a small penalty: it rises only when the
added terms' nested-model $F$ statistic exceeds 1. Like $R^2$, AIC and
BIC, it is computed from the fit to the same 98 rows, so it is an
in-sample summary, not a held-out measure. Chapter 11 develops likelihood
fitting further for binomial outcomes; this local Normal-model
calculation does not need it.

\Needspace{9\baselineskip}

| Model | $k$ | $R^2$ | Adjusted $R^2$ | AIC | BIC |
|:--|--:|--:|--:|--:|--:|
| m0 | 3 | .3296 | .3226 | 526.5052 | 534.2601 |
| m1 | 4 | .4996 | .4891 | 499.8328 | 510.1727 |
| m2 | 7 | .6362 | .6164 | 474.6040 | 492.6988 |

: In-sample summaries for comparable income models. {#tbl-ch8b-criteria}

A criterion printed for a log-income model is not comparable with these
raw-income values, because changing the response scale changes the
likelihood's reference measure; a common-scale comparison would need the
corresponding transformation adjustment.

No residual graph establishes independence: the occupations share one
historical labour market, and clustered or repeated records would need
dependence-aware uncertainty, which HC3 alone does not supply. Rules
for choosing a causal adjustment set, about confounders, mediators, and
colliders, apply to a declared causal target (Chapter 14), not to
association or prediction.

```{r}
#| label: ch8-anchor-check
#| include: false
local({
  had_seed <- exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
  if (had_seed) old_seed <- get(".Random.seed", envir = .GlobalEnv)
  on.exit({
    if (had_seed) assign(".Random.seed", old_seed, envir = .GlobalEnv)
    else if (exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE))
      rm(".Random.seed", envir = .GlobalEnv)
  })
  check <- function(actual, expected, label) {
    if (!isTRUE(all.equal(unname(round(as.numeric(actual), 4)),
                         as.numeric(expected), tolerance = 1e-9)))
      stop("Chapter 8 numeric anchor: ", label)
  }
  stopifnot(nrow(dat) == 98, identical(as.integer(table(dat$type)), c(44L,31L,23L)),
            identical(levels(dat$type), c("bc","prof","wc")), df.residual(m2)==92)
  check(c(mean(dat$education),sd(dat$education),cor(dat$education,dat$prestige),
          cor(dat$education,dat$women)),c(10.7951,2.7489,.8665,.0619),"records")
  fits <- list(m0,m_w,m1,m2)
  check(sapply(fits,function(f) coef(f)["education"]),
        c(.883,.9292,-.2179,.1164),"same-row slopes")
  check(sapply(fits,function(f) summary(f)$r.squared),
        c(.3296,.5641,.4996,.6362),"training R squared")
  check(c(coef(m_w)["women"],sqrt(diag(vcov(m_w)))[c("education","women")]),
        c(-.0654,.1044,.0091),"education and women")
  check(c(coef(lm(prestige~education,dat))[2],coef(lm(women~education,dat))[2]),
        c(5.3884,.7069),"omitted-variable slopes")
  check(coef(m2),c(.4205,.1164,.1406,-.0535,.3242,.2353),"full coefficients")
  check(sqrt(diag(vcov(m2)))[-1],c(.2752,.0354,.0097,1.4631,.9848),"full SEs")
  check(c(sigma(m2),sigma(m2)^2,summary(m2)$adj.r.squared),
        c(2.6187,6.8575,.6164),"full fit")
  check(c(cor(r_E,r_Y),sd(r_E),sd(dat$education)/sd(r_E)),
        c(.0441,.9661,2.8455),"residual variation")
  check(coef(summary(m2))["education",],c(.1164,.2752,.423,.6732),"partial inference")
  check(coef(summary(lm(r_Y~r_E)))[2,],c(.1164,.2694,.4321,.6666),"naive residual fit")
  check(c(var(dat$education),vcov(m2)["education","education"]),
        c(7.5567,.0757),"two covariance objects")
  check(confint(m2,"education"),c(-.4302,.6631),"classical interval")
  check(c(hc3_se,coef(m2)["education"]+c(-1,1)*qt(.975,92)*hc3_se),
        c(.3823,-.6429,.8758),"HC3 interval")
  check(c(summary(m2)$fstatistic[1],summary(m0)$fstatistic[1]),
        c(32.1747,47.1956),"overall F")
  stopifnot(abs(sum(support_weights)-1)<1e-10,
    max(abs(colSums(as.matrix(support_rows)*support_weights)-c(11,45,50)))<1e-10)
  check(car::vif(m2)[c("education","prestige","women"),1],
        c(8.0968,5.17,1.3007),"VIF")
  check(c(anova(m2,ch8_curve)$F[2],anova(m2,ch8_curve)$`Pr(>F)`[2],
          coef(ch8_curve)["education"]),c(6.736,.011,-.0298),"curvature sensitivity")
  check(c(cooks.distance(m2)["general.managers"],
          coef(update(m2,data=dat[rownames(dat)!="general.managers",]))["education"]),
        c(.4467,.4372),"deletion sensitivity")
  check(predict(m2,profile,interval="confidence"),c(5.5886,4.4865,6.6908),"profile mean")
  check(predict(m2,profile,interval="prediction"),c(5.5886,.2722,10.9051),"profile new case")
  for (g in levels(dat$type)) {
    changed <- dat; changed$type <- factor(g,levels=levels(dat$type))
    a <- colMeans(model.matrix(~education+prestige+women+type,changed))
    se <- sqrt(drop(t(a)%*%vcov(m2)%*%a)); est <- mean(predict(m2,changed))
    target <- switch(g,bc=c(6.7811,.6942,5.4023,8.1598),
       prof=c(7.1053,.9011,5.3156,8.895),wc=c(7.0164,.6177,5.7896,8.2432))
    check(c(est,se,est+c(-1,1)*qt(.975,92)*se),target,paste("standardization",g))
  }
  check(tapply(dat$income_k,dat$type,mean),c(5.3741,10.5595,5.0523),"group means")
  check(c(coef(lm(score~strategy+prior,ch8a_sim))[2:3],
          sqrt(diag(vcov(lm(score~strategy+prior,ch8a_sim))))[2:3],
          coef(lm(score~strategy,ch8a_sim))[2],
          sqrt(diag(vcov(lm(score~strategy,ch8a_sim))))[2]),
        c(3.0259,4.3991,.5325,.5282,5.9971,.4466),"one simulated sample")
  check(c(colMeans(ch8a_slopes),apply(ch8a_slopes,2,sd)),
        c(5.2581,2.001,.5952,.6707),"repeated samples")
  cv <- function(fold) {
    pred <- matrix(NA_real_,nrow(dat),2)
    for (k in sort(unique(fold))) {
      tr <- dat[fold!=k,];te <- dat[fold==k,]
      a <- lm(income_k~education,tr)
      b <- lm(income_k~education+prestige+women+type,tr)
      stopifnot(all(table(tr$type)>0),a$rank==2,b$rank==6)
      pred[fold==k,] <- cbind(predict(a,te),predict(b,te))
    }
    list(pred=pred,mae=colMeans(abs(pred-dat$income_k)))
  }
  set.seed(80806);fold_check <- sample(rep(seq_len(10),length.out=nrow(dat)))
  ans <- cv(fold_check)
  stopifnot(fold_check[1]==9,sum(fold_check==9)==9)
  check(ans$pred[1,],c(8.9461,11.5948),"first held-out predictions")
  check(ans$mae,c(2.4031,1.5412),"pooled MAE")
  check(sqrt(colMeans((ans$pred-dat$income_k)^2)),c(3.5051,2.7111),"pooled RMSE")
  set.seed(80807)
  resplit <- replicate(200,{
    out <- cv(sample(rep(seq_len(10),length.out=nrow(dat))))
    out$mae[2]-out$mae[1]
  })
  check(c(mean(resplit),range(resplit)),c(-.8311,-.914,-.741),"resplit sensitivity")
  check(c(sqrt(mean(residuals(m2)^2)),as.numeric(logLik(m2)),AIC(m2),BIC(m2)),
        c(2.5373,-230.302,474.604,492.6988),"likelihood criteria")
})
```

## References {.unnumbered}

::: {#refs}
:::
