---
title: "How to Read a Quantitative Study as an Argument"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-cap: true
    tbl-cap: true
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(dev = "cairo_pdf")
}

draw_argument <- function() {
  old <- par(mar = rep(0.15, 4), xpd = NA)
  on.exit(par(old))
  plot.new()
  plot.window(xlim = c(0, 19.5), ylim = c(0, 9.8))

  centres <- c(4.25, 8.15, 12.65, 17.05)
  widths <- c(2.70, 3.05, 3.15, 3.25)
  rows <- list(stat = c(7.05, 8.62), sub = c(0.75, 2.32))
  fills <- c("#DCEAF3", "#DDEEDB", "#F8EDC7", "#F5D9C8")
  labels <- list(
    stat = c("Recorded\nevidence", "Statistical\njustification",
             "Statistical\ninference", "Bounded\nstatistical claim"),
    sub = c("Situated\nevidence", "Substantive\njustification",
            "Substantive\ninference", "Bounded\nsubstantive claim")
  )

  component <- function(x, width, y, code, label, fill) {
    rect(x - width / 2, y[1], x + width / 2, y[2],
         col = fill, border = "#203744", lwd = 1.15)
    text(x, mean(y) + 0.31, code, font = 2, cex = 0.92,
         col = "#203744")
    text(x, mean(y) - 0.31, label, cex = 0.67, col = "#29353B")
  }

  for (row_name in names(rows)) {
    y <- rows[[row_name]]
    prefix <- if (row_name == "stat") "STAT-" else "SUB-"
    for (i in seq_along(centres)) {
      component(
        centres[i], widths[i], y,
        paste0(prefix, c("E", "J", "I", "C")[i]),
        labels[[row_name]][i], fills[i]
      )
    }
    y_mid <- mean(y)
    text(6.10, y_mid, "+", font = 2, cex = 1.22, col = "#344A55")
    arrows(9.82, y_mid, 10.93, y_mid, length = 0.08,
           lwd = 1.35, col = "#344A55")
    arrows(14.37, y_mid, 15.28, y_mid, length = 0.08,
           lwd = 1.35, col = "#344A55")
  }

  relations <- c(
    "situates", "augments and\ncontextualizes",
    "interprets", "bounds"
  )
  for (i in seq_along(centres)) {
    segments(centres[i], rows$stat[1], centres[i], 5.45,
             col = "#9A5A35", lwd = 1.05)
    rect(centres[i] - 1.28, 4.28, centres[i] + 1.28, 5.43,
         col = "white", border = "#C28C69", lwd = 0.85)
    text(centres[i], 4.855, relations[i], cex = 0.68,
         col = "#7A4B2A")
    arrows(centres[i], 4.25, centres[i], rows$sub[2],
           length = 0.075, col = "#9A5A35", lwd = 1.05)
  }

  text(0.15, mean(rows$stat), "STATISTICAL\nLAYER",
       adj = c(0, 0.5), font = 2, cex = 0.66, col = "#244A63")
  text(0.15, mean(rows$sub), "SUBSTANTIVE\nLAYER",
       adj = c(0, 0.5), font = 2, cex = 0.66, col = "#244A63")
  for (i in seq_along(centres)) {
    text(
      centres[i], 9.22,
      c("EVIDENCE", "JUSTIFICATION", "INFERENCE", "CLAIM")[i],
      font = 2, cex = 0.73, col = "#56656C"
    )
  }
}

draw_workflow <- function() {
  old <- par(mar = rep(0.2, 4), xpd = NA)
  on.exit(par(old))
  plot.new()
  plot.window(xlim = c(0, 14), ylim = c(0, 8))

  rect(0.15, 4.15, 13.85, 7.60, col = "#EEF5F8", border = "#9BB6C4")
  rect(0.15, 0.25, 13.85, 3.70, col = "#F1F6EF", border = "#A7BFA2")
  text(0.38, 7.30, "PHASE 1 - UNDERSTAND AND PLAN | iterative Stages 1-3",
       adj = 0, font = 2, cex = 0.67, col = "#244A63")
  text(0.38, 3.40, "PHASE 2 - PRODUCE AND COMPLETE | iterative Stages 4-6",
       adj = 0, font = 2, cex = 0.65, col = "#466B43")

  x <- c(0.48, 5.05, 9.62)
  width <- 3.88
  labels <- c(
    "Review problem\nand literature",
    "Frame question and\ncandidate answer",
    "Align design, measures,\nestimand, evidence,\nand inference plan",
    "Produce or acquire\nand document evidence",
    "Estimate, check,\nand analyse",
    "Contextualize, interpret,\nand complete the\nbounded answer"
  )

  stage_box <- function(x0, y0, number, label) {
    rect(x0, y0, x0 + width, y0 + 1.62, col = "white",
         border = "#344A55", lwd = 1)
    text(x0 + 0.18, y0 + 1.34, paste0("STAGE ", number), adj = 0,
         font = 2, cex = 0.60, col = "#203744")
    text(x0 + width / 2, y0 + 0.66, label, cex = 0.66,
         col = "#29353B")
  }

  for (i in 1:3) stage_box(x[i], 4.65, i, labels[i])
  bottom_stages <- c(6, 5, 4)
  for (i in 1:3) {
    stage_box(x[i], 0.75, bottom_stages[i], labels[bottom_stages[i]])
  }

  for (i in 1:2) {
    arrows(x[i] + width + 0.02, 5.46, x[i + 1] - 0.02, 5.46,
           length = 0.07, col = "#4F6875")
  }
  for (i in 3:2) {
    arrows(x[i] - 0.02, 1.56, x[i - 1] + width + 0.02, 1.56,
           length = 0.07, col = "#557052")
  }

  stage_right <- x[3] + width / 2
  arrows(stage_right, 4.63, stage_right, 2.39, length = 0.07,
         col = "#6B6B6B")
  arrows(x[3] + width / 2, 6.58, x[1] + width / 2, 6.58,
         length = 0.07, lty = 2, col = "#7A4B2A")
  text(mean(c(x[1], x[3])) + width / 2, 6.84,
       "revise and realign before evidence production",
       cex = 0.65, col = "#7A4B2A")
  arrows(x[1] + width / 2, 0.64, x[3] + width / 2, 0.64,
         length = 0.07, lty = 2, col = "#7A4B2A")
  text(mean(c(x[1], x[3])) + width / 2, 0.40,
       "verify, reanalyse, and recalibrate within Phase 2",
       cex = 0.65, col = "#7A4B2A")
}
```

{{< pagebreak >}}

::: {.callout-tip title="Chapter box"}
**Central question:** How can a reader identify and trace the support connecting
one quantitative research question to the answer a study reports, together with
its stated bounds?

**Focal components:** One argument with linked statistical and substantive
layers. Within each layer, Evidence and Justification support Inference, which
supports a bounded Claim. The worked reading completes STAT-E through STAT-C
and SUB-E through SUB-C for one question--answer pair. Six quality-evaluation
questions help readers judge the resulting support, while a six-stage,
two-phase workflow shows how researchers construct it.

**Main evidence:** A hypothetical randomized special-practice study with 120
consenting eligible students assigned 60 per format, complete outcome
follow-up, a common 0--40 delayed assessment, an estimated 3.1-point mean
difference, a 95% confidence interval from 0.9 to 5.3 points, and a documented,
course-specific two-point benchmark for potential educational importance. The
example is used first as a completed argument to read and then as an
argument-construction cycle to follow.

**Scope:** This basic tutorial is designed for a simple study with one
primary quantitative question, or for several independent questions that can
each be addressed as a separate question--answer pair. Reading quantitative
studies is its primary use. Quality evaluation is a distinct secondary use,
and the construction workflow is the auxiliary use for conducting research.
Studies with coordinated questions whose estimands, methods, or
interpretations interact are developed in *Extending the Argument-Based
Approach to Quantitative Studies*.
:::

## A Quantitative Study as a Question--Answer Contract

### From an empirical problem to a quantitative question

Empirical research uses observations to address questions about the world.
**Quantitative research** is the broader process of formulating questions,
planning a study, producing or acquiring numerical observations, analysing
them, and making bounded claims. A **quantitative study** is the particular
inquiry that readers encounter through a report or paper and, where available,
its protocol, appendices, data documentation, code, and other related
materials.

The distinction matters because a paper is a reorganized record rather than a
diary of the research process. A decision described in the Methods section may
have been made before data collection, during analysis, or during revision.
Important support may appear in a protocol or supplement rather than the main
paper. Reading well therefore requires recovering the reasoning across
locations without mistaking manuscript order for research chronology.

Consider a university that is choosing between two ways of helping students
practise course material. Its practical concern might be expressed as:

> Does a different kind of practice improve learning?

This is not yet a sufficiently clear quantitative question. *Learning* might
mean an immediate quiz, delayed performance, course completion, confidence, or
something else. *Students* might mean volunteers in one class, all eligible
students at one university, or students in a wider system. *Improve* might ask
for a description, association, prediction, or causal comparison.

A useful quantitative question identifies, in readable language:

- whom or what the question concerns;
- the outcome or phenomenon to be learned about;
- the comparison or relationship of interest; and
- the kind of answer sought.

A fuller research specification can then record the unit, population, setting,
period, variables, estimand, intended claim mode, and planned analysis.
Keeping the main question readable while documenting this fuller specification
prevents the statistical method from quietly becoming the research question.

Four common question modes create different answer obligations. The mode comes
from what the question asks, not from the name of the statistical procedure.

| Mode | Beginner question form | Example estimand | Typical support question |
|:-----|:-----------------------|:---------------|:-------------------------|
| **Descriptive** | What is present, how much is there, or how is it distributed? | A mean or proportion for a named group, place, and period | Are the cases and measurements suitable for what is being described? |
| **Associational** | How do measured variables differ or vary together? | A difference or relationship for named variables and cases | Does the comparison describe an association without implying causation? |
| **Predictive** | How well can an outcome be predicted for new cases, settings, or times? | A specified performance measure for a stated prediction rule in a defined population of new cases, settings, or times | Was prediction checked on suitable observations not used to fit the procedure? |
| **Causal** | What would differ under one intervention or exposure rather than another? | An average outcome contrast under two named alternatives | Do the design and assumptions permit the intended causal comparison? |

: *Four question modes, example estimands, and support questions.* {#tbl-question-modes tbl-colwidths="[14,29,25,32]"}

The intended claim mode should match the question mode. A causal question seeks
a causal answer, for example, but causal wording is warranted only when the
design and justification support it. If they do not, the study may provide a
narrower causal answer that keeps the same mode and estimand, or an inconclusive
answer to the causal question. An associational answer belongs to a different
pair rather than quietly replacing the causal question.

Mode is different from analytic status. A descriptive analysis can be primary
or exploratory; a causal analysis can also be primary, secondary, or
exploratory. These labels report how the analysis entered the study, not what
kind of answer its evidence can support.

Every mode also needs an **estimand**. It is a precisely specified target
quantity: what the study seeks to learn for named units or a population,
conditions or a comparison, an outcome, a summary, and a scope
[@lundberg2021]. For a predictive question, the estimand may be a specified
performance measure for a stated prediction rule in a defined population of
new cases, settings, or times; an arbitrary prediction for one individual is
not itself the estimand. An estimand may answer a question directly or
contribute to its answer
when several estimands are required. In the worked study, the estimand is the
mean difference in delayed-assessment scores that would result for the
participating students under assignment to Special rather than Usual Practice.
The observed 3.1-point difference is an **estimate** obtained from the evidence
for that estimand. Later chapters refine these terms for particular methods; the
beginner's task is to distinguish what the study seeks from what it observed.

### Questions, hypotheses, results, and answers

A **research question** states what the study seeks to learn. A
**hypothesis** is an optional prospective candidate answer. A **result** is a
quantity produced by analysis, such as a mean, difference, interval, or
prediction. A **claim** is a statement the researchers make after interpreting
the result. These terms are related, but they are not interchangeable.

The order of responsibility is:

1. The question states the answer obligation.
2. A hypothesis may record an expectation before the focal evidence is
   examined.
3. Analysis produces one or more results.
4. Reasoning connects those results to an eventual bounded answer, which may
   itself be inconclusive.

A hypothesis is not required for every worthwhile study. Descriptive and
exploratory questions can be valuable without one. A result contributes to an
answer, but it does not determine that answer's meaning, support, or scope.

Depending on the question and evidence, a reported answer may be positive,
negative, close to no difference, conditional, or inconclusive. These are
ordinary descriptions, not a second classification system. Whatever its
direction, the answer should preserve magnitude, uncertainty, assumptions, and
scope. A planned analysis can be inconclusive, and an exploratory analysis can
still report a carefully bounded answer.

The practical sequence is therefore not “question, calculation, conclusion.”
It is question, documented support, analytical result, interpretation, and
bounded answer. A reader who keeps these roles distinct is less likely to treat
a hypothesis as a finding or a numerical result as a complete conclusion.

### The question--answer contract

A finalized quantitative study can be read as a
**question--answer contract**:

> Each stated quantitative question creates an obligation to provide a
> traceable answer no stronger than the available evidence and justification
> permit.

The word *contract* is a metaphor. It is not a legal agreement, and it does not
assume that every paper fulfils its obligations. A paper may report a positive,
negative, close-to-no-difference, or inconclusive answer, or it may fail to
provide one; Section 4 separately evaluates what answer the accessible support
warrants.

The unit of this contract is a **question--answer pair**: one research question
together with one separately reportable question-facing answer obligation.
Hypotheses, coefficients, tests, contrasts, and other results stay inside that
pair's argument when they only help discharge the same obligation; a separately
owed conclusion creates another pair, and several pairs require several
arguments even when they share support. When a paper states an aim rather than
an explicit question, the reader writes a question from the aim and headline
result and marks it as reader-supplied rather than author-stated.

This basic tutorial also applies when a study contains several **independent
questions** that can be handled separately. Here, *independent* means separable
for the purpose of reading their arguments: answering one question does not
require another question's result to specify its estimand or interpret its
claim.
When questions must be coordinated--for example, when an interaction qualifies
an average effect--cross-question alignment and synthesis are required. Those
extensions belong to *Extending the Argument-Based Approach to Quantitative
Studies*. The present worked example concentrates on one pair.

The metaphor is deliberately demanding in two directions. It requires authors
to make support visible, but it also requires readers to distinguish absence
from the paper from absence in the research. A paper may omit information that
exists in a protocol or supplement. Conversely, a sentence in a Methods
section does not prove that the reported procedure was adequate or followed.

### Reading Studies, Evaluating Quality, and Conducting Research

The same question-to-answer support is encountered from different positions.

**Reading studies** is the primary use. A reader identifies the question,
locates the reported answer, and traces the support available in the accessible
record. The reader should not assume that every relevant decision is visible or
that every reported decision was prospective.

**Evaluating quality** is the secondary use. It follows and depends on reading:
after identifying the argument, an evaluator asks whether its question,
evidence, justification, inference, transitions, and final answer remain
aligned, adequate, faithful, and appropriately limited. Evaluation is not part
of reading's definition. Section 4 introduces a small common set of questions
for making those judgments.

**The construction workflow** is the auxiliary use for conducting research. It
is prospective and iterative. Researchers
formulate questions, choose measures and comparisons, anticipate threats,
produce evidence, analyse it, and revise what they can claim. The eventual
argument is not normally written as eight labelled components during the
study, but those logical obligations are being built throughout the work.
Section 5 organizes much of that work in a six-stage construction workflow.

Reporting matters because it makes the support inspectable, but it is not a
fourth organizing activity here. A good report does more than state a result.
It connects the question to the population, unit, design, measure, estimand,
analysis, uncertainty, interpretation, and reservations. Related materials can
strengthen this record by showing prospective plans, detailed measurement
information, processing decisions, code, and sensitivity analyses.

The approach is not a complete research-management or study-appraisal system.

These activities create complementary responsibilities. Authors should make the
support traceable without requiring readers to reverse-engineer every decision.
Readers should examine the complete accessible record without assuming that a
conventional heading guarantees a logical function. A tidy final paper can
conceal an iterative process, while a careful process can be difficult to
evaluate if reporting is incomplete.

### Where the support may be found

Conventional sections help navigation, but they are not the logical components
of an argument. One justification may be distributed across an introduction,
protocol, methods section, and supplement. One methods paragraph may contribute
to several components. Reporting standards improve traceability, but readers
still have to recover the function performed by each piece of information
[@appelbaum2018].

```{=latex}
\Needspace{0.48\textheight}
```

| Logical role | Possible study location | Information a reader may seek |
|:-------------|:------------------------|:------------------------------|
| **Question and candidate answer** | Abstract, introduction, or protocol | Purpose, quantitative question, intended claim mode, hypothesis when present, and important planned choices |
| **Evidence and its production** | Methods, participant flow, instrument documentation, data dictionary | Population and unit, recruitment, assignment or exposure, measurement, timing, implementation, exclusions, and completeness |
| **Design and justification** | Protocol, Methods, analysis plan, supplement | Why the evidence, comparison, and uncertainty procedure can support the intended move |
| **Analysis and inference** | Results, code, or analytical supplement | Group summaries, difference, uncertainty, and relevant checks |
| **Bounded answer and reservations** | Discussion and conclusion | Substantive meaning, alternatives, scope, and unresolved reservations |

: *Where support for a question--answer pair may appear.* {#tbl-study-locations tbl-colwidths="[25,31,44]"}

Location is not logical role. A sample description, for example, helps identify
the empirical evidence, may help justify reference to a target population, and should
constrain the final claim. The argument-based approach asks what work the
information does, wherever it appears.

## One Argument with Linked Layers

### One argument for one question--answer pair

The argument-based approach reads the contract through one central rule:

> Each question--answer pair has **one argument with linked layers**.

For the observable-variable scope of this tutorial, the argument has two
layers:

- the **statistical layer (STAT)** concerns what the recorded observations
  support about a defined estimand; and
- the **substantive layer (SUB)** concerns what that statistical support means
  for the people, practices, institutions, or other phenomena named in the
  question.

The statistical layer is the minimum. A purely statistical or methodological
question may be answered there. A substantive empirical question requires the
substantive layer even when the calculation is elementary. The need for the
substantive layer depends on the question, not on whether the method appears
simple or advanced.

The layers belong to the same argument. They are parallel in form but distinct
in function. The statistical layer does not determine substantive meaning. The
substantive layer cannot erase uncertainty, unsupported conditions, or design
limitations. A responsible answer must keep both forms of reasoning connected.

### Four argument components

An **argument component** names a logical function in the argument a study
presents. Its **content** is the study-specific information or reasoning that
performs that function; a **documentary trace** is the passage, table, protocol,
or other accessible record through which we identify that content. For example,
a Methods paragraph may document how participants were assigned and give a
reason for comparing their outcomes. The paragraph can provide traces for two
functions. Its location under Methods does not determine either function.

Within each layer, four components perform different jobs:

- **Evidence (E)** is what was observed together with relevant information
  about how the observations were produced.
- **Justification (J)** records the reasons the study offers for treating the
  evidence and chosen method as support for the intended inference.
- **Inference (I)** is the reasoned move that applies the justification to the
  evidence by bringing together the estimand, result, uncertainty, relevant
  checks, and analytic status.
- **bounded Claim (C)** is the claim the study reports together with the bounds
  it states. Whether those bounds are adequate is evaluated later.

Calling something **Justification** identifies the role it plays, not whether
it succeeds. A mistaken or incomplete reason still belongs in J when the study
offers it to support an inference. Reading records that reason; evaluation
judges its adequacy. Likewise, a trace of a reported action does not prove that
the action occurred or was performed well. Justification can also invoke a
principle or assumption, so it need not describe a research action at all.

The word **bounded** names the component's intended function. Ideally, its force
and scope retain the uncertainty, assumptions, design, alternatives, and
population limits on which it depends. While reading, however, record the claim
and whatever bounds the study actually states, even when they appear incomplete
or too broad. Deciding whether they are responsible belongs to evaluation.

**Bounded Claim** names an argument component; the **reported answer** is what
the study says in response to the question. Evaluation may confirm that answer
as the **supported bounded answer**, narrow it, replace it with an inconclusive
answer, or find no supportable answer. For a substantive empirical question,
the reported answer is normally expressed by SUB-C. *Worked* describes the
example used to practise reading and is not another answer type.

The four functions appear in both layers, producing eight components.

| Code | Argument component | Core function |
|:-----|:-------------------|:--------------|
| **STAT-E** | Statistical-layer Evidence | Recorded observations plus information about recruitment, assignment or exposure, measurement, timing, implementation, exclusions, and completeness |
| **STAT-J** | Statistical-layer Justification | Reasons the study offers for using its evidence, comparison, analytical procedure, and uncertainty method to support the statistical move |
| **STAT-I** | Statistical-layer Inference | Estimand, result, uncertainty, and relevant checks brought together |
| **STAT-C** | Statistical-layer bounded Claim | Statistical proposition the study reports, together with its stated conditions and uncertainty |
| **SUB-E** | Substantive-layer Evidence | STAT-E situated as people or entities, variables, conditions, setting, implementation, and period |
| **SUB-J** | Substantive-layer Justification | Substantive reasons the study offers to augment and contextualize STAT-J through literature, measurement meaning, implementation, alternatives, and scope |
| **SUB-I** | Substantive-layer Inference | STAT-I interpreted through SUB-E and SUB-J in substantive units and context |
| **SUB-C** | Substantive-layer bounded Claim | Question-facing claim and stated bounds; it normally expresses the reported answer and may be inconclusive |

: *Argument components in the statistical and substantive layers.* {#tbl-argument-components tbl-colwidths="[13,30,57]"}

STAT-E is not merely a spreadsheet of numbers. It also includes the information
needed to understand how the numbers arose. Estimates, confidence intervals,
tests, and diagnostics are derived through analysis; they belong within
STAT-I rather than the empirical evidence itself.

STAT-J is not a ritual list of assumptions. It contains the study's stated or
implicit-but-traceable reasons connecting its evidence and estimand to the
comparison, analytical procedure, and uncertainty calculation. For a causal
question, this includes the study's reasons for treating its design and
assumptions as a basis for the intended alternative-condition comparison. Some
reasons may come from design, some from statistical conditions, and some from
substantive knowledge.

STAT-I is more than calculation. It connects the declared estimand to the
estimate, interval, checks, and analytic status. STAT-C records the statistical
proposition the study reports together with its stated conditions.

SUB-E does not introduce a second dataset. It identifies what the recorded
units, variables, conditions, setting, and period mean in the world named by
the question. SUB-J then augments and contextualizes STAT-J with the study's
substantive reasons concerning those meanings, implementation, alternatives,
and scope. Evaluation later asks whether those reasons are adequate.

SUB-I interprets the statistical result in meaningful units and considers
alternatives. SUB-C records the question-facing answer the study reports and
the limits it states. For a substantive question, the reported answer therefore
appears at SUB-C, not in a numerical result alone.

The mode-specific prompts introduced in @tbl-question-modes span the components
rather than belonging to STAT-J alone. Representation engages E and J,
predictive validation enters I, and avoiding causal overstatement also
constrains C.

**What function does a passage perform?** The special-practice study developed
in Section 3 illustrates the distinction:

| Information in the study | Contribution to the argument |
|:-------------------------|:-----------------------------|
| Participants were allocated to the two practice formats using a random sequence. | **STAT-E:** a reported feature of how the observations were produced. |
| The study invokes randomized assignment as the basis for comparing outcomes under the assigned formats. | **STAT-J:** the reason offered for making that comparison. |
| The study reports a mean difference and its uncertainty interval for the assigned-format contrast. | **STAT-I:** analytical results connected to the declared estimand and procedure. |

: *Different functions of information about the same comparison.* {#tbl-function-traces tbl-colwidths="[51,49]"}

These are function distinctions, not exclusive filing rules. The first
statement may also provide an implicit trace of STAT-J when the comparison and
the reason for invoking randomization are identifiable from the record. A
sentence such as “We used SPSS” names software but, by itself, gives no reason
why a particular comparison or uncertainty method supports the intended
inference. STAT-J therefore does not collect every sentence about methods.

The substantive distinction works similarly. An assessment guide's account of
the course topics covered helps identify **SUB-E**, what the scores are about.
The study's reasons for treating that coverage as evidence of delayed
course-material performance contribute to **SUB-J**. Interpreting the
estimated difference and its uncertainty in those terms belongs to **SUB-I**.
Background literature belongs in SUB-J when it supplies a reason for the
particular interpretation or scope, rather than merely discussing the topic.

### Logical support is not a workflow

Within either layer, the logical relationship is:

$$
E + J \longrightarrow I \longrightarrow C.
$$

Evidence and Justification work together to support Inference; Inference then
supports the bounded Claim. This is a **logical support order**, not the order
in which research activities must occur.

The arrows show the support relation the study presents. They are not an
evaluation verdict that the support is adequate.

Justification may be developed before evidence is collected, and later checks
may lead researchers to reconsider an analysis or narrow a claim. The logical
diagram asks what supports what in the completed argument. It does not claim
that research proceeds in a straight line from left to right. Section 5 returns
to the activities and revisions involved in constructing the argument.

This support order reinforces the earlier distinction between a result and an
answer. Even a correctly calculated estimate contributes only part of STAT-I;
it is not by itself STAT-C, SUB-I, or the final SUB-C.

Likewise, a *p*-value alone supplies neither direction, magnitude, nor
precision, and *nonsignificant* does not mean *no effect* [@wasserstein2016].
Chapter 3 develops that reading.

### How the two layers are linked

@fig-core-argument displays one argument as two linked rows. The visual
symmetry makes the common component pattern easy to see. The vertical
connections, however, do different work.

```{r}
#| label: fig-core-argument
#| fig-cap: "One argument with linked statistical and substantive layers."
#| fig-alt: "Two aligned rows labeled statistical layer and substantive layer. In each row, Evidence plus Justification lead to Inference and then bounded Claim. Downward arrows are labeled situates, augments and contextualizes, interprets, and bounds."
#| fig-width: 7.4
#| fig-height: 4.7
#| fig-pos: H

draw_argument()
```

| Shared function | Cross-layer relation | What the relation does |
|:----------------|:---------------------|:-----------------------|
| **Evidence** | STAT-E to SUB-E **situates** | Identifies what the units, variables, conditions, setting, and period represent while retaining how the evidence was produced |
| **Justification** | STAT-J to SUB-J **augments and contextualizes** | Retains statistical and design support while adding the study's substantive reasons about theory, measurement meaning, implementation, alternatives, benchmarks, and scope |
| **Inference** | STAT-I to SUB-I **interprets** | Expresses the estimate or prediction, uncertainty, checks, and sensitivity in substantive units and context |
| **Claim** | STAT-C to SUB-C **bounds** | Carries the statistical conditions and limitations into the study's substantive answer |

: *Role-specific transitions across the linked layers.* {#tbl-cross-layer-relations tbl-colwidths="[18,26,56]"}

The evidence transition **situates**. A row of scores becomes evidence about
students completing a particular assessment under specified conditions.

The justification transition **augments and contextualizes**. The substantive
layer retains the statistical and design warrant while adding the study's
reasons about measurement meaning, implementation, alternatives, benchmarks,
and scope. Whether those reasons suffice is evaluated later.

The inference transition **interprets**. A difference of 3.1 score points gains
meaning only on a defined scale and in relation to uncertainty, setting, and
possible alternatives.

The claim transition **bounds**. It carries the conditions and limits of the
statistical claim into the substantive answer rather than automatically turning
a participant-specific comparison into a claim about every student, every
implementation, or an unobserved construct called learning.

While reading, trace whether magnitude, uncertainty, conditions, analytic
status, and reservations remain visible across these connections. Section 4
calls the corresponding quality-evaluation question **Transition fidelity**.
Context may add meaning; it may not change what the statistical record says.

The downward arrows in @fig-core-argument show how corresponding components
are connected as readers trace the argument. They do not mean that
substantive reasoning begins only after statistical analysis. The substantive
question should shape the population, measures, design, and estimand
from the beginning. Assessment meaning, for example, belongs in prospective
planning as well as later interpretation.

Nor is each vertical relation an automatic one-to-one conversion. One
statistical component may need several kinds of substantive support, and one
piece of contextual information may contribute to several components. The
parallel layout is a memory aid for logical roles, not an algorithm that turns
numbers into meaning.

This tutorial synthesis is informed by Toulmin's warrants and qualified claims,
LeMire's statistical-to-substantive reasoning, and Kane's linked inferential
support, without reproducing any source as a one-to-one architecture or
presenting the statistical and substantive layers as separate arguments
[@toulmin2003; @lemire2010; @kane2013].

## The Special-Practice Study: A Complete Worked Reading

### The study record

Consider a hypothetical study in one first-year course. The instructors have
developed a four-session **Special Practice** format in which students work
through carefully sequenced questions before seeing full solutions. The
existing **Usual Practice** format uses the same subject matter and total study
time but presents worked examples before students complete related questions.
*Special* is only a neutral label for the hypothetical format; it does not
assume that the format is better.

One hundred and twenty consenting students who met the study's eligibility
requirements are randomly assigned, 60 to each format. The formats use matched
content and total study time, and every student completes the assigned
four-session format. Two weeks after the final session, every participant
completes the same 0--40 assessment under a common scoring procedure. The paper
identifies the course as the participant source but does not report the size of
the course roster, how many students were screened or invited, or how many were
lost before consent.

Assume that the accessible study record contains the paper, a protocol, an
assessment-and-scoring guide, and a short implementation appendix. Before
outcome access, the protocol labels the assignment question primary and
specifies a two-point difference as potentially important for this course. It
records the rationale: the assessment guide awards two points for one correctly
completed application task, and the course team expected adopting Special
Practice to require only modest additional preparation. This gives the
benchmark a prospective, course-specific documentary trace; evaluation still
asks whether the rationale is persuasive.

The study asks:

> Among the 120 participating students in this course, what difference in mean
> delayed-assessment score would result from assignment to Special Practice
> rather than Usual Practice?

The question is causal but deliberately narrow. It concerns assignment, these
participants, this course, this implementation, and a particular delayed
assessment. It does not ask whether every form of special practice improves
all aspects of learning for all students.

The student is the unit of assignment, measurement, and analysis. The direct
study population is the 120 consenting eligible participants. The estimand is
their mean effect of assignment to Special rather than Usual Practice. The
intended outcome is a bounded causal answer about these formats as implemented.
Naming those features before viewing the result keeps the answer obligation
stable.

The paper reports that the Special Practice group has a mean score of 27.1 with a standard
deviation of 6.0, while the Usual Practice group has a mean of 24.0 with a
standard deviation of 6.4. The estimated difference is therefore 3.1 points.
A two-sided 95% Welch two-sample *t* interval, rounded to one decimal place,
runs from 0.9 to 5.3 points.

The interval summarizes uncertainty about the mean assignment contrast under
the stated method; it is not a range of individual student scores. Later
chapters develop interval construction.

The full interval is above zero, but it contains values below and above the
prespecified two-point benchmark. Reading records both features: the authors
report a positive assignment effect, while also saying that the result does not
resolve whether the effect reaches their importance benchmark. Evaluation later
asks whether that reported answer and benchmark are adequately supported.

### Reading the statistical layer

**STAT-E: What was recorded, and how was it produced?**

The evidence includes more than 120 outcome scores. It includes the assignment
record, group sizes, delivery records, assessment timing and scoring, consent,
and complete follow-up. These details identify the comparison actually
created.

**STAT-J: What reasons does the study give for the statistical move?**

Individual random assignment supports comparing the assigned groups for these
participants. Matching total study time and subject matter helps define the
difference between formats. Common assessment procedures support score
comparability. Complete follow-up removes one important attrition concern.
The interval additionally requires an uncertainty procedure appropriate to the
randomized two-group comparison and the observed spread.

Randomization does not solve every problem. It does not establish that the
assessment measures every aspect of learning. It does not show that delivery
was identical to the written guide. It does not make the volunteers
representative of other courses.

Randomization creates comparability in expectation rather than guaranteeing
identical groups on every observed characteristic in this one realization.
The causal support comes from the assignment mechanism.

**STAT-I: How does the study connect the estimand, result, uncertainty, checks,
and analytic status?**

The observed assigned-group means are 27.1 and 24.0. Their estimated difference
is 3.1 points, with a 95% confidence interval from 0.9 to 5.3. The protocol
identifies this as the primary analysis and records the two-point benchmark.
The direction is positive throughout the interval, while the magnitudes
compatible with the data and method include values below and above that
benchmark.

Analysis computes the group summaries, difference, spread, and interval.
STAT-I connects those results to the declared estimand, uncertainty procedure,
relevant checks, and primary status; it is the inference rather than another
numerical output.

**STAT-C: What statistical claim does the study report, with what bounds?**

> For these 120 randomized participants, Special-minus-Usual assignment is
> estimated to increase the mean delayed-assessment score by 3.1 points, with a
> 95% confidence interval from 0.9 to 5.3 points; whether the contrast reaches
> the prespecified two-point importance benchmark remains unresolved.

This statement retains the assigned comparison, participant group, magnitude,
scale, and uncertainty. It remains a statistical claim because it does not yet
establish what the score means educationally or how far the finding should
travel.

### Reading the substantive layer

**SUB-E: What do the recorded observations represent?**

The observations concern consenting students in one first-year course,
assigned to two four-session practice formats, and assessed two weeks later on
a 0--40 course-linked measure. The result represents performance on that
assessment under this implementation. The study describes it as delayed
course-material performance rather than every possible form of learning.

**SUB-J: What substantive reasons augment and contextualize the statistical
justification?**

The study draws on the meaning of the two formats, the assessment blueprint,
common administration and scoring, implementation records, and prior
educational reasoning. These reasons augment and contextualize the statistical
warrant by connecting the assigned comparison to delayed course-material
performance under this implementation. The protocol also records the study's
course-specific rationale for treating two points as a threshold of potential
importance: the scoring guide awards two points for one correctly completed
application task, and the course team expected only modest additional
preparation. This does not make two points a universal threshold. Whether the
rationale is adequate is asked only in Section 4.

This justification also identifies limits. Consent and one course do not
support broad population generalization. A delayed assessment is evidence about
defined performance, not every possible form of learning. The assignment
effect is not automatically the effect of attending every session.

**SUB-I: What does the statistical result mean in context?**

The estimate is 3.1 points on a 0--40 assessment scale. That scale gives the
number a measurement context, but it does not determine practical importance.
Educational importance also depends on assessment content, score consequences,
feasible costs, and comparison with a benchmark specified independently of the
observed result.

**SUB-C: What answer does the study report, and what bounds does it state?**

> Under the implemented formats and common assessment conditions, assignment
> to Special rather than Usual Practice increased mean delayed-assessment
> performance for these 120 participating students by an estimated 3.1 points
> on the 0--40 scale, 95% CI [0.9, 5.3]. The result does not resolve whether the
> effect reaches the prespecified two-point importance benchmark. The study does not by itself establish
> the effect for other courses, other implementations, students who did not
> participate, or broader forms of learning.

### Reading the representation path

STAT-E also records how units move from the group named in the question to the
cases used in analysis:

```text
target population -> frame or source -> invited or eligible units
                  -> observed or responding units -> analytic sample
```

- The **target population** is the group to which the question and proposed
  answer refer: here, the 120 consenting eligible participants in this course.
- The **frame or source** is the list, setting, or record system from which units
  could enter: the paper identifies one course, but a complete roster or frame
  is **not located**.
- The **invited or eligible units** are those approached or meeting entry rules:
  the 120 participants met the rules, but the numbers screened, invited, or
  declining are **not located**.
- The **observed or responding units** are those from whom the relevant
  observations were obtained: all 120 randomized participants supplied delayed
  scores.
- The **analytic sample** is the set actually used for the result: the same 120
  participants because follow-up is complete.

Not every study has every step. Administrative data may begin with a source
system, and a census may not use a sampling frame. Mark a genuinely inapplicable
step **not applicable**; mark an applicable but unreported step **not located**.
Neither label authorizes the reader to invent what happened.

```{=latex}
\Needspace{0.70\textheight}
```

### The eight components together

| Component | Component content | Documentary trace(s) | Documentary status | Conflict flag |
|:----------|:------------------|:---------------------|:-------------------|:--------------|
| **STAT-E** | Assignment, delayed scores, scoring and timing, format delivery, consent, and complete follow-up for 120 participants | Paper Methods and Results; implementation appendix | Stated in the paper; delivery detail located in related materials | None located |
| **STAT-J** | The study's reasons for using the randomized assigned-group comparison, matched conditions, common scoring, and Welch interval | Paper Methods; protocol; implementation appendix | Stated in the paper; interval method located in related materials | None located |
| **STAT-I** | Primary estimand; means 27.1 and 24.0; estimate 3.1; 95% CI [0.9, 5.3]; two-point benchmark | Paper Results; protocol | Stated in the paper; status and benchmark located in related materials | None located |
| **STAT-C** | Reported participant-specific assignment contrast, interval, and unresolved benchmark conclusion | Paper Results | Stated in the paper | None located |
| **SUB-E** | Consenting eligible students in one course, two matched-time formats, and one delayed course assessment | Paper Methods; assessment guide | Stated in the paper; assessment detail located in related materials | None located |
| **SUB-J** | The study's reasons connecting format content, implementation, assessment meaning, and the course-specific two-point benchmark to delayed course-material performance | Introduction; protocol; implementation appendix; assessment guide | Stated in the paper; benchmark rationale located in related materials | None located |
| **SUB-I** | Interpretation of the estimate and interval on the 0--40 delayed-performance scale, including the benchmark | Discussion; protocol | Stated in the paper; benchmark located in related materials | None located |
| **SUB-C** | Reported causal answer limited to participants, implementation, measure, interval, and unresolved importance | Abstract and Discussion | Stated in the paper | None located |

: *Reading record for one special-practice argument.* {#tbl-special-components tbl-colwidths="[10,36,21,25,8]"}

The table separates a component's logical function and study-specific content
from the place where that content can be inspected. The same passage may trace
several components, and one component may be distributed across sources. The
four documentary statuses are **stated in the paper**, **located in related
materials**, **implicit but traceable**, and **not located**. A contradiction
across materials is recorded separately. These descriptions are not quality
scores. If a needed reason is not found, write **not located** rather than
*absent*; a premise supplied by the reader is only a proposed repair and does
not become support in the argument being read.

The same rule protects an unsuccessful justification from being misrecorded as
missing. If the paper explicitly gives a reason that does not support its
inference, its documentary status is still **stated in the paper**. Its
adequacy is a separate judgment under Section 4's **Justification adequacy**.

Recognizing a function requires statistical and substantive knowledge, which
the rest of this course develops. During reading, that knowledge can make a
conventional connection explicit when it is already traceable in the record.
It cannot supply an undocumented study-specific premise, such as successful
allocation concealment or complete attendance. Identifying the reason offered
and deciding whether it is appropriate or sufficient are different tasks. A
reader may notice a weakness immediately; keeping the two uses separate means
recording that judgment separately, not postponing thought until a second pass.

### Following the transitions

The component pairs are related through four distinct moves:

- **STAT-E to SUB-E situates:** “120 scores” becomes evidence about consenting
  students, two practice formats, one course, and one delayed assessment.
- **STAT-J to SUB-J augments and contextualizes:** the authors carry the
  statistical and design warrant forward while adding reasons about assessment meaning,
  implementation, prior reasoning, and scope.
- **STAT-I to SUB-I interprets:** the 3.1-point contrast and interval are
  expressed on the 0--40 scale and considered in relation to educational
  benchmarks.
- **STAT-C to SUB-C bounds:** the conditions and limitations of the
  participant-specific assignment contrast remain attached to the educational
  causal answer.

The reported substantive answer retains the interval and does not expand the
sample, while the authors treat educational importance as requiring contextual
support beyond the statistical result. Section 4 evaluates whether each
transition preserves both forms of discipline.

### Locating the supporting record

A reader should distinguish documentary presence from adequacy. Using the
general locations in @tbl-study-locations, the worked study might place its
question in an introduction or protocol; assignment, delivery, outcome, and
completion records in Methods and related materials; its estimate and interval
in Results; and the interpretation of meaning, alternatives, magnitude, and
scope in the Discussion. Support for one component may span several locations.

Use the four documentary statuses listed above and flag contradictions
separately. It is better to seek a **documentary trace** than “proof of a
component.” Presence in a document does not establish adequacy. Failure to
locate a trace does not establish that the underlying work never occurred. The
reader's question is narrower: what support can be found in the accessible
study record?

The completed reading therefore records the authors' reported causal answer
with its participant, implementation, measurement, uncertainty, and benchmark
bounds. It does not yet decide whether every part is warranted.

## Evaluating the Quality of a Question-Level Argument

Reading identifies the parts and traces their support. Quality evaluation asks
how well those parts work together. It is the secondary use after reading, not
another argument layer and not a numerical score.

### Six questions focus the evaluation

Six questions provide a common vocabulary for this chapter and the chapters
that follow. The short family labels help memory; they do not replace the six
question names.

| Family | Quality-evaluation question | Primary object | Beginner question |
|:-------|:----------------------------|:---------------|:------------------|
| **Alignment** | **Question alignment** | Relations among the question, estimand, design, evidence, analysis, and intended answer | Do they remain directed at the same question? |
| **Support** | **Evidence integrity** | Observations, contextual records, and how they were produced and represented | Are the records trustworthy, traceable, and represented appropriately? |
| **Support** | **Justification adequacy** | Design, measurement, model, and contextual reasons | Do the stated reasons warrant the proposed moves? |
| **Support** | **Inferential adequacy** | Inference in both layers: estimation or prediction, checking, uncertainty, sensitivity, and substantive interpretation | Do these support the statistical and substantive inferences? |
| **Translation** | **Transition fidelity** | Links between the statistical and substantive layers | Does each transition preserve meaning, uncertainty, design status, and scope? |
| **Translation** | **Claim calibration** | The bounded Claims, especially the final substantive answer | Is each claim no stronger or broader than its support? |

: *Six questions for evaluating one question-level argument.* {#tbl-basic-evaluation tbl-colwidths="[15,25,31,29]"}

The first question asks about alignment. The next three examine the support
within and around the layers. The final two examine how faithfully that support
travels into substantive meaning and the final answer. Inferential adequacy
examines reasoning within either layer; Transition fidelity examines what is
retained when moving between them; Claim calibration examines the strength and
scope of the resulting answer. A problem can have
downstream consequences under more than one question. The evaluator should
identify where the problem originates, then explain those consequences rather
than count the same weakness several times.

### Applying all six questions

The special-practice study provides a first complete application. The
judgments below concern the hypothetical record as stated; a real study may
provide insufficient information to judge some entries.

| Quality-evaluation question | Application to the special-practice argument |
|:----------------------------|:---------------------------------------------|
| **Question alignment** | **Adequate for this purpose.** The participant-specific causal question, randomized assignment, delayed assessment, mean-difference estimand, primary analysis, two-point benchmark, and reported answer address the same comparison. |
| **Evidence integrity** | **Adequate for this purpose.** Assignment, delivery, scoring, and complete outcomes are documented for 120 consenting students; the record does not establish representation beyond that group. |
| **Justification adequacy** | **Adequate for this purpose.** Random assignment supports the narrow assignment comparison, and common assessment procedures support score comparability. The protocol's assessment-task and modest additional-preparation rationale gives the two-point benchmark a course-specific basis, but not universal authority. Assessment meaning and one-course recruitment limit wider interpretation [@aeraApaNcme2014; @kane2013]. |
| **Inferential adequacy** | **Adequate for this purpose.** The primary estimate is accompanied by group spread and a Welch 95% interval; the interval supports direction but crosses the prespecified two-point benchmark. |
| **Transition fidelity** | **Adequate for this purpose.** The substantive interpretation retains the 0--40 delayed-assessment scale, participant group, randomized assignment, estimate, interval, setting, and implementation. |
| **Claim calibration** | **Adequate for this purpose.** The reported answer uses causal language about assignment for these participants, retains the interval and unresolved benchmark, and does not claim effects for every student, implementation, or form of learning. |

: *Quality evaluation of the special-practice argument.* {#tbl-special-evaluation tbl-colwidths="[27,73]"}

These judgments form a reasoned profile, not six independent grades. For
example, weak measurement support originates under Justification adequacy. It
can then damage Transition fidelity and require a narrower Claim. Recording
that path is more useful than subtracting three points for one problem.

Depending on the record and purpose, an evaluator may write **adequate for
this purpose**, **needs revision**, **insufficient information to judge**, or **not
applicable** beside an individual evaluation question. The third label marks
missing support without implying that the underlying research activity never
occurred. Reserve **not assessable from the accessible record** for the
whole-pair conclusion introduced below.

For this record, the **supported bounded answer** matches the reported answer:

> Under the implemented formats and common assessment conditions, assignment
> to Special rather than Usual Practice increased mean delayed-assessment
> performance for these 120 participants by an estimated 3.1 points on the
> 0--40 scale, 95% CI [0.9, 5.3]. Whether the effect reaches the prespecified
> two-point importance benchmark remains unresolved.

The whole-pair evaluation conclusion is therefore **supported as stated**. This
conclusion is separate from the six question-level judgments and from the
documentary statuses recorded during reading.

### Evidence integrity includes representation

Evidence integrity requires attention to whom or what the observations
represent, not only whether values were entered correctly. Section 3 established
that the study's course source is stated but its roster, invitation counts, and
pre-consent losses are not located. Evaluation therefore supports the narrow
participant-level estimand and scope but not an extension to all students in the course or
university. At each applicable transition, ask who or what was lost, added, or
excluded and how that change affects STAT-E and the scope of the answer.

Random sampling and random assignment answer different questions. Random
sampling primarily supports representation of a population. Random assignment
primarily supports a causal comparison of assigned conditions for the study
participants. A study can have one without the other. A narrow interval cannot
repair a poorly represented target population.

### Recurring constraints belong inside the answer

The six questions repeatedly direct attention to four practical constraints.
Readers should check:

- whether the final answer stays with the people or entities, place, period,
  and observations actually represented;
- whether its substantive words match what the recorded measures can mean;
- whether its verbs express only the interpretation supported by the design;
  and
- whether it retains relevant uncertainty, assumptions, alternatives, and
  unresolved reservations.

These constraints are not four additional questions and need not map
one-to-one onto the six quality-evaluation questions. A single measurement problem, for
example, may weaken justification, distort the substantive transition, and
force a narrower claim.

In the worked study, the point estimate is 3.1, but the interval ranges from
0.9 to 5.3. “Exactly 3.1 points” ignores uncertainty, while “at least five
points” conflicts with much of the interval. Calling the difference
educationally important would additionally require a defensible benchmark.
Reservations therefore belong in the reasoning and wording of the answer, not
in a detached limitations paragraph.

### Four evaluation conclusions

Quality evaluation should lead to an explicit outcome rather than a vague list of
strengths and weaknesses.

An answer is **supported as stated** when the question, linked layers, and
reported bounds are adequate for the question as stated. This does not mean the
study is perfect; it means the supported bounded answer matches what was
reported. A correctly reported inconclusive answer can therefore be supported
as stated.

An answer is **supported only after narrowing or revision** when a defensible
revision still directly answers the same question without changing its mode or
estimand. For example, random assignment among volunteers may support a causal
assignment effect for participants while the paper concludes that the format
works for all university students. Narrowing the population wording repairs the
answer without changing the question.

An answer is **not supported** when the accessible record supports no defensible
answer to the original question. Changing a causal question into an
associational one, or replacing its estimand, is not narrowing: it creates a
different or revised question--answer pair and workflow cycle.

An answer is **not assessable from the accessible record** when the record is
insufficient to decide what answer the support warrants. This conclusion does
not prove that the researchers never performed the work, and it should not be
silently converted into either adequacy or failure.

These outcomes keep criticism connected to the contract. The reader states not
only what is missing or weak, but how that condition changes the strongest
answer the accessible study can support.

::: {.callout-note title="Same result, different design"}
Suppose the means, 3.1-point difference, and interval are unchanged, but
students chose their practice format instead of being randomly assigned. Hold
the original causal question fixed. The observed comparison no longer supplies
the design warrant needed for a causal answer, so that causal answer is **not
supported** by the described record. Statistical adjustment alone would not
restore the missing warrant.

The same records may support a bounded association between chosen format and
delayed score for these participants. That statement answers a separately
stated **associational** question and therefore belongs to a different
question--answer pair; it is not a weaker answer to the original causal
question. A causal interpretation of an observational comparison would require
defensible design-and-assumption support beyond statistical adjustment alone.
:::

Authors have the first responsibility to evaluate and document the quality of
their argument. Collaborators and supervisors can test it while research is in
progress; reviewers, editors, and readers test the accessible record
independently. The six questions therefore support self-evaluation, but they
are not only a self-evaluation tool.

Chapter 1a retains these six questions and adds **Cross-question coherence**
for studies whose separately owed answers must be coordinated.

## Constructing the Argument: A Six-Stage Workflow

### Components and stages do different work

Argument components describe the logical roles that support an answer. A
**workflow stage** describes an activity used to construct or revise that
support. Components answer *what supports the claim?*; stages answer *how was
that support built?* They have a many-to-many relationship: one stage can
develop several components, and one component can depend on work performed
across several stages.

The six stages below organize much of the question-to-answer work within the
broader quantitative research process. They do not exhaust ethics, governance,
stakeholder engagement, project management, data stewardship, dissemination,
or archiving. The cycle adapts the iterative character of established
statistical-enquiry guidance [@wild1999; @asa2016gaise].

### Six stages in two phases

| Phase and stage | Main activity | Special-practice application |
|:----------------|:--------------|:--------------------------------|
| **Understand and plan -- 1. Review the problem, setting, and relevant literature** | Develop prospective statistical and substantive justification | Review prior evidence about special practice, delayed performance, assessment meaning, additional preparation, plausible alternatives, and defensible importance benchmarks |
| **Understand and plan -- 2. Frame the question and, where useful, a candidate answer** | Declare the population, comparison, outcome, claim mode, and provisional limits on the claim | Ask the participant-specific assignment question and frame a candidate answer about delayed-assessment performance, not learning everywhere |
| **Understand and plan -- 3. Align design, measurement, estimand, evidence-production or acquisition plan, and inferential plan** | Specify the design, measures, estimand, evidence-production or acquisition plan, any stopping rule, comparison, uncertainty, and checks | Plan random assignment of 120 participants, the common 0--40 delayed assessment, mean difference, Welch interval, course-specific two-point benchmark and rationale, delivery records, complete follow-up, and fixed sample size |
| **Produce and complete -- 4. Produce or acquire, and document, empirical and contextual evidence** | Produce the planned evidence or obtain and designate the question-specific analysis records; preserve the context needed for SUB-E | Retain consent, assignment, delivery, scoring, outcome, follow-up, deviation, and setting records for the planned 120 participants |
| **Produce and complete -- 5. Estimate, check, and analyse** | Develop STAT-I and STAT-C while checking the planned comparison and evidence record | Calculate the group means, 3.1-point contrast, 95% CI [0.9, 5.3], and relevant scoring, follow-up, and implementation checks |
| **Produce and complete -- 6. Contextualize, interpret, and complete the bounded answer** | Complete or refine SUB-J, SUB-I, and SUB-C without inventing post hoc reasons | Use the study's substantive reasons, interpret the result on the delayed-assessment scale, retain uncertainty and design meaning, and state the participant- and implementation-specific answer |

: *A six-stage construction cycle applied to the simple special-practice study.* {#tbl-basic-workflow tbl-colwidths="[28,31,41]"}

The first phase makes the intended argument coherent before evidence production.
The second phase produces the evidence, analyses it, and completes the bounded
answer. In the advanced tutorial, prospective alignment and cross-question
synthesis become separate stages because coordinated questions create further
obligations.

```{r}
#| label: fig-basic-workflow
#| fig-cap: "A six-stage, two-phase workflow cycle for constructing one question-level argument."
#| fig-alt: "Stages 1 through 3 move from left to right in the understand-and-plan phase. A solid downward arrow connects Stage 3 to Stage 4. Stages 4 through 6 then move from right to left in the produce-and-complete phase. Dashed return arrows remain within their respective phases; no return arrow crosses from Phase 2 to Phase 1."
#| fig-width: 7.4
#| fig-height: 4.25
#| fig-pos: H

draw_workflow()
```

### Iteration stays within a phase

Each argument is constructed within one documented workflow cycle. Iteration
is permitted only within Phase 1, among Stages 1--3, and within Phase 2, among
Stages 4--6. A Phase 1 return revises the prospective question, candidate
answer, or plan before evidence production or question-specific acquisition
begins. A Phase 2 return verifies or
reworks evidence documentation, checking, analysis, and interpretation without
pretending to redesign what has already happened. The handoff from Stage 3 to
Stage 4 is therefore one-way within a cycle.

The **Phase 1 anchor** comprises the question and its mode, design, measurement,
estimand, evidence-production or acquisition plan, stopping rule, and inferential
plan. A prospectively documented candidate answer may be contradicted by the
result without starting a new cycle. For an existing data source, Stage 4 means
obtaining or designating the question-specific analysis records or extract; it
need not be the first time anyone accessed the source. First access to relevant
outcomes may precede, coincide with, or follow that handoff. Whether a decision
was prospective depends on whether it preceded the evidence relevant to that
decision, not merely on its stage number.

Three cases keep revision honest:

1. **Normal completion:** the planned analysis produces a bounded answer,
   including an inconclusive answer when appropriate, and the cycle closes.
2. **Phase 2 repair:** an outcome-coding error may be checked and corrected, or
   a prespecified sensitivity analysis run, while the Phase 1 anchor remains
   unchanged.
3. **Additional or revised cycle:** a late additional question begins its own
   explicitly exploratory cycle. A newly preferred outcome selected after
   relevant evidence has been examined also creates a separate exploratory
   question--answer pair and cycle; it cannot silently replace the planned
   pair. Changing an estimand, design, or inferential commitment gives the
   original cycle an explicit outcome and begins a new or revised cycle.

If the planned measure cannot answer the original question, record that cycle's
outcome as **not answerable as planned** and begin the redesigned study as a new
or revised cycle.

A new cycle may reuse literature, measures, records, code, and lessons, but
their provenance and timing remain visible. Reuse does not automatically make a
new decision prospective, and the original cycle is not erased.

::: {.callout-note title="Three different tasks"}
**Reading** identifies and traces the argument the study presents. **Quality
evaluation** judges that already-read argument when evaluation is wanted. The
**construction workflow** shows how researchers build and revise its support.
None is another argument layer.
:::

Researchers can use the six quality-evaluation questions while constructing an argument,
but the questions are not additional workflow stages. They test the quality of
what a stage produces rather than describe another activity in the sequence.

## Summary and Reading Bridge

### Read first; evaluate when needed

Reading asks what argument the study presents. Identify the question, mode,
estimand, answer obligation, and reported answer; identify the content performing
the eight component functions; locate documentary traces and record their
statuses and any conflicts; and follow logical support within and across the
linked layers. Do not fill a documentary gap with a premise of your own.

Quality evaluation is the separate secondary use. When it is needed, apply
the six questions to the argument already read, state the supported bounded
answer, and record one of four conclusions: **supported as stated**, **supported
only after narrowing or revision**, **not supported**, or **not assessable from
the accessible record**.

The construction workflow is the auxiliary use for conducting research. The
six-stage workflow helps researchers build and revise the support; it is not a
reading procedure and does not recover the historical order merely from the
finished paper.

### Compact glossary

**Question--answer contract.** The study's obligation to connect each
quantitative question to a traceable reported outcome; the obligation may be
met well, met by an inconclusive answer, or not met.

**Question--answer pair.** One research question and one separately reportable
question-facing answer obligation.

**Question mode.** The kind of answer sought: descriptive, associational,
predictive, or causal. Mode is not the same as analytic status.

**Estimand and estimate.** The estimand specifies what the study seeks to
learn; the estimate is the numerical result obtained from observed evidence
for it. A question may require one or more estimands.

**Argument and layer.** An argument is the complete question-level reasoning
for one pair; a layer is a coordinated statistical or substantive part of that
one argument.

**Argument component.** Evidence, Justification, Inference, or bounded Claim as
a logical function within a layer.

**Bounded Claim and answers.** C records the study's claim and stated bounds.
Reading records the reported answer; evaluation determines the supported
bounded answer.

**Documentary trace.** An inspectable passage, table, protocol, code file,
record, or related material through which component content can be located.

**Documentary status.** Stated in the paper, located in related materials,
implicit but traceable, or not located; conflicts are flagged separately.

**Representation path.** The applicable transitions from target population
through a source or frame, eligible or invited units, observed or responding
units, and the analytic sample.

**Quality-evaluation question.** One of six named questions used to evaluate
Question alignment, Evidence integrity, Justification adequacy, Inferential
adequacy, Transition fidelity, or Claim calibration.

### Reading bridge

The argument-based approach is a tutorial synthesis rather than a new theory of
statistical inference. Several sources deepen different parts of it:

- Toulmin's *The Uses of Argument* explains why evidence needs warrants and why
  claims retain qualifications and reservations [@toulmin2003].
- LeMire offers a statistical precedent connecting parameter-level reasoning
  with substantive interpretation [@lemire2010].
- Kane and the *Standards for Educational and Psychological Testing* develop
  argument-based reasoning for proposed score interpretations and uses
  [@kane2013; @aeraApaNcme2014].
- Lundberg, Johnson, and Stewart explain why defining an estimand helps
  connect theory, evidence, design-based support, and estimation [@lundberg2021].
- *Introduction to Modern Statistics* gives an accessible introduction to
  cases, variables, populations, sampling, assignment, and study design
  [@cetinkaya2024].
- Reporting standards show why a paper and its related materials should make
  design, analysis, interpretation, and limitations inspectable
  [@appelbaum2018].

Shadish, Cook, and Campbell's established validity terminology distinguishes
statistical conclusion, internal, construct, and external validity, especially
for causal studies [@shadish2002]. Those types illuminate several concerns in
this chapter, but the six questions are organized for reading a question-level
argument. They are not renamed validity types or a competing validity taxonomy.

The next tutorial, *Extending the Argument-Based Approach to Quantitative
Studies*, extends the six-stage cycle to seven stages for coordinated questions
and develops the same six quality-evaluation questions plus Cross-question coherence,
design complications, and
regression-based methods. It is optional: Chapters 2--14 assume only the basic
concepts taught in this chapter. Any later chapter that needs a method-specific
refinement introduces it locally rather than sending readers backward to the
advanced extension.

```{=latex}
\Needspace{8\baselineskip}
\phantomsection
```

## Exercises {.unnumbered}

These exercises form a modular bank. Each prompt contains the information
needed for its task. Select the exercises that suit the class; the labels
indicate the kind of thinking required rather than a fixed order.

```{=latex}
\Needspace{10\baselineskip}
\phantomsection
```

<!-- prompt-id: CH01-E01 -->
### Exercise 1: Find the question--answer pair (Foundational) {.unnumbered #ex-ch01-01}

A report states:

> Researchers studied 100 students in one university course. They asked
> whether assignment to weekly planning messages rather than usual course
> messages would change the mean final-assessment score. The planning-message
> group averaged 78 points and the usual-message group averaged 73 points on a
> 100-point assessment. The report concluded, “Weekly planning messages
> improve achievement for every university student.” The excerpt does not say
> how students were assigned or report an interval for the difference.

Complete the following tasks:

1. Identify the research question, numerical result, and reported answer.
2. Classify the question as descriptive, associational, predictive, or causal.
3. State the estimand in ordinary language. Remember that the observed
   five-point difference is an estimate, not the estimand itself.
4. Name one design fact needed before treating the difference as an effect of
   the messages.
5. Identify one way the reported answer is broader than the study record.

<!-- END EX01-01 PROMPT -->

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

<!-- prompt-id: CH01-E02 -->
### Exercise 2: Distinguish component functions (Foundational) {.unnumbered #ex-ch01-02}

The following statements come from the special-practice study. For each
statement, first choose the **statistical (STAT)** or **substantive (SUB)**
layer. Then choose its main function: **Evidence (E), Justification (J),
Inference (I),** or **bounded Claim (C)**. Use each of the eight codes once.

1. The records contain the assignment, delivery, scoring, and delayed scores
   for all 120 participants.
2. Random assignment and common scoring support comparing the two assigned
   groups for these participants.
3. For the protocol's primary participant-level assignment estimand, the group
   means are 27.1 and 24.0; their estimated difference is 3.1 points, 95% CI
   [0.9, 5.3], with the interval above zero but crossing the two-point benchmark.
4. For the randomized participants, the estimated Special-minus-Usual mean
   score difference is 3.1 points, 95% CI [0.9, 5.3].
5. The records concern consenting students in one course who completed two
   defined practice formats and a delayed 0--40 assessment.
6. The assessment blueprint and implementation records support interpreting
   the scores as delayed performance on the assessed course material and give
   the two-point benchmark a course-specific rationale, not a universal
   importance threshold.
7. In course terms, the estimated assignment effect is a 3.1-point increase in
   mean delayed-assessment performance, with compatible values from 0.9 to 5.3
   points under the stated method.
8. Under this implementation, assignment to Special Practice increased mean
   delayed-assessment performance for these participants; effects on other
   students or broader learning are not established.

For one statement, briefly explain why its wording signals the function you
selected. Then consider Statement 2: assume its randomization procedure appears
in the protocol but not in the paper. Assign the documentary status and explain
why the protocol is a trace rather than the component itself.

<!-- END EX01-02 PROMPT -->

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

<!-- prompt-id: CH01-E03 -->
### Exercise 3: Complete the linked layers (Integrative) {.unnumbered #ex-ch01-03}

Suppose a randomized study conducted during one professional-development
workshop reports the following:

- 80 participants were assigned equally to two feedback formats;
- at the end of the workshop, all completed the same immediate 20-point quiz;
- the new-format mean was 15.2 and the usual-format mean was 14.1;
- the estimated difference was 1.1 points, 95% CI [-0.2, 2.4];
- assignment and scoring procedures were documented;
- the planned interval procedure was used, with no consequential problem found
  in the reported scoring, implementation, or analysis checks; and
- a 1.5-point difference was prospectively identified as potentially
  important.

Using only the information provided, complete the linked-layer map:

| Code | What belongs here? |
|:-----|:-------------------|
| STAT-E | |
| STAT-J | |
| STAT-I | |
| STAT-C | |
| SUB-E | |
| SUB-J | |
| SUB-I | |
| SUB-C | |

What directions and importance conclusions remain compatible with the interval
from -0.2 to 2.4? Write STAT-C without saying either “the format worked” or
“there was no effect.” In SUB-C, preserve the interval and distinguish immediate
quiz performance from durable professional learning. Then apply **Inferential
adequacy** to the reported inference and **Claim calibration** to your drafted
claims, giving one short judgment for each.

<!-- END EX01-03 PROMPT -->

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

<!-- prompt-id: CH01-E04 -->
### Exercise 4: Evaluate and repair a conclusion (Integrative) {.unnumbered #ex-ch01-04}

In one first-year course, 120 consenting students were randomly assigned to
Special or Usual Practice. Everyone completed the same delayed 0--40
assessment. The Special-minus-Usual difference in group means was 3.1 points,
95% CI [0.9, 5.3]. Before outcome access, the protocol specified a two-point
difference as potentially important because the scoring guide awarded two
points for one correctly completed application task and adoption was expected
to require only modest additional preparation. The abstract concludes:

> Special Practice is an effective way to improve learning. Students gained
> 3.1 points, proving that the method should replace usual practice in
> universities.

1. For each quality-evaluation question, record one short judgment about the
   conclusion.

| Quality-evaluation question | What do you notice? |
|:----------------------------|:--------------------|
| Question alignment | |
| Evidence integrity | |
| Justification adequacy | |
| Inferential adequacy | |
| Transition fidelity | |
| Claim calibration | |

2. Identify one problem where it first arises and one consequence it creates
   later in the argument.
3. Write a replacement conclusion of no more than two sentences. Retain the
   supported causal comparison, but do not turn a difference between group
   means into the gain experienced by every student.
4. State the supported bounded answer and choose one overall evaluation
   conclusion: **supported as stated**, **supported only after narrowing or
   revision**, **not supported**, or **not assessable from the accessible
   record**. Explain your choice in one sentence.

<!-- END EX01-04 PROMPT -->

```{=latex}
\Needspace{6\baselineskip}
\phantomsection
```

<!-- prompt-id: CH01-E05 -->
### Exercise 5: Map and revise a workflow cycle (Integrative) {.unnumbered #ex-ch01-05}

**Part A: Map the work.** Assign each activity to Stage 1--6 and identify its
phase.

A. Estimate the group difference and interval, then check the evidence record.
B. Review the practical problem, setting, and relevant prior studies.
C. Interpret the result in course context and complete the bounded answer.
D. Frame the participant-specific question and a candidate answer.
E. Produce or acquire the evidence and preserve assignment, delivery, scoring,
   outcome, and follow-up records.
F. Align random assignment, the delayed assessment, the mean-difference estimand,
   evidence-production plan and fixed sample size, and the uncertainty plan.

**Part B: Classify six cases.** For each case, choose **normal completion**,
**iteration within Phase 1**, **iteration within Phase 2**, **an additional
exploratory cycle**, or **a revised cycle**, and give one reason.

1. Before recruitment, a pilot reveals that the practice instructions are
   ambiguous.
2. During analysis, the team discovers an outcome-coding error and verifies the
   original records before recalculating the estimate.
3. After examining the outcomes, the team adds a new attendance question while
   retaining the original assignment question.
4. After examining the outcomes, the team prefers a different outcome and
   wants it to replace the planned primary outcome.
5. The planned analysis is completed as specified, but its interval leaves both
   benefit and harm compatible with the record; the team reports an
   inconclusive answer.
6. During Phase 2, the team discovers that the planned delayed assessment was
   administered immediately instead. It cannot answer the original
   delayed-performance question, so the team plans a redesigned study.

For Cases 3, 4, and 6, state what the additional or revised cycle must make
visible. Give one example of material that may be reused and one kind of status
that does not transfer automatically.

<!-- END EX01-05 PROMPT -->

```{=latex}
\Needspace{8\baselineskip}
\phantomsection
```

## References {.unnumbered}

\begingroup
\footnotesize

::: {#refs}
:::

\endgroup
