---
title: "Estimates, Uncertainty, and Statistical Tests"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{12\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \usepackage{xcolor}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false

if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
n_special <- 60; n_usual <- 60
mean_special <- 27.1; mean_usual <- 24.0
sd_special <- 6.0; sd_usual <- 6.4
estimate <- mean_special - mean_usual
standard_error <- sqrt(sd_special^2 / n_special + sd_usual^2 / n_usual)
welch_df <- (sd_special^2 / n_special + sd_usual^2 / n_usual)^2 /
  ((sd_special^2 / n_special)^2 / (n_special - 1) +
   (sd_usual^2 / n_usual)^2 / (n_usual - 1))
test_statistic <- estimate / standard_error
p_value <- 2 * pt(-abs(test_statistic), df = welch_df)
confidence_interval <- estimate + c(-1, 1) * qt(.975, df = welch_df) * standard_error
ink <- "#243447"; blue <- "#2B6F8A"; blue_light <- "#DDECF2"
gold <- "#B7791F"; red <- "#A23B3B"; grey <- "#6B7280"
```

{{< pagebreak >}}

::: {.callout-note title="Chapter overview"}
Chapter 2 introduced population and sample, parameter and statistic,
sampling distribution and standard error. We now use those ideas to
specify what a study seeks to estimate and to quantify uncertainty about
the estimate obtained. A continuing comparison leads from an estimation
rule to its standard error, a confidence interval, and a statistical
test. We transfer the reasoning to scores measured twice on the same
students and to categorical counts, with school comparisons developed in
the reference appendix. We then bring magnitude, practical importance, error
rates, and several comparisons together in a qualified conclusion.
:::

A first-year course wanted to know whether a new practice format helps
students remember what they have learned. Its instructors recruited 120
consenting students and randomly assigned 60 to Special Practice and 60
to Usual Practice. Both formats covered the same content over four
sessions with equal total study time. Two weeks later, all 120 completed
the same assessment, scored from 0 to 40. Special Practice averaged 27.1
points and Usual Practice 24.0, a difference of 3.1000 points. This study is
invented for the chapter.

The difference is easy to calculate. Deciding what it tells us takes
more work. A different assignment could have produced a different
difference, even with the same participants. We need to understand that
variation to judge how precisely this study estimates its intended
comparison. We also need the outcome's scale and the course's purpose to
judge whether a difference of this size matters. The calculations below
develop those two judgments together.

## From a Sample to an Estimate {#sec-ch03-estimation}

### The study and its estimated difference

The study asks about the **mean effect of assignment to Special rather
than Usual Practice on the delayed assessment among these 120
participants**. Each part matters. Assignment is the intervention being
compared; attending every session would be a different comparison. A mean
effect summarizes the participants and allows individual responses to
differ. The two-week assessment measures a specified outcome at a
specified time. An effect on durable learning or among students elsewhere
would require additional information.

The course team also chose a practical benchmark before seeing the
outcomes. It treated two points as the smallest average difference that
would change its recommendation: the marking scheme awards two points
for one completed application task, and the new format adds little
preparation time. This gives the benchmark a concrete, course-specific
rationale. We will return to it after calculating uncertainty.

| Item | Special Practice | Usual Practice |
|:--|--:|--:|
| Assigned students | 60 | 60 |
| Delayed assessments observed | 60 | 60 |
| Delayed-assessment mean | 27.1 | 24.0 |
| Delayed-assessment SD | 6.0 | 6.4 |

: Observed summaries for the invented special-practice study. {#tbl-ch03-observed tbl-colwidths="[50,25,25]"}

The standard deviations describe differences among students' scores
within each group. Scores vary by roughly six points around their group
means, while the two means differ by 3.1000 points. That comparison of
spreads is useful description, but the uncertainty of a *difference
between means* is a separate quantity. Its calculation begins by naming
the target and the rule used to estimate it.

### Estimand, estimator, estimation, and estimate

Chapter 2 distinguished the population and sample, introduced parameter
and statistic, and explained sampling distributions and standard error.
We now use those ideas to distinguish the quantity sought, the
estimation rule, and the estimate obtained.

An **estimand** is the precisely specified quantity the question seeks.
For this study, it is the mean assignment effect just described. To make
the target explicit, let $Y_i(1)$ be student $i$'s delayed score if
assigned to Special Practice and $Y_i(0)$ the score if assigned to Usual
Practice. These are the student's two **potential outcomes**. Only the
one corresponding to the actual assignment is observed; Chapter 1 called
the unobserved alternative the counterfactual outcome. The target is

$$
\Delta\mu=\frac{1}{120}\sum_{i=1}^{120}\{Y_i(1)-Y_i(0)\}.
$$

Inside the braces is one student's assignment effect. The sum adds all
120 effects, and division by 120 takes their mean. Written this way, $\Delta\mu$ is a mean
difference over these 120 participants, not a population quantity. The notation allows different students to
have different effects. We assume the two formats have a consistent
meaning and one student's assignment does not change another student's
outcome in a relevant way.

An **estimator** is a rule for using observed data to approximate this
target. Here the rule subtracts the Usual assigned-group mean from the
Special assigned-group mean. **Estimation** is the process of applying
and checking that rule and characterizing its uncertainty. The
**estimate** is the numerical value obtained:

$$
\widehat{\Delta\mu}=\bar y_1-\bar y_0=27.1-24.0=3.1000\text{ points}.
$$

The bar denotes a group mean, and the hat marks an estimate. The rule
would stay the same under another admissible assignment, although the
observed outcomes and the two group means could change. Reversing the
subtraction gives $-3.1000$; it expresses the reverse comparison using the
same information.

The observed difference exactly describes the two realized means. That
description is known. The participant mean assignment effect remains
unknown because no student reveals both potential outcomes. This is why
an exact piece of arithmetic can still be an uncertain estimate.

| Object | Working definition | In this study | Common reading error |
|:--|:--|:--|:--|
| Estimand | The precisely specified quantity the question seeks | Mean delayed-score difference caused by Special rather than Usual assignment among the 120 randomized participants | Naming only "the treatment effect" |
| Estimator | The rule that would be applied in each admissible repetition | Special-group mean minus Usual-group mean | Treating the rule as the observed result |
| Estimation | Applying and checking the estimator for the specified estimand and characterizing the estimate and its uncertainty | Apply the mean-difference rule, verify its inputs and conditions, and quantify uncertainty | Treating estimation as another number |
| Estimate | The number the estimator produced in this study | $27.1-24.0=3.1000$ points | Reporting a number without units or direction |
| Parameter | The fixed feature named by the estimand under the chosen population or process interpretation | The participant-specific mean assignment contrast | Assuming it is the sample estimate or a universal truth |
| Statistic | A numerical function of observed data | A group mean, the 3.1000-point estimate, or the 2.7372 test statistic | Assuming every statistic estimates the same thing |
| Standard error | The modeled spread of an estimator across the declared repetitions | 1.1325 points for the mean-difference estimator | Calling it variation among students |

: Inferential objects in the special-practice study. {#tbl-ch03-inferential-objects tbl-colwidths="[13,29,33,25]"}

The table distinguishes objects by their function. A statistic is any
numerical function of the observed data; a group mean, an estimated SE,
and a test statistic all qualify. A parameter is fixed within the
specified population or process interpretation. Calling it fixed does
not make it universal: changing the participant set, outcome, time, or
condition comparison may change the estimand and its parameter.

### Name the repetition and working model

Uncertainty needs a description of what could have varied. In a sampling
question, different people might have been selected from a population.
In the special-practice study, the primary target concerns these same
120 participants. The relevant repetition assigns exactly 60 of them to
each format again. Their potential outcomes remain fixed; the assigned
groups and observed outcomes can change.

Random assignment makes the difference of assigned-group means useful
for estimating the participant effect. It does not turn these consenting
students into a representative sample of all first-year students.
Inference about a wider population would need a stated model or sampling
justification and an explicit extension of the target.

The numerical analysis uses the familiar **Welch procedure** as an
approximate working calibration [@welch1947]. It allows different
observed group variances. Its formulas also occur in independent-sample
comparisons, but complete assignment of 60 out of 120 students does not
make the two assigned-group means independent. Sections 2 and 3 explain
what the formula estimates here and what its approximation requires.
Appendix A gives the design-based variance identity and the distinction
between an average-effect null and a sharp null.

### Expectation, bias, and precision

Imagine applying the estimator over all repetitions allowed by the
design or model. Its expectation, in Chapter 2's sense, is the probability-weighted
average of the values it would produce across those repetitions. If an estimator averages 3.4 points when
the target is 3.0, its bias is $3.4-3.0=0.4000$ points. This describes the
rule's systematic displacement, not the error of one particular result.

Under complete random assignment and the stated potential-outcome
conditions, each assigned-group mean is centered on the corresponding
participant mean potential outcome. Their difference is therefore
unbiased for the participant mean assignment effect:

$$
\operatorname{Bias}(\bar Y_1-\bar Y_0)
=E_{\mathrm{assign}}(\bar Y_1-\bar Y_0)-\Delta\mu=0.
$$

The equality does not say that this estimate, 3.1000, equals the unknown
effect. **Precision** concerns how tightly the estimator's values cluster
across assignments:

$$
SE(\bar Y_1-\bar Y_0)=SD(\bar Y_1-\bar Y_0\text{ across possible assignments}).
$$

An unbiased rule can be imprecise; a tightly concentrated rule can be
biased. A third useful consideration is how its behavior changes under
unusual observations or departures from its conditions. A small SE under
an unsuitable model can give unwarranted confidence. Appendix A develops
the bias--variance trade-off using mean squared error.

## Using Sampling Variation to Quantify Uncertainty {#sec-ch03-se}

### From the visible sampling pattern to an SE formula

Chapter 2 Section 4 generated 5,000 independent waiting-time samples at
each of three sample sizes. The waiting-time model had mean and SD both
equal to 10 minutes. The sample means varied less when a sample contained
more observations. Increasing the number of simulated repetitions made
their distribution easier to see; increasing the size of each sample
changed the distribution's spread.

We can now explain that spread algebraically. Suppose $Y_1,\ldots,Y_n$
are independent observations from the same distribution, with finite
variance $\sigma^2$. The sum has variance $n\sigma^2$, because
independent variances add. The mean divides that sum by $n$; multiplying
a variable by $1/n$ multiplies its variance by $1/n^2$. Thus

$$
\operatorname{Var}(\bar Y)=\frac{n\sigma^2}{n^2}=\frac{\sigma^2}{n},
\qquad SE(\bar Y)=\frac{\sigma}{\sqrt n},
\qquad \widehat{SE}(\bar Y)=\frac{s}{\sqrt n}.
$$

The last expression estimates the unknown population SD with the sample
SD. The estimated SE is calculated from one sample; the SE it estimates
describes the mean across repetitions. Both are standard deviations,
but of different distributions from the individual waiting times.

For the waiting-time model, the theoretical SEs are
$10/\sqrt5=4.4721$, $10/\sqrt{30}=1.8257$, and
$10/\sqrt{100}=1.0000$ minutes. Chapter 2's empirical values, about 4.54,
1.79, and 1.01, approximate those targets. Simulation noise explains
small discrepancies. The square-root relationship also explains why
halving an SE usually requires four times as many independent
observations under the same model.

### Uncertainty for a difference in means

For two independent samples, uncertainty comes from both means. Their
variances add when one mean is subtracted from the other:

$$
\operatorname{Var}(\bar Y_1-\bar Y_0)
=\frac{\sigma_1^2}{n_1}+\frac{\sigma_0^2}{n_0},
\qquad
\widehat{SE}(\bar y_1-\bar y_0)
=\sqrt{\frac{s_1^2}{n_1}+\frac{s_0^2}{n_0}}.
$$

The minus sign changes the direction of a difference, not the sign of a
variance contribution. Independence removes the covariance term; the
next subsection shows what happens when that condition fails.

Return now to the fixed-participant assignment study. Its estimated
variance has the same numerical form, with a different design-based
interpretation. Substitution gives

$$
\begin{aligned}
\widehat{SE}(\bar y_1-\bar y_0)
&=\sqrt{\frac{6.0^2}{60}+\frac{6.4^2}{60}}
=\sqrt{\frac{36}{60}+\frac{40.96}{60}}\\
&=\sqrt{0.6000+0.682666\ldots}=1.1325\text{ points}.
\end{aligned}
$$

Each variance is in squared assessment points. Division by a group size
gives its contribution to uncertainty, and the final square root
returns to points. Keep unrounded quantities while calculating and
report 1.1325 points in the chapter's result.

Across assignments of these participants, the exact variance contains a
subtraction for variability in individual effects. We cannot recover
that variability, because no student reveals both potential outcomes.
The squared SE above omits this subtraction and estimates the remaining
terms. It estimates an upper bound on assignment variance *on average*,
with equality when individual effects are constant. This does not mean
every realized SE is too large. Appendix A shows the identity that
supports this statement [@bai2024].

The working reference also needs adequate information in each group and
no dominating observations. Random assignment alone does not make the
Welch reference exact. @fig-repeated-estimates illustrates the spread
using 500 Normal draws centered at 3.1 with SE 1.1325. The center is set
to the observed value to make the display concrete; it is not a claim
that the unknown estimand equals 3.1. These are simulated values of the
same estimator, not additional study evidence or the enumerated
randomization distribution.

```{r}
#| label: fig-repeated-estimates
#| fig-cap: "An illustrative simulation of repeated Special-minus-Usual estimates. The estimand and estimator remain fixed; realized estimates vary, and their modeled spread is the standard error."
#| fig-alt: "Histogram of 500 possible mean-difference estimates centered near 3.1 assessment points. A vertical line marks the fixed estimand. A horizontal arrow extends one standard error to each side of it, spanning two standard errors in total."
#| fig-width: 7.2
#| fig-height: 3.5
#| fig-pos: H

set.seed(30301)
repeated_estimates <- rnorm(500, mean = estimate, sd = standard_error)
old <- par(mar = c(4.1, 4.3, 1.0, 0.8))
hist(repeated_estimates, breaks = 24, probability = TRUE, col = blue_light, border = "white",
     xlab = "Special-minus-Usual estimate (assessment points)", ylab = "Density", main = "",
     col.axis = ink, col.lab = ink)
abline(v = estimate, col = gold, lwd = 2.4)
arrows(estimate - standard_error, 0.025, estimate + standard_error, 0.025,
       code = 3, angle = 90, length = 0.06, col = blue, lwd = 2)
text(estimate, 0.355, "fixed estimand", col = gold, font = 2, cex = 0.8)
text(estimate, 0.045, "fixed estimand +/- 1 SE", col = blue, cex = 0.78, pos = 3)
par(old)
```

### Pairing, clustering, and sampling without replacement

Dependence changes a variance calculation. For any two quantities with
finite variances,

$$
\operatorname{Var}(U-V)=\operatorname{Var}(U)+\operatorname{Var}(V)
-2\operatorname{Cov}(U,V).
$$

If paired measurements tend to be high or low together, subtracting
within a pair removes some shared variation. In a separate illustration,
suppose both measurements have SD 4, their covariance is 12, and there
are 25 independent pairs. Let $d$ be the within-pair difference. Then

$$
\begin{aligned}
\operatorname{Var}(d)&=4^2+4^2-2(12)=8.0000,\\
\widehat{SE}(\bar d)&=\sqrt{\frac{8}{25}}=0.5657,\\
\text{ignoring the pairing: }&\sqrt{\frac{16+16}{25}}=1.1314.
\end{aligned}
$$

The reduction comes from the positive covariance, not from a change in
the number of observations. Actual paired-data analysis computes each
difference and uses the SD of those differences. Negative covariance
would change the result in the opposite direction.

An invented paired-school feedback illustration provides a different
form of dependence. It measured 480 students in 24 classes within
12 schools, assigning one
class in each school to each condition. Its uncertainty follows the 12
within-school contrasts, which [Appendix B](#app-ch03-school-pairs)
works through. Counting 480
distinct students does not establish 480 independent assignments.

How much information data provide depends on the estimand, dependence,
weighting, and design. Repeated rows for the same person can add less
information than equally many independent observations. In simple random
sampling without replacement, observing a large fraction of a finite
population leaves less uncertainty about its mean.
[Appendix A](#app-ch03-dependence) gives the finite-population correction and explains a
precision-equivalent effective sample size under a stated comparison
model.

### Concentration, shape, and the conditions behind them

Two ideas explain different features of Chapter 2's repeated means. The
**law of large numbers** describes concentration: under its conditions,
a sample mean gets close to the population mean as sample size grows.
The **central limit theorem** describes shape: a suitably centered and
scaled sample mean approaches a Normal distribution under conditions
such as independent, identically distributed observations with finite
variance. The original waiting times can remain strongly skewed.

These results concern increasing sample size, not repeatedly simulating
the same small sample. They also require conditions. Strong dependence,
extreme tails, or an observation that dominates the mean can make a
familiar Normal approximation poor. There is no universal sample-size
cutoff that settles the issue for every distribution and design.

A running mean makes the concentration claim visible. In
@fig-running-mean, each point averages all the exponential waiting times
accumulated so far. The population mean stays at ten minutes while the
sample mean changes as observations arrive.

```{r}
#| label: fig-running-mean
#| fig-pos: H
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "A running mean from 400 independent exponential waiting times with population mean ten minutes."
#| fig-alt: "The running sample mean falls near five minutes early in the sequence, then approaches the horizontal ten-minute population-mean line and ends at 9.7207 minutes. Individual waiting times retain their skewed distribution."
#| fig-width: 6.4
#| fig-height: 3.0

local({
  set.seed(60204)
  waiting <- rexp(400, rate = 1 / 10)
  running_mean <- cumsum(waiting) / seq_along(waiting)
  saved_par <- par(mar = c(4.1, 4.3, .7, .7))
  on.exit(par(saved_par))
  plot(seq_along(waiting), running_mean, type = "l", lwd = 1.5,
       col = "#2B6F8A", ylim = range(c(running_mean, 10)),
       xlab = "Number of draws included",
       ylab = "Running mean (minutes)",
       col.axis = "#243447", col.lab = "#243447")
  abline(h = 10, lty = 2, lwd = 1.6, col = "#A23B3B")
  legend("bottomright", c("Running mean", "Population mean = 10"),
         col = c("#2B6F8A", "#A23B3B"), lty = c(1, 2),
         lwd = c(1.5, 1.6), bty = "n", cex = .85)
})
```

The running mean finishes at 9.7207 minutes. Its early shortfall does
not change the population mean, and the next observation can move the
average toward or away from ten. The individual waiting times remain
skewed. This is one accumulating sequence; to examine the CLT's shape
claim, we need a distribution of means across repeated samples of the
same size, as in Chapter 2.

The SE formula answers how wide a sampling distribution is under a
model. An approximate shape then helps choose an interval or test. Keep
those two jobs distinct: a correct spread calculation alone supplies
no tail probability. [Appendix A](#app-ch03-limits) gives the convergence
notation and fuller conditions.

## Confidence Intervals {#sec-ch03-intervals}

### Estimate, SE, critical value, and interval width

A confidence interval combines an estimate with a multiple of its SE.
For the symmetric procedures used here,

$$
\text{estimate}\pm\text{critical value}\times\text{standard error}.
$$

The SE supplies a scale. A **critical value** says how far to extend in
SE units to obtain the intended coverage under a reference distribution.
Chapter 2 introduced the standard Normal distribution, centered at zero
with SD one. Its .975 quantile is about 1.9600: .025 of the probability
lies beyond that point, and another .025 lies below $-1.9600$.

A **t distribution** is also symmetric around zero but has heavier
tails, especially with few degrees of freedom. In a one-mean procedure,
these tails account for estimating the spread. As degrees of freedom
increase, the t reference approaches the standard Normal. An estimated
spread does not make every statistic exactly t-distributed: the exact
one-mean result requires independent Normal sampling, while Welch's
reference is approximate.

The 95% reference values below leave .025 in each tail. At low degrees
of freedom, estimating spread produces a larger t critical value; the
larger-df reference is closer to Normal.

| Reference | 95% critical value | Role in the examples |
|:--|--:|:--|
| Standard Normal | 1.9600 | Mean with a population SD known independently of the sample, under a suitable Normal model |
| t with 24 df | 2.0639 | One mean with SD estimated from 25 independent Normal observations |
| t with 117.5119 df | 1.9804 | Approximate Welch reference for the supplied Special Practice summaries |

: Reference choice and critical values. {#tbl-ch03-critical-values tbl-colwidths="[25,20,55]"}

The procedure and its conditions determine the reference. Choosing the
smaller critical value merely to narrow an interval would change the
procedure without justification.

Chapter 2 used a **z score**, $z_i=(x_i-\mu)/\sigma$, to locate
one observation in population SD units. Sample standardization,
$(x_i-\bar x)/s$, describes an observation relative to its sample.
Neither operation makes the original distribution Normal. For a mean,
the quantity being standardized is instead its estimator, so the
denominator is an SE:

$$
z=\frac{\bar x-\mu_0}{\sigma/\sqrt n},\qquad
t=\frac{\bar x-\mu_0}{s/\sqrt n},\qquad df=n-1.
$$

The **one-sample z test** uses a population SD known independently of
the sample. The **one-sample t test** estimates that SD with $s$.
Its exact t reference requires independent observations from a Normal
population; suitable distributional conditions and enough information
can support an approximation more broadly. A large sample does not turn
an unknown SD into a known one or remove dependence.

@fig-ch03-z-t compares references for the same standardized statistic.
With a statistic of 2.0000, the two-sided probabilities are .0455 for
Normal, .0766 for t with 9 df, and .0482 for t with 99 df. The change
comes from the reference tails. Changing a study's sample size can also
change its SE and observed statistic, which is a separate comparison.

```{r}
#| label: fig-ch03-z-t
#| fig-pos: H
#| echo: false
#| results: hide
#| fig-cap: "Normal and t references at the same location and scale of the test statistic. The heavier t tails are most visible with few degrees of freedom."
#| fig-alt: "Overlaid standard Normal and t curves with 9 and 99 degrees of freedom. Vertical lines mark minus and plus 2; the t curve with 9 degrees of freedom has the largest probability beyond them."
#| fig-width: 6.8
#| fig-height: 3.5
z_grid <- seq(-4, 4, length.out = 801)
plot(z_grid, dnorm(z_grid), type = "l", lwd = 2,
     xlab = "Standardized test statistic", ylab = "Density")
lines(z_grid, dt(z_grid, 9), col = "#A23B3B", lwd = 2)
lines(z_grid, dt(z_grid, 99), col = "#2B6F8A", lwd = 2, lty = 2)
abline(v = c(-2, 2), lty = 3)
legend("topright", c("Normal", "t: df = 9", "t: df = 99"),
       col = c("black", "#A23B3B", "#2B6F8A"),
       lty = c(1, 1, 2), lwd = 2, bty = "n")
c(Normal = 2 * pnorm(-2), t9 = 2 * pt(-2, 9),
  t99 = 2 * pt(-2, 99))
```

For the special-practice procedure, the Welch degrees of freedom are
117.5119 and the 95% critical value is 1.9804. The fractional degrees
of freedom describe this reference approximation; they are not a count
of students. Appendix A provides the full formula. The interval is

$$
\begin{aligned}
(\bar y_1-\bar y_0)\pm t_{.975,\,117.5119}\,\widehat{SE}(\bar y_1-\bar y_0)
&\approx3.1000\pm1.9804(1.1325)\\
&\approx3.1000\pm2.2429\\
&\approx[0.8571,\ 5.3429]\text{ points}.
\end{aligned}
$$

The endpoints retain the Special-minus-Usual
direction and assessment-point units. They allow positive effects near
one point and effects above five points. The data therefore give a
useful estimate without pinpointing its magnitude.

For a separate illustration, suppose observations are independently
sampled from the same Normal population, with $n=25$,
$\bar x=52$, and $s=5$. To estimate its population mean, use
$\widehat{SE}(\bar X)=5/\sqrt{25}=1.0000$ point. With 24 degrees of
freedom and the supplied critical value 2.064,

$$
52\pm2.064(1.0000)=[49.9360,\ 54.0640].
$$

The mechanics are the same, but the estimand and repetition have changed
from a participant assignment effect to a population mean under sampling.

### Coverage belongs to the repeated procedure

A 95% confidence procedure aims to cover the fixed estimand in 95% of
its repeated applications under its conditions. Across repetitions, the
observations, estimate, and endpoints vary; the target stays fixed.

@fig-interval-coverage shows 40 applications of a simplified Normal
procedure with a known SE. The estimate is drawn from that model, and
the interval adds and subtracts $1.96SE$. Most intervals cross the
fixed target line; a few miss. A batch of 40 need not contain exactly
38 successes even when the procedure has 95% coverage. The figure's
coverage describes this simplified model.

```{r}
#| label: fig-interval-coverage
#| fig-cap: "Forty intervals from one calibrated illustrative procedure. The estimand remains fixed while estimates and intervals vary; occasional misses are compatible with 95% long-run coverage."
#| fig-alt: "Forty horizontal confidence intervals are plotted against a vertical estimand line at 3.1000 points. Most blue intervals cross the estimand; a small number of red intervals do not."
#| fig-width: 7.2
#| fig-height: 3.3
#| fig-pos: H

set.seed(30306)
coverage_estimate <- rnorm(40, mean = estimate, sd = standard_error)
coverage_lower <- coverage_estimate - qnorm(.975) * standard_error
coverage_upper <- coverage_estimate + qnorm(.975) * standard_error
coverage_hit <- coverage_lower <= estimate & estimate <= coverage_upper
old <- par(mar = c(4.6, 4.6, 1.0, 0.8), mgp = c(2.8, 0.8, 0))
plot(NA, xlim = range(c(coverage_lower, coverage_upper, estimate)), ylim = c(0.5, 40.5),
     xlab = "Interval for the mean assignment contrast", ylab = "Repeated study",
     yaxt = "n", col.axis = ink, col.lab = ink)
abline(v = estimate, col = gold, lwd = 2)
for (i in seq_len(40)) {
  colour <- if (coverage_hit[i]) blue else red
  segments(coverage_lower[i], i, coverage_upper[i], i, col = colour, lwd = 1.4)
  points(coverage_estimate[i], i, pch = 16, col = colour, cex = 0.55)
}
axis(2, at = c(1, 10, 20, 30, 40), las = 1, col.axis = ink)
legend("topright", legend = c("covers estimand", "misses estimand"),
       col = c(blue, red), lwd = 2, bty = "n", cex = 0.78)
par(old)
```

Once the actual interval has been calculated, the frequentist coverage
statement does not assign a 95% probability to its fixed estimand.
Neither does it say that 95% of students score between 0.8571 and 5.3429, or
that 95% of repeated estimates fall inside this particular interval.
It describes the success rate of the *interval-producing procedure*.

For interpretation, read the endpoints as values compatible with the
observations under the procedure at the stated level. This does not
make all values inside equally plausible or values just outside
impossible. The interval expresses a range of uncertainty, with no
natural cliff at its rounded endpoints.

### What a narrower interval can and cannot resolve

Holding the model and data fixed, a higher confidence level needs a
wider interval. More outcome variability generally widens it, while
more relevant information narrows it. Changes in the design or variance
estimator can also change its width. A useful target width depends on
which differences would alter a substantive conclusion.

For the Special Practice comparison, changing only the confidence level
gives the following intervals. Both use the same 3.1000-point estimate,
unrounded SE, degrees of freedom, and observations.

| Confidence level | Critical value | Margin | Interval in points | Full width |
|:--|--:|--:|:--|--:|
| 90% | 1.6579 | 1.8777 | [1.2223, 4.9777] | 3.7554 |
| 95% | 1.9804 | 2.2429 | [0.8571, 5.3429] | 4.4857 |

: Changing confidence while keeping the estimate and SE fixed. {#tbl-ch03-confidence-levels tbl-colwidths="[18,18,16,30,18]"}

Predict the changes before comparing the rows: the smaller critical value
narrows the 90% interval without adding observations or reducing its SE.
The narrower interval has a lower coverage target. Both intervals lie
above zero and span the two-point benchmark.

The special-practice interval lies above zero but crosses the two-point
benchmark. It can support a positive mean assignment effect under the
procedure while leaving practical importance unresolved. We develop
that joint reading in Section 5.

::: {.callout-note title="An interval describes particular uncertainty"}
The interval quantifies the uncertainty represented by its procedure.
A narrower interval can coexist with systematic selection, poor
measurement, unmodeled dependence, unavailable outcomes, or an
unsupported generalization. Those problems need information about
the study and suitable analysis, in addition to precision.
:::

For example, measuring the wrong outcome very consistently can yield
a small SE for the wrong target. Adding many pupils from a few
jointly assigned classes does not reproduce the information from
the same number of independent assignments. Precision helps only
when the procedure remains aligned with the question and design.

## Statistical Tests and Worked Comparisons {#sec-ch03-tests}

### Hypotheses, test statistics, reference distributions, and tail areas

A statistical hypothesis makes a proposition about a target. Begin
with the separate one-mean example: $H_0:\mu=50$ says the population
mean equals 50 points, and $H_A:\mu\ne50$ allows either direction.
The sample has $n=25$, $\bar x=52$ and $s=5$. Its statistic is

$$
t=\frac{52-50}{5/\sqrt{25}}=2.0000,\quad df=24,\quad
p=2P(T_{24}\ge2)=.0569.
$$

The numerator is the observed departure from the null mean. The
denominator is the estimated SE of the mean, so the departure is two
estimated SEs. The t reference with 24 df supplies the probability in
both tails. Equivalently, subtracting 50 from every score creates a
centered variable whose null mean is zero; the question stays the same.

For two independent populations, write $H_0:\Delta\mu=0$, where
$\Delta\mu=\mu_1-\mu_0$ and the subscripts identify the two groups.
For Special Practice, $\Delta\mu$ is the fixed participants' mean
assignment effect defined in Section 1.2, and its null preserves that
target:

$$
H_0:\Delta\mu=0\qquad\text{versus}\qquad H_A:\Delta\mu\ne0.
$$

The null says the participant mean assignment effect is zero; the
alternative allows either a positive or a negative effect. A test
is the procedure for assessing this proposition, while a test result
is the numerical output obtained from the data. A common statistic is

$$
\text{test statistic}=
\frac{\text{estimate}-\text{null value}}
{\text{standard error under the test procedure}}.
$$

The numerator measures departure in outcome units. The denominator
expresses how large that departure is relative to the procedure's
uncertainty. Here

$$
t=\frac{\widehat{\Delta\mu}-0}{\widehat{SE}(\bar y_1-\bar y_0)}
\approx2.7372,\qquad
p=2P(T_{117.5119}\ge|t_{\mathrm{obs}}|)\approx.0072.
$$

The estimate is about 2.7372 estimated SEs above zero. The symbol
$T_{117.5119}$ denotes a variable with the reference t distribution.
Using the unrounded statistic and degrees of freedom, its upper-tail
probability is .0036. The
two-sided test adds the equally extreme negative tail, giving .0072.
Software or a reference table supplies the area under the approximate
Welch working procedure.

A **p-value** is this conditional tail probability. If the zero-effect
null and the procedure's conditions hold, a result at least this
incompatible would occur with probability about .0072. The calculation
starts by assuming the null for its reference; it does not calculate
the probability that the null itself is true [@wasserstein2016].

A **significance level**, $\alpha$, is an advance threshold for a
decision rule. A level-.05 rule rejects when its calibrated p-value is
below .05; .0072 meets that rule. Calling the result statistically
significant records this comparison. It supplies neither the size of
the effect nor a study-specific probability of having made an error.

The focal test is two-sided because either direction counts. A
one-sided test needs an alternative whose direction was chosen before
the result was inspected. A large effect in the opposite direction
does not support that alternative. Selecting whichever tail favors
the observed estimate changes the testing procedure and its error
calibration.

Different t procedures use different units and variance calculations.
Their shared name does not make paired and independent data
interchangeable.

| Procedure | Data and null | SE and reference |
|:--|:--|:--|
| One-sample t | One quantitative sample; $H_0:\mu=\mu_{\mathrm{null}}$ | $s/\sqrt n$; $df=n-1$ |
| Welch independent-samples t | Two independent samples; $H_0:\mu_1-\mu_0=0$ | $\sqrt{s_1^2/n_1+s_0^2/n_0}$; Welch--Satterthwaite df |
| Pooled independent-samples t | Two independent samples with a common population variance; same mean null | $s_p\sqrt{1/n_1+1/n_0}$; $df=n_1+n_0-2$ |
| Paired-samples t | Correctly matched observations, independent pairs; $H_0:\mu_D=0$ | $s_D/\sqrt n$ on pair differences; $df=n-1$ |

: Named t tests and the observations that carry information. {#tbl-ch03-t-procedures tbl-colwidths="[24,40,36]"}

The **pooled t test** combines the two within-group variance estimates:

$$
s_p^2=\frac{(n_1-1)s_1^2+(n_0-1)s_0^2}{n_1+n_0-2}.
$$

It is the **equal-variance t test**: it assumes a common population
variance, uses $n_1+n_0-2$ df, and when that assumption holds it is the
more powerful of the two procedures, with an exact t reference and a
slightly narrower interval for the same data. Welch's **unequal-variances
t test** allows the group variances to differ; its estimated df are
smaller, so it gives up a little power in exchange for protection when the
variances are unequal, and it is the safer default when a common variance
cannot be justified from the design. For the equal-size Special Practice
groups the two statistics coincide at 2.7372; only the df differ (118
versus 117.5119), and both p-values display as .0072. Equal sample sizes
do not establish equal variances, and choosing a procedure solely from a
preliminary variance-test p-value obscures the design and model; state
the choice with its reason. Pairing comes from how observations were obtained;
its t procedure analyzes within-pair differences. Chapter 4's term
*pairwise comparisons* refers to several group contrasts, a different
use of the word.

| Mistranslation | Why it fails | Repair |
|:--|:--|:--|
| "There is a 0.72% chance the null is true." | The p-value conditions on the null; it does not assign probability to it | Under the zero-contrast null and procedure, results at least this incompatible have probability .0072 |
| "There is a 0.72% chance the result happened by chance." | Chance is not a complete model; the result already happened | Name the null, design, statistic, and reference process |
| "The study has a 99.28% chance of replicating." | Replication depends on the effect, design, sample, and analysis | Report the current estimate and uncertainty; discuss replication separately |
| "The effect is large or important because $p=.0072$." | A p-value is not an effect-size or importance measure | Interpret 3.1000 points and its interval against a defensible benchmark |
| "The study's false-positive probability is 0.72%." | A long-run error rule and a data-dependent p-value differ | State any prespecified decision rule; avoid a posterior error claim |
| "The p-value keeps its usual meaning after this outcome or model was selected from inspected results." | The selection rule is part of the procedure and can alter calibration | Report the family of analyses and whether the comparison was chosen in advance; interpret the selected result accordingly |

: Six mistranslations of the displayed $p=.0072$ and their repairs. {#tbl-ch03-p-mistranslations tbl-colwidths="[30,32,38]"}

The repairs in the table retain what the test can answer while directing
other questions to the information they require. Replication needs a
new-study model; practical importance needs an outcome scale and a
benchmark. Reporting the estimate and interval alongside the p-value
makes these distinctions easier to maintain.

### A matched interval and test

A two-sided test at the 5% level built from the same procedure as a
95% interval agrees with that interval about zero. Here the estimator,
SE, reference distribution, sidedness, and comparison all agree.
The test rejects zero precisely when the interval excludes it.
@fig-test-interval places the tail area and interval beside one another.

```{r}
#| label: fig-test-interval
#| fig-cap: "The two-sided test and 95% interval for the same contrast, built from the same procedure. The null value, standard error, and model remain aligned."
#| fig-alt: "Two panels. The first shows a t reference distribution with red tails beyond plus and minus 2.7372. The second shows a 3.1000-point estimate and a 95 percent interval from 0.8571 to 5.3429; the zero null value lies outside."
#| fig-width: 7.2
#| fig-height: 3.2
#| fig-pos: H

old <- par(mfrow = c(1, 2), mar = c(4.0, 3.8, 2.2, 0.7))
x <- seq(-4.2, 4.2, length.out = 800); y <- dt(x, df = welch_df)
plot(x, y, type = "l", lwd = 2, col = ink, xlab = "Test statistic under the null",
     ylab = "Density", main = "Null-model reference")
right <- x[x >= abs(test_statistic)]; left <- x[x <= -abs(test_statistic)]
polygon(c(right, rev(right)), c(dt(right, welch_df), rep(0, length(right))), col = "#F4C7C7", border = NA)
polygon(c(left, rev(left)), c(dt(left, welch_df), rep(0, length(left))), col = "#F4C7C7", border = NA)
lines(x, y, lwd = 2, col = ink)
abline(v = c(-test_statistic, test_statistic), col = red, lwd = 1.7)
text(test_statistic, 0.12, paste("observed", formatC(test_statistic, format = "f", digits = 4)), pos = 2, col = red, cex = 0.76)
plot(NA, xlim = c(-1, 7), ylim = c(0.5, 1.5), yaxt = "n",
     xlab = "Special-minus-Usual contrast (points)", ylab = "", main = "Estimate and interval")
abline(v = 0, col = grey, lty = 2)
segments(confidence_interval[1], 1, confidence_interval[2], 1, col = blue, lwd = 4)
points(estimate, 1, pch = 21, bg = gold, col = ink, cex = 1.5)
text(estimate, 1.20, formatC(estimate, format = "f", digits = 4), col = ink, cex = 0.8)
text(0, 0.72, "null = 0", col = grey, cex = 0.75)
par(old)
```

Algebra explains the agreement: zero lies outside
$\widehat{\Delta\mu}\pm t^*\widehat{SE}$ when
$|\widehat{\Delta\mu}|/\widehat{SE}>t^*$. A different SE, tail rule,
reference, or multiplicity adjustment can break the correspondence.
Two mechanically aligned calculations can also share an inadequate
model. Agreement checks their construction; it does not validate their
conditions.

For the separate one-mean example, testing $\mu=50$ gives
$t=(52-50)/1=2.0000$. This is smaller than the supplied two-sided critical
value 2.064, and the interval [49.9360, 54.0640] includes 50. The
procedure therefore fails to reject 50 at the 5% level. That outcome
does not establish that the mean equals 50: other values remain
compatible as well.

\Needspace{25\baselineskip}

### Paired-samples and independent-samples t tests {#sec-ch03-paired}

Eight invented students completed the same-score-scale assessment
before and after a short practice session. These are separate teaching
records from Special Practice. Define each student's difference as
$d_i=\mathrm{after}_i-\mathrm{before}_i$, preserving the person and
subtraction direction before calculating a summary.

| Student | Before | After | Difference |
|:--|--:|--:|--:|
| 1 | 50 | 49 | -1 |
| 2 | 54 | 54 | 0 |
| 3 | 48 | 49 | 1 |
| 4 | 60 | 62 | 2 |
| 5 | 55 | 57 | 2 |
| 6 | 58 | 61 | 3 |
| 7 | 62 | 66 | 4 |
| 8 | 53 | 58 | 5 |

: Invented paired student scores; these are not Special Practice records. {#tbl-ch03-student-pairs tbl-colwidths="[25,25,25,25]"}

The differences sum to 16, so their mean is $16/8=2.0000$ points.
Their squared deviations from two sum to 28. The SD therefore uses
seven degrees of freedom:

$$
\bar d=2.0000,\qquad s_d=\sqrt{28/7}=2.0000,\qquad
\widehat{SE}(\bar d)=2/\sqrt8=0.7071.
$$

Under an independent-pair model, the population mean change is
$\mu_D$. Testing a zero mean change and constructing the corresponding
95% interval gives

$$
H_0:\mu_D=0,\qquad
t=\frac{2.0000-0}{2/\sqrt8}=2.8284,\quad df=7,\quad p=.0255,
$$

$$
\bar d\pm t_{.975,7}\frac{s_d}{\sqrt8}
\approx2.0000\pm2.3646(0.7071)
\approx[0.3280,\ 3.6720]\text{ points}.
$$

The calculation retains unrounded values. An exact t reference requires
independent Normal pair differences; with other distributions its
approximation needs justification. The result describes mean change
under the stated sampling model. Without an untreated comparison,
the session cannot be separated from time trends, repeated testing or
other concurrent changes.

The variance identity in Section 2.3 explains why pairing matters:

$$
\operatorname{Var}(Y_{\mathrm{after}}-Y_{\mathrm{before}})
=\operatorname{Var}(Y_{\mathrm{after}})
+\operatorname{Var}(Y_{\mathrm{before}})
-2\operatorname{Cov}(Y_{\mathrm{after}},Y_{\mathrm{before}}).
$$

Treating the eight before scores and eight after scores as independent
groups would impose an inappropriate independence structure on these
paired records and omit their within-student covariance. Its p-value
cannot justify that change. For genuinely independent groups, Welch or the pooled
procedure instead follows the group variances and their conditions.

The school-pair example in [Appendix B](#app-ch03-school-pairs)
extends this logic to class means. It prints all class means and works
through the calculation: twelve within-school contrasts, rather than
480 individual scores, carry the assignment uncertainty. Chapter 4
recalls the purpose of blocking.

### Transfer from means to categorical counts {#sec-categorical-comparison}

The same question--estimate--reference sequence can assess a categorical
comparison. Consider an invented teaching model with 100 independently
sampled learners. Participation format and completion are recorded for
each learner. The format was observed rather than randomly assigned,
so the question concerns association.

| Participation format | Completed | Did not complete | Total |
|:--|--:|--:|--:|
| Online | 32 | 18 | 50 |
| In person | 18 | 32 | 50 |
| Total | 50 | 50 | 100 |

: Invented counts for the first Pearson chi-square calculation. {#tbl-ch03-counts tbl-colwidths="[34,22,28,16]"}

Among the 50 Online learners, 32 completed: $32/50=.6400$. Among the
50 In-person learners, 18 completed: $18/50=.3600$. The observed
Online-minus-In-person difference is .2800, or 28.0000 percentage points.
Each denominator is its own format's total.

The null says format and completion are independent in the stipulated
population or process. Under that restriction both formats have a common
completion probability. Estimating it from all records gives
$50/100=.5000$. A cell's **expected count** equals its row total times
column total divided by the total number of records:

$$
E_{rc}=\frac{(\text{row total})(\text{column total})}{n}
=\frac{50(50)}{100}=25.
$$

Every cell has expected count 25 here. This pattern preserves the
observed margins while imposing the null relationship. An expected
count is a model-based reference; in other tables it can be fractional.

| Cell | Observed $O$ | Expected $E$ | $(O-E)^2/E$ |
|:--|--:|--:|--:|
| Online, completed | 32 | 25 | 1.9600 |
| Online, did not complete | 18 | 25 | 1.9600 |
| In person, completed | 18 | 25 | 1.9600 |
| In person, did not complete | 32 | 25 | 1.9600 |

: Expected counts and contributions for the invented table. {#tbl-ch03-expected-counts tbl-colwidths="[40,18,18,24]"}

Pearson's statistic adds squared observed-minus-expected departures,
each scaled by its expected count:

$$
X^2=\sum_{r,c}\frac{(O_{rc}-E_{rc})^2}{E_{rc}}
=4(1.9600)=7.8400.
$$

With the margins accounted for, specifying one cell in a two-by-two
table fixes the remaining three. Thus $df=(2-1)(2-1)=1$. The
nonnegative squared discrepancies become larger as observations depart
from the null pattern, so the test uses the upper tail of a chi-square
reference. Its probability is $p=.0051$.

Under the stipulated model, the table gives evidence against independence.
The proportions give the direction and magnitude that the squared
statistic discards: completion is 28.0000 percentage points more frequent
in the Online group. This observed association does not establish that
assigning a format would change completion.

The approximation requires independent observational units, mutually
exclusive cell coding and adequate expected counts. Expected counts of
25 do not themselves establish independent sampling or a causal design.
Chapter 6 reproduces the uncorrected Pearson calculation in R; software
defaults can use a continuity correction for two-by-two tables, producing
a related but different calculation. Exercise 7 supplies a separate
200-record table for transferring this reasoning.

The UCI intention-by-pass table in Chapter 2 has a different source,
population and denominator. Chapter 6's categorical reference develops
its inference and coding sensitivity. The simple invented table here
keeps the first expected-count calculation completely visible.

## Magnitude, Uncertainty, and Several Results {#sec-ch03-interpretation}

### Magnitude and practical importance

The special-practice estimate, interval, test, and benchmark answer
different parts of the substantive question. The raw difference is
3.1000 points. Its approximate interval [0.8571, 5.3429] lies above zero and
agrees with the two-sided $p=.0072$ result. It also spans the
course-specific two-point importance criterion.

Consequently, a positive mean assignment effect is supported under
the procedure, while its practical importance remains unresolved.
Effects below the criterion and effects clearly above it are both
compatible with the data. The estimate exceeding two points does not
settle this issue. Nor would an interval crossing zero automatically
mean that all effects of practical interest had been ruled out.

A **standardized mean difference** expresses a raw difference relative
to a named SD. One familiar denominator pools within-group variation:

$$
s_{\mathrm{pooled}}=
\sqrt{\frac{(n_1-1)s_1^2+(n_0-1)s_0^2}{n_1+n_0-2}},
\qquad d=\frac{\bar x_1-\bar x_0}{s_{\mathrm{pooled}}}.
$$

Using the exact supplied practice summaries gives
$s_{\mathrm{pooled}}=6.2032$ and $d=.4997$. The mean difference is
about half of this particular within-group SD. Hedges' correction
adjusts $d$ for small-sample bias under its reference model; here its
approximation gives $g=.4966$. [Appendix B](#app-ch03-effect-scales)
develops that correction and the regression connections.

Changing the denominator changes the comparison's scale. It does not
broaden the participant scope or supply a universal importance threshold.
For the current course decision, assessment points and the two-point
benchmark remain the most direct terms. Chapter 4 develops effect-size
interpretation across group comparisons.

### Errors, power, and precision planning

A decision rule can make two kinds of error. A **Type I error** rejects
a true null. A calibrated level-.05 rule limits this probability under
its null conditions. A **Type II error** fails to reject under a
specified alternative. If that probability is $\beta$, the rule's
**power** is $1-\beta$, its probability of rejecting under that
alternative.

The possible decisions and model states form the following table.
Use *do not reject*, because the decision does not reveal that the null
is true.

| Decision | Null is true | Specified alternative is true |
|:--|:--|:--|
| Do not reject $H_0$ | Correct nonrejection; probability $1-\alpha$ for an exactly calibrated continuous test | Type II error; probability $\beta$ at this alternative |
| Reject $H_0$ | Type I error; probability $\alpha$ for that calibrated test | Correct rejection; power $1-\beta$ at this alternative |

: Decisions and error probabilities under specified states of the model. {#tbl-ch03-errors tbl-colwidths="[22,39,39]"}

For an approximately calibrated or conservative procedure, nominal
$\alpha$ need not equal the actual null rejection probability.
$\beta$ changes across alternative effects. Both error probabilities
condition on a model state; neither is the probability that the decision
in this completed study is wrong.

Power .80 at a particular effect means that 80% of repetitions would
reject under that effect, design, and test. It leaves a .2000 chance of
failing to reject there. It is neither an 80% probability that a
completed study is correct nor a prediction that its exact finding
will recur. The effect magnitude, outcome variability, sample size,
dependence, sidedness, and decision rule all affect power. A study
has a power curve across alternative effects, not one context-free
power value. [Appendix B](#app-ch03-thousand-studies) uses a bounded
1,000-study illustration to distinguish the false-positive fraction
among rejections from the Type I error probability.

Precision planning asks for an interval narrow enough for the intended
use. Power planning asks for a stated chance of rejection at a meaningful
alternative under a specified procedure. Both require inputs about anticipated variation,
measurement quality, allocation, and dependence. For a class-assigned
study, additional independently assigned classes can supply information
that adding more students within existing classes cannot replace.
A sample-size justification may use desired precision, power,
feasibility, or a combination; its stated purpose explains what the
study is planned to make sufficiently informative.

For a separate independent two-group study, suppose each group has anticipated SD 6.2 points and
we want a Normal-approximation 95% interval with half-width 2. With
equal group sizes $n$, the half-width is
$h=z^*\sigma\sqrt{2/n}$. Solving for $n$ gives

$$
n_{\mathrm{per\ group}}\approx2\left(\frac{z^*\sigma}{h}\right)^2
=2\left(\frac{1.96\times6.2}{2}\right)^2
=2(6.0760)^2=73.8356.
$$

Round upward to 74 observed outcomes per group. The calculation uses a
stipulated SD and independent-sampling approximation; it does not
redesign the fixed 120-participant assignment study.

The same plan illustrates why precision and power answer different
questions. At 74 per group, its working SE is
$6.2\sqrt{2/74}=1.0193$. Using the full Normal critical value for a
two-sided .05 test of zero, the estimate must exceed 1.9977 or fall
below $-1.9977$ to reject. Both tails contribute to the power calculation.

@fig-ch03-errors-power, adapted from @biau2010, keeps that rejection
rule fixed under two possible truths. At a true difference of zero, the
two shaded rejection tails total $\alpha=.0500$. At a true difference of
three points, the middle nonrejection area is $\beta=.1627$ and both
rejection tails together give power .8373.

```{r}
#| label: fig-ch03-errors-power
#| fig-pos: H
#| echo: false
#| results: hide
#| fig-cap: "A fixed decision rule under two possible population differences. At zero, rejection has probability alpha; at a true three-point difference, the middle region has probability beta and both outer regions together give power. Both panels use the same rejection boundaries."
#| fig-alt: "Two Normal density panels with standard error 1.0193 and rejection boundaries at minus and plus 1.9977. Null tails total .05; under a true difference of three, the middle area is .1627 and both rejection tails total .8373."
#| fig-width: 7.2
#| fig-height: 3.6
plan_se <- 6.2 * sqrt(2 / 74)
cutoff <- qnorm(.975) * plan_se
grid <- seq(-5, 8, length.out = 1001)
old <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (delta in c(0, 3)) {
  dens <- dnorm(grid, mean = delta, sd = plan_se)
  plot(grid, dens, type = "l", lwd = 2, ylim = c(0, .42),
       xlab = "Estimated mean difference", ylab = "Density",
       main = paste("True difference =", delta))
  for (side in c(-1, 1)) {
    at <- if (side == -1) grid < -cutoff else grid > cutoff
    polygon(c(grid[at], rev(grid[at])),
            c(dens[at], rep(0, sum(at))),
            col = "#BBDDEB", border = NA)
  }
  if (delta == 3) {
    at <- abs(grid) <= cutoff
    polygon(c(grid[at], rev(grid[at])),
            c(dens[at], rep(0, sum(at))),
            col = "#F4C7C7", border = NA)
  }
  lines(grid, dens, lwd = 2)
  abline(v = c(-cutoff, cutoff), lty = 2)
  if (delta == 0) {
    legend("topright", "alpha = .0500", fill = "#BBDDEB",
           bty = "n", cex = .75)
  } else {
    legend("topleft", c("beta = .1627", "power = .8373"),
           fill = c("#F4C7C7", "#BBDDEB"), bty = "n", cex = .75)
  }
}
par(old)
beta <- pnorm(cutoff, 3, plan_se) - pnorm(-cutoff, 3, plan_se)
c(SE = plan_se, cutoff = cutoff, beta = beta, power = 1 - beta)
```

| True difference in the planning model | Rejection probability |
|:--|--:|
| 2 points | .5009 |
| 3 points | .8373 |

: Power for the same 74-per-group planning model. {#tbl-ch03-planning-power tbl-colwidths="[65,35]"}

The design chosen for a two-point interval half-width has about one-half
power at a true two-point difference, rather than automatically achieving
an .80 target. A larger true difference is easier to distinguish from
zero with the same spread. The model, alternative, and rejection rule
must accompany each power value. [Appendix B](#app-ch03-planning) gives
the probability calculation.

Planning also needs a plausible spread, allocation, independently
informative units, dependence structure, procedure, sidedness, threshold,
measurement quality, and feasible follow-up. If 10% of outcomes are
expected to be unavailable independently of relevant outcomes,
$74/.90=82.2222$ suggests recruiting at least 83 per group to obtain
74 on average. This is no guarantee of complete data and does not repair
bias from selective attrition. Unequal allocation or clustering can change
the information supplied by a given number of participants.

### Several comparisons and when they were chosen

One analysis can be part of a larger search. A researcher might inspect
several outcomes, subgroups, models, exclusion rules, or stopping
points, then report the most favorable result. The reported tail area
can still be calculated, but a calibration for one advance-specified
test does not automatically cover that selection process
[@simmons2011]. Explain which decisions were made before and which
after examining the relevant results.

Advance choice and multiplicity control do different jobs. Advance
choice makes the intended comparison and the scope for selection
visible. Multiplicity control addresses a defined **family** of
comparisons. Five planned tests can still form a family needing
adjustment. A comparison noticed later can still be informative when
its exploratory origin and appropriate uncertainty are reported.

The **familywise error rate** is the probability of at least one
Type I error in a defined family. If five true-null tests are
independent and each uses .05, the chance of no false rejection is
$.95^5$. Thus

$$
1-(1-.05)^5=1-.95^5=.2262.
$$

The independence condition is essential to that calculation. For a
familywise target of .05, **Bonferroni** instead allocates

$$
\alpha_{\mathrm{each}}=\frac{.05}{5}=.0100.
$$

The probability of one or more false rejections is no larger than
the sum of their individual probabilities. The five allowances sum
to .05, so this guarantee does not require independent tests, provided
the individual tests have their stated calibration. Adjustment cannot
repair an invalid individual procedure or an undisclosed wider search.

Selection can also change the behavior of an otherwise unbiased
estimator. Consider a Normal sampling model with true difference 1 point
and known SE 2 points. A two-sided .05 test of zero using the stipulated
critical value 1.96 selects estimates above 3.9200 or below $-3.9200$.

```{r}
#| label: fig-ch03-selection
#| fig-pos: H
#| fig-cap: "A significance filter retains both tails of a Normal estimate distribution centered at the true difference of one point, with known SE two points."
#| fig-alt: "Normal density centered at one. Selected estimates below minus 3.9200 have the wrong sign; selected estimates above 3.9200 substantially exceed the true effect."
#| fig-width: 6.8
#| fig-height: 3.1
local({
  saved_par <- par(mar = c(4.2, 4.2, 1.0, .7))
  on.exit(par(saved_par))
  x <- seq(-7, 9, length.out = 1601)
  plot(x, dnorm(x, 1, 2), type = "l", lwd = 2, col = ink,
       xlab = "Estimated difference (points)", ylab = "Density")
  for (selected in list(x[x < -3.92], x[x > 3.92])) {
    polygon(c(selected, rev(selected)),
            c(dnorm(selected, 1, 2), rep(0, length(selected))),
            col = "#F4C7C7", border = NA)
  }
  lines(x, dnorm(x, 1, 2), lwd = 2, col = ink)
  abline(v = 1, col = gold, lwd = 2)
  abline(v = c(-3.92, 3.92), col = red, lty = 2)
  text(1, .18, "true difference = 1", pos = 4, cex = .8, col = gold)
  text(-3.92, .065, "-3.9200", pos = 2, cex = .8, col = red)
  text(3.92, .105, "3.9200", pos = 4, cex = .8, col = red)
})
```

The selected positive estimates are much larger than the truth; the
selected negative estimates have the wrong sign. The underlying estimand
stays at one point, while selection changes which estimates are retained.
The estimator's original unbiasedness and its interval's unconditional
coverage need not hold within this selected set. This connects the timing
of analysis choices to the comparison families developed in Chapter 4.

Chapter 4 applies this idea to intervals for several contrasts. Tukey,
Dunnett, and Scheffé use other calibrations for particular families;
they are not special cases of Bonferroni. The family should follow the
question and the comparisons considered, rather than whichever rule
produces the preferred result.

### Report the result with its conditions and scope

| Item | Special Practice | Usual Practice | Comparison or status |
|:--|--:|--:|:--|
| Assigned students | 60 | 60 | Random assignment |
| Delayed assessments observed | 60 | 60 | Complete follow-up |
| Delayed-assessment mean | 27.1 | 24.0 | 3.1000-point mean difference |
| Delayed-assessment SD | 6.0 | 6.4 | Variation among student scores |
| Standard error of the mean difference | — | — | 1.1325 points |
| 95% confidence interval | — | — | [0.8571, 5.3429] points |
| Two-sided test from the same procedure | — | — | $t(117.5119)=2.7372$, $p=.0072$ |
| Benchmark for potential importance | — | — | 2.0 points; rationale documented before outcome access, course-specific |

: The invented special-practice study and its results. {#tbl-ch03-study-record tbl-colwidths="[30,16,16,38]"}

The completed table assembles quantities already developed. A useful
results paragraph explains how they relate, rather than substituting
the phrase “Special Practice worked” for the comparison:

> Among the 120 consenting first-year students randomly assigned to
> Special Practice or Usual Practice, the mean delayed-assessment
> score was 3.1000 points higher under Special Practice (27.1 versus
> 24.0 on the 0--40 assessment), with an approximate 95% interval
> from 0.8571 to 5.3429 points and $p=.0072$ for the two-sided zero-effect
> test. The comparison and the two-point benchmark were chosen
> before the outcomes were seen. The interval excludes zero but
> spans the benchmark, supporting a positive direction while
> leaving practical importance unresolved. The result concerns
> assignment among these participants, this assessment two weeks
> after four sessions, and the Welch working procedure. Attendance,
> mechanisms, longer-term learning, and students elsewhere require
> their own evidence.

Random assignment supports the participant comparison, and complete
follow-up avoids missing outcomes for this assessment. Its outcome and
participant scope remain those stated in the report. This combination of estimates, conditions, and
scope makes the result interpretable [@appelbaum2018].

Suppose the researchers now replace the primary assessment because a
different one gives a larger difference. They have changed the outcome
and selected it after inspection. The report should identify the
original result, the new comparison, the timing and reason for the
change, and the analyses considered. The new result can motivate
further work; its ordinary p-value cannot be presented as if it arose
from the original advance-specified procedure.

## Reading Bridge {#sec-ch03-bridge}

We have followed one estimate from its estimand and estimation rule
through a stated repetition, an estimated SE, a confidence interval,
and a test. Paired student scores showed why within-person differences
carry the information; the school reference extends this to assignment
units. The categorical transfer showed how expected counts and a reference
distribution extend the same reasoning. Magnitude, practical
importance, error rates, and the comparisons examined then shaped
one qualified conclusion.

Chapter 4 compares several groups at once. It develops a model of
group means, decomposes variation into between-group and within-group
parts, and uses their mean squares to form an F ratio. Specific
contrasts and their simultaneous intervals follow. A two-factor
example then shows how main effects and interaction work together.
The SE, interval, testing, and comparison-family ideas developed here
will be used at each step.

### Key terms {.unnumbered}

Use this short reference to revisit the main concepts as needed. The
Part One introduction gives the broader reference and the principal
home of each concept, including the sampling ideas developed in
Chapter 2.

| Term and common alternatives | Meaning |
|:--|:--|
| Estimate | A numerical approximation to an estimand, calculated from the available data. A point estimate expresses the approximation as one value. |
| Estimator | The rule used to calculate an estimate from data. Its bias and precision concern how the rule behaves across the repetitions specified by the study's procedure or model. |
| Bias (of an estimator) | The difference between an estimator's expected value over the specified repetitions and the estimand. An unbiased estimator has bias zero; precision describes its spread, a separate quality. |
| Confidence interval (CI) | An interval expressing uncertainty about an estimand. Under its conditions, the procedure covers that estimand at its stated rate across repetitions; the rate describes the procedure. |
| Critical value | The cut-point of a reference distribution, such as the Normal or t, selected by a confidence or significance level. It multiplies the SE in an interval and marks the rejection region of a test. |
| Coverage | The proportion of repetitions in which an interval procedure contains the estimand under its conditions. The confidence level is the procedure's target coverage, not a probability for one realized interval. |
| Null hypothesis | The statistical proposition used as the reference for a test, such as equality of specified population means. The alternative hypothesis states the departure being considered. |
| Test statistic | A quantity calculated from data to assess departure from a null hypothesis. The test compares it with a reference distribution justified under that null and stated conditions. |
| One-sample z test; one-sample t test | Both compare a sample mean with a stated null mean using its SE; z uses a known process SD under its model, whereas t estimates that SD and uses the appropriate t reference. Neither choice follows an automatic sample-size cutoff. |
| Independent-samples t test; Welch test; pooled t test | Independent-group tests compare means. Welch allows unequal variances with an estimated reference df; the pooled procedure adds the equal-variance condition. |
| Paired-samples t test | A one-sample t test of within-pair differences against a specified mean difference, with independent pairs under the stated model. |
| p-value | Under a specified null hypothesis and conditions, the probability of a test result at least as extreme as the observed one, using the test's stated tail rule. |
| Significance level (alpha) | The threshold used for rejecting a null hypothesis, chosen to control the probability of rejecting a true null under the test's conditions. This error is called a Type I error. |
| Expected count | The count a cell of a two-way table would have under a stated null pattern such as independence, given the observed margins: row total times column total, divided by the overall total. |
| Pearson's chi-square statistic ($X^2$) | The sum over cells of squared differences between observed and expected counts, each divided by its expected count, compared with a chi-square reference with $(R-1)(C-1)$ degrees of freedom under the stated model. |
| Power (statistical power) | The probability of rejecting the null under a specified alternative, design, and test. Failure to reject under that alternative is a Type II error; power is one minus its probability. |
| Familywise error rate (FWER) | The probability of at least one Type I error among a defined family of tests. Procedures such as Bonferroni allocate the error allowance across that family. |

: Selected concepts for estimates, uncertainty, and tests. {#tbl-ch03-key-terms tbl-colwidths="[28,72]"}

### Further reading {.unnumbered}

The GAISE College Report emphasizes statistical investigation,
conceptual understanding, and interpretation [@asa2016gaise].
*Introduction to Modern Statistics* develops inference through
repetition, with separate treatments of paired means and two-way
tables [@cetinkaya2024]. [Penn State STAT 200, Lesson 4](https://online.stat.psu.edu/stat200/Lesson04)
connects interval construction with its interpretation. The ASA
p-value statement [@wasserstein2016] and [Moving to a World Beyond
$p<.05$](https://doi.org/10.1080/00031305.2019.1583913) explain why
uncertainty and context deserve more attention than a threshold label.


## Exercises {.unnumbered}

These exercises use the information supplied and a standard calculator. They practise calculation, interpretation, and the conditions behind a conclusion.

<!-- prompt-id: CH03-E01 -->
### Exercise 1: Name the inferential objects (Foundational) {.unnumbered #ex-ch03-01}

A report states:

> The estimand was the mean Special-minus-Usual assignment contrast in
> delayed-assessment score among the 120 randomized participants. In every
> admissible repetition, the analyst would subtract the Usual assigned-group
> mean from the Special assigned-group mean. The observed difference was
> 3.1000 points, with an estimated standard error of 1.1325 points.

Identify:

1. the estimand;
2. the estimator;
3. the estimation process;
4. the estimate and one statistic used in obtaining it;
5. the standard error and what varies in its repeated-process interpretation;
6. the parameter; and
7. two bounds that must remain in the answer.

Explain why the estimate and the parameter should not be given the same label.

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH03-E02 -->
### Exercise 2: Calculate SE and distinguish it from SD (Foundational) {.unnumbered #ex-ch03-02}

The Special and Usual Practice groups each contain 60 students. Their
respective delayed-score SDs are 6.0 and 6.4 points. The reported uncertainty
calculation is

$$
\widehat{SE}=\sqrt{\frac{s_1^2}{n_1}+\frac{s_0^2}{n_0}}.
$$

1. Substitute the group sizes and SDs to check the reported SE of 1.1325 points.
2. Which quantities describe variation among scores, and which describes
   estimated variation of the mean difference across possible assignments?
3. Why does the SE retain assessment-point units?
4. If effective information increased while the design and score variation
   remained comparable, would the SD or SE usually become smaller?
5. Name two defects that a smaller standard error would not repair.
6. Does random assignment make the Welch reference an exact randomization
   distribution? Explain briefly.

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH03-E03 -->
### Exercise 3: Interpret interval coverage (Foundational) {.unnumbered #ex-ch03-03}

A student writes:

> The 95% confidence interval [0.8571, 5.3429] means there is a 95% probability that
> the true mean assignment effect lies between 0.8571 and 5.3429 points. If the
> method is valid, every 95% interval should contain the true effect.

Rewrite the statement using repeated-procedure language. Then explain:

1. why some intervals from a valid 95% procedure miss the fixed estimand;
2. what the endpoints contribute to reading this study; and
3. three sources of uncertainty or bias that the conventional interval may
   omit.

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH03-E04 -->
### Exercise 4: Translate a p-value (Foundational) {.unnumbered #ex-ch03-04}

For the matching two-sided zero-contrast test, *t*(117.5119) = 2.7372 and
*p* = .0072. Consider four interpretations:

A. There is a 0.72% probability that the null hypothesis is true.

B. There is a 99.28% probability that Special Practice will work again.

C. Under the zero-contrast null model and the stated procedure, results at
least as incompatible as the observed one have probability about .0072.

D. The 3.1000-point effect is educationally important because *p* = .0072.

Choose the defensible interpretation and diagnose each rejected one. Then use
an estimate of 3.1, an SE of 1.13255, and a supplied 95% critical value of
1.98036 to:

1. check $t=(3.1-0)/1.13255$;
2. calculate $3.1\pm1.98036(1.13255)$ and round the endpoints to four decimal places;
3. explain why exclusion of zero agrees with the matching two-sided *p* < .05
   result, without implying exact randomization calibration; and
4. write one sentence reporting estimate, interval, and p-value without using
   **prove**, **accept**, or **worked**.

The reference tail area is supplied as *p* = .0072; you do not need to
calculate a *t*-distribution probability by hand.

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH03-E05 -->
### Exercise 5: Interpret one published active-learning result (Integrative) {.unnumbered #ex-ch03-05}

Deslauriers and colleagues (2019) used a randomized crossover study in introductory
college physics: students were randomly assigned to two groups, and each
group experienced both instructional conditions in successive class meetings,
in opposite orders. In a regression model, performance on a test of learning was
0.46 SD higher under active instruction than under passive instruction,
*p* < .001. The brief result supplied here does not report a confidence
interval [@deslauriers2019].

Use the supplied description to:

1. mark the participant set, condition contrast, outcome, estimate, units,
   p-value, and design;
2. explain what the estimate and p-value each tell you;
3. identify the information supplied and what else would help you understand
   the result; and
4. state a bounded one-sentence version of the reported answer.

Then explain which design and analysis details would help judge the uncertainty
calculation and scope. State what further information you would seek in the
Methods or supplement and name one broader claim this excerpt cannot support.



\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH03-E06 -->
### Exercise 6: Repair the special-practice conclusion (Integrative) {.unnumbered #ex-ch03-06}

Rewrite:

> Special Practice worked, *p* < .05.

Use this study record (the same one summarized in the chapter's study-and-results table):
120 consenting students in one course were randomly assigned, 60 per group,
to four sessions of Special or Usual Practice with equal total study time.
All completed a common 0--40 assessment two weeks later. Special had mean
27.1 (SD 6.0), and Usual had mean 24.0 (SD 6.4). The prespecified primary
comparison gave 3.1000 points, 95% CI [0.8571, 5.3429], approximate Welch
*t*(117.5119) = 2.7372, *p* = .0072. Before outcome access, the protocol specified
the two-point benchmark as potentially important: the course team treated two
points as the smallest average difference that would change its recommendation,
because the marking scheme awards two points to one completed application task
and the new format would add little preparation time. Your revision must
retain:

1. the participant set and assignment comparison;
2. the outcome, time, direction, and raw units;
3. the estimate and 95% interval;
4. the reported numerical p-value only after the estimate and interval;
5. the choice of the primary comparison before outcome access;
6. what random assignment and complete follow-up support;
7. the interval's relation to the prespecified, course-specific two-point
   benchmark for potential importance;
   and
8. at least three scope boundaries.

After the rewrite, choose three phrases and explain what each repairs in the
original sentence.

After viewing the outcomes, the researchers decide that a different assessment
should replace the planned assessment as primary because it gives a larger
difference. Explain how this changes the comparison and its interpretation.
State what information should be reported about the original and new analyses,
what advance-planning claim can no longer be made, and why the new assessment's
p-value cannot be read as if that assessment had been chosen before inspection.

<!-- prompt-id: CH03-E07 -->
### Exercise 7: Check expected counts and a chi-square result (Foundational) {.unnumbered #ex-ch03-07}

An invented study records participation format and completion. It is
distinct from the UCI student data.

| Participation format | Completed | Did not complete | Total |
|:--|--:|--:|--:|
| Online | 84 | 36 | 120 |
| In person | 63 | 17 | 80 |
| Total | 147 | 53 | 200 |

: Participation-format outcomes for Exercise 7. {#tbl-ch03-ex07-counts tbl-colwidths="[34,22,28,16]"}

The expected counts under independence are 88.2000, 31.8000, 58.8000,
and 21.2000.
The Online/Completed contribution is $(84-88.2000)^2/88.2000=0.2000$; the
four contributions (0.2000, 0.5547, 0.3000, 0.8321) sum to
$X^2=1.8868$ on $(2-1)(2-1)=1$ df, giving $p=.1696$.
Report calculated results to four decimal places. The completion proportions are
$84/120=.7000$ and $63/80=.7875$.

1. Reproduce the Online/Completed expected count, its contribution, and
   the degrees of freedom, explaining what an expected count represents.
2. Identify the cells in which the observed count exceeds the expected count.
3. State the omnibus null in words.
4. Write one conclusion that distinguishes the observed proportion difference,
   evidence against independence, association magnitude, and causation.
5. Explain why $p=.1696$ is not proof that format and completion are independent.


## Appendix A: Probability and Repetition Foundations {.unnumbered #app-probability}

This appendix develops the probability and variance results behind the
chapter's uncertainty calculations. Each result belongs to a stated
repetition: drawing observations from a model, sampling a finite list, or
reassigning the same participants. Keeping that repetition visible explains
both the usefulness of a formula and the limits of its application.

### Probability, models and expectation {.unnumbered #app-ch03-conditioning}

Chapter 2 introduces conditional probability, independence and the
distribution catalogue. Recall the distinction needed here. An
empirical distribution describes observed values; a probability model
assigns chances to possible values; a sampling distribution describes a
statistic over the stated repetitions. A conditional probability
$P(A\mid B)=P(A\cap B)/P(B)$ uses only outcomes satisfying $B$.
Independence of two events means $P(A\cap B)=P(A)P(B)$; independence
of observations is a separate condition in the SE calculations.

Chapter 2 defined the expectation of a discrete random variable as the
probability-weighted mean $E(X)=\sum_x xP(X=x)$; for a continuous
variable an integral replaces the sum. For example, the Binomial model with ten independent
trials and constant success probability .70 has expected count
$10(.70)=7.0000$ and variance $10(.70)(.30)=2.1000$. Seven is an
average over the model's repetitions, not a prediction for every ten
trials. We now use expectation and covariance to derive uncertainty.

### Variance, sample variance, and averages {.unnumbered #app-ch03-variance}

The variance of a random variable is its expected squared distance from
its expectation: $\operatorname{Var}(Y)=E\{(Y-E(Y))^2\}$.
For fixed weights $a_i$ and observations with finite second moments,

$$
\operatorname{Var}\left(\sum_i a_iY_i\right)
=\sum_i a_i^2\operatorname{Var}(Y_i)
+2\sum_{i<j}a_ia_j\operatorname{Cov}(Y_i,Y_j).
$$

Here covariance is the expected product of two centered observations, and
$i<j$ includes each distinct pair once. Expanding the square of the
weighted sum gives the squared terms and the cross-products in this
identity. For an average, every weight is $1/n$, giving

$$
\operatorname{Var}(\bar Y)=\frac{1}{n^2}
\left\{\sum_i\operatorname{Var}(Y_i)
+2\sum_{i<j}\operatorname{Cov}(Y_i,Y_j)\right\}.
$$

Independent observations have zero covariance. If their common variance
is $\sigma^2$, the expression becomes $n\sigma^2/n^2=\sigma^2/n$.
For a difference, weights $1$ and $-1$ instead give
$\operatorname{Var}(U-V)=\operatorname{Var}(U)+\operatorname{Var}(V)
-2\operatorname{Cov}(U,V)$. These two applications explain the mean,
independent-group, and paired-difference calculations in the chapter.

Estimating $\sigma^2$ introduces the sample variance,
$s^2=\sum_i(Y_i-\bar Y)^2/(n-1)$. The deviations satisfy
$\sum_i(Y_i-\bar Y)=0$: once $n-1$ deviations are specified, the
last is fixed. This constraint helps explain the denominator, but its
unbiasedness requires a calculation. For independent observations with
common mean $\mu$ and variance $\sigma^2$,

$$
\sum_i(Y_i-\bar Y)^2
=\sum_i(Y_i-\mu)^2-n(\bar Y-\mu)^2.
$$

The expectation of the first term on the right is $n\sigma^2$.
The sample mean is unbiased and has variance $\sigma^2/n$, so the
second term has expectation $n(\sigma^2/n)=\sigma^2$. Consequently,

$$
E(s^2)=\frac{n\sigma^2-\sigma^2}{n-1}=\sigma^2.
$$

For the three observed values 2, 4 and 6, the mean is 4 and the
squared deviations sum to 8. The sample variance is $8/(3-1)=4.0000$,
whereas the squared-deviation average $8/3=2.6667$ uses a different
denominator.

Normality is unnecessary for this expectation result; the independence,
common mean and variance, and finite-variance conditions matter. Unbiased
estimation of the variance does not imply that its square root $s$ is
unbiased for $\sigma$.

The observed-data version also prepares Chapter 4's sums of squares. For
any proposed center $a$,
$\sum_i(y_i-a)^2=\sum_i(y_i-\bar y)^2+n(\bar y-a)^2$.
The last term is nonnegative and vanishes at $a=\bar y$, so the mean
minimizes the sum of squared deviations. This **least-squares** property
is an algebraic fact about the observed values.

### Concentration and limiting sampling shape {.unnumbered #app-ch03-limits}

The running-mean illustration in Section 2.4 separates concentration of
one accumulating sequence from the sampling-distribution shape across
repetitions. The formal statements make their conditions explicit.

For independent, identically distributed observations with finite absolute
mean, $E(|Y|)<\infty$, the **law of large numbers** states

$$
\bar Y_n\xrightarrow{p}\mu\quad\text{as }n\to\infty.
$$

Convergence in probability means that, for every fixed positive tolerance
$\varepsilon$, $P(|\bar Y_n-\mu|>\varepsilon)$ tends to zero.
It describes concentration near the mean, rather than a requirement that
every successive average move closer to it.

For independent, identically distributed observations with finite,
nonzero variance $\sigma^2$, the **central limit theorem** states

$$
\frac{\sqrt n(\bar Y_n-\mu)}{\sigma}
=\frac{\bar Y_n-\mu}{\sigma/\sqrt n}
\xrightarrow{d}N(0,1).
$$

Convergence in distribution means that the centered mean, measured in its
own SE units, approaches a standard Normal distribution. The theorem
concerns the sampling shape of the mean, not the distribution of individual
waiting times. Strong skewness or heavy tails can slow this approximation;
dependence may require a different theorem and SE. A universal sample-size
cutoff cannot certify the approximation. Neither limiting result repairs
selection, measurement, missing-data, or design problems.

### Dependence, finite populations, and effective information {.unnumbered #app-ch03-dependence}

The covariance terms in [the variance identities](#app-ch03-variance) show how dependence changes
precision. Positive total covariance makes a mean more variable than the
independent-observation formula suggests. Negative total covariance can
make it less variable. Independence implies zero covariance when second
moments exist; zero covariance alone allows nonlinear dependence.

Chapter 2's [correlation-bound calculation](chapter_02.qmd#sec-ch02-correlation-bound)
shows why a standardized covariance lies between $-1$ and $1$. Here the
covariance terms matter because they change the uncertainty of a mean or
a difference.

Sampling without replacement from a fixed list produces dependence even
when selection is random. For simple random sampling without replacement
from $N$ fixed population values, let $\bar Y_N$ be the population mean
and define the population variance using divisor $N$:

$$
\sigma^2=\frac{1}{N}\sum_{i=1}^N(Y_i-\bar Y_N)^2,
\qquad
SE(\bar Y)=\frac{\sigma}{\sqrt n}\sqrt{\frac{N-n}{N-1}}.
$$

The **finite-population correction**, $\sqrt{(N-n)/(N-1)}$, accounts
for the decreasing number of unobserved units; the sampling uncertainty
is zero at a census. Other selection schemes introduced in Chapter 2
require their own design-based variance calculation. Sampling without
replacement alone does not establish this formula.

An equivalent convention uses $S_N^2$, the same population squared
deviations divided by $N-1$. Since $S_N^2=N\sigma^2/(N-1)$, it gives

$$
SE(\bar Y)=\sqrt{1-\frac nN}\frac{S_N}{\sqrt n}.
$$

With this convention, the correction for 100 of 20,000 units is
$\sqrt{1-.0050}\approx.9975$; with 400 it is $\sqrt{1-.0200}\approx.9899$.
The choice of population-variance divisor explains the different-looking
factors; both formulas give the same SE. This correction for the sampling
fraction serves a different purpose from the sample-variance denominator
$n-1$.

A **precision-equivalent effective sample size** expresses a comparison
with a specified independent-sampling benchmark. If that benchmark gives
variance $\sigma^2/n$, define $n_{\mathrm{eff}}$ by
$\operatorname{Var}_{\mathrm{design}}(\bar Y)=\sigma^2/n_{\mathrm{eff}}$.
For equal clusters of size $m$, equal weights, common marginal variance,
within-cluster correlation $\rho$, and independent clusters,
$n_{\mathrm{eff}}=n/[1+(m-1)\rho]$. With $m=20$ and $\rho=.05$,
the denominator is 1.9500, so these observations supply about half the
mean-estimation precision of equally many independent observations.
Different weighting or dependence changes the comparison; an efficient
design can yield $n_{\mathrm{eff}}>n$. This quantity is specific to the
estimand and variance benchmark. It is neither the number of assignment
units nor a universal replacement for degrees of freedom.

### Random assignment and reference calibration {.unnumbered #app-ch03-assignment}

For the fixed $N=120$ Special Practice participants, let $S_1^2$ and
$S_0^2$ be the potential-outcome variances under the two assignments,
with divisor $N-1$. Let $S_{\mathrm{eff}}^2$ be the variance of the individual
effects $Y_i(1)-Y_i(0)$, using the same divisor. Under complete random
assignment of $n_1$ and $n_0$ participants, consistent conditions and
no relevant interference [@bai2024],

$$
\operatorname{Var}_{\mathrm{assign}}(\bar Y_1-\bar Y_0)
=\frac{S_1^2}{n_1}+\frac{S_0^2}{n_0}-\frac{S_{\mathrm{eff}}^2}{N}.
$$

The assigned groups are complementary, so their means need not be
independent. Only one potential outcome is observed per participant,
leaving the individual-effect variance unidentified. The observed
within-group variances estimate the corresponding potential-outcome
variances. Consequently,

$$
E_{\mathrm{assign}}\left(\frac{s_1^2}{n_1}+\frac{s_0^2}{n_0}\right)
=\operatorname{Var}_{\mathrm{assign}}(\bar Y_1-\bar Y_0)+\frac{S_{\mathrm{eff}}^2}{N}.
$$

The familiar summed variance is conservative in expectation under this
design, with equality when individual effects are constant. A particular
estimate need not be too large. This expectation identity alone neither
makes a Welch t reference exact nor guarantees conservative finite-sample
interval coverage. For genuinely independent samples the same SE formula
has another justification: their mean variances add with zero covariance.

The **Welch--Satterthwaite degrees of freedom** are

$$
\nu=\frac{(s_1^2/n_1+s_0^2/n_0)^2}
{(s_1^2/n_1)^2/(n_1-1)+(s_0^2/n_0)^2/(n_0-1)}.
$$

Using $36/60$ and $40.96/60$ gives $\nu=117.5119$ and critical value
1.9804. Fractional df describe the reference approximation, not fractional
people. Welch's independent-sample procedure allows unequal variances
[@welch1947]; its application to the participant assignment remains the
chapter's working approximation.

A zero average effect allows positive and negative individual effects
to cancel. The stronger **sharp null** says every participant's effect
is zero. A randomization test of that stronger proposition uses the actual
assignment mechanism, rather than obtaining exactness from a Normal or t
reference. The cited randomized-experiment reading develops its full
derivation; the distinction is enough to interpret this chapter's working
mean comparison [@bai2024].

For a formal coverage statement, write $C(D)$ for the interval from a
possible data record $D$. Then $P\{\theta\in C(D)\}$ concerns a fixed
estimand across the declared repetitions. Exact 95% coverage equals .95
under its model, approximate coverage approaches that target under
justified conditions, and conservative coverage is at least .95.
Other quantities, such as an unknown variance, can affect this probability.
The model and design therefore accompany the coverage claim; it assigns
no probability to the fixed estimand after observing one interval.

### Bias, variance, and mean squared error {.unnumbered #app-ch03-mse}

Mean squared error combines an estimator's systematic displacement and
its variability under one stated repetition:

$$
\operatorname{MSE}(\widehat\theta)
=E\{(\widehat\theta-\theta)^2\}
=\operatorname{Var}(\widehat\theta)
+\operatorname{Bias}(\widehat\theta)^2.
$$

To see the identity, write the estimation error as
$\widehat\theta-E(\widehat\theta)
+E(\widehat\theta)-\theta$ and square it. The first part
has expectation zero, so the expected cross-product vanishes. The
remaining terms are the variance and squared bias.

An unbiased estimator with SD 2 has MSE $2^2+0^2=4.0000$. An estimator with
SD 1 and bias 1.5 has MSE $1^2+1.5^2=3.2500$. The second has lower
average squared error in this comparison, despite its bias. That judgment
depends on the common estimand, repetition, and squared-error criterion.
Precision alone cannot establish accuracy, and a favorable MSE under a
model cannot repair an unsuitable outcome or target population.

## Appendix B: Inferential Extensions {.unnumbered #app-procedures}

These extensions connect the chapter's worked calculations to related
procedures and planning choices. The target quantity and the independent
or assigned units determine the relevant formula. Chapter 4 develops the
comparison of several means and comparison families in greater detail.

### A compact guide to common procedures {.unnumbered #app-ch03-procedure-guide}

The named procedures in Section 4.1 connect the observed units to
their SE and reference. Here we connect their formulas and distinguish
estimation from a null restriction. Many intervals have the form
$\widehat\theta\pm c\widehat{SE}(\widehat\theta)$, where $c$ is a
critical value. Many test statistics have the form
$(\widehat\theta-\theta_0)/SE_0(\widehat\theta)$, where the
subscript zero reminds us that the SE may be calculated under a null
restriction. The two expressions match only when their estimators,
restrictions, reference distributions, and sidedness are compatible.



Under the independent Normal model for one mean,

$$
\bar Y\pm t_{1-\alpha/2,n-1}\frac{s}{\sqrt n},\qquad
t=\frac{\bar Y-\mu_0}{s/\sqrt n}.
$$

For $n$ independent pairs, apply these formulas to the pair differences:
replace $\bar Y$ by $\bar D$, $s$ by $s_D$, and $\mu_0$ by
$\mu_{D,0}$. Normality for an exact $t$ reference concerns the
differences. For two independent groups, use the sum of the two mean
variances and the Welch degrees of freedom in
[Appendix A](#app-ch03-assignment).
For a population SD known independently of the sample, a suitable Normal
model instead gives a $z$ reference for a mean; such knowledge is unusual
in applied social research.

For $n$ independent binary observations with a common probability $p$,
$SE(\widehat p)=\sqrt{p(1-p)/n}$. Estimation commonly replaces
$p$ by $\widehat p$; a test of $p=p_0$ may use $p_0$ in its null
SE. A correct variance formula alone does not ensure that a Normal
interval has reliable coverage for small or boundary proportions.

Sidedness also changes the procedure. For a continuous symmetric $t$ or
$z$ reference, let $p_2$ be the usual two-sided p-value. If the statistic
points in an alternative direction chosen before inspecting the relevant
results, its one-sided p-value is $p_2/2$. If it points in the opposite
direction, it is $1-p_2/2$. The special-practice statistic therefore gives
approximately .0036 for a positive alternative and .9964 for a
negative alternative; its reported procedure remains two-sided. This
halving shortcut need not hold for asymmetric or discrete tests. A
one-sided level-.05 test matches a one-sided 95% confidence bound, with a
different cutoff from a two-sided 95% interval.

### Resampling and uncertainty {.unnumbered #app-ch03-resampling}

A **bootstrap distribution** approximates an estimator's variability by
repeatedly resampling from observed data. Its center is usually near the
observed estimate. In a simple case bootstrap for independent observations:

1. Identify the independent observational unit and the original number of units.
2. Draw that many units with replacement from the observed units.
3. Apply the same estimator to the resampled data.
4. Repeat the resampling and estimation many times.
5. Use the distribution of estimates to examine variability or construct an appropriate interval.

The bootstrap SE is the sample SD of the resulting estimates. It is
distinct from the SD of individual observations and from the uncertainty
caused by using only finitely many simulation repetitions. Preserve the
study structure: resample pairs together, or clusters when independent
clusters are the appropriate units. Resampling individual students from
clustered classes can misrepresent uncertainty. The number resampled in
the simple algorithm is the observed unit count, not a substituted
precision-equivalent effective sample size.

A **null distribution** instead describes a test statistic under specified
null restrictions and other conditions. A centered $t$ statistic may
have a zero-centered null reference; a chi-square or $F$ statistic does
not. A bootstrap distribution centered near an observed difference is
therefore not automatically a null distribution for testing zero. A
randomization test must reproduce the allowed assignment mechanism under
the null being tested. Resampling can represent an appropriate repetition;
it cannot recover unobserved cases or repair biased selection, invalid
measurement, or an unsuitable estimand.

### Paired schools: a comparison of class means {.unnumbered #app-ch03-school-pairs}

Each of twelve participating schools contributes two classes of twenty
students. Within each school, one class is randomly assigned to
explanatory feedback and the other to usual feedback. All 480 students
complete the same delayed 0--40 assessment. These are invented records;
the schools are participating schools rather than a probability sample
of all schools. The class is the assignment unit, and each school
supplies one explanatory-minus-usual contrast.

| School | Usual mean | Explanatory mean | Difference |
|:--|--:|--:|--:|
| A | 21 | 19 | -2 |
| B | 22 | 22 | 0 |
| C | 20 | 21 | 1 |
| D | 23 | 24 | 1 |
| E | 24 | 26 | 2 |
| F | 22 | 24 | 2 |
| G | 25 | 28 | 3 |
| H | 24 | 27 | 3 |
| I | 23 | 27 | 4 |
| J | 26 | 30 | 4 |
| K | 25 | 30 | 5 |
| L | 22 | 29 | 7 |

: Invented class means under within-school assignment. {#tbl-ch03-school-pairs tbl-colwidths="[20,25,30,25]"}

The twelve differences sum to 30. Weighting schools equally gives
$\bar d=30/12=2.5000$. Equal class sizes also make student weights
equal in this example. The squared deviations from 2.5 sum to 63:

$$
s_d=\sqrt{63/11}=2.3932,\qquad
\widehat{SE}(\bar d)=\frac{\sqrt{63/11}}{\sqrt{12}}=0.6908.
$$

The zero-mean contrast test uses
$t=2.5/(\sqrt{63/11}/\sqrt{12})=3.6187$, with 11 df and $p=.0040$.
The interval is

$$
\bar d\pm t_{.975,11}\widehat{SE}(\bar d)
=2.5000\pm2.2009852\frac{\sqrt{63/11}}{\sqrt{12}}
\approx[0.9795,\ 4.0205]\text{ points}.
$$

This approximate t calibration represents variation among the twelve
school contrasts. Implemented within-school assignment, the common outcome
and follow-up, and no relevant interference between classes support the
interpretation. It does not make these schools representative of a wider
population. Replacing 12 by 480 in the SE denominator would discard the
assignment design.

This is a **blocked assignment**: each school forms a block, and the
two conditions are allocated within every school. Comparisons then use
the within-school differences. Blocking can remove school-level variation
from those contrasts. **Stratified sampling** instead selects observations
within population strata to determine who enters a sample. It addresses
selection into the study; blocking here addresses allocation after entry.
The similar use of groups does not make either mechanism imply the other.
Chapter 4 recalls this design when discussing ANOVA conditions.

### Error probabilities across 1,000 studies {.unnumbered #app-ch03-thousand-studies}

Consider stipulated expected counts across many studies, not empirical
counts about education research. Suppose 900 of 1,000 studies have a
true null and 100 have the specified alternative. Every test has
$\alpha=.05$ under its null model and power .80 at that alternative.

| Model state | Reject | Do not reject | Total |
|:--|--:|--:|--:|
| True null | 45 | 855 | 900 |
| Specified alternative | 80 | 20 | 100 |
| Total | 125 | 875 | 1,000 |

: Expected decisions under stipulated prevalence, alpha and power. {#tbl-ch03-thousand tbl-colwidths="[40,20,20,20]"}

The null-state Type I error probability is $45/900=.0500$; at the
alternative the Type II error probability is $20/100=.2000$. Among
the expected 125 rejections, however, $45/125=.3600$ arise from true
nulls. This false-positive fraction has a different denominator.
Changing the stipulated frequency of true nulls or the power changes it.
It is neither an observed p-value nor a probability to assign directly
to the null in one completed study.

### Standardized effects and regression connections {.unnumbered #app-ch03-effect-scales}

A standardized mean difference expresses a raw difference relative to a
named spread. For two independent groups, one familiar choice is

$$
\begin{aligned}
s_{\mathrm{pooled}}&=
\sqrt{\frac{(n_1-1)s_1^2+(n_0-1)s_0^2}{n_1+n_0-2}},\\
d&=\frac{\bar y_1-\bar y_0}{s_{\mathrm{pooled}}},\qquad
g\approx d\left(1-\frac{3}{4(n_1+n_0)-9}\right).
\end{aligned}
$$

The pooled SD combines the within-group sums of squares. Hedges' $g$
applies a small-sample bias correction to $d$; the displayed multiplier
approximates the usual correction under its reference model
[@lakens2013]. The special-practice summaries give
$s_{\mathrm{pooled}}=\sqrt{38.4800}\approx6.2032$, $d=.4997$, and
$g\approx.4966$. These scale summaries accompany the 3.1000-point raw
difference; they do not change its participant scope or certify its
educational importance.

A within-person standardized change could use the SD of change scores,
the pretest SD, or another reference spread. These denominators express
different comparisons. Standardization can assist comparison across
measures, but it removes the original unit, and conventional size labels
are not universal importance criteria. Chapter 4 applies effect-size
reasoning to comparisons among several groups.

Several familiar estimates also have regression representations. A
one-sample mean is the fitted intercept in an intercept-only model; the
mean paired change is the intercept when the outcome is the within-pair
difference. With $X_i=1$ for Special assignment and $0$ for Usual,

$$
Y_i=\beta_0+\beta_1X_i+\varepsilon_i
$$

fits the Usual mean as $\widehat\beta_0$ and the
Special-minus-Usual difference as $\widehat\beta_1$. This is an
algebraic equality of ordinary-least-squares coefficients. Uncertainty
still requires the appropriate variance estimator and design; the
ordinary equal-variance regression SE is not automatically the Welch SE.
The representation does not assert a constant individual treatment effect
or create a population estimand. Later chapters develop the additional
assumptions and uses of regression.

### Power, precision, selection, and several comparisons {.unnumbered #app-ch03-planning}

Section 5.2 develops the planning inputs, the 74-per-group example, and
the distinction between interval width and power. This appendix supplies
the algebra, probability calculation, and further consequences.

Precision planning begins with a target interval width in meaningful
units. In a separate model with two independent, equally sized groups,
common planning SD $\sigma$, and a Normal critical value $z^*$,
the SE of the difference is $\sigma\sqrt{2/n}$. Setting the desired
half-width to $h=z^*\sigma\sqrt{2/n}$ and solving for $n$ gives

$$
n_{\mathrm{per\ group}}\approx
2\left(\frac{z^*\sigma}{h}\right)^2
=2\left(\frac{1.96(6.2)}{2}\right)^2
=73.8356.
$$

Rounding upward gives 74 observed outcomes per group. A finite-t
critical value, estimated rather than known spread, unequal allocation,
or clustering can change the requirement. The main discussion explains
the attrition allowance. Halving the target half-width approximately
quadruples the required sample size under this model.

For the power comparison, write the estimator as
$\widehat{\Delta\mu}\sim N(\Delta\mu,SE^2)$, where $SE=6.2\sqrt{2/74}$,
and let $c=z_{.975}SE$. The two-sided test rejects when the estimate is
below $-c$ or above $c$. With $\Phi$ denoting the standard Normal
cumulative distribution function,

$$
\operatorname{Power}(\Delta\mu)
=\Phi\left(\frac{-c-\Delta\mu}{SE}\right)
 +1-\Phi\left(\frac{c-\Delta\mu}{SE}\right).
$$

Using the unrounded $SE$ and $z_{.975}$ gives $c=1.9977$ and the
two powers in Section 5.2. At $\Delta\mu=0$, this expression is the .05
Type I error probability. A power curve repeats the calculation across
plausible alternatives, retaining the same design and test.

The significance filter in Section 5.3 changes the distribution being
summarized. If $S$ denotes selection, the conditional mean
$E(\widehat{\Delta\mu}\mid S)$ need not equal
$E(\widehat{\Delta\mu})$, and
$P\{\Delta\mu\in C(D)\mid S\}$ need not equal the interval's unconditional
coverage. Selection affects what is retained without changing the
underlying estimand.

The relevant choices can extend beyond the tests in a reported table:
outcomes, time points, subgroups, transformations, coding rules,
exclusions, model specifications, repeated looks, and stopping decisions
all provide opportunities for selection [@simmons2011]. A useful report
identifies which choices preceded inspection, which followed it, and
which results are available. Exploratory findings can motivate further
study, while their interpretation should preserve how they were found.

For $m$ tests with valid individual null p-values, Bonferroni's rule
uses $\alpha/m$ per test. The union bound gives

$$
P(\text{at least one false rejection})
\le\sum_{j\in\mathcal H_0}P(\text{false rejection in test }j)
\le m\frac{\alpha}{m}=\alpha,
$$

where $\mathcal H_0$ indexes the null hypotheses that are actually true.
The bound allows dependent tests and also applies when some nulls are
false. Five tests with a familywise target .05 therefore use .0100 each.
By contrast, five *independent*, true-null tests each conducted at .05
have probability $1-.95^5=.2262$ of at least one false rejection.
Dependence changes that product calculation.

The guarantee concerns the declared family and valid component procedures;
it cannot automatically cover an undocumented search. For corresponding
two-sided intervals, Bonferroni uses confidence level $1-\alpha/m$ for
each member to obtain at least $1-\alpha$ simultaneous coverage.
Chapter 4 applies this idea to specific contrasts and develops other
procedures suited to different comparison families. Tukey, Dunnett, and
Scheffé procedures have their own reference calculations and are not
special cases of Bonferroni.


## References {.unnumbered}

\small

::: {#refs}
:::

\normalsize
