---
title: "Simple Linear Regression"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \widowpenalty=10000
          \clubpenalty=10000
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AtBeginDocument{\renewcommand{\boldsymbol}[1]{\symbf{#1}}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                     fig.pos = "H", collapse = TRUE, comment = "#>")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 10, width = 68)
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 11))
# Display helper for authored tables: round once to four decimals, give a
# column the fewest common decimals that show every value, a true minus
# sign, and a bound for a nonzero value that rounds to zero.
show_value <- function(x) {
  shown <- format(round(x, 4), digits = 15, scientific = FALSE, trim = TRUE)
  tiny <- x != 0 & round(x, 4) == 0
  shown[tiny] <- ifelse(x[tiny] > 0, "< 0.0001", "> \u22120.0001")
  unname(sub("^-", "\u2212", shown))
}
```

\clearpage

::: {.callout-note title="Chapter overview"}
This chapter reads regression as prediction with quantified uncertainty.
From a reported claim about education and income we state a question
about 102 occupations, plot the records and fit a straight line, read
its slope, intercept, fitted values and residuals, and see how least
squares chooses it and how well it fits. A process model then gives the
slope a standard error and an interval, repeated samples of a known
process show how fitted slopes vary, and prediction comes at three
levels: a point, a mean response and one new case. Checks of form,
spread, influence and support qualify these answers. The chapter closes
with regression toward the mean, in Galton's heights of fathers and sons
and in a hypothetical class's exam scores. By the end you will be able
to fit and read a simple regression in R, say what its descriptive and
model-based answers concern, and recognize regression toward the mean in
a before-and-after claim. An appendix develops the mathematics.
:::

Suppose a report says, "Another year of education increases income by about
900 dollars." Before asking whether the number is significant, ask whose
education and whose income were measured: individuals, households and
occupations give three different comparisons.

\Needspace{6\baselineskip}

Our question is narrower:

> Among the 102 recorded Canadian occupations, how did occupation-average
> income differ across the supported range of occupation-average education in
> 1971, as summarized by a fitted straight line?

One row is an **occupation**, not a worker. An occupation with higher
average education need not contain the same people as one with lower
average education, so a comparison of their averages does not describe
what another year of schooling would do to one person's income.
Aggregate and individual relationships can differ substantially
[@robinson1950].

Regression can be read as **prediction with quantified uncertainty**.
The fitted line predicts an occupation's average income from its average
education; the residuals show how far each record lies from its
prediction; and, once a process model is declared, standard errors and
intervals say how uncertain the slope and the predictions are. The
chapter builds on Chapter 5's covariance, correlation and mean inference
in R and on Chapter 6's indicator coefficients for group means.

## Education and Income Across Occupations {#sec-ch7-question}

### The occupation records

The data are distributed as *carData::Prestige* [@carData2022; @fox2019].
The package documentation attributes education and income to the *Census of
Canada* (1971), volume 3, part 6, pages 19-1 to 19-21. We treat that
documentation as a source record, not as evidence that the occupations
form a probability sample, and the calculations are our reanalysis.

| Feature | Meaning in this chapter |
|:--|:--|
| Unit and record count | 102 occupations; each occupation receives equal weight |
| Education, $x$ | Average years of education of occupational incumbents in 1971 |
| Income, $y$ | Average income of incumbents, divided by 1,000; thousand 1971 Canadian dollars |
| Missingness for the fit | No missing education or income values; no modeled rows excluded |
| Observed support | Education: 6.38--15.97 years; income: 0.611--25.879 thousand dollars |
| Descriptive estimand | Least-squares slope summarizing these 102 fixed occupation records |
| Limits of the source record | No probability-sampling basis or individual-level analysis established here |

: Record, measurements, and scope for the focal analysis. {#tbl-ch7-record tbl-colwidths="[30,70]"}

Each occupation counts once, however many people it employs, so the
analysis does not describe an average worker's experience; and 1971
dollars are not today's purchasing power.

### A first plot and a first fit

```{r}
#| label: ch7-data-visible
#| echo: true
prestige <- carData::Prestige
prestige$income_k <- prestige$income / 1000
library(ggplot2)
```

The package-qualified name `carData::Prestige` selects the supplied
dataset, and dividing income by 1,000 changes its units without changing
any occupation.

```{r}
#| label: fig-ch7-scatter
#| echo: true
#| fig-cap: "Education and income in the 102 occupation records, before fitting a line."
#| fig-alt: "Occupation-average income generally rises with occupation-average education, with considerable scatter and some especially high-income occupations."
#| fig-width: 6.3
#| fig-height: 3.8
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, size = 2) +
  labs(x = "Average education (years)",
       y = "Average income (thousand 1971 dollars)")
```

Income rises with education, with much scatter. General managers
average 12.26 years of education and 25.879 thousand dollars; university
teachers average 15.97 years and 12.480. More education does not order
every pair of occupations by income, so a line can summarize the overall
pattern but cannot describe every occupation equally well. We save the
straight-line fit as `model` and its coefficients as `b`.

```{r}
#| label: ch7-prepare-visible
#| echo: true
model <- lm(income_k ~ education, data = prestige)
b <- coef(model)
coef(model)
```

The formula puts the outcome left of the tilde and the predictor to its
right. `lm()` finds the coefficients; it does not choose a population,
establish a causal design or justify a confidence interval.

## The Fitted Line and Least Squares {#sec-ch7-line}

### Fitted values, slope and residuals

A straight line assigns each observed predictor value $x_i$ a **fitted
value** $\widehat y_i$, the line's prediction for record $i$:

$$
\widehat y_i=\widehat\beta_0+\widehat\beta_1x_i,
\qquad e_i=y_i-\widehat y_i.
$$

The subscript $i$ indexes a record and the hat marks a fitted quantity.
The **residual** $e_i$ is observed minus fitted income: positive for an
occupation above the line, negative below it. It is computed from the
fitted line, unlike the unobservable error of a process model (Section 3).

For the occupation records, the line is

$$
\widehat{\mathrm{income}}=-2.8536+0.8988\,\mathrm{education}.
$$

With income in thousands of dollars, the slope says that fitted income
differs by 0.8988 thousand dollars, about 899 dollars, per one-year
difference in occupation-average education; a two-year difference
corresponds to $2\widehat\beta_1\approx1.7976$ thousand dollars.

For university teachers, the fitted coefficients before rounding give

$$
\widehat y=\widehat\beta_0+\widehat\beta_1(15.97)
\approx11.5005,\qquad
e=12.4800-\widehat y\approx0.9795,
$$

so their recorded income lies about 980 dollars above their prediction.
`predict()` applies the saved line to a record or to a new education
value without refitting:

```{r}
#| label: ch7-teacher-visible
#| echo: true
teacher <- prestige["university.teachers", ]
teacher_fit <- predict(model, teacher)
c(fitted = unname(teacher_fit),
  residual = teacher$income_k - unname(teacher_fit),
  at_12 = unname(predict(model, data.frame(education = 12))))
```

An occupation averaging 12 years of education has fitted income 7.9322
thousand dollars. @tbl-ch7-slope-readings sets the slope's reading
beside two sentences a report might also write about it.

\Needspace{12\baselineskip}

| A sentence about the slope 0.8988 | What it compares | Do these records support it? |
|:--|:--|:--|
| Occupations whose average education is one year higher have fitted average income about 899 dollars higher | Different occupations, along the fitted line | Yes, as a description of the 102 records over 6.38--15.97 years |
| If an occupation's average education rose by one year, its average income would rise by about 899 dollars | One occupation before and after a change | No: it needs records of occupations over time and a design that isolates the change |
| Another year of schooling raises a worker's income by about 899 dollars | One person before and after more schooling | No: it needs records of individuals and a causal design; occupation averages cannot supply either |

: Three readings of one slope; only the first is a comparison these records make. {#tbl-ch7-slope-readings tbl-colwidths="[40,26,34]"}

Only the first sentence compares what the records compare: occupations
that differ in average education. The second describes a change within
one occupation over time and the third a change within one person; the
records observe neither [@robinson1950].

The intercept, $-2.8536$, is the line's fitted income at zero years of
education. Zero lies far outside the observed range, so the intercept
only places the line; no recorded occupation has negative income.
Section 2.5 moves the intercept to a supported education value.

```{r}
#| label: fig-ch7-line-residual
#| fig-cap: "The least-squares line passes through the mean point; the marked vertical segment is the university-teacher residual."
#| fig-alt: "A fitted line crosses the mean education and income. At 15.97 years, a vertical segment joins the fitted value to the higher university-teacher observation."
#| fig-width: 6.3
#| fig-height: 2.8
teacher <- prestige["university.teachers", ]
teacher$fit <- predict(model, teacher)
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, colour = "grey45") +
  geom_abline(intercept = b[1], slope = b[2], linewidth = .8) +
  annotate("point", x = mean(prestige$education),
    y = mean(prestige$income_k), shape = 18, size = 3) +
  geom_segment(data = teacher, aes(xend = education, yend = fit),
               linewidth = 1) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)")
```

### How least squares chooses the line

Ordinary least squares (**OLS**) chooses the intercept and slope that minimize
the **residual sum of squares**:

$$
RSS=\sum_{i=1}^{n}(y_i-\widehat\beta_0-\widehat\beta_1x_i)^2.
$$

Squaring stops positive and negative residuals from cancelling and gives
large residuals more weight. "Best" means best by this criterion among
straight lines, not best for every research purpose. Two centered sums
give the solution:

$$
S_{xx}=\sum_i(x_i-\bar x)^2,\qquad
S_{xy}=\sum_i(x_i-\bar x)(y_i-\bar y),
$$

$$
\widehat\beta_1=\frac{S_{xy}}{S_{xx}}
=\frac{s_{xy}}{s_x^2}
=r\frac{s_y}{s_x},
\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x.
$$

Here $\bar x,\bar y$ are means, $s_x,s_y$ sample standard deviations,
$s_{xy}$ the sample covariance and $r$ Pearson's correlation. The
variance and covariance both divide by $n-1$, which cancels in their
ratio. The slope carries thousand dollars per education year; the
correlation has no units. For these data,

$$
\widehat\beta_1=\frac{675.8041}{751.8852}\approx0.8988,
\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x\approx-2.8536.
$$

Chapter 5's covariance and correlation matrices give the same slope. For
the 102 complete education--income pairs, in education--income order,

$$
\mathbf S=\begin{pmatrix}7.4444&6.6911\\6.6911&18.0279\end{pmatrix},
\qquad
\mathbf R=\begin{pmatrix}1&.5776\\.5776&1\end{pmatrix}.
$$

Dividing the covariance, in thousand-dollar-years, by the education
variance, in years squared, leaves thousand dollars per year; the
unit-free correlation times the income SD over the education SD gives
the same slope. Both routes return 0.8988 when every entry comes from the
same rows, since entries formed by separate pairwise deletion can
describe different occupations. The Computer Lab carries out the centered
sums and both matrix routes.

A slope needs education to vary: if every occupation had the same
education, $S_{xx}=0$ and no slope could be identified. The line passes
through $(\bar x,\bar y)$, since substituting $\bar x$ into the intercept
formula gives $\bar y$. With an intercept, the residuals also sum to zero
and their products with $x$ sum to zero; otherwise the line could be
shifted or tilted to reduce RSS. These are algebraic properties of OLS, not tests of
independence or linearity. **Appendix A** derives them.

### A three-record example

Three teaching records, $(x,y)=(1,2),(2,4),(3,3)$, show the whole
calculation at once. Their means are $\bar x=2$ and $\bar y=3$.
@tbl-ch7-tiny lays out the centered values, their squares and products,
the fitted values and the residuals, with column sums in the last row.

\Needspace{12\baselineskip}

| $x$ | $y$ | $x-\bar x$ | $y-\bar y$ | $(x-\bar x)^2$ | $(x-\bar x)(y-\bar y)$ | $\widehat y$ | $e$ | $e^2$ |
|--:|--:|--:|--:|--:|--:|--:|--:|--:|
| 1 | 2 | $-1$ | $-1$ | 1 | 1 | 2.5 | $-0.5$ | 0.25 |
| 2 | 4 | 0 | 1 | 0 | 0 | 3.0 | 1.0 | 1.00 |
| 3 | 3 | 1 | 0 | 1 | 0 | 3.5 | $-0.5$ | 0.25 |
| Sum |  | 0 | 0 | 2 | 1 | 9.0 | 0.0 | 1.50 |

: The complete least-squares calculation for three records; the column sums give $S_{xx}$, $S_{xy}$ and $RSS$. {#tbl-ch7-tiny tbl-colwidths="[9,9,11,11,12,17,11,10,10]"}

The sums give $S_{xx}=2$ and $S_{xy}=1$, so

$$
\widehat\beta_1=\frac{S_{xy}}{S_{xx}}=\frac12=0.5,\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x=3-0.5(2)=2.
$$

The fitted values 2.5, 3.0 and 3.5 lie on this line, the residuals sum
to zero, and their squares sum to $RSS=1.5$. No Normality assumption is
needed; a probability model enters only when we ask about repeated-sample
uncertainty.

### Comparing candidate lines

Least squares also lets us compare the fitted line with alternatives.
Make every candidate pass through $(\bar x,\bar y)$ and vary only its
slope $b_1$; then

$$
RSS(b_1)=RSS(\widehat\beta_1)
       +S_{xx}(b_1-\widehat\beta_1)^2.
$$

The added term is zero only at the OLS slope. @tbl-ch7-candidate-lines
and @fig-ch7-candidate-lines check five lines on the occupation records;
they are alternative summaries of the same records, not repeated samples.

```{r}
#| label: ch7-candidate-lines-code
x <- prestige$education
y <- prestige$income_k
candidate_slopes <- unname(b[2]) + c(-0.6, -0.3, 0, 0.3, 0.6)
candidate_intercepts <- mean(y) - candidate_slopes * mean(x)
ols_rss <- sum(resid(model)^2)
predictor_spread <- sum((x - mean(x))^2)
candidate_rss <- ols_rss +
  predictor_spread * (candidate_slopes - b[2])^2
candidate_lines <- data.frame(
  Intercept = candidate_intercepts,
  Slope = candidate_slopes,
  RSS = candidate_rss
)
```

\Needspace{9\baselineskip}

```{r}
#| label: tbl-ch7-candidate-lines
#| tbl-cap: "Candidate lines through the mean point and their residual sums of squares."
#| echo: false
knitr::kable(data.frame(lapply(candidate_lines, show_value)),
  align = "r")
```

```{r}
#| label: fig-ch7-candidate-lines
#| echo: false
#| fig-cap: "Five candidate lines through the sample mean. The OLS line has the smallest residual sum of squares."
#| fig-alt: "Five lines with different slopes cross at the mean occupation point. The highlighted OLS line runs centrally through the data; steeper and flatter alternatives have larger residual sums of squares."
#| fig-width: 6.3
#| fig-height: 3.8
candidate_grid <- expand.grid(
  education = seq(min(x), max(x), length.out = 100),
  Candidate = seq_along(candidate_slopes)
)
candidate_grid$income <- candidate_intercepts[candidate_grid$Candidate] +
  candidate_slopes[candidate_grid$Candidate] * candidate_grid$education
candidate_grid$Line <- ifelse(
  candidate_grid$Candidate == 3, "OLS", "Alternative"
)
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, colour = "grey55", size = 1.4) +
  geom_line(data = candidate_grid,
            aes(y = income, group = Candidate, colour = Line,
                linewidth = Line)) +
  scale_colour_manual(values = c(Alternative = "grey65", OLS = "black")) +
  scale_linewidth_manual(values = c(Alternative = .45, OLS = 1)) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)", colour = NULL, linewidth = NULL)
```

Every alternative has a larger RSS than the OLS line, and slopes equally
far above and below 0.8988 add the same excess. Because large vertical
discrepancies weigh heavily, high-income occupations can pull the line,
a feature of the criterion rather than evidence that they are wrong.

### Centering and changes of units

The fitted equation can be written

$$
\widehat y_i=\bar y+\widehat\beta_1(x_i-\bar x).
$$

Centering education at its mean makes the intercept $\bar y=6.7979$
thousand dollars, the fitted income at average education; centering at
12 makes it 7.9322. The slope, fitted values, residuals and $R^2$ stay
the same: centering moves the reference point, not the line. The
Computer Lab refits the centered line and checks the fitted values.
Changing income from thousands of dollars to dollars multiplies the
intercept, slope, residual SD and coefficient SEs by 1,000 and leaves
$t$, $p$, $r$ and $R^2$ unchanged; a unit change is not new evidence.

### How well does the line fit? {#sec-ch7-fit}

A line can show a clear trend and still leave large differences between
occupations. For OLS with an intercept, the total variation of income
around its mean splits into two sums of squares:

$$
TSS=\sum_i(y_i-\bar y)^2=RegSS+RSS,
\qquad RegSS=\sum_i(\widehat y_i-\bar y)^2.
$$

For the occupation records, $1820.8134=607.4214+1213.3920$, in squared
thousand dollars. These are sums of squared deviations, not variances;
the split holds because the fitted deviations and the residuals have a
zero cross-product. The coefficient of determination is

$$
R^2=1-\frac{RSS}{TSS}
=1-\frac{1213.3920}{1820.8134}\approx.3336.
$$

The line accounts for a proportion .3336 of the squared variation of
income around its mean. "Accounts for" is algebraic: it does not mean
that education causes or determines income. With an intercept and one
predictor, $R^2=r^2$, here $.5776^2\approx.3336$.

The residual standard deviation, in thousand dollars, is

$$
s_e=\sqrt{\frac{RSS}{n-2}}
=\sqrt{\frac{1213.392}{100}}=3.4834.
$$

Fitting an intercept and a slope leaves $n-2$ residual degrees of
freedom. Under the process model of Section 3, $s_e^2$ is an unbiased
estimate of the error variance and $s_e$ estimates the error SD
(Appendix A); without that model, $s_e$ measures residual scatter. It is
not the slope's standard error. `sigma()` returns it, and `summary()`
reports $R^2$:

```{r}
#| label: ch7-fit-summaries-visible
#| echo: true
c(residual_SD = sigma(model), R_squared = summary(model)$r.squared)
```

@tbl-ch7-quantities sets four summaries side by side. Here and below,
$SE$ denotes an estimated standard error; the last row belongs to the
process model of Section 3, not to the completed calculation on the 102
records.

| Quantity | Occupation result | What it describes |
|:--|--:|:--|
| Slope | 0.8988 | Change in fitted thousand-dollar income per education year |
| Residual SD | 3.4834 | Remaining income scatter on the outcome scale |
| $R^2$ | .3336 | Proportion of centered squared variation accounted for in these records |
| Classical slope SE | 0.1270 | Model-based repetition variability of the slope, under the next assumptions |

: Different summaries have different units and jobs. {#tbl-ch7-quantities tbl-colwidths="[24,22,54]"}

\Needspace{16\baselineskip}

## Inference for the Slope {#sec-ch7-inference}

### The slope's standard error and confidence interval

The slope 0.8988 is not waiting to be discovered: we computed it from all
102 records named in the question. A confidence interval answers a
different question, about a process that could generate new outcomes.
Suppose occupation incomes arose independently from a process with
conditional mean

$$
E(Y_i\mid X_i=x_i)=\beta_0+\beta_1x_i.
$$

The estimand $\beta_1$ is now the slope of that hypothetical process, not
the fixed slope of the records, and repetition means new outcomes at the
same education values, not a random sample of occupations. Write the
process as

$$
Y_i=\beta_0+\beta_1x_i+\varepsilon_i,\qquad
E(\varepsilon_i\mid\mathbf X)=0,\qquad
\operatorname{Var}(\varepsilon_i\mid\mathbf X)=\sigma^2.
$$

For the exact $t$ procedure the errors are also independent and Normal
with this common variance, and $S_{xx}>0$. Neither education nor income
needs a Normal distribution of its own: the conditions concern the errors
around the line. Independence comes from how the records arose, not from
a good-looking residual plot.

```{r}
#| label: fig-ch7-conditional
#| fig-cap: "What the process conditions claim, drawn on the occupation records: at each education value, a Normal distribution of incomes centred on the line, with the same spread everywhere."
#| fig-alt: "The 102 occupation records and the fitted line rising from about 3 to 12 thousand dollars over education 6 to 16. At 8, 10, 12 and 14 years, an identical bell-shaped curve stands sideways on the line, spreading about 10 thousand dollars above and below it; the curve at 8 years extends below zero income, while several high-income records at 12 years and above lie beyond the curves' upper tails."
#| fig-width: 6.3
#| fig-height: 3.4
curve_sd <- sigma(model)
curves <- do.call(rbind, lapply(c(8, 10, 12, 14), function(x0) {
  mu <- unname(predict(model, data.frame(education = x0)))
  income <- seq(mu - 3 * curve_sd, mu + 3 * curve_sd, length.out = 120)
  data.frame(at = x0, income = income,
    education = x0 + 1.4 * dnorm(income, mu, curve_sd) /
      dnorm(0, 0, curve_sd))
}))
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, colour = "grey60", size = 1.3) +
  geom_abline(intercept = b[1], slope = b[2], linewidth = .8) +
  geom_segment(data = unique(curves["at"]),
    aes(x = at, xend = at,
      y = b[1] + b[2] * at - 3 * curve_sd,
      yend = b[1] + b[2] * at + 3 * curve_sd),
    colour = "grey40", linewidth = .3, inherit.aes = FALSE) +
  geom_path(data = curves, aes(education, income, group = at),
    linewidth = .7, inherit.aes = FALSE) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)")
```

@fig-ch7-conditional draws what these conditions claim, using the fitted
line and $s_e=3.4834$ in place of the unknown $\beta_0$, $\beta_1$ and
$\sigma$: at every education value, a Normal distribution of incomes
centred on the line, with one common spread. The records do not look
like that everywhere. At higher education they spread more widely, and
several incomes lie beyond the upper tails; at 8 years the curve reaches
below zero, a sign that a Normal model for a nonnegative outcome is an
approximation, which Section 5 meets again.

Substituting the process into the slope formula shows where slope
uncertainty comes from. With $\mathbf X$ the predictor values we
condition on, and since $\sum_i(x_i-\bar x)=0$,

$$
\widehat\beta_1-\beta_1
=\frac{\sum_i(x_i-\bar x)\varepsilon_i}{S_{xx}}.
$$

The fitted slope misses $\beta_1$ because the realized errors need not
cancel once weighted by distance from mean education. Independence and
common variance give

$$
\operatorname{Var}(\widehat\beta_1\mid\mathbf X)
=\frac{\sigma^2\sum_i(x_i-\bar x)^2}{S_{xx}^2}
=\frac{\sigma^2}{S_{xx}},
$$

which needs no Normality; Normality enters the exact $t$ reference.
Estimating $\sigma$ by $s_e$ gives the standard error and the 95%
interval:

$$
SE(\widehat\beta_1)=\frac{s_e}{\sqrt{S_{xx}}}
=\frac{3.4834}{\sqrt{751.8852}}=0.127,
$$

$$
\widehat\beta_1\ \pm\ t_{.975,n-2}SE(\widehat\beta_1)
\approx0.8988\pm1.984(0.127)
\approx[0.6468,\ 1.1508].
$$

Here 1.984 is the 97.5th percentile of $t$ with 100 degrees of freedom.
More error variation raises the SE, and a wider predictor spread lowers
it. The test of $H_0:\beta_1=0$ gives
$t=\widehat\beta_1/SE(\widehat\beta_1)\approx7.0753$ and a two-sided
$p<.0001$: evidence against the zero-slope model under the stated
procedure. It is not the probability that the null is true, evidence of
causation or a measure of practical importance. The interval describes
what the procedure achieves over repetitions, which Section 4 makes
visible, not a 95% probability for the fixed parameter. The coefficient
table gives the estimate, SE and test, and `confint()` the interval, at
its default level .95:

```{r}
#| label: ch7-coefficient-inference-visible
#| echo: true
coef(summary(model))
confint(model, "education")
```

The process conditions are a proposed justification, not facts created
by the formula; Section 6 examines them on the occupation records.

### The slope test and the correlation test {#ch07-correlation-test}

Chapter 5 tested a zero Pearson correlation. For an OLS line with an
intercept, fitted to the same complete pairs,

$$
\begin{aligned}
t_{\mathrm{slope}}
&=\frac{\widehat\beta_1}{s_e/\sqrt{S_{xx}}}
=r\sqrt{\frac{n-2}{1-r^2}}
=t_{\mathrm{Pearson}},\\
F_{\mathrm{regression}}
&=t_{\mathrm{slope}}^2
=\frac{R^2}{1-R^2}(n-2).
\end{aligned}
$$

The first identity follows from the covariance form of the slope and
$RSS=(1-r^2)TSS$; the second compares the intercept-only model with the
line, and squaring the signed $t$ gives the nondirectional $F$. For the
occupation line both $t$ statistics are 7.0753 on 100 df, and the $F$
statistic that `summary()` prints is $F(1,100)=50.0598=t^2$, all with
$p<.0001$. They are one comparison written three ways, not three pieces
of evidence, and the Computer Lab reproduces all three.

The equality is algebra, and it needs the same intercept, rows, weights
and coefficient-covariance convention: an HC3 slope test, which
estimates the slope's SE without assuming equal error variance
(Section 6), need not equal the ordinary $F$. For randomly sampled
bivariate-Normal pairs a zero population correlation, $\rho=0$, also
means a zero conditional-mean slope, but the occupation records have no
sampling warrant, so here the Pearson test only reproduces the process
calculation.

### A binary predictor and the two-sample t test {#ch07-binary-predictor}

A predictor need not be continuous. Chapter 4's invented feedback study
gave each of its 24 pupils a score out of 20, twelve under massed and
twelve under spaced practice. Code the schedule $x_i=0$ for massed and
$x_i=1$ for spaced, and fit the same kind of line. With only two
predictor values, the least-squares line joins the two group means
(@fig-ch7-binary): the intercept is the massed mean and the slope the
difference of means,

$$
\widehat\beta_0=\bar y_0=8,\qquad
\widehat\beta_1=\bar y_1-\bar y_0=12-8=4.
$$

The residual SD is the pooled SD, $s_e=4.156$ on $24-2=22$ df, and the
slope SE becomes
$SE(\widehat\beta_1)=s_e\sqrt{1/n_0+1/n_1}=4.156\sqrt{1/12+1/12}=1.6967$,
so $t=2.3575$ on 22 df, $p=.0277$ and the interval is
$[0.4813,\ 7.5187]$. These are exactly the equal-variance two-sample $t$
test and interval of Chapters 3 and 5, written as a line, with
$R^2=.2017$ and $F(1,22)=5.5579=t^2$; the Computer Lab reproduces the
interval with `t.test()`.

\Needspace{12\baselineskip}

```{r}
#| label: ch7-binary-predictor
#| echo: true
score <- c(4, 5, 7, 8, 4, 5, 7, 8, 6, 7, 9, 10, 10, 11, 13, 14,
           8, 9, 11, 12, 16, 17, 19, 20)
spaced <- rep(rep(c(0, 1), each = 4), 3)
two_groups <- lm(score ~ spaced)
coef(summary(two_groups))
```

```{r}
#| label: fig-ch7-binary
#| fig-cap: "With a 0/1 predictor, the least-squares line joins the two group means: the intercept is the massed mean and the slope the difference in means."
#| fig-alt: "Twenty-four invented scores plotted at 0 for massed and 1 for spaced practice, massed scores from 4 to 12 and spaced from 7 to 20. A line joins the massed mean 8 to the spaced mean 12, both marked with filled diamonds."
#| fig-width: 5.6
#| fig-height: 3
two_group_data <- data.frame(spaced, score)
group_means <- data.frame(spaced = c(0, 1),
  score = tapply(score, spaced, mean))
ggplot(two_group_data, aes(spaced, score)) +
  geom_point(shape = 1, size = 1.8,
    position = position_jitter(width = .04, height = 0, seed = 12345)) +
  geom_abline(intercept = coef(two_groups)[1],
    slope = coef(two_groups)[2], linewidth = .8) +
  geom_point(data = group_means, shape = 18, size = 4.5) +
  scale_x_continuous(breaks = c(0, 1),
    labels = c("0 (massed)", "1 (spaced)")) +
  coord_cartesian(xlim = c(-.25, 1.25)) +
  labs(x = "Practice schedule", y = "Score (out of 20)")
```

Welch's test (Chapter 3) uses the separate group SDs, 2.3741 and 5.3767,
and gives 15.1322 df and $p=.0323$; the line instead assumes one common
error variance. The HC3 covariance of Section 6 keeps the line and gives
a slope SE of 1.7721. A factor with more than two levels needs more than
one 0/1 column; Chapter 9 develops that coding, its contrasts and the
link to the analysis of variance of Chapter 4.

\Needspace{16\baselineskip}

## Slopes Across Repeated Samples {#sec-ch7-repetition}

### One sample from a known line

Section 3 described repetition in words; a simulation makes it visible.
Generate 20 outcomes at predictor values equally spaced from 6 to 16 from
a process we write ourselves:

$$
Y_i=2+0.5x_i+\varepsilon_i,\qquad
\varepsilon_i\overset{\mathrm{ind}}\sim\mathcal N(0,1).
$$

The true intercept is 2, the true slope 0.5 and the error SD 1; these are
a teaching construction, not occupation records.

```{r}
#| label: ch7-known-sample-code
#| echo: true
set.seed(123)
known_data <- data.frame(x = seq(6, 16, length.out = 20))
known_data$y <- 2 + 0.5 * known_data$x + rnorm(20, sd = 1)
known_model <- lm(y ~ x, data = known_data)
```

`set.seed()` fixes the random start, so the same 20 errors are drawn each
time, and `rnorm()` draws them with SD 1. @tbl-ch7-known-sample reports
the fit.

\Needspace{10\baselineskip}

```{r}
#| label: tbl-ch7-known-sample
#| tbl-cap: "One fitted sample from a process with known coefficients and error SD."
#| echo: false
known_summary <- coef(summary(known_model))
knitr::kable(data.frame(
  Quantity = c("Intercept", "Slope", "Residual SD"),
  Estimate = show_value(c(coef(known_model), sigma(known_model))),
  `Model SE` = c(show_value(known_summary[, "Std. Error"]), ""),
  check.names = FALSE
), align = c("l", "r", "r"))
```

The residual SD has no SE in the table because it is a scale estimate,
not a coefficient.

```{r}
#| label: fig-ch7-core-true-fitted
#| echo: false
#| fig-cap: "Known and fitted lines for one simulated dataset. The true process is programmed; the fitted line responds to the realized errors."
#| fig-alt: "Twenty simulated points surround the known line with intercept two and slope one half. A solid fitted line and dashed true line differ slightly."
ggplot2::ggplot(known_data, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(size = 1.6) +
  ggplot2::geom_abline(
    intercept = 2, slope = 0.5, linetype = "dashed", linewidth = 0.8
  ) +
  ggplot2::geom_abline(
    intercept = coef(known_model)[1],
    slope = coef(known_model)[2], linewidth = 0.8
  ) +
  ggplot2::labs(x = "Simulated predictor", y = "Simulated outcome")
```

In one sample the fitted slope and intercept miss their programmed
values. The error is the deviation from the true line and the residual
the deviation from the fitted line; residuals are tied together by their
zero-sum constraints, so a residual plot is an imperfect window on the
errors.

### Repeated samples and slope precision

Now hold 40 predictor values fixed and regenerate only the errors, 2,000
times, the generate--estimate--repeat route of Chapters 5 and 6. The
study size is 40; the repetition count, 2,000, only refines the
simulation's description of that study. Each repetition draws new
errors, adds them to the same 40 process means, fits a line, records its
slope and SE, and checks whether its 95% interval contains 0.5. The
Computer Lab runs the loop.

```{r}
#| label: ch7-repeated-slopes-code
#| echo: false
set.seed(321)
rep_x <- seq(6, 16, length.out = 40)
repetitions <- data.frame(
  slope = numeric(2000), se = numeric(2000), covered = logical(2000)
)
for (j in seq_len(nrow(repetitions))) {
  rep_y <- 2 + 0.5 * rep_x + rnorm(40, sd = 1)
  rep_fit <- lm(rep_y ~ rep_x)
  repetitions$slope[j] <- coef(rep_fit)[2]
  repetitions$se[j] <- coef(summary(rep_fit))[2, "Std. Error"]
  rep_ci <- confint(rep_fit)[2, ]
  repetitions$covered[j] <- rep_ci[1] <= 0.5 &&
    rep_ci[2] >= 0.5
}
repetition_summary <- c(
  mean_slope = mean(repetitions$slope),
  empirical_SD = sd(repetitions$slope),
  mean_estimated_SE = mean(repetitions$se),
  theoretical_SD = 1 / sqrt(sum((rep_x - mean(rep_x))^2)),
  empirical_coverage = mean(repetitions$covered)
)
```

\Needspace{9\baselineskip}

```{r}
#| label: tbl-ch7-repeated-slopes
#| tbl-cap: "Slope behavior across 2,000 repetitions of the known process."
#| echo: false
knitr::kable(data.frame(
  Quantity = c("Mean estimated slope", "SD of estimated slopes",
               "Mean estimated SE", "Theoretical slope SD",
               "Fraction of intervals covering 0.5"),
  Value = replace(show_value(unname(repetition_summary)), 5,
    sub("^0", "", show_value(unname(repetition_summary))[5]))
))
```

```{r}
#| label: fig-ch7-core-repetition
#| echo: false
#| fig-cap: "Slopes in 2,000 repetitions of the known process. The vertical reference is the true slope, 0.5."
#| fig-alt: "A roughly symmetric histogram of repeated fitted slopes is centered near one half; its spread represents repeated-sample slope variation."
#| fig-pos: H
#| fig-width: 6.3
#| fig-height: 2.8
ggplot2::ggplot(repetitions, ggplot2::aes(x = slope)) +
  ggplot2::geom_histogram(bins = 35, color = "white", fill = "grey50") +
  ggplot2::geom_vline(xintercept = 0.5, linewidth = 0.8, linetype = "dashed") +
  ggplot2::labs(x = "Estimated slope", y = "Number of repetitions")
```

The SD of the 2,000 slopes estimates their repeated-sample spread, and
each fitted SE estimates that spread from one sample; on average the two
agree closely. Coverage belongs to the interval procedure, not to one
realized interval, and with 2,000 repetitions its Monte Carlo SE at a
true .95 is $\sqrt{.95(.05)/2000}\approx0.0049$, so the observed .9455
lies within one Monte Carlo SE of .95.

The theoretical slope SD is $\sigma/\sqrt{S_{xx}}$.
@tbl-ch7-precision-conditions changes one feature of the design at a
time.

```{r}
#| label: ch7-precision-conditions-code
baseline_x <- seq(6, 16, length.out = 40)
more_x <- seq(6, 16, length.out = 160)
narrow_x <- seq(8.5, 13.5, length.out = 40)
baseline_sd <- 1 / sqrt(sum((baseline_x - mean(baseline_x))^2))
more_rows_sd <- 1 / sqrt(sum((more_x - mean(more_x))^2))
narrow_sd <- 1 / sqrt(sum((narrow_x - mean(narrow_x))^2))
more_noise_sd <- 2 / sqrt(sum((baseline_x - mean(baseline_x))^2))
precision_conditions <- data.frame(
  Condition = c("Baseline", "More rows", "Narrower support", "More noise"),
  n = c(40, 160, 40, 40), Lower = c(6, 6, 8.5, 6),
  Upper = c(16, 16, 13.5, 16),
  `Error SD` = c(1, 1, 1, 2),
  `Slope SD` = c(baseline_sd, more_rows_sd, narrow_sd, more_noise_sd),
  check.names = FALSE
)
```

\Needspace{9\baselineskip}

```{r}
#| label: tbl-ch7-precision-conditions
#| tbl-cap: "How sample size, predictor support, and error SD affect theoretical slope precision."
#| echo: false
precision_conditions$`Slope SD` <- show_value(
  precision_conditions$`Slope SD`)
knitr::kable(precision_conditions)
```

Halving the predictor range or doubling $\sigma$ doubles the slope SD;
quadrupling the points over the same range about halves it, not exactly,
because the two grids' variances differ slightly. Taking `x <- 1:n` with
a growing $n$ would change size and range together. More rows do not
remove confounding, measurement error, dependence or misspecification;
these simulations show behavior only under the mechanism programmed into
them.

\Needspace{9\baselineskip}

## Prediction {#sec-ch7-prediction}

At a chosen education value $x_0$ the line gives one number, which can
answer three questions of increasing uncertainty: what the line
predicts, what the mean income of occupations at $x_0$ is, and what one
new occupation's income will be. The last two need the process model of
Section 3.

### Point prediction

The **point prediction** at $x_0$ is the fitted line's value,

$$
\widehat y_0=\widehat\beta_0+\widehat\beta_1x_0,
$$

built from the estimates, not the unknown parameters. At $x_0=12$ it is
7.9322 thousand dollars, the value `predict()` returned in Section 2.1.
By itself a point prediction says nothing about its uncertainty, and
the same number serves two different targets: it estimates the mean
income of occupations at $x_0$ and predicts one new occupation's income.

### The mean response and its confidence interval

Under the process model, occupations with education $x_0$ have mean
income $\mu_0=\beta_0+\beta_1x_0$, the **mean response**, a process
parameter. The point prediction estimates it, $\widehat\mu_0=\widehat y_0$,
with standard error

$$
SE(\widehat\mu_0)
=s_e\sqrt{\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}}
=3.4834\sqrt{\frac1{102}+
 \frac{(12-10.738)^2}{751.8852}}=0.3803
$$

at 12 years. The 95% **confidence interval for the mean response** is
$7.9322\pm1.984(0.3803)\approx[7.1776,\ 8.6868]$. The SE is smallest at
$\bar x$, grows with distance from it, and shrinks as independent records
accumulate: it describes uncertainty about an average, not the spread of
individual occupations. `predict()` returns the fit and the two limits:

```{r}
#| label: ch7-interval-mean
#| echo: true
predict(model, data.frame(education = 12), interval = "confidence")
```

### A new case and its prediction interval

One new occupation at $x_0$ has income $Y_0=\mu_0+\varepsilon_0$, a
random outcome with its own error. Predicting it by $\widehat\mu_0$ makes
an error with two parts,

$$
Y_0-\widehat\mu_0
=\varepsilon_0-(\widehat\mu_0-\mu_0),
$$

the new case's own error and the error in the estimated mean. When the
new error is independent of the fitted data and comes from the same
common-variance process, their variances add:

$$
\operatorname{Var}(Y_0-\widehat\mu_0\mid\mathbf X,x_0)
=\sigma^2+\operatorname{Var}(\widehat\mu_0\mid\mathbf X,x_0).
$$

Substituting the mean-response variance and estimating $\sigma$ by $s_e$
gives

$$
SE_{\mathrm{new}}
=s_e\sqrt{1+\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}}
=3.5041,
$$

and the 95% **prediction interval**
$7.9322\pm1.984(3.5041)\approx[0.9802,\ 14.8842]$. The extra 1 is the new
outcome's own error variance, in units of $\sigma^2$; unlike the
mean-response SE, it does not shrink as the estimation sample grows.

```{r}
#| label: ch7-interval-new
#| echo: true
predict(model, data.frame(education = 12), interval = "prediction")
```

### Confidence and prediction intervals compared

```{r}
#| label: fig-ch7-intervals
#| fig-cap: "Confidence intervals for the mean response (dark band) and prediction intervals for one new occupation (light band), from the same fitted line."
#| fig-alt: "The mean-confidence band is narrowest near average education. The new-case prediction band is much wider throughout the observed education range."
#| fig-width: 6.3
#| fig-height: 3.3
grid <- data.frame(education = seq(min(x), max(x), length.out = 150))
cm <- predict(model, grid, interval = "confidence")
pi <- predict(model, grid, interval = "prediction")
grid$fit <- cm[, 1]
grid$cl <- cm[, 2]; grid$cu <- cm[, 3]
grid$pl <- pi[, 2]; grid$pu <- pi[, 3]
ggplot(grid, aes(education, fit)) +
  geom_ribbon(aes(ymin = pl, ymax = pu), fill = "grey90") +
  geom_ribbon(aes(ymin = cl, ymax = cu), fill = "grey65") +
  geom_line(linewidth = .7) +
  geom_point(data = prestige, aes(y = income_k), shape = 1, size = 1.3) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)")
```

@fig-ch7-intervals draws both intervals across the observed education
range. The prediction band is much wider everywhere, because it carries a
new occupation's own scatter as well as the uncertainty about the line;
the confidence band carries only the latter. Both widen away from
$\bar x$. Below about 11 years the prediction band's lower limit falls
below zero, which exposes the limits of a Normal model for a nonnegative
outcome rather than forecasting negative income. Outside 6.38--15.97
years the software still returns numbers, but the records cannot check
them. Neither interval measures how well the line predicts records it
has not seen; that held-out question belongs to Chapter 8. Appendix A
derives the mean-response variance from the coefficient variances and
their covariance, and gives both intervals at four education values.

## Diagnostic Checks {#sec-ch7-check}

### Residuals, influence and unequal spread

The fitted line exists whether or not any process model holds. Claims
that it describes a linear conditional mean, that its interval has the
stated coverage, or that it predicts new occupations well need more
support, and a check earns its place when its result could change one of
those claims.

```{r}
#| label: fig-ch7-checks
#| fig-cap: "Residual pattern and influence invite different follow-up questions."
#| fig-alt: "The left panel shows curvature and uneven scatter in residuals. The right panel shows Cook's distances, with physicians most influential for this fit."
#| fig-width: 6.4
#| fig-height: 2.6
op <- par(mfrow = c(1, 2), mar = c(4, 4, 1, 1), mgp = c(2.5, .7, 0))
plot(fitted(model), resid(model), xlab = "Fitted income", ylab = "Residual",
     pch = 1, cex = .7)
abline(h = 0, col = "grey60")
lines(lowess(fitted(model), resid(model)), lwd = 2)
plot(cooks.distance(model), type = "h", xlab = "Occupation row",
     ylab = "Cook's distance")
check_most <- which.max(cooks.distance(model))
text(check_most, cooks.distance(model)[check_most], "physicians",
     pos = 4, cex = .7)
par(op)
```

A curve in the residuals (@fig-ch7-checks, left) suggests that one
straight conditional mean leaves systematic variation, and a residual
spread that changes with the fitted value challenges the common-variance
SE. Leverage concerns an unusual predictor value; influence concerns how
much the fit changes when a record is left out. A large residual, high
leverage and high influence are related but different properties.

\Needspace{7\baselineskip}

For a straight line with an intercept, the leverage and Cook's distance
of record $i$ are

$$
h_i=\frac1n+\frac{(x_i-\bar x)^2}{S_{xx}},
\qquad
D_i=\frac{e_i^2}{2s_e^2}\frac{h_i}{(1-h_i)^2}.
$$

Leverage grows with distance from $\bar x$; Cook's distance combines it
with the residual to summarize how far the fitted values move when record
$i$ is omitted (@fig-ch7-checks, right). It prompts a look at provenance
and a sensitivity analysis, not automatic deletion.

Adding education squared lets the fit bend. This same-record comparison
is an exploratory check, not a new final model:

```{r}
#| label: ch7-check-models
#| echo: true
quad <- lm(income_k ~ education + I(education^2), data = prestige)
most <- which.max(cooks.distance(model))
without <- lm(income_k ~ education, data = prestige[-most, ])
c(full = unname(coef(model)[2]), without = unname(coef(without)[2]))
coef(summary(quad))["I(education^2)", ]
```

```{r}
#| label: fig-ch7-curvature
#| fig-cap: "Straight and quadratic fits to the same 102 records. The curve is an exploratory sensitivity, not a selected final model."
#| fig-alt: "The quadratic fit bends upward toward higher education, unlike the constant-slope straight line. Both fits use all 102 occupation records on the same income scale."
#| fig-width: 6.3
#| fig-height: 3.7
check_grid <- data.frame(
  education = seq(min(x), max(x), length.out = 200)
)
check_curves <- rbind(
  data.frame(check_grid, Fit = "Straight line",
             Income = predict(model, check_grid)),
  data.frame(check_grid, Fit = "Quadratic sensitivity",
             Income = predict(quad, check_grid))
)
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, colour = "grey50", size = 1.5) +
  geom_line(data = check_curves,
            aes(y = Income, linetype = Fit), linewidth = .8) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)", linetype = NULL)
```

The quadratic coefficient, 0.1342, has $p=.0059$ under that model's
classical calculation: a reason to look at the curvature rather than a
model-selection rule, since the spread is unequal and the comparison was
chosen after seeing the records. Omitting physicians, the record with
the largest Cook's distance (0.3983), moves the slope from 0.8988 to
0.7982. Nothing shows that physicians were recorded wrongly; the deletion
is a sensitivity disclosure, not a replacement sample.

\Needspace{7\baselineskip}

An HC3 heteroskedasticity-consistent covariance gives a slope SE of
0.1491 rather than 0.1270 and, with the same $t_{100}$ multiplier, an
approximate interval $[0.6031,\ 1.1945]$. The line does not change. HC3
relaxes the equal-variance assumption for independent records; it does
not repair curvature, dependence, confounding or measurement error
[@white1980; @mackinnon1985; @long2000; @zeileis2004].

\Needspace{10\baselineskip}

```{r}
#| label: ch7-hc3-visible
#| echo: true
V_hc3 <- sandwich::vcovHC(model, type = "HC3")
c(classical = coef(summary(model))["education", "Std. Error"],
  HC3 = sqrt(V_hc3["education", "education"]))
```

Later chapters go further: Chapter 8, Appendix B, builds the HC3 matrix
from its parts; Chapter 9, Appendix A, defines what a least-squares line
estimates when the conditional mean is not linear, and its Section 5.3
shows that measurement error in the predictor attenuates a slope. Nothing
here relies on them. @tbl-ch7-diagnostic-consequences collects the five
checks and what each can change.

\Needspace{20\baselineskip}

| Check | What needs to be examined | Consequence |
|:--|:--|:--|
| Form | Residual curve and a curved-fit comparison | A straight line may remain a description but fail as a constant conditional-mean slope account |
| Spread | Residual variation over fitted values | Classical SEs may need revision; HC3 is a variance sensitivity, not a universal repair |
| Influence | Source record and result with/without a consequential case | Disclose sensitivity; correct errors only when verified |
| Independence | How occupations and measurements were generated or related | A plot cannot establish independent process realizations |
| Support | Actual education range and density of observations | Restrict comparisons; avoid extrapolation and individual-level claims |

: Diagnostic questions and their consequences. {#tbl-ch7-diagnostic-consequences tbl-colwidths="[16,38,46]"}

\Needspace{24\baselineskip}

## Regression Toward the Mean {#sec-ch7-rtm}

### The line in standard units

Francis Galton, who gave regression its name, measured the adult heights
of parents and their grown children in 205 families in the 1880s
[@galton1886]. We use James Hanley's transcription of Galton's notebook
[@hanley2004], distributed in the R package HistData [@friendly2025], and
keep one row per family: among the 179 families with at least one son,
one son was drawn at random from each, so every father appears once and
the rows describe separate families rather than brothers who share a
father. Heights are in inches, mostly recorded to the nearest half inch.
The small extract `galton1886_sons.csv` is
supplied with the chapter under this note: *Galton's family heights
(published 1886), transcribed by J. A. Hanley and distributed in the R
package HistData under the GPL; one son per family drawn for QMSBR. The
measurements are historical records in the public domain; the extract is
redistributed with this attribution.*

Chapter 2 measured a value's distance from its mean in standard units,
$z_x=(x-\bar x)/s_x$, and Section 2.2 wrote the slope as
$\widehat\beta_1=r\,s_y/s_x$. Substituting that slope into the line
through the mean point, $\widehat y-\bar y=\widehat\beta_1(x-\bar x)$,
and dividing by $s_y$ gives the line in standard units:

$$
\frac{\widehat y-\bar y}{s_y}=r\,\frac{x-\bar x}{s_x},
\qquad\text{that is,}\qquad
\widehat z_y=r\,z_x .
$$

The line passes through the origin with slope $r$. Unless the
correlation is perfect, $\lvert r\rvert<1$, so a case $z_x$ SDs from the
mean on the predictor is predicted only $r\,z_x$ SDs from the mean on
the outcome. Galton called this regression toward mediocrity; we call it
**regression toward the mean**.

```{r}
#| label: ch7-galton-visible
#| echo: true
galton <- read.csv("galton1886_sons.csv")
sapply(galton[c("father", "son")], sd)
c(r = cor(galton$father, galton$son),
  z_slope = unname(coef(lm(scale(son) ~ scale(father),
    data = galton))[2]))
```

`sapply()` applies `sd()` to both columns, and `scale()` converts a
column to standard units. Fathers average 69.0989 inches (SD 2.5466),
sons 69.2140 inches (SD 2.6475), and $r=.4269$. In inches the fitted slope is
0.4438 inch of son's height per inch of father's height; in standard
units it is .4269, which is $r$. A father 1.6 SD above the fathers' mean
has a son predicted $.4269\times1.6=0.6831$ SD above the sons' mean:
taller than average, but less than half as far out.

```{r}
#| label: fig-ch7-rtm
#| fig-cap: "Galton's 179 fathers and sons in standard units. The fitted line has slope $r=.4269$, flatter than the dashed line of slope 1; the filled points are the mean positions of the fathers more than 1 SD above and below their mean, with their sons."
#| fig-alt: "A cloud of 179 points, fathers' heights in standard units from about −2.8 to 3.7 across and sons' from about −3.5 to 2.8 up. A dashed line of slope 1 and a solid fitted line of slope about 0.43 cross at the origin. Two filled diamonds, near (1.6, 0.7) and (−1.5, −0.6), sit close to the solid line, inside the dashed one."
#| fig-width: 5.4
#| fig-height: 4.6
z_father <- as.numeric(scale(galton$father))
z_son <- as.numeric(scale(galton$son))
r_galton <- cor(galton$father, galton$son)
galton_z <- data.frame(z_father, z_son)
tall <- z_father > 1
short <- z_father < -1
group_means <- data.frame(
  z_father = c(mean(z_father[tall]), mean(z_father[short])),
  z_son = c(mean(z_son[tall]), mean(z_son[short])))
ggplot(galton_z, aes(z_father, z_son)) +
  geom_point(shape = 1, colour = "grey55", size = 1.4) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
    linewidth = .7) +
  geom_abline(intercept = 0, slope = r_galton, linewidth = .9) +
  geom_point(data = group_means, shape = 18, size = 4.5) +
  annotate("text", x = 3.7, y = c(-2.6, -3.2), hjust = 1, size = 3.6,
    label = c("solid: fitted line, slope r = .4269",
      "dashed: slope 1")) +
  coord_equal(xlim = c(-3.6, 3.8), ylim = c(-3.6, 3.8)) +
  labs(x = "Father's height (standard units)",
       y = "Son's height (standard units)")
```

In @fig-ch7-rtm the dashed line of slope 1 marks where a son would sit if
he were exactly as many SDs from his mean as his father. The fitted line
is flatter, lying below the dashed line to the right of the mean and
above it to the left: tall fathers have sons predicted less tall, and
short fathers sons predicted less short, in standard units. The two filled points, the
average positions of the fathers more than 1 SD above and below their
mean with their sons, sit near the fitted line, well inside the dashed
one.

### Extreme groups and the unchanged spread

@tbl-ch7-rtm-groups groups the men in bands of 1 SD, first by the fathers'
heights and then by the sons'. Each band includes its upper bound, so the
four bands of each block hold all 179 men.

\Needspace{14\baselineskip}

| Selected on | Band (SDs) | Men | Mean $z$ of the selected height | Mean $z$ of the other height | $r$ times the selected mean |
|:--|:--|--:|--:|--:|--:|
| Fathers | Above $+1$ | 24 | 1.6072 | 0.7029 | 0.6862 |
| Fathers | 0 to $+1$ | 60 | 0.4978 | 0.2251 | 0.2125 |
| Fathers | $-1$ to 0 | 65 | $-0.3463$ | $-0.2058$ | $-0.1479$ |
| Fathers | Below $-1$ | 30 | $-1.5310$ | $-0.5668$ | $-0.6536$ |
| Sons | Above $+1$ | 31 | 1.4057 | 0.6224 | 0.6001 |
| Sons | 0 to $+1$ | 60 | 0.4650 | 0.1431 | 0.1985 |
| Sons | $-1$ to 0 | 61 | $-0.4610$ | $-0.0697$ | $-0.1968$ |
| Sons | Below $-1$ | 27 | $-1.6057$ | $-0.8751$ | $-0.6855$ |

: Groups of Galton's fathers and sons in standard units, selected on either height. {#tbl-ch7-rtm-groups tbl-colwidths="[12,14,8,22,22,22]"}

The 24 fathers more than 1 SD above their mean average $z=1.6072$, or
73.1917 inches; their sons average 0.7029, or 71.0750 inches, close to
the line's prediction $.4269\times1.6072=0.6862$. The 30 short fathers
mirror them, and in every row the other height's mean lies between zero
and the selected mean, near $r$ times it. Selecting on the sons gives the
same movement backward in time: the 31 sons more than 1 SD above their
mean, at 1.4057, had fathers averaging only 0.6224. No force pulls
heights toward the average; a force could not act backward in time. The
imperfect correlation guarantees only that the line's predictions, in
standard units, lie nearer the mean than the values they are predicted
from. Groups follow those predictions on average when the mean of one
height is roughly a straight line in the other, as it is in these
families; even then each group's mean scatters around its prediction, so
no particular extreme group is guaranteed to sit nearer the mean.

If tall and short fathers both have sons nearer the mean, are the sons
less spread out than their fathers? No: the sons' SD, 2.6475 inches, is
slightly larger than the fathers', 2.5466. This is the **regression
paradox**. The fitted values carry the part of a son's height that his
father's height predicts, and spread only $r\,s_y=1.1303$ inches; the
residuals carry the rest, with spread $s_y\sqrt{1-r^2}=2.3941$ inches
(both SDs with the $n-1$ divisor; `sigma()` divides by $n-2$ and gives
2.4008). Variances add, not SDs: $1.1303^2+2.3941^2=2.6475^2$, which is
Section 2.6's decomposition divided by $n-1$, in which the fitted part
holds the share $R^2=r^2=.1823$. Some sons of average fathers are very tall or very
short, so the sons' generation is as varied as the fathers'. Regression
toward the mean describes predictions for selected groups, not a
population growing more uniform.

### A hypothetical class: luck that does not repeat

Heights show the pattern; a second case, invented for this chapter,
shows one mechanism behind it. Imagine a large class of 2,000 students who sit a
midterm and a final, with scores generated so that nothing between the
exams has any effect: each student has a stable true score $T_i$, and
each exam adds independent luck of the same size,

$$
M_i=T_i+\varepsilon_{i1},\qquad F_i=T_i+\varepsilon_{i2},\qquad
T_i\sim\mathcal N(65,\,8^2),\qquad
\varepsilon_{i1},\,\varepsilon_{i2}\overset{\mathrm{ind}}\sim\mathcal N(0,\,6^2).
$$

Chapter 1 described reliability as the consistency of scores; here it is
the share of score variance that comes from the true score,
$8^2/(8^2+6^2)=.64$ on both exams. Only the true score repeats, so .64
is also the expected correlation of the two exams, the slope of the line
in standard units that the sample $r$ estimates. The class is large so
that its group means carry little chance variation. Four lines of R
generate the scores:

```{r}
#| label: ch7-true-score-visible
#| echo: true
set.seed(12345)
true_score <- rnorm(2000, mean = 65, sd = 8)
midterm <- true_score + rnorm(2000, mean = 0, sd = 6)
final <- true_score + rnorm(2000, mean = 0, sd = 6)
```

```{r}
#| label: fig-ch7-true-score
#| fig-cap: "A hypothetical class of 2,000 students, each score a stable true score plus independent luck. The top and bottom tenth on the midterm (diamonds) score nearer the class mean on the final, at their unchanged mean true score (dashed); the boxes, the middle half of the class, span about the same range of scores on the final as on the midterm."
#| fig-alt: "Two grey boxes for the whole class's midterm and final scores, each spanning about 58 to 72 points around a dotted class mean near 65. A line falls from a diamond at about 82 on the midterm to one at about 77 on the final, landing on a dashed line at about 77 labelled as that group's mean true score; a mirror line rises from about 48 to about 53, landing on a dashed line at about 53."
#| fig-width: 6.3
#| fig-height: 3.3
exams <- c("Midterm", "Final")
scores <- data.frame(exam = factor(rep(exams, each = 2000),
  levels = exams), score = c(midterm, final))
top <- midterm > quantile(midterm, .9)
bottom <- midterm < quantile(midterm, .1)
groups <- data.frame(group = rep(c("top", "bottom"), each = 2),
  exam = factor(rep(exams, 2), levels = exams),
  score = c(mean(midterm[top]), mean(final[top]),
    mean(midterm[bottom]), mean(final[bottom])))
levels_y <- c(mean(true_score[top]), mean(true_score[bottom]),
  mean(c(midterm, final)))
ggplot(scores, aes(exam, score)) +
  geom_boxplot(width = .3, colour = "grey65", fill = "grey96",
    outlier.shape = NA, coef = 0) +
  annotate("segment", x = .7, xend = 2.3, y = levels_y,
    yend = levels_y, linetype = c("dashed", "dashed", "dotted"),
    colour = c("black", "black", "grey45"), linewidth = .6) +
  geom_line(data = groups, aes(group = group), linewidth = .9) +
  geom_point(data = groups, shape = 18, size = 4.5) +
  annotate("text", x = .66, y = groups$score[c(1, 3)], hjust = 1,
    size = 3.4, label = c("top tenth\non the midterm",
      "bottom tenth\non the midterm")) +
  annotate("text", x = 2.34, y = levels_y, hjust = 0, size = 3.4,
    label = c("their mean\ntrue score", "their mean\ntrue score",
      "class mean")) +
  coord_cartesian(xlim = c(.2, 2.8), ylim = c(40, 90)) +
  labs(x = NULL, y = "Exam score (points)")
```

The simulated exams correlate at $r=.6509$, close to the expected .64. In
@fig-ch7-true-score the
200 students in the top tenth on the midterm averaged 82.3217 points, and
76.5335 on the final: still well above the class mean, but nearer it.
Their true scores average 76.5186, so the midterm overstated them.
Students picked for a high score are mostly strong but, on average, also
lucky that day; on the final their luck is drawn afresh, and the group
lands on its true level. The bottom tenth mirror them, rising from
47.6984 to 53.3550 toward a mean true score of 53.2729. No true score
changed, and no one was praised, tutored or discouraged. The boxes show
the other half of the story: the class is as spread out on the final (SD
10.0853) as on the midterm (SD 9.8406), because fresh luck carries other
students into the extremes. The Computer Lab reruns the simulation with
less luck.

### The regression fallacy

Regression toward the mean becomes a fallacy when the movement it
predicts is credited to a cause. Tversky and Kahneman [-@tversky1974]
recount a conversation with flight instructors: a trainee praised for an
unusually smooth landing tended to land less well next time, and a
trainee reprimanded for a rough landing tended to do better, so the
instructors concluded that praise hurts and criticism helps.

The hypothetical class explains the pattern without either conclusion.
A landing reflects the trainee's skill and the luck of that attempt, so
two landings by one trainee are correlated, but not perfectly. The
landings that drew comment were selected for being extreme, and the next
landing is predicted nearer the trainee's usual level whatever the
instructor says. Crediting that movement to the feedback is the
**regression fallacy**.

The same pattern appears whenever cases are selected for an extreme first
measurement and measured again: pupils chosen for low test scores and
given extra help, patients enrolled when their symptoms peak. Such
groups move toward the mean on average even if nothing done in between
has any effect. A comparison group selected the same way and measured at
the same two times shows how much movement selection alone produces;
assigning the help at random among the selected cases does better still,
since both groups then regress alike and their difference estimates the
help's effect [@barnett2005]. Chapter 14, Section 4.1,
applies this reasoning to selection on a noisy baseline.

## Reading Bridge {#sec-ch7-summary}

A simple regression starts with a unit and a comparison. Its fitted
values are the line's predictions for the records and its residuals the
records' distances from them; least squares puts the line through the
mean point, with the slope set by covariance and predictor variation.
Across the 102 recorded occupations, fitted income rises by 0.8988
thousand 1971 Canadian dollars, about 899 dollars, per year of
occupation-average education over the observed 6.38--15.97 years: a
bounded description of the records, not an individual, causal or
population return. The residual SD, $R^2$, the slope's standard error,
and the mean-response and new-case intervals answer different questions,
and the checks say which answers the records support. That is the
chapter's through-line: the line predicts, and, once a process model is
declared, standard errors and intervals say how uncertain its slope and
predictions are. In standard units
the line predicts $r$ times a case's distance from the mean, so groups
selected for extreme values are predicted nearer the mean on a second
measurement while the spread stays; a before-and-after gain in such a group needs a
comparison selected the same way before it is read as an effect.
Supplementary A reads a short report of the occupation line as an
argument.

Chapter 8 first refits the education-only line on the 98 occupations
with a recorded type, so that the change from 102 to 98 records is not
mistaken for an effect of adjustment; it then adds prestige, percentage
of women and occupation type, and asks a separate held-out prediction
question. Chapter 9 collects the regression toolbox, from factors with
several levels to robust and resampled uncertainty, and its Section 7
(Project STAR) reads a randomized class-size experiment whose
uncertainty follows its school-clustered design [@finn1990;
@mosteller1995]. Chapter 10 extends the factorial interactions of
Chapters 4 and 6 to regression predictors.

### Key terms {.unnumbered}

Use this short reference to revisit the regression concepts introduced
here. The Part One Concept Reference gives the broader reference and the
principal home of each concept; fitted values, residuals, standard errors,
intervals, standardized values, and the reading vocabulary are explained
in Chapters 1--4.

| Term and common alternatives | Meaning |
|:--|:--|
| Simple linear regression (regression line) | A straight-line summary of how one outcome differs across one predictor: fitted value equals intercept plus slope times the predictor. Its descriptive use concerns the records fitted; a model-based use needs a declared process. |
| Intercept | The fitted outcome at predictor value zero. It places the line vertically; when zero is outside the observed support, it is a mathematical anchor rather than a supported prediction. Centering the predictor moves it to a supported reference point. |
| Slope (regression coefficient) | The difference in fitted outcome per one-unit difference in the predictor, in outcome units per predictor unit. It compares units that differ in the predictor; it is not a change within one unit unless a design supports that reading. |
| Ordinary least squares (OLS) | The rule that chooses the intercept and slope minimizing the residual sum of squares. With an intercept, the fitted line passes through the mean point and the residuals sum to zero; the slope equals covariance divided by predictor variance. |
| Residual sum of squares (RSS) | The sum of squared residuals from a fitted line, in squared outcome units. Total squared variation about the mean splits into a regression part and RSS when the model has an intercept. |
| Residual standard deviation (residual standard error) | The square root of RSS divided by its residual degrees of freedom, in outcome units; with an intercept and one predictor the degrees of freedom are $n-2$. Under the process model it estimates the error SD. |
| R-squared (coefficient of determination) | One minus RSS divided by total squared variation: the proportion of centered squared outcome variation accounted for by the fit in these records. In simple regression with an intercept it equals the squared correlation. |
| Confidence interval for a mean response (mean-response interval) | An interval for the process mean outcome at a stated predictor value, using the SE of the fitted value. It concerns an average, narrows near the predictor mean, and does not cover individual outcomes. |
| Prediction interval (new-case interval) | An interval for one new outcome at a stated predictor value from the same process. Its variance adds the error variance to the fitted-mean variance, so it is wider than the mean-response interval and keeps a nonzero floor as the sample grows. |
| Leverage | A measure of how unusual a record's predictor value is relative to the others; it grows with distance from the predictor mean and appears in the SE of a fitted value and in influence measures. |
| Cook's distance (influence) | A summary of how much the fitted values change when one record is omitted, combining its residual and its leverage. It prompts inspection of provenance and a disclosed sensitivity, not automatic deletion. |
| Heteroskedasticity-consistent covariance (HC3; sandwich estimator) | A coefficient covariance estimate that allows unequal error variance across predictor values under independence. It changes SEs and intervals, not the fitted coefficients, and does not repair curvature, dependence, measurement error, or confounding. |
| Extrapolation (predictor support) | Using a fitted line at predictor values outside the observed range or where observations are sparse. Software returns a number there; the data do not support it. |
| Regression toward the mean (regression to the mean) | In standard units a case's predicted outcome is $r$ times its predictor value, $\widehat z_y=r\,z_x$, so with $\lvert r\rvert<1$ cases extreme on one measurement are predicted nearer the mean on the other, and a group selected for an extreme noisy first value is expected to move toward the mean on a second measurement, on average rather than in every group. The outcome's spread does not shrink (the regression paradox). |
| Regression fallacy | Attributing to a treatment, a talent or another cause the movement toward the mean that selection on an extreme noisy measurement produces by itself. A comparison group selected the same way, or random assignment, separates a real change from regression toward the mean. |

: Selected concepts for simple linear regression. {#tbl-ch07-key-terms tbl-colwidths="[28,72]"}

### Further reading {.unnumbered}

For another accessible account of lines, residuals, and fit, see
[OpenIntro's single-predictor regression tutorial](https://openintro-ims.netlify.app/model-slr).
For a compact executable route from a fitted model to intervals and checks, see
the [ISLR authors' Chapter 3 R lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch3-linreg-lab.html)
[@james2021].
Consult
[the carData reference manual](https://cran.r-project.org/web/packages/carData/carData.pdf)
[@carData2022] for data documentation, @barnett2005 on regression to the
mean in repeated measurements, and @fox2016, @fox2019, and @gelman2020 for
broader regression treatments.

\Needspace{18\baselineskip}

## Exercises {.unnumbered}

The prompts below use **Foundational** and **Integrative** levels. Where
rounded summaries are supplied, answers within 0.02 in the stated units are
sufficient. Exercises 1--6 draw on the main chapter.

<!-- prompt-id: CH07-E01 -->
### Exercise 1: Read the unit and the line (Foundational) {.unnumbered #ex-ch07-01}

<!-- exercise-id: ch07-ex01 -->

A teaching report describes 80 occupations with
$\widehat{\mathrm{income}}=-1.5+0.7\,\mathrm{education}$, where income is in
thousand dollars and average education ranges from 7 to 16 years.
Interpret the slope and the intercept. Calculate the fitted income at 12 years
and the residual for an occupation at 12 years with income 9.
Explain why the slope does not give an individual's gain from another year
of schooling.

<!-- prompt-id: CH07-E02 -->
### Exercise 2: Connect covariance and the fitted line (Foundational) {.unnumbered #ex-ch07-02}

<!-- exercise-id: ch07-ex02 -->

A set of records has $\bar x=4,\bar y=10,S_{xx}=20,S_{xy}=30$.
Calculate the OLS slope and intercept, and the fitted value at $x=5$.
Show that the line passes through the mean point. If $s_x=2$ and
$s_y=5$, calculate the correlation implied by these summaries.
What would prevent fitting a slope if every $x$ were equal?

<!-- prompt-id: CH07-E03 -->
### Exercise 3: Separate fit and uncertainty (Foundational) {.unnumbered #ex-ch07-03}

<!-- exercise-id: ch07-ex03 -->

An intercept-and-slope regression has $n=22$, $RSS=80$, $TSS=144$,
$S_{xx}=100$, and fitted slope 0.8. Calculate residual SD, $R^2$, and the
classical slope SE. Use $t_{.975,20}=2.086$ to obtain a 95% slope interval.
State the process assumptions needed for that interval and explain why it
does not by itself establish that the line fits closely or identifies a cause.

<!-- prompt-id: CH07-E04 -->
### Exercise 4: Mean or new case? (Foundational) {.unnumbered #ex-ch07-04}

<!-- exercise-id: ch07-ex04 -->

Under the stated independent normal-error constant-variance model, a fitted
value at $x_0=\bar x$ is 12, $n=25$, and $s_e=3$.
Use $t_{.975,23}=2.069$. Calculate the SE and 95% interval for the conditional
mean, then the corresponding prediction SE and interval for one independent
new case at the same $x_0$. Explain the extra 1 in the prediction expression.
Would either interval validate the model in a different population?

\Needspace{23\baselineskip}

<!-- prompt-id: CH07-E05 -->
### Exercise 5: Tell regression toward the mean from an effect (Integrative) {.unnumbered #ex-ch07-05}

<!-- exercise-id: ch07-ex05 -->

In an invented study, a school records a reading test in September and
again in June for all 200 of its pupils. Both tests have mean 50 and SD
10, and their correlation is .6. The 20 pupils with the lowest September
scores, averaging 32 points, receive extra tutoring, and their June mean
is 42.8 points. These figures include the tutored pupils' June scores, so
by themselves they cannot show what those scores would have been without
tutoring. The school's records from earlier years, when no pupil was
tutored, supply that benchmark: the two tests again had means 50, SDs 10
and correlation .6, and the mean June score of pupils with the same
September score followed a straight line in that score. Assume that this
year's pupils would have followed the same pattern without tutoring. (1)
Express the tutored group's September mean in standard units, and give
the June mean that this pattern's line in standard units predicts for
pupils with that September score, in standard units and in points. (2)
The school reports that tutoring raised the group by 10.8 points. Name
the error in reading the whole gain as an effect of tutoring, and say how
much of the gain regression toward the mean predicts by itself under
this pattern. (3) The report adds that, since the lowest
scorers caught up, the June scores of all 200 pupils must be less spread
out than September's. Split a June score into the part that the
September score predicts and the part it does not, give the SD of each
part, and explain why regression toward the mean leaves the June SD at 10.
(4) Describe a comparison that would separate an effect of tutoring from
regression toward the mean.

<!-- prompt-id: CH07-E06 -->

\Needspace{16\baselineskip}

### Exercise 6: Respond to sensitivity (Integrative) {.unnumbered #ex-ch07-06}

<!-- exercise-id: ch07-ex06 -->

In a descriptive analysis of 102 occupations, the slope is 0.9.
A valid, high-income occupation has the largest Cook's distance; removing it
gives 0.79. A curved-fit check changes the shape, and an HC3 calculation changes
the slope SE from 0.13 to 0.15 without changing the line.

What does each check tell you? What does it not tell you? Explain why deleting
the occupation automatically is inappropriate. Distinguish a variance repair
for the same estimand from replacing a constant process slope with a different
estimand. Write a short sensitivity disclosure for the descriptive analysis.

{{< pagebreak >}}

## Appendix A: Extended regression mathematics and interval derivations {.unnumbered #sec-ch7-app-a}

This appendix derives the identities the main text uses, keeping
identities about the observed records apart from claims about
hypothetical repetitions. Occupation calculations use the same 102 rows,
education in years and income in thousand 1971 Canadian dollars.

### Derivatives, normal equations, and the minimizing line {.unnumbered}

Start with a fitted model that predicts one common center $m$ for every
observation. Chapter 2 Section 2.2 states that the mean minimizes the
sum of squared deviations, and Chapter 3 Appendix A derives the identity
behind it:

$$
L(m)=\sum_i(y_i-m)^2
=\sum_i(y_i-\bar y)^2+n(m-\bar y)^2.
$$

Write $y_i-m=(y_i-\bar y)+(\bar y-m)$ and expand the square. The cross
term is zero because $\sum_i(y_i-\bar y)=0$. For $n>0$, the last term
is nonnegative and vanishes only at $m=\bar y$, so the mean uniquely
minimizes this criterion.

The three outcomes $2,4,3$ from Section 2.3 have mean 3 and
$L(m)=2+3(m-3)^2$, so their best constant line has loss 2. Letting the
fitted value depend on the three predictor values reduces RSS to 1.5,
as @tbl-ch7-tiny showed; neither result requires Normal errors or a
probability sample. Chapter 6's likelihood example chose a parameter by
optimizing a different criterion; here we develop the least-squares line
algebraically.

The intercept-and-slope criterion has two coefficients to determine, so
we set its two partial derivatives to zero and then verify that the
resulting line is the minimum. For candidate intercept $b_0$ and slope
$b_1$, define

$$
Q(b_0,b_1)=\sum_{i=1}^n(y_i-b_0-b_1x_i)^2.
$$

Differentiating with respect to each candidate coefficient gives

$$
\frac{\partial Q}{\partial b_0}
=-2\sum_i(y_i-b_0-b_1x_i),\qquad
\frac{\partial Q}{\partial b_1}
=-2\sum_i x_i(y_i-b_0-b_1x_i).
$$

At the minimum both derivatives are zero. The resulting **normal equations**
are

$$
n\widehat\beta_0+\widehat\beta_1\sum_i x_i=\sum_i y_i,
\qquad
\widehat\beta_0\sum_i x_i+\widehat\beta_1\sum_i x_i^2
=\sum_i x_iy_i.
$$

The first equation gives
$\widehat\beta_0=\bar y-\widehat\beta_1\bar x$.
Substitution into the second yields

$$
\widehat\beta_1
=\frac{\sum_i x_iy_i-n\bar x\bar y}
       {\sum_i x_i^2-n\bar x^2}
=\frac{S_{xy}}{S_{xx}}.
$$

Here $S_{xx}=\sum_i(x_i-\bar x)^2>0$ is required: a slope cannot be
identified when every predictor value is the same. Because sample covariance
and variance use the same denominator $n-1$,

$$
\widehat\beta_1
=\frac{s_{xy}}{s_x^2}
=r\frac{s_y}{s_x},
\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x.
$$

In the occupation data,

$$
\widehat\beta_1
=\frac{675.8041}{751.8852}\approx0.8988,
\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x\approx-2.8536.
$$

Unlike a correlation, the slope changes when a variable's units change.
Standardizing both variables makes their SDs one, so the slope becomes
$r$; Section 7 reads that line.

To verify that the stationary point is the unique minimum, write any other
line as $\widehat\beta_0+d_0+(\widehat\beta_1+d_1)x_i$. The normal
equations make the cross term with the OLS residuals vanish, leaving

$$
Q(\widehat\beta_0+d_0,\widehat\beta_1+d_1)
=RSS+n(d_0+d_1\bar x)^2+d_1^2S_{xx}.
$$

With $S_{xx}>0$, every different line has larger RSS. If a candidate line is
constrained to pass through $(\bar x,\bar y)$, its excess RSS is simply
$S_{xx}(b_1-\widehat\beta_1)^2$. Section 2.4 makes this
comparison visible with five candidate lines.

### Residual constraints and projection geometry {.unnumbered}

At the OLS solution the derivative equations become
$\sum_i e_i=0$ and $\sum_i x_ie_i=0$. Thus the residuals have zero
sample covariance with the predictor and mean zero **by construction**.
They are not evidence that the unknown population errors are independent,
normally distributed, or conditionally mean zero.

For the geometry, let $\mathbf X$ be the $n\times2$ matrix with first
column all ones and second column containing the predictor. Bold
$\mathbf X$ is a matrix, distinct from a single predictor value $x_i$.
The normal equations can be written

$$
\mathbf X^{\mathsf T}(\mathbf y-\mathbf X\widehat{\boldsymbol\beta})
=\mathbf0,
\qquad
\widehat{\boldsymbol\beta}
=(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}\mathbf y.
$$

The fitted vector is the projection of $\mathbf y$ onto the two-dimensional
space spanned by the intercept and predictor columns. The residual vector
is perpendicular to that space. Writing
$\mathbf H=\mathbf X(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}$,
we have $\widehat{\mathbf y}=\mathbf H\mathbf y$ and
$\mathbf e=(\mathbf I-\mathbf H)\mathbf y$.

Because the model includes an intercept, both $\widehat{\mathbf y}$ and
the constant vector $\bar y\mathbf1$ lie in the fitted space. Consequently,

$$
\begin{aligned}
\sum_i(y_i-\bar y)^2
&=\sum_i\{(\widehat y_i-\bar y)+e_i\}^2\\
&=\sum_i(\widehat y_i-\bar y)^2+\sum_i e_i^2,
\end{aligned}
$$

since $\sum_i(\widehat y_i-\bar y)e_i=0$. This proves
$TSS=RegSS+RSS$, an algebraic decomposition of observed variation, not
a division into causal contributions. For these records,
$1820.8134=607.4214+1213.3920$, in squared thousand-dollar units.

Also $RegSS=\widehat\beta_1^2S_{xx}=S_{xy}^2/S_{xx}$. Dividing by
$TSS=S_{yy}$ proves $R^2=S_{xy}^2/(S_{xx}S_{yy})=r^2$ for simple
regression **with an intercept**. The identities need not hold for a
no-intercept fit. When fitted values are nonconstant,
$\operatorname{cor}(y,\widehat y)^2=R^2$ as well.

There are two fitted directions and $n-2$ residual directions. Under a
correct linear mean with independent constant-variance errors,
$E(RSS\mid\mathbf X)=(n-2)\sigma^2$, so $s_e^2=RSS/(n-2)$ is an unbiased
estimate of the error variance. Its square root $s_e$ estimates the
error SD in the outcome's units but is not itself exactly unbiased for
$\sigma$.

### From an error model to slope and intercept uncertainty {.unnumbered}

The 102-record least-squares line is already determined by those records.
The next derivation concerns a **separate model-based question**: the
slope of a process that could generate new, independent occupation
outcomes at the observed education values. Assume

$$
Y_i=\beta_0+\beta_1x_i+\varepsilon_i,
\quad E(\varepsilon_i\mid\mathbf X)=0,
\quad \operatorname{Var}(\boldsymbol\varepsilon\mid\mathbf X)
=\sigma^2\mathbf I.
$$

These are working conditions, not consequences of fitting `lm()`. The
slope error is

$$
\widehat\beta_1-\beta_1
=\frac{\sum_i(x_i-\bar x)\varepsilon_i}{S_{xx}}.
$$

Independence and common variance then give

$$
\operatorname{Var}(\widehat\beta_1\mid\mathbf X)
=\frac{\sigma^2}{S_{xx}},
\qquad
SE(\widehat\beta_1)=\frac{s_e}{\sqrt{S_{xx}}}.
$$

Because $\widehat\beta_0=\bar Y-\widehat\beta_1\bar x$ and
$\operatorname{Cov}(\bar Y,\widehat\beta_1\mid\mathbf X)=0$,

$$
\operatorname{Var}(\widehat\beta_0\mid\mathbf X)
=\sigma^2\left(\frac1n+\frac{\bar x^2}{S_{xx}}\right),
\qquad
\operatorname{Cov}(\widehat\beta_0,\widehat\beta_1\mid\mathbf X)
=-\frac{\bar x\sigma^2}{S_{xx}}.
$$

Here $S_{xx}=\sum_i(x_i-\bar x)^2$, as in Section 2.2. The covariance is
why simply adding the variances of an uncentered intercept and slope
would be wrong; with the predictor centered at its mean the two
estimates are uncorrelated, and the intercept SE is $s_e/\sqrt n$.

Under the stated linear-mean, uncorrelated, equal-variance conditions, OLS
is the best linear unbiased estimator: it has the smallest variance among
estimators linear in the outcomes and unbiased for the coefficients. This
Gauss--Markov statement does not require normal errors and does not establish
causal identification, valid measurement, or population coverage [@fox2016;
@kutner2005].

Add a jointly normal error vector conditional on the predictor values. Then
the standardized coefficient error is normal, $RSS/\sigma^2$ has a
$\chi^2_{n-2}$ distribution, and these quantities are independent. Their
ratio gives

$$
\frac{\widehat\beta_j-\beta_j}{SE(\widehat\beta_j)}
\sim t_{n-2},\qquad j=0,1.
$$

That is the basis of exact classical coverage for
$\widehat\beta_j\pm t_{.975,n-2}SE(\widehat\beta_j)$ and of the
corresponding null test. For the occupation slope the calculation is
$0.8988\pm1.984(0.127)=[0.6468,1.1508]$ thousand dollars per year; the
diagnostics of Section 6 do not support treating its nominal coverage as
established here.

These derivations condition on the predictor values, which need not have
been fixed by an experiment; a model with random predictors uses the same
conditional argument when its error conditions hold. Neither route shows
that the occupations were probability sampled, and neither requires
education or income to be Normal.

### Why a new-case interval has an additional variance term {.unnumbered}

For a supported value $x_0$, define the model mean
$\mu_0=\beta_0+\beta_1x_0$ and its estimate
$\widehat\mu_0=\widehat\beta_0+\widehat\beta_1x_0$.
Substituting the coefficient variances and covariance above gives

$$
\begin{aligned}
\operatorname{Var}(\widehat\mu_0\mid\mathbf X)
&=\operatorname{Var}(\widehat\beta_0)
 +x_0^2\operatorname{Var}(\widehat\beta_1)
 +2x_0\operatorname{Cov}(\widehat\beta_0,\widehat\beta_1)\\
&=\sigma^2\left\{\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}\right\}.
\end{aligned}
$$

The conditioning on $\mathbf X$ is suppressed in the first line. A
confidence interval for this mean uses the square root with $s_e$ for
$\sigma$. A new occupation outcome is instead $Y_0=\mu_0+\varepsilon_0$;
if its error is independent of the fitted data and comes from the same
Normal, equal-variance process,

$$
\operatorname{Var}(Y_0-\widehat\mu_0\mid\mathbf X,x_0)
=\sigma^2\left\{1+\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}\right\}.
$$

The extra $1$ is new outcome variation, which would remain even if the
model mean were known. Both intervals use the same $t_{.975,n-2}$
multiplier. The table applies both formulas at four supported education
values, the same calculation as Section 5's two `predict()` calls.

\Needspace{12\baselineskip}

| Education | Fitted | Mean lower | Mean upper | Case lower | Case upper |
|--:|--:|--:|--:|--:|--:|
| 8 | 4.3369 | 3.3651 | 5.3087 | $-2.6420$ | 11.3158 |
| 10 | 6.1345 | 5.4254 | 6.8437 | $-0.8127$ | 13.0818 |
| 12 | 7.9322 | 7.1776 | 8.6868 | 0.9802 | 14.8842 |
| 15 | 10.6286 | 9.3550 | 11.9022 | 3.6013 | 17.6559 |

: Mean-response and new-case intervals at four supported education values. {#tbl-ch07-app-a-intervals tbl-colwidths="[16,16,17,17,17,17]"}

At 10 years the point prediction is 6.1345 thousand dollars, the mean
interval [5.4254, 6.8437] and the new-occupation interval
[$-0.8127$, 13.0818]. Cutting the negative endpoint off at zero would not
produce a valid 95% interval; a transformation or a model that respects
the boundary would be a different analysis with its own checks.

None of these formulas includes uncertainty from a wrong mean function,
biased measurement or a shift to another labour market. They are
pointwise model intervals, not a simultaneous band or an out-of-sample
validation, and their target is a new occupation-level record from the
declared process, never an individual worker's income.

## References {.unnumbered}

::: {#refs}
:::
