---
title: "Multiple Regression II: Categorical Predictors, Model Comparison, Nonlinearity, and Robust Uncertainty"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - ../koma_heading_tags.tex
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{16\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AtBeginDocument{\renewcommand{\boldsymbol}[1]{\symbf{#1}}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.pos = "!ht")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 7, width = 68)
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 11))
```

\clearpage

::: {.callout-note title="Chapter overview"}
Chapter 8 wrote one regression model for the occupation records. This
chapter extends it in two directions: Sections 1--4 change or compare
the mean part; Sections 5--7 recompute uncertainty when spread differs
or pupils share a school. By the end, you will be able to read indicator
coefficients and a declared contrast, explain covariance adjustment,
compare models and predictors, interpret curvature and a logged outcome,
and choose an uncertainty calculation for a stated reason. The occupation
records, a randomized reading study, and Project STAR connect each
method with the question it answers and the conditions it needs.
:::

Chapter 8 asked how income differs with education when prestige, the
percentage of women, and occupational type are held fixed. A report on
the same 98 records usually needs more. How do the three types compare?
Does a group of predictors add anything once the others are in? Is a
straight line in prestige adequate, and should income be analysed on its
own scale? How far can the standard errors be trusted when professional
incomes spread much more widely? We answer these questions on the same
records, with income in thousand 1971 Canadian dollars and the type
levels *bc*, *prof*, and *wc*. Two further studies join where these
records cannot serve: a randomized reading study with a pretest (Section
2) and Project STAR (Section 7).

Recall Chapter 8's model, with the conditions Chapter 7, Section 1.1,
first stated for one predictor:

$$
Y_i=\beta_0+\sum_j\beta_jx_{ij}+\varepsilon_i,\qquad
E(\varepsilon_i\mid\mathbf x_i)=0,\qquad
\operatorname{Var}(\varepsilon_i\mid\mathbf x_i)=\sigma^2.
$$

Its **mean part** gives the expected outcome at stated predictor values.
Its **error part** says that errors average zero at every predictor
profile, share one variance, and are independent across records; Normal
errors additionally supply the exact $t$ and $F$ references. For the
occupation model `m2`, with blue-collar as reference, the mean part is

$$
E(\mathrm{income}_k\mid e,p,w,\mathrm{type})=\beta_0+\beta_Ee+\beta_Pp
+\beta_Ww+\delta_{\mathrm{prof}}D_{\mathrm{prof}}
+\delta_{\mathrm{wc}}D_{\mathrm{wc}}.
$$

Here $e$, $p$, and $w$ are education, prestige, and the percentage of
women; the two $D$ columns indicate professional and white-collar types.

We work on these two parts in turn: indicators, covariance adjustment,
model comparisons and transformations first; unequal spread and
school-level dependence second. Short R demonstrations connect the model
to its fitted comparisons and uncertainty.

\clearpage

## Predictors That Are Categories {#sec-ch9-factors}

### One indicator, then several

Chapter 7, Section 3.3, fitted a single 0/1 predictor to 24 invented
feedback-by-practice scores. The fitted line passed through the two group
means, its slope was their difference, 4 with SE 1.6967 and $t=2.3575$ on
22 df, and that test was the equal-variance two-sample test written as a
regression. The same idea extends to more than two groups.

A factor with $J$ levels enters a regression as $J-1$ indicator columns.
One level is omitted and becomes the **reference**: the intercept is its
fitted mean, and each indicator coefficient is a difference from it. With
three occupation types and blue-collar as the reference,

$$
\widehat y_g=\widehat\beta_0+\widehat\delta_{\mathrm{prof}}D_{\mathrm{prof}}
+\widehat\delta_{\mathrm{wc}}D_{\mathrm{wc}},\qquad
\widehat\mu_{\mathrm{bc}}=\widehat\beta_0,\quad
\widehat\mu_{\mathrm{prof}}=\widehat\beta_0+\widehat\delta_{\mathrm{prof}},\quad
\widehat\mu_{\mathrm{wc}}=\widehat\beta_0+\widehat\delta_{\mathrm{wc}}.
$$

Two indicators, not three, carry three means. A third column would equal
the intercept column minus the other two, so the design matrix would lose
full rank and no coefficient would be separately estimable. Which level
is omitted is a choice of coordinates, not a choice of model.

The working data `d` retain the 98 occupations in `carData::Prestige`
with a recorded type. Income divided by 1,000 gives `income_k` in
thousand 1971 Canadian dollars, and the factor order is blue-collar,
professional, then white-collar, so blue-collar is the reference. On
these same rows, `m0`, `m1`, and `m2` name the education-only, the
education-and-prestige, and the full additive models:

```{r}
#| label: ch9-records-prepare
#| include: false
prestige <- carData::Prestige
prestige$income_k <- prestige$income / 1000
d <- prestige[!is.na(prestige$type), ]
d$type <- factor(d$type, levels = c("bc", "prof", "wc"))
```

```{r}
#| label: ch9-records
#| echo: true
m0 <- lm(income_k ~ education, data = d)
m1 <- lm(income_k ~ education + prestige, data = d)
m2 <- lm(income_k ~ education + prestige + women + type, data = d)
```

The 98 records hold 44 blue-collar, 31 professional, and 23 white-collar
occupations, and the full model fits six coefficients, leaving 92
residual degrees of freedom. The rows were chosen because their records
are complete, not sampled at random from a labour market [@carData2022;
@fox2019].

### What the indicator coefficients mean

Fit income on type alone and read the coefficients against
@fig-ch9-indicators. A blue-collar occupation carries 0 in both indicator
columns, a professional one a 1 in the first, and a white-collar one a 1
in the second, so each type's fitted value is the intercept plus that
type's coefficient.

```{r}
#| label: ch9-group-means
#| echo: true
group_fit <- lm(income_k ~ type, data = d)
coef(group_fit)
```

```{r}
#| label: fig-ch9-indicators
#| fig-width: 6.4
#| fig-height: 3.1
#| fig-cap: "Income of the 98 occupations by type. Bars mark the group means; the dashed line is the blue-collar mean, the intercept, and the arrows are the two indicator coefficients, 5.1853 and −0.3218."
#| fig-alt: "Jittered open circles of income in thousand dollars for 44 blue-collar, 31 professional and 23 white-collar occupations, with a thick bar at each group mean, about 5.37, 10.56 and 5.05. A dashed horizontal line at the blue-collar mean runs across the plot. An upward arrow from the dashed line to the professional mean is labelled 5.1853 higher and a short downward arrow to the white-collar mean is labelled 0.3218 lower. Professional incomes spread from about 4.6 to 25.9 thousand dollars; the other two types stay below 9."
type_means <- data.frame(x = 1:3,
  mean = as.numeric(tapply(d$income_k, d$type, mean)))
ggplot2::ggplot(d, ggplot2::aes(as.numeric(type), income_k)) +
  ggplot2::geom_hline(yintercept = type_means$mean[1], linetype = 2,
    colour = "grey45") +
  ggplot2::geom_point(shape = 1, colour = "grey55",
    position = ggplot2::position_jitter(width = .12, height = 0,
      seed = 12345)) +
  ggplot2::geom_segment(data = type_means, ggplot2::aes(x = x - .28,
    xend = x + .28, y = mean, yend = mean), linewidth = 1.1) +
  ggplot2::annotate("segment", x = c(2.4, 3.4), xend = c(2.4, 3.4),
    y = type_means$mean[1], yend = type_means$mean[2:3],
    arrow = ggplot2::arrow(length = ggplot2::unit(.15, "cm"))) +
  ggplot2::annotate("text", x = c(2.46, 3.46), y = c(9.9, 6.9),
    label = c("5.1853 higher", "0.3218 lower"), hjust = 0, size = 3.4) +
  ggplot2::scale_x_continuous(breaks = 1:3, limits = c(.6, 3.9),
    labels = c("Blue-collar (44)", "Professional (31)",
      "White-collar (23)")) +
  ggplot2::labs(x = NULL, y = "Income (thousand dollars)") +
  ggplot2::theme(panel.grid.major.x = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank())
```

The coefficients are 5.3741, 5.1853, and $-0.3218$. The intercept is the
blue-collar mean, the dashed line in the figure, and each coefficient is
an arrow from that line to another type's mean: professional occupations
average 5.1853 thousand dollars more, white-collar occupations 0.3218
less. The fitted means, 5.3741, 10.5595, and 5.0523, are exactly the
observed means, for the reason Chapter 8, Appendix A, gave: with one free
parameter per group, least squares returns each group's own mean. A
group-only regression is therefore a description of three means, with no
adjustment in it. The figure also shows the wide spread of professional
incomes, from about 4.6 to 25.9 thousand dollars, which Section 5 takes
up.

Keep this description apart from the full model. In `m2` the
professional and white-collar coefficients are 0.3242 and 0.2353: they
compare fitted incomes at the *same* education, prestige, and percentage
of women. The raw professional advantage of 5.1853 and the adjusted
0.3242 answer different questions, and neither is "the effect of
occupational type".

### Changing the reference level

**Releveling** makes a different level the reference. Refit the
group-only model with white-collar as the reference:

```{r}
#| label: ch9-relevel
#| echo: true
group_fit_wc <- lm(income_k ~ relevel(type, ref = "wc"), data = d)
coef(group_fit_wc)
```

The coefficients are now 5.0523, 0.3218, and 5.5071: the intercept is the
white-collar mean, and the two coefficients measure the blue-collar and
professional means from it. The fitted values and residuals are unchanged
(Chapter 8, Appendix A); the same three means are reported in new
coordinates. Releveling `m2` in the same way gives intercept 0.6559,
blue-collar $-0.2353$, and professional 0.0889, while the education,
prestige, and women coefficients stay exactly as they were.

Centering a continuous predictor (Chapter 7, Section 2.5, and Chapter
8, Appendix A), subtracting a constant such as its mean, is the same
kind of change: it moves the profile the intercept
describes and leaves every fitted value, residual, and substantive
contrast alone. Neither operation adds information, such as the
cross-type support the records lack, and "the professional coefficient"
always means professional minus whichever level is currently the
reference.

### One-way analysis of variance as a regression

Chapter 4, Sections 2.2--2.4, compared one common mean with separate
group means through between-group and within-group variation. In the
group-only regression, equal means says that both indicator coefficients
are zero. The between-group sum of squares is $SS_B=595.9562$ on
$J-1=2$ degrees of freedom. The within-group sum of squares is the
group-only model's residual sum of squares, $SS_W=1138.123$ on
$N-J=95$ degrees of freedom. The familiar calculation is

$$
F=\frac{MS_B}{MS_W}=\frac{SS_B/(J-1)}{SS_W/(N-J)}
=\frac{595.9562/2}{1138.123/95}=24.8725,\qquad
\eta^2=\frac{SS_B}{SS_T}=.3437.
$$

The regression and ANOVA describe the same fitted means, so eta squared
is also the group-only regression's $R^2$. Chapter 6, Sections 1.3 and
3.2, fitted and compared these kinds of models in R. Here the reduced
model has only an intercept and the fuller one has two indicators;
Section 3.1 generalizes their comparison.

```{=latex}
\Needspace{13\baselineskip}
```

```{r}
#| label: ch9-anova-identity
#| echo: true
anova(lm(income_k ~ 1, data = d), group_fit)
```

`aov()` returns the same statistic from the same fitted means, residual
sum of squares, and degrees of freedom. The equality is algebraic: it
holds whether or not group membership was randomized, so it supplies no
causal warrant of its own. As in Chapter 4, the omnibus $F$ says that the
three means are not all equal but not which differ, and a comparison
chosen after seeing it needs Chapter 4's family calibration. A declared comparison needs no significant omnibus result first; the
next subsection gives its calculation.

### A contrast and its standard error

A **linear contrast** writes a comparison as a linear combination of
coefficients, $C=\mathbf c^{\mathsf T}\boldsymbol\beta$. For a comparison
declared before the results are read, the estimate is that same
combination of the fitted coefficients, and its variance uses the whole
coefficient covariance matrix $\widehat{\mathbf V}$:

$$
\widehat C=\mathbf c^{\mathsf T}\widehat{\boldsymbol\beta},\qquad
\widehat{\operatorname{Var}}(\widehat C)=\mathbf c^{\mathsf T}\widehat{\mathbf V}\mathbf c,\qquad
SE(\widehat C)=\sqrt{\mathbf c^{\mathsf T}\widehat{\mathbf V}\mathbf c}.
$$

For a difference of two coefficients, the variance is the sum of their
two variances minus twice their covariance. The group-only comparison
makes the covariance visible. Professional minus white-collar is
$\widehat\delta_{\mathrm{prof}}-\widehat\delta_{\mathrm{wc}}=5.5071$,
with $\mathbf c=(0,1,-1)^{\mathsf T}$:

```{r}
#| label: ch9-contrast
#| echo: true
c_vec <- c(0, 1, -1)
V <- vcov(group_fit)
c(estimate = sum(c_vec * coef(group_fit)),
  SE = sqrt(drop(c_vec %*% V %*% c_vec)))
```

Both coefficients measure a group mean from the *same blue-collar mean*.
If a sample overstates that reference mean, both coefficients fall
together. Their covariance, 0.2723, is the variance of the shared mean,
$MS_W/44$. It drops out when the two coefficients are subtracted:

$$
\widehat{\operatorname{Var}}(\widehat C)
=0.6587+0.7932-2(0.2723)=0.9073
=MS_W\left(\frac1{31}+\frac1{23}\right).
$$

Thus the SE is 0.9525 and, with $t_{.975,95}=1.9853$, the pointwise 95%
interval is $[3.6161,\ 7.3982]$. Omitting the covariance gives an SE of
1.2049; adding the two SEs gives 1.7022, which is not the SE of this
contrast. The covariance is part of the comparison, not a correction
added after it.

The same formula applies to `m2` with
$\mathbf c=(0,0,0,0,1,-1)^{\mathsf T}$: the estimate is 0.0889, its SE
1.08, and its pointwise 95% interval $[-2.0561,\ 2.2338]$
($t_{.975,92}=1.9861$). Sections 1.2--1.3 explain why the raw 5.5071
becomes 0.0889 when education, prestige, and the percentage of women are
held fixed. This interval covers zero and leaves the adjusted comparison
unresolved; it is a working-model calculation among types with thin
joint support (Chapter 8). Exercise 2 practises its covariance-based
variance.

The vector $\mathbf c$ states the question. One entry set to 1 returns
an ordinary coefficient; weights on type coefficients with intercept
weight zero express contrasts between group means in Chapter 4's sense.
The vector changes with the reference coding, but a named comparison
does not: the releveled `m2` in Section 1.3 gave 0.0889 directly.
Declaring a comparison before reading its result governs the calibration,
not whether it is a contrast. A comparison selected after inspecting the
output needs the family calibration of Chapter 4, Section 3.3, whose R
calculations appear in Chapter 6, Section 2.

## Adjusted Group Comparisons: The Analysis of Covariance {#sec-ch9-contrasts}

### A regression with one common slope {#ch09-ancova}

The occupation types differ in many ways at once. A randomized study with
a baseline score lets the group question be asked more sharply. Baumann
and colleagues randomly assigned 68 fourth-grade pupils in one school to
three reading-instruction conditions [@baumann1992]: DRA, a directed
reading activity from the basal programme; DRTA, a directed
reading-thinking activity; and think-aloud strategy instruction. Each
pupil took an error-detection test of comprehension monitoring before
and after the lessons. `carData::Baumann` holds the 66 pupils with a
recorded first posttest, 22 per group. We use the first pretest and
posttest, with DRA as the reference; the investigators also analysed
further outcomes and both pretests.

The **analysis of covariance** adds the pretest to the group comparison
as one more predictor with a single slope:

$$
Y_{ij}=\mu_j+\beta(X_{ij}-\bar X)+\varepsilon_{ij},\qquad
\bar Y_j^{\,\mathrm{adj}}=\bar Y_j-\widehat\beta(\bar X_j-\bar X).
$$

Here $Y_{ij}$ and $X_{ij}$ are the posttest and pretest of pupil $i$ in
group $j$. The group terms $\mu_j$ shift the line up or down, and the
pretest enters once, so every group's line has the same slope $\beta$. An
**adjusted mean** is a group's fitted mean at a stated pretest value. At
the grand pretest mean $\bar X$ it is the observed posttest mean
$\bar Y_j$ moved along the common slope by the group's pretest departure
from $\bar X$, the second equation.

The working data `reading` hold these 66 pupils, all with recorded first
pretests and posttests. The package's group codes `Basal`, `DRTA` and
`Strat` are labelled DRA, DRTA and Think-aloud, with DRA as the reference.
The prediction below evaluates each group's line at the grand pretest
mean.

```{=latex}
\Needspace{20\baselineskip}
```

```{r}
#| label: ch9-baumann-prepare
#| include: false
reading <- carData::Baumann
reading$group <- factor(reading$group,
  levels = c("Basal", "DRTA", "Strat"),
  labels = c("DRA", "DRTA", "Think-aloud"))
```

```{r}
#| label: ch9-baumann
#| echo: true
ancova <- lm(post.test.1 ~ group + pretest.1, data = reading)
coef(ancova)
predict(ancova, interval = "confidence",
  newdata = data.frame(group = levels(reading$group),
    pretest.1 = mean(reading$pretest.1)))
```

```{r}
#| label: fig-ch9-ancova
#| fig-width: 6.4
#| fig-height: 3.9
#| fig-cap: "Posttest against pretest for the 66 Baumann pupils. The parallel lines are the common-slope fit over the common pretest support 6--14; each filled symbol marks a group's observed pretest and posttest means, and its arrow follows the group's line to the grand pretest mean 9.7879, where the open symbol marks the adjusted mean."
#| fig-alt: "Grey points for DRA (circles), DRTA (triangles) and Think-aloud (squares) pupils, pretest from 4 to 16 and posttest from 1 to 15, slightly jittered. Three parallel fitted lines with slope about 0.69 run from pretest 6 to 14: DRTA highest, Think-aloud in the middle and DRA lowest. A vertical line marks the grand pretest mean 9.79, where open symbols mark the adjusted means. The DRA mean, a filled circle at pretest 10.5 and posttest 6.68, moves down its line to 6.19; the Think-aloud mean, at 9.14 and 7.77, moves up to 8.22; the DRTA mean, at 9.73 and 9.77, barely moves, to 9.81, so its filled and open triangles almost coincide."
grand_pretest <- mean(reading$pretest.1)
line_grid <- expand.grid(pretest.1 = c(6, 14),
  group = levels(reading$group))
line_grid$fitted <- predict(ancova, newdata = line_grid)
shift <- data.frame(group = levels(reading$group),
  x = as.numeric(tapply(reading$pretest.1, reading$group, mean)),
  y = as.numeric(tapply(reading$post.test.1, reading$group, mean)))
shift$adjusted <- predict(ancova, newdata = data.frame(
  group = shift$group, pretest.1 = grand_pretest))
ggplot2::ggplot(reading, ggplot2::aes(pretest.1, post.test.1,
    shape = group, linetype = group)) +
  ggplot2::geom_point(colour = "grey60", size = 1.6,
    position = ggplot2::position_jitter(width = .15, height = .15,
      seed = 12345)) +
  ggplot2::geom_line(data = line_grid, ggplot2::aes(y = fitted),
    linewidth = .7) +
  ggplot2::geom_vline(xintercept = grand_pretest, colour = "grey40",
    linewidth = .4) +
  ggplot2::geom_segment(data = shift, ggplot2::aes(x = x, y = y,
    xend = grand_pretest, yend = adjusted), inherit.aes = FALSE,
    linewidth = .9,
    arrow = ggplot2::arrow(length = ggplot2::unit(.22, "cm"))) +
  ggplot2::geom_point(data = shift, ggplot2::aes(x, y), size = 2.8) +
  ggplot2::geom_point(data = shift, ggplot2::aes(grand_pretest, adjusted),
    shape = c(1, 2, 0), size = 3.8, stroke = .9, show.legend = FALSE) +
  ggplot2::scale_shape_manual(values = c(16, 17, 15)) +
  ggplot2::scale_linetype_manual(values = c(1, 2, 4)) +
  ggplot2::scale_x_continuous(breaks = seq(4, 16, 2)) +
  ggplot2::labs(x = "Pretest score", y = "Posttest score", shape = NULL,
    linetype = NULL) +
  ggplot2::theme(legend.position = "bottom",
    panel.grid.minor = ggplot2::element_blank())
```

Read @fig-ch9-ancova from each filled symbol, a group's observed pretest
and posttest means, to the open symbol on the vertical line, its adjusted
mean. DRA happened to start highest, 0.7121 points above the grand
pretest mean of 9.7879, so following its line to the grand mean lowers
its posttest mean by $0.6932\times0.7121$, from 6.6818 to 6.1882.
Think-aloud started 0.6515 points below and rises from 7.7727 to 8.2243,
while DRTA, almost at the grand mean, barely moves, from 9.7727 to
9.8147. The fit gives the common slope 0.6932 (SE 0.1015), differences
from DRA of 3.6266 (SE 0.7362) for DRTA and 2.0362 (SE 0.7450) for
Think-aloud, and a residual mean square of 5.894 on 62 degrees of
freedom. Tested together after the pretest, the two group coefficients
give $F(2,62)=12.1689$, $p<.0001$, a comparison of named models as in
Chapter 6, Section 3.2, whose statistic Section 3.1 below develops. The
adjusted means have 95% intervals
$[5.1435,\ 7.2329]$, $[8.7800,\ 10.8495]$, and $[7.1813,\ 9.2674]$.

Three conditions carry that reading. The adjusted means need a **common
reference**, here the grand pretest mean. The reference must lie inside
every group's observed pretest range, the **common support**: the
pretests run from 4 to 16, 6 to 16, and 4 to 14, so the lines are drawn
over 6 to 14, and a comparison read outside that band would rest on
extrapolation. And the comparison needs a stated **averaging
population**: for this additive model, predicting every group at the
grand pretest mean gives the same values as predicting every group at
each of the 66 pupils' pretests and averaging. Because the lines are
parallel, the adjusted differences, 3.6266 and 2.0362, are the same at
any pretest value inside the common support; the adjusted means are not,
which is why the reference value is stated.

### Three analyses of a pretest--posttest study {#ch09-ancova-pretest}

A randomized pretest--posttest study is often analysed in one of three
ways: compare the posttest means, compare the gains (posttest minus
pretest), or adjust for the pretest as above. The three are one
regression that differs only in the pretest coefficient $\gamma$:

$$
Y_i-\gamma X_i=\beta_0+\delta_{\mathrm{DRTA}}D_{\mathrm{DRTA},i}
+\delta_{\mathrm{TA}}D_{\mathrm{TA},i}+\varepsilon_i,\qquad
\gamma=0,\ 1,\ \text{or estimated}.
$$

With $\gamma=0$ the model compares posttest means, with $\gamma=1$ it
compares gains, and with $\gamma$ estimated it is the analysis of
covariance. In R, `offset()` enters the pretest with its coefficient
fixed at 1:

```{r}
#| label: ch9-three-analyses
#| echo: true
post_only <- lm(post.test.1 ~ group, data = reading)
gain <- lm(post.test.1 ~ group + offset(pretest.1), data = reading)
```

Under random assignment the expected pretest difference between groups is
zero. Subtracting a fixed multiple of the pretest therefore preserves the
average group-comparison target. Under the common-slope working model,
covariance adjustment estimates that same comparison while estimating
the pretest slope from the records. Random assignment covered the 68 pupils assigned, while
the analyses use the 66 with a recorded first posttest, so the argument
also needs a complete-record condition, for example that the two
posttests went missing for reasons unrelated to what those pupils would
have scored,
which the design does not guarantee and the records cannot check.

\Needspace{14\baselineskip}

| Analysis | Pretest coefficient | DRTA minus DRA (SE) | Think-aloud minus DRA (SE) | Residual SD (df) |
|:--|:--|:--|:--|:--|
| Posttest only | 0 | 3.0909 (0.9614) | 1.0909 (0.9614) | 3.1885 (63) |
| Gain score | 1 | 3.8636 (0.7779) | 2.4545 (0.7779) | 2.5799 (63) |
| Analysis of covariance | 0.6932, estimated | 3.6266 (0.7362) | 2.0362 (0.7450) | 2.4278 (62) |

: Three analyses of the Baumann records as one regression. {#tbl-ch09-three-analyses tbl-colwidths="[24,18,20,20,18]"}

The three analyses differ in precision and in how this sample's chance
pretest imbalance enters. The pretest means are 10.5000, 9.7273, and
9.1364, so DRA had a head start of 0.7727 points over DRTA, about a
quarter of the pretest SD of 3.0205 across all 66 pupils. The size of
that gap is what matters. A significance test of it would add nothing:
random assignment already makes any baseline gap a chance gap, so the
test would examine a null hypothesis known to be true. The gap accounts
exactly for the spread of the estimates in @tbl-ch09-three-analyses:
each DRTA estimate is the posttest-only difference plus $\gamma$ times
DRA's head start, $3.0909+\gamma(10.5-9.7273)$, which gives 3.0909,
3.8636, and 3.6266 (3.6265 from the rounded inputs).

Estimating the coefficient leaves the smallest residual SD, 2.4278
against 3.1885 and 2.5799, and the smallest standard errors, because the
pretest predicts the posttest with a slope near 0.7 rather than 0 or 1.
The choice belongs in the analysis plan, made before the outcomes are
seen, not after comparing the three estimates.

\Needspace{16\baselineskip}

### Parallel slopes and the limits of adjustment {#ch09-ancova-limits}

A common slope is a condition, not a finding. The check adds the
group-by-pretest product block, which lets each group have its own slope,
and compares the two named models on the same records (Chapter 6,
Section 3.2; Section 3.1 below develops the general statistic):

```{=latex}
\Needspace{11\baselineskip}
```

```{r}
#| label: ch9-parallel
#| echo: true
varying_slopes <- lm(post.test.1 ~ group * pretest.1, data = reading)
anova(ancova, varying_slopes)
```

The product block gives $F(2,60)=1.5054$, $p=.2302$, with group slopes of
0.4663 (DRA), 0.6868 (DRTA), and 0.8768 (Think-aloud). These 66 records do
not distinguish the two models, which is not the same as showing that the
slopes are equal. When slopes do differ, an adjusted difference depends on
the pretest value at which it is read; Chapter 10 develops that
varying-slope model as an interaction.

What adjustment supplies depends on how the groups were formed. In groups
formed by random assignment, the pretest only sharpens a comparison that
the design already warrants. In observational groups, a fitted comparison
at a common covariate value cannot balance what was not measured. In
Section 1.2 the professional advantage fell from 5.1853 to 0.3242 at
fixed education, prestige, and percentage of women, but neither number
was an effect of type. A
covariate measured with error can leave part of the
baseline difference unadjusted (Chapter 14, Section 5.1); and a covariate
recorded after the groups formed can remove part of the effect being
estimated.

**Lord's paradox** makes the point with intact groups that differ at
baseline [@lord1967]. The gain-score analysis asks whether the groups
changed by different amounts; the covariance analysis asks whether pupils
with the same pretest ended differently. The two can reach opposite
conclusions from the same records, partly through the regression toward
the mean of Chapter 7, Section 7.3: groups that start apart each regress
toward their own mean. Neither analysis says which question is
the causal one; the design does, and Chapter 14 returns to what a design
supplies that adjustment cannot.

## Comparing Models and Predictors {#sec-ch9-comparison}

### Comparing residual sums of squares

Chapter 4 compared one common mean with group means (Sections 2.2--2.3)
and additive with interaction models (Section 4.4); Chapter 6 fitted
those comparisons in R (Sections 1.3 and 3.2). This section develops the
general question on the occupation records: what does each mean part
permit, and what restrictions distinguish them? Two models are **nested**
when the
reduced model is the fuller model with some coefficients set to zero, and
both are fitted to the same records and outcome. Extra coefficients
never raise the residual sum of squares; the question is whether they
lower it by more than noise alone would. The **extra-sum-of-squares**
statistic compares the two residual sums of squares:

$$
F=\frac{(RSS_{R}-RSS_{F})/q}{RSS_{F}/(n-p)}.
$$

The numerator is the drop that the $q$ extra coefficients buy, per
coefficient. The denominator is the fuller model's residual mean square,
with $p$ its number of coefficients: the same $s^2$ that appears in every
classical standard error. When the reduced-model restrictions hold, under the Normal linear
model with independent common-variance errors, the ratio has an exact
central $F$ reference with $q$ and $n-p$ degrees of freedom.

```{r}
#| label: fig-ch9-rss
#| fig-width: 6.4
#| fig-height: 2.6
#| fig-cap: "Residual sum of squares as columns are added to the occupation model on the same 98 records; each label gives the drop that the added column or block buys."
#| fig-alt: "Horizontal bars of residual sum of squares for five nested models. The intercept-only bar is labelled RSS 1734.0792; adding education drops it by 571.5324, adding prestige by 294.8931, adding the percentage of women by 236.3266, and adding the two type columns by only 0.4375, so the last two bars look the same length."
rss_path <- data.frame(model = c("Intercept only", "+ education",
    "+ prestige", "+ women", "+ type (2 columns)"),
  rss = sapply(list(lm(income_k ~ 1, data = d), m0, m1,
    lm(income_k ~ education + prestige + women, data = d), m2),
    deviance))
rss_path$label <- paste(c("RSS", rep("drop", nrow(rss_path) - 1)),
  format(round(c(rss_path$rss[1], -diff(rss_path$rss)), 4), digits = 15,
    scientific = FALSE, trim = TRUE))
rss_path$model <- factor(rss_path$model, levels = rev(rss_path$model))
ggplot2::ggplot(rss_path, ggplot2::aes(rss, model)) +
  ggplot2::geom_col(width = .6, fill = "grey75") +
  ggplot2::geom_text(ggplot2::aes(label = label), hjust = -.1,
    size = 3.3) +
  ggplot2::scale_x_continuous(limits = c(0, 2150), expand = c(0, 0)) +
  ggplot2::labs(x = "Residual sum of squares", y = NULL)
```

@fig-ch9-rss shows what each addition buys in the occupation models.
Education buys 571.5324 of the total 1734.0792, and prestige a further
294.8931. The percentage of women then buys 236.3266 for one coefficient,
while the two type columns buy only 0.4375 for two. Against the full
model's residual mean square of 6.8575, the type block's 0.4375 over two
coefficients gives $F(2,92)=0.0319$. Section 3.2 sets this beside the
women-and-type comparison in @tbl-ch09-blocks. The
subtraction means something only when both models use the same rows and
the same outcome, which is why every comparison in this chapter runs on
the same 98 records.

The **overall** regression test is the special case whose reduced model
is the intercept alone, with all $p-1$ non-intercept coefficients set to
zero. Chapter 8, Section 3.3, gave its formula and read it from
`summary(m2)$fstatistic`: $F(5,92)=32.1747$ for the full model and $F(1,96)=47.1956$
for the education-only model on the same records. The larger value is
not evidence that `m0` is better, because the two statistics test
different restrictions. Neither checks the mean function, and neither
says how a fitted rule would predict new records (Chapter 13). The same
statistic can also test any $q$ linear restrictions on one fitted model
without fitting the reduced model; Appendix A develops that form.

### Testing a block of coefficients

A **block test** applies the preceding statistic to a declared set of
coefficients jointly. Two blocks ask different questions on these same
98 records: do women and type add to education and prestige, and does
type add once women is already present?

```{r}
#| label: ch9-blocks
#| echo: true
#| results: hide
no_type <- lm(income_k ~ education + prestige + women, data = d)
anova(m1, m2)
anova(no_type, m2)
```

Adding the percentage of women and the two type indicators to `m1` tests
three restrictions against the full model's 92 residual degrees of
freedom; dropping type from the full model tests two restrictions against
the same denominator.

| Added | df | $F$ | $p$ |
|:--|--:|--:|--:|
| Women and type | 3 | 11.5088 | $<.0001$ |
| Type after women | 2 | 0.0319 | .9686 |

: Two declared adjusted block comparisons. {#tbl-ch09-blocks tbl-colwidths="[40,15,20,25]"}

The first block is strongly supported: $F(3,92)=11.5088$ against a 5%
critical value of 2.7036. The second is not: $F(2,92)=0.0319$, $p=.9686$,
against 3.0954. Joint evidence for the first block does not show that
each added column contributes; the percentage of women could carry the
whole block. The second result does not show that occupational type is
irrelevant: once education, prestige, and the percentage of women are in
the equation, the two type indicators reduce the residual sum of squares
by almost nothing *on these 98 records*, where the types barely overlap
on the other predictors.

A single `anova(m2)` answers a different question. It gives
**sequential** sums of squares: each line tests a term given only the
terms above it in the formula, so the answers depend on the order of the
formula, and only the last line matches a block test. Section 5.2 gives
the robust version of the type block.

### Information criteria: AIC and BIC

A nested $F$ test compares two models when one contains the other.
Information criteria can also rank models that are not nested, using a
different quantity: the maximized likelihood, penalized for the number of
fitted parameters.

Chapter 8, Appendix B, built that likelihood for these models by carrying
Chapter 6's binomial likelihood for 32 successes in 50 trials over to
independent Normal outcomes around the regression. Maximizing it over
the coefficients reproduces least squares.
Its scale estimate divides the residual sum of squares by $n$,
$\sqrt{RSS/n}=2.5373$, where the residual SD reported by `lm()` divides
by the residual degrees of freedom, $\sqrt{RSS/(n-p)}=2.6187$. AIC adds
$2k$ to $-2\log\widehat L$ and BIC adds $k\log n$, where $k$ counts the
fitted coefficients **and** the residual variance, so `m2` has $k=7$,
not 6. One call returns all three quantities:

```{r}
#| label: ch9-likelihood
#| echo: true
c(logLik = logLik(m2), AIC = AIC(m2), BIC = BIC(m2))
```

For `m2` the call returns $-230.3020$, 474.6040, and 492.6988.
Chapter 8, Appendix B, tabulates `m0`, `m1`, and `m2`. The new use here
compares candidates that are **not nested**: income on education and the
percentage of women (AIC 486.3288, BIC 496.6687), and income on prestige
and the percentage of women (AIC 469.3428, BIC 479.6827). Neither model
contains the other, so no nested $F$ compares them. Both have $k=4$;
the penalties cancel, and the criteria prefer prestige with women because
it fits these records better.

Set `m2` beside that pair: its AIC 474.604 and BIC 492.6988 are both
higher than the prestige-and-women model's. Its three further
coefficients, education and the two type indicators, do not buy enough
fit to pay their penalty. Section 3.2's type block and Section 3.4's
education comparison show why. Adjusted $R^2$ gives the same order:
.5549, .6257, and .6164 for education with women, prestige with women,
and `m2`. A ranking is always relative to the candidates offered.

The two criteria charge differently for parameters. For the seven
parameters of `m2`, BIC charges $7\log 98=32.0948$ instead of AIC's 14,
because at $n=98$ the factor $\log n=4.585$ exceeds 2, so the criteria
can disagree; a disagreement shows how strongly the extra parameters are
charged, not a tie to break by preference. The criteria are comparable
only among models fitted to the same records, with the same outcome on
the same scale, under one likelihood convention; Section 4.2 shows a
change of scale breaking the comparison. Both are calculated on the records that fitted the models,
so neither is a held-out error, and both are comparative: a smaller AIC
says one candidate is preferred to another, not that either describes
the records well [@akaike1974; @schwarz1978]. Chapter 11 develops
likelihood fully.

### What each predictor adds

Raw slopes use different units. The education slope, 0.1164 thousand
dollars per year, looks similar to the prestige slope, 0.1406 per rating
point, but a year and a point are different changes. Their predictor SDs
are 2.7489 years and 17.0949 points; the percentage of women's SD is
31.382 percentage points and income's is 4.2281 thousand dollars. Two
summaries answer different questions:

$$
\beta_j^{*}=\widehat\beta_j\frac{s_{x_j}}{s_y},\qquad
\Delta R^2_j=R^2_{\mathrm{full}}-R^2_{\mathrm{without}\ j}.
$$

A **standardized coefficient**, or beta weight, rescales a partial slope
into outcome-SD units per predictor SD. The values are 0.0757 for
education, 0.5685 for prestige, and $-0.3971$ for women. This differs
from Chapter 8's standardization over named profiles: that operation
averages fitted outcomes over a stated population, while these beta
weights change the units of slopes. Their values depend on the spread
of the predictors in these particular records.

A **drop-one $R^2$** asks how much of the full model's explained variation
disappears when one predictor is removed. From the full $R^2=.6362$,
the drops are .0007 for education, .0625 for prestige, and .1212 for
women. Each is Section 3.2's one-column block comparison on a different
scale: the corresponding $F(1,92)$ values are 0.179, 15.807, and
30.6593, each the square of its coefficient's $t$. Education's near-zero
drop describes the same limited unique contribution as its wide
coefficient interval.

Neither summary measures importance. Restricted predictor ranges can
shrink standardized coefficients. Overlapping predictor information
means that the drop-one values need not sum to the joint $R^2$; a
predictor that duplicates another can contribute almost nothing uniquely
while still being substantively relevant. Neither summary establishes a
causal ranking or held-out value. Name the comparison before reading
the number.

## Changing a Predictor's Form or the Outcome's Scale {#sec-ch9-forms}

### Adding a squared term

In every model so far, a continuous predictor's slope was the same at
every value of that predictor. Chapter 8, Section 4.3, refitted the full
model with a squared prestige term, and that refit is the worked example
here. It was a check on functional form declared before it was run, not
a search across candidate terms for the smallest $p$-value; a $p$-value
read after such a search would not mean what it appears to mean. A
squared term is one kind of **polynomial term**, a power of a predictor
entered as an ordinary extra column, here `I(prestige^2)`, and the model
is still linear in its coefficients. What changes is the reading:

$$
E(Y\mid x,\ldots)=\beta_0+\beta_1x+\beta_2x^2+\cdots,\qquad
\frac{\partial E(Y)}{\partial x}=\beta_1+2\beta_2x,\qquad
x^{*}=-\frac{\beta_1}{2\beta_2}.
$$

The predictor's slope now depends on its own value, $\beta_1+2\beta_2x$,
and is zero at the turning point $x^{*}$.

```{=latex}
\Needspace{18\baselineskip}
```

```{r}
#| label: ch9-quadratic
#| echo: true
m2q <- lm(income_k ~ education + prestige + I(prestige^2) + women + type,
  data = d)
anova(m2, m2q)
```

```{r}
#| label: fig-ch9-quadratic
#| fig-width: 6.4
#| fig-height: 3.4
#| fig-cap: "Component-plus-residual display for prestige in the model with a squared prestige term. The solid curve is the fitted quadratic component and the dashed line the linear component of the additive model; the labels give the curve's slope at prestige 30, 50 and 70."
#| fig-alt: "Open grey circles for the 98 occupations, prestige from 17.3 to 87.2 on the horizontal axis and component plus residual from about −6 to 18 on the vertical axis. A solid curve rises slowly at low prestige and more steeply at high prestige; a dashed straight line crosses it twice. Black dots on the curve at prestige 30, 50 and 70 carry the labels slope 0.0697, slope 0.1665 and slope 0.2633."
b_q <- coef(m2q)
component <- function(x) b_q["prestige"] * x + b_q["I(prestige^2)"] * x^2
centre <- mean(component(d$prestige))
partial_data <- data.frame(prestige = d$prestige,
  partial = component(d$prestige) - centre + residuals(m2q))
curve_grid <- data.frame(prestige = seq(17.3, 87.2, length.out = 200))
curve_grid$quadratic <- component(curve_grid$prestige) - centre
curve_grid$linear <- coef(m2)["prestige"] *
  (curve_grid$prestige - mean(d$prestige))
marks <- data.frame(prestige = c(30, 50, 70))
marks$partial <- component(marks$prestige) - centre
marks$label <- paste("slope", format(round(
  b_q["prestige"] + 2 * b_q["I(prestige^2)"] * marks$prestige, 4),
  digits = 15, scientific = FALSE, trim = TRUE))
ggplot2::ggplot(partial_data, ggplot2::aes(prestige, partial)) +
  ggplot2::geom_point(shape = 1, colour = "grey65") +
  ggplot2::geom_line(data = curve_grid, ggplot2::aes(y = linear),
    linetype = 2) +
  ggplot2::geom_line(data = curve_grid, ggplot2::aes(y = quadratic),
    linewidth = .8) +
  ggplot2::geom_point(data = marks, size = 2.6) +
  ggplot2::geom_label(data = marks, ggplot2::aes(y = partial + 3.2,
    label = label), size = 3.2, linewidth = 0) +
  ggplot2::labs(x = "Prestige", y = "Prestige component plus residual")
```

The quadratic coefficient is 0.0024 with SE 0.0009 and $t=2.5954$; the
one-restriction block test gives $F(1,91)=6.736$ with $p=.011$; and the
criteria move from 474.6040 and 492.6988 to 469.6058 and 490.2855. The
education coefficient changes from 0.1164 to $-0.0298$, a reminder that
each coefficient is defined relative to the other columns present.

@fig-ch9-quadratic, Chapter 8's component-plus-residual display with the
fitted curve added, shows what the squared term claims. With linear
coefficient $-0.0754$ and quadratic coefficient 0.0024, the curve's slope
is 0.0697 thousand dollars per prestige point at prestige 30, 0.1665 at
50, and 0.2633 at 70: income rises with prestige throughout the records,
and faster at the top, where the straight line falls below several
high-prestige occupations. The turning point, 15.5906, lies below the
smallest observed prestige of 17.3, so the falling branch of the parabola
is outside the records. The coordinate $-0.0754$ is the slope at prestige
zero, a value no occupation has, so it is not a "negative prestige
effect".

Centering fixes that reading without changing the curve, because it is a
change of coordinates. Prestige and its square correlate .9849; centering
prestige at its mean, 47.3276, before squaring lowers that correlation to
.3216. The quadratic coefficient and its SE stay 0.0024 and 0.0009, and
every fitted value stays the same. The linear coefficient becomes 0.1536
with SE 0.0347: the slope at mean prestige, a different quantity from
the slope at prestige zero, whose SE is about 2.6 times as large. The
smaller SE belongs to a slope read in the middle of the records, not to a
sharper estimate of the old one.

### A log-transformed outcome

Changing a predictor's form keeps the question. Changing the outcome's
scale changes it.

$$
\log Y_i=\beta_0+\sum_j\beta_jx_{ij}+\varepsilon_i,\qquad
100\,(e^{\beta_j}-1)\ \text{percent per unit of } x_j .
$$

A coefficient in a log-income model is an additive difference in expected
log income. Exponentiating turns it into a multiplicative difference in
geometric-mean income under that model, and $100(e^{\widehat\beta}-1)$
expresses the multiplier as a percentage. The geometric mean is not the
arithmetic mean, so this is not automatically a statement about average
income in dollars.

```{r}
#| label: ch9-log-outcome
#| echo: true
m_log <- lm(log(income) ~ education + prestige + women + type, data = d)
100 * (exp(coef(m_log)[c("education", "prestige", "women")]) - 1)
```

\Needspace{10\baselineskip}

| Term | Log coefficient | Percent | Lower | Upper |
|:--|--:|--:|--:|--:|
| education | 0.0076 | 0.7661 | $-4.2041$ | 5.9941 |
| prestige | 0.0207 | 2.0965 | 1.4351 | 2.7622 |
| women | $-0.0085$ | $-0.8430$ | $-1.0189$ | $-0.6667$ |

: Log-model contrasts on a geometric-mean percent scale. {#tbl-ch09-log tbl-colwidths="[22,22,18,19,19]"}

One more prestige point is associated with about 2.0965 percent higher
geometric-mean income at the same education, percentage of women, and
type, with interval $[1.4351,\ 2.7622]$ percent; one more percentage point
of women with about $-0.843$ percent. The two type coefficients, $-0.0137$
and 0.0708 on the log scale, are $-1.3584$ and 7.3341 percent. The
intervals in @tbl-ch09-log are the classical log-scale intervals with both
endpoints transformed, which is legitimate because the transformation is
monotone.

The criteria do not carry across this change. This model's AIC is 8.091
against 474.604 for `m2`, and the difference means nothing: the two
likelihoods are densities of different response variables. Keep the
outcome and the question fixed inside any comparison, and treat a change
of scale as a new question rather than as data cleaning [@fox2019].

On the log scale, the professional residual SD is 0.3276, against
0.1937 for blue-collar and 0.1607 for white-collar occupations. On the
income scale the corresponding SDs are 4.3242, 1.0437, and 1.0243.
The new scale evens the spread only partly, and also changes the
question. The next section keeps the question and changes the uncertainty.

## Robust Standard Errors for Unequal Spread {#sec-ch9-hc3}

The first half changed or compared the mean part. This half keeps that
mean part and revisits the error conditions: classical standard errors
rested on independent records and one common error variance. We first
reconsider unequal spread, then dependence within schools. These sections
can be read after Chapter 8 using the same occupation model.

### Unequal spread and HC3 standard errors

Every classical standard error so far assumed one common error variance
for all 98 occupations. Keep the same records and fitted mean, but allow
each record to contribute a different error variance. **Must every
coefficient's SE grow? Which fitted quantities would change?** Use the
residual pattern and the SE comparison below to check the prediction.

@fig-ch9-spread shows why one common variance is doubtful here.

```{r}
#| label: fig-ch9-spread
#| fig-width: 6.4
#| fig-height: 3
#| fig-cap: "Residuals of the full occupation model by type, with each type's residual SD."
#| fig-alt: "Jittered open circles of residuals in thousand dollars, one column per type, around a horizontal line at zero. Blue-collar and white-collar residuals stay between about −2.4 and 2.6, with SDs 1.0437 and 1.0243; professional residuals run from about −7.8 to 14.2, with SD 4.3242."
spread <- data.frame(type = as.numeric(d$type), residual = residuals(m2))
type_sd <- data.frame(type = 1:3, label = paste("SD", format(round(
  tapply(spread$residual, d$type, sd), 4), digits = 15, scientific = FALSE,
  trim = TRUE)))
ggplot2::ggplot(spread, ggplot2::aes(type, residual)) +
  ggplot2::geom_hline(yintercept = 0, colour = "grey45") +
  ggplot2::geom_point(shape = 1, position = ggplot2::position_jitter(
    width = .12, height = 0, seed = 12345)) +
  ggplot2::geom_text(data = type_sd, ggplot2::aes(type, 15.5,
    label = label), size = 3.3) +
  ggplot2::scale_x_continuous(breaks = 1:3,
    labels = c("Blue-collar", "Professional", "White-collar")) +
  ggplot2::labs(x = NULL, y = "Residual (thousand dollars)") +
  ggplot2::theme(panel.grid.major.x = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank())
```

The residuals of professional occupations spread about four times as
widely as the others: their SD is 4.3242 thousand dollars against 1.0437
for blue-collar and 1.0243 for white-collar occupations. A single $s^2$
averages these spreads, so it overstates the uncertainty contributed by
some rows and understates it for others.

A short graph shows the same comparison: `resid(m2)` supplies the fitted
residuals, and `type` puts each occupation in its group. The finished
figure adds a zero line, labels and a small horizontal displacement so
that overlapping points remain visible.

```{r}
#| label: ch9-spread-demo
#| echo: true
#| eval: false
library(ggplot2)
ggplot(d, aes(type, resid(m2))) + geom_point()
```

HC3 replaces the single pooled variance by each row's squared residual,
adjusted for its **leverage**: how unusual its predictor profile is
relative to the other records. That adjustment allows for the smaller
fitted residuals high-leverage rows can produce. The coefficient
covariance combines these row contributions with the same predictor
information used in the fit [@white1980; @mackinnon1985; @long2000;
@zeileis2004]. Appendix A derives the estimator.

The following call compares the two SE calculations on the same `m2`.
`vcov()` supplies the classical coefficient covariance, `diag()` extracts
its variances, and their square roots are the SEs. `vcovHC()` replaces
that covariance by HC3; the fitted coefficients are unchanged.

```{r}
#| label: ch9-hc3-demo
#| echo: true
#| eval: false
V_hc3 <- sandwich::vcovHC(m2, type = "HC3")
cbind(classical = sqrt(diag(vcov(m2))),
      HC3 = sqrt(diag(V_hc3)))
```

The professional coefficient's SE grows by about 40%, from 1.4631 to
2.0928, while the white-collar SE grows by about 7%, from 0.9848 to
1.0548. The education SE grows from 0.2752 to 0.3823, with robust 95%
interval $[-0.6429,\ 0.8758]$: wider, with the same reading of an
imprecise comparison. The SE for the percentage of women *falls*, from
0.0097 to 0.0079. The direction depends on how the residual spread and
predictor profiles combine; wide spread in one group alone cannot tell
us that every SE will rise.

The prediction therefore needs two distinctions. HC3 changes the
covariance, so it can raise or lower an SE; every coefficient, fitted
value and residual stays exactly what it was. Switching estimators after seeing which gives the preferred
conclusion turns a stated method into a selection. For Section 1.5's
group-only contrast, professional incomes spread about 2.8 times as
widely as the least spread type, and HC3 raises the SE from 0.9525 to
1.0733.

### A robust block test

A joint Wald test measures how far the fitted type coefficients lie
from zero in units of their estimated covariance, divided by the number
of restrictions. Appendix A gives its matrix form; with the classical
covariance it reproduces Section 3.2's block $F$. Substituting $\widehat{\mathbf V}_{\mathrm{HC3}}$
for $\widehat{\mathbf V}$ gives an approximate Wald version of the same
block test: for the type block, 0.0345 with $p=.9661$, against the
classical 0.0319 and .9686. Here the substitution changes nothing of
substance, because the type coefficients sit close to the restriction,
each within a quarter of its classical SE of zero, so both statistics are
near zero and far below the .05 critical value, 3.0954. A change of
covariance matters most when a statistic lies near its critical value.
The two also differ in kind: the classical statistic has an exact
$F_{2,92}$ reference under the Normal linear model, while the robust one
has only an approximate reference, justified as the sample grows, not
the exact finite-sample result of Section 3.2's test.

### Limits of robust standard errors

HC3 has one job: under a linear conditional mean and independent
observations, it estimates the coefficient covariance without assuming a
common error variance. Several problems lie outside that job.

- **Curvature.** If the mean function is wrong, the coefficients estimate
  a different quantity (Appendix A names it), and a robust covariance
  describes that quantity's variability without making it the quantity
  of interest.
- **Dependence.** Rows that are not independent break the estimator's own
  derivation. Section 7 shows a robust row-level SE and a naive one
  agreeing closely while both understate the uncertainty by nearly half.
- **Selection and causal identification.** Selection and missing outcomes
  change what the records represent, and causal identification comes
  from the design or from an argument about the conditioning set.
- **Measurement error.** With one predictor, independent classical
  measurement error attenuates its slope toward zero by the predictor's
  reliability; with several predictors the direction of bias can differ.
  A different covariance estimator does not repair measurement error
  [@cohen2003]. Independent mean-zero error in the outcome widens the
  residual spread and the standard errors but leaves the slope's target
  unchanged. Chapter 14, Section 5.1, develops attenuation.
- **Thin support.** The type contrast's robust SE of 1.1827 is still the
  uncertainty of a comparison the records barely observe.

A report therefore names the covariance estimator and the reason for it,
and keeps these limits attached to the answer.

## Resampling the Records {#sec-ch9-bootstrap}

### Choosing what to resample

Every uncertainty statement so far came from a formula: a coefficient
covariance matrix, classical or robust, propagated through a linear
combination. The **bootstrap** offers a different approximation: refit
the model to many resamples of the records and read the spread of the
refitted estimates directly [@efron1979; @efron1993; @davison1997]. What
to resample is an assumption, not a default.

Chapter 3, Appendix B, described the procedure: draw records with
replacement, refit, repeat, and read the spread of the estimates. Here we
choose **case resampling**, drawing whole occupation records, for two
reasons. Their predictors were not designed, so records and predictors
are naturally treated as sampled together. And their residual spread
differs by type (@fig-ch9-spread), which resampling whole rows preserves
without one common error distribution [@freedman1981]. Appendix B
compares the residual and wild schemes, which hold the design fixed and
model the errors, on these same targets.

### Case resampling for the full model

Resample the 98 rows with replacement, refit `m2`, and keep the education
slope. Each selected row brings its outcome and all its predictors with
it. One repetition can be shown in three lines: `rows` names the selected
records, and `update()` refits the same model formula on those records.

```{r}
#| label: ch9-bootstrap-one
#| echo: true
#| eval: false
set.seed(90909)
rows <- sample.int(nrow(d), replace = TRUE)
coef(update(m2, data = d[rows, ]))["education"]
```

Repeat this operation $B=2000$ times and collect the slopes in
`boot_education`. Their SD estimates the slope's sampling SE; their
2.5th and 97.5th percentiles supply the percentile interval. The seed
90909 and repetition count belong in the record because another set of
resamples gives slightly different summaries.

```{r}
#| label: ch9-bootstrap
#| include: false
set.seed(90909)
B <- 2000
boot_education <- replicate(B, {
  rows <- sample.int(nrow(d), replace = TRUE)
  coef(lm(income_k ~ education + prestige + women + type,
    data = d[rows, ]))["education"]
})
```

```{r}
#| label: ch9-bootstrap-summary
#| echo: true
#| eval: false
sd(boot_education)
quantile(boot_education, c(.025, .975))
```

A histogram makes the repeated slopes visible. Its vertical scale is
density, so it can be compared with the Normal reference curve added in
@fig-ch9-bootstrap. The simpler call shows the same resampling
histogram without the reference curve and interval lines.

```{r}
#| label: ch9-bootstrap-plot
#| echo: true
#| eval: false
ggplot(data.frame(slope = boot_education), aes(slope)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40)
```

```{r}
#| label: fig-ch9-bootstrap
#| fig-width: 6.4
#| fig-height: 3
#| fig-cap: "The 2,000 case-resampled education slopes (seed 90909). Solid lines mark the percentile interval; the dashed curve is the Normal distribution with the classical SE 0.2752."
#| fig-alt: "Histogram of refitted education slopes from about −1.2 to 1.1, centred near 0.1 and slightly longer on the left, with solid vertical lines at −0.683 and 0.744. A dashed Normal curve centred at 0.1164 is narrower and taller than the histogram."
normal_curve <- data.frame(x = seq(-1.5, 1.5, length.out = 300))
normal_curve$density <- dnorm(normal_curve$x, coef(m2)["education"],
  sqrt(vcov(m2)["education", "education"]))
ggplot2::ggplot(data.frame(slope = boot_education),
    ggplot2::aes(slope)) +
  ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = 40, fill = "grey80", colour = "white") +
  ggplot2::geom_line(data = normal_curve, ggplot2::aes(x, density),
    linetype = 2) +
  ggplot2::geom_vline(xintercept = quantile(boot_education,
    c(.025, .975)), linewidth = .7) +
  ggplot2::labs(x = "Education slope in a resample (thousand dollars per year)",
    y = "Density")
```

@fig-ch9-bootstrap is the resampling distribution itself. The 2,000
refitted slopes spread more widely than the Normal curve built from the
classical SE: their SD, the bootstrap SE, is 0.3699, and their middle
95%, the percentile interval, runs from $-0.683$ to 0.744. The same
resamples give the other two targets of @tbl-ch09-bootstrap, the
professional-minus-white-collar contrast and the professional
coefficient. A resample could omit a whole occupational type and lose a
type coefficient; all 2,000 resamples here contained all three types, so
all 2,000 refits are usable.

\Needspace{9\baselineskip}

| Target | Estimate | Classical SE | HC3 SE | Bootstrap SE | Percentile 2.5% | Percentile 97.5% |
|:--|--:|--:|--:|--:|--:|--:|
| Education slope | 0.1164 | 0.2752 | 0.3823 | 0.3699 | $-0.6830$ | 0.7440 |
| prof $-$ wc | 0.0889 | 1.0800 | 1.1827 | 1.1187 | $-1.9613$ | 2.4313 |
| prof coefficient | 0.3242 | 1.4631 | 2.0928 | 1.9926 | $-3.1416$ | 4.5443 |

: Model-based, HC3, and case-bootstrap uncertainty for three targets in the full occupation model (seed 90909, $B=2000$). {#tbl-ch09-bootstrap tbl-colwidths="[22,12,14,12,14,13,13]"}

Read the table one row at a time. The education slope's bootstrap SE of
0.3699 sits between the classical 0.2752 and the HC3 0.3823: three
calculations, one conclusion that this comparison is imprecise. The type
contrast's bootstrap SE of 1.1187 exceeds the classical 1.0800, with
percentile interval $[-1.9613,\ 2.4313]$. The professional coefficient
moves most, from a classical 1.4631 to 1.9926, close to the HC3 2.0928:
with only 31 professional occupations spreading as widely as
@fig-ch9-spread shows, a resample's composition matters.

Agreement between HC3 and the bootstrap is not independent confirmation:
both work on the same 98 records. And a different seed would move the
last decimal of a bootstrap summary, which is why the seed and $B=2000$
are reported.

### Bootstrap intervals and their limits

The **normal-approximation interval** uses the resampling distribution
only for its spread: the original estimate plus and minus 1.96 bootstrap
standard errors. For the professional-minus-white-collar contrast it is
$[-2.1038,\ 2.2815]$, beside the percentile interval
$[-1.9613,\ 2.4313]$. The endpoints tend to be close when the resampling
distribution is approximately Normal and centered near the estimate. Here the
three distributions have skewness $-0.3891$, 0.2635, and 0.3785, read by
their sign as in Chapter 2, Section 2.1: the education slopes have the
slightly longer left tail visible in the figure, the other two a slightly
longer right tail, and the asymmetry is mild enough that the two kinds of
endpoint agree closely. For a strongly skewed or biased resampling
distribution, **bootstrap-$t$** and **BCa** intervals are the
higher-order alternatives; Appendix B defines both in words.

What resampling assumes matters more than which interval is chosen.
Drawing occupations with replacement treats them as independent and
exchangeable, and there are reasons to doubt both: the occupations share
one historical labour market, where common conditions can move related
incomes together; the 98 rows were chosen by complete records rather than
sampled at random; and their measurements are aggregates over workers.
Resampling does not repair thin cross-type support, correct the
measurement problems Chapter 8 documented, or make an observational
comparison causal. And when a design has clusters, strata, or weights,
the resampling unit has to follow the design rather than the rows, which
is where the next section begins.

## Clustered Data: Project STAR {#sec-ch9-star}

```{=latex}
\hypertarget{sec-ch9-star}{}
```

```{r}
#| label: ch9-star-prepare
data("STAR", package = "AER")
star_vars <- c("stark", "readk", "lunchk", "gender", "ethnicity",
  "experiencek", "schoolidk")
star_two_arm <- !is.na(STAR$stark) & STAR$stark %in% c("regular", "small")
star_df <- droplevels(STAR[star_two_arm &
  complete.cases(STAR[, star_vars]), star_vars])
star_df$stark <- relevel(star_df$stark, ref = "regular")
star_observed <- droplevels(STAR[star_two_arm & !is.na(STAR$readk),
  c("stark", "readk", "schoolidk")])
star_observed$stark <- relevel(star_observed$stark, ref = "regular")
```

The occupation records were observational, and the reading study was a
small experiment in one school. Project STAR is a large experiment of a
different shape: kindergarten pupils were randomized to class types
within their schools. The randomization supports reading the class-type
comparison as an assignment comparison, and the schools make pupils'
outcomes dependent, which the uncertainty calculation has to follow.

### The study and its question

The question is specific:

> Among eligible kindergarten pupils in the participating Tennessee
> schools, what was the mean reading-score difference under assignment to
> a small rather than a regular class?

The outcome is `readk`, an end-of-kindergarten reading scale score, and
the field `stark` records the kindergarten class-type condition. Reading
the small-versus-regular comparison as an assignment comparison rests on
the documented protocol, not on the column name or the `lm()` call:
random assignment occurred **within** participating schools [@finn1990;
@mosteller1995]. The regular-with-aide arm lies outside this two-arm
comparison and is excluded by design.

Three limits come with the question. Class **type** is not a measured
headcount, so the analysis estimates no dose response in pupils per
room. The distributed field does not verify each pupil's original
assignment; an assignment-effect reading also requires linking the
recorded type to the assignment record and checking switches, which this
teaching reanalysis does not do. And the target is the pupil-weighted
mean assignment effect for the eligible cohort, which a common
coefficient from a complete-record model reaches only through a further
link: a common effect across schools or an explicit weighting argument.

The data are distributed as `AER::STAR`, accompanying *Applied
Econometrics with R* and following the version used by Stock and Watson
[@kleiber2008; @stock2007]. They are loaded from the package, and the
analysis is retrospective rather than preregistered.

### Which records enter the analysis

A record can be absent from an analysis for two different reasons, and
they are kept apart. Pupils who entered the study after kindergarten have
no kindergarten condition: their absence is **structural ineligibility**.
Eligible kindergarten pupils whose reading score or covariates were not
recorded are **missing**, and missingness can distort a comparison. Only
the variables a calculation uses enter its complete-record rule, so a
missing mathematics score (`mathk`) does not exclude a usable reading
record.

\Needspace{12\baselineskip}

| Stage | $n$ |
|:--|--:|
| Distributed records | 11,598 |
| Kindergarten condition recorded | 6,325 |
| Eligible small or regular | 4,094 |
| Eligible, reading observed | 3,745 |
| Complete primary records | 3,735 |
| Primary regular | 2,001 |
| Primary small | 1,734 |
| Primary schools | 79 |

: STAR eligibility and reading-analysis completeness. {#tbl-ch09-star-sample tbl-colwidths="[60,40]"}

The 5,273 records without a kindergarten condition are later entrants,
not failed measurements; Chapter 2, Section 1.3, explains why such
ineligibility and missing values need different denominators. Among the
4,094 eligible two-arm records, 349 lack a reading score. Ten more of
the 3,745 observed-reading records lack a covariate, leaving 3,735
primary records in 79 schools. Complete-record inference assumes that,
given the declared covariates and within arm and school, whether a score
was recorded does not depend on the score itself. Similar missing
percentages across arms do not establish that condition. The 3,745
observed-reading records remain a declared sensitivity analysis.

### The arm difference, unadjusted and adjusted

Section 1.1's identity applies unchanged. With $D_i=1$ for the small arm
and 0 for regular,

$$
\widehat Y_i=\widehat\beta_0+\widehat\delta D_i,\qquad
\widehat\beta_0=\bar Y_{D=0},\qquad
\widehat\delta=\bar Y_{D=1}-\bar Y_{D=0}=440.5474-434.7323=5.8151.
$$

On the 3,745 observed-reading records the slope reproduces the arm
difference of 5.8151 reading points exactly, as Chapter 7's binary
predictor did for 24 pupils. The identity is algebraic; randomization
adds the reason to read the difference as an assignment comparison.

Two cautions attach to the number. Treating the rows as independent with
a common variance, the slope's $t$ test equals the pooled two-sample
test, here $t=5.6039$, which ignores that pupils sit in schools (Section
7.4). And raw arm means weight each school by its own arm-specific pupil
count, so the pooled difference equals the pupil-weighted cohort target
only under common assignment probabilities and comparable observation
patterns, or after explicit standardization to target school weights.

The adjusted model adds free-lunch status, gender, ethnicity, teacher
experience, and school indicators. The school indicators match the
blocks within which assignment happened; the other covariates are the
declared adjustment set, not a substitute for randomization.

```{r}
#| label: ch9-star-models
#| echo: true
star_unadjusted <- lm(readk ~ stark, data = star_df)
star_primary <- lm(readk ~ stark + lunchk + gender + ethnicity +
  experiencek + schoolidk, data = star_df)
```

On the common 3,735 primary rows, the unadjusted difference is 5.8955 and
the adjusted coefficient is 6.6575. Adjustment can improve precision when
the added variables predict reading. It is not needed to remove baseline
confounding that a well-implemented randomization already prevents; this
is Sections 2.2--2.3's point about randomized groups, on a larger design.
Adjustment also does not automatically preserve the causal target: when
the covariates were recorded, which rows enter, and what the fitted
comparison compares still need scrutiny.

Chapter 8, Section 2.2, kept model comparisons on identical rows. The
earlier 5.8151 comes from 3,745 rows, so it is not the same-row
baseline for 6.6575; comparing the two would mix a change of records with
a change of model. The same-row pair is 5.8955 against 6.6575. Similar
unadjusted and adjusted coefficients are consistent with successful
randomization; they do not show that implementation, missingness, or
weighting problems are absent.

### Clustered standard errors

Pupils are the outcome units. Schools are the randomization blocks and
the approximately independent units for the uncertainty calculation; the
schools themselves were not assigned to arms. Chapter 3, Section 2.3,
showed that the uncertainty of a school-blocked comparison follows its
within-school comparisons, and @fig-ch9-star-schools shows those
comparisons for STAR.

```{r}
#| label: fig-ch9-star-schools
#| fig-width: 6.4
#| fig-height: 3
#| fig-cap: "Small-minus-regular mean reading difference in each of the 78 schools that observed both class types, ordered by size. The dashed line is their unweighted mean, 6.7094."
#| fig-alt: "Dot plot of 78 school differences, one dot per school, rising from about −32 to 59 reading points in an S-shaped column; 25 dots lie left of the solid zero line, and a dashed line marks the mean, about 6.71."
school_means <- with(star_observed,
  tapply(readk, list(schoolidk, stark), mean))
school_diff <- sort(school_means[, "small"] - school_means[, "regular"])
ggplot2::ggplot(data.frame(difference = school_diff,
    school = seq_along(school_diff)), ggplot2::aes(difference, school)) +
  ggplot2::geom_vline(xintercept = 0, colour = "grey45") +
  ggplot2::geom_vline(xintercept = mean(school_diff), linetype = 2) +
  ggplot2::geom_point(size = 1.4) +
  ggplot2::labs(x = "Small minus regular mean reading score in the school",
    y = "Schools, ordered") +
  ggplot2::theme(axis.text.y = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank())
```

In the 78 schools that observed pupils in both arms, the small-class
advantage ranges from about $-32$ to 59 reading points, and it is
negative in 25 schools. Pupils in one school share whatever made their
school's difference large or small: its teachers, its intake, its
classrooms. The 3,745 pupils therefore do not supply 3,745 independent
pieces of information about the arm difference; the schools do. An
average of 78 independent differences with SD 15.4192 has SE
$15.4192/\sqrt{78}=1.7459$, close to the clustered standard errors below
and far from the row-level 1.0377.

A **cluster-robust** covariance builds this into the robust covariance
of Section 5.1: the design component stays outside, and only the middle
changes. Here $\mathbf X$ is the design matrix, with one row per pupil
and one column per coefficient; $\mathbf X_g$ and $\mathbf e_g$ hold
school $g$'s rows and residuals. The score $\mathbf s_g=\mathbf X_g^{\mathsf T}\mathbf e_g$ sums the
residual contributions **within** the school before an outer product is
taken, and the CR1/HC1 covariance is

$$
\widehat{\mathbf V}_{\mathrm{CR1}}=c_G(\mathbf X^{\mathsf T}\mathbf X)^{-1}
\Big(\sum_{g=1}^{G}\mathbf s_g\mathbf s_g^{\mathsf T}\Big)(\mathbf X^{\mathsf T}\mathbf X)^{-1},
\qquad \mathbf s_g=\mathbf X_g^{\mathsf T}\mathbf e_g,\qquad
c_G=\frac{G}{G-1}\frac{n-1}{n-p}.
$$

The within-school sums are what distinguish this from treating every
pupil as independent, and $c_G$ is a finite-cluster correction. With
$G=79$ schools, intervals use a $t_{78}$ critical value, 1.9908
[@zeileis2020].

```{=latex}
\Needspace{10\baselineskip}
```

```{r}
#| label: ch9-star-cluster
#| echo: true
star_V <- sandwich::vcovCL(star_primary, cluster = star_df$schoolidk,
  type = "HC1", cadjust = TRUE)
sqrt(star_V["starksmall", "starksmall"])
```

\Needspace{12\baselineskip}

| Model | $n$ | Estimate | Row-level SE | HC3 SE | Clustered SE | Clustered interval |
|:--|--:|--:|--:|--:|--:|:--|
| Observed: unadjusted | 3,745 | 5.8151 | 1.0377 | 1.0416 | 1.8476 | [2.1368, 9.4935] |
| Primary: unadjusted | 3,735 | 5.8955 | 1.0399 | 1.0439 | 1.8491 | [2.2143, 9.5767] |
| Primary: adjusted | 3,735 | 6.6575 | 0.9161 | 0.9509 | 1.6974 | [3.2782, 10.0368] |
| Observed: school-adjusted | 3,745 | 6.6273 | 0.9504 | 0.9860 | 1.7797 | [3.0842, 10.1703] |

: Class-type comparisons with row-level, HC3, and school-clustered standard errors; intervals use the clustered SE with $t_{78}$. {#tbl-ch09-star-estimates tbl-colwidths="[24,8,10,11,9,12,26]"}

@tbl-ch09-star-estimates holds four distinct calculations, each with its
own records and specification. Each clustered interval uses the school
identifiers from that model's own rows. For the observed-reading
comparison, the row-level SE is 1.0377 and HC3 gives 1.0416, while
school clustering gives 1.8476, nearly twice as large. Every row shows
the same pattern. Unequal spread across pupils was not the main issue;
dependence within schools was. Appendix A explains how the within-school
contributions enter the covariance.

These are cluster-asymptotic approximations, which improve as the number
of schools grows; they are not exact randomization inference or a
multilevel model. The design, not a residual plot, chooses the
independence unit. And 79 clusters do not by themselves certify the
approximation: the balance of information across schools and the
influence of particular schools also matter.

### Reporting the result and its limits

A chapter-authored report can now state:

> On 3,735 complete two-arm kindergarten records, the common adjusted
> small-versus-regular class-type coefficient was 6.6575 reading points,
> with approximate school-clustered interval [3.2782, 10.0368]. Its causal
> mean-effect interpretation relies on documented within-school
> randomization, a verified link from recorded class type to original
> assignment, suitable complete-record selection, independent-school
> uncertainty, and a common-effect or target-matching weighting argument.
> It concerns assigned class type, one kindergarten reading outcome, and
> participating Tennessee schools, not realized headcount or a portable
> policy dose response.

As in Chapter 8, Section 7, the report connects statistical functions
(the sample and model, the design and covariance reasoning, the estimate
with its interval, and the bounded answer) with substantive ones (the
pupils, test and schools, assignment and its qualifications, and the
substantive claim).

Judging the report separates what is *stated* from what is
*demonstrated*. The randomization protocol is documented in the study's
own publications; the selection and weighting conditions and the link
from recorded class type to original assignment are stated here, not
shown. Where a necessary link cannot be assessed, the report names the
limitation rather than assuming it away or quietly changing the target.
The design literature discusses implementation and interpretation issues
that one coefficient cannot settle [@krueger1999; @hanushek1999;
@nye2000; @chetty2011], and the result does not transport by itself to
other schools, periods, outcomes, or education systems. Those limits
belong inside the answer; a crossed significance boundary is not what
makes a result [@gelman2006; @gelman2020]. Chapter 14 returns to
dependence as a design fact, reuses this calculation as its example, and
asks what a design supplies that adjustment cannot.

```{r}
#| label: ch9-numerical-checks
#| include: false
local({
  near <- function(x, y, tol = 5e-5)
    stopifnot(length(x) == length(y), all(abs(unname(x) - y) < tol))
  stopifnot(nrow(d) == 98, df.residual(m2) == 92)
  near(c(deviance(group_fit), anova(lm(income_k ~ 1, d), group_fit)$F[2],
    summary(group_fit)$r.squared), c(1138.1230, 24.8725, .3437))
  c0 <- c(0, 1, -1)
  c2 <- c(0, 0, 0, 0, 1, -1)
  near(c(sum(c0 * coef(group_fit)),
    sqrt(drop(c0 %*% vcov(group_fit) %*% c0)),
    sum(c2 * coef(m2)), sqrt(drop(c2 %*% vcov(m2) %*% c2))),
    c(5.5071, .9525, .0889, 1.0800))
  candidates <- list(lm(income_k ~ education + women, d),
    lm(income_k ~ prestige + women, d), m2)
  near(sapply(candidates, AIC), c(486.3288, 469.3428, 474.6040))
  near(sapply(candidates, BIC), c(496.6687, 479.6827, 492.6988))
  near(coef(m2)[2:4] * sapply(d[c('education','prestige','women')], sd) /
    sd(d$income_k), c(.0757, .5685, -.3971))
  near(sapply(c('education','prestige','women'), function(v)
    summary(m2)$r.squared - summary(update(m2,
      as.formula(paste('. ~ . -', v))))$r.squared), c(.0007,.0625,.1212))
  near(tapply(residuals(m_log), d$type, sd), c(.1937,.3276,.1607))
  near(c(sd(boot_education), quantile(boot_education,c(.025,.975))),
    c(.3699,-.6830,.7440))
  set.seed(90909)
  draws <- replicate(2000, {
    rows <- sample.int(nrow(d), replace=TRUE)
    b <- coef(lm(income_k ~ education + prestige + women + type,d[rows,]))
    c(b['education'], b['typeprof']-b['typewc'], b['typeprof'])
  })
  stopifnot(all(is.finite(draws)))
  near(apply(draws,1,sd), c(.3699,1.1187,1.9926))
  near(coef(ancova)[-1],c(3.6266,2.0362,.6932))
  near(anova(ancova,varying_slopes)$F[2],1.5054)
  stopifnot(nrow(star_observed)==3745,nrow(star_df)==3735,
    nlevels(star_df$schoolidk)==79,length(school_diff)==78)
  fits <- list(lm(readk ~ stark,star_observed),star_unadjusted,
    star_primary,lm(readk ~ stark + schoolidk,star_observed))
  clusters <- list(star_observed$schoolidk,star_df$schoolidk,
    star_df$schoolidk,star_observed$schoolidk)
  rows <- t(sapply(seq_along(fits),function(i) {
    f <- fits[[i]]
    get_se <- function(V) sqrt(V['starksmall','starksmall'])
    c(coef(f)['starksmall'],get_se(vcov(f)),
      get_se(sandwich::vcovHC(f,type='HC3')),
      get_se(sandwich::vcovCL(f,cluster=clusters[[i]],
        type='HC1',cadjust=TRUE)))
  }))
  expected <- rbind(c(5.8151,1.0377,1.0416,1.8476),
    c(5.8955,1.0399,1.0439,1.8491),c(6.6575,.9161,.9509,1.6974),
    c(6.6273,.9504,.9860,1.7797))
  near(as.vector(rows),as.vector(expected))
})
```

## Reading Bridge {#sec-ch9-summary}

A factor enters a regression as indicators, and the one-way analysis of
variance is that regression; a contrast's uncertainty needs the
covariance between coefficients, not the sum of two standard errors. The
analysis of covariance adds a pretest: the posttest-only, gain-score, and
adjusted analyses differ only in the pretest coefficient, and adjustment
sharpens a randomized comparison without supplying what randomization
supplies. Nested models on the same records and outcome are compared
through their residual sums of squares, and information criteria rank
models by penalized in-sample fit, never by held-out error. A squared
predictor has no single slope, and a logged outcome asks a new question
on a geometric-mean scale. HC3, the case bootstrap, and a clustered
covariance each change the uncertainty for a stated reason and leave the
coefficients alone; none repairs curvature, missing outcomes, weighting,
or the absence of a design. In Project STAR, schools rather than pupils
were the approximately independent units, and the standard error nearly
doubled.

In the additive models, a group comparison is the same at every value
of the other predictors; Section 2.3 checked this common-slope
requirement for the reading study. Chapter 10 lets a
comparison vary with another predictor and develops how to read, probe,
and report such an interaction.

\Needspace{14\baselineskip}

### Key terms {.unnumbered}

Use this short reference to revisit the concepts introduced here. The
Part One Concept Reference gives the broader reference and the principal
home of each concept; indicator coding, partial slopes, coefficient
covariance, and HC3 are explained in Chapters 7 and 8, and regression
toward the mean in Chapter 7.

| Term and common alternatives | Meaning |
|:--|:--|
| ANCOVA (analysis of covariance) | A regression comparing groups while including a quantitative covariate with one common slope; its adjusted group means are read at a stated covariate value within common support. The posttest-only, gain-score and ANCOVA analyses are one regression whose pretest coefficient is 0, 1 or estimated; adjustment cannot supply what random assignment supplies. |
| Homogeneity of regression slopes (parallel slopes) | ANCOVA's condition that the outcome--covariate slope is the same in every group, checked by testing the group-by-covariate product block; when the slopes differ, the adjusted difference depends on the covariate value at which it is read. |
| Nested-model $F$ test (extra sum of squares; block test) | Compares a reduced model with a fuller one fitted to the same records and outcome: the RSS reduction per added coefficient, divided by the fuller model's residual mean square, referred to $F$ with the number of restrictions and the fuller model's residual df. It tests the added block jointly, not each coefficient separately. |
| Information criteria (AIC; BIC) | Penalized in-sample fit: $-2\log\widehat L$ plus $2k$ (AIC) or $k\log n$ (BIC), with $k$ counting the coefficients and the residual variance. Smaller is preferred among models fitted to the same records and outcome scale; neither is held-out error. |
| Standardized coefficient (beta weight) | A partial slope rescaled by the predictor and outcome SDs, in outcome-SD units per predictor SD, with the other predictors held fixed. It differs from standardization over named profiles and does not measure causal importance. |
| Polynomial term (quadratic term; curvature) | A power of a predictor added as a further column so that the fitted mean can curve. The predictor's comparison then depends on its value; centering before squaring reduces collinearity without changing the fitted curve. |
| Log-transformed outcome (geometric-mean scale) | Fitting the logarithm of the outcome, so that a coefficient is an additive difference in expected log outcome and $100(e^{\beta}-1)$ percent is a multiplicative difference in geometric means. A changed outcome scale is a changed question, and its criteria are not comparable with the original scale's. |
| Bootstrap (case resampling; percentile interval) | Refitting the model to many resamples of the records drawn with replacement and using the spread of the refitted estimates as an SE, or their middle 95% as a percentile interval. It treats the resampled units as independent and exchangeable; it does not repair dependence, thin support, measurement error, or an observational design. |
| Cluster-robust standard error (school-clustered covariance) | A coefficient covariance that sums residual contributions within clusters, such as schools, before combining them, allowing dependence within clusters under an approximation that treats clusters as independent. It changes uncertainty, not the coefficient, and does not repair missing outcomes or weighting. |

: Selected concepts for the regression toolbox. {#tbl-ch09-key-terms tbl-colwidths="[28,72]"}

### Further reading {.unnumbered}

- The [OpenIntro multiple-predictor and model-selection tutorial](https://openintro-ims.netlify.app/model-mlr)
  gives concrete indicator interpretations and criterion-based reading. We
  use its equation-first device without adopting automatic stepwise
  selection.
- The [ISLR regression lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch3-linreg-lab.html)
  and the [ISLR resampling lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch5-resample-lab.html)
  connect fitting, factors, criteria, and resampling in one workflow
  [@james2021].
- Fox [@fox2016] and Kutner and colleagues [@kutner2005] develop the
  linear-model algebra, the restriction statistic, and the diagnostics
  in this chapter's Appendix A.
- Efron and Tibshirani [@efron1993] and Davison and Hinkley [@davison1997]
  are the standard bootstrap references, including the bootstrap-$t$ and
  BCa intervals named in Section 6.3.
- The [sandwich documentation](https://cran.r-project.org/web/packages/sandwich/sandwich.pdf)
  specifies the HC and CR covariance calls used here [@zeileis2020].

## Exercises {.unnumbered}

Use the supplied numbers; rounding within 0.02 is acceptable unless a
different precision is requested. These prompts do not require the
appendices.

\Needspace{12\baselineskip}

<!-- prompt-id: CH09-E01 -->
### Exercise 1: Read indicator coefficients (Foundational) {.unnumbered #ex-ch09-01}

A group-only regression of income (thousand dollars) on occupation type
has observed group means 5.3741 (blue-collar, 44 records), 10.5595
(professional, 31), and 5.0523 (white-collar, 23). (1) With blue-collar as
reference, give the intercept and the two indicator coefficients. (2) Give
the intercept and both coefficients with white-collar as reference. (3)
Explain why the fitted values are the same under both codings. (4) The
full additive model's professional coefficient is 0.3242. Explain why it
differs from the professional coefficient in (1) and why neither is "the
effect of occupational type".

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E02 -->
### Exercise 2: A contrast needs covariance (Foundational) {.unnumbered #ex-ch09-02}

In the full additive model the professional and white-collar coefficients
are 0.3242 and 0.2353, with estimated variances 2.1406 and 0.9698 and
covariance 0.9720, on 92 residual df. (1) The
professional-minus-white-collar contrast is 0.0889. Calculate its
variance, and explain why the covariance enters with a minus sign and what
a positive covariance does to the result. (2) The contrast's SE is 1.08
and its 95% interval is $[-2.0561,\ 2.2338]$. A student instead adds the
two coefficient SEs (1.4631 and 0.9848). State what that number is and why
it is wrong. (3) Explain what the interval does and does not say about the
two types.

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E03 -->
### Exercise 3: A nested comparison from residual sums of squares (Foundational) {.unnumbered #ex-ch09-03}

On the same 98 records, the model with education and prestige has RSS
867.6537 on 95 df, and the full model adding the percentage of women and
the two type indicators has RSS 630.8897 on 92 df. The 5% critical value
of $F(3,92)$ is 2.7036. (1) Calculate the $F$ statistic for the added
block. (2) State the decision at 5% and what it does and does not say
about the three added columns. (3) Explain why the sequential sums of
squares from a single `anova()` call on the full model would not answer
this question.

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E04 -->
### Exercise 4: What information criteria compare (Foundational) {.unnumbered #ex-ch09-04}

The full additive model has maximized log-likelihood $-230.302$ on 98
records, with AIC 474.6040 and BIC 492.6988; the model with education and
prestige only has AIC 499.8328 and BIC 510.1727. (1) The criteria count
seven fitted parameters for the full model. Say what the seven are, and
why BIC charges the full model more for them than AIC does
($\log 98=4.585$). (2) Which model do the criteria favour, and on what
condition is that comparison meaningful? (3) A log-income version of the
full model has AIC 8.091. Explain why this number cannot be compared with
474.604.

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E05 -->
### Exercise 5: Percent and curvature (Integrative) {.unnumbered #ex-ch09-05}

Two changed models of the occupation records raise different questions.
(1) In a log-income model the prestige coefficient is 0.0207. Give the
percent difference in geometric-mean income per prestige point and say
what "geometric mean" adds to the reading. (2) In an income model with a
squared prestige term, the fitted slope is 0.0697 thousand dollars per
prestige point at prestige 30, 0.1665 at 50 and 0.2633 at 70, and the
turning point is at prestige 15.5906; prestige ranges from 17.3 to 87.2.
Explain why the model has no single prestige slope, and what the position
of the turning point tells you. (3) State in one sentence what the curve
claims about income and prestige over the observed range, and one reason
that claim is a sensitivity check rather than a discovery.

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E06 -->
### Exercise 6: Choose the uncertainty calculation (Integrative) {.unnumbered #ex-ch09-06}

The full model of the same 98 occupation records gives the following
standard errors; its coefficients and fitted values are held fixed.

| Coefficient | Classical SE | HC3 SE |
|:--|--:|--:|
| Education | 0.2752 | 0.3823 |
| Percentage of women | 0.0097 | 0.0079 |

(1) A student predicts that allowing unequal spread must make every
SE larger. Use the supplied output to assess that prediction and explain
what changes when the covariance estimator changes.

In Project STAR, the small-minus-regular reading difference on 3,745
observed records is 5.8151 points. The SE is 1.0377 treating pupils as
independent, 1.0416 with HC3, and 1.8476 clustering within 79 schools.
The corresponding independent-row and clustered 95% intervals are
[3.7807, 7.8496] and [2.1368, 9.4935]. Pupils and teachers were assigned
to class types within schools. (2) Explain which uncertainty calculation
respects this design, why row-level HC3 barely changes the SE here, and
what remains unchanged when the clustered calculation replaces it.
(3) Name two problems that none of these covariance changes repairs.

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E07 -->
### Exercise 7: One randomized study, three analyses (Integrative) {.unnumbered #ex-ch09-07}

In the Baumann records, 66 pupils in three randomized instruction groups
have a pretest and a posttest. With DRA as reference, the DRTA difference
is 3.0909 (SE 0.9614) from the posttest alone, 3.8636 (SE 0.7779) from the
gain score, and 3.6266 (SE 0.7362) from the analysis of covariance, whose
pretest coefficient is 0.6932; the DRA and DRTA pretest means are 10.5000
and 9.7273. (1) Write the three analyses as one regression and give the
pretest coefficient each uses. (2) Explain why all three estimate the same
average difference under random assignment, and show how the pretest gap
accounts for the differences between the three estimates. (3) Say which
analysis you would report, and when that choice has to be made. (4)
Suppose the groups had been intact classes that differed at the pretest.
Explain what changes, naming the paradox that can arise.

## Appendix A: Matrix Geometry, Restrictions, and Sandwich Constructions {.unnumbered #app-ch9-a}

This appendix derives the results that Sections 3.1--3.2, 5.2, and 7.4 used,
including the HC3 estimator that Chapter 8, Appendix B, stated, and
names the target least squares estimates when a linear mean is only an
approximation. The exercises do not require it.

### Projection and the normal equations {.unnumbered}

The outcome vector $\mathbf y$ stacks the $n$ observed outcomes in one
column. The $n\times p$ **design matrix** $\mathbf X$ stacks the
corresponding predictor rows: one row per record and one column per
coefficient, including a column of ones for the intercept. The $p$-entry
vector $\boldsymbol\beta$ contains those coefficients. For `m2`, these
are 98 outcomes, a $98\times6$ design matrix, and six coefficients.

Assume $\mathbf X$ has full column rank: none of its columns can be
formed from the others, so $\mathbf X^{\mathsf T}\mathbf X$ is
invertible. The least-squares criterion is
$S(\boldsymbol\beta)=(\mathbf y-\mathbf X\boldsymbol\beta)^{\mathsf T}(\mathbf y-\mathbf X\boldsymbol\beta)$.
Expanding gives
$S=\mathbf y^{\mathsf T}\mathbf y-2\boldsymbol\beta^{\mathsf T}\mathbf X^{\mathsf T}\mathbf y
+\boldsymbol\beta^{\mathsf T}\mathbf X^{\mathsf T}\mathbf X\boldsymbol\beta$,
and differentiating with respect to $\boldsymbol\beta$ yields the normal
equations

$$
\mathbf X^{\mathsf T}\mathbf X\widehat{\boldsymbol\beta}=\mathbf X^{\mathsf T}\mathbf y,
\qquad\text{equivalently}\qquad
\mathbf X^{\mathsf T}(\mathbf y-\mathbf X\widehat{\boldsymbol\beta})=\mathbf 0 .
$$

Solving gives the estimator and its fitted and residual vectors:

$$
\widehat{\boldsymbol\beta}
=(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}\mathbf y,
\qquad
\widehat{\mathbf y}=\mathbf H\mathbf y,
\qquad
\mathbf e=(\mathbf I-\mathbf H)\mathbf y,
$$

where $\mathbf H=\mathbf X(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}$
is the **hat matrix** and $\mathbf I$ is the $n\times n$ identity matrix.
The second normal equation is the geometry: $\mathbf e$ is orthogonal to
every column of $\mathbf X$, and therefore to their fitted linear
combination $\widehat{\mathbf y}$. The matrix $\mathbf H$ is symmetric
($\mathbf H^{\mathsf T}=\mathbf H$) and idempotent
($\mathbf H^2=\mathbf H$): projecting an outcome onto the space spanned
by the predictor columns, then projecting again, gives the same fit.
Its trace is $p$, which is why the leverages $h_{ii}$ sum to the number
of fitted coefficients; for `m2` they sum to 6.

With an intercept, $\mathbf e$ is also orthogonal to the constant
vector, so the centered fitted values and residuals are orthogonal.
Writing $y_i-\bar y=(\widehat y_i-\bar y)+e_i$, the cross-product term
vanishes when the squares are summed:

$$
\underbrace{\sum_i(y_i-\bar y)^2}_{TSS}
=\underbrace{\sum_i(\widehat y_i-\bar y)^2}_{RegSS}
+\underbrace{\sum_i e_i^2}_{RSS}.
$$

This is the projection basis for the regression sum-of-squares
decomposition. Without an intercept, the centered decomposition does
not follow automatically.

Two consequences used earlier follow immediately. Any full-rank recoding
$\mathbf X\mathbf A$, with $\mathbf A$ invertible, spans the same column
space, so $\mathbf H$ and $\widehat{\mathbf y}$ are unchanged: releveling a
factor or centering a predictor cannot alter a single fitted value. Adding
a genuinely new column enlarges the space and can change every fitted
value. And under the classical error model,
$\widehat{\mathbf V}=s^2(\mathbf X^{\mathsf T}\mathbf X)^{-1}$ with
$s^2=RSS/(n-p)$, which is where every classical standard error in this
chapter came from. Solving the normal equations directly for `m2`
reproduces `lm()`'s coefficients to about $10^{-13}$. This equation explains the estimator, not its
implementation: `lm()` uses a QR decomposition rather than forming the
inverse, a more stable numerical route [@fox2016; @fox2019].

### General linear restrictions {.unnumbered}

Let $\mathbf R$ be $q\times p$ of rank $q$ and consider
$\mathbf R\boldsymbol\beta=\mathbf r$. Since $\widehat{\boldsymbol\beta}$
has covariance $\sigma^2(\mathbf X^{\mathsf T}\mathbf X)^{-1}$ under the
classical model, the vector $\mathbf R\widehat{\boldsymbol\beta}-\mathbf r$
has covariance
$\sigma^2\mathbf R(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf R^{\mathsf T}$.
Standardizing by that covariance and dividing by $q$ gives

$$
F=\frac{(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)^{\mathsf T}
[\mathbf R(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf R^{\mathsf T}]^{-1}
(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)}{q\,s^2}.
$$

When the null restriction $\mathbf R\boldsymbol\beta=\mathbf r$ holds,
under the classical Normal model, this has an exact central
$F_{q,\,n-p}$ reference: the numerator and denominator quadratic forms
are independent chi-squares. Under an alternative the numerator is
noncentral.

It is the same number the extra-sum-of-squares formula produces.
Minimizing $S(\boldsymbol\beta)$ subject to
$\mathbf R\boldsymbol\beta=\mathbf r$ with a Lagrange multiplier gives the
restricted estimator
$\widetilde{\boldsymbol\beta}=\widehat{\boldsymbol\beta}
-(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf R^{\mathsf T}
[\mathbf R(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf R^{\mathsf T}]^{-1}
(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)$, a correction applied
only in the directions the restriction touches. Substituting it into the
criterion shows that the increase in residual sum of squares is

$$
RSS_R-RSS_F=(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)^{\mathsf T}
[\mathbf R(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf R^{\mathsf T}]^{-1}
(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r),
$$

which is exactly the numerator of the restriction statistic. Fitting the
reduced model and differencing, or evaluating the quadratic form in the
unrestricted fit, are two routes to one number; for the two type
coefficients of `m2` both return 0.0319. Replacing
$s^2(\mathbf X^{\mathsf T}\mathbf X)^{-1}$ by a robust covariance in the
middle of the expression gives the Wald statistic of Section 5.2. The
algebra survives the substitution; the exact reference distribution does
not.

### The HC3 sandwich {.unnumbered}

Chapter 8, Appendix B, stated the HC3 estimator; this subsection shows
where it comes from. Write the coefficient error as
$\widehat{\boldsymbol\beta}-\boldsymbol\beta
=(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}\boldsymbol\varepsilon$.
Under independent rows with possibly different variances $\sigma_i^2$, its
exact covariance is

$$
(\mathbf X^{\mathsf T}\mathbf X)^{-1}
\Big(\sum_i\sigma_i^2\,\mathbf x_i\mathbf x_i^{\mathsf T}\Big)
(\mathbf X^{\mathsf T}\mathbf X)^{-1},
$$

a *sandwich* with the design component as bread and the error-scale
component as filling. The classical estimator substitutes a common $s^2$,
which collapses the filling to $s^2\mathbf X^{\mathsf T}\mathbf X$. The HC
family substitutes each row's own squared residual instead. Because
$E(e_i^2)=\sigma_i^2(1-h_{ii})$ under the classical model, the raw residual
understates the error at high-leverage rows; HC3 corrects by dividing by
$(1-h_{ii})^2$, which is deliberately slightly conservative and performs
well in small samples [@mackinnon1985; @long2000]. Built directly from
these pieces, the estimator reproduces `sandwich::vcovHC(m2, type = "HC3")`
to about $10^{-12}$.

### The CR1 cluster sandwich {.unnumbered}

Cluster robustness changes the filling, not the bread. Partition the rows
into $G$ clusters and let $\mathbf X_g$ and $\mathbf e_g$ hold cluster
$g$'s rows and residuals. The cluster score
$\mathbf s_g=\mathbf X_g^{\mathsf T}\mathbf e_g$ sums the within-cluster
contributions *before* the outer product is taken, so the estimator of
Section 7.4 retains the within-cluster residual correlations that the
independent-row filling discards. If every cluster held one row, the
filling would reduce to the HC0 form, which is why a row-level robust SE
is the special case that assumes away the dependence.

The asymptotics are in $G$, not in $n$: 3,745 pupils in 79 schools supply
79 approximately independent pieces of information for this covariance,
which is why the clustered SE of 1.8476 is so much larger than the
row-level 1.0377 and why a $t_{G-1}$ reference is used. For the
observed-reading model the finite-cluster factor is $c_G=1.0131$, and
building the estimator from the school scores reproduces
`sandwich::vcovCL(..., type = "HC1", cadjust = TRUE)` to about $10^{-13}$.

### What least squares estimates when the mean is not linear {.unnumbered #app-ch09-projection-target}

A linear regression is often fitted where the conditional mean of the
outcome is not exactly linear in the predictors, as Section 4.1's curved
prestige component shows. Least squares then still estimates a defined
target: the **best linear projection**, the coefficients that minimize
expected squared error over the population the records represent,

$$
\begin{gathered}
(\beta_0^*,\beta_1^*)=\arg\min_{a,b}E\{(Y-a-bX)^2\},\qquad
\beta_1^*=\frac{\operatorname{Cov}(X,Y)}{\operatorname{Var}(X)}, \\
\boldsymbol\beta^*=\{E(\mathbf x\mathbf x^{\mathsf T})\}^{-1}E(\mathbf x Y).
\end{gathered}
$$

With one predictor the target slope is the population covariance over the
predictor's variance, the population version of Chapter 7's
covariance-to-slope identity; with several, it is the population version
of the normal equations. The target is well defined, but it is an
average: the projection slope of income on prestige in the additive model
averages the curve's steeper and flatter stretches, weighted by where the
occupations' prestige values fall. Its value therefore depends on the
distribution of the predictors, and a sample with a different prestige
mix would have a different target even with the same curve. It is a
declared alternative target, not a rescue for a misspecified mean. HC3
standard errors remain appropriate for it under independent sampling,
because the curvature the linear fit misses behaves like one more source
of unequal spread [@fox2016].

## Appendix B: Bootstrap Schemes and Intervals {.unnumbered #app-ch9-b}

Section 6 used case resampling and said why. @tbl-ch09-schemes sets it
beside the other two schemes on the same model, the same three targets,
the same seed, and $B=2000$.

| Scheme | What is resampled | Assumption carried | Education slope | prof $-$ wc | prof coefficient |
|:--|:--|:--|--:|--:|--:|
| Case | Whole records | Records independent and exchangeable | 0.3699 | 1.1187 | 1.9926 |
| Residual | Rescaled residuals, moved across records; design fixed | One error distribution for every record | 0.2820 | 1.1081 | 1.4861 |
| Wild | Each record's own rescaled residual, sign flipped at random; design fixed | Each record keeps its own error scale | 0.3673 | 1.1151 | 1.9767 |

: Three bootstrap schemes for the full occupation model, with each scheme's bootstrap SE for three targets (seed 90909, $B=2000$). {#tbl-ch09-schemes tbl-colwidths="[11,27,24,13,12,13]"}

Both design-fixed schemes rebuild outcomes from the fitted values.
Residual resampling draws leverage-rescaled residuals $e_i/\sqrt{1-h_{ii}}$
with replacement and adds them to the fitted values, which treats the
rescaled residuals as exchangeable across rows. The wild bootstrap keeps
each row's own rescaled residual and multiplies it by $\pm1$ with equal
probability, so each row retains its own error scale.

Read the table by column. For the education slope the residual scheme is
the outlier, 0.2820, close to the classical 0.2752, as expected when
both assume one common error distribution. The wild bootstrap, which
does not, lands at 0.3673, near the case result and HC3's 0.3823. For
the professional coefficient the pattern is stronger: 1.9926, 1.4861,
and 1.9767. When the schemes disagree, the disagreement is about the
error model, and it is a reason to say which scheme was used.

Section 6.3 defined the percentile interval, the middle 95% of the
resampling distribution, and the normal-approximation interval, which
uses that distribution only for its spread and so ignores its skewness.
Two higher-order intervals are worth describing in words.

- The **bootstrap-$t$ interval** resamples a studentized quantity, the
  resampled estimate minus the original estimate, divided by the
  resample's own standard error, and inverts the empirical quantiles of
  that ratio. It needs a standard error inside every resample, and it is
  more accurate when the studentized quantity is closer to pivotal than
  the estimate itself.
- The **BCa interval** adjusts the percentile endpoints for median bias
  and for an acceleration constant that captures how the standard error
  changes with the parameter. It is the usual recommendation when the
  resampling distribution is visibly skewed [@efron1993; @davison1997].

For the three targets of Section 6 the skewness values $-0.3891$, 0.2635,
and 0.3785 are mild, so percentile and normal-approximation endpoints agree
closely. The higher-order intervals would need their own calculation:
mild skewness alone does not establish how much their endpoints change.

## References {.unnumbered}

::: {#refs}
:::
