---
title: "Description and Basic Inference with R"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: true
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \widowpenalty=10000
          \clubpenalty=10000
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{12\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AtBeginDocument{\renewcommand{\boldsymbol}[1]{\symbf{#1}}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                     collapse = TRUE, comment = "#>")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 10, width = 65)
```


::: {.callout-note title="Chapter overview"}
This chapter introduces R and uses it to make the descriptions of
Chapter 2 and the mean inference of Chapter 3 executable. We read and
select observations, describe distributions and groups, and let R
generate repeated samples so that the variation of an estimator becomes
visible. We then see what an interval and a test compute: an estimate,
its standard error, a reference distribution and a critical value.
Relationships receive the same treatment, from a scatterplot and a
covariance to correlation matrices, a selected correlation test and
one-control linear adjustment. Each procedure appears with the one R
call that produces it and a figure or table that shows how to read the
result; the Computer Lab carries the fuller practice. A short Quarto
report connects the final answer to the code that produced it.
:::

Three students scored 50, 60 and 70 points. Their mean is
$(50+60+70)/3=60$ points. R gives the same result by applying a
function to a vector:

```{r}
#| label: ch05-scores
scores <- c(50, 60, 70)
mean(scores)
```

`c()` collects the numbers into a **vector**, an ordered collection of
values. The assignment symbol `<-` gives the vector the name `scores`,
making it an **object** that later commands can use without retyping
the values. `mean(scores)` supplies the vector to the **function**
`mean()`, which returns 60; the `[1]` before the answer only numbers
the first printed position. This answer describes the three recorded
scores directly. Using it to estimate another group's mean would add a
target and an account of how these observations inform that target.

The chapter assumes no previous R course. Familiar browser examples from
Chapters 3--4 provide a starting point; a saved script now lets us run a
connected analysis again. The Computer Lab gives short setup checks and
practice with the same calculations.

## Read and Prepare Observations {#sec-ch5-inspect}

[]{#sec-ch5-start}
[]{#ch05-data}

Programming examples use short, meaningful names such as `group`,
`score` and `id`; when records are changed, `d0` holds the starting data
and `dat` the working copy.

### R, RStudio, a script, and Quarto

**R** performs statistical calculations and draws graphs [@rcore2025].
**RStudio** provides an editor for writing commands and a Console for
running them. Commands saved in a `.R` **script** record how an answer
was obtained: running the script in order recreates every object in a
fresh session, which a screenshot or a remembered sequence of clicks
cannot do. **Quarto** combines prose, mathematics, executable code,
tables, figures and references in a `.qmd` source and renders a report
[@quarto2026]; Section 6 shows what that connection accomplishes.

Four ideas carry the rest of the chapter. An object is a named value,
vector, table or result held in the session. A vector holds values of
one basic type: numbers, text or the logical values `TRUE` and `FALSE`.
A function performs an operation on the inputs it receives, and an
**argument** names one of those inputs or options, as in
`mean(x = scores)`; this computing sense of argument is distinct from
a study's argument in Chapter 1's approach. A **package** collects
functions and data. All statistical calculations and plots here use
**base R**, the stats, graphics, utils and datasets packages supplied
with R. Additional report support, such as knitr and rmarkdown, lets
Quarto execute R and belongs with the reporting example; Chapter 6
adds a small plotting vocabulary after this base-R route. The Computer
Lab's setup check covers launching RStudio, saving a script and
finding Help, with installation in its last reference, and its first
task practises selecting values by position and by condition.

### A data frame is a table of cases and variables

A **data frame** is a table whose columns can hold different types of
values. In Chapter 2's data-matrix notation, $x_{ij}$ is case $i$'s
value on variable $j$; a data frame keeps that arrangement while
allowing text and category columns beside numbers. Its layout should
make the observational unit clear. In the next table a row represents
one student; the columns give an identifier, a baseline score and a
follow-up score, and student B's follow-up score is unavailable. A row
can instead record one person on one occasion or under one condition,
so count the represented units as well as the rows.

```{r}
#| label: ch05-small-records
small <- data.frame(student = c("A", "B", "C"),
  baseline = c(50, 60, 70), followup = c(55, NA, 78))
small
mean(small$followup, na.rm = TRUE)
both <- complete.cases(small)
mean(small$followup[both] - small$baseline[both])
```

The `$` operator selects a named column. `NA` marks an unavailable
value; it is not a zero. An ordinary `mean()` of the follow-up column
would return `NA`, signalling that a value it needs is missing, and the
argument `na.rm = TRUE` asks R to set the missing value aside. The
available follow-up mean is therefore $(55+78)/2=66.5$ points, a
description of two recorded scores. Treating B's missing score as zero
would answer a different question with an invented measurement.

A change needs both measurements of the same student. `complete.cases()`
marks the rows with no missing value, A and C, and the last line
averages their changes, 5 and 8 points, giving a complete-pair mean
change of 6.5 points. Subtracting the two separately available means
happens to give $66.5-60=6.5$ as well, but the baseline mean uses three
students and the follow-up mean two. Their agreement here is a
coincidence of these numbers, not evidence that the two calculations
used the same students; Exercise 3 changes one score and the two
calculations part. Chapter 2's zero-grade example shows why a
measurement definition and codebook must support any decision to
convert zeros to missing values.

### Inspect, select, and derive a variable

The [airquality documentation](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/airquality.html)
describes daily New York measurements from May to September 1973.
A row is a day. `Ozone` is ozone in parts per billion (ppb), `Wind`
wind speed in miles per hour, `Temp` maximum temperature in degrees
Fahrenheit, and `Solar.R` solar radiation in Langleys. `Month` and
`Day` identify dates within that period. Inspect the table before
choosing a calculation:

```{r}
#| label: ch05-air-inspection
air <- airquality
dim(air)
ozone <- air$Ozone[!is.na(air$Ozone)]
length(ozone)
```

`dim()` reports 153 rows and six columns. `is.na()` asks whether each
ozone entry is missing, `!` reverses the answers, and the square
brackets keep only the recorded values: 116 of the 153 days, with 37
missing. The mean of `ozone` will therefore describe 116 days, whereas
the percentage of days with a measurement uses all 153 as its
denominator, $100(116/153)=75.817\%$. Every summary has a denominator,
and naming it is part of the result.

```{r}
#| label: ch05-operations
#| include: false
july <- air[air$Month == 7 & !is.na(air$Ozone), ]
air$temp_C <- (air$Temp - 32) * 5 / 9
c(nrow(july), mean(july$Ozone), mean(air$temp_C))
```

Three operations change a dataset in different ways, and a report
should say which one it performed. @tbl-ch05-operations sets them side
by side on the ozone records.

| Operation | Example on the ozone records | What changes | Result |
|:--|:--|:--|:--|
| Select rows | July days with recorded ozone: `air[air$Month == 7 & !is.na(air$Ozone), ]` | The cases summarized | 26 days, mean 59.1154 ppb |
| Transform a variable | Celsius temperature: `(air$Temp - 32) * 5 / 9` | The units; all 153 days stay | Mean 25.4902 degrees Celsius |
| Recode a variable | Month codes 5 to 9 as a factor with month names | What a value means | Five named groups in calendar order |

: Three operations on the same records. {#tbl-ch05-operations tbl-colwidths="[16,40,26,18]"}

Selecting July answers a question about a different set of days; its
mean, 59.1154 ppb, describes those 26 days only. Converting temperature
keeps every day and changes the units. Recoding changes meaning: the
month codes 5 to 9 are labels, and an average code would not be an
average month. A relationship adds a further rule, because it needs
jointly recorded measurements. Ozone and solar radiation are both
recorded on 111 days, so any calculation using the pair describes those
111. Complete records define a calculation's cases; they do not
establish random sampling.

### Recorded cases and a proposed population

The 153 days form a fixed historical record for a particular city and
period. An observed mean can describe its recorded measurements fully.
A question about the mean ozone of all New York summer days, the 37
days without a measurement, or other years treats these observations
as evidence about a wider population. The arithmetic might begin with
the same mean, but the second use adds assumptions about selection,
time and measurement. Section 3 makes those assumptions visible when
moving from an observed SD to the modelled uncertainty of a mean.

Chapter 2 distinguished the schemes by which units enter a study.
Filtering recorded values in R describes which observations enter a
particular calculation; it supplies no random sampling mechanism. The
invented school system of Chapter 2, 20,000 students in schools of
different types, gives a compact reminder:

\Needspace{14\baselineskip}

| Scheme | Selection from the school system |
|:--|:--|
| Simple random sampling | Every set of 100 students from the register has the same chance. |
| Systematic sampling | Choose a random start, then every 200th student; the ordering of the list matters. |
| Stratified random sampling | Select a probability sample within every school-type stratum. |
| Cluster sampling | Randomly select schools and include their eligible students. |
| Multistage sampling | Randomly select schools, then students within those schools. |

: Recognize the sampling procedure before calculating. {#tbl-ch05-sampling-schemes tbl-colwidths="[30,70]"}

Probability sampling gives eligible units known nonzero inclusion
probabilities, which need not be equal. An open invitation, convenience
recruitment or quotas alone do not supply those probabilities, and
frame coverage and nonresponse still matter after probability
selection. A simple random sample of 100 from the register gives each
student inclusion probability $100/20{,}000=.005$; the Computer Lab
draws such a sample with R. Sampling strata concern selection, whereas
Chapter 3's school blocks concern the allocation of conditions; cluster
sampling and the assignment of whole clusters are different design
decisions, even when both produce records grouped within schools.
General unequal-probability estimation and the SEs for all survey
designs extend beyond the methods developed here.

## Describe Distributions and Groups {#sec-ch5-summaries}

[]{#ch05-descriptions}

### Shape, centre, and spread

A distribution shows how the values are arranged. Before reducing the
116 ozone measurements to numbers, look at their shape. A histogram
groups nearby values into intervals and counts the observations in each
interval; a boxplot compresses their location and spread.

```{r}
#| label: fig-ch05-ozone
#| echo: !expr c(2, 3)
#| fig-cap: "Recorded ozone values on 116 days, displayed as a histogram and a boxplot."
#| fig-alt: "The histogram has a long right tail reaching 168 ppb. The horizontal boxplot places the median near the lower values and marks several high observations beyond its upper whisker."
#| fig-width: 6.2
#| fig-height: 3.4
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 2, .5))
hist(ozone, breaks = seq(0, 180, by = 20), xlab = "Ozone (ppb)", main = "")
boxplot(ozone, horizontal = TRUE, xlab = "Ozone (ppb)")
par(oldpar)
```

The `breaks` argument sets the histogram's interval boundaries, and
`main = ""` leaves out the default title. Other boundaries can reveal
or conceal detail while the observations stay the same. Most days have lower values, with a long right tail reaching 168
ppb. The boxplot marks the median and a box between its hinges; each
whisker reaches the most distant value within 1.5 box lengths of its
hinge. Points beyond the whiskers invite examination, and the display
alone does not justify removing them.

Chapter 2 used the mean and median to describe location, and the
variance, SD and IQR to describe spread. For observations
$x_1,\ldots,x_n$,

$$
\bar x=\frac{1}{n}\sum_{i=1}^{n}x_i,\qquad
s^2=\frac{\sum_{i=1}^{n}(x_i-\bar x)^2}{n-1},\qquad
s=\sqrt{s^2},\qquad IQR=Q_3-Q_1.
$$

The variance measures each value's deviation from the mean, squares the
deviations so that positive and negative ones cannot cancel, adds them
and divides by $n-1$; the square root returns the SD to the original
units. For the three scores the deviations from 60 are $-10$, 0 and
10, their squares sum to 200, and $s^2=200/2=100$ squared points, so
$s=10$ points. The divisor $n-1$ reflects that deviations around a
sample mean sum to zero, so once $n-1$ of them are known the last is
determined; under independent sampling from a common distribution it
also makes $s^2$ unbiased for the process variance, a property that
Section 3.3 shows by repetition. For ozone, one call returns the
summaries:

```{r}
#| label: ch05-ozone-summaries
c(mean = mean(ozone), median = median(ozone), SD = sd(ozone),
  IQR = IQR(ozone))
```

The mean, 42.1293 ppb, exceeds the median, 31.5 ppb, as the long right
tail leads us to expect. The SD, 32.9879 ppb, describes spread around
the mean; the quartiles 18.00 and 63.25 ppb give an IQR of 45.25 ppb for
the middle half of the ordered values; and the range is 1 to 168 ppb.
Each summary keeps a different feature, which @tbl-ch05-score-sets
shows with four invented sets of three scores.

| Scores | Mean | Median | SD |
|:--|--:|--:|--:|
| 50, 60, 70 | 60 | 60 | 10.0000 |
| 50, 60, 100 | 70 | 60 | 26.4575 |
| 58, 60, 62 | 60 | 60 | 2.0000 |
| 40, 60, 80 | 60 | 60 | 20.0000 |

: What the mean, the median and the SD each keep. {#tbl-ch05-score-sets tbl-colwidths="[34,22,22,22]"}

Read the rows in pairs. Replacing 70 by 100 moves the mean from 60 to
70 and raises the SD from 10.0000 to 26.4575, while the median stays at 60,
still the middle ordered value: the median resists one extreme value,
which does not make either centre universally preferable. The last two
sets balance at the same centre and differ tenfold in spread, so a
report of equal means would conceal what distinguishes them. This is
why a group comparison keeps counts, centres and spreads together.

Quartiles also depend on a stated rule. R's default, `type = 7`, places
the quartile at position $h=1+(n-1)p$ among the ordered values and
interpolates between neighbours; Chapter 2 took the medians of the two
halves. @tbl-ch05-quartile-rules applies both rules to the values 1, 3,
3 and 5.

| Rule | $Q_1$ | $Q_3$ | IQR |
|:--|--:|--:|--:|
| Chapter 2: medians of the two halves | 2.0 | 4.0 | 2 |
| R's default, `type = 7` | 2.5 | 3.5 | 1 |

: Two quartile rules applied to the same four values. {#tbl-ch05-quartile-rules tbl-colwidths="[52,16,16,16]"}

The observations are the same; the rules differ. Type 7 uses positions
1.75 and 3.25, so each quartile moves part of the way between
neighbouring values, and its IQR is half the Chapter 2 value. A
reproducible report names the rule when it affects the answer, and a
boxplot's hinges need not equal either rule's quartiles.

### Group summaries and weighted means

For the ozone records, the month codes identify groups of days. A
**factor** stores such categories with named levels in a stated order,
and `tapply()` applies a function within each level:

```{r}
#| label: ch05-monthly
air_observed <- air[!is.na(air$Ozone), ]
air_observed$month <- factor(air_observed$Month, levels = 5:9,
  labels = c("May", "June", "July", "August", "September"))
tapply(air_observed$Ozone, air_observed$month, mean)
```

`5:9` generates the integers 5 to 9; `levels` states their order and
`labels` their names. A group mean needs its count and spread beside
it, which @tbl-ch05-monthly supplies.

| Month | Recorded days | Mean ozone | SD ozone |
|:--|--:|--:|--:|
| May | 26 | 23.6154 | 22.2244 |
| June | 9 | 29.4444 | 18.2079 |
| July | 26 | 59.1154 | 31.6358 |
| August | 26 | 59.9615 | 39.6812 |
| September | 29 | 31.4483 | 24.1418 |

: Recorded ozone by month; means and SDs are in ppb. {#tbl-ch05-monthly tbl-colwidths="[28,24,24,24]"}

June's mean describes only nine recorded days, and its count matters
when we place it among the other monthly means.

```{r}
#| label: fig-ch05-months
#| echo: false
#| fig-cap: "Recorded ozone distributions by month; each box uses that month's available measurements."
#| fig-alt: "July and August have higher median ozone than May, June, and September, with different spreads and only nine observed days in June."
#| fig-width: 6
#| fig-height: 3.4
boxplot(Ozone ~ month, data = air_observed,
        xlab = "Month", ylab = "Ozone (ppb)", col = "grey90")
```

The boxes show the higher centres in July and August together with the
spread within each month, which the monthly means alone cannot show.
These differences summarize the record; they do not by themselves
isolate an effect of season from other changes over the period.

The **grand mean** over individual observations weights each group mean
by its count:

$$
\bar x=\frac{\sum_g n_g\bar x_g}{\sum_g n_g}.
$$

Chapter 2's two invented training groups make the weighting visible.
Group A's completion times are 1 and 3 minutes and Group B's 3, 3, 5
and 9, with means 2 and 5 minutes. The six learners' mean is 4 minutes,
because Group B supplies four of the six values; the unweighted mean of
the two group means, 3.5 minutes, describes an average group instead.
The same rule applied to the five monthly ozone means returns 42.1293
ppb, the mean of all 116 recorded days. For June and July alone, with 9
and 26 recorded days, the recorded-day mean is 51.4857 ppb and the
equal-month mean 44.2799 ppb. Neither is the correct one in general:
the question decides whether days or months receive equal weight.

### Counts, proportions, and a binary indicator

The [mtcars documentation](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html)
describes 32 car-model records from the 1974 *Motor Trend* magazine,
covering 1973--1974 models. Each row is a car model. The variable `mpg`
gives fuel economy in miles per US gallon, while `am` codes automatic
transmission as 0 and manual transmission as 1. Use 20 mpg as a
teaching threshold, label the transmission categories, and form their
two-way table:

```{r}
#| label: ch05-transmission-table
cars <- mtcars
cars$transmission <- factor(cars$am, levels = c(0, 1),
  labels = c("Automatic", "Manual"))
tab <- table(cars$transmission, cars$mpg >= 20)
tab
100 * prop.table(tab, margin = 1)
```

The table counts 15 automatic models below the threshold and 4 at or
above it, against 3 and 10 manual models. `prop.table()` turns counts
into proportions; `margin = 1` makes each row its own denominator, and
multiplying by 100 gives percentages. One table answers several
questions, and each question has its own denominator
(@tbl-ch05-denominators).

\Needspace{9\baselineskip}

| Question | Denominator | Percentage |
|:--|:--|--:|
| Automatic models meeting the threshold | 19 automatic models | 21.0526 |
| Manual models meeting the threshold | 13 manual models | 76.9231 |
| Threshold-meeting models that are manual | 14 threshold-meeting models | 71.4286 |

: Three percentages from one table, each with its denominator. {#tbl-ch05-denominators tbl-colwidths="[46,34,20]"}

The first two rows condition on transmission and differ by 55.8704
percentage points. The third reverses the condition: among the 14
models that meet the threshold, 71.4286% are manual. "Meeting the
threshold among manual models" and "manual among threshold-meeting
models" answer different questions, and naming the denominator keeps
them apart. The percentages describe an association among the 32
recorded models, whose transmission groups also differ in weight and
engine; Chapter 6 applies inference to counts, and neither direction
of conditioning identifies what changing a transmission would do.

A binary indicator has one further property: its mean is the proportion
coded 1. Of the 32 models, 14 meet the threshold, so the indicator
`cars$mpg >= 20` has mean $14/32=.4375$. The 0/1 `am` variable is such
an indicator, and Section 5.3 uses it in a point-biserial correlation.

### Transformations and the Normal model

A linear transformation changes location and units predictably. If
$y_i=a+bx_i$, the mean becomes $a+b\bar x$, the variance $b^2s^2$ and
the SD $|b|s$. The Celsius conversion has $a=-32(5/9)$ and $b=5/9$, so
the mean temperature of 77.8824 and SD of 9.4653 degrees Fahrenheit
become 25.4902 and 5.2585 degrees Celsius: subtracting the constant
moves the mean only, and the factor $5/9$ rescales both.

A standardized value measures distance from the mean in SD units. For
an ozone reading of 100 ppb,

$$
z_i=\frac{x_i-\bar x}{s},\qquad
z_{100\text{ ppb}}=\frac{100-42.1293}{32.9879}=1.7543.
$$

Standardizing changes location and scale while preserving shape:
right-skewed ozone stays right-skewed. This individual $z$ divides by
the observations' SD; Section 4's statistic for a mean divides by an
SE instead. R's Normal functions give model probabilities: `pnorm()`
returns a cumulative probability, and `lower.tail = FALSE` asks for the
probability above the value.

```{r}
#| label: ch05-normal-model
pnorm(100, mean = mean(ozone), sd = sd(ozone), lower.tail = FALSE)
```

A Normal model with the ozone mean and SD puts probability .0397 above
100 ppb. @tbl-ch05-normal-check sets this and one more model
probability beside what the 116 recorded days show.

\Needspace{8\baselineskip}

| Event for one recorded day | Normal model | Observed |
|:--|--:|--:|
| Ozone of at least 100 ppb | .0397 | .0603 |
| Ozone within one SD of the mean | .6827 | .7241 |

: A Normal model with the ozone mean and SD, against the 116 recorded values. {#tbl-ch05-normal-check tbl-colwidths="[52,24,24]"}

Seven of the 116 days reach 100 ppb, a larger share than the model
gives, and more days lie within one SD of the mean than the model
expects. The model misses the long right tail and the concentration of
low values. A function evaluates the model it is given; it does not
certify that the model describes the data, and the familiar
68--95--99.7 rule belongs to the Normal model, not to every
standardized dataset.

For a separate, stipulated score model $X\sim N(100,15^2)$, a
probability for a range is a difference of cumulative probabilities,
and `qnorm()` reverses the question by returning the value at a stated
cumulative probability:

```{r}
#| label: ch05-score-model
pnorm(120, mean = 100, sd = 15) - pnorm(100, mean = 100, sd = 15)
qnorm(c(.025, .975), mean = 100, sd = 15)
```

The probability of a score between 100 and 120 is .4088. The central
95% of scores under the model lies between 70.6005 and 129.3995; these
limits describe individual scores under the model, not a confidence
interval for an unknown mean. The same functions give .5000 below 100 and
.0038 above 140.

```{r}
#| label: fig-ch05-normal-area
#| echo: false
#| fig-cap: "Probability is an area under a stipulated Normal density."
#| fig-alt: "The score range from 100 to 120 is shaded under a Normal curve with mean 100 and SD 15. Its area is .4088."
#| fig-width: 6.2
#| fig-height: 3.3
curve(dnorm(x, mean = 100, sd = 15), 40, 160,
      xlab = "Score", ylab = "Density")
shade_x <- seq(100, 120, length.out = 200)
polygon(c(100, shade_x, 120),
        c(0, dnorm(shade_x, 100, 15), 0), col = "grey80")
```

The curve's height is a density. Probability belongs to an area, such
as the shaded region in @fig-ch05-normal-area, rather than to the height
at one exact score.

### A compact distribution catalogue in R {#ch05-distributions}

Chapter 2 connected variable forms to probability models and drew them
(Chapter 2, Section 2.5). R names the functions for each family by one
pattern: a `d` function gives a mass at a discrete value or a density
height, `p` the cumulative probability $P(X\le x)$, `q` a cumulative
quantile, and `r` observations generated from the model. The
parameters identify which member of a family is meant.

\Needspace{22\baselineskip}

| Model | Object and parameters | Base-R family |
|:--|:--|:--|
| Bernoulli | One 0/1 response, probability $p$ of 1 | `binom`, with `size=1` |
| Binomial | Success count in $m$ independent trials with common $p$ | `dbinom`, `pbinom`, `qbinom`, `rbinom` |
| Categorical | One label among categories with stated probabilities | `sample()` with `prob`; a table of masses |
| Poisson | Nonnegative count with model mean $\lambda$ | `dpois`, `ppois`, `qpois`, `rpois` |
| Normal | Continuous value with mean $\mu$ and SD $\sigma$ | `dnorm`, `pnorm`, `qnorm`, `rnorm` |
| Lognormal | Positive value whose log is Normal; parameters on the log scale | `dlnorm`, `plnorm`, `qlnorm`, `rlnorm` |
| Logistic | Continuous symmetric value with location $\mu_L$ and positive scale $s_L$ | `dlogis`, `plogis`, `qlogis`, `rlogis` |
| Student t | Symmetric reference distribution with specified df | `dt`, `pt`, `qt`, `rt` |

: Distribution functions respect the model's support and parameters. {#tbl-ch05-models tbl-colwidths="[16,49,35]"}

For example, $X\sim\operatorname{Binomial}(20,.3)$ counts successes in
20 independent trials with success probability .3. The four prefixes
ask four questions of this one model:

```{r}
#| label: ch05-binomial-functions
c(d = dbinom(4, size = 20, prob = .3),
  p = pbinom(4, size = 20, prob = .3),
  q = qbinom(.95, size = 20, prob = .3))
set.seed(20517)
rbinom(6, size = 20, prob = .3)
```

```{r}
#| label: fig-ch05-dpqr
#| echo: false
#| fig-cap: "Four questions asked of one binomial model, 20 trials with success probability .3. (a) The mass at 4 (black bar) is what `dbinom` returns; the masses at 0 to 4 together (dark bars) are what `pbinom` returns; the triangles below the axis are the six counts from `rbinom`. (b) The cumulative probability first reaches .95 at 9, the value `qbinom` returns."
#| fig-alt: "Left panel: bars of binomial probability for 0 to 20 successes, highest near 6, with the bar at 4 in black, the bars at 0 to 3 dark grey, and six triangles below the axis at 3, 5, 6, 6, 6 and 8. Right panel: the cumulative probability as a step function, crossing the dashed .95 line at 9."
#| fig-width: 6.5
#| fig-height: 3.1
k <- 0:20
mass <- dbinom(k, size = 20, prob = .3)
cumulative <- pbinom(k, size = 20, prob = .3)
set.seed(20517)
drawn <- table(rbinom(6, size = 20, prob = .3))
oldpar <- par(mfrow = c(1, 2), mar = c(4.2, 4, 2, .5), mgp = c(2.3, .7, 0))
plot(k, mass, type = "h", lwd = 7, lend = 1,
     col = ifelse(k == 4, "black", ifelse(k < 4, "grey45", "grey82")),
     ylim = c(-.03, .25), yaxt = "n", xlab = "Successes in 20 trials",
     ylab = "Probability mass", main = "(a) Mass: d, p and r")
axis(2, at = seq(0, .2, by = .05))
text(10, .13, "d = .1304\n(black bar)", pos = 4, cex = .8)
text(-.6, .235, "p = .2375 (dark bars 0 to 4)", pos = 4, cex = .8)
for (v in names(drawn)) points(rep(as.numeric(v), drawn[v]),
  -.01 - .008 * (seq_len(drawn[v]) - 1), pch = 17, cex = .7)
text(9, -.018, "r: six draws", pos = 4, cex = .8)
plot(k, cumulative, type = "s", ylim = c(0, 1),
     xlab = "Successes in 20 trials", ylab = "Cumulative probability",
     main = "(b) Cumulative: q")
points(k, cumulative, pch = 16, cex = .6)
abline(h = .95, lty = 2)
segments(9, 0, 9, cumulative[10], lty = 3)
text(9.3, .5, "q = 9: first count\nreaching .95", pos = 4, cex = .8)
par(oldpar)
```

@fig-ch05-dpqr reads the output. The probability of exactly four
successes is .1304, one bar; the probability of at most four is .2375,
the sum of five bars. The .95 quantile is 9 because the cumulative
probability is .8867 at 8 and .9520 at 9: for a discrete model the
quantile is the first count whose cumulative probability reaches the
requested level, which it need not attain exactly. The six generated
values, 6, 8, 6, 3, 5 and 6, are counts from six separate sets of 20
trials; the first argument of `rbinom()` is the number of counts
generated, whereas `size = 20` is the number of trials in each. The
seed makes the draws reproducible; another seed changes the draws and
leaves the model probabilities unchanged.

The same vocabulary works across the other families, with care over
what their parameters mean. Lognormal `meanlog` and `sdlog` describe the
logarithm, not the mean and SD of the positive values. The logistic
distribution is continuous and differs from both lognormal and binary
outcomes; its cumulative function supplies a probability curve in
Chapter 11. Categorical probabilities refer to labels, and a cumulative
probability requires an explicit ordering of those labels. The Computer
Lab runs these families and draws the catalogue.

## Sampling Variation and Standard Errors {#sec-ch5-sampling}

[]{#ch05-sampling}

### A target, an estimator, and a realized estimate

Chapter 2's invented school system contained 20,000 students, and
repeated simple random samples of 100 gave different means while the
population mean stayed fixed. We now change the generating process:
instead of selecting from a fixed population, R generates independent
observations from a probability distribution whose parameters we
choose. Let the model be Normal with mean 50 and SD 10, and generate
30 draws:

```{r}
#| label: ch05-one-sample
set.seed(60506)
one_sample <- rnorm(30, mean = 50, sd = 10)
c(mean = mean(one_sample), SD = sd(one_sample))
```

`rnorm()` takes the number of observations, the model mean and the
model SD; the result is one sample from that model.

\Needspace{8\baselineskip}

| Quantity | Model value (parameter) | This sample (statistic) |
|:--|--:|--:|
| Mean | $\mu=50$ | $\bar x=49.5691$ |
| SD | $\sigma=10$ | $s=7.0051$ |

: One sample of 30 draws from a declared Normal model (seed 60506). {#tbl-ch05-parameter-statistic tbl-colwidths="[24,38,38]"}

The **parameters** are fixed features of the model; the **statistics**
are calculated from the generated observations and vary with the
draws. Running `rnorm()` again without resetting the seed gives another
sample and generally different statistics, while resetting the seed
recreates the same draws. A seed controls the repeatability of the
numerical experiment; it supplies no evidence that the Normal model
describes a real study. Nor does knowing the generating SD in a
simulation make a sample SD a known population SD in an actual study.

A simulation gives us a known target, but a study usually has a target
we cannot observe directly. Recall Chapter 3's invented
special-practice study. It assigned 120 participants, 60 to Special
Practice and 60 to Usual Practice, and measured delayed-assessment
scores; the group means were 27.1 and 24.0 points, with SDs 6.0 and
6.4. The **estimand** is the mean effect of assignment to Special
rather than Usual Practice among these 120 participants. Write $Y_{i1}$
for participant $i$'s potential score under Special assignment and
$Y_{i0}$ for the score under Usual assignment. The target is

$$
\Delta\mu=\frac{1}{n}\sum_{i=1}^{n}\{Y_{i1}-Y_{i0}\},
$$

where the braces hold one participant's effect and division by $n=120$
takes the mean over the 120 assigned participants. Each participant
reveals only the outcome under the assigned condition, so no individual
effect is ever recorded and the effects cannot simply be averaged. The
**estimator**, the difference in assigned-group means
$\widehat{\Delta\mu}=\bar y_1-\bar y_0$, uses the observed outcomes;
**estimation** applies that rule and checks its inputs; the
**estimate** is the resulting $27.1-24.0=3.1$ points. The repetition
behind its uncertainty reassigns the same 120 participants, exactly 60
to each condition: the participants and their potential outcomes stay
fixed while the assigned groups and observed outcomes vary. That
differs from repeatedly sampling new people.

| Object | Working definition | In the special-practice study | In R |
|:--|:--|:--|:--|
| Estimand | The precisely specified quantity the question seeks | Mean delayed-score difference caused by Special rather than Usual assignment among the 120 participants | Not an object; stated in words before any code |
| Estimator | The rule applied in each admissible repetition | Special-group mean minus Usual-group mean | An expression such as `mean_special - mean_usual` |
| Estimate | The number the estimator produced | $27.1-24.0=3.1$ points | The value that expression returns |
| Statistic | A numerical function of observed data | A group mean, the 3.1-point estimate | Any computed summary |
| Parameter | A fixed feature named by the estimand under the chosen interpretation | The participant-specific mean assignment contrast | Unknown here; Section 3.3 changes the target to a known waiting-time variance |
| Standard error | The modelled spread of the estimator across the declared repetitions | 1.1325 points (Section 3.4) | Computed from the group SDs and counts |

: Inferential objects with their R counterparts. {#tbl-ch05-inferential-objects tbl-colwidths="[13,27,34,26]"}

@tbl-ch05-inferential-objects separates the scientific target from its
computational representation. Storing a result in an object named
`estimate` does not define its estimand; that definition comes from the
question, the participants, the comparison and the outcome.

### The first repeated-sample calculation

Return to Chapter 2's waiting-time model: independent exponential waits
with mean and SD both 10 minutes, many short and a few long. R's
`rexp()` uses the **rate**, the reciprocal of the mean, so
`rate = 1 / 10` gives a ten-minute mean. One call repeats the whole
sampling operation, drawing five waits and averaging them, 5,000 times:

```{r}
#| label: ch05-replicate
set.seed(60203)
means_5 <- replicate(5000, mean(rexp(5, rate = 1 / 10)))
c(mean = mean(means_5), SE = sd(means_5))
```

`replicate()` evaluates its second argument repeatedly and collects the
results, so `means_5` holds 5,000 sample means, not 5,000 individual
waits. Their mean, 10.0141 minutes, lies close to the model mean, and
their SD, 4.5395 minutes, describes how much a mean of five waits
varies. These are the values of Chapter 2's five-wait row, and Chapter
2's sampling-means figure (Chapter 2, Section 4.2) shows how the
distribution narrows at larger sample sizes; the Computer Lab
reproduces that figure. A statistic's **standard error** is its SD
across the stated repetitions. For independent observations from a
common distribution with SD $\sigma$, the mean's model SE is

$$
SE(\bar Y)=\frac{\sigma}{\sqrt n},
$$

and the SD of simulated means estimates it; we call that SD the
*empirical SE*. A simulation has two sizes, and they do different
jobs: the study size $n$, the number of waits in each sample, and the
number of repetitions $B$, the number of simulated samples.
@fig-ch05-repetitions changes one at a time.

```{r}
#| label: ch05-repetitions
#| include: false
means_30 <- replicate(5000, mean(rexp(30, rate = 1 / 10)))
set.seed(60203)
means_5_again <- replicate(5000, mean(rexp(5, rate = 1 / 10)))
means_120 <- replicate(5000, mean(rexp(120, rate = 1 / 10)))
```

```{r}
#| label: fig-ch05-repetitions
#| echo: false
#| fig-cap: "More repetitions or larger studies. Each histogram collects simulated sample means from the waiting-time model; the dashed line marks the model mean, 10 minutes. (a) The first 200 of the 5,000 means of 30 waits; (b) all 5,000; (c) 5,000 means of 120 waits."
#| fig-alt: "Three histograms on common axes from 4 to 18 minutes. The first, from 200 means, is ragged; the second, from 5,000 means of the same study size, is smooth with the same spread; the third, from 5,000 means of 120 waits, is about half as wide."
#| fig-width: 6.5
#| fig-height: 2.9
panels <- list(means_30[1:200], means_30, means_120)
titles <- c("(a) n = 30, B = 200", "(b) n = 30, B = 5,000",
            "(c) n = 120, B = 5,000")
oldpar <- par(mfrow = c(1, 3), mar = c(4, 3.6, 2, .4), mgp = c(2.2, .7, 0))
for (j in 1:3) {
  hist(panels[[j]], breaks = seq(4, 18, by = .5), probability = TRUE,
       ylim = c(0, .5), col = "grey85", main = titles[j],
       xlab = "Sample mean (minutes)", ylab = "Density")
  abline(v = 10, lty = 2)
}
par(oldpar)
```

\Needspace{9\baselineskip}

| Panel | Study size $n$ | Repetitions $B$ | Mean of sample means | Empirical SE | Model SE |
|:--|--:|--:|--:|--:|--:|
| (a) | 30 | 200 | 9.6698 | 1.7484 | 1.8257 |
| (b) | 30 | 5,000 | 9.9626 | 1.7878 | 1.8257 |
| (c) | 120 | 5,000 | 9.9922 | 0.9214 | 0.9129 |

: Repetitions and study size change different things (seed 60203). {#tbl-ch05-repetitions tbl-colwidths="[10,16,16,20,19,19]"}

From panel (a) to panel (b) only the number of repetitions changes. The
histogram becomes smoother and its empirical SE settles near the model
SE $10/\sqrt{30}=1.8257$ minutes, which has not changed: more
repetitions show the same sampling distribution more steadily. From
panel (b) to panel (c) the study size quadruples, so the model SE halves
to $10/\sqrt{120}=0.9129$ minutes and the histogram narrows. The
density scale makes each bar's area the fraction of generated means in
its interval, so the three panels are directly comparable. With
finitely many repetitions, an empirical SE need not equal its model
value exactly.

Keep three quantities distinct. The 10-minute SD describes individual
waits. The SE describes how sample means vary at a stated study size.
A particular sample's **error**, such as $11.3-10=1.3$ minutes for an
illustrative mean of 11.3, is one realized departure from the target;
an SE summarizes the distribution of possible estimates and does not
reveal the error of a particular one. Halving an SE under this model
takes four times as many observations.

### Bias and precision

An estimator's **expectation** is its probability-weighted average
across the declared repetitions, and its **bias** is that expectation
minus the target. Chapter 3 illustrated bias with an estimator
averaging 3.4 when the target was 3, a bias of 0.4. **Precision**
concerns how tightly the possible values cluster: an estimator can be
centred correctly and still vary widely from one sample to the next.

To compare the two variance divisors, change the target to the variance
of the waiting-time process, 100 squared minutes. For each of 5,000
independent samples of five waits, compute one variance with divisor
$n-1$ and one with divisor $n$; the Computer Lab carries the
simulation, and @tbl-ch05-bias-evidence reports its averages.

\Needspace{8\baselineskip}

| Variance rule | Mean over 5,000 samples | Expectation under the model |
|:--|--:|--:|
| Divide by $n-1$ | 101.2241 | 100 |
| Divide by $n$ | 80.9793 | 80 |

: Variance estimates from independent exponential samples of five waits (seed 60501). {#tbl-ch05-bias-evidence tbl-colwidths="[34,33,33]"}

The expectations in the last column follow from

$$
\operatorname{Bias}(\widehat\theta)=E(\widehat\theta)-\theta,\qquad
E(s^2)=\sigma^2,\qquad
E\left\{\frac{1}{n}\sum_{i=1}^{n}(X_i-\bar X)^2\right\}=\frac{n-1}{n}\sigma^2.
$$

The $n-1$ rule is unbiased under the independent common-distribution
model; the $n$ rule has downward bias $\sigma^2/n=20$ squared minutes
here. The simulation average 101.2241 can differ from 100 even for the
unbiased rule, because 5,000 samples approximate the expectation rather
than enumerating every possible sample. Both rules can be far from 100
in a single sample. The $n$ estimate is always four-fifths of the
$n-1$ estimate here, so it varies less across repetitions while being
centred on the wrong value: an apparently stable estimate can
systematically miss its target, which is why precision and bias are
reported separately.

### SE follows the unit and the comparison {#sec-ch5-reshape}

An SE follows the observational units that can vary independently. The
[sleep documentation](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/sleep.html)
describes extra hours of sleep, relative to a control period, under two
drugs for ten people. `ID` identifies a person, `group` a drug and
`extra` the extra hours; negative values mean less sleep than under
control. Twenty measurement rows therefore represent ten people, and
alignment by person comes before subtraction:

```{r}
#| label: ch05-sleep-differences
sleep_wide <- reshape(sleep, idvar = "ID", timevar = "group",
  direction = "wide")
sleep_d <- sleep_wide$extra.2 - sleep_wide$extra.1
c(mean = mean(sleep_d), SD = sd(sleep_d),
  SE = sd(sleep_d) / sqrt(length(sleep_d)))
```

`reshape()` uses the ID to place each person's two measurements on one
row, as `extra.1` and `extra.2`; before reshaping an unfamiliar file,
check that each person--condition key occurs once, the Computer Lab's
first step in Round 2. The subtraction is **drug 2 minus drug 1**. The
ten differences have mean 1.58 hours and SD 1.23 hours, and because the
same people enter both conditions, the mean difference equals the
difference of the condition means, $2.33-0.75=1.58$. Their variance
keeps the pairing:

$$
s_d^2=s_2^2+s_1^2-2s_{12},
$$

where $s_{12}$ is the covariance of the two measurements.
@tbl-ch05-paired-se sets the two possible SE calculations for the same
ten people side by side.

\Needspace{8\baselineskip}

| Calculation for the ten people | Variance used | SE of the mean difference (hours) |
|:--|:--|--:|
| Paired: the ten within-person differences | $s_d^2=1.5129$ | 0.3890 |
| As if the two conditions were independent groups | $s_1^2+s_2^2=3.2006+4.0090$ | 0.8491 |

: Two SE calculations for the same ten people. {#tbl-ch05-paired-se tbl-colwidths="[44,32,24]"}

The covariance of the two measurements is positive, 2.8483: people who
gain more sleep under one drug tend to gain more under the other. The
paired calculation removes that shared variation; the independent-groups
calculation leaves it in and more than doubles the SE. The number of
people is ten in both rows, so the variance calculation, not the count,
is what changed.

For a single mean of independent observations from a common process,
the sample SD estimates the process SD and the estimated SE is
$\widehat{SE}(\bar Y)=s/\sqrt n$. For the 116 ozone values it is
$32.9879/\sqrt{116}=3.0628$ ppb, **under a working assumption of
independent observations from a common process**. The historical daily
record does not establish that assumption: nearby days could be
related and the process could change over the summer. The formula
returns a number; its inferential use needs a justification for the
model and the target.

For two independent groups, each mean contributes its own variance
term. The special-practice summaries give

$$
\widehat{SE}(\bar y_1-\bar y_0)=\sqrt{\frac{6.0^2}{60}+\frac{6.4^2}{60}}
=1.1325\text{ points},
$$

which, with 60 participants in each group, equals the pooled SE that
Chapter 3 uses (Chapter 3, Section 2.2). Its interpretation in this
fixed-participant assignment study differs from independent sampling.
Exactly 60 of the 120 participants are assigned to each group, so the
two assigned-group means are dependent, and the exact assignment
variance subtracts a term for variation in individual assignment
effects, which the observed outcomes cannot recover. The squared SE
above omits that term and estimates an upper bound on the assignment
variance *on average*; it need not exceed the true variance in every
realized study. Chapter 3 develops this distinction and the approximate
t calibrations that Section 4 reproduces.

Clustered assignment needs the same attention to independent units.
Chapter 3, Section 2.3,
develops the school-blocked feedback example: 480 students in 24
classes within 12 schools, with one class per format in each school.
Each of its twelve explanatory-minus-usual class-mean contrasts,
$-2$, 0, 1, 1, 2, 2, 3, 3, 4, 4, 5 and 7, is one independent unit; they
have mean 2.5 points, SD 2.3932 points and SE 0.6908 points. Counting
480 student identifiers would not establish 480 independent
assignments.

Sampling without replacement also changes uncertainty. If $n$ members
are selected by simple random sampling from a finite population of
size $N$, whose variance $\sigma^2$ uses denominator $N$, then

$$
SE_{\text{finite}}=\frac{\sigma}{\sqrt n}\sqrt{\frac{N-n}{N-1}},
$$

estimated by $(s/\sqrt n)\sqrt{1-n/N}$. Observing a larger fraction of
a fixed population leaves less uncertainty about its mean, and
observing everyone leaves none. Chapter 3, Section 2.3, introduces the
correction, which belongs to simple random sampling without replacement
rather than to every way of selecting from a finite population. Across
these settings, the estimand, dependence, weighting and design decide
the uncertainty calculation; row counts alone cannot.

### Concentration and sampling shape

Two results explain the changing distributions of sample means. The
**law of large numbers** concerns concentration: under its conditions,
larger samples make the mean more likely to lie close to the population
mean. The **central limit theorem** concerns shape: under conditions
such as independent, identically distributed observations with finite
variance, the centred and scaled sample mean approaches a Normal
distribution. A running mean shows the first idea along one sequence
of draws: after draw $k$, average the first $k$ waits.

```{r}
#| label: fig-ch05-running-mean
#| echo: false
#| fig-cap: "One running mean from independent exponential waits with model mean ten minutes."
#| fig-alt: "The running mean fluctuates irregularly around ten; it is 9.7207 after 400 observations. Adding a draw can move it closer to or farther from the target."
#| fig-width: 6.2
#| fig-height: 3.3
set.seed(60204)
running_waits <- rexp(400, rate = 1 / 10)
running_mean <- cumsum(running_waits) / seq_along(running_waits)
plot(running_mean, type = "l", xlab = "Number of observations",
     ylab = "Running mean (minutes)", ylim = range(c(running_mean, 10)))
abline(h = 10, lty = 2)
```

The running means are 9.0830 minutes after 10 draws, 7.9852 after 50,
8.8373 after 100 and 9.7207 after 400. The path is irregular: adding
observations can move the mean farther from 10 for a while.
Concentration is a probabilistic statement as the sample grows, not a
promise of steady improvement along one sequence or a guarantee at a
particular size.

The histograms of @fig-ch05-repetitions and of Chapter 2's
sampling-means figure change shape as well as width: means of larger
samples become more nearly symmetric, although individual waits keep
their exponential distribution. Concentration describes closeness to
the target; the central limit theorem describes the sampling shape
after centring and scaling. Neither makes individual observations
Normal. Strong dependence, extreme tails or a dominating observation
can make a familiar approximation poor, and there is no universal
sample-size cutoff. Chapter 3, Section 2.4, states both results with
their conditions; the Computer Lab applies the reasoning to a second
skewed process.

## Intervals and Tests for Means {#ch05-mean-inference}

### Estimate, SE, critical value, and interval {#ch05-z-t}

An estimate answers a numerical question; an interval describes
uncertainty about its target under a stated procedure. Before asking R
for either, name the outcome and its units, the included observations,
the parameter and the repetition that makes an SE meaningful: the model
and design determine the calculation, and a function's default cannot
supply them. The common construction from Chapter 3 is

$$
\text{interval}=\widehat\theta\pm c^*
\widehat{\operatorname{SE}}(\widehat\theta),\qquad
T=\frac{\widehat\theta-\theta_0}
{\widehat{\operatorname{SE}}(\widehat\theta)}.
$$

The estimate and SE have outcome units. Multiplying the SE by the
**critical value** $c^*$ gives a margin in those units, and the test
statistic expresses the departure from a null value $\theta_0$ in SE
units. The confidence level, the tail convention and the reference
distribution determine the critical value and the p-value.

For a mean of independent Normal observations whose SD $\sigma$ is
stipulated independently, the SE is $\sigma/\sqrt n$ and the statistic
has a standard Normal reference. When the SD is estimated from the
same observations, the SE is $s/\sqrt n$ and the reference is t with
$n-1$ df, whose heavier tails reflect the uncertainty of the estimated
scale [@student1908]; Chapter 3, Section 3.1, draws the two references.
The critical values that leave .025 in the upper tail show the
difference:

```{r}
#| label: ch05-critical-values
c(z = qnorm(.975), t5 = qt(.975, 5), t9 = qt(.975, 9),
  t30 = qt(.975, 30))
```

The multipliers are 1.96 for the Normal reference and 2.5706, 2.2622
and 2.0423 for t with 5, 9 and 30 df. A large sample makes the two
references similar; it does not make an estimated SD a known one.

The 32 recorded car models have mean fuel economy 20.0906 mpg. Take the
null process mean to be 20 mpg and let R run the one-sample t
procedure:

```{r}
#| label: ch05-one-t
t.test(mtcars$mpg, mu = 20)
```

The printout names the method, gives the statistic with its df and the
p-value, states the two-sided alternative, and reports the 95% interval
and the estimated mean. The returned object also stores these parts
(`$estimate`, `$stderr`, `$statistic`, `$parameter` for the df,
`$p.value` and `$conf.int`), which the Computer Lab extracts. Suppose,
as a second and explicitly different working assumption, that the
process SD were **known independently to be 5 mpg**; this teaching
value is not an estimate from `mtcars`. @tbl-ch05-one-mean sets the
two procedures side by side.

| Procedure | SE | Statistic | df | Two-sided p | 95% interval |
|:--|--:|--:|--:|--:|:--|
| Known-SD z, $\sigma=5$ | 0.8839 | 0.1025 | -- | .9183 | [18.3582, 21.8230] |
| Estimated-SD t | 1.0654 | 0.0851 | 31 | .9328 | [17.9177, 22.2636] |

: One mean, two explicitly different uncertainty assumptions. {#tbl-ch05-one-mean tbl-colwidths="[27,10,12,5,16,30]"}

The t row uses the observed SD, 6.0269, and the critical value 2.0395.
Both its SE and its multiplier exceed those of the known-SD row, so its
interval is wider. Both intervals contain 20 mpg, matching their
tests' failure to reject that null at .05; this matching holds when
interval and test share a method, assumptions and calibration. A
p-value is a tail probability under the null and those conditions, not
the probability that the null is true, and including 20 establishes
neither exact equality nor practical equivalence. An interval for a
process mean also differs from a range containing 95% of individual
cars. These calculations illustrate working process models: the
historical collection of car models does not become a probability
sample when `t.test()` runs, and a broader inference needs a supported
target process with appropriate dependence and distribution conditions.

### Supplied summaries and a planned direction

Chapter 3's one-mean district illustration, on which it first builds
intervals and tests, supplies only summaries: $n=25$, mean 52 and
sample SD 5 under independent Normal sampling. They are enough, so we
use them directly rather than inventing observations to pass into
`t.test()`. For the null mean 50, the estimated SE is $5/\sqrt{25}=1$
and the statistic is $t=(52-50)/1=2$ on 24 df. `pt()` turns the
statistic into tail areas:

```{r}
#| label: ch05-summary-p
c(two_sided = 2 * pt(2, df = 24, lower.tail = FALSE),
  upper = pt(2, df = 24, lower.tail = FALSE))
```

The two-sided p-value is .0569; the upper-tail p-value for the
alternative $\mu>50$ is .0285. They differ because they answer
different questions, and a directional alternative must come from the
question before the result is seen. For a negative statistic the
upper-tail probability would exceed .5, so halving a two-sided p-value
blindly would give the wrong tail. The two-sided 95% interval is
[49.9361, 54.0639], using R's critical value 2.0639, and it contains
50, consistent with the two-sided test; the upper-tail test does not
correspond to that interval. Chapter 3 used the t-table value 2.064,
which gives [49.936, 54.064]; a task that stipulates that rounded
critical value keeps those endpoints. The difference is rounding, not
a changed conclusion.

### Paired observations and complete pairs {#ch05-paired}

The sleep measurements return us to a comparison whose unit is a
person: ten complete pairs, not twenty independent people. Their
condition means are 0.75 and 2.33 hours, and the mean of their aligned
differences is 1.58 hours. Nine differences are positive and one is
zero, ranging from 0 to 4.6 hours.

```{r}
#| label: fig-ch05-pairs
#| echo: false
#| fig-cap: "Two measurements for each of ten people."
#| fig-alt: "Lines connect drug 1 and drug 2 for each person; nine rise and one is flat."
#| fig-width: 6.2
#| fig-height: 3.4
matplot(1:2, t(as.matrix(sleep_wide[c("extra.1", "extra.2")])),
  type = "b", lty = 1, pch = 16, col = "grey40",
  xaxt = "n", xlab = "Drug", ylab = "Extra sleep (hours)")
axis(1, at = 1:2, labels = c("1", "2"))
```

The inferential target is a process mean difference $\mu_D$. Under a
working model of independent Normal differences between people,
$s_d/\sqrt n$ estimates the SE of their mean and the t reference has
$n-1$ df. Normality concerns the differences, and the dependence
between one person's two measurements is kept in the difference, as
@fig-ch05-pairs shows line by line. The argument `paired = TRUE` tells
R to subtract measurements at corresponding positions, which the
ID-based reshaping of Section 3.4 made correct:

\Needspace{16\baselineskip}

```{r}
#| label: ch05-paired-test
t.test(sleep_wide$extra.2, sleep_wide$extra.1, paired = TRUE)
```

The output gives $t(9)=4.0621$, $p=.0028$ and the 95% interval [0.7001,
2.4599] hours around the mean difference 1.58. A paired t calculation
is a one-sample t calculation on the within-person differences, so the
same result follows from `t.test(sleep_d)`. Raising the confidence to
99% widens the interval to [0.3159, 2.8441] while the observations and
the estimate stay the same.

With missing measurements, keep the rows that have both values before
forming differences; removing missing entries separately from each
column could align different people. Complete-pair analysis also
changes who contributes and may need a missingness justification: a
software exclusion is not that justification. The interval concerns a
process mean, not a guarantee of benefit for every person, and ten
differences give limited protection against non-Normal tails or
influential cases. Neither the paired call nor the dataset establishes
random sampling, treatment order, absence of carryover or the other
conditions needed for a causal interpretation. Chapter 3's
eight-student illustration (Chapter 3, Section 5.2) and its school
contrasts apply the same form; the Computer Lab runs both.

### Two independent groups and Welch's method {#ch05-welch}

Now compare separate automatic and manual car models. Let
$\Delta\mu=\mu_A-\mu_M$ be the automatic-minus-manual process mean mpg
difference. Under an independent-groups working model the two means
contribute separate variance terms. Welch's method estimates both and
allows the group variances to differ [@welch1947]; it is the last of
Chapter 3's t procedures (Chapter 3, Section 5.4), following the
equal-variance test. With $a=s_A^2/n_A$ and $b=s_M^2/n_M$,

$$
\widehat{\Delta\mu}=\bar y_A-\bar y_M,\qquad
\widehat{\operatorname{SE}}=\sqrt{a+b},\qquad
\nu=\frac{(a+b)^2}{a^2/(n_A-1)+b^2/(n_M-1)}.
$$

The Welch--Satterthwaite df $\nu$ account approximately for the two
estimated variance terms and need not be an integer. Extracting the two
groups in a stated order makes the subtraction unambiguous:

\Needspace{18\baselineskip}

```{r}
#| label: ch05-welch-test
auto <- mtcars$mpg[mtcars$am == 0]
manual <- mtcars$mpg[mtcars$am == 1]
t.test(auto, manual)
```

The 19 automatic and 13 manual models have means 17.1474 and 24.3923
mpg, a difference of $-7.2449$ mpg. The output reports $t=-3.7671$ on
$\nu=18.3323$ df, $p=.0014$ and the 95% interval [$-11.2802$,
$-3.2097$]. The negative sign describes the declared order; reversing
it would negate the estimate and exchange the negated endpoints, with
the same two-sided p-value. R's two-sample `t.test()` uses Welch's test
by default, and the argument `var.equal = TRUE` gives Chapter 3's
equal-variance (pooled) test, which adds a common-variance assumption.
@tbl-ch05-welch-pooled compares the two on the same records.

\Needspace{8\baselineskip}

| Procedure | SE | df | Two-sided p | 95% interval |
|:--|--:|--:|--:|:--|
| Welch's unequal-variance t | 1.9232 | 18.3323 | .0014 | [$-11.2802$, $-3.2097$] |
| Equal-variance (pooled) t | 1.7644 | 30 | .0003 | [$-10.8484$, $-3.6415$] |

: The same automatic-minus-manual estimate, $-7.2449$ mpg, under two variance assumptions. {#tbl-ch05-welch-pooled tbl-colwidths="[34,12,14,14,26]"}

The estimate and the observations are unchanged; the variance
assumption and the calibration differ. The equal-variance t reference
is exact for independent Normal groups with a common variance, whereas
Welch's reference remains approximate even for independent Normal
groups with unequal variances. As Chapter 3 explains, the choice comes
from the design and subject knowledge before the analysis: a smaller
p-value, or a preliminary variance test, is not a reason to adopt the
common-variance assumption. Neither method resolves the differences in
weight, engine and other features that accompany transmission, so
these observational comparisons do not isolate a transmission effect.

Chapter 3's special-practice comparison has the same structure. Its
main result used the equal-variance test, $t(118)=2.7372$ with 95%
interval [0.8572, 5.3428] points, and its Section 5.4 set it beside
Welch's test, which has the same SE and statistic, $\nu=117.5119$ and
$p=.0072$. With 60 participants in each group only the df differ, and
the two intervals differ in the fourth decimal. The estimand and
repetition remain those of Section 3.1, and Section 3.4's reading of
the squared SE as an upper bound still holds; the Computer Lab
reproduces the Welch calculation from the supplied summaries. Either
interval includes effects below a stated two-point importance benchmark
as well as above it, so evidence against a zero effect supports less
than a claim that the mean improvement exceeds two points. The
supplied rounded summaries are also distinct from the distributed
CSV's slightly different SDs, and the two should not be silently
substituted for one another.

### Coverage belongs to the procedure {#ch05-coverage}

What makes an interval procedure a 95% procedure? A known process lets
us see. Let independent differences be Normal with mean 1.5 hours and
SD 1.2 hours; these are teaching constants, not a reconstruction of the
sleep study. Each repeated study draws fresh differences, estimates
their mean and SE, and forms a 95% t interval; recalculating one
interval on the same observations would reproduce that interval, not
test the procedure. The Computer Lab carries the loop over 2,000
studies at each of three sample sizes, and @tbl-ch05-coverage reports
its results.

\Needspace{11\baselineskip}

| n | Mean of estimates | Empirical SE | Model SE | Coverage |
|--:|--:|--:|--:|--:|
| 10 | 1.5015 | 0.3766 | 0.3795 | .9515 |
| 20 | 1.5006 | 0.2591 | 0.2683 | .9510 |
| 40 | 1.5014 | 0.1886 | 0.1897 | .9460 |

: Coverage of 95% t intervals from the Normal process, using 2,000 repetitions at each sample size and resetting seed 60610 for each row. {#tbl-ch05-coverage tbl-colwidths="[10,25,25,20,20]"}

The empirical SE is the SD of the generated means and the model SE is
$1.2/\sqrt n$; coverage is the fraction of generated intervals that
contain 1.5. Under the declared Normal model the t procedure has exact
95% coverage, so the simulated fractions differ from .95 only through
the finite number of studies. That numerical uncertainty has its own
**Monte Carlo SE**: each study's coverage indicator is a Bernoulli
outcome, and at coverage .95 with $B=2000$,

$$
\operatorname{MCSE}(\widehat p_{\mathrm{cover}})
=\sqrt{\frac{.95(.05)}{2000}}=0.0049.
$$

All three fractions lie within one MCSE of .95. Increasing $B$ reduces
the simulation noise, quadrupling it halves the MCSE. Increasing $n$
supplies more information in each study and narrows the intervals,
from 10 to 40 halving the model SE, without raising the nominal 95%
coverage above 95%. The individual-difference SD, the mean's SE and the
simulation's MCSE describe different sources of variation. Chapter 3's
coverage display shows the same fixed-parameter reasoning: for one
observed interval the parameter is inside or outside, and the 95%
describes repeated use of the procedure.

## Correlations and Their Inference {#sec-ch5-relationships}

[]{#ch05-correlations}

### Begin with the plotted observations

Mean comparisons group observations by a category. A relationship
between two quantitative variables begins instead with pairs of values.
For the 32 car models, `wt` measures weight in thousands of pounds and
`mpg` fuel economy in miles per US gallon. All 32 rows record both, so
the graph and every numerical description use the same rows:

```{r}
#| label: fig-ch05-car-pairs
#| fig-cap: "Mileage and weight for 32 recorded car models."
#| fig-alt: "Heavier models generally have lower mpg, with substantial scatter around the decreasing pattern and possible curvature."
#| fig-width: 6.2
#| fig-height: 3.5
plot(mtcars$wt, mtcars$mpg, pch = 16,
  xlab = "Weight (thousand pounds)", ylab = "Miles per gallon")
```

Read direction, strength and form before reducing the display to a
coefficient. Heavier models generally have lower mpg. Inspect unusual
observations, the range of weights, possible subgroups and changes in
spread: a coefficient compresses these features, and the plot shows
what that number can usefully describe.

Covariance averages the products of paired deviations. On the same $n$
observations, sample covariance and Pearson correlation are

$$
s_{xy}=\frac{\sum_{i=1}^n(x_i-\bar x)(y_i-\bar y)}{n-1},\qquad
r_{xy}=\frac{s_{xy}}{s_xs_y}.
$$

A pair contributes positively when both values lie above their means or
both below, and negatively when they lie on opposite sides. Covariance
keeps the product of the two units; dividing by the two SDs gives a
unit-free coefficient between $-1$ and 1 for variables with nonzero
spread. For $x=(1,2,3)$ and $y=(2,4,6)$, the deviation products sum to
4, the covariance is $4/(3-1)=2$, and the SDs are 1 and 2, so $r=1$:
all points lie on one increasing line.

```{r}
#| label: ch05-cov-cor
c(covariance = cov(mtcars$wt, mtcars$mpg),
  r = cor(mtcars$wt, mtcars$mpg))
```

The car covariance is $-5.1167$ thousand-pounds times mpg and $r=-.8677$.
Recording weight in pounds multiplies the covariance by 1,000, to
$-5116.6847$ pounds times mpg, and multiplies the weight SD by 1,000 as
well, so the correlation stays $-.8677$. Adding a constant changes
neither statistic; multiplying one variable by a negative constant
reverses both signs.

A small correlation need not mean little relationship. The five points
$x=(-2,-1,0,1,2)$ and $y=x^2=(4,1,0,1,4)$ in @fig-ch05-curve follow an
exact curve, yet their covariance and correlation are zero.

```{r}
#| label: fig-ch05-curve
#| echo: false
#| fig-cap: "An exact curved relationship with zero Pearson correlation."
#| fig-alt: "Five points at x equal to minus two, minus one, zero, one and two have y values four, one, zero, one and four, forming a U."
#| fig-width: 5.5
#| fig-height: 2.8
x_curve <- -2:2
plot(x_curve, x_curve^2, type = "b", pch = 16,
     xlab = "x", ylab = expression(y == x^2))
```

The cross-products at $x$ and $-x$ cancel by symmetry, although knowing
$x$ determines $y$. Pearson correlation describes linear association,
so "the variables are unrelated" would overstate this result. Chapter
2's correlation gallery (Chapter 2, Section 3.3) shows realized samples
with different directions and strengths; the Computer Lab builds it.

### Covariance and correlation matrices {#ch05-matrices}

A covariance matrix extends the pairwise calculation to several
variables: its diagonal holds variances and its off-diagonal entries
covariances. A correlation matrix standardizes each entry, with ones on
the diagonal for variables with positive variance. Both describe one
common set of rows. For fuel economy, gross horsepower (`hp`) and
weight, all 32 rows are complete, and three lines produce both
matrices.

```{r}
#| label: ch05-matrices
X <- mtcars[c("mpg", "hp", "wt")]
S <- cov(X)
R <- cov2cor(S)
```

`mtcars[c("mpg", "hp", "wt")]` selects three columns by name, `cov()`
returns the covariance matrix and `cov2cor()` rescales it into the
correlation matrix. @tbl-ch05-covariance and @tbl-ch05-correlation give
both in the same row and column order.

| Variable | mpg | hp | wt |
|:--|--:|--:|--:|
| mpg | 36.3241 | $-320.7321$ | $-5.1167$ |
| hp | $-320.7321$ | 4700.8669 | 44.1927 |
| wt | $-5.1167$ | 44.1927 | 0.9574 |

: Sample covariance matrix for the same 32 car models. {#tbl-ch05-covariance}

\Needspace{9\baselineskip}

| Variable | mpg | hp | wt |
|:--|--:|--:|--:|
| mpg | 1.0000 | $-.7762$ | $-.8677$ |
| hp | $-.7762$ | 1.0000 | .6587 |
| wt | $-.8677$ | .6587 | 1.0000 |

: Corresponding Pearson correlation matrix. {#tbl-ch05-correlation}

The mpg variance has squared-mpg units, and its covariance with
horsepower has mpg-times-horsepower units; the correlations remove
those units. If $\mathbf D$ is the diagonal matrix of the positive
sample SDs, the two conversions are

$$
\mathbf R=\mathbf D^{-1}\mathbf S\mathbf D^{-1},\qquad
\mathbf S=\mathbf D\mathbf R\mathbf D.
$$

Correlations alone therefore cannot recover the covariance matrix: the
SDs and their units are also needed, and the Computer Lab rebuilds
$\mathbf S$ from $\mathbf R$ and the SDs. A constant column has zero SD
and an undefined correlation, so it must be detected before
converting. Common rows matter when measurements are missing. Computing
each entry from a different pairwise subset can produce a matrix that
describes no single joint sample and is not positive semidefinite, so
it can imply an impossible negative variance for a combination of the
variables. For airquality's ozone, solar radiation and wind, the
individual recorded counts are 116, 146 and 153, but the common set
contains 111 days.

### Pearson, rank, and point-biserial coefficients

Pearson correlation uses the measured scales and describes linear
association. **Spearman correlation** applies Pearson's formula to the
ranks and describes monotone association: values move consistently in
one direction, even along a curve. **Kendall correlation** compares
concordant and discordant orderings of pairs, with an adjustment for
ties. For mpg and horsepower:

```{r}
#| label: ch05-spearman
c(Pearson = cor(X$mpg, X$hp),
  Spearman = cor(X$mpg, X$hp, method = "spearman"))
```

```{r}
#| label: fig-ch05-rank
#| echo: false
#| fig-cap: "Mileage and horsepower for the 32 models, (a) on the recorded scales and (b) as ranks. The curved decline in (a) weakens the linear coefficient; ranks straighten a monotone pattern."
#| fig-alt: "Left panel: mpg falls steeply with horsepower up to about 150 and then levels off, a curved decreasing pattern. Right panel: the ranks of the same models fall along a nearly straight decreasing band."
#| fig-width: 6.5
#| fig-height: 3
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 2, .5), mgp = c(2.3, .7, 0))
plot(X$hp, X$mpg, pch = 16, xlab = "Gross horsepower",
     ylab = "Miles per gallon", main = "(a) Recorded values")
plot(rank(X$hp), rank(X$mpg), pch = 16, xlab = "Horsepower rank",
     ylab = "Mpg rank", main = "(b) Ranks")
par(oldpar)
```

Mileage falls steeply as horsepower rises and then levels off
(@fig-ch05-rank). The decline is monotone but curved, so the linear
coefficient, $-.7762$, is weaker than the rank coefficient, $-.8947$,
which sees only the ordering. A **point-biserial correlation** is
Pearson's coefficient with one variable coded 0/1; @tbl-ch05-coefficients
collects the four coefficients.

| Coefficient | What it describes | mtcars value |
|:--|:--|--:|
| Pearson, mpg and hp | Linear association on the recorded scales | $-.7762$ |
| Spearman, mpg and hp | Monotone association: Pearson's coefficient on ranks | $-.8947$ |
| Kendall, mpg and hp | Concordant minus discordant ordering of pairs, adjusted for ties | $-.7428$ |
| Point-biserial, `am` and mpg | Pearson's coefficient with manual coded 1 and automatic 0 | .5998 |

: Four coefficients, what each describes, and its value. {#tbl-ch05-coefficients tbl-colwidths="[28,54,18]"}

The different magnitudes do not make the largest coefficient the best
description in general: the question and the plot decide which
association matters. Tied horsepower values keep tied ranks, and the
ties also affect test calibration. The point-biserial .5998 corresponds
to manual and automatic means of 24.3923 and 17.1474 mpg; reversing the
0/1 coding reverses its sign and leaves the same two groups. For more
than two categories, arbitrary numerical labels do not define a
meaningful scale for correlation, so the coding must follow the
variable's role.

### A selected correlation test {#ch05-correlation-tests}

The descriptive weight--mpg correlation concerns the recorded models.
Inference adds a target process correlation $\rho$ and a supporting
model. Suppose we selected the two-sided null $H_0:\rho=0$ before
examining the result, with level $\alpha=.05$. Under independent
bivariate-Normal sampling the zero-correlation statistic

$$
t=r\sqrt{\frac{n-2}{1-r^2}},\qquad df=n-2,
$$

has an exact $t_{n-2}$ reference. One call carries out the test on the
same 32 pairs:

\Needspace{15\baselineskip}

```{r}
#| label: ch05-pearson-test
cor.test(mtcars$wt, mtcars$mpg)
```

The output gives $r=-.8677$, $t(30)=-9.559$ and a p-value so small that
R prints it in scientific notation, $p<.0001$. The result is evidence
against zero process correlation under the stated reference model; a
small p-value does not establish that model or turn the historical car
records into a probability sample. Normality is not needed to calculate
the descriptive $r$, whereas independence, form, influential cases and
the observed range matter for inference.

The printed 95% interval, [$-.9338$, $-.7441$], uses Fisher's
transformation, a different approximation from the exact zero-null
test:

$$
CI_\rho=\tanh\left\{\operatorname{atanh}(r)
\pm\frac{z_{.975}}{\sqrt{n-3}}\right\}.
$$

Why transform? @fig-ch05-fisher simulates 5,000 studies of 32 pairs
whose process correlation is $-.8677$.

```{r}
#| label: fig-ch05-fisher
#| echo: false
#| fig-cap: "Sample correlations from 5,000 simulated studies of 32 pairs with process correlation $-.8677$. (a) On the correlation scale the distribution is skewed, with a long tail toward zero and a hard limit at $-1$; (b) on the Fisher scale it is nearly Normal, and the curve is the Normal density with SD $1/\\sqrt{29}$. Dashed lines mark the process value on each scale."
#| fig-alt: "Left panel: a histogram of simulated correlations between about minus .97 and minus .6, peaked near minus .88 with a long right tail. Right panel: the same values after the inverse hyperbolic tangent form a symmetric histogram matched by a Normal curve."
#| fig-width: 6.5
#| fig-height: 3
set.seed(12345)
rho <- -.8677
r_sim <- replicate(5000, {
  x <- rnorm(32)
  y <- rho * x + sqrt(1 - rho^2) * rnorm(32)
  cor(x, y)
})
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 2, .5), mgp = c(2.3, .7, 0))
hist(r_sim, breaks = seq(-1, -.55, by = .01), probability = TRUE,
     col = "grey85", xlab = "Sample correlation r",
     main = "(a) Correlation scale")
abline(v = rho, lty = 2)
hist(atanh(r_sim), breaks = 40, probability = TRUE, col = "grey85",
     xlab = "atanh(r)", main = "(b) Fisher scale")
curve(dnorm(x, atanh(rho), 1 / sqrt(29)), add = TRUE, lwd = 2)
abline(v = atanh(rho), lty = 2)
par(oldpar)
```

Near $-1$ the sample correlations are squeezed against the bound and
trail off toward zero, so a margin of equal size on both sides of $r$
would misplace the interval. On the Fisher scale the same values are
nearly Normal, with SD 0.1831 against the model's
$1/\sqrt{29}=0.1857$. The interval is therefore built there and both
endpoints are transformed back, which makes it asymmetric about $r$:
it extends 0.0662 below $-.8677$ and 0.1236 above. The interval needs
at least four complete pairs and remains approximate even under
bivariate Normality, and substantial nonlinearity can make linear
correlation a poor answer to the substantive question whatever the
interval's width.

Other questions have tests of their own, each reported with its native
statistic (@tbl-ch05-association-tests).

\Needspace{11\baselineskip}

| Question and test | Native statistic | Two-sided p |
|:--|:--|:--|
| Pearson, weight and mpg | $t(30)=-9.559$ | $<.0001$ |
| Spearman, mpg and horsepower (approximate) | $S=10337.2903$ | $<.0001$ |
| Kendall, mpg and horsepower (approximate) | $z=-5.871$ | $<.0001$ |
| Point-biserial, `am` and mpg | $t(30)=4.1061$ | .0003 |

: Four tests of association and the statistics R reports for them. {#tbl-ch05-association-tests tbl-colwidths="[46,30,24]"}

Spearman's S and Kendall's z are not Pearson t statistics, and a report
names the statistic it quotes. The horsepower values contain ties, so
the approximate rank calibrations are requested explicitly with
`exact = FALSE`; the rank tests supply no Fisher interval. A selected
question justifies one test. Choosing whichever method gives the
smallest p-value changes the testing procedure, and a matrix of
unplanned tests creates a comparison family, whose calibration Chapter
4, Section 3.3, develops and Chapter 6 carries out in R.

The point-biserial statistic equals the statistic of Chapter 3's
equal-variance (pooled) two-sample t test, with manual first to match
the positive coding. Its supporting model is independent, conditionally
Normal outcomes with a common variance in the two groups; a binary
variable cannot itself be bivariate Normal with the outcome. The
equivalence is to the equal-variance procedure and generally does not
hold for Welch's test; Chapter 7 connects it to a regression slope.

### One-control partial correlation {#ch05-partial}

Chapter 2 distinguished changing the cases from adjusting a
relationship on the same cases. For variables $X$, $Y$ and $Z$, the
**partial correlation** of $X$ and $Y$ controlling linearly for $Z$ is

$$
r_{XY\cdot Z}=\frac{r_{XY}-r_{XZ}r_{YZ}}
{\sqrt{(1-r_{XZ}^2)(1-r_{YZ}^2)}}.
$$

Chapter 2's supplied practice-time, score and baseline correlations,
$r_{XY}=.6$, $r_{XZ}=.5$ and $r_{YZ}=.6$, give .433. For mpg and
horsepower controlling for weight, the correlations of
@tbl-ch05-correlation give $-.547$. A second route shows what the
adjustment does: fit a straight line from weight to each variable, keep
the **residuals**, the observed values minus their fitted values, and
correlate what remains.

```{r}
#| label: ch05-partial-residuals
resid_mpg <- residuals(lm(mpg ~ wt, data = X))
resid_hp <- residuals(lm(hp ~ wt, data = X))
cor(resid_mpg, resid_hp)
```

`lm()` fits the straight line its formula names, the outcome left of
`~` and the predictor right of it, with an intercept; `residuals()`
returns what remains after that fitted linear relationship, for the
same 32 rows.

```{r}
#| label: fig-ch05-partial
#| echo: false
#| fig-cap: "Mileage and horsepower after separately fitting their linear relationships with weight."
#| fig-alt: "The two residual vectors from the same 32 car models retain a negative association, with partial correlation minus .547."
#| fig-width: 6.2
#| fig-height: 3.5
plot(resid_hp, resid_mpg, pch = 16,
     xlab = "Horsepower residual after weight adjustment",
     ylab = "Mpg residual after weight adjustment")
```

Both routes give $-.547$: the mpg--horsepower association weakens from
$-.7762$ once each variable's linear relationship with weight is
removed, and a negative association remains (@fig-ch05-partial). The
adjustment uses the same 32 rows; it does not exclude heavy cars. It
removes the specified linear relationships and nothing more: not all
confounding, omitted nonlinear relationships or measurement problems,
and it does not establish a causal effect of horsepower. Chapter 8
develops adjustment for several controls.

If the partial coefficient is the preselected inferential question, the
reference must account for fitting the control. With $k$ controls and
$n$ common observations the classical statistic is

$$
t=r_{XY\cdot Z}\sqrt{\frac{n-k-2}{1-r_{XY\cdot Z}^{\,2}}},\qquad df=n-k-2,
$$

which here gives $t=-3.5187$ on 29 df, $p=.0015$. This exact classical
calibration requires independent observations, full-rank linear
adjustment and a Normal conditional-error model, or a multivariate
Normal model. Applying an ordinary `cor.test()` to the two residual
vectors would use $n-2=30$ df and ignore the fitted control.

A sensitivity analysis asks a different question. Restricting the car
record to weights of at most 5,000 pounds changes its cases and gives
$r=-.8792$ between weight and mpg; removing the three ozone days above
120 ppb changes the mean to 39.4867, the SD to 28.8942 and the median to
30. These are consequences of stated restrictions, not reasons by
themselves to delete observations, and linear adjustment and case
selection remain distinct even when both change a reported association.

## Report the Answer {#sec-ch5-report}

[]{#ch05-report}

### A result connects numbers to a question

A statistical report returns from the function output to the question.
Name the people or records, the outcome and its units, the comparison,
the estimate and the uncertainty; then state the conditions and scope
that support the interpretation. The sleep result can be reported as
follows:

> For the ten recorded people, extra sleep under drug 2 averaged
> 1.58 hours more than under drug 1. The SD of the ten within-person
> differences was 1.23 hours. Under an independent Normal-differences
> working model, the mean difference has SE 0.389 hours and a 95% t
> interval of [0.7001, 2.4599] hours ($t(9)=4.0621$, two-sided $p=.0028$).
> This uncertainty concerns the stated process mean. A causal account
> also requires information about treatment assignment, order and carryover.

The first two sentences describe the recorded people. The next adds
model-based uncertainty for a named target, rather than a range of
individual responses. The final sentence identifies the information a
causal claim would need. A successful computation supplies the result;
the study setting supplies the argument for using it [@peng2011].

### Connect a report to its source

A saved script records calculations. A Quarto source places them beside
the prose that explains them, so rendering executes the code and
inserts its output. The prepared two-file example uses a short `.qmd`
source and an accompanying bibliography. Its paired analysis uses these
same data, subtraction direction, model and base-R commands: one chunk
computes the result, an inline value inserts a number into a sentence,
and a table or graph helps readers see what it describes.

During the final Computer Lab round, render the prepared source, change
the confidence level from .95 to .99 and revise the corresponding
sentence. The interval becomes [0.3159, 2.8441] while the estimate stays
1.58 hours. That small change connects an input, a calculation, its
output and its interpretation. The lab reference explains the header,
chunks, cross-references, citations and report-support installation,
and its first HTML render requires no PDF typesetting setup.

When a data-selection rule or a method changes, computed values update
automatically; the interpretation still needs to be reread. A
reproducible report keeps question, data, code, output and explanation
consistent. Rendering shows that a document can be produced, while the
evidence and reasoning determine what its claims establish.

## Reading Bridge {#sec-ch5-summary}

When a study reports a mean, interval or association, identify what was
recorded, which observations entered the calculation, the statistic
or comparison and its units. Next locate the repetition or model that
supports the SE, reference distribution, interval or test. Finally
separate the numerical result from the assumptions needed for population,
practical or causal claims. R reproduces a computation; the study's
evidence and design support its interpretation.

This chapter connected base-R descriptions to sampling variation,
mean inference and selected correlation inference. Chapter 6 applies
the same reading pattern to counts and group models, introduces a small
ggplot2 plotting vocabulary, and examines fitting criteria and procedures
across repetitions. Chapters 7--8 extend correlations, comparisons and
linear adjustment into regression arguments.

### Key terms {.unnumbered}

Use this short reference to revisit the computing, simulation and
matrix terms explained here. The Part One Concept Reference gives the
principal home of the descriptive and inferential terms this chapter
applies, which Chapters 2--3 explain.

| Term and common alternatives | Meaning |
|:--|:--|
| Object (R object) | A named value, vector, table, model result, or other item stored in an R session. Assignment with `<-` names an object; the Environment pane lists the current objects. |
| Vector | A sequence of values of one basic type: numeric, character, or logical. `NA` marks an unavailable value; it is not zero. A data frame is a table whose columns are vectors of equal length, one row per case. |
| Function and argument (R function) | A function performs an operation on supplied inputs; an argument names an input or an option, such as `na.rm = TRUE`. An R argument is distinct from a study argument in the argument-based approach. |
| Package (library) | An installed collection of functions and data, such as stats or graphics, which are supplied with R. Installing an additional package makes it available on the computer; `library()` attaches it for the session; `package::function()` names one exported function explicitly. |
| Script and Quarto source | A saved `.R` script records commands so an analysis can be rerun; a `.qmd` source connects prose, mathematics, executable code, tables, figures, and references, and renders to PDF, HTML, or Word. |
| Simulation and seed (Monte Carlo simulation) | Generating data from a declared process to see what a statistic or procedure does across repetitions. A seed makes the random draws repeatable; it does not make the process true for an actual study. |
| Monte Carlo standard error | The numerical uncertainty of a simulated probability or mean that comes from using finitely many repetitions; more repetitions shrink it. It is distinct from a study's SE, which depends on the study's sample size. |

: Computing and simulation terms. {#tbl-ch05-computing-terms tbl-colwidths="[28,72]"}

| Term and common alternatives | Meaning |
|:--|:--|
| Covariance matrix | A square table of variances on the diagonal and covariances off the diagonal for a stated set of variables and observations. |
| Correlation matrix | A square table of pairwise correlation coefficients for a stated set of variables and observations; its diagonal is 1 for variables with positive variance. |

: Data-summary terms. {#tbl-ch05-matrix-terms tbl-colwidths="[28,72]"}

### Further reading {.unnumbered}

R's reference system [@rcore2025] and local Help for `mean`, `sd`,
`quantile`, `t.test`, `cor`, `cov2cor` and `cor.test` describe their
arguments, defaults and computational conventions. The original papers
by Student [-@student1908] and Welch [-@welch1947] explain the classical
mean procedures developed here. Quarto's documentation [@quarto2026]
gives examples of executable reports, and Peng [-@peng2011] connects
reproducible computation with broader scientific reporting.

## Exercises {.unnumbered}

Use the stated records and each prompt's supplied summaries. Explain the
statistical meaning as well as any calculation, and report numerical
results to the requested precision.

<!-- prompt-id: CH05-E01 -->

### Exercise 1: Count the right observations (Foundational) {.unnumbered #ex-ch05-01}

A table contains 40 students with one baseline and one follow-up row each,
plus one duplicate follow-up row. There are 81 rows. Explain the difference
between a row count and a student count. If four students have missing
follow-up scores, give the denominator for the mean observed follow-up score
after the documented duplicate is resolved. State what the mean describes.

<!-- END EX05-01 PROMPT -->

<!-- prompt-id: CH05-E02 -->

### Exercise 2: Center, spread and a changed observation (Foundational) {.unnumbered #ex-ch05-02}

Section 2.1's three recorded scores, 50, 60 and 70 points, have mean and
median 60 and sample SD 10 points. Suppose the lowest score had
instead been 20.

1. Before calculating, predict whether the mean, the median and the SD
   rise, fall or stay the same, and give a reason for each.
2. Calculate the new mean and sample SD (denominator $n-1$), reporting
   the SD to four decimal places. Compare this SD with the one Section
   2.1 reports after 70 is replaced by 100, and explain the comparison.
3. Which center would you report as a typical score, and does the low
   value alone justify removing it?

<!-- END EX05-02 PROMPT -->

<!-- prompt-id: CH05-E03 -->

### Exercise 3: A missing score is not zero (Foundational) {.unnumbered #ex-ch05-03}

In Section 1.2, students A, B and C have baseline scores 50, 60 and 70
and follow-up scores 55, missing and 78, and the complete-pair mean
change and the difference of the separate available means both equal
6.5 points. Suppose instead that B's follow-up score is 66 and C's
is missing.

1. Before calculating, predict whether the two quantities still agree
   and, if not, which is larger. Base the reason on which students
   enter each mean.
2. Calculate both quantities and state each denominator.
3. Explain what recording C's missing score as 0 would claim, and why
   the agreement in Section 1.2 did not show that the two calculations
   used the same students.

<!-- END EX05-03 PROMPT -->

<!-- prompt-id: CH05-E04 -->

### Exercise 4: Choose the group denominator (Foundational) {.unnumbered #ex-ch05-04}

In Section 2.3, 4 of 19 automatic and 10 of 13 manual car models meet a
fuel-economy threshold: 21.0526% and 76.9231% within the transmission
groups, and 71.4286% of the 14 threshold-meeting models are manual.

1. A report states, "71.4286% of manual models meet the threshold."
   Name the denominator the sentence implies and the one the 71.4286%
   uses, then correct the sentence.
2. An invented second record has the same 13 manual models and 38
   automatic models, 8 of which meet the threshold. Before calculating,
   predict which of the three percentages change, and why. Then
   calculate the percentage of threshold-meeting models that are manual.

<!-- END EX05-04 PROMPT -->

<!-- prompt-id: CH05-E05 -->

### Exercise 5: Read a relationship before evaluating it (Integrative) {.unnumbered #ex-ch05-05}

Section 5.1 plots five paired observations, $x=(-2,-1,0,1,2)$ and
$y=(4,1,0,1,4)$, whose covariance and Pearson correlation are both zero.
Describe the scatterplot, and use the signs of the paired deviations to
explain why the covariance is zero. A report says, "The correlation is
zero, so the variables are unrelated." First identify its numerical
result and reported conclusion, then evaluate whether that conclusion
follows. What happens to covariance and correlation if $y$ is
multiplied by 10?

<!-- END EX05-05 PROMPT -->

<!-- prompt-id: CH05-E06 -->

### Exercise 6: Report paired differences (Integrative) {.unnumbered #ex-ch05-06}

Four people have condition-1 values 1, 2, 4 and 5 and corresponding
condition-2 values 2, 4, 4 and 8 hours. Their condition-2-minus-condition-1
differences are 1, 2, 0 and 3 hours, with mean 1.5 hours and sample SD
1.291 hours.

1. Calculate the two condition means and confirm that their difference
   equals the mean of the differences.
2. Suppose the condition-2 values were attached to the wrong people, in
   the order 8, 4, 4 and 2. Before calculating anything, predict whether
   the mean difference and the SD of the differences change, and explain
   which summary depends on keeping each person's two values together.
3. Write a descriptive result that names the people, direction and
   units, and explain why it does not say that every person gained
   1.5 hours.

<!-- END EX05-06 PROMPT -->

<!-- prompt-id: CH05-E07 -->

### Exercise 7: Simulate a sampling distribution and read its SE (Integrative) {.unnumbered #ex-ch05-07}

The waiting-time simulation in Section 3.2 reports, for $n=30$, a mean
of sample means of 9.9626 and an empirical SE of 1.7878 against a model
SE of 1.8257.

1. Explain what one histogram bar counts and why the empirical and model
   SEs differ.
2. Without running R, give the model SE for $n=120$ and for $n=480$ and
   explain the pattern.
3. Say what changes, and what does not, if the number of repetitions is
   raised from 5,000 to 20,000.
4. Rerun from the same seed with 120 waits in each sample, the
   simulation gives an empirical SE of 0.9214 minutes. Compare it with
   your model SE for $n=120$, and explain why the two need not agree
   exactly.
5. The running mean and the histograms of repeated sample means show
   different changes as sample size increases. Explain concentration
   toward a target and increasingly Normal standardized sampling shape,
   and say whether either makes individual waiting times Normal.

<!-- END EX05-07 PROMPT -->

<!-- prompt-id: CH05-E08 -->

### Exercise 8: Name the estimand, estimator, and estimate, then read the SE (Integrative) {.unnumbered #ex-ch05-08}

The special-practice study assigned 60 students to each of two formats;
the delayed-assessment means were 27.1 and 24.0 with SDs 6.0 and 6.4.

1. State the estimand, the estimator, and the estimate in one sentence
   each, and name the repetition behind the uncertainty.
2. Section 3.4 calculates the SE of the mean difference from these
   summaries as 1.1325 points. Explain why its square estimates an
   upper bound on the assignment variance rather than the variance
   itself.
3. For the ten sleep differences (SD 1.23 hours), the SE of the mean
   difference is 0.389 hours. Using the two condition SDs, 1.7890 and
   2.0022, as if the two sets of condition measurements were independent
   gives 0.8491 hours instead. Explain why pairing gives the smaller SE
   for the same ten people.
4. A separate simulation estimates a waiting-time variance of 100.
   Across 5,000 samples of five waits, variances using the $n-1$
   denominator average 101.2241 and those using $n$ average 80.9793.
   Explain which rule is unbiased, why 101.2241 need not equal 100 in
   this finite run, and why unbiasedness does not ensure that one
   estimate is close to the target.

<!-- END EX05-08 PROMPT -->

<!-- prompt-id: CH05-E09 -->

### Exercise 9: An interval and its question (Foundational) {.unnumbered #ex-ch05-09}

Twenty-five independent paired differences have mean 3 and SD 5,
so the estimated SE of their mean is 1 and the statistic for a zero
process mean is $t=3$. Under the Normal-difference model use
$t_{.975,24}=2.064$ to calculate the 95% interval, and explain what
$t=3$ says about the observed mean. Then name the appropriate
procedures for (a) the same 20 people measured twice and (b)
independent groups of 20 and 25 without equal variances. State why the
interval is not a range for individual differences.

<!-- END EX05-09 PROMPT -->

<!-- prompt-id: CH05-E10 -->

### Exercise 10: A selected relationship (Integrative) {.unnumbered #ex-ch05-10}

Section 5 analyses the common 32 rows of mpg, hp and wt. Their
correlations are $-.7762$ for mpg and hp, $-.8677$ for mpg and wt and
.6587 for hp and wt; the Spearman coefficient for mpg and hp is
$-.8947$; and `cor.test()` for the preselected weight--mpg question
reports $t=-9.559$ on 30 df, $p<.0001$ and the 95% interval
$[-.9338,-.7441]$.

1. Explain which entries of the covariance matrix carry measurement
   units, and why the correlation matrix alone cannot recover the
   covariance matrix.
2. State which association the Pearson and the Spearman coefficients
   for mpg and hp each describe, and use the scatterplot of mpg against
   horsepower in Section 5.3 to explain why the rank coefficient is
   larger in magnitude.
3. Calculate the partial correlation of mpg and hp controlling linearly
   for weight from the supplied correlations, give the df of its
   classical test, and explain why the adjustment alone does not
   establish a causal effect.
4. Report the preselected weight--mpg result with its estimate,
   interval and the conditions supporting its reference, and explain
   why the interval is not symmetric about $-.8677$.

<!-- END EX05-10 PROMPT -->

## References {.unnumbered}
