---
title: "Multiple Regression II: Categorical Predictors, Model Comparison, Nonlinearity, and Robust Uncertainty"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.pos = "H")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 7, width = 68)
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 11))
```

\clearpage

::: {.callout-note title="Chapter overview"}
Chapters 7 and 8 fitted lines and additive models and read them as
conditional comparisons. Along the way they used tools they could only
name: a factor coded as 0/1 columns, a contrast between two categories, a
joint test of several added coefficients, an information criterion, a
squared predictor, a logged outcome, a covariance estimate that allows
unequal spread, and a covariance that respects pupils sitting in schools.
This chapter develops each of them once, on the same 98 occupation
records, and then on Project STAR, where a randomized design supplies a
different warrant. By the end of the chapter you will be able to code and
read a factor with several levels, estimate a contrast with the right
uncertainty, compare nested models and read AIC and BIC, change a
predictor's form or an outcome's scale and say what question changed,
choose between classical, HC3, cluster-robust, and bootstrap uncertainty
for a stated reason, and say what none of these calculations repairs.
:::

A regression report rarely stops at one additive equation. It compares
groups that are categories, asks whether a set of added variables matters
jointly, ranks candidate models, lets a relationship curve or a scale
change, and reports uncertainty calculated in more than one way. Each of
these steps has a small mathematics of its own and a temptation of its
own: reading a coefficient as "the effect of type", adding two standard
errors, comparing criteria across different outcomes, or treating a robust
standard error as a repair for a design problem.

We keep the occupation records of Chapter 8, with their 98 complete rows,
income in thousand 1971 Canadian dollars, and the additive model of
education, prestige, percentage of women, and type, so that every tool is
seen on numbers already read. Nothing here is a new study. The chapter's
contribution is to give a scattered set of operations one home, one
vocabulary, and one set of records, so that the difference between
changing a model, changing a question, and changing an uncertainty
calculation stays visible. The chapter closes with Project STAR, where the
uncertainty calculation has to follow the design that produced the data.

Each visible R example follows from the earlier visible preparation in
this chapter; Chapters 7 and 8 supply familiarity, not an existing R
session. The type levels remain *bc*, *prof*, and *wc*, the full additive
formula remains `income_k ~ education + prestige + women + type`, and
every adjusted type comparison names education, prestige, and the
percentage of women as its conditioning set.

\clearpage

## Predictors That Are Categories {#sec-ch9-factors}

### One indicator, then several

Chapter 7, Section 3.3, fitted a single 0/1 predictor to 24 invented
feedback-by-practice scores. The fitted line passed through the two group
means, its slope was the difference between them, 4.0000 with SE 1.6967
and $t=2.3575$ on 22 df, and that test was the equal-variance two-sample
test written as a regression. Nothing about that arrangement is special to
two groups.

A factor with $J$ levels enters a regression as $J-1$ indicator columns.
One level is omitted and becomes the **reference**: the intercept is its
fitted mean, and each indicator coefficient is a difference from it. With
three occupation types and blue-collar as the reference,

$$
\widehat y_g=\widehat\alpha+\widehat\delta_{\mathrm{prof}}D_{\mathrm{prof}}
+\widehat\delta_{\mathrm{wc}}D_{\mathrm{wc}},\qquad
\widehat\mu_{\mathrm{bc}}=\widehat\alpha,\quad
\widehat\mu_{\mathrm{prof}}=\widehat\alpha+\widehat\delta_{\mathrm{prof}},\quad
\widehat\mu_{\mathrm{wc}}=\widehat\alpha+\widehat\delta_{\mathrm{wc}}.
$$

Two indicators, not three, carry three means: the third column would be
the intercept column minus the other two, and the design matrix would
lose full rank. The choice of reference is a choice of coordinates, not a
choice of model.

We re-create the Chapter 8 objects visibly so that every later calculation
in this chapter has a stated origin. The 98 rows are the type-complete
occupations; `m0`, `m1`, and `m2` are the education-only, the
education-and-prestige, and the full additive models.

```{=latex}
\Needspace{20\baselineskip}
```

```{r}
#| label: ch9-records
#| echo: true
prestige <- carData::Prestige
prestige$income_k <- prestige$income / 1000
d <- prestige[!is.na(prestige$type), ]
d$type <- factor(d$type, levels = c("bc", "prof", "wc"))
m0 <- lm(income_k ~ education, data = d)
m1 <- lm(income_k ~ education + prestige, data = d)
m2 <- lm(income_k ~ education + prestige + women + type, data = d)
c(rows = nrow(d), residual_df = df.residual(m2))
```

The 98 retained records hold 44 blue-collar, 31 professional, and 23
white-collar occupations. The full model fits six coefficients, leaving 92
residual degrees of freedom. Complete-case selection is how these rows
were chosen; it is not random sampling from a labour market
[@carData2022; @fox2019].

### Indicators recover fitted group means

Before adjusting for anything, fit income on type alone. Printing the
model matrix for one row of each type makes the coding visible rather than
assumed, and placing the fitted values beside the observed group means
shows what the indicators are doing.

```{=latex}
\Needspace{24\baselineskip}
```

```{r}
#| label: ch9-group-means
#| echo: true
type_levels <- data.frame(type = factor(c("bc", "prof", "wc"),
  levels = levels(d$type)))
model.matrix(~ type, data = type_levels)
group_fit <- lm(income_k ~ type, data = d)
cbind(type_levels,
      observed = as.numeric(tapply(d$income_k, d$type, mean)),
      fitted = predict(group_fit, type_levels))
coef(group_fit)
```

The blue-collar row carries zeros in both indicator columns, so its fitted
value is the intercept alone. The observed means are 5.3741, 10.5595, and
5.0523 thousand dollars, and the fitted values reproduce them exactly. The
coefficients are 5.3741, 5.1853, and $-0.3218$: the intercept is the
blue-collar mean, and each indicator coefficient is that type's mean minus
the blue-collar mean.

This is not an accident of these data. Least squares minimizes the sum of
squared deviations, and with one free parameter per group the minimizing
value within each group is that group's own mean. A group-only regression
is therefore a description of three means with no adjustment in it at all.

Keep that description apart from the adjusted comparison of the full
model. In `m2` the professional and white-collar coefficients are 0.3242
and 0.2353: they compare fitted incomes at the *same* education, prestige,
and percentage of women, not the observed incomes of differently composed
groups. The observed professional advantage of 5.1853 thousand dollars
and the adjusted 0.3242 are answers to different questions, and neither is
"the effect of occupational type".

### Releveling and centering change coordinates, not the fit

Releveling makes a different level the reference. Refit the group-only
model with white-collar as reference and compare the fitted values.

```{=latex}
\Needspace{14\baselineskip}
```

```{r}
#| label: ch9-relevel
#| echo: true
group_fit_wc <- lm(income_k ~ relevel(type, ref = "wc"), data = d)
coef(group_fit_wc)
max(abs(fitted(group_fit) - fitted(group_fit_wc)))
```

The coefficients are now 5.0523, 0.3218, and 5.5071: the intercept is the
white-collar mean, and the two indicator coefficients are the blue-collar
and professional means measured from it. The fitted values agree with the
blue-collar coding to about $10^{-13}$, the size of floating-point
rounding. The same three means are being reported in a different
coordinate system.

The full model behaves the same way. Releveling `m2` to white-collar gives
intercept 0.6559, blue-collar $-0.2353$, and professional 0.0889, while
the education, prestige, and women coefficients are unchanged to the last
printed digit. Centering a continuous predictor is the analogous
operation: subtracting a constant moves the profile the intercept refers
to, and leaves every fitted value, residual, and fixed substantive
contrast alone.

Two consequences are worth keeping. First, a coefficient's size is not
interpretable without the reference it is measured from; "the professional
coefficient" is short for "professional minus whichever level is currently
omitted". Second, neither releveling nor centering creates information.
Chapter 8 documented how thin the joint support across types is at common
education and prestige values; changing coordinates does not add a single
observed comparison.

### The one-way analysis of variance of Chapter 4 as a regression

Chapter 4 tested equality of several group means with a one-way analysis
of variance. In the group-only regression, that null is the null that both
indicator coefficients are zero, so the reduced model is the
intercept-only model and the comparison is a nested-model comparison:

$$
F=\frac{(RSS_{\mathrm{reduced}}-RSS_{\mathrm{full}})/2}
{RSS_{\mathrm{full}}/(98-3)}
=\frac{(1734.0792-1138.1230)/2}{1138.1230/95}=24.8725.
$$

The reduced model's residual sum of squares is the total sum of squares
about the grand mean, 1734.0792; the group-only model leaves 1138.1230 on
95 degrees of freedom. Two restrictions are tested, and the denominator is
the full model's residual mean square.

```{=latex}
\Needspace{26\baselineskip}
```

```{r}
#| label: ch9-anova-identity
#| echo: true
null_fit <- lm(income_k ~ 1, data = d)
anova(null_fit, group_fit)
type_aov <- aov(income_k ~ type, data = d)
c(regression_F = anova(null_fit, group_fit)$F[2],
  anova_F = summary(type_aov)[[1]][["F value"]][1],
  max_fitted_difference = max(abs(fitted(group_fit) - fitted(type_aov))))
```

`aov()` returns the same statistic from the same fitted means, the same
within-group residual sum of squares, the same two restrictions, and the
same denominator degrees of freedom, and its fitted values are identical.
The equality is algebraic. It holds whether or not group membership was
randomized, and it therefore supplies no causal warrant of its own; what
randomization would add is a reason to read the compared means as
assignment effects, which Section 7 takes up.

The Chapter 4 cautions travel with the identity. An omnibus $F$ says that
the three means are not all equal under the stated reference distribution;
it does not say which differ, and any comparison chosen after looking at
this result still needs the family calibration of Chapter 4. The next
section gives the calculation that a single declared comparison requires.

## Contrasts and Adjusted Group Comparisons {#sec-ch9-contrasts}

### A contrast needs coefficient covariance

A **contrast** is a declared linear combination of coefficients,
$C=c'\boldsymbol\beta$, chosen before the comparison is read. Its estimate
is the same combination of the fitted coefficients, and its variance is
the corresponding quadratic form in the coefficient covariance matrix:

$$
\widehat C=c'\widehat{\boldsymbol\beta},\qquad
\widehat{\operatorname{Var}}(\widehat C)=c'\widehat{\mathbf V}c,\qquad
SE(\widehat C)=\sqrt{c'\widehat{\mathbf V}c}.
$$

Estimated coefficients are correlated, because the predictors they belong
to are correlated. The quadratic form is exactly the bookkeeping that
tracks those correlations: for a difference of two coefficients it gives
$\operatorname{Var}(b_1)+\operatorname{Var}(b_2)-2\operatorname{Cov}(b_1,b_2)$.
Take professional minus white-collar in `m2`, with the coefficient order
intercept, education, prestige, women, professional, white-collar, so that
$c=(0,0,0,0,1,-1)'$.

```{=latex}
\Needspace{20\baselineskip}
```

```{r}
#| label: ch9-contrast
#| echo: true
c_vec <- c(0, 0, 0, 0, 1, -1)
V <- vcov(m2)
C_hat <- sum(c_vec * coef(m2))
C_var <- drop(c_vec %*% V %*% c_vec)
c(estimate = C_hat, variance = C_var, SE = sqrt(C_var))
C_hat + c(-1, 1) * qt(.975, df.residual(m2)) * sqrt(C_var)
```

The estimate is $0.3242-0.2353=0.0889$ thousand dollars. Its variance
assembles the two coefficient variances and their covariance,

$$
2.1406+0.9698-2(0.9720)=1.1664,
$$

so the SE is 1.0800 and, with $t_{.975,92}=1.9861$, the interval is
$[-2.0561,\ 2.2338]$.

Two wrong calculations are worth naming because both are common. Adding
the two standard errors gives $1.4631+0.9848=2.4479$, which is not the
standard error of anything here; it is an upper bound that the positive
covariance between the two coefficients makes badly loose. Adding the two
variances and discarding the covariance gives 3.1104 and an SE of 1.7636,
also wrong. Only the quadratic form is the variance of the declared
comparison.

The same machinery covers any declared comparison, because $c$ is simply
the recipe written as a vector. Setting one entry to 1 and the rest to 0
returns an ordinary coefficient and its usual SE. Setting the two type
entries to $1/2$ each compares the average of the two named types with
the blue-collar reference. Whenever the non-intercept entries sum to
zero, the combination is a contrast in the Chapter 4 sense and does not
depend on which level is the reference; the comparison above is such a
combination, which is why Section 1.3 found the same 0.0889 as the
professional coefficient of the releveled fit. What never changes is that
the variance is the quadratic form, computed from the same
$\widehat{\mathbf V}$ that `vcov()` returns.

Declaring the comparison in advance matters as much as computing it
correctly. Chapter 4's family calibration applies to comparisons chosen
after seeing an omnibus result, and a contrast selected because it looked
large in the fitted coefficients is exactly such a choice. The calculation
in this subsection assumes one comparison named before the model was read.

Read the interval for what it is. It is a working-model benchmark for an
adjusted comparison: professional against white-collar occupations at the
same education, prestige, and percentage of women. It is not a
difference of observed group means, and its width does not overcome the
limited cross-type support that Chapter 8 documented. An interval can be
arithmetically correct and still describe a comparison that the records
barely observe.

### Adjusted block tests

A contrast tests one declared combination. A **block test** asks whether a
set of coefficients is jointly zero, and it is the nested-model comparison
of Section 1.4 with a different reduced model. Two blocks are worth
testing here, and they are different questions.

```{=latex}
\Needspace{12\baselineskip}
```

```{r}
#| label: ch9-blocks
#| echo: true
no_type <- lm(income_k ~ education + prestige + women, data = d)
women_type <- anova(m1, m2)
type_after <- anova(no_type, m2)
```

Adding the percentage of women and the two type indicators to `m1` tests
three restrictions against the full model's 92 residual degrees of
freedom. Dropping type from the full model tests two restrictions against
the same denominator.

\Needspace{10\baselineskip}

| Added | df | $F$ | $p$ |
|:--|--:|--:|--:|
| Women and type | 3 | 11.5088 | $<.0001$ |
| Type after women | 2 | 0.0319 | .9686 |

: Two declared adjusted block comparisons. {#tbl-ch09-blocks tbl-colwidths="[40,15,20,25]"}

The first block is strongly supported: $F(3,92)=11.5088$ against a 5%
critical value of 2.7036. The second is not: $F(2,92)=0.0319$ with
$p=.9686$ against a critical value of 3.0954.

Neither result says what it is often read as saying. Joint evidence for
the first block does not establish that each of the three added columns
contributes; the percentage of women could be carrying the whole block.
And the second result is not proof that occupational type is irrelevant.
It says that, once education, prestige, and the percentage of women are in
the equation, the two type indicators add no detectable reduction in
residual sum of squares *on these 98 records*, where the types barely
overlap on the other predictors. Absence of detectable joint contribution
is not evidence of no difference.

Two further distinctions matter. `anova(m2)` on a single model gives
**sequential** sums of squares in formula order: each line tests a term
given only the terms above it, so the answers depend on the order in which
the formula was written. Those lines are not order-free tests of every
term, and only the last one coincides with the corresponding block test.
Separately, Welch-type and robust-covariance tests are different
procedures with different reference distributions, not variants of this
formula; Section 5 gives the robust version of exactly this type block.

### Adjusted means under a common slope: the Baumann reference {#ch09-ancova}

Chapter 6 introduced the analysis of covariance on a separate invented
twelve-pupil example with a pretest: a group comparison made at a common
baseline value under a common slope. The real-record reference for that
idea belongs here, because it is the same indicator-and-contrast
machinery with one continuous covariate.

Baumann and colleagues studied 68 fourth-grade pupils in one rural school
under three reading-instruction conditions [@baumann1992]. The
`carData::Baumann` distribution holds the 66 pupils with a recorded first
posttest, 22 per group. The teaching analysis uses the first pretest and
first posttest only, and relabels the package's `Basal`, `DRTA`, and
`Strat` as DRA, DRTA, and Think-aloud, with DRA as the reference. The
model assumes one common pretest slope across groups:

$$
Y_{ij}=\mu_j+\beta(X_{ij}-\bar X)+\varepsilon_{ij},\qquad
\bar Y_j^{\,\mathrm{adj}}=\bar Y_j-\widehat\beta(\bar X_j-\bar X).
$$

The adjusted mean is the observed group mean moved along the common slope
by the group's pretest advantage or deficit relative to the grand pretest
mean.

```{=latex}
\Needspace{20\baselineskip}
```

```{r}
#| label: ch9-baumann
#| echo: true
reading <- carData::Baumann
reading$group <- factor(reading$group,
  levels = c("Basal", "DRTA", "Strat"),
  labels = c("DRA", "DRTA", "Think-aloud"))
ancova <- lm(post.test.1 ~ group + pretest.1, data = reading)
summary(ancova)$coefficients
reading_means <- tapply(reading$post.test.1, reading$group, mean)
pretest_means <- tapply(reading$pretest.1, reading$group, mean)
adjusted <- reading_means - coef(ancova)["pretest.1"] *
  (pretest_means - mean(reading$pretest.1))
adjusted
varying_slopes <- lm(post.test.1 ~ group * pretest.1, data = reading)
anova(ancova, varying_slopes)
```

The common pretest slope is 0.6932 with SE 0.1015. Against DRA, the DRTA
coefficient is 3.6266 (SE 0.7362) and the Think-aloud coefficient is
2.0362 (SE 0.7450); the residual mean square is 5.8940 on 62 degrees of
freedom. The group pretest means are 10.5000, 9.7273, and 9.1364 against a
grand mean of 9.7879, so DRA starts above and Think-aloud below the common
reference point.

Adjustment moves the raw posttest means 6.6818, 9.7727, and 7.7727 to
6.1882, 9.8147, and 8.2243. The DRA mean falls because that group's
pretest advantage of 0.7121 points is removed at the common slope; the
Think-aloud mean rises for the mirror-image reason. The adjusted values
answer a question the raw means do not: what these groups' posttest means
would be if each had the grand pretest mean, *under the common-slope
model*.

Two conditions carry that reading, and both need saying. The adjusted
means need a **common reference** — here the grand pretest mean — and an
**averaging population** over which the comparison is made. Averaging the
fitted model over the common empirical pretest distribution returns the
same three values, because the model is linear in the pretest and the same
distribution is used for every group; that equality is why the simple
shift formula suffices.

The common-slope condition is checkable. Comparing the additive model with
the model that lets each group have its own pretest slope gives
$F(2,60)=1.5054$, $p=.2302$: on these records, the varying-slope model
does not improve the fit detectably. That is a check, not a
demonstration that the slopes are equal, and it is where this chapter
stops. Chapter 10 develops the varying-slope model itself as an
interaction, with conditional slopes and their uncertainty.

## Compare Nested Models {#sec-ch9-comparison}

### The extra-sum-of-squares statistic

Sections 1.4 and 2.2 were two instances of one statistic. Let the reduced
model be obtained from the full model by setting $q$ coefficients to zero,
with both models fitted to the same records and the same outcome. Then

$$
F=\frac{(RSS_{R}-RSS_{F})/q}{RSS_{F}/(n-p)},\qquad
F_{\mathrm{overall}}=\frac{RegSS/(p-1)}{RSS/(n-p)},\quad
RegSS=\sum_i(\widehat y_i-\bar y)^2.
$$

The numerator is the extra sum of squares the added block explains, per
added coefficient. The denominator is the full model's residual mean
square, the same $s^2$ that appears in every coefficient standard error.
Under the normal linear model with independent common-variance errors, the
ratio has an $F_{q,\,n-p}$ reference.

Both earlier tests are instances. The one-way identity of Section 1.4 sets
$q=2$ and $n-p=95$, with $RSS_R=1734.0792$ from the intercept-only model
and $RSS_F=1138.1230$ from the group-only model, giving
$(595.9562/2)/(1138.1230/95)=24.8725$. The adjusted type block of Section
2.2 sets $q=2$ and $n-p=92$, with $RSS_R=631.3272$ from the model without
type and $RSS_F=630.8897$ from `m2`, giving
$(0.4375/2)/(630.8897/92)=0.0319$. The women-and-type block sets $q=3$
with $RSS_R=867.6537$, giving $(236.7640/3)/6.8575=11.5088$. The same
three ingredients appear every time: how much residual sum of squares the
block buys, how many coefficients it spends, and what a unit of unexplained
variation costs in the fuller model.

Notice what must be held fixed for the subtraction to mean anything. The
two models must use the same rows and the same outcome, or the difference
of residual sums of squares is not a reduction attributable to the added
block; and the reduced model must be obtainable from the fuller one by
restrictions, or there is no null to test. Chapter 8's complete-case
selection is why every comparison in this chapter runs on the same 98
records.

The **overall** regression test is the special case in which the reduced
model is the intercept alone, so every non-intercept coefficient is
restricted to zero. It is the line R prints at the foot of `summary()`.

```{=latex}
\Needspace{12\baselineskip}
```

```{r}
#| label: ch9-overall-f
#| echo: true
summary(m2)$fstatistic
summary(m0)$fstatistic
```

The full model gives $F(5,92)=32.1747$; the education-only model on the
same 98 records gives $F(1,96)=47.1956$. Two remarks keep these in
proportion. They test different joint restrictions in different models, so
the larger value for `m0` is not evidence that `m0` is the better model.
And neither statistic verifies the mean function, the error model, or the
independence of the rows; a large overall $F$ is compatible with a badly
misspecified equation. Nor does either say how a fitted rule would perform
on records it has not seen, which is a separate question with a separate
calculation (Chapter 13).

### Restrictions in matrix form

The same statistic has a compact matrix form, and the matrix notation is
worth one pass because it makes clear what "the same records" is doing in
every comparison above. Let $\mathbf X$ hold $n$ rows and $p$ coefficient
columns including the intercept, with full column rank. Minimizing
$(\mathbf y-\mathbf X\boldsymbol\beta)'(\mathbf y-\mathbf X\boldsymbol\beta)$
gives the normal equations
$\mathbf X'(\mathbf y-\mathbf X\widehat{\boldsymbol\beta})=\mathbf 0$ and
hence
$\widehat{\boldsymbol\beta}=(\mathbf X'\mathbf X)^{-1}\mathbf X'\mathbf y$,
$\widehat{\mathbf y}=\mathbf H\mathbf y$ with
$\mathbf H=\mathbf X(\mathbf X'\mathbf X)^{-1}\mathbf X'$, and
$\widehat{\mathbf V}=s^2(\mathbf X'\mathbf X)^{-1}$. For $q$ independent
linear restrictions $\mathbf R\boldsymbol\beta=\mathbf r$,

$$
F=\frac{(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)'\,
[\mathbf R(\mathbf X'\mathbf X)^{-1}\mathbf R']^{-1}\,
(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)}{q\,s^{2}}\sim F_{q,\,n-p}.
$$

Choosing the rows of $\mathbf R$ that select the two type coefficients,
with $\mathbf r=\mathbf 0$, reproduces the adjusted type block test of
Section 2.2 without fitting a second model.

```{=latex}
\Needspace{24\baselineskip}
```

```{r}
#| label: ch9-matrix
#| echo: true
X <- model.matrix(m2)
beta_hat <- solve(crossprod(X), crossprod(X, d$income_k))
c(columns = ncol(X), rank = qr(X)$rank,
  max_coefficient_difference = max(abs(beta_hat - coef(m2))),
  max_normal_equation_error = max(abs(crossprod(X, residuals(m2)))))
R <- rbind(c(0, 0, 0, 0, 1, 0), c(0, 0, 0, 0, 0, 1))
Rb <- R %*% coef(m2)
drop(t(Rb) %*% solve(R %*% vcov(m2) %*% t(R)) %*% Rb) / nrow(R)
```

The model matrix has six columns and rank 6, so the two indicators plus
the intercept are not collinear. Solving the normal equations directly
reproduces `lm()`'s coefficients to within about $10^{-13}$, and the
residuals are orthogonal to every column of $\mathbf X$ to the same
precision. The restriction statistic returns 0.0319, the value the
`anova()` comparison gave.

Two practical points follow. `lm()` does not form
$(\mathbf X'\mathbf X)^{-1}$; it uses a QR decomposition, which is
numerically better behaved. The displayed matrix equation explains what
the estimator *is*, not how R computes it [@fox2016; @fox2019]. And
because $\widehat{\mathbf y}=\mathbf H\mathbf y$ is a projection onto the
column space of $\mathbf X$, any full-rank recoding of a factor spans the
same space and leaves the fitted vector alone — which is Section 1.3's
releveling result stated geometrically. Appendix A carries the derivation.

### A local Normal likelihood and information criteria

Nested $F$ tests compare two models when one contains the other.
Information criteria rank models that need not be nested, using a
different quantity: the maximized likelihood, penalized for the number of
fitted parameters. Chapter 6 built a likelihood once, for 32 successes in
50 binomial trials, by treating the probability of the observed data as a
function of the unknown parameter. The linear-model version is the same
idea with a continuous density.

Treat the predictor rows as fixed and the outcomes as independent Normal
draws with mean $\mathbf x_i'\boldsymbol\beta$ and SD $\sigma$. The
log-likelihood, and its value at the maximizing arguments, are

$$
\ell(\boldsymbol\beta,\sigma)=-n\log\sigma-\frac n2\log(2\pi)
-\frac{1}{2\sigma^2}\sum_i(y_i-\mathbf x_i'\boldsymbol\beta)^2,\qquad
\ell(\widehat{\boldsymbol\beta},\widehat\sigma_{\mathrm{ML}})
=-\frac n2\left\{\log(2\pi)+1+\log(RSS/n)\right\}.
$$

Only the final term of the first expression involves $\boldsymbol\beta$,
and it is $-RSS/(2\sigma^2)$. Maximizing over the coefficients therefore
*minimizes* the residual sum of squares: least squares and maximum
likelihood give the same coefficients under this error model. They do not
give the same scale estimate. The maximizing $\sigma$ is
$\sqrt{RSS/n}=2.5373$, while the residual SD reported by `lm()` is
$\sqrt{RSS/(n-p)}=2.6187$, which divides by the residual degrees of
freedom instead. The criteria use the first.

$$
AIC=-2\log\widehat L+2k,\qquad BIC=-2\log\widehat L+k\log n.
$$

Here $k$ counts the fitted coefficients **and** the residual variance, so
`m2` has $k=7$, not 6.

```{=latex}
\Needspace{22\baselineskip}
```

```{r}
#| label: ch9-likelihood
#| echo: true
ll <- logLik(m2)
sigma_ml <- sqrt(mean(residuals(m2)^2))
manual_ll <- sum(dnorm(d$income_k, mean = fitted(m2), sd = sigma_ml,
  log = TRUE))
c(sigma_ML = sigma_ml, residual_SD = sigma(m2),
  logLik = as.numeric(ll), manual = manual_ll, k = attr(ll, "df"),
  AIC = AIC(m2), BIC = BIC(m2))
```

`logLik(m2)` returns $-230.3020$, and summing the Normal log-densities of
the 98 observed incomes at the fitted means and the ML scale returns the
same number. The criteria follow arithmetically:
$AIC=2(230.3020)+2(7)=474.6040$ and $BIC=460.6040+7\log 98=492.6988$, with
$\log 98=4.5850$.

\Needspace{12\baselineskip}

| Model | $k$ | $R^2$ | Adjusted $R^2$ | AIC | BIC |
|:--|--:|--:|--:|--:|--:|
| `m0` | 3 | .3296 | .3226 | 526.5052 | 534.2601 |
| `m1` | 4 | .4996 | .4891 | 499.8328 | 510.1727 |
| `m2` | 7 | .6362 | .6164 | 474.6040 | 492.6988 |

: In-sample summaries for comparable income models. {#tbl-ch09-criteria tbl-colwidths="[16,10,16,20,19,19]"}

Smaller is preferred under either criterion, and both prefer `m2` here.
The conditions for reading the table are strict and easy to breach. The
three models must be fitted to the same records, with the same outcome on
the same scale, under the same likelihood convention; Section 4.2 shows
what happens when the last of these fails. BIC's heavier penalty,
$k\log n$ against AIC's $2k$, means the two can disagree, and a
disagreement is information about how strongly the extra parameters are
being charged, not a tie to be broken by preference.

Above all, neither criterion is a held-out error. Both are calculated on
the same records that fitted the model, and both are comparative rather
than absolute: a smaller AIC says one candidate is preferred to another,
not that either describes the records well [@akaike1974; @schwarz1978].
Chapter 11 develops likelihood properly, for models where least squares
is no longer available; this section needs only the local Normal case.

## Change the Form of a Predictor or the Scale of the Outcome {#sec-ch9-forms}

### A polynomial term

Chapter 8, Section 4.3, refitted the full model with a squared prestige
term as a disclosed sensitivity check on functional form. Here that refit
is the worked example, with its block test, its criteria, and the reading
a curved term forces.

A squared term is an ordinary extra column, `I(prestige^2)`, and the
model is still linear in its coefficients. What changes is the reading:

$$
E(Y\mid x,\ldots)=\beta_0+\beta_1x+\beta_2x^2+\cdots,\qquad
\frac{\partial E(Y)}{\partial x}=\beta_1+2\beta_2x,\qquad
x^{*}=-\frac{\beta_1}{2\beta_2}.
$$

The predictor no longer has a single slope. It has a slope that depends on
its own value, and a turning point where that slope is zero.

```{=latex}
\Needspace{30\baselineskip}
```

```{r}
#| label: ch9-quadratic
#| echo: true
m2q <- lm(income_k ~ education + prestige + I(prestige^2) + women + type,
  data = d)
coef(summary(m2q))[c("prestige", "I(prestige^2)"), ]
anova(m2, m2q)
c(AIC = AIC(m2q), BIC = BIC(m2q),
  turning_point = -coef(m2q)["prestige"] / (2 * coef(m2q)["I(prestige^2)"]))
d$prestige_c <- d$prestige - mean(d$prestige)
m2qc <- lm(income_k ~ education + prestige_c + I(prestige_c^2) + women +
  type, data = d)
c(cor(d$prestige, d$prestige^2), cor(d$prestige_c, d$prestige_c^2),
  max(abs(fitted(m2q) - fitted(m2qc))))
```

The quadratic coefficient is 0.0024 with SE 0.0009 and $t=2.5954$; the
one-restriction block test gives $F(1,91)=6.7360$ with $p=.0110$; and the
criteria move from 474.6040 and 492.6988 to 469.6058 and 490.2855. Adding
the term also changes the education coefficient from 0.1164 to $-0.0298$,
a reminder that coefficients in an additive model are defined relative to
the other columns present.

Read the curve by evaluating its slope. With $\beta_1=-0.0754$ and
$\beta_2=0.0024$, the fitted slope $-0.0754+2(0.0024)x$ is 0.0697 at
prestige 30, 0.1665 at 50, and 0.2633 at 70: income rises with prestige
throughout, and rises faster at the top. The turning point 15.5906 lies
below the smallest observed prestige of 17.3, so the descending branch of
the parabola is entirely outside the records. Reporting a "negative
prestige coefficient" from this model would be a misreading of the
coordinate $-0.0754$, which is the slope at prestige zero, a value no
occupation has.

Centering repairs a second coordinate problem. Prestige and its square
correlate .9849, which inflates both coefficients' standard errors;
centering prestige before squaring lowers that correlation to .3216 and
changes the linear coefficient to 0.1536, now the slope at the mean
prestige. The two fits have identical fitted values to about $10^{-14}$.
Centering changes the coordinates in which the curve is described; it does
not change the curve.

Finally, keep the status of the exercise straight. This is a disclosed
sensitivity check on functional form, declared before the refit, and its
result is that a modest curvature is detectable on these records. It is
not a search across candidate terms for the smallest $p$-value, and a
$p$-value read after such a search would not mean what it appears to mean.

### A log-transformed outcome

Changing a predictor's form keeps the question. Changing the outcome's
scale changes the question, and that difference is the point of this
subsection.

$$
\log Y_i=\beta_0+\sum_j\beta_jx_{ij}+\varepsilon_i,\qquad
100\,(e^{\beta_j}-1)\ \text{percent per unit of } x_j .
$$

A coefficient in a log-income model is an additive difference in expected
log income. Exponentiating turns it into a multiplicative difference in
geometric mean income under that model, and
$100(e^{\widehat\beta}-1)$ expresses that multiplier as a percentage. The
geometric mean is not the arithmetic mean, so this is not automatically a
statement about average income in dollars.

```{=latex}
\Needspace{18\baselineskip}
```

```{r}
#| label: ch9-log-outcome
#| echo: true
m_log <- lm(log(income) ~ education + prestige + women + type, data = d)
terms <- c("education", "prestige", "women")
log_ci <- confint(m_log)[terms, ]
data.frame(term = terms, log_coefficient = coef(m_log)[terms],
  percent = 100 * (exp(coef(m_log)[terms]) - 1),
  lower = 100 * (exp(log_ci[, 1]) - 1),
  upper = 100 * (exp(log_ci[, 2]) - 1), row.names = NULL)
```

\Needspace{12\baselineskip}

| Term | Log coefficient | Percent | Lower | Upper |
|:--|--:|--:|--:|--:|
| education | 0.0076 | 0.7661 | $-4.2041$ | 5.9941 |
| prestige | 0.0207 | 2.0965 | 1.4351 | 2.7622 |
| women | $-0.0085$ | $-0.8430$ | $-1.0189$ | $-0.6667$ |

: Log-model contrasts on a geometric-mean percent scale. {#tbl-ch09-log tbl-colwidths="[22,22,18,19,19]"}

One more prestige point is associated with about 2.0965 percent higher
geometric mean income at the same education, percentage of women, and
type, with interval $[1.4351,\ 2.7622]$ percent. One more percentage point
of women is associated with about $-0.8430$ percent. The two type
coefficients, $-0.0137$ and 0.0708 on the log scale, are $-1.3584$ and
7.3341 percent. The intervals are the classical log-scale intervals with
both endpoints transformed, which is legitimate because the
transformation is monotone; they inherit the log-scale working model and
supply no new warrant.

The criteria will not travel across this change. This model's AIC is
8.0910 against `m2`'s 474.6040, and the difference says nothing at all:
the two likelihoods are densities of different response variables, so
their reference measures differ and their values are not on a common
scale. A valid comparison across a transformed outcome needs the Jacobian
adjustment that converts one density to the other. The simpler discipline
is the one to keep: fix the outcome and the question inside any comparison,
and treat a scale change as a new question rather than as data cleaning
[@fox2019].

### Coefficient scales: standardized coefficients and unique contributions

Two summaries are often asked for once several predictors are in an
equation, and both are more limited than their names suggest.

$$
\beta_j^{*}=\widehat\beta_j\frac{s_{x_j}}{s_y},\qquad
\Delta R^2_j=R^2_{\mathrm{full}}-R^2_{\mathrm{without}\ j}.
$$

A **standardized coefficient**, or beta weight, rescales a partial slope
into outcome-SD units per predictor SD. A **drop-one $R^2$** reports how
much of the explained variation disappears when one predictor is removed
from the full model.

```{=latex}
\Needspace{14\baselineskip}
```

```{r}
#| label: ch9-coefficient-scales
#| echo: true
coef(m2)[2:4] * sapply(d[c("education", "prestige", "women")], sd) /
  sd(d$income_k)
summary(m2)$r.squared -
  summary(lm(income_k ~ prestige + women + type, data = d))$r.squared
```

The beta weights are 0.0757, 0.5685, and $-0.3971$ for education,
prestige, and the percentage of women. Education's drop-one $R^2$ is
.0007: removing it from the full model costs essentially nothing in fit,
which is the same message the education coefficient's wide interval
carried in Chapter 8.

Both quantities are rescalings of an in-sample fit, and neither measures
importance. A standardized coefficient depends on the predictor's observed
spread in these particular records, so a restricted range shrinks it for
reasons that have nothing to do with the relationship. It also differs
from the standardization over named profiles that Chapter 8 used, which
averages fitted values over a declared distribution rather than dividing
by an SD. The drop-one change describes only the *unique* contribution:
because predictor information overlaps, the separate drop-one values need
not sum to the joint $R^2$, and a predictor that duplicates another can
show a near-zero value while mattering. Neither number establishes a
causal ranking or predicts held-out performance.

## Robust Uncertainty Under Unequal Spread {#sec-ch9-hc3}

### The HC3 covariance

Chapter 7 used a robust standard error once, on the simple regression
line, where the education slope's SE moved from 0.1270 to 0.1491. The
estimator behind that number is worth stating, because the rest of this
chapter turns on what it does and does not change.

The classical covariance $s^2(\mathbf X'\mathbf X)^{-1}$ assumes one
common error variance for every row. The heteroskedasticity-consistent
family replaces the single $s^2$ by each row's own squared residual,
sandwiched between two copies of the design component:

$$
\widehat{\mathbf V}_{\mathrm{HC3}}=(\mathbf X'\mathbf X)^{-1}
\left\{\sum_i \mathbf x_i\mathbf x_i'
\frac{e_i^2}{(1-h_{ii})^2}\right\}(\mathbf X'\mathbf X)^{-1}.
$$

The outer factors are the design component; the inner sum is the
residual-scale component. Dividing by $(1-h_{ii})^2$ inflates the
contribution of high-leverage rows, which is why HC3 is the usual
small-sample recommendation in this family [@white1980; @mackinnon1985;
@long2000; @zeileis2004].

```{=latex}
\Needspace{16\baselineskip}
```

```{r}
#| label: ch9-hc3
#| echo: true
V_hc3 <- sandwich::vcovHC(m2, type = "HC3")
cbind(classical = sqrt(diag(vcov(m2))), HC3 = sqrt(diag(V_hc3)))
coef(m2)["education"] + c(-1, 1) * qt(.975, 92) *
  sqrt(V_hc3["education", "education"])
```

\Needspace{14\baselineskip}

| Coefficient | Estimate | Classical SE | HC3 SE |
|:--|--:|--:|--:|
| Intercept | 0.4205 | 1.9573 | 2.7741 |
| education | 0.1164 | 0.2752 | 0.3823 |
| prestige | 0.1406 | 0.0354 | 0.0387 |
| women | $-0.0535$ | 0.0097 | 0.0079 |
| prof | 0.3242 | 1.4631 | 2.0928 |
| wc | 0.2353 | 0.9848 | 1.0548 |

: Classical and HC3 standard errors for the full occupation model. {#tbl-ch09-hc3 tbl-colwidths="[28,24,24,24]"}

Read the table by column and then by row. The estimate column is
unchanged: HC3 replaces a covariance estimate, not a fit, so every
coefficient, fitted value, and residual is exactly what it was. The
education SE grows from 0.2752 to 0.3823, and its 95% interval on the
robust covariance is $[-0.6429,\ 0.8758]$, wider than the classical one
but leading to the same reading of an imprecisely estimated comparison.

The women row is the instructive one: its SE *falls* from 0.0097 to
0.0079. A robust covariance is not a penalty applied to every coefficient.
It is a different estimate of the same target, and where the classical
common-variance assumption happened to overstate a coefficient's
variability, the robust version is smaller. Reporting only the
coefficients whose intervals widened, or switching estimators after seeing
which choice gives the preferred conclusion, turns a stated method into a
selection.

### A block test with a robust covariance

The restriction statistic of Section 3.2 needs only a coefficient vector
and a covariance matrix. Substituting
$\widehat{\mathbf V}_{\mathrm{HC3}}$ for $\widehat{\mathbf V}$ gives an
approximate Wald version of the same block test.

```{=latex}
\Needspace{20\baselineskip}
```

```{r}
#| label: ch9-hc3-block
#| echo: true
wald_F <- function(fit, V, which) {
  b <- coef(fit)[which]
  drop(t(b) %*% solve(V[which, which]) %*% b) / length(which)
}
type_cols <- c("typeprof", "typewc")
c(classical = wald_F(m2, vcov(m2), type_cols),
  HC3 = wald_F(m2, V_hc3, type_cols))
pf(wald_F(m2, V_hc3, type_cols), 2, 92, lower.tail = FALSE)
```

The type block gives 0.0319 with the classical covariance and 0.0345 with
HC3, $p=.9661$ against the classical .9686. Here the substitution changes
nothing of substance, which is the ordinary case when the coefficients are
far from the restriction.

The reference distribution is where the two differ in kind. The classical
statistic has an exact $F_{2,92}$ distribution under the stated normal
linear model. The robust version has only an approximate reference,
justified as the sample grows, and $F$ with the same degrees of freedom is
a convention rather than a derivation. Quoting the robust $p$-value to
several digits, or treating it as the same object as the classical one
because both are labelled $F$, overstates what the calculation delivers.
Section 2.2's classical test and this one answer the same question with
different warrants.

### What HC3 does not repair

HC3 has one job. Under a maintained linear conditional mean and
independent observations, it estimates the coefficient covariance without
assuming a common error variance. That is all.

It does not repair curvature: if the mean function is wrong, the
coefficients estimate a different quantity and a robust covariance
describes that different quantity's variability more honestly, without
making it the quantity of interest. It does not repair dependence: rows
that are not independent break the estimator's own derivation, and Section
7 shows a case where a robust row-level SE and a naive one agree closely
while both understate the uncertainty by nearly half. It does not repair
measurement error, selection, or missing outcomes, which change what the
records represent. It does not supply causal identification, which comes
from design or from an argument about the conditioning set. And it does
not create joint support: the type contrast's robust SE of 1.1827 is still
the uncertainty of a comparison the records barely observe.

The practical response is to say which estimator was used and why, and to
keep the answer's limits attached to it. Chapter 8's advice holds here:
report the named comparison, its form and influence sensitivities, and the
covariance calculation, rather than treating any one of them as a pass/fail
certificate.

## Resample the Records {#sec-ch9-bootstrap}

### Why resample, and which scheme

Every uncertainty statement so far came from a formula: a coefficient
covariance matrix, classical or robust, propagated through a linear
combination. Resampling offers a different approximation. Refit the model
to many resamples of the records, and read the spread of the refitted
estimates directly [@efron1979; @efron1993; @davison1997]. Where the
formula and the empirical resampling distribution disagree, the
disagreement is itself informative.

The choice of what to resample is an assumption, not a default.

- **Case resampling** draws whole records with replacement and refits. It
  conditions on nothing and inherits whatever unequal spread the rows
  contain, but it requires the resampling unit to match the sampling unit,
  with units independent and exchangeable.
- **Residual resampling** holds the design fixed, resamples rescaled
  residuals, and adds them back to the fitted values. It assumes
  identically distributed errors — the assumption HC3 exists to relax.
- **The wild bootstrap** also holds the design fixed but multiplies each
  rescaled residual by an independent mean-zero, unit-variance draw, so
  every record keeps its own error scale. It is the residual scheme
  repaired for unequal variance.

This chapter uses case resampling, for two stated reasons. The occupation
predictors were not designed: no one assigned education or prestige to an
occupation, so a joint-sampling view of rows is the natural one. And the
residual spread differs visibly by type, so the identically-distributed
errors behind residual resampling are not available [@freedman1981].
Appendix B carries out the other two schemes on the same targets for
comparison.

### Case resampling for the full model

Resample the 98 rows with replacement, refit `m2`, and record three
targets: the education slope, the professional-minus-white-collar
contrast, and the professional coefficient. The seed and the number of
resamples $B$ belong in the record, so both appear in the code; $B$ counts
bootstrap resamples here, as it counted simulation repetitions in
Chapter 5.

```{=latex}
\Needspace{40\baselineskip}
```

```{r}
#| label: ch9-bootstrap
#| echo: true
set.seed(90909)
B <- 2000
coef_names <- names(coef(m2))
draws <- matrix(NA_real_, B, 3,
  dimnames = list(NULL, c("education", "prof_minus_wc", "prof")))
for (b in seq_len(B)) {
  rows <- sample.int(nrow(d), replace = TRUE)
  fit_b <- lm(income_k ~ education + prestige + women + type,
    data = d[rows, ])
  b_hat <- coef(fit_b)
  if (all(coef_names %in% names(b_hat)) && !anyNA(b_hat[coef_names])) {
    draws[b, ] <- c(b_hat["education"], b_hat["typeprof"] - b_hat["typewc"],
      b_hat["typeprof"])
  }
}
colSums(is.finite(draws))
rbind(estimate = c(coef(m2)["education"], C_hat, coef(m2)["typeprof"]),
  classical_SE = c(sqrt(V["education", "education"]), sqrt(C_var),
    sqrt(V["typeprof", "typeprof"])),
  HC3_SE = c(sqrt(V_hc3["education", "education"]),
    sqrt(drop(c_vec %*% V_hc3 %*% c_vec)),
    sqrt(V_hc3["typeprof", "typeprof"])),
  bootstrap_SE = apply(draws, 2, sd, na.rm = TRUE),
  lower = apply(draws, 2, quantile, .025, na.rm = TRUE, names = FALSE),
  upper = apply(draws, 2, quantile, .975, na.rm = TRUE, names = FALSE))
```

The printed matrix carries one blank column name: the middle target is a
contrast rather than a coefficient, so `C_hat` arrives without a name. Its
column is the professional-minus-white-collar comparison of Section 2.1,
and the table below labels all three targets explicitly.

The guard inside the loop matters. A resample could omit an entire
occupational type, and the refit would then drop a coefficient; such a
draw must be recorded as missing rather than silently misaligned. Here all
2,000 resamples contained all three types, so all 2,000 refits are usable,
and that count is part of the result rather than an assumption.

\Needspace{12\baselineskip}

| Target | Estimate | Classical SE | HC3 SE | Bootstrap SE | Percentile 2.5% | Percentile 97.5% |
|:--|--:|--:|--:|--:|--:|--:|
| Education slope | 0.1164 | 0.2752 | 0.3823 | 0.3699 | $-0.6830$ | 0.7440 |
| prof $-$ wc | 0.0889 | 1.0800 | 1.1827 | 1.1187 | $-1.9613$ | 2.4313 |
| prof coefficient | 0.3242 | 1.4631 | 2.0928 | 1.9926 | $-3.1416$ | 4.5443 |

: Model-based, HC3, and case-bootstrap uncertainty for three targets in the full occupation model (seed 90909, $B=2000$). {#tbl-ch09-bootstrap tbl-colwidths="[22,12,14,12,14,13,13]"}

Read the table one row at a time. The education slope's bootstrap SE of
0.3699 sits between the classical 0.2752 and the HC3 0.3823, with
percentile interval $[-0.6830,\ 0.7440]$: three calculations, one
conclusion that this comparison is imprecise. The type contrast's
bootstrap SE of 1.1187 exceeds the classical 1.0800, with percentile
interval $[-1.9613,\ 2.4313]$. The professional coefficient shows the
largest movement, from a classical 1.4631 to 1.9926, close to the HC3
2.0928; with only 31 professional occupations, a resample's composition
matters, and the formula that assumes a common error variance is the most
optimistic of the three.

Agreement between HC3 and the bootstrap is not independent confirmation.
Both are sensitivity procedures applied to the same 98 records, and both
inherit the same limitations of those records; agreement means the two
approximations behave similarly here, not that the answer has been
replicated. Monte Carlo variation also means that the last decimal of a
bootstrap summary carries no weight: a different seed would move it, which
is exactly why the seed and $B=2000$ are reported.

### Which interval, and what resampling does not repair

Two intervals come directly from the resampling distribution. The
**percentile interval** takes its 2.5th and 97.5th quantiles, as in the
table above. The **normal-approximation interval** takes the original
estimate plus and minus 1.96 bootstrap standard errors. The two agree when
the resampling distribution is roughly symmetric; here the three
distributions have skewness $-0.3891$, 0.2635, and 0.3785, mild enough
that the endpoints agree closely. For a strongly skewed or biased
resampling distribution, **bootstrap-$t$** intervals, which resample a
studentized quantity, and **BCa** intervals, which correct for bias and
for skewness-driven acceleration, are the higher-order alternatives;
Appendix B defines both in words.

What resampling cannot do is more important than which interval to
choose. Drawing occupations with replacement treats them as independent
and exchangeable draws, which they are not: they share one historical
labour market, and their measurements are aggregates from different years.
Resampling does not repair the thin cross-type support, correct the
measurement problems Chapter 8 documented, or convert an observational
comparison into a causal one. And if a design contained clusters, strata,
or weights, the resampling unit would have to follow the design rather
than the rows — which is the subject of the next section.


## Dependence in the Design: Project STAR {#sec-ch9-star}

```{=latex}
\hypertarget{sec-ch9-star}{}
```

Every calculation so far has used one set of 98 observational records, and
every qualification has been about what those records cannot support. This
section changes the design. Project STAR randomized kindergarten pupils to
class types within their schools, which supplies a warrant no amount of
adjustment can manufacture — and a dependence structure that the
uncertainty calculation must follow.

### A different design supplies a different warrant

The question is specific:

> Among eligible kindergarten pupils in the participating Tennessee
> schools, what was the mean reading-score difference under assignment to
> a small rather than a regular class?

The outcome is `readk`, an end-of-kindergarten reading scale score. The
field `stark` records the kindergarten class-type condition. Reading its
small-versus-regular comparison as an assignment comparison rests on the
documented experimental protocol, not on the column name or on the
`lm()` call: random assignment occurred **within** participating schools
[@finn1990; @mosteller1995]. The regular-with-aide arm is outside this
declared two-arm comparison and is excluded by design, not dropped for
convenience.

Three limits belong with the question from the start. Class **type** is
not a measured headcount, so this analysis estimates no dose response in
pupils per room. The distributed field does not independently verify each
pupil's original randomized assignment; an assignment-effect reading
additionally requires linking the recorded type to the assignment record
and checking switches and implementation deviations, which this teaching
reanalysis does not supply. And the scientific target is the
pupil-weighted mean assignment effect for the eligible cohort in
participating schools, while a common coefficient from a
complete-record model is an estimator that needs further links — a
common effect across schools, or an explicit weighting argument — to
reach that target.

The data are distributed as `AER::STAR`, accompanying *Applied
Econometrics with R* and following the version used by Stock and Watson
[@kleiber2008; @stock2007]. They are loaded from the package rather than
redistributed here, and the analysis is retrospective rather than
preregistered.

### Separate eligibility from missingness

Two different reasons for a record's absence from an analysis are easy to
merge and must not be. Some distributed rows belong to pupils who entered
the study after kindergarten: they have no kindergarten condition at all,
and their absence is **structural ineligibility**. Other rows belong to
eligible kindergarten pupils whose reading score or covariates were not
recorded: that is **missingness**, and it can distort a comparison.

Only variables a calculation actually uses may enter its complete-record
rule. In particular `mathk` must not exclude an otherwise usable reading
record.

```{=latex}
\Needspace{22\baselineskip}
```

```{r}
#| label: ch9-star-prepare
#| echo: true
data("STAR", package = "AER")
star_vars <- c("stark", "readk", "lunchk", "gender", "ethnicity",
  "experiencek", "schoolidk")
star_k_recorded <- !is.na(STAR$stark)
star_two_arm <- star_k_recorded & STAR$stark %in% c("regular", "small")
star_complete <- complete.cases(STAR[, star_vars])
star_df <- droplevels(STAR[star_two_arm & star_complete, star_vars])
star_df$stark <- relevel(star_df$stark, ref = "regular")
star_observed <- droplevels(STAR[star_two_arm & !is.na(STAR$readk), c(
  "stark", "readk", "schoolidk")])
star_observed$stark <- relevel(star_observed$stark, ref = "regular")
c(observed_reading = nrow(star_observed), complete_covariates = nrow(
  star_df))
```

\Needspace{16\baselineskip}

| Stage | $n$ |
|:--|--:|
| Distributed records | 11,598 |
| Kindergarten condition recorded | 6,325 |
| Eligible small or regular | 4,094 |
| Eligible, reading observed | 3,745 |
| Complete primary records | 3,735 |
| Primary regular | 2,001 |
| Primary small | 1,734 |
| Primary schools | 79 |

: STAR eligibility and reading-analysis completeness. {#tbl-ch09-star-sample tbl-colwidths="[60,40]"}

Of 11,598 distributed records, 5,273 have no kindergarten condition; they
are later entrants, not 5,273 failed measurements. The remaining 6,325
have a recorded kindergarten arm, and across all three arms 536 lack a
reading score. Restricting to the two declared arms leaves 4,094 eligible
records, of which 349 lack reading — 188 in the regular arm and 161 in the
small arm — leaving 3,745 with an observed reading score, 2,006 regular
and 1,739 small.

Similar missing percentages in the two arms are reassuring but prove
nothing: the absent scores could differ systematically within either arm,
and no observed value can reveal that. Complete-record inference here
requires an assumption that, given the declared covariates and within arm
and school, observation does not depend on the reading score that would
have been recorded. Stating the assumption is not verifying it.

Among the observed-reading records across all arms, 38 lack at least one
of lunch status, gender, ethnicity, and teacher experience; within the two
focal arms, ten do. The primary common-covariate sample therefore has
3,735 pupils, 2,001 regular and 1,734 small, in 79 schools. The
observed-reading sample keeps all 3,745 focal records and is used as a
declared sensitivity, not as an alternative headline.

### The binary slope recovers the observed arm means

Section 1.1's identity applies unchanged. With $D_i=1$ for the small arm
and 0 for regular,

$$
\widehat Y_i=\widehat\alpha+\widehat\delta D_i,\qquad
\widehat\alpha=\bar Y_{D=0},\qquad
\widehat\delta=\bar Y_{D=1}-\bar Y_{D=0}=440.5474-434.7323=5.8151.
$$

```{=latex}
\Needspace{20\baselineskip}
```

```{r}
#| label: ch9-star-means
#| echo: true
star_observed_unadjusted <- lm(readk ~ stark, data = star_observed)
star_arm_means <- tapply(star_observed$readk, star_observed$stark, mean)
c(regular_mean = unname(star_arm_means["regular"]),
  small_mean = unname(star_arm_means["small"]),
  small_minus_regular = unname(star_arm_means["small"] -
    star_arm_means["regular"]),
  regression_difference = unname(coef(
    star_observed_unadjusted)["starksmall"]))
```

On these 3,745 records the difference is 5.8151 reading points, and the
regression slope reproduces it exactly, as Chapter 7's binary predictor
did on 24 pupils. The identity is algebraic and requires no randomization;
what randomization adds is the reason to read the difference as an
assignment comparison.

Two cautions attach to the number itself. Under independent
common-variance rows, the slope's $t$ test equals the pooled two-sample
test, here $t=5.6039$; that row-level reference ignores the fact that
pupils sit in schools, which the next subsection addresses. And raw arm
means weight each school by its own arm-specific pupil count, so the
pooled difference equals the pupil-weighted cohort target only under
common assignment probabilities and comparable observation patterns, or
after explicit standardization to target school weights.

### Compare two specifications on identical primary rows

The adjusted model adds free-lunch status, gender, ethnicity, teacher
experience, and school indicators. School indicators respect the
assignment blocks; the remaining covariates are the declared adjustment
specification, not a substitute for randomization.

```{=latex}
\Needspace{18\baselineskip}
```

```{r}
#| label: ch9-star-models
#| echo: true
star_unadjusted <- lm(readk ~ stark, data = star_df)
star_primary <- lm(readk ~ stark + lunchk + gender + ethnicity +
  experiencek + schoolidk, data = star_df)
star_sensitivity <- lm(readk ~ stark + schoolidk, data = star_observed)
c(unadjusted = nobs(star_unadjusted), adjusted = nobs(star_primary),
  observed_reading = nobs(star_sensitivity))
```

On the common 3,735 primary rows, the unadjusted difference is 5.8955 and
the adjusted coefficient is 6.6575. Adjustment can improve precision when
the added variables are prognostic for reading, and school indicators
match the blocks in which assignment happened. It is not needed to remove
baseline confounding that a successfully implemented randomization already
prevents, and it does not automatically preserve a causal target: timing,
selection, and the fitted comparison still need scrutiny.

The earlier 5.8151 comes from 3,745 rows and must not serve as the
same-row unadjusted baseline for 6.6575. Comparing 5.8151 with 6.6575
confounds two changes — a change of records and a change of
specification — and the whole discipline of this chapter is to keep those
apart. The same-row pair to compare is 5.8955 against 6.6575.

Similar unadjusted and adjusted coefficients are consistent with
successful randomization. They are not proof that implementation,
missingness, or weighting problems are absent, and a common coefficient
still carries its own implicit weights if the effect varies across schools
or pupil profiles.

### Cluster uncertainty follows the design

Pupils are the outcome units. Schools are the randomization blocks and the
approximately independent units for this covariance calculation; the
schools themselves were not assigned to arms. Pupils sharing a school —
and, within it, a classroom — may have correlated outcomes, and a
covariance estimate that ignores this understates the uncertainty of an
arm comparison.

For school $g$, form its score $\mathbf s_g=\mathbf X_g'\mathbf e_g$ by
summing the residual contributions **within** the school before taking an
outer product. The CR1/HC1 cluster-robust covariance is then

$$
\widehat{\mathbf V}_{\mathrm{CR1}}=c_G(\mathbf X'\mathbf X)^{-1}
\Big(\sum_{g=1}^{G}\mathbf s_g\mathbf s_g'\Big)(\mathbf X'\mathbf X)^{-1},
\qquad \mathbf s_g=\mathbf X_g'\mathbf e_g,\qquad
c_G=\frac{G}{G-1}\frac{n-1}{n-p}.
$$

The within-school cross-products are exactly what distinguishes this from
treating every pupil as independent, and $c_G$ is the finite-cluster
correction. With $G=79$ schools, the interval below uses a $t_{78}$
critical value, 1.9908 [@zeileis2020].

```{=latex}
\Needspace{30\baselineskip}
```

```{r}
#| label: ch9-star-cluster
#| echo: true
star_V <- sandwich::vcovCL(star_primary, cluster = star_df$schoolidk,
  type = "HC1", cadjust = TRUE)
star_b <- unname(coef(star_primary)["starksmall"])
star_se <- sqrt(star_V["starksmall", "starksmall"])
star_G <- nlevels(star_df$schoolidk)
c(estimate = star_b, SE = star_se)
star_b + c(-1, 1) * qt(.975, star_G - 1) * star_se
c(row_level = sqrt(vcov(star_observed_unadjusted)["starksmall", "starksmall"]),
  HC3 = sqrt(sandwich::vcovHC(star_observed_unadjusted,
    type = "HC3")["starksmall", "starksmall"]),
  clustered = sqrt(sandwich::vcovCL(star_observed_unadjusted,
    cluster = star_observed$schoolidk, type = "HC1",
    cadjust = TRUE)["starksmall", "starksmall"]))
```

\Needspace{12\baselineskip}

| Model | $n$ | Estimate | SE | Interval |
|:--|--:|--:|--:|:--|
| Observed: unadjusted | 3,745 | 5.8151 | 1.8476 | [2.1368, 9.4935] |
| Primary: unadjusted | 3,735 | 5.8955 | 1.8491 | [2.2143, 9.5767] |
| Primary: adjusted | 3,735 | 6.6575 | 1.6974 | [3.2782, 10.0368] |
| Observed: school-adjusted | 3,745 | 6.6273 | 1.7797 | [3.0842, 10.1703] |

: Class-type comparisons and school-clustered intervals. {#tbl-ch09-star-estimates tbl-colwidths="[34,12,16,14,24]"}

The table holds four distinct calculations, each with its own records and
its own specification; each row's interval uses the school identifiers
from that model's own rows. Changing the model or its records is a
different act from changing the covariance calculation, and the table
keeps them separable.

The three standard errors printed at the end of the chunk make Section 5's
point concrete. For the same 3,745-record unadjusted comparison, treating
pupils as independent rows gives SE 1.0377; HC3 gives 1.0416, barely
different; school clustering gives 1.8476, nearly twice as large. Unequal
spread across pupils was not this design's problem. Dependence within
schools was, and no row-level robust covariance addresses it. The adjusted
primary model shows the same pattern: its row-level SE is 0.9161 and its
HC3 SE 0.9509, against the clustered 1.6974.

These are cluster-asymptotic approximations with 79 clusters, not exact
randomization inference and not a multilevel model. A residual plot cannot
choose the independence unit — the design does. And a count of 79 clusters
does not certify the approximation: the balance of information across
clusters and the influence of particular schools also matter.

### Read the bounded answer, then evaluate its support

A chapter-authored report can now state:

> On 3,735 complete two-arm kindergarten records, the common adjusted
> small-versus-regular class-type coefficient was 6.6575 reading points,
> with approximate school-clustered interval [3.2782, 10.0368]. Its causal
> mean-effect interpretation relies on documented within-school
> randomization, a verified link from recorded class type to original
> assignment, suitable complete-record selection, independent-school
> uncertainty, and a common-effect or target-matching weighting argument.
> It concerns assigned class type, one kindergarten reading outcome, and
> participating Tennessee schools, not realized headcount or a portable
> policy dose response.

Read the report's functions before judging it, applying Chapter 1's
argument components as Chapters 7 and 8 did. The sample audit and the
model code carry **STAT-E** content; the design and covariance reasoning
supplies **STAT-J**; the estimate and its clustered interval supply
**STAT-I**; and the first sentence is the bounded **STAT-C** answer.
**SUB-E** situates the pupils, the test, the cohort, and the school
context; **SUB-J** adds the assignment record together with the selection
and weighting qualifications; **SUB-I** reads the coefficient as
assignment-related only under that support; and **SUB-C** bounds the
substantive claim. These are linked functions within one argument, not two
arguments.

Then separate what is *stated* from what is *demonstrated*. The report
states the selection and weighting conditions; stating them does not
establish that they hold. The randomization protocol is documented in the
study's own publications, not shown by this coefficient table. The link
from the recorded class-type field to the original assignment is asserted
here and not verified. Where a necessary link cannot be assessed, the
honest move is to report the limitation rather than to invent it or to
quietly substitute a different target.

The design literature documents implementation and interpretation issues
that a single coefficient cannot settle [@krueger1999; @hanushek1999;
@nye2000; @chetty2011]. Nor does this result transport by itself to
nonparticipating schools, another period, another outcome, or another
education system. Those limits belong inside the answer, not in a list
appended after an unrestricted claim — and a crossed significance boundary
is not what makes a result [@gelman2006; @gelman2020].

Chapter 14 returns to dependence as a design fact and reuses this
calculation as its example. What a design supplies that adjustment cannot
is the subject there; what this section supplies is the discipline of
letting the covariance calculation follow the unit at which the design
acted.

## Reading Bridge {#sec-ch9-summary}

A factor enters a regression as indicators, and its fitted means,
contrasts, and omnibus test are the analysis of variance written as a
model; a contrast's uncertainty needs the covariance between coefficients,
not the sum of two standard errors. Nested models compare on the same
records and the same outcome by an extra-sum-of-squares $F$, and
information criteria rank them by penalized in-sample fit, never by
held-out error. A squared predictor gives a comparison that depends on the
predictor's value; a logged outcome changes the question to a
geometric-mean scale, and its criteria no longer speak to the original
one. HC3, the case bootstrap, and a clustered covariance each change the
uncertainty calculation for a stated reason and leave the fitted
coefficients alone; none repairs curvature, missing outcomes, weighting,
or the absence of a design. Project STAR showed the uncertainty following
the design: schools, not pupils, were the approximately independent units,
and the difference was nearly a factor of two.

Chapter 10 asks a different question of the same equations. Every
comparison in this chapter was the same at every value of the other
predictors — that is what "additive" means, and Section 2.3's common-slope
check was the first test of it. Chapter 10 lets a comparison itself vary
with another predictor, and develops how to read, probe, and report such
an interaction.

### Key terms {.unnumbered}

Use this short reference to revisit the concepts introduced here. The Part
One introduction gives the broader reference and the principal home of
each concept; indicator coding, partial slopes, coefficient covariance,
and HC3 are explained in Chapters 7 and 8.

| Term and common alternatives | Meaning |
|:--|:--|
| Nested-model $F$ test (extra sum of squares; block test) | Compares a reduced model with a fuller one fitted to the same records and outcome: the RSS reduction per added coefficient, divided by the fuller model's residual mean square, referred to $F$ with the number of restrictions and the fuller model's residual df. It tests the added block jointly, not each coefficient separately. |
| Information criteria (AIC; BIC) | Penalized in-sample fit: $-2\log\widehat L$ plus $2k$ (AIC) or $k\log n$ (BIC), with $k$ counting the coefficients and the residual variance. Smaller is preferred among models fitted to the same records and outcome scale; neither is held-out error. |
| Polynomial term (quadratic term; curvature) | A power of a predictor added as a further column so that the fitted mean can curve. The predictor's comparison then depends on its value; centering before squaring reduces collinearity without changing the fitted curve. |
| Log-transformed outcome (geometric-mean scale) | Fitting the logarithm of the outcome, so that a coefficient is an additive difference in expected log outcome and $100(e^{\beta}-1)$ percent is a multiplicative difference in geometric means. A changed outcome scale is a changed question, and its criteria are not comparable with the original scale's. |
| Standardized coefficient (beta weight) | A partial slope rescaled by the predictor and outcome SDs, in outcome-SD units per predictor SD, with the other predictors held fixed. It differs from standardization over named profiles and does not measure causal importance. |
| Bootstrap (case resampling; percentile interval) | Refitting the model to many resamples of the records drawn with replacement and using the spread of the refitted estimates as an SE, or their middle 95% as a percentile interval. It treats the resampled units as independent and exchangeable; it does not repair dependence, thin support, measurement error, or an observational design. |
| Cluster-robust standard error (school-clustered covariance) | A coefficient covariance that sums residual contributions within clusters, such as schools, before combining them, allowing dependence within clusters under an approximation that treats clusters as independent. It changes uncertainty, not the coefficient, and does not repair missing outcomes or weighting. |

: Selected concepts for the regression toolbox. {#tbl-ch09-key-terms tbl-colwidths="[28,72]"}

### Further reading {.unnumbered}

- The [OpenIntro multiple-predictor and model-selection tutorial](https://openintro-ims.netlify.app/model-mlr)
  gives concrete indicator interpretations and criterion-based reading. We
  use its equation-first device without adopting automatic stepwise
  selection.
- The [ISLR regression lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch3-linreg-lab.html)
  and the [ISLR resampling lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch5-resample-lab.html)
  connect fitting, factors, criteria, and resampling in one workflow
  [@james2021].
- Fox [@fox2016] and Kutner and colleagues [@kutner2005] develop the
  linear-model algebra, the restriction statistic, and the diagnostics
  behind this chapter's matrix section.
- Efron and Tibshirani [@efron1993] and Davison and Hinkley [@davison1997]
  are the standard bootstrap references, including the bootstrap-$t$ and
  BCa intervals named in Section 6.3.
- The [sandwich documentation](https://cran.r-project.org/web/packages/sandwich/sandwich.pdf)
  specifies the HC and CR covariance calls used here [@zeileis2020].

## Exercises {.unnumbered}

Use the supplied numbers; rounding within 0.02 is acceptable unless a
different precision is requested. These prompts do not require the
appendices.

\Needspace{18\baselineskip}

<!-- prompt-id: CH09-E01 -->
### Exercise 1: Read indicator coefficients (Foundational) {.unnumbered #ex-ch09-01}

A group-only regression of income (thousand dollars) on occupation type
has observed group means 5.3741 (blue-collar, 44 records), 10.5595
(professional, 31), and 5.0523 (white-collar, 23). (1) With blue-collar as
reference, give the intercept and the two indicator coefficients. (2) Give
the intercept and both coefficients with white-collar as reference. (3)
Explain why the fitted values are the same under both codings. (4) The
full additive model's professional coefficient is 0.3242. Explain why it
differs from the professional coefficient in (1) and why neither is "the
effect of occupational type".

\Needspace{18\baselineskip}

<!-- prompt-id: CH09-E02 -->
### Exercise 2: A contrast needs covariance (Foundational) {.unnumbered #ex-ch09-02}

In the full additive model the professional and white-collar coefficients
are 0.3242 and 0.2353, with estimated variances 2.1406 and 0.9698 and
covariance 0.9720; the residual df are 92 and $t_{.975,92}=1.9861$. (1)
Estimate the professional-minus-white-collar contrast and its variance.
(2) Give its SE and 95% interval. (3) A student adds the two SEs (1.4631
and 0.9848) instead. State what that number is and why it is wrong. (4)
Explain what the interval does and does not say about the two types.

\Needspace{20\baselineskip}

<!-- prompt-id: CH09-E03 -->
### Exercise 3: A nested comparison from residual sums of squares (Foundational) {.unnumbered #ex-ch09-03}

On the same 98 records, the model with education and prestige has RSS
867.6537 on 95 df, and the full model adding the percentage of women and
the two type indicators has RSS 630.8897 on 92 df. The 5% critical value
of $F(3,92)$ is 2.7036. (1) Calculate the $F$ statistic for the added
block. (2) State the decision at 5% and what it does and does not say
about the three added columns. (3) Explain why the sequential sums of
squares from a single `anova()` call on the full model would not answer
this question.

\Needspace{18\baselineskip}

<!-- prompt-id: CH09-E04 -->
### Exercise 4: Information criteria by hand (Foundational) {.unnumbered #ex-ch09-04}

The full model's maximized log-likelihood is $-230.3020$ on 98 records,
with seven fitted parameters (six coefficients and the residual variance),
and $\log 98=4.584967$. (1) Calculate AIC and BIC. (2) The model with
education and prestige only has AIC 499.8328. Which model do the criteria
favour, and on what condition is that comparison meaningful? (3) A
log-income version of the full model has AIC 8.0910. Explain why this
number cannot be compared with your answer to (1).

\Needspace{22\baselineskip}

<!-- prompt-id: CH09-E05 -->
### Exercise 5: Percent and curvature (Integrative) {.unnumbered #ex-ch09-05}

(1) In a log-income model the prestige coefficient is 0.0207. Give the
percent difference in geometric-mean income per prestige point and say
what "geometric mean" adds to the reading. (2) In an income model with a
squared prestige term, the prestige coefficients are $-0.0754$ (linear) and
0.0024 (squared); prestige ranges from 17.3 to 87.2. Give the fitted slope
at prestige 30, 50, and 70 and locate the turning point. (3) State in one
sentence what the curve claims about income and prestige over the observed
range, and one reason that claim is a sensitivity check rather than a
discovery.

\Needspace{22\baselineskip}

<!-- prompt-id: CH09-E06 -->
### Exercise 6: Choose the uncertainty calculation (Integrative) {.unnumbered #ex-ch09-06}

In Project STAR the small-minus-regular kindergarten reading difference on
3,745 observed records is 5.8151 points. Three standard errors are
available: 1.0377 treating pupils as independent, 1.0416 with HC3, and
1.8476 clustering pupils within their 79 schools; use $t_{.975}=1.9606$
for the first two and 1.9908 for the third. (1) Give the interval under
the first and the third calculation. (2) Say which one a report should use
and why, naming the unit at which assignment happened. (3) Name two
problems that none of the three calculations repairs.

## Appendix A: Matrix Geometry, Restrictions, and Sandwich Constructions {.unnumbered #app-ch9-a}

This appendix derives the results that Sections 3.2 and 5.1 used. It is
reference material: the exercises do not require it, and the main route is
complete without it. Complete R reproductions of every construction below
belong to the Computer Lab tasks on normal equations, on manual HC3, and
on the joint Wald helper.

### Projection and the normal equations {.unnumbered}

Let $\mathbf X$ be $n\times p$ with full column rank, so $\mathbf X'\mathbf X$
is invertible. The least-squares criterion is
$S(\boldsymbol\beta)=(\mathbf y-\mathbf X\boldsymbol\beta)'(\mathbf y-\mathbf X\boldsymbol\beta)$.
Expanding gives
$S=\mathbf y'\mathbf y-2\boldsymbol\beta'\mathbf X'\mathbf y
+\boldsymbol\beta'\mathbf X'\mathbf X\boldsymbol\beta$, and differentiating
with respect to $\boldsymbol\beta$ yields the normal equations

$$
\mathbf X'\mathbf X\widehat{\boldsymbol\beta}=\mathbf X'\mathbf y,
\qquad\text{equivalently}\qquad
\mathbf X'(\mathbf y-\mathbf X\widehat{\boldsymbol\beta})=\mathbf 0 .
$$

The second form is the geometry: the residual vector is orthogonal to
every column of $\mathbf X$. Hence
$\widehat{\mathbf y}=\mathbf X\widehat{\boldsymbol\beta}=\mathbf H\mathbf y$
with $\mathbf H=\mathbf X(\mathbf X'\mathbf X)^{-1}\mathbf X'$, and
$\mathbf H$ is symmetric and idempotent — the orthogonal projection onto
the column space of $\mathbf X$. Its trace is $p$, which is why the
leverages $h_{ii}$ sum to the number of fitted coefficients; for `m2` they
sum to 6.

Two consequences used earlier follow immediately. Any full-rank recoding
$\mathbf X\mathbf A$, with $\mathbf A$ invertible, spans the same column
space, so $\mathbf H$ and $\widehat{\mathbf y}$ are unchanged: releveling
a factor or centering a predictor cannot alter a single fitted value.
Adding a genuinely new column enlarges the space and can change every
fitted value. And under the classical error model,
$\widehat{\mathbf V}=s^2(\mathbf X'\mathbf X)^{-1}$ with
$s^2=RSS/(n-p)$, which is where every classical standard error in this
chapter came from. Section 3.2's check confirmed both the coefficients and
the orthogonality to about $10^{-13}$ on the occupation model.

### General linear restrictions {.unnumbered}

Let $\mathbf R$ be $q\times p$ of rank $q$ and consider
$\mathbf R\boldsymbol\beta=\mathbf r$. Since
$\widehat{\boldsymbol\beta}$ has covariance
$\sigma^2(\mathbf X'\mathbf X)^{-1}$ under the classical model, the vector
$\mathbf R\widehat{\boldsymbol\beta}-\mathbf r$ has covariance
$\sigma^2\mathbf R(\mathbf X'\mathbf X)^{-1}\mathbf R'$. Standardizing by
that covariance and dividing by $q$ gives

$$
F=\frac{(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)'\,
[\mathbf R(\mathbf X'\mathbf X)^{-1}\mathbf R']^{-1}\,
(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)}{q\,s^{2}},
$$

which has an exact $F_{q,\,n-p}$ reference under normal errors, because
the numerator and denominator quadratic forms are independent chi-squares.
This is the same number the extra-sum-of-squares formula produces: fitting
the restricted model and differencing residual sums of squares is an
algebraically equivalent route. Taking the two rows of $\mathbf R$ that
select `typeprof` and `typewc`, with $\mathbf r=\mathbf 0$, returned
0.0319 in Section 3.2 — the adjusted type block test, obtained without
fitting the reduced model.

The equivalence with the extra-sum-of-squares route is worth seeing once.
Minimizing $S(\boldsymbol\beta)$ subject to
$\mathbf R\boldsymbol\beta=\mathbf r$ with a Lagrange multiplier gives the
restricted estimator
$\widetilde{\boldsymbol\beta}=\widehat{\boldsymbol\beta}
-(\mathbf X'\mathbf X)^{-1}\mathbf R'
[\mathbf R(\mathbf X'\mathbf X)^{-1}\mathbf R']^{-1}
(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)$, a correction applied
only in the directions the restriction touches. Substituting it into the
criterion shows that the increase in residual sum of squares is
$RSS_R-RSS_F=(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)'
[\mathbf R(\mathbf X'\mathbf X)^{-1}\mathbf R']^{-1}
(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)$, which is exactly the
numerator above. Fitting the reduced model and differencing, or evaluating
the quadratic form in the unrestricted fit, are two routes to one number.

Replacing $s^2(\mathbf X'\mathbf X)^{-1}$ by a robust covariance in the
middle of this expression gives the Wald statistic of Section 5.2. The
algebra survives the substitution; the exact reference distribution does
not.

### The HC3 sandwich {.unnumbered}

Write the coefficient error as
$\widehat{\boldsymbol\beta}-\boldsymbol\beta
=(\mathbf X'\mathbf X)^{-1}\mathbf X'\boldsymbol\varepsilon$. Under
independent rows with possibly different variances $\sigma_i^2$, its exact
covariance is

$$
(\mathbf X'\mathbf X)^{-1}
\Big(\sum_i\sigma_i^2\,\mathbf x_i\mathbf x_i'\Big)
(\mathbf X'\mathbf X)^{-1},
$$

a *sandwich* with the design component as bread and the error-scale
component as filling. The classical estimator substitutes a common $s^2$,
which collapses the filling to $s^2\mathbf X'\mathbf X$. The HC family
substitutes each row's own squared residual instead. Because
$E(e_i^2)=\sigma_i^2(1-h_{ii})$ under the classical model, the raw
residual understates the error at high-leverage rows; HC3 corrects by
dividing by $(1-h_{ii})^2$, which is deliberately slightly conservative
and performs well in small samples [@mackinnon1985; @long2000].

Building the estimator directly — inverting $\mathbf X'\mathbf X$ for the
bread and forming the cross-product of the rows of $\mathbf X$ scaled by
$e_i/(1-h_{ii})$ for the filling — reproduces
`sandwich::vcovHC(m2, type = "HC3")` to within about $10^{-12}$ on the
occupation model, which is floating-point agreement.

### The CR1 cluster sandwich {.unnumbered}

Cluster robustness changes the filling, not the bread. Partition the rows
into $G$ clusters and let $\mathbf X_g$ and $\mathbf e_g$ hold cluster
$g$'s rows and residuals. The cluster score
$\mathbf s_g=\mathbf X_g'\mathbf e_g$ sums the within-cluster
contributions *before* the outer product is taken, so the estimator
retains the within-cluster residual correlations that the independent-row
filling discards:

$$
\widehat{\mathbf V}_{\mathrm{CR1}}=c_G(\mathbf X'\mathbf X)^{-1}
\Big(\sum_{g=1}^{G}\mathbf s_g\mathbf s_g'\Big)(\mathbf X'\mathbf X)^{-1},
\qquad c_G=\frac{G}{G-1}\,\frac{n-1}{n-p}.
$$

If every cluster held one row, the filling would reduce to the HC0 form,
which is why a row-level robust SE is the special case that assumes away
the dependence. The asymptotics are in $G$, not in $n$: 3,745 pupils in 79
schools supply 79 independent pieces of information for this covariance,
which is why the clustered SE of 1.8476 is so much larger than the
row-level 1.0377 and why a $t_{G-1}$ reference is used. Constructing the
score matrix by summing $\mathbf X\odot\mathbf e$ within school, applying
$c_G=1.0131$, and sandwiching it reproduces
`sandwich::vcovCL(..., type = "HC1", cadjust = TRUE)` to within about
$10^{-13}$ on the observed-reading STAR model.

## Appendix B: Bootstrap Schemes and Intervals {.unnumbered #app-ch9-b}

Section 6 used case resampling and said why. This appendix runs the other
two schemes on the same model, the same three targets, and the same seed,
and defines the two higher-order intervals named there. The lab tasks on
resampling variants and on manual covariance reproduction carry the full
implementations.

### Residual and wild resampling {.unnumbered}

Both schemes hold the design matrix fixed and rebuild outcomes from the
fitted values. Residual resampling draws leverage-rescaled residuals
$e_i/\sqrt{1-h_{ii}}$ with replacement and adds them to the fitted values,
which assumes the rescaled residuals are exchangeable across rows. The
wild bootstrap keeps each row's own rescaled residual and multiplies it by
an independent mean-zero, unit-variance draw — here $\pm1$ with equal
probability — so each row retains its own error scale.

```{=latex}
\Needspace{26\baselineskip}
```

```{r}
#| label: ch9b-schemes
#| echo: true
set.seed(90909)
X <- model.matrix(m2); e <- residuals(m2); h <- hatvalues(m2)
fit_draw <- function(y) {
  b <- coef(lm(y ~ X - 1))
  c(b["Xeducation"], b["Xtypeprof"] - b["Xtypewc"], b["Xtypeprof"])
}
residual_draws <- t(replicate(B, fit_draw(fitted(m2) +
  sample(e / sqrt(1 - h), replace = TRUE))))
wild_draws <- t(replicate(B, fit_draw(fitted(m2) +
  e / sqrt(1 - h) * sample(c(-1, 1), nrow(d), replace = TRUE))))
rbind(case = apply(draws, 2, sd, na.rm = TRUE),
  residual = apply(residual_draws, 2, sd), wild = apply(wild_draws, 2, sd))
```

For the education slope the three schemes give 0.3699 (case), 0.2820
(residual), and 0.3673 (wild). The residual scheme is the outlier, and its
value sits close to the classical 0.2752 — which is what one should
expect, since both assume one common error distribution. The wild
bootstrap, which does not, lands near the case result and near HC3's
0.3823. For the professional coefficient the pattern repeats more
strongly: 1.9926, 1.4861, and 1.9767. When the schemes disagree, the
disagreement is about the error model, not about the data, and it is a
reason to say which scheme was used.

### Percentile, normal, and higher-order intervals {.unnumbered}

Four interval definitions are worth distinguishing in words.

- The **percentile interval** takes the 2.5th and 97.5th quantiles of the
  resampling distribution of the estimate.
- The **normal-approximation interval** takes the original estimate plus
  and minus 1.96 bootstrap standard errors. It uses the resampling
  distribution only for its spread, so it is insensitive to skewness.
- The **bootstrap-$t$ interval** resamples a studentized quantity,
  $(\widehat\theta^{*}-\widehat\theta)/SE^{*}$, and inverts its empirical
  quantiles. It needs a standard error inside every resample, which is
  expensive, and it is more accurate when the studentized quantity is
  closer to pivotal than the estimate itself.
- The **BCa interval** adjusts the percentile endpoints for median bias
  and for an acceleration constant that captures how the standard error
  changes with the parameter. It is the usual recommendation when the
  resampling distribution is visibly skewed [@efron1993; @davison1997].

For the three targets of Section 6 the skewness values $-0.3891$, 0.2635,
and 0.3785 are mild, so percentile and normal-approximation endpoints agree
closely and the higher-order corrections would change little. That
agreement is a property of this example, not a general licence.

### Extended practice: resampling choices (Integrative) {.unnumbered}

1. Repeat the case bootstrap of Section 6.2 for the women coefficient in
   `m2`, with the same seed and $B$. Report the bootstrap SE and
   percentile interval, compare them with the classical 0.0097 and the
   HC3 0.0079, and explain what it means that the classical SE is the
   largest of the three here. State what the comparison does not show.
2. Resample the STAR observed-reading records in two ways: drawing pupils
   with replacement within each school, and drawing whole schools with
   replacement. Report the resulting SEs for the small-minus-regular
   difference beside the row-level 1.0377 and the clustered 1.8476.
   Explain which resampling unit the design supports and why, and say
   what neither scheme repairs.

## References {.unnumbered}

::: {#refs}
:::
