---
title: "Understanding Data, Variation, and Sampling"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-cap: true
    tbl-cap: true
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "88%"
)
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(dev = "cairo_pdf")
}

options(digits = 3, width = 72, scipen = 4)
library(ggplot2)
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      plot.title.position = "plot"
    )
)

## The curated UCI secondary-school record of Cortez and Silva (2008),
## reduced to the columns this chapter analyses and embedded so that the
## chapter renders without an external data file. Values are unchanged;
## pass reproduces the documented cut point of 10 on the 0-20 scale.
sp_absences <- c(
 6,4,10,2,4,10,0,6,0,0,0,4,2,2,0,4,6,4,16,4,0,0,2,0,2,14,2,4,4,16,0,0,0,
 0,0,0,2,7,2,8,25,8,2,0,14,8,12,4,2,2,2,2,6,0,6,8,0,4,2,2,6,6,4,2,0,2,4,
 4,2,12,0,0,2,2,54,6,8,0,2,12,2,4,10,4,2,6,4,4,12,18,0,4,4,0,6,2,2,2,6,
 0,14,0,4,26,0,10,8,2,6,4,6,0,6,10,8,2,2,0,20,6,2,6,2,18,0,0,0,2,0,8,0,
 0,12,16,0,0,0,0,0,0,0,8,2,2,0,0,0,2,0,0,0,6,8,0,0,2,8,6,2,4,0,6,0,2,0,
 16,4,0,0,0,0,2,0,0,4,4,2,4,10,4,10,2,0,56,14,12,2,0,6,4,10,0,12,8,0,0,
 4,8,24,0,2,6,4,18,6,28,5,10,6,6,10,13,0,15,12,2,22,13,3,4,2,0,2,0,0,16,
 10,2,14,10,14,4,14,2,18,10,4,20,2,0,14,2,0,0,0,6,4,16,8,0,0,6,4,0,0,2,
 6,12,8,0,21,2,1,4,0,13,2,8,10,0,15,4,2,2,2,6,75,22,15,8,30,19,1,4,4,2,
 5,6,6,9,11,0,12,6,8,4,0,10,0,5,14,0,0,0,20,8,0,38,0,18,0,20,3,22,14,40,
 0,9,0,2,23,12,3,1,0,3,3,8,7,4,2,7,0,0,0,16,12,0,7,4,4,0,11,0,4,7,9,0,0,
 10,8,2,7,4,4,0,4,2,4,0,0,2,0,0,0,4,0,0,0,10,4,3,8,14,0,2,4,4,0,17,4,5,
 2,0,14,2,7,0,0,0,11,3,3,0,5
)

sp_grade_period_2 <- c(
 6,5,8,14,10,15,12,5,18,15,8,12,14,10,16,14,14,10,5,10,14,15,15,13,9,9,
 12,16,11,12,11,16,16,10,14,7,16,16,12,13,10,12,18,8,10,8,12,19,15,7,13,
 13,11,10,13,9,15,15,10,16,11,8,10,9,10,15,13,7,9,16,15,10,6,12,12,9,11,
 11,8,5,12,10,6,15,10,9,7,14,10,6,7,17,6,10,13,10,15,9,14,9,7,17,13,6,
 18,11,8,18,13,15,19,10,13,19,9,15,13,14,7,13,15,14,13,11,7,13,10,8,4,
 18,0,0,13,11,0,0,0,0,12,16,9,9,11,14,0,11,7,11,6,9,5,13,10,0,11,8,12,8,
 15,12,6,9,0,10,8,11,10,15,7,14,5,15,11,7,11,9,13,5,8,10,8,13,17,9,13,
 12,12,15,7,9,12,8,8,9,14,15,15,9,18,9,16,10,9,6,10,9,7,12,9,7,8,12,13,
 7,10,15,6,6,7,10,6,5,16,13,13,8,15,11,8,10,13,11,9,13,7,9,13,12,11,7,
 12,11,0,12,0,18,12,8,5,15,8,10,9,9,12,9,12,11,14,9,18,8,12,9,10,17,9,
 10,9,0,9,14,11,14,10,12,9,9,8,11,8,9,12,9,9,10,18,12,14,13,11,15,12,18,
 13,12,9,8,13,15,10,11,12,17,14,12,18,9,12,10,9,12,11,10,13,11,8,10,11,
 11,13,9,11,14,15,12,15,10,9,14,8,14,0,8,9,15,13,8,15,10,12,10,15,8,10,
 13,15,10,15,13,7,13,7,8,11,9,13,12,10,16,13,12,11,15,11,10,13,6,10,12,
 7,12,11,5,18,8,14,9,15,10,14,6,11,5,5,9,5,5,9,5,9,16,8,12,9
)

sp_final_grade <- c(
 6,6,10,15,10,15,11,6,19,15,9,12,14,11,16,14,14,10,5,10,15,15,16,12,8,8,
 11,15,11,11,12,17,16,12,15,6,18,15,11,13,11,12,18,11,9,6,11,20,14,7,13,
 13,10,11,13,10,15,15,9,16,11,11,9,9,10,15,12,6,8,16,15,10,5,14,11,10,
 10,11,10,5,12,11,6,15,10,8,6,14,10,7,8,18,6,10,14,10,15,10,14,8,5,17,
 14,6,18,11,8,18,13,16,19,10,13,19,9,16,14,13,8,13,15,15,13,13,8,12,11,
 9,0,18,0,0,12,11,0,0,0,0,12,15,0,9,11,13,0,11,0,11,0,10,0,14,10,0,12,8,
 13,10,15,12,0,7,0,10,7,12,10,16,0,14,0,16,10,0,9,9,11,6,9,11,8,12,17,8,
 12,11,11,15,9,10,13,9,8,10,14,15,16,10,18,10,16,10,10,6,11,9,7,13,10,7,
 8,13,14,8,10,15,4,8,8,10,6,0,17,13,14,7,15,12,9,12,14,11,9,13,6,10,13,
 12,11,0,12,12,0,12,0,18,13,8,5,15,8,10,8,8,12,8,13,11,14,0,18,8,12,9,0,
 17,10,11,10,0,9,14,11,14,10,12,9,9,8,10,8,10,12,10,11,11,19,12,14,15,
 11,15,13,18,14,11,0,8,14,16,11,10,14,18,13,12,18,8,12,10,0,13,11,11,13,
 11,0,9,10,11,13,9,11,15,15,11,16,10,9,14,8,14,0,0,0,15,13,0,17,10,11,0,
 15,0,10,14,16,9,15,13,8,13,8,8,11,9,13,11,10,16,13,12,10,15,12,10,13,0,
 10,11,9,12,11,5,19,10,15,10,15,10,14,7,10,0,5,10,6,0,8,0,9,16,7,10,9
)

sp_higher_no <- c(
 79,128,149,151,154,161,164,165,166,167,240,248,251,253,270,277,312,353,
 362,372
)

sp_school_ms <- c(
 350,351,352,353,354,355,356,357,358,359,360,361,362,363,364,365,366,
 367,368,369,370,371,372,373,374,375,376,377,378,379,380,381,382,383,
 384,385,386,387,388,389,390,391,392,393,394,395
)

students <- data.frame(
  student_id = sprintf("U%03d", seq_along(sp_absences)),
  school = "GP",
  absences = as.integer(sp_absences),
  grade_period_2 = as.integer(sp_grade_period_2),
  final_grade = as.integer(sp_final_grade),
  higher_education = "yes",
  stringsAsFactors = FALSE
)
students$school[sp_school_ms] <- "MS"
students$higher_education[sp_higher_no] <- "no"
students$pass <- ifelse(students$final_grade >= 10, "Yes", "No")

grade_zero <- students$final_grade == 0
students_no_grade_zero <- students[!grade_zero, ]
stopifnot(
  sum(grade_zero) == 38L,
  all(students$absences[grade_zero] == 0),
  sum(students$grade_period_2[grade_zero] > 0) == 25L
)
grade_zero_sensitivity <- c(
  n = nrow(students_no_grade_zero),
  mean = mean(students_no_grade_zero$final_grade),
  median = median(students_no_grade_zero$final_grade),
  variance = var(students_no_grade_zero$final_grade),
  sd = sd(students_no_grade_zero$final_grade),
  IQR = IQR(students_no_grade_zero$final_grade),
  covariance = cov(
    students_no_grade_zero$absences,
    students_no_grade_zero$final_grade
  ),
  correlation = cor(
    students_no_grade_zero$absences,
    students_no_grade_zero$final_grade
  )
)

grade_summary <- data.frame(
  Summary = c("Mean", "Median", "Sample variance", "Sample standard deviation", "IQR"),
  Value = c(
    mean(students$final_grade),
    median(students$final_grade),
    var(students$final_grade),
    sd(students$final_grade),
    IQR(students$final_grade)
  ),
  Units = c("grade points", "grade points", "squared grade points",
            "grade points", "grade points")
)

same_mean <- data.frame(
  Pattern = rep(c("Concentrated", "Spread", "Right-skewed"), each = 9),
  Value = c(
    8, 9, 9, 10, 10, 10, 11, 11, 12,
    2, 4, 6, 8, 10, 12, 14, 16, 18,
    5, 6, 7, 8, 9, 10, 11, 13, 21
  ),
  Index = rep(seq_len(9), 3)
)

set.seed(60202)
pattern_x <- seq(-2, 2, length.out = 80)
relationship_patterns <- rbind(
  data.frame(
    X = pattern_x,
    Y = 1.2 + 1.1 * pattern_x + rnorm(80, 0, .45),
    Pattern = "Positive linear"
  ),
  data.frame(
    X = pattern_x,
    Y = 1.2 - 1.1 * pattern_x + rnorm(80, 0, .45),
    Pattern = "Negative linear"
  ),
  data.frame(
    X = pattern_x,
    Y = pattern_x^2 + rnorm(80, 0, .25),
    Pattern = "Curvilinear"
  ),
  data.frame(
    X = runif(80, -2, 2),
    Y = rnorm(80, 1.2, 1.1),
    Pattern = "No clear pattern"
  )
)
relationship_patterns$Pattern <- factor(
  relationship_patterns$Pattern,
  levels = c("Positive linear", "Negative linear", "Curvilinear", "No clear pattern")
)

normal_grid <- data.frame(Z = seq(-3.4, 3.4, length.out = 500))
normal_grid$Density <- dnorm(normal_grid$Z)

set.seed(60203)
repetitions <- 5000L
sample_sizes <- c(5L, 30L, 100L)
sampling_means <- do.call(rbind, lapply(sample_sizes, function(n) {
  data.frame(
    Sample_size = factor(n, levels = sample_sizes),
    Sample_mean = replicate(repetitions, mean(rexp(n, rate = 1 / 10)))
  )
}))
sampling_split <- split(sampling_means$Sample_mean, sampling_means$Sample_size)
sampling_summary <- do.call(rbind, lapply(sampling_split, function(x) {
  c(
    Center = mean(x),
    Empirical_SE = sd(x),
    Skewness = mean((x - mean(x))^3) / sd(x)^3
  )
}))
sampling_summary <- data.frame(
  n = sample_sizes,
  Theoretical_SE = 10 / sqrt(sample_sizes),
  sampling_summary,
  row.names = NULL,
  check.names = FALSE
)

set.seed(60204)
lln_draws <- rexp(400, rate = 1 / 10)
lln <- data.frame(
  Draw = seq_along(lln_draws),
  Running_mean = cumsum(lln_draws) / seq_along(lln_draws)
)
```

{{< pagebreak >}}

::: {.callout-tip title="Chapter box"}
**Central question.** Among the 395 students recorded in two Portuguese
secondary schools, how strong is the linear relationship between recorded
absences and final mathematics grade, as summarized by Pearson's correlation,
and what visible features, coding sensitivity, and scope limits must accompany
that summary?

**Focal concepts.** Represented observations, distributions, graphical and
numerical summaries, relationships, and repeated-sampling variation.

**Main evidence.** The documented two-school student-performance record, in
which one row represents one student and both focal variables are present for
all 395 rows [@cortez2008data].

**Defensible scope.** An exploratory, observational, local description. The
record does not estimate an effect of changing attendance and does not by itself
support generalization to other students.

Reading the focal associational argument is the **primary use** of this chapter;
the evidence for it is analysed descriptively. Quality evaluation follows
reading as a separate and dependent **secondary use**. For researchers, the
discussion also supports selected stages of the construction workflow as an
**auxiliary use**.

Chapter 1 is the sole conceptual prerequisite. Chapter 1a is optional and is
not needed to read or use this chapter.
:::

## Start with the Question, Not the Statistic

A short school report might say:

> Recorded absences were almost uncorrelated with final mathematics grades, so
> attendance does not matter for achievement.

The reported number in the first clause could be correct while the sentence as
a whole is not. A coefficient summarizes only a specified feature of recorded
values. The second clause changes the language from a local description to a
much broader conclusion about what matters. Before accepting the sentence, ask
four questions:

1. What exactly is being asked?
2. What do the available records represent?
3. What pattern is visible, and how should it be summarized?
4. What answer remains after design and scope constraints are retained?

These questions keep the study question visible while we learn descriptive
tools. They also prevent the statistic from deciding the question after the
result is known.

### The focal question and estimand

The focal question asks how two measured variables vary together, so its mode
is **associational**. It is analysed descriptively within a fixed analytic
record. Its **estimand** is the declared linear association: the Pearson
correlation between recorded absences and final mathematics grade for the
defined group of 395 students. An estimand is the precisely specified quantity
that would answer the question if it could be known. Here all 395 rows contain
numeric entries; taking them literally, the calculated correlation, $r=.034$,
is the value of that finite-record estimand under the recorded coding. No
sampling error is attached to this complete-record description. Numeric
completeness does not establish coding validity, and the zero-grade pattern
examined below remains unresolved. Because the records are not established as a
probability sample from a wider population, this value is not automatically an
estimate of a wider-population correlation.

A graph remains necessary. Pearson's $r$ compresses a two-variable pattern
to one number and targets linear form. It cannot show whether observations form
a curve or clusters, whether most absences occupy a narrow range, or whether a
few observations strongly influence the result. The scatterplot and smooth
serve as checks on the scalar summary. Covariance explains the
unstandardized cross-product behind $r$. These tools serve one declared
estimand; they do not create extra question--answer pairs.

A wider-population correlation or the causal effect of preventing absence would
be a different estimand. Each would require additional support. A reader should
therefore identify the question, estimand, and reported answer before deciding
what a result means.

### A compact study record

The data were adapted from a documented student-performance resource based on
two Portuguese secondary schools [@cortez2008data]. @tbl-study-record gives
the minimum record needed before calculation.

| Feature | Recorded information | Why it matters |
|:--------|:---------------------|:---------------|
| Case and unit | One row per student; 395 rows | One point in the scatterplot represents one recorded student |
| Focal explanatory variable | Absences, recorded as a count from 0 to 75 | A large count is unusual but not automatically an error |
| Focal response variable | Final mathematics grade, recorded from 0 to 20 | The bounded scale constrains interpretation |
| Numeric-entry completeness | Both focal fields contain numeric values in all 395 rows | The literal recorded-value denominator is 395; coding validity remains separate |
| Source and design | Observational school records; no probability sample or attendance intervention is established | Population and causal interpretations require support not supplied here |
| Analytic status | Exploratory local description | The graph and summaries describe the accessible record |

: *Compact record for the focal absence--grade description.* {#tbl-study-record tbl-colwidths="[20,35,45]"}

The original source documents variables and study context, but a publicly
available file is not the same as a random sample from an unlimited population.
Nor does a complete pair of columns guarantee that the measurements capture
every substantive meaning a reader might attach to the words *attendance* and
*achievement*.

### Reading support before evaluating it

Chapter 1 taught a reader to locate a question--answer pair, identify one
argument with linked statistical and substantive layers, and trace Evidence,
Justification, Inference, and bounded Claim. This chapter applies that approach.
Sections 1--5 accumulate relevant component content and representative
documentary traces. Only after the argument is visible will we ask selected
quality-evaluation questions. Reading records what the accessible study says
and where its support can be found; it does not silently repair or grade the
study.

## What the Data Represent

A dataset is not merely a rectangular collection of numbers. Each row, column,
code, and missing entry represents a decision about what was observed and how
it was recorded. These decisions determine what later summaries can mean.

### Cases, units, and occasions

A **case** is the entity or occasion represented by one row. The
**observational unit** is the unit on which information is observed. Sometimes
they are the same; sometimes a person supplies several rows across visits,
trials, or waves. If each of 120 students is recorded before and after a course,
240 rows do not represent 240 independent students.

In the focal record, one row represents one student and the student is the
observational unit. There is one occasion per row. The identifier distinguishes
rows; it is not a score to average. @tbl-student-preview makes those
roles concrete.

```{r}
#| label: tbl-student-preview
#| tbl-cap: "Selected fields from six records in the student-performance data."
#| tbl-colwidths: [16, 16, 22, 22, 18]

preview <- students[1:6, c(
  "student_id", "school", "absences", "final_grade", "pass"
)]
names(preview) <- c("Student ID", "School", "Absences", "Final grade", "Pass")
knitr::kable(preview, booktabs = TRUE, align = c("l", "l", "r", "r", "l"))
```

### Variable form and substantive role

A **variable** records a feature that can differ across cases. A **value** is
the recorded entry for one case. Variable form and variable role answer
different questions:

- **Form** asks how values are represented: quantitative, nominal categorical,
  or ordered categorical.
- **Role** asks what the variable does in the present question: explanatory,
  response, grouping, or identifier.
- **Scale and units** ask what a difference or ratio can mean.

Absences and final grade are quantitative in recorded form. Absences plays the
explanatory role and final grade the response role for the focal associational
question. Those roles do not establish causal direction. Another question
could reverse them or use either variable only to define groups.

| Recorded feature | Form | Role in the focal question | Interpretive caution |
|:-----------------|:-----|:---------------------------|:---------------------|
| Student ID | Nominal label | Identifier: links and checks records | Numerical-looking labels are not quantities |
| School | Nominal categorical | Possible grouping information | Codes name categories; their order has no arithmetic meaning |
| Absences | Quantitative count | Explanatory variable | Counts are bounded below and strongly right-skewed |
| Final grade | Quantitative bounded score | Response variable | A score is not identical to every meaning of achievement |
| Pass | Binary categorical | Derived grouping variable | Its definition depends on the documented cut point |

: *Recorded form and substantive role are different properties of a variable.* {#tbl-variable-form-role tbl-colwidths="[20,20,24,36]"}

### Coding, missingness, and the denominator

Coding translates observations into analyzable values. A code can preserve or
distort substantive meaning. The numbers 1 through 4 may label ordered study
time categories without establishing equal-hour intervals. A value of zero may
mean a genuine measured zero, an absence, a score below a threshold, or a
missing-value code. The documentation must decide which.

The focal file contains 38 final-grade zeros. The UCI repository identifies the
final grade as a value from 0 to 20 and reports no missing values
[@cortez2008data]. Yet every zero-grade row also records zero absences, and 25
have nonzero second-period grades from 4 to 10. The accessible documentation
does not explain that joint pattern. We retain the values for the primary
as-recorded description, flag their meaning as unresolved, and use an explicit
deletion sensitivity rather than silently treating them as either invalid
zeros or missing values.

**Missingness** changes the set of cases supporting a summary. If 1,240 clinic
visits contain 36 missing waiting times and 81 missing return statuses, the
complete-case denominator is not determined by subtracting 36 and 81 unless the
overlap is known. If 10 visits are missing both, the denominator with both
variables observed is

$$
1240-(36+81-10)=1133.
$$

Always report the denominator that actually supports the graph, table, or
coefficient. A source record, observed record, and analytic sample can differ.
Both focal fields contain numeric entries in all 395 student rows. Even so, how
the denominator works when values are absent is important because later studies
will rarely be this tidy.

### Follow the representation path

The representation path asks how the analytic cases came to stand for the
people, occasions, or events named in a claim. @fig-representation-path shows
the minimum sequence.

```{r}
#| label: fig-representation-path
#| fig-cap: "A representation path separates the intended population from the analytic sample."
#| fig-alt: "Five boxes run from target population through frame or source, eligible or invited units, observed or responding units, and analytic sample. A note says to mark nonapplicable steps instead of inventing them."
#| fig-height: 2.5
#| out-width: "94%"

old <- par(mar = rep(.15, 4), xpd = NA)
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 2.2))
xs <- seq(1, 9, length.out = 5)
labs <- c(
  "Target\npopulation",
  "Frame or\nsource",
  "Eligible or\ninvited",
  "Observed or\nresponding",
  "Analytic\nsample"
)
for (i in seq_along(xs)) {
  rect(xs[i] - .78, .85, xs[i] + .78, 1.62,
       col = c("#E9F1F5", "#DDEAF0", "#E8E6F3", "#E6F0E2", "#F7E8D5")[i],
       border = "#28424F", lwd = 1.2)
  text(xs[i], 1.235, labs[i], cex = .73)
  if (i < length(xs)) {
    arrows(xs[i] + .8, 1.235, xs[i + 1] - .8, 1.235,
           length = .08, col = "#50646D", lwd = 1.2)
  }
}
text(5, .35, "Mark a step not applicable; do not invent a sampling step.",
     cex = .75, col = "#5A4141")
par(old)
```

For the focal resource, the two schools and their recorded students provide a
source and analytic record. The accessible documentation does not establish a
probability sample from a wider target population. We should therefore leave
unsupported links open instead of assuming that 395 cases represent all
secondary students.

A small evidence audit before calculation is often enough:

1. What does one row represent?
2. Which variables, scales, roles, and units serve the question?
3. Which values need documentary checking?
4. Which cases enter each summary, and what is its denominator?
5. What path connects the analytic sample to the scope of the claim?

These are reading questions about represented evidence. They do not yet decide
whether the study's support is adequate.

## See and Summarize a Distribution

A **distribution** describes how a variable's values are arranged across cases.
A useful beginner sequence is: identify the variable, display the observations,
describe visible structure, choose center and spread, and say what the summary
leaves unresolved.

### See shape before reducing it

@fig-grade-distribution displays all 395 final grades. The graph makes
several features visible: the score is bounded between 0 and 20, centered near
11, discrete rather than continuous, and has a visible group at zero.

```{r}
#| label: fig-grade-distribution
#| fig-cap: "Distribution of final mathematics grades for the 395 recorded students."
#| fig-alt: "A histogram of grades from zero to twenty is centered near eleven, bounded at both ends, and shows a visible group at zero."
#| fig-height: 3.4

ggplot(students, aes(x = final_grade)) +
  geom_histogram(
    binwidth = 1,
    boundary = -.5,
    fill = "#B9D5E5",
    color = "white"
  ) +
  geom_vline(xintercept = mean(students$final_grade),
             color = "#9B3D45", linetype = 2, linewidth = .7) +
  geom_vline(xintercept = median(students$final_grade),
             color = "#315B45", linetype = 3, linewidth = .7) +
  scale_x_continuous(breaks = seq(0, 20, 2)) +
  labs(
    x = "Final mathematics grade (0--20)",
    y = "Number of students"
  )
```

Describe a distribution with four kinds of language:

- **Shape:** symmetry or skewness, clusters, gaps, and discreteness.
- **Center:** where a typical or balancing value lies.
- **Spread:** how widely observations vary.
- **Unusual observations:** values or patterns needing contextual checking.

An unusual value is not automatically a mistake. Its recorded meaning and
provenance should be investigated before it is changed or removed.

### Five summaries with distinct jobs

The main route uses exactly five numerical summaries: mean, median, sample
variance, sample standard deviation, and interquartile range. They are a
selected family, not the **five-number summary** of minimum, first quartile,
median, third quartile, and maximum.

For observations $x_1,\ldots,x_n$, $n$ is their number and $x_i$ is observation
$i$. The symbol $\sum$ means "add the indicated terms":
$\sum_{i=1}^n x_i=x_1+x_2+\cdots+x_n$. The mean, sample variance, sample
standard deviation, and IQR are

$$
\bar{x}=\frac{1}{n}\sum_{i=1}^{n}x_i,\qquad
s^2=\frac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar{x})^2,
$$

$$
s=\sqrt{s^2},\qquad
\operatorname{IQR}=Q_3-Q_1.
$$

The **median** is the middle ordered observation when $n$ is odd and the
average of the two middle observations when $n$ is even. The first quartile
$Q_1$ marks approximately the 25th percentile; the third quartile $Q_3$ marks
approximately the 75th. They bracket the middle half of the distribution.
For the small hand examples, take the median of each ordered half, excluding
the overall middle observation when $n$ is odd. Software may interpolate
between values instead; the large-data summaries here use R's default
interpolation convention. These rules can differ slightly in small datasets.

| Summary | Interpretive job | Units | Main caution |
|:--------|:-----------------|:------|:-------------|
| Mean $\bar{x}$ | Arithmetic balance point; center used by much later algebra | Original units | Can move strongly with skewness or extreme values |
| Median | Ordered midpoint | Original units | Does not describe spread or tail shape |
| Sample variance $s^2$ | Sum of squared deviations divided by $n-1$ | Squared units | Magnitude is hard to read directly in original units |
| Sample standard deviation $s$ | Typical scale of deviation around the mean | Original units | Can be influenced by extreme values |
| IQR | Width of the middle half, $Q_3-Q_1$ | Original units | Does not show what happens in the tails |

: *Five selected summaries and the questions they answer.* {#tbl-five-summaries tbl-colwidths="[19,33,17,31]"}

To make the formulas concrete, consider four illustrative completion times:
1, 3, 3, and 5 minutes. The sum is $1+3+3+5=12$, so
$\bar x=12/4=3$ minutes. The middle two values are both 3, giving median 3.
The lower-half median is $Q_1=(1+3)/2=2$ and the upper-half median is
$Q_3=(3+5)/2=4$, so $\operatorname{IQR}=4-2=2$ minutes.

| Observation $x_i$ | Deviation $x_i-3$ | Squared deviation $(x_i-3)^2$ |
|-----------------:|-----------------:|----------------------------:|
| 1 | $-2$ | 4 |
| 3 | 0 | 0 |
| 3 | 0 | 0 |
| 5 | 2 | 4 |
| **Sum** | **0** | **8** |

: *Working from observations to squared-deviation spread.* {#tbl-worked-variance tbl-colwidths="[25,30,45]"}

Thus $s^2=8/(4-1)=8/3=2.67$ squared minutes, and
$s=\sqrt{8/3}=1.63$ minutes. Keep the unrounded variance while taking its
square root. Variance measures spread on a squared scale; SD restores the
original units.

Why $n-1$? Deviations from their own mean must sum to zero. Here, once the
first three deviations $-2,0,0$ are known, the last must be 2. In general,
only $n-1$ deviations can vary freely: one **degree of freedom** is used by
estimating the center. This zero-sum constraint explains the lost degree of
freedom, but is not by itself a proof of unbiasedness. Section 5 supplies the
sampling-model justification after defining expectation.

For a fixed observed distribution, dividing the same squared-deviation sum by
$n$ instead gives its mean squared deviation ($8/4=2$ here). Both conventions
can describe spread when clearly labelled. We use the conventional sample
variance $s^2$ throughout; the name and denominator do not establish a random
sample or justify population claims. Nor is $n-1$ the finite-population
correction discussed in Appendix C.

@tbl-grade-summary applies all five summaries to final grade.

```{r}
#| label: tbl-grade-summary
#| tbl-cap: "Five numerical summaries of final mathematics grade."
#| tbl-colwidths: [32, 24, 28]

grade_print <- grade_summary
grade_print$Value <- c("10.42", "11.00", "20.99", "4.58", "6.00")
knitr::kable(
  grade_print,
  names = c("Summary", "Value", "Units"),
  booktabs = TRUE,
  align = c("l", "r", "l")
)
```

The mean is 10.42 and the median is 11. The sample variance is 20.99 squared
grade points, while the standard deviation is 4.58 grade points. The IQR is 6
grade points. These numbers summarize the recorded distribution; they do not
make the score scale continuous, validate the zeros, or establish a probability
model.

Omitting the 38 zero-grade rows only as a coding sensitivity leaves
`r unname(grade_zero_sensitivity["n"])` records: mean
`r sprintf("%.2f", grade_zero_sensitivity["mean"])`, median
`r sprintf("%.0f", grade_zero_sensitivity["median"])`, variance
`r sprintf("%.2f", grade_zero_sensitivity["variance"])`, SD
`r sprintf("%.2f", grade_zero_sensitivity["sd"])`, and IQR
`r sprintf("%.0f", grade_zero_sensitivity["IQR"])`. These are diagnostic
quantities, not replacement summaries.

### Equal means can describe different empirical worlds

@fig-same-mean shows three constructed distributions with the same mean
of 10. Their shapes and spreads are plainly different.

```{r}
#| label: fig-same-mean
#| fig-cap: "Three constructed distributions share a mean of 10 but differ in shape and spread."
#| fig-alt: "Faceted dot plots show a concentrated distribution around ten, a broad evenly spaced distribution, and a right-skewed distribution. Each has mean ten."
#| fig-height: 3.0

ggplot(same_mean, aes(x = Value, y = 0)) +
  geom_dotplot(binaxis = "x", stackdir = "up", binwidth = .55,
               fill = "#6C9EAE", color = "#28424F") +
  geom_vline(xintercept = 10, linetype = 2, color = "#9B3D45") +
  facet_wrap(~ Pattern, nrow = 1) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(x = "Observed value")
```

A center cannot replace a distribution. Report mean with standard deviation
when that pair serves the question and the visible shape. Report median with
IQR when resistant summaries communicate center and spread more honestly. In
either case, retain a graph. Resistant summaries are not repairs for invalid
measurement, selection bias, or unexplained exclusions.

### A brief Normal-model and z-score bridge

The **Normal distribution** is a symmetric, bell-shaped probability model
located by a mean $\mu$ and scaled by a standard deviation $\sigma$. It is
not the default shape of every dataset. The **standard Normal distribution** has
mean 0 and standard deviation 1.

If a modeled variable follows $X\sim N(\mu,\sigma^2)$, then

$$
Z=\frac{X-\mu}{\sigma}\sim N(0,1).
$$

For observed sample values, a standardized score is often written

$$
z_i=\frac{x_i-\bar{x}}{s}.
$$

It records how far an observed value lies from the sample mean in sample
standard-deviation units. For example, a sample of commute times has mean
26 minutes and SD 6 minutes. A 38-minute commute has
$z=(38-26)/6=12/6=2$: it is two sample standard deviations above the sample
mean. Model standardization would instead use the declared model parameters
$\mu$ and $\sigma$. Subtracting a center changes location; dividing by a
positive SD changes scale, not shape. Computing z scores does **not** make a
skewed dataset Normal.

```{r}
#| label: fig-standard-normal
#| fig-cap: "The standard Normal model centers distance at zero and measures it in standard-deviation units."
#| fig-alt: "A symmetric bell curve is centered at zero with vertical markers at minus one, zero, and one standard deviation."
#| fig-height: 2.9

ggplot(normal_grid, aes(x = Z, y = Density)) +
  geom_area(fill = "#DDEAF0", color = "#325A6A", linewidth = .7) +
  geom_vline(xintercept = c(-1, 0, 1),
             linetype = c(3, 2, 3),
             color = c("#6A6A6A", "#9B3D45", "#6A6A6A")) +
  annotate("text", x = c(-1, 0, 1), y = c(.06, .43, .06),
           label = c("z = -1", "z = 0", "z = 1"), size = 3.2) +
  scale_x_continuous(breaks = -3:3) +
  labs(x = "Standardized value, z", y = "Density")
```

This is a preview of a probability distribution, not a normality test or a
tail-area exercise. Section 5 distinguishes the observed data distribution
from a modeled probability distribution and from a sampling distribution.

### Choose a graph and summary from the question

| Variable forms | Useful first display | Initial graph-reading focus |
|:---------------|:---------------------|:---------------------------|
| One quantitative variable | Dotplot, histogram, or boxplot | What shape, center, spread, and unusual observations are visible? |
| Quantitative and quantitative | Scatterplot | What form, direction, strength, support, and unusual observations are visible? |
| Quantitative and categorical | Grouped distributions | How do distributions differ across groups? |
| Categorical and categorical | Counts with declared conditional percentages | Which denominator answers the comparison question? |

: *A graph-and-summary decision guide.* {#tbl-graph-choice tbl-colwidths="[26,31,43]"}

A graph is evidence, not decoration. Numerical summaries are purposeful
reductions of that evidence. A good description names what the chosen reduction
reveals and what it omits.

## See and Summarize a Relationship

A relationship asks how values of one variable are arranged with values of
another. The first display follows the variable forms and the question. This
section develops a quantitative--quantitative relationship because it serves
the focal absence--grade question. Grouped distributions and categorical
percentages are introduced later when they can receive full treatment.

### Form before coefficient

Four small displays in @fig-relationship-forms show why *form* comes
before a coefficient.

```{r}
#| label: fig-relationship-forms
#| fig-cap: "Linear, curvilinear, and patternless relationships require different descriptions."
#| fig-alt: "Four panels show a positive line, a negative line, a U-shaped curve, and no clear pattern."
#| fig-height: 4.0

ggplot(relationship_patterns, aes(x = X, y = Y)) +
  geom_point(alpha = .55, size = 1.2, color = "#2F6073") +
  geom_smooth(method = "loess", se = FALSE, linewidth = .8, color = "#A24B2A") +
  facet_wrap(~ Pattern, nrow = 2, scales = "free_y") +
  labs(x = "X", y = "Y")
```

Describe a scatterplot by:

1. **form**: linear, curved, clustered, or otherwise structured;
2. **direction**: increasing, decreasing, or mixed;
3. **strength**: how tightly observations follow the form;
4. **support**: where observations actually occur; and
5. **unusual observations**: points or groups that may change the summary.

*Nonlinear* or *curvilinear* describes form. It does not mean random,
unimportant, or causal. A strong U-shaped relationship can have Pearson's
correlation near zero because its negative and positive linear tendencies
cancel.

### Covariance and Pearson's correlation

For paired observations $(x_i,y_i)$, the sample covariance is

$$
s_{xy}=\frac{1}{n-1}\sum_{i=1}^{n}
(x_i-\bar{x})(y_i-\bar{y}).
$$

Positive covariance means paired deviations tend to have the same sign;
negative covariance means they tend to have opposite signs. Its magnitude
depends on the units of both variables. Pearson's correlation standardizes
covariance:

$$
r=\frac{s_{xy}}{s_xs_y}
=\frac{\sum_i(x_i-\bar{x})(y_i-\bar{y})}
{\sqrt{\sum_i(x_i-\bar{x})^2\sum_i(y_i-\bar{y})^2}}.
$$

The second form follows because the $n-1$ factors cancel. Consider four paired
records with $x=(1,3,3,5)$ practice hours and $y=(2,2,4,4)$ score points.
Both means are 3. The $x$ deviations are $-2,0,0,2$ and the $y$ deviations
are $-1,-1,1,1$. Their cross-products are $2,0,0,2$, summing to 4. Hence

$$
s_{xy}=\frac{4}{3},\quad s_x=\sqrt{\frac83},\quad
s_y=\sqrt{\frac43},\quad
r=\frac{4}{\sqrt{8\times4}}=0.707.
$$

Covariance has hour-point units; dividing by the two SDs cancels those units.
If hours become minutes, each $x$ deviation and $s_x$ is multiplied by 60.
Covariance therefore becomes $60(4/3)=80$ minute-points, but the same factor
cancels in $r$, which stays .707.

Equivalently, write $z_{xi}=(x_i-\bar x)/s_x$ and
$z_{yi}=(y_i-\bar y)/s_y$. Then

$$
r=\frac{1}{n-1}\sum_{i=1}^n z_{xi}z_{yi}.
$$

These are standardized cross-products, with divisor **$n-1$, not $n$**, because
the z scores use sample SDs. Positive products support a positive linear
association and negative products support a negative one. The bound
$-1\le r\le1$ follows from a property of these squared-deviation sums, not
merely from averaging; Appendix C gives a short explanation. Correlation is
unchanged by positive rescaling, reverses sign if one scale is reversed, and
is undefined if either variable has zero SD. Independence implies zero
covariance when second moments exist, but zero covariance does not imply
independence: the curved example already shows why.

| Feature | Sample covariance $s_{xy}$ | Pearson correlation $r$ |
|:--------|:-------------------------------|:--------------------------|
| What it summarizes | Sum of paired-deviation products divided by $n-1$ | Standardized linear direction and strength |
| Units | Product of the two variables' units | None |
| Range | Unbounded; scale dependent | From $-1$ to $1$ |
| Effect of changing measurement units | Magnitude changes | Unchanged by positive linear rescaling |
| What it cannot reveal alone | Curvature, clusters, restricted support, or influential observations | The same omissions; near zero is not proof of no relationship |

: *Covariance and Pearson's correlation have connected but different jobs.* {#tbl-covariance-correlation tbl-colwidths="[23,38,39]"}

For the focal records, the sample covariance is 1.26
absence-count-by-grade-point units and Pearson's $r=.034$. Converting absences
from counts to a positively rescaled unit would change the covariance magnitude
but not $r$. Neither number should be read without the graph.

### The focal relationship

@fig-absence-grade preserves all 395 paired observations. Jitter makes
overlapping integer-valued points visible; it does not change the values used
for the coefficient.

```{r}
#| label: fig-absence-grade
#| fig-cap: "Recorded absences and final mathematics grade in the observational two-school record."
#| fig-alt: "A jittered scatterplot shows most students below twenty absences, a concentrated set at zero absences and zero final grade, a few much larger absence counts up to seventy-five, and little overall linear pattern. A smooth line is nearly flat over the dense region."
#| fig-height: 4.1

ggplot(students, aes(x = absences, y = final_grade)) +
  geom_jitter(width = .35, height = .18, alpha = .48,
              color = "#28536B", size = 1.35) +
  geom_smooth(method = "loess", se = FALSE, color = "#A24B2A",
              linewidth = .9) +
  scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60, 75)) +
  scale_y_continuous(breaks = seq(0, 20, 5), limits = c(-.5, 20.5)) +
  labs(
    x = "Recorded absences (count)",
    y = "Final mathematics grade (0--20)"
  )
```

Most observations occupy low absence counts, several much larger counts extend
the support, and the overall linear pattern is weak. The smooth does not reveal
a strong simple curve that would overturn the near-zero linear summary, but it
is a descriptive check rather than proof that no localized or subgroup pattern
exists.

The concentrated zero set is consequential. Omitting the 38 unresolved
zero-grade rows leaves `r unname(grade_zero_sensitivity["n"])` records and
changes the correlation from $.034$ to
$`r sprintf("%.3f", grade_zero_sensitivity["correlation"])`$. The literal
result is therefore not robust to this coding pattern.
That sensitivity does not establish that omission is correct; it establishes
that the pattern must remain visible in the answer.

A bounded statistical description is:

> Across these 395 paired records, the calculated Pearson correlation between
> absences and final grade was $r=.034$, indicating almost no overall linear
> association. The scatterplot shows dense overlap at low absence counts and a
> few much larger counts. Omitting the 38 source-coded zero grades as a
> diagnostic changes the correlation to $r=-.213$; because the available
> documentation does not explain why every zero grade is paired with zero
> absences, the as-recorded coefficient does not settle a coding-insensitive
> direction or exclude every nonlinear or subgroup pattern.

The substantive interpretation must retain the observational design, recorded
measures, and local scope:

> In these two-school records as coded, absence counts and final mathematics
> grades show little overall linear association. A diagnostic omission of the
> 38 zero-grade records yields a modest negative association, so the unresolved
> coding context is material. Neither result establishes that attendance is
> unimportant, that changing attendance would leave grades unchanged, or that
> the same pattern holds for other students.

### One estimand, complementary checks

The graph, smooth, covariance, and correlation play different supporting roles
in one argument. The declared question-facing association--the Pearson
correlation for the defined group of 395 records--is the estimand. The
calculated correlation, $r=.034$, gives its value for the as-recorded analytic
set. The graph and smooth check whether that scalar summary hides visible
form or support problems. Covariance exposes the scale-dependent quantity
standardized by $r$.

The zero-grade diagnostic changes the analytic set and therefore the
finite-record quantity. It is a sensitivity check on the interpretation of the
primary as-recorded estimand, not a second estimate of that identical estimand.
Adopting the exclusion as the target analysis would require an explicitly
revised record definition and justification.

Several displays, summaries, and checks serving the same estimand do not create
additional question--answer pairs or arguments. A question can also need more
than one estimand: parallel statistical strands can then contribute to its
one integrated substantive answer. A separate argument is needed when the
study owes a separate substantive answer to a different question, not merely
because it reports another statistic. This focal question declares just one
estimand.

### Three necessary criteria for a causal claim

A causal interpretation must address three necessary criteria:

1. **Association.** The proposed cause and outcome differ or co-vary in the
   relevant comparison.
2. **Temporal order.** The proposed cause precedes the outcome over a defensible
   time scale.
3. **Credible handling of alternative explanations.** The design and analysis
   make confounding, selection, measurement, history, and other plausible
   explanations sufficiently implausible for the claim being made.

These are criteria, not three mechanical proofs. The focal records do not
establish when every absence occurred relative to the assessed outcome and do
not credibly rule out alternative explanations. Even a visible association
would not be enough for a causal answer.

For example, prior academic difficulty, motivation, or illness could influence
both later attendance and final performance, creating a **common-cause**
explanation. Earlier low grades might instead lead to later disengagement and
absence, illustrating **reverse direction**. Students represented in these two
schools may also differ from students elsewhere, so **selection** can limit the
scope of the pattern. These are candidate structures to investigate, not facts
established by this file.

Association here means a difference or relationship in the relevant causal
comparison, not necessarily a nonzero Pearson correlation or a statistically
significant result. A near-zero overall correlation can hide a nonlinear
relationship or different effects in subgroups; it does not rule out causation.

**Random assignment** is a powerful design route for the third criterion
because it balances pre-assignment alternatives in expectation and establishes
assignment before an outcome. It does not automatically prove faithful
intervention delivery, valid measurement, harmless attrition or noncompliance,
absence of interference, or transport to other settings. **Random selection**
addresses population representation instead. Confusing the two changes what a
study can claim.

## From One Dataset to Repeated Samples

The focal record addresses a finite-record associational question with
descriptive analysis. Many studies instead use a sample to learn about a wider
population or compare outcomes under a chance-based design. Readers then need
to distinguish three distributions and understand why a statistic varies
across repeated samples.

### Three distributions answer three questions

| Object | What varies | Question answered | Example |
|:-------|:------------|:------------------|:--------|
| Data distribution | Values across observed cases | What pattern is present in these observations? | The 395 recorded grades |
| Probability distribution | Possible outcomes under a stated model | What could happen, and with what modeled chances? | A Normal model or a fair-die model |
| Sampling distribution | A statistic across comparable repeated samples or assignments | How would this statistic vary across repetitions? | Sample means from repeated random samples of size 30 |

: *Data, probability, and sampling distributions answer different questions.* {#tbl-three-distributions tbl-colwidths="[20,25,30,25]"}

A data histogram contains observed individual values. A Normal curve represents
a model for possible values. A sampling distribution contains statistics, such
as sample means, produced across comparable repetitions. A Normal-looking
sampling distribution does not imply that the observed individual values are
Normal.

A probability statement is incomplete without a generating story. The story
might be random sampling, randomized assignment, repeated measurement error, or
an explicitly assumed model. The arrows in @fig-sampling-sequence
describe such a process; they do not assert that the focal school records were
randomly sampled from every population a reader can imagine.

```{r}
#| label: fig-sampling-sequence
#| fig-cap: "Repeated sampling links a defined population or chance process to a sampling distribution."
#| fig-alt: "A defined population or process leads to one sample and one statistic. Repeating that process produces many statistics whose distribution is the sampling distribution."
#| fig-height: 3.0
#| out-width: "92%"

old <- par(mar = rep(.15, 4), xpd = NA)
plot.new()
plot.window(xlim = c(0, 12), ylim = c(0, 4))
box <- function(x, y, w, h, label, fill) {
  rect(x - w/2, y - h/2, x + w/2, y + h/2,
       col = fill, border = "#28424F", lwd = 1.2)
  text(x, y, label, cex = .78)
}
box(1.7, 2, 2.5, 1.0, "Defined population\nor chance process", "#E2EEF3")
box(5.1, 3, 2.0, .9, "One sample", "#E5EBDD")
box(8.1, 3, 2.0, .9, "One statistic", "#F4E5CF")
box(5.1, 1, 2.0, .9, "Many comparable\nsamples", "#E5EBDD")
box(9.7, 1, 3.0, .9, "Sampling distribution\nof the statistic", "#F4E5CF")
arrows(3.0, 2.2, 4.05, 2.8, length = .08, col = "#52666F")
arrows(6.12, 3, 7.05, 3, length = .08, col = "#52666F")
arrows(3.0, 1.8, 4.05, 1.2, length = .08, col = "#52666F")
arrows(6.12, 1, 8.15, 1, length = .08, col = "#52666F")
text(6.8, .35, "Repeat the same stated process", cex = .72, col = "#5A4141")
par(old)
```

### Population, sample, parameter, statistic, and estimand

These terms describe different parts of the question and evidence. They should
not be treated as interchangeable vocabulary.

| Concept | Beginner meaning | Important distinction |
|:--------|:-----------------|:----------------------|
| Population | Defined collection of units, occasions, or outcomes about which a question is posed | Must be named; it is not automatically everyone |
| Sample | Units or outcomes selected or observed under a stated process | The analytic sample may be smaller than the observed record |
| Parameter | Fixed, usually unknown feature of a defined population or probability model | It may be the estimand when the question targets it |
| Statistic | Numerical function calculated from sample data | It becomes an estimate when used to learn about an estimand |
| Estimand | Precisely specified quantity the question seeks | What the study seeks, not the observed numerical answer |
| Estimator | Rule applied to each comparable sample to estimate an estimand | The same rule can produce different estimates |
| Estimate | Observed numerical result used for an estimand | One realized statistic, not its sampling distribution |
| Sampling distribution | Distribution of a specified statistic across comparable repetitions | Its observations are statistics, not people or scores |
| Sampling variability | How a statistic differs across repetitions | Repeat-to-repeat variation, not a data-entry mistake |
| Realized sampling error | Difference between one estimate and its estimand under the stated model | Usually unknown; not systematic bias or measurement error |
| Standard error | Standard deviation of a statistic's sampling distribution, or an estimate of that spread | Variability of a statistic, not spread among individuals |

: *A linked family of sampling concepts.* {#tbl-sampling-concepts tbl-colwidths="[19,39,42]"}

Suppose a defined population contains all 20,000 students enrolled in a system
on a specified date. Its mean grade $\mu$ is a parameter. If the question seeks
that mean, $\mu$ is also the estimand. A probability sample of 100 students
produces a sample mean $\bar{x}$, a statistic used as an estimate. The
**estimator** is the rule "add the selected grades and divide by 100."
Before a sample is drawn, $\bar X$ denotes the value that this rule will
produce; after observing it, $\bar x$ denotes its realized value. Suppose the
observed mean is 10.6 grade points. Repeating the same sample design produces
many sample means, perhaps 10.6, 11.2, 10.9, and so on; their distribution is
the sampling distribution. Its SD is the SE, not the SD among 100 students.

If we knew that the population mean were 11.0, the **realized sampling error**
would be $\bar x-\mu=10.6-11.0=-0.4$ grade points. The negative sign means
this estimate falls below the estimand. In practice that difference is
usually unknown because $\mu$ is unknown. **Sampling variability** describes the
spread across all repetitions. **Systematic bias** concerns a procedure that
tends to miss the estimand in a direction. Undercoverage, nonresponse,
dependence, invalid measurement, and processing mistakes are nonsampling
problems; larger $n$ does not automatically repair them.

For the model illustration that follows, suppose observations are independent
draws from the same distribution, with finite mean $\mu$ and finite variance
$\sigma^2$. Its center and spread can be written

$$
\mu=\mathbb E[X],\qquad
\sigma^2=\operatorname{Var}(X)=\mathbb E[(X-\mu)^2].
$$

Here $\mathbb E$ means **expectation**, the probability-weighted average over
possible outcomes, not an Evidence component such as STAT-E. Variance is the
expected squared distance from the model mean. This is an explicit sampling
model for teaching; it is not a newly assumed population model for the 395
school records. A simple random sample without replacement from a finite list
is a different design, treated briefly in Appendix C.

We can now finish the $n-1$ explanation. Under these independent,
identically distributed, finite-variance conditions, $\mathbb E[s^2]=\sigma^2$:
the conventional sample variance is **unbiased** for the model variance.
Across repeated samples its average equals $\sigma^2$; a particular $s^2$
can be too high or too low. Dividing the squared-deviation sum by $n$ instead
has expectation $(n-1)\sigma^2/n$ and is downward biased for that model
variance. The zero-sum mnemonic and this sampling property answer different
questions; Appendix C provides the short derivation [@moore2014].

### Repeated sampling, standard error, and stabilization

To see repeated sampling clearly, imagine a stable right-skewed population with
mean 10 and standard deviation 10. Take independent random samples of size
$n$, calculate the mean, replace the sample, and repeat 5,000 times.
@fig-sampling-means and @tbl-clt-summary summarize the resulting sample
means.

```{r}
#| label: fig-sampling-means
#| fig-cap: "Sampling distributions of the mean from a right-skewed population narrow and become more symmetric as sample size grows."
#| fig-alt: "Three histograms of five thousand sample means at sample sizes five, thirty, and one hundred share an axis. They remain centered near ten, become narrower, and become more symmetric."
#| fig-height: 3.7

ggplot(sampling_means, aes(x = Sample_mean)) +
  geom_histogram(bins = 38, fill = "#B9D5E5", color = "white") +
  geom_vline(xintercept = 10, linetype = 2, color = "#9B3D45") +
  facet_wrap(~ Sample_size, nrow = 1, labeller = label_both) +
  coord_cartesian(xlim = c(0, 28)) +
  labs(
    x = "Sample mean across repeated samples",
    y = "Number of repetitions"
  )
```

```{=latex}
\Needspace{10\baselineskip}
```

```{r}
#| label: tbl-clt-summary
#| tbl-cap: "Repeated-sampling summaries under the stated right-skewed population model."
#| tbl-colwidths: [14, 22, 23, 22, 19]

sampling_print <- sampling_summary
sampling_print$Center <- sprintf("%.2f", sampling_print$Center)
sampling_print$Empirical_SE <- sprintf("%.2f", sampling_print$Empirical_SE)
sampling_print$Theoretical_SE <- sprintf("%.2f", sampling_print$Theoretical_SE)
sampling_print$Skewness <- sprintf("%.2f", sampling_print$Skewness)
names(sampling_print) <- c(
  "Sample size", "Model SE", "Mean of sample means",
  "Empirical SE", "Skewness"
)
knitr::kable(sampling_print, booktabs = TRUE, align = c("r", "r", "r", "r", "r"))
```

The sample means remain centered near 10. Under the stated independent,
identically distributed model, averaging $n$ observations with expectation
$\mu$ still has expectation $\mu$. Independence makes their variances add,
and dividing a sum by $n$ divides its variance by $n^2$:

$$
\mathbb E[\bar X]=\mu,\qquad
\operatorname{Var}(\bar X)=\frac{n\sigma^2}{n^2}
=\frac{\sigma^2}{n},\qquad
SE(\bar X)=\frac{\sigma}{\sqrt n}.
$$

The **Model SE** column uses that last formula: $10/\sqrt5=4.47$ and
$10/\sqrt{100}=1.00$. The **Empirical SE** column is the observed SD of the
5,000 simulated sample means. Its close, but not exact, agreement with Model
SE is the intended check; a finite simulation introduces some fluctuation.
Neither column is the SD of individual observations, which stays 10 in the
generating model.

Quadrupling sample size halves the SE because $\sqrt{4n}=2\sqrt n$.
When $\sigma$ is unknown, $s/\sqrt n$ estimates the mean's SE under these
conditions. Positively related observations, such as repeated visits from one
person, supply less independent information: the iid formula can then be too
small. A larger row count is not a substitute for the independence assumption.

The **law of large numbers** says, under independent, identically distributed
sampling with a finite mean, that the sample mean becomes increasingly likely
to lie close to the population mean as $n$ grows. In @tbl-clt-summary,
increasing concentration around 10 is the visible lesson.
The law does not force later observations to compensate for earlier ones, make
individual values less variable, or turn a large convenience sample into a
representative one.

The decreasing skewness in the table illustrates a different lesson: in this
example, the sampling distribution of the mean becomes more nearly Normal as
$n$ grows. Under independent, identically distributed sampling with finite,
nonzero variance, the **central limit theorem** explains this approximate
Normal shape after centering the means and scaling by their SE. It supplies
no universal sample-size cutoff. **Concentration** and **approximate shape**
are not the same claim.
Neither theorem repairs poor measurement, dependence, selection, or causal
design.

No interval or p-value is used to convert the fixed observational school record
into a population inference. These concepts prepare the repeated-sampling
foundation needed when later studies develop estimators, estimates, standard
errors, intervals, and tests more fully.

### Complete the bounded answer and make the argument visible

@tbl-ch02-argument is a reading checkpoint, not an evaluation scorecard.
The four rows are argument components. The linked columns show their statistical
and substantive content, followed by representative documentary traces. A
single source may support several components; a component may have several
traces.

| Component | Statistical layer | Substantive layer | Representative documentary trace and status |
|:----------|:------------------|:------------------|:---------------------------------------------|
| Evidence | 395 numeric pairs for absences and final grade; visible distributions; 38 zero grades all paired with zero absences | Two-school student records; documented 0--20 final-grade scale and recorded absence counts; unexplained joint zero pattern | Adapted data file and variable documentation: **located in related materials**; zero-pattern explanation: **not located**; displayed pattern: generated in this chapter from that file |
| Justification | The declared finite-record Pearson correlation matches the requested linear association; the paired-record denominator and coding rule identify the calculation's scope | Recorded variables are interpreted as absence counts and final mathematics grade, within the documented schools and scale | Analysis rationale: stated in this chapter; source meanings: **located in related materials** |
| Inference | The plot and $r=.034$ indicate almost no overall linear association as recorded; $r=-.213$ after the diagnostic omission shows material sensitivity to the zero pattern | The numerical and graphical results are interpreted as local observational descriptions with unresolved coding context, not intervention effects | Figure and summaries: generated in this chapter; zero-pattern explanation: **not located**; interpretation: stated in this chapter |
| Bounded Claim | Almost no overall linear association in all 395 source-coded records, with a modest negative diagnostic after omitting zero grades; no preferred deletion or exclusion of hidden forms | The as-recorded direction does not settle a coding-insensitive attendance--performance relationship; no conclusion that attendance is unimportant, causal, or population-wide | Opening overclaim: hypothetical wording introduced in this chapter; revised sensitivity-bounded answer: constructed in this chapter; no source-paper contradiction is claimed |

: *Reading one associational argument, analysed descriptively, with linked statistical and substantive layers.* {#tbl-ch02-argument tbl-colwidths="[14,27,29,30]"}

The statistical layer establishes what was calculated and inferred from the
record. The substantive layer builds on and contextualizes that support. Its
Justification augments and contextualizes the statistical justification by
connecting the recorded variables, school setting, and question; it does not
duplicate or automatically strengthen the statistical work. The transitions
must preserve magnitude, visible form, design status, and scope.

The opening school report is hypothetical, so this worked example has no source
paper to which the analysis, figures, interpretation, or revised answer can be
attributed. The external documentary traces used here are the adapted data file
and its source documentation; the chapter generates and interprets its own
analysis. In a real study, if a relevant component cannot be found, record
**not located** as the documentary status. That label does not mean the work was
never done. The pair-level phrase **not assessable from the accessible record**
belongs to quality evaluation, not documentary tracing.

::: {.callout-note title="Secondary use: selected quality-evaluation questions"}
Only after reading the argument do we evaluate its support. Four of Chapter
1's six questions are especially informative here:

| Quality-evaluation question | Judgment and reason |
|:----------------------------|:--------------------|
| **Evidence integrity** | **Insufficient information to judge.** The underlying measurement cannot be settled: numeric entries, documented ranges, and the denominator are traceable, but the sources do not explain why all 38 final-grade zeros pair with zero absences. The as-recorded analysis and sensitivity preserve rather than resolve that limitation, and neither establishes wider representation. |
| **Inferential adequacy** | **Adequate for this purpose.** For the as-recorded estimand after reporting sensitivity, the graph, smooth, covariance, and primary correlation address visible form and the declared all-record quantity; the diagnostic omission shows that a coding-insensitive interpretation remains unsupported. |
| **Transition fidelity** | **Needs revision.** The opening move from near-zero recorded correlation to what attendance “matters” for drops the observational and local limits. |
| **Claim calibration** | **Needs revision.** “Attendance does not matter” is stronger and broader than a near-zero finite-record linear association supports. |

These judgments form a profile, not a summed score. The pair-level evaluation
conclusion is **supported only after narrowing or revision**: the same
associational question can be answered defensibly by replacing the opening
overclaim with the bounded local answer above, provided that the unresolved
zero-pattern sensitivity remains attached.
:::

{{< pagebreak >}}

::: {.callout-note title="Auxiliary use: selected construction stages"}
For a researcher conducting a similar study, this chapter illuminates parts of
Stages 3--6:

- **Stage 3:** Align design, measurement, estimand, evidence-production or
  acquisition plan, and inferential plan.
- **Stage 4:** Produce or acquire, and document, empirical and contextual
  evidence.
- **Stage 5:** Estimate, check, and analyse.
- **Stage 6:** Contextualize, interpret, and complete the bounded answer.

Here Stage 5 includes the graph-first description.

Adding jitter or a complementary display after checking the data is a
within-Phase-2 repair: the Phase 1 anchor has not changed. Retroactively
changing the question to a causal effect changes the anchored question, mode,
estimand, and design obligation. The existing cycle must receive an explicit
outcome, and an additional or revised cycle is required. Defensible elements
may be reused only with their provenance and changed status visible.
:::

## Summary and Reading Bridge

A careful reading of an associational question analysed descriptively follows
the support from question to bounded answer:

1. State the question, estimand, and intended answer mode.
2. Identify what each row, variable, value, scale, and denominator represents.
3. See the observations before reducing them.
4. Choose graphs and summaries that fit the question, variable forms, and
   visible shape.
5. Distinguish linear association from curvilinear form and from causation.
6. Distinguish a statistic in one sample from its sampling distribution.
7. Carry measurement, design, uncertainty, and scope limits into the bounded
   answer.

For the focal record, the answer is modest: absences and final grades show
almost no overall linear association among all 395 source-coded records, while
a diagnostic omission of the 38 zero grades gives a modest negative
association. Because the documentation does not explain that joint zero
pattern, neither result licenses the stronger claims that attendance is
unimportant, that changing attendance would not affect grades, or that the same
result holds elsewhere.

Reading the argument is the primary use. Evaluating its quality is a dependent
secondary use, illustrated by four selected questions. The construction
workflow is an auxiliary use for researchers and was invoked only where the
chapter's descriptive work illuminates Stages 3--6.

### Essential glossary

**Case.** The entity or occasion represented by one row.

**Distribution.** The arrangement of values across observed cases, possible
model outcomes, or repeated statistics, depending on the object named.

**Estimand.** The precisely specified quantity the question seeks.

**Estimate.** The observed numerical result used to learn about an estimand.

**IQR.** The distance $Q_3-Q_1$, covering the middle half of ordered
observations.

**Pearson's correlation.** A unit-free summary of linear direction and strength
between two quantitative variables.

**Sample variance and standard deviation.** Squared and original-unit measures,
respectively, of spread around the sample mean.

**Sampling distribution.** The distribution of a specified statistic over
comparable repeated samples or assignments.

**Standard error.** The standard deviation of a statistic's sampling
distribution, or an estimate of that spread.

### Further reading

- The ASA GAISE College Report emphasizes statistical problem solving through
  questions, data, analysis, and interpretation [@asa2016gaise].
- OpenIntro's free materials connect data graphics, numerical summaries, and
  foundations for inference [@cetinkaya2024].
- [NIST's exploratory-data guidance](https://www.itl.nist.gov/div898/handbook/eda/section1/eda15.htm)
  emphasizes graphical examination before model reduction.
- The UCI student-performance source documents the variables and provenance of
  the worked record [@cortez2008data].

## Exercises {.unnumbered}

The exercises are modular and code independent. Use only the evidence provided
unless a prompt explicitly invites outside material.

<!-- prompt-id: CH02-E01 -->
### Exercise 1: Identify the evidence structure (Foundational) {.unnumbered #ex-ch02-01}

A clinic report contains one row per visit. Some patients appear in several
rows. The file records patient ID, clinic, visit date, waiting time in minutes,
and whether the patient returned within 30 days. Of 1,240 visits, waiting time
is missing for 36, return status is missing for 81, and 10 visits are missing
both.

Identify:

1. the case represented by one row;
2. the observational unit or occasion;
3. one identifier, one quantitative variable, and one categorical variable;
4. why 1,240 visits should not be described as 1,240 independent patients;
5. the complete-case denominator for an analysis using waiting time and return
   status; and
6. one question that must be answered before generalizing beyond these visits.

<!-- prompt-id: CH02-E02 -->
### Exercise 2: See and summarize one distribution (Foundational) {.unnumbered #ex-ch02-02}

Eight recorded completion times, in minutes, are:

> 2, 4, 4, 4, 5, 5, 7, 9

1. Sketch a dotplot or ordered display.
2. Calculate the mean, median, sample variance, sample standard deviation, and
   IQR. For quartiles, use the medians of the lower and upper halves.
3. State the units of the variance and standard deviation.
4. If the question asks for a resistant description of typical time and spread,
   choose a pair of summaries and explain why.
5. Name one feature that the chosen pair does not show.

<!-- prompt-id: CH02-E03 -->
### Exercise 3: Read a Normal model and standardized value (Foundational) {.unnumbered #ex-ch02-03}

A reference model for a score is Normal with mean 50 and standard deviation 10.
A learner receives 65.

1. Calculate and interpret the learner's z score.
2. Distinguish this model-based standardization from using a sample mean and
   sample standard deviation.
3. Explain why transforming every observed score to z scores would not make a
   skewed observed distribution Normal.
4. State one question that cannot be answered without additional probability
   information or a Normal table.

<!-- prompt-id: CH02-E04 -->
### Exercise 4: Compare a graph, covariance, and correlation (Foundational) {.unnumbered #ex-ch02-04}

A scatterplot of weekly practice hours and assessment score follows a clear
U-shaped curve. The sample covariance is 0.80 hour-points and Pearson's
correlation is .03.

1. Why can the graph show a strong relationship while $r$ is near zero?
2. What does the positive sign of the covariance summarize?
3. Why is the covariance magnitude not directly comparable across measurement
   units?
4. If hours are converted to minutes, what happens to the covariance and to
   Pearson's correlation?
5. Write one bounded descriptive sentence that respects the graph.

<!-- prompt-id: CH02-E05 -->
### Exercise 5: Apply the three causal criteria (Integrative) {.unnumbered #ex-ch02-05}

An observational study finds that students who voluntarily attend tutoring
have higher end-of-term scores than students who do not. Attendance is recorded
during the term; the outcome is measured at the end.

1. Which of the three necessary causal criteria are partly visible from this
   information?
2. Give two plausible alternative explanations.
3. Explain how random assignment to a tutoring offer could strengthen the
   design.
4. Name two problems random assignment would not automatically solve.
5. Write an associational answer supported by the stated evidence.

<!-- prompt-id: CH02-E06 -->
### Exercise 6: Interpret population, sample, and repeated sampling (Integrative) {.unnumbered #ex-ch02-06}

A school system defines its population as all 20,000 enrolled students on
1 September. It draws a simple random sample of 100 students without replacement and
uses their mean attendance rate to estimate the population mean. Across many
such samples, the mean of the sample means is 91.0%, their standard deviation is
1.2 percentage points, and the realized sample mean is 89.7%.

1. Identify the population, sample, parameter, statistic, estimand, estimator, estimate,
   sampling distribution, sampling variability, realized sampling error, and
   standard error.
2. If the true population mean is 91.0%, calculate the realized sampling error.
3. If the sample size increased from 100 to 400, use the square-root rule to
   approximate the new SE. Explain why this is an approximation for sampling
   without replacement, rather than an exact independent-sampling result.
4. State why a larger sample would not repair an invalid attendance measure or
   an incomplete sampling frame.
5. Relate the increasing concentration of sample means to the law of large
   numbers without using a compensation story.

<!-- prompt-id: CH02-E07 -->
### Exercise 7: Read and then evaluate a bounded argument (Integrative) {.unnumbered #ex-ch02-07}

A report asks: “Among 260 volunteers in an online training course, how strong
is the linear relationship between logged study minutes and final quiz score?”
It documents one row per volunteer, 18 missing quiz scores, a Pearson
correlation of .41 among 242 complete pairs, and a scatterplot showing a
moderate positive linear pattern with two high-leverage observations. The
report concludes: “More study time raises scores for online learners
everywhere.” The course platform's data dictionary is available, but the
recruitment notice and analysis script are not located.

1. State the question--answer pair and its intended estimand for all 260
   volunteers, preserving what the report actually asks and claims.
2. Fill the Evidence, Justification, Inference, and bounded Claim rows for the
   statistical and substantive layers.
3. Identify representative documentary traces and give each a documentary
   status. Record any contradiction separately.
4. State the report's answer before judging it.
5. Apply Evidence integrity, Inferential adequacy, Transition fidelity, and
   Claim calibration using Chapter 1's judgment labels. At this evaluation
   stage, explain whether the complete-pair correlation identifies the
   260-volunteer estimand without additional assumptions.
6. Give one pair-level evaluation conclusion for the original 260-volunteer
   pair and state what, if anything, the accessible record supports. If you
   formulate an answer for the 242 complete pairs, state whether it belongs to
   the original pair or to an additional or revised pair.
7. Is checking the two high-leverage observations a within-Phase-2 repair, or
   does it require an additional or revised cycle? Explain.
8. Would changing the question to the causal effect of assigning extra study
   time preserve the Phase 1 anchor? Explain.

## Appendix A: Conditional Probability and Independence {.unnumbered}

Conditional probability changes the reference set. For events $A$ and $B$,
$A\cap B$ means "both $A$ and $B$ occur," and $P(A\mid B)$ means "the
probability of $A$ given that $B$ occurred."

$$
P(A\mid B)=\frac{P(A\cap B)}{P(B)},\qquad P(B)>0.
$$

The denominator is the number or probability of cases satisfying $B$, not the
whole record. Suppose a study records the following support and pass counts:

| Recorded school support | Passed | Did not pass | Total |
|:------------------------|-------:|-------------:|------:|
| Yes | 14 | 37 | 51 |
| No | 218 | 126 | 344 |
| Total | 232 | 163 | 395 |

: *Hypothetical two-way display of support and pass status for denominator practice.* {#tbl-conditional-example tbl-colwidths="[38,20,22,20]"}

The joint empirical proportion for support and passing is $14/395=.0354$
(3.54%). The conditional proportion passing among support recipients is
$14/51=.2745$ (27.45%). The marginal proportion passing is
$232/395=.5873$ (58.73%). Each answers a different question. These are empirical
probabilities for a uniformly selected record from this table, not automatically
population probabilities.

Events are independent in a probability model when

$$
P(A\cap B)=P(A)P(B),
$$

or, when the conditioning probability is positive,
$P(A\mid B)=P(A)$. An observed difference between proportions is descriptive.
Whether it reflects chance, bias, confounding, or a stable population
difference requires further support. Empirical independence in one finite table
is also different from assuming independence in a probability model.

## Appendix B: Random Variables and Probability-Model Detail {.unnumbered}

A **random variable** assigns a number to each outcome of a stated chance
process. A probability model should name its repeatable story, possible
outcomes, parameters, and conditions.

For example, let $X$ be the number of successful submissions in 10
hypothetical independent attempts, each with constant success probability
$p=.70$. Then

$$
X\sim\operatorname{Binomial}(10,.70),\qquad
P(X=x)={10\choose x}(.70)^x(.30)^{10-x}.
$$

The expression ${10\choose x}$ counts the different ways to place $x$
successes among 10 attempts. For exactly nine successes, any one attempt could
be the failure, giving 10 possible arrangements:
$P(X=9)=10(.70)^9(.30)=.1211$, about 12.1%. For this model,
$\mathbb E[X]=10(.70)=7$ and
$\operatorname{Var}(X)=10(.70)(.30)=2.1$, so its SD is
$\sqrt{2.1}=1.45$ successes. The expectation is a model average, not a promise
that a particular set of 10 attempts contains seven successes.

The model is plausible only if the number of attempts is fixed, each outcome is
meaningfully binary, trials are independent for the purpose at hand, and the
success probability remains constant. Real repeated attempts may violate these
conditions through learning, fatigue, feedback, or shared conditions.
Simulation can expose the implications of assumptions; it cannot validate the
assumptions from the data alone.

A Normal model likewise has a story and parameters. Its symmetric density may
be useful for a measurement model or sampling approximation, but the density
does not prove that observed values are Normal. Tail probabilities answer
model-based questions only after the model and its conditions are defended.

## Appendix C: Stabilization, Sampling Shape, and Formal Results {.unnumbered}

The main text used one repeated-sampling table to distinguish concentration
from approximate sampling shape. A running mean gives another view of
stabilization.

```{r}
#| label: fig-running-mean
#| fig-cap: "One running mean from independent right-skewed draws stabilizes near the population mean of 10."
#| fig-alt: "The running mean varies widely early and gradually settles near a horizontal line at ten by four hundred draws."
#| fig-height: 3.0

ggplot(lln, aes(x = Draw, y = Running_mean)) +
  geom_line(color = "#2F6073", linewidth = .65) +
  geom_hline(yintercept = 10, linetype = 2, color = "#9B3D45") +
  labs(x = "Number of draws included", y = "Running sample mean")
```

Under independent, identically distributed sampling with a finite absolute
mean ($\mathbb E[|X|]<\infty$),
the law of large numbers formalizes the concentration of $\bar X_n$ near
$\mu$ as $n$ increases: $\bar X_n\xrightarrow{p}\mu$. The arrow means
**convergence in probability**: for any fixed positive tolerance, the chance
of falling outside it tends to zero as $n$ grows. It does not say that
short-run imbalances must reverse.

Under common conditions including finite nonzero variance $\sigma^2$, the
central limit theorem describes the standardized sampling shape:

$$
\frac{\bar X_n-\mu}{\sigma/\sqrt{n}}\xrightarrow{d}N(0,1).
$$

The arrow means **convergence in distribution**: the centered mean, expressed
in its own SE units, approaches a standard Normal distribution as $n$ grows.
This is the sampling analogue of a z score, not a transformation of the raw
data into a Normal population. The exact
conditions and the quality of an approximation matter. Dependence, very heavy
tails, complex sampling, or a statistic other than the mean can require
different results.

Here **standardized** means mathematical centering and scaling used to compare
sampling shapes. It is unrelated to the substantive standardization choices
sometimes used in regression. Neither theorem repairs selection, dependence,
measurement, processing, or causal-design weaknesses.

### Why the sample-variance correction works {.unnumbered}

For independent observations $X_1,\ldots,X_n$ with common mean $\mu$ and
variance $\sigma^2$, the squared-deviation decomposition is

$$
\sum_{i=1}^n(X_i-\bar X)^2
=\sum_{i=1}^n(X_i-\mu)^2-n(\bar X-\mu)^2.
$$

Taking expectations gives $n\sigma^2-n(\sigma^2/n)=(n-1)\sigma^2$.
The first term sums $n$ expected squared distances from the population mean;
the second uses the variance of the unbiased sample mean. Dividing by $n-1$
therefore gives $\mathbb E[s^2]=\sigma^2$. This is the expectation calculation
that the zero-sum mnemonic alone does not provide.

The corresponding observed-data identity, for any proposed center $a$, is
$\sum_i(x_i-a)^2=\sum_i(x_i-\bar x)^2+n(\bar x-a)^2$.
Its final term is nonnegative and is zero at $a=\bar x$. Thus the mean
minimizes the sum of squared deviations: the **least-squares center**. This
property prepares the logic of regression without requiring a population model.

### Why correlation is bounded {.unnumbered}

When both sample SDs are positive, $\sum z_{xi}^2=\sum z_{yi}^2=n-1$.
Because a sum of squared numbers cannot be negative,
$\sum_i(z_{xi}-z_{yi})^2=2(n-1)(1-r)\ge0$, giving $r\le1$.
Likewise, $\sum_i(z_{xi}+z_{yi})^2=2(n-1)(1+r)\ge0$, giving $r\ge-1$.
Equality requires perfectly aligned or reversed standardized observations.

### When the independent-sampling SE changes {.unnumbered}

For observations with finite variances, the more general expression is

$$
\operatorname{Var}(\bar X)=\frac{1}{n^2}
\left[\sum_i\operatorname{Var}(X_i)
+2\sum_{i<j}\operatorname{Cov}(X_i,X_j)\right].
$$

The second sum runs over each distinct pair of observations once. Independence
makes all of those covariance terms zero. Positive total covariance increases
the variance relative to the independent case; repeated records from the same
person can therefore yield less precision than their row count suggests.

Sampling without replacement from a fixed list is another case. For a simple
random sample of $n$ units from $N$ units, let $S_N^2$ be the list's squared
deviations from its mean divided by $N-1$. Then

$$
SE(\bar X)=\sqrt{1-\frac nN}\,\frac{S_N}{\sqrt n}.
$$

The square-root factor is the **finite-population correction**: uncertainty
vanishes when every unit is observed. It is unrelated to the $n-1$ denominator
in the sample variance. With 100 of 20,000 students, the factor is
$\sqrt{1-.005}=.9975$; with 400, it is $\sqrt{1-.02}=.9899$.
Ignoring these small corrections makes the simple square-root rule a close
approximation here, but not an exact statement for the without-replacement
design [@moore2014].

## References {.unnumbered}
