---
title: "Descriptive Statistics with R"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-cap: true
    tbl-cap: true
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{env/Shaded/before}{\par\begin{minipage}{\linewidth}}
          \AddToHook{env/Shaded/after}{\end{minipage}\par}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                     collapse = TRUE, comment = "#>")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 4, width = 65)
```

{{< pagebreak >}}

::: {.callout-tip title="Chapter box"}
**Central question:** What do the included observations show about a
distribution, a group comparison or a relationship, and how can we reproduce
and explain that descriptive answer with simple R?

**Focal concepts:** Observations and denominators; mean, median, variance,
SD, quartiles and IQR; graphical distributions; grouped summaries and
conditional percentages; covariance, correlation and paired differences;
elementary R objects, the base R/tidyverse approach and integrated reporting.

**Main evidence:** Small score tables introduce the calculations. Built-in
`airquality` provides 153 historical days, including 116 recorded ozone
values; `mtcars` supplies 32 car-model records; `sleep` provides paired
measurements for ten patients.

**Scope:** Chapter 2 supplies descriptive foundations and Chapter 1 the
reading approach. No previous R programming is required. Statistics organizes
the examples; the lab supplies setup and operating practice. Chapter 6 adds
uncertainty under stated sampling or model assumptions, and Chapter 7 connects
variation and covariance to regression.
:::

# Questions, Records and a First Summary {#sec-ch5-start}

## State what the number describes

A descriptive result answers a question about specified observations.
Before calculating, name the row unit, variable, units and included records.
For three recorded scores of 50, 60 and 70 points, the arithmetic mean is

$$
\bar x=\frac{50+60+70}{3}=60\text{ points}.
$$

R can reproduce that calculation with two short lines:

```{r}
scores <- c(50, 60, 70)
mean(scores)
```

`c()` collects the values into a vector; `<-` gives it the name `scores`.
`mean(scores)` asks a function to calculate its mean. The returned 60 is a
summary of these three observations. A population interpretation needs an
additional question and justification. There is no need for a p-value to
know the mean of the recorded values.

Chapter 2 explained what the summary means. Here the formula, the R command
and the interpretation describe the same calculation. We use that sequence
throughout the chapter: question, observations, calculation and answer.

## R, RStudio and a saved script

R is a language and platform for statistical computing: it stores data,
calculates statistics and draws displays. Statistical programming begins
with short instructions such as the two-line mean calculation above.
An **object** is a named value, vector, table or other result stored by R.
A **function** performs an operation on supplied inputs.

RStudio is an integrated development environment (IDE). Its Source editor
holds a saved `.R` script; the Console runs R commands. The Environment
pane lists current objects, and the Plots and Help panes show figures and
documentation. Saving a script keeps the instructions so that they can be
read, changed and run again. Comments begin with `#`. In printed output,
`[1]` marks the position of the first displayed element; it is not an answer
or a command to type. The lab demonstrates the interface and file steps.

## Values, vectors and function arguments

A vector contains values of one basic type. Scores are **numeric**;
student names such as `"A"` are **character** values; a comparison returns
**logical** values, `TRUE` or `FALSE`. Quotation marks distinguish text
from an object name. `NA` denotes an unavailable value. Putting a word into
a numeric vector, as in `c(50, "absent")`, converts it to character data;
use `NA` for a missing score and retain its explanation separately.

```{r}
#| label: ch5-vector-basics
scores + 5
scores[c(1, 3)]
scores > 55
scores[scores > 55]
```

The first operation adds five to each score, giving 55, 65 and 75; it
does not replace `scores` unless the result is assigned. Positions start
at 1, so the second line selects 50 and 70. The condition gives `FALSE`,
`TRUE`, `TRUE`, and the final line keeps 60 and 70. These are different
observations from the original three-score summary.

Arithmetic uses `+`, `-`, `*`, `/` and `^`. Parentheses specify order:
`2 + 3 * 4` gives 14, whereas `(2 + 3) * 4` gives 20. This matters when
translating a statistical formula into R. Name intermediate quantities while
translating a formula; doing so makes the calculation easier to read.

An **R function argument** specifies a function's input or an option. This
programming term is distinct from a **study argument**: the complete
question-level reasoning for one question--answer pair. For example,
`mean(scores)` and `mean(x = scores)` use the same input; `round(60.25,
digits = 1)` requests one decimal place. Keep intermediate results
unrounded and round the reported answer. R is case-sensitive: `scores`
and `Scores` are different names. Type `?mean` in the Console and read
the help page's purpose, arguments and examples. We introduce the
missing-value argument `na.rm` beside its first calculation below.

## Base R and a small selection of tidyverse tools

**Base R** here means the functions supplied with an ordinary R
installation, including its standard statistics packages. The
**tidyverse** is a family of additional packages with related conventions.
They work within the same R language. The course combines direct base R
calculations with a small selection from three packages:

| Tool | Role in this course |
|:--|:--|
| Base R | Arithmetic, vectors, tables, indexing and ordinary statistics |
| ggplot2 | Graphics from data, mappings and layers |
| dplyr | Selected table operations, grouped summaries and joins |
| tidyr | Long/wide reshaping, including matching paired measurements |

Installing a package makes it available on the computer. `library()`
loads an installed package for the current session. The expression
`dplyr::filter()` names a function and its package explicitly, even
without loading the package's function names. The lab explains both
forms. `ggplot2` is a package; `ggplot()` is one of its functions.

We show one small selection in both styles in @sec-ch5-inspect,
then use the clearest short expression for each statistical task.
There is no requirement to implement every analysis in two styles or
to load the entire tidyverse. Base R remains the main tool for the
descriptive calculations.

## Where Quarto fits

A script stores calculations. **Quarto** connects those calculations
with prose, mathematics, tables, figures and references in a `.qmd`
source document. It can render that source to PDF, HTML or Word.
PDF is the main professional-report format here, with HTML as a useful
second format. RStudio can edit the source and start a render; Quarto
is a separate application. For R chunks it uses **knitr** to run code
and insert results. PDF typesetting also needs a TeX installation,
such as TinyTeX.

@sec-ch5-report shows a small report source. The lab collects
one-time setup at the beginning and gives the optional production steps
later. Understanding these roles belongs in the chapter; producing a
complete report need not occupy the classroom statistics session.

# Select Observations and Preserve the Measurement {#sec-ch5-inspect}

## Rows, columns and missing values

A data frame is a table whose columns may contain different types of values.
Here each row is a student; the two score columns are measurements on that
same student. Programme codes label categories, not quantities.

```{r}
small <- data.frame(
  student = c("A", "B", "C"),
  program_code = c(1, 2, 2),
  baseline = c(50, 60, 70),
  followup = c(55, NA, 78)
)
small
```

`NA` means a value is unavailable. It is not zero. `$` selects a column
by name, so `small$baseline` is the vector of three baseline scores.
The available follow-up mean uses two observations:

```{r}
mean(small$followup, na.rm = TRUE)
sum(!is.na(small$followup))
```

The mean is 66.5 points and its denominator is two. `is.na()` identifies
missing entries; `!` reverses a logical condition. `na.rm = TRUE` asks the
function to exclude missing values from this calculation. It does not
recover their values or make the included students representative of all three.

For change, subtract within the same student:

```{r}
small$change <- small$followup - small$baseline
mean(small$change, na.rm = TRUE)
```

The changes are 5 and 8, giving 6.5 points for two complete pairs. In this
particular table, subtracting the separate available-case means also gives
6.5, but the baseline mean uses three students. Numerical agreement does
not establish matching denominators. A paired calculation should explicitly
retain the same people for both measurements.

## Inspect the observations needed for a summary

The built-in `airquality` data set records daily measurements in New York from
May to September 1973. The [dataset documentation](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/airquality.html)
defines ozone in parts per billion (ppb), wind in miles per hour and maximum
daily temperature in degrees Fahrenheit. A row is a day, not a person.

```{r}
aq <- airquality
head(aq)
names(aq)
dim(aq)
str(aq)
sum(is.na(aq$Ozone))
required_columns <- c("Ozone", "Solar.R", "Month", "Temp")
stopifnot(all(required_columns %in% names(aq)))
stopifnot(all(aq$Month >= 5 & aq$Month <= 9))
```

There are 153 rows and six columns, including 37 missing ozone values.
`head()` shows the first rows, `names()` gives the column names,
`dim()` gives the row and column counts, and `str()` displays the structure
and column types. `summary(aq)` gives a first statistical overview. These
complementary views help identify the measurements before calculation.
The two `stopifnot()` checks stop execution if a required column is absent or
if a month code falls outside the documented May--September range. Passing
these checks shows only that two declared structural expectations hold; it
does not prove that every value is accurate or that the record is suitable for
the question.
Keep an explicit set for the ozone question:

```{r}
ozone_rows <- !is.na(aq$Ozone)
aq_observed <- aq[ozone_rows, ]
ozone <- aq_observed$Ozone
length(ozone)
100 * length(ozone) / nrow(aq)
```

In `table[rows, columns]`, leaving the column position blank retains all
columns. The ozone mean will use 116 recorded values. The recorded-value
percentage is $100(116/153)=75.8\%$, using all days as its denominator.
Different questions can require different denominators on the same table.

For two variables, select rows with both measurements. A missing value in
an unrelated variable need not remove a row. For example:

```{r}
both <- complete.cases(aq[c("Ozone", "Solar.R")])
sum(both)
```

This comparison has 111 complete days. State the variables required by a
calculation before choosing its rows. Complete records are an inclusion
rule, not evidence of random sampling or absence of selection bias.

## One selection in base R and dplyr

Both styles must answer the same question using the same missing-value
rule. To keep July days with recorded ozone:

```{r}
#| label: ch5-base-tidy-selection
july_rule <- aq$Month == 7 & !is.na(aq$Ozone)
july_base <- aq[july_rule, ]
july_tidy <- dplyr::filter(aq, Month == 7, !is.na(Ozone))
nrow(july_base)
nrow(july_tidy)
mean(july_tidy$Ozone)
```

Each result has the same 26 days and mean ozone 59.12 ppb. `==` tests
equality, `&` requires both conditions, and `!` reverses the missingness
test. Base indexing names columns through `aq$`; `filter()` reads the
column names within its supplied table. Its comma-separated conditions
must all hold.

The native pipe `|>` passes its left-hand result as the first argument
of the function on the right. Thus `aq |> dplyr::filter(Month == 7,
!is.na(Ozone))` expresses the same selection. A short pipe can make a
sequence readable, but a direct call is often enough. It differs from
ggplot's `+`, which adds a graphical layer. Keep the statistical inclusion
rule explicit whichever notation is used.

## Transform units without changing the measurement

Convert maximum temperature to Celsius while retaining the original:

```{r}
aq$temp_c <- (aq$Temp - 32) * 5 / 9
mean(aq$Temp)
mean(aq$temp_c)
sd(aq$Temp)
sd(aq$temp_c)
```

For $Y=a+bX$, the mean becomes $a+b\bar x$, the SD becomes $|b|s_X$
and the variance becomes $b^2s_X^2$. Subtracting 32 shifts the location;
multiplying by $5/9$ changes the numerical scale. The same daily maximum
temperatures have been expressed in different units. All 153 temperatures
are recorded, so these means use more days than the ozone mean.

# Describe a Distribution {#sec-ch5-summaries}

## Center and variation answer different questions

For observed values $x_1,\ldots,x_n$, the mean and sample variance are

$$
\bar x=\frac{1}{n}\sum_i x_i,\qquad
s^2=\frac{\sum_i(x_i-\bar x)^2}{n-1},\qquad s=\sqrt{s^2}.
$$

The mean locates the center. The squared deviations quantify dispersion;
their sum is divided by $n-1$ in the sample-variance convention used by R.
The deviations sum to zero, leaving $n-1$ freely varying deviations.
Under independent sampling from a common finite-variance distribution,
this denominator makes $s^2$ unbiased for the process variance. Using it
on a fixed file does not establish that sampling model.

For 50, 60 and 70, the squared deviations sum to 200. Thus $s^2=100$
square points and $s=10$ points. Variance has squared units; SD has the
original units. If the intended descriptive variance divides by the full
record count $n$, name that convention: here it is $200/3$.

```{r}
mean(scores)
sum((scores - mean(scores))^2)
var(scores)
sd(scores)
```

The median is the middle ordered value, or the mean of the middle two
when the count is even. It is less sensitive than the mean to a very large
observation. Replacing 70 by 100 changes the mean to 70 while the median
remains 60; the SD rises to about 26.46.

```{r}
changed_scores <- c(50, 60, 100)
mean(changed_scores)
median(changed_scores)
sd(changed_scores)
```

An extreme observation is a feature to investigate, not automatically an
error to delete. Check its measurement and relevance before changing the
analytic records. A distribution can have the same mean as another while
having a different spread or shape.

## Quartiles, IQR and the numerical convention

The first and third quartiles locate the lower and upper quarters of an
ordered distribution. The interquartile range is $IQR=Q_3-Q_1$ and describes
the spread of its central half. The range uses the two most extreme values
and is especially sensitive to them.

```{r}
mean(ozone)
median(ozone)
sd(ozone)
quantile(ozone, probs = c(0, .25, .5, .75, 1), type = 7)
IQR(ozone, type = 7)
```

The 116 ozone values have mean 42.13 ppb, median 31.5 ppb and SD 32.99 ppb.
Their minimum, first quartile, median, third quartile and maximum are
1, 18, 31.5, 63.25 and 168 ppb. The IQR is 45.25 ppb.

R's default `type = 7` uses interpolation. For sorted values, its position
is $h=1+(n-1)p$. At $p=.25$ for `c(1, 3, 3, 5)`, $h=1.75$, giving
$Q_1=1+.75(3-1)=2.5$. At $p=.75$, $Q_3=3+.25(5-3)=3.5$, so the IQR
is 1. Chapter 2's medians-of-halves calculation gives an IQR of 2.
State the convention when reproducing a small-sample answer.

## Read the shape alongside the numbers

```{r}
#| label: fig-ch5-ozone
#| fig-cap: "The distribution of the 116 recorded ozone values."
#| fig-alt: "Ozone values cluster at the lower end, with a long right tail extending to 168 parts per billion."
#| fig-width: 6
#| fig-height: 3.2
library(ggplot2)
ggplot(aq_observed, aes(x = Ozone)) +
  geom_histogram(binwidth = 10, color = "white") +
  labs(x = "Ozone (ppb)", y = "Days")
```

`ggplot()` identifies the data, `aes()` maps a variable to an axis,
`geom_histogram()` groups values into bins and `labs()` supplies labels.
The `+` adds a plot layer. The right tail helps explain why the mean exceeds
the median. Changing bin width changes the display, not the observations
or their arithmetic mean.

```{r}
#| label: fig-ch5-ozone-box
#| fig-cap: "A boxplot summarizes the center, middle spread and unusually high ozone values."
#| fig-alt: "The median is nearer the lower end of the ozone distribution and several high values appear beyond the upper whisker."
#| fig-width: 6
#| fig-height: 2.8
ggplot(aq_observed, aes(x = Ozone, y = "Recorded days")) +
  geom_boxplot() +
  labs(x = "Ozone (ppb)", y = NULL)
```

The box displays quartiles and the median. Each whisker reaches the most
distant observation no more than 1.5 times the IQR below $Q_1$ or above
$Q_3$; values beyond those fences are drawn separately. These are display
flags, not a
decision to remove data. Small-
sample boxplot conventions may differ from a hand quartile calculation.
Always retain the units and inspect the values behind an unusual point.

# Compare Groups and Conditional Percentages {#sec-ch5-groups}

## Read the group count beside its mean

To compare recorded ozone across months, label the month codes and calculate
each group's count, mean and SD. `factor()` marks a categorical variable;
`tapply(values, groups, function)` applies one statistical function to each
group. This short form keeps the variable and grouping explicit.

```{r}
aq_observed$month <- factor(aq_observed$Month,
  levels = 5:9, labels = c("May", "June", "July", "August", "September"))
table(aq_observed$month)
tapply(ozone, aq_observed$month, mean)
tapply(ozone, aq_observed$month, sd)
```

The counts are 26, 9, 26, 26 and 29. Their means are approximately 23.62,
29.44, 59.12, 59.96 and 31.45 ppb. June's summary describes nine recorded
days, not all 30 June days. A gap between the observed monthly means does
not by itself isolate an effect of season or establish population uncertainty.

The same grouped operation can be made explicit with a short dplyr sequence:

```{r}
monthly_ozone <- aq_observed |>
  dplyr::group_by(month) |>
  dplyr::summarise(
    n = dplyr::n(),
    mean_ozone = mean(Ozone),
    sd_ozone = sd(Ozone),
    .groups = "drop"
  )
monthly_ozone
```

`group_by(month)` marks which rows belong together; it does not yet collapse
the data. `summarise()` then returns one row per month, `n()` counts the rows
contributing under the current ozone inclusion rule, and `.groups = "drop"`
removes the grouping structure from the result. The new row therefore means a
month summary, not an individual day, and its `n` is the relevant denominator.

```{r}
#| label: fig-ch5-months
#| fig-cap: "Recorded ozone distributions by month; each box uses that month's available measurements."
#| fig-alt: "July and August have higher median ozone than May, June and September, with different spreads and only nine observed days in June."
#| fig-width: 6
#| fig-height: 3.4
ggplot(aq_observed, aes(x = month, y = Ozone)) +
  geom_boxplot() +
  labs(x = "Month", y = "Ozone (ppb)")
```

The overall mean of individual observations weights each group mean by
its contributing count:

$$
\bar x=\frac{\sum_g n_g\bar x_g}{\sum_g n_g}.
$$

Two observations with mean 20 and eight with mean 50 have overall mean
$(2\times20+8\times50)/10=44$. The unweighted mean of 20 and 50 is 35;
it gives equal weight to groups. These summaries answer different questions.

## Choose the denominator of a conditional percentage

The [mtcars documentation](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html)
describes 32 car models from 1973-74. `mpg` is miles per US gallon and
`am` is automatic (0) or manual (1) transmission. Define 20 mpg as a
teaching threshold and compare its frequency by transmission:

```{r}
cars <- mtcars
cars$transmission <- factor(cars$am, levels = c(0, 1),
  labels = c("Automatic", "Manual"))
cars$mpg20 <- factor(cars$mpg >= 20, levels = c(FALSE, TRUE),
  labels = c("Below 20", "At least 20"))
tab <- table(cars$transmission, cars$mpg20)
tab
```

There are 19 automatic models (15 below 20 and 4 at least 20) and
13 manual models (3 below 20 and 10 at least 20). All 32 have the needed
measurements. A row percentage conditions on transmission:

```{r}
100 * prop.table(tab, margin = 1)
100 * prop.table(tab, margin = 2)
```

Among automatic models, $100(4/19)=21.05\%$ meet the threshold; among manual
models, $100(10/13)=76.92\%$ do. The manual-minus-automatic difference is
55.87 percentage points. Among the 14 models meeting the threshold,
$100(10/14)=71.43\%$ are manual. The latter uses a column denominator and
answers a different question. The row and column labels explain the output
from `margin = 1` and `margin = 2`.

These are observed associations among car models. Cars with different
transmissions also differ in weight and other features. Chapter 6 adds a
chi-square reference under a stated model; that reference does not identify
the causal effect of changing a transmission.

# Describe Quantitative Relationships {#sec-ch5-association}

## Covariance retains units; correlation standardizes them

For the same $n$ complete pairs of numerical measurements,

$$
s_{XY}=\frac{\sum_i(x_i-\bar x)(y_i-\bar y)}{n-1},\qquad
r=\frac{s_{XY}}{s_Xs_Y}.
$$

Covariance averages the cross-products of centered observations with the
sample denominator. Positive cross-products occur when the two measurements
are on the same side of their respective means. Covariance has product units;
correlation is unitless and lies between -1 and 1 when both SDs are positive.

For $x=(1,2,3)$ and $y=(2,4,6)$, the centered cross-products sum to 4,
so covariance is $4/(3-1)=2$. Their SDs are 1 and 2, giving $r=1$.

```{r}
pair_rows <- complete.cases(cars[c("wt", "mpg")])
car_pairs <- cars[pair_rows, ]
x <- car_pairs$wt
y <- car_pairs$mpg
length(x)
cov(x, y)
cor(x, y)
```

Weight is recorded in thousands of pounds. The 32 pairs have covariance
$-5.1167$ (thousand pounds times mpg) and correlation $-.8677$. This
describes a strong negative linear association in these records.

\Needspace{24\baselineskip}

```{r}
#| label: fig-ch5-car-pairs
#| fig-cap: "Weight and fuel economy for the same 32 car-model records used in the correlation."
#| fig-alt: "Heavier car models generally have lower fuel economy, with variation around the negative pattern."
#| fig-width: 6
#| fig-height: 3.4
ggplot(car_pairs, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(x = "Weight (thousand pounds)", y = "Fuel economy (mpg)")
```

Converting weight to pounds multiplies its covariance with mpg by 1000,
but leaves the correlation unchanged. Adding a constant changes neither
covariance nor correlation. Multiplying one variable by a negative constant
reverses the correlation's sign. Correlation requires paired measurements,
not independently sorted columns; use the same complete rows for the
covariance and both SDs.

## A small correlation can miss a strong curved pattern

Consider five fixed pairs on a curve:

```{r}
curve <- data.frame(x = c(-2, -1, 0, 1, 2))
curve$y <- curve$x^2
cov(curve$x, curve$y)
cor(curve$x, curve$y)
```

Both covariance and correlation are zero. Nevertheless, each $y$ is exactly
$x^2$. Positive and negative centered cross-products cancel by symmetry.
The coefficient summarizes linear association; it does not summarize every
possible relationship.

```{r}
#| label: fig-ch5-curve
#| fig-cap: "Five observations with zero Pearson correlation and an exact curved relationship."
#| fig-alt: "The points at x equal to minus two, minus one, zero, one and two form a U, with y values four, one, zero, one and four."
#| fig-width: 5.5
#| fig-height: 3.1
ggplot(curve, aes(x, y)) +
  geom_point() +
  geom_line() +
  labs(x = "X", y = "Y = X squared")
```

Read form, direction, spread and unusual observations from the display,
then interpret the numerical coefficient in that context. A correlation
test in Chapter 6 addresses a model's zero linear-correlation question;
it cannot recover a curved pattern discarded by the coefficient.

# Describe Paired Differences {#sec-ch5-reshape}

## Align the measurements before subtracting

The [sleep documentation](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/sleep.html)
describes extra hours of sleep relative to control under two drugs for ten
people. `ID` identifies a person, `group` identifies the drug condition and
`extra` is the measured extra sleep. One original row is a person-condition
measurement. Twenty condition rows therefore represent ten people.

```{r}
table(sleep$ID, sleep$group)
sleep_wide <- tidyr::pivot_wider(sleep, id_cols = ID,
  names_from = group, values_from = extra, names_prefix = "drug_")
sleep_wide$difference <- sleep_wide$drug_2 - sleep_wide$drug_1
head(sleep_wide, 3)
```

`pivot_wider()` puts each ID's two measurements on one row. Its arguments
name the person, condition and measurement. This operation preserves pairing
even if the original rows are not ordered by condition. It does not resolve
duplicate measurements for an ID-condition combination; those need review.
The first difference is $1.9-.7=1.2$ hours. Drug 2 minus drug 1 is the
chosen direction throughout Chapters 5 and 6. These are not baseline and
follow-up measurements, and negative `extra` means less sleep than control.

## Describe the distribution of differences

```{r}
complete_pair <- complete.cases(sleep_wide[c("drug_1", "drug_2")])
difference <- sleep_wide$difference[complete_pair]
length(difference)
mean(difference)
sd(difference)
range(difference)
sum(difference > 0)
```

All ten pairs are complete. Their differences sum to 15.8 hours, giving a
mean of 1.58 hours and an SD of 1.23 hours. Nine are positive, one is zero and
none is negative; the range is 0 to 4.6 hours. The mean describes the average
difference and the SD describes variation among the differences.

When both condition means use the same people,

$$
\overline{y_2-y_1}=\bar y_2-\bar y_1,
\qquad
s_D^2=s_2^2+s_1^2-2s_{12}.
$$

The condition means are .75 and 2.33 hours, whose difference is 1.58.
Their SDs are about 1.79 and 2.00, and their covariance is about 2.85.
The positive within-person covariance reduces the variance of differences
relative to the sum of the marginal variances. R can check the identity:

```{r}
paired <- sleep_wide[complete_pair, ]
mean(paired$drug_2) - mean(paired$drug_1)
var(paired$drug_2) + var(paired$drug_1) -
  2 * cov(paired$drug_2, paired$drug_1)
var(difference)
```

The equality requires the same pairs for all terms. Subtracting two
available-case means can compare different people when a measurement is
missing. Correlation between conditions and the mean condition difference
answers a different question: closely related measurements need not
have equal means.

```{r}
#| label: fig-ch5-paired
#| fig-cap: "Extra sleep under two drugs for the same ten people; the diagonal marks equal condition values."
#| fig-alt: "Nine paired points lie above the equality diagonal and one lies on it, with varying distances from equality."
#| fig-width: 6
#| fig-height: 3.5
ggplot(sleep_wide, aes(x = drug_1, y = drug_2)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = 2) +
  labs(x = "Drug 1: extra sleep (hours)",
       y = "Drug 2: extra sleep (hours)")
```

The diagonal is an equality reference, not a fitted regression line.
An observed mean of 1.58 hours does not mean every person has that difference.
Chapter 6 asks a further question about a process mean and introduces its
SE and confidence interval under stated conditions.

# Report the Descriptive Answer {#sec-ch5-report}

## Connect the number to the observations

A short descriptive report can combine center, spread and scope:

> Ozone was recorded on 116 of 153 days in the New York May-September 1973
> dataset. Among those recorded days, mean ozone was 42.13 ppb, median
> 31.5 ppb, SD 32.99 ppb and IQR 45.25 ppb using R's type-7 quartiles.
> The distribution had a right tail. Monthly means were accompanied by
> their available counts; June contributed nine observed values.

Read the reported answer first. The data and inclusion rule identify the
evidence, the definitions justify the arithmetic, and the calculations
support a claim about the included days. A separate quality evaluation asks
whether the definitions, records and scope fit the question. The report
does not establish the ozone distribution for all unrecorded days, later
years or other cities. Computing more decimal places would not resolve
those representation questions.

Keep the commands producing a report in a short script. A useful report
names the data, variables, units, missing-value rule, denominator and
statistical convention. A Quarto document can connect those commands to
prose and figures; the optional report and lab show the operating steps.

## Combine a descriptive result into a Quarto report

If your instructor provides the optional classroom companion, obtain
`quarto_example.qmd` and `quarto_example.bib` from the course space and save
them beside each other. The source illustrates a report containing text,
mathematics, R code, tables and graphics; the bibliography stores the reference
records. Its data are built into R, so no separate data file is required.

A source begins with a YAML header. A compact version states its title,
output formats and bibliography:

```yaml
---
title: "Recorded ozone in New York"
format:
  pdf: default
  html: default
bibliography: quarto_example.bib
---
```

The source body uses ordinary Markdown and mathematical notation:

```markdown
## Data and summary

We describe days with a recorded ozone measurement.
The observed-value mean is
$$
\bar{x} = \frac{1}{n}\sum_{i=1}^{n}x_i.
$$
```

An executable R chunk opens with three backticks followed immediately by
`{r}`, and closes with three backticks. Place these option lines and commands
between those fences:

```r
#| label: tbl-ozone-summary
#| tbl-cap: "Observed ozone summary."
aq <- airquality
observed <- aq[!is.na(aq$Ozone), ]
n_observed <- nrow(observed)
mean_ozone <- mean(observed$Ozone)
knitr::kable(
  data.frame(n = n_observed, mean_ppb = mean_ozone),
  digits = 2
)
```

When rendered, the chunk produces 116 observations, a mean of 42.13 ppb,
and a numbered table. In the source sentence, put the engine marker `r`, one
space, and `round(mean_ozone, 2)` between single backticks to insert the
computed mean.
Write `Table @tbl-ozone-summary` to refer to the labelled table by its current
number. A later chunk can draw the histogram with ggplot2 and use a `fig-`
label in the same way. A citation key such as `[@peng2011]` connects a sentence
to its bibliography entry.

The optional example declares an online APA citation style, so it needs
no separate local style file. Its full header adds presentation settings.
Rendering runs the calculations and produces the chosen document format.
When a calculation changes, render again so that the numbers, table,
figure and explanation can be read together. A report still needs a
clear question, denominator, units and an appropriately limited answer.
The lab gives the exact PDF and HTML commands and a short guided edit.

## Summary and Reading Bridge {#sec-ch5-summary}

The statistical sequence is question, observations, summary, display and
bounded interpretation. Simple R makes that sequence reproducible. Different
summaries answer different questions: SD describes dispersion, conditional
percentages depend on their denominator, correlation concerns linear form
and a paired difference concerns measurements on the same unit.

Chapter 2 supplied these descriptive ideas. Chapter 6 adds sampling or
working-model uncertainty to an explicitly stated estimand, using the same
paired example and basic group comparisons. Chapter 7 connects covariance
and variance to the regression slope; Chapter 8 requires consistent analytic
records when comparing adjusted models. None of these steps makes a
descriptive record summary automatically representative or causal.

Use local help such as `?sd`, `?quantile` and `?cor` for computational
conventions, and @rcore2025 for R's reference system. The chapter explains
the statistical reasons; the lab provides practice with the short commands.

\Needspace{8\baselineskip}
\phantomsection

# Exercises {.unnumbered}

Use the stated records and conventions. Explain the statistical meaning as
well as the calculation. Report numerical results to the requested precision.

{{< pagebreak >}}

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH05-E01 -->

## Exercise 1: Count the right observations (Foundational) {.unnumbered #ex-ch05-01}

A table contains 40 students with one baseline and one follow-up row each,
plus one duplicate follow-up row. There are 81 rows. Explain the difference
between a row count and a student count. If four students have missing
follow-up scores, give the denominator for the mean observed follow-up score
after the documented duplicate is resolved. State what the mean describes.

<!-- END EX05-01 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH05-E02 -->

## Exercise 2: Center, spread and a changed observation (Foundational) {.unnumbered #ex-ch05-02}

Three recorded scores are 50, 60 and 70 points. Calculate the mean, median,
sample variance and sample SD. Replace 70 by 100 and calculate the new mean,
median and SD. Explain what changed statistically. Use the sample-variance
denominator $n-1$ and report SDs to two decimal places.

<!-- END EX05-02 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH05-E03 -->

## Exercise 3: A missing score is not zero (Foundational) {.unnumbered #ex-ch05-03}

Students A, B and C have baseline scores 50, 60 and 70, and corresponding
follow-up scores 55, missing and 78. Calculate the available follow-up mean,
the complete-pair mean change and the difference of the separate available
means. State each denominator. Explain why equality of two numerical answers
would not establish that they used the same student sets.

<!-- END EX05-03 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH05-E04 -->

## Exercise 4: Choose the group denominator (Foundational) {.unnumbered #ex-ch05-04}

A table classifies 19 automatic car models, of which 4 meet a fuel-economy
threshold, and 13 manual models, of which 10 meet it. Calculate the percentage
meeting the threshold within each transmission group and their manual-minus-
automatic percentage-point difference. Then calculate the percentage of
threshold-meeting models that are manual. Explain the different denominators.

<!-- END EX05-04 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH05-E05 -->

## Exercise 5: Read a relationship before evaluating it (Integrative) {.unnumbered #ex-ch05-05}

For five paired observations, $x=(-2,-1,0,1,2)$ and $y=(4,1,0,1,4)$.
Calculate covariance and Pearson correlation and describe the scatterplot.
A report says, "The correlation is zero, so the variables are unrelated."
First identify its numerical result and reported conclusion, then evaluate
whether that conclusion follows. What happens to covariance and correlation
if $y$ is multiplied by 10?

<!-- END EX05-05 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH05-E06 -->

## Exercise 6: Report paired differences (Integrative) {.unnumbered #ex-ch05-06}

Four people have condition-1 values 1, 2, 4 and 5 and corresponding
condition-2 values 2, 4, 4 and 8 hours. Calculate the mean and sample SD
of condition 2 minus condition 1, and verify the difference-of-means identity.
Write a descriptive result that names the people, direction and units.
Explain why it does not say that every person gained the mean difference.

<!-- END EX05-06 PROMPT -->


\Needspace{8\baselineskip}
\phantomsection

# Appendix A: Supporting Table Operations {.unnumbered #sec-ch5-app-a}

These optional operations support more complicated data layouts. The main
statistical route requires only the short ID-based pivot already shown.

\Needspace{6\baselineskip}
\phantomsection

## Return to a long layout when it is useful {.unnumbered}

For a display or calculation that uses one outcome column and a condition
column, `pivot_longer()` reverses this change of layout:

```{r}
sleep_long <- tidyr::pivot_longer(
  sleep_wide[c("ID", "drug_1", "drug_2")],
    cols = c(drug_1, drug_2),
    names_to = "drug",
    values_to = "extra"
  )
head(sleep_long)
```

`cols` names the measurement columns to stack; `names_to` and
`values_to` name the two resulting columns. Selecting the two drug
columns first keeps the derived difference out of the stack: it is a
comparison, not a third drug condition. The long result again has
20 measurement rows.

\Needspace{6\baselineskip}
\phantomsection

## Add information from a lookup table {.unnumbered}

A **join** combines information using a shared identifier or **key**.
Suppose programme labels are held in a separate lookup rather than
written directly into a factor:

```{r}
student_codes <- small[c("student", "program_code")]
program_lookup <- data.frame(
  program_code = c(1, 2),
  program = c("Standard", "Practice")
)
student_codes
program_lookup
labelled_students <- dplyr::left_join(
  student_codes, program_lookup, by = "program_code"
)
labelled_students
```

The argument `by` names the shared key. A `left_join()` keeps the rows
from the left table and adds matching information from the right.
There are still three students. Students B and C both receive the
Practice label because both have code 2.

This is a **many-to-one** relationship: many students may have one code,
but each code should identify one row in the lookup. A simple count helps
check that expectation:

```{r}
table(program_lookup$program_code)
```

Both codes occur once. An unmatched code in the student table would remain
in a left join with a missing programme label. That differs from a repeated
lookup key, which can multiply rows.

For example, suppose students A and B attend school S1 and student C
attends S2. If a school lookup has two S1 rows and one S2 row, an
unrestricted left join produces five rows: two for A, two for B and
one for C. It has not created two more students. Check the school lookup
key and resolve what the repeated rows mean; taking the first three
joined rows would arbitrarily lose information.

When this expectation is central to a task, `left_join()` can also
express it with the argument `relationship = "many-to-one"`. It reports
an error if a left-hand row matches more than one right-hand row.
The declaration makes the expected relationship explicit; the key's
meaning still comes from the study.


\Needspace{6\baselineskip}
\phantomsection

## Files and a short reproducible report {.unnumbered}

A saved R script records the analytical steps. For a simple CSV, use
`read.csv()` and inspect the imported columns before calculating. The student
lab provides an optional read/write example and Quarto instructions. A report
can combine prose, equations, code, tables and plots. Keep its variable
definitions, inclusion rule and interpretation visible; a successful render
is not evidence of an adequate statistical claim.

\Needspace{8\baselineskip}
\phantomsection

# References {.unnumbered}

::: {#refs}
:::
