---
title: "Reading a Simple Linear Regression"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-cap: true
    tbl-cap: true
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 4, width = 74)
library(ggplot2)
library(carData)
theme_set(theme_minimal(base_size = 11))
prestige <- carData::Prestige
prestige$occupation <- rownames(prestige)
prestige$income_k <- prestige$income / 1000
x <- prestige$education
y <- prestige$income_k
n <- length(y)
Sxx <- sum((x - mean(x))^2)
Sxy <- sum((x - mean(x)) * (y - mean(y)))
model <- lm(income_k ~ education, data = prestige)
TSS <- sum((y - mean(y))^2)
RSS <- sum(resid(model)^2)
se <- summary(model)$sigma
b <- coef(model)
r <- cor(x, y)
hc3 <- sandwich::vcovHC(model, type = "HC3")
se_hc3 <- sqrt(diag(hc3))
crit <- qt(.975, n - 2)
quad <- lm(income_k ~ education + I(education^2), data = prestige)
most <- which.max(cooks.distance(model))
without <- lm(income_k ~ education, data = prestige[-most, ])
stopifnot(n == 102, !anyNA(prestige[c("education", "income_k")]),
          abs(b[2] - Sxy / Sxx) < 1e-10)
```

{{< pagebreak >}}

::: {.callout-tip title="Chapter box"}
**Central question:** How can we read a fitted regression line, explain its
calculation and distinguish its descriptive answer from model-based inference?

**Focal concepts:** Slope, intercept, fitted value and residual; least squares
and covariance; residual SD and R-squared; slope SE and confidence interval;
repeated sampling and precision; mean-response and new-case intervals; form,
spread, influence and support; reading a reported argument before evaluating it.

**Main evidence:** The 102 Canadian occupations in `carData::Prestige`, with
income in thousand 1971 Canadian dollars, plus small declared teaching
examples and a simulation with known coefficients.

**Scope:** Chapters 2–4 supply statistical foundations and Chapters 5–6 supply
elementary R. Chapter 1 supplies the argument vocabulary; Chapter 1a is
optional. The nine sections retain the ordinary regression method and its
analytical explanations. Appendices A–B add extended derivations, matrix
covariance and a projection target. The lab supplies operating practice;
Chapter 8 adds adjustment and the full Project STAR design transfer.
:::

**Target map.** Sections 1--3 stay with the descriptive least-squares slope for
the 102 fixed occupation records. Sections 4--5 explicitly open a separate
hypothetical process question whose estimand is a constant conditional-mean
slope, $\beta_1$. Section 6 distinguishes a fitted mean at a specified education
value from a new occupation's outcome; Section 7 checks form, spread, influence,
and support. Section 8 then reads and evaluates the chapter's reported
finite-record argument, and Section 9 summarizes the route. This map prevents a
model-based interval from being silently attached to the descriptive answer.

# A Question About Occupations and a Line Through Their Records {#sec-ch7-question}

## Begin with the unit and the question

Suppose a report says, "Another year of education increases income by about
900 dollars." Before examining its significance, ask whose education and whose
income were measured. The same number could refer to individuals, households,
or occupational averages. Those are different comparisons.

\Needspace{6\baselineskip}

Our question is deliberately narrower:

> Among the 102 recorded Canadian occupations, how did occupation-average
> income differ across the supported range of occupation-average education in
> 1971, as summarized by a fitted straight line?

One row represents an **occupation**, not a worker. An occupation with higher
average education need not contain the same people as an occupation with lower
average education. A comparison between their averages does not describe what
would happen to one person's income after another year of schooling.
Aggregate and individual relationships can differ substantially [@robinson1950].

The data are distributed as *carData::Prestige* [@carData2022; @fox2019].
The package documentation attributes education and income to the *Census of
Canada* (1971), volume 3, part 6, pages 19-1 to 19-21.
We read the package documentation as a source record, not as evidence that
these occupations form a probability sample. The chapter's calculations and
teaching report are our reanalysis, not statements attributed to an original
research paper.

| Feature | Meaning in this chapter |
|:--|:--|
| Unit and record count | 102 occupations; each occupation receives equal weight |
| Education, $x$ | Average years of education of occupational incumbents in 1971 |
| Income, $y$ | Average income of incumbents, divided by 1,000; thousand 1971 Canadian dollars |
| Missingness for the fit | No missing education or income values; no modeled rows excluded |
| Observed support | Education: 6.38--15.97 years; income: 0.611--25.879 thousand dollars |
| Descriptive estimand | Least-squares slope summarizing these 102 fixed occupation records |
| Limits of the source record | No probability-sampling basis or individual-level analysis established here |

: Record, measurements, and scope for the focal analysis. {#tbl-ch7-record}

Equal weighting matters. A small occupation and a large occupation contribute
one row each. Without suitable counts and a different estimand, this analysis
does not describe an average worker's experience. Historical dollars also
cannot be read as contemporary purchasing power.

## Look before fitting

```{r}
#| label: fig-ch7-scatter
#| fig-cap: "Education and income in the 102 occupation records, before fitting a line."
#| fig-alt: "Occupation-average income generally rises with occupation-average education, with considerable scatter and some especially high-income occupations."
#| fig-width: 6.3
#| fig-height: 3.8
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, size = 2) +
  labs(x = "Average education (years)",
       y = "Average income (thousand 1971 dollars)")
```

The scatterplot gives the line a job: summarize an overall upward pattern
amid substantial variation. It also shows why the line cannot describe every
occupation equally well. General managers, for example, have an average
education of 12.26 years and an average income of 25.879 thousand dollars;
university teachers have corresponding values of 15.97 and 12.480.
More education does not order every pair of occupations by income.

Here is the short R preparation used throughout. Package installation and
project setup are developed in the computing chapters. The data come from the
named `carData` package, so this analysis requires no separate classroom file.

\Needspace{16\baselineskip}

```{r}
#| label: ch7-prepare-visible
#| echo: true
#| eval: false
prestige <- carData::Prestige
prestige$income_k <- prestige$income / 1000
library(ggplot2)
ggplot(prestige, aes(education, income_k)) + geom_point() +
  labs(x = "Mean education (years)", y = "Mean income (thousand dollars)")
model <- lm(income_k ~ education, data = prestige)
x <- prestige$education
y <- prestige$income_k
b <- coef(model)
coef(model)
```

The formula puts the outcome to the left of the tilde and the predictor to its
right. The command finds coefficients. It does not choose a population,
establish a causal design, or justify a confidence interval.

# Read the Fitted Line and Understand How OLS Finds It {#sec-ch7-line}

## Fitted values, slopes, and residuals

A straight-line summary assigns each observed predictor value $x_i$ a
**fitted value** $\widehat y_i$:

$$
\widehat y_i=\widehat\beta_0+\widehat\beta_1x_i,
\qquad e_i=y_i-\widehat y_i.
$$

The subscript $i$ indexes a record. The hat marks a fitted quantity.
The **residual** $e_i$ is observed income minus fitted income. A positive
residual places an occupation above the line; a negative residual places it
below. A residual is calculated after fitting and is not the same as an
unobservable error from a population model.

For the occupation records, the line is

$$
\widehat{\mathrm{income}}=-2.8536+0.8988\,\mathrm{education}.
$$

Income is still in thousands of dollars. The slope says that the fitted
income differs by 0.8988 thousand dollars, about 899 dollars, for a one-year
difference in occupation-average education. A two-year difference along this
line corresponds to $2(0.8988)=1.7976$ thousand dollars. These are comparisons
on the fitted line, not predicted gains from changing a worker's education.

For university teachers,

$$
\widehat y=-2.8536+0.8988(15.97)=11.5002,
\qquad e=12.4800-11.5002=0.9798.
$$

Their recorded average income is about 980 dollars above the fitted value.
Using the unrounded coefficients gives essentially the same rounded answer.
For an occupation averaging 12 years of education, the fitted value is
$-2.8536+0.8988(12)=7.9320$ thousand dollars.

The intercept places the line vertically: it is its fitted income at
education zero. Zero is outside the observed range, so the negative intercept
is not evidence that any recorded occupation has negative average income.
Centering education at 12 would change the intercept to 7.9322 while leaving
the slope and all fitted values unchanged.

```{r}
#| label: fig-ch7-line-residual
#| fig-cap: "The least-squares line passes through the mean point; the marked vertical segment is the university-teacher residual."
#| fig-alt: "A fitted line crosses the mean education and income. At 15.97 years, a vertical segment joins the fitted value to the higher university-teacher observation."
#| fig-width: 6.3
#| fig-height: 3.8
teacher <- prestige["university.teachers", ]
teacher$fit <- predict(model, teacher)
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, colour = "grey45") +
  geom_abline(intercept = b[1], slope = b[2], linewidth = .8) +
  annotate("point", x = mean(x), y = mean(y), shape = 18, size = 3) +
  geom_segment(data = teacher, aes(xend = education, yend = fit),
               linewidth = 1) +
  labs(x = "Average education (years)",
       y = "Average income (thousand 1971 dollars)")
```

## Why this particular line?

Ordinary least squares (**OLS**) chooses the intercept and slope that minimize
the **residual sum of squares**:

$$
RSS=\sum_{i=1}^{n}(y_i-\widehat\beta_0-\widehat\beta_1x_i)^2.
$$

The summation sign says to add the squared residual from every record.
Squaring prevents positive and negative deviations from cancelling and gives
large deviations more influence on the criterion. "Best" here means best by
this criterion among straight lines with an intercept, not best for every
research purpose.

Two centered sums lead directly to the solution:

$$
S_{xx}=\sum_i(x_i-\bar x)^2,\qquad
S_{xy}=\sum_i(x_i-\bar x)(y_i-\bar y),
$$

$$
\widehat\beta_1=\frac{S_{xy}}{S_{xx}}
=\frac{s_{xy}}{s_x^2}
=r\frac{s_y}{s_x},
\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x.
$$

Here $\bar x,\bar y$ are means; $s_x,s_y$ are sample standard deviations;
$s_{xy}$ is sample covariance; and $r$ is Pearson's correlation.
The variance and covariance both divide their centered sums by $n-1$, so
that divisor cancels in their ratio. The slope has income-per-education units;
correlation has no units.

For these data,

$$
\widehat\beta_1=\frac{675.8041}{751.8852}=0.8988,
\qquad
\widehat\beta_0=6.7979-(0.8988)(10.7380)\approx-2.8534.
$$

The final small rounding difference is avoided by using unrounded summaries.
Positive covariance produces a positive slope, provided education varies at all.
If every occupation had the same education, $S_{xx}=0$ and a slope could
not be identified.

The fitted line passes through $(\bar x,\bar y)$, because substituting
$\bar x$ into the intercept formula gives $\bar y$.
With an intercept, the residuals also sum to zero and their products with
$x$ sum to zero. Otherwise the line could be shifted or tilted to reduce
RSS. These are algebraic properties of OLS, not tests of independence or
linearity. Full normal-equation and projection-matrix derivations are retained
in **Appendix A**.

## A tiny calculation you can inspect completely

Consider the teaching records $(x,y)=(1,2),(2,4),(3,3)$. Their means are
$\bar x=2,\bar y=3$, so

$$
S_{xx}=(-1)^2+0^2+1^2=2,\qquad
S_{xy}=(-1)(-1)+(0)(1)+(1)(0)=1.
$$

Thus the slope is $1/2=0.5$, the intercept is $3-0.5(2)=2$, and the fitted
values are 2.5, 3, and 3.5. The residuals are $-0.5,1,-0.5$, which sum to
zero, and $RSS=0.25+1+0.25=1.5$. No normality assumption is needed for
these calculations. A stochastic model becomes necessary when we ask about
repeated-sample uncertainty.

## Compare candidate lines with the OLS minimum

The least-squares criterion also lets us compare the fitted line with
alternatives. To keep the comparison transparent, make every candidate line
pass through $(\bar x,\bar y)$ and vary only its slope. For such a line,

$$
RSS(b_1)=RSS(\widehat\beta_1)
       +S_{xx}(b_1-\widehat\beta_1)^2.
$$

The added term is zero only at the OLS slope and positive for every other
slope when $S_{xx}>0$. The following table checks five deliberately chosen
lines on the occupation records. These are alternative summaries of the same
records, not repeated samples.

```{r}
#| label: ch7-candidate-lines-code
#| echo: true
candidate_slopes <- unname(b[2]) + c(-0.6, -0.3, 0, 0.3, 0.6)
candidate_intercepts <- mean(y) - candidate_slopes * mean(x)
ols_rss <- sum(resid(model)^2)
predictor_spread <- sum((x - mean(x))^2)
candidate_rss <- ols_rss +
  predictor_spread * (candidate_slopes - b[2])^2
candidate_lines <- data.frame(
  Intercept = candidate_intercepts,
  Slope = candidate_slopes,
  RSS = candidate_rss
)
```

Each operation follows the displayed identity. R applies the same arithmetic
to all five candidate slopes, so no function or loop is needed here.

```{r}
#| label: tbl-ch7-candidate-lines
#| tbl-cap: "Candidate lines through the mean point and their residual sums of squares."
#| echo: false
knitr::kable(candidate_lines, digits = 3)
```

```{r}
#| label: fig-ch7-candidate-lines
#| echo: false
#| fig-cap: "Five candidate lines through the sample mean. The OLS line has the smallest residual sum of squares."
#| fig-alt: "Five lines with different slopes cross at the mean occupation point. The highlighted OLS line runs centrally through the data; steeper and flatter alternatives have larger residual sums of squares."
#| fig-width: 6.3
#| fig-height: 3.8
candidate_grid <- expand.grid(
  education = seq(min(x), max(x), length.out = 100),
  Candidate = seq_along(candidate_slopes)
)
candidate_grid$income <- candidate_intercepts[candidate_grid$Candidate] +
  candidate_slopes[candidate_grid$Candidate] * candidate_grid$education
candidate_grid$Line <- ifelse(
  candidate_grid$Candidate == 3, "OLS", "Alternative"
)
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, colour = "grey55", size = 1.4) +
  geom_line(data = candidate_grid,
            aes(y = income, group = Candidate, colour = Line,
                linewidth = Line)) +
  scale_colour_manual(values = c(Alternative = "grey65", OLS = "black")) +
  scale_linewidth_manual(values = c(Alternative = .45, OLS = 1)) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)", colour = NULL, linewidth = NULL)
```

The comparison also explains why high-income occupations can exert substantial
influence: squaring gives large vertical discrepancies extra weight. That is
a feature of the criterion, not evidence that those occupations are invalid.

## Centering changes the reference point, not the line

The fitted equation can be written

$$
\widehat y_i=\bar y+\widehat\beta_1(x_i-\bar x).
$$

Centering education at its observed mean makes the intercept equal to
$\bar y=6.7979$ thousand dollars, the fitted income at average education.
It leaves the slope, fitted values, residuals, RSS, and $R^2$ unchanged.
Centering at 12 instead makes the intercept the already computed fitted value
7.9322. Thus centering can make the intercept interpretable without changing
the comparison represented by the line.

```{r}
#| label: ch7-centering-visible
#| echo: true
#| eval: false
prestige$education_c <- prestige$education - mean(prestige$education)
model_centered <- lm(income_k ~ education_c, data = prestige)
coef(model_centered)
max(abs(fitted(model_centered) - fitted(model)))
```

Changing income from thousands of dollars to dollars would multiply the
intercept, slope, residual SD, and coefficient SEs by 1,000, while leaving
$t$ statistics, $p$ values, correlation, and $R^2$ unchanged. A unit change
is not new evidence or a stronger relationship.

# Describe Fit Before Making Model-Based Inferences {#sec-ch7-fit}

## How much variation remains?

A fitted line can show a clear linear trend and still leave large differences
between occupations unexplained by its fitted values.

For OLS with an intercept,

$$
TSS=\sum_i(y_i-\bar y)^2=RegSS+RSS,
\qquad RegSS=\sum_i(\widehat y_i-\bar y)^2.
$$

For the occupation records, $1820.8134=607.4214+1213.3920$.
This is a decomposition of squared deviations, in squared thousand-dollar
units. It holds because the cross-product of fitted deviations and residuals
is zero.

The coefficient of determination is

$$
R^2=1-\frac{RSS}{TSS}
=1-\frac{1213.3920}{1820.8134}=0.3336.
$$

The fitted line accounts for about 33.4% of the variation in recorded income
around its mean, according to this squared-deviation measure.
"Accounts for" or "explains" in this algebraic sense does not mean causes.
For simple regression with an intercept, $R^2=r^2$:
$0.5776^2\approx0.3336$. The identity need not hold for other models or a
no-intercept fit.

The residual standard deviation is

$$
s_e=\sqrt{\frac{RSS}{n-2}}
=\sqrt{\frac{1213.3920}{100}}=3.4834.
$$

Its units are thousand dollars, the same as income. Fitting an intercept and
slope imposes two independent residual constraints, leaving $n-2$ residual
degrees of freedom. Under the constant-variance model introduced next,
$s_e$ estimates the error standard deviation. Without that model it remains
a scaled measure of residual scatter; it is not the standard error of the
slope or a guaranteed typical prediction error in new records.

| Quantity | Occupation result | What it describes |
|:--|--:|:--|
| Slope | 0.8988 | Change in fitted thousand-dollar income per education year |
| Residual SD | 3.4834 | Remaining income scatter on the outcome scale |
| $R^2$ | 0.3336 | Proportion of centered squared variation accounted for in these records |
| Classical slope SE | 0.1270 | Model-based repetition variability of the slope, under the next assumptions |

: Different summaries have different units and jobs. {#tbl-ch7-quantities}

# Make and Read a Model-Based Inference {#sec-ch7-inference}

## Announce the inferential extension before its interval

The finite-record slope is not waiting to be discovered in a larger population:
we have computed it for all 102 records named in the focal question.
A confidence interval answers an additional question.

For a **separable teaching extension**, suppose occupations' outcomes arose
independently from a process with conditional mean

$$
E(Y_i\mid X_i=x_i)=\beta_0+\beta_1x_i.
$$

The estimand $\beta_1$ is now a constant conditional-mean slope in that
hypothetical process, not the fixed slope of these 102 records. Repetition
means generating new outcomes from that process at the stated predictor values
and refitting. It does not mean that the documented occupations were randomly
sampled. Changing the estimand changes the question--answer pair.

Write the process as

$$
Y_i=\beta_0+\beta_1x_i+\varepsilon_i,\qquad
E(\varepsilon_i\mid X)=0,\qquad
\operatorname{Var}(\varepsilon_i\mid X)=\sigma^2.
$$

For the exact classical $t$ procedure, errors are independent and conditionally
normal with this common variance, and $S_{xx}>0$. The $x_i$ values need not
be normal; neither must the marginal distribution of $Y$. It is the
conditional error distribution, not the histogram of every variable, that
appears in the model. Independence is supported by how records arise, not by a
good-looking residual plot.

To see where slope uncertainty comes from, substitute the process equation
into the OLS slope formula. Here $X$ denotes the predictor values on which
we condition. Since $\sum_i(x_i-\bar x)=0$,

$$
\widehat\beta_1-\beta_1
=\frac{\sum_i(x_i-\bar x)\varepsilon_i}{S_{xx}}.
$$

The fitted slope differs from the process slope because the realized errors
need not cancel when weighted by distance from mean education. Independence
makes their covariance terms zero; common variance gives

$$
\operatorname{Var}(\widehat\beta_1\mid X)
=\frac{\sigma^2\sum_i(x_i-\bar x)^2}{S_{xx}^2}
=\frac{\sigma^2}{S_{xx}}.
$$

This variance identity does not require normal errors. Normality enters the
exact $t$ reference. Taking the square root and estimating $\sigma$ by $s_e$
gives the slope standard error:

$$
SE(\widehat\beta_1)=\frac{s_e}{\sqrt{S_{xx}}}
=\frac{3.4834}{\sqrt{751.8852}}=0.1270,
$$

$$
\widehat\beta_1\ \pm\ t_{.975,n-2}SE(\widehat\beta_1)
=0.8988\pm1.984(0.1270)
=[0.6468,\ 1.1508].
$$

The critical value 1.984 is the 97.5th percentile of a $t$ distribution with
100 degrees of freedom. More error variation raises the SE; greater predictor
spread lowers it when the other quantities are held fixed. More observations
usually improve precision through their contribution to $S_{xx}$, but
dependent observations do not supply independent information merely by
increasing the row count.

The null test $H_0:\beta_1=0$ uses
$t=0.8988/0.1270=7.075$ and a two-sided $p<.001$.
This is evidence against that zero-slope model under the stated procedure.
It is not the probability that the null is true, the probability that the
data occurred by chance, evidence of causation, or a measure of practical
importance. The interval describes the range produced by this uncertainty
method, not a 95% probability assigned to the fixed parameter after observing
this interval. The next section makes the repeated-sample interpretation
visible before prediction is introduced.

The assumptions are a proposed justification, not facts created by the formula.
Curvature, changing spread, or dependence can invalidate the exact coverage
claim. The diagnostic section teaches the checks before the reported argument
is evaluated.

# See Sampling Uncertainty Through Repetition {#sec-ch7-repetition}

## A known-truth line and one fitted sample

Generate 20 outcomes at predictor values equally spaced from 6 to 16:

$$
Y_i=2+0.5x_i+\varepsilon_i,\qquad
\varepsilon_i\overset{\mathrm{ind}}\sim\mathcal N(0,1).
$$

These units and values are a teaching construction, not occupation records.
The true intercept is 2, true slope is 0.5, and true error SD is 1.

```{r}
#| label: ch7-known-sample-code
#| echo: true
#| results: hide
set.seed(123)
known_data <- data.frame(x = seq(6, 16, length.out = 20))
known_data$y <- 2 + 0.5 * known_data$x + rnorm(20, sd = 1)
known_model <- lm(y ~ x, data = known_data)
coef(known_model)
sigma(known_model)
```

`coef()` returns the fitted intercept and slope; `sigma()` returns the
residual SD. The table adds the coefficient SEs from `summary(known_model)`.

\Needspace{10\baselineskip}

```{r}
#| label: tbl-ch7-known-sample
#| tbl-cap: "One fitted sample from a process with known coefficients and error SD."
#| echo: false
known_summary <- coef(summary(known_model))
knitr::kable(data.frame(
  Quantity = c("Intercept", "Slope", "Residual SD"),
  Estimate = sprintf("%.3f", c(coef(known_model), sigma(known_model))),
  `Model SE` = c(sprintf("%.3f", known_summary[, "Std. Error"]), ""),
  check.names = FALSE
), align = c("l", "r", "r"))
```

The last cell is blank because residual SD is a scale estimate, not a
coefficient with a coefficient standard error.

```{r}
#| label: fig-ch7-core-true-fitted
#| echo: false
#| fig-cap: "Known and fitted lines for one simulated dataset. The true process is programmed; the fitted line responds to the realized errors."
#| fig-alt: "Twenty simulated points surround the known line with intercept two and slope one half. A solid fitted line and dashed true line differ slightly."
ggplot2::ggplot(known_data, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(size = 1.6) +
  ggplot2::geom_abline(
    intercept = 2, slope = 0.5, linetype = "dashed", linewidth = 0.8
  ) +
  ggplot2::geom_abline(
    intercept = coef(known_model)[1],
    slope = coef(known_model)[2], linewidth = 0.8
  ) +
  ggplot2::labs(x = "Simulated predictor", y = "Simulated outcome")
```

The fitted slope and intercept need not equal their programmed values in
one sample. The unobserved error is the deviation from the true mean;
the residual is the observed deviation from the fitted line. Even when
the errors are independent, fitted residuals are linked by their zero-sum
and zero-covariance constraints. A residual plot is an imperfect diagnostic
window on error behavior, not a direct sample of observed true errors.

## Repeated samples and the sources of slope precision

The next experiment holds 40 predictor values fixed and regenerates only
the errors, 2,000 times. It records slopes, estimated standard errors,
and whether each classical interval covers the known slope 0.5.

\Needspace{30\baselineskip}

```{r}
#| label: ch7-repeated-slopes-code
#| echo: true
set.seed(321)
rep_x <- seq(6, 16, length.out = 40)
repetitions <- data.frame(
  slope = numeric(2000), se = numeric(2000), covered = logical(2000)
)
for (j in seq_len(nrow(repetitions))) {
  rep_y <- 2 + 0.5 * rep_x + rnorm(40, sd = 1)
  rep_fit <- lm(rep_y ~ rep_x)
  repetitions$slope[j] <- coef(rep_fit)[2]
  repetitions$se[j] <- coef(summary(rep_fit))[2, "Std. Error"]
  rep_ci <- confint(rep_fit)[2, ]
  repetitions$covered[j] <- rep_ci[1] <= 0.5 &&
    rep_ci[2] >= 0.5
}
repetition_summary <- c(
  mean_slope = mean(repetitions$slope),
  empirical_SD = sd(repetitions$slope),
  mean_estimated_SE = mean(repetitions$se),
  theoretical_SD = 1 / sqrt(sum((rep_x - mean(rep_x))^2)),
  empirical_coverage = mean(repetitions$covered)
)
```

\Needspace{9\baselineskip}

```{r}
#| label: tbl-ch7-repeated-slopes
#| tbl-cap: "Slope behavior across 2,000 repetitions of the known process."
#| echo: false
knitr::kable(data.frame(
  Quantity = c("Mean estimated slope", "SD of estimated slopes",
               "Mean estimated SE", "Theoretical slope SD",
               "Fraction of intervals covering 0.5"),
  Value = unname(repetition_summary)
), digits = 4)
```

```{r}
#| label: fig-ch7-core-repetition
#| echo: false
#| fig-cap: "Slopes in 2,000 repetitions of the known process. The vertical reference is the true slope, 0.5."
#| fig-alt: "A roughly symmetric histogram of repeated fitted slopes is centered near one half; its spread represents repeated-sample slope variation."
#| fig-pos: H
ggplot2::ggplot(repetitions, ggplot2::aes(x = slope)) +
  ggplot2::geom_histogram(bins = 35, color = "white", fill = "grey50") +
  ggplot2::geom_vline(xintercept = 0.5, linewidth = 0.8, linetype = "dashed") +
  ggplot2::labs(x = "Estimated slope", y = "Number of repetitions")
```

The empirical SD of the slopes estimates their repeated-sample spread. Each
fitted SE tries to estimate that spread using just one sample. Coverage is a
property of the repeated interval-producing procedure, not a probability
assigned to a fixed parameter after one interval is realized. The empirical
coverage will not be exactly .95: at true .95 coverage, its Monte Carlo SE
with 2,000 independent repetitions is
$\sqrt{.95(.05)/2000}\approx.0049$.

The theoretical slope SD is $\sigma/\sqrt{S_{xx}}$. More noise increases
it; greater predictor spread decreases it. More observations typically
decrease it when the predictor distribution is held comparable. To isolate
each feature, change one at a time:

\Needspace{18\baselineskip}

```{r}
#| label: ch7-precision-conditions-code
#| echo: true
#| results: hide
baseline_x <- seq(6, 16, length.out = 40)
more_x <- seq(6, 16, length.out = 160)
narrow_x <- seq(8.5, 13.5, length.out = 40)
baseline_sd <- 1 / sqrt(sum((baseline_x - mean(baseline_x))^2))
more_rows_sd <- 1 / sqrt(sum((more_x - mean(more_x))^2))
narrow_sd <- 1 / sqrt(sum((narrow_x - mean(narrow_x))^2))
more_noise_sd <- 2 / sqrt(sum((baseline_x - mean(baseline_x))^2))
c(baseline = baseline_sd, more_rows = more_rows_sd,
  narrower_support = narrow_sd, more_noise = more_noise_sd)
```

The first three lines define the predictor grids. Each subsequent line
substitutes a noise SD and a grid's spread into the same formula.

```{r}
#| echo: false
precision_conditions <- data.frame(
  Condition = c("Baseline", "More rows", "Narrower support", "More noise"),
  n = c(40, 160, 40, 40), Lower = c(6, 6, 8.5, 6),
  Upper = c(16, 16, 13.5, 16),
  `Error SD` = c(1, 1, 1, 2),
  `Slope SD` = c(baseline_sd, more_rows_sd, narrow_sd, more_noise_sd),
  check.names = FALSE
)
```

\Needspace{9\baselineskip}

```{r}
#| label: tbl-ch7-precision-conditions
#| tbl-cap: "How sample size, predictor support, and error SD affect theoretical slope precision."
#| echo: false
knitr::kable(precision_conditions, digits = 4)
```

Halving the predictor range at the same sample size doubles the slope SD;
doubling $\sigma$ also doubles it. Quadrupling the number of equally spaced
points over the same range approximately halves it; the small difference
from an exact half comes from the finite grids' variances. Using `x <- 1:n`
while increasing $n$ would change both sample size and predictor range.
More rows do not automatically remove confounding, measurement error,
dependence, or misspecification. These simulations establish behavior only
under the mechanisms programmed into them.

\Needspace{9\baselineskip}

# Distinguish a Mean Response from a New-Case Prediction {#sec-ch7-prediction}

Slope uncertainty concerns one process coefficient. Prediction introduces a
different distinction: estimating a conditional mean versus anticipating one
new outcome from the same declared process.

## A fitted mean is not a new occupation

At $x_0=12$, the fitted income is 7.9322 thousand dollars. Under the model,
this estimates the conditional mean for occupations with that average
education. The standard error for the fitted mean is

$$
SE(\widehat y_0)
=s_e\sqrt{\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}}.
$$

Substitution gives

$$
3.4834\sqrt{\frac1{102}+
 \frac{(12-10.7380)^2}{751.8852}}=0.3803.
$$

The resulting 95% mean interval is
$7.9322\pm1.984(0.3803)=[7.1776,8.6868]$.

A new occupation's outcome also varies around its process mean.
Write $\mu_0=\beta_0+\beta_1x_0$, $\widehat\mu_0=\widehat y_0$, and
$Y_0=\mu_0+\varepsilon_0$. The prediction error has two parts:

$$
Y_0-\widehat\mu_0
=\varepsilon_0-(\widehat\mu_0-\mu_0).
$$

If this new error is independent of the fitted data and comes from the same
common-variance process, the two variance contributions add:

$$
\operatorname{Var}(Y_0-\widehat\mu_0\mid X,x_0)
=\sigma^2+\operatorname{Var}(\widehat\mu_0\mid X,x_0).
$$

The first term is new outcome variation; the second is uncertainty in the
estimated mean. Substituting the fitted-mean variance and estimating $\sigma$
by $s_e$ gives

$$
SE_{\mathrm{new}}
=s_e\sqrt{1+\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}}
=3.5041.
$$

Thus the 95% prediction interval is approximately
$7.9322\pm1.984(3.5041)=[0.9802,14.8842]$.
The extra 1 is the new outcome's own error variance, expressed as a multiple
of $\sigma^2$; it does not disappear as the estimation sample grows.

```{r}
#| label: fig-ch7-intervals
#| fig-cap: "Mean-response intervals and new-case prediction intervals under the same working model."
#| fig-alt: "The mean-confidence band is narrowest near average education. The new-case prediction band is much wider throughout the observed education range."
#| fig-width: 6.3
#| fig-height: 3.8
grid <- data.frame(education = seq(min(x), max(x), length.out = 150))
cm <- predict(model, grid, interval = "confidence")
pi <- predict(model, grid, interval = "prediction")
grid$fit <- cm[, 1]
grid$cl <- cm[, 2]; grid$cu <- cm[, 3]
grid$pl <- pi[, 2]; grid$pu <- pi[, 3]
ggplot(grid, aes(education, fit)) +
  geom_ribbon(aes(ymin = pl, ymax = pu), fill = "grey90") +
  geom_ribbon(aes(ymin = cl, ymax = cu), fill = "grey65") +
  geom_line(linewidth = .7) +
  geom_point(data = prestige, aes(y = income_k), shape = 1, size = 1.3) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)")
```

Both intervals get wider away from $\bar x$, even within the observed range.
Outside that range, numerical output is not a warrant for extrapolation.
A lower prediction bound below zero also exposes the limitations of an
unrestricted normal-error model for a nonnegative outcome; it is not a
plausible negative-income forecast.

```{r}
#| label: ch7-interval-code
#| echo: true
#| eval: false
confint(model)
predict(model, data.frame(education = 12), interval = "confidence")
predict(model, data.frame(education = 12), interval = "prediction")
```

A separately owed new-case prediction answer is a separable question, not
automatically another strand of the original descriptive question.
Moreover, a model prediction interval is not a test of a learning algorithm
on unseen records. Held-out prediction performance is a distinct estimand,
developed with multiple regression. Appendix A retains the interval
derivations and further supported-value comparisons.


# Check Form, Spread, Influence, and Support {#sec-ch7-check}

## Checks should change an interpretation or a decision

The fitted line does not have to satisfy a stochastic model to exist.
But claims that it describes a linear conditional mean, that its classical
interval has exact coverage, or that it predicts new occupations reliably
need additional support.

```{r}
#| label: fig-ch7-checks
#| fig-cap: "Residual pattern and influence invite different follow-up questions."
#| fig-alt: "The left panel shows curvature and uneven scatter in residuals. The right panel shows Cook's distances, with physicians most influential for this fit."
#| fig-width: 6.4
#| fig-height: 3.5
op <- par(mfrow = c(1, 2), mar = c(4, 4, 1, 1), mgp = c(2.5, .7, 0))
plot(fitted(model), resid(model), xlab = "Fitted income", ylab = "Residual",
     pch = 1, cex = .7)
abline(h = 0, col = "grey60")
lines(lowess(fitted(model), resid(model)), lwd = 2)
plot(cooks.distance(model), type = "h", xlab = "Occupation row",
     ylab = "Cook's distance")
text(most, cooks.distance(model)[most], "physicians", pos = 4, cex = .7)
par(op)
```

A curve in residuals suggests that one straight conditional mean leaves
systematic variation. A changing residual spread challenges a common-variance
uncertainty calculation. Leverage concerns unusual predictor positions;
influence concerns how much the fitted model changes when a record's
contribution changes. A large residual, high leverage, and high influence are
related but different properties.

For a straight line with an intercept, leverage is

$$
h_i=\frac1n+\frac{(x_i-\bar x)^2}{S_{xx}}.
$$

It grows as a predictor value moves away from $\bar x$. Cook's distance then
combines leverage and residual magnitude:

$$
D_i=\frac{e_i^2}{2s_e^2}\frac{h_i}{(1-h_i)^2}.
$$

Cook's distance summarizes how much the fitted values change when case $i$
is omitted, relative to residual variation and the two fitted coefficients.
It is a prompt to inspect provenance and sensitivity, not an automatic
deletion cutoff.

For these data, adding education squared lets the fitted relationship bend.
The following same-record comparison makes the change in shape visible. It
is an exploratory check, not a retrospectively prespecified new final model.

```{r}
#| label: ch7-check-models
#| echo: true
quad <- lm(
  income_k ~ education + I(education^2), data = prestige
)
most <- which.max(cooks.distance(model))
without <- lm(income_k ~ education, data = prestige[-most, ])
```

```{r}
#| label: fig-ch7-curvature
#| fig-cap: "Straight and quadratic fits to the same 102 records. The curve is an exploratory sensitivity, not a selected final model."
#| fig-alt: "The quadratic fit bends upward toward higher education, unlike the constant-slope straight line. Both fits use all 102 occupation records on the same income scale."
#| fig-width: 6.3
#| fig-height: 3.7
check_grid <- data.frame(
  education = seq(min(x), max(x), length.out = 200)
)
check_curves <- rbind(
  data.frame(check_grid, Fit = "Straight line",
             Income = predict(model, check_grid)),
  data.frame(check_grid, Fit = "Quadratic sensitivity",
             Income = predict(quad, check_grid))
)
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, colour = "grey50", size = 1.5) +
  geom_line(data = check_curves,
            aes(y = Income, linetype = Fit), linewidth = .8) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)", linetype = NULL)
```

The quadratic coefficient has $p=.0059$ under that model's classical
reference calculation. This is a reason to examine the visible curvature,
not a sufficient model-selection rule: unequal spread and choosing the
comparison after inspecting the records also qualify that number.

Removing physicians, the record with the largest Cook's
distance (0.398), changes the slope from 0.8988 to 0.7982.
Nothing in that result proves physicians were recorded incorrectly. The
deletion is a sensitivity calculation; it is not our replacement analytic
sample or a licence to hide a valid unusual occupation.

\Needspace{7\baselineskip}

An HC3 heteroskedasticity-consistent covariance calculation gives a slope SE
of 0.1491 rather than 0.1270. Using the same $t_{100}$ critical value as an
approximate sensitivity procedure gives $[0.6031,1.1945]$.
The line does not change. This procedure relaxes a variance assumption under
suitable independent-observation conditions; it does not repair curvature,
dependence, confounding, or measurement error [@white1980; @mackinnon1985].
Its finite-sample interval is not the exact classical normal-model interval.
The short R calculation makes the covariance choice explicit:

```{r}
#| label: ch7-hc3-visible
#| echo: true
V_hc3 <- sandwich::vcovHC(model, type = "HC3")
c(classical = coef(summary(model))["education", "Std. Error"],
  HC3 = sqrt(V_hc3["education", "education"]))
```

**Appendix B** retains the matrix sandwich construction and the optional
best-linear-projection target under curvature.

\Needspace{20\baselineskip}

| Check | What needs to be examined | Consequence |
|:--|:--|:--|
| Form | Residual curve and a curved-fit comparison | A straight line may remain a description but fail as a constant conditional-mean slope account |
| Spread | Residual variation over fitted values | Classical SEs may need revision; HC3 is a variance sensitivity, not a universal repair |
| Influence | Source record and result with/without a consequential case | Disclose sensitivity; correct errors only when verified |
| Independence | How occupations and measurements were generated or related | A plot cannot establish independent process realizations |
| Support | Actual education range and density of observations | Restrict comparisons; avoid extrapolation and individual-level claims |

: Diagnostic questions and their consequences. {#tbl-ch7-diagnostic-consequences}

# Read the Reported Argument, Then Evaluate Its Quality {#sec-ch7-reading}

## Read the reported question and answer

To apply Chapter 1, we need a report, not merely an output table.
The following **chapter-authored teaching report** is the accessible record.
Its numbered paragraphs provide trace locations. They are not quotations from
the original data source.

::: {.callout-note title="Teaching report: an occupation-level description"}
**R1 -- Question and evidence.** We described the education--income relationship
in the 102 *carData::Prestige* occupation records. Income was expressed in
thousand 1971 Canadian dollars. All records had both modeled measurements.
Each occupation was weighted equally; the observed education range was
6.38--15.97 years.

**R2 -- Offered reasons.** The upward scatterplot motivated a straight-line
summary. OLS was chosen because it minimizes squared vertical discrepancies.
The aim was a summary of these records, not a constant causal effect or an
individual-level comparison. The source documentation identifies the
measurements as historical occupational averages. No probability-sampling
justification was available in the record consulted.

**R3 -- Analysis and interpretation.** The fitted slope was 0.8988 thousand
dollars per education year, with $R^2=.3336$ and residual SD 3.4834 thousand
dollars. The residual pattern and a curved-fit sensitivity qualified the line
as an incomplete description. The fit and sensitivities were exploratory.
The descriptive result itself was not assigned sampling uncertainty.

**R4 -- Reported answer.** Across these 102 recorded occupations, higher average
education accompanied higher average income overall. The least-squares line
summarized that pattern by about 899 additional 1971 dollars of fitted
occupation-average income per education year over the observed range. It did
not describe every occupation closely or establish an individual or causal
return to education.
:::

The report deliberately separates its descriptive conclusion from the
hypothetical inferential route in Sections 4--6. A reader should not attach the
model-based interval to R4 as though R4 named a population slope.

## Locate the component content and its traces

Apply Chapter 1's functions to the report's actual content. A documentary
trace identifies where that content can be inspected; it does not establish
that the content is adequate.

| Function | Content read in this report | Trace and documentary status |
|:--|:--|:--|
| STAT-E | Records, measurements, scaling, and complete sample | R1; stated in the paper |
| STAT-J | Reasons for OLS as a descriptive, noncausal summary | R2; stated in the paper |
| STAT-I | Slope, fit, sensitivity, and exploratory status | R3; stated in the paper |
| STAT-C | Positive but incomplete description over recorded support | R4; stated in the paper |
| SUB-E | Historical occupation averages, not individual measurements | R1 and data note; stated in the paper and located in related materials |
| SUB-J | Measurement meanings supporting an occupation-level reading | R2 and data note; stated in the paper and located in related materials |
| SUB-I | Meaning of about 899 dollars per education year | R3--R4; stated in the paper |
| SUB-C | Bounded historical answer, without a causal or individual return | R4; stated in the paper |

: Reading one argument with linked statistical and substantive layers. {#tbl-ch7-traces}

Here "the paper" means the supplied teaching report. Chapter 1's statuses
remain **stated in the paper**, **located in related materials**, **implicit
but traceable**, and **not located**; flag conflicts separately. Not located
does not mean absent everywhere. One passage may serve several functions.
The focal slope uses one statistical strand; fit summaries and checks support
that strand rather than automatically becoming separately owed answers.

For a faulty R2 variant, consider "We used R, so our slope is an unbiased causal
effect." Its offered reason remains located STAT-J content, not "not located."
Reading records that reason and the reported answer; evaluation judges whether
the reason supports the claim. Do not silently replace the report's answer
with a better one.

\Needspace{8\baselineskip}

## Evaluate the argument already read

| Quality-evaluation question | Application to the reported descriptive argument |
|:--|:--|
| Question alignment | Does the answer remain about the 102 occupations and the declared straight-line summary? |
| Evidence integrity | Are the source, aggregate unit, scaling, exclusions, and unusual records traceable? |
| Justification adequacy | Are the offered reasons adequate for a descriptive line, rather than stretched into random sampling or causal identification? |
| Inferential adequacy | Do the calculation and substantive reasoning support the reported pattern while retaining curvature and scatter? |
| Transition fidelity | Does the interpretation preserve occupation-level measurements, units, and exploratory status? |
| Claim calibration | Does the answer retain historical scope, supported range, and the limits of a linear summary? |

: Evaluating the argument already read. {#tbl-ch7-quality}

These are six questions, not six components or a total score.
A local judgment can be adequate for this purpose, needs revision,
insufficient information to judge, or not applicable. The judgment belongs
to evaluation, not to the documentary-status column of the reading table.

For the deliberately limited R4 answer, the overall conclusion is
**supported as stated**: it reports a checkable finite-record summary and
acknowledges what the line does not represent. Missing details about national
sampling do not invalidate arithmetic on the named records, but would matter
for a national generalization.

The additional model-based extension is less secure. Its exact classical
uncertainty justification is not established merely by calculating an
interval, and the diagnostics challenge its linear-mean and common-variance
conditions. Retaining the descriptive answer does not rescue that distinct
process claim. A stochastic best-linear-projection slope under curvature
would be another explicitly defined estimand with suitable random-$X$
conditions and uncertainty, not an unannounced relabelling of $\beta_1$.

The opening claim about an individual's return to education is **not
supported** by this record. Replacing it with R4 changes the unit and mode of
the question; that is a revised question--answer pair, not cosmetic narrowing
of the same claim. A causal comparison needs different support, not a different software command.

In Project STAR [@finn1990; @mosteller1995], a binary-regression slope equals a
difference in observed
class-condition means. That arithmetic alone supplies no causal warrant.
The within-school assignment design matters, together with the link from
recorded class type to original assignment, missing outcomes, school weights,
and dependence among pupils. This teaching reanalysis does not independently
verify the assignment link. Schools are assignment blocks and covariance
clusters, not the units assigned to class conditions. The full
[Chapter 8, Section 8: Project STAR](`r if (knitr::is_latex_output()) "chapter_08.pdf#sec-ch8-star" else "chapter_08.qmd#sec-ch8-star"`) works through
these distinctions and separates the observed-reading sample from the smaller
covariate-complete sample. It is a different question, not a prerequisite for
the bounded occupation answer.

## Auxiliary construction callback

For this regression, **Understand and plan** would declare the occupation unit,
historical scope, finite-record slope, variables, and proposed checks before
fitting. **Produce and complete** would preserve the source record, estimate
and check the line, then report the bounded occupation-level answer. This is a
construction callback, not a reading sequence. Revising uncertainty for the
same question may repair Phase 2; changing the unit, outcome, mode, or estimand
starts a new or revised cycle. A retrospective reanalysis does not become
prospective merely because the workflow is described afterward.

# Summary and Reading Bridge {#sec-ch7-summary}

A simple regression starts with a comparison and a unit, not a coefficient
table. Its fitted values and residuals are calculations on records.
OLS connects the slope to covariance and predictor variation; its intercept
places the line through the two means. $R^2$, residual SD, slope uncertainty,
and new-case variation answer different questions.

Across the 102 recorded occupations, the fitted line rises by 0.8988 thousand
1971 Canadian dollars---about 899 dollars---for each additional year of
occupation-average education over the observed 6.38--15.97-year range. This is
a bounded finite-record summary, not an individual, causal, or population
return.

Known-truth repetition shows why fitted slopes vary, what their standard error
estimates, and why 95% coverage belongs to a procedure rather than to one fixed
parameter after the data are observed. Noise, predictor spread, sample size,
and dependence affect precision in different ways.

Reading an argument locates the offered support and reported answer.
Quality evaluation then asks whether that support is adequate for the
declared claim. The auxiliary workflow explains how such an analysis and
record can be constructed. None of these activities permits a finite-record
slope to acquire a population interpretation simply because software printed
a standard error.

For another accessible account of lines, residuals, and fit, see
[OpenIntro's single-predictor regression tutorial](https://openintro-ims.netlify.app/model-slr).
For a compact executable route from a fitted model to intervals and checks, see
the [ISLR authors' Chapter 3 R lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch3-linreg-lab.html).
These provide complementary practice; the present chapter additionally makes
the finite-record and model-based questions explicit. Consult
[the carData reference manual](https://cran.r-project.org/web/packages/carData/carData.pdf)
for data documentation, and @fox2019 and @gelman2020 for broader regression
treatments.

Chapter 8 next asks conditional-comparison and coordinated
association/prediction questions. It begins adjustment by refitting the
education-only model on the same 98 type-complete occupation records: moving
from 102 to 98 is a sample change, not an adjustment effect. A short
indicator-variable bridge connects regression to group comparisons; fuller
ANOVA algebra stays optional. Interaction, where one comparison depends on
another predictor, receives its own later treatment.

# Exercises {.unnumbered}

The prompts below use **Foundational** and **Integrative** levels. Where
rounded summaries are supplied, answers within 0.02 in the stated units are
sufficient. No optional appendix is needed for Exercises 1--6.

<!-- prompt-id: CH07-E01 -->
## Exercise 1: Read the unit and the line (Foundational) {.unnumbered #ex-ch07-01}

<!-- exercise-id: ch07-ex01 -->

A teaching report describes 80 occupations with
$\widehat{\mathrm{income}}=-1.5+0.7\,\mathrm{education}$, where income is in
thousand dollars and average education ranges from 7 to 16 years.
Interpret the slope and the intercept. Calculate the fitted income at 12 years
and the residual for an occupation at 12 years with income 9.0.
Explain why the slope does not give an individual's gain from another year
of schooling.

<!-- prompt-id: CH07-E02 -->
## Exercise 2: Connect covariance and the fitted line (Foundational) {.unnumbered #ex-ch07-02}

<!-- exercise-id: ch07-ex02 -->

A set of records has $\bar x=4,\bar y=10,S_{xx}=20,S_{xy}=30$.
Calculate the OLS slope and intercept, and the fitted value at $x=5$.
Show that the line passes through the mean point. If $s_x=2$ and
$s_y=5$, calculate the correlation implied by these summaries.
What would prevent fitting a slope if every $x$ were equal?

<!-- prompt-id: CH07-E03 -->
## Exercise 3: Separate fit and uncertainty (Foundational) {.unnumbered #ex-ch07-03}

<!-- exercise-id: ch07-ex03 -->

An intercept-and-slope regression has $n=22$, $RSS=80$, $TSS=144$,
$S_{xx}=100$, and fitted slope 0.8. Calculate residual SD, $R^2$, and the
classical slope SE. Use $t_{.975,20}=2.086$ to obtain a 95% slope interval.
State the process assumptions needed for that interval and explain why it
does not by itself establish that the line fits closely or identifies a cause.

<!-- prompt-id: CH07-E04 -->
## Exercise 4: Mean or new case? (Foundational) {.unnumbered #ex-ch07-04}

<!-- exercise-id: ch07-ex04 -->

Under the stated independent normal-error constant-variance model, a fitted
value at $x_0=\bar x$ is 12, $n=25$, and $s_e=3$.
Use $t_{.975,23}=2.069$. Calculate the SE and 95% interval for the conditional
mean, then the corresponding prediction SE and interval for one independent
new case at the same $x_0$. Explain the extra 1 in the prediction expression.
Would either interval validate the model in a different population?

<!-- prompt-id: CH07-E05 -->
## Exercise 5: Locate first, evaluate second (Integrative) {.unnumbered #ex-ch07-05}

<!-- exercise-id: ch07-ex05 -->

A report states: "We used a regression program, so the coefficient is a causal
effect. Our 60 records are school averages. The fitted slope is 4 test-score
points per additional average study hour. The observed study-hour range is
2--6." No assignment or sampling documentation is provided.

Identify an offered justification and its documentary status. State the
reported claim before evaluating it. Apply Justification adequacy and
Transition fidelity. Explain whether an individual causal conclusion is
supported, and identify a narrower descriptive question that the fitted line
could address without pretending it was the original causal question.

<!-- prompt-id: CH07-E06 -->
## Exercise 6: Respond to sensitivity (Integrative) {.unnumbered #ex-ch07-06}

<!-- exercise-id: ch07-ex06 -->

In a descriptive analysis of 102 occupations, the slope is 0.90.
A valid, high-income occupation has the largest Cook's distance; removing it
gives 0.79. A curved-fit check changes the shape, and an HC3 calculation changes
the slope SE from 0.13 to 0.15 without changing the line.

What does each check tell you? What does it not tell you? Explain why deleting
the occupation automatically is inappropriate. Distinguish a variance repair
for the same estimand from replacing a constant process slope with a different
estimand. Write a short sensitivity disclosure for the descriptive analysis.

# Appendix A: Extended regression mathematics and interval derivations {.unnumbered #sec-ch7-app-a}

The main text uses the identities needed to read a fitted line. This appendix
derives them and distinguishes identities about the observed records from
claims about hypothetical repetitions. All occupation calculations retain
the same 102 rows, education in years, and income in thousand 1971 Canadian
dollars. Recreating these objects makes the appendix code usable independently
of the main-text examples.

```{r}
#| label: ch7app-a-data
#| echo: true
ch7app_data <- carData::Prestige
ch7app_data$occupation <- rownames(ch7app_data)
ch7app_data$income_k <- ch7app_data$income / 1000
ch7app_model <- lm(income_k ~ education, data = ch7app_data)
ch7app_x <- ch7app_data$education
ch7app_y <- ch7app_data$income_k
ch7app_n <- nrow(ch7app_data)
ch7app_Sxx <- sum((ch7app_x - mean(ch7app_x))^2)
ch7app_Sxy <- sum(
  (ch7app_x - mean(ch7app_x)) * (ch7app_y - mean(ch7app_y))
)
```

## Derivatives, normal equations, and the minimizing line {.unnumbered}

For candidate intercept $b_0$ and slope $b_1$, define

$$
Q(b_0,b_1)=\sum_{i=1}^n(y_i-b_0-b_1x_i)^2.
$$

Squared residuals cannot cancel and give large vertical discrepancies greater
weight. Differentiating with respect to each candidate coefficient gives

$$
\frac{\partial Q}{\partial b_0}
=-2\sum_i(y_i-b_0-b_1x_i),\qquad
\frac{\partial Q}{\partial b_1}
=-2\sum_i x_i(y_i-b_0-b_1x_i).
$$

At the minimum both derivatives are zero. The resulting **normal equations**
are

$$
n\widehat\beta_0+\widehat\beta_1\sum_i x_i=\sum_i y_i,
\qquad
\widehat\beta_0\sum_i x_i+\widehat\beta_1\sum_i x_i^2
=\sum_i x_iy_i.
$$

The first equation gives
$\widehat\beta_0=\bar y-\widehat\beta_1\bar x$.
Substitution into the second yields

$$
\widehat\beta_1
=\frac{\sum_i x_iy_i-n\bar x\bar y}
       {\sum_i x_i^2-n\bar x^2}
=\frac{S_{xy}}{S_{xx}}.
$$

Here $S_{xx}=\sum_i(x_i-\bar x)^2>0$ is required: a slope cannot be
identified when every predictor value is the same. Because sample covariance
and variance use the same denominator $n-1$,

$$
\widehat\beta_1
=\frac{s_{xy}}{s_x^2}
=r\frac{s_y}{s_x},
\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x.
$$

In the occupation data,

$$
\widehat\beta_1
=\frac{675.8041}{751.8852}=0.8988,
\qquad
\widehat\beta_0=6.7979-0.8988(10.7380)\approx-2.8534.
$$

The ratio has units of thousand dollars per education year. Unlike a
correlation, the slope changes when a variable's units change. Standardizing
both variables makes their sample standard deviations one, so the simple
intercept-model slope becomes $r$.

To verify that the stationary point is the unique minimum, write any other
line as $\widehat\beta_0+d_0+(\widehat\beta_1+d_1)x_i$. The normal
equations make the cross term with the OLS residuals vanish, leaving

$$
Q(\widehat\beta_0+d_0,\widehat\beta_1+d_1)
=RSS+n(d_0+d_1\bar x)^2+d_1^2S_{xx}.
$$

With $S_{xx}>0$, every different line has larger RSS. If a candidate line is
constrained to pass through $(\bar x,\bar y)$, its excess RSS is simply
$S_{xx}(b_1-\widehat\beta_1)^2$. The numbered method route makes this
comparison visible with five candidate lines.

## Residual constraints and projection geometry {.unnumbered}

At the OLS solution the derivative equations become
$\sum_i e_i=0$ and $\sum_i x_ie_i=0$. Thus the residuals have zero
sample covariance with the predictor and mean zero **by construction**.
They are not evidence that the unknown population errors are independent,
normally distributed, or conditionally mean zero.

For the geometry, let $\mathbf X$ be the $n\times2$ matrix with first
column all ones and second column containing the predictor. Bold
$\mathbf X$ is a matrix, distinct from a single predictor value $x_i$.
The normal equations can be written

$$
\mathbf X^{\mathsf T}(\mathbf y-\mathbf X\widehat{\boldsymbol\beta})
=\mathbf0,
\qquad
\widehat{\boldsymbol\beta}
=(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}\mathbf y.
$$

The fitted vector is the projection of $\mathbf y$ onto the two-dimensional
space spanned by the intercept and predictor columns. The residual vector
is perpendicular to that space. Writing
$\mathbf H=\mathbf X(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}$,
we have $\widehat{\mathbf y}=\mathbf H\mathbf y$ and
$\mathbf e=(\mathbf I-\mathbf H)\mathbf y$.

Because the model includes an intercept, both $\widehat{\mathbf y}$ and
the constant vector $\bar y\mathbf1$ lie in the fitted space. Consequently,

$$
\begin{aligned}
\sum_i(y_i-\bar y)^2
&=\sum_i\{(\widehat y_i-\bar y)+e_i\}^2\\
&=\sum_i(\widehat y_i-\bar y)^2+\sum_i e_i^2,
\end{aligned}
$$

since $\sum_i(\widehat y_i-\bar y)e_i=0$. This proves
$TSS=RegSS+RSS$, an algebraic decomposition of observed variation, not
a division into causal contributions. For these records,
$1820.8134=607.4214+1213.3920$, in squared thousand-dollar units.

Also $RegSS=\widehat\beta_1^2S_{xx}=S_{xy}^2/S_{xx}$. Dividing by
$TSS=S_{yy}$ proves $R^2=S_{xy}^2/(S_{xx}S_{yy})=r^2$ for simple
regression **with an intercept**. The identities need not hold for a
no-intercept fit. When fitted values are nonconstant,
$\operatorname{cor}(y,\widehat y)^2=R^2$ as well.

There are two fitted directions and $n-2$ residual directions. Under a
correct linear mean with independent constant-variance errors,
$E(RSS\mid\mathbf X)=(n-2)\sigma^2$. Thus
$s_e^2=RSS/(n-2)$ estimates the error variance, while $s_e$ estimates its
standard deviation in the outcome's units. The square root is not itself
an exactly unbiased estimator of $\sigma$. Without that stochastic model,
$s_e$ remains a computable residual-scale summary, not proof of a common
population error variance.

## From an error model to slope and intercept uncertainty {.unnumbered}

The 102-record least-squares line is already determined by those records.
The next derivation concerns a **separate model-based question**: what
constant conditional-mean coefficient generated hypothetical independent
occupation outcomes at the observed education values? Assume

$$
Y_i=\beta_0+\beta_1x_i+\varepsilon_i,
\quad E(\varepsilon_i\mid\mathbf X)=0,
\quad \operatorname{Var}(\boldsymbol\varepsilon\mid\mathbf X)
=\sigma^2\mathbf I.
$$

These are offered working conditions, not consequences of the package's
availability or of fitting `lm()`. The slope error is

$$
\widehat\beta_1-\beta_1
=\frac{\sum_i(x_i-\bar x)\varepsilon_i}{S_{xx}}.
$$

Independence and common variance then give

$$
\operatorname{Var}(\widehat\beta_1\mid\mathbf X)
=\frac{\sigma^2}{S_{xx}},
\qquad
SE(\widehat\beta_1)=\frac{s_e}{\sqrt{S_{xx}}}.
$$

Because $\widehat\beta_0=\bar Y-\widehat\beta_1\bar x$ and
$\operatorname{Cov}(\bar Y,\widehat\beta_1\mid\mathbf X)=0$,

$$
\operatorname{Var}(\widehat\beta_0\mid\mathbf X)
=\sigma^2\left(\frac1n+\frac{\bar x^2}{S_{xx}}\right),
\qquad
\operatorname{Cov}(\widehat\beta_0,\widehat\beta_1\mid\mathbf X)
=-\frac{\bar x\sigma^2}{S_{xx}}.
$$

This covariance explains why simply adding the variances of an uncentered
intercept and slope prediction would be wrong. If the predictor is centered
at its observed mean, the intercept and slope estimates are uncorrelated
under the classical model, and the intercept SE is $s_e/\sqrt n$.

Under the stated linear-mean, uncorrelated, equal-variance conditions, OLS
is the best linear unbiased estimator: it has the smallest variance among
estimators linear in the outcomes and unbiased for the coefficients. This
Gauss--Markov statement does not require normal errors and does not establish
causal identification, valid measurement, or population coverage [@fox2016;
@kutner2005].

Add a jointly normal error vector conditional on the predictor values. Then
the standardized coefficient error is normal, $RSS/\sigma^2$ has a
$\chi^2_{n-2}$ distribution, and these quantities are independent. Their
ratio gives

$$
\frac{\widehat\beta_j-\beta_j}{SE(\widehat\beta_j)}
\sim t_{n-2},\qquad j=0,1.
$$

That is the basis of exact classical conditional coverage for
$\widehat\beta_j\pm t_{.975,n-2}SE(\widehat\beta_j)$, and of the
corresponding null test. For the occupation slope the calculation is
$0.8988\pm1.9840(0.1270)=[0.6468,1.1508]$ thousand dollars per year.
It illustrates the model-based procedure; the diagnostics in Section 7
do not support treating its nominal exact coverage as established here.

Fixed-$X$ derivations condition on the predictor values; those values need
not literally have been experimentally fixed. A random-$X$ conditional-mean
model can use the same conditional argument when its error conditions hold.
Neither route demonstrates that these occupations were probability sampled.
Normality of the marginal distribution of education or income is not required.
The assumptions concern conditional errors, not the two marginal histograms.

## Why a new-case interval has an additional variance term {.unnumbered}

For a supported value $x_0$, define the model mean
$\mu_0=\beta_0+\beta_1x_0$ and its estimate
$\widehat\mu_0=\widehat\beta_0+\widehat\beta_1x_0$.
Substituting the coefficient variances and covariance above gives

$$
\begin{aligned}
\operatorname{Var}(\widehat\mu_0\mid\mathbf X)
&=\operatorname{Var}(\widehat\beta_0)
 +x_0^2\operatorname{Var}(\widehat\beta_1)
 +2x_0\operatorname{Cov}(\widehat\beta_0,\widehat\beta_1)\\
&=\sigma^2\left\{\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}\right\}.
\end{aligned}
$$

The common conditioning on $\mathbf X$ is suppressed in the first line.
A confidence interval for this mean uses the square root of this expression
with $s_e$ replacing $\sigma$.

A new occupation outcome is instead $Y_0=\mu_0+\varepsilon_0$.
If its error is independent of the fitted data and comes from the same
normal, equal-variance process,

$$
\operatorname{Var}(Y_0-\widehat\mu_0\mid\mathbf X,x_0)
=\sigma^2\left\{1+\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}\right\}.
$$

The extra $1$ is new outcome variation; it does not disappear even if the
model mean were known. Both standard errors use the same $t_{.975,n-2}$
critical value for their classical 95% intervals under the full model.

\Needspace{20\baselineskip}

```{r}
#| label: ch7app-a-prediction
#| echo: true
#| results: hide
ch7app_new <- data.frame(education = c(8, 10, 12, 15))
ch7app_mean_intervals <- predict(
  ch7app_model, newdata = ch7app_new, interval = "confidence"
)
ch7app_case_intervals <- predict(
  ch7app_model, newdata = ch7app_new, interval = "prediction"
)
ch7app_mean_intervals
ch7app_case_intervals
```

The table puts both sets of limits beside the common fitted values.

```{r}
#| echo: false
knitr::kable(data.frame(
  Education = ch7app_new$education,
  Fitted = ch7app_mean_intervals[, "fit"],
  Mean_lower = ch7app_mean_intervals[, "lwr"],
  Mean_upper = ch7app_mean_intervals[, "upr"],
  Case_lower = ch7app_case_intervals[, "lwr"],
  Case_upper = ch7app_case_intervals[, "upr"]
), digits = 2)
```

At 10 years, the point prediction is 6.13 thousand dollars, the mean interval
is [5.43, 6.84], and the new-occupation interval is [-0.81, 13.08]. The
negative endpoint exposes the lower-tail limitation of this unbounded
normal-linear model. Cutting it off at zero does not produce a validated
95% interval. An outcome transformation or a model respecting the boundary
would change the analysis and would need its own interpretation and checks.

Both intervals widen away from $\bar x$; their narrowest point is the mean
education value. With more comparable independent observations and stable
predictor spread, mean uncertainty tends to shrink. The new-case interval
retains a nonzero error-variation floor. None of these formulas includes
uncertainty from an inappropriate mean function, biased measurement, or a
shift to another labour market. They are pointwise model intervals, not a
simultaneous band or empirical out-of-sample validation. Their target is a
new occupation-level record from the declared process, never an individual
worker's income.

# Appendix B: Extended covariance and projection targets {.unnumbered #sec-ch7-app-b}

This appendix retains optional matrix and target-level extensions. The
known-truth and repeated-sample demonstrations now appear in the numbered
method route, where they explain the ordinary slope standard error and
coverage interpretation. Nothing programmed in those simulations validates
the empirical occupation model.


## Unequal variance and the HC3 sandwich calculation {.unnumbered}

For independent observations, a heteroskedasticity-consistent covariance
estimate allows the error variance to differ across predictor values.
It changes estimated coefficient uncertainty, not the OLS coefficients.
Let $\mathbf x_i=(1,x_i)^{\mathsf T}$ denote a design column for case $i$,
and let $h_i=\mathbf x_i^{\mathsf T}(\mathbf X^{\mathsf T}\mathbf X)^{-1}
\mathbf x_i$ be its leverage. In simple regression with an intercept,

$$
h_i=\frac1n+\frac{(x_i-\bar x)^2}{S_{xx}}.
$$

HC3 uses the leverage-adjusted squared residual
$w_i=e_i^2/(1-h_i)^2$ inside

$$
\widehat{\operatorname{Var}}_{HC3}(\widehat{\boldsymbol\beta})
=(\mathbf X^{\mathsf T}\mathbf X)^{-1}
\left(\sum_i w_i\mathbf x_i\mathbf x_i^{\mathsf T}\right)
(\mathbf X^{\mathsf T}\mathbf X)^{-1}.
$$

The two outer factors are often called the sandwich's bread and the middle
factor its meat. The original heteroskedasticity-consistent construction uses
case-specific squared residuals; HC3 inflates contributions from high-leverage
cases to address finite-sample behavior [@white1980; @mackinnon1985;
@long2000]. It does not reveal each case's unknown true variance.

```{r}
#| label: ch7app-b-hc3
#| echo: true
ch7app_X <- model.matrix(ch7app_model)
ch7app_e <- resid(ch7app_model)
ch7app_h <- hatvalues(ch7app_model)
ch7app_w <- (ch7app_e / (1 - ch7app_h))^2
ch7app_bread <- solve(crossprod(ch7app_X))
ch7app_meat <- t(ch7app_X) %*% (ch7app_X * ch7app_w)
ch7app_vcov_hc3 <- ch7app_bread %*% ch7app_meat %*% ch7app_bread
ch7app_se_hc3 <- sqrt(diag(ch7app_vcov_hc3))
ch7app_critical <- qt(0.975, df.residual(ch7app_model))
ch7app_hc3_ci <- cbind(
  lower = coef(ch7app_model) - ch7app_critical * ch7app_se_hc3,
  upper = coef(ch7app_model) + ch7app_critical * ch7app_se_hc3
)
c(
  classical_slope_SE = coef(summary(ch7app_model))["education", "Std. Error"],
  HC3_slope_SE = ch7app_se_hc3["education"]
)
ch7app_hc3_ci["education", ]
```

The slope SE changes from 0.127 to 0.149 thousand dollars per year. Using
the same $t_{100}$ critical value yields [0.603, 1.195], an **approximate
sensitivity interval**, not an exact normal-theory $t$ interval under
heteroskedasticity. A maintained implementation such as
`sandwich::vcovHC(ch7app_model, type = "HC3")` is preferable in applied
work once the calculation is understood [@zeileis2004].

The comparison challenges the common-variance uncertainty calculation.
HC3 does not correct a nonlinear mean, dependent occupations, confounding,
or errors in measurement. Under the classical measurement-error setup,
independent noise added to a true predictor attenuates the population
simple-regression slope, while independent mean-zero outcome noise increases
error variation without changing that slope [@cohen2003; @fox2016]. Residual
plots cannot recover the true predictor or establish measurement validity.

## Diagnostic limits and a projection target {.unnumbered}

The numbered route now contains the ordinary curvature, leverage, influence,
and HC3 sensitivity work. Its same-record comparison preserves the original
outcome scale and analytic sample. The appendix keeps the more technical target
that becomes relevant if a constant conditional-mean slope is abandoned.

Neither a residual plot nor a quadratic comparison can certify independent
occupations. Dependence requires evidence about how occupational categories
and their measurements were produced. A quantile plot can flag tail behavior
relevant to normal-theory inference, but cannot establish the sampling
mechanism. Extrapolation is likewise a support problem: the observed education
range is 6.38--15.97 years. A line can return a value at 20 years without
making that value empirically supported.

Curvature leaves the finite-record OLS slope well defined but challenges
its interpretation as a constant conditional-mean contrast. A **best linear
projection** is a distinct stochastic target,

$$
(\beta_0^*,\beta_1^*)
=\arg\min_{a,b}E\{(Y-a-bX)^2\},
\qquad
\beta_1^*=\frac{\operatorname{Cov}(X,Y)}{\operatorname{Var}(X)},
$$

when the required moments exist and $\operatorname{Var}(X)>0$. Its residual
is orthogonal to 1 and $X$ in the population, but need not have zero mean
at every $X=x$. With independent identically distributed pairs and suitable
moment conditions, robust covariance methods can support asymptotic
inference for that projection target. Such a target depends on the joint
distribution of occupations and education and is not a constant
conditional-mean derivative. It must be declared as a new or revised
model-based question; it cannot retroactively receive an exact classical
interval from the rejected linear-mean model. The descriptive conclusion
about the observed least-squares line does not require making that extension.

# References {.unnumbered}

::: {#refs}
:::
