---
title: "Binary Logistic Regression"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    documentclass: scrartcl
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    latex-auto-install: false
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - ../koma_heading_tags.tex
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{16\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AtBeginDocument{\renewcommand{\boldsymbol}[1]{\symbf{#1}}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---


```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.pos = "!ht")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 7, width = 68)
library(ggplot2)
theme_set(theme_minimal(base_size = 11))
d0 <- carData::Cowles
dat <- d0
dat$event <- as.integer(dat$volunteer == "yes")
dat$sex <- factor(dat$sex, levels = c("female", "male"))
m0 <- glm(event ~ 1, data = dat, family = binomial)
m1 <- glm(event ~ extraversion, data = dat, family = binomial)
m2 <- glm(event ~ sex + neuroticism + extraversion, data = dat, family = binomial)
m3 <- glm(event ~ sex + neuroticism * extraversion, data = dat, family = binomial)
mq <- glm(event ~ sex + neuroticism + extraversion + I(extraversion^2), data = dat, family = binomial)
md <- glm(event ~ sex + neuroticism + extraversion,
  data = dat[-which.max(cooks.distance(m2)), ], family = binomial)
set.seed(1010)
n <- 1500
hours <- round(runif(n, 0, 20), 1)
tutor <- rbinom(n, 1, .35)
prob <- plogis(-2.2 + .30 * hours + .60 * tutor)
sim <- data.frame(pass = rbinom(n, 1, prob), hours, tutor)
ms <- glm(pass ~ hours + tutor, data = sim, family = binomial)
ml <- lm(pass ~ hours + tutor, data = sim)
set.seed(1012)
repcoef <- replicate(500, {
  hours <- round(runif(300, 0, 20), 1)
  tutor <- rbinom(300, 1, .35)
  pass <- rbinom(300, 1, plogis(-2.2 + .30 * hours + .60 * tutor))
  m <- glm(pass ~ hours + tutor, family = binomial)
  c(coef(m)[2:3], cover = abs(coef(m)[2] - .30) <= 1.96 * sqrt(vcov(m)[2, 2]))
})
phat <- fitted(m2)
bin <- cut(phat, quantile(phat, seq(0, 1, .05)), include.lowest = TRUE)
bins <- data.frame(fit = as.numeric(tapply(phat, bin, mean)),
  resid = as.numeric(tapply(dat$event - phat, bin, mean)), n = as.numeric(table(bin)))
bins$se <- sqrt(bins$fit * (1 - bins$fit) / bins$n)
dec <- cut(phat, quantile(phat, seq(0, 1, .1)), include.lowest = TRUE)
cal <- data.frame(fit = as.numeric(tapply(phat, dec, mean)),
  obs = as.numeric(tapply(dat$event, dec, mean)), n = as.numeric(table(dec)))
cal$se <- sqrt(cal$obs * (1 - cal$obs) / cal$n)
eligible <- which((dat$extraversion <= 8 & dat$event == 0) |
  (dat$extraversion >= 17 & dat$event == 1))
set.seed(1011)
sep <- dat[sort(sample(eligible, 60)), ]
sepfit <- glm(event ~ extraversion, data = sep, family = binomial)
stopifnot(nrow(dat) == 1421, sum(dat$event) == 597,
  abs(as.numeric(logLik(m2)) + 953.03064262484) < .00001,
  abs(coef(m2)["extraversion"] - .0663253) < .000001)
```

{{< pagebreak >}}

::: {.callout-note title="Chapter overview"}
A yes-or-no outcome has a mean that is a probability. This chapter builds
a model for that probability, connects it to odds and log-odds, and explains
why the same odds ratio can imply very different probability differences.
A known-process certification example makes the model's behavior visible;
a study of student volunteering supplies the empirical analysis. Maximum
likelihood, model comparisons and diagnostic displays lead to a report
whose event, scale, uncertainty and scope are explicit. The appendices
extend numerical fitting and simulation without supplying a prerequisite
for the main analysis.
:::

Cowles and Davis recorded personality scores and volunteering for an
experiment among 1,421 introductory psychology students [@cowles1987].
There were 597 volunteers and 824 non-volunteers. The question is how the
probability of volunteering differs across extraversion scores, and how
that comparison changes after allowing for recorded sex and neuroticism.
A department interested in recruitment needs an answer on a usable scale:
a detectable association need not separate future volunteers from other
students well enough to guide a decision.

The distributed records are `carData::Cowles` [@carData2022; @fox2019].
They contain one row per student, with no missing values in these fields.
The extract supplies no probability-sampling or classroom-clustering
information. The fitted comparisons therefore describe these recorded
students under a working independent-row model; generalization and causal
interpretation require additional design information.



Displayed results are rounded to four decimal places, with trailing zeros
dropped unless they align a comparison. Calculations using different
rounding or software settings may differ slightly.

## A Binary Outcome and Its Probability {#ch11-binary}

### Define the event and the comparison {#ch11-event}

A binary outcome has two categories. A binary model treats one as the
**event**, coded 1, and the other as the **non-event**, coded 0. Naming
the event determines what every coefficient and predicted probability
means. Here the event is volunteering, and the **prevalence**, the
proportion of records that are events, is .4201. The recorded non-event
means only that the student did not volunteer; it does not tell us whether
the student declined, was unavailable, or never received the invitation.
Those distinctions concern measurement, before any statistical fit.



```{r}
#| eval: false
#| echo: true
d0 <- carData::Cowles
dat <- d0
dat$event <- as.integer(dat$volunteer == "yes")
table(dat$sex, dat$event)
mean(dat$event)
```



| Recorded sex | Volunteers | Non-volunteers | Proportion volunteering | Odds of volunteering |
|:--|--:|--:|--:|--:|
| Female | 349 | 431 | .4474 | .8097 |
| Male | 248 | 393 | .3869 | .6310 |

: One observed table supports more than one scale of comparison. {#tbl-ch11-crude tbl-colwidths="[22,16,18,22,22]"}

The female-minus-male probability difference is .0605; the ratio of their
odds is 1.2832. Both describe the same table, but they answer different
questions. The first compares proportions directly. The second compares
events relative to non-events in each group. Neither adjusts for the two
personality scores or identifies a consequence of changing a person's sex.

Write $\pi(x)=P(Y=1\mid X=x)$. Three possible comparisons of units
one predictor point apart are

$$
\pi(x+1)-\pi(x),\qquad
\frac{\pi(x+1)}{\pi(x)},\qquad
\frac{\pi(x+1)/\{1-\pi(x+1)\}}{\pi(x)/\{1-\pi(x)\}}.
$$

They are the probability difference, risk ratio and odds ratio. A model
with a constant logistic slope holds the third constant across $x$;
the other two generally vary. An estimand for a binary outcome must name
its scale as well as the units and comparison.

### A probability model and its link {#ch11-glm}

Applying least squares to a 0/1 outcome gives a **linear probability
model**. Its slope is already on a probability-difference scale, which
can be useful over a limited range. Its weakness as a general probability
model is visible before fitting: an unrestricted line can leave $[0,1]$,
and one fixed probability difference cannot remain plausible near both
boundaries. Moreover a binary outcome has variance $\pi(1-\pi)$, so its
conditional variance changes with its mean. Ordinary constant-variance
standard errors are inappropriate; robust uncertainty can address that
variance calculation but cannot keep a straight mean function in range.
Normal-error small-sample $t$ theory does not apply to two-valued errors.

The logistic alternative declares the outcome model and mean function
separately:

$$
Y_i\mid \mathbf{x}_i\sim\operatorname{Bernoulli}(\pi_i),\quad
E(Y_i\mid \mathbf{x}_i)=\pi_i,\quad
\operatorname{Var}(Y_i\mid \mathbf{x}_i)=\pi_i(1-\pi_i),
$$

$$
\eta_i=\beta_0+\sum_{j=1}^{p}\beta_jx_{ij},\qquad
\operatorname{logit}(\pi_i)=\log\frac{\pi_i}{1-\pi_i}=\eta_i,
\qquad \pi_i=\frac{1}{1+e^{-\eta_i}}.
$$

Here $\eta_i$ is the **linear predictor**, an additive combination of
predictors. The **link** connects the conditional mean to that predictor;
it transforms the probability, not the observed 0 or 1. The inverse logit
turns any finite $\eta_i$ into a probability strictly between 0 and 1.
It passes through .5 at $\eta_i=0$ and is steepest there. The model is
linear in log-odds and curved in probability. It has no separate residual
variance parameter because the Bernoulli mean determines its variance.

A response distribution, a linear predictor and a link form the generalized
linear model (GLM) framework [@nelder1972]. Changing the response and link
extends this framework to other outcomes, while retaining the distinction
between model scale and reporting scale.


| Component | Chapter 11 | Chapter 12, nominal or ordered | Chapter 12, counts |
|:--|:--|:--|:--|
| Outcome | Binary event | Nominal or ordered category | Count, with an exposure where needed |
| Response distribution | Bernoulli or binomial | Multinomial | Poisson or negative binomial |
| Link | Logit | Generalized or cumulative logit | Log |
| Exponentiated coefficient | Odds ratio | Relative-risk or cumulative odds ratio | Rate ratio |

: Only the response distribution and the link change. {#tbl-ch11-glm-components tbl-colwidths="[24,24,28,24]"}



### Conditions and the scope of interpretation {#ch11-conditions}

| Condition | What it supplies or limits |
|:--|:--|
| AS1: a Bernoulli outcome with a well-defined event | The meaning of $\pi_i$ and its variance; coding must represent the question |
| AS2: the log-odds is linear in the entered terms | The conditional comparison represented by each coefficient; curves or products require explicit terms |
| AS3: independent rows given the predictors | The likelihood product and its usual standard errors; repeated and clustered rows need dependence-aware methods |
| AS4: predictors are measured and coded as intended | The interpretation of a comparison; measurement error can change the relationship being estimated |
| AS5: supported profiles, overlap and no separating predictor combination | A defensible comparison and a finite maximum-likelihood estimate |
| AS6: no perfect linear dependence among predictor columns | Unique coefficient coordinates; near dependence can make estimates imprecise |

: Conditions of the working logistic model. {#tbl-ch11-conditions tbl-colwidths="[44,56]"}

There is no homoskedasticity or Normal-residual condition. The mean model
is checked on the logit scale, and the study design supplies information
about independence. Observing both events and non-events in the dataset
is necessary but does not rule out a combination of predictors that
separates them perfectly. Nor must both outcomes appear at every exact
continuous predictor value. Support means that the proposed profiles and
contrasts are represented well enough to make their interpretation credible.

A coefficient compares units agreeing on the other included predictors.
It becomes a causal effect only with an adequate design and identification
argument. A further complication is **noncollapsibility**: a conditional
odds ratio can differ from a marginal odds ratio even without confounding,
because averaging nonlinear probabilities does not preserve an odds ratio
[@agresti2018]. Movement after adjustment alone therefore does not diagnose
confounding. A standardized probability comparison names an averaging
population explicitly and can make the substantive magnitude clearer.

## Three Scales: Probability, Odds, and Log-Odds {#ch11-scales}

### Translate one quantity three ways {#ch11-conversions}

A probability compares events with all units. Odds compare events with
non-events. Log-odds put that positive ratio on an additive, unbounded scale:

$$
\text{odds}=\frac{\pi}{1-\pi},\qquad
\pi=\frac{\text{odds}}{1+\text{odds}},\qquad
\text{log-odds}=\log\frac{\pi}{1-\pi}.
$$

At probability .5, events and non-events are equally common: odds are 1
and log-odds are 0. At probability .75, there are three events per
non-event. The three scales contain the same information; they organize
comparisons differently.



```{r}
#| eval: false
#| echo: true
p <- c(.05, .20, .42, .50, .75, .90)
cbind(probability = p, odds = p / (1 - p), log_odds = qlogis(p))
```


| Probability | Odds | Log-odds |
|--:|--:|--:|
| .05 | .0526 | $-2.9444$ |
| .20 | .2500 | $-1.3863$ |
| .42 | .7241 | $-.3228$ |
| .50 | 1.0000 | .0000 |
| .75 | 3.0000 | 1.0986 |
| .90 | 9.0000 | 2.1972 |

: Probability, odds, and log-odds for selected probabilities; the odds scale is multiplicative around 1 and the log-odds scale additive around 0. {#tbl-ch11-scales tbl-colwidths="[34,33,33]"}



Odds exceed probabilities for $0<\pi<1$, with a widening gap toward
1. Complementary probabilities have log-odds equal in magnitude and
opposite in sign. These features explain why an event reversal inverts
an odds ratio, and why odds and risk ratios agree approximately only
when both compared probabilities are small.

### Equal logit steps, unequal probability steps {#ch11-curve}

The inverse-logit function is the cumulative distribution function of a
standard logistic variable, evaluated at $\eta$. This mathematical
connection offers a threshold-crossing intuition without changing the
observed Bernoulli outcome model. A one-unit step in $\eta$ has the same
size everywhere on the logit scale; its probability consequence depends
on the starting point.



```{r}
#| eval: false
#| echo: true
plogis(c(-3, -2, 0, 1))
diff(plogis(c(-3, -2)))
diff(plogis(c(0, 1)))
```



```{r}
#| label: fig-ch11-logit-curves
#| fig-cap: "Equal steps on the linear-predictor scale give unequal probability differences."
#| fig-alt: "An S-shaped inverse-logit curve rises through probability one half at zero. A one-unit step near the centre raises probability by .2311, compared with .0718 in the left tail."
#| fig-height: 3.6
#| fig-width: 6.2
g <- data.frame(eta = seq(-6, 6, length.out = 401))
g$p <- plogis(g$eta)
ggplot(g, aes(eta, p)) + geom_line(linewidth = .8) +
  geom_hline(yintercept = .5, linetype = 3) +
  annotate("segment", x = c(-3, 0), xend = c(-2, 1),
    y = plogis(c(-3, 0)), yend = plogis(c(-3, 0)), linetype = 2) +
  annotate("segment", x = c(-2, 1), xend = c(-2, 1),
    y = plogis(c(-3, 0)), yend = plogis(c(-2, 1)), linewidth = 1.2) +
  annotate("text", x = c(-3.3, 1.8), y = c(.19, .65),
    label = c("difference .0718", "difference .2311"), size = 3.5) +
  labs(x = "Linear predictor", y = "Event probability")
```



Moving from $-3$ to $-2$ raises probability by .0718; moving from 0
to 1 raises it by .2311, 3.2191 times as much. A statement that one more
unit adds a fixed number of percentage points is therefore incomplete
for a logistic model. Coefficients and their uncertainty are convenient
on the logit scale, multiplicative comparisons on the odds scale, and
most practical decisions on the probability scale.

## Fit by Maximum Likelihood {#ch11-likelihood}

### From one probability to a coefficient vector {#ch11-loglik}

For a common probability, 32 successes in 50 fixed trials produce a
binomial likelihood maximized at .64. Here each student's probability
can differ through the predictor values, but the principle is the same:
hold the observed data fixed and ask which parameter values make them
most probable.

A small example makes the likelihood's job concrete. Three invented
observations have predictor values 0, 1 and 2 and outcomes 1, 0 and 1.
Compare two proposed models, A with logit $-1+x$ and B with logit $-2+x$.
The likelihood uses the probability of what happened: $p$ for an event
and $1-p$ for a non-event.

\Needspace{10\baselineskip}

| $x$ | $y$ | A: event | A: observed | B: event | B: observed |
|--:|--:|--:|--:|--:|--:|
| 0 | 1 | .2689 | .2689 | .1192 | .1192 |
| 1 | 0 | .5000 | .5000 | .2689 | .7311 |
| 2 | 1 | .7311 | .7311 | .5000 | .5000 |

: Each observed outcome contributes its own probability. {#tbl-ch11-small-likelihood tbl-colwidths="[8,8,21,21,21,21]"}

Predict which model has the larger product before calculating. Model A
helps both events but makes the non-event less probable. Its total
log-likelihood is nevertheless larger: $-2.3197$ against $-3.1333$.
Improving total fit need not improve every observation. Neither candidate
has yet been claimed to maximize likelihood.

```{r}
#| eval: false
#| echo: true
x <- c(0, 1, 2); y <- c(1, 0, 1)
sum(dbinom(y, 1, plogis(-1 + x), log = TRUE))
sum(dbinom(y, 1, plogis(-2 + x), log = TRUE))
```

With independent rows,

$$
L(\boldsymbol\beta)=\prod_{i=1}^{n}\pi_i^{y_i}(1-\pi_i)^{1-y_i},\qquad
\ell(\boldsymbol\beta)=\sum_{i=1}^{n}\{y_i\log\pi_i+(1-y_i)\log(1-\pi_i)\},
$$

where $\pi_i=\operatorname{logit}^{-1}(\mathbf{x}_i^{\mathsf T}\boldsymbol\beta)$ and
$\mathbf{x}_i$ includes the intercept entry 1. Each observed event contributes
its fitted probability and each non-event its fitted complement. Larger
log-likelihood means better agreement for these observations. The
likelihood is a function of parameters with the data fixed, not a
probability distribution over those parameters.

**Maximum likelihood** chooses the coefficient vector maximizing this
function. Its score equations are

$$
\frac{\partial\ell}{\partial\boldsymbol\beta}
=\sum_i \mathbf{x}_i(y_i-\pi_i)=0.
$$

The equation for the intercept says that the residuals sum to zero:
fitted probabilities total the observed event count. Aggregate calibration
is therefore built into a converged, unpenalized logistic fit with an
intercept; it cannot by itself establish a good model. The remaining
equations need numerical solution because the probabilities depend
nonlinearly on all coefficients.



```{r}
#| eval: false
#| echo: true
m2 <- glm(event ~ sex + neuroticism + extraversion,
  data = dat, family = binomial(link = "logit"))
logLik(m2)
sum(fitted(m2))
```



The adjusted volunteering fit has log-likelihood $-953.0306$.
Directly summing its Bernoulli contributions gives the same value, and
its fitted probabilities sum to 597, the observed number of volunteers.
Appendix A connects this numerical fitting to scalar boundary cases,
multiple starting values and local curvature.

### Curvature, uncertainty and convergence {#ch11-irls}

`glm()` uses **iteratively reweighted least squares** (IRLS): it forms
a working linear regression with adjusted outcomes and weights
$\widehat\pi_i(1-\widehat\pi_i)$, updates the coefficients, and repeats.
For a full-rank design without separation the logistic log-likelihood has
a unique finite maximum. A stopping flag says the algorithm met its
numerical criterion; the data and model still require scrutiny. This fit
reports convergence after four iterations.

Before reading a covariance matrix, consider what the curve is measuring.
In an invented two-group study, ten observations have $x=-1$ and ten have
$x=1$. Use the one-coefficient model $\operatorname{logit}(p)=b x$, whose
intercept is fixed at zero. If the groups contain 2 and 8 events, the
likelihood peaks at $b=\log4=1.3863$: moving either way makes the observed
pattern less probable. If instead they contain 0 and 10 events, increasing
$b$ keeps improving the likelihood. Every finite curve leaves some chance
of a contrary outcome, so the maximum is approached only as $b$ diverges.
This second pattern is **complete separation**.

```{r}
#| label: fig-ch11-likelihood-shapes
#| fig-cap: "Changing the event counts changes whether the likelihood has a finite maximum."
#| fig-alt: "For ten observations at each of x minus one and plus one, two and eight events give a curved peak at b 1.3863. Zero and ten events give a curve increasing toward zero without a finite peak."
#| fig-height: 3.6
#| fig-width: 6.2
b <- seq(-1, 6, length.out = 401)
finite <- 4 * plogis(-b, log.p = TRUE) + 16 * plogis(b, log.p = TRUE)
peak <- 4 * log(.2) + 16 * log(.8)
g <- rbind(data.frame(b, relative = finite - peak, counts = "2 and 8 events"),
  data.frame(b, relative = 20 * plogis(b, log.p = TRUE), counts = "0 and 10 events"))
g$counts <- factor(g$counts, levels = c("2 and 8 events", "0 and 10 events"))
ggplot(g, aes(b, relative)) + geom_line(linewidth = .8) +
  geom_hline(yintercept = 0, linetype = 3) + facet_wrap(~counts) +
  geom_point(data = data.frame(b = log(4), relative = 0, counts = "2 and 8 events"), size = 2) +
  labs(x = "Log-odds coefficient b", y = "Relative log-likelihood")
```

The dashed zero line represents the best attainable log-likelihood in the
finite case and the limiting value in the separated case. A flat tail is
not evidence for a small coefficient; here it shows why a finite maximum
is absent. Section 6.3 returns to this issue in the volunteering records.

Near the maximum, curvature describes how quickly fit deteriorates as
a coefficient moves. Flat curvature allows many neighboring values and
therefore larger standard errors. With design matrix $\mathbf X$ and diagonal
weight matrix $\mathbf W$, the model-based covariance is

$$
\widehat{\mathbf V}=(\mathbf{X}^{\mathsf T}\mathbf{W}\mathbf{X})^{-1},\qquad
\mathbf W=\operatorname{diag}\{\widehat\pi_i(1-\widehat\pi_i)\},\qquad
SE(\widehat\beta_j)=\sqrt{\widehat{\mathbf V}_{jj}}.
$$

A weight is largest at .5 and small near either boundary. Precision also
depends on predictor spread and overlap, so neither total sample size nor
event count alone determines it. Independent-row standard errors cannot
repair dependence or measurement error.

A **Wald interval** is $\widehat\beta_j\pm1.96SE$ for 95% coverage
under the large-sample Normal approximation, then exponentiated if the
reported quantity is an odds ratio. A **profile-likelihood interval**
instead keeps values not rejected by an appropriate likelihood-ratio
comparison while optimizing the other coefficients. It can reflect
asymmetry the Wald interval misses. The two approaches here differ by at
most .0015 on the odds-ratio scale; they can diverge markedly in small
samples or near separation [@hosmer2013].



```{r}
#| eval: false
#| echo: true
m2$converged
m2$iter
exp(confint.default(m2))
exp(confint(m2))
```



A declared linear combination $C=\mathbf{c}^{\mathsf T}\boldsymbol\beta$ has estimated
variance $\mathbf{c}^{\mathsf T}\widehat{\mathbf V}\mathbf c$. The off-diagonal covariance terms matter when
combining coefficients, including for a predicted logit or a conditional
slope. The familiar contrast rule remains; its covariance now comes from
likelihood curvature.

### A known process: certification success {#ch11-simulation}

Consider an invented certification course. Passing is the event; study
hours span 0 to 20, and tutoring is an indicator with probability .35,
independent of hours in this constructed process. Declare

$$
\operatorname{logit}\{P(\text{pass})\}=-2.2+.3\,\text{hours}+.6\,\text{tutor}.
$$

At seed 1010, 1,500 students give 961 passes. Fitting the declared model
returns $-2.2271$, .3024 and .5264, with all three generating values
inside their Wald intervals. This is recovery under a known process,
not evidence that an empirical study follows that process.



```{r}
#| eval: false
#| echo: true
set.seed(1010)
hours <- round(runif(1500, 0, 20), 1)
tutor <- rbinom(1500, 1, .35)
pass <- rbinom(1500, 1, plogis(-2.2 + .30 * hours + .60 * tutor))
glm(pass ~ hours + tutor, family = binomial)
```


| Study hours | Probability | Probability at one more hour | Probability difference | Odds ratio |
|--:|--:|--:|--:|--:|
| 2 | .1680 | .2142 | .0462 | 1.3499 |
| 7 | .4750 | .5498 | .0748 | 1.3499 |
| 12 | .8022 | .8455 | .0434 | 1.3499 |
| 17 | .9478 | .9608 | .0130 | 1.3499 |

: One constant odds ratio, four different probability differences, in the known generating process. {#tbl-ch11-sim-nonconstant tbl-colwidths="[18,20,26,22,14]"}



For students without tutoring, the declared one-hour odds ratio is
$e^{.3}=1.3499$ everywhere, yet the probability difference is .0748
at seven hours and .013 at seventeen. The starting probability makes
the practical benefit different even before sampling variation enters.
The estimated hours odds ratio, 1.3531, closely recovers that truth.

Repeating the declared generation at seed 1012 for 500 samples of 300
gives an hours coefficient averaging .3068 with SD .0359 and empirical
Wald coverage .948. The Monte Carlo SE of that coverage is .0099, so
this run is compatible with .95 coverage. Finite-sample logistic estimates
can be biased away from zero; simulation separates that tendency from
variation in one sample. Tutoring's coefficient varies much more here
(SD .3309): the two predictors have different scales and information,
so their coefficient SDs should not be read as a general comparison of
binary and continuous predictors.

The same generated data also let us compare mean functions honestly.
The fitted linear probability model gives .0492 per hour but produces
193 fitted values above 1. Its fitted range is .1253 to 1.1861.



```{r}
#| label: fig-ch11-sim-lpm
#| fig-cap: "Two mean functions fitted to the same invented certification records, for students without tutoring."
#| fig-alt: "A logistic curve levels off below probability one, while the fitted straight line crosses one before twenty study hours."
#| fig-height: 3.6
#| fig-width: 6.2
g <- data.frame(hours = seq(0, 20, length.out = 201), tutor = 0)
p <- rbind(data.frame(hours = g$hours, fit = predict(ms, g, type = "response"), model = "Logistic"),
  data.frame(hours = g$hours, fit = predict(ml, g), model = "Linear probability"))
ggplot(p, aes(hours, fit, linetype = model)) + geom_line(linewidth = .8) +
  geom_hline(yintercept = 1, linetype = 3) +
  labs(x = "Study hours", y = "Fitted probability", linetype = NULL) + theme(legend.position = "bottom")
```



This example deliberately uses a logistic truth and approaches the
boundaries. It shows why an unbounded line can fail; it does not show
that logistic regression must be right for every dataset. The volunteering
analysis below supplies the complementary case in which both mean
functions are numerically close over the observed range.

## Read Coefficients on Interpretable Scales {#ch11-reading}

### What an odds ratio compares {#ch11-odds-ratio}

If a predictor enters additively, increasing it by one unit while holding
the others fixed changes the log-odds by $\beta_j$:

$$
\operatorname{logit}\{\pi(x_j+1)\}-\operatorname{logit}\{\pi(x_j)\}=\beta_j,
\qquad
\frac{\pi(x_j+1)/\{1-\pi(x_j+1)\}}{\pi(x_j)/\{1-\pi(x_j)\}}=e^{\beta_j}.
$$

First fit extraversion alone. The coefficient is .0656 (SE .0141), giving
an odds ratio of 1.0678 and a 95% interval [1.0386, 1.0978]. Each extra
extraversion point is associated with 1.0678 times the odds of volunteering
rather than not volunteering. Five points compound multiplicatively to
1.3883, rather than adding five copies of the odds ratio.



```{r}
#| eval: false
#| echo: true
m1 <- glm(event ~ extraversion, data = dat, family = binomial)
exp(coef(m1))
exp(confint.default(m1))
```



Four qualifications make this a complete comparison: the predictor's
**units**, any factor's **reference category**, the **event**, and the
**adjustment set**. Changing a unit changes the contrast; changing the
event or reference changes its direction; adding controls changes the
conditional comparison. The intercept $-1.1394$ describes extraversion
zero, outside the observed 2--23 range, so it is a mathematical anchor
rather than a supported student profile.

An odds ratio is not a risk ratio. At a prevalence near .42, saying
"1.5 times the odds" as "50 percent more likely" can substantially
exaggerate a probability comparison. The approximation is useful only
when the event probabilities in both groups are small.

The local slope on the probability scale is

$$
\frac{\partial\pi}{\partial x_j}=\beta_j\pi(1-\pi).
$$

Its absolute value is bounded by $|\beta_j|/4$, attained at .5.
This also bounds the magnitude of a one-unit finite probability difference
when the logit is additive in that predictor. The derivative is a local
rate, not exactly the difference between two profiles one full unit apart;
compute both fitted probabilities when that finite contrast is wanted.

### Reverse the event, preserve the underlying probabilities {#ch11-reversal}

Before looking at new output, predict the result of coding non-volunteering
as 1. The probability must become $1-\widehat\pi_i$, the log-odds must
change sign, and the odds ratio must become its reciprocal. The simple
extraversion coefficient becomes $-.0656$ and its odds ratio .9365.
Their signed $z$ statistics reverse too. Standard errors, two-sided
p-values, deviance and McFadden's pseudo $R^2$ stay the same. After
translating back to the original event, every student still has the same
fitted probability of volunteering.



```{r}
#| eval: false
#| echo: true
dat$no <- 1 - dat$event
rev <- glm(no ~ extraversion, data = dat, family = binomial)
exp(coef(rev))
max(abs(fitted(rev) + fitted(m1) - 1))
```



Changing a predictor factor's reference instead changes coefficient
coordinates while retaining the same outcome event and fitted
probabilities. For a binary factor the focal comparison reverses, but
its intercept moves and other coefficient changes depend on the model's
terms. These recodings change what a printed coefficient names, not the
fitted probability surface.

### Adjusted results and probabilities at supported profiles {#ch11-multiple}

The multiple model is

$$
\log\frac{\pi_i}{1-\pi_i}=\beta_0+\beta_1\,\mathrm{male}_i+\beta_2N_i+\beta_3E_i,
$$

where $N$ and $E$ are the recorded neuroticism and extraversion scores.
Female is the reference sex category. The fitted table attaches each
coefficient to that event, coding and adjustment set.


| Term | Estimate (log-odds) | SE | $z$ | Odds ratio | OR 95% lower | OR 95% upper |
|:--|--:|--:|--:|--:|--:|--:|
| Intercept | $-1.1165$ | .2491 | $-4.4829$ | .3274 | .2010 | .5335 |
| Sex: male versus female | $-.2352$ | .1112 | $-2.1151$ | .7904 | .6357 | .9829 |
| Neuroticism | .0064 | .0114 | .5602 | 1.0064 | .9842 | 1.0290 |
| Extraversion | .0663 | .0143 | 4.6512 | 1.0686 | 1.0391 | 1.0989 |

: Multiple logistic regression of volunteering on sex, neuroticism, and extraversion; Wald 95% intervals, female as the reference category, event coded as volunteering. {#tbl-ch11-fit2 tbl-colwidths="[28,16,10,12,12,11,11]"}



Among students of the same recorded sex and neuroticism, an extra
extraversion point is associated with an odds ratio of 1.0686, close to
the unadjusted 1.0678. Male students have .7904 times the odds of female
students of the same personality scores. Neuroticism has no detected
adjusted association in this additive model ($p=.5754$), but its interval
admits odds ratios from .9842 to 1.0290 per point; a large p-value does
not prove no relationship.

Choose profiles before interpreting their estimates: a female student at
neuroticism 11, its median, and at extraversion 10, 13 or 15, its quartiles.
These values sit within the observed support. Prediction first gives the
logit and its SE; transform the two confidence limits to probability.



```{r}
#| eval: false
#| echo: true
grid <- data.frame(sex = "female", neuroticism = 11,
  extraversion = c(10, 13, 15))
pred <- predict(m2, grid, type = "link", se.fit = TRUE)
plogis(cbind(pred$fit - 1.96 * pred$se.fit,
  pred$fit, pred$fit + 1.96 * pred$se.fit))
```


| Extraversion | Fitted log-odds | SE (log-odds) | Fitted probability | 95% lower | 95% upper |
|--:|--:|--:|--:|--:|--:|
| 10 | $-.3833$ | .0830 | .4053 | .3668 | .4451 |
| 13 | $-.1843$ | .0742 | .4541 | .4183 | .4903 |
| 15 | $-.0516$ | .0813 | .4871 | .4474 | .5269 |

: Fitted probability of volunteering for a female student at the median neuroticism score, at the quartiles of extraversion, with intervals built on the log-odds scale and transformed. {#tbl-ch11-predictions tbl-colwidths="[16,18,18,20,14,14]"}



The probability rises from .4053 to .4871 across the middle half of
extraversion, a difference of .0818. Its interval is asymmetric because
it is transformed from the logit scale. This is uncertainty about a
conditional probability, not a prediction interval for an individual
student: the next student's outcome remains either 0 or 1. Comparing two
separate intervals is also not a test of their difference; that contrast
uses their joint covariance.

### Profile and average probability comparisons {#ch11-average}

A **case-specific comparison** fixes the other predictor values and
compares fitted probabilities at two profiles. An **averaged comparison**
computes the contrast for each recorded student's other predictor values
and then averages it. This makes the averaging population explicit rather
than treating one artificial "average student" as representative.







Averaging is a substantive choice even when the model is already fixed.
Consider the invented model $p(x,u)=\operatorname{logit}^{-1}(x+u)$ and
compare $u=1$ with $u=0$. One population has equal numbers at $x=-.5$
and $.5$; another has equal numbers at $x=-4$ and 4. Both have mean zero.
Predict which population has the larger average probability difference
from the vertical gaps in @fig-ch11-averaging.

```{r}
#| label: fig-ch11-averaging
#| fig-pos: "H"
#| fig-cap: "The same probability model gives different average differences in two populations."
#| fig-alt: "Both panels show the same pair of logistic curves. Equal-weight profiles near zero have average difference .22; profiles at minus four and plus four have average .0204. Both sets have mean zero, where the difference is .2311."
#| fig-height: 3.6
#| fig-width: 6.2
g <- expand.grid(x = seq(-5, 5, length.out = 301), u = 0:1,
  population = c("Central profiles", "Tail profiles"))
g$p <- plogis(g$x + g$u)
loc <- data.frame(x = c(-.5, .5, -4, 4),
  population = rep(c("Central profiles", "Tail profiles"), each = 2))
loc$lo <- plogis(loc$x); loc$hi <- plogis(loc$x + 1)
ggplot(g, aes(x, p, linetype = factor(u))) + geom_line(linewidth = .7) +
  geom_segment(data = loc, aes(x = x, xend = x, y = lo, yend = hi),
    inherit.aes = FALSE, linewidth = 1.2) +
  geom_rug(data = loc, aes(x = x), inherit.aes = FALSE, sides = "b") +
  facet_wrap(~population) +
  labs(x = "Other predictor x", y = "Event probability", linetype = "u") +
  theme(legend.position = "bottom")
```

The central profiles give average difference .22 and the tail profiles
give .0204. In contrast, evaluating both at the common mean $x=0$ gives
.2311. The coefficient of $u$ and its conditional odds ratio $e$ are
identical throughout. The average changes because the populations place
different weight on steep and flat portions of the same probability
curves. A profile at the average predictor is therefore not interchangeable
with an average over the actual profiles. These are model comparisons,
not causal effects supplied by the invented equation.

```{r}
#| eval: false
#| echo: true
x <- c(-.5, .5)
mean(plogis(x + 1) - plogis(x))
x <- c(-4, 4)
mean(plogis(x + 1) - plogis(x))
plogis(1) - plogis(0)
```

Return now to the recorded volunteering profiles.

```{r}
#| eval: false
#| echo: true
plus <- dat
plus$extraversion <- plus$extraversion + 1
mean(predict(m2, plus, type = "response") - fitted(m2))
```

The average one-point extraversion difference is .0159. The local
maximum bound is .0166, and an additive linear probability fit gives
.0158. Here its fitted values remain within [.2261, .6188], near the
middle of the logistic curve, so the two mean functions agree closely.
The familiar least-squares constant-variance uncertainty still needs
correction, and a new predictor range can expose the line's boundaries.
The one-point standardization also moves the single student at score 23
to 24, just outside the recorded range; that small extrapolation is part
of this explicitly defined average, not evidence about students at 24.

Assigning each observed profile first the female and then the male
indicator gives an averaged male-minus-female probability difference of
$-.0562$. This is about 5.6 percentage points, whereas the odds ratio
.7904 describes odds about 21 percent lower. The distinction changes
what a reader imagines about actual volunteering rates.



```{r}
#| label: fig-ch11-fitted
#| fig-cap: "Adjusted volunteering probabilities at neuroticism 11, with pointwise Wald 95% bands and observed score support."
#| fig-alt: "Two gently rising curves overlap in uncertainty bands, with the female curve higher. Bands widen at sparse extremes; the rug is dense in the middle of the extraversion range."
#| fig-height: 3.6
#| fig-width: 6.2
g <- expand.grid(extraversion = seq(2, 23, length.out = 201), sex = c("female", "male"))
g$neuroticism <- 11
p <- predict(m2, g, type = "link", se.fit = TRUE)
g$p <- plogis(p$fit); g$lo <- plogis(p$fit - 1.96 * p$se.fit); g$hi <- plogis(p$fit + 1.96 * p$se.fit)
ggplot(g, aes(extraversion, p)) + geom_ribbon(aes(ymin = lo, ymax = hi, group = sex), alpha = .16) +
  geom_line(aes(linetype = sex), linewidth = .8) +
  geom_rug(data = dat, aes(extraversion, y = NULL), sides = "b", alpha = .12) +
  labs(x = "Extraversion score", y = "Fitted probability of volunteering", linetype = "Recorded sex") +
  theme(legend.position = "bottom")
```



The two lines are parallel on the logit scale because the additive model
contains no product term. Their probability separation varies. Additivity
belongs to a named scale, a distinction that matters when we next allow
conditional contrasts to differ.

## Compare and Test {#ch11-compare}

### Deviance and a declared model comparison {#ch11-deviance}

The **deviance** compares a fitted model with the saturated model that
fits every observation exactly. For individual binary observations the
saturated log-likelihood is zero, so

$$
D=-2\ell(\widehat{\boldsymbol\beta}),\qquad
G^2=D_A-D_B\ \dot\sim\ \chi^2_q
\quad\text{for regular nested models }A\subset B.
$$

Here $q$ counts the restrictions defining the smaller model. The
comparison requires the same rows and event, finite well-identified
estimates, and an adequate large-sample approximation. The absolute
ungrouped binary deviance has no simple chi-square goodness-of-fit reading;
its difference can test a declared nested comparison.



```{r}
#| eval: false
#| echo: true
anova(m0, m2, test = "LRT")
small <- glm(event ~ sex + extraversion, data = dat, family = binomial)
anova(small, m2, test = "LRT")
```


| Comparison | Added | df | $G^2$ | $p$ |
|:--|:--|--:|--:|--:|
| Intercept only versus full | Sex, neuroticism, extraversion | 3 | 27.4447 | $<.0001$ |
| Without neuroticism versus full | Neuroticism | 1 | .3139 | .5753 |

: Two declared likelihood-ratio comparisons on the same 1,421 rows and the same event. {#tbl-ch11-lrt tbl-colwidths="[36,30,10,12,12]"}



The full block reduces deviance from 1933.5060 to 1906.0613. Its statistic
27.4447 exceeds the 5% critical value 7.8147 on three df. This establishes
information in the block under the model, not a separate association for
each member. Removing neuroticism alone costs only .3139 on one df,
with $p=.5753$.

### Wald tests and likelihood comparisons {#ch11-wald-lrt}

A single-coefficient Wald test uses $z=\widehat\beta/SE$, or $z^2$
with a chi-square reference on one df. The likelihood-ratio test instead
compares the best fits allowed by two models. For neuroticism the two
agree: $z=.5602$, $z^2=.3138$, $p=.5754$, against $G^2=.3139$,
$p=.5753$. Close agreement is expected when the likelihood is nearly
quadratic near its maximum. With strong asymmetry or unstable estimates,
the Wald approximation can be misleading. Separation below shows an
extreme case in which its statistic shrinks despite a perfect pattern.

### Criteria rank models; they do not certify them {#ch11-criteria}

For $k$ estimated coefficients and $n$ records,

$$
AIC=-2\ell+2k,\qquad BIC=-2\ell+k\log n,\qquad
R^2_{\mathrm{McF}}=1-\frac{\ell}{\ell_0}.
$$

There is no separately estimated variance parameter in the logistic
model. AIC and BIC penalize in-sample likelihood by model size; smaller
values are preferred among models fitted to the same records and outcome
[@akaike1974; @schwarz1978]. McFadden's pseudo $R^2$ compares with the
intercept-only maximum $\ell_0$. It is not a proportion of variance
explained and should not be read against linear-model $R^2$ benchmarks.



```{r}
#| eval: false
#| echo: true
AIC(m0, m1, m2)
BIC(m0, m1, m2)
1 - as.numeric(logLik(m2) / logLik(m0))
```


| Model | $k$ | Deviance | AIC | BIC | McFadden pseudo $R^2$ |
|:--|--:|--:|--:|--:|--:|
| Intercept only | 1 | 1933.5060 | 1935.5060 | 1940.7651 | .0000 |
| Extraversion | 2 | 1911.4838 | 1915.4838 | 1926.0020 | .0114 |
| Sex, neuroticism, extraversion | 4 | 1906.0613 | 1914.0613 | 1935.0977 | .0142 |

: Model comparison for the volunteering analysis; both criteria are comparative and neither certifies adequacy. {#tbl-ch11-criteria tbl-colwidths="[34,8,16,14,14,14]"}



AIC prefers the adjusted model; BIC prefers extraversion alone. At
$n=1421$, BIC charges $\log n=7.2591$ per coefficient against AIC's
2, explaining the difference. Neither criterion is a test or a held-out
prediction score. The small McFadden value .0142 is compatible with a
detectable association and little ability to distinguish volunteers.

### A product term changes a conditional contrast {#ch11-product}

Allow the extraversion contrast to vary with neuroticism:

$$
\operatorname{logit}(\pi_i)=\beta_0+\beta_1\mathrm{male}_i+\beta_2N_i+
\beta_3E_i+\beta_4N_iE_i,\qquad
\frac{\partial\operatorname{logit}(\pi)}{\partial E}=\beta_3+\beta_4N.
$$

The product coefficient is $-.0086$ (SE .0029). The nested likelihood
comparison gives $G^2(1)=8.6213$, $p=.0033$. The lower-order extraversion
coefficient .1668 is its contrast at neuroticism zero, not a summary
for every student. The additive model's undetected neuroticism term and
the detected product answer different questions, so hierarchical model
construction retains both lower-order terms.



```{r}
#| eval: false
#| echo: true
m3 <- glm(event ~ sex + neuroticism * extraversion,
  data = dat, family = binomial)
anova(m2, m3, test = "LRT")
```


| Neuroticism | Log-odds contrast | SE | Odds ratio | OR 95% lower | OR 95% upper | Students within one point |
|--:|--:|--:|--:|--:|--:|--:|
| 8 | .0984 | .0184 | 1.1034 | 1.0644 | 1.1438 | 276 |
| 11 | .0727 | .0146 | 1.0755 | 1.0451 | 1.1067 | 302 |
| 15 | .0385 | .0170 | 1.0393 | 1.0052 | 1.0746 | 242 |

: Conditional extraversion odds ratios at the quartiles of neuroticism, each comparing students differing by one extraversion point at the stated neuroticism score. {#tbl-ch11-conditional tbl-colwidths="[13,18,11,14,15,15,14]"}


```{r}
#| label: fig-ch11-interaction
#| fig-cap: "The same product-term fit on log-odds and probability scales, for female students."
#| fig-alt: "Three straight lines on the logit scale cross near extraversion thirteen and are steepest at neuroticism eight. Their probability curves have the same crossing but unequal changes across the range."
#| fig-height: 3.6
#| fig-width: 6.2
g <- expand.grid(extraversion = seq(2, 23, length.out = 201), neuroticism = c(8, 11, 15)); g$sex <- "female"
g$logit <- predict(m3, g); g$prob <- plogis(g$logit)
p <- rbind(data.frame(g[, 1:2], value = g$logit, scale = "Log-odds"),
  data.frame(g[, 1:2], value = g$prob, scale = "Probability"))
ggplot(p, aes(extraversion, value, linetype = factor(neuroticism))) + geom_line(linewidth = .8) +
  facet_wrap(~scale, scales = "free_y") + labs(x = "Extraversion score", y = "Fitted value", linetype = "Neuroticism") +
  theme(legend.position = "bottom")
```



The table evaluates $\beta_3+\beta_4N$ at supported quartiles, with
SEs from $\mathbf{c}^{\mathsf T}\widehat{\mathbf V}\mathbf c$. At extraversion 13, the probabilities for
neuroticism 8 and 15 are nearly equal (.4519 and .4512), but a one-point
extraversion difference is .0245 at the former and .0096 at the latter.
Conversely, different probability differences can arise with a constant
odds ratio simply because baseline probabilities differ. An interaction
claim therefore names its scale. Chapter 10 develops centering and
probing; the new issue here is the nonlinear transfer between scales.

## Check Calibration, Fit, and Influence {#ch11-checks}

### Form and observed support {#ch11-form}

An observed proportion at each extraversion score makes both the pattern
and its support visible. An **empirical logit**
$\widetilde\eta=\log\{(y+.5)/(n-y+.5)\}$ adds half a count to keep
groups with zero or all events finite. It is a plotting device rather
than a replacement for the fitted Bernoulli likelihood.



```{r}
#| eval: false
#| echo: true
aggregate(event ~ extraversion, data = dat, mean)
anova(m2, mq, test = "LRT")
```



```{r}
#| label: fig-ch11-proportions
#| fig-cap: "Observed volunteering by score: empirical proportions and logits, with point area showing the number of students."
#| fig-alt: "Most students lie in the middle scores, where proportions and empirical logits rise gently. Score twenty-two contains five volunteers and score twenty-three one non-volunteer; their extreme proportions have very little support."
#| fig-height: 3.6
#| fig-width: 6.2
e <- aggregate(event ~ extraversion, dat, length); names(e)[2] <- "n"
e$yes <- aggregate(event ~ extraversion, dat, sum)$event
e$p <- e$yes / e$n; e$logit <- log((e$yes + .5) / (e$n - e$yes + .5))
p <- rbind(data.frame(e, value = e$p, scale = "Proportion"), data.frame(e, value = e$logit, scale = "Empirical logit"))
ggplot(p, aes(extraversion, value)) + geom_point(aes(size = n), alpha = .7) + scale_size_area(max_size = 6) +
  facet_wrap(~scale, scales = "free_y") + labs(x = "Extraversion score", y = NULL, size = "Students") + theme(legend.position = "bottom")
```



Score 13 has 151 students. Score 22 has five, all volunteers; score 23
has one non-volunteer. Their extreme empirical logits, 2.3979 and
$-1.0986$, deserve less weight than the densely populated middle.
The observed proportions need not increase at every score: sampling
variation around an increasing conditional mean remains.

The grouped pattern gives no compelling curvature, and adding squared
extraversion gives $G^2(1)=1.4257$, $p=.2325$. This does not prove
linearity in the logit; it says this particular alternative is not clearly
distinguished by the available records. A one-predictor empirical-logit
display is also a marginal diagnostic, not a complete check of the
adjusted model's functional form.

### Residuals and influence {#ch11-residuals}

Raw residuals are $y_i-\widehat\pi_i$. At a fitted probability they
take only $1-\widehat\pi_i$ or $-\widehat\pi_i$, so the raw plot has
two lines rather than a Normal cloud. Pearson residuals divide by
$\sqrt{\widehat\pi_i(1-\widehat\pi_i)}$; signed deviance residuals
have squares summing to deviance. None is expected to be Normal.

A **binned residual plot** instead groups nearby fitted probabilities
and compares their average residuals with zero. The displayed pointwise
bands, $\pm2\sqrt{\bar\pi(1-\bar\pi)/n_{\mathrm{bin}}}$, are a rough
visual reference for groups with similar fitted probabilities; they are
not a simultaneous test and ignore parameter-estimation effects.



```{r}
#| label: fig-ch11-binned
#| fig-cap: "Mean residuals in twenty fitted-probability bins, with approximate pointwise two-SE reference bands."
#| fig-alt: "Twenty mean residuals fluctuate on both sides of zero without an evident trend. One lies outside the approximate two-standard-error bands."
#| fig-height: 3.6
#| fig-width: 6.2
ggplot(bins, aes(fit, resid)) + geom_hline(yintercept = 0) +
  geom_line(aes(y = 2 * se), linetype = 2) + geom_line(aes(y = -2 * se), linetype = 2) +
  geom_point(size = 2) + labs(x = "Mean fitted probability", y = "Mean residual")
```



One bin falls outside its reference band. That is compatible with chance
variation and does not certify the form. The largest Cook's distance is
.0051. Removing that row moves the extraversion coefficient from .0663
to .0679, with a maximum coefficient change of .0304 elsewhere. A deletion
refit asks whether a conclusion depends on one observation; it supplies
no automatic reason to discard it. The principal analysis retains the row.



```{r}
#| eval: false
#| echo: true
which.max(cooks.distance(m2))
coef(m2)
coef(md)
```



### Separation: a strong pattern with no finite estimate {#ch11-separation}

A sparse category supplies little information. **Complete separation**
is more specific: a predictor combination places every event on one side
and every non-event on the other. With **quasi-complete separation**, some
observations lie on the separating boundary. In these circumstances one
or more coefficients can increase in magnitude without reaching a finite
likelihood maximum [@albert1984]. A number printed by software can be
merely the point at which iteration stopped.

Make this failure visible with a deliberately constructed subset, not a
substantive analysis. Retain non-volunteers scoring at most 8 and
volunteers scoring at least 17, then sample 60 eligible rows at seed
1011. There is a gap between the two outcomes by construction.



```{r}
#| eval: false
#| echo: true
glm(event ~ extraversion, data = sep, family = binomial)
```



| Output feature | Constructed separated fit | What it signals here |
|:--|--:|:--|
| Extraversion coefficient | 5.5928 | An iteration moving toward an infinite limit |
| Standard error | 8014.9006 | Nearly flat local information in the escaping direction |
| Wald $z$ and two-sided $p$ | .0007; .9994 | A failed Wald approximation, not absence of the observed pattern |
| Residual deviance | $5.642\times10^{-10}$ | Almost perfect fit on the selected rows |
| Fitted probabilities | about $2.22\times10^{-16}$ and 1 | Numerical boundary values |
| Iterations | 25 | Far more than the four needed by the regular adjusted fit |

: Symptoms of this complete-separation example, read together. {#tbl-ch11-separation tbl-colwidths="[26,25,49]"}

The selected rows reproduce the separated pattern in
@fig-ch11-likelihood-shapes: increasing the slope can approach perfect
classification without a finite likelihood peak.

The tiny Wald statistic is particularly instructive: it does not negate
the perfect ordering. The coefficient and SE have both become enormous,
and the usual local approximation is unusable. These are characteristic
symptoms in this example, not a checklist every separated model must
satisfy. Quasi-complete separation need not make the whole deviance zero,
and a warning is neither necessary nor sufficient for diagnosis.

The full data offer an unwarned example. If scores 22 and 23 are separate
factor levels, one contains five events and the other one non-event.
Their coefficients become 15.2161 (SE 394.7749) and $-13.916$
(SE 882.7434). Combining the substantively adjacent upper bands as
18+ gives 130 students with 71 events and a finite coefficient .8352
(SE .2222). That demonstrates what the sparse bands were doing; such a
merge still requires a meaningful category definition, not merely a
wish for smaller standard errors.


| Option | What it does | When it is appropriate |
|:--|:--|:--|
| Collapse categories | Merges a sparse level with a substantively adjacent one | The merged levels are defensibly similar on subject-matter grounds |
| Drop the term | Removes the separating predictor | The predictor is not central to the question and its removal is reported |
| Report the boundary honestly | States the direction, notes that the magnitude is not identified, and reports any unbounded profile limit | The separating pattern is itself the finding |
| Penalized likelihood | Adds a penalty giving finite estimates with reduced small-sample bias, as in Firth's method | A finite estimate is needed and the penalty is disclosed |
| Collect more data | Fills the empty cell | The design permits it |

: Options when a logistic model shows separation. {#tbl-ch11-separation-options tbl-colwidths="[24,40,36]"}



Firth's penalized likelihood is one way to obtain finite estimates with
reduced small-sample bias [@firth1993]. It is a different estimation rule,
whose penalty must be stated. It does not make the original infinite
maximum-likelihood estimate finite. More generally, a finite estimate
should be interpreted only where the data support the intended comparison.

### Calibration and discrimination {#ch11-calibration}

**Calibration** asks whether assigned probabilities agree with observed
event frequencies: among students assigned about .4, do roughly 40%
volunteer? **Discrimination** asks whether probabilities tend to rank
volunteers above non-volunteers. They answer distinct questions. A
constant fitted prevalence has perfect aggregate agreement but ranks
nobody; a rule can rank well and systematically overstate probabilities.



```{r}
#| label: fig-ch11-calibration
#| fig-cap: "In-sample calibration by fitted-probability decile; vertical segments show one observed-proportion SE."
#| fig-alt: "Nine of ten groups lie reasonably near the equality line. The fifth decile is assigned about .413 but has an observed proportion about .297."
#| fig-height: 3.6
#| fig-width: 6.2
ggplot(cal, aes(fit, obs)) + geom_abline(intercept = 0, slope = 1, linetype = 2) +
  geom_linerange(aes(ymin = obs - se, ymax = obs + se)) + geom_point(size = 2) +
  coord_equal(xlim = c(.25, .60), ylim = c(.25, .60)) +
  labs(x = "Mean fitted probability", y = "Observed proportion volunteering")
```



The fifth decile is assigned .4126 but has observed proportion .2966,
a gap of .1161 or 3.0596 observed-proportion SEs. This deserves
investigation rather than dismissal. Because the group was noticed among
ten and the model was fitted to these same rows, this descriptive SE
comparison is not a calibrated global test of fit. Aggregate agreement
from the intercept says nothing about agreement within each decile.



```{r}
#| label: fig-ch11-discrimination
#| fig-cap: "Fitted probabilities overlap strongly between volunteers and non-volunteers."
#| fig-alt: "Two boxplots overlap almost entirely across fitted probabilities from about .24 to .62, showing weak descriptive separation."
#| fig-height: 3.1
#| fig-width: 6.2
ggplot(data.frame(p = phat, outcome = factor(dat$event, labels = c("Non-volunteer", "Volunteer"))), aes(outcome, p)) +
  geom_boxplot(width = .45) + labs(x = NULL, y = "Fitted probability of volunteering")
```



Fitted probabilities average .4313 among volunteers and .412 among the
rest, a difference of .0192, and span only [.2418, .6235]. The boxes
overlap strongly. Extraversion is detectably associated with volunteering,
yet the fitted probabilities separate these students poorly. A small
coefficient p-value is not evidence of useful prediction [@shmueli2010].
Both displays reuse the fitting data and are optimistic as performance
assessments. Chapter 13 develops held-out evaluation, including AUC,
Brier score, log loss and threshold choice.

{{< pagebreak >}}

### What changed under sensitivity analysis {#ch11-sensitivity}


| Specification | Extraversion odds ratio per point |
|:--|--:|
| Simple model, extraversion only | 1.0678 |
| Additive model, sex and neuroticism added | 1.0686 |
| Additive model, most influential row removed | 1.0703 |
| Additive model plus quadratic extraversion, local at extraversion 13 | 1.0745 |
| Product-term model, at median neuroticism | 1.0755 |

: Sensitivity of the extraversion odds ratio to modeling and sample choices; each row refits the model with one feature changed. {#tbl-ch11-sensitivity tbl-colwidths="[70,30]"}



The quadratic entry is a local derivative-based odds ratio at score 13;
it is not exactly the finite 13-to-14 contrast. The product entry fixes
neuroticism at its median. These definitions keep the compared quantities
clear. The values remain close, but every specification uses the same
underlying study, measurements and design. Stability across them cannot
establish accurate volunteering records, valid personality measurement,
independent classrooms or absence of unmeasured common causes.

## From Analysis to a Defensible Claim {#ch11-claim}

### Report the magnitude and the limit together {#ch11-report}

The event and analytic unit come before any ratio. The model, predictor
units and reference categories make the comparison reproducible. A
probability-scale result then says what the ratio means in this study,
while checks and uncertainty qualify the conclusion.

::: {.callout-tip title="A bounded reporting example"}
Among 1,421 recorded introductory psychology students, the volunteering
proportion was .4201. Adjusting for recorded sex and neuroticism, each
extraversion point was associated with 1.0686 times the odds of
volunteering rather than not volunteering (95% CI [1.0391, 1.0989]). The
averaged one-point probability difference was .0159 over the observed
profiles. For a female student at neuroticism 11, fitted probabilities
were .4053 [.3668, .4451] at extraversion 10 and .4871 [.4474, .5269]
at 15. The additive neuroticism association was not detected, although a
neuroticism-by-extraversion product improved fit and showed a weaker
extraversion contrast at higher neuroticism. A quadratic check and a
single-row deletion gave similar local comparisons. Sparse extreme scores
limited their own interpretation; one calibration decile showed a material
departure, and fitted probabilities overlapped strongly by outcome. These
are conditional associations among recorded students. They establish
neither a causal effect of personality nor useful prediction in a new cohort.
:::

The report can contain a null comparison, a detected interaction and weak
discrimination together. They address different aspects of the research
question. Describing them faithfully is more useful than converting a
coefficient table into a general claim that the model "worked".

### Recognize the claim a number supports {#ch11-attractive}

| Tempting sentence | What the evidence supports |
|:--|:--|
| "Volunteering is about 7% more likely per extraversion point." | Odds are 1.0686 times as large; the averaged probability difference is .0159. |
| "Neuroticism has no relationship with volunteering." | The additive adjusted interval admits small associations in both directions; the product term asks another question. |
| "The product coefficient is the extraversion effect." | Evaluate the conditional contrast at a stated neuroticism value and name its scale. |
| "The small p-value means the model predicts well." | A coefficient test, calibration check and held-out performance assessment answer different questions. |
| "The separated fit found a very large, imprecise effect." | Ordinary maximum likelihood has no finite estimate for that separated magnitude. |

: Interpretation depends on the quantity and its conditions. {#tbl-ch11-attractive tbl-colwidths="[43,57]"}

A reusable report makes seven elements recoverable: event and unit;
model and coding; a coefficient-scale estimate with uncertainty; a
probability-scale comparison; evidence for a declared term or block;
adequacy and sensitivity; and the claim's boundary. They need not appear
as seven ritual sentences. Their function is to let a reader judge what
was estimated and why its interpretation is warranted.

The empirical analyses use all Cowles records, except the explicitly
labelled deletion and constructed separation checks. The certification
sample uses seed 1010, its repetition uses 1012, and the separated subset
uses 1011. More than two categories, event counts, dependence, severe
measurement error and sparse events each require a reconsidered model
or inferential calculation, not a stronger reading of this coefficient
table.

## Reading Bridge

A logistic model connects a Bernoulli event to a linear predictor through
the logit link. Maximum likelihood supplies estimates and uncertainty;
probabilities give the substantive magnitude. Event coding, support,
separation and the distinction between calibration and discrimination
remain part of interpreting every result. Chapter 12 extends the same
framework to nominal categories, ordered categories and counts, each with
a new response structure and a new meaning for an exponentiated coefficient.

## Key terms {.unnumbered}

This reference collects the concepts introduced here. The Part One Concept
Reference locates the broader ideas, including likelihood, coefficient
covariance and model comparison.


| Term and common alternatives | Meaning or essential distinction |
| :-- | :-- |
| Event and non-event (the category coded 1) | The category a binary model treats as occurring, named together with the category coded 0. Reversing the event reciprocates the odds ratios and complements fitted event probabilities; probabilities for the original event agree after relabelling. |
| Prevalence (event rate) | The proportion of analytic rows that are events. Coefficient precision also depends on conditional probabilities, predictor spread and overlap; the rare-outcome approximation of an odds ratio by a risk ratio depends on the risks being compared. |
| Generalized linear model (GLM) | A response distribution, a linear predictor and a link connecting the conditional mean to the linear predictor; changing the response distribution and link gives the models of Chapters 11 and 12. |
| Link function; logit link and inverse logit (logistic function) | The logit is the log of the odds and its inverse the S-shaped function bounded in $(0,1)$, equal to one half and steepest at a linear predictor of zero. |
| Odds and log-odds | The odds are $\pi/(1-\pi)$ and the log-odds their logarithm; the three scales describe one quantity with null values 1 and 0, and the log-odds scale is symmetric under reversing the event. |
| Odds ratio (OR) | The factor $e^{\beta_j}$ by which the odds differ between units differing by one unit in a predictor and agreeing on the others; it depends on the units, reference, event and adjustment set, is not a risk ratio unless the event is rare, and implies no constant probability difference. |
| Noncollapsibility (of the odds ratio) | A conditional odds ratio generally differs from the marginal one even with no confounding, so a coefficient that moves under adjustment is not by itself evidence of confounding. |
| Bernoulli likelihood and score equations (binary-outcome likelihood) | The probability of the observed data as a function of the parameters, maximized to estimate them; for a binary outcome it is a product of Bernoulli probabilities whose score equations have no closed form, and a model with an intercept reproduces the observed event total. |
| Iteratively reweighted least squares (IRLS) | The algorithm that repeatedly solves a weighted linear regression with weights $\widehat\pi_i(1-\widehat\pi_i)$ until the coefficients stop changing; its convergence flag and iteration count belong in the record and establish no adequacy. |
| Deviance and the likelihood-ratio test ($G^2$) | For ungrouped binary data the deviance is $-2$ times the maximized log-likelihood and carries no goodness-of-fit reading alone, while the difference between nested models on the same rows and event is referred to chi-square. |
| Wald test and profile-likelihood interval | The Wald statistic squares the estimate-to-SE ratio while a profile interval collects the values the log-likelihood does not reject; they diverge for rare events, small samples or near-infinite estimates. |
| McFadden pseudo $R^2$ | One minus the ratio of the fitted to the intercept-only maximized log-likelihood: bounded in $[0,1]$, not a proportion of variance explained, and routinely small for useful models. |
| Empirical logit | The log-odds of a grouped event count with one half added to each count, plotted against a predictor to check linearity of the log-odds in that predictor as entered. |
| Binned residual | The mean raw residual within a bin of rows sorted by fitted probability, against a two-standard-error band; it replaces the residual-versus-fitted plot, which shows only two curved bands here. |
| Separation (complete; quasi-complete) | A predictor pattern separates events from non-events completely or on a boundary, so at least one coefficient has no finite ordinary maximum-likelihood estimate. Diverging coefficients, large standard errors and convergence warnings can be symptoms; no one warning or numerical pattern is required. |
| In-sample calibration and discrimination checks (fitted-probability checks) | Calibration asks whether fitted probabilities match observed event frequencies, discrimination whether they rank events above non-events; the two answer distinct questions, and honest evaluation of new-case performance requires held-out data. |
| Linear probability model (LPM) | Ordinary least squares on a 0/1 outcome, whose coefficients are probability differences but whose mean function is unbounded and whose conditional variance cannot be constant. |
| Averaged probability difference (average marginal comparison) | The mean, over the analytic rows with their own other predictor values, of the difference between fitted probabilities at two values of one predictor. |

: Concepts introduced in this chapter. {#tbl-ch11-key-terms tbl-colwidths="[30,70]"}



## Further reading {.unnumbered}

Agresti [@agresti2018] and Hosmer, Lemeshow and Sturdivant [@hosmer2013]
develop logistic models and diagnostics. Gelman, Hill and Vehtari
[@gelman2020] emphasize probability-scale interpretation. The R
`glm()`, `family()` and `predict.glm()` documentation explains the
implementation [@rCoreLogistic; @rcore2025].

## Exercises {.unnumbered}


<!-- prompt-id: CH11-E01 -->

### Exercise 1: Read a logistic result on three scales (Foundational). {.unnumbered}

A
logistic regression of whether a student volunteered for an experiment
on an extraversion score in integer points, adjusting for recorded sex
and a neuroticism score, gives an extraversion coefficient of .0663 with
standard error .0143. The event is volunteering, the reference sex
category is female, extraversion runs from 2 to 23, and the prevalence
is .4201. (1) Give the odds ratio and its Wald 95% interval, and write
one sentence interpreting them with the units, the event and the
adjustment set explicit. (2) Give the odds ratio for a five-point
difference. (3) Give the divide-by-four bound on the probability
difference per point, compute the local rate of change at a fitted
probability of .12 from $\partial\pi/\partial x=\beta\,\pi(1-\pi)$, and
explain why it must be smaller than the bound. (4) Name three
conclusions the interval excluding 1 does not establish.

<!-- END exercise-id: CH11-E01 -->


<!-- prompt-id: CH11-E02 -->

### Exercise 2: Declare the event, then reverse it (Foundational). {.unnumbered}

A
researcher codes program *completion* as the event and reports an odds
ratio of .62 per additional hour of paid weekly employment, with a 95%
interval of [.51, .75]. (1) State in one sentence what this odds ratio
compares, including the event and the units. (2) Give the odds ratio and
interval the same fitted model would report with *non-completion* as the
event, showing the arithmetic. (3) State which of these change under the
recoding and which do not: the coefficient signs, the standard errors,
the $z$ statistics, the p-values, the deviance, the fitted probability
of completion for a given student, and McFadden's pseudo $R^2$. (4) The
researcher also relevels the employment-status factor so that "employed"
is the reference, and a reviewer complains that the results changed.
Explain in what sense they did and did not change.

<!-- END exercise-id: CH11-E02 -->


<!-- prompt-id: CH11-E03 -->

### Exercise 3: Odds ratios are not probability differences (Foundational). {.unnumbered}

Two studies report the same adjusted odds ratio of
2. In Study A the comparison group's event probability is .04; in
Study B it is .45. (1) For each study compute the exposed group's event
probability, the probability difference and the risk ratio. (2) Explain
why the odds ratio and the risk ratio nearly coincide in one study and
diverge in the other. (3) A policy brief states that "an odds ratio of
2 means the risk is doubled" for both. Say where this is approximately
acceptable and where it is not, and quantify the error. (4) Explain why
noncollapsibility means an adjusted odds ratio can differ from an
unadjusted one even when the added predictor is unrelated to the focal
predictor. (5) Recommend, with reasons, which scale you would lead with
in a report for school administrators, and which you would also supply.
(6) In the invented model $p(x,u)=\operatorname{logit}^{-1}(x+u)$,
compare $u=1$ with $u=0$. Population C has equal numbers at $x=-.5$ and
$.5$; Population D has equal numbers at $x=-4$ and 4. Before
calculating, predict which has the larger average probability
difference. Explain why their common mean $x=0$ and common conditional
odds ratio do not make these averages equal.

<!-- END exercise-id: CH11-E03 -->


<!-- prompt-id: CH11-E04 -->

### Exercise 4: Deviance, Wald, and the criteria by hand (Foundational). {.unnumbered}

On 1,421 rows with the same event, an intercept-only logistic model has
deviance 1933.506 and a model with sex, neuroticism and extraversion
has deviance 1906.0613 with four fitted coefficients; the 5% critical
value of $\chi^2_3$ is 7.8147 and $\log 1421=7.259116$. The neuroticism
coefficient is .0064 with standard error .0114. (1) Calculate the
likelihood-ratio statistic for the three-predictor block, give its df,
and state the decision at 5% and what it does and does not say about the
three added columns. (2) Calculate AIC and BIC for the fuller model and
say which model each criterion prefers, given the single-predictor
model's AIC 1915.4838 and BIC 1926.002. (3) Calculate the Wald
statistic for neuroticism, compare it with the likelihood-ratio value
.3139, and say when the two would be expected to disagree. (4) Explain
why a deviance of 1906.0613 on its own says nothing about fit.

<!-- END exercise-id: CH11-E04 -->


<!-- prompt-id: CH11-E05 -->

### Exercise 5: Evaluate a product term, do not read it (Integrative). {.unnumbered}

A
logistic model of volunteering contains sex, neuroticism $N$,
extraversion $E$ and their product, with $\widehat\beta_E=.1668$ and
$\widehat\beta_{NE}=-.0086$; neuroticism runs from 0 to 24 with
quartiles 8, 11 and 15, the likelihood-ratio test of the product gives
$G^2(1)=8.6213$, $p=.0033$, and neuroticism alone had $p=.5754$ in the
additive model. (1) Give the conditional extraversion log-odds contrast
and odds ratio at each quartile. (2) Explain what .1668 refers to and
why it is not "the extraversion effect". (3) Explain how a term with no
detected additive association can participate in a detected product, and
why the additive term is kept. (4) At extraversion 13 a female student's
fitted probability is .4519 at neuroticism 8 and .4512 at 15, and the
one-point probability differences are .0245 and .0096. Explain why the
probability and log-odds pictures can agree here and disagree elsewhere,
and what a report must name before claiming an interaction.

<!-- END exercise-id: CH11-E05 -->


<!-- prompt-id: CH11-E06 -->

### Exercise 6: Read a separated output and a calibration display (Integrative). {.unnumbered}

A logistic regression fitted to 60 rows returns a
coefficient of 5.5928 with standard error 8014.9006, $p=.9994$, a
residual deviance of $5.642\times10^{-10}$, fitted probabilities of
essentially 0 and 1, and 25 iterations against the four a well-behaved
model needed. A separate, well-behaved model has one calibration decile
with mean fitted probability .4126 against an observed proportion of
.2966, whose standard error is .0379, while the other nine lie within
1.4 standard errors; its fitted probabilities span .2418 to .6235. (1)
Name the pattern producing the first output and say why the Wald test
reports no evidence for the strongest possible pattern. (2) State what a
report can and cannot say about that coefficient, and name two remedies
with the condition each requires. (3) Express the calibration gap in
standard errors and say how much weight one such decile deserves among
ten. (4) State what the fitted-probability range says about
discrimination, why that differs from calibration, and why both are
optimistic when computed on the fitting rows.
(5) Sketch the qualitative likelihood shape for a finite estimate and
for the separated example. Explain why a finite stopping point and a
very small change in likelihood do not establish a finite maximum.

<!-- END exercise-id: CH11-E06 -->



## Appendix A: Likelihood Mathematics and Numerical Checks {#ch11-likelihood-appendix .unnumbered}

A numerical optimum, a finite identified parameter and an adequate scientific model are different claims. The following two examples make their checks concrete.

### Interior and boundary solutions {.unnumbered}

For $s$ successes in $n>0$ independent Bernoulli trials with common
probability $p$,

$$
L(p)=\binom{n}{s}p^{s}(1-p)^{n-s},\qquad
\ell(p)=\log\binom{n}{s}+s\log p+(n-s)\log(1-p).
$$

For $0<s<n$, differentiating on $0<p<1$ gives

$$
\ell'(p)=\frac{s}{p}-\frac{n-s}{1-p},\qquad \widehat p=\frac{s}{n}.
$$

The binomial coefficient is constant in $p$, so recording only the count
rather than the particular observed 0/1 pattern does not change the
maximizing $p$: the Bernoulli product $p^{s}(1-p)^{n-s}$ and the binomial
likelihood share a maximizer. With only an intercept in a logistic model,
$p=\operatorname{logit}^{-1}(\beta_0)$, so the interior solution is
$\widehat p=s/n$ with
$\widehat\beta_0=\operatorname{logit}(s/n)$. At $s=0$ or $s=n$ the
likelihood is maximized at a boundary probability of 0 or 1 and there is
no finite intercept estimate, which is the one-parameter form of the
separation problem Section 6.3 develops for predictor patterns; the
analogy motivates, and does not replace, that section's conditions.
Evaluate the boundaries explicitly, because an interior numerical search
can return a point near an endpoint without ever evaluating it.

The 14 of 32 `mtcars`
records reaching 20 mpg supply a transparent illustration under a separately stipulated binomial model. Counting those outcomes does not establish
independent Bernoulli sampling of car models; the records and their
illustrative qualification are those of Chapters 5 and 6, and no new
extract or file is created.


```{r}
#| eval: false
#| echo: true
nll <- function(p) -dbinom(14, 32, p, log = TRUE)
optimize(nll, c(0, 1))
14 / 32
```


Both displayed proportions are .4375 to four decimals; the small
difference from exactly 14/32 is optimization tolerance. `dbinom()`
includes the combinatorial term, which does not move the maximizer, and
its `log = TRUE` form avoids taking a logarithm after underflow.


```{r}
#| eval: false
#| echo: true
dbinom(0, 32, 0, log = TRUE)
dbinom(32, 32, 1, log = TRUE)
qlogis(c(0, 14/32, 1))
```


The three fitted probabilities are 0, .4375 and 1, with minimum negative
log-likelihoods 0, 1.9588 and 0. `qlogis()` returns negative and positive
infinity at the two boundaries: these are exact mathematical results, not
rounded small positive probabilities, and displaying them at four decimals
would misrepresent them. The intercept-only logistic fit returns
$-.2513$, which is $\operatorname{logit}(.4375)$, closing the loop between
the scalar binomial case and the regression model of Section 3.

### Several parameters and local curvature {.unnumbered}

Under a hypothetical independent Normal model for a quantitative outcome,

$$
-\ell(\mu,\sigma)=\frac n2\log(2\pi)+n\log\sigma
+\frac{\sum_i(y_i-\mu)^2}{2\sigma^2},\qquad \sigma>0 .
$$

Parameterize $\sigma=e^{\eta}$, so an unconstrained search in $\eta$ keeps
the scale positive. The parameters are then $(\mu,\eta)$ rather than
$(\mu,\sigma)$, which also fixes the scale on which a numerical Hessian is
reported. With positive observed variation the analytic maximizers are

$$
\widehat\mu=\bar y,\qquad
\widehat\sigma_{\mathrm{ML}}=\sqrt{\frac1n\sum_i(y_i-\bar y)^2},
$$

so the maximum likelihood scale divides by $n$ where `sd()` divides by
$n-1$; this is the same distinction Chapter 9 makes before AIC and BIC.
All-equal data require separate treatment, because the Normal likelihood
is unbounded as $\sigma$ tends to zero and no finite positive-scale
maximizer exists. The `mpg` values below have positive variation. This
illustration concerns a different model and a generic optimizer: it does
not claim that car outcomes follow a Normal process, and it does not claim
that `glm()` uses BFGS.


```{r}
#| eval: false
#| echo: true
loss <- function(par) {
  -sum(dnorm(mtcars$mpg, par[1], exp(par[2]), log = TRUE))
}
start <- c(mean(mtcars$mpg) - 1, log(sd(mtcars$mpg)))
a <- optim(start, loss, method = "BFGS", hessian = TRUE)
a$convergence
eigen(a$hessian, symmetric = TRUE)$values
```


| Check | Reproduced anchor | Meaning |
|:--|:--|:--|
| Analytic mean and ML scale | 20.0906; 5.9320 | Closed-form benchmark on the recorded data; `sd()` gives 6.0269 |
| First-start estimates | 20.0908; 5.9320 | Numerical tolerance can move the last displayed digit even when convergence is reported |
| Both convergence codes | 0 | The algorithm's stopping criterion was met |
| First negative log-likelihood | 102.3778 | The same objective is compared under both starts |
| First minus second objective | About $-5.99\times10^{-7}$ | Near equality supports reproducibility of the numerical solution; it is not a test of the model |
| Hessian eigenvalues in $(\mu,\eta)$ | About 64.0006 and .9094 | Positive local curvature of the negative log-likelihood at the returned point |

: Numerical and analytic values have different roles. {#tbl-ch11-optimizer tbl-colwidths="[28,26,46]"}

A stopping code alone establishes neither
agreement at the reporting precision, nor a global optimum, nor
identification, nor scientific adequacy: compare starts and objective
values, use an analytic benchmark where one exists, and read the returned
local curvature. Section 3.2's IRLS report applies the same discipline to
the fitting routine actually used, and Section 6.3 shows its symptoms when
no maximum exists. **Extended practice.** Compare an interior binomial
case with the zero-success and all-success cases, checking endpoint
likelihoods explicitly and saying which constraint changes; for the Normal
example compare the two starts with the analytic solution and state what
the convergence codes and Hessian do and do not establish. 




## Appendix B: Changed Conditions in the Generating Process {.unnumbered}

The certification process in Section 3.3 is a reference, not a universal
performance guarantee. The three changes below retain the hours coefficient
.3 and alter one feature at a time. A smaller study reduces information;
a lower intercept changes where observations lie on the logistic curve;
a narrower range removes contrast in hours. None changes the declared
coefficient being estimated.

| Condition | Change from the baseline | Consequence to investigate |
|:--|:--|:--|
| Smaller study | Reduce 1,500 students to 100 | More sampling variation and greater risk of unstable estimates |
| Lower baseline | Set the intercept to $-5$ | More probabilities near zero, reducing typical Bernoulli information |
| Narrower hours | Generate hours uniformly from 8 to 12 | Less predictor spread for identifying the hours slope |

: One changed feature per simulation condition. {#tbl-ch11-sim-conditions tbl-colwidths="[22,34,44]"}

For each condition, generate 500 independent datasets and retain the hours
estimate, its standard error, the Wald interval and any fitting warning.
The estimate distribution, its mean and SD, and coverage of .3 answer
different questions: sampling stability, finite-sample bias, precision and
interval calibration. Record estimates with absolute magnitude above 5
and separation warnings rather than silently deleting those replications.
A warning is a screening signal, not a complete separation diagnosis.

A baseline of $-5$ does not make every profile's event rare: at twenty
hours without tutoring the logit is 1, so the probability is .7311.
Describe it as a lower-baseline process and inspect its actual prevalence.
When comparing coverage across 500 repetitions, Monte Carlo variation
matters: near .95 coverage its SE is about .0097. An apparent difference
of a few thousandths is not compelling evidence of changed calibration.

Finally fit a linear probability model to one generated dataset from each
condition and count fitted values outside $[0,1]$. A narrower range can
make the line's boundary problem less visible while making its slope less
precise. Boundary protection, sampling precision and correct interpretation
are distinct reasons for preferring or questioning a model.

## References {.unnumbered}

