---
title: "Group Comparisons with Analysis of Variance"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \widowpenalty=10000
          \clubpenalty=10000
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{12\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \usepackage{xcolor}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.pos = "!ht", collapse = TRUE, comment = "#>")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
# Display helper: round once to four decimals and drop trailing zeros;
# a vector shares the fewest decimals that show every member.
num4 <- function(x) format(round(x, 4), digits = 15, scientific = FALSE,
  trim = TRUE)
sound_levels <- c("Silence", "Instrumental music", "Music with lyrics")
sound_data <- data.frame(
  student = sprintf("S%02d", 1:30),
  sound = factor(rep(sound_levels, each = 10), levels = sound_levels),
  score = c(8, 9, 9, 11, 12, 13, 13, 14, 15, 16,
            6, 8, 8, 9, 9, 11, 11, 12, 12, 14,
            4, 5, 6, 7, 8, 8, 9, 10, 11, 12))
one_way <- aov(score ~ sound, data = sound_data)
feedback_data <- data.frame(
  student = 1:24,
  feedback = factor(rep(c("Rereading", "Answer feedback",
    "Explanatory feedback"), each = 8),
    levels = c("Rereading", "Answer feedback", "Explanatory feedback")),
  practice = factor(rep(rep(c("Massed", "Spaced"), each = 4), 3),
    levels = c("Massed", "Spaced")),
  score = c(4, 5, 7, 8, 4, 5, 7, 8,
    6, 7, 9, 10, 10, 11, 13, 14,
    8, 9, 11, 12, 16, 17, 19, 20))
full_fit <- aov(score ~ feedback * practice, data = feedback_data)
# The three-way study of Section 4.7: the 24 students above, quizzed one
# week after practice, and 24 more quizzed right after practice, whose
# cells lie at their means minus 2, minus 1, plus 1 and plus 2.
right_after_scores <- rep(c(9, 11, 12, 14, 16, 18), each = 4) +
  rep(c(-2, -1, 1, 2), 6)
timing_data <- rbind(
  data.frame(feedback_data, timing = "One week later"),
  data.frame(student = 25:48, feedback = feedback_data$feedback,
    practice = feedback_data$practice, score = right_after_scores,
    timing = "Right after practice"))
timing_data$timing <- factor(timing_data$timing,
  levels = c("Right after practice", "One week later"))
three_fit <- aov(score ~ feedback * practice * timing, data = timing_data)
```

```{r}
#| label: ch04-oneway-check
#| include: false
# Stops the render if a printed one-way anchor no longer matches its data.
f4 <- function(x) sprintf("%.4f", x)
one_way_table <- anova(one_way)
welch_check <- oneway.test(score ~ sound, data = sound_data,
  var.equal = FALSE)
tukey_check <- TukeyHSD(one_way, which = "sound")$sound
two_groups <- droplevels(subset(sound_data, sound != "Instrumental music"))
t_two <- t.test(score ~ sound, data = two_groups, var.equal = TRUE)
group_mean <- ave(sound_data$score, sound_data$sound)
doubled_check <- anova(aov(score ~ sound, data = transform(sound_data,
  score = group_mean + 2 * (score - group_mean))))
stopifnot(
  f4(tapply(sound_data$score, sound_data$sound, sd)) ==
    c("2.7080", "2.4037", "2.5820"),
  f4(one_way_table$`F value`[1]) == "6.0674",
  f4(one_way_table$`Pr(>F)`[1]) == "0.0067",
  f4(one_way_table$`Mean Sq`[2]) == "6.5926",
  f4(c(welch_check$statistic, welch_check$parameter[2],
    welch_check$p.value)) == c("5.5144", "17.9561", "0.0136"),
  f4(tukey_check[, "lwr"]) == c("-4.8470", "-6.8470", "-4.8470"),
  f4(tukey_check[, "upr"]) == c("0.8470", "-1.1530", "0.8470"),
  f4(tukey_check[, "p adj"]) == c("0.2085", "0.0047", "0.2085"),
  f4(c(t_two$statistic, t_two$p.value)) == c("3.3806", "0.0033"),
  f4(c(doubled_check$`F value`[1], doubled_check$`Pr(>F)`[1])) ==
    c("1.5169", "0.2375"))
# Anchors of the what-moves-F table, the F-reference figure, the
# familywise figure, the printed family critical values and the
# practice-ignored display.
gm <- ave(sound_data$score, sound_data$sound)
halved <- transform(sound_data, score = score - (gm - 10) / 2)
halved_tab <- anova(aov(score ~ sound, data = halved))
v <- tapply(sound_data$score, sound_data$sound, var)
F_twenty <- (20 * 8 / 2) / (19 * sum(v) / 57)
critical <- c(qt(.975, 27), qt(1 - .05 / 4, 27), sqrt(2 * qf(.95, 2, 27)))
F_crit <- c(qf(.95, 2, 27), qf(.95, 2, 8), qf(.95, 5, 27))
m <- 1:10
se_pair <- sqrt(178 / 27 * (1 / 10 + 1 / 10))
ignore_tab <- anova(aov(score ~ feedback, data = feedback_data))
stopifnot(
  f4(c(halved_tab$`Mean Sq`[1], halved_tab$`F value`[1],
    halved_tab$`Pr(>F)`[1])) == c("10.0000", "1.5169", "0.2375"),
  f4(c(F_twenty, 160 / (160 + 19 * sum(v)))) == c("12.1348", "0.2986"),
  pf(F_twenty, 2, 57, lower.tail = FALSE) < .0001,
  f4(critical) == c("2.0518", "2.3734", "2.5900"),
  f4(c(-2, -4) - critical[2] * se_pair) == c("-4.7253", "-6.7253"),
  f4(c(-2, -4) + critical[2] * se_pair) == c("0.7253", "-1.2747"),
  f4(F_crit) == c("3.3541", "4.4590", "2.5719"),
  f4(1 - (1 - .05)^c(3, 5)) == c("0.1426", "0.2262"),
  all(1 - (1 - .05 / m)^m <= .05 + 1e-12),
  f4(c(ignore_tab$`Mean Sq`[2], ignore_tab$`F value`[1],
    ignore_tab$`Pr(>F)`[1])) == c("10.4762", "12.2182", "0.0003"),
  f4(tapply(feedback_data$score, feedback_data$feedback, sd)) ==
    c("1.6903", "2.7255", "4.5981"),
  f4(258 / 29) == "8.8966")
# Anchors of Sections 2.1, 2.3, 4.5 and 4.7: the pairwise-test rate, the
# model figure's SD, the declared contrast's interval, the three-way
# cell means and the entries of the side-by-side table.
three_tab <- anova(three_fit)
two_tab <- anova(full_fit)
se_dd <- sqrt(60 / 18 * (1 / 4 + 1 / 4 + 1 / 4 + 1 / 4))
se_simple <- sqrt(60 / 18 * (1 / 4 + 1 / 4))
t18 <- qt(.975, 18)
stopifnot(
  f4(1 - ptukey(qt(.975, 27) * sqrt(2), 3, 27)) == "0.1193",
  f4(sqrt(178 / 27)) == "2.5676",
  f4(c(se_dd, t18, 4 - t18 * se_dd, 4 + t18 * se_dd)) ==
    c("1.8257", "2.1009", "0.1643", "7.8357"),
  f4(c(4 - t18 * se_simple, 4 + t18 * se_simple)) ==
    c("1.2877", "6.7123"),
  identical(as.vector(with(timing_data,
    tapply(score, list(feedback, practice, timing), mean))),
    c(9, 12, 16, 11, 14, 18, 6, 8, 10, 6, 12, 18)),
  f4(tapply(timing_data$score, timing_data$timing, mean)) ==
    c("13.3333", "10.0000"),
  isTRUE(all.equal(two_tab$`Sum Sq`, c(256, 96, 64, 60))),
  two_tab$Df == c(2, 1, 2, 18),
  f4(two_tab$`F value`[1:3]) == c("38.4000", "28.8000", "9.6000"),
  three_tab$Df == c(2, 1, 1, 2, 2, 1, 2, 36),
  f4(three_tab$`Sum Sq`) == c("450.6667", "108.0000", "133.3333",
    "32.0000", "2.6667", "12.0000", "32.0000", "120.0000"),
  f4(sum(three_tab$`Sum Sq`)) == "890.6667",
  f4(three_tab$`Mean Sq`) == c("225.3333", "108.0000", "133.3333",
    "16.0000", "1.3333", "12.0000", "16.0000", "3.3333"),
  f4(three_tab$`F value`[1:7]) == c("67.6000", "32.4000", "40.0000",
    "4.8000", "0.4000", "3.6000", "4.8000"),
  all(three_tab$`Pr(>F)`[1:3] < .0001),
  f4(three_tab$`Pr(>F)`[4:7]) == c("0.0142", "0.6733", "0.0658", "0.0142"))
# Anchors of tbl-ch04-multiway: the familywise rate at seven tests, the
# counts of rows, cells and error df, and the df of the four-way rows.
k <- 2:4
cells <- c(6, 12, 24)
four_df <- c(2, 1, 1, 1, 2, 2, 2, 1, 1, 1, 2, 2, 2, 1, 2)
stopifnot(
  f4(1 - (1 - .05)^7) == "0.3017",
  2^k - 1 == c(3, 7, 15),
  4 * cells - cells == c(18, 36, 72),
  sum(four_df) == 24 - 1, length(four_df) == 15,
  choose(12, 2) == 66)
```

\clearpage

::: {.callout-note title="Chapter overview"}
This chapter compares means by examining variation. Two invented studies
provide the observations: a one-way experiment on background sound while
studying carries the first analyses, and a two-factor experiment
crossing feedback with practice schedule carries the factorial analysis,
which asks how a comparison changes across practice schedules. We follow
the one-way ANOVA from individual scores to its F distribution,
distinguish its omnibus question from a particular contrast, and explain
how the comparison family determines an interval's coverage. Cell means
and simple effects then reveal what main effects can conceal. A worked
three-way table, set beside the two-way table of the same study, shows
how a third factor adds rows and how a two-way pattern can itself
change, and what a fourth factor would add. By the end, you will be able
to read an ANOVA table and the F distribution behind it, interpret a
contrast and its uncertainty within a family of comparisons, and connect
a factorial pattern to the design and model conditions that support its
interpretation.
:::

Two **invented studies** supply the observations. In the first, thirty
first-year university students each read the same passage for 20
minutes and then answer a 20-item reading-comprehension quiz. Each
student reads in an individual session and is randomly assigned to one
of three background-sound conditions, ten to each: *Silence*,
*Instrumental music*, or *Music with lyrics* playing during the reading.
In the second, twenty-four students study the same fractions material
under one of three kinds of feedback and one of two practice schedules,
four different students in each combination, and take a common 20-item
quiz one week after their final practice. These design details and all
scores are invented for teaching.

Our first question concerns the background-sound study: how do the
three sound-condition means compare? Chapter 3 built, step by step, the
tools for one comparison: an estimate and its standard error, a
confidence interval, a hypothesis test with its p-value, and the t tests
for one sample, paired samples and two independent samples. For two
independent groups sharing one spread, the equal-variance t test pooled
the two group variances. We extend that reasoning to several groups,
keeping the difference between an overall pattern and particular
comparisons visible.

Our later question concerns the feedback-by-practice study: how does
the feedback comparison depend on practice schedule? Section 4 answers
it, and Section 4.7 adds a third factor, the timing of the quiz.

## Comparison Questions and Designs {#sec-ch04-design}

### Outcomes, factors and levels

Both outcomes count correct answers. A comprehension score is the
number of reading-comprehension items answered correctly, and a quiz
score is the number of fractions items answered correctly; each runs
from 0 to 20, and higher values mean more correct answers. A four-point
difference therefore means four more correct answers on average; it
does not by itself describe every aspect of reading comprehension or of
learning fractions.

A **factor** in analysis of variance is a categorical variable defining
groups or conditions. Its categories are **levels**. Background sound is
one factor with three levels, so its analysis is **one-way**. The second
study crosses two factors, feedback with three levels and practice
schedule with two, so its analysis is **two-way** (Section 4). One-way
and two-way refer to the number of factors, not to the number of
levels: a one-way analysis can compare three, four or more groups, and a
third or fourth crossed factor gives a three-way or four-way layout
(Section 4.7). This use of *factor* differs from a latent factor used to
represent shared indicator variation in a measurement model.

\Needspace{10\baselineskip}

| Sound condition | Comprehension scores of ten different students |
|:--|:--|
| Silence | 8, 9, 9, 11, 12, 13, 13, 14, 15, 16 |
| Instrumental music | 6, 8, 8, 9, 9, 11, 11, 12, 12, 14 |
| Music with lyrics | 4, 5, 6, 7, 8, 8, 9, 10, 11, 12 |

: Invented reading-comprehension scores, ten different students per sound condition, on a 0--20 scale. {#tbl-ch04-sound-scores tbl-colwidths="[30,70]"}

Each score in @tbl-ch04-sound-scores belongs to a different student:
the ten scores of one sound condition describe ten different students,
not repeated measurements, and the student is both the stipulated
assignment unit and the measurement unit. The scores are fixed teaching
values. They place the three group means two points apart with similar,
unequal spreads, so every calculation reproduces by hand, but they do
not establish a Normal model, equal variances or a real effect of
background sound. The complete course data file, `background_sound.csv`,
holds the same table in one row per student, with columns `student`,
`sound` and `score`; students S01--S30 appear ten to each condition in
the order Silence, Instrumental music, Music with lyrics, and Silence is
the reference condition for the comparisons of Section 3.

### Design determines the relevant comparison

Random allocation is stipulated in each invented story. If implemented
in a real study, it could support a causal comparison among the
participating students, subject to adherence, relevant interference and
missing outcomes. It would not make the participants a random sample of
all students, and the data table alone establishes neither the
implementation of assignment nor representativeness.

Different designs can produce superficially similar tables. If students
who heard one sound condition read together in one room, they would
share that session's influences, such as noise from a corridor or a
disruption partway through the reading. If whole classes received
conditions, students within a class would share influences and the
class could be the assignment unit. If each student were measured twice,
the two outcomes would form a pair. For a real study, identify the
allocation, any repeated measurements, shared sessions or classroom
clustering, and any losses before choosing the error term: an
uncertainty formula must reflect these connections rather than treat
every row as an independent assignment.

\Needspace{12\baselineskip}

## One-Way ANOVA {#sec-mean-comparison}

### Grouped observations and the idea of ANOVA

The three sound conditions have means 12, 10 and 8 comprehension points
(@tbl-ch04-group-summary). Each has ten scores, and the sample SDs,
2.7080, 2.4037 and 2.5820, are similar. The ranges overlap widely: some
students who read with lyrics playing, scoring 11 or 12, outscored some
who read in silence, scoring 8 or 9. A difference between means
summarizes group distributions; it does not say that every student in
one group outperformed every student in another.

\Needspace{12\baselineskip}

| Sound condition | n | Mean | SD | Range |
|:--|--:|--:|--:|:--|
| Silence | 10 | 12 | 2.7080 | 8--16 |
| Instrumental music | 10 | 10 | 2.4037 | 6--14 |
| Music with lyrics | 10 | 8 | 2.5820 | 4--12 |

: Descriptions of the thirty background-sound scores. {#tbl-ch04-group-summary tbl-colwidths="[36,10,18,18,18]"}

Why not compare the conditions two at a time? Three conditions give
three pairwise differences, and a t test of each at the .05 level asks
three questions. If all three model means were equal, the chance that
at least one of the three tests rejects wrongly would be about .12,
more than twice the .05 that each test states. The three tests are not
independent, because each group mean enters two of them; three
independent tests would give $1-.95^3=.1426$. Chapter 3, Section 7.3,
noted that several tests raise the chance of at least one false
rejection. Analysis of variance therefore asks the overall question
first, in one test at .05: do the three means differ at all? Particular
comparisons follow, declared in advance or calibrated as a family
(Section 3.3).

How can one test compare three means at once?
@fig-ch04-separation-spread draws the thirty scores in its left panel:
each box shows a group's median and middle half, and the points keep
every student visible, which a bar chart of the three means would hide.
The right panel is a constructed version with every score's distance
from its group mean doubled, so the group means and counts stay fixed
while the spread doubles. The same two-point steps between the means
stand out against the narrow spread on the left and are lost in the
wide spread on the right. That is the idea of analysis of variance: it
judges how far apart the group means lie relative to how widely scores
vary within the groups [@cetinkaya2024; @kutner2005]. Section 2.3 turns
this comparison into the F statistic and returns to the right panel with
numbers.

```{r}
#| label: fig-ch04-separation-spread
#| fig-cap: "The idea of ANOVA: the same group means with different within-group spread. Left: the thirty invented scores. Right: a constructed version in which every score's distance from its group mean is doubled, so the group means and counts are unchanged, although the medians that the boxes mark can shift. Boxes show each group's median and middle half; each point is a student, jittered sideways so that equal scores stay visible."
#| fig-alt: "Two panels of vertical boxplots with jittered points for Silence, Instrumental music and Music with lyrics, on a 0 to 20 score axis. In the left panel the invented scores span 4 to 16, with medians 12.5, 10 and 8. In the right panel every deviation from the group mean is doubled, so the scores span 0 to 20, the boxes are about twice as tall and the Silence median moves to 13, while the group means stay at 12, 10 and 8."
#| fig-width: 7
#| fig-height: 3.4
#| out-width: 100%
group_mean <- ave(sound_data$score, sound_data$sound)
doubled <- transform(sound_data,
  score = group_mean + 2 * (score - group_mean))
panel <- function(dat, title) {
  boxplot(score ~ sound, data = dat, ylim = c(0, 20), las = 1,
    names = c("Silence", "Instrumental", "Lyrics"),
    col = "grey92", border = "grey40", xlab = "", cex.axis = .85,
    ylab = "Comprehension score (0-20)", main = title)
  set.seed(12345)
  stripchart(score ~ sound, data = dat, vertical = TRUE,
    method = "jitter", jitter = .12, pch = 16, cex = .7,
    add = TRUE)
}
par(mfrow = c(1, 2), mar = c(3, 4.2, 2.6, 1), cex.main = 1)
panel(sound_data, "Invented scores")
panel(doubled, "Same means, deviations doubled")
par(mfrow = c(1, 1))
```

### The group-mean model and the partition of variation

Write $Y_{ij}$ for the random score of student $i$ in sound condition $j$.
Here $j$ indexes a group, rather than a data-file column. The one-way
model is

$$
Y_{ij}=\mu_j+\varepsilon_{ij}.
$$

The unknown $\mu_j$ is the group mean in the model, and
$\varepsilon_{ij}$ is that student's unknown departure from it. After
fitting the group means, the **fitted value** is
$\widehat y_{ij}=\bar y_j$, and the **residual** is
$e_{ij}=y_{ij}-\bar y_j$. A residual is calculated from observed scores
and fitted means; a model error is a departure from an unknown model
mean. This distinction will recur in regression.

The comparison concerns two explanations of the same thirty scores.
A **one-mean model**, $Y_{ij}=\mu+\varepsilon_{ij}$, gives every sound
condition the same mean. The **group-mean model** lets each condition
have its own mean, $Y_{ij}=\mu_j+\varepsilon_{ij}$. Both allow students
to differ around their model mean. The second model can reproduce the
first by setting its three means equal. Fitting separate means can
therefore reduce the residual variation even when the population means
are equal; a smaller residual sum of squares alone cannot settle the
comparison. Section 2.3 judges that improvement relative to the
remaining within-group variation and the two extra mean parameters.

The observed grand mean is $\bar y=10$, the average of the three
group means because the groups are equally sized. Every score's
distance from the grand mean splits into two parts, the group's
departure from the grand mean and the student's residual:

$$
y_{ij}-\bar y=(\bar y_j-\bar y)+(y_{ij}-\bar y_j).
$$

For the Music-with-lyrics student who scored 4, $4-10=(8-10)+(4-8)$:
the group mean lies two points below the grand mean, and this student
another four points below the group mean. The first part is shared by
every student in the group; the second differs from student to
student.

[]{#app-ch04-small-data}Squaring the parts and adding over all thirty
students turns the split into a partition of the total squared
variation, $SS_T=SS_B+SS_W$, where

$$
\begin{aligned}
SS_T&=\sum_{j=1}^J\sum_{i=1}^{n_j}(y_{ij}-\bar y)^2,\\
SS_B&=\sum_{j=1}^Jn_j(\bar y_j-\bar y)^2,\\
SS_W&=\sum_{j=1}^J\sum_{i=1}^{n_j}(y_{ij}-\bar y_j)^2.
\end{aligned}
$$

The **total sum of squares** $SS_T$ measures variation around the one
grand mean. The **between-group sum of squares** $SS_B$ counts each
group's departure once for every student in that group, so it grows as
the group means separate. The **within-group sum of squares** $SS_W$
adds the squared residuals around the group means. The partition holds
because, within each group, the cross-products
$2(\bar y_j-\bar y)(y_{ij}-\bar y_j)$ sum to zero: the group's departure
is the same for every student, and the residuals of a group sum to zero.
For the background-sound study, $SS_B=80$, $SS_W=178$ and
$SS_T=258$, in squared comprehension points. The Instrumental-music
group adds nothing to $SS_B$, because its mean equals the grand mean,
although its scores still vary. The partition is arithmetic; it needs no
Normality or independence assumption. Those conditions enter when the
partition is used for inference, in Section 2.3.

### Degrees of freedom, mean squares and the F test {#sec-ch04-f-test}

The F statistic puts the idea of @fig-ch04-separation-spread into one
number, the variation among the group means divided by the variation
within the groups, each measured as an average squared departure. A sum
of squares grows with both the size of the departures and the number of
them, so each is divided by its **degrees of freedom** (df), the number
of departures that can vary freely, before the two sources are compared.
The F statistic has two kinds of df.

- The **between-groups df**, $J-1=2$, are the numerator df. The three
  group means can take any values, but once the grand mean is fixed
  their departures from it, weighted by group size, sum to zero, so
  only two of them are free.
- The **within-groups df**, $N-J=27$, are the denominator df. The ten
  residuals of each group sum to zero, so nine of them are free in each
  group, and three groups give 27.

The two kinds add to the total df, $N-1=29$, because the thirty
departures from the grand mean sum to zero. Dividing each sum of squares
by its df gives a **mean square**,

$$
MS_B=\frac{SS_B}{J-1},\qquad MS_W=\frac{SS_W}{N-J},\qquad
F=\frac{MS_B}{MS_W}.
$$

The within-group mean square is the pooled variance of Chapter 3,
Section 2.2, extended from two groups to $J$: each group's sample
variance $s_j^2$ is weighted by its own df,

$$
MS_W=\frac{\sum_{j=1}^J(n_j-1)s_j^2}{\sum_{j=1}^J(n_j-1)}.
$$

With equal group sizes it is the plain average of the group variances,
here 7.3333, 5.7778 and 6.6667. @tbl-ch04-one-way-anova gathers the
parts.

\Needspace{9\baselineskip}

| Source | SS | df | What the df count | MS | F | p |
|:--|--:|--:|:--|--:|--:|--:|
| Sound condition | 80 | 2 | Group means free to vary, $J-1$ | 40.0000 | 6.0674 | .0067 |
| Within groups | 178 | 27 | Residuals free to vary after fitting three means, $N-J$ | 6.5926 | | |
| Total | 258 | 29 | Departures from the grand mean free to vary, $N-1$ | | | |

: One-way ANOVA for the 30 students of the background-sound study, with what each df counts. {#tbl-ch04-one-way-anova tbl-colwidths="[20,10,5,39,10,9,7]"}

[]{#sec-ch04-anova-table}Read the table as one sentence about the
observations, $\text{Data}=\text{fit}+\text{residual}$. The
sound-condition row is the fit, what the three group means account for;
the within-groups row is the residual, what is left around them; and the
total row is the data, measured around the grand mean. The sums of
squares add across the rows, $80+178=258$, and so do the df, $2+27=29$,
which makes a quick check of any printed table. $MS_W=6.5926$ estimates
the common error variance of the model; it is smaller than the total
sample variance, $258/29=8.8966$, because it leaves out the separation
of the means. Statistical software fits the same model from the thirty
records and often prints a longer layout, with rows such as an intercept
and an uncorrected total that serve the fitting rather than the
comparison question; the Computer Lab reads one such layout row by row.

\Needspace{10\baselineskip}

| Fitted mean model | Fitted values | Residual SS | Residual df |
|:--|:--|--:|--:|
| One common mean | 10 for every student | 258 | 29 |
| Three group means | 12, 10 and 8 for the respective conditions | 178 | 27 |

: The same thirty scores fitted by one mean or three group means. {#tbl-ch04-oneway-models tbl-colwidths="[28,42,15,15]"}

Allowing the means to differ removes 80 squared score points from the
residual variation, the between-group SS in the ANOVA table, while
using two more mean parameters. Its mean square is 40, compared with
6.5926 left within groups, giving the same F of 6.0674. Thus the
between/within reading and the model-comparison reading explain one
test. The reference distribution asks how large this improvement could
be when one common mean and the stated error conditions hold. It does
not ask whether adding parameters improves the observed fit at all.

Write $\mu_S$, $\mu_I$ and $\mu_L$ for the model means of Silence,
Instrumental music and Music with lyrics. The omnibus null hypothesis is
$H_0:\mu_S=\mu_I=\mu_L$; the alternative says that at least two of these
means differ. Judging whether 6.0674 is large needs the distribution
that the statistic would follow if the null were true.

Three conditions support this reference distribution;
@fig-ch04-model draws the last two. **Independent observations**: one student's
departure from a group mean carries no information about another
student's departure. Here that rests on the design (Section 1.2), with
students individually assigned, each reading in an individual session
and scored once; no display of the scores can establish it.
**Approximately Normal errors**: the departures around each group mean,
not the pooled mixture of all thirty scores, have a roughly Normal
shape. **A common variance**: one error variance $\sigma^2$ describes
every group, which is what allows $MS_W$ to pool the three group
variances. In model shorthand, the errors $\varepsilon_{ij}$ are
independent Normal with mean zero and common variance $\sigma^2$.

```{r}
#| label: fig-ch04-model
#| fig-cap: "The one-way model for the background-sound study: the scores of each condition vary Normally around that condition's model mean, with one spread common to all three. The curves are drawn at the observed means 12, 10 and 8 with the common SD estimated by the square root of the within-group mean square, 2.5676."
#| fig-alt: "Three bell-shaped Normal curves of identical width on a comprehension-score axis from 0 to 20, centred at 8 for Music with lyrics, 10 for Instrumental music and 12 for Silence, each marked at its centre by a dotted line, with a legend naming the three conditions; the curves overlap widely."
#| fig-width: 7
#| fig-height: 2.3
#| out-width: 100%
sd_common <- sqrt(178 / 27)
centre <- c(12, 10, 8)
curve_col <- c("#202020", "#2B6CA3", "#9B4922")
xs <- seq(0, 20, length.out = 400)
par(mar = c(3.6, 4.6, .6, .8), mgp = c(2.2, .7, 0))
plot(NA, xlim = c(0, 20), ylim = c(0, .19), las = 1, yaxs = "i",
  xlab = "Comprehension score", ylab = "")
title(ylab = "Density", line = 3)
for (g in 1:3) {
  lines(xs, dnorm(xs, centre[g], sd_common), lwd = 2, lty = g,
    col = curve_col[g])
  segments(centre[g], 0, centre[g], dnorm(0, 0, sd_common),
    lty = 3, col = curve_col[g])
}
legend("topleft", c("Silence", "Instrumental music", "Music with lyrics"),
  lwd = 2, lty = 1:3, col = curve_col, bty = "n", cex = .8)
```

Under the null, both mean squares estimate the same error variance, so
their ratio tends to lie near 1; the numerator also grows with the
separation of the model means, so a large ratio is evidence against
equality. Under the three conditions and the null, F follows an
**F distribution**, one of a family of right-skewed distributions on
positive values, each indexed by its two df, numerator first: here
$F(2,27)$. The left panel of @fig-ch04-f-reference shows three members
of the family and marks each one's 5% point, the value beyond which 5%
of the distribution lies. With few denominator df, as in $F(2,8)$, the
error variance is poorly estimated and large ratios occur more often by
chance: its curve looks much like that of $F(2,27)$, but its heavier
tail moves the 5% point from 3.3541 to 4.4590. With more numerator df,
as in $F(5,27)$, the distribution gathers near 1 and its 5% point falls
to 2.5719.

```{r}
#| label: fig-ch04-f-reference
#| fig-cap: "F distributions and the one-way statistic. Left: three F distributions, each indexed by its numerator and denominator df; a vertical mark in each curve's colour and line type, labelled with its value, shows the point beyond which 5% of that distribution lies. Right: the right tail of F(2, 27), enlarged. The shaded areas together are the .05 beyond the critical value 3.3541; the dark area beyond the observed statistic 6.0674 is the p-value, .0067."
#| fig-alt: "Left panel, titled F distributions and 5% points: three right-skewed density curves over F values from 0 to 6. F(2, 27) and F(2, 8) both start near 1 at zero and decline almost together, F(2, 8) with a slightly heavier right tail; F(5, 27) rises from zero to a peak of about 0.73 near 0.6 and then declines. A labelled vertical mark in each curve's colour and line type stands at its 5% point: 2.5719 for F(5, 27), 3.3541 for F(2, 27) and 4.4590 for F(2, 8), so the heavier tail of F(2, 8) shows as a mark further to the right. Right panel: the F(2, 27) density from 2 to 9 on a vertical scale enlarged to 0.10. A dotted line marks the critical value 3.3541 and the area to its right is shaded light grey; a dashed line marks the observed F = 6.0674, and the dark area to its right is labelled p = .0067."
#| fig-width: 7
#| fig-height: 3.3
#| out-width: 100%
F_obs <- summary(one_way)[[1]][1, "F value"]
F_crit <- qf(.95, 2, 27)
df_pairs <- list(c(2, 27), c(2, 8), c(5, 27))
curve_col <- c("black", "#1F5A94", "#9B4922")
crit <- sapply(df_pairs, function(d) qf(.95, d[1], d[2]))
mark_top <- c(.34, .22, .46)
par(mfrow = c(1, 2), mar = c(4, 4.2, 2.4, .8), cex.main = .95)
xx <- seq(.001, 6, length.out = 600)
plot(NA, xlim = c(0, 6), ylim = c(0, 1), las = 1, xlab = "F statistic",
  ylab = "Density", main = "F distributions and 5% points")
for (k in 1:3) lines(xx, df(xx, df_pairs[[k]][1], df_pairs[[k]][2]),
  lwd = 2, lty = k, col = curve_col[k])
segments(crit, 0, crit, mark_top, lwd = 2, lty = 1:3, col = curve_col)
text(crit + .06, mark_top, num4(crit), adj = c(0, .3),
  col = curve_col, cex = .8)
legend("topright", lwd = 2, lty = 1:3, col = curve_col, bty = "n",
  cex = .8, sprintf("F(%d, %d)", sapply(df_pairs, `[`, 1),
    sapply(df_pairs, `[`, 2)))
xt <- seq(2, 9, length.out = 400)
plot(xt, df(xt, 2, 27), type = "n", ylim = c(0, .1), las = 1,
  xlab = "F statistic", ylab = "Density",
  main = "Right tail of F(2, 27), enlarged")
t1 <- seq(F_crit, 9, length.out = 200)
polygon(c(F_crit, t1, 9), c(0, df(t1, 2, 27), 0), col = "grey85",
  border = NA)
t2 <- seq(F_obs, 9, length.out = 200)
polygon(c(F_obs, t2, 9), c(0, df(t2, 2, 27), 0), col = "grey30",
  border = NA)
lines(xt, df(xt, 2, 27), lwd = 2)
segments(c(F_crit, F_obs), 0, c(F_crit, F_obs), c(.075, .035),
  lty = c(3, 2))
text(F_crit, .08, paste("Critical value", num4(F_crit)), adj = c(0, 0),
  cex = .75)
text(F_obs + .1, .04, sprintf("Observed\nF = %s\np = .0067", num4(F_obs)),
  adj = c(0, 0), cex = .75)
par(mfrow = c(1, 1))
```

The dark area in the right panel of @fig-ch04-f-reference, .0067, is
read in the same way as the tail areas and p-values of Chapter 3,
Section 4. Only the right tail counts, because greater separation of the
means produces a larger ratio. Under the null and the model conditions,
a ratio at least as large as 6.0674 occurs with that probability. The
observed F lies beyond the critical value 3.3541, so at a prespecified
.05 significance level the equality null is rejected; comparing F with
the critical value and comparing p with .05 give the same decision. This
is evidence about the three-mean pattern, not a probability that the
null is true and not an identification of which particular comparisons
differ.

The box below states, in the four lines used since Chapter 3, what this
F test addresses, what it rests on, what the data can check and what the
result supports.

[]{#ch04-justify-oneway}

::: {.callout-tip title="Justifying the analysis"}
- **Target:** whether the model means of the three sound conditions,
  $\mu_S$, $\mu_I$ and $\mu_L$, are all equal for the thirty students of
  the background-sound study, on the 0--20 comprehension scale; the
  observed means 12, 10 and 8 estimate them.
- **Conditions:**
  - *Secured by design:* individual random assignment, ten students to
    each condition, and one score per student from an individual
    session, so the scores are independent and a difference among the
    means reads as an effect of sound among these students, not in a
    wider population (Sections 1.2 and 2.3).
  - *Model assumptions:* Normal errors with one common variance (Section
    2.3), estimated by $MS_W=6.5926$ on 27 df.
  - *Choices and defaults:* all thirty students, none missing; a .05
    level fixed in advance; the right tail of $F(2,27)$; the classical
    calibration that pools the group variances.
- **Check:** the group SDs, 2.7080, 2.4037 and 2.5820, are similar, the
  largest 1.1266 times the smallest, with ten students in each group,
  and the boxplots show no outlying score; markedly unequal SDs would
  call for Welch's one-way ANOVA, chosen in advance (Section 5.2). Ten
  scores per group say little about shape, and independence rests on
  the design.
- **Bound:** $F(2,27)=6.0674$, $p=.0067$ is evidence at the .05 level
  against equal means for these thirty students under the model. It
  does not say which means differ or by how much (Section 3), and,
  with invented scores, says nothing about real studying.
:::

With equal group sizes $n$, the numerator of F shows directly what the
statistic compares:

$$
F=\frac{n\sum_{j=1}^J(\bar y_j-\bar y)^2/(J-1)}{MS_W}.
$$

Three quantities move it, the same three that moved Chapter 3's t
statistic (Section 4.4): the separation of the group means, the group
size $n$ and the within-group spread in $MS_W$.
@tbl-ch04-what-moves-f changes one of them at a time; its second row is
the right panel of @fig-ch04-separation-spread.

\Needspace{11\baselineskip}

| Version of the study | $n$ | Group means | $MS_B$ | $MS_W$ | F | p |
|:--|--:|:--|--:|--:|--:|--:|
| The thirty invented scores | 10 | 12, 10, 8 | 40 | 6.5926 | 6.0674 | .0067 |
| Spread doubled | 10 | 12, 10, 8 | 40 | 26.3704 | 1.5169 | .2375 |
| Separation halved | 10 | 11, 10, 9 | 10 | 6.5926 | 1.5169 | .2375 |
| Group size doubled | 20 | 12, 10, 8 | 80 | 6.5926 | 12.1348 | <.0001 |

: What moves the F statistic: one change at a time to the background-sound study. The last three rows are constructed variations of the invented scores, not further studies; the last has 2 and 57 df, the others 2 and 27. {#tbl-ch04-what-moves-f tbl-colwidths="[34,7,17,11,11,11,9]"}

Doubling every distance from the group mean doubles each group's SD and
multiplies $MS_W$ by four; halving every mean's distance from the grand
mean divides $MS_B$ by four. Each change alone quarters F, from 6.0674
to 1.5169, and moves p from .0067 to .2375: F reads separation relative
to spread, never either alone. Doubling the group size with the same
means and SDs leaves every boxplot looking the same, yet doubles $MS_B$
and so F: more students supply more information about the same pattern.
Section 2.4 shows that eta squared barely responds to that third change.

### Two groups, direction and magnitude

With two groups, $MS_W$ is Chapter 3's pooled variance $s_p^2$, so the
classical one-way F test and Chapter 3's **equal-variance two-sample t
test** (Section 5.3) share one model, one pooled variance and one
residual df, and $F(1,N-2)=t_{N-2}^2$. Chapter 3 labels its two groups 1
and 0; this chapter numbers groups $j=1,\ldots,J$, so its two-group case
uses 1 and 2.

For Silence and Music with lyrics alone, the pooled variance is 7
on 18 df, and $t(18)=3.3806$ and $F(1,18)=11.4286=3.3806^2$ have the
same two-sided $p=.0033$. Squaring removes direction, so the signed
difference, Silence minus Music with lyrics, 4 points, carries it.
This matched fit reads with the [*Justifying the analysis*
box](chapter_03.qmd#ch03-justify-equal-variance) that Chapter 3 gave the
equal-variance t test; only the target changes, to the
Silence-minus-Music-with-lyrics difference for these twenty students,
whose SDs, 2.708 and 2.582, are similar. The two-group refit pools
7.0000 on 18 df rather than the three-group $MS_W=6.5926$ on 27, because
it leaves out the Instrumental group, and it answers a narrower question
than the three-group F: one pairwise comparison rather than the whole
pattern. [Appendix A](#app-ch04-pooled-identity) derives the identity,
and Section 5.2 returns to groups whose spreads may differ.

Magnitude begins on the comprehension scale. The observed
Instrumental-music-minus-Silence difference is $-2$ points and the
Music-with-lyrics-minus-Silence difference is $-4$: two and four
fewer items answered correctly, on average, than in silence. Whether a
difference of this size matters for studying needs a substantive
criterion, not only a p-value. One descriptive summary of the overall
separation is **eta squared**,

$$
\eta^2=\frac{SS_B}{SS_T}=\frac{80}{258}=.3101:
$$

group means account for .3101 of the observed squared variation in these
thirty records. It is the fit's share of the table's total, the quantity
that later chapters call the $R^2$ of a fitted model. This is a fraction
of a sample decomposition, not the proportion of comprehension caused by
background sound. Eta squared and F answer different questions: in the
last row of @tbl-ch04-what-moves-f, twenty students per group double F
while eta squared barely moves, from .3101 to .2986, because the means
account for about the same share of the variation while the larger study
gives stronger evidence that they differ. A small p-value therefore does
not signal a large difference. Because sample means differ even when
model means are equal, eta squared slightly overstates the share in the
model; omega squared adjusts for that and gives .2525 here
([Appendix B](#app-ch04-effect-sizes)). A standardized mean difference,
explained in [Chapter 3, Section 7.1](chapter_03.qmd#sec-ch03-interpretation),
divides a difference by a stated SD; dividing it by an individual-score
SD gives a different quantity from dividing it by the SE of that
difference.

\Needspace{12\baselineskip}

## Contrasts and Multiple Comparisons {#sec-ch04-contrasts}

### Contrasts and their weights

The omnibus question asks whether all three model means agree. A
**contrast** states a particular comparison as a weighted combination
of means whose weights sum to zero:

$$
C=\sum_{j=1}^Jc_j\mu_j,\qquad \sum_{j=1}^Jc_j=0,
\qquad \widehat C=\sum_{j=1}^Jc_j\bar y_j.
$$

The symbol $C$ here names a contrast, and $\widehat C$ its estimate.
The weights specify both the comparison and its direction.

The study plan declares two questions, each music condition against
Silence: Instrumental music minus Silence and Music with lyrics minus
Silence. A negative estimate means fewer items answered correctly than
in silence. A broader question compares the equal average of the two
music conditions with Silence. In the declared order Silence,
Instrumental music, Music with lyrics, the three comparisons have these
weights:

\Needspace{12\baselineskip}

| Comparison | Weights | Estimate |
|:--|:--|--:|
| Instrumental music minus Silence | $(-1,1,0)$ | $-2$ |
| Music with lyrics minus Silence | $(-1,0,1)$ | $-4$ |
| Equal average of the two music conditions minus Silence | $(-1,1/2,1/2)$ | $-3$ |

: Three contrast meanings in the background-sound study. {#tbl-ch04-weights tbl-colwidths="[55,25,20]"}

Only the first two form the declared family whose intervals Section 3.3
calibrates together. The third gives the two music conditions equal
weight; its $-3$ hides that they fall 2 and 4 points below Silence, so it
does not replace both separate comparisons when both are required. The
weights also fix the scale: weights such as $(-2,1,1)$ double the
estimate and its SE and report twice a mean difference, whereas
$(-1,1/2,1/2)$ keeps the estimate in comprehension points. With unequal
group sizes, equal weights and weights proportional to student counts
answer different questions ([Appendix B.1](#app-ch04-weighting)).

### The standard error of a contrast

The SE of a contrast extends the two-group SE of Chapter 3,
Section 2.2. A difference between two means is the contrast with
weights $+1$ and $-1$, and Chapter 3's pooled SE can be written with
those weights visible:

$$
s_p\sqrt{\frac{1}{n_1}+\frac{1}{n_0}}
=\sqrt{s_p^2\left(\frac{1^2}{n_1}+\frac{(-1)^2}{n_0}\right)}.
$$

Under the independent common-variance model, each estimated group mean
has variance $\sigma^2/n_j$, and the variance of a weighted sum of
independent means uses the **squared weights**. Opposite signs define a
difference; they do not make independent sources of uncertainty cancel.
Several groups make two changes to Chapter 3's form: the weights $c_j$
can describe any comparison, and $s_p^2$ is replaced by $MS_W$, the
variance pooled over all $J$ groups (Section 2.3). Then

$$
\widehat{SE}(\widehat C)=
\sqrt{MS_W\sum_{j=1}^J\frac{c_j^2}{n_j}}.
$$

For each difference with Silence,

$$
\widehat{SE}=
\sqrt{\frac{178}{27}\left(\frac{1}{10}+\frac{1}{10}\right)}
=1.1483\text{ points}.
$$

Each contrast uses the three-group estimate 6.5926 and its 27 error df,
although it uses only two of the three means: under the common-variance
model all three groups estimate the same $\sigma^2$, so pooling them
uses more information than the two-group refit of Section 2.4, whose
7.0000 on 18 df gives an SE of 1.1832. For the equal-average contrast,
the squared weights $1$, $1/4$ and $1/4$ give $\widehat{SE}=0.9944$.
This smaller SE belongs to a different estimand, an average of two
means compared with the reference; it does not make either separate
comparison more precise. The
[NIST contrast explanation](https://www.itl.nist.gov/div898/handbook/prc/section4/prc436.htm)
develops the same weighted-mean variance calculation.

### Families of comparisons and the familywise error rate {#sec-ch04-families}

Section 2.1 showed that several tests of one study raise the chance of
at least one false rejection, so a reader needs to know how many
comparisons were made and when each was chosen. A **planned
comparison** is declared before the results are seen, as this study's
plan declares each music condition against Silence. A **post hoc
comparison** is chosen after the results are seen, often because it
looks large, such as the largest of the three pairwise differences.
[Chapter 3, Section 7.3](chapter_03.qmd#sec-ch03-selection),
showed that a result chosen because it looks large tends to overstate
its target; the same selection problem arises among the several
comparisons of one study.

Each comparison tested at level .05 has a **per-comparison error rate**,
also called the testwise rate, of .05: when its null hypothesis is true,
it rejects wrongly with probability .05. A reader of the whole report,
however, cares whether any of its comparisons is a false rejection. The
**familywise error rate** is the probability of at least one Type I
error (Chapter 3, Section 7.2) among all the comparisons of a family.
For $m$ independent comparisons whose null hypotheses are all true, each
tested at .05, it is $1-(1-.05)^m$, which @fig-ch04-familywise draws
against $m$: already .1426 for three comparisons and .2262 for five.

To hold the familywise rate at .05, **Bonferroni's method** tests each
of the $m$ comparisons at level $.05/m$, or builds each interval with
confidence $1-.05/m$. Its guarantee needs no independence, because the
probability that at least one of several events occurs is at most the
sum of their probabilities:

$$
P(\text{at least one false rejection})
\le\sum_{k=1}^mP(\text{false rejection }k)
=m\cdot\frac{.05}{m}=.05.
$$

[]{#app-ch04-families}The lower line of @fig-ch04-familywise stays at or
below .05 for every $m$. The price is width: each interval is wider and
each comparison has less power, so a family should contain the
comparisons that matter rather than every comparison that could be
made. Bonferroni's method suits any declared set; other methods are
calibrated for particular families and give narrower intervals when
their family fits the question (@tbl-ch04-families) [@tukey1953;
@kutner2005].

```{r}
#| label: fig-ch04-familywise
#| fig-cap: "The familywise error rate for m hypothetical independent comparisons whose null hypotheses are all true, each tested at .05 (upper line) or at Bonferroni's .05/m (lower line)."
#| fig-alt: "Two lines over the number of comparisons m, from 1 to 10. The upper line, each comparison tested at .05, rises from .05 at m = 1 through .0975 and .1426, marked at m = 3, and .1855 and .2262, marked at m = 5, to .4013 at m = 10. The lower line, Bonferroni's allocation of .05/m to each comparison, stays just below a dotted reference line at .05 for every m."
#| fig-width: 7
#| fig-height: 2.5
#| out-width: 100%
m <- 1:10
each <- 1 - (1 - .05)^m
bonferroni <- 1 - (1 - .05 / m)^m
par(mar = c(4, 4.4, 1, 1))
plot(m, each, type = "b", pch = 16, lwd = 2, ylim = c(0, .45), las = 1,
  xaxt = "n", xlab = "Number of comparisons in the family, m",
  ylab = "Familywise error rate")
axis(1, at = m)
abline(h = .05, col = "grey55", lty = 3)
lines(m, bonferroni, type = "b", pch = 1, lwd = 2, lty = 2,
  col = "#1F5A94")
points(c(3, 5), each[c(3, 5)], pch = 16, cex = 1.7)
text(c(3, 5), each[c(3, 5)], sub("^0", "", num4(each[c(3, 5)])),
  pos = 2, offset = .8, cex = .85)
legend("topleft", c("Each comparison tested at .05",
  "Bonferroni: each tested at .05/m"),
  pch = c(16, 1), lty = 1:2, lwd = 2, col = c("black", "#1F5A94"),
  bty = "n", cex = .85)
```

\Needspace{16\baselineskip}

| Family of comparisons | When it fits | Method that holds the familywise rate |
|:--|:--|:--|
| One comparison | The plan names it in advance as the study's only question | None needed: the ordinary 95% interval, $\widehat C\pm t_{.975}\,\widehat{SE}$ |
| A declared set of $m$ comparisons | The plan names the set before the data are seen | Bonferroni: each comparison at $.05/m$ |
| Each group against one reference | The plan compares every condition with a control or reference condition | Dunnett's method |
| All pairs of groups | Every pairwise difference matters, or a pair is picked after the means are seen | Tukey's method |
| Every contrast, including averages of groups | Any comparison may be formed after the data are seen | Scheffé's method |

: Families of comparisons and the methods that hold the familywise error rate for each. {#tbl-ch04-families tbl-colwidths="[28,42,30]"}

The wider the family, the more comparisons its intervals must cover
together, and the larger the critical value. For one difference between
two sound conditions, on 27 error df, the value for a single planned
comparison is 2.0518 and Scheffé's value for every contrast of the three
means is 2.5900; the declared pair, Tukey's all pairs and Bonferroni's
three pairs fall in between. The family comes from the comparisons for
which coverage is owed, fixed before looking at which interval excludes
zero. An omnibus rejection is not required before a planned comparison,
and it does not by itself protect comparisons chosen afterward.

The study plan declares two comparisons, each music condition against
Silence, so the family is that pair, and Bonferroni gives each interval
97.5% confidence, with .0125 in each tail: the critical value is
$t_{.9875;27}=2.3734$. @tbl-ch04-contrasts gives the two intervals,
$\widehat C\pm2.3734\,\widehat{SE}$. Both estimates use the Silence
mean, a dependence that the union bound allows
([Appendix B.3](#app-ch04-dependence)).

\Needspace{10\baselineskip}

| Comparison | Estimate | SE | Family lower | Family upper |
|:--|--:|--:|--:|--:|
| Instrumental music minus Silence | $-2$ | 1.1483 | $-4.7253$ | 0.7253 |
| Music with lyrics minus Silence | $-4$ | 1.1483 | $-6.7253$ | $-1.2747$ |

: Bonferroni 95% simultaneous intervals for the two declared comparisons with Silence. {#tbl-ch04-contrasts tbl-colwidths="[36,16,16,16,16]"}

[]{#ch04-justify-contrasts}These intervals read with the
[one-way box](#ch04-justify-oneway) of Section 2.3, with three changes.
The target becomes two contrasts in comprehension points, $\mu_I-\mu_S$
and $\mu_L-\mu_S$, rather than the equality of all three means; the
choices add the declared family and Bonferroni's calibration, fixed
before the scores were seen, which only the analysis record can confirm;
and the bound belongs to this declared pair, not to comparisons chosen
after the means were inspected.

The omnibus F and these intervals answer different questions. The F
test says that the three model means are not all equal; the intervals
say which declared differences are established and how large they may
be. Jointly, Music with lyrics minus Silence is negative, between
$-6.7253$ and $-1.2747$ points. The Instrumental-music interval,
$[-4.7253,0.7253]$, includes zero: it is compatible with a small
benefit, no difference or a loss of several points, so it establishes
no equivalence, which would need a declared margin and an analysis
designed for it. These are calculations for invented scores, not a
finding about studying with music.

## Multi-Way (Factorial) ANOVA {#sec-ch04-two-way}

Analysis of variance with two or more crossed factors is **multi-way**,
or **factorial**, ANOVA; this section works the two-way case in full
with the feedback-by-practice study and closes, in Section 4.7, with a
worked three-way analysis and the structure of four-way designs.

### The two-way design and what interaction means {#sec-ch04-interaction-idea}

In the feedback-by-practice study, twenty-four students study the same
fractions material and complete a common 20-item quiz one week after
their final practice. *Rereading* offers the worked material again;
*Answer feedback* supplies the correct answer after an attempt;
*Explanatory feedback* supplies the answer with a worked explanation.
*Massed* practice occurs in one session and *Spaced* practice in two
shorter sessions with the same total practice time. Random allocation
places four different students in each combination of feedback and
practice (@tbl-ch04-feedback-scores).

\Needspace{10\baselineskip}

| Feedback condition | Massed practice | Spaced practice |
|:--|:--|:--|
| Rereading | 4, 5, 7, 8 | 4, 5, 7, 8 |
| Answer feedback | 6, 7, 9, 10 | 10, 11, 13, 14 |
| Explanatory feedback | 8, 9, 11, 12 | 16, 17, 19, 20 |

: Invented fractions-quiz scores, four different students per cell, on a 0--20 scale. {#tbl-ch04-feedback-scores tbl-colwidths="[38,31,31]"}

A **cell** contains the cases that share one level of each factor;
Answer feedback with Spaced practice is one cell. Crossing three
feedback levels with two practice levels gives six cells, and four
students in every cell make the layout **balanced**, as ten students
per condition balance the one-way layout. Each score belongs to a
different student, the stipulated assignment and measurement unit: the
four scores of a cell are not repeated measurements, and the Massed and
Spaced scores are not paired. Each cell's scores lie 2 and 1 points
below and above its mean, a repeated pattern that simplifies the
arithmetic; it is a teaching construction, not evidence of Normal
errors, equal variances or an educational effect. The course data file
`feedback_practice.csv` holds the same scores in one row per student,
with columns `student`, `feedback`, `practice` and `score`, ordered by
feedback, then practice, then student.

A two-way ANOVA compares a numerical score across combinations of two
factors. Chapter 3's two-way frequency table (Section 6) instead counts
cases in combinations of two categorical variables: *two-way* describes
a layout, not a shared statistical test.

Two factors **interact** when the effect of one factor differs across
the levels of the other. Here is one invented illustration, outside the
chapter's two studies. Suppose a short vocabulary preview is given
before a reading passage, and that students are also grouped by whether
they began with weak or strong vocabulary. The preview raises
comprehension scores substantially for the students who began weak, and
barely changes them for the students who began strong. The effect of the
preview *depends on* prior vocabulary: that dependence is the
interaction. Describing only an average preview effect would tell a
reader that the preview helps somewhat, and would conceal that it helps
one group a great deal and the other hardly at all.

Studying two factors together in one crossed design has three
advantages over running two separate one-factor studies.

1. **It is more efficient.** Every student contributes to both factor
   comparisons at once, because the levels of one factor are spread
   across the levels of the other. A crossed design therefore needs
   fewer cases per condition than a series of separate one-factor
   studies aimed at the same comparisons.
2. **It reduces residual variation.** A second factor that influences
   the outcome moves variation out of the error term and into the fit.
   Section 4.4 shows what leaving practice out does to these scores.
3. **It makes interaction examinable.** Only a crossed design supports
   the question of whether a comparison changes across the other
   factor's levels. A series of one-factor studies supplies no
   estimate of that change, however carefully each one is run.

Practice schedule therefore gives each feedback comparison a context.
We can compare feedback conditions within Massed practice, within Spaced
practice, or after averaging over the two schedules, and these are
distinct quantities. A **main effect** compares one factor's levels
after averaging across the other factor with stated weights. A
**simple effect** makes the comparison at a specified level of the
other factor. **Interaction** means that a comparison changes across
those levels.

### Main effects and interaction through graphs

In a two-factor design either main effect, the interaction, or any
combination of the three can be present. Six schematic two-by-two
patterns make the whole range visible before we return to the full
three-by-two data. Each line in @fig-ch04-interaction-patterns is one
level of a two-level feedback factor, Feedback 1 or Feedback 2; the
horizontal positions are Massed and Spaced practice. These are
mean-only drawings, not additional observed studies, and they carry no
significance claim.

```{r}
#| label: fig-ch04-interaction-patterns
#| fig-cap: "Six schematic patterns of cell means, covering the whole range of main-effect and interaction combinations. The two lines are two levels of a feedback factor; the drawings illustrate averages and changing differences, without significance claims."
#| fig-alt: "Six panels. No effects: Feedback 1 and Feedback 2 coincide as one flat line at 10. One main effect: two separated flat lines at 8 and 12. Two main effects: parallel rising lines. Interaction only: crossing lines with equal marginal means. Interaction qualifies: both lines rise from 8, Feedback 2 more steeply. Interaction reverses: Feedback 1 rises from 8 to 10 while Feedback 2 falls from 16 to 10."
#| fig-width: 7
#| fig-height: 3.9
#| out-width: 100%
patterns <- list(
  "No effects" = matrix(c(10, 10, 10, 10), 2, byrow = TRUE),
  "One main effect" = matrix(c(8, 8, 12, 12), 2, byrow = TRUE),
  "Two main effects" = matrix(c(8, 10, 12, 14), 2, byrow = TRUE),
  "Interaction only" = matrix(c(8, 14, 14, 8), 2, byrow = TRUE),
  "Interaction qualifies" = matrix(c(8, 10, 8, 14), 2, byrow = TRUE),
  "Interaction reverses" = matrix(c(8, 10, 16, 10), 2, byrow = TRUE))
par(mfrow = c(2, 3), mar = c(3.6, 3.6, 2.6, .8), mgp = c(2.2, .7, 0))
for (nm in names(patterns)) {
  matplot(1:2, t(patterns[[nm]]), type = "b", pch = c(16, 1),
    lty = c(1, 2), col = c("black", "grey40"),
    ylim = c(6, 20), xlim = c(.85, 2.15), xaxt = "n",
    xlab = "Practice", ylab = "Mean score", main = nm)
  axis(1, at = 1:2, labels = c("Massed", "Spaced"))
  legend("topleft", c("Feedback 1", "Feedback 2"), pch = c(16, 1),
    lty = c(1, 2), col = c("black", "grey40"), bty = "n",
    cex = .75, horiz = TRUE, x.intersp = .5)
}
par(mfrow = c(1, 1))
```

Read each pattern by taking differences, not only by recognizing a
shape. Define a schedule difference as Spaced minus Massed and a
difference of differences as Feedback 2's schedule difference minus
Feedback 1's.

\Needspace{16\baselineskip}

| Pattern | Feedback 1: Massed, Spaced | Feedback 2: Massed, Spaced | Schedule differences 1, 2 | Difference of differences |
|:--|:--|:--|:--|--:|
| No effects | 10, 10 | 10, 10 | 0, 0 | 0 |
| One main effect | 8, 8 | 12, 12 | 0, 0 | 0 |
| Two main effects | 8, 10 | 12, 14 | 2, 2 | 0 |
| Interaction only | 8, 14 | 14, 8 | 6, $-6$ | $-12$ |
| Interaction qualifies | 8, 10 | 8, 14 | 2, 6 | 4 |
| Interaction reverses | 8, 10 | 16, 10 | 2, $-6$ | $-8$ |

: Arithmetic behind the six schematic patterns. {#tbl-ch04-patterns tbl-colwidths="[25,19,19,19,18]"}

The first three panels have a zero difference of differences. Nothing
varies in the first; in the second, Feedback 2 exceeds Feedback 1 by 4
points under either schedule, one main effect; in the third, Spaced
also exceeds Massed by 2 points everywhere, two main effects. Parallel
lines express those constant comparisons.

The last three panels have a nonzero difference of differences. In
the crossing panel, both feedback marginal means are 11 and both
practice marginal means are 11 when cells receive equal weight. Both
averaged comparisons are zero despite large conditional differences,
which is why a small main effect need not mean that a factor is
irrelevant at every level of another factor.

In the qualifying panel the schedule comparison is 2 for Feedback 1 and
6 for Feedback 2. Both are positive, so the averaged 4-point schedule
comparison points the right way for both feedback levels but describes
neither exactly. The lines need not cross for interaction to be
present. In the reversing panel the schedule comparison is 2 for
Feedback 1 and $-6$ for Feedback 2, so the averaged comparison of $-2$
points the wrong way for Feedback 1. Reporting only the two main effects
would then mislead a reader about every conditional comparison. Examine
the interaction first, and read a main effect as an average whose parts
are still visible. These drawings establish only the numerical
patterns: significance also needs information about variation, sample
size and the reference distribution.

### Cells, marginal means and the full data

Turn to the six cells of the feedback-by-practice study, whose units
Section 4.1 described: this analysis uses six separate groups of four,
where the background-sound analysis used three groups of ten.

Write $\bar y_{jk}$ for a cell mean, with $j=1,2,3$ indexing feedback
and $k=1,2$ indexing practice. A dot replaces an index that has been
averaged over: $\bar y_{j\cdot}$ averages a feedback level over
practice, and $\bar y_{\cdot k}$ averages a practice level over
feedback. An **effect** in this layout is a marginal mean's departure
from the grand mean, 10. @tbl-ch04-cell-means gives the cell means, the
equally weighted marginal means and the effects.

\Needspace{15\baselineskip}

| Feedback | Massed | Spaced | Feedback mean | Feedback effect |
|:--|--:|--:|--:|--:|
| Rereading | 6 | 6 | 6 | $-4$ |
| Answer feedback | 8 | 12 | 10 | 0 |
| Explanatory feedback | 10 | 18 | 14 | 4 |
| Practice mean | 8 | 12 | Grand mean 10 | |
| Practice effect | $-2$ | 2 | | |

: Cell means, equally weighted marginal means and effects for the complete invented study. {#tbl-ch04-cell-means tbl-colwidths="[27,13,13,26,21]"}

Each marginal mean averages its cell means with equal weights; with four
students in every cell it is also the mean of that condition's students.
The feedback effects, $-4$, 0 and 4, and the practice effects, $-2$ and
2, each sum to zero. The practice comparison, Spaced minus Massed,
averages 4 points over the three feedback conditions, but its
components, the schedule differences within the rows, are 0,
4 and 8. A four-point main effect is therefore a meaningful
stated average that describes none of the conditional comparisons except
Answer feedback's.

```{r}
#| label: fig-ch04-cell-means
#| fig-cap: "Quiz cell means across practice schedules in the full 3 × 2 example. Each line joins means of different students; the schedule differences are 0, 4 and 8 points."
#| fig-alt: "Rereading stays at 6 from Massed to Spaced practice. Answer feedback rises from 8 to 12. Explanatory feedback rises from 10 to 18. The nonparallel lines show that the schedule comparison changes across feedback conditions."
#| fig-width: 7
#| fig-height: 3.1
#| out-width: 100%
par(mar = c(4, 4, 1, 1), cex = .9)
with(feedback_data,
  interaction.plot(practice, feedback, score, type = "b",
    pch = c(16, 1, 17), lty = 1:3,
    col = c("#202020", "#2B6CA3", "#9B4922"),
    xlab = "Practice schedule", ylab = "Mean quiz score",
    trace.label = "Feedback", ylim = c(0, 20)))
```

The connected lines in @fig-ch04-cell-means organize cell means. They
do not show individual students moving from one condition to another.
The nonparallel pattern describes interaction in the observed means;
the full model supplies the uncertainty calculation for an inferential
claim.

### Additivity and the full model {#sec-ch04-full-model}

If feedback and practice combined additively, every cell mean would be
the grand mean plus its feedback effect plus its practice effect, and
the schedule difference would be the same in every row.
@tbl-ch04-interaction-departures sets these **additive predictions**
beside the observed cell means. The difference between an observed cell
mean and its additive prediction is an **interaction departure**.

\Needspace{12\baselineskip}

| Cell | Observed mean | Additive prediction | Interaction departure |
|:--|--:|--:|--:|
| Rereading, Massed | 6 | 4 | 2 |
| Rereading, Spaced | 6 | 8 | $-2$ |
| Answer feedback, Massed | 8 | 8 | 0 |
| Answer feedback, Spaced | 12 | 12 | 0 |
| Explanatory feedback, Massed | 10 | 12 | $-2$ |
| Explanatory feedback, Spaced | 18 | 16 | 2 |

: Observed cell means, additive predictions from the grand mean and both effects, and interaction departures. {#tbl-ch04-interaction-departures tbl-colwidths="[37,19,22,22]"}

Answer feedback's two cells sit exactly on their predictions, but the
other four do not: Rereading gains less from spacing than the average,
and Explanatory feedback more, by 2 points in each cell. Agreement in
one row does not make the whole layout additive.

The full mean model includes such departures:

$$
Y_{ijk}=\mu+\alpha_j+\beta_k+(\alpha\beta)_{jk}
        +\varepsilon_{ijk}.
$$

Here $i=1,\ldots,4$ identifies a student within a cell, $j=1,2,3$
identifies feedback and $k=1,2$ identifies practice. The grand model
mean is $\mu$; $\alpha_j$ and $\beta_k$ are the factor effects;
$(\alpha\beta)_{jk}$ is the interaction departure; and
$\varepsilon_{ijk}$ is the individual error around the cell mean.
These indexed effects are distinct from the unindexed error
probabilities $\alpha$ and $\beta$ used in hypothesis testing.

In the balanced decomposition, main effects sum to zero, and
interaction departures sum to zero along each row and each column, so
they carry only the pattern left after the two main effects. With two
columns, a row's second departure is the negative of its first, and the
three first-column departures, 2, 0 and $-2$, sum to zero, so only two
of them are free. Thus the interaction has $(3-1)(2-1)=2$ df. It
describes a two-dimensional pattern, rather than one particular
four-cell difference.

[]{#app-ch04-two-way-arithmetic}Each score's distance from the grand
mean now splits into four parts, a feedback effect, a practice effect,
an interaction departure and the student's residual. Squaring the parts
and adding over all 24 students gives, in this balanced layout as in
Section 2.2, four sums of squares with matching df:

$$
SS_T=SS_A+SS_B+SS_{AB}+SS_E,\qquad
df_T=df_A+df_B+df_{AB}+df_E.
$$

In this two-way table A denotes feedback and B practice, so $SS_B$ now
means the practice component, not the between-group component of the
one-way table, and $SS_E$ is the within-cell error:

$$
\begin{aligned}
SS_A&=2n\sum_j(\bar y_{j\cdot}-\bar y)^2=256,\\
SS_B&=3n\sum_k(\bar y_{\cdot k}-\bar y)^2=96,\\
SS_{AB}&=n\sum_j\sum_k
 (\bar y_{jk}-\bar y_{j\cdot}-\bar y_{\cdot k}+\bar y)^2
 =64,\\
SS_E&=\sum_j\sum_k\sum_i(y_{ijk}-\bar y_{jk})^2=60.
\end{aligned}
$$

Each component counts its effects or departures once for every student
who shares them. The partition reads
$476=256+96+64+60$ on $23=2+1+2+18$ df: each of
the six cells uses one mean, so the error has $24-6=18$ df, and the fit
part is $256+96+64=416$ on five df.

\Needspace{14\baselineskip}

| Source | SS | df | MS | F | p |
|:--|--:|--:|--:|--:|--:|
| Feedback (A) | 256 | 2 | 128.0000 | 38.4 | <.0001 |
| Practice (B) | 96 | 1 | 96.0000 | 28.8 | <.0001 |
| A × B | 64 | 2 | 32.0000 | 9.6 | .0015 |
| Within cells | 60 | 18 | 3.3333 | | |
| Total | 476 | 23 | | | |

: Full two-way ANOVA using all 24 students. {#tbl-ch04-two-way-anova}

Read this table as in [Section 2.3](#sec-ch04-anova-table): the three
source rows are the fit, the within-cells row the residual and the total the
data, and the sums of squares and df add across the rows. Read the
interaction row first, because it decides how the two main-effect rows
can be read.

\Needspace{11\baselineskip}

| Fitted mean model | Comparison it allows | Residual SS | Residual df |
|:--|:--|--:|--:|
| Additive: four mean parameters | One practice difference shared by all feedback conditions | 124 | 20 |
| Full: six cell means | A separate practice difference in each feedback condition | 60 | 18 |

: Two mean models fitted to the same twenty-four students. {#tbl-ch04-factorial-models tbl-colwidths="[28,42,15,15]"}

This comparison repeats the logic of Section 2.3. The additive model
restricts all three practice differences to be equal. The full model
relaxes those restrictions, using two more mean parameters. The
resulting reduction in residual SS, 64, is the interaction SS; its
mean square is 32, compared with the full model's residual mean square
of 3.3333. That is the interaction F of 9.6. The two extra df count
departures from a common practice difference, rather than two
additional groups of students. Both models use the same outcomes and
observations. The full model supplies the denominator, so unexplained
interaction is not put back into the error estimate used to test it.

The within-cells row of @tbl-ch04-two-way-anova holds only the variation of
students around their own cell means. @fig-ch04-ignore-practice shows
what happens to that row if practice is left out of the model, so that
each feedback condition becomes one group of eight students. This is
not an analysis the design calls for; it shows what the second factor
removes from the error term.

```{r}
#| label: fig-ch04-ignore-practice
#| fig-cap: "The 24 feedback-by-practice scores analysed two ways. Left: practice is left out, so each feedback condition is one group of eight; filled circles mark Massed and triangles Spaced students. Right: the six cells of the full model, with a bar at each cell mean. Leaving practice out moves the practice and interaction variation into the within-group term."
#| fig-alt: "Two panels on a 0 to 20 quiz-score axis. Left: three boxplots of eight scores each. Rereading runs from 4 to 8, its Massed and Spaced scores identical; Answer feedback runs from 6 to 14; Explanatory feedback runs from 8 to 20, its Massed scores 8 to 12 and Spaced scores 16 to 20 forming two clusters. Beneath it: within-group MS 10.4762 on 21 df, feedback F(2, 21) = 12.2182. Right: the same scores in six cells of four, each spanning four points around its cell mean: 6 and 6 for Rereading, 8 and 12 for Answer feedback, 10 and 18 for Explanatory feedback, Massed then Spaced. Beneath it: within-cell MS 3.3333 on 18 df, feedback F(2, 18) = 38.4."
#| fig-width: 7
#| fig-height: 3.9
#| out-width: 100%
short <- c("Rereading", "Answer", "Explanatory")
symbol <- ifelse(feedback_data$practice == "Massed", 16, 2)
par(mfrow = c(1, 2), mar = c(5.4, 4.2, 2.4, .8), cex.main = .95)
boxplot(score ~ feedback, data = feedback_data, ylim = c(0, 20), las = 1,
  names = short, col = "grey92", border = "grey40", cex.axis = .85,
  xlab = "", ylab = "Quiz score (0-20)",
  main = "Practice ignored: 3 groups of 8")
points(as.integer(feedback_data$feedback) +
    ifelse(feedback_data$practice == "Massed", -.13, .13),
  feedback_data$score, pch = symbol, cex = .8)
legend("topleft", c("Massed", "Spaced"), pch = c(16, 2), bty = "n",
  cex = .8)
mtext("Within-group MS 10.4762 on 21 df", side = 1, line = 2.6, cex = .8)
mtext("Feedback F(2, 21) = 12.2182", side = 1, line = 3.6, cex = .8)
cell <- interaction(feedback_data$practice, feedback_data$feedback)
cell_mean <- tapply(feedback_data$score, cell, mean)
plot(as.integer(cell), feedback_data$score, pch = symbol, cex = .8,
  xlim = c(.5, 6.5), ylim = c(0, 20), xaxt = "n", las = 1, xlab = "",
  ylab = "Quiz score (0-20)", main = "Full model: 6 cells of 4")
segments(1:6 - .3, cell_mean, 1:6 + .3, cell_mean, lwd = 2.5,
  col = "#9B4922")
axis(1, at = 1:6, labels = rep(c("M", "S"), 3), cex.axis = .8, padj = -.8)
axis(1, at = c(1.5, 3.5, 5.5), labels = short, tick = FALSE, line = .6,
  cex.axis = .85)
mtext("Within-cell MS 3.3333 on 18 df", side = 1, line = 3, cex = .8)
mtext("Feedback F(2, 18) = 38.4", side = 1, line = 4, cex = .8)
```

Leaving practice out changes neither the feedback means, 6, 10 and 14,
nor the feedback sum of squares, 256; it changes the error term. The
practice and interaction components stay inside the within-group
variation, which becomes $220=60+96+64$ on 21 df, so the within-group
mean square, 10.4762, is more than three times the full model's 3.3333,
and the feedback F falls from 38.4000 to 12.2182, $p=.0003$. The left
panel shows where the extra variation comes from: each group mixes two
schedules, so the Explanatory group splits into two clusters eight
points apart and its SD, 4.5981, is 2.7203 times Rereading's 1.6903,
whose two schedules do not differ. One pooled variance would
misdescribe all three groups. A factor that influences the outcome
belongs in the model, which is the second advantage of a crossed design
in Section 4.1.

Every displayed F divides its source mean square by the same
$MS_E=60/18=10/3$. The corresponding null hypotheses address
different features of the model:

\Needspace{12\baselineskip}

| Source | Null hypothesis | Comparison represented |
|:--|:--|:--|
| Feedback | $\mu_{R\cdot}=\mu_{A\cdot}=\mu_{E\cdot}$ | Equal feedback means, each averaging the two schedules equally |
| Practice | $\mu_{\cdot M}=\mu_{\cdot S}$ | Equal schedule means, each averaging the three feedback levels equally |
| Interaction | $(\alpha\beta)_{jk}=0$ for every cell | Each feedback comparison is constant across schedules, equivalently each schedule difference is the same across feedback levels |

: Three null hypotheses with their averaging rules. {#tbl-ch04-factorial-nulls tbl-colwidths="[17,37,46]"}

Under the conditions of Section 2.3, now applied to the six cells, the
interaction result $F(2,18)=9.6$, $p=.0015$, gives evidence against an
additive pattern. Both main effects also have small p-values, but they
remain averaged comparisons; reading them as “each factor works equally
well everywhere” would discard the interaction.

The clean component decomposition uses the balanced layout. With
unequal cell counts, the averaging target and the model comparison
need more care; different software sums of squares can address
different hypotheses. The [weighting reference](#app-ch04-weighting)
explains the first distinction without changing these data.

\Needspace{28\baselineskip}

[]{#ch04-justify-factorial}

::: {.callout-tip title="Justifying the analysis"}
- **Target:** the six cell means $\mu_{jk}$ of the feedback-by-practice
  layout for the 24 students, in quiz points. The three F tests ask
  whether the equally weighted feedback means are equal, whether the two
  equally weighted practice means are equal, and whether every
  interaction departure is zero.
- **Conditions:**
  - *Secured by design:* individual random assignment to the six cells,
    four different students in each, one score per student (Section
    4.1; independence as in Section 2.3).
  - *Model assumptions:* Normal errors with one common variance in every
    cell (Section 2.3), estimated by $MS_E=3.3333$ on 18 df. The model
    includes the interaction, so it does not assume additivity; reading
    a main effect as the effect at every level would.
  - *Choices and defaults:* the full model with interaction; equal cell
    weights for the marginal means; a .05 level fixed in advance; with
    four students in every cell, every sums-of-squares convention gives
    the same table.
- **Check:** every cell has residuals $-2$, $-1$, 1 and 2, so the six
  spreads agree by construction; residuals fanning out as the fitted
  mean grows would question the common variance, and four scores per
  cell say little about shape. The interaction row, $F(2,18)=9.6$,
  $p=.0015$, with the cell-mean plot, shows that additivity fails for
  these scores.
- **Bound:** under the model, the evidence is against an additive
  pattern. The main effects, $F(2,18)=38.4$ and $F(1,18)=28.8$, both
  $p<.0001$, describe equally weighted averages; the schedule
  differences within the three feedback conditions, 0, 4 and 8 points,
  show what those averages conceal. The scores are invented, so the
  result describes a calculation, not an educational finding.
:::

### Simple effects and a declared interaction contrast

The Answer-feedback schedule comparison, Spaced minus Massed, is a
**simple effect**: $12-8=4$ points. Under the full model its SE is
$\sqrt{(10/3)(1/4+1/4)}=1.291$, and with 18 error df a pointwise 95%
interval is $[1.2877,6.7123]$. The two means come from different
students, so this is an independent-cell calculation, not a paired one.

The study plan also declared one interaction comparison before any
scores were seen: whether spacing helps more under Answer feedback
than under Rereading. In model means the target is

$$
C=(\mu_{A,S}-\mu_{A,M})-(\mu_{R,S}-\mu_{R,M}),
$$

one **difference of differences**, estimated as $(12-8)-(6-6)=4$.
Its four cell means come from different students and carry weights 1,
$-1$, $-1$ and 1, so squaring the weights gives
$\widehat{SE}(\widehat C)=\sqrt{(10/3)(1/4+1/4+1/4+1/4)}=1.8257$.
Declared in advance as the study's one interaction question, the
contrast is a family of one, so its 95% interval uses the ordinary
multiplier $t_{.975;18}=2.1009$ and is $[0.1643,7.8357]$. It excludes
zero: under the model, spacing helps more under Answer feedback than
under Rereading, by between 0.1643 and 7.8357 quiz points. This interval
reads with the [factorial box](#ch04-justify-factorial) of Section 4.4,
with two changes: the target becomes this one contrast of four cell
means, and the choices add its declaration before the scores were seen.

As in Section 3.3, the history of the comparison sets its family. The
interaction test, $F(2,18)=9.6$, $p=.0015$, says that some departure
from additivity exists but not which one, and had this contrast been
picked after inspecting the plot, its interval would need a family that
covers every comparison the plot could have suggested, with a larger
multiplier than 2.1009. Each interval here likewise covers its one
stated comparison, not every simple effect or contrast a reader might
inspect afterward.

### A connected interpretation

Three rules order the reading of a factorial result: plot the cell
means first; examine the interaction before the main effects; and when
an interaction is present, report the cell means and conditional
comparisons, reading each main effect as a stated average. Applied to
this study, the schedule differences are 0, 4 and 8 quiz points for
Rereading, Answer feedback and Explanatory feedback; the interaction
test gives $F(2,18)=9.6$, $p=.0015$, under the stated conditions; and
the declared Answer-versus-Rereading difference of schedule differences
is 4 points, with 95% interval $[0.1643,7.8357]$. The feedback and
practice marginal means summarize explicit equal weights, and the cell
means show what those averages conceal.

These conclusions use the 24 students of the feedback-by-practice study
and 18 error df, whereas the one-way conclusions of Sections 2 and 3
concern the background-sound study. They are analyses of invented data,
not evidence that a teaching method succeeds in a population of
students.

### Three and four factors {#sec-ch04-multiway}

The students of Section 4.1 took their quiz one week after their final
practice. Suppose the study had also allocated 24 further students in
the same way, four to each feedback-by-practice cell, and quizzed them
right after their final practice. With quiz timing as a third crossed
factor, the 48 students fill twelve cells, a $3\times2\times2$ three-way
layout with four different students in each; we take all 48 to have been
randomly allocated. The added cell means are Rereading 9 (Massed) and 11
(Spaced), Answer feedback 12 and 14, and Explanatory feedback 16 and 18;
each added cell's four scores lie at its mean minus 2, minus 1, plus 1
and plus 2, the pattern of Section 4.1. The course data file
`feedback_practice_timing.csv` holds all 48 scores, with columns
`student`, `feedback`, `practice`, `timing` and `score`: students 1--24
are the one-week students of `feedback_practice.csv`, and students
25--48 were quizzed right after practice. @fig-ch04-three-way draws the
cell means for the two timings.

```{r}
#| label: fig-ch04-three-way
#| fig-cap: "Cell means of the three-way study. Each panel is the feedback-by-practice plot for one quiz timing, with each line joining means of different students. Right after practice the lines are parallel, so Answer's schedule difference minus Rereading's is 0; one week later they fan out, and the same difference of differences is 4. That change across timing is a three-way interaction."
#| fig-alt: "Two panels of cell means on a 0 to 20 quiz-score axis, with practice schedule on the horizontal axis and one line per feedback condition. Left panel, quiz right after practice: Rereading rises from 9 to 11, Answer feedback from 12 to 14 and Explanatory feedback from 16 to 18, three parallel lines; beneath it, Spaced minus Massed: 2, 2 and 2. Right panel, quiz one week later: Rereading stays at 6, Answer feedback rises from 8 to 12 and Explanatory feedback from 10 to 18, so the lines fan out; beneath it, Spaced minus Massed: 0, 4 and 8."
#| fig-width: 7
#| fig-height: 3.2
#| out-width: 100%
line_col <- c("#202020", "#2B6CA3", "#9B4922")
note <- c("Spaced minus Massed: 2, 2 and 2",
  "Spaced minus Massed: 0, 4 and 8")
par(mfrow = c(1, 2), mar = c(5.2, 4.2, 2.4, .8), cex.main = .95)
for (p in 1:2) {
  half <- timing_data[timing_data$timing == levels(timing_data$timing)[p], ]
  means <- with(half, tapply(score, list(feedback, practice), mean))
  matplot(1:2, t(means), type = "b", pch = c(16, 1, 17),
    lty = 1:3, col = line_col, ylim = c(0, 20), xlim = c(.8, 2.2),
    xaxt = "n", las = 1, xlab = "", ylab = "Mean quiz score",
    main = paste("Quiz", tolower(levels(timing_data$timing)[p])))
  axis(1, at = 1:2, labels = c("Massed", "Spaced"))
  mtext(note[p], side = 1, line = 2.6, cex = .8)
  if (p == 1) legend("bottomright", c("Rereading", "Answer", "Explanatory"),
    pch = c(16, 1, 17), lty = 1:3, col = line_col, bty = "n", cex = .8)
}
par(mfrow = c(1, 1))
```

Read each panel as @fig-ch04-cell-means was read. Right after practice
the three lines are parallel: spacing adds 2 points under every kind of
feedback, so Answer's schedule difference minus Rereading's is
$2-2=0$, and this panel shows no two-way interaction. One week later
the lines fan out: spacing adds 0, 4 and 8 points, and the same
difference of differences is $4-0=4$. The feedback-by-practice
interaction therefore differs between the two timings, and that change
is a **three-way interaction**: a two-way pattern that changes across
the levels of a third factor. One three-way contrast is the change in
the difference of differences, $4-0=4$ points, a difference of
differences of differences.

The three-way model adds to the two-way model of Section 4.4 a timing
effect and every interaction that contains timing. With $l$ indexing
timing,

$$
Y_{ijkl}=\mu+\alpha_j+\beta_k+\gamma_l+(\alpha\beta)_{jk}
+(\alpha\gamma)_{jl}+(\beta\gamma)_{kl}+(\alpha\beta\gamma)_{jkl}
+\varepsilon_{ijkl},
$$

and each set of effects becomes one row of the ANOVA table.
@tbl-ch04-two-three-way sets the two-way analysis of the 24 one-week
students beside the three-way analysis of all 48, with A for feedback and B
for practice, as in Section 4.4, and C for timing.

\Needspace{16\baselineskip}

+------------------+----------------------+------------------------------------------------+
| Source           | Two-way table        | Three-way table                                |
|                  +------+-------+-------+------+-----------+-----------+-------+---------+
|                  | df   | SS    | F     | df   | SS        | MS        | F     | p       |
+:=================+:====:+======:+======:+:====:+==========:+==========:+======:+========:+
| Feedback (A)     | 2    | 256   | 38.4  | 2    | 450.6667  | 225.3333  | 67.6  | <.0001  |
+------------------+------+-------+-------+------+-----------+-----------+-------+---------+
| Practice (B)     | 1    | 96    | 28.8  | 1    | 108.0000  | 108.0000  | 32.4  | <.0001  |
+------------------+------+-------+-------+------+-----------+-----------+-------+---------+
| Timing (C)       | —    | —     | —     | 1    | 133.3333  | 133.3333  | 40.0  | <.0001  |
+------------------+------+-------+-------+------+-----------+-----------+-------+---------+
| A × B            | 2    | 64    | 9.6   | 2    | 32.0000   | 16.0000   | 4.8   | .0142   |
+------------------+------+-------+-------+------+-----------+-----------+-------+---------+
| A × C            | —    | —     | —     | 2    | 2.6667    | 1.3333    | 0.4   | .6733   |
+------------------+------+-------+-------+------+-----------+-----------+-------+---------+
| B × C            | —    | —     | —     | 1    | 12.0000   | 12.0000   | 3.6   | .0658   |
+------------------+------+-------+-------+------+-----------+-----------+-------+---------+
| A × B × C        | —    | —     | —     | 2    | 32.0000   | 16.0000   | 4.8   | .0142   |
+------------------+------+-------+-------+------+-----------+-----------+-------+---------+
| Within cells     | 18   | 60    |       | 36   | 120.0000  | 3.3333    |       |         |
+------------------+------+-------+-------+------+-----------+-----------+-------+---------+
| Total            | 23   | 476   |       | 47   | 890.6667  |           |       |         |
+------------------+------+-------+-------+------+-----------+-----------+-------+---------+

: The two-way ANOVA of the 24 students quizzed one week after practice beside the three-way ANOVA of all 48 students; a dash marks a row that the two-way table does not have. {#tbl-ch04-two-three-way}

Four rows are new. The timing row, C, compares the two timings averaged
over the six feedback-by-practice cells: scores right after practice
average 13.3333 against 10 a week later, $F(1,36)=40$. The other three
new rows contain timing: A × C asks whether the feedback comparison
changes with timing, and hardly finds it does, $F(2,36)=0.4$; B × C
compares the average spacing benefit, 2 points right after practice
against 4 a week later, $F(1,36)=3.6$, $p=.0658$; and A × B × C, the
three-way row, compares the whole feedback-by-practice pattern between
the timings.

The effect rows that the two tables share change their meaning: in the
three-way table each averages over the two timings and rests on all 48
students. The feedback row compares feedback means of 8.0, 11.5 and
15.5, each averaged over timing and resting on 16 students; the
one-week means 6, 10 and 14 rest on 8. The averaged means spread
slightly less, but twice as many students stand behind each one, so the
feedback sum of squares is larger. The practice row compares the
spacing benefit averaged over timing, 3 points, the average of the 2
right after practice and the 4 a week later, with 24 students per
schedule instead of 12. Because the within-cell mean square is 3.3333
in both tables, each F follows its row's mean square: the practice F
rises from 28.8 to 32.4 although the spacing benefit it tests falls
from 4 points to 3, because each schedule mean now rests on twice as
many students.

In the three-way table the A × B row is the feedback-by-practice
interaction averaged over the two timings: its sum of squares falls from 64 in the
one-week study to 32, because right after practice the lines are
parallel. The three-way row, $F(2,36)=4.8$, $p=.0142$, says that the
feedback-by-practice pattern differs between the timings, so the
averaged row describes neither panel: averaged over timing, Answer's
schedule difference minus Rereading's is 2 points, against 0 right after
practice and 4 a week later. A multi-way table is therefore read from
the highest-order row down. Section 4.4 read the interaction row before
the main effects, because a main effect averages over a pattern that the
interaction says is changing; one level up the rule is the same. With a
three-way interaction present, report the two-way pattern within each
level of the third factor, and read a lower-order row alone only when
the design and the evidence make the higher-order interactions
negligible [@kutner2005].

The within-cells row keeps its meaning. Its df grow with the students,
from $24-6=18$ to $48-12=36$, while its mean square stays 3.3333,
because every cell's residuals are $-2$, $-1$, 1 and 2 in both
analyses; in each table every F divides its row's mean square by that
one within-cell estimate.

A fourth factor extends the table in the same way. Suppose the study
also crossed the reading medium, paper or screen, with the twelve cells,
each student reading the practice material in one medium, again with
four different students in each of now 24 cells, 96 students in all. A
**four-way interaction** would mean that the three-way pattern itself
differs between the two media: for example, that the pattern of
@fig-ch04-three-way holds for students who read on paper, while for
students who read on screen the lines stay parallel at both timings.
@tbl-ch04-multiway lists the rows of each design's table, with D for
the medium.

\Needspace{16\baselineskip}

| Rows of the table | Two-way, A\ ×\ B | Three-way, A\ ×\ B\ ×\ C | Four-way, A\ ×\ B\ ×\ C\ ×\ D, with the df of each row |
|:--|:--|:--|:--|
| Main effects | A, B | A, B, C | A\ (2), B\ (1), C\ (1), D\ (1) |
| Two-way interactions | A\ ×\ B | A\ ×\ B, A\ ×\ C, B\ ×\ C | A\ ×\ B\ (2), A\ ×\ C\ (2), A\ ×\ D\ (2), B\ ×\ C\ (1), B\ ×\ D\ (1), C\ ×\ D\ (1) |
| Three-way interactions | --- | A\ ×\ B\ ×\ C | A\ ×\ B\ ×\ C\ (2), A\ ×\ B\ ×\ D\ (2), A\ ×\ C\ ×\ D\ (2), B\ ×\ C\ ×\ D\ (1) |
| Four-way interaction | --- | --- | A\ ×\ B\ ×\ C\ ×\ D\ (2) |
| Source rows, $2^k-1$ | 3 | 7 | 15, with 23 df in all |
| Cells; students at four per cell | 6; 24 | 12; 48 | 24; 96 |
| Within-cells df, students minus cells | 18 | 36 | 72 |

: The rows of the two-way, three-way and four-way tables as quiz timing and then reading medium are added to the feedback-by-practice design, four students per cell, with the df of each four-way row. {#tbl-ch04-multiway tbl-colwidths="[25,13,21,41]"}

A design with $k$ factors has $k$ main effects and one interaction for
every pair, triple and larger set of factors, $2^k-1$ rows of effects
in all: 3, 7 and 15. A row's df multiply the df of its factors, so every
row that contains feedback has 2 df and every other row 1; the effect
df of the four-way table add to the cells minus one, 23, and its
within-cells df are the students minus the cells, 72. Each added
two-level factor doubles the cells, so four students per cell means 24,
then 48, then 96 students.

Added factors have costs. A three-way contrast combines eight cell
means where a difference of differences combines four, so with the same
students per cell and the same error variance its SE is $\sqrt2$, about
1.4, times as large, and matching the precision of a difference of
differences takes twice as many students per cell. The seven F tests of
a three-factor table, each at .05, also form a family in the sense of
[Section 3.3](#sec-ch04-families): for seven independent tests whose
null hypotheses are all true, the chance of at least one false
rejection is .3017, where the upper line of @fig-ch04-familywise stands
at $m=7$, and the fifteen tests of a four-factor table raise it
further. More cells also make unequal or empty cells more likely:
unequal counts make each marginal mean depend on its weights
([Appendix B.1](#app-ch04-weighting)), and an empty cell leaves some
interactions without an estimate.

A study with several factors therefore declares in advance the few
comparisons it was built for, such as whether the spacing benefit under
Answer feedback grows between the two timings, and reads the other rows
and cells as description. Twelve cells allow 66 pairwise comparisons and
many more contrasts, so a comparison found by searching them needs a
family that covers the search, as Section 3.3 showed for three groups.

## Checking the Conditions {#sec-ch04-conditions}

### What the design and the data can show {#sec-ch04-deeper-checks}

Section 2.3 stated the conditions where the F test first relied on
them, and each *Justifying the analysis* box recorded what the data
could check; @tbl-ch04-conditions gathers them with what the design or
the data can show about each and the response when one fails. A
condition is judged from the design, subject knowledge and displays such
as @fig-ch04-residuals together, not from a sequence of Normality or
variance tests, which have least power in the small samples where the
choice matters most [@kutner2005].

\Needspace{20\baselineskip}

| Condition | What the design or the data can show | Response if it fails |
|:--|:--|:--|
| Independent observations | Only the design record: individual assignment, individual sessions, one score per student. No display of the scores can establish it | An analysis that represents the dependence: comparisons within blocks, as in Chapter 3's paired schools (Section 2.3), or models for classes and repeated measurements. Treating dependent rows as independent misstates every SE; positive clustering makes it too small, by about the square root of the design effect of Chapter 2, Section 4.4 |
| Roughly Normal errors around the group or cell means | A residual display and a Normal quantile plot of the residuals, not of the pooled scores; small groups reveal only pronounced departures | With moderate skew and larger groups the F reference remains a reasonable approximation; with small groups and strong skew or a dominating value, disclose the weaker approximation and examine the influential scores |
| One common variance | The group or cell SDs beside the group sizes; a fan in the residuals as the fitted mean grows | Welch's one-way ANOVA, chosen before the analysis (Section 5.2); for a contrast, an SE built from its groups' own variances, which for a pair is Chapter 3's Welch t |
| Additivity, when a main effect is read as the effect at every level | The interaction row and the cell-mean plot | Report cell means and simple effects, and read each main effect as a stated average (Section 4) |

: Conditions of the chapter's analyses, what can show whether each holds, and the response if it fails. {#tbl-ch04-conditions tbl-colwidths="[22,38,40]"}

```{r}
#| label: fig-ch04-residuals
#| fig-cap: "Residuals of the one-way fit to the background-sound scores. Left: each student's residual, the score minus the group mean, by condition, jittered sideways so that equal values stay visible. Right: the thirty residuals against Normal quantiles, with a reference line."
#| fig-alt: "Left panel: three vertical strips of ten points, for Silence, Instrumental music and Music with lyrics, each spread from -4 to 4 around a dashed line at zero, with similar spread in every group. Right panel: a Normal quantile plot of the thirty residuals; the points rise in whole-number steps and follow a dashed reference line, the residuals of -4 and of 4 at the two ends sitting flatter than the line, and no point is far from it."
#| fig-width: 7
#| fig-height: 3.3
#| out-width: 100%
res <- residuals(one_way)
par(mfrow = c(1, 2), mar = c(4, 4.2, 2.2, .8), cex.main = .95)
set.seed(12345)
stripchart(res ~ sound_data$sound, vertical = TRUE, method = "jitter",
  jitter = .12, pch = 16, cex = .8, las = 1, cex.axis = .85,
  group.names = c("Silence", "Instrumental", "Lyrics"),
  ylab = "Residual (score minus group mean)",
  main = "Residuals by condition")
abline(h = 0, lty = 2)
qqnorm(res, pch = 16, cex = .8, las = 1, main = "Normal quantile plot",
  xlab = "Normal quantile", ylab = "Residual")
qqline(res, lty = 2)
par(mfrow = c(1, 1))
```

For the background-sound study the residuals spread similarly around
zero in all three groups, each from $-4$ to 4, and in the quantile plot
they follow the line, with no outlying value; the residuals of $-4$ and
4 at the two ends sit a little flatter than a Normal tail. Thirty
residuals can show only pronounced departures, so the display is
consistent with the model rather than proof of it, and independence
still rests on the design. The factorial scores were built with
residuals $-2$, $-1$, 1 and 2 in every cell, a teaching construction
that no display could treat as evidence.

Neither invented dataset has missing scores; a real report compares
assigned and analysed counts and explains known losses, because missing
outcomes can change who contributes to a comparison and call for
assumptions beyond the ANOVA model. Keep three claims distinct in every
report: the observed means and sums of squares describe these records;
intervals and tests express uncertainty under a stated model; and
causal or population interpretations need the support that Section 1.2
describes.

### When spreads differ: Welch's one-way ANOVA {#sec-ch04-welch}

When a common variance is doubtful, **Welch's one-way ANOVA** is the
counterpart of Welch's unequal-variance t test in Chapter 3,
Section 5.4. It keeps each group's own variance instead of pooling them
into $MS_W$, weights each group mean by the precision implied by its own
variance and size, $n_j/s_j^2$, and uses an approximate F reference
whose denominator df are estimated from the group variances
[@welch1951]; with two groups its F is the square of Welch's t.
[]{#ch04-justify-welch}It reads with the
[one-way box](#ch04-justify-oneway) of Section 2.3, with two changes:
the model assumption of one common variance is dropped, so each group
mean needs only a near-Normal sampling distribution, and the calibration
changes with it. For the background-sound study it gives
$F(2,17.9561)=5.5144$, $p=.0136$, beside the classical
$F(2,27)=6.0674$, $p=.0067$: with similar spreads and equal group sizes
the two calibrations lead to the same conclusion at .05, although
Welch's p is about twice as large. Its F is slightly smaller, and its
denominator df fall from 27 to 17.9561 because each weight rests on a
variance estimated from its own ten scores. Welch's
method has the scope of one factor: clustering still needs an analysis
that represents the design, and crossed factors with unequal cell
spreads take each stated contrast's SE from the cells' own variances.

The classical F test applies one pooled variance to every group mean.
Chapter 3, Section 5.4, showed for two groups that a pooled variance
misstates the variance of a small group's mean when spreads differ.
@tbl-ch04-welch-consequence shows what this does with three groups: each
row simulates 50,000 studies of three Normal groups with equal means and
the stated SDs, so every rejection at level .05 rejects a true null.

```{r}
#| label: ch04-welch-sim
#| include: false
# Recomputes the rates printed in tbl-ch04-welch-consequence and stops the
# render if any printed value no longer matches.
simulate_design <- function(n, s, B = 50000) {
  set.seed(12345)
  J <- length(n); N <- sum(n)
  means <- vars <- matrix(NA, nrow = B, ncol = J)
  for (j in 1:J) {
    x <- matrix(rnorm(B * n[j], mean = 0, sd = s[j]), nrow = n[j])
    means[, j] <- colMeans(x)
    vars[, j] <- apply(x, 2, var)
  }
  grand <- drop(means %*% n) / N
  MS_B <- drop((means - grand)^2 %*% n) / (J - 1)
  MS_W <- drop(vars %*% (n - 1)) / (N - J)
  w <- sweep(1 / vars, 2, n, "*")
  W <- rowSums(w)
  weighted_mean <- rowSums(w * means) / W
  S <- drop((1 - w / W)^2 %*% (1 / (n - 1)))
  F_welch <- rowSums(w * (means - weighted_mean)^2) / (J - 1) /
    (1 + 2 * (J - 2) / (J^2 - 1) * S)
  nu <- (J^2 - 1) / (3 * S)
  c(classical = mean(MS_B / MS_W > qf(.95, df1 = J - 1, df2 = N - J)),
    welch = mean(F_welch > qf(.95, df1 = J - 1, df2 = nu)))
}
welch_rates <- rbind(simulate_design(c(10, 20, 40), c(8, 4, 4)),
  simulate_design(c(10, 20, 40), c(4, 4, 8)),
  simulate_design(c(20, 20, 20), c(8, 4, 4)))
stopifnot(sprintf("%.4f", welch_rates) ==
  c("0.1572", "0.0126", "0.0610", "0.0511", "0.0499", "0.0502"),
  sprintf("%.4f", 1504 / 67) == "22.4478",
  sprintf("%.4f", 1504 / 67 / 10) == "2.2448")
```

\Needspace{14\baselineskip}

| Group sizes | Population SDs | Classical F | Welch's one-way ANOVA |
|:--|:--|--:|--:|
| 10, 20 and 40 | 8, 4 and 4: the smallest group varies most | .1572 | .0511 |
| 10, 20 and 40 | 4, 4 and 8: the largest group varies most | .0126 | .0499 |
| 20, 20 and 20 | 8, 4 and 4 | .0610 | .0502 |

: Rejection rates of a true null of equal means at nominal level .05, from 50,000 simulated studies per design. {#tbl-ch04-welch-consequence tbl-colwidths="[18,42,20,20]"}

When the smallest group varies most, the variance that pooling targets,
22.4478, gives that group's mean a variance of 2.2448 instead of the
6.4000 its own SD implies, and the classical F rejects the true null in
.1572 of the studies, more than three times the nominal .05; when the
largest group varies most it rejects in only .0126, and with equal group
sizes it rises only to .0610. Welch's one-way ANOVA stays near .05 in all
three designs, which is why unequal group sizes with spreads that may
differ call for it.

As in Chapter 3, choose between the common-variance and Welch
calibrations before the analysis, from the design and subject knowledge,
rather than from a preliminary variance test or from whichever p-value
is smaller [@delacre2017; @delacre2019]. Many statistical programs
report Welch's t by default for two groups; that default is Chapter 3's
last t procedure, not the equal-variance test whose square equals the
classical F of Section 2.4. Chapter 6 runs the Welch alternative beside
the common-variance calculation.

## Reading Bridge {#sec-ch04-bridge}

Comparing several means extends the estimate-and-SE reasoning of
Chapter 3. ANOVA separates sources of variation to ask an omnibus
question; a contrast states a narrower question and its uncertainty.
Main effects average across factor levels, while cell means and
simple effects show the conditional pattern, and a third factor adds
rows to the table and can change a two-way pattern itself. A joint
interaction test and one declared interaction contrast answer different
questions, and the history of a comparison sets the family its interval
must cover.

The same group-mean model can be written using indicator variables,
as Chapter 6 does in R. A measured baseline can also enter an adjusted
model, called analysis of covariance (ANCOVA), which Chapter 9,
Section 2, develops as regression after Chapters 7--8 develop the
regression mathematics, several controls and standardized
predictions. Baseline adjustment accounts for a measured
source of score variation and compares fitted outcomes at a stated common
baseline value, or averages them over the same stated baseline distribution.
The groups need adequate overlap in those baseline values, often called
common support, for those comparisons to be informed by observations.
Adjustment is not a general cure for confounding or missing outcomes.

**Blocking** makes condition comparisons within groups of similar
units. Chapter 3, Section 2.3, illustrates it with paired schools: in
each of twelve schools one class is assigned to each format, so the
uncertainty follows twelve independent within-school comparisons rather
than 480 students. Sampling strata, which organize selection into a
sample, and allocation blocks, which organize treatment comparisons,
have different roles. The three sound conditions and the
feedback-by-practice cells of this chapter instead contain distinct,
individually assigned students. Repeated measurements and classroom
clustering change the covariance structure and the relevant error term;
the shape of a data table cannot settle these design questions.
Chapters 5--6 now make the statistical foundations executable while
keeping the question, records and interpretation beside each
calculation.

### Key terms {.unnumbered}

Use this reference to revisit the concepts principally explained here.
The Part One Concept Reference collects earlier terms, including
estimand, SE and confidence interval.

```{=latex}
\small
```

| Term and common alternatives | Meaning |
|:--|:--|
| Factor (grouping factor) | In ANOVA, a categorical variable defining groups or conditions; its categories are levels. This meaning differs from a latent factor used to model shared indicator variation. |
| Analysis of variance (ANOVA) | A method for studying mean differences through sources of variation. One-way ANOVA uses one grouping factor, and multi-way (factorial) ANOVA uses two or more crossed factors, as two-way and three-way ANOVA do; each factor can have several levels. |
| Fitted value | The value a fitted model assigns to an observed case. In a one-way group-mean model, it is that case's group mean. |
| Residual | An observed outcome minus its fitted value. A residual is calculated from the fitted model; an error is the corresponding unknown departure in the population model. |
| Mean square (MS) | A sum of squares divided by its degrees of freedom. In classical ANOVA, the within-group mean square estimates the common error variance under the model's conditions. |
| F statistic (F ratio) | In classical ANOVA, a ratio of the mean square for a tested source to the appropriate error mean square. Its null F reference depends on the model and both degrees of freedom. |
| Omnibus test | A test of one joint null hypothesis about a whole pattern, such as equality of all group means. Rejecting it does not identify which comparisons differ or how large they are. |
| Eta squared | The proportion of total observed variation attributed to a specified source in an ANOVA decomposition. In one-way ANOVA, it is between-group divided by total sum of squares. |
| Contrast (linear contrast) | A weighted combination of means whose weights sum to zero, expressing a specific comparison. The means, weights, and target population determine the quantity compared. |
| Familywise error rate (FWER) | The probability of at least one Type I error among a defined family of tests. Procedures such as Bonferroni allocate the error allowance across that family. |
| Simultaneous confidence intervals | A set of intervals calibrated so that, under their conditions, all of them cover their estimands together at the stated familywise rate; Bonferroni, Tukey, and Dunnett are different calibrations for different families. |
| Cell mean and marginal mean | In a two-factor layout, a cell mean is the mean of the cases sharing one level of each factor; a marginal mean averages the cell means across the levels of the other factor, with stated weights. |
| Main effect | A comparison across one factor's levels averaged over the other factor's levels, using stated weights. It remains an average when the conditional pattern differs across levels. |
| Interaction | A pattern in which a comparison across one factor's levels changes with the other factor's level. A difference of differences expresses one such comparison. |
| Simple effect | A comparison across one factor's levels at a specified level of the other factor. It describes a conditional comparison within the factorial design. |
| Three-way interaction (higher-order interaction) | A pattern in which a two-way interaction changes across the levels of a third factor. Each added factor multiplies the cells and adds interactions, so the highest-order interaction is read first and a few comparisons are declared in advance. |

: Selected concepts for group comparisons and ANOVA. {#tbl-ch04-key-terms tbl-colwidths="[28,72]"}

```{=latex}
\normalsize
```

### Further reading {.unnumbered}

- *Introduction to Modern Statistics* develops the transition from
  grouped observations to [inference for many means](https://openintro-ims.netlify.app/inference-many-means)
  [@cetinkaya2024].
- [Penn State STAT 200, Lesson 10](https://online.stat.psu.edu/stat200/Lesson10)
  provides an introductory ANOVA treatment; its
  [factorial-design lesson](https://online.stat.psu.edu/stat503/Lesson05)
  develops main effects and interaction.
- The NIST handbook explains [contrasts and their SEs](https://www.itl.nist.gov/div898/handbook/prc/section4/prc436.htm)
  and [Bonferroni simultaneous intervals](https://www.itl.nist.gov/div898/handbook/prc/section4/prc473.htm).

## Exercises {.unnumbered}

The printed scores and the supplied tables and results support each
interpretation, with a calculator for the few calculations that the
reasoning needs. Retain full precision until reporting calculated
results to four decimal places. State the relevant records, conditions
and comparison direction in each interpretation.

<!-- prompt-id: CH04-E01 -->
### Exercise 1: Read grouped observations (Foundational) {.unnumbered #ex-ch04-01}

Use the thirty background-sound scores of the invented study, printed
in Section 1.

1. The group means are 12, 10 and 8, and the sample SDs are 2.7080,
   2.4037 and 2.5820. Use them with the ranges to compare the centers,
   spreads and overlap of the three groups, and say what an SD of
   about 2.5 points describes for these students.
2. Suppose every student's distance from their group mean were
   doubled, so that the three means stay at 12, 10 and 8. Before
   calculating, say whether the evidence against equal means would
   become stronger or weaker, and why. Then compare the supplied
   results, $F(2,27)=6.0674$, $p=.0067$ for the actual scores and
   $F(2,27)=1.5169$, $p=.2375$ for the doubled version, and explain
   why F falls to one quarter.
3. The three SDs are similar. Explain why that similarity does not
   establish that a common-variance Normal model will hold in a real
   study.
4. Identify the assignment and measurement units stipulated in this
   invented study. Explain what the scores alone cannot establish
   about causal interpretation or generalization.

<!-- prompt-id: CH04-E02 -->
### Exercise 2: Read the one-way table (Foundational) {.unnumbered #ex-ch04-02}

Use Exercise 1's scores. There are $J=3$ groups and $N=30$ students,
and the classical analysis assumes independent Normal errors with a
common variance. Statistical software reports this table:

| Source | SS | df | MS | F | p |
|:--|--:|--:|--:|--:|--:|
| Sound condition | 80 | 2 | 40.0000 | 6.0674 | .0067 |
| Within groups | 178 | 27 | 6.5926 | | |
| Total | 258 | 29 | | | |

1. Explain what the between-groups df, 2, and the within-groups df,
   27, each count, and why they add to the total df, 29. Name the
   reference distribution of F under the null.
2. Explain why the Instrumental-music group contributes nothing to
   $SS_B$ although its scores vary. Say what $MS_W$ estimates and why
   it is smaller than the total sample variance, $258/29=8.8966$.
3. State the omnibus null. Interpret $p=.0067$ at $\alpha=.05$,
   including the model conditions and what the conclusion leaves
   unanswered about particular comparisons.
4. Without refitting, predict how $SS_B$, $SS_W$ and F would change
   if (a) every Music-with-lyrics score rose by 2 points, or (b) all
   thirty scores rose by 5 points. Then check prediction (a) against
   the supplied result $SS_B=26.6667$, $F(2,27)=2.0225$, $p=.1519$,
   and say what change (a) makes to the conclusion of part 3.

<!-- prompt-id: CH04-E03 -->
### Exercise 3: Define and estimate contrasts (Foundational) {.unnumbered #ex-ch04-03}

The sound-condition means are 12, 10 and 8 in the order Silence,
Instrumental music, Music with lyrics. Each group has ten students and
the common-variance estimate is $MS_W=178/27$, with 27 residual df.

1. Identify which weight vectors define contrasts: $(-1,1,0)$,
   $(-1,0,1)$, $(-1,1/2,1/2)$ and $(1/3,1/3,1/3)$. State the
   ordinary-language question represented by each contrast.
2. The vectors $(-1,1,0)$, $(-1,0,1)$ and $(-1,1/2,1/2)$ give
   estimates $-2$, $-4$ and $-3$ with SEs 1.1483, 1.1483 and 0.9944,
   each from $\sqrt{MS_W\sum_j c_j^2/n_j}$. Explain, through the squared
   weights, why the equal-average contrast has the smallest SE, and
   why all three SEs use the full three-group error estimate.
3. Explain why the equal-average contrast cannot replace both separate
   music-versus-Silence comparisons when both are required.
4. For a weighting illustration only, suppose the Instrumental-music
   and Music-with-lyrics means stay at 10 and 8 but their counts are 10
   and 30. The Silence mean stays at 12. Compare the equally weighted
   music-average difference with the student-weighted difference. State
   the question answered by each. These illustrative counts do not
   change the chapter's actual balanced dataset.

<!-- prompt-id: CH04-E04 -->
### Exercise 4: Interpret comparison families (Integrative) {.unnumbered #ex-ch04-04}

In the invented background-sound study, thirty first-year students were
individually and randomly assigned to three sound conditions, ten to a
condition, and each answered the 0--20 comprehension quiz once after
reading the same passage. The group means are 12 for Silence, 10 for
Instrumental music and 8 for Music with lyrics. Under the independent
common-variance Normal model, $MS_W=178/27$ on 27 error df, so each
difference between two of these means has
$SE=\sqrt{(178/27)(1/10+1/10)}=1.148267616$; this SE and the critical
values below carry nine decimals so that the printed four-decimal
limits can be reproduced by hand. Two reports give the same estimate,
$-4$ comprehension points for Music with lyrics minus Silence, but the
comparison has a different history in each:

- **Report A.** Before any scores were seen, the study plan declared
  a family of two comparisons, each music condition against Silence:
  Instrumental music minus Silence and Music with lyrics minus Silence,
  with 95% simultaneous coverage for the pair.
- **Report B.** No comparison was declared in advance. After
  inspecting the three means, an analyst reports the largest pairwise
  difference, Music with lyrics minus Silence.

The Bonferroni critical value for two comparisons is 2.373417201, and
Report A's 95% simultaneous intervals are $[-4.7253,0.7253]$ for
Instrumental music minus Silence and $[-6.7253,-1.2747]$ for Music
with lyrics minus Silence. Tukey's 95% all-pairs intervals are
$[-4.847,0.847]$ for Instrumental music minus Silence,
$[-6.847,-1.153]$ for Music with lyrics minus Silence and
$[-4.847,0.847]$ for Music with lyrics minus Instrumental music.

1. Before calculating, state what differs between the two reports and
   what effect the difference has on the family of comparisons that
   the uncertainty statement must cover. Give your reasons.
2. For Report A, explain how dividing a .05 familywise error
   allowance between the two intervals and their tails leads to the
   quantile .9875 behind its critical value. Interpret the two
   intervals together, and explain why an interval containing zero
   does not establish equivalence.
3. For Report B, identify the family from which the comparison was
   selected and the interval warranted for Music with lyrics minus
   Silence. Compare it with Report A's interval for the same
   comparison, say which one describes the uncertainty under Report
   B's history and why, and state what each history lets you conclude
   about the comparison. A Bonferroni bound over Report B's family
   uses $t_{1-.05/6;27}=2.552458806$: calculate its half-width and
   explain why it exceeds both Report A's half-width, 2.7253, and
   Tukey's, 2.8470.
4. Report A's two comparisons share the Silence mean. Explain why
   this creates dependence between the two estimates and why it does
   not invalidate the Bonferroni bound (Appendix B.3).

\Needspace{16\baselineskip}

<!-- prompt-id: CH04-E05 -->
### Exercise 5: Read the factorial pattern (Foundational) {.unnumbered #ex-ch04-05}

Use the 24 feedback-by-practice scores of the invented study, printed
in Section 4.1. There are four different students per cell,
individually allocated to the six combinations.

1. Identify the outcome, the two factors and their levels, the
   assignment unit, the number of cells and the meaning of balance.
2. Leave practice aside for a moment, so that each feedback condition
   is one group of eight students. The three feedback means are then
   6, 10 and 14, and the two practice means 8 and 12; explain why they
   equal the equally weighted marginal means here. Analysed this way,
   the feedback groups have SDs 1.6903, 2.7255 and 4.5981, a within-group mean
   square of 10.4762 on 21 df and $F(2,21)=12.2182$, whereas the full
   two-way model has $MS_E=3.3333$ on 18 df and feedback
   $F(2,18)=38.4000$. Explain where the extra within-group variation
   comes from and why the Explanatory group's SD is the largest.
3. The additive prediction for Explanatory-Spaced, the grand mean
   plus the Explanatory-feedback and Spaced-practice effects, is
   16, and the observed cell mean is 18. Explain what the
   interaction departure of 2 says about spacing under
   Explanatory feedback, and why the other departures in its row and
   column must balance it.
4. The Spaced-minus-Massed differences are 0, 4 and 8 points for
   Rereading, Answer feedback and Explanatory feedback, so Answer's
   schedule difference minus Rereading's is 4 points. Explain how
   these values appear in the cell-mean plot, and what the plot would
   show if the three differences were equal.
5. Explain why the four-cell difference in part 4 is one interaction
   contrast, while the full three-by-two interaction has two df.
   Explain why the connected lines do not represent paired students.

<!-- prompt-id: CH04-E06 -->
### Exercise 6: Distinguish joint and declared results (Integrative) {.unnumbered #ex-ch04-06}

For the full feedback-by-practice model, the ANOVA table gives
feedback $F(2,18)=38.4$, practice $F(1,18)=28.8$ and interaction
$F(2,18)=9.6$, with p-values below .0001, below .0001 and .0015,
and error mean square $MS_E=60/18=10/3$ on 18 df. Assume independent
common-variance Normal errors. The critical value in parts 3 and 4
carries extra decimals so that the printed limits can be reproduced by
hand.

1. State in words the joint null hypothesis that each of the three
   rows tests, the order in which to read the rows, and the reason
   for that order.
2. The schedule marginal means are 8 and 12, but the conditional
   schedule differences are 0, 4 and 8. Explain what the practice
   main effect averages and why it does not describe every condition.
3. Answer's 4-point schedule difference has $SE=\sqrt{5/3}$, and
   with $t_{.975,18}=2.10092204$ its pointwise 95% interval is
   $[1.2877,6.7123]$. State the family limitation of this
   single-comparison interval.
4. The study plan declared one interaction comparison before any
   scores were seen: Answer's schedule difference minus Rereading's,
   estimated as 4. Its SE from four independent cell means of size 4
   and $MS_E=10/3$ is 1.8257, and with the same critical value its 95%
   interval is $[0.1643,7.8357]$. Explain why this single-comparison
   interval is warranted for the declared contrast, and what it says
   about spacing under the two kinds of feedback.
5. Explain why the interaction test's $p=.0015$ does not by itself
   establish the direction in part 4, and why the interval of part 4
   would need a wider family had the same contrast been picked after
   the plot was inspected.

<!-- prompt-id: CH04-E07 -->
### Exercise 7: Repair a claim about teaching (Integrative) {.unnumbered #ex-ch04-07}

Repair this claim about the invented study:

> Spaced practice improves every student's learning by four points,
> whatever feedback is provided. The significant interaction proves
> that Answer feedback benefits more from spacing than Rereading,
> and the results apply to all students.

All 24 quiz scores are observed. The stipulated design allocates
four different students to each cell. The equal-weight practice
main effect is 4 points; conditional schedule differences
are 0, 4 and 8. The interaction test gives
$F(2,18)=9.6$, $p=.0015$. The study plan declared one interaction
comparison in advance, Answer's schedule difference minus
Rereading's; it is 4, with 95% interval $[0.1643,7.8357]$.

1. Identify at least four words or claims that exceed the
   supplied statistical and design information.
2. Write a bounded replacement retaining the conditional pattern,
   the joint interaction result, the declared estimate and interval,
   the 0--20 quiz scale, and the invented nature of the example.
3. Explain what random allocation could support in a real study
   and what it would not establish about population sampling or
   individual treatment effects.
4. Explain how class-level assignment, repeated observations or
   missing outcomes would affect the information needed for the
   analysis. Do not invent losses in the present dataset. State
   why four observations per cell limit diagnostic reassurance.

<!-- prompt-id: CH04-E08 -->
### Exercise 8: Relate two-group ANOVA to the t test (Integrative) {.unnumbered #ex-ch04-08}

Use the Silence and Music-with-lyrics scores printed in Section 1,
which belong to twenty different students. For these two groups alone
the pooled within-group variance is 7 on 18 df. Silence minus Music
with lyrics is 4 points, with $SE=\sqrt{7(1/10+1/10)}=1.1832$ and
pooled $t=3.3806$ on 18 df, and a one-way ANOVA of the same two groups
gives $F(1,18)=11.4286$; both two-sided p-values are .0033.

1. Explain why F equals $t^2$ here and why the two p-values agree.
2. Say what F loses that t and the signed difference keep.
3. The three-group analysis of Section 2.3 estimates the error
   variance as 6.5926 on 27 df. Explain why the two-group fit gives 7
   on 18 df instead, and which question each fit answers.

<!-- prompt-id: CH04-E09 -->
### Exercise 9: Match the analysis to the design (Integrative) {.unnumbered #ex-ch04-09}

In the background-sound study, the Silence and Music-with-lyrics
scores printed in Section 1 belong to twenty different students, each
randomly assigned to one condition.

1. Suppose instead that each of ten students read one passage in
   silence and a comparable passage with lyrics, in a random order.
   Explain why a paired analysis would use within-student differences
   and why an independent-group calculation, such as Exercise 8's,
   would no longer be the appropriate starting point. Name the
   dependence information needed to calculate its SE.
2. Explain how assigning one class to each condition within each
   school creates a blocked comparison. Identify the comparison
   units in Chapter 3's paired-school example (Section 2.3) and
   distinguish an allocation block from a sampling stratum.
3. A researcher proposes measuring a fractions pretest before
   assigning feedback in the feedback-by-practice study. Explain what
   baseline adjustment adds, why an adjusted comparison needs a
   specified common value or averaging distribution, and why
   adjustment alone does not establish a causal interpretation.

## Appendix A: Deriving the Matched Pooled-t Identity {.unnumbered #app-ch04-pooled-identity}

[]{#app-ch04-calculations}

Section 2.4 states this identity and its conditions; the derivation
shows where it comes from. For two independent groups, let
$\widehat{\Delta\mu}=\bar y_1-\bar y_2$ and $N=n_1+n_2$. Their grand mean is

$$
\bar y=\frac{n_1\bar y_1+n_2\bar y_2}{N}.
$$

Subtracting it from each group mean gives

$$
\bar y_1-\bar y=\frac{n_2}{N}\widehat{\Delta\mu},
\qquad
\bar y_2-\bar y=-\frac{n_1}{N}\widehat{\Delta\mu}.
$$

Consequently,

$$
\begin{aligned}
SS_B
&=n_1\left(\frac{n_2}{N}\widehat{\Delta\mu}\right)^2
  +n_2\left(-\frac{n_1}{N}\widehat{\Delta\mu}\right)^2\\
&=\frac{n_1n_2}{N}\widehat{\Delta\mu}^2.
\end{aligned}
$$

There is one numerator df, so $MS_B=SS_B$. Define the pooled
variance $s_p^2=SS_W/(N-2)$, which is Chapter 3's pooled variance and
equals $MS_W$ when $J=2$. Because
$n_1n_2/N=1/(1/n_1+1/n_2)$,

$$
F=\frac{SS_B}{s_p^2}
=\frac{\widehat{\Delta\mu}^2}{s_p^2(1/n_1+1/n_2)}=t^2,
$$

where $t=\widehat{\Delta\mu}/\{s_p\sqrt{1/n_1+1/n_2}\}$ is the
equal-variance t statistic.

The equality is algebraic for these matched definitions; the equal
p-values add the test conditions and a two-sided t alternative. It
connects the one-way F test to the equal-variance t test of Chapter 3,
Section 5.3. The paired-samples t test (Chapter 3, Section 5.2) and
Welch's t test (Chapter 3, Section 5.4) have their own SEs and df, so
this algebra supplies neither; Welch's several-group counterpart is
the Welch one-way ANOVA of Section 5.2.

## Appendix B: Reference Notes on Weights, Effect Sizes and a Shared Reference Mean {.unnumbered #app-ch04-comparisons}

### B.1 Unequal-cell weights {.unnumbered #app-ch04-weighting}

In the chapter's balanced data, equal-cell and student-weighted
marginal means coincide. For unequal cell counts, define them
separately:

$$
\bar y_{j\cdot}^{\text{students}}
=\frac{\sum_kn_{jk}\bar y_{jk}}{\sum_kn_{jk}},
\qquad
\bar y_{j\cdot}^{\text{equal cells}}
=\frac1K\sum_k\bar y_{jk}.
$$

The first describes the mixture of students observed in that
feedback group. The second gives each of the $K$ practice conditions
the same weight. Neither is automatically the right target for
every question.

For an illustration, retain Answer's cell means 8 and 12 but
imagine counts of 2 Massed and 6 Spaced students. The equally
weighted mean is $(8+12)/2=10$; the student-weighted mean is
$(2\times8+6\times12)/8=11$. These are hypothetical weights,
not changes to the actual four-per-cell dataset. A comparison of
feedback groups using different schedule mixtures can combine a
feedback comparison with a weighting difference. Adjusted means in
later chapters make their common profile or averaging distribution
explicit for this reason.

### B.2 Effect-size denominators {.unnumbered #app-ch04-effect-sizes}

Section 2.4 used one omnibus summary, eta squared. Software reports
several related summaries whose denominators differ, so each needs its
name and its denominator before it is interpreted.

\Needspace{14\baselineskip}

| Summary | Formula | Denominator | Value here |
|:--|:--|:--|--:|
| Eta squared, one-way | $SS_B/SS_T$ | All observed variation | .3101 |
| Omega squared, one-way | $\dfrac{SS_B-(J-1)MS_W}{SS_T+MS_W}$ | All observed variation plus one error mean square; the numerator first removes the variation expected from noise | .2525 |
| Eta squared for a factorial term | $SS_{\text{term}}/SS_T$ | All observed variation, including the other factors | Interaction .1345 |
| Partial eta squared | $\dfrac{SS_{\text{term}}}{SS_{\text{term}}+SS_E}$ | The term and the error only | Interaction .5161 |

: Four effect-size summaries and their denominators. {#tbl-ch04-effect-sizes tbl-colwidths="[24,25,37,14]"}

Here $\eta^2=80/258=.3101$, $\omega^2=\{80-2(178/27)\}/(258+178/27)=.2525$,
and the interaction's two summaries are $64/476=.1345$ and
$64/(64+60)=.5161$. Omega squared subtracts the between-group
variation that group means show even when the model means are equal;
it rests on the common-variance model, can fall below zero in a small
sample, and like eta squared says nothing about causation. The two
factorial rows show why the name matters: the same interaction is
.5161 of its partial denominator but .1345 of all the variation,
because the partial version leaves the feedback and practice
components out. In one-way ANOVA the two eta squared versions agree,
since the total holds only the factor and error. The raw mean
differences remain necessary for understanding the outcome scale.

### B.3 Two comparisons that share a reference mean {.unnumbered #app-ch04-dependence}

The two declared comparisons of Section 3.3, Instrumental music minus
Silence and Music with lyrics minus Silence, both subtract the Silence
mean, so their estimates are positively related: a Silence mean that
happens to fall high lowers both differences together, and their
covariance under independent group means is
$\operatorname{Var}(\bar Y_S)=\sigma^2/10$. The union bound needs no
independence, so the joint coverage of at least 95% stands; it
describes repeated use of the whole procedure, not a probability
assigned to the two realized intervals. Dunnett's method, calibrated
for exactly this family of comparisons with one reference, would give
slightly narrower intervals; Bonferroni's arithmetic is the more
transparent. The
[NIST Bonferroni reference](https://www.itl.nist.gov/div898/handbook/prc/section4/prc473.htm)
explains the joint-coverage bound.

## References {.unnumbered}

::: {#refs}
:::
