---
title: "Group Comparisons and Statistical Modelling with R"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: true
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \widowpenalty=10000
          \clubpenalty=10000
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\linespread{0.95}\selectfont\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{8\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---


```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                     collapse = TRUE, comment = "#>")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 10, width = 65)
```

\clearpage

::: {.callout-note title="Chapter overview"}
Chapter 5 made estimates, intervals and tests executable in base R. Here we
return to Chapter 4's group comparisons and use R to connect a fitted model
with the question it answers. We begin with three means, compare selected
differences, and then examine a six-cell pattern. A separate example adds a
pretest so that a group comparison can hold baseline performance fixed.

Counts require a different response model. They also provide a small example
of likelihood, which makes a fitting criterion explicit. Finally, simulation
examines procedures under a stated generating process, and planning relates
sample size to either power or interval width. Throughout, a useful result
keeps its data, target, uncertainty calculation and conditions together.
:::

Chapter 5 made estimates, intervals and tests executable for one mean and
for one correlation. Studies rarely stop at one mean. Twelve pupils of the
invented feedback study practised in the massed schedule, four in each of
three feedback conditions, and their three group means on the 20-item
assessment are 6, 8 and 10 correct answers. Means that differ in a record
differ for more than one reason, so the question is what a stated model and
its conditions let us say about three means at once, and what a single
answer to that question leaves unsettled. These pupils are authored
teaching inputs allocated by stipulation, so each analysis shows what a
procedure reports, not what feedback does.

## Fit and read group means {#ch06-group-models}

### Begin with the question and records

Does mean assessment performance differ across three feedback conditions?
Chapter 4 developed this question with an invented study of 24 pupils.
Each pupil was individually assigned to one combination of feedback
(Rereading, Answer feedback or Explanatory feedback) and practice
(Massed or Spaced). The outcome is the number correct on a 20-item assessment.
Four pupils contribute scores to each cell in @tbl-ch06-feedback-data.

| Feedback | Massed scores | Spaced scores |
|:--|:--|:--|
| Rereading | 4, 5, 7, 8 | 4, 5, 7, 8 |
| Answer feedback | 6, 7, 9, 10 | 10, 11, 13, 14 |
| Explanatory feedback | 8, 9, 11, 12 | 16, 17, 19, 20 |

: The same 24 authored scores as Chapter 4. {#tbl-ch06-feedback-data tbl-colwidths="[35,32,33]"}

These values were constructed to make the calculations transparent. They are
teaching inputs, not empirical evidence that one feedback condition works
better. Each score comes from a different pupil; the four scores within a cell
are not repeated measurements of one pupil. Repeated patterns in their
deviations were deliberately chosen and cannot establish assumptions about a
population.

The following code creates the complete data. The factor levels specify the
display order and the reference group for subsequent model coefficients.
`subset()` selects the twelve pupils assigned to Massed practice; the selected
rows still represent twelve distinct pupils.

```{r}
#| label: ch06-feedback-setup
feedback_data <- data.frame(
  student = 1:24,
  feedback = factor(rep(c("Rereading", "Answer feedback",
    "Explanatory feedback"), each = 8),
    levels = c("Rereading", "Answer feedback",
      "Explanatory feedback")),
  practice = factor(rep(rep(c("Massed", "Spaced"),
    each = 4), 3), levels = c("Massed", "Spaced")),
  score = c(4,5,7,8, 4,5,7,8, 6,7,9,10, 10,11,13,14,
    8,9,11,12, 16,17,19,20))
massed <- subset(feedback_data, practice == "Massed")
table(massed$feedback)
with(massed, tapply(score, feedback, mean))
with(massed, tapply(score, feedback, sd))
```

The Massed means are 6, 8 and 10 correct answers, with four pupils in each
group and a sample SD of 1.8257 in every group. Their grand mean is 8.
Before asking whether a difference is large relative to its uncertainty,
we can see the difference on the outcome's own scale: Answer feedback exceeds
Rereading by 2 points and Explanatory feedback exceeds it by 4 points.

### Extend a graph one useful component at a time

Chapter 5 used base-R plotting functions. The package `ggplot2` provides
another way to describe a graph: name a data frame, map its variables to
visual properties, and add the kinds of marks that should be drawn. Its
incremental advantage is that another layer or mapping can extend the same
description without rebuilding the graph from separate drawing commands.
Load this one package after its installation has been checked in the
Computer Lab.

```{r}
#| label: ch06-first-ggplot
#| fig-width: 6.4
#| fig-height: 2.9
#| fig-alt: "The first ggplot places the twelve Massed pupil scores in three vertical groups, labelled Rereading, Answer feedback and Explanatory feedback."
library(ggplot2)
p <- ggplot(massed, aes(feedback, score))
p + geom_point()
```

`ggplot(massed, ...)` supplies the records. Inside `aes()`, the feedback
variable maps to horizontal position and score maps to vertical position.
`geom_point()` adds one point for each record. The object `p` stores the
data and mapping; the expression with `+` adds a plotting component.

We can now add the three group means as larger diamonds. `aggregate()` uses
the same twelve scores to make one mean per feedback group. The second point
layer inherits the mapping from `p` but uses this three-row data frame.

```{r}
#| label: fig-ch06-feedback-groups
#| fig-width: 6.4
#| fig-height: 3.5
#| fig-cap: "Individual Massed scores and their three group means."
#| fig-alt: "Four scores in each of Rereading, Answer feedback and Explanatory feedback lie around means 6, 8 and 10. Larger diamonds mark the means; all groups have the same constructed spread."
group_means <- aggregate(score ~ feedback, massed, mean)
p + geom_point(size = 2) +
  geom_point(data = group_means, shape = 18, size = 4,
    colour = "black") +
  labs(x = "Feedback", y = "Number correct (out of 20)") +
  theme_minimal(base_size = 11)
```

The fixed colour is an argument outside `aes()`. Later, placing a variable
inside `aes(colour = ...)` will give different colours to its levels and
create a legend. `labs()` names the axes, and `theme_minimal()` changes their
presentation. Neither changes the observations or the means. In
@fig-ch06-feedback-groups the raw points keep the denominator and within-group
variation visible beside the means.

### Fit the group model and read its decomposition

For pupil $i$ in feedback group $j$, the one-way model is

$$
Y_{ij}=\mu_j+\varepsilon_{ij}.
$$

Here $\mu_j$ is the group's process mean and $\varepsilon_{ij}$ is its pupil's
deviation from that mean. Fitting by least squares estimates each $\mu_j$
with the corresponding sample mean, so $\widehat Y_{ij}=\bar Y_j$.
The residual is $e_{ij}=y_{ij}-\bar y_j$: an observed score minus its fitted
group mean. For the first group the fitted values are all 6 and the residuals
are $-2,-1,1,2$.

The model lets us separate differences between groups from differences among
pupils within a group. Chapter 4's decomposition becomes

$$
\begin{aligned}
SS_T&=\sum_{j}\sum_i(y_{ij}-\bar y)^2=SS_B+SS_W,\\
SS_B&=\sum_j n_j(\bar y_j-\bar y)^2,\\
SS_W&=\sum_j\sum_i(y_{ij}-\bar y_j)^2.
\end{aligned}
$$

The total measures deviation from the grand mean. The between-group term
measures deviation of the three group means from that grand mean, weighted
by their group counts. The within-group term measures deviation of pupils
from their own fitted group means. In this example,
$SS_B=4(6-8)^2+4(8-8)^2+4(10-8)^2=32$ and each group contributes
$(-2)^2+(-1)^2+1^2+2^2=10$ to $SS_W$, giving $SS_W=30$.

The formula `score ~ feedback` tells `aov()` to explain score using the
feedback factor. `anova()` extracts the decomposition from the fitted
object; `fitted()` and `resid()` expose its predicted values and residuals.

```{r}
#| label: ch06-oneway-fit
one_fit <- aov(score ~ feedback, data = massed)
anova(one_fit)
fitted(one_fit)
resid(one_fit)
```

Three fitted group means use three parameters. Consequently the within-group
degrees of freedom are $12-3=9$. The between-group component has $3-1=2$
degrees of freedom, and the total has $12-1=11$. Dividing each sum of squares
by its degrees of freedom gives a mean square. The within-group mean square
$MS_W=30/9=3.3333$ estimates the common error variance.

| Source | SS | df | MS | F | p |
|:--|--:|--:|--:|--:|--:|
| Feedback | 32.0000 | 2 | 16.0000 | 4.8000 | .0381 |
| Within groups | 30.0000 | 9 | 3.3333 | | |
| Total | 62.0000 | 11 | | | |

: One-way ANOVA on the twelve Massed pupils. {#tbl-ch06-oneway}

The null hypothesis is $H_0:\mu_R=\mu_A=\mu_E$. Under independent Normal
errors with a common variance, the ratio

$$
F=\frac{MS_B}{MS_W}=\frac{16}{30/9}=4.8000
$$

has an $F_{2,9}$ reference distribution when that null holds. The p-value
.0381 is the probability, under this null model, of an F statistic at least
this large. At a .05 threshold it provides evidence against equal process
means. It does not say that every pair differs, that all feedback improves
performance, or that the invented result establishes a real intervention
effect. The next section asks the individual comparison questions.

The observed share of score variation associated with these group means is

$$
\eta^2=\frac{SS_B}{SS_T}=\frac{32}{62}=.5161.
$$

This is an in-sample variation summary, accompanying differences expressed in
score points. It is neither the probability that feedback caused a result
nor a replacement for a confidence interval. Appendix A gives the related
bias-adjusted omega-squared calculation, with its denominator and interpretation.

### Express the same fit with indicators

A factor with three levels can also enter a linear model as two indicator
variables. Let $D_A=1$ for Answer feedback and 0 otherwise, and let $D_E=1$
for Explanatory feedback and 0 otherwise. Rereading has both indicators 0.
The fitted equation is

$$
\widehat y=6+2D_A+4D_E.
$$

Substituting the indicators gives 6 for Rereading, $6+2=8$ for Answer feedback
and $6+4=10$ for Explanatory feedback. The intercept is the reference mean;
each other coefficient is a difference from that mean. The factor order in
the original setup makes Rereading the reference level.

```{r}
#| label: ch06-indicator-fit
indicator_fit <- lm(score ~ feedback, data = massed)
coef(indicator_fit)
anova(indicator_fit)
```

`lm()` and `aov()` describe the same set of fitted means and give the same
residuals and ANOVA here. The formula names the outcome and explanatory
variable; the study's design supplies the independence assumption. This
indicator representation will make the adjusted group comparison easier to
read. Chapter 7 develops a continuous predictor's slope, and Chapter 8
extends the conditional interpretation to several predictors.

## Ask comparisons within stated families {#ch06-contrast-families}

### Specify the comparison and its standard error

Suppose the two questions concern Answer feedback versus Rereading and
Explanatory feedback versus Rereading, among Massed pupils. In the order
Rereading, Answer feedback, Explanatory feedback, their coefficient vectors
are $(-1,1,0)$ and $(-1,0,1)$. Each vector sums to zero. A quantity

$$
C=\sum_jc_j\mu_j,\qquad \sum_jc_j=0,
$$

is a contrast of the process means. Its estimate replaces process means with
the observed group means:
$\widehat C=\sum_jc_j\bar y_j$. Thus our estimates are 2 and 4 points.
Calling a comparison a contrast specifies this mathematical form; it does
not say whether the comparison was planned before seeing the results or
selected afterward. That is additional information about the analysis.

For independent groups with a common error variance, Chapter 3's variance
rule for a linear combination gives

$$
SE(\widehat C)=\sqrt{MS_W\sum_j\frac{c_j^2}{n_j}}.
$$

Each comparison uses two of the means, so both have estimated SE
$\sqrt{(30/9)(1/4+1/4)}=1.2910$ points. The third group still contributes
to the common error estimate and its 9 residual degrees of freedom.
This matters when comparing these results with an analysis restricted to
only two groups.

### Calibrate the two-comparison family

An individual 95% interval uses a .05 error probability for that one
comparison. To cover both declared comparisons with at least 95% familywise
confidence, Bonferroni allocates .025 to each two-sided comparison. Each tail
then receives $.05/(2\times2)=.0125$. The calibration uses the $t$
reference distribution with the model's 9 residual degrees of freedom
[@student1908], giving the critical value

$$
t_{1-.05/(2\times2),9}=2.6850.
$$

The calculations retain the same estimates and SEs but multiply each SE by
this larger critical value. The following code extracts the common variance
estimate from the fitted object and calculates both results from unrounded
inputs.

```{r}
#| label: ch06-bonferroni
mse <- deviance(one_fit) / df.residual(one_fit)
means <- with(massed, tapply(score, feedback, mean))
estimate <- c(Answer = unname(means[2] - means[1]),
  Explain = unname(means[3] - means[1]))
se <- sqrt(mse * (1/4 + 1/4))
critical <- qt(1 - .05/(2*2), df.residual(one_fit))
raw_p <- 2 * pt(-abs(estimate/se), df.residual(one_fit))
cbind(estimate, SE = se, lower = estimate-critical*se,
  upper = estimate+critical*se,
  adjusted_p = p.adjust(raw_p, method = "bonferroni"))
```

| Comparison | Estimate | SE | Simultaneous interval | Adjusted p |
|:--|--:|--:|:--|--:|
| Answer minus Rereading | 2.0000 | 1.2910 | [$-1.4663$, 5.4663] | .3115 |
| Explanatory minus Rereading | 4.0000 | 1.2910 | [0.5337, 7.4663] | .0255 |

: Bonferroni results for the declared family of two comparisons. {#tbl-ch06-bonferroni tbl-colwidths="[32,13,13,28,14]"}

Multiplying each raw p-value by two and capping at one gives the corresponding
Bonferroni-adjusted p-value. Explanatory minus Rereading is distinguishable
from zero with this familywise calibration, whereas Answer minus Rereading
is not. Its interval allows both a small negative difference and a positive
difference of several points. This uncertainty is compatible with the earlier
omnibus rejection: the omnibus test and each contrast address different
statements.

### Compare with an all-pairs family

If the question includes every pair, there are three comparisons. Tukey's
procedure uses the joint distribution of pairwise mean differences under the
common-variance Normal-error model. `TukeyHSD()` supplies those simultaneous
intervals and their adjusted p-values [@tukey1953].

```{r}
#| label: ch06-tukey
TukeyHSD(one_fit, "feedback", conf.level = .95)
```

The output labels each subtraction direction. Explanatory minus Rereading is
4, with interval [0.3955, 7.6045] and adjusted p .0310. Answer minus Rereading
and Explanatory minus Answer both equal 2, with intervals [$-1.6045$, 5.6045]
and adjusted p .3150. These limits differ from the preceding intervals
because the declared family and its calibration differ. An interval should
therefore be read together with the comparison and family it represents.

An omnibus rejection does not by itself protect any comparison selected
after examining the means. Conversely, a separately specified contrast can
have its own valid uncertainty procedure without requiring omnibus rejection
as an automatic preliminary gate. The relevant information is the question,
selection process and calibration, rather than a universal sequence of
significance decisions.

### Check what changes when the variance model changes

The common-variance model assumes that within-group variability estimates
one shared error variance. In a real study, design information and observed
patterns help assess that working model. A small normality or variance test
cannot certify it, especially with four pupils per group. Here the identical
SDs were deliberately constructed.

For independent groups whose error variances may differ, Welch's one-way
procedure provides an alternative [@welch1947]. R selects it by default in
`oneway.test()` [@rcore2025]:

```{r}
#| label: ch06-welch-oneway
oneway.test(score ~ feedback, data = massed)
```

It gives $F(2,6)=4.3200$, p .0688. Although the observed group SDs happen to
be equal, Welch's small-sample correction and reference degrees of freedom
differ from the pooled-variance ANOVA. Its uncertainty calculation is built
for a different variance model. The two p-values are therefore results to
explain through their assumptions and calibration, rather than votes on
which numerical conclusion to prefer. With small groups, the model choice
can matter appreciably.

A useful exact identity connects a two-group pooled t test with a two-group
ANOVA: $F=t^2$, with the same two-sided p-value. It requires the same records,
equal-variance denominator and group comparison. R's default Welch t test
does not meet that specification. Nor is our three-group contrast the same
calculation, because it uses residual information from the third group.
Appendix A reproduces the matched identity explicitly.

## Compare a two-way pattern {#ch06-factorial-comparisons}

### Read cells and margins before fitting

We now return to all 24 pupils and ask how the Spaced-minus-Massed comparison
changes across feedback conditions. A cell is one combination of feedback
and practice. Computing cell counts and means first makes the six cells
and their denominators visible.

```{r}
#| label: ch06-cell-summaries
with(feedback_data, table(feedback, practice))
cell_mean <- with(feedback_data,
  tapply(score, list(feedback, practice), mean))
cell_mean
rowMeans(cell_mean)
colMeans(cell_mean)
mean(feedback_data$score)
```

| Feedback | Massed | Spaced | Equally weighted margin | Spaced minus Massed |
|:--|--:|--:|--:|--:|
| Rereading | 6 | 6 | 6 | 0 |
| Answer feedback | 8 | 12 | 10 | 4 |
| Explanatory feedback | 10 | 18 | 14 | 8 |
| Practice margin | 8 | 12 | 10 | 4 |

: Cell means and equally weighted margins of the balanced design. {#tbl-ch06-cell-means tbl-colwidths="[31,12,12,23,22]"}

Each cell has four pupils, so averaging the two practice cells equally gives
the feedback group's overall mean. Similarly, each practice margin averages
equally across the three feedback conditions. In an unbalanced design, a
margin must specify its weights: the observed mix of pupils and an equally
weighted target can answer different questions.

The new graph uses the same data/mapping/layer idea as the earlier group
plot. Here a colour identifies each feedback condition and `group` identifies
the means to connect. The line joins separate cell means; it does not trace
individual pupils across repeated measurements.

```{r}
#| label: fig-ch06-feedback-interaction
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: "The practice difference depends on feedback."
#| fig-alt: "Rereading joins 6 to 6, Answer feedback joins 8 to 12, and Explanatory feedback joins 10 to 18 from Massed to Spaced. The lines have different slopes."
cell_plot <- aggregate(score ~ feedback + practice,
  feedback_data, mean)
ggplot(cell_plot, aes(practice, score, colour = feedback,
  group = feedback)) + geom_line() + geom_point(size = 2) +
  labs(x = "Practice", y = "Mean number correct",
    colour = "Feedback") +
  theme_minimal(base_size = 11)
```

`colour = feedback` is inside `aes()` because a variable now determines
colour. The mapping also creates the legend. The observed changes are 0,
4 and 8 points, producing nonparallel lines in
@fig-ch06-feedback-interaction. This describes interaction in the observed
means. Inference about a process interaction additionally requires a model
and a measure of residual variation.

### Fit the full model and identify its tests

Chapter 4 wrote the two-way model as

$$
Y_{ijk}=\mu+\alpha_j+\beta_k+(\alpha\beta)_{jk}
  +\varepsilon_{ijk},
$$

where $j$ indexes feedback, $k$ indexes practice and $i$ indexes pupils
within a cell. With the usual sum-to-zero constraints, $\mu$ is the grand
mean, $\alpha_j$ and $\beta_k$ describe departures of the margins from it,
and $(\alpha\beta)_{jk}$ describes the remaining cell departure. The
constraints make this representation identifiable: the effects cannot all
shift freely while leaving the same fitted means.

There is one grand-mean parameter, $3-1=2$ feedback degrees of freedom,
$2-1=1$ practice degree of freedom, and $(3-1)(2-1)=2$ interaction degrees
of freedom. Together these describe six fitted cell means. The model
therefore has $24-6=18$ residual degrees of freedom.

In an R formula, `feedback * practice` includes both main effects and their
interaction. It expands to `feedback + practice + feedback:practice`.

```{r}
#| label: ch06-factorial-fit
two_fit <- aov(score ~ feedback * practice,
  data = feedback_data)
anova(two_fit)
```

| Source | SS | df | MS | F | p |
|:--|--:|--:|--:|--:|--:|
| Feedback | 256.0000 | 2 | 128.0000 | 38.4000 | < .0001 |
| Practice | 96.0000 | 1 | 96.0000 | 28.8000 | < .0001 |
| Interaction | 64.0000 | 2 | 32.0000 | 9.6000 | .0015 |
| Residual | 60.0000 | 18 | 3.3333 | | |
| Total | 476.0000 | 23 | | | |

: Full balanced factorial decomposition. {#tbl-ch06-factorial}

The feedback test concerns equality of feedback margins averaged equally
over practice. The practice test concerns equality of practice margins
averaged equally over feedback. The interaction test asks whether the
practice difference is the same across all feedback conditions. With
independent Normal errors of common variance, its null reference is
$F_{2,18}$; the observed value 9.6000 and p .0015 provide evidence against
that common-difference null.

The balanced layout makes the source components orthogonal: the variation
attributed to one component does not compete with another component for
the same fitted direction. A single-model `anova()` table ordinarily uses
sequential sums of squares, entering terms in formula order. This convenient
balanced decomposition should not be extended automatically to unbalanced
or adjusted data. Later in this chapter, we compare explicitly named nested
models on identical rows to specify adjusted questions.

The feedback and practice margins remain meaningful descriptions, but the
interaction clarifies their limits. For example, the 4-point overall practice
difference averages a 0-point, a 4-point and an 8-point difference. A result
paragraph should convey this varying pattern when the scientific question
concerns how practice works across feedback conditions.

### Work the selected difference of differences completely

Consider Answer feedback's practice change minus Rereading's practice change:

$$
\widehat{DD}=(12-8)-(6-6)=4.
$$

The contrast has coefficients $(1,-1,-1,1)$ in the cell order
Rereading–Massed, Rereading–Spaced, Answer–Massed, Answer–Spaced.
Its four cell means come from different pupils. Under the common-variance
model, their variances therefore add after multiplying by the squared
coefficients:

$$
SE(\widehat{DD})=
\sqrt{MS_E\left(\frac14+\frac14+\frac14+\frac14\right)}
=1.8257.
$$

The omnibus interaction spans two independent directions. For simultaneous
confidence over all contrasts in that two-dimensional interaction space,
Scheffe's method uses the multiplier

$$
\sqrt{2F_{.95;2,18}}=2.6663.
$$

The complete calculation keeps the cell subtraction, residual variance and
family calibration visible:

```{r}
#| label: ch06-selected-interaction
dd <- (cell_mean["Answer feedback", "Spaced"] -
  cell_mean["Answer feedback", "Massed"]) -
  (cell_mean["Rereading", "Spaced"] -
  cell_mean["Rereading", "Massed"])
mse_two <- deviance(two_fit) / df.residual(two_fit)
se_dd <- sqrt(mse_two * 4/4)
scheffe <- sqrt(2 * qf(.95, 2, 18))
c(estimate = dd, SE = se_dd, multiplier = scheffe)
dd + c(-1, 1) * scheffe * se_dd
```

The resulting interval is $4\pm2.6663(1.8257)=[-0.8680,8.8680]$.
It includes zero even though the interaction F test rejects its null.
The F test asks whether both interaction directions are zero jointly;
this interval addresses one selected direction while allowing for the
entire family. Other directions can carry substantial evidence, and a
simultaneous interval for this direction can remain broad. Neither result
cancels the other.

Three related quantities help distinguish the questions. Answer feedback's
simple practice effect is 4, with SE 1.2910 and pointwise t interval
[1.2877, 6.7123]. The selected DD also happens to equal 4, but its SE is
1.8257 because it compares two practice effects. Its pointwise interval is
[0.1643, 7.8357], whereas its simultaneous interval is [$-0.8680$, 8.8680].
Appendix A shows both DD calculations using the same estimate and SE.

A simple effect includes the overall practice component as well as an
interaction component. It is therefore outside the interaction-only family
used above. The shared numerical estimate does not give it the DD's meaning,
SE or familywise guarantee. State which effect was selected, which family
was considered and which interval was used before interpreting a result.

## Adjust a group comparison for baseline {#ch06-ancova}

### State the new data and the conditional question

Suppose we want to compare group means at a common pretest score. The earlier
24-pupil dataset has no pretest. We therefore introduce a separate authored
example of twelve pupils in @tbl-ch06-adjustment-data. These are new teaching
inputs, with individual assignment to feedback stipulated to occur after the
pretest. Each row records one pupil's two assessments, both counting correct
answers out of 20. The pupil is the assignment and measurement unit; distinct
pupils supply the independent error units in the working model.

| Pupil | Feedback | Pretest | Posttest |
|--:|:--|--:|--:|
| 1 | Rereading | 4 | 5 |
| 2 | Rereading | 6 | 9 |
| 3 | Rereading | 8 | 11 |
| 4 | Rereading | 10 | 11 |
| 5 | Answer feedback | 6 | 9 |
| 6 | Answer feedback | 8 | 13 |
| 7 | Answer feedback | 10 | 15 |
| 8 | Answer feedback | 12 | 15 |
| 9 | Explanatory feedback | 5 | 10 |
| 10 | Explanatory feedback | 7 | 14 |
| 11 | Explanatory feedback | 9 | 16 |
| 12 | Explanatory feedback | 11 | 16 |

: Fixed inputs for a separate baseline-adjustment illustration. {#tbl-ch06-adjustment-data tbl-colwidths="[12,48,20,20]"}

Missing values are absent by construction. The regular residual pattern
makes adjustment arithmetic easy to follow, but cannot demonstrate Normal
errors, independence or a common slope in a real study. The observed pretest
ranges are 4–10, 6–12 and 5–11, so their common range is 6–10. Pretest 8
lies within each group's observed range. Comparing fitted means there is
more supportable than extending a group's line far beyond its observations.

```{r}
#| label: ch06-adjustment-setup
feedback_adjustment <- data.frame(
  student = 1:12,
  feedback = factor(rep(c("Rereading", "Answer feedback",
    "Explanatory feedback"), each = 4),
    levels = c("Rereading", "Answer feedback",
      "Explanatory feedback")),
  pretest = c(4,6,8,10, 6,8,10,12, 5,7,9,11),
  posttest = c(5,9,11,11, 9,13,15,15, 10,14,16,16))
feedback_adjustment$baseline <- feedback_adjustment$pretest - 8
aggregate(cbind(pretest, posttest) ~ feedback,
  feedback_adjustment, mean)
```

The raw posttest means are 9, 13 and 14, while the pretest means are 7, 9
and 8. A raw group mean averages over that group's own baselines. Our new
question fixes pretest at 8 and compares the fitted posttest means there.
The difference between these questions explains why the adjusted differences
can differ from the raw differences.

### Describe and fit a common-slope model

Analysis of covariance, or ANCOVA, uses a linear model that includes both
group indicators and quantitative covariates. Define the centered baseline
$x_c=x-8$. With the same indicator meanings as before, the model is

$$
Y_i=\beta_0+\beta_x x_{ci}+\beta_A D_{Ai}+\beta_E D_{Ei}
+\varepsilon_i.
$$

At $x=8$, the centered baseline is zero. The intercept is therefore
Rereading's fitted mean at pretest 8. The two group coefficients are
differences from Rereading at that same pretest. The common slope $\beta_x$
describes the fitted change in mean posttest for a one-point pretest change
within any group. Graphically, the model uses three parallel lines.

Centering changes the intercept's reference point, while leaving the fitted
values unchanged. Choosing 8 also makes the intercept refer to a value
supported by observations in all three groups. Chapter 7 develops how a
slope is fitted, and Chapter 8 develops conditioning on several predictors;
here the line gives a concrete first adjusted comparison.

```{r}
#| label: ch06-adjustment-fit
unadjusted <- lm(posttest ~ feedback,
  data = feedback_adjustment)
adjusted <- lm(posttest ~ baseline + feedback,
  data = feedback_adjustment)
coef(summary(adjusted))
confint(adjusted)
```

| Term | Estimate | SE | 95% pointwise interval |
|:--|--:|--:|:--|
| Rereading at pretest 8 | 10.0000 | 0.6325 | [8.5416, 11.4584] |
| Pretest slope | 1.0000 | 0.1581 | [0.6354, 1.3646] |
| Answer minus Rereading | 2.0000 | 0.9220 | [$-0.1260$, 4.1260] |
| Explanatory minus Rereading | 4.0000 | 0.8803 | [1.9699, 6.0301] |

: Common-slope fit, with residual MSE 1.5000 on 8 df. {#tbl-ch06-ancova-coefficients tbl-colwidths="[40,16,16,28]"}

`coef(summary(adjusted))` returns estimates, SEs, t statistics and two-sided
p-values for the individual zero-coefficient tests. The adjusted Answer
difference has p .0619 and the Explanatory difference has p .0019.
`confint()` uses the residual t reference to form the corresponding intervals.
These intervals are pointwise: a simultaneous conclusion about a family
requires an explicitly chosen family adjustment.

### Read fitted means at the same baseline

To obtain a fitted mean for each group at pretest 8, construct one profile
per group with `baseline = 0`. `predict()` evaluates the fitted equation at
each profile and returns an interval for its conditional mean.

```{r}
#| label: ch06-adjusted-means
profiles <- data.frame(baseline = 0,
  feedback = levels(feedback_adjustment$feedback))
predict(adjusted, newdata = profiles,
  interval = "confidence")
```

The adjusted means are 10, 12 and 14. Their 95% pointwise confidence
intervals are [8.5416, 11.4584], [10.5416, 13.4584] and
[12.5879, 15.4121], respectively. They estimate group means at a stated
baseline, rather than ranges containing individual future pupils. Prediction
for a new pupil would include that pupil's additional residual variation;
Chapter 8 develops the distinction.

The adjustment can also be read directly from the raw means. For the
common-slope fit,

$$
\widehat\mu_j(8)=\bar y_j-\widehat\beta_x(\bar x_j-8).
$$

With slope 1, Rereading's adjusted mean is $9-(7-8)=10$; Answer feedback's
is $13-(9-8)=12$; Explanatory feedback's is $14-(8-8)=14$.
Thus the raw differences of 4 and 5 points become conditional differences
of 2 and 4 points. Adjustment compares groups at a common baseline rather
than at their different average baselines.

@fig-ch06-ancova shows this geometry. The fitted segments are restricted
to the shared observed range 6–10, and the dashed line marks pretest 8.
The points still show the full range of each group's data.

```{r}
#| label: fig-ch06-ancova
#| echo: false
#| fig-width: 6.4
#| fig-height: 3.8
#| fig-cap: "Compare fitted group means at a shared baseline."
#| fig-alt: "The twelve pretest/posttest points appear by feedback group. Three parallel fitted segments over common baseline support 6 to 10 pass through means 10, 12 and 14 at the dashed baseline 8 line."
line_grid <- expand.grid(pretest = seq(6, 10, length.out = 20),
  feedback = levels(feedback_adjustment$feedback))
line_grid$baseline <- line_grid$pretest - 8
line_grid$fitted <- predict(adjusted, newdata = line_grid)
ggplot(feedback_adjustment,
  aes(pretest, posttest, colour = feedback)) +
  geom_point(size = 2) +
  geom_line(data = line_grid, aes(y = fitted)) +
  geom_vline(xintercept = 8, linetype = 2) +
  labs(x = "Pretest (out of 20)", y = "Posttest (out of 20)",
    colour = "Feedback") + theme_minimal(base_size = 11)
```

The Computer Lab's adjustment reference constructs the plotting grid and
changes the supported baseline value. The statistical calculation remains
the printed `predict()` call: specify the profiles first, then interpret the
fitted means at those profiles.

### Compare named nested models on identical records

Two questions about this fit require different smaller models. First, does
baseline add information after accounting for the feedback groups? Compare
the groups-only fit, `unadjusted`, with the larger fit, `adjusted`. Second,
do the group indicators jointly add information after accounting for
baseline? Compare a baseline-only fit with the same larger fit.

```{r}
#| label: ch06-nested-tests
baseline_only <- lm(posttest ~ baseline,
  data = feedback_adjustment)
anova(unadjusted, adjusted)
anova(baseline_only, adjusted)
```

For nested models fitted to the same outcome and rows, let $RSS_0$ and
$df_0$ denote the smaller model's residual sum of squares and degrees of
freedom, and let $RSS_1$ and $df_1$ denote the larger model's. Then

$$
F=\frac{(RSS_0-RSS_1)/(df_0-df_1)}{RSS_1/df_1}.
$$

The numerator measures the improvement per added parameter. The denominator
is the residual variance estimate from the larger model. Here the larger
model estimates four coefficients from twelve pupils, leaving 8 residual
degrees of freedom and $RSS_1/df_1=12/8=1.5000$.

Adding baseline to groups reduces RSS from 72 on 9 df to 12 on 8 df, giving
$F(1,8)=40.0000$, p .0002. Adding the group indicators to baseline alone
reduces RSS from 43.0588 on 10 df to 12 on 8 df, giving
$F(2,8)=10.3529$, p .0060. The latter tests both adjusted group coefficients
jointly. As with the earlier omnibus tests, it does not establish that each
individual adjusted comparison differs from zero.

The same-row condition is essential. If different predictors contain missing
values, automatic omission can fit the smaller and larger models to
different pupils. Their RSS difference would then mix a model change with
a data change. Explicit model comparison also clarifies the question more
reliably than treating every entry in a single sequential ANOVA table as an
order-free adjusted test.

### Check the fit and state its scope

The classical tests and intervals use independent errors, an appropriate
linear common-slope conditional mean, a common conditional error variance,
and a Normal-error reference. A graph of the fitted relationship helps
locate unsupported extrapolation or a changing group comparison. A residual
plot examines departures left after fitting; a Normal-quantile plot compares
their shape with Normal quantiles.

```{r}
#| label: fig-ch06-residuals
#| echo: false
#| fig-width: 6.4
#| fig-height: 3.1
#| fig-cap: "Residual checks describe the deliberately regular ANCOVA construction."
#| fig-alt: "Residuals are approximately plus or minus one at the fitted values. The Normal-quantile panel displays the same two-level pattern, illustrating why an authored arithmetic example cannot validate a Normal population model."
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 1, .5))
plot(fitted(adjusted), resid(adjusted), xlab = "Fitted posttest",
  ylab = "Residual", pch = 1)
abline(h = 0, lty = 2)
qqnorm(resid(adjusted), main = "", pch = 1)
qqline(resid(adjusted))
par(oldpar)
```

In @fig-ch06-residuals, residuals near $-1$ and 1 reflect the deliberately
regular construction. The display illustrates what to examine; these
twelve chosen values cannot validate a Normal population model. Similarly,
a nonsignificant test for a group-by-baseline term would not prove equal
slopes. Such a term asks whether the adjusted group comparison changes with
baseline; later chapters develop that moderation question.

In real data, report the observed denominators and exclusions as well as
the fitted model. Random allocation does not restore unavailable outcomes.
Adjustment alone does not identify a causal effect in observational data,
repair missing outcomes, or change the population sampled. Here the
assignment story and values are invented; the result demonstrates an
analysis, not an intervention finding. Chapter 8 develops several controls,
covariance-based fitted quantities and comparisons averaged over a named
population. Chapter 9 carries the separate Baumann adjusted-mean reference
as an independent extension.

## Analyse proportions and counts {#ch06-counts}

### Keep the event and denominator visible

A score model is not the appropriate starting point for every outcome.
Chapter 3 introduced the invented completion table in
@tbl-ch06-completion-counts. It contains aggregate counts for 100 records,
classified by learning format and completion. These records have no
pupil-level linkage to either feedback dataset above. The counts themselves
establish neither random allocation nor representative sampling.

| Format | Completed | Not completed | Total |
|:--|--:|--:|--:|
| Online | 32 | 18 | 50 |
| In person | 18 | 32 | 50 |
| Total | 50 | 50 | 100 |

: The same authored completion counts as Chapter 3. {#tbl-ch06-completion-counts}

For the Online group, the event is completion and the denominator is its
50 records. Under an independent Bernoulli model with a common completion
probability $p$, the count of completions has a binomial distribution. The
estimated probability is $\widehat p=32/50=.6400$. If records were clustered
within dependent classes or came from repeated observations of the same
people, this ordinary independent-unit uncertainty model would require
modification.

Suppose the benchmark is $p_0=.5$. The score test standardizes the difference
using the variance under that null:

$$
z=\frac{\widehat p-p_0}{\sqrt{p_0(1-p_0)/50}}
=\frac{.64-.5}{\sqrt{.5(.5)/50}}=1.9799.
$$

Squaring gives $X^2=z^2=3.9200$, compared with a chi-squared distribution
on one degree of freedom. The procedure uses a large-sample reference;
the expected numbers of successes and failures under this null are both 25.

\Needspace{17\baselineskip}

```{r}
#| label: ch06-one-proportion
one_prop <- prop.test(32, 50, p = .5, correct = FALSE)
one_prop$estimate
one_prop$statistic
one_prop$p.value
one_prop$conf.int
```

The two-sided p-value is .0477. The estimated completion probability is
.6400, with 95% Wilson score interval [.5014, .7586]. `correct = FALSE`
explicitly selects the uncorrected score procedure; accepting R's default
continuity correction would specify a different calculation. The call and
the reported result should identify which procedure was used.

The score interval contains the candidate probabilities that their own
two-sided score tests would retain [@wilson1927]. Each candidate supplies
the variance for its null calculation. This inversion explains why the
interval is linked to the score test and is not simply
$\widehat p\pm1.96\sqrt{\widehat p(1-\widehat p)/n}$, the Wald interval.
Appendix A gives the Wilson endpoints directly. The p-value assesses the
stipulated benchmark; it is not the probability that $p=.5$ is true or a
measure of the completion rate's practical importance.

### Compare the two groups through their full table

Now ask whether completion probability differs by format. This uses both
rows and a different null: format and completion are independent in the
working population model, equivalently equal completion probabilities in
this two-group setting. Constructing labelled counts prevents an accidental
change in the event or group direction.

\Needspace{28\baselineskip}

```{r}
#| label: ch06-two-way-counts
counts <- matrix(c(32, 18, 18, 32), nrow = 2,
  byrow = TRUE,
  dimnames = list(Format = c("Online", "In person"),
    Result = c("Completed", "Not completed")))
counts
prop.table(counts, margin = 1)
count_test <- chisq.test(counts, correct = FALSE)
count_test$expected
count_test$statistic
count_test$p.value
count_test$residuals
```

`margin = 1` divides each row by its own total. Completion rates are .6400
for Online and .3600 for In person, a difference of .2800, or 28.0000
percentage points, in the Online-minus-In-person direction. Under independence the
fitted expected count in row $j$, column $k$ is

$$
E_{jk}=\frac{(\text{row }j\text{ total})
(\text{column }k\text{ total})}{N}.
$$

All four expected counts here equal $50\times50/100=25$. Pearson's
statistic accumulates the squared discrepancies relative to those expected
counts:

$$
X^2=\sum_{j,k}\frac{(O_{jk}-E_{jk})^2}{E_{jk}}
=4\left(\frac{7^2}{25}\right)=7.8400.
$$

For a $J$-by-$K$ table the usual independence reference has
$(J-1)(K-1)$ degrees of freedom. Here $df=1$ and p .0051. The observed
Pearson statistic $X^2$ is a number calculated from the table; $\chi^2_1$
names its approximate reference distribution under the null model.

The Pearson residuals are $(O-E)/\sqrt E$: 1.4, $-1.4$, $-1.4$, 1.4
in row order. Their signs locate the pattern: more Online completions and
fewer Online non-completions than the independence fit predicts, with the
opposite departures for In person. Squaring and adding these residuals
recovers $X^2$. They are related through the table margins, so their signs
do not constitute four independent cell tests.

The approximation concerns expected counts under the fitted null model,
along with independent observational units and a sampling model appropriate
to the table. Looking only at observed cell counts misses part of this
check. Sparse expected cells may require an alternative reference or a more
careful model. In the present example all expected counts are 25, supporting
the ordinary chi-squared approximation under the stated unit model.

The one-proportion test asked whether Online completion probability equals
a fixed .5 benchmark. The complete-table test asks whether the two format
probabilities are equal; it estimates their common probability under the
null from the full table. Their different p-values answer different
questions. Neither test establishes that changing format would cause
completion to change without relevant design information.

### Use alternatives as changes to a stated analysis {#ch06-categorical-reference}

The Computer Lab's categorical reference compares the two-proportion
interface, a continuity-corrected calculation, Fisher's conditional test
[@fisher1935] and a standardized association summary, Cramer's V
[@cramer1946], using the same table.
These alternatives keep their own conditions and uncertainty calculations.
For example, the two-proportion difference interval is a large-sample
interval for a difference, while the preceding Wilson interval concerns
one probability. A shared software function name does not make the
intervals identical procedures.

That reference also continues Chapters 2–3's observational student
intention-by-pass example. Its complete primary calculation keeps zero
grades, states the original denominators and compares the outcome under
a documented source-coding sensitivity analysis. This is a separate
application with its own provenance. The small invented completion table
above supplies the first complete procedure; the reference shows how
method choice and unresolved source information matter in an existing
dataset.

## Fit by an objective function and likelihood {#ch06-likelihood}

### Fix the data and vary a candidate parameter

The fitted means in Section 1 were chosen by a rule: minimize the sum
of squared deviations from the observations. We can inspect this rule
with the four Massed Rereading scores, 4, 5, 7 and 8. These observations
stay fixed. What changes is the proposed center, denoted by $c$.

An **objective function** assigns a numerical criterion to each candidate
parameter value for fixed data. For a center fitted by least squares,
the objective is

$$
Q(c)=\sum_{i=1}^{4}(y_i-c)^2
     =(4-c)^2+(5-c)^2+(7-c)^2+(8-c)^2.
$$

At $c=6$, the squared deviations sum to $4+1+1+4=10.0000$
squared score points. At either 5 or 7, they sum to 14.0000. Chapter 2
showed why the sample mean minimizes this criterion. Here the mean is
6.0000 points. A numerical search lets us reproduce the minimizing
value without writing it into the fitting instruction.

```{r}
#| label: ch06-scalar-squared-loss
#| echo: true
rereading <- subset(massed, feedback == "Rereading")$score
loss <- function(center) sum((rereading - center)^2)
mean_fit <- optimize(loss, interval = range(rereading))
c(center = mean_fit$minimum, squared_loss = mean_fit$objective)
c(loss_at_5 = loss(5), loss_at_6 = loss(6), loss_at_7 = loss(7))
```

`function(center)` makes an R function whose input is a candidate
center. Subtracting it from the vector forms four deviations; squaring
and summing returns one number. `optimize()` searches for a minimum
of that function over the interval supplied. Here `range(rereading)`
gives $[4,8]$, which contains the sample mean. The objective is a
single smooth bowl over this interval, so the scalar search has one
minimum to locate.

The two returned quantities have different meanings. `minimum` is
the **parameter location**, 6.0000 points; `objective` is the function
value there, 10.0000 squared points. The latter is a sum of squared
deviations, not an SE or an estimate of the mean. The search interval
also belongs to the fitting instruction: excluding the best center
would restrict which answer the search could return. Numerical stopping
tolerance concerns the accuracy of that search; statistical uncertainty
concerns what an estimate would do across the repetitions justified by
the study.

These least-squares identities are arithmetic. An uncertainty statement
for the fitted mean adds a model or design and its conditions. Chapter 7
extends this criterion from one center to a line with an intercept and
slope. The Computer Lab reference varies the search interval while
keeping the observations fixed.

### Change the fitting criterion for a binomial outcome

For the completion example, the observations are 32 completions in 50
Online trials. Let $p$ denote a candidate probability of completion.
Under the stipulated model of independent trials with a common
probability, the number completed follows a binomial distribution.
The model assigns probability

$$
P_p(X=32)={50\choose32}p^{32}(1-p)^{18}.
$$

In a probability calculation, we choose a value of $p$ and consider
which counts might occur. For fitting, we have already observed the
count: 32 stays fixed, as does the total of 50. We now evaluate the
same expression at different candidate values of $p$. Viewed in this
direction, it is the **likelihood function**,

$$
L(p)={50\choose32}p^{32}(1-p)^{18},\qquad 0\le p\le1.
$$

The binomial coefficient counts the ways 32 completions can occur
among 50 trials. It is constant as $p$ varies. In this binomial model, likelihood is the
probability of the observed data under each candidate model parameter;
it is not a probability distribution over that parameter. In
particular, a high likelihood at a candidate $p$ does not give the
probability that $p$ is true.

**Maximum likelihood estimation** chooses the parameter value that
makes the observed data most likely under the declared model. The
logarithm is increasing, so maximizing the log-likelihood gives the
same parameter value. For $0<p<1$,

$$
\ell(p)=\log {50\choose32}+32\log p+18\log(1-p).
$$

Logs turn a product of powers into a sum and make calculation easier.
The binomial coefficient affects the reported function value, although
it cannot change the maximizing $p$. Here the interior maximum can also
be found analytically. Differentiating with respect to the candidate
probability gives

$$
\ell'(p)=\frac{32}{p}-\frac{18}{1-p}
         =\frac{32-50p}{p(1-p)}.
$$

Setting this slope to zero gives $32-50p=0$, and therefore
$\widehat p=32/50=.6400$. The denominator is positive inside $(0,1)$.
The slope is positive below .64 and negative above .64: the curve rises
to this point and then falls. This verifies a maximum rather than
merely finding a flat point.

To use the default minimization operation in `optimize()`, supply the
**negative** log-likelihood. A smaller negative log-likelihood means a
larger log-likelihood, so its minimum locates the same fitted probability.

\Needspace{12\baselineskip}

```{r}
#| label: ch06-scalar-binomial-mle
#| echo: true
negative_loglik <- function(p) {
  -dbinom(32, size = 50, prob = p, log = TRUE)
}
likelihood_fit <- optimize(negative_loglik, interval = c(0, 1))
c(numerical = likelihood_fit$minimum, analytic = 32/50,
  negative_loglik = likelihood_fit$objective)
negative_loglik(.5)
```

`dbinom()` evaluates the binomial probability for the fixed count 32.
With `log = TRUE`, it computes the log probability directly. The
numerical and analytic estimates both round to .6400. The small
numerical difference from exactly .64 reflects the search tolerance.
At the fitted probability the negative log-likelihood is 2.1465,
compared with 4.1330 at $p=.5$. The smaller value is the better fit
according to this criterion. The parameter estimate .6400 is a
completion probability; the criterion value 2.1465 is not.

```{r}
#| label: fig-ch06-likelihood
#| echo: false
#| fig-pos: H
#| fig-width: 6.1
#| fig-height: 3.4
#| fig-cap: "The binomial log-likelihood is largest at the observed proportion."
#| fig-alt: "For 32 successes in 50 fixed trials, the log-likelihood curve peaks at candidate probability .64. A dashed line at .5 intersects the curve at a lower value."
likelihood_grid <- data.frame(p = seq(.05, .95, length.out = 200))
likelihood_grid$loglik <- dbinom(32, size = 50,
  prob = likelihood_grid$p, log = TRUE)
ggplot(likelihood_grid, aes(p, loglik)) + geom_line() +
  geom_vline(xintercept = .64) +
  geom_vline(xintercept = .5, linetype = 2) +
  labs(x = "Candidate completion probability", y = "Log-likelihood") +
  theme_minimal(base_size = 11)
```

The horizontal axis contains candidate parameter values, and the
vertical axis gives the log-likelihood of the same observed count at
each candidate. Thus the curve displays a fitting criterion, not a
sampling distribution or a distribution of probabilities over $p$.
The peak identifies the fitted value. The score test and interval in
Section 5 answer additional questions about a proposed null and the
uncertainty of estimation under their stated calibration.

Finally, check the boundaries of the parameter space. With both
completions and non-completions observed, $p=0$ and $p=1$ each give
likelihood zero, so the maximum is interior. If all 50 trials instead
had zero completions, the maximum would be at $p=0$; if all completed,
it would be at $p=1$. An interior numerical search can approach such
an endpoint without returning it exactly. The endpoint values must
therefore be examined explicitly in those cases. Chapter 11 develops
these boundary questions further when likelihood is used to fit
logistic regression.

## Examine procedures by repetition and plan precision {#ch06-simulation}

The preceding analyses started with observed data and produced estimates,
intervals and tests. We can also begin with a declared generating model,
create data from it, and examine what an analysis does over repeated studies.
Chapter 5 used this approach to examine sampling variation and interval
coverage. Here the questions concern an omnibus ANOVA procedure and the
sample size for a specified comparison.

### Follow one generated study before repeating it

Imagine three independent groups whose outcomes follow Normal distributions,
each with standard deviation $\sqrt{10/3}=1.8257$ points. First set all three
population means to 8, so that the equal-means null hypothesis is true.
These values define a hypothetical world in which we know the truth. The
constructed feedback scores used earlier help motivate the scale of the
example; they do not establish that this generating model describes real
pupils.

Begin with four observations per group and analyse one generated study.

\Needspace{15\baselineskip}

```{r}
#| label: ch06-one-generated-anova
#| echo: true
set.seed(60630)
n_per_group <- 4
sim_group <- factor(rep(1:3, each = n_per_group))
sim_score <- rnorm(3*n_per_group,
  mean = rep(c(8, 8, 8), each = n_per_group), sd = sqrt(10/3))
anova(lm(sim_score ~ sim_group))
```

`set.seed()` makes this particular sequence of generated observations
reproducible. `rep(1:3, each = n_per_group)` creates four labels for each
of three groups, and `factor()` tells the model to treat those labels as
categories. The second `rep()` supplies the population mean for each of
the 12 observations. `rnorm()` draws independent outcomes with those means
and the declared common SD.

The formula `sim_score ~ sim_group` fits a separate mean for each group;
`anova()` compares that fit with a common mean. Its table has 2 group degrees
of freedom and $12-3=9$ error degrees of freedom. For these generated data,
$SS_{\mathrm{between}}=7.5977$ and $SS_{\mathrm{within}}=32.9851$, giving
$F=1.0365$ and $p=.3935$. Applying the rule $p<.05$ therefore does not reject
the equal-means null in this study. Although the three generating means are
identical, the three sample means vary, so the between-group sum of squares
is positive. The estimated error variance, $MS_E=3.6650$, also varies around
the generating variance $10/3$.

One generated study cannot describe a procedure's long-run behaviour. The
Computer Lab puts the same calculation in a small function: its arguments
are the group size, the three generating means and the common SD; its result
is one omnibus p-value. Repeating that function gives a collection of
p-values to compare with .05.

@tbl-ch06-simulation reports 2,000 studies for each of four conditions.
The null conditions use means $(8,8,8)$; the alternative conditions use
$(6,8,10)$. Each condition retains independent Normal observations and
the same common SD. For reproducibility, the lab restarts at seed 60630
and runs the conditions in this order: four per group under the null,
four under the alternative, 12 under the null, then 12 under the alternative,
without resetting the seed between conditions.

| Pupils per group | Null rejection rate | MCSE | Power at means 6/8/10 | MCSE |
|--:|--:|--:|--:|--:|
| 4 | .0560 | 0.0051 | .6545 | 0.0106 |
| 12 | .0500 | 0.0049 | .9990 | 0.0007 |

: Operating characteristics in the declared generating models, $B=2000$ studies per condition. {#tbl-ch06-simulation}

The **null rejection rate** estimates the Type I error probability: how
often this .05 omnibus procedure rejects when all three population means
are equal. Both estimated rates are close to .05. The **power** column
estimates the probability of rejecting that same null under the specified
unequal means. It rises from .6545 with four pupils per group to .9990 with
12. This is power against the particular means 6, 8 and 10, with the stated
spread and independence; a smaller separation would give different power.

Each rate is itself an estimate based on finitely many repetitions. If
$\hat q$ is a simulated rejection proportion from $B$ independent studies,
its approximate Monte Carlo standard error is

$$
MCSE(\hat q)=\sqrt{\frac{\hat q(1-\hat q)}{B}}.
$$

For the estimated power .6545, the calculation is:

```{r}
#| label: ch06-power-mcse
#| echo: true
sqrt(.6545 * (1 - .6545) / 2000)
```

The result, 0.0106, describes the numerical uncertainty from using 2,000
simulations. A fresh set of 2,000 simulated studies can give a somewhat
different power estimate. This MCSE is distinct from a pupil outcome's
SD or the SE of a group mean or contrast within one study. A rate very
close to one, such as .9990, also deserves care: only two of the 2,000
studies failed to reject, so its tiny plug-in MCSE rests on very few
non-rejections.

Increasing the number of pupils $n$ changes the information within a study,
its error degrees of freedom and its ability to detect the specified
difference. Increasing the number of repetitions $B$ makes the simulated
rate more precise; it leaves the study being simulated unchanged. The
Computer Lab develops these changes in the repetition task and its
reference extension. Simulation describes a procedure under the declared
model; agreement with its expected behaviour does not establish that the
model fits a particular real study.

### Plan for a specified comparison and precision target

The omnibus simulation asks whether some population group means differ.
A future study may instead be planned to estimate or detect one particular
difference. Return to the two comparisons with Rereading from the one-way
example: Answer feedback versus Rereading, and Explanatory feedback versus
Rereading. Keep this declared **two-comparison Bonferroni family**, with
family significance level .05, even when calculating power for just one
of its comparisons.

Suppose the comparison of interest has a population difference of
$\delta=2$ points. Plan equal group sizes $n$, independent Normal outcomes
and common SD $\sigma=\sqrt{10/3}$. The planning SE for a difference of two
group means is $\sigma\sqrt{2/n}$. All three groups contribute to the
pooled error variance, so its degrees of freedom are $3(n-1)$, even though
the contrast compares only two of the means.

Bonferroni assigns two-sided significance level $.05/2=.025$ to each
comparison. Dividing that amount between its two tails puts .0125 in each
tail. The rejection cutoffs are therefore $-c$ and $c$, where

$$
c=t_{1-.05/(2\times2),\,3(n-1)}=t_{.9875,\,3(n-1)}.
$$

Under the zero-difference null, the test statistic has the central
$t$ distribution already used for the comparison. Under a specified
nonzero difference, its distribution moves towards that difference.
The **noncentral $t$ distribution** describes this statistic when its
numerator has a nonzero mean and its denominator still estimates the
common spread. Its noncentrality parameter expresses the population
difference in planning SE units:

$$
\lambda=\frac{\delta}{\sigma\sqrt{2/n}}.
$$

For this alternative, power is the area beyond the same two rejection
cutoffs: $P(T<-c)+P(T>c)$ for a noncentral $t$ variable with $3(n-1)$
degrees of freedom and noncentrality $\lambda$. These are probabilities
over future studies, rather than a p-value from observed data. Both tails
belong to the two-sided procedure, even when the planning difference is
positive.

```{r}
#| label: ch06-contrast-power
#| echo: true
contrast_power <- function(n, delta = 2, sigma = sqrt(10/3)) {
  df <- 3*(n-1)
  critical <- qt(1 - .05/(2*2), df)
  ncp <- delta / (sigma * sqrt(2/n))
  pt(-critical, df, ncp = ncp) +
    pt(critical, df, ncp = ncp, lower.tail = FALSE)
}
round(c(n16 = contrast_power(16), n17 = contrast_power(17)), 4)
```

The function accepts a proposed group size and returns power. Its first
`pt()` call obtains the lower-tail probability; the second obtains the
upper-tail probability. At $n=17$, the planning SE is 0.6262, the error
degrees of freedom are 48, the critical value is 2.3139, and
$\lambda=2/0.6262\approx3.1937$. Summing the two tail areas gives power
.8073. Checking successive integer sizes gives 17 pupils per group,
51 in total, as the first size reaching .80 power; 16 gives .7793.

This is the power of **one specified comparison within the declared
family**. It is neither the probability that both comparisons reject nor
the power of the omnibus F test. For example, with 12 pupils per group
the specified two-point comparison has power .6328, whereas the omnibus
simulation in @tbl-ch06-simulation gives estimated power .9990 against
the complete pattern $(6,8,10)$. The hypotheses and rejection rules differ,
so these numbers answer different questions.

A precision target concerns the interval's width rather than its
probability of excluding zero. Suppose the aim is a simultaneous interval
whose planned half-width is no more than one point. The family and
degrees of freedom stay the same, giving

$$
\text{planned half-width}
=t_{.9875,\,3(n-1)}\,\sigma\sqrt{\frac{2}{n}}.
$$

This calculation substitutes the planning SD $\sigma$ for the error SD
that the future study will estimate. Evaluating the two sizes either side
of the threshold gives:

```{r}
#| label: ch06-planned-half-width
#| echo: true
n_width <- c(34, 35)
round(data.frame(
  n = n_width,
  planned_half_width = qt(1 - .05/(2*2), 3*(n_width-1)) *
    sqrt(2*(10/3)/n_width)
), 4)
```

The first group size meeting this planning target is 35, or 105 pupils in
total: the planned half-width is 0.9929 points, compared with 1.0078 at 34.
The realised interval uses the observed $\sqrt{MS_E}$ and therefore has a
random width. Choosing 35 does not guarantee that its half-width will be
at most one in every study. @fig-ch06-planning shows why a power target and
an interval-width target can select quite different sample sizes.

```{r}
#| label: fig-ch06-planning
#| echo: false
#| fig-width: 6.4
#| fig-height: 3.1
#| fig-cap: "Power and planned interval width impose different sample-size targets."
#| fig-alt: "For the same two-comparison three-group procedure, power rises above .8 at 17 pupils per group. Planned simultaneous interval half-width falls below one point at 35 pupils per group."
n_grid <- 4:50
powers <- sapply(n_grid, contrast_power)
half_width <- qt(1-.05/(2*2), 3*(n_grid-1)) *
  sqrt(2*(10/3)/n_grid)
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 1, .5))
plot(n_grid, powers, type = "l", xlab = "Pupils per group",
  ylab = "Power", ylim = c(0, 1))
abline(h = .8, v = 17, lty = 2)
plot(n_grid, half_width, type = "l", xlab = "Pupils per group",
  ylab = "Planned half-width (points)")
abline(h = 1, v = 35, lty = 2)
par(oldpar)
```

Finally, distinguish the desired number of available outcomes from the
number recruited. If outcomes are expected to be available for 90% of
recruits, the 17-per-group power target gives:

```{r}
#| label: ch06-recruitment-allowance
#| echo: true
ceiling(17 / .9)
```

Recruiting 19 per group gives 57 recruits and an expected 17.1 available
outcomes per group under that retention assumption. It is an average
allowance, not a guarantee of 17 outcomes in every group or a correction
for bias from systematically missing outcomes.

The Computer Lab reference extension supplies the full sample-size grid
and changes the target difference, comparison family and retention rate.
Its two-group `power.t.test()` example declares a separate two-group study:
that procedure has different error degrees of freedom and cannot silently
replace this three-group calculation. Choose a planning difference from
the substantive question and relevant prior evidence. As Chapter 3's
significance-filter example showed, selecting statistically significant
results can favour exaggerated estimates; an observed difference from
such a selection is not automatically a suitable planning alternative.

## Reading Bridge {#ch06-reading-bridge}

The chapter compared group means, separated omnibus questions from
calibrated contrasts, and read a comparison that changes across another
factor. A separate baseline example showed what an adjusted comparison
holds fixed. Counts required a different response model; likelihood and an
objective function made the fitting criterion explicit. Repetition examined
procedures under a declared generating model, while power and precision
described different planning aims.

When reading a study, identify the question, analysed records, model,
estimate and uncertainty procedure before evaluating their adequacy.
For a group comparison, locate its subtraction direction and any covariates
held fixed. For a test or interval, identify the null, reference distribution
and comparison family. For a simulation or planning result, locate the
generating assumptions and the particular procedure being evaluated.
These details connect the reported numbers to the claims they can support.

Chapter 7 develops how a continuous-predictor line is fitted and read.
Chapter 8 extends adjustment and distinguishes its conditional,
standardized and predictive questions. Chapter 11 returns to likelihood for
binary regression. Appendix B describes these connections more
specifically.

\Needspace{28\baselineskip}

### Key terms and notation {.unnumbered}

ANOVA, contrasts, interactions and comparison families retain the meanings
introduced in Chapter 4. The following concepts were introduced or developed
through their R applications here.

| Term | Definition |
|:--|:--|
| ANCOVA (analysis of covariance) | A linear model comparing groups while including quantitative covariates; its adjusted comparison is defined at stated covariate values or over a stated covariate distribution. |
| Model formula (R formula) | Notation such as `score ~ group` naming an outcome and the variables that explain it for a fitting function. It does not state whether observations are independent or paired; that comes from the design. |
| Objective function | A numerical criterion evaluated at candidate parameter values for fixed data; fitting seeks a stated minimum or maximum. |
| Likelihood and log-likelihood | The probability or density of the observed data under a declared model, viewed as a function of its parameters; the log-likelihood is its logarithm. |
| Maximum likelihood estimation (MLE) | Choosing parameter values that maximize the likelihood, equivalently the log-likelihood, over the stated parameter space. |

: Selected concepts introduced or developed here. {#tbl-ch06-terms tbl-colwidths="[30,70]"}

The notation follows the local question. $C=\sum_jc_j\mu_j$ is a contrast,
whereas $L(\theta)$ is a likelihood function of a parameter. Similarly, Type II error probability
$\beta$ differs from a model coefficient such as $\beta_x$.
Observed Pearson $X^2$ differs from the reference distribution $\chi^2$.
In simulations, $n$ counts observations per stated study or group and $B$
counts repeated studies. An SE describes an estimator's sampling variation;
an MCSE describes finite-simulation uncertainty in a computed summary.

### Further reading {.unnumbered}

For the linear-model basis of ANOVA, contrasts and covariance adjustment,
see @kutner2005. @tukey1953 develops simultaneous comparison ideas, and
@wilson1927 provides the score-interval construction for a proportion.
@cohen1988 discusses power and the dependence of a planning result on its
chosen effect and procedure, and @cetinkaya2024 develops the
simulation-based treatment of operating characteristics used in Section 7.
R's documentation gives the arguments and returned components for `aov`,
`lm`, `TukeyHSD`, `oneway.test`, `prop.test`, `chisq.test` and `optimize`
[@rcore2025]. The plotting chapters of
@wickham2023 develop the data, mapping and layer vocabulary used in the
small `ggplot2` examples here.

## Exercises {.unnumbered}

Use the chapter's printed data. Where a prompt supplies rounded
summaries, answers within 0.02 in the stated units are sufficient unless
that prompt requests a different precision. Keep each interpretation
beside its target, denominator and working conditions.

<!-- prompt-id: CH06-E01 -->

### Exercise 1: Read and reconstruct the one-way fit {.unnumbered #ex-ch06-01}

Use the twelve Massed scores. Find the three means and grand mean. Reconstruct
the between-group and within-group sums of squares, their degrees of freedom
and the F statistic. State the equal-means null, explain what its p-value
addresses under the working model, and calculate eta-squared.

<!-- END EX06-01 PROMPT -->

<!-- prompt-id: CH06-E02 -->

### Exercise 2: Keep the comparison and its family together {.unnumbered #ex-ch06-02}

For the same scores, compare Answer feedback and Explanatory feedback separately
with Rereading. Write both coefficient vectors, estimates and SEs. Construct
the 95% Bonferroni intervals for these two comparisons and compare them with
the corresponding all-pairs Tukey intervals. Explain why their limits differ.

<!-- END EX06-02 PROMPT -->

<!-- prompt-id: CH06-E03 -->

### Exercise 3: Read interaction as a change in comparison {.unnumbered #ex-ch06-03}

Use all 24 feedback/practice scores. Produce the six cell means and the
interaction plot. Calculate each feedback group's Spaced-minus-Massed change.
Fit the full two-way model and report the interaction F test. For Answer's
change minus Rereading's change, calculate the estimate, SE and 95% Scheffe
interval over the two-dimensional interaction family. Explain how that
interval can include zero when the omnibus interaction is significant.

<!-- END EX06-03 PROMPT -->

<!-- prompt-id: CH06-E04 -->

### Exercise 4: State what adjustment holds fixed {.unnumbered #ex-ch06-04}

Use the separately printed twelve-pupil pretest/posttest data. Compare the
raw group means with the fitted means at pretest 8 under the common-slope
model. Reproduce the Answer-minus-Rereading estimate and its pointwise
interval. Compare the baseline-only and full models on identical records.
Explain the conditioning target, the common-slope assumption and why these
authored results are not empirical evidence about feedback.

<!-- END EX06-04 PROMPT -->

<!-- prompt-id: CH06-E05 -->

### Exercise 5: Preserve the event, denominator and expected counts {.unnumbered #ex-ch06-05}

In the invented table, Online has 32 completions and 18 non-completions;
In person has 18 completions and 32 non-completions. For Online, test p=.5 with the uncorrected score
procedure and report its 95% score interval. For the complete table, compute
all expected counts and Pearson's X-squared statistic. Explain why the two
tests answer different questions and what design information an effect claim
would additionally need.

<!-- END EX06-05 PROMPT -->

<!-- prompt-id: CH06-E06 -->

### Exercise 6: Distinguish study size, repetitions and planning targets {.unnumbered #ex-ch06-06}

Use the chapter's ANOVA simulation table. Explain the changes from n=4 to
n=12 under the null and alternative models. Calculate MCSE for the .6545
power estimate with B=2000. Explain what increasing B changes. Then compare
the n=17 power target with the n=35 planned-width target for the two-comparison
family; say what each target controls and why a realised interval can have a
different width from its planning value.

<!-- END EX06-06 PROMPT -->

<!-- prompt-id: CH06-E07 -->

### Exercise 7: Distinguish an estimate from its fitting criterion {.unnumbered #ex-ch06-07}

Keep 32 successes in 50 trials fixed. Write the binomial likelihood as a
function of p, find its analytic maximizer, and reproduce it using the
negative-log-likelihood and `optimize()`. State the returned probability and
objective value, compare the criterion at p=.5, and explain why a likelihood
curve is not a probability distribution over p.

<!-- END EX06-07 PROMPT -->

<!-- prompt-id: CH06-E08 -->

### Exercise 8: Write a qualified result paragraph {.unnumbered #ex-ch06-08}

Choose the one-way comparisons, selected interaction contrast or adjusted
group comparison from this chapter. Write a short paragraph giving the
question, data and denominator, estimate, interval and its family or
pointwise status, working conditions and scope. Identify one tempting claim
that the numerical result does not establish, and explain the missing support.

<!-- END EX06-08 PROMPT -->

\clearpage

## Appendix A: Alternative calculations on the same data {.unnumbered #ch06-appendix-a}

These short calculations expose relationships behind the main procedures.
They use the objects created in the chapter. The Computer Lab references
provide the corresponding variations and complete starting setups.

### Direct sums of squares and omega-squared {.unnumbered}

The fitted object's residual sum of squares can be checked directly from
the scores and group means. The following calculation reconstructs all
three sums of squares and the observed eta-squared summary.

```{r}
#| label: ch06-app-direct-ss
grand <- mean(massed$score)
ss_between <- sum(4 * (means-grand)^2)
ss_within <- sum(resid(one_fit)^2)
ss_total <- sum((massed$score-grand)^2)
c(SS_B = ss_between, SS_W = ss_within,
  SS_T = ss_total, eta2 = ss_between/ss_total)
```

The values 32, 30 and 62 reproduce the main table, with
$32+30=62$ and $\eta^2=.5161$. For this one-way model, a conventional
bias-adjusted variance-share estimator is

$$
\widehat\omega^2=
\frac{SS_B-(J-1)MS_W}{SS_T+MS_W}
=\frac{32-2(30/9)}{62+30/9}=.3878.
$$

```{r}
#| label: ch06-app-omega
(ss_between - 2*mse) / (ss_total + mse)
```

Subtracting the error contribution from the numerator makes the estimate
smaller here. In a sample with little between-group variation, that numerator
can be negative. A negative estimate is a consequence of the bias adjustment,
not a negative population variance share. Report the convention used if a
display truncates it to zero. As with eta-squared, omega-squared does not
replace the outcome-scale differences, their intervals or the design
conditions for interpreting them.

### Match the two-group t and F calculations {.unnumbered}

Restrict the Massed records to Rereading and Answer feedback. A pooled
equal-variance t test and a two-group ANOVA then use the same eight pupils,
two fitted means and six residual degrees of freedom.

\Needspace{20\baselineskip}

```{r}
#| label: ch06-app-t-squared
two_groups <- droplevels(subset(massed,
  feedback %in% c("Rereading", "Answer feedback")))
pooled <- t.test(score ~ feedback, data = two_groups,
  var.equal = TRUE)
two_group_fit <- aov(score ~ feedback, data = two_groups)
c(t = unname(pooled$statistic),
  t_squared = unname(pooled$statistic)^2,
  p = pooled$p.value)
anova(two_group_fit)
```

The t statistic is $-1.5492$ because this formula call subtracts Answer's
mean from Rereading's. Its square is $F=2.4000$, and both tests
give p .1723. Squaring loses the sign but preserves the evidence against
equal means for this matched procedure. Welch is a different calibration
and does not generally share this identity; the equal group sizes and
equal observed variances here happen to give the same numerical result.
Using different records or the three-group residual denominator changes
the calculation being compared.

### Change the interval calibration for the selected DD {.unnumbered}

The selected Answer-minus-Rereading practice DD stays 4 with SE 1.8257.
Only the multiplier changes between its pointwise t interval and the
simultaneous Scheffe interval for the two-dimensional interaction family.

```{r}
#| label: ch06-app-dd-intervals
dd + c(-1, 1) * qt(.975, 18) * se_dd
dd + c(-1, 1) * scheffe * se_dd
```

The pointwise interval [0.1643, 7.8357] excludes zero; the simultaneous
interval [$-0.8680$, 8.8680] includes it. Selecting the narrower interval after
seeing which conclusion it yields does not preserve the broader family's
guarantee. Choose and describe the inferential question and calibration.

For comparison, Answer's simple practice effect uses only two means:

\Needspace{10\baselineskip}

```{r}
#| label: ch06-app-simple-effect
simple_answer <- cell_mean["Answer feedback", "Spaced"] -
  cell_mean["Answer feedback", "Massed"]
se_simple <- sqrt(mse_two * (1/4 + 1/4))
c(estimate = simple_answer, SE = se_simple)
simple_answer + c(-1, 1) * qt(.975, 18) * se_simple
```

Its pointwise interval is [1.2877, 6.7123]. The smaller SE reflects a
different contrast. It does not acquire a simultaneous guarantee over
simple effects from the interaction-only Scheffe calculation.

### Compute the Wilson endpoints directly {.unnumbered}

For one proportion, write $z=z_{.975}$, $n=50$ and $\widehat p=32/50$.
Inverting the score inequality gives a center and half-width

$$
\begin{aligned}
\text{center}&=\frac{\widehat p+z^2/(2n)}{1+z^2/n},\\
\text{half-width}&=
\frac{z\sqrt{\widehat p(1-\widehat p)/n+z^2/(4n^2)}}{1+z^2/n}.
\end{aligned}
$$

```{r}
#| label: ch06-app-wilson
phat <- 32/50
z <- qnorm(.975)
center <- (phat + z^2/(2*50)) / (1 + z^2/50)
half <- z * sqrt(phat*(1-phat)/50 + z^2/(4*50^2)) /
  (1 + z^2/50)
center + c(-1, 1)*half
```

The endpoints [.5014, .7586] reproduce the uncorrected one-proportion
`prop.test()` interval. The altered center and half-width are consequences
of score-test inversion. The denominator and Bernoulli working model remain
those of the original 50 Online records.

\Needspace{26\baselineskip}

## Appendix B: What later chapters develop {.unnumbered #ch06-appendix-b}

Chapter 7 extends squared loss from a fitted center to a line. Its appendix
recalls the mean-loss identity, derives the line's normal equations and
relates fitted values and residuals through projection. It also develops
uncertainty for the slope and fitted mean. The scalar numerical search here
supplies a first fitting example; those later derivations explain the linear
model's structure.

Chapter 8 extends adjustment to several controls. It distinguishes a
coefficient conditional on specified predictors from a fitted mean at a
profile, a prediction for a new case, and an average of fitted values over
a named population. Coefficient covariances help calculate uncertainty for
these quantities. The chapter also develops common support and the difference
between observational and randomized-design interpretations. Chapter 9's
separate Baumann ANCOVA reference retains the original study's outcome
definitions, observed denominators and missing-data qualifications; it is
distinct from the twelve authored pupils analysed here.

Chapter 11 develops likelihood for binary regression. Its reference material
works through exact boundary estimates and their logit limits, and a
Normal likelihood with more than one parameter. Numerical fitting then
requires attention to parameter constraints, starting values and convergence
checks. The scalar 32-of-50 calculation here introduces the objective and
its maximizer while keeping those more extensive issues with their later
applications.

## References {.unnumbered}
