---
title: "Simple Linear Regression"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \widowpenalty=10000
          \clubpenalty=10000
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AtBeginDocument{\renewcommand{\boldsymbol}[1]{\symbf{#1}}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                     fig.pos = "H", collapse = TRUE, comment = "#>")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 10, width = 68)
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 11))
# Display helper for authored tables: round once to four decimals, give a
# column the fewest common decimals that show every value, a true minus
# sign, and a bound for a nonzero value that rounds to zero.
show_value <- function(x) {
  shown <- format(round(x, 4), digits = 15, scientific = FALSE, trim = TRUE)
  tiny <- x != 0 & round(x, 4) == 0
  shown[tiny] <- ifelse(x[tiny] > 0, "< 0.0001", "> \u22120.0001")
  unname(sub("^-", "\u2212", shown))
}
```

\clearpage

::: {.callout-note title="Chapter overview"}
This chapter reads a fitted line. Starting from a reported claim about
education and income, we state an occupation-level question, plot the
records, and fit a straight line to them. We read the slope, intercept,
fitted values, and residuals; see how least squares chooses the line and
how the slope is covariance divided by predictor variation; and describe
how much variation the line leaves. We then declare a separate process
question, derive the slope's standard error and interval, and watch
fitted slopes vary across repetitions of a known process. We distinguish
a mean response from one new outcome, check form, spread, influence, and
support, and close with Galton's heights of fathers and sons, where the
line in standard units explains regression toward the mean. By the end
of the chapter you will be able to fit and read a simple regression in R,
state what its descriptive and model-based answers concern, and recognize
regression toward the mean in a before-and-after claim. An appendix
develops the mathematics further.
:::

Suppose a report says, "Another year of education increases income by about
900 dollars." Before examining its significance, ask whose education and whose
income were measured. The same number could refer to individuals, households,
or occupational averages. Those are different comparisons.

\Needspace{6\baselineskip}

Our question is deliberately narrower:

> Among the 102 recorded Canadian occupations, how did occupation-average
> income differ across the supported range of occupation-average education in
> 1971, as summarized by a fitted straight line?

One row represents an **occupation**, not a worker. An occupation with higher
average education need not contain the same people as an occupation with lower
average education. A comparison between their averages does not describe what
would happen to one person's income after another year of schooling.
Aggregate and individual relationships can differ substantially [@robinson1950].

Chapter 5 connected covariance and correlation to recorded values and
made mean and correlation inference executable in R. Chapter 6 fitted
group means through indicator coefficients in a model formula. Here we
develop the fitted line for a continuous predictor: how its slope is
calculated, what its uncertainty requires, and which comparisons and
predictions the records support.

## A Question About Occupations and a Line Through Their Records {#sec-ch7-question}

### Begin with the unit and the question

The data are distributed as *carData::Prestige* [@carData2022; @fox2019].
The package documentation attributes education and income to the *Census of
Canada* (1971), volume 3, part 6, pages 19-1 to 19-21.
We read the package documentation as a source record, not as evidence that
these occupations form a probability sample. The chapter's calculations are
our reanalysis, not statements attributed to an original research paper.

| Feature | Meaning in this chapter |
|:--|:--|
| Unit and record count | 102 occupations; each occupation receives equal weight |
| Education, $x$ | Average years of education of occupational incumbents in 1971 |
| Income, $y$ | Average income of incumbents, divided by 1,000; thousand 1971 Canadian dollars |
| Missingness for the fit | No missing education or income values; no modeled rows excluded |
| Observed support | Education: 6.38--15.97 years; income: 0.611--25.879 thousand dollars |
| Descriptive estimand | Least-squares slope summarizing these 102 fixed occupation records |
| Limits of the source record | No probability-sampling basis or individual-level analysis established here |

: Record, measurements, and scope for the focal analysis. {#tbl-ch7-record tbl-colwidths="[30,70]"}

Equal weighting matters. A small occupation and a large occupation contribute
one row each. Without suitable counts and a different estimand, this analysis
does not describe an average worker's experience. Historical dollars also
cannot be read as contemporary purchasing power.

### Look before fitting

```{r}
#| label: ch7-data-visible
#| echo: true
prestige <- carData::Prestige
prestige$income_k <- prestige$income / 1000
library(ggplot2)
```

The package-qualified name `carData::Prestige` selects the supplied dataset.
Dividing income by 1,000 changes its units while preserving every occupation.
The Computer Lab supplies installation and operating practice if needed; the
occupation example requires no separate data file.

```{r}
#| label: fig-ch7-scatter
#| echo: true
#| fig-cap: "Education and income in the 102 occupation records, before fitting a line."
#| fig-alt: "Occupation-average income generally rises with occupation-average education, with considerable scatter and some especially high-income occupations."
#| fig-width: 6.3
#| fig-height: 3.8
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, size = 2) +
  labs(x = "Average education (years)",
       y = "Average income (thousand 1971 dollars)")
```

The scatterplot gives the line a job: summarize an overall upward pattern
amid substantial variation. It also shows why the line cannot describe every
occupation equally well. General managers, for example, have an average
education of 12.26 years and an average income of 25.879 thousand dollars;
university teachers have corresponding values of 15.97 and 12.480.
More education does not order every pair of occupations by income.

The scatterplot motivates fitting a straight-line summary. Save that fit
as `model`, and keep its coefficients in `b` for calculations that follow.

```{r}
#| label: ch7-prepare-visible
#| echo: true
model <- lm(income_k ~ education, data = prestige)
b <- coef(model)
coef(model)
```

The formula puts the outcome to the left of the tilde and the predictor to its
right. The command finds coefficients. It does not choose a population,
establish a causal design, or justify a confidence interval.

## Read the Fitted Line and Understand How OLS Finds It {#sec-ch7-line}

### Fitted values, slopes, and residuals

A straight-line summary assigns each observed predictor value $x_i$ a
**fitted value** $\widehat y_i$:

$$
\widehat y_i=\widehat\beta_0+\widehat\beta_1x_i,
\qquad e_i=y_i-\widehat y_i.
$$

The subscript $i$ indexes a record. The hat marks a fitted quantity.
The **residual** $e_i$ is observed income minus fitted income. A positive
residual places an occupation above the line; a negative residual places it
below. A residual is calculated after fitting and is not the same as an
unobservable error from a population model.

For the occupation records, the line is

$$
\widehat{\mathrm{income}}=-2.8536+0.8988\,\mathrm{education}.
$$

Income is still in thousands of dollars. The slope says that the fitted
income differs by 0.8988 thousand dollars, about 899 dollars, for a one-year
difference in occupation-average education. A two-year difference along this
line corresponds to $2\widehat\beta_1\approx1.7976$ thousand dollars.

For university teachers, use the fitted coefficients before rounding:

$$
\widehat y=\widehat\beta_0+\widehat\beta_1(15.97)
\approx11.5005,\qquad
e=12.4800-\widehat y\approx0.9795.
$$

Their recorded average income is about 980 dollars above the fitted value.
The printed regression equation abbreviates the coefficients; the stored
model retains the digits needed for these calculations. R reproduces the
fitted value and signed residual directly:

```{r}
#| label: ch7-teacher-visible
#| echo: true
teacher <- prestige["university.teachers", ]
teacher_fit <- predict(model, teacher)
c(fitted = unname(teacher_fit),
  residual = teacher$income_k - unname(teacher_fit),
  at_12 = unname(predict(model, data.frame(education = 12))))
```

For an occupation averaging 12 years of education, the fitted value is
7.9322 thousand dollars. `predict()` applies the saved line to the named
education value; it does not refit the model.

@tbl-ch7-slope-readings sets the slope's reading beside two sentences a
report might also write about it.

\Needspace{12\baselineskip}

| A sentence about the slope 0.8988 | What it compares | Do these records support it? |
|:--|:--|:--|
| Occupations whose average education is one year higher have fitted average income about 899 dollars higher | Different occupations, along the fitted line | Yes, as a description of the 102 records over 6.38--15.97 years |
| If an occupation's average education rose by one year, its average income would rise by about 899 dollars | One occupation before and after a change | No: it needs records of occupations over time and a design that isolates the change |
| Another year of schooling raises a worker's income by about 899 dollars | One person before and after more schooling | No: it needs records of individuals and a causal design; occupation averages cannot supply either |

: Three readings of one slope; only the first is a comparison these records make. {#tbl-ch7-slope-readings tbl-colwidths="[40,26,34]"}

Only the first sentence compares what the records compare: occupations
that differ in average education. The second describes a change within
one occupation over time and the third a change within one person. The
records observe neither, and a comparison between occupation averages
cannot stand in for either one [@robinson1950]. The slope supports the
first reading and is silent about the other two.

The intercept places the line vertically: it is its fitted income at
education zero. Zero is outside the observed range, so the negative intercept
is not evidence that any recorded occupation has negative average income.
Centering education at 12 would change the intercept to 7.9322 while leaving
the slope and all fitted values unchanged.

```{r}
#| label: fig-ch7-line-residual
#| fig-cap: "The least-squares line passes through the mean point; the marked vertical segment is the university-teacher residual."
#| fig-alt: "A fitted line crosses the mean education and income. At 15.97 years, a vertical segment joins the fitted value to the higher university-teacher observation."
#| fig-width: 6.3
#| fig-height: 3.8
teacher <- prestige["university.teachers", ]
teacher$fit <- predict(model, teacher)
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, colour = "grey45") +
  geom_abline(intercept = b[1], slope = b[2], linewidth = .8) +
  annotate("point", x = mean(prestige$education),
    y = mean(prestige$income_k), shape = 18, size = 3) +
  geom_segment(data = teacher, aes(xend = education, yend = fit),
               linewidth = 1) +
  labs(x = "Average education (years)",
       y = "Average income (thousand 1971 dollars)")
```

### Why this particular line?

Ordinary least squares (**OLS**) chooses the intercept and slope that minimize
the **residual sum of squares**:

$$
RSS=\sum_{i=1}^{n}(y_i-\widehat\beta_0-\widehat\beta_1x_i)^2.
$$

The summation sign says to add the squared residual from every record.
Squaring prevents positive and negative deviations from cancelling and gives
large deviations more influence on the criterion. "Best" here means best by
this criterion among straight lines with an intercept, not best for every
research purpose.

Two centered sums lead directly to the solution:

$$
S_{xx}=\sum_i(x_i-\bar x)^2,\qquad
S_{xy}=\sum_i(x_i-\bar x)(y_i-\bar y),
$$

$$
\widehat\beta_1=\frac{S_{xy}}{S_{xx}}
=\frac{s_{xy}}{s_x^2}
=r\frac{s_y}{s_x},
\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x.
$$

Here $\bar x,\bar y$ are means; $s_x,s_y$ are sample standard deviations;
$s_{xy}$ is sample covariance; and $r$ is Pearson's correlation.
The variance and covariance both divide their centered sums by $n-1$, so
that divisor cancels in their ratio. The slope has income-per-education units;
correlation has no units.

For these data,

$$
\widehat\beta_1=\frac{675.8041}{751.8852}\approx0.8988,
\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x\approx-2.8536.
$$

Chapter 5's covariance and correlation matrices give the same calculation
in a reusable form. For the same 102 complete education--income pairs, in
education--income order,

$$
\mathbf S=\begin{pmatrix}7.4444&6.6911\\6.6911&18.0279\end{pmatrix},
\qquad
\mathbf R=\begin{pmatrix}1&.5776\\.5776&1\end{pmatrix}.
$$

The covariance matrix records education variance in years squared, income
variance in squared thousand dollars, and their covariance in
thousand-dollar-years. Dividing covariance by education variance cancels
the common $n-1$ divisor and gives the slope's thousand-dollars-per-year
units. Correlation removes units; multiplying it by the income SD divided
by the education SD restores those same slope units. Computed from the
unrounded entries, both routes return 0.8988: they reproduce the
centered-sum calculation rather than define a different fitted line. The
Computer Lab carries out the centered sums and both matrix routes from the
records.

The shared records matter as much as the formula. Entries formed by
separate pairwise deletion can describe different sets of occupations;
they need not reconstruct a regression fitted on one common set. No such
missingness occurs in these two columns.

Positive covariance produces a positive slope, provided education varies at all.
If every occupation had the same education, $S_{xx}=0$ and a slope could
not be identified.

The fitted line passes through $(\bar x,\bar y)$, because substituting
$\bar x$ into the intercept formula gives $\bar y$.
With an intercept, the residuals also sum to zero and their products with
$x$ sum to zero. Otherwise the line could be shifted or tilted to reduce
RSS. These are algebraic properties of OLS, not tests of independence or
linearity. Full normal-equation and projection-matrix derivations are retained
in **Appendix A**.

### A tiny calculation you can inspect completely

Three teaching records, $(x,y)=(1,2),(2,4),(3,3)$, show the whole
least-squares calculation at once. Their means are $\bar x=2$ and
$\bar y=3$. @tbl-ch7-tiny lays out the centered values, their squares and
products, the fitted values and the residuals; its last row sums each
column.

\Needspace{12\baselineskip}

| $x$ | $y$ | $x-\bar x$ | $y-\bar y$ | $(x-\bar x)^2$ | $(x-\bar x)(y-\bar y)$ | $\widehat y$ | $e$ | $e^2$ |
|--:|--:|--:|--:|--:|--:|--:|--:|--:|
| 1 | 2 | $-1$ | $-1$ | 1 | 1 | 2.5 | $-0.5$ | 0.25 |
| 2 | 4 | 0 | 1 | 0 | 0 | 3.0 | 1.0 | 1.00 |
| 3 | 3 | 1 | 0 | 1 | 0 | 3.5 | $-0.5$ | 0.25 |
| Sum |  | 0 | 0 | 2 | 1 | 9.0 | 0.0 | 1.50 |

: The complete least-squares calculation for three records; the column sums give $S_{xx}$, $S_{xy}$ and $RSS$. {#tbl-ch7-tiny tbl-colwidths="[9,9,11,11,12,17,11,10,10]"}

The sums give $S_{xx}=2$ and $S_{xy}=1$, so

$$
\widehat\beta_1=\frac{S_{xy}}{S_{xx}}=\frac12=0.5,\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x=3-0.5(2)=2.
$$

The fitted values 2.5, 3.0 and 3.5 lie on this line. The residuals sum to
zero, as the OLS properties above require, and their squares sum to
$RSS=1.5$. No normality assumption is needed for these calculations. A
stochastic model becomes necessary when we ask about repeated-sample
uncertainty.

### Compare candidate lines with the OLS minimum

The least-squares criterion also lets us compare the fitted line with
alternatives. To keep the comparison transparent, make every candidate line
pass through $(\bar x,\bar y)$ and vary only its slope. For such a line,

$$
RSS(b_1)=RSS(\widehat\beta_1)
       +S_{xx}(b_1-\widehat\beta_1)^2.
$$

The added term is zero only at the OLS slope and positive for every other
slope when $S_{xx}>0$. The following table checks five deliberately chosen
lines on the occupation records. These are alternative summaries of the same
records, not repeated samples.

```{r}
#| label: ch7-candidate-lines-code
x <- prestige$education
y <- prestige$income_k
candidate_slopes <- unname(b[2]) + c(-0.6, -0.3, 0, 0.3, 0.6)
candidate_intercepts <- mean(y) - candidate_slopes * mean(x)
ols_rss <- sum(resid(model)^2)
predictor_spread <- sum((x - mean(x))^2)
candidate_rss <- ols_rss +
  predictor_spread * (candidate_slopes - b[2])^2
candidate_lines <- data.frame(
  Intercept = candidate_intercepts,
  Slope = candidate_slopes,
  RSS = candidate_rss
)
```

```{r}
#| label: tbl-ch7-candidate-lines
#| tbl-cap: "Candidate lines through the mean point and their residual sums of squares."
#| echo: false
knitr::kable(data.frame(lapply(candidate_lines, show_value)),
  align = "r")
```

```{r}
#| label: fig-ch7-candidate-lines
#| echo: false
#| fig-cap: "Five candidate lines through the sample mean. The OLS line has the smallest residual sum of squares."
#| fig-alt: "Five lines with different slopes cross at the mean occupation point. The highlighted OLS line runs centrally through the data; steeper and flatter alternatives have larger residual sums of squares."
#| fig-width: 6.3
#| fig-height: 3.8
candidate_grid <- expand.grid(
  education = seq(min(x), max(x), length.out = 100),
  Candidate = seq_along(candidate_slopes)
)
candidate_grid$income <- candidate_intercepts[candidate_grid$Candidate] +
  candidate_slopes[candidate_grid$Candidate] * candidate_grid$education
candidate_grid$Line <- ifelse(
  candidate_grid$Candidate == 3, "OLS", "Alternative"
)
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, colour = "grey55", size = 1.4) +
  geom_line(data = candidate_grid,
            aes(y = income, group = Candidate, colour = Line,
                linewidth = Line)) +
  scale_colour_manual(values = c(Alternative = "grey65", OLS = "black")) +
  scale_linewidth_manual(values = c(Alternative = .45, OLS = 1)) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)", colour = NULL, linewidth = NULL)
```

Every alternative has a larger RSS than the OLS line, and slopes the same
distance above and below 0.8988 add the same excess, as the identity
predicts. The comparison also explains why high-income occupations can exert
substantial influence: squaring gives large vertical discrepancies extra
weight. That is a feature of the criterion, not evidence that those
occupations are invalid.

### Centering changes the reference point, not the line

The fitted equation can be written

$$
\widehat y_i=\bar y+\widehat\beta_1(x_i-\bar x).
$$

Centering education at its observed mean makes the intercept equal to
$\bar y=6.7979$ thousand dollars, the fitted income at average education.
It leaves the slope, fitted values, residuals, RSS, and $R^2$ unchanged.
Centering at 12 instead makes the intercept the already computed fitted value
7.9322. Thus centering can make the intercept interpretable without changing
the comparison represented by the line. The Computer Lab refits the centered
line and checks that every fitted value is unchanged.

Changing income from thousands of dollars to dollars would multiply the
intercept, slope, residual SD, and coefficient SEs by 1,000, while leaving
$t$ statistics, $p$ values, correlation, and $R^2$ unchanged. A unit change
is not new evidence or a stronger relationship.

### How much variation remains? {#sec-ch7-fit}

A fitted line can show a clear linear trend and still leave large differences
between occupations unexplained by its fitted values.

For OLS with an intercept,

$$
TSS=\sum_i(y_i-\bar y)^2=RegSS+RSS,
\qquad RegSS=\sum_i(\widehat y_i-\bar y)^2.
$$

For the occupation records, $1820.8134=607.4214+1213.3920$.
This is a decomposition of squared deviations, in squared thousand-dollar
units. It holds because the cross-product of fitted deviations and residuals
is zero.

The coefficient of determination is

$$
R^2=1-\frac{RSS}{TSS}
=1-\frac{1213.3920}{1820.8134}\approx.3336.
$$

The fitted line accounts for 33.3599% of the variation in recorded income
around its mean, according to this squared-deviation measure.
"Accounts for" or "explains" in this algebraic sense does not mean causes.
For simple regression with an intercept, $R^2=r^2$:
$.5776^2\approx.3336$. The identity need not hold for other models or a
no-intercept fit.

The residual standard deviation is

$$
s_e=\sqrt{\frac{RSS}{n-2}}
=\sqrt{\frac{1213.392}{100}}=3.4834.
$$

Its units are thousand dollars, the same as income. Fitting an intercept and
slope imposes two independent residual constraints, leaving $n-2$ residual
degrees of freedom. Under the constant-variance model introduced next,
$s_e$ estimates the error standard deviation. Without that model it remains
a scaled measure of residual scatter; it is not the standard error of the
slope or a guaranteed typical prediction error in new records.

`sigma()` returns the residual SD of a fitted model, and `summary()`
reports its $R^2$ among the other fit summaries:

```{r}
#| label: ch7-fit-summaries-visible
#| echo: true
c(residual_SD = sigma(model), R_squared = summary(model)$r.squared)
```

Here and below, $SE$ denotes an estimated standard error. The final row
anticipates the separately declared process in the next section; it is
not uncertainty about the already computed finite-record slope.

| Quantity | Occupation result | What it describes |
|:--|--:|:--|
| Slope | 0.8988 | Change in fitted thousand-dollar income per education year |
| Residual SD | 3.4834 | Remaining income scatter on the outcome scale |
| $R^2$ | .3336 | Proportion of centered squared variation accounted for in these records |
| Classical slope SE | 0.1270 | Model-based repetition variability of the slope, under the next assumptions |

: Different summaries have different units and jobs. {#tbl-ch7-quantities tbl-colwidths="[24,22,54]"}

\Needspace{16\baselineskip}

## Make and Read a Model-Based Inference {#sec-ch7-inference}

### Announce the inferential extension before its interval

The finite-record slope is not waiting to be discovered in a larger population:
we have computed it for all 102 records named in the focal question.
A confidence interval answers an additional question.

For a **separable teaching extension**, suppose occupations' outcomes arose
independently from a process with conditional mean

$$
E(Y_i\mid X_i=x_i)=\beta_0+\beta_1x_i.
$$

The estimand $\beta_1$ is now a constant conditional-mean slope in that
hypothetical process, not the fixed slope of these 102 records. Repetition
means generating new outcomes from that process at the stated predictor values
and refitting. It does not mean that the documented occupations were randomly
sampled. Changing the estimand changes the question--answer pair.

Write the process as

$$
Y_i=\beta_0+\beta_1x_i+\varepsilon_i,\qquad
E(\varepsilon_i\mid\mathbf X)=0,\qquad
\operatorname{Var}(\varepsilon_i\mid\mathbf X)=\sigma^2.
$$

For the exact classical $t$ procedure, errors are independent and conditionally
normal with this common variance, and $S_{xx}>0$. The $x_i$ values need not
be normal; neither must the marginal distribution of $Y$. It is the
conditional error distribution, not the histogram of every variable, that
appears in the model. Independence is supported by how records arise, not by a
good-looking residual plot.

```{r}
#| label: fig-ch7-conditional
#| fig-cap: "What the process conditions claim, drawn on the occupation records: at each education value, a Normal distribution of incomes centred on the line, with the same spread everywhere."
#| fig-alt: "The 102 occupation records and the fitted line rising from about 3 to 12 thousand dollars over education 6 to 16. At 8, 10, 12 and 14 years, an identical bell-shaped curve stands sideways on the line, spreading about 10 thousand dollars above and below it; the curve at 8 years extends below zero income, while several high-income records at 12 years and above lie beyond the curves' upper tails."
#| fig-width: 6.3
#| fig-height: 3.4
curve_sd <- sigma(model)
curves <- do.call(rbind, lapply(c(8, 10, 12, 14), function(x0) {
  mu <- unname(predict(model, data.frame(education = x0)))
  income <- seq(mu - 3 * curve_sd, mu + 3 * curve_sd, length.out = 120)
  data.frame(at = x0, income = income,
    education = x0 + 1.4 * dnorm(income, mu, curve_sd) /
      dnorm(0, 0, curve_sd))
}))
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, colour = "grey60", size = 1.3) +
  geom_abline(intercept = b[1], slope = b[2], linewidth = .8) +
  geom_segment(data = unique(curves["at"]),
    aes(x = at, xend = at,
      y = b[1] + b[2] * at - 3 * curve_sd,
      yend = b[1] + b[2] * at + 3 * curve_sd),
    colour = "grey40", linewidth = .3, inherit.aes = FALSE) +
  geom_path(data = curves, aes(education, income, group = at),
    linewidth = .7, inherit.aes = FALSE) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)")
```

@fig-ch7-conditional draws what these conditions claim, using the fitted
line and the residual SD, 3.4834, in place of the unknown $\beta_0$,
$\beta_1$ and $\sigma$. At every education value the model places a Normal
distribution of incomes centred on the line, with one common spread. The
records do not look like that everywhere: at higher education they spread
more widely, and several incomes lie far beyond the upper tails, which
Section 6 examines. At 8 years the curve reaches below zero, a first sign
that a Normal model for a nonnegative outcome is an approximation, which
Section 5 meets again.

To see where slope uncertainty comes from, substitute the process equation
into the OLS slope formula. Here $\mathbf X$ denotes the predictor values on which
we condition. Since $\sum_i(x_i-\bar x)=0$,

$$
\widehat\beta_1-\beta_1
=\frac{\sum_i(x_i-\bar x)\varepsilon_i}{S_{xx}}.
$$

The fitted slope differs from the process slope because the realized errors
need not cancel when weighted by distance from mean education. Independence
makes their covariance terms zero; common variance gives

$$
\operatorname{Var}(\widehat\beta_1\mid\mathbf X)
=\frac{\sigma^2\sum_i(x_i-\bar x)^2}{S_{xx}^2}
=\frac{\sigma^2}{S_{xx}}.
$$

This variance identity does not require normal errors. Normality enters the
exact $t$ reference. Taking the square root and estimating $\sigma$ by $s_e$
gives the slope standard error:

$$
SE(\widehat\beta_1)=\frac{s_e}{\sqrt{S_{xx}}}
=\frac{3.4834}{\sqrt{751.8852}}=0.127,
$$

$$
\widehat\beta_1\ \pm\ t_{.975,n-2}SE(\widehat\beta_1)
\approx0.8988\pm1.984(0.127)
\approx[0.6468,\ 1.1508].
$$

The critical value 1.984 is the 97.5th percentile of a $t$ distribution with
100 degrees of freedom. More error variation raises the SE; greater predictor
spread lowers it when the other quantities are held fixed. More observations
usually improve precision through their contribution to $S_{xx}$, but
dependent observations do not supply independent information merely by
increasing the row count.

The null test $H_0:\beta_1=0$ uses
$t=\widehat\beta_1/SE(\widehat\beta_1)\approx7.0753$ and a two-sided
$p<.0001$.
This is evidence against that zero-slope model under the stated procedure.
It is not the probability that the null is true, the probability that the
data occurred by chance, evidence of causation, or a measure of practical
importance. The interval describes the range produced by this uncertainty
method, not a 95% probability assigned to the fixed parameter after observing
this interval. The next section makes the repeated-sample interpretation
visible before prediction is introduced.

The coefficient table gives the estimate, its SE, and the test; the interval
call returns the corresponding two-sided limits. The function's default
confidence level is .95.

```{r}
#| label: ch7-coefficient-inference-visible
#| echo: true
coef(summary(model))
confint(model, "education")
```

The assumptions are a proposed justification, not facts created by the formula.
Curvature, changing spread, or dependence can invalidate the exact coverage
claim. Section 6 examines these conditions on the occupation records.

### Connect the Pearson and slope tests {#ch07-correlation-test}

Chapter 5 tested a zero Pearson correlation. The simple regression now
lets us connect that familiar statistic to the slope calculation. For an
OLS line with an intercept, fitted to the same complete pairs,

$$
\begin{aligned}
t_{\mathrm{slope}}
&=\frac{\widehat\beta_1}{s_e/\sqrt{S_{xx}}}
=r\sqrt{\frac{n-2}{1-r^2}}
=t_{\mathrm{Pearson}},\\
F_{\mathrm{regression}}
&=t_{\mathrm{slope}}^2
=\frac{R^2}{1-R^2}(n-2).
\end{aligned}
$$

The first identity follows by substituting the covariance expression for
the slope and using $RSS=(1-r^2)TSS$. The second compares the
intercept-only mean model with the intercept-and-slope model: their
numerator degrees of freedom differ by one. The signed $t$ retains the
slope's direction; squaring it gives the nondirectional $F$ statistic.

For the occupation line, Chapter 5's Pearson test and the slope test both
give $t=7.0753$ on 100 df, and the $F$ statistic that `summary()` prints
is $F(1,100)=50.0598=t^2$; the matched two-sided $t$ and upper-tail $F$
p-values are below .0001. These entries express the same
one-parameter comparison, not three independent pieces of evidence. The
Computer Lab reproduces all three from the fitted model.

Equality of the statistics is algebraic. Their exact classical references
use the independent, constant-variance Normal error model already stated.
For randomly sampled bivariate-Normal pairs, a zero population correlation
is also equivalent to a zero linear conditional-mean slope. The documented
occupation records provide no probability-sampling warrant: here the
Pearson call reproduces the declared process calculation numerically,
rather than adding a new population-correlation claim.

Match the intercept, rows, weights, and covariance convention before
comparing outputs. An HC3 slope test uses a different covariance estimate;
it need not equal the ordinary ANOVA $F$ calculation. Section 6 uses that
change to examine uncertainty while keeping the fitted line fixed.

### A binary predictor: the two-group comparison as a line {#ch07-binary-predictor}

A predictor need not be continuous. Chapter 4's invented feedback study
recorded a practice schedule for each of its 24 pupils, twelve under
massed practice and twelve under spaced practice, with a score out of 20;
Chapter 6 fitted the same scores by feedback condition with a model
formula. Code the schedule as $x_i=0$ for massed and $x_i=1$ for spaced,
and fit the same kind of line. With only two predictor values, the
least-squares line passes through the two group means
(@fig-ch7-binary): the intercept is the massed mean and the slope is the
difference of means,

$$
\widehat\beta_0=\bar y_0=8,\qquad
\widehat\beta_1=\bar y_1-\bar y_0=12-8=4.
$$

The residual SD is the pooled SD of the equal-variance procedure,
$s_e=4.156$ on $24-2=22$ df, and the slope SE formula becomes
$SE(\widehat\beta_1)=s_e\sqrt{1/n_0+1/n_1}=4.156\sqrt{1/12+1/12}=1.6967$,
so $t=4/1.6967=2.3575$ on 22 df with $p=.0277$ and interval
$[0.4813,\ 7.5187]$, which the Computer Lab reproduces with `t.test()`.
These are exactly the equal-variance two-sample $t$
test and interval for a difference in means (Chapters 3 and 5): a
regression on a 0/1 predictor is that procedure written as a line. The
matched statistics hold as well, with $R^2=.2017$ and
$F(1,22)=5.5579=t^2$.

\Needspace{14\baselineskip}

```{r}
#| label: ch7-binary-predictor
#| echo: true
score <- c(4, 5, 7, 8, 4, 5, 7, 8, 6, 7, 9, 10, 10, 11, 13, 14,
           8, 9, 11, 12, 16, 17, 19, 20)
spaced <- rep(rep(c(0, 1), each = 4), 3)
two_groups <- lm(score ~ spaced)
coef(summary(two_groups))
```

```{r}
#| label: fig-ch7-binary
#| fig-cap: "With a 0/1 predictor, the least-squares line joins the two group means: the intercept is the massed mean and the slope the difference in means."
#| fig-alt: "Twenty-four invented scores plotted at 0 for massed and 1 for spaced practice, massed scores from 4 to 12 and spaced from 7 to 20. A line joins the massed mean 8 to the spaced mean 12, both marked with filled diamonds."
#| fig-width: 5.6
#| fig-height: 3
two_group_data <- data.frame(spaced, score)
group_means <- data.frame(spaced = c(0, 1),
  score = tapply(score, spaced, mean))
ggplot(two_group_data, aes(spaced, score)) +
  geom_point(shape = 1, size = 1.8,
    position = position_jitter(width = .04, height = 0, seed = 12345)) +
  geom_abline(intercept = coef(two_groups)[1],
    slope = coef(two_groups)[2], linewidth = .8) +
  geom_point(data = group_means, shape = 18, size = 4.5) +
  scale_x_continuous(breaks = c(0, 1),
    labels = c("0 (massed)", "1 (spaced)")) +
  coord_cartesian(xlim = c(-.25, 1.25)) +
  labs(x = "Practice schedule", y = "Score (out of 20)")
```

Welch's unequal-variance test (Chapter 3) uses the separate group SDs,
2.3741 and 5.3767, and gives 15.1322 df and $p=.0323$. The one-line
regression assumes a common error variance; that is the model's
assumption, not a fact about the pupils. The HC3 covariance of Section 6
keeps the fitted line and gives a slope SE of 1.7721 under the same
independence assumption. A factor with more than two levels needs more
than one 0/1 column; Chapter 9 develops that coding, its contrasts, and
the equivalence with the analysis of variance of Chapter 4.

\Needspace{16\baselineskip}

## See Sampling Uncertainty Through Repetition {#sec-ch7-repetition}

### A known-truth line and one fitted sample

Generate 20 outcomes at predictor values equally spaced from 6 to 16:

$$
Y_i=2+0.5x_i+\varepsilon_i,\qquad
\varepsilon_i\overset{\mathrm{ind}}\sim\mathcal N(0,1).
$$

These units and values are a teaching construction, not occupation records.
The true intercept is 2, true slope is 0.5, and true error SD is 1.

```{r}
#| label: ch7-known-sample-code
#| echo: true
set.seed(123)
known_data <- data.frame(x = seq(6, 16, length.out = 20))
known_data$y <- 2 + 0.5 * known_data$x + rnorm(20, sd = 1)
known_model <- lm(y ~ x, data = known_data)
```

`set.seed()` fixes the random start, so the same 20 errors are drawn each
time, and `rnorm()` draws them with SD 1. @tbl-ch7-known-sample reports
the fitted coefficients, their SEs and the residual SD of `known_model`.

\Needspace{10\baselineskip}

```{r}
#| label: tbl-ch7-known-sample
#| tbl-cap: "One fitted sample from a process with known coefficients and error SD."
#| echo: false
known_summary <- coef(summary(known_model))
knitr::kable(data.frame(
  Quantity = c("Intercept", "Slope", "Residual SD"),
  Estimate = show_value(c(coef(known_model), sigma(known_model))),
  `Model SE` = c(show_value(known_summary[, "Std. Error"]), ""),
  check.names = FALSE
), align = c("l", "r", "r"))
```

The last cell is blank because residual SD is a scale estimate, not a
coefficient with a coefficient standard error.

```{r}
#| label: fig-ch7-core-true-fitted
#| echo: false
#| fig-cap: "Known and fitted lines for one simulated dataset. The true process is programmed; the fitted line responds to the realized errors."
#| fig-alt: "Twenty simulated points surround the known line with intercept two and slope one half. A solid fitted line and dashed true line differ slightly."
ggplot2::ggplot(known_data, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(size = 1.6) +
  ggplot2::geom_abline(
    intercept = 2, slope = 0.5, linetype = "dashed", linewidth = 0.8
  ) +
  ggplot2::geom_abline(
    intercept = coef(known_model)[1],
    slope = coef(known_model)[2], linewidth = 0.8
  ) +
  ggplot2::labs(x = "Simulated predictor", y = "Simulated outcome")
```

The fitted slope and intercept need not equal their programmed values in
one sample. The unobserved error is the deviation from the true mean;
the residual is the observed deviation from the fitted line. Even when
the errors are independent, fitted residuals are linked by their zero-sum
and zero-covariance constraints. A residual plot is an imperfect diagnostic
window on error behavior, not a direct sample of observed true errors.

### Repeated samples and the sources of slope precision

Recall the generate--estimate--repeat route from Chapters 5 and 6.
The next experiment holds 40 predictor values fixed and regenerates only
the errors, 2,000 times. The study size is 40; the repetition count is
2,000. Increasing the latter refines the simulation's description of the
same study-level procedure. It records slopes, estimated standard errors,
and whether each classical interval covers the known slope 0.5.

For one repetition, generate new independent errors, add them to the same
40 process means, fit a line, and record its slope and estimated SE. Check
whether its interval contains the programmed slope, then start the next
repetition with new errors. Holding the predictor grid fixed isolates
outcome variation; it does not resample the historical occupations. The Computer
Lab provides the complete loop and a small change in the repetition count.

```{r}
#| label: ch7-repeated-slopes-code
#| echo: false
set.seed(321)
rep_x <- seq(6, 16, length.out = 40)
repetitions <- data.frame(
  slope = numeric(2000), se = numeric(2000), covered = logical(2000)
)
for (j in seq_len(nrow(repetitions))) {
  rep_y <- 2 + 0.5 * rep_x + rnorm(40, sd = 1)
  rep_fit <- lm(rep_y ~ rep_x)
  repetitions$slope[j] <- coef(rep_fit)[2]
  repetitions$se[j] <- coef(summary(rep_fit))[2, "Std. Error"]
  rep_ci <- confint(rep_fit)[2, ]
  repetitions$covered[j] <- rep_ci[1] <= 0.5 &&
    rep_ci[2] >= 0.5
}
repetition_summary <- c(
  mean_slope = mean(repetitions$slope),
  empirical_SD = sd(repetitions$slope),
  mean_estimated_SE = mean(repetitions$se),
  theoretical_SD = 1 / sqrt(sum((rep_x - mean(rep_x))^2)),
  empirical_coverage = mean(repetitions$covered)
)
```

\Needspace{9\baselineskip}

```{r}
#| label: tbl-ch7-repeated-slopes
#| tbl-cap: "Slope behavior across 2,000 repetitions of the known process."
#| echo: false
knitr::kable(data.frame(
  Quantity = c("Mean estimated slope", "SD of estimated slopes",
               "Mean estimated SE", "Theoretical slope SD",
               "Fraction of intervals covering 0.5"),
  Value = replace(show_value(unname(repetition_summary)), 5,
    sub("^0", "", show_value(unname(repetition_summary))[5]))
))
```

```{r}
#| label: fig-ch7-core-repetition
#| echo: false
#| fig-cap: "Slopes in 2,000 repetitions of the known process. The vertical reference is the true slope, 0.5."
#| fig-alt: "A roughly symmetric histogram of repeated fitted slopes is centered near one half; its spread represents repeated-sample slope variation."
#| fig-pos: H
#| fig-width: 6.3
#| fig-height: 2.8
ggplot2::ggplot(repetitions, ggplot2::aes(x = slope)) +
  ggplot2::geom_histogram(bins = 35, color = "white", fill = "grey50") +
  ggplot2::geom_vline(xintercept = 0.5, linewidth = 0.8, linetype = "dashed") +
  ggplot2::labs(x = "Estimated slope", y = "Number of repetitions")
```

The empirical SD of the slopes estimates their repeated-sample spread. Each
fitted SE tries to estimate that spread using just one sample. Coverage is a
property of the repeated interval-producing procedure, not a probability
assigned to a fixed parameter after one interval is realized. The empirical
coverage will not be exactly .95: at true .95 coverage, its Monte Carlo SE
with 2,000 independent repetitions is
$\sqrt{.95(.05)/2000}\approx0.0049$.

The theoretical slope SD is $\sigma/\sqrt{S_{xx}}$. More noise increases
it; greater predictor spread decreases it. More observations typically
decrease it when the predictor distribution is held comparable.
@tbl-ch7-precision-conditions changes one feature at a time and applies the
formula to each design.

```{r}
#| label: ch7-precision-conditions-code
baseline_x <- seq(6, 16, length.out = 40)
more_x <- seq(6, 16, length.out = 160)
narrow_x <- seq(8.5, 13.5, length.out = 40)
baseline_sd <- 1 / sqrt(sum((baseline_x - mean(baseline_x))^2))
more_rows_sd <- 1 / sqrt(sum((more_x - mean(more_x))^2))
narrow_sd <- 1 / sqrt(sum((narrow_x - mean(narrow_x))^2))
more_noise_sd <- 2 / sqrt(sum((baseline_x - mean(baseline_x))^2))
precision_conditions <- data.frame(
  Condition = c("Baseline", "More rows", "Narrower support", "More noise"),
  n = c(40, 160, 40, 40), Lower = c(6, 6, 8.5, 6),
  Upper = c(16, 16, 13.5, 16),
  `Error SD` = c(1, 1, 1, 2),
  `Slope SD` = c(baseline_sd, more_rows_sd, narrow_sd, more_noise_sd),
  check.names = FALSE
)
```

\Needspace{9\baselineskip}

```{r}
#| label: tbl-ch7-precision-conditions
#| tbl-cap: "How sample size, predictor support, and error SD affect theoretical slope precision."
#| echo: false
precision_conditions$`Slope SD` <- show_value(
  precision_conditions$`Slope SD`)
knitr::kable(precision_conditions)
```

Halving the predictor range at the same sample size doubles the slope SD;
doubling $\sigma$ also doubles it. Quadrupling the number of equally spaced
points over the same range approximately halves it; the small difference
from an exact half comes from the finite grids' variances. Using `x <- 1:n`
while increasing $n$ would change both sample size and predictor range.
More rows do not automatically remove confounding, measurement error,
dependence, or misspecification. These simulations establish behavior only
under the mechanisms programmed into them.

\Needspace{9\baselineskip}

## Distinguish a Mean Response from a New-Case Prediction {#sec-ch7-prediction}

Slope uncertainty concerns one process coefficient. Prediction introduces a
different distinction: estimating a conditional mean versus anticipating one
new outcome from the same declared process.

### A fitted mean is not a new occupation

At $x_0=12$, the fitted income is 7.9322 thousand dollars. Under the model,
this estimates the conditional mean for occupations with that average
education. The standard error for the fitted mean is

$$
SE(\widehat y_0)
=s_e\sqrt{\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}}.
$$

Substitution gives

$$
3.4834\sqrt{\frac1{102}+
 \frac{(12-10.738)^2}{751.8852}}=0.3803.
$$

The resulting 95% mean interval is
$7.9322\pm1.984(0.3803)\approx[7.1776,8.6868]$.

A new occupation's outcome also varies around its process mean.
Write $\mu_0=\beta_0+\beta_1x_0$, $\widehat\mu_0=\widehat y_0$, and
$Y_0=\mu_0+\varepsilon_0$. The prediction error has two parts:

$$
Y_0-\widehat\mu_0
=\varepsilon_0-(\widehat\mu_0-\mu_0).
$$

If this new error is independent of the fitted data and comes from the same
common-variance process, the two variance contributions add:

$$
\operatorname{Var}(Y_0-\widehat\mu_0\mid\mathbf X,x_0)
=\sigma^2+\operatorname{Var}(\widehat\mu_0\mid\mathbf X,x_0).
$$

The first term is new outcome variation; the second is uncertainty in the
estimated mean. Substituting the fitted-mean variance and estimating $\sigma$
by $s_e$ gives

$$
SE_{\mathrm{new}}
=s_e\sqrt{1+\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}}
=3.5041.
$$

Thus the 95% prediction interval is approximately
$7.9322\pm1.984(3.5041)\approx[0.9802,14.8842]$.
The extra 1 is the new outcome's own error variance, expressed as a multiple
of $\sigma^2$; it does not disappear as the estimation sample grows.

```{r}
#| label: fig-ch7-intervals
#| fig-cap: "Mean-response intervals and new-case prediction intervals under the same working model."
#| fig-alt: "The mean-confidence band is narrowest near average education. The new-case prediction band is much wider throughout the observed education range."
#| fig-width: 6.3
#| fig-height: 3.3
grid <- data.frame(education = seq(min(x), max(x), length.out = 150))
cm <- predict(model, grid, interval = "confidence")
pi <- predict(model, grid, interval = "prediction")
grid$fit <- cm[, 1]
grid$cl <- cm[, 2]; grid$cu <- cm[, 3]
grid$pl <- pi[, 2]; grid$pu <- pi[, 3]
ggplot(grid, aes(education, fit)) +
  geom_ribbon(aes(ymin = pl, ymax = pu), fill = "grey90") +
  geom_ribbon(aes(ymin = cl, ymax = cu), fill = "grey65") +
  geom_line(linewidth = .7) +
  geom_point(data = prestige, aes(y = income_k), shape = 1, size = 1.3) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)")
```

Both intervals get wider away from $\bar x$, even within the observed range.
Outside that range, numerical output is not a warrant for extrapolation.
A lower prediction bound below zero also exposes the limitations of an
unrestricted normal-error model for a nonnegative outcome; it is not a
plausible negative-income forecast.

```{r}
#| label: ch7-interval-code
#| echo: true
predict(model, data.frame(education = 12), interval = "confidence")
predict(model, data.frame(education = 12), interval = "prediction")
```

A separately owed new-case prediction answer is a separable question, not
automatically another strand of the original descriptive question.
Moreover, a model prediction interval is not a test of a learning algorithm
on unseen records. Held-out prediction performance is a distinct estimand,
developed with multiple regression. Appendix A retains the interval
derivations and further supported-value comparisons.


## Check Form, Spread, Influence, and Support {#sec-ch7-check}

### Checks should change an interpretation or a decision

The fitted line does not have to satisfy a stochastic model to exist.
But claims that it describes a linear conditional mean, that its classical
interval has exact coverage, or that it predicts new occupations reliably
need additional support.

```{r}
#| label: fig-ch7-checks
#| fig-cap: "Residual pattern and influence invite different follow-up questions."
#| fig-alt: "The left panel shows curvature and uneven scatter in residuals. The right panel shows Cook's distances, with physicians most influential for this fit."
#| fig-width: 6.4
#| fig-height: 2.6
op <- par(mfrow = c(1, 2), mar = c(4, 4, 1, 1), mgp = c(2.5, .7, 0))
plot(fitted(model), resid(model), xlab = "Fitted income", ylab = "Residual",
     pch = 1, cex = .7)
abline(h = 0, col = "grey60")
lines(lowess(fitted(model), resid(model)), lwd = 2)
plot(cooks.distance(model), type = "h", xlab = "Occupation row",
     ylab = "Cook's distance")
check_most <- which.max(cooks.distance(model))
text(check_most, cooks.distance(model)[check_most], "physicians",
     pos = 4, cex = .7)
par(op)
```

A curve in residuals suggests that one straight conditional mean leaves
systematic variation. A changing residual spread challenges a common-variance
uncertainty calculation. Leverage concerns unusual predictor positions;
influence concerns how much the fitted model changes when a record's
contribution changes. A large residual, high leverage, and high influence are
related but different properties.

\Needspace{7\baselineskip}

For a straight line with an intercept, leverage is

$$
h_i=\frac1n+\frac{(x_i-\bar x)^2}{S_{xx}}.
$$

It grows as a predictor value moves away from $\bar x$. Cook's distance then
combines leverage and residual magnitude:

$$
D_i=\frac{e_i^2}{2s_e^2}\frac{h_i}{(1-h_i)^2}.
$$

Cook's distance summarizes how much the fitted values change when case $i$
is omitted, relative to residual variation and the two fitted coefficients.
It is a prompt to inspect provenance and sensitivity, not an automatic
deletion cutoff.

For these data, adding education squared lets the fitted relationship bend.
The following same-record comparison makes the change in shape visible. It
is an exploratory check, not a retrospectively prespecified new final model.

```{r}
#| label: ch7-check-models
#| echo: true
quad <- lm(income_k ~ education + I(education^2), data = prestige)
most <- which.max(cooks.distance(model))
without <- lm(income_k ~ education, data = prestige[-most, ])
c(full = unname(coef(model)[2]), without = unname(coef(without)[2]))
coef(summary(quad))["I(education^2)", ]
```

```{r}
#| label: fig-ch7-curvature
#| fig-cap: "Straight and quadratic fits to the same 102 records. The curve is an exploratory sensitivity, not a selected final model."
#| fig-alt: "The quadratic fit bends upward toward higher education, unlike the constant-slope straight line. Both fits use all 102 occupation records on the same income scale."
#| fig-width: 6.3
#| fig-height: 3.7
check_grid <- data.frame(
  education = seq(min(x), max(x), length.out = 200)
)
check_curves <- rbind(
  data.frame(check_grid, Fit = "Straight line",
             Income = predict(model, check_grid)),
  data.frame(check_grid, Fit = "Quadratic sensitivity",
             Income = predict(quad, check_grid))
)
ggplot(prestige, aes(education, income_k)) +
  geom_point(shape = 1, colour = "grey50", size = 1.5) +
  geom_line(data = check_curves,
            aes(y = Income, linetype = Fit), linewidth = .8) +
  labs(x = "Average education (years)",
       y = "Income (thousand 1971 dollars)", linetype = NULL)
```

The quadratic coefficient has $p=.0059$ under that model's classical
reference calculation. This is a reason to examine the visible curvature,
not a sufficient model-selection rule: unequal spread and choosing the
comparison after inspecting the records also qualify that number.

Removing physicians, the record with the largest Cook's
distance (0.3983), changes the slope from 0.8988 to 0.7982.
Nothing in that result proves physicians were recorded incorrectly. The
deletion is a sensitivity calculation; it is not our replacement analytic
sample or a licence to hide a valid unusual occupation.

\Needspace{7\baselineskip}

An HC3 heteroskedasticity-consistent covariance calculation gives a slope SE
of 0.1491 rather than 0.1270. Using the same $t_{100}$ critical value as an
approximate sensitivity procedure gives $[0.6031,1.1945]$.
The line does not change. This procedure relaxes a variance assumption under
suitable independent-observation conditions; it does not repair curvature,
dependence, confounding, or measurement error [@white1980; @mackinnon1985;
@long2000; @zeileis2004].
Its finite-sample interval is not the exact classical normal-model interval.
The short R calculation makes the covariance choice explicit:

\Needspace{10\baselineskip}

```{r}
#| label: ch7-hc3-visible
#| echo: true
V_hc3 <- sandwich::vcovHC(model, type = "HC3")
c(classical = coef(summary(model))["education", "Std. Error"],
  HC3 = sqrt(V_hc3["education", "education"]))
```

Later chapters take these questions further. Chapter 8, Appendix B,
assembles the HC3 matrix from its parts. Chapter 9, Appendix A, defines
the target that a least-squares line estimates when the conditional mean
is not linear, and its Section 5.3 shows that measurement error in the
predictor attenuates a slope, which no covariance choice repairs. Nothing
in this chapter relies on any of them.

\Needspace{20\baselineskip}

| Check | What needs to be examined | Consequence |
|:--|:--|:--|
| Form | Residual curve and a curved-fit comparison | A straight line may remain a description but fail as a constant conditional-mean slope account |
| Spread | Residual variation over fitted values | Classical SEs may need revision; HC3 is a variance sensitivity, not a universal repair |
| Influence | Source record and result with/without a consequential case | Disclose sensitivity; correct errors only when verified |
| Independence | How occupations and measurements were generated or related | A plot cannot establish independent process realizations |
| Support | Actual education range and density of observations | Restrict comparisons; avoid extrapolation and individual-level claims |

: Diagnostic questions and their consequences. {#tbl-ch7-diagnostic-consequences tbl-colwidths="[16,38,46]"}

\Needspace{24\baselineskip}

## Regression Toward the Mean {#sec-ch7-rtm}

### A line in standard units

Francis Galton, who gave regression its name, measured the adult heights
of parents and their grown children in 205 families in the 1880s
[@galton1886]. We use James Hanley's transcription of Galton's notebook
[@hanley2004], distributed in the R package HistData [@friendly2025], and
keep one row per family: among the 179 families with at least one son,
one son was drawn at random from each, so every father appears once and
the rows describe separate families. Heights are in inches, mostly
recorded to the nearest half inch. The small extract `galton1886_sons.csv` is
supplied with the chapter under this note: *Galton's family heights
(published 1886), transcribed by J. A. Hanley and distributed in the R
package HistData under the GPL; one son per family drawn for QMSBR. The
measurements are historical records in the public domain; the extract is
redistributed with this attribution.*

Chapter 2 measured a value's distance from its mean in standard units,
$z_x=(x-\bar x)/s_x$, and Section 2.2 wrote the slope as
$\widehat\beta_1=r\,s_y/s_x$. Substituting that slope into the line
through the mean point, $\widehat y-\bar y=\widehat\beta_1(x-\bar x)$, and
dividing both sides by $s_y$ gives the line in standard units:

$$
\frac{\widehat y-\bar y}{s_y}=r\,\frac{x-\bar x}{s_x},
\qquad\text{that is,}\qquad
\widehat z_y=r\,z_x .
$$

The line passes through the origin with slope $r$. Unless the correlation
is perfect, $\lvert r\rvert<1$, so a case $z_x$ SDs from the mean on the
predictor is predicted only $r\,z_x$ SDs from the mean on the outcome,
each measured in its own SDs. Galton called this regression toward
mediocrity; we call it **regression toward the mean**.

```{r}
#| label: ch7-galton-visible
#| echo: true
galton <- read.csv("galton1886_sons.csv")
sapply(galton[c("father", "son")], sd)
c(r = cor(galton$father, galton$son),
  z_slope = unname(coef(lm(scale(son) ~ scale(father),
    data = galton))[2]))
```

`sapply()` applies `sd()` to both columns, and `scale()` converts a column
to standard units. The fathers average 69.0989 inches with SD 2.5466, the
sons 69.2140 inches with SD 2.6475, and their correlation is $r=.4269$. In
inches the fitted slope is 0.4438 inch of son's height per inch of
father's height; fitted to the standardized heights it is .4269, which is
$r$, as the equation says. A father 1.6 SD above the fathers' mean has a
son predicted $.4269\times1.6=0.6831$ SD above the sons' mean: still
taller than average, but less than half as far out.

```{r}
#| label: fig-ch7-rtm
#| fig-cap: "Galton's 179 fathers and sons in standard units. The fitted line has slope $r=.4269$, flatter than the dashed line of slope 1; the filled points are the mean positions of the fathers more than 1 SD above and below their mean, with their sons."
#| fig-alt: "A cloud of 179 points, fathers' heights in standard units from about −2.8 to 3.7 across and sons' from about −3.5 to 2.8 up. A dashed line of slope 1 and a solid fitted line of slope about 0.43 cross at the origin. Two filled diamonds, near (1.6, 0.7) and (−1.5, −0.6), sit close to the solid line, inside the dashed one."
#| fig-width: 5.4
#| fig-height: 4.6
z_father <- as.numeric(scale(galton$father))
z_son <- as.numeric(scale(galton$son))
r_galton <- cor(galton$father, galton$son)
galton_z <- data.frame(z_father, z_son)
tall <- z_father > 1
short <- z_father < -1
group_means <- data.frame(
  z_father = c(mean(z_father[tall]), mean(z_father[short])),
  z_son = c(mean(z_son[tall]), mean(z_son[short])))
ggplot(galton_z, aes(z_father, z_son)) +
  geom_point(shape = 1, colour = "grey55", size = 1.4) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
    linewidth = .7) +
  geom_abline(intercept = 0, slope = r_galton, linewidth = .9) +
  geom_point(data = group_means, shape = 18, size = 4.5) +
  annotate("text", x = 3.7, y = c(-2.6, -3.2), hjust = 1, size = 3.6,
    label = c("solid: fitted line, slope r = .4269",
      "dashed: slope 1")) +
  coord_equal(xlim = c(-3.6, 3.8), ylim = c(-3.6, 3.8)) +
  labs(x = "Father's height (standard units)",
       y = "Son's height (standard units)")
```

In @fig-ch7-rtm the dashed line of slope 1 marks where a son would sit if
he were exactly as many SDs from his mean as his father. The fitted line
is flatter. To the right of the mean it lies below the dashed line and to
the left above it: tall fathers have sons predicted less tall, and short
fathers sons predicted less short, in standard units. The two filled
points mark the average positions of the fathers more than 1 SD above and
below their mean, with their sons; both sit near the fitted line, well
inside the dashed one.

### Extreme groups move toward the mean; the spread does not

@tbl-ch7-rtm-groups groups the men in bands of 1 SD, first by the fathers'
heights and then by the sons'. Each band includes its upper bound, so the
four bands of each block hold all 179 men.

\Needspace{14\baselineskip}

| Selected on | Band (SDs) | Men | Mean $z$ of the selected height | Mean $z$ of the other height | $r$ times the selected mean |
|:--|:--|--:|--:|--:|--:|
| Fathers | Above $+1$ | 24 | 1.6072 | 0.7029 | 0.6862 |
| Fathers | 0 to $+1$ | 60 | 0.4978 | 0.2251 | 0.2125 |
| Fathers | $-1$ to 0 | 65 | $-0.3463$ | $-0.2058$ | $-0.1479$ |
| Fathers | Below $-1$ | 30 | $-1.5310$ | $-0.5668$ | $-0.6536$ |
| Sons | Above $+1$ | 31 | 1.4057 | 0.6224 | 0.6001 |
| Sons | 0 to $+1$ | 60 | 0.4650 | 0.1431 | 0.1985 |
| Sons | $-1$ to 0 | 61 | $-0.4610$ | $-0.0697$ | $-0.1968$ |
| Sons | Below $-1$ | 27 | $-1.6057$ | $-0.8751$ | $-0.6855$ |

: Groups of Galton's fathers and sons in standard units, selected on either height. {#tbl-ch7-rtm-groups tbl-colwidths="[12,14,8,22,22,22]"}

Read the first block row by row. The 24 fathers more than 1 SD above
their mean average $z=1.6072$, or 73.1917 inches; their sons average
0.7029, or 71.0750 inches, close to the line's prediction
$.4269\times1.6072=0.6862$. The 30 short fathers mirror them. In every
row the other height's mean lies between zero and the selected mean,
near $r$ times it; bands nearer the middle move less, because they start
nearer the mean.

The second block selects on the sons, and the movement toward the mean
appears again, now running backward in time: the 31 sons more than 1 SD
above their mean, at 1.4057, had fathers averaging only 0.6224. No force
pulls heights toward the average from one generation to the next; such a
force could not act backward in time. An imperfect correlation is
enough for the fitted line: its predictions lie closer to the mean, in
standard units, than the values they are predicted from. For groups of
men, the pattern also needs the mean of one height to follow a straight
line in the other, as it roughly does in these families; then selecting
men for an extreme value of one height selects, on average, men whose
other height is less extreme. On average is the operative phrase: a
group's observed mean scatters around its prediction, so an imperfect
correlation does not guarantee that every group selected for an extreme
value sits nearer the mean on the other height.

If tall fathers have sons nearer the mean and short fathers do too,
should the sons' heights be less spread out than their fathers'? They are
not: the sons' SD, 2.6475 inches, is slightly larger than the fathers',
2.5466. This is the **regression paradox**. The fitted values carry only
the part of a son's height that his father's height predicts, and they
spread only $r\,s_y=1.1303$ inches. The residuals carry the rest, with
spread $s_y\sqrt{1-r^2}=2.3941$ inches (both SDs with the $n-1$ divisor;
`sigma()` divides by $n-2$ and gives 2.4008). The two parts combine
through their variances, not their SDs: $1.1303^2+2.3941^2=2.6475^2$,
Section 2.6's decomposition of total variation, in which the fitted part
holds the share $R^2=r^2=.1823$. Some sons of average fathers are very
tall or very short, so the sons' generation is as varied as the
fathers'. Regression toward the mean describes predictions for selected
groups, not a population growing more uniform.

### The regression fallacy

Regression toward the mean becomes a fallacy when the movement it
predicts is credited to a cause. Tversky and Kahneman [-@tversky1974] recount a conversation with flight instructors: a trainee praised for an
unusually smooth landing tended to land less well next time, and a
trainee reprimanded for a rough landing tended to do better, so the
instructors concluded that praise hurts and criticism helps.

The line in standard units explains the pattern without either
conclusion. A landing's quality reflects the trainee's skill and the
chance conditions of that attempt, so two landings by the same trainee
are correlated, but not perfectly. The landings that drew comment were
selected for being extreme, and the next landing is predicted nearer the
trainee's usual level whatever the instructor says: the praise and the
criticism coincided with regression toward the mean. Crediting that
movement to the feedback is the **regression fallacy**.

The same pattern appears whenever cases are selected for an extreme first
measurement and measured again: pupils chosen for their low test scores
and given extra help, patients enrolled when their symptoms peak, schools
placed on a list after their worst year. On the second measurement such
groups move toward the mean on average even if nothing done in between
has any effect. A comparison group selected the same way and measured at the same
two times shows how much movement selection alone produces. Assigning the
help at random among the selected cases does better still, since both
groups then regress alike and their difference estimates the help's
effect [@barnett2005]. Chapter 14, Section 4.1, applies this reasoning to
selection on a noisy baseline.

## Reading Bridge {#sec-ch7-summary}

A simple regression starts with a comparison and a unit. Its fitted values
and residuals are calculations on records. Least squares connects the
slope to covariance and predictor variation and puts the line through the
mean point. Residual SD, $R^2$, slope uncertainty, and new-case variation
answer different questions; the matched Pearson, slope, and $F$ statistics
connect calculations without supplying a sampling design.

Across the 102 recorded occupations, fitted income rises by 0.8988 thousand
1971 Canadian dollars---about 899 dollars---per additional year of
occupation-average education over the observed 6.38--15.97-year range.
This is a bounded description of the records, not an individual, causal,
or population return. Known-truth repetition showed how a separately
declared process gives slope uncertainty and coverage. In standard units
the line predicts $r$ times a case's distance from the mean, so groups
selected for extreme values are predicted nearer the mean on another
measurement while the spread stays the same; a before-and-after gain in
such a group needs a comparison selected the same way before it is read
as an effect. Supplementary A reads a short report of the occupation line
as an argument, locating its reasons before judging its answer.

Chapter 8 refits the education-only line on the same 98 type-complete
occupations before adding prestige, percentage of women, and occupation
type. Moving from 102 to 98 changes the records; it is not an adjustment
effect. The chapter develops conditional comparisons and standardized
fitted quantities, and asks a separate held-out prediction question.
Chapter 9 collects the regression toolbox these chapters use only
briefly, from factors with several levels to robust and resampled
uncertainty. Chapter 9, Section 7 (Project STAR), then reads a randomized
class-size experiment whose uncertainty follows its school-clustered
design [@finn1990; @mosteller1995]; Chapter 10 extends the factorial
interactions from Chapters 4 and 6 to comparisons involving regression
predictors.

### Key terms {.unnumbered}

Use this short reference to revisit the regression concepts introduced
here. The Part One Concept Reference gives the broader reference and the
principal home of each concept; fitted values, residuals, standard errors,
intervals, standardized values, and the reading vocabulary are explained
in Chapters 1--4.

| Term and common alternatives | Meaning |
|:--|:--|
| Simple linear regression (regression line) | A straight-line summary of how one outcome differs across one predictor: fitted value equals intercept plus slope times the predictor. Its descriptive use concerns the records fitted; a model-based use needs a declared process. |
| Intercept | The fitted outcome at predictor value zero. It places the line vertically; when zero is outside the observed support, it is a mathematical anchor rather than a supported prediction. Centering the predictor moves it to a supported reference point. |
| Slope (regression coefficient) | The difference in fitted outcome per one-unit difference in the predictor, in outcome units per predictor unit. It compares units that differ in the predictor; it is not a change within one unit unless a design supports that reading. |
| Ordinary least squares (OLS) | The rule that chooses the intercept and slope minimizing the residual sum of squares. With an intercept, the fitted line passes through the mean point and the residuals sum to zero; the slope equals covariance divided by predictor variance. |
| Residual sum of squares (RSS) | The sum of squared residuals from a fitted line, in squared outcome units. Total squared variation about the mean splits into a regression part and RSS when the model has an intercept. |
| Residual standard deviation (residual standard error) | The square root of RSS divided by its residual degrees of freedom, in outcome units; with an intercept and one predictor the degrees of freedom are $n-2$. Under the process model it estimates the error SD. |
| R-squared (coefficient of determination) | One minus RSS divided by total squared variation: the proportion of centered squared outcome variation accounted for by the fit in these records. In simple regression with an intercept it equals the squared correlation. |
| Confidence interval for a mean response (mean-response interval) | An interval for the process mean outcome at a stated predictor value, using the SE of the fitted value. It concerns an average, narrows near the predictor mean, and does not cover individual outcomes. |
| Prediction interval (new-case interval) | An interval for one new outcome at a stated predictor value from the same process. Its variance adds the error variance to the fitted-mean variance, so it is wider than the mean-response interval and keeps a nonzero floor as the sample grows. |
| Leverage | A measure of how unusual a record's predictor value is relative to the others; it grows with distance from the predictor mean and appears in the SE of a fitted value and in influence measures. |
| Cook's distance (influence) | A summary of how much the fitted values change when one record is omitted, combining its residual and its leverage. It prompts inspection of provenance and a disclosed sensitivity, not automatic deletion. |
| Heteroskedasticity-consistent covariance (HC3; sandwich estimator) | A coefficient covariance estimate that allows unequal error variance across predictor values under independence. It changes SEs and intervals, not the fitted coefficients, and does not repair curvature, dependence, measurement error, or confounding. |
| Extrapolation (predictor support) | Using a fitted line at predictor values outside the observed range or where observations are sparse. Software returns a number there; the data do not support it. |
| Regression toward the mean (regression to the mean) | In standard units a case's predicted outcome is $r$ times its predictor value, $\widehat z_y=r\,z_x$, so with $\lvert r\rvert<1$ cases extreme on one measurement are predicted nearer the mean on the other, and a group selected for an extreme noisy first value is expected to move toward the mean on a second measurement, on average rather than in every group. The outcome's spread does not shrink (the regression paradox). |
| Regression fallacy | Attributing to a treatment, a talent or another cause the movement toward the mean that selection on an extreme noisy measurement produces by itself. A comparison group selected the same way, or random assignment, separates a real change from regression toward the mean. |

: Selected concepts for simple linear regression. {#tbl-ch07-key-terms tbl-colwidths="[28,72]"}

### Further reading {.unnumbered}

For another accessible account of lines, residuals, and fit, see
[OpenIntro's single-predictor regression tutorial](https://openintro-ims.netlify.app/model-slr).
For a compact executable route from a fitted model to intervals and checks, see
the [ISLR authors' Chapter 3 R lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch3-linreg-lab.html)
[@james2021].
These provide complementary practice; the present chapter additionally makes
the finite-record and model-based questions explicit. Consult
[the carData reference manual](https://cran.r-project.org/web/packages/carData/carData.pdf)
[@carData2022] for data documentation, @barnett2005 on regression to the
mean in repeated measurements, and @fox2016, @fox2019, and @gelman2020 for
broader regression treatments.

\Needspace{18\baselineskip}

## Exercises {.unnumbered}

The prompts below use **Foundational** and **Integrative** levels. Where
rounded summaries are supplied, answers within 0.02 in the stated units are
sufficient. Exercises 1--6 draw on the main chapter.

<!-- prompt-id: CH07-E01 -->
### Exercise 1: Read the unit and the line (Foundational) {.unnumbered #ex-ch07-01}

<!-- exercise-id: ch07-ex01 -->

A teaching report describes 80 occupations with
$\widehat{\mathrm{income}}=-1.5+0.7\,\mathrm{education}$, where income is in
thousand dollars and average education ranges from 7 to 16 years.
Interpret the slope and the intercept. Calculate the fitted income at 12 years
and the residual for an occupation at 12 years with income 9.
Explain why the slope does not give an individual's gain from another year
of schooling.

<!-- prompt-id: CH07-E02 -->
### Exercise 2: Connect covariance and the fitted line (Foundational) {.unnumbered #ex-ch07-02}

<!-- exercise-id: ch07-ex02 -->

A set of records has $\bar x=4,\bar y=10,S_{xx}=20,S_{xy}=30$.
Calculate the OLS slope and intercept, and the fitted value at $x=5$.
Show that the line passes through the mean point. If $s_x=2$ and
$s_y=5$, calculate the correlation implied by these summaries.
What would prevent fitting a slope if every $x$ were equal?

<!-- prompt-id: CH07-E03 -->
### Exercise 3: Separate fit and uncertainty (Foundational) {.unnumbered #ex-ch07-03}

<!-- exercise-id: ch07-ex03 -->

An intercept-and-slope regression has $n=22$, $RSS=80$, $TSS=144$,
$S_{xx}=100$, and fitted slope 0.8. Calculate residual SD, $R^2$, and the
classical slope SE. Use $t_{.975,20}=2.086$ to obtain a 95% slope interval.
State the process assumptions needed for that interval and explain why it
does not by itself establish that the line fits closely or identifies a cause.

<!-- prompt-id: CH07-E04 -->
### Exercise 4: Mean or new case? (Foundational) {.unnumbered #ex-ch07-04}

<!-- exercise-id: ch07-ex04 -->

Under the stated independent normal-error constant-variance model, a fitted
value at $x_0=\bar x$ is 12, $n=25$, and $s_e=3$.
Use $t_{.975,23}=2.069$. Calculate the SE and 95% interval for the conditional
mean, then the corresponding prediction SE and interval for one independent
new case at the same $x_0$. Explain the extra 1 in the prediction expression.
Would either interval validate the model in a different population?

\Needspace{23\baselineskip}

<!-- prompt-id: CH07-E05 -->
### Exercise 5: Tell regression toward the mean from an effect (Integrative) {.unnumbered #ex-ch07-05}

<!-- exercise-id: ch07-ex05 -->

In an invented study, a school records a reading test in September and
again in June for all 200 of its pupils. Both tests have mean 50 and SD
10, and their correlation is .6. The 20 pupils with the lowest September
scores, averaging 32 points, receive extra tutoring, and their June mean
is 42.8 points. These figures include the tutored pupils' June scores, so
by themselves they cannot show what those scores would have been without
tutoring. The school's records from earlier years, when no pupil was
tutored, supply that benchmark: the two tests again had means 50, SDs 10
and correlation .6, and the mean June score of pupils with the same
September score followed a straight line in that score. Assume that this
year's pupils would have followed the same pattern without tutoring. (1)
Express the tutored group's September mean in standard units, and give
the June mean that this pattern's line in standard units predicts for
pupils with that September score, in standard units and in points. (2)
The school reports that tutoring raised the group by 10.8 points. Name
the error in reading the whole gain as an effect of tutoring, and say how
much of the gain regression toward the mean predicts by itself under
this pattern. (3) The report adds that, since the lowest
scorers caught up, the June scores of all 200 pupils must be less spread
out than September's. Split a June score into the part that the
September score predicts and the part it does not, give the SD of each
part, and explain why regression toward the mean leaves the June SD at 10.
(4) Describe a comparison that would separate an effect of tutoring from
regression toward the mean.

<!-- prompt-id: CH07-E06 -->

\Needspace{16\baselineskip}

### Exercise 6: Respond to sensitivity (Integrative) {.unnumbered #ex-ch07-06}

<!-- exercise-id: ch07-ex06 -->

In a descriptive analysis of 102 occupations, the slope is 0.9.
A valid, high-income occupation has the largest Cook's distance; removing it
gives 0.79. A curved-fit check changes the shape, and an HC3 calculation changes
the slope SE from 0.13 to 0.15 without changing the line.

What does each check tell you? What does it not tell you? Explain why deleting
the occupation automatically is inappropriate. Distinguish a variance repair
for the same estimand from replacing a constant process slope with a different
estimand. Write a short sensitivity disclosure for the descriptive analysis.

{{< pagebreak >}}

## Appendix A: Extended regression mathematics and interval derivations {.unnumbered #sec-ch7-app-a}

The main text uses the identities needed to read a fitted line. This appendix
derives them and distinguishes identities about the observed records from
claims about hypothetical repetitions. All occupation calculations retain
the same 102 rows, education in years, and income in thousand 1971 Canadian
dollars.

### Derivatives, normal equations, and the minimizing line {.unnumbered}

Start with a fitted model that predicts one common center $m$ for every
observation. Chapter 2 Section 2.2 states that the mean minimizes the
sum of squared deviations, and Chapter 3 Appendix A derives the identity
behind it:

$$
L(m)=\sum_i(y_i-m)^2
=\sum_i(y_i-\bar y)^2+n(m-\bar y)^2.
$$

Write $y_i-m=(y_i-\bar y)+(\bar y-m)$ and expand the square. The cross
term is zero because $\sum_i(y_i-\bar y)=0$. For $n>0$, the last term
is nonnegative and vanishes only at $m=\bar y$, so the mean uniquely
minimizes this criterion.

The three outcomes $2,4,3$ from Section 2.3 have mean 3 and
$L(m)=2+3(m-3)^2$, so their best constant line has loss 2. Allowing the
fitted value to depend on the three predictor values reduces RSS to 1.5,
as @tbl-ch7-tiny showed. Neither result
requires Normal errors or a probability sample. Chapter 6's likelihood
example chose a parameter by optimizing a different stated criterion;
here we develop the least-squares line algebraically. An uncertainty or
causal claim still needs its own supporting conditions.

The intercept-and-slope criterion has two coefficients to determine.
Instead of setting one derivative to zero, we set its two partial
derivatives to zero and verify that the resulting line is the minimum.

For candidate intercept $b_0$ and slope $b_1$, define

$$
Q(b_0,b_1)=\sum_{i=1}^n(y_i-b_0-b_1x_i)^2.
$$

Squared residuals cannot cancel and give large vertical discrepancies greater
weight. Differentiating with respect to each candidate coefficient gives

$$
\frac{\partial Q}{\partial b_0}
=-2\sum_i(y_i-b_0-b_1x_i),\qquad
\frac{\partial Q}{\partial b_1}
=-2\sum_i x_i(y_i-b_0-b_1x_i).
$$

At the minimum both derivatives are zero. The resulting **normal equations**
are

$$
n\widehat\beta_0+\widehat\beta_1\sum_i x_i=\sum_i y_i,
\qquad
\widehat\beta_0\sum_i x_i+\widehat\beta_1\sum_i x_i^2
=\sum_i x_iy_i.
$$

The first equation gives
$\widehat\beta_0=\bar y-\widehat\beta_1\bar x$.
Substitution into the second yields

$$
\widehat\beta_1
=\frac{\sum_i x_iy_i-n\bar x\bar y}
       {\sum_i x_i^2-n\bar x^2}
=\frac{S_{xy}}{S_{xx}}.
$$

Here $S_{xx}=\sum_i(x_i-\bar x)^2>0$ is required: a slope cannot be
identified when every predictor value is the same. Because sample covariance
and variance use the same denominator $n-1$,

$$
\widehat\beta_1
=\frac{s_{xy}}{s_x^2}
=r\frac{s_y}{s_x},
\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x.
$$

In the occupation data,

$$
\widehat\beta_1
=\frac{675.8041}{751.8852}\approx0.8988,
\qquad
\widehat\beta_0=\bar y-\widehat\beta_1\bar x\approx-2.8536.
$$

The ratio has units of thousand dollars per education year. Unlike a
correlation, the slope changes when a variable's units change. Standardizing
both variables makes their sample standard deviations one, so the simple
intercept-model slope becomes $r$; Section 7 reads that line.

To verify that the stationary point is the unique minimum, write any other
line as $\widehat\beta_0+d_0+(\widehat\beta_1+d_1)x_i$. The normal
equations make the cross term with the OLS residuals vanish, leaving

$$
Q(\widehat\beta_0+d_0,\widehat\beta_1+d_1)
=RSS+n(d_0+d_1\bar x)^2+d_1^2S_{xx}.
$$

With $S_{xx}>0$, every different line has larger RSS. If a candidate line is
constrained to pass through $(\bar x,\bar y)$, its excess RSS is simply
$S_{xx}(b_1-\widehat\beta_1)^2$. Section 2.4 makes this
comparison visible with five candidate lines.

### Residual constraints and projection geometry {.unnumbered}

At the OLS solution the derivative equations become
$\sum_i e_i=0$ and $\sum_i x_ie_i=0$. Thus the residuals have zero
sample covariance with the predictor and mean zero **by construction**.
They are not evidence that the unknown population errors are independent,
normally distributed, or conditionally mean zero.

For the geometry, let $\mathbf X$ be the $n\times2$ matrix with first
column all ones and second column containing the predictor. Bold
$\mathbf X$ is a matrix, distinct from a single predictor value $x_i$.
The normal equations can be written

$$
\mathbf X^{\mathsf T}(\mathbf y-\mathbf X\widehat{\boldsymbol\beta})
=\mathbf0,
\qquad
\widehat{\boldsymbol\beta}
=(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}\mathbf y.
$$

The fitted vector is the projection of $\mathbf y$ onto the two-dimensional
space spanned by the intercept and predictor columns. The residual vector
is perpendicular to that space. Writing
$\mathbf H=\mathbf X(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}$,
we have $\widehat{\mathbf y}=\mathbf H\mathbf y$ and
$\mathbf e=(\mathbf I-\mathbf H)\mathbf y$.

Because the model includes an intercept, both $\widehat{\mathbf y}$ and
the constant vector $\bar y\mathbf1$ lie in the fitted space. Consequently,

$$
\begin{aligned}
\sum_i(y_i-\bar y)^2
&=\sum_i\{(\widehat y_i-\bar y)+e_i\}^2\\
&=\sum_i(\widehat y_i-\bar y)^2+\sum_i e_i^2,
\end{aligned}
$$

since $\sum_i(\widehat y_i-\bar y)e_i=0$. This proves
$TSS=RegSS+RSS$, an algebraic decomposition of observed variation, not
a division into causal contributions. For these records,
$1820.8134=607.4214+1213.3920$, in squared thousand-dollar units.

Also $RegSS=\widehat\beta_1^2S_{xx}=S_{xy}^2/S_{xx}$. Dividing by
$TSS=S_{yy}$ proves $R^2=S_{xy}^2/(S_{xx}S_{yy})=r^2$ for simple
regression **with an intercept**. The identities need not hold for a
no-intercept fit. When fitted values are nonconstant,
$\operatorname{cor}(y,\widehat y)^2=R^2$ as well.

There are two fitted directions and $n-2$ residual directions. Under a
correct linear mean with independent constant-variance errors,
$E(RSS\mid\mathbf X)=(n-2)\sigma^2$. Thus
$s_e^2=RSS/(n-2)$ estimates the error variance, while $s_e$ estimates its
standard deviation in the outcome's units. The square root is not itself
an exactly unbiased estimator of $\sigma$. Without that stochastic model,
$s_e$ remains a computable residual-scale summary, not proof of a common
population error variance.

### From an error model to slope and intercept uncertainty {.unnumbered}

The 102-record least-squares line is already determined by those records.
The next derivation concerns a **separate model-based question**: what
constant conditional-mean coefficient generated hypothetical independent
occupation outcomes at the observed education values? Assume

$$
Y_i=\beta_0+\beta_1x_i+\varepsilon_i,
\quad E(\varepsilon_i\mid\mathbf X)=0,
\quad \operatorname{Var}(\boldsymbol\varepsilon\mid\mathbf X)
=\sigma^2\mathbf I.
$$

These are offered working conditions, not consequences of the package's
availability or of fitting `lm()`. The slope error is

$$
\widehat\beta_1-\beta_1
=\frac{\sum_i(x_i-\bar x)\varepsilon_i}{S_{xx}}.
$$

Independence and common variance then give

$$
\operatorname{Var}(\widehat\beta_1\mid\mathbf X)
=\frac{\sigma^2}{S_{xx}},
\qquad
SE(\widehat\beta_1)=\frac{s_e}{\sqrt{S_{xx}}}.
$$

Because $\widehat\beta_0=\bar Y-\widehat\beta_1\bar x$ and
$\operatorname{Cov}(\bar Y,\widehat\beta_1\mid\mathbf X)=0$,

$$
\operatorname{Var}(\widehat\beta_0\mid\mathbf X)
=\sigma^2\left(\frac1n+\frac{\bar x^2}{S_{xx}}\right),
\qquad
\operatorname{Cov}(\widehat\beta_0,\widehat\beta_1\mid\mathbf X)
=-\frac{\bar x\sigma^2}{S_{xx}}.
$$

This covariance explains why simply adding the variances of an uncentered
intercept and slope prediction would be wrong. If the predictor is centered
at its observed mean, the intercept and slope estimates are uncorrelated
under the classical model, and the intercept SE is $s_e/\sqrt n$.

Under the stated linear-mean, uncorrelated, equal-variance conditions, OLS
is the best linear unbiased estimator: it has the smallest variance among
estimators linear in the outcomes and unbiased for the coefficients. This
Gauss--Markov statement does not require normal errors and does not establish
causal identification, valid measurement, or population coverage [@fox2016;
@kutner2005].

Add a jointly normal error vector conditional on the predictor values. Then
the standardized coefficient error is normal, $RSS/\sigma^2$ has a
$\chi^2_{n-2}$ distribution, and these quantities are independent. Their
ratio gives

$$
\frac{\widehat\beta_j-\beta_j}{SE(\widehat\beta_j)}
\sim t_{n-2},\qquad j=0,1.
$$

That is the basis of exact classical conditional coverage for
$\widehat\beta_j\pm t_{.975,n-2}SE(\widehat\beta_j)$, and of the
corresponding null test. For the occupation slope the calculation is
$0.8988\pm1.984(0.127)=[0.6468,1.1508]$ thousand dollars per year.
It illustrates the model-based procedure; the diagnostics in Section 6
do not support treating its nominal exact coverage as established here.

Fixed-$X$ derivations condition on the predictor values; those values need
not literally have been experimentally fixed. A random-$X$ conditional-mean
model can use the same conditional argument when its error conditions hold.
Neither route demonstrates that these occupations were probability sampled.
Normality of the marginal distribution of education or income is not required.
The assumptions concern conditional errors, not the two marginal histograms.

### Why a new-case interval has an additional variance term {.unnumbered}

For a supported value $x_0$, define the model mean
$\mu_0=\beta_0+\beta_1x_0$ and its estimate
$\widehat\mu_0=\widehat\beta_0+\widehat\beta_1x_0$.
Substituting the coefficient variances and covariance above gives

$$
\begin{aligned}
\operatorname{Var}(\widehat\mu_0\mid\mathbf X)
&=\operatorname{Var}(\widehat\beta_0)
 +x_0^2\operatorname{Var}(\widehat\beta_1)
 +2x_0\operatorname{Cov}(\widehat\beta_0,\widehat\beta_1)\\
&=\sigma^2\left\{\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}\right\}.
\end{aligned}
$$

The common conditioning on $\mathbf X$ is suppressed in the first line.
A confidence interval for this mean uses the square root of this expression
with $s_e$ replacing $\sigma$.

A new occupation outcome is instead $Y_0=\mu_0+\varepsilon_0$.
If its error is independent of the fitted data and comes from the same
normal, equal-variance process,

$$
\operatorname{Var}(Y_0-\widehat\mu_0\mid\mathbf X,x_0)
=\sigma^2\left\{1+\frac1n+\frac{(x_0-\bar x)^2}{S_{xx}}\right\}.
$$

The extra $1$ is new outcome variation; it does not disappear even if the
model mean were known. Both standard errors use the same $t_{.975,n-2}$
critical value for their classical 95% intervals under the full model.

The table applies both formulas at four supported education values, the
same calculation as Section 5's two `predict()` calls, and puts both sets
of limits beside the common fitted values.

\Needspace{12\baselineskip}

| Education | Fitted | Mean lower | Mean upper | Case lower | Case upper |
|--:|--:|--:|--:|--:|--:|
| 8 | 4.3369 | 3.3651 | 5.3087 | $-2.6420$ | 11.3158 |
| 10 | 6.1345 | 5.4254 | 6.8437 | $-0.8127$ | 13.0818 |
| 12 | 7.9322 | 7.1776 | 8.6868 | 0.9802 | 14.8842 |
| 15 | 10.6286 | 9.3550 | 11.9022 | 3.6013 | 17.6559 |

: Mean-response and new-case intervals at four supported education values. {#tbl-ch07-app-a-intervals tbl-colwidths="[16,16,17,17,17,17]"}

At 10 years, the point prediction is 6.1345 thousand dollars, the mean
interval is [5.4254, 6.8437], and the new-occupation interval is
[$-0.8127$, 13.0818]. The
negative endpoint exposes the lower-tail limitation of this unbounded
normal-linear model. Cutting it off at zero does not produce a validated
95% interval. An outcome transformation or a model respecting the boundary
would change the analysis and would need its own interpretation and checks.

Both intervals widen away from $\bar x$; their narrowest point is the mean
education value. With more comparable independent observations and stable
predictor spread, mean uncertainty tends to shrink. The new-case interval
retains a nonzero error-variation floor. None of these formulas includes
uncertainty from an inappropriate mean function, biased measurement, or a
shift to another labour market. They are pointwise model intervals, not a
simultaneous band or empirical out-of-sample validation. Their target is a
new occupation-level record from the declared process, never an individual
worker's income.

## References {.unnumbered}

::: {#refs}
:::
