---
title: "How to Read a Quantitative Study as an Argument"
subtitle: "Argument Chapter A"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

references:
  - id: bloom2015wfh
    type: article-journal
    title: "Does working from home work? Evidence from a Chinese experiment"
    author:
      - family: Bloom
        given: Nicholas
      - family: Liang
        given: James
      - family: Roberts
        given: John
      - family: Ying
        given: Zhichun Jenny
    issued:
      date-parts: [[2015]]
    container-title: The Quarterly Journal of Economics
    volume: 130
    issue: 1
    page: 165-218
    DOI: 10.1093/qje/qju032
  - id: realinho2022
    type: article-journal
    title: "Predicting student dropout and academic success"
    author:
      - family: Realinho
        given: Valentim
      - family: Machado
        given: Jorge
      - family: Baptista
        given: Luís
      - family: Martins
        given: Mónica V.
    issued:
      date-parts: [[2022]]
    container-title: Data
    volume: 7
    issue: 11
    page: 146
    DOI: 10.3390/data7110146
  - id: uci697
    type: dataset
    title: "Predict students' dropout and academic success"
    author:
      - family: Realinho
        given: Valentim
      - family: Vieira Martins
        given: Mónica
      - family: Machado
        given: Jorge
      - family: Baptista
        given: Luís
    issued:
      date-parts: [[2021]]
    publisher: UCI Machine Learning Repository
    DOI: 10.24432/C5MC89

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    documentclass: scrartcl
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    latex-auto-install: false
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - ../koma_heading_tags.tex
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{16\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AtBeginDocument{\renewcommand{\boldsymbol}[1]{\symbf{#1}}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup-argument-a
#| include: false
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev="cairo_pdf")
# Verify the unchanged simplified student analysis.
d0 <- read.csv("realinho2022_students.csv")
dat <- d0
dat$drop <- as.integer(dat$recorded_outcome == "Dropout")
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
set.seed(12345)
idx <- sort(c(sample(which(dat$drop == 0), floor(.7 * sum(dat$drop == 0))),
              sample(which(dat$drop == 1), floor(.7 * sum(dat$drop == 1)))))
fit <- glm(drop ~ admission_grade + age, data=dat[idx, ], family=binomial())
p <- predict(fit, newdata=dat[-idx, ], type="response")
err <- mean((p-dat$drop[-idx])^2)
base <- mean((mean(dat$drop[idx])-dat$drop[-idx])^2)
stopifnot(nrow(dat)==4424, length(idx)==3096,
          abs(err-.200011325828)<1e-9, abs(base-.218150879563)<1e-9,
          sum(p>=.5 & dat$drop[-idx]==1)==73)
occ <- carData::Prestige
reg <- lm(I(income/1000) ~ education, data=occ)
stopifnot(nrow(occ)==102, abs(coef(reg)[2]-.8988128)<1e-7,
          abs(summary(reg)$r.squared-.3336)<.0001,
          abs(summary(reg)$sigma-3.4834)<.0001)
diagram_start <- function() {
  par(mar=c(.1,.1,.1,.1)); plot.new()
  plot.window(xlim=c(0,1),ylim=c(0,1))
}
support_row <- function(labels, y, color="grey95", functions=FALSE) {
  x <- seq(.12,.88,length.out=length(labels))
  for (i in seq_along(x)) {
    rect(x[i]-.105,y-.10,x[i]+.105,y+.10,col=color,border="grey35")
    text(x[i],y,labels[i],cex=.78)
    if (i<length(x)) {
      if (functions && i==1) text(mean(x[1:2]),y,"+")
      else arrows(x[i]+.11,y,x[i+1]-.11,y,length=.06)
    }
  }; invisible(x)
}
```
{{< pagebreak >}}

::: {.callout-note title="Chapter overview"}
Read one research question and the answer a study offers. Follow the specific
information supporting that answer across the paper, protocol, measurements
and analytical materials. Distinguish a procedure from its rationale, an
analysis from its claim, and identifying an argument from judging its quality.
A documented construction history shows how researchers produce that support.
The principal case is an invented feedback experiment; shorter prediction and
regression reports show how the reading transfers. All necessary records are
supplied here. This optional argument tutorial requires no programming and is
not a prerequisite for the numbered methods chapters.
:::

A published study asks a question and offers an answer. Between them are
observations, the circumstances in which they arose, reasons for trusting a
comparison, analytical work, and an interpretation. Much of that support lies
outside the sentence reporting the main result. A protocol may explain
allocation, an instrument may reveal what an outcome measures, and a supplement
may show what changed when an unusual observation was omitted. Learning to
connect those records is part of understanding quantitative research.

The invented feedback study will provide a complete example. Its reported
assignment effect is 2.5 points, with a 95% interval from 1 to 4 points.
Reading those numbers alone cannot tell us whether the study compared students
or classes, what the score represents, or why the comparison could support an
effect of assignment. The following sections make these connections explicit,
then give less structured cases in which you must decide which connection matters.

## From a Research Question to a Traceable Argument

### Start with what the study asks and answers

Consider two extracts from the invented study. Their labels are illustrative
locations in the teaching record, not quotations from an actual publication.

> **Paper, aim:** We examined whether assigning explanatory rather than usual
> feedback changes average delayed achievement for the 480 eligible students
> in 24 participating classes.
>
> **Protocol, allocation:** Twelve classes were randomly assigned to each
> format across the full pool of 24 classes. The comparison uses equally
> weighted class means and an uncertainty procedure reflecting class assignment.

The aim can be restated as a question: *What is the average effect of assigning
explanatory rather than usual feedback on delayed achievement for these eligible
students?* This is a reader's formulation of the stated aim. It names a causal
comparison, an outcome and a scope. An aim such as "improve education" would
need these details before it could identify a quantitative question.

The **estimand** is the target quantity, here the student-average assignment
effect for the eligible cohort. The **estimate** is the reported 2.5-point
result. The **reported answer** is the study's conclusion about that question.
It can be cautious or overstated; the reader first identifies the answer that
was actually given. Replacing it with a more defensible statement is an
evaluation and revision, not a discovery about what the original sentence said.

A **question-answer pair** gives an argument its task. Evidence and methods
are relevant through their contribution to this task. A well-computed coefficient
can still concern the wrong outcome, follow-up time or unit. Conversely, a
carefully supported answer can remain inconclusive. A research argument need
not end with a positive result to succeed.

### Four functions in two linked layers

An **argument component** names a function of information. Evidence supplies
observations and their production record; Justification supplies invoked
reasons and conditions; Inference connects those contributions through analysis
and reasoning; a bounded Claim states the conclusion with its limits.
The statistical and substantive layers are coordinated parts of one argument.
The statistical layer follows the numerical target and analysis; the substantive
layer keeps their meaning connected to the people, practices and question.

| Component | Statistical function | Substantive function | Connection |
|:--|:--|:--|:--|
| **Evidence (E)** | **STAT-E:** information about the observations, variables, and procedures through which the analysis data were produced or acquired | **SUB-E:** information about the people, setting, activities, and substantive meaning represented by those data | SUB-E **situates** STAT-E |
| **Justification (J)** | **STAT-J:** reasons and conditions invoked to support the intended statistical comparison and inference | **SUB-J:** domain-based reasons supporting measurement meaning, implementation, explanation, and applicability | SUB-J **augments and contextualizes** STAT-J |
| **Inference (I)** | **STAT-I:** the analytical reasoning and work connecting evidence and justification to statistical conclusions, including estimates, uncertainty, comparisons, and checks | **SUB-I:** interpretation of that statistical inference in substantive terms, retaining the relevant uncertainty and conditions | SUB-I **interprets** STAT-I |
| **Bounded claim (C)** | **STAT-C:** the reported statistical conclusion about the estimand, with uncertainty, conditions, and scope | **SUB-C:** the reported question-facing answer, with its substantive scope, qualifications, and unresolved matters | SUB-C **bounds** the substantive reach of STAT-C |

: Four component functions and their linked meanings. {#tbl-supa-components tbl-colwidths="[13,31,31,25]"}

In the feedback study, class assignments and assessment records contribute
Evidence. The reason that random allocation supports the assigned-group
comparison contributes Justification. Computing the mean contrast, its interval
and influence checks contributes Inference. Saying that assignment increased
average assessed performance in these classes contributes a bounded Claim.
These functions are not section headings: Methods can contain both a procedure
and its rationale, and Results can contain both a calculation and a conclusion.

```{r}
#| label: fig-supa-functions
#| fig-cap: "Four functions in two linked layers. Evidence plus Justification support Inference, which supports a bounded Claim; vertical links identify the substantive contribution."
#| fig-alt: "Substantive Evidence, Justification, Inference and Claim form the top row. The statistical functions form the bottom row. Downward links say situates, augments and contextualizes, interprets, and bounds substantive reach. Horizontal links show E plus J supporting I, and I supporting C. These are logical support links, not causal effects or a time sequence."
#| fig-width: 7
#| fig-height: 3.4
#| fig-pos: H
diagram_start()
xx <- support_row(paste0("SUB-",c("E","J","I","C")),.8,"#E8F1E6",TRUE)
support_row(paste0("STAT-",c("E","J","I","C")),.2,"#E4EEF5",TRUE)
for(i in 1:4) {
  segments(xx[i],.69,xx[i],.62,col="grey40")
  arrows(xx[i],.38,xx[i],.31,length=.06,col="grey40")
}
text(xx,.50,c("situates","augments and\ncontextualizes","interprets",
             "bounds\nsubstantive reach"),cex=.71)
```

The score's content situates the statistical measurement. Reasons why the
assessment represents the taught objectives augment its interpretation.
Reading the interval in course terms interprets the numerical result. Limiting
the answer to assessed performance in these classes bounds its substantive
reach. Numerical statements can do substantive work: "one application task"
is a domain interpretation of two score points. The layers are not numbers
versus words, and substantive reasoning starts when the question and measures
are chosen, not only after calculations finish.

### Connect a function to an inspectable location

**Traceable information** is specific study content through which a component's
function can be identified. A **documentary location** is where that content
can be inspected: a paragraph, table, protocol item, codebook entry or analysis
file. "Research design" names a topic. "Twelve classes were randomly assigned
to each format across the full class pool" identifies a particular procedure.

The useful connection is **question-answer pair → component function → specific
information → documentary location**. You can follow it in either direction.
Starting from Evidence, look for how groups arose; starting from an allocation
paragraph, identify its contribution to evidence production. Keep enough of the
location for another reader to find the same information.

Professional knowledge helps recognize the contribution. Knowing what a
confidence interval concerns helps identify which estimate's uncertainty is
reported. Recognizing a reason does not certify its quality. "Our software
estimates causal effects" is still a reason the authors offered, although
running software cannot supply a causal design. Reading records the offered
reason; evaluation judges it.

When information remains unclear, explain the particular gap. "The article
reports 480 scores but the participant-flow table reports 460; neither explains
the difference" is more informative than an overall label. A missing account
in accessible materials does not establish that the action never occurred.
Do not fill the gap with a procedure you would have preferred. These habits
make a reading useful without turning every paper into a compulsory checklist.

## Tracing Information Through E, J, I, and C in Both Layers

The following tables collect fifteen recurring information types: four for
Evidence, four for Justification, five for Inference and two for Claim. They
are reference aids, not an exhaustive list to memorize. The information in an
actual reading is the particular study's account. Return to a row when a
statistical term or an unfamiliar analysis makes its function difficult to see.

### Statistical Evidence: observations and their production

STAT-E includes observations and the record of their production.
@tbl-supa-stat-e distinguishes the structure of the records,
measurement, selection and implementation, and the observations
actually analyzed.

```{=latex}
\Needspace{13\baselineskip}
```

An *observed variable* is a characteristic recorded for study units. In the
feedback-format example, each student has an assessment score and a
recorded feedback format. The score is numerical, so a mean and a
measure of variation can help describe it. Practice format is
categorical, so counts or percentages show how students are distributed
across the categories. The recorded values contribute to STAT-E;
calculations that summarize them contribute to STAT-I.

A variable's *role* concerns how it enters a particular analysis. An *outcome*
(or dependent variable) is what the analysis describes, explains, or
predicts. If the question asks how assessment scores differ between
formats, score is the outcome and feedback format identifies the
comparison groups. A predictor, such as a prior assessment score, can
help model the outcome. Papers may use "independent variable" for a
predictor or an experimentally manipulated condition. Identify the role
from the question and analysis, then follow the design and justification
to understand whether the relationship receives a causal interpretation.
The same variable can serve a different role in another analysis.

```{=latex}
\Needspace{12\baselineskip}
\begingroup
\small
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.08}
```

| Traceable information | What to identify | Common cases | Likely documentary location |
|:----------------------|:-----------------|:-------------|:----------------------------|
| Data structure, units, and variable roles | What a record represents; repeated or grouped observations; outcome and predictor roles. | Students are recorded within classes; assessment score is the outcome and assigned format is the grouping variable. | Data description; Measures; dictionary |
| Measurement scales, instruments, and procedures | The response scale, instrument, scoring, and administration actually reported. | Responses use ordered categories; an assessment is scored from 0 to 40 and administered two weeks after instruction. | Measures; instrument; scoring guide; Procedures |
| Design and implementation information | Whom the study concerns; sample selection or recruitment; assignment or exposure; comparison structure and timing. | Classes are recruited from schools and randomly assigned to formats; another study observes existing exposures without assigning them. | Participants; Design; Procedures; protocol |
| Analysis sample, exclusions, and missing data | Which units and observations entered each analysis, their counts, and reported omissions or exclusions. | Some recruited participants lack follow-up scores; an analysis uses the records meeting a stated inclusion rule. | Participant flow; Results; data-processing record |

: Statistical Evidence: Traceable Information {#tbl-supa-stat-e tbl-colwidths="[24,27,30,19]"}

*Note.* Identify both the number entering the study and the number
included in each analysis. A participant may contribute several
observations, so distinguish participant counts from observation counts.

```{=latex}
\endgroup
```

**Measurement information is classified by its use.** The instrument,
scoring, and administration record contribute to STAT-E. Reasons for
using those measurements in a statistical procedure belong in STAT-J.
What the scores mean substantively belongs in SUB-E, while reasons
supporting that meaning belong in SUB-J. To identify Justification,
look for the reasons given for using or interpreting the instrument.

### Statistical Justification: invoked reasons and conditions

STAT-J identifies the reasons and conditions the study invokes for its
statistical reasoning. @tbl-supa-stat-j separates support from the
design, sample-size planning, method choice, and assumptions.

```{=latex}
\Needspace{12\baselineskip}
\begingroup
\small
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.08}
```

| Traceable information | What to identify | Common cases | Likely documentary location |
|:----------------------|:-----------------|:-------------|:----------------------------|
| Design-based support for inference | How sampling, assignment, or comparison design is invoked to support the intended inference. | Probability sampling is invoked for population inference; random assignment is invoked to support an assigned-group comparison. | Design rationale; sampling plan; protocol |
| Sample-size, power, and precision rationale | Why the amount of evidence was planned in relation to an inferential goal. | Planned sample size aims at a stated confidence-interval width or power for a specified effect. | Sample-size calculation; protocol; planning supplement |
| Rationale for selecting statistical methods and models | Reasons connecting the chosen procedure to the estimand, measurement scale, or data structure. | Welch's method is chosen because it allows unequal group variances; within-class dependence is allowed because students share classes. | Statistical Analysis; analysis plan |
| Statistical assumptions invoked | Conditions the study relies on for its comparison, model, or uncertainty calculation. | Independence between observations is assumed for a standard-error calculation; a model assumes a linear relationship. | Statistical Analysis; assumptions statement; technical supplement |

: Statistical Justification: Traceable Information {#tbl-supa-stat-j tbl-colwidths="[24,27,30,19]"}

*Note.* Alongside the named plan or method, identify the reasons invoked
for its use. Evaluate those reasons' adequacy separately.

```{=latex}
\endgroup
```

```{=latex}
\Needspace{13\baselineskip}
```

**Sampling and assignment are distinct parts of study design.** Sampling
concerns which units enter a study; assignment concerns which conditions
they receive. Random sampling provides a basis for population inference
under the sampling design, contributing to generalizability or external
validity. Random assignment supports a causal comparison of assigned
conditions, contributing to internal validity. Both forms of support
depend on the relevant implementation and analysis conditions. A study
may use either form of randomization or both. To trace assignment, look
for the allocation procedure used to form the groups. A protocol
documents the planned procedure; implementation records document
what was carried out.

**A procedure and its rationale do different work.** "Students were
randomly assigned" records a production procedure in STAT-E.
"Random assignment provides the basis for estimating the effect of
assignment from the comparison" identifies the reason invoked in
STAT-J. Locate this connection in an explicit rationale or by following
the documented links among design, estimand, and analysis.

```{=latex}
\Needspace{8\baselineskip}
```

Likewise, "100 participants were analyzed" records the analysis sample.
"The sample size was planned to achieve the desired precision" supplies
a planning rationale. *Precision* concerns estimation uncertainty;
*power* concerns the probability of rejecting a null hypothesis under
a specified alternative and procedure. The reader identifies the stated
goal before considering whether the planning was adequate.

A statistical *model* represents selected relationships or variation
mathematically. When a study invokes an assumption as a condition
supporting inference, that contribution belongs in STAT-J. The
specification actually fitted, and the reported analytical checks,
contribute to STAT-I. A passage can describe the fitted model and
invoke its assumptions as support; identify what each contribution
does. "We used SPSS" helps identify the software used for computation.
Look for the method-selection rationale to identify the reasons for
the analytical choice.

### Statistical Inference: analytical work and its consequences

STAT-I is often the most demanding component to understand, both
technically and in its statistical reasoning, especially in large-scale,
complex studies. It traces the central statistical information of the
argument: how evidence, analyzed under the study's stated conditions and
justification, supports a statistical conclusion. Understanding this
transition requires reasoning about variation, uncertainty, models, and
alternative analyses. Some of these ideas challenge everyday intuition,
so learning to follow the reasoning is as important as learning the
procedures. Much of the statistical reasoning developed in later chapters
therefore relates directly to STAT-I. Its contribution becomes clear
through its connections with evidence production, justification,
substantive interpretation, and the bounded answer.

STAT-I connects the estimand with the analysis performed, its results,
uncertainty, and relevant checks. *Model building* specifies the
statistical relationships to be estimated. *Diagnostics* examine how
the fitted model and estimation procedure behave; *sensitivity analysis*
examines how results change under alternative analytical choices.
A *statistical hypothesis* expresses a proposition to be examined,
such as a zero mean difference. @tbl-supa-stat-i connects these
activities to their traceable information and reported results.

```{=latex}
\Needspace{12\baselineskip}
\begingroup
\small
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.08}
```

| Traceable information | What to identify | Common cases | Likely documentary location |
|:----------------------|:-----------------|:-------------|:----------------------------|
| Describing data and relationships | How observations and associations are summarized. | Means and standard deviations describe scores; a correlation describes a linear association between two numerical variables. | Descriptive tables; Results; figures |
| Model building and estimation | Model specification in equations, diagrams, or syntax; connection to declared estimands; estimation procedure, computational implementation, and point estimates. | A regression equation specifies score differences by feedback format; software estimates its coefficients and the declared contrast. | Statistical Analysis; model diagram; Results; syntax or code |
| Model diagnostics, modification, and sensitivity analysis | Fit and computational checks; model or analysis changes; sensitivity to analytical choices or priors; predictive validation and its results. | Residuals suggest a curved relationship; an alternative prior changes estimates; held-out observations assess prediction. | Diagnostic figures; analytical supplement; validation results; code |
| Statistical uncertainty and interval estimates | The reported uncertainty for an estimate or prediction, including its procedure and stated level. | A standard error and confidence interval accompany a mean difference; a prediction interval accompanies a new-case prediction. | Results; interval tables; figures |
| Statistical hypotheses, testing, and comparison | Hypotheses or restrictions examined; tests and comparison criteria; competing models where relevant; test statistics, p-values, or other comparison results. | A test examines a zero mean difference or equal coefficients; competing models are compared by a test or predictive performance. | Hypothesis statement; Statistical Analysis; Results; model-comparison table |

: Statistical Inference: Traceable Information {#tbl-supa-stat-i tbl-colwidths="[24,27,30,19]"}

*Note.* Some descriptive questions are answered by summaries alone.
For analyses and modifications, record whether they were prospectively
specified, exploratory, or revised when known. "Primary" describes
priority. Separately describe when and how the analysis was specified or
changed, and whether relevant outcomes had already been examined.

```{=latex}
\endgroup
```

**Connect a calculation to the question it addresses.** A group mean
difference may estimate the declared contrast between feedback formats.
For a question about how two numerical variables vary together, a
correlation summarizes their linear association. Regression describes
an outcome in relation to one or more predictors; the coefficients
address particular relationships in the fitted model. The reported
calculation and its role in answering the question belong together
when you read STAT-I.

**Follow model building through to the estimates.** An equation or a
statistical path diagram might represent assessment score as an outcome
related to feedback format and prior score. Identify how the model
represents the declared estimand, which could be a coefficient or a
contrast calculated from several coefficients. Then follow the
estimation procedure to its point estimates. Computational information
may appear in R scripts, software syntax, menu settings, or a Methods
description; each can help identify what was implemented.

Specifying the model and fitting it contribute to STAT-I. Reasons for
choosing that specification or estimation method contribute to STAT-J.
Likewise, a diagram expressing the fitted statistical relationships
helps trace STAT-I, while a diagram explaining a substantive mechanism
may contribute to SUB-J. The diagram's contribution depends on what
it represents in this study.

```{=latex}
\Needspace{10\baselineskip}
```

**Distinguish variation in the observations from estimation uncertainty.**
Students' assessment scores vary, and a standard deviation describes
that variation. An estimated mean difference also has uncertainty:
its standard error describes how the estimate would vary under the
specified sampling, assignment, or model conditions. A confidence
interval concerns the mean difference. For example, an interval from
1 to 5 points expresses uncertainty about the difference between means,
rather than the range of individual students' scores.

A p-value addresses a null hypothesis under the test's assumptions;
it is not the probability that the null is true. These concepts receive
fuller statistical treatment later. For now, identify what each output
concerns and follow the reported uncertainty into the substantive
interpretation and the study's conclusions.

**Follow the checks, changes, and their consequences.** Diagnostics can
examine model fit, unusual observations, or whether the estimation
algorithm reached a stable solution. A modification changes the model
or analysis, for example by adding a curved relationship after examining
residuals. Record what changed, why the study says it changed, and the
result. A stated reason for the change may also contribute to STAT-J.

Sensitivity analysis asks how estimates or conclusions change under
alternative choices or assumptions. In a Bayesian analysis, a *prior
distribution* expresses information or assumptions about model
parameters before incorporating the current data. Repeating an analysis
with alternative priors can show how sensitive its findings are to those
choices. The prior used is part of the model specification; reasons
invoked for choosing it can contribute to STAT-J. Predictive validation
asks how the model predicts outcomes,
for example for held-out observations. Checks, modifications, and
sensitivity analyses can inform one another. During reading, identify
the study's reported checks and assessments; quality evaluation then
considers whether they provide adequate support.

**Distinguish specifying a restriction from examining it.** A model
*constraint* restricts the values parameters may take, for example by
setting a coefficient to zero or requiring two coefficients to be equal.
Using such a constraint to specify a model belongs with model building.
Examining the restriction through a test or comparison belongs with
statistical hypotheses, testing, and comparison. Some constraints instead
serve to define a model's scale or make its parameters estimable; identify
the role of each constraint from the study's account.

Model comparisons can also use criteria other than hypothesis tests,
such as prediction error on held-out observations. Record which models
were compared, by what criterion, and with what result. The same
analysis may contribute both predictive validation and a comparison;
identify these connected uses without treating them as separate analyses.

### Statistical Claim: the conclusion and its scope

STAT-C records the statistical proposition the study concludes, together
with its stated qualifications. @tbl-supa-stat-c uses familiar
ideas from the analysis in the different function of stating an answer.

```{=latex}
\Needspace{18\baselineskip}
\begingroup
\small
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.08}
```

| Traceable information | What to identify | Common cases | Likely documentary location |
|:----------------------|:-----------------|:-------------|:----------------------------|
| Reported statistical conclusion | What is concluded about direction, magnitude, or pattern, including a testing conclusion when reported. | The study concludes that mean scores are higher in one group; it describes an association as negative. | Results summary; Abstract; Discussion |
| Remaining uncertainty and scope of the conclusion | What remains unresolved and which units, comparisons, conditions, or periods the conclusion covers. | Whether an effect reaches a stated size remains unresolved; the conclusion is restricted to participating units. | Results interpretation; limitations; Conclusion |

: Statistical Claim: Traceable Information {#tbl-supa-stat-c tbl-colwidths="[24,27,30,19]"}

```{=latex}
\endgroup
```

**Connect analytical outputs to the statistical conclusion.** A calculated mean
difference with its interval contributes to STAT-I. "The comparison
indicates a positive mean difference for the participating students"
states a conclusion in STAT-C. A reported claim of statistical
significance may be part of C; the p-value producing that claim remains
an analytical result in I. Include a significance claim when the study
makes one, and record the answer with its reported bounds. Keep any
reader-proposed revisions separate. "We investigated whether the
groups differ" identifies the question; look for the reported finding
to identify its answer.

### Following substantive meaning across the linked layers

The substantive layer builds on the statistical layer and connects it
to the people, measures, practices, and settings named by the question.
Its information is more domain-dependent and can be directly documented:
look for what an assessment covers or what an intervention involved.
@tbl-supa-sub retains four component rows;
the cases illustrate different possible studies.

```{=latex}
\Needspace{12\baselineskip}
\begingroup
\small
\setlength{\tabcolsep}{3pt}
\renewcommand{\arraystretch}{1.08}
```

| Component | Traceable information | What to identify | Common cases | Likely documentary location |
|:----------|:----------------------|:-----------------|:-------------|:----------------------------|
| **SUB-E** | People and setting, measure meanings, and implemented practices | What the recorded units, scores, conditions, or exposures represent in the substantive domain. | A course assessment covers application tasks; guided practice consists of worked examples and feedback. | Participants; Measures; instrument; intervention description |
| **SUB-J** | Theory, mechanisms, and reasons supporting interpretation | Reasons invoked about measurement, implementation, comparability, alternatives, practical benchmarks, or applicability elsewhere. | Assessment coverage supports the score interpretation; an importance benchmark is linked to expected benefits and effort. | Introduction; measurement rationale; protocol; Discussion |
| **SUB-I** | Contextual interpretation of statistical inference and uncertainty | How findings and uncertainty relate to substantive meanings, expectations, benchmarks, or proposed explanations. | An interval spans gains below and above a practical benchmark; the reported pattern is related to a proposed mechanism. | Results interpretation; Discussion |
| **SUB-C** | Question-facing answer, scope, and qualifications | The reported answer to the substantive question or answerable aim, with its stated limits and unresolved matters. | A practice improved delayed course performance in the studied setting, while the importance of the gain remains unresolved. | Abstract; Discussion; Conclusion |

: Substantive Components: Traceable Information {#tbl-supa-sub tbl-colwidths="[12,21,24,26,17]"}

*Note.* A numerical statement can perform a substantive function when
it expresses domain meaning. Identify the interpretation during reading
and assess its support during quality evaluation.

```{=latex}
\endgroup
```

**Identify what a measure means and the reasons offered for that meaning.**
A description of an assessment's application tasks tells you what the
assessment covers, contributing to SUB-E. An explanation that these
tasks cover the course objectives offers a reason for interpreting the
score as course performance, contributing to SUB-J. Follow the
instrument name to its description and the reasons given for
interpreting its scores.

The same distinction helps with theory and mechanisms. A paper might
describe a theory in its Introduction and later invoke one of its
predictions to explain why feedback could improve performance.
Identify the particular connection offered as substantive support.
This makes the theory's role in SUB-J visible, beyond simply recording
that the paper mentions it. Apply the same reasoning to implementation,
alternative explanations, and applicability elsewhere.

**SUB-I retains the statistical uncertainty.** When interpreting an
interval against a practical benchmark, retain its full range and the
possibilities it leaves open. Distinguish a pattern described as
consistent with a proposed mechanism from evidence establishing that
mechanism.
SUB-C then states the study's question-facing answer and its
qualifications. Readers identify these interpretations and claims;
quality evaluation asks whether the support warrants them.

```{=latex}
\Needspace{24\baselineskip}
```



A condition may be invoked even though the study has not checked it, and a
check may later be invoked as support for another inference. Identify the use
made of the information. Assuming independent classrooms is Justification for
an uncertainty calculation. Examining dependence indicators and reporting the
result is Inference. Using that result as a reason for a subsequent comparison
gives it an additional justificatory role.

For missing outcomes, the number absent is Evidence; the assumption used to
analyze incomplete records is Justification; imputation or a sensitivity
analysis and its results are Inference. A method's name alone does not settle
what missingness mechanism would make its use defensible [@littlerubin2019].
Sensitivity analysis can make the conclusion's dependence on an unverifiable
assumption visible [@nrc2010]. It need not remove that dependence.


## Reading One Argument Across a Paper and Its Related Materials

### The complete feedback packet

The following invented packet expands the same supplied feedback result. It
adds documentary detail without generating a second experiment. The estimate
and interval are stipulated teaching results; no set of student records is
claimed to reproduce them. The class allocation is across the whole pool,
not one class of each format forced within every school.

> **F1 -- Abstract.** We asked whether assignment to explanatory feedback
> changes average delayed achievement for eligible students in participating
> classes. Assignment increased average assessed course-topic performance
> under these conditions; the practical size remains uncertain.
>
> **F2 -- Methods.** Twelve schools contributed two classes each, with 20
> eligible students per class. Before a six-week program, 12 of the 24 classes
> were randomly assigned to explanatory feedback and 12 to usual feedback
> across the full class pool. All 480 students supplied the common 0--40
> assessment two weeks after the program and were analyzed by class assignment.
>
> **F3 -- Protocol.** Random allocation supplies the basis for the assignment
> comparison. Equally weighted class means give students equal weight because
> all classes have 20 eligible students. The uncertainty procedure reflects
> assignment of classes. The inference assumes implementation of allocation,
> a common outcome definition, complete follow-up and no influence of one
> class's assignment on another class's outcomes.
>
> **F4 -- Assessment and implementation supplement.** Both formats used the
> same topics and scheduled time. Explanatory feedback added explanations of
> errors; usual feedback supplied correctness feedback. Delivery logs documented
> the activities and separate materials. The assessment blueprint linked tasks
> to taught topics, and scorers worked without knowledge of format. The curriculum
> team judged two points, full credit on one application task, worth considering
> against the additional preparation.
>
> **F5 -- Results and analytical supplement.** The estimated student-average
> assignment effect was 2.5 points, approximate 95% CI [1, 4], with uncertainty
> reflecting class assignment. Checks confirmed score ranges, allocation,
> denominators and complete outcomes. Recalculating the contrast with each
> school omitted in turn left its point estimate positive. These are influence
> checks; the full eligible cohort remains the basis of the primary answer.
>
> **F6 -- Discussion.** Assignment improved average assessed performance on
> the taught topics in these classes and conditions. The compatible magnitudes
> include gains below and above the local two-point benchmark. Broader learning,
> durable effects beyond this follow-up and effects for every individual remain
> unresolved.

The claim in F1 is brief. F5 supplies its uncertainty, and F6 explains its
scope. Reading these together preserves what the abstract alone leaves implicit.
The protocol's conditions also matter: the words "random assignment" cannot
substitute for the allocation actually used or a matching uncertainty procedure.

### Follow the comparison through both layers

| Quantity or feature | Supplied result | Meaning in the argument |
|:--|:--|:--|
| Eligible students / assigned classes | 480 / 24 | Students define the target; classes receive assignment |
| Explanatory-minus-usual estimate | 2.5 points | Estimated student-average assignment effect |
| Approximate 95% interval | [1, 4] points | Uncertainty reflecting class assignment |
| Local importance benchmark | 2 points | One application task, considered against preparation effort |
| Assessment occasion | Two weeks after a six-week program | Delayed course-topic performance at the specified follow-up |

: The result and the information needed to interpret it. {#tbl-supa-feedback-results tbl-colwidths="[29,23,48]"}

The distinction between students and classes connects several components.
The recorded score belongs to a student; allocation belongs to a class; the
estimand weights students equally; and equal class sizes make equally weighted
class means compatible with that target. Treating all 480 rows as independently
assigned would change the inferential justification. The number of records in
a file does not determine the number of independent assignments.

| Function | Particular contribution | Location |
|:--|:--|:--|
| STAT-E | Class allocation, eligible records, scoring occasion and complete analysis sample | F2, with implementation in F4 |
| STAT-J | Allocation rationale, equal-size weighting and conditions of the class-based uncertainty procedure | F3 |
| STAT-I | Estimated contrast and interval; coding, denominator and influence checks | F5 |
| STAT-C | Positive eligible-cohort average assignment-effect conclusion, with uncertainty | F1, F5--F6 |
| SUB-E | Students, course topics, feedback activities and delayed assessment | F2 and F4 |
| SUB-J | Blueprint, common conditions, delivery records and the local benchmark's rationale | F4 |
| SUB-I | Interval read against the application-task benchmark at this follow-up | F5--F6 |
| SUB-C | Bounded answer about assessed course-topic performance, with broader claims unresolved | F6 |

: A completed reading of the feedback argument. {#tbl-supa-feedback-traces tbl-colwidths="[15,63,22]"}

One document can contribute to several functions, but locating it twice does
not create two independent pieces of evidence. The blueprint's description
of tasks helps situate the outcome; its alignment with course objectives is
also invoked as a reason for interpreting it. The statistical result supports
a mean effect; it does not show that every student's score increased by 2.5.

::: {.callout-note title="Predict, then explain: keep the numbers, change the design"}
Suppose the same reported estimate and interval appear in a different study
where teachers selected their preferred feedback format. Before reading on,
predict which part of the argument would change first and whether the same
causal answer would remain warranted.
:::

The allocation account changes Evidence and removes the random-assignment
reason in Justification. Selection could be related to prior achievement,
teacher resources or other causes of outcomes. The numerical table can remain
the same while its causal warrant changes. Its interval also needs a stated
uncertainty justification for the new design; copying the numbers does not
carry the old calibration with them. A descriptive difference remains
reportable, but it answers a different question from an assignment effect.

### A missing explanation and a conflicting count

Two variants of the packet show why identification and judgment remain distinct.
In the first, F3 is omitted from the paper but available in its protocol. The
rationale is accessible: state that the protocol explains weighting and
assignment-based uncertainty. Publication outside the article does not by
itself weaken the reason, although readers need a way to find it.

In the second variant, F2 says all 480 students were analyzed, while a supplied
flow table says 460 completed follow-up. The discrepancy is unresolved; the
packet gives no reconciliation. State both accounts and their locations. Do
not silently choose the number that makes the result easier to interpret.
The target still concerns 480 eligible students, while the actual analysis
membership and handling of missing outcomes need clarification. This may
limit Evidence integrity and the ability to judge the whole answer.

The task is specific: which students are absent from which calculation, why,
and how were their outcomes handled? Neither "the report has a supplement"
nor "missing data were handled statistically" answers it. A useful reading
identifies what further information could resolve the question.

### A shorter transfer: predicting recorded student outcomes

The real data of Realinho and colleagues contain 4,424 students from the
Polytechnic Institute of Portalegre, Portugal [@realinho2022; @uci697]. Outcomes
were recorded at the normal degree duration. Dropout can include changing
course or institution; Enrolled means still enrolled at that endpoint; Graduate
means completed. The binary teaching analysis compares recorded Dropout with
the other categories combined, while preserving all three labels in the data.

The question is whether admission grade and age at enrollment improve
prediction of that recorded outcome for held-out students from the same source.
Both predictors are available at enrollment. Semester results would be too late
for this intended prediction time. The teaching analysis uses a fixed,
stratified approximately 70% training split and fits a logistic rule; this is
a simplified analysis, not a reproduction of the source authors' classifiers.

| Records | Dropout | Enrolled | Graduate | Total |
|:--|--:|--:|--:|--:|
| Full release | 1,421 | 794 | 2,209 | 4,424 |
| Training | 994 | 562 | 1,540 | 3,096 |
| Held out | 427 | 232 | 669 | 1,328 |

: Counts supporting the prediction comparison. {#tbl-supa-prediction-records}

The fitted rule assigns a probability to each held-out student's recorded
Dropout status. Average squared probability error is 0.2000, against 0.2182
for assigning everyone the training-data dropout proportion; lower is better.
These errors compare probability predictions, rather than count correct alerts.
At a probability cutoff of 0.5, the fitted rule flags 73 of the 427 held-out
Dropout records, or 17.096%. A better average probability score therefore
coexists with missing most eventual Dropout records at this threshold.

Outcome definitions, predictors, split and denominators are Evidence. Their
availability at the intended time and separation of fitting from assessment
supply reasons in Justification. The fit, probabilities, error comparison and
threshold calculation are Inference. The reported Claim is that this simple
rule improves average probability prediction on the internal held-out records,
while this cutoff identifies only a small fraction of recorded Dropout.

In substantive terms, an alert is about this administrative outcome, not
necessarily departure from higher education. The released data have no
cohort-year field for an independent later-cohort assessment. Predicting well
within this split neither establishes performance for a future cohort nor
shows that an intervention prompted by an alert would help. Those are
separately owed questions with different evidence requirements.

### Focused passages from three other settings

**Working from home.** Bloom and colleagues' experiment randomly assigned
eligible volunteer employees, 131 to home work and 118 to office work
(Appendix Table A.2, p. 214). The nine-month arrangement and four home-working
days per week are described on p. 169. Table II, column 4, p. 187 reports a
weekly log-call analysis for 134 order takers, using employee and week effects
and uncertainty clustered by employee; the authors interpret the result on
p. 188 as about 13% more weekly calls [@bloom2015wfh]. These are the paper's
rounded reporting choices. The 134 analysis employees are not a contradictory
count of all 249 assigned employees: follow the outcome and employee group
for the selected analysis. Allocation documents production; its invoked causal
reason belongs to Justification; a regression and its reported comparison
belong to Inference. The [author's materials](https://nbloom.people.stanford.edu/research)
allow a reader to inspect the underlying reports. Repeated weekly outcomes
also explain why counting weeks as independent employees would be a mistake.

**Confidence and belonging.** In an invented study, three 0--100 confidence
indicators ask about explaining, applying and checking learning tasks; three
belonging indicators concern acceptance, inclusion and participation. The
questions, scoring and administration supply measurement records. Their
content situates the scores in a substantive domain. Reasons why these items
represent the intended constructs provide substantive Justification. A scoring
rule alone cannot establish that interpretation. Reading these contributions
requires no factor analysis in this chapter.

**Language progress.** An invented student scores 58, 65, 68 and 76 at equally
spaced occasions; another scores 62, 63, 67 and 68. Their average gains per
interval are 6 and 2, but successive observed gains are 7, 3, 8 and 1, 4, 1.
A constant average is not a constant observed trajectory. Student and occasion
identifiers document the repeated structure. A reason for modeling within-student
dependence is distinct from the data layout. A missing assessment is also
different from a student who was never observed or eligible for this study.

### Follow the organization and the reporting

Two auxiliary tasks help when a reading becomes difficult. **Identifying
organization** asks which result belongs to which question, what information
is shared, and which results support rather than replace the focal answer.
The feedback influence checks support the one assignment-effect argument;
each omitted-school calculation does not create a new substantive question.
A separate question about preparation burden would need its own stated answer
unless it had been defined as a strand required for one integrated decision.

**Following reporting** connects an abstract's conclusion to a Results table's
interval and a Discussion's scope. F1, F5 and F6 form that chain. The task
adds no new component: it follows how the existing contributions are expressed
and located. A concise abstract and a qualified Discussion need to be read
together, but the fuller passage does not excuse a plainly contradictory
headline. Argument B develops these tasks when several answers depend on one
another. Here they organize a reading without adding another taxonomy.

### Read a reported regression argument

The following teaching report concerns 102 historical occupation records from
`carData::Prestige` [@carData2022]. It is written for this tutorial. R1--R4 are
locations within the teaching report, not quotations from the dataset's authors.
A slope describes how fitted occupation-average income changes with education;
$R^2$ describes the fraction of recorded income variation accounted for by the
line, and residual SD describes remaining vertical scatter in income units.
These local reminders support the reading; no regression calculation is required.

::: {.callout-note title="Teaching report: an occupation-level description"}
> **R1 -- Question and evidence.** We described the education--income
> relationship in the 102 *carData::Prestige* occupation records. Income
> was expressed in thousand 1971 Canadian dollars. All records had both
> modeled measurements. Each occupation was weighted equally; the observed
> education range was 6.38--15.97 years.
>
> **R2 -- Offered reasons.** The upward scatterplot motivated a
> straight-line summary. OLS was chosen because it minimizes squared
> vertical discrepancies. The aim was a summary of these records, not a
> constant causal effect or an individual-level comparison. The source
> documentation identifies the measurements as historical occupational
> averages. No probability-sampling justification was available in the
> record consulted.
>
> **R3 -- Analysis and interpretation.** The fitted slope was 0.8988
> thousand dollars per education year, with $R^2=.3336$ and residual SD
> 3.4834 thousand dollars. The residual pattern and a curved-fit
> sensitivity qualified the line as an incomplete description. The fit and
> sensitivities were exploratory. The descriptive result itself was not
> assigned sampling uncertainty.
>
> **R4 -- Reported answer.** Across these 102 recorded occupations, higher
> average education accompanied higher average income overall. The
> least-squares line summarized that pattern by about 899 additional 1971
> dollars of fitted occupation-average income per education year over the
> observed range. It did not describe every occupation closely or establish
> an individual or causal return to education.
:::

| Function | Content read in this report | Where the information is found |
|:--|:--|:--|
| STAT-E | Records, measurements, scaling, and complete sample | R1 |
| STAT-J | Reasons for OLS as a descriptive, noncausal summary | R2 |
| STAT-I | Slope, fit, sensitivity, and exploratory status | R3 |
| STAT-C | Positive but incomplete description over recorded support | R4 |
| SUB-E | Historical occupation averages, not individual measurements | R1 and the dataset documentation |
| SUB-J | Measurement meanings supporting an occupation-level reading | R2 and the dataset documentation |
| SUB-I | Meaning of about 899 dollars per education year | R3--R4 |
| SUB-C | Bounded historical answer, without a causal or individual return | R4 |

: Reading one regression argument with linked statistical and substantive layers. {#tbl-supa-regression-traces tbl-colwidths="[12,42,46]"}

R4 names the 102 recorded occupations and the descriptive line. It does not
claim a population slope, so the reader attaches no confidence interval to that
answer. The fitted slope is one statistical strand; fit summaries and checks
support its interpretation rather than supply separately owed substantive answers.

Now replace R2's reason with "We used R, so our slope is an unbiased causal
effect." The offered reason is identifiable STAT-J content even though it is
inadequate. Reading records it as given. The next section judges whether it
supports the claim. A reader does not silently supply random assignment or
replace the report's question with a better one.


## Evaluating the Quality of the Question-Level Argument

### Six questions, with reasons rather than scores

Evaluation depends on the argument that has been read. Its six questions focus
attention on a particular weakness or strength; they are not six new components
and their judgments are not a total quality score. Some questions overlap in
the consequence they reveal. Name where a problem first arises, then explain
what it changes later in the argument.

| Group | Evaluation question | What to consider |
|:--|:--|:--|
| **Alignment** | **Question alignment** | Whether the question, estimand, design, evidence, analysis, and answer remain directed at the same aim |
| **Support** | **Evidence integrity** | Whether the observations and contextual records, their production, and their representation are trustworthy, traceable, and appropriate |
| **Support** | **Justification adequacy** | Whether the statistical, design, measurement, and contextual reasons warrant the proposed moves |
| **Support** | **Inferential adequacy** | Whether reasoning within both layers is adequate, including estimation, uncertainty, checks, sensitivity, and substantive interpretation |
| **Translation** | **Transition fidelity** | Whether the links between layers retain meaning, uncertainty, design status, and scope |
| **Translation** | **Claim calibration** | Whether the reported answer's strength and breadth match its support |

: Six questions for evaluating one question-level argument. {#tbl-supa-evaluation tbl-colwidths="[17,28,55]"}

A local judgment is **adequate for this purpose**, **needs revision**,
**insufficient information to judge**, or **not applicable**. Accompany it with
a reason: "the complete cohort supports the stated denominator" is informative;
"Evidence integrity: adequate" alone does not show what was inspected.

The whole question-answer pair can be **supported as stated**, **supported
only after narrowing or revision**, **not supported**, or **not assessable from
the accessible record**. These outcomes distinguish a warranted answer from an
unwarranted one and an unresolved documentary gap from a demonstrated failure.
An inconclusive answer can be supported as stated. A study need not discover
a benefit to provide a defensible answer to its question.

### Evaluate the feedback claim and its variants

The baseline feedback answer retains the eligible cohort, assigned classes,
assessment meaning and follow-up. Its interval is above zero but crosses the
two-point benchmark. Within the supplied conditions, the positive average
assignment-effect conclusion is supported as stated, with practical size
uncertain. That judgment is about this bounded answer, not a certificate that
the study has answered every educational question.

| Evaluation question | Application to the feedback packet |
|:--|:--|
| Question alignment | The answer remains about the eligible-cohort assignment effect at the stated occasion |
| Evidence integrity | Allocation, common scoring, denominators and complete follow-up are documented |
| Justification adequacy | The class-assignment comparison and equal-size weighting address the target, under the stated implementation and no-between-class-influence conditions |
| Inferential adequacy | The stated uncertainty reflects assignment; checks inform the calculation without establishing every assumption |
| Transition fidelity | Two score points retain their course-specific meaning; a cohort-average effect does not become each student's gain |
| Claim calibration | A positive delayed-assessment conclusion retains an unresolved practical magnitude and limited reach |

: Applying evaluation to the supplied feedback answer. {#tbl-supa-feedback-quality tbl-colwidths="[28,72]"}

Now consider "the format produces substantial, lasting improvements in all
forms of learning for every student." The result has not changed. Transition
fidelity needs revision because an assessment of taught topics has become all
learning and an average has become every individual. Claim calibration needs
revision because the two-week follow-up has become indefinitely lasting and
practical importance has been settled despite the interval crossing the
benchmark. Narrowing that overstatement back to the declared cohort-average
question preserves a supported answer. Claiming an individual response instead
would introduce a different estimand requiring different evidence.

For the conflicting-count variant, Evidence integrity is unresolved. If the
missing outcomes could change the comparison and their handling is unknown,
the whole answer may be not assessable from the accessible record. This differs
from a documented coding error that demonstrably changes the result: then the
calculation and answer need revision. Different concerns call for different
remedies, not a common penalty score.

::: {.callout-note title="Predict, then explain: the estimate stays the same"}
A separate invented study reports the same 2.5-point estimate but an interval
from $-1$ to 6 points. Its local benchmark is still two points. What changes in
the conclusion about direction and importance, and what stays unchanged?
:::

The point estimate stays positive, but the interval now includes a negative
effect, zero and gains both below and above the benchmark. The study cannot
settle direction or practical size at the stated level. Reporting that
uncertainty can be a supported answer. "No effect" would discard compatible
positive and negative values; "2.5 points proves a meaningful benefit" would
discard the uncertainty. The interval concerns an average effect, not the range
of individual student effects.

### Evaluate the prediction promise

The student prediction comparison supports a limited promise: lower average
squared probability error than the training-prevalence rule on this internal
held-out set. Evidence integrity concerns the outcome coding and disjoint fit
and assessment records. Justification adequacy concerns predictor availability
at enrollment and an assessment population relevant to the intended use.
Inferential adequacy concerns the specified error measure and any uncertainty
or stability needed for a stronger comparative claim.

The 0.5 cutoff identifies 73 of 427 Dropout records. Claim calibration therefore
requires a distinction between improved probability predictions and useful
screening. A threshold could be changed, but deciding its value requires the
consequences of missed cases and false alerts. An appropriate future evaluation
would preserve the chosen rule and threshold before measuring performance in
the new setting. Trying thresholds on the same assessment set and reporting
only the best would spend that set on selection.

A future-cohort promise remains unassessed because the released record does not
permit a later-cohort split. A causal claim that acting on alerts prevents
dropout is also unassessed by this prediction analysis. More elaborate software
cannot replace the missing time information or evidence about intervention
consequences. The way forward is to identify the additional question and obtain
evidence suited to it, rather than make the same error score carry it.

### Evaluate the regression answer already read

| Quality-evaluation question | Application to the reported descriptive argument |
|:--|:--|
| Question alignment | Does the answer remain about the 102 occupations and the declared straight-line summary? |
| Evidence integrity | Are the source, aggregate unit, scaling, exclusions, and unusual records traceable? |
| Justification adequacy | Are the offered reasons adequate for a descriptive line, rather than stretched into random sampling or causal identification? |
| Inferential adequacy | Do the calculation and substantive reasoning support the reported pattern while retaining curvature and scatter? |
| Transition fidelity | Does the interpretation preserve occupation-level measurements, units, and exploratory status? |
| Claim calibration | Does the answer retain historical scope, supported range, and the limits of a linear summary? |

: Evaluating the regression argument already read. {#tbl-supa-regression-quality tbl-colwidths="[30,70]"}

R4 is supported as stated: it reports a checkable summary of named records and
retains their unit and historical scope. Missing information about national
sampling does not invalidate arithmetic on those named records. It would matter
for a proposed national generalization. No fit summary establishes that the
line describes every occupation closely; R3 acknowledges substantial scatter
and a curved-fit sensitivity.

A model-based extension is a separate commitment. An interval for a process
slope needs an appropriate mean model and uncertainty assumptions. The residual
pattern and curvature qualify that extension. Changing only the covariance
estimator might address a variance condition; it cannot make a straight
conditional-mean relationship true or provide random sampling.

The news claim "another year of education increases income by about 900 dollars"
is not supported by these records [@robinson1950]. It moves from occupation
averages to individuals and from description to causation. Replacing it with R4
changes the question-answer pair, not merely the tone of the conclusion.
A causal comparison needs a design or identification argument. In Project STAR,
for example, the within-school random assignment supplies a reason for causal
comparison; the fact that a binary-regression slope equals a group-mean
difference does not itself supply that reason [@finn1990; @mosteller1995].

## From a Reading to a Documented Construction Cycle

### Six stages organize the work

A finished article is organized for communication. Its polished order does not
prove when researchers made their decisions. A **construction workflow**
describes how an argument is produced, whereas the component map describes
how its support functions. A Methods paragraph may report a plan, a later
repair or a post-inspection choice; dated records help distinguish them.

The six-stage cycle has two phases. **Understand and plan** develops a question
and a coherent way to answer it. **Produce and complete** obtains the
question-specific evidence, analyzes it and completes a bounded answer.
Substantive reasons belong in both phases: measurement meaning and practical
benchmarks should not be invented after the preferred result appears.

| Phase and stage | Feedback construction example |
|:--|:--|
| Understand and plan: 1. Review the problem, setting, and relevant literature | Review what feedback activities could support learning and what earlier assessments measured |
| 2. Frame the question and, where useful, a candidate answer | Ask about the eligible-cohort average assignment effect on delayed course-topic performance |
| 3. Align design, measurement, estimand, evidence-production or acquisition plan, and inferential plan | Commit to class allocation, equal student weighting, a common delayed assessment, complete follow-up and matching uncertainty; document the two-point rationale |
| Produce and complete: 4. Produce or acquire, and document, empirical and contextual evidence | Preserve allocation, delivery, scoring, follow-up and analysis-membership records |
| 5. Estimate, check, and analyse | Estimate the contrast and interval; check coding, denominators and influential schools |
| 6. Contextualize, interpret, and complete the bounded answer | Interpret the interval against its course benchmark and retain scope and unresolved practical size |

: A six-stage construction cycle for the feedback question. {#tbl-supa-cycle-stages tbl-colwidths="[45,55]"}

```{r}
#| label: fig-supa-cycle
#| fig-cap: "The six-stage cycle has one forward handoff between planning and production. Iteration stays within each phase; changing anchored commitments starts a revised or additional cycle."
#| fig-alt: "Review, Frame and Align form the first phase, Understand and plan. Produce or acquire and document, Estimate check and analyse, and Contextualize interpret and complete form the second phase, Produce and complete. A one-way arrow links Align to Produce. Iteration is allowed within a phase; no return arrow crosses phases within this cycle."
#| fig-width: 7
#| fig-height: 4
#| fig-pos: H
diagram_start()
text(.5,.96,"Understand and plan",cex=.9)
support_row(c("1. Review","2. Frame","3. Align"),.78,"#E4EEF5")
text(.5,.48,"Produce and complete",cex=.9)
support_row(c("4. Produce/acquire\nand document","5. Estimate, check\nand analyse",
  "6. Contextualize,\ninterpret, complete"),.25,"#E8F1E6")
segments(.88,.67,.88,.57); segments(.88,.57,.12,.57)
arrows(.12,.57,.12,.36,length=.07)
text(.5,.61,"One-way handoff",cex=.75)
text(.5,.06,"Iteration within each phase; changed commitments start another cycle",cex=.69)
```

### Read the decision history, not just the final order

The following chronology is an invented extension of the feedback protocol.
It teaches how timing contributes to interpretation; it adds no new results.

| Decision time | Documented action | What the record allows a reader to conclude |
|:--|:--|:--|
| Before recruitment | The team specifies the eligible-cohort question, assessment occasion and two-point benchmark | These commitments precede relevant outcomes |
| During planning | A pilot reveals ambiguous delivery instructions; the team clarifies them and realigns the plan | A refinement within Phase 1, before production for this question |
| At production | The allocation list, class membership and scoring records are preserved | The plan can be compared with implementation |
| During analysis | A duplicated score entry is found; the original record is checked and the affected calculation corrected | A Phase 2 repair, retaining the original estimand and commitments |
| After inspecting outcomes | The team adds a question about baseline-score subgroups | An additional exploratory question and cycle, not a retrospectively planned primary question |
| At completion | The original answer retains the 2.5-point estimate and [1, 4] interval with practical size uncertain | Normal completion can preserve an unresolved practical judgment |

: A decision history distinguishes refinement, repair and a new question. {#tbl-supa-history tbl-colwidths="[19,41,40]"}

The **Phase 1 anchor** records the question and mode, estimand, design,
measurement, evidence-production or acquisition plan, inferential plan and,
where relevant, a stopping rule. The one-way handoff makes changes to those
commitments visible. It is a convention for documenting research, not a claim
that investigators can never learn from outcomes and redesign a study.

For existing data, Stage 4 can mean obtaining or designating the records for
this question, rather than the first time anyone has opened the original
file. A decision is prospective with respect to relevant evidence only if its
timing supports that claim. Calling a later reanalysis "Phase 1" does not erase
prior outcome access. Conversely, a documented revision before relevant
outcome access need not be called exploratory merely because it is a revision.

### Completion, repair and a revised question

Three endings make the record honest. **Normal completion** answers the
planned question, possibly inconclusively or against an expectation.
**Phase 2 repair** corrects work while retaining its anchored question and
commitments. An **additional or revised cycle** makes a changed question,
estimand or inferential commitment explicit and gives the earlier cycle a
visible outcome. If the intended delayed assessment was administered immediately,
the original delayed-performance question is not answerable as planned. The
immediate scores can support another question with its own scope.

Changing a score-coding error is a repair; replacing the planned outcome after
seeing a disappointing result is a revised question and must preserve that
timing. A prespecified sensitivity can be run within the planned cycle. A new
sensitivity motivated by a surprising diagnostic can also be documented as
such; it does not acquire prospective status by being sensible. If it changes
a prior inferential commitment, record a revised cycle rather than overwrite
the original.

A new cycle can reuse literature, records, measures and code. Reuse saves work,
but does not transfer a causal warrant, population scope or prospective status
automatically. For the occupation report, planning would declare historical
occupation units, a finite-record slope and checks. Completion would preserve
the records, estimate and examine the line, and report the bounded description.
Changing to an individual causal return would require a different question and
evidence, rather than a different coefficient label.

Researchers can use the six evaluation questions during construction to judge
what a stage produces. The questions do not become workflow stages. Reading
identifies support; evaluation judges its adequacy; construction documents how
support is developed. Keeping these tasks distinct makes room for judgment
without requiring a ritual sequence every time a paper is opened.

## Summary and Reading Bridge

A useful reading connects the question and reported answer to particular
information at inspectable locations. Procedures, invoked reasons, analysis
and conclusions can appear in the same paragraph while doing different work.
The substantive layer follows what those contributions mean for the study's
people, measures and setting. An inadequate offered reason is still part of
the argument to be read; its adequacy is a separate judgment.

The feedback packet showed how allocation, weighting, assessment meaning and
uncertainty support one bounded answer. The student example distinguished an
internal probability comparison from a useful alert or future-cohort promise.
The occupation report separated a finite-record description from individual
causal claims. In each case, changing one fact can change the support even
when the numerical output stays the same.

This approach is a teaching synthesis, informed by Toulmin's warrants and
qualifications [@toulmin2003], LeMire's connection between statistical testing
and substantive reasoning [@lemire2010], and Kane's argument-based account of
interpretations and uses [@kane2013]. Work on estimands [@lundberg2021] and
quantitative reporting [@appelbaum2018] develops the link between a target and
the inspectable record. These sources complement statistical method knowledge;
the component labels cannot replace that knowledge.

Argument B extends the reading to several statistical strands required for
one answer and to separate questions whose support depends on one another.
Each argument chapter supplies its own local recap and cases. Courses can
select one without requiring a running case carried through every chapter.

### Key terms {.unnumbered}

The Part One Concept Reference brings the shared terms together. This table
selects the terms developed in this argument chapter.

| Concept group | Term and common alternatives | Meaning or essential distinction | Main explanation |
|:--|:--|:--|:--|
| Argument-based reading | Argument component | One of the four functions, evidence, justification, inference, and bounded claim, that a piece of study information performs in the statistical or the substantive layer. A reader identifies the function before judging its adequacy. | Arg. A, Sec. 2 |
| Argument-based reading | Documentary location | The place in a report or its related materials, such as Methods, a protocol, a codebook, or an analytical supplement, where the traceable information for a component is found. | Arg. A, Sec. 1 |
| Argument-based reading | Evaluation question | One of the six questions on alignment, support, and translation used to judge the quality of a question-level argument after it has been read. Each receives a reasoned judgment, not a score. | Arg. A, Sec. 4 |

: Selected terms introduced in Argument Chapter A. {#tbl-supa-key-terms tbl-colwidths="[18,21,44,17]"}

## Exercises {.unnumbered}

Each prompt supplies its own record. Explain the consequential connection; a
complete component map is needed only when the question asks for one.

<!-- prompt-id: CH01-E01 -->
### Exercise 1: Find the question and the reported answer (Foundational) {.unnumbered #ex-ch01-01}

An invented abstract states: "We examined whether assigning weekly planning
messages rather than usual course messages changes the average final-assessment
score for 100 consenting students in one course. The planning group averaged
78 and the usual group 73 on a 100-point assessment. We conclude that planning
messages improve achievement for every university student." The excerpt gives
no allocation procedure or interval.

1. Formulate the research question from the aim, identifying your wording as
   a reader's formulation. Locate the numerical result and reported answer.
2. Identify the question's mode and estimand. Distinguish the five-point
   estimate from that target.
3. Name one specific piece of supporting information to seek and a plausible
   documentary location. State one way the reported answer exceeds the record.

<!-- END EX01-01 PROMPT -->

<!-- prompt-id: CH01-E02 -->
### Exercise 2: Distinguish a procedure, a reason and a result (Foundational) {.unnumbered #ex-ch01-02}

An invented report contains these passages:

- **M1:** Twelve of 24 classes were randomly allocated to each feedback format.
- **P1:** Random allocation supports comparing the assigned formats for the
  participating classes; uncertainty must reflect class assignment.
- **R1:** The estimated eligible-cohort average assignment effect is 2.5 points,
  approximate 95% CI [1, 4].
- **D1:** Assignment increased average assessed performance in these classes;
  practical magnitude remains uncertain against the two-point benchmark.

1. Identify the main statistical function of each passage and its location.
2. Explain why locating P1 in a protocol rather than the paper does not change
   its function. Is its location itself a judgment of adequacy?
3. In a variant, P1 says "Random assignment makes these classes representative
   of all schools." Identify its function first, then explain the problem with
   the offered reason.

<!-- END EX01-02 PROMPT -->

<!-- prompt-id: CH01-E03 -->
### Exercise 3: Follow an interval into an answer (Integrative) {.unnumbered #ex-ch01-03}

A separate invented workshop study randomly assigns 80 participants equally
to two feedback formats. Everyone completes the same immediate 20-point quiz.
The estimated assigned-group mean difference is 1.1 points, 95% CI [-0.2, 2.4].
The documented uncertainty procedure matches assignment. Before outcomes were
examined, a 1.5-point difference was identified as potentially important.

1. State the estimate, uncertainty and substantive benchmark, explaining what
   each contributes. Does the interval describe individual participants' gains?
2. Which directions and practical-size conclusions remain compatible with the
   interval? Write a bounded answer of no more than two sentences.
3. Explain what additional evidence would be needed to turn this answer about
   an immediate quiz into a claim about durable professional learning.

<!-- END EX01-03 PROMPT -->

<!-- prompt-id: CH01-E04 -->
### Exercise 4: Read incomplete or conflicting materials (Integrative) {.unnumbered #ex-ch01-04}

In an invented feedback report, Methods says all 480 eligible students were
included. A linked flow table says 460 supplied follow-up scores. The analysis
code is unavailable. The protocol explains class allocation and equal-size
weighting, but does not explain missing outcomes.

1. State what each source contributes and the specific unresolved difference.
   Which information would you seek next, and why?
2. Explain why the protocol is useful support even though it does not resolve
   this discrepancy. Do two references to the protocol create independent evidence?
3. Give a reasoned local evaluation and explain when "not assessable from the
   accessible record" is preferable to declaring the whole answer unsupported.

<!-- END EX01-04 PROMPT -->

<!-- prompt-id: CH01-E05 -->
### Exercise 5: Predict the consequence of a changed design (Integrative) {.unnumbered #ex-ch01-05}

A supplied invented report gives a 2.5-point mean difference and 95% interval
[1, 4] for 480 students in 24 classes, and concludes that assigning explanatory
feedback improved average delayed performance in those classes. In Version A,
classes were randomly assigned and uncertainty reflects that allocation. In
Version B, teachers chose their preferred format; the numerical table is
identical, and no adjustment or identification argument is supplied.

1. Predict whether the same causal answer remains supported in Version B.
   Explain which changed fact matters before referring to the interval.
2. Apply Justification adequacy and Claim calibration selectively. Distinguish
   a numerical difference from the support for its causal interpretation and
   its uncertainty calculation.
3. If you replace the causal answer with a description of the observed groups,
   are you only narrowing the same answer or answering a different question?
4. For Version A, change only the interval to [-1, 6]. Explain what changes in
   the conclusion about direction and the two-point importance benchmark.

<!-- END EX01-05 PROMPT -->

<!-- prompt-id: CH01-E06 -->
### Exercise 6: Read a documented construction history (Integrative) {.unnumbered #ex-ch01-06}

An invented team makes these decisions about its planned delayed-assessment
feedback study:

A. Before recruitment, a pilot reveals ambiguous instructions; the team clarifies
   delivery and realigns its plan.

B. During analysis, a verified duplicated score is corrected and the declared
   comparison is recalculated.

C. After viewing outcomes, the team chooses an immediate quiz instead of its
   planned delayed assessment because the result looks more favorable.

D. The declared analysis is completed, but its interval leaves meaningful
   benefit and harm compatible; the team reports that uncertainty.

1. Classify each decision as refinement within Phase 1, repair within Phase 2,
   a revised/additional cycle, or normal completion. Give a reason.
2. State what the record of C must preserve. Name one useful resource that can
   be reused and one status that does not transfer automatically.
3. If the delayed assessment was never administered, what should be recorded
   about the original question? Explain why an orderly final Methods section
   cannot establish that decisions were prospective.

<!-- END EX01-06 PROMPT -->

<!-- prompt-id: CH01-E07 -->
### Exercise 7: Locate the reasons, then judge the answers (Integrative) {.unnumbered #ex-ch01-07}

The report below is invented for this exercise. It is the whole
report; no assignment or sampling documentation accompanies it.

**S1.** We describe how average weekly study time relates to average
test score in 60 secondary schools, using each school's recorded
averages for one term. School-average study time ranged from 2 to 6
hours a week, and every school had both averages.

**S2.** A scatterplot of the 60 schools showed a roughly straight
upward pattern, so we summarized it with a least-squares line. We used
a regression program, so the coefficient is a causal effect.

**S3.** The fitted slope was 4 test-score points per additional hour of
school-average weekly study time.

**S4.** Across these 60 schools, schools with more average study time
had higher average scores: the least-squares line summarized the
pattern by about 4 points of fitted school-average score per
additional hour over the observed 2--6 hours.

**S5.** Each additional hour a student studies raises that student's
score by 4 points.

1. Locate the reasons the report offers and its reported answers. Name
   the function each performs and the paragraph where you find it.
2. Before judging it, say what function the sentence "We used a
   regression program, so the coefficient is a causal effect" performs.
   Then apply Justification adequacy to it.
3. Judge S4 as the answer to a descriptive question about these 60
   schools. Is it supported as stated? Does it need an interval?
4. Apply Transition fidelity and Claim calibration to S5. Is it
   supported? Explain whether replacing S5 with S4 narrows the same
   answer or answers a different question.

<!-- END EX01-07 PROMPT -->

## References {.unnumbered}

::: {#refs}
:::
