---
title: "Multiple Regression I: Conditioning, Checks, and Coordinated Questions"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.pos = "H")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 7, width = 68)
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 11))
```

\clearpage

::: {.callout-note title="Chapter overview"}
One study can owe several answers. This chapter keeps the occupation
records of Chapter 7 and asks three questions of the 98 with a recorded
type: how income varies with education across them, how it differs with
education at the same prestige, percentage of women, and type, and
whether a fuller learning rule predicts held-out occupations better than
the education-only rule. We develop what a partial slope compares and why
it can differ from the marginal slope, estimate it with a stated working
model, check support, form, influence, and precision, and calculate a
fitted profile, its two intervals, and a standardized quantity. We then
answer the prediction question with shared held-out folds, read the three
reported answers and their reasons before evaluating each and their
synthesis. By the end of the chapter
you will be able to fit and read an additive multiple regression in R,
keep changes of records, conditioning, coordinates, and uncertainty
apart, judge a held-out comparison, and evaluate coordinated answers
without merging them.
:::

A report may ask whether education and income move together, whether that
relationship remains when other occupational characteristics are held
constant, and whether those characteristics help predict income. These
questions can inform one another, but answering one does not discharge the
other two obligations.

We use a **chapter-authored retrospective teaching study**, not a published
study with a newly discovered analysis plan. Its records are the
*Prestige* data distributed with the carData package
[@carData2022; @fox2019]. The analyses and short report below are our
teaching record; the package documentation supplies the data descriptions,
not the claims written for this chapter.

A rounded coefficient equation explains a comparison; evaluate the
unrounded fitted model when checking its last digits. Each visible R
example follows from the earlier visible preparation in this chapter;
Chapter 7 supplies familiarity, not an existing R session. Chapter 5's
one-control partial correlation and Chapter 6's common-slope ANCOVA also
prepare this analysis: we now handle several controls, track the full
model's uncertainty, and distinguish a conditional comparison from an
average over named profiles.

\clearpage

## Three Questions in the Same Occupation Records {#sec-ch8-questions}

### One study can owe several answers

```{=latex}
\Needspace{20\baselineskip}
```

| Pair and role | Question | Estimand and owed answer |
|:--|:--|:--|
| **1. Contextual description** | How does income vary linearly with education across the 98 retained occupations? | The finite-record marginal least-squares slope; a descriptive answer for these records |
| **2. Primary conditional association** | How does income differ with education at the same prestige, percentage of women, and type, within support? | The linear conditional-mean coefficient $\beta_E$ under the declared working model; fitted $b_E$ is its estimate from the 98 retained records |
| **3. Secondary prediction analysis** | Does the full additive learning rule predict eligible same-source occupation records with lower expected absolute error than the education-only rule? | Difference in expected absolute error of the two learning procedures at the declared training sizes; an internal validation answer |

: Three separately owed answers in the occupation teaching study. {#tbl-ch8-pairs tbl-colwidths="[22,36,42]"}

These are three question--answer pairs, each with **one argument with linked
layers**. They are not three statistical strands of one argument: the report
owes separate conclusions about marginal association, conditional
association, and prediction. If several estimands were jointly necessary
for *one* integrated conclusion, they could instead be parallel statistical
strands. Merely printing several coefficients or diagnostics does not create
either several questions or several strands.

### Keep the records, time, and units visible

A row represents an **occupation**, not an individual worker. Income is the
occupation's average annual income in 1971 Canadian dollars; we divide it
by 1,000. Education is average years of education, and *women* is the
percentage of women in the occupation. Prestige ratings were collected in
a mid-1960s survey, whereas the income and education summaries come from
1971. Occupational type is blue-collar (*bc*), professional/managerial/
technical (*prof*), or white-collar (*wc*). These historical and aggregate
measurements are not interchangeable with current individual earnings.

The earlier simple-regression analysis used all 102 occupations with
education and income recorded. The declared full variable set leaves 98
complete records: athletes, newsboys, babysitters, and farmers have missing
type. The retained type counts are 44, 31, and 23 respectively. We keep the
original row order, which will also make the validation split reproducible.

```{=latex}
\Needspace{20\baselineskip}
```

```{r}
#| label: ch8-records
#| echo: true
all_occ <- carData::Prestige
all_occ$occupation <- rownames(all_occ)
all_occ$income_k <- all_occ$income / 1000
all_occ$type <- factor(all_occ$type, levels = c("bc", "prof", "wc"))
vars <- c("income_k", "education", "prestige", "women", "type")
d <- all_occ[complete.cases(all_occ[, vars]), ]
c(original = nrow(all_occ), retained = nrow(d))
table(d$type)
```

We refit the education-only model on these **same 98 records** before
comparing adjustment specifications. Moving from 102 to 98 changes which
records are described. It must not be presented as an adjustment effect.
Complete-case selection makes the computations comparable; it does not
establish that the omitted occupations are irrelevant or that these records
are a probability sample of Canadian occupations.

### Declare the model question before reporting an interval

Pair 1 is finite-record description. Its slope is exactly calculable once
the records and least-squares rule are specified.

Pair 2 has one intended estimand: the education coefficient $\beta_E$ in the
declared linear conditional-mean working model. The fitted coefficient $b_E$
is its estimate from the 98 retained records. Although $b_E$ is exactly
calculable once the rows and fitting rule are fixed, that fact does not create
a second finite-record estimand. A confidence interval additionally posits
that occupation outcomes could vary independently around the stated linear
conditional mean at the recorded predictors. This is not a claim that the
dataset came from a verified random sample or that income was randomly
assigned.

For the classical interval we additionally assume a common error variance;
normal errors make its finite-sample $t$ reference exact. Robust uncertainty
relaxes the common-variance requirement under its own large-sample
conditions. It does not create independence or linearity. We will check how
credible and stable the working-model reading is before judging its answer.

A **best linear projection** can instead summarize a curved conditional
relationship under a specified distribution of predictors. That is a
different interpretation, not a rescue silently adopted after discovering
curvature. The core uses the stated linear conditional-mean interpretation;
Appendix B explains the alternative and its limitations.

Pair 3 will declare a different hypothetical repetition: train a rule on
eligible, exchangeable same-source records, then predict a new record
without using its outcome in training. Its justification cannot be supplied
by the coefficient interval in Pair 2.

## From Marginal to Conditional Comparisons {#sec-ch8-comparisons}

### What a partial slope compares

Chapter 6's adjusted group coefficients compared fitted outcomes at the
same baseline value. The idea extends to a continuous predictor and
several controls: specify what changes, hold the other predictor values
fixed, and subtract the two fitted means.

Let $e$ be education, $p$ prestige, $w$ percentage of women, and $t$ type.
The fitted additive equation is

$$
\widehat m(e,p,w,t)
 =b_0+b_Ee+b_Pp+b_Ww
   +b_{\mathrm{prof}}D_{\mathrm{prof}}
   +b_{\mathrm{wc}}D_{\mathrm{wc}}.
$$

An indicator $D$ is 1 for its named category and 0 otherwise. Blue-collar is
the reference type, with both indicators zero. This model has six
coefficients, including its intercept.

Take two profiles with the same $p,w,t$ and education differing by one year.
Subtract their fitted means:

$$
\begin{aligned}
&\widehat m(e+1,p,w,t)-\widehat m(e,p,w,t)\\
&\qquad=b_E(e+1)-b_Ee=b_E.
\end{aligned}
$$

All the common terms cancel. Thus $b_E$ compares modeled occupation-level
income at the **same prestige, percentage of women, and type**. A two-year
difference corresponds to $2b_E$ under this additive model.

“Holding constant” describes this mathematical comparison. It does not say
that education was changed experimentally, that two observed occupations
match exactly on every other characteristic, or that unmeasured differences
are controlled. An additive slope also assumes the same education
comparison across the included predictor values. Allowing it to vary with
another predictor is the interaction question taken up later.

### Why the marginal slope can differ

A short known-process example makes the distinction visible. Suppose

$$
Y=10+2X+5Z+\varepsilon,
\qquad E(\varepsilon\mid X,Z)=0,
$$

with finite variances and
$\operatorname{Cov}(X,Z)/\operatorname{Var}(X)=0.65$.
The conditional $X$ coefficient is 2. The marginal linear slope is

$$
\frac{\operatorname{Cov}(X,Y)}{\operatorname{Var}(X)}
=2+5\frac{\operatorname{Cov}(X,Z)}{\operatorname{Var}(X)}
=2+5(0.65)=5.2500.
$$

The marginal comparison incorporates how $Z$ moves with $X$ in this
specified process. The conditional comparison holds $Z$ fixed. Neither
arithmetic result alone says what would happen if a researcher intervened
on $X$. Calling 5.2500 “biased” requires naming an estimand for which it is
being used incorrectly; as a marginal linear summary, 5.2500 is the right
answer here. Appendix A retains the derivation and simulation.

More generally, adding predictors can change the comparison, but it does
not **always** change the scientific estimand. In a randomized study,
appropriately specified covariate adjustment may estimate the same declared
average assignment effect more precisely. Releveling a factor changes
coefficient coordinates without changing fitted values. A different
uncertainty estimator may change the interval without changing the fitted
coefficient.

### A small indicator bridge

The full occupation fit, displayed to four decimal places, is

$$
\begin{aligned}
\widehat{\mathrm{income}}_k
={}&0.4205+0.1164e+0.1406p-0.0535w\\
 &+0.3242D_{\mathrm{prof}}+0.2353D_{\mathrm{wc}}.
\end{aligned}
$$

Consider a white-collar profile with education 11, prestige 45, and women 50.
Here $D_{\mathrm{prof}}=0$ and $D_{\mathrm{wc}}=1$. Using the full fitted
coefficients in the substitution gives

$$
\begin{aligned}
\widehat m_*
={}&\hat\beta_0+11\hat\beta_E+45\hat\beta_P
+50\hat\beta_W+\hat\delta_{\mathrm{wc}}\\
\approx{}&5.5886.
\end{aligned}
$$

That is a fitted mean of about 5,589 dollars on the historical income scale,
not a guarantee for any occupation or worker. Can these three predictor
values occur together within the observed white-collar covariate cloud?
Separate minimum and maximum values cannot answer that question.

Here is a checkable **joint-support witness**. A convex combination is a
weighted average with nonnegative weights summing to one. Four retained
white-collar records can be combined to recover education 11, prestige 45,
and women 50 simultaneously. The source spelling *receptionsts* is retained
in the code so that the record can be located exactly.

Use the supplied weights to check the three weighted averages. The order of
the weights matches the four named records. `weighted.mean()` performs the
same weighted-average calculation for each variable; finding these weights
is not a prerequisite for interpreting their evidence.

\Needspace{23\baselineskip}

```{r}
#| label: ch8-profile-support
#| echo: true
support_names <- c("radio.tv.announcers", "tellers.cashiers",
                   "receptionsts", "sales.supervisors")
support_rows <- d[support_names, c("education", "prestige", "women")]
support_weights <- c(0.245243736035, 0.230972611567,
                     0.226143656937, 0.297639995461)
d[support_names, "type"]
sum(support_weights)
c(education = weighted.mean(support_rows$education, support_weights),
  prestige = weighted.mean(support_rows$prestige, support_weights),
  women = weighted.mean(support_rows$women, support_weights))
```

\Needspace{11\baselineskip}

```{r}
#| label: tbl-ch8-profile-support
#| echo: false
#| tbl-cap: "Four observed white-collar records recover the declared profile."
support_display <- data.frame(
  Occupation = c("Radio/TV announcers", "Tellers/cashiers",
                 "Receptionists", "Sales supervisors"),
  support_rows,
  Weight = sub("^0", "", sprintf("%.4f", unname(support_weights))),
  row.names = NULL
)
knitr::kable(support_display,
  col.names = c("Occupation", "Education", "Prestige", "Women", "Weight"),
  digits = c(0, 2, 1, 2, 4), align = c("l", "r", "r", "r", "r"))
```

The supplied weights are all positive and sum to one; their weighted
coordinates recover $(11,45,50)$ within $10^{-10}$ using the supplied reproduction constants in the R code. The table
rounds weights for reading. This establishes membership in the joint
covariate hull **within white-collar type**, using all three continuous
predictors in the full model. It does not show that many occupations lie
near this profile, that observed occupations match exactly, or that a
causal comparison is supported. Local data may still be sparse.

The intercept refers to zero education, prestige, and women in the
reference type. It helps construct the fitted surface; it need not describe
a plausible occupation. Likewise, one type coefficient is not “the effect
of occupational type” as a whole. Appendix A retains the group-mean and releveling calculation; Chapter 9
develops contrasts, block tests, and the analysis-of-variance identity.

## Estimate the Conditional Association {#sec-ch8-estimate}

### Compare specifications on the same rows

```{r}
#| label: ch8-fits
#| echo: true
m0 <- lm(income_k ~ education, data = d)
m1 <- lm(income_k ~ education + prestige, data = d)
m2 <- lm(income_k ~ education + prestige + women + type, data = d)
coef(m2)
```

| Records and specification | Education slope | In-sample $R^2$ | Role here |
|:--|--:|--:|:--|
| Earlier 102, education only | 0.8988 | .3336 | Earlier result; not the adjustment baseline |
| Same 98, education only | 0.8830 | .3296 | Pair 1's marginal description |
| Same 98, education and prestige | $-0.2179$ | .4996 | Intermediate comparison |
| Same 98, full declared predictors | 0.1164 | .6362 | Pair 2's estimate of $\beta_E$ |

: The model sequence separates sample composition from adjustment. {#tbl-ch8-sequence tbl-colwidths="[36,18,16,30]"}

*Note.* Slopes are thousand 1971 Canadian dollars per education year.
These are fitted summaries, not causal effects. No sampling interval is
attached to Pair 1's finite-record estimand.

Education and prestige are highly correlated here, $r=.8665$. Once prestige
is included, the education comparison uses variation not linearly accounted
for by prestige. Adding women and type changes the conditioning set again.
The movement from 0.8830 to $-0.2179$ to 0.1164 is not a narrative in
which a true effect is successively “explained away.” It is a sequence
of named comparisons in a small observational record.

The increase in $R^2$ describes training-data fit. Adding predictors cannot
increase the least-squares residual sum of squares on the same rows when
the models are nested. Better training fit does not establish better
prediction on records left out of fitting.

### Partial regression gives the same coefficient {#ch08-partial-regression}

Multiple regression still uses Chapter 7's covariance-to-slope reasoning.
Chapter 5 removed one control's fitted linear component from each of two
variables before correlating their residuals. Here we remove several
controls. First regress education on prestige, percentage of women, and
type, retaining its residuals $r_{E,i}$. Do the same for income, obtaining
$r_{Y,i}$. Both auxiliary regressions include an intercept and use exactly
the same 98 rows as the full fit. Then

$$
b_E=\frac{\sum_i r_{E,i}r_{Y,i}}{\sum_i r_{E,i}^2}
    =r_{EY\cdot Z}\frac{s_{r_Y}}{s_{r_E}}.
$$

Here $Z$ denotes the control columns and $r_{EY\cdot Z}$ is the correlation
of the two residual vectors. The numerator measures remaining linear
co-variation; the denominator measures remaining education variation.
This partial-regression identity gives the full model's education
coefficient. Its correlation is unitless; multiplying by the residual-SD
ratio expresses the slope in thousand dollars per education year.

Uncertainty must also account for the fitted controls. Prestige and women
supply one column each, while type supplies two indicator columns. With
education and the intercept, the full model therefore estimates $p=6$
coefficients from $n=98$ rows. Its residual degrees of freedom are
$n-p=92$. Using its residual sum of squares, $RSS_{\mathrm{full}}$, gives

$$
SE(b_E)=\sqrt{\frac{RSS_{\mathrm{full}}/(n-p)}
                         {\sum_i r_{E,i}^2}},
\qquad
t_E=r_{EY\cdot Z}\sqrt{\frac{n-p}{1-r_{EY\cdot Z}^2}}.
$$

The slope identity is algebraic. The displayed SE estimates uncertainty
under the independent, constant-variance error model; Normal errors give
the exact $t_{92}$ reference for $H_0:\beta_E=0$. The following calculation
keeps those full-model degrees of freedom.

\Needspace{15\baselineskip}

```{r}
#| label: ch8-partial
#| echo: true
r_E <- residuals(lm(education ~ prestige + women + type, data = d))
r_Y <- residuals(lm(income_k ~ prestige + women + type, data = d))
partial_r <- cor(r_E, r_Y)
full_df <- df.residual(m2)
partial_se <- sqrt((sum(residuals(m2)^2) / full_df) / sum(r_E^2))
partial_t <- partial_r * sqrt(full_df / (1 - partial_r^2))
c(slope = sum(r_E * r_Y) / sum(r_E^2), partial_r = partial_r,
  SE = partial_se, df = full_df, t = partial_t,
  p_value = 2 * pt(-abs(partial_t), full_df))
```

The slope is 0.1164, partial correlation .0441, SE 0.2752, $t=0.4230$,
and two-sided $p=.6732$. These reproduce the education row of the full
model. A fresh `lm(r_Y ~ r_E)` or ordinary `cor.test(r_E, r_Y)` instead
uses 96 df: its default treats the residuals as if fitting the controls
had used no coefficients. Use the original `m2` for uncertainty, or the
explicit full-model calculation above. The HC3 calculation in the next
subsection retains the slope and uses a different covariance estimate.

This distinction also helps separate two meanings of covariance. Chapter
5's `cov()` describes the spread and co-variation of recorded variables.
The regression function `vcov()` estimates the covariance of the fitted
coefficients. Compare them on the same 98 rows:

\Needspace{14\baselineskip}

```{r}
#| label: ch8-covariance-objects
#| echo: true
data_cov <- cov(d[c("income_k", "education", "prestige", "women")])
coefficient_cov <- vcov(m2)
dim(data_cov)
dim(coefficient_cov)
c(education_data_variance = data_cov["education", "education"],
  education_coefficient_variance =
    coefficient_cov["education", "education"],
  education_SE = sqrt(coefficient_cov["education", "education"]))
```

The recorded-data matrix is 4 by 4; this numeric summary excludes the type
factor. Its diagonal entries are sample variances and its off-diagonal
entries are sample covariances. Education variance is 7.5567 years squared.
The coefficient matrix is 6 by 6, including the intercept and two type
coefficients. Its diagonal entries are squared coefficient SEs and its
off-diagonal entries describe joint coefficient uncertainty. The education
coefficient's variance is 0.0757; its square root is SE 0.2752 thousand
dollars per education year. These matrices answer different questions and
are not interchangeable in a fitted-mean or contrast interval. Section 5
uses the coefficient covariance in $x_*'\widehat Vx_*$.

```{r}
#| label: fig-ch8-partial
#| fig-pos: H
#| fig-cap: "Marginal and residualized income--education comparisons in the same 98 occupations."
#| fig-alt: "The marginal scatterplot has a positive fitted line. After linearly removing prestige, women, and type from both variables, the fitted line is nearly flat and substantial residual variation remains."
#| fig-width: 6.4
#| fig-height: 3.5
oldpar <- par(mfrow = c(1, 2), mar = c(4.3, 4.2, 1, .5), mgp = c(2.7, .8, 0))
plot(d$education, d$income_k, pch = 1, xlab = "Education (years)",
     ylab = "Income (thousand dollars)")
abline(m0, col = "#245C78", lwd = 2)
plot(r_E, r_Y, pch = 1, xlab = "Residual education (years)",
     ylab = "Residual income (thousand dollars)")
abline(a = 0, b = coef(m2)["education"], col = "#245C78", lwd = 2)
par(oldpar)
```

Residualization does not remove all substantive differences between
occupations. It removes the **fitted linear components** of the named
variables. Chapter 9 develops the matrix version.

### Read a coefficient interval under its stated model

The hypothetical linear conditional-mean model introduced in
@sec-ch8-questions
has six estimated coefficients, so its residual degrees of freedom are
$98-6=92$. For education the classical calculation is

$$
\begin{aligned}
b_E\pm t_{.975,92}SE(b_E)
 &=0.1164\pm1.9861(0.2752)\\
 &\approx[-0.4302,\ 0.6631].
\end{aligned}
$$

The result is compatible, under those assumptions, with negative and
positive conditional slopes on this scale. It does not demonstrate that
education is unimportant, that the slope equals zero, or that adjustment
has removed an education effect. No substantive equivalence margin was
declared, and the interval includes differences of several hundred dollars
per year of education. The data and model therefore leave the direction
and practically relevant magnitude unresolved.

```{r}
#| label: ch8-uncertainty
#| echo: true
confint(m2, "education")
hc3_covariance <- sandwich::vcovHC(m2, type = "HC3")
hc3_se <- sqrt(hc3_covariance["education", "education"])
coef(m2)["education"] + c(-1, 1) * qt(.975, 92) * hc3_se
```

HC3 gives SE 0.3823 and an approximate interval $[-0.6429,0.8758]$ using
the displayed $t_{92}$ convention. It allows heteroskedasticity under
independent-observation and other regularity conditions; it is not an exact
small-sample guarantee. Both intervals leave broad uncertainty. Neither
accounts for unmodeled dependence, selection, measurement error, or a wrong
conditional-mean form.

If a test is reported, its target belongs with this inference: the
two-sided education-coefficient test addresses $H_0:\beta_E=0$ in the stated
full model, not “education has no effect.” Its classical $t=0.4230$ and
$p=.6732$ do not prove the null. We do not use this test to choose which of
the three questions deserves an answer.

\Needspace{12\baselineskip}

## Check Support, Form, Influence, and Precision {#sec-ch8-check}

Checking asks what the model leaves unexplained and what follows for the
answer. It is not a ritual in which each graph receives a pass or fail.
Residuals, support, coefficient precision, and sensitivity each answer a
different diagnostic question.

### Read residual, scale, tail, and influence views together

A residual is an observed deviation from a fitted value, not the unobserved
population error. A standardized residual uses a common residual-scale
estimate; an externally studentized residual uses a scale estimate with that
observation omitted. Leverage measures an unusual predictor configuration.
Influence combines predictor and outcome information [@cook1977]. The
first extraction collects these quantities for every retained occupation
and lists the five with the largest Cook's distance.

```{r}
#| label: ch8-diagnostics
#| echo: true
ch8_diag <- data.frame(
  occupation = rownames(d), fitted = fitted(m2),
  residual = residuals(m2), standardized = rstandard(m2),
  studentized = rstudent(m2), leverage = hatvalues(m2),
  cooks = cooks.distance(m2)
)
ch8_top <- order(ch8_diag$cooks, decreasing = TRUE)[1:5]
ch8_diag[ch8_top, ]
```

```{r}
#| label: fig-ch8-diagnostics
#| echo: false
#| fig-cap: "Four complementary diagnostic views for the full additive model."
#| fig-alt: >
#|   Residual and scale plots show structure and changing spread.
#|   A normal quantile plot departs in the right tail. The final
#|   panel relates leverage and studentized residuals, with point
#|   sizes indicating Cook's distance.
#| fig-width: 6.7
#| fig-height: 6.2
local({
  old <- par(no.readonly = TRUE)
  on.exit(par(old))
  par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
  z <- ch8_diag
  plot(z$fitted, z$residual, pch = 1,
       xlab = "Fitted income", ylab = "Residual",
       main = "(a) Mean pattern")
  abline(h = 0, lty = 2)
  lines(lowess(z$fitted, z$residual), lwd = 2)
  plot(z$fitted, sqrt(abs(z$standardized)), pch = 1,
       xlab = "Fitted income", ylab = "Sqrt |standardized|",
       main = "(b) Scale pattern")
  lines(lowess(z$fitted, sqrt(abs(z$standardized))), lwd = 2)
  qqnorm(z$studentized, pch = 1, main = "(c) Normal reference")
  qqline(z$studentized, lwd = 2)
  plot(z$leverage, z$studentized, pch = 1,
       cex = .8 + 2 * sqrt(z$cooks / max(z$cooks)),
       xlab = "Leverage", ylab = "Studentized residual",
       main = "(d) Influence")
  abline(h = 0, lty = 2)
})
```

The smooth and changing spread challenge a constant conditional-mean and
variance description. The right tail matters to exact small-sample
normal-theory inference. Neither feature prevents calculation of OLS. Large
residual, high leverage, and high influence are different features; none is
an automatic record-invalidity verdict. No residual graph can establish that
occupations are independent draws from a wider population.

### Inspect joint support and coefficient precision

Joint support concerns predictor combinations, not merely separate minimum
and maximum values. The convex-combination witness in
@sec-ch8-comparisons establishes support for the focal white-collar
profile across all three continuous predictors. The marginal ranges below ask
a different, coarser question: how far does each variable extend within each
recorded type?

```{=latex}
\Needspace{8\baselineskip}
```

```{r}
#| label: tbl-ch8-support
#| echo: false
#| tbl-cap: "Predictor ranges within each recorded occupation type."
ch8_ranges <- do.call(rbind, lapply(split(d, d$type), function(g) {
  data.frame(n = nrow(g),
    Education = sprintf("%.2f--%.2f",
                        min(g$education), max(g$education)),
    Prestige = sprintf("%.1f--%.1f",
                       min(g$prestige), max(g$prestige)),
    Women = sprintf("%.2f--%.2f", min(g$women), max(g$women)))
}))
knitr::kable(cbind(Type = rownames(ch8_ranges), ch8_ranges),
  row.names = FALSE)
```

Blue-collar education ends at 10.93 years, below the professional minimum
of 11.09. A same-education comparison between those types necessarily crosses
that gap. A model can return both predictions there; software output does not
create observations. Lack of support limits what can be learned or interpreted
about a defined comparison rather than making the estimand cease to exist.

The Computer Lab reference on conditional comparisons supplies the grid and
plotting code used here. A conditional grid can also avoid pretending that an unrestricted rectangle
is supported. For *m1*, hold prestige at 30, 50, or 70 and plot education only
over values observed within 7.5 prestige points of each target. The width is
an explicit display convention, not a formal support guarantee.

```{r}
#| label: ch8-grid
#| echo: false
#| results: hide
ch8_grid <- data.frame()
for (p0 in c(30, 50, 70)) {
  nearby <- d[abs(d$prestige - p0) <= 7.5, ]
  one_grid <- data.frame(
    education = seq(min(nearby$education), max(nearby$education),
                    length.out = 40),
    prestige = p0, Profile = paste("Prestige", p0)
  )
  ch8_grid <- rbind(ch8_grid, one_grid)
}
ch8_grid$fit <- predict(m1, newdata = ch8_grid)
```

```{r}
#| label: fig-ch8-grid
#| echo: false
#| fig-cap: "Neighborhood-restricted fitted profiles from the education-and-prestige model."
#| fig-alt: >
#|   Three gently decreasing parallel lines represent fitted income
#|   across education at prestige 30, 50, and 70. Their education
#|   ranges differ, reflecting the declared neighborhoods.
#| fig-width: 6.1
#| fig-height: 3.5
ggplot2::ggplot(ch8_grid,
  ggplot2::aes(education, fit, linetype = Profile)) +
  ggplot2::geom_line(linewidth = .8) +
  ggplot2::labs(x = "Education (years)",
    y = "Fitted income (thousand 1971 dollars)") +
  ggplot2::theme(legend.position = "bottom")
```

The parallel lines share *m1*'s education slope. This display conditions on
prestige only, not on the full set for Pair 2. It teaches a transparent plotting
choice; it does not prove support for the full-model profile.

Education's variance inflation factor makes the collinearity calculation
checkable. Regress education on the remaining predictors; the auxiliary
$R^2$ measures how much of its variation they account for:

```{r}
#| label: ch8-vif-first
#| echo: true
education_auxiliary <- lm(education ~ prestige + women + type, data = d)
education_r2 <- summary(education_auxiliary)$r.squared
education_vif <- 1 / (1 - education_r2)
c(auxiliary_R2 = education_r2, VIF = education_vif,
  SE_multiplier = sqrt(education_vif))
```

$$
VIF_E=\frac{1}{1-R^2_{E\mid P,W,T}}\approx8.0968,
\qquad \sqrt{VIF_E}\approx2.8455.
$$

Under the classical model, holding residual variance and total education
variation fixed, its coefficient SE is multiplied by
$\sqrt{8.0968}\approx2.8455$ relative to an orthogonal design. This is a
comparison of variance formulas, not a claim that deleting other predictors
will divide the actual SE by 2.8455. Deletion changes the comparison and can
change residual variance. VIF is not an automatic deletion threshold, and
near collinearity does not by itself create bias under the maintained
zero-conditional-mean model.

### Use targeted sensitivities to bound the answer

A component-plus-residual plot adds the fitted prestige contribution back to
each residual, $e_i+\hat\beta_P P_i$. The smooth bends upward at high prestige
relative to the straight fitted component. That pattern motivates a declared
form sensitivity; it does not prove that one chosen curve is the true process.

```{r}
#| label: fig-ch8-partial-residual
#| echo: false
#| fig-cap: "The prestige component plus residual in the full additive model."
#| fig-alt: >
#|   A component-plus-residual scatterplot has a positive trend.
#|   A curved smooth bends upward at high prestige relative to the
#|   straight fitted component.
#| fig-width: 6.1
#| fig-height: 3.5
ch8_partial <- data.frame(
  prestige = d$prestige,
  component = residuals(m2) + coef(m2)["prestige"] * d$prestige
)
ggplot2::ggplot(ch8_partial, ggplot2::aes(prestige, component)) +
  ggplot2::geom_point(shape = 1) +
  ggplot2::geom_smooth(method = "lm", se = FALSE,
                      colour = "black", linetype = 2) +
  ggplot2::geom_smooth(method = "loess", se = FALSE,
                      colour = "black") +
  ggplot2::labs(x = "Prestige", y = "Component plus residual")
```

\Needspace{26\baselineskip}

```{r}
#| label: ch8-form-sensitivity
#| echo: true
ch8_curve <- update(m2, . ~ . + I(prestige^2))
anova(m2, ch8_curve)
c(full_education = unname(coef(m2)["education"]),
  curve_education = unname(coef(ch8_curve)["education"]))
```

The classical nested comparison gives $F(1,91)=6.7360$ and $p=.0110$;
education changes from 0.1164 to $-0.0298$. In the enlarged equation,
the prestige slope depends on prestige through
$\beta_P+2\beta_{P^2}P$. This is a focused sensitivity for Pair 2, not a
validation-tuned replacement for the frozen prediction rule in Pair 3.

The record for general managers has the largest Cook's distance, about 0.4467. Its high
income relative to the model is not proof of error. A transparent deletion
refit asks how dependent the answer is on that record:

```{r}
#| label: ch8-influence-sensitivity
#| echo: true
ch8_without_top <- lm(
  income_k ~ education + prestige + women + type,
  data = d[-ch8_top[1], ]
)
c(full_education = unname(coef(m2)["education"]),
  omit_one_education = unname(coef(ch8_without_top)["education"]))
```

The education estimate moves to 0.4372. Preserve the full-data estimate and
disclose the sensitivity. Verify coding, measurement, and substantive
distinctiveness before considering exclusion; a more attractive $p$ value is
not an exclusion rule. The refit is a diagnostic perturbation, not a new
research question or evidence that the reduced sample is preferable.

```{=latex}
\Needspace{42\baselineskip}
```

| Check | What is visible here | Consequence for the answer |
|:--|:--|:--|
| Functional form | Adding a prestige-squared term changes the education slope to about $-0.0298$ | Treat the additive form as a working approximation; do not silently reinterpret it as an exact conditional law |
| Spread | HC3 raises the education SE from 0.2752 to 0.3823 | Show sensitivity; a robust SE does not repair all model problems |
| Influence | Removing the record for general managers, which has the largest Cook's distance, gives an education slope about 0.4372 | Inspect the record and disclose sensitivity; do not delete it merely to obtain a preferred result |
| Collinearity | Education's auxiliary $R^2$ is .8765 | Little education variation remains after the other predictors; separate conditional coefficients can be imprecise |
| Joint support | Professional and blue-collar education ranges do not overlap | Cross-type comparisons at common covariates may depend heavily on extrapolation |
| Dependence and scope | Occupational aggregates have shared economic and classification contexts | Independence and same-source transport are assumptions, not verified by a residual plot |

: Checks change the bounded answer rather than producing one omnibus pass/fail score. {#tbl-ch8-checks tbl-colwidths="[19,38,43]"}

Appendix B develops Cook's distance, coefficient covariance, generalized
VIF, changed-outcome interpretation, and model-comparison criteria. The
Computer Lab's conditional-comparisons reference supplies the longer diagnostic and
matrix reproductions. The first calculations needed to interpret the answer
are already in this section.

## Move from Coefficients to Fitted and Standardized Quantities {#sec-ch8-quantities}

The profile and standardization calculations in this section are **didactic
extensions of the fitted model**, not extra question-facing answers owed by
the occupation study. Pair 2 remains about $\beta_E$. A report that promoted
one of these quantities to its own substantive conclusion would need to state
the corresponding question, estimand, justification, and bounds.

### A fitted mean is not prediction performance

Start with a complete input row and ask R for the two interval types.
`interval = "confidence"` concerns the model mean at that profile;
`interval = "prediction"` concerns one new outcome at the same profile.
Both use the full fitted model and its stated classical assumptions.

\Needspace{17\baselineskip}

```{r}
#| label: ch8-profile-predict-check
#| echo: true
profile <- data.frame(education = 11, prestige = 45, women = 50,
                     type = factor("wc", levels = levels(d$type)))
mean_check <- predict(m2, newdata = profile, interval = "confidence")
new_check <- predict(m2, newdata = profile, interval = "prediction")
mean_check
new_check
```

For the white-collar profile in @sec-ch8-comparisons, collect the
coefficient multipliers in $x_*=(1,11,45,50,0,1)'$. Their order is intercept,
education, prestige, women, professional indicator, and white-collar indicator.
The fitted mean and its estimated classical variance are below. A prime denotes
transpose (`t()` in R). Recall the two covariance objects in Section 3.2:
`vcov(m2)` is the estimated covariance matrix of the six fitted
coefficients. Its variances and covariances combine the uncertainty in the
terms used to calculate this fitted mean.

$$
\widehat m_*=x_*'\hat\beta,\qquad
\widehat{\operatorname{Var}}(\widehat m_*)=x_*'\widehat Vx_*,
\quad \widehat V=\operatorname{vcov}(m2).
$$

The variance includes the coefficient variances **and their covariances**.
For example, two terms $a\hat\beta_j+b\hat\beta_k$ contribute
$a^2\widehat{\operatorname{Var}}(\hat\beta_j)+
b^2\widehat{\operatorname{Var}}(\hat\beta_k)+
2ab\widehat{\operatorname{Cov}}(\hat\beta_j,\hat\beta_k)$.
Adding coefficient SEs, or ignoring covariance, is not the same calculation.
Take the square root only after calculating the variance.

```{=latex}
\Needspace{19\baselineskip}
```

```{r}
#| label: ch8-profile-covariance
#| echo: true
x_star <- c(1, 11, 45, 50, 0, 1)
V <- vcov(m2)
profile_fit <- sum(x_star * coef(m2))
profile_var <- drop(t(x_star) %*% V %*% x_star)
s2 <- sum(residuals(m2)^2) / df.residual(m2)
critical <- qt(.975, df.residual(m2))
profile_mean_ci <- profile_fit +
  c(-1, 1) * critical * sqrt(profile_var)
profile_new_pi <- profile_fit +
  c(-1, 1) * critical * sqrt(s2 + profile_var)
rbind(fitted_mean = profile_mean_ci, new_case = profile_new_pi)
```

The fitted mean is 5.5886; its estimated variance is 0.3080 and
SE is $\sqrt{\widehat{\operatorname{Var}}(\widehat m_*)}\approx0.5549$, in thousand-dollar units for the SE.
Variance has squared units. The classical fitted-mean interval is therefore
$[4.4865,6.6908]$ thousand dollars.

For **one new outcome**, its own error variation adds $s^2=6.8575$:
the SE is $\sqrt{s^2+x_*'\widehat Vx_*}$, giving the much wider interval
$[0.2722,10.9051]$. This calculation uses the independent normal
common-variance conditional-mean model; it is not a promise established by
the diagnostics. Replacing $\widehat V$ with HC3 alone does not estimate an
arbitrary new outcome's conditional error distribution.

Neither interval tells us how accurately a learning rule predicts held-out
records on average. That requires Pair 3.

### Standardize over a named distribution

A conditional fitted value uses one specified profile. A standardized fitted
value averages across a named distribution. To see the operation, copy the
98 observed profiles, change their type to white-collar, predict each
fitted value, then take the mean. Each continuous profile has equal weight.

\Needspace{11\baselineskip}

```{r}
#| label: ch8-standardization-first
#| echo: true
white_profiles <- d
white_profiles$type <- factor("wc", levels = levels(d$type))
white_predictions <- predict(m2, newdata = white_profiles)
mean(white_predictions)
```

This is an average of model predictions after a specified profile change,
not the observed white-collar group mean. For type $g$ and fixed
nonnegative weights $w_i$ summing to one,

$$
\hat\mu_g^{\mathrm{std}}=\sum_i w_i x_{ig}'\hat\beta
=a_g'\hat\beta,\qquad a_g=\sum_i w_i x_{ig}.
$$

Here $x_{ig}$ is the design vector after assigning type $g$ to observed
continuous profile $i$. With empirical weights $w_i=1/98$, this is
standardization over these 98 profiles, not over a national population.
Treating those profiles and weights as fixed gives estimated variance
$a_g'\widehat V a_g$.

For the white-collar quantity, `model.matrix()` gives the six coefficient
multipliers for every changed profile. `colMeans()` averages each multiplier
to obtain $a_{\mathrm{wc}}$. Its uncertainty is a direct application of
the coefficient covariance calculation from Section 5.1:

```{r}
#| label: ch8-standardization-covariance
#| echo: true
white_X <- model.matrix(~ education + prestige + women + type,
  data = white_profiles)
white_a <- colMeans(white_X)
white_fit <- sum(white_a * coef(m2))
white_var <- drop(t(white_a) %*% vcov(m2) %*% white_a)
white_se <- sqrt(white_var)
c(estimate = white_fit, SE = white_se)
white_fit + c(-1, 1) * qt(.975, df.residual(m2)) * white_se
```

The result is 7.0164 thousand dollars, with SE 0.6177 and interval
[5.7896, 8.2432]. The variance includes the covariance between predictions
through their shared fitted coefficients. Averaging individual interval
endpoints would give a different calculation. Repeating this operation for
each type produces the following table; the Computer Lab reference on common-profile
averaging supplies the full loop and its output.

| Type | Estimate | SE | Lower | Upper |
|:--|--:|--:|--:|--:|
| bc | 6.7811 | 0.6942 | 5.4023 | 8.1598 |
| prof | 7.1053 | 0.9011 | 5.3156 | 8.8950 |
| wc | 7.0164 | 0.6177 | 5.7896 | 8.2432 |

: Empirical-profile standardization under the full additive model. {#tbl-ch8-standardized tbl-colwidths="[20,20,20,20,20]"}

Because *m2* is linear and additive in the continuous predictors, its average
fitted value equals its fitted value at their means. This is an algebraic
special case: with squared terms, interactions, or a nonlinear response
function, prediction at means need not equal mean prediction. Constant
additive type contrasts remain unchanged by common weights, although absolute
standardized levels can change.

The overall education mean is 10.7951 years. Relabelling every profile
professional or white-collar creates some unsupported combinations. Averaging
them does not solve that problem. The white-collar profile used above is a
different, locally supported conditional quantity; it is not a three-type
comparison. The intervals condition on the empirical profiles and use the
classical covariance. Estimating a new population's covariate distribution
would add uncertainty not included here.

Standardization does not identify a causal effect by itself and does not
validate out-of-sample performance. Pair 3 requires a learning rule and
held-out outcomes.

## Ask and Answer the Separate Prediction Question {#sec-ch8-prediction}

### Define the learning rule, new case, and loss

A learning rule says how training data become a fitted prediction equation.
Our two rules are fixed: fit education-only OLS, or fit the full additive
OLS equation with education, prestige, women, and type. Both use the same
outcome and records. OLS **fits** by minimizing squared training errors;
we separately choose **mean absolute error (MAE)** to evaluate the
resulting predictions in interpretable income units.

A prediction opportunity here means a hypothetical same-source occupation
with all declared predictors available before its income outcome is used.
It does not mean predicting future earnings from information that would
only become available afterward. The records do not establish that this
hypothetical exchangeability assumption holds for new countries, periods,
individual workers, or occupational types absent from the dictionary.

For held-out outcomes $y_i$ and predictions $\widehat y_i^{(-)}$ fitted
without those outcomes,

$$
MAE=\frac{1}{n_{\mathrm{test}}}
       \sum_{i\in\mathrm{test}}\left|y_i-\widehat y_i^{(-)}\right|.
$$

The superscript reminds us that the case was held out of fitting. A smaller
number means smaller average absolute prediction error on the evaluated
cases. It does not say that every case was predicted better.

### Keep held-out outcomes out of training

We use shared ten-fold cross-validation. The 98 records are divided into
ten folds; each fold takes a turn as the held-out set. Fit both equations
on the other records, predict the held-out cases, then combine the 98
case-level errors.

\Needspace{24\baselineskip}

| Decision | Frozen teaching protocol | Why it matters |
|:--|:--|:--|
| Candidates | Education-only and the declared full additive OLS rule | No search for the best-looking model after validation |
| Folds | Shared balanced random ten-fold assignment, seed 80806 | Both rules face the same held-out cases |
| Training sizes | Eight fits use 88 records; two use 89 | Risk concerns those training sizes, not a model trained on all 98 |
| Preparation | Same 98 complete records; fixed income/1,000 conversion and type dictionary | No outcome-informed preprocessing; selection still limits scope |
| Estimability | Every training set must contain all three types and have full column rank | Stop rather than silently substitute a different rule |
| Summary | Pool absolute errors over all 98 held-out cases | Each case receives equal weight; folds of different size do not |
| Sensitivity | 200 new resplits of the same records, seed 80807 | Describes sensitivity to splitting, not new independent studies |

: The internal-validation protocol was fixed before this teaching comparison. {#tbl-ch8-validation tbl-colwidths="[20,42,38]"}

This is a retrospective teaching reanalysis of familiar historical data.
“Fixed before this comparison” is not a claim of historical preregistration
or that analysts had never seen the outcomes. The candidates are not revised
after inspecting the primary errors; the association sensitivities in
@sec-ch8-check are not covert additions to the validated candidate set.

\Needspace{30\baselineskip}

```{r}
#| label: ch8-one-fold
#| echo: true
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
set.seed(80806)
fold <- sample(rep(seq_len(10), length.out = nrow(d)))
held_fold <- fold[1]
train <- d[fold != held_fold, ]
held <- d[fold == held_fold, ]
fold_simple <- lm(income_k ~ education, data = train)
fold_full <- lm(income_k ~ education + prestige + women + type,
                data = train)
table(train$type)
c(training = nrow(train), held_out = nrow(held),
  simple_rank = fold_simple$rank, full_rank = fold_full$rank)
first_case <- d[1, , drop = FALSE]
first_predictions <- c(
  Education = unname(predict(fold_simple, first_case)),
  Full = unname(predict(fold_full, first_case))
)
rbind(prediction = first_predictions,
      absolute_error = abs(first_predictions - first_case$income_k))
```

The check should show all three types, 89 training records and nine held-out
records. The fitted ranks are 2 and 6, the numbers of coefficient columns.
Proceed with the declared comparison only when its training data can estimate
both rules. The Computer Lab reference on all-fold prediction checks this for
every fold.

This is one fold of the frozen protocol, with 89 training records and 9
held-out records. Government administrators, the first source record, is
in fold 9 and was excluded from both fits. The displayed code predicts
that record; repeat the same train/predict/error steps across all ten folds
to obtain the all-98 comparison below. One record's error is not that summary.

The Computer Lab reference on all-fold prediction provides the complete ten-fold
function and 200-resplit code with their interpretation; Chapter 13 develops
held-out validation fully. The visible one-fold example needs
only the preceding core data definitions, not that helper. No held-out
outcome is used to estimate a coefficient. If we introduced learned scaling,
imputation, or variable selection, each step would have to be fitted using
training data only. A fixed unit conversion such as dollars to thousands is
not an estimated preprocessing step.

```{r}
#| label: fig-ch8-folds
#| fig-cap: "Shared held-out cases let two fixed learning rules face the same prediction task."
#| fig-alt: "A training-record box branches to education-only OLS and full OLS. Both arrows lead to predictions for the same held-out records, then to paired absolute errors. The diagram is repeated over ten folds."
#| fig-width: 6.3
#| fig-height: 3.3
par(mar = c(.2, .2, .2, .2))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 5))
boxtext <- function(x1, y1, x2, y2, label, fill = "#EFF4F7") {
  rect(x1, y1, x2, y2, col = fill, border = "#426278")
  text((x1 + x2)/2, (y1 + y2)/2, label, cex = .9)
}
boxtext(.1, 1.8, 2.3, 3.2, "Training records\n88 or 89")
boxtext(3.1, 3.1, 6.2, 4.4, "Education-only OLS")
boxtext(3.1, .6, 6.2, 1.9, "Full additive OLS")
boxtext(7, 1.8, 9.9, 3.2, "Same held-out cases\nPaired absolute errors")
arrows(2.3, 2.8, 3.1, 3.6, length = .08)
arrows(2.3, 2.2, 3.1, 1.4, length = .08)
arrows(6.2, 3.6, 7, 2.8, length = .08)
arrows(6.2, 1.4, 7, 2.2, length = .08)
text(5, .12, "Repeat across ten folds; pool the 98 case-level errors.", cex = .9)
```

### Work one error, then read the comparison

For government administrators, the first source record, observed income is
12.351 thousand dollars. Its held-out fold is 9. The full-rule prediction
is 11.5948, giving

$$
|12.351-11.5948|=0.7562.
$$

The education-only prediction is 8.9461, giving
$|12.351-8.9461|=3.4049$.
This one record favors the full rule. We still need all 98 errors:

| Learning rule | Held-out MAE (thousand dollars) |
|:--|--:|
| Education-only OLS | 2.4031 |
| Full additive OLS | 1.5412 |
| Difference: full minus education-only | $-0.8619$ |

: Case-weighted absolute errors from the frozen shared-fold comparison. {#tbl-ch8-mae tbl-colwidths="[60,40]"}

The full rule's average held-out absolute error is about 862 dollars lower
in this split. Because we subtract the two errors for **each same case**
before averaging, difficulty differences between validation cases do not
come from using different test sets for the two rules.

What is the performance estimand? Eight test folds contain 10 cases whose
predictions were trained on 88; two contain 9 cases trained on 89.
Consequently the pooled quantity targets the declared mixture

$$
R_{\mathrm{mix}}=\frac{80}{98}R_{88}+\frac{18}{98}R_{89},
$$

where $R_m$ is expected absolute error of the specified learning rule
trained on $m$ eligible same-source records and used on a new one under
the stated repetition. Pair 3 compares the two rules' $R_{\mathrm{mix}}$.
It does not directly estimate performance of the single equation fitted
to all 98 observed records.

The 200 resplits give full-minus-education MAE differences from $-0.9140$ to
$-0.7410$, all favoring the full rule. This supports stability to the examined
split choices. It is **not a 95% interval**: the same cases are reused,
training sets overlap, and the results are not independent studies.
We do not test a mean of ten fold errors as if those errors were independent.
Uncertainty across newly drawn training samples is not quantified here.
The estimated risk advantage is therefore provisional; its sign is not
established for the hypothetical source population by the resplit range.

The comparison supports lower internal held-out error for the full rule in
these records and under this protocol. It does not isolate the predictive
contribution of education: the rules differ by prestige, women, and type
together. Nor does it establish external transport, individual earnings
prediction, or a causal education effect.

\Needspace{14\baselineskip}

## Read the Coordinated Arguments, Then Evaluate and Synthesize {#sec-ch8-coordinate}

### The chapter-authored study record

The following report consolidates the preceding analysis. Treat its
paragraphs, tables, equations, and code as an accessible teaching record,
not as statements copied from the source investigators.

> **R0 -- Common record and declared scope.** We analyzed the 98 complete
> occupation records described in @sec-ch8-questions. Income was
> scaled to thousand
> 1971 Canadian dollars. Education and income were historical occupational
> averages; prestige came from an earlier rating survey. We retained the
> same rows for all three questions and make no claim of verified probability
> sampling or randomized education.
>
> **R1 -- Marginal answer.** Across these records the education-only
> least-squares slope was 0.8830 thousand dollars per year. This is a
> descriptive linear summary: occupations with more education tended to
> have higher average income. It is not an individual or causal effect.
>
> **R2 -- Conditional answer.** The intended estimand was the linear
> conditional-mean education coefficient $\beta_E$, comparing education at the
> same prestige, percentage of women, and occupational type. Its fitted
> estimate was $b_E=0.1164$, with a classical 95% interval [$-0.4302$, 0.6631]
> under the independent linear conditional-mean working model with common
> variance. HC3 gave [$-0.6429$, 0.8758]. Influence, functional-form, and support
> checks limit that model's interpretation. The direction and practically
> relevant magnitude remain unresolved under these calculations; the result
> does not establish no conditional relationship.
>
> **R3 -- Prediction answer.** The two fixed OLS learning rules were compared
> with the shared ten-fold protocol in @sec-ch8-prediction. The full
> rule had MAE
> 1.5412 versus 2.4031 for education-only, difference $-0.8619$. The examined
> resplits all favored the full rule. This is internal evidence about
> learning procedures trained on 88/89 eligible records under the stated
> same-source exchangeability assumptions. Lower expected error is a
> provisional conclusion because new-sample uncertainty is not quantified;
> this is not external validation of a final 98-record equation.
>
> **R4 -- Synthesis.** Education has a positive marginal linear association
> with income in these historical records. Its conditional association is
> imprecisely resolved by the specified additive model, while the full
> learning rule predicts the held-out records more accurately than the
> education-only rule. These conclusions answer different questions and
> do not establish that changing education would change income.

### Read the functions and their traces

A component is a logical function; the record contains study-specific
content serving that function. A passage can carry several functions.
For example, R0 identifies records for STAT-E and supplies historical
context for SUB-E. It becomes J content only when it offers a reason
connecting those records to a particular inference. “We used R” does not
by itself supply that reason.

| Function | Shared or pair-specific content | Trace in this teaching record |
|:--|:--|:--|
| **STAT-E** | Named 98 rows, variables, exclusions, and computations' inputs | R0; the row audit and import code in @sec-ch8-questions |
| **SUB-E** | Occupations rather than workers; historical dollars, education, ratings, and types | R0; variable/source descriptions |
| **STAT-J** | Pair 1: fixed-record least-squares definition. Pair 2: declared stochastic model and uncertainty conditions. Pair 3: held-out fitting, fixed rules, loss, and exchangeability | R1--R3; model declaration; validation protocol and code |
| **SUB-J** | Reasons connecting those measurements and comparisons to historical occupational income, with limits on individual, current, and causal interpretations | R0--R3; unit/time explanations and prediction-opportunity definition |
| **STAT-I** | Pair 1's 0.8830 slope; Pair 2's 0.1164 estimate, intervals, coefficient test and checks; Pair 3's case-level errors, MAEs, and split sensitivity | @sec-ch8-estimate through @sec-ch8-prediction; R1--R3 |
| **STAT-C** | Descriptive positive marginal slope; unresolved conditional slope under the model; smaller internal error for the full rule under the protocol | R1, R2, and R3 respectively |
| **SUB-I** | Interpret years, income units, conditioning, and prediction errors in the named historical occupation setting | Worked substitutions and error calculation; R1--R3 |
| **SUB-C** | Each question-facing answer with its own limits; none supplies a causal or individual-earnings answer | R1--R3; their distinctions preserved in R4 |

: Reading the linked-layer functions without three duplicate maps. {#tbl-ch8-traces tbl-colwidths="[15,47,38]"}

SUB-E situates evidence; SUB-J **augments and contextualizes** statistical
justification; SUB-I interprets the inference; SUB-C bounds the substantive
answer. Sharing a dataset does not make Pair 2's independent-error model
into Pair 3's validation procedure.

For each supporting statement, record what the report says and where the
reader can find it: the report itself, the data documentation, or the
executable calculation record. State gaps and conflicts in ordinary prose.
For example, R2 states the independence assumption, but this teaching record
does not supply evidence establishing independence under the original
occupation-selection process. A stated assumption and evidence for it are
different pieces of information.

Similarly, the income-unit conversion makes the interpretation of 0.8830
thousand dollars as about 883 dollars traceable without inventing an
empirical premise. No conflicting source is identified in this defined
record; a real study may contain incompatible descriptions that a reader
should record and investigate. Here, a reason's documentary status means
what has been stated or located, and which support is missing. It is not a
quality score or a four-category classification to learn.

Professional knowledge helps identify what function a trace serves. It
does not authorize adding an unreported design feature. If an author gave
the invalid reason “all rows are in the same dataset, therefore they are
independent,” it would still be **stated J**, whose adequacy is judged
separately. Reading records the claim actually offered, not a repaired
claim the reader wishes had been offered.

### Evaluate each argument separately

We now change from identifying the reported reasoning to evaluating its
quality. The six Chapter 1 questions apply to each argument. They are not
six scores to total, and not every concern needs to be counted repeatedly.

| Quality-evaluation question | Application to these arguments |
|:--|:--|
| **Question alignment** | R1 matches its finite-record slope. R2 names $\beta_E$, its fitted estimate, the conditioning set, and the model extension. R3 names the learning procedure, loss, and training sizes rather than substituting a coefficient test. |
| **Evidence integrity** | The 102-to-98 accounting and common rows are explicit. Missing type still limits what is represented. CV excludes held-out outcomes from fitting; source selection is not repaired by that procedure. |
| **Justification adequacy** | Least squares is sufficient for the declared finite description. Pair 2's model conditions and Pair 3's same-source exchangeability remain assumptions with incomplete empirical support. Statistical computation alone cannot establish them. |
| **Inferential adequacy** | The slope, intervals, error arithmetic, and comparisons are reproducible. Neither zero-crossing nor a large p-value proves absence; repeated folds are not independent replication. Substantive reasoning must retain those distinctions. |
| **Transition fidelity** | Occupation-level average income remains occupation-level average income; partial adjustment is not a causal intervention, and model fit is not held-out performance. |
| **Claim calibration** | R1 remains finite and historical; R2 reports unresolved magnitude/direction rather than no relationship; R3 remains internal and procedure-specific. No claim extends to individual workers or modern populations. |

: Apply the six questions before evaluating the study-level synthesis. {#tbl-ch8-quality tbl-colwidths="[28,72]"}

At an individual question, useful judgments include **adequate for this
purpose**, **needs revision**, **insufficient information to judge**, or
**not applicable**, with a reason. For example, the record is adequate to
check Pair 3's held-out computations but gives insufficient information to
establish exchangeability for an external population. These judgments assess adequacy; they are separate from recording
what the report says and where its support can be found.

\Needspace{6\baselineskip}

Each whole pair then receives its own conclusion:

- **Pair 1: supported as stated.** The answer is an accurate bounded
  description of the retained records.
- **Pair 2: supported as stated**, as the explicitly conditional,
  unresolved working-model answer about $\beta_E$ actually reported. This does not certify
  the model or establish a population relationship. A claim that the true
  national conditional slope is zero would not be supported by this record.
- **Pair 3: supported as stated**, as a qualified internal comparison with
  assumptions and limits retained. External superiority of the final
  all-record fitted equation is **not assessable from the accessible record**;
  it is not the reported answer being accepted.

The other available whole-pair conclusions are **supported only after
narrowing or revision**, **not supported**, and **not assessable from the
accessible record**. A correctly reported inconclusive answer can be
supported as stated. Evaluating what the study actually says prevents a
bounded answer from being rejected merely because it does not answer a
stronger question.

### Add Cross-question coherence for the synthesis

For coordinated questions, ask one additional question:
**Does the synthesis preserve the distinct questions, estimands, methods,
analytic statuses, uncertainty, and limits of the separate answers?**
This is **Cross-question coherence**, introduced locally here. Supplementary B
develops it further but is not needed to apply it.

R4 is coherent because it reports three distinct outcomes. A positive
marginal slope, an unresolved conditional slope, and better internal
prediction can coexist. They do not contradict one another.

Now consider the alternative sentence:

> Once the other predictors were controlled, education had no effect;
> nevertheless, the significant full regression proves it will predict
> workers' incomes accurately.

Reading first identifies the offered inference and claim. Evaluation then
finds several failures: “no effect” substitutes a causal null for the
conditional-association answer; an omnibus test is not validation; and
occupations become workers. No omnibus significance test can replace the
held-out errors or rescue the change of unit. The corrected synthesis is
R4, not a fourth overarching argument that erases the separate verdicts.

### Auxiliary construction callback

The construction workflow supports researchers carrying out the study; it
is not a mandatory reading sequence. Each question retains its own
six-stage cycle from Chapter 1. For coordinated questions, Supplementary B's
extension aligns the plans at Stage 3 and adds Stage 7 for synthesis.

In **Understand and plan**, researchers review the problem and literature,
frame questions and candidate answers, and align design, measurement,
estimands, evidence acquisition, and inference. Here, “separate questions”
means separately owed conclusions. If the obligation were one integrated
conclusion, jointly necessary estimands would be planned as strands instead.

The extended **Stage 3, Plan and align prospectively across questions**,
specifies common rows and units but distinct marginal, model-based, and
validation plans. In **Produce and complete**, **Stage 4, Produce or acquire
and document evidence**, designates the question-specific records;
**Stage 5, Estimate, check, and analyse**, performs the computations;
and **Stage 6, Contextualize, interpret, and complete the bounded answer**, finishes each answer with its actual limits.
**Stage 7, Give explicit outcomes and synthesize**, records all outcomes
and brings them together without creating an omnibus argument.

Iteration remains within a phase. A same-anchor error correction or
justified uncertainty repair can reopen Phase 2 work. A newly preferred
outcome, question mode, or estimand requires a new or revised cycle, with
reused records where appropriate and its timing disclosed. A newly chosen
causal question cannot be “narrowed” into association and treated as answered.
This teaching reanalysis does not retrospectively claim that the historical
data were collected for our three questions.

The occupation study supplies no randomized-design warrant. A second application
now shows how covariate adjustment can seek the same assignment-effect estimand
when design and weighting support that link.

```{=latex}
\Needspace{22\baselineskip}
```

## Reading Bridge {#sec-ch8-summary}

Multiple regression adds conditional comparisons, not automatic causal
control. Subtracting fitted means showed what a coefficient compares;
residualization connected it to covariance; residual, influence,
support, and precision checks had different consequences for a bounded
answer; a fitted profile and a standardized quantity were didactic
extensions of the conditional model; and held-out errors answered a
separately owed question. The occupation study owed three answers: a
finite-record marginal description, a working-model conditional interval
with unresolved magnitude, and a qualified internal prediction comparison.
Shared records made the questions coordinated without making their
justifications identical. The next regression question is different:
might the education comparison itself vary with another predictor rather
than remain the same additive slope? Chapter 9 first collects the tools this chapter used only briefly: factors
with several levels and their contrasts, nested model comparison and
information criteria, changed predictor and outcome forms, and robust,
clustered, and bootstrap uncertainty, and it transfers the reading
discipline to Project STAR, where a randomized design supplies a different
warrant. Chapter 10 then extends the
interaction reasoning from Chapters 4 and 6 to these regression comparisons.

### Key terms {.unnumbered}

Use this short reference to revisit the concepts introduced here. The
Part One introduction gives the broader reference and the principal home
of each concept; the simple-regression, interval, ANOVA, and reading
vocabulary used in this chapter is explained in Chapters 1--4 and 7.

| Term and common alternatives | Meaning |
|:--|:--|
| Multiple regression (additive model) | A fitted equation in which the outcome's mean is the sum of an intercept and one term per predictor. Each coefficient compares fitted means at the same values of the other predictors; the additive form assumes each comparison is the same across those values. |
| Partial slope (conditional coefficient; adjusted coefficient) | The difference in fitted outcome per one-unit difference in one predictor, holding the model's other predictors fixed. Its meaning depends on the conditioning set named with it; it is not an intervention effect or a claim about unmeasured differences. |
| Indicator variable (dummy variable) and reference category | A 0/1 variable marking membership in one category of a factor; a factor with $J$ levels needs $J-1$ indicators, and the omitted level is the reference. The intercept refers to the reference category; each indicator coefficient is a difference from it at the same other predictors. Releveling changes coordinates, not fitted values. |
| Partial regression (residualization; added-variable form) | Regressing both the outcome and one predictor on the other predictors, then relating their residuals. The residual slope equals the partial slope; the residual correlation is the partial correlation, unit-free. |
| Coefficient covariance matrix | The estimated variances and covariances of the fitted coefficients (`vcov()` in R). Any linear combination of coefficients, such as a fitted mean or a contrast, has variance $c'\widehat Vc$; adding SEs or ignoring covariances is a different, incorrect calculation. |
| Variance inflation factor (VIF) | One divided by one minus the $R^2$ from regressing a predictor on the other predictors. Its square root multiplies that coefficient's SE relative to an orthogonal design, holding residual variance fixed; it describes precision, not bias, and is not a deletion rule. |
| Joint support (covariate hull) | Whether a combination of predictor values lies within the region where observations actually occur, not merely within each variable's separate range. A fitted value outside joint support is extrapolation even if each coordinate is observed somewhere. |
| Standardization over a named distribution (model-averaged fitted value) | Averaging the model's fitted values after setting one predictor to a chosen value for every profile in a declared set with fixed weights. Its uncertainty uses the coefficient covariance; it is not an observed group mean, a causal effect, or held-out performance. |
| Learning rule (learning procedure) | A fixed recipe that turns training records into a prediction equation, such as fitting a declared OLS formula. Prediction performance concerns the rule at stated training sizes, not one final fitted equation. |
| Cross-validation (held-out validation) | Dividing records into folds, fitting a rule on the other folds, predicting the held-out fold, and pooling the errors so that no outcome contributes to its own prediction. Resplits describe sensitivity to the partition, not new independent studies. |
| Mean absolute error (MAE) | The average absolute difference between held-out outcomes and their predictions, in outcome units. It is an evaluation loss chosen separately from the squared-error criterion that OLS minimizes. |
| Cross-question coherence | The additional evaluation question for coordinated answers: does the synthesis preserve the distinct questions, estimands, methods, uncertainty, and limits of the separate answers rather than merging them into one omnibus claim? |

: Selected concepts for multiple regression across coordinated questions. {#tbl-ch08-key-terms tbl-colwidths="[28,72]"}

### Further reading {.unnumbered}

- The [OpenIntro multiple-predictor tutorial](https://openintro-ims.netlify.app/model-mlr)
  offers concrete equations and indicator interpretations. We use that
  equation-first teaching device without adopting automatic stepwise
  selection as this chapter's route.
- The [ISLR regression lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch3-linreg-lab.html)
  connects fitting, coefficients, checks, and predictions [@james2021].
- The [ISLR cross-validation lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch5-resample-lab.html)
  develops the train/predict/check sequence [@james2021]. Our fixed rules,
  MAE, and training-size qualifications are specified for this teaching
  example.
- The [carData documentation](https://cran.r-project.org/web/packages/carData/carData.pdf)
  [@carData2022; @fox2019] supplies the Prestige variable meanings and
  provenance. The coordinated study and R0--R4 report are chapter-authored,
  not claims attributed to that documentation.
- Fox [@fox2016] and Kutner and colleagues [@kutner2005] develop the
  linear-model algebra behind partial slopes, indicators, and diagnostics.

Supplementary B develops coordination and its auxiliary construction
workflow in greater detail.

## Exercises {.unnumbered}

Use the supplied numbers; rounding within 0.02 thousand dollars is
acceptable unless a different precision is requested. These prompts do not
require Supplementary B or the appendices.

\Needspace{18\baselineskip}

<!-- prompt-id: CH08-E01 -->
### Exercise 1: Keep the three answers distinct (Foundational) {.unnumbered #ex-ch08-01}

A report promises: (a) a description of education and income across 98
occupations, (b) an education comparison at the same prestige, women, and
type, and (c) a comparison of two prediction rules on held-out records.

1. Explain why these are three question--answer pairs rather than three
   statistical strands serving one integrated answer.
2. For each, name its estimand in plain language and one kind of output
   that could contribute to its answer.
3. Does sharing the same 98 records establish the same justification for
   all three? Give one concrete reason.

\Needspace{18\baselineskip}

<!-- prompt-id: CH08-E02 -->
### Exercise 2: Name what changes (Foundational) {.unnumbered #ex-ch08-02}

Classify each change as primarily changing the records, the comparison,
the coefficient coordinates, or the uncertainty calculation. State whether
it necessarily creates a new scientific question.

1. Refit education-only regression after excluding four missing-type records.
2. Add prestige, women, and type to the education-only association equation.
3. Change the reference category from blue-collar to white-collar.
4. Keep the fitted equation but replace classical SEs with HC3 SEs.

Finally, explain why appropriately specified adjustment in a randomized
study need not change its declared average assignment-effect estimand.

<!-- prompt-id: CH08-E03 -->
### Exercise 3: Interpret and calculate a fitted profile (Foundational) {.unnumbered #ex-ch08-03}

Use this rounded hypothetical equation, with income in thousand dollars:

$$
\widehat y=1+0.20e+0.10p-0.04w+0.30D_{\mathrm{prof}}
           +0.15D_{\mathrm{wc}}.
$$

1. Calculate the fitted mean for a white-collar profile with $e=12$,
   $p=40$, and $w=50$.
2. Calculate its modeled difference from a profile with $e=10$ and the
   same other values. Interpret the result in dollars.
3. What must be checked before calling this a comparison supported by the
   observed data?
4. Why is neither calculation an estimate of the effect of giving a worker
   two more years of education?

\Needspace{18\baselineskip}

<!-- prompt-id: CH08-E04 -->
### Exercise 4: Read held-out errors (Foundational) {.unnumbered #ex-ch08-04}

For four held-out occupations, observed incomes are $(5,8,10,13)$ thousand
dollars. The education-only predictions are $(4,10,7,12)$; full-rule
predictions are $(6,8,9,11)$. Each prediction was fitted without the
corresponding outcome.

1. Calculate the four absolute errors and MAE for each rule.
2. Calculate the average paired difference, full minus education-only.
3. Does the better average rule have to be better for every occupation?
4. Explain why this comparison is neither a coefficient interval nor
   validation of a model subsequently fitted to all available records.

\Needspace{22\baselineskip}

<!-- prompt-id: CH08-E05 -->
### Exercise 5: Locate a reason before judging it (Integrative) {.unnumbered #ex-ch08-05}

An accessible report states:

> We retained the same 98 occupations for both equations. This makes the
> observations independent. The full equation has higher training
> $R^2$, so we conclude it predicts new occupations more accurately.

The report provides import code and fitted-model output, but no validation
split, held-out predictions, or study-specific evidence about dependence.

1. Identify the reasons offered, the inference, and the claim.
2. Record the documentary status of the independence reason and the
   held-out prediction evidence separately. Do not rewrite “stated” as
   “not located” merely because a reason is wrong.
3. Apply Justification adequacy, Inferential adequacy, and Claim calibration.
4. Give a whole-pair evaluation conclusion for the prediction answer and
   name the additional work it would require.

\Needspace{25\baselineskip}

<!-- prompt-id: CH08-E06 -->
### Exercise 6: Evaluate the study-level synthesis (Integrative) {.unnumbered #ex-ch08-06}

Use the occupation study's reported results: marginal slope 0.883;
fitted conditional estimate $b_E=0.116$ of $\beta_E$, with classical
interval [$-0.430$, 0.663];
and held-out MAEs 2.403 for education-only versus 1.541 for the full rule.

A summary says:

> Education does not matter after adjustment, but the model proves that
> increasing workers' education raises income and predicts earnings well
> in other countries.

1. Read the claims actually offered before evaluating them.
2. Identify which separate answer obligation is being replaced or exceeded
   by each clause.
3. Apply Cross-question coherence and write a corrected three-sentence
   synthesis that retains the three reported answers and their limits.
4. If the authors now want a causal effect of education, is that a repair
   of their existing uncertainty calculation or a new/revised question-level
   cycle? Explain.

## Appendix A: Adjustment, Coding, and Group-Comparison Mathematics {.unnumbered #app-ch8-a}

This appendix develops the mathematics behind @sec-ch8-comparisons.
Appendix B retains technical reproduction and sensitivity; Chapter 9
develops contrasts, block tests, adjusted means, the matrix form, and the
Project STAR transfer.
The occupation calculations use the same 98-row object *d* and the
models *m0*, *m1*, and *m2* defined in the main text. They do not introduce
another sample. The type codes remain *bc*, *prof*, and *wc*.

### Why the marginal slope includes another association {.unnumbered}

Suppose a population process is

$$
Y=\beta_0+\beta_X X+\beta_Z Z+\varepsilon,
\qquad E(\varepsilon\mid X,Z)=0,
$$

with finite variances and $\operatorname{Var}(X)>0$. The marginal
least-squares slope is a covariance divided by a variance. Substituting
the outcome equation gives

$$
\begin{aligned}
\frac{\operatorname{Cov}(X,Y)}{\operatorname{Var}(X)}
&=\frac{\beta_X\operatorname{Var}(X)
+\beta_Z\operatorname{Cov}(X,Z)
+\operatorname{Cov}(X,\varepsilon)}{\operatorname{Var}(X)}\\
&=\beta_X+\beta_Z
\frac{\operatorname{Cov}(X,Z)}{\operatorname{Var}(X)}.
\end{aligned}
$$ {#eq-ch8a-omission}

The zero-error-covariance condition removes the final term. The
discrepancy therefore involves both the association between $X$ and $Z$
and $Z$'s coefficient in the outcome model. Neither ingredient alone
determines its size. If the target is the conditional coefficient,
omission creates bias relative to that target. If the target is the
marginal slope, that marginal quantity is not a biased version of an
unstated conditional target. Neither interpretation supplies a causal
warrant for observational variables.

Changing a specification does not universally change a scientific
estimand. Different estimators can target the same declared quantity,
as Chapter 9's randomized adjustment example, Project STAR, illustrates.
Releveling a factor merely changes coordinates. By contrast, the
marginal and conditional coefficients in this example name different
comparisons.

### A known process and repeated samples {.unnumbered}

Let independent standard normal variables $X$ and $U$ generate
$Z=.65X+\sqrt{1-.65^2}U$. Then both predictors have variance one and
their correlation is .65. Independently generate a normal error with
SD 6, and set

$$
Y=50+2X+5Z+\varepsilon.
$$ {#eq-ch8a-sim-process}

The marginal slope is $2+5(.65)=5.2500$; the conditional $X$ slope is 2.
The generated sample below has 250 observations and uses seed 82108.
Its marginal slope is 5.9971 and conditional strategy slope is 3.0259.
The strategy and prior-attainment labels are fictional teaching
interpretations. The Computer Lab reference on conditional comparisons supplies
the complete generator, fitted models, and plotting code for these figures.

```{=latex}
\Needspace{24\baselineskip}
```

```{r}
#| label: ch8a-one-sample
#| echo: false
#| results: hide
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
set.seed(82108)
ch8a_n <- 250
ch8a_x <- rnorm(ch8a_n)
ch8a_z <- .65 * ch8a_x + sqrt(1 - .65^2) * rnorm(ch8a_n)
ch8a_y <- 50 + 2 * ch8a_x + 5 * ch8a_z +
  rnorm(ch8a_n, sd = 6)
ch8a_sim <- data.frame(
  strategy = ch8a_x, prior = ch8a_z, score = ch8a_y
)
ch8a_marginal <- lm(score ~ strategy, data = ch8a_sim)
ch8a_adjusted <- lm(score ~ strategy + prior, data = ch8a_sim)
c(marginal = unname(coef(ch8a_marginal)["strategy"]),
  conditional = unname(coef(ch8a_adjusted)["strategy"]))
```

```{r}
#| label: fig-ch8a-predictors
#| fig-pos: H
#| echo: false
#| fig-cap: "Joint predictor support in one generated sample."
#| fig-alt: >
#|   A diagonal cloud shows positively related strategy and prior
#|   attainment. The central region has more nearby comparisons
#|   than the sparse corners.
#| fig-width: 6.1
#| fig-height: 3.5
ggplot2::ggplot(ch8a_sim, ggplot2::aes(strategy, prior)) +
  ggplot2::geom_point(shape = 1, alpha = .65) +
  ggplot2::geom_smooth(method = "lm", se = FALSE) +
  ggplot2::labs(x = "Strategy (standardized)",
                y = "Prior attainment (standardized)")
```

```{r}
#| label: fig-ch8a-outcome
#| fig-pos: H
#| echo: false
#| fig-cap: "A marginal line does not hold prior attainment fixed."
#| fig-alt: >
#|   Score increases with strategy. Higher prior-attainment values,
#|   represented by darker points, tend to accompany higher scores
#|   and strategy. The black line is the marginal fitted relation.
#| fig-width: 6.1
#| fig-height: 3.5
ggplot2::ggplot(
  ch8a_sim, ggplot2::aes(strategy, score, colour = prior)
) +
  ggplot2::geom_point(alpha = .8) +
  ggplot2::geom_smooth(
    ggplot2::aes(group = 1), method = "lm", se = FALSE,
    colour = "black"
  ) +
  ggplot2::scale_colour_gradient(low = "grey75", high = "black") +
  ggplot2::labs(x = "Strategy (standardized)", y = "Score",
                colour = "Prior attainment")
```

The first plot displays the support needed for a conditional
comparison. The second shows that prior attainment varies along the
marginal relation. A line calculated outside the observed predictor
cloud is not made well supported by the existence of its equation.

Now use a different, explicitly stated sample size: 150 observations in
each of 1,000 repetitions. This is a distinct repeated-study design, with seed 82109; it is not
repetition of the 250-observation draw. The same Computer Lab reference supplies
the full simulation and density-plot code.

```{=latex}
\Needspace{16\baselineskip}
```

```{r}
#| label: ch8a-repetition
#| echo: false
#| results: hide
set.seed(82109)
ch8a_one_rep <- function() {
  x <- rnorm(150)
  z <- .65 * x + sqrt(1 - .65^2) * rnorm(150)
  y <- 50 + 2 * x + 5 * z + rnorm(150, sd = 6)
  c(Marginal = unname(coef(lm(y ~ x))["x"]),
    Conditional = unname(coef(lm(y ~ x + z))["x"]))
}
ch8a_slopes <- t(replicate(1000, ch8a_one_rep()))
```

```{r}
#| label: tbl-ch8a-repeated
#| echo: false
#| tbl-cap: "Targets and results across 1,000 generated samples."
knitr::kable(data.frame(
  Model = colnames(ch8a_slopes), Target = c(5.25, 2),
  Mean = colMeans(ch8a_slopes),
  SD = apply(ch8a_slopes, 2, sd), row.names = NULL
), digits = 4)
```

```{r}
#| label: fig-ch8a-repeated
#| echo: false
#| fig-pos: H
#| fig-cap: "Sampling distributions around two different targets."
#| fig-alt: >
#|   Separate panels show marginal slopes centered near 5.25 and
#|   conditional slopes centered near 2. Dashed lines mark their
#|   respective population targets.
#| fig-width: 6.1
#| fig-height: 3.8
ch8a_long <- data.frame(
  Slope = c(ch8a_slopes[, 1], ch8a_slopes[, 2]),
  Model = rep(c("Marginal", "Conditional"), each = 1000)
)
ch8a_targets <- data.frame(
  Model = c("Marginal", "Conditional"), Target = c(5.25, 2)
)
ggplot2::ggplot(ch8a_long, ggplot2::aes(Slope)) +
  ggplot2::geom_density(fill = "grey80") +
  ggplot2::geom_vline(
    data = ch8a_targets, ggplot2::aes(xintercept = Target),
    linetype = 2
  ) +
  ggplot2::facet_wrap(~Model, scales = "free_x") +
  ggplot2::labs(x = "Estimated strategy slope", y = "Density")
```

Both procedures center near their own targets. The conditional
estimates are more variable here: separating correlated predictors
uses less residual predictor variation. This is a result of the
specified process, not a general promise that adjustment always lowers
precision. Prognostic covariates and their relations to the focal
predictor both matter. Simulation demonstrates consequences of its
assumptions; it cannot certify the observational occupation study.

\Needspace{12\baselineskip}

### Indicators recover fitted group means {.unnumbered}

Start with a model containing only the three-level type factor. With
blue-collar as reference,

$$
\widehat y_g=\hat\alpha+
\hat\delta_{\mathrm{prof}}D_{\mathrm{prof}}+
\hat\delta_{\mathrm{wc}}D_{\mathrm{wc}}.
$$

The fitted means are $\hat\mu_{\mathrm{bc}}=\hat\alpha$,
$\hat\mu_{\mathrm{prof}}=\hat\alpha+\hat\delta_{\mathrm{prof}}$,
and $\hat\mu_{\mathrm{wc}}=\hat\alpha+\hat\delta_{\mathrm{wc}}$.
OLS minimizes squared deviations separately within each group, so these
equal the observed group means. The hats distinguish fitted summaries
from hypothetical population means.

```{r}
#| label: ch8a-group-means
#| echo: true
ch8a_groups <- data.frame(
  type = factor(c("bc", "prof", "wc"), levels = levels(d$type))
)
model.matrix(~type, data = ch8a_groups)
ch8a_group_fit <- lm(income_k ~ type, data = d)
cbind(ch8a_groups,
      observed = as.numeric(tapply(d$income_k, d$type, mean)),
      fitted = predict(ch8a_group_fit, ch8a_groups))
coef(ch8a_group_fit)
```

This group-only description is not the adjusted type comparison in
*m2*. In *m2*, indicators compare fitted values at the same education,
prestige, and percentage women. The factor is one term represented by
two columns; it is neither an ordered numeric score nor one selected
reference-category coefficient.

Releveling changes the intercept and indicator coefficients, not the
fitted values, residuals, or a fixed substantive contrast.

\Needspace{14\baselineskip}

The Computer Lab reference on conditional comparisons reproduces this releveling
operation and checks the unchanged fitted values with complete code and
output.

Releveling reproduces the same fit up to floating-point noise. No new question has appeared merely
because the coordinates changed. Centering continuous predictors also
changes the intercept's reference profile without changing fitted
values. It cannot create cross-type support.

### The overall F test {.unnumbered}

The overall regression test sets all $p-1$ non-intercept coefficients to
zero. Under the same normal-error conditions as the coefficient tests, its
statistic is

$$
F_{\mathrm{overall}}=
\frac{RegSS/(p-1)}{RSS/(n-p)},
\qquad RegSS=\sum_i(\hat y_i-\bar y)^2.
$$

\Needspace{10\baselineskip}

```{r}
#| label: ch8a-overall-f
#| echo: true
summary(m2)$fstatistic
summary(m0)$fstatistic
```

The full model gives $F(5,92)=32.1747$; the education-only model on the
same 98 records gives $F(1,96)=47.1956$. Each tests its stated joint
restriction. Neither verifies the mean function or shows how well the
fitted rule will predict a held-out outcome. Chapter 9 develops the
general nested comparison behind this statistic, its matrix form, and the
contrasts, block tests, and adjusted group means that a factor with
several levels invites.

### Extended practice: coding and known processes (Integrative) {.unnumbered}

1. Recover fitted *m2* values for the three types at education 11,
   prestige 50, and women 30. Treat this as algebra, then evaluate
   whether all three combinations have observed support. Relevel
   type to white-collar and verify a fixed fitted contrast is unchanged.
2. Reproduce the repeated-sample process at $n=150$. Change exactly
   one of the correlation's sign, its value to zero, or the omitted
   variable's coefficient to zero. Derive both targets, run at least
   1,000 repetitions with a recorded seed, and compare bias and SD
   relative to each target. A further comparison uses separated
   predictor clusters and explicitly recalculates their covariance.

## Appendix B: Technical Reproduction and Sensitivity {.unnumbered #app-ch8-b}

The numbered route now carries the ordinary support, diagnostic,
standardization, and targeted-sensitivity lessons. This appendix retains
mathematical detail and worked results; the Computer Lab supplies the longer
computational reproductions. It uses *d*, *m0*,
*m1*, and *m2* without changing their source rows. Classical intervals are
conditional-model benchmarks; technical refinements do not create missing
design, measurement, independence, or transport warrants.

\Needspace{12\baselineskip}

### Cook's distance and full deletion accounting {.unnumbered}

The main route interprets residual, leverage, and influence views and shows
the education-slope sensitivity. This section supplies the corresponding
Cook's distance formula and complete coefficient accounting.

With $p$ coefficient columns including the intercept, leverage is
$h_{ii}=x_i'(X'X)^{-1}x_i$. Cook's distance can be calculated as

$$
D_i=\frac{e_i^2}{p s^2}\frac{h_{ii}}{(1-h_{ii})^2}
=\frac{(\hat\beta-\hat\beta_{(-i)})'X'X
(\hat\beta-\hat\beta_{(-i)})}{p s^2}.
$$

The second expression measures a coefficient change using the design
matrix's metric, not an unweighted sum of squared coefficients with
incompatible units.

```{=latex}
\Needspace{12\baselineskip}
```

```{r}
#| label: tbl-ch8b-influence
#| echo: false
#| tbl-cap: "The five largest Cook's distances in the full fit."
knitr::kable(ch8_diag[ch8_top,
  c("occupation", "residual", "leverage", "cooks")],
  digits = 4, row.names = FALSE)
```

The record for general managers has the largest Cook's distance, approximately 0.4467.
The main route explains why this observation is inspected and retained.

\Needspace{14\baselineskip}

```{r}
#| label: tbl-ch8b-deletion-coefficients
#| echo: false
#| tbl-cap: "Coefficient sensitivity to the most influential record."
knitr::kable(data.frame(
  Term = names(coef(m2)), Full = coef(m2),
  Omit_one = coef(ch8_without_top),
  Difference = coef(ch8_without_top) - coef(m2), row.names = NULL
), digits = 4, row.names = FALSE)
```

The type coordinates also move. The table records a diagnostic perturbation
of the fit, not six new research questions or evidence that the reduced
sample is preferable.

### HC3 changes covariance, not the records or target {.unnumbered}

For independent observations, the HC3 covariance estimator is

$$
\widehat V_{\mathrm{HC3}}=(X'X)^{-1}
\left\{\sum_i x_ix_i'\frac{e_i^2}{(1-h_{ii})^2}\right\}
(X'X)^{-1}.
$$

The outer factors are the familiar design component. The middle
component allows different residual scales and adjusts for leverage
[@long2000; @zeileis2004]. It leaves the OLS coefficients unchanged.

The Computer Lab reference on conditional comparisons supplies the manual matrix
reproduction of this expression and compares it with `vcovHC()`. The
first classical/HC3 comparison in Section 3.3 already supplies the
education interval; the table gives the full coefficient context.

```{r}
#| label: tbl-ch8b-hc3
#| echo: false
#| tbl-cap: "Coefficient estimates and two covariance summaries."
knitr::kable(data.frame(
  Term = names(coef(m2)), Estimate = coef(m2),
  Classical_SE = sqrt(diag(vcov(m2))),
  HC3_SE = sqrt(diag(hc3_covariance)), row.names = NULL
), digits = 4, row.names = FALSE)
```

Education's HC3 SE is 0.3823 rather than 0.2752. Using $t_{92}$ as
an approximate reference gives $[-0.6429,0.8758]$. This is not an
exact finite-sample HC3 coverage result. Women's HC3 SE is smaller
than its classical counterpart: robust covariance need not widen
every interval.

For a maintained linear conditional mean, HC3 addresses unequal
variance under independent observations. For a random-design
best-linear-projection target, a suitable sandwich covariance can
support large-sample projection inference without making the mean
exactly linear. Neither interpretation repairs measurement error,
dependence, selection, causal identification, or absent joint support.

\Needspace{13\baselineskip}

### VIF describes coefficient precision {.unnumbered}

For a single predictor column, classical conditional variance can be
written

$$
\operatorname{Var}(\hat\beta_j\mid X)=
\frac{\sigma^2}{\sum_i(X_{ij}-\bar X_j)^2}
\frac{1}{1-R_j^2}.
$$

Here $R_j^2$ comes from regressing that predictor on the remaining
columns, including the intercept. The variance inflation factor is
$VIF_j=1/(1-R_j^2)$; its square root is an SE multiplier, holding the
residual variance and focal predictor variation fixed [@kutner2005].

The first education VIF calculation is in Section 4.2. The Computer Lab reference
extends it to `car::vif(m2)` and the two-column type term, with the code and
output kept beside the dimension qualification below.

Education gives $R_j^2=.8765$, VIF 8.0968, and SE multiplier
2.8455. The predictor correlation between education and prestige
is .8665, but VIF uses all the other columns, not just that pair.

The two-column type term has a generalized VIF. Its raw GVIF is
7.7001; the dimension-adjusted value
$GVIF^{1/(2\,df)}=1.6658$ is on a comparable scale to the square
root of a one-column VIF [@fox2019]. Do not compare a raw two-column
GVIF with an ordinary one-column VIF as if their dimensions matched.

Exact collinearity creates rank deficiency. Near collinearity leaves
coefficients estimable but potentially imprecise. Under the maintained
zero-conditional-mean model it does not itself bias OLS. A cutoff of
5 or 10 neither establishes all assumptions nor justifies removing a
variable required by the question. Measurement problems also need
their own evidence: the familiar simple attenuation story requires
classical error assumptions and is not a universal rule for every
coefficient in a multiple-predictor model.

### A local Normal likelihood and in-sample comparison criteria {.unnumbered}

Chapter 6 varied the candidate probability for a fixed binomial count of
32 successes in 50 trials. Here the data remain fixed again, but the
response is continuous and the parameters are several regression
coefficients plus an error scale. For a declared independent Normal
linear model, the **likelihood** is the joint density of the observed
outcomes at their fixed predictor values, viewed as a function of those
parameters. It is not a probability assigned to a parameter. A continuous
density value also differs from the probability of an exact observed
vector. The **log-likelihood** adds log densities across the independent
records.

Write the predictor row for record $i$ as $x_i'$ and let $\sigma>0$ be the
error SD. Before fitting, the log-likelihood is

$$
\ell(\beta,\sigma)=-n\log\sigma-\frac n2\log(2\pi)
-\frac{1}{2\sigma^2}\sum_i(y_i-x_i'\beta)^2.
$$

At any fixed positive scale, only the last term changes with the
coefficients. Maximizing it over $\beta$ therefore minimizes the residual
sum of squares: the Normal maximum-likelihood coefficients equal the OLS
coefficients. This connects Chapter 7's squared-loss criterion to this
particular distributional model. Least squares itself did not require
that model to calculate a minimizing line.

Maximizing over scale gives
$\hat\sigma_{\mathrm{ML}}=\sqrt{RSS/n}$. The residual SD used for classical
coefficient uncertainty is $s=\sqrt{RSS/(n-p)}$, with $p$ counting the
regression coefficients. The two denominators serve different estimation
purposes; substitute the ML scale when checking the fitted likelihood.
At the fitted coefficients and ML scale,

$$
\ell(\hat\beta,\hat\sigma_{\mathrm{ML}})
=-\frac n2\left\{\log(2\pi)+1+\log(RSS/n)\right\}.
$$

R's `logLik()` returns this maximized value for an `lm()` fit by default.
We can check its meaning directly by adding Normal log densities with
the fitted means and maximum-likelihood scale. Two information criteria
combine this maximized log-likelihood with a parameter-count penalty:

$$
AIC=-2\log\hat L+2k,\qquad
BIC=-2\log\hat L+k\log n.
$$

Here $k$ counts the regression coefficients **and** the estimated residual
variance. For the full occupation model, six coefficient columns plus
that variance give $k=7$. Smaller values are preferred under each
criterion; neither is observed held-out MAE [@akaike1974; @schwarz1978].

```{r}
#| label: ch8b-likelihood-first
#| echo: true
ch8b_ll <- logLik(m2)
ch8b_n <- nobs(m2)
ch8b_k <- attr(ch8b_ll, "df")
ch8b_sigma_ml <- sqrt(mean(residuals(m2)^2))
ch8b_manual_ll <- sum(dnorm(d$income_k,
  mean = fitted(m2), sd = ch8b_sigma_ml, log = TRUE))
c(sigma_ML = ch8b_sigma_ml, residual_SD = sigma(m2),
  logLik = as.numeric(ch8b_ll), manual = ch8b_manual_ll, k = ch8b_k,
  AIC = -2 * as.numeric(ch8b_ll) + 2 * ch8b_k,
  BIC = -2 * as.numeric(ch8b_ll) + ch8b_k * log(ch8b_n))
```

\Needspace{7\baselineskip}

```{r}
#| label: ch8b-likelihood-criteria-check
#| echo: true
c(AIC_R = AIC(m2), BIC_R = BIC(m2))
```

The ML scale is 2.5373 and the residual SD is 2.6187. The two
log-likelihood calculations agree at $-230.3020$. The full model's
AIC is 474.6040 and BIC is 492.6988. The table compares the three declared
models using the same 98 rows, response scale, and likelihood convention.
The Computer Lab reference on common-profile averaging supplies the repeated
criterion-table construction. Chapter 11 develops likelihood fitting
further for binomial outcomes and numerical calculations; that later
treatment is not needed to read this local Normal-model calculation.

\Needspace{10\baselineskip}

| Model | k | R2 | Adjusted_R2 | AIC | BIC |
|:--|:--|--:|--:|--:|--:|
| m0 | 3 | .3296 | .3226 | 526.5052 | 534.2601 |
| m1 | 4 | .4996 | .4891 | 499.8328 | 510.1727 |
| m2 | 7 | .6362 | .6164 | 474.6040 | 492.6988 |

: In-sample summaries for comparable income models. {#tbl-ch8b-criteria}

The raw AIC/BIC values for income and log income must not be compared
as if they used the same response density. A scale change changes the
likelihood's reference measure; a valid common-scale comparison would
need the corresponding transformation adjustment. The simpler lesson
here is to preserve the outcome and question in each comparison.

No residual graph establishes independence. The occupations share
one historical labour market, and metadata do not establish a
probability sample of a wider population. For students within schools
or repeated observations, the design may require dependence-aware
uncertainty. HC3 alone is not such a repair.

The practical response is consequential rather than pass/fail: retain
the named comparison, report its form and influence sensitivities,
distinguish classical from robust uncertainty, and bound interpretation
by measurement, support, and source. Causal adjustment-set rules about
confounders, mediators, and colliders apply to a declared causal target;
they are not universal requirements for association or prediction.

### Extended practice: reproduction and critique (Integrative) {.unnumbered}

1. Compute the white-collar conditional profile (11, 45, 50) and the
   empirical-profile standardized value, with appropriately labelled
   intervals. Name each target and assess support. Propose different
   weights and explain which part of the target changes.
2. Rebuild the 102-to-98 audit and three-model sequence from a clean
   session. Preserve names, units, type levels, and source row order.
   Add one supported display and a bounded occupation-level report.
3. A school-level analysis reports a slope 0.40, classical SE 0.18,
   and HC3 SE 0.31. Its residual smooth curves, spread increases,
   deleting one influential school changes the slope to $-0.05$, and
   sectors have little prior-score overlap. Assess the claim:
   "The causal effect is robust because VIFs are below 10; deleting
   the unusual school will make the model reliable." Match each
   finding to its consequence and propose a transparent sensitivity.

   \Needspace{10\baselineskip}

4. Choose another documented public dataset with a continuous outcome,
   two continuous predictors, and a factor with at least three levels.
   Declare an additive question and analytic sample before fitting.
   Document source, licence, units, missingness, coding, support,
   conditional interpretation, a block comparison, checks, and one
   sensitivity. Read the reported reasons and answer, then separately
   evaluate them. Do not require every changed estimator to imply a
   changed scientific estimand. Record versions and stochastic seeds.

## References {.unnumbered}

::: {#refs}
:::
