---
title: "Reading Multiple Regression Across Coordinated Questions"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-cap: true
    tbl-cap: true
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 4, width = 68)
library(ggplot2)
theme_set(theme_minimal(base_size = 11))
all_occ <- carData::Prestige
all_occ$occupation <- rownames(all_occ)
all_occ$income_k <- all_occ$income / 1000
all_occ$type <- factor(all_occ$type, levels = c("bc", "prof", "wc"))
vars <- c("income_k", "education", "prestige", "women", "type")
d <- all_occ[complete.cases(all_occ[, vars]), ]
m0 <- lm(income_k ~ education, data = d)
m1 <- lm(income_k ~ education + prestige, data = d)
m2 <- lm(income_k ~ education + prestige + women + type, data = d)
profile <- data.frame(education = 11, prestige = 45, women = 50,
                      type = factor("wc", levels = levels(d$type)))
x_res <- residuals(lm(education ~ prestige + women + type, data = d))
y_res <- residuals(lm(income_k ~ prestige + women + type, data = d))
cv_compare <- function(data, fold) {
  pred <- matrix(NA_real_, nrow(data), 2)
  for (k in sort(unique(fold))) {
    train <- data[fold != k, ]
    test <- data[fold == k, ]
    stopifnot(length(unique(train$type)) == 3)
    fit0 <- lm(income_k ~ education, data = train)
    fit2 <- lm(income_k ~ education + prestige + women + type,
               data = train)
    stopifnot(fit0$rank == 2, fit2$rank == 6,
              nobs(fit0) == nrow(train), nobs(fit2) == nrow(train))
    pred[fold == k, 1] <- predict(fit0, newdata = test)
    pred[fold == k, 2] <- predict(fit2, newdata = test)
  }
  stopifnot(all(is.finite(pred)))
  error <- abs(pred - data$income_k)
  list(predictions = pred, absolute_errors = error,
       MAE = colMeans(error),
       difference = mean(error[, 2] - error[, 1]))
}
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
set.seed(80806)
fold <- sample(rep(seq_len(10), length.out = nrow(d)))
cv <- cv_compare(d, fold)
stopifnot(nrow(d) == 98, df.residual(m2) == 92,
          abs(cv$difference + .8618922149) < 1e-8)
```

\pagebreak

::: {.callout-tip title="Chapter box"}
**Central question:** How can one study answer marginal-association,
conditional-association and prediction questions without treating their
answers or supporting reasons as interchangeable?

**Focal concepts:** Common analytic records; partial slopes and indicators;
coefficient covariance; support, form and influence; fitted and standardized
quantities; held-out prediction error; separate arguments and their synthesis.

**Main evidence:** The 98 complete occupation records from
`carData::Prestige`, followed by Project STAR's kindergarten reading
comparison. Their different units and designs support different conclusions.

**Scope:** Chapter 7 supplies simple regression and Chapters 2–6 supply the
statistical and R foundations. Chapter 1 supplies the reading approach;
coordination is introduced locally and Chapter 1a remains optional. All nine
sections, including STAR, form the analytical route. Appendices A–B retain
optional algebra and technical reproduction. Reading comes before separate
quality evaluation; construction is an auxiliary use.
:::

**Core roadmap.** Section 1 separates three questions and their owed answers.
Sections 2--5 develop the conditional-association question from coefficient
meaning through estimation, checks, and selected fitted quantities. Section 6
asks the separate prediction question, and Section 7 reads, evaluates, and
synthesizes the three occupation arguments. Section 8 then transfers the
reading discipline to Project STAR's randomized design; Section 9 summarizes
the route. The appendices are not required to understand the conceptual core;
Appendix B is needed only to reproduce the complete all-fold and 200-resplit
computations.

# Three Questions in the Same Occupation Records {#sec-ch8-questions}

## One study can owe several answers

A report may ask whether education and income move together, whether that
relationship remains when other occupational characteristics are held
constant, and whether those characteristics help predict income. These
questions can inform one another, but answering one does not discharge the
other two obligations.

We use a **chapter-authored retrospective teaching study**, not a published
study with a newly discovered analysis plan. Its records are the
*Prestige* data distributed with the carData package. The analyses and short
report below are our teaching record; the package documentation supplies
the data descriptions, not the claims written for this chapter.

```{=latex}
\Needspace{20\baselineskip}
```

| Pair and role | Question | Estimand and owed answer |
|:--|:--|:--|
| **1. Contextual description** | How does income vary linearly with education across the 98 retained occupations? | The finite-record marginal least-squares slope; a descriptive answer for these records |
| **2. Primary conditional association** | How does income differ with education at the same prestige, percentage of women, and type, within support? | The linear conditional-mean coefficient $\beta_E$ under the declared working model; fitted $b_E$ is its estimate from the 98 retained records |
| **3. Secondary prediction analysis** | Does the full additive learning rule predict eligible same-source occupation records with lower expected absolute error than the education-only rule? | Difference in expected absolute error of the two learning procedures at the declared training sizes; an internal validation answer |

: Three separately owed answers in the occupation teaching study. {#tbl-ch8-pairs tbl-colwidths="[22,36,42]"}

These are three question--answer pairs, each with **one argument with linked
layers**. They are not three statistical strands of one argument: the report
owes separate conclusions about marginal association, conditional
association, and prediction. If several estimands were jointly necessary
for *one* integrated conclusion, they could instead be parallel statistical
strands. Merely printing several coefficients or diagnostics does not create
either several questions or several strands.

## Keep the records, time, and units visible

A row represents an **occupation**, not an individual worker. Income is the
occupation's average annual income in 1971 Canadian dollars; we divide it
by 1,000. Education is average years of education, and *women* is the
percentage of women in the occupation. Prestige ratings were collected in
a mid-1960s survey, whereas the income and education summaries come from
1971. Occupational type is blue-collar (*bc*), professional/managerial/
technical (*prof*), or white-collar (*wc*). These historical and aggregate
measurements are not interchangeable with current individual earnings.

The earlier simple-regression analysis used all 102 occupations with
education and income recorded. The declared full variable set leaves 98
complete records: athletes, newsboys, babysitters, and farmers have missing
type. The retained type counts are 44, 31, and 23 respectively. We keep the
original row order, which will also make the validation split reproducible.

```{=latex}
\Needspace{20\baselineskip}
```

```{r}
#| label: ch8-records
#| echo: true
all_occ <- carData::Prestige
all_occ$income_k <- all_occ$income / 1000
all_occ$type <- factor(all_occ$type, levels = c("bc", "prof", "wc"))
vars <- c("income_k", "education", "prestige", "women", "type")
d <- all_occ[complete.cases(all_occ[, vars]), ]
c(original = nrow(all_occ), retained = nrow(d))
table(d$type)
```

We refit the education-only model on these **same 98 records** before
comparing adjustment specifications. Moving from 102 to 98 changes which
records are described. It must not be presented as an adjustment effect.
Complete-case selection makes the computations comparable; it does not
establish that the omitted occupations are irrelevant or that these records
are a probability sample of Canadian occupations.

## Declare the model question before reporting an interval

Pair 1 is finite-record description. Its slope is exactly calculable once
the records and least-squares rule are specified.

Pair 2 has one intended estimand: the education coefficient $\beta_E$ in the
declared linear conditional-mean working model. The fitted coefficient $b_E$
is its estimate from the 98 retained records. Although $b_E$ is exactly
calculable once the rows and fitting rule are fixed, that fact does not create
a second finite-record estimand. A confidence interval additionally posits
that occupation outcomes could vary independently around the stated linear
conditional mean at the recorded predictors. This is not a claim that the
dataset came from a verified random sample or that income was randomly
assigned.

For the classical interval we additionally assume a common error variance;
normal errors make its finite-sample $t$ reference exact. Robust uncertainty
relaxes the common-variance requirement under its own large-sample
conditions. It does not create independence or linearity. We will check how
credible and stable the working-model reading is before judging its answer.

A **best linear projection** can instead summarize a curved conditional
relationship under a specified distribution of predictors. That is a
different interpretation, not a rescue silently adopted after discovering
curvature. The core uses the stated linear conditional-mean interpretation;
Appendix B explains the alternative and its limitations.

Pair 3 will declare a different hypothetical repetition: train a rule on
eligible, exchangeable same-source records, then predict a new record
without using its outcome in training. Its justification cannot be supplied
by the coefficient interval in Pair 2.

# From Marginal to Conditional Comparisons {#sec-ch8-comparisons}

## What a partial slope compares

Let $e$ be education, $p$ prestige, $w$ percentage of women, and $t$ type.
The fitted additive equation is

$$
\widehat m(e,p,w,t)
 =b_0+b_Ee+b_Pp+b_Ww
   +b_{\mathrm{prof}}D_{\mathrm{prof}}
   +b_{\mathrm{wc}}D_{\mathrm{wc}}.
$$

An indicator $D$ is 1 for its named category and 0 otherwise. Blue-collar is
the reference type, with both indicators zero. This model has six
coefficients, including its intercept.

Take two profiles with the same $p,w,t$ and education differing by one year.
Subtract their fitted means:

$$
\begin{aligned}
&\widehat m(e+1,p,w,t)-\widehat m(e,p,w,t)\\
&\qquad=b_E(e+1)-b_Ee=b_E.
\end{aligned}
$$

All the common terms cancel. Thus $b_E$ compares modeled occupation-level
income at the **same prestige, percentage of women, and type**. A two-year
difference corresponds to $2b_E$ under this additive model.

“Holding constant” describes this mathematical comparison. It does not say
that education was changed experimentally, that two observed occupations
match exactly on every other characteristic, or that unmeasured differences
are controlled. An additive slope also assumes the same education
comparison across the included predictor values. Allowing it to vary with
another predictor is the interaction question taken up later.

## Why the marginal slope can differ

A short known-process example makes the distinction visible. Suppose

$$
Y=10+2X+5Z+\varepsilon,
\qquad E(\varepsilon\mid X,Z)=0,
$$

with finite variances and
$\operatorname{Cov}(X,Z)/\operatorname{Var}(X)=0.65$.
The conditional $X$ coefficient is 2. The marginal linear slope is

$$
\frac{\operatorname{Cov}(X,Y)}{\operatorname{Var}(X)}
=2+5\frac{\operatorname{Cov}(X,Z)}{\operatorname{Var}(X)}
=2+5(0.65)=5.25.
$$

The marginal comparison incorporates how $Z$ moves with $X$ in this
specified process. The conditional comparison holds $Z$ fixed. Neither
arithmetic result alone says what would happen if a researcher intervened
on $X$. Calling 5.25 “biased” requires naming an estimand for which it is
being used incorrectly; as a marginal linear summary, 5.25 is the right
answer here. Appendix A retains the derivation and simulation.

More generally, adding predictors can change the comparison, but it does
not **always** change the scientific estimand. In a randomized study,
appropriately specified covariate adjustment may estimate the same declared
average assignment effect more precisely. Releveling a factor changes
coefficient coordinates without changing fitted values. A different
uncertainty estimator may change the interval without changing the fitted
coefficient.

## A small indicator bridge

The full occupation fit, rounded for substitution, is

$$
\begin{aligned}
\widehat{\mathrm{income}}_k
={}&0.420524+0.116434e+0.140605p-0.053505w\\
 &+0.324193D_{\mathrm{prof}}+0.235331D_{\mathrm{wc}}.
\end{aligned}
$$

Consider a white-collar profile with education 11, prestige 45, and women 50.
Here $D_{\mathrm{prof}}=0$ and $D_{\mathrm{wc}}=1$:

$$
\begin{aligned}
\widehat{\mathrm{income}}_k
={}&0.420524+0.116434(11)+0.140605(45)\\
 &-0.053505(50)+0+0.235331\\
\approx{}&5.589.
\end{aligned}
$$

That is a fitted mean of about 5,589 dollars on the historical income scale,
not a guarantee for any occupation or worker. Can these three predictor
values occur together within the observed white-collar covariate cloud?
Separate minimum and maximum values cannot answer that question.

Here is a checkable **joint-support witness**. A convex combination is a
weighted average with nonnegative weights summing to one. Four retained
white-collar records can be combined to recover education 11, prestige 45,
and women 50 simultaneously. The source spelling *receptionsts* is retained
in the code so that the record can be located exactly.

Use the supplied weights to check the three weighted averages. The order of
the weights matches the four named records. `weighted.mean()` performs the
same weighted-average calculation for each variable; finding these weights
is not a prerequisite for interpreting their evidence.

\Needspace{23\baselineskip}

```{r}
#| label: ch8-profile-support
#| echo: true
support_names <- c("radio.tv.announcers", "tellers.cashiers",
                   "receptionsts", "sales.supervisors")
support_rows <- d[support_names, c("education", "prestige", "women")]
support_weights <- c(0.245243736035, 0.230972611567,
                     0.226143656937, 0.297639995461)
d[support_names, "type"]
sum(support_weights)
c(education = weighted.mean(support_rows$education, support_weights),
  prestige = weighted.mean(support_rows$prestige, support_weights),
  women = weighted.mean(support_rows$women, support_weights))
```

\Needspace{11\baselineskip}

```{r}
#| label: tbl-ch8-profile-support
#| echo: false
#| tbl-cap: "Four observed white-collar records recover the declared profile."
support_display <- data.frame(
  Occupation = c("Radio/TV announcers", "Tellers/cashiers",
                 "Receptionists", "Sales supervisors"),
  support_rows, Weight = unname(support_weights), row.names = NULL
)
knitr::kable(support_display, digits = c(0, 2, 1, 2, 6))
```

The supplied weights are all positive and sum to one; their weighted
coordinates recover $(11,45,50)$ within $10^{-10}$ using the displayed
precision. The table rounds weights for reading. This establishes membership in the joint
covariate hull **within white-collar type**, using all three continuous
predictors in the full model. It does not show that many occupations lie
near this profile, that observed occupations match exactly, or that a
causal comparison is supported. Local data may still be sparse.

The intercept refers to zero education, prestige, and women in the
reference type. It helps construct the fitted surface; it need not describe
a plausible occupation. Likewise, one type coefficient is not “the effect
of occupational type” as a whole. Appendix A retains the full factor,
group-mean, contrast-covariance, and classical ANOVA bridge. That material
extends the compact indicator calculation without becoming the main route.

# Estimate the Conditional Association {#sec-ch8-estimate}

## Compare specifications on the same rows

```{r}
#| label: ch8-fits
#| echo: true
m0 <- lm(income_k ~ education, data = d)
m1 <- lm(income_k ~ education + prestige, data = d)
m2 <- lm(income_k ~ education + prestige + women + type, data = d)
coef(m2)
```

| Records and specification | Education slope | In-sample $R^2$ | Role here |
|:--|--:|--:|:--|
| Earlier 102, education only | 0.899 | .334 | Earlier result; not the adjustment baseline |
| Same 98, education only | 0.883 | .330 | Pair 1's marginal description |
| Same 98, education and prestige | -0.218 | .500 | Intermediate comparison |
| Same 98, full declared predictors | 0.116 | .636 | Pair 2's estimate of $\beta_E$ |

: The model sequence separates sample composition from adjustment. {#tbl-ch8-sequence}

*Note.* Slopes are thousand 1971 Canadian dollars per education year.
These are fitted summaries, not causal effects. No sampling interval is
attached to Pair 1's finite-record estimand.

Education and prestige are highly correlated here, $r=.866$. Once prestige
is included, the education comparison uses variation not linearly accounted
for by prestige. Adding women and type changes the conditioning set again.
The movement from .883 to -.218 to .116 is not a narrative in which a true
effect is successively “explained away.” It is a sequence of named
comparisons in a small observational record.

The increase in $R^2$ describes training-data fit. Adding predictors cannot
increase the least-squares residual sum of squares on the same rows when
the models are nested. Better training fit does not establish better
prediction on records left out of fitting.

## Partial regression gives the same coefficient

Multiple regression still uses the covariance reasoning from simple
regression. First regress education on the other predictors and retain its
residuals $r_{E,i}$. Do the same for income, obtaining $r_{Y,i}$. Both
auxiliary regressions include an intercept. Then

$$
b_E=\frac{\sum_i r_{E,i}r_{Y,i}}{\sum_i r_{E,i}^2}.
$$

The numerator measures remaining linear co-variation; the denominator
measures remaining education variation. This formula is the partial
regression identity, not a new estimation method with a new question.

```{r}
#| label: ch8-partial
#| echo: true
r_E <- residuals(lm(education ~ prestige + women + type, data = d))
r_Y <- residuals(lm(income_k ~ prestige + women + type, data = d))
sum(r_E * r_Y) / sum(r_E^2)
cor(r_E, r_Y)
```

The slope is 0.1164 and the partial correlation is .0441. The latter is
unitless; it is not the slope expressed in dollars. Both summarize the same
residualized comparison in different units.

```{r}
#| label: fig-ch8-partial
#| fig-pos: H
#| fig-cap: "Marginal and residualized income--education comparisons in the same 98 occupations."
#| fig-alt: "The marginal scatterplot has a positive fitted line. After linearly removing prestige, women, and type from both variables, the fitted line is nearly flat and substantial residual variation remains."
#| fig-width: 6.4
#| fig-height: 3.5
oldpar <- par(mfrow = c(1, 2), mar = c(4.3, 4.2, 1, .5), mgp = c(2.7, .8, 0))
plot(d$education, d$income_k, pch = 1, xlab = "Education (years)",
     ylab = "Income (thousand dollars)")
abline(m0, col = "#245C78", lwd = 2)
plot(r_E, r_Y, pch = 1, xlab = "Residual education (years)",
     ylab = "Residual income (thousand dollars)")
abline(a = 0, b = coef(m2)["education"], col = "#245C78", lwd = 2)
par(oldpar)
```

Residualization does not remove all substantive differences between
occupations. It removes the **fitted linear components** of the named
variables. Appendix A develops the matrix version.

## Read a coefficient interval under its stated model

The hypothetical linear conditional-mean model introduced in
@sec-ch8-questions
has six estimated coefficients, so its residual degrees of freedom are
$98-6=92$. For education the classical calculation is

$$
\begin{aligned}
b_E\pm t_{.975,92}SE(b_E)
 &=0.1164\pm1.9861(0.2752)\\
 &\approx[-0.430,\ 0.663].
\end{aligned}
$$

The result is compatible, under those assumptions, with negative and
positive conditional slopes on this scale. It does not demonstrate that
education is unimportant, that the slope equals zero, or that adjustment
has removed an education effect. No substantive equivalence margin was
declared, and the interval includes differences of several hundred dollars
per year of education. The data and model therefore leave the direction
and practically relevant magnitude unresolved.

```{r}
#| label: ch8-uncertainty
#| echo: true
confint(m2, "education")
hc3_covariance <- sandwich::vcovHC(m2, type = "HC3")
hc3_se <- sqrt(hc3_covariance["education", "education"])
coef(m2)["education"] + c(-1, 1) * qt(.975, 92) * hc3_se
```

HC3 gives SE 0.3823 and an approximate interval $[-0.643,0.876]$ using
the displayed $t_{92}$ convention. It allows heteroskedasticity under
independent-observation and other regularity conditions; it is not an exact
small-sample guarantee. Both intervals leave broad uncertainty. Neither
accounts for unmodeled dependence, selection, measurement error, or a wrong
conditional-mean form.

If a test is reported, its target belongs with this inference: the
two-sided education-coefficient test addresses $H_0:\beta_E=0$ in the stated
full model, not “education has no effect.” Its classical $t=0.423$ and
$p=.673$ do not prove the null. We do not use this test to choose which of
the three questions deserves an answer.

\Needspace{12\baselineskip}

# Check Support, Form, Influence, and Precision {#sec-ch8-check}

Checking asks what the model leaves unexplained and what follows for the
answer. It is not a ritual in which each graph receives a pass or fail.
Residuals, support, coefficient precision, and sensitivity each answer a
different diagnostic question.

## Read residual, scale, tail, and influence views together

A residual is an observed deviation from a fitted value, not the unobserved
population error. A standardized residual uses a common residual-scale
estimate; an externally studentized residual uses a scale estimate with that
observation omitted. Leverage measures an unusual predictor configuration.
Influence combines predictor and outcome information [@cook1977].

```{r}
#| label: ch8-diagnostics
#| echo: true
ch8_diag <- data.frame(
  occupation = rownames(d), fitted = fitted(m2),
  residual = residuals(m2), standardized = rstandard(m2),
  studentized = rstudent(m2), leverage = hatvalues(m2),
  cooks = cooks.distance(m2)
)
ch8_top <- order(ch8_diag$cooks, decreasing = TRUE)[1:5]
```

```{r}
#| label: fig-ch8-diagnostics
#| echo: false
#| fig-cap: "Four complementary diagnostic views for the full additive model."
#| fig-alt: >
#|   Residual and scale plots show structure and changing spread.
#|   A normal quantile plot departs in the right tail. The final
#|   panel relates leverage and studentized residuals, with point
#|   sizes indicating Cook's distance.
#| fig-width: 6.7
#| fig-height: 6.2
local({
  old <- par(no.readonly = TRUE)
  on.exit(par(old))
  par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
  z <- ch8_diag
  plot(z$fitted, z$residual, pch = 1,
       xlab = "Fitted income", ylab = "Residual",
       main = "(a) Mean pattern")
  abline(h = 0, lty = 2)
  lines(lowess(z$fitted, z$residual), lwd = 2)
  plot(z$fitted, sqrt(abs(z$standardized)), pch = 1,
       xlab = "Fitted income", ylab = "Sqrt |standardized|",
       main = "(b) Scale pattern")
  lines(lowess(z$fitted, sqrt(abs(z$standardized))), lwd = 2)
  qqnorm(z$studentized, pch = 1, main = "(c) Normal reference")
  qqline(z$studentized, lwd = 2)
  plot(z$leverage, z$studentized, pch = 1,
       cex = .8 + 2 * sqrt(z$cooks / max(z$cooks)),
       xlab = "Leverage", ylab = "Studentized residual",
       main = "(d) Influence")
  abline(h = 0, lty = 2)
})
```

The smooth and changing spread challenge a constant conditional-mean and
variance description. The right tail matters to exact small-sample
normal-theory inference. Neither feature prevents calculation of OLS. Large
residual, high leverage, and high influence are different features; none is
an automatic record-invalidity verdict. No residual graph can establish that
occupations are independent draws from a wider population.

## Inspect joint support and coefficient precision

Joint support concerns predictor combinations, not merely separate minimum
and maximum values. The convex-combination witness in
@sec-ch8-comparisons establishes support for the focal white-collar
profile across all three continuous predictors. The marginal ranges below ask
a different, coarser question: how far does each variable extend within each
recorded type?

```{=latex}
\Needspace{8\baselineskip}
```

```{r}
#| label: tbl-ch8-support
#| echo: false
#| tbl-cap: "Predictor ranges within each recorded occupation type."
ch8_ranges <- do.call(rbind, lapply(split(d, d$type), function(g) {
  data.frame(n = nrow(g),
    Education = sprintf("%.2f--%.2f",
                        min(g$education), max(g$education)),
    Prestige = sprintf("%.1f--%.1f",
                       min(g$prestige), max(g$prestige)),
    Women = sprintf("%.2f--%.2f", min(g$women), max(g$women)))
}))
knitr::kable(ch8_ranges)
```

Blue-collar education ends at 10.93 years, below the professional minimum
of 11.09. A same-education comparison between those types necessarily crosses
that gap. A model can return both predictions there; software output does not
create observations. Lack of support limits what can be learned or interpreted
about a defined comparison rather than making the estimand cease to exist.

A conditional grid can also avoid pretending that an unrestricted rectangle
is supported. For *m1*, hold prestige at 30, 50, or 70 and plot education only
over values observed within 7.5 prestige points of each target. The width is
an explicit display convention, not a formal support guarantee.

```{r}
#| label: ch8-grid
#| echo: true
ch8_grid <- data.frame()
for (p0 in c(30, 50, 70)) {
  nearby <- d[abs(d$prestige - p0) <= 7.5, ]
  one_grid <- data.frame(
    education = seq(min(nearby$education), max(nearby$education),
                    length.out = 40),
    prestige = p0, Profile = paste("Prestige", p0)
  )
  ch8_grid <- rbind(ch8_grid, one_grid)
}
ch8_grid$fit <- predict(m1, newdata = ch8_grid)
```

```{r}
#| label: fig-ch8-grid
#| echo: false
#| fig-cap: "Neighborhood-restricted fitted profiles from the education-and-prestige model."
#| fig-alt: >
#|   Three gently decreasing parallel lines represent fitted income
#|   across education at prestige 30, 50, and 70. Their education
#|   ranges differ, reflecting the declared neighborhoods.
#| fig-width: 6.1
#| fig-height: 3.5
ggplot2::ggplot(ch8_grid,
  ggplot2::aes(education, fit, linetype = Profile)) +
  ggplot2::geom_line(linewidth = .8) +
  ggplot2::labs(x = "Education (years)",
    y = "Fitted income (thousand 1971 dollars)") +
  ggplot2::theme(legend.position = "bottom")
```

The parallel lines share *m1*'s education slope. This display conditions on
prestige only, not on the full set for Pair 2. It teaches a transparent plotting
choice; it does not prove support for the full-model profile.

Education's variance inflation factor makes the collinearity calculation
checkable:

$$
VIF_E=\frac{1}{1-R^2_{E\mid P,W,T}}
     =\frac{1}{1-.876495}\approx8.097.
$$

Under the classical model, holding residual variance and total education
variation fixed, its coefficient SE is multiplied by
$\sqrt{8.097}\approx2.846$ relative to an orthogonal design. This is a
comparison of variance formulas, not a claim that deleting other predictors
will divide the actual SE by 2.846. Deletion changes the comparison and can
change residual variance. VIF is not an automatic deletion threshold, and
near collinearity does not by itself create bias under the maintained
zero-conditional-mean model.

## Use targeted sensitivities to bound the answer

A component-plus-residual plot adds the fitted prestige contribution back to
each residual, $e_i+\hat\beta_P P_i$. The smooth bends upward at high prestige
relative to the straight fitted component. That pattern motivates a declared
form sensitivity; it does not prove that one chosen curve is the true process.

```{r}
#| label: fig-ch8-partial-residual
#| echo: false
#| fig-cap: "The prestige component plus residual in the full additive model."
#| fig-alt: >
#|   A component-plus-residual scatterplot has a positive trend.
#|   A curved smooth bends upward at high prestige relative to the
#|   straight fitted component.
#| fig-width: 6.1
#| fig-height: 3.5
ch8_partial <- data.frame(
  prestige = d$prestige,
  component = residuals(m2) + coef(m2)["prestige"] * d$prestige
)
ggplot2::ggplot(ch8_partial, ggplot2::aes(prestige, component)) +
  ggplot2::geom_point(shape = 1) +
  ggplot2::geom_smooth(method = "lm", se = FALSE,
                      colour = "black", linetype = 2) +
  ggplot2::geom_smooth(method = "loess", se = FALSE,
                      colour = "black") +
  ggplot2::labs(x = "Prestige", y = "Component plus residual")
```

\Needspace{26\baselineskip}

```{r}
#| label: ch8-form-sensitivity
#| echo: true
ch8_curve <- update(m2, . ~ . + I(prestige^2))
anova(m2, ch8_curve)
c(full_education = unname(coef(m2)["education"]),
  curve_education = unname(coef(ch8_curve)["education"]))
```

The classical nested comparison gives $F(1,91)=6.736$ and $p=.011$;
education changes from 0.116434 to -0.029800. In the enlarged equation,
the prestige slope depends on prestige through
$\beta_P+2\beta_{P^2}P$. This is a focused sensitivity for Pair 2, not a
validation-tuned replacement for the frozen prediction rule in Pair 3.

The record for general managers has the largest Cook's distance, about 0.447. Its high
income relative to the model is not proof of error. A transparent deletion
refit asks how dependent the answer is on that record:

```{r}
#| label: ch8-influence-sensitivity
#| echo: true
ch8_without_top <- lm(
  income_k ~ education + prestige + women + type,
  data = d[-ch8_top[1], ]
)
c(full_education = unname(coef(m2)["education"]),
  omit_one_education = unname(coef(ch8_without_top)["education"]))
```

The education estimate moves to 0.437189. Preserve the full-data estimate and
disclose the sensitivity. Verify coding, measurement, and substantive
distinctiveness before considering exclusion; a more attractive $p$ value is
not an exclusion rule. The refit is a diagnostic perturbation, not a new
research question or evidence that the reduced sample is preferable.

```{=latex}
\Needspace{42\baselineskip}
```

| Check | What is visible here | Consequence for the answer |
|:--|:--|:--|
| Functional form | Adding a prestige-squared term changes the education slope to about -0.030 | Treat the additive form as a working approximation; do not silently reinterpret it as an exact conditional law |
| Spread | HC3 raises the education SE from .275 to .382 | Show sensitivity; a robust SE does not repair all model problems |
| Influence | Removing the record for general managers, which has the largest Cook's distance, gives an education slope about .437 | Inspect the record and disclose sensitivity; do not delete it merely to obtain a preferred result |
| Collinearity | Education's auxiliary $R^2$ is .8765 | Little education variation remains after the other predictors; separate conditional coefficients can be imprecise |
| Joint support | Professional and blue-collar education ranges do not overlap | Cross-type comparisons at common covariates may depend heavily on extrapolation |
| Dependence and scope | Occupational aggregates have shared economic and classification contexts | Independence and same-source transport are assumptions, not verified by a residual plot |

: Checks change the bounded answer rather than producing one omnibus pass/fail score. {#tbl-ch8-checks tbl-colwidths="[19,38,43]"}

Appendix B retains the full Cook's distance and deletion accounting, manual
HC3 covariance reproduction, generalized VIF, changed-outcome models, and
model-comparison criteria. These are technical extensions of checks already
interpretable here, not prerequisites hidden outside the numbered route.

# Move from Coefficients to Fitted and Standardized Quantities {#sec-ch8-quantities}

The profile and standardization calculations in this section are **didactic
extensions of the fitted model**, not extra question-facing answers owed by
the occupation study. Pair 2 remains about $\beta_E$. A report that promoted
one of these quantities to its own substantive conclusion would need to state
the corresponding question, estimand, justification, and bounds.

## A fitted mean is not prediction performance

Start with a complete input row and ask R for the two interval types.
`interval = "confidence"` concerns the model mean at that profile;
`interval = "prediction"` concerns one new outcome at the same profile.
Both use the full fitted model and its stated classical assumptions.

\Needspace{17\baselineskip}

```{r}
#| label: ch8-profile-predict-check
#| echo: true
profile <- data.frame(education = 11, prestige = 45, women = 50,
                     type = factor("wc", levels = levels(d$type)))
mean_check <- predict(m2, newdata = profile, interval = "confidence")
new_check <- predict(m2, newdata = profile, interval = "prediction")
mean_check
new_check
```

For the white-collar profile in @sec-ch8-comparisons, collect the
coefficient multipliers in $x_*=(1,11,45,50,0,1)'$. Their order is intercept,
education, prestige, women, professional indicator, and white-collar indicator.
The fitted mean and its estimated classical variance are below. A prime denotes
transpose (`t()` in R); `vcov(m2)` is the estimated covariance matrix of the
fitted coefficients, not the covariance matrix of the observed variables.

$$
\widehat m_*=x_*'\hat\beta,\qquad
\widehat{\operatorname{Var}}(\widehat m_*)=x_*'\widehat Vx_*,
\quad \widehat V=\operatorname{vcov}(m2).
$$

The variance includes the coefficient variances **and their covariances**.
For example, two terms $a\hat\beta_j+b\hat\beta_k$ contribute
$a^2\widehat{\operatorname{Var}}(\hat\beta_j)+
b^2\widehat{\operatorname{Var}}(\hat\beta_k)+
2ab\widehat{\operatorname{Cov}}(\hat\beta_j,\hat\beta_k)$.
Adding coefficient SEs, or ignoring covariance, is not the same calculation.
Take the square root only after calculating the variance.

```{=latex}
\Needspace{19\baselineskip}
```

```{r}
#| label: ch8-profile-covariance
#| echo: true
x_star <- c(1, 11, 45, 50, 0, 1)
V <- vcov(m2)
profile_fit <- sum(x_star * coef(m2))
profile_var <- drop(t(x_star) %*% V %*% x_star)
s2 <- sum(residuals(m2)^2) / df.residual(m2)
critical <- qt(.975, df.residual(m2))
profile_mean_ci <- profile_fit +
  c(-1, 1) * critical * sqrt(profile_var)
profile_new_pi <- profile_fit +
  c(-1, 1) * critical * sqrt(s2 + profile_var)
rbind(fitted_mean = profile_mean_ci, new_case = profile_new_pi)
```

The fitted mean is 5.588634; its estimated variance is 0.307962 and
SE is $\sqrt{0.307962}=0.554943$, in thousand-dollar units for the SE.
Variance has squared units. The classical fitted-mean interval is therefore
$[4.486,6.691]$ thousand dollars.

For **one new outcome**, its own error variation adds $s^2=6.857497$:
the SE is $\sqrt{s^2+x_*'\widehat Vx_*}$, giving the much wider interval
$[0.272,10.905]$. This calculation uses the independent normal
common-variance conditional-mean model; it is not a promise established by
the diagnostics. Replacing $\widehat V$ with HC3 alone does not estimate an
arbitrary new outcome's conditional error distribution.

Neither interval tells us how accurately a learning rule predicts held-out
records on average. That requires Pair 3.

## Standardize over a named distribution

A conditional fitted value uses one specified profile. A standardized fitted
value averages across a named distribution. To see the operation, copy the
98 observed profiles, change their type to white-collar, predict each
fitted value, then take the mean. Each continuous profile has equal weight.

\Needspace{11\baselineskip}

```{r}
#| label: ch8-standardization-first
#| echo: true
white_profiles <- d
white_profiles$type <- factor("wc", levels = levels(d$type))
white_predictions <- predict(m2, newdata = white_profiles)
mean(white_predictions)
```

This is an average of model predictions after a specified profile change,
not the observed white-collar group mean. For type $g$ and fixed
nonnegative weights $w_i$ summing to one,

$$
\hat\mu_g^{\mathrm{std}}=\sum_i w_i x_{ig}'\hat\beta
=a_g'\hat\beta,\qquad a_g=\sum_i w_i x_{ig}.
$$

Here $x_{ig}$ is the design vector after assigning type $g$ to observed
continuous profile $i$. With empirical weights $w_i=1/98$, this is
standardization over these 98 profiles, not over a national population.
Treating those profiles and weights as fixed gives estimated variance
$a_g'\widehat V a_g$.

The supplied loop repeats the same change--predict--average operation for
each type. `model.matrix()` gives the six coefficient multipliers for every
changed profile; `colMeans()` averages each multiplier to obtain $a_g$.
The variance calculation includes covariance between all predictions through
their shared fitted coefficients. Averaging individual interval endpoints
does not give an interval for this standardized mean.

\Needspace{29\baselineskip}

```{r}
#| label: ch8-standardization
#| echo: true
ch8_std <- data.frame(Type = levels(d$type), Estimate = NA_real_,
                      SE = NA_real_)
V <- vcov(m2)
for (j in seq_len(nrow(ch8_std))) {
  changed_profiles <- d
  changed_profiles$type <- factor(
    ch8_std$Type[j], levels = levels(d$type)
  )
  ch8_std$Estimate[j] <- mean(predict(m2, newdata = changed_profiles))
  X <- model.matrix(~ education + prestige + women + type,
                    data = changed_profiles)
  a <- colMeans(X)
  ch8_std$SE[j] <- sqrt(drop(t(a) %*% V %*% a))
}
critical <- qt(.975, df.residual(m2))
ch8_std$Lower <- ch8_std$Estimate - critical * ch8_std$SE
ch8_std$Upper <- ch8_std$Estimate + critical * ch8_std$SE
```

```{=latex}
\Needspace{8\baselineskip}
```

```{r}
#| label: tbl-ch8-standardized
#| echo: false
#| tbl-cap: "Empirical-profile standardization under the full additive model."
knitr::kable(ch8_std, digits = 3, row.names = FALSE)
```

Because *m2* is linear and additive in the continuous predictors, its average
fitted value equals its fitted value at their means. This is an algebraic
special case: with squared terms, interactions, or a nonlinear response
function, prediction at means need not equal mean prediction. Constant
additive type contrasts remain unchanged by common weights, although absolute
standardized levels can change.

The overall education mean is 10.795 years. Relabelling every profile
professional or white-collar creates some unsupported combinations. Averaging
them does not solve that problem. The white-collar profile used above is a
different, locally supported conditional quantity; it is not a three-type
comparison. The intervals condition on the empirical profiles and use the
classical covariance. Estimating a new population's covariate distribution
would add uncertainty not included here.

Standardization does not identify a causal effect by itself and does not
validate out-of-sample performance. Pair 3 requires a learning rule and
held-out outcomes.

# Ask and Answer the Separate Prediction Question {#sec-ch8-prediction}

## Define the learning rule, new case, and loss

A learning rule says how training data become a fitted prediction equation.
Our two rules are fixed: fit education-only OLS, or fit the full additive
OLS equation with education, prestige, women, and type. Both use the same
outcome and records. OLS **fits** by minimizing squared training errors;
we separately choose **mean absolute error (MAE)** to evaluate the
resulting predictions in interpretable income units.

A prediction opportunity here means a hypothetical same-source occupation
with all declared predictors available before its income outcome is used.
It does not mean predicting future earnings from information that would
only become available afterward. The records do not establish that this
hypothetical exchangeability assumption holds for new countries, periods,
individual workers, or occupational types absent from the dictionary.

For held-out outcomes $y_i$ and predictions $\widehat y_i^{(-)}$ fitted
without those outcomes,

$$
MAE=\frac{1}{n_{\mathrm{test}}}
       \sum_{i\in\mathrm{test}}\left|y_i-\widehat y_i^{(-)}\right|.
$$

The superscript reminds us that the case was held out of fitting. A smaller
number means smaller average absolute prediction error on the evaluated
cases. It does not say that every case was predicted better.

## Keep held-out outcomes out of training

We use shared ten-fold cross-validation. The 98 records are divided into
ten folds; each fold takes a turn as the held-out set. Fit both equations
on the other records, predict the held-out cases, then combine the 98
case-level errors.

\Needspace{24\baselineskip}

| Decision | Frozen teaching protocol | Why it matters |
|:--|:--|:--|
| Candidates | Education-only and the declared full additive OLS rule | No search for the best-looking model after validation |
| Folds | Shared balanced random ten-fold assignment, seed 80806 | Both rules face the same held-out cases |
| Training sizes | Eight fits use 88 records; two use 89 | Risk concerns those training sizes, not a model trained on all 98 |
| Preparation | Same 98 complete records; fixed income/1,000 conversion and type dictionary | No outcome-informed preprocessing; selection still limits scope |
| Estimability | Every training set must contain all three types and have full column rank | Stop rather than silently substitute a different rule |
| Summary | Pool absolute errors over all 98 held-out cases | Each case receives equal weight; folds of different size do not |
| Sensitivity | 200 new resplits of the same records, seed 80807 | Describes sensitivity to splitting, not new independent studies |

: The internal-validation protocol was fixed before this teaching comparison. {#tbl-ch8-validation tbl-colwidths="[20,42,38]"}

This is a retrospective teaching reanalysis of familiar historical data.
“Fixed before this comparison” is not a claim of historical preregistration
or that analysts had never seen the outcomes. The candidates are not revised
after inspecting the primary errors; the association sensitivities in
@sec-ch8-check are not covert additions to the validated candidate set.

\Needspace{30\baselineskip}

```{r}
#| label: ch8-one-fold
#| echo: true
set.seed(80806)
fold <- sample(rep(seq_len(10), length.out = nrow(d)))
held_fold <- fold[1]
train <- d[fold != held_fold, ]
held <- d[fold == held_fold, ]
fold_simple <- lm(income_k ~ education, data = train)
fold_full <- lm(income_k ~ education + prestige + women + type,
                data = train)
table(train$type)
c(training = nrow(train), held_out = nrow(held),
  simple_rank = fold_simple$rank, full_rank = fold_full$rank)
first_case <- d[1, , drop = FALSE]
first_predictions <- c(
  Education = unname(predict(fold_simple, first_case)),
  Full = unname(predict(fold_full, first_case))
)
rbind(prediction = first_predictions,
      absolute_error = abs(first_predictions - first_case$income_k))
```

The check should show all three types, 89 training records and nine held-out
records. The fitted ranks are 2 and 6, the numbers of coefficient columns.
Proceed with the declared comparison only when its training data can estimate
both rules. The full reproduction in Appendix B checks this for every fold.

This is one fold of the frozen protocol, with 89 training records and 9
held-out records. Government administrators, the first source record, is
in fold 9 and was excluded from both fits. The displayed code predicts
that record; repeat the same train/predict/error steps across all ten folds
to obtain the all-98 comparison below. One record's error is not that summary.

Appendix B provides the complete ten-fold function and 200-resplit code as
optional computational reproduction. The visible one-fold example needs
only the preceding core data definitions, not that helper. No held-out
outcome is used to estimate a coefficient. If we introduced learned scaling,
imputation, or variable selection, each step would have to be fitted using
training data only. A fixed unit conversion such as dollars to thousands is
not an estimated preprocessing step.

```{r}
#| label: fig-ch8-folds
#| fig-cap: "Shared held-out cases let two fixed learning rules face the same prediction task."
#| fig-alt: "A training-record box branches to education-only OLS and full OLS. Both arrows lead to predictions for the same held-out records, then to paired absolute errors. The diagram is repeated over ten folds."
#| fig-width: 6.3
#| fig-height: 3.3
par(mar = c(.2, .2, .2, .2))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 5))
boxtext <- function(x1, y1, x2, y2, label, fill = "#EFF4F7") {
  rect(x1, y1, x2, y2, col = fill, border = "#426278")
  text((x1 + x2)/2, (y1 + y2)/2, label, cex = .9)
}
boxtext(.1, 1.8, 2.3, 3.2, "Training records\n88 or 89")
boxtext(3.1, 3.1, 6.2, 4.4, "Education-only OLS")
boxtext(3.1, .6, 6.2, 1.9, "Full additive OLS")
boxtext(7, 1.8, 9.9, 3.2, "Same held-out cases\nPaired absolute errors")
arrows(2.3, 2.8, 3.1, 3.6, length = .08)
arrows(2.3, 2.2, 3.1, 1.4, length = .08)
arrows(6.2, 3.6, 7, 2.8, length = .08)
arrows(6.2, 1.4, 7, 2.2, length = .08)
text(5, .12, "Repeat across ten folds; pool the 98 case-level errors.", cex = .9)
```

## Work one error, then read the comparison

For government administrators, the first source record, observed income is
12.351 thousand dollars. Its held-out fold is 9. The full-rule prediction
is 11.594819, giving

$$
|12.351-11.594819|=0.756181.
$$

The education-only prediction is 8.946130, giving
$|12.351-8.946130|=3.404870$.
This one record favors the full rule. We still need all 98 errors:

| Learning rule | Held-out MAE (thousand dollars) |
|:--|--:|
| Education-only OLS | 2.403 |
| Full additive OLS | 1.541 |
| Difference: full minus education-only | -0.862 |

: Case-weighted absolute errors from the frozen shared-fold comparison. {#tbl-ch8-mae}

The full rule's average held-out absolute error is about 862 dollars lower
in this split. Because we subtract the two errors for **each same case**
before averaging, difficulty differences between validation cases do not
come from using different test sets for the two rules.

What is the performance estimand? Eight test folds contain 10 cases whose
predictions were trained on 88; two contain 9 cases trained on 89.
Consequently the pooled quantity targets the declared mixture

$$
R_{\mathrm{mix}}=\frac{80}{98}R_{88}+\frac{18}{98}R_{89},
$$

where $R_m$ is expected absolute error of the specified learning rule
trained on $m$ eligible same-source records and used on a new one under
the stated repetition. Pair 3 compares the two rules' $R_{\mathrm{mix}}$.
It does not directly estimate performance of the single equation fitted
to all 98 observed records.

The 200 resplits give full-minus-education MAE differences from -0.914 to
-0.741, all favoring the full rule. This supports stability to the examined
split choices. It is **not a 95% interval**: the same cases are reused,
training sets overlap, and the results are not independent studies.
We do not test a mean of ten fold errors as if those errors were independent.
Uncertainty across newly drawn training samples is not quantified here.
The estimated risk advantage is therefore provisional; its sign is not
established for the hypothetical source population by the resplit range.

The comparison supports lower internal held-out error for the full rule in
these records and under this protocol. It does not isolate the predictive
contribution of education: the rules differ by prestige, women, and type
together. Nor does it establish external transport, individual earnings
prediction, or a causal education effect.

# Read the Coordinated Arguments, Then Evaluate and Synthesize {#sec-ch8-coordinate}

## The chapter-authored study record

The following report consolidates the preceding analysis. Treat its
paragraphs, tables, equations, and code as an accessible teaching record,
not as statements copied from the source investigators.

> **R0 -- Common record and declared scope.** We analyzed the 98 complete
> occupation records described in @sec-ch8-questions. Income was
> scaled to thousand
> 1971 Canadian dollars. Education and income were historical occupational
> averages; prestige came from an earlier rating survey. We retained the
> same rows for all three questions and make no claim of verified probability
> sampling or randomized education.
>
> **R1 -- Marginal answer.** Across these records the education-only
> least-squares slope was 0.883 thousand dollars per year. This is a
> descriptive linear summary: occupations with more education tended to
> have higher average income. It is not an individual or causal effect.
>
> **R2 -- Conditional answer.** The intended estimand was the linear
> conditional-mean education coefficient $\beta_E$, comparing education at the
> same prestige, percentage of women, and occupational type. Its fitted
> estimate was $b_E=0.116$, with a classical 95% interval [-0.430, 0.663]
> under the independent linear conditional-mean working model with common
> variance. HC3 gave [-0.643, 0.876]. Influence, functional-form, and support
> checks limit that model's interpretation. The direction and practically
> relevant magnitude remain unresolved under these calculations; the result
> does not establish no conditional relationship.
>
> **R3 -- Prediction answer.** The two fixed OLS learning rules were compared
> with the shared ten-fold protocol in @sec-ch8-prediction. The full
> rule had MAE
> 1.541 versus 2.403 for education-only, difference -0.862. The examined
> resplits all favored the full rule. This is internal evidence about
> learning procedures trained on 88/89 eligible records under the stated
> same-source exchangeability assumptions. Lower expected error is a
> provisional conclusion because new-sample uncertainty is not quantified;
> this is not external validation of a final 98-record equation.
>
> **R4 -- Synthesis.** Education has a positive marginal linear association
> with income in these historical records. Its conditional association is
> imprecisely resolved by the specified additive model, while the full
> learning rule predicts the held-out records more accurately than the
> education-only rule. These conclusions answer different questions and
> do not establish that changing education would change income.

## Read the functions and their traces

A component is a logical function; the record contains study-specific
content serving that function. A passage can carry several functions.
For example, R0 identifies records for STAT-E and supplies historical
context for SUB-E. It becomes J content only when it offers a reason
connecting those records to a particular inference. “We used R” does not
by itself supply that reason.

| Function | Shared or pair-specific content | Trace in this teaching record |
|:--|:--|:--|
| **STAT-E** | Named 98 rows, variables, exclusions, and computations' inputs | R0; the row audit and import code in @sec-ch8-questions |
| **SUB-E** | Occupations rather than workers; historical dollars, education, ratings, and types | R0; variable/source descriptions |
| **STAT-J** | Pair 1: fixed-record least-squares definition. Pair 2: declared stochastic model and uncertainty conditions. Pair 3: held-out fitting, fixed rules, loss, and exchangeability | R1--R3; model declaration; validation protocol and code |
| **SUB-J** | Reasons connecting those measurements and comparisons to historical occupational income, with limits on individual, current, and causal interpretations | R0--R3; unit/time explanations and prediction-opportunity definition |
| **STAT-I** | Pair 1's .883 slope; Pair 2's .116 estimate, intervals, coefficient test and checks; Pair 3's case-level errors, MAEs, and split sensitivity | @sec-ch8-estimate through @sec-ch8-prediction; R1--R3 |
| **STAT-C** | Descriptive positive marginal slope; unresolved conditional slope under the model; smaller internal error for the full rule under the protocol | R1, R2, and R3 respectively |
| **SUB-I** | Interpret years, income units, conditioning, and prediction errors in the named historical occupation setting | Worked substitutions and error calculation; R1--R3 |
| **SUB-C** | Each question-facing answer with its own limits; none supplies a causal or individual-earnings answer | R1--R3; their distinctions preserved in R4 |

: Reading the linked-layer functions without three duplicate maps. {#tbl-ch8-traces tbl-colwidths="[15,47,38]"}

SUB-E situates evidence; SUB-J **augments and contextualizes** statistical
justification; SUB-I interprets the inference; SUB-C bounds the substantive
answer. Sharing a dataset does not make Pair 2's independent-error model
into Pair 3's validation procedure.

The four documentary statuses from Chapter 1 remain useful. Here,
**stated in the paper** means stated in the chapter-authored report;
**located in related materials** covers the linked source documentation
and executable calculation record; **implicit but traceable** marks an
identifiable link whose premise is actually in the record; and
**not located** records a missing trace in the accessible material.
A separate conflict flag records inconsistent sources.

For example, R2 states the independence assumption. Evidence establishing
that assumption for the original occupation-selection process is **not
located** in this teaching record. Those are different entries: a stated
assumption is not a demonstrated sampling design. Conversely, the
translation of .883 thousand dollars into about 883 dollars is implicit
but traceable through the stated unit conversion; no new empirical premise
is invented. No conflict is identified in the defined teaching record;
a conflict search in a real study could have a different result.

Professional knowledge helps identify what function a trace serves. It
does not authorize adding an unreported design feature. If an author gave
the invalid reason “all rows are in the same dataset, therefore they are
independent,” it would still be **stated J**, whose adequacy is judged
separately. Reading records the claim actually offered, not a repaired
claim the reader wishes had been offered.

## Evaluate each argument separately

We now change from identifying the reported reasoning to evaluating its
quality. The six Chapter 1 questions apply to each argument. They are not
six scores to total, and not every concern needs to be counted repeatedly.

| Quality-evaluation question | Application to these arguments |
|:--|:--|
| **Question alignment** | R1 matches its finite-record slope. R2 names $\beta_E$, its fitted estimate, the conditioning set, and the model extension. R3 names the learning procedure, loss, and training sizes rather than substituting a coefficient test. |
| **Evidence integrity** | The 102-to-98 accounting and common rows are explicit. Missing type still limits what is represented. CV excludes held-out outcomes from fitting; source selection is not repaired by that procedure. |
| **Justification adequacy** | Least squares is sufficient for the declared finite description. Pair 2's model conditions and Pair 3's same-source exchangeability remain assumptions with incomplete empirical support. Statistical computation alone cannot establish them. |
| **Inferential adequacy** | The slope, intervals, error arithmetic, and comparisons are reproducible. Neither zero-crossing nor a large p-value proves absence; repeated folds are not independent replication. Substantive reasoning must retain those distinctions. |
| **Transition fidelity** | Occupation-level average income remains occupation-level average income; partial adjustment is not a causal intervention, and model fit is not held-out performance. |
| **Claim calibration** | R1 remains finite and historical; R2 reports unresolved magnitude/direction rather than no relationship; R3 remains internal and procedure-specific. No claim extends to individual workers or modern populations. |

: Apply the six questions before evaluating the study-level synthesis. {#tbl-ch8-quality tbl-colwidths="[28,72]"}

At an individual question, useful judgments include **adequate for this
purpose**, **needs revision**, **insufficient information to judge**, or
**not applicable**, with a reason. For example, the record is adequate to
check Pair 3's held-out computations but gives insufficient information to
establish exchangeability for an external population. These judgments are
not documentary statuses.

\Needspace{6\baselineskip}

Each whole pair then receives its own conclusion:

- **Pair 1: supported as stated.** The answer is an accurate bounded
  description of the retained records.
- **Pair 2: supported as stated**, as the explicitly conditional,
  unresolved working-model answer about $\beta_E$ actually reported. This does not certify
  the model or establish a population relationship. A claim that the true
  national conditional slope is zero would not be supported by this record.
- **Pair 3: supported as stated**, as a qualified internal comparison with
  assumptions and limits retained. External superiority of the final
  all-record fitted equation is **not assessable from the accessible record**;
  it is not the reported answer being accepted.

The other available whole-pair conclusions are **supported only after
narrowing or revision**, **not supported**, and **not assessable from the
accessible record**. A correctly reported inconclusive answer can be
supported as stated. Evaluating what the study actually says prevents a
bounded answer from being rejected merely because it does not answer a
stronger question.

## Add Cross-question coherence for the synthesis

For coordinated questions, ask one additional question:
**Does the synthesis preserve the distinct questions, estimands, methods,
analytic statuses, uncertainty, and limits of the separate answers?**
This is **Cross-question coherence**, introduced locally here. Chapter 1a
develops it further but is not needed to apply it.

R4 is coherent because it reports three distinct outcomes. A positive
marginal slope, an unresolved conditional slope, and better internal
prediction can coexist. They do not contradict one another.

Now consider the alternative sentence:

> Once the other predictors were controlled, education had no effect;
> nevertheless, the significant full regression proves it will predict
> workers' incomes accurately.

Reading first identifies the offered inference and claim. Evaluation then
finds several failures: “no effect” substitutes a causal null for the
conditional-association answer; an omnibus test is not validation; and
occupations become workers. No omnibus significance test can replace the
held-out errors or rescue the change of unit. The corrected synthesis is
R4, not a fourth overarching argument that erases the separate verdicts.

## Auxiliary construction callback

The construction workflow supports researchers carrying out the study; it
is not a mandatory reading sequence. Each question retains its own
six-stage cycle from Chapter 1. For coordinated questions, Chapter 1a's
extension aligns the plans at Stage 3 and adds Stage 7 for synthesis.

In **Understand and plan**, researchers review the problem and literature,
frame questions and candidate answers, and align design, measurement,
estimands, evidence acquisition, and inference. Here, “separate questions”
means separately owed conclusions. If the obligation were one integrated
conclusion, jointly necessary estimands would be planned as strands instead.

The extended **Stage 3, Plan and align prospectively across questions**,
specifies common rows and units but distinct marginal, model-based, and
validation plans. In **Produce and complete**, **Stage 4, Produce or acquire
and document evidence**, designates the question-specific records;
**Stage 5, Estimate, check, and analyse**, performs the computations;
and **Stage 6, Contextualize, justify, interpret, and complete each bounded
substantive answer**, finishes each answer with its actual limits.
**Stage 7, Give explicit outcomes and synthesize**, records all outcomes
and brings them together without creating an omnibus argument.

Iteration remains within a phase. A same-anchor error correction or
justified uncertainty repair can reopen Phase 2 work. A newly preferred
outcome, question mode, or estimand requires a new or revised cycle, with
reused records where appropriate and its timing disclosed. A newly chosen
causal question cannot be “narrowed” into association and treated as answered.
This teaching reanalysis does not retrospectively claim that the historical
data were collected for our three questions.

The occupation study supplies no randomized-design warrant. A second application
now shows how covariate adjustment can seek the same assignment-effect estimand
when design and weighting support that link.

```{=latex}
\Needspace{22\baselineskip}
```

# Project STAR: A Randomized-Design Transfer {#sec-ch8-star}

```{=latex}
\hypertarget{sec-ch8-star}{}
```

This is the full STAR transfer signposted from Chapter 7. It first connects
binary regression to observed arm means, then compares adjusted and
unadjusted specifications on identical primary rows. These are related
calculations, not interchangeable estimates from one unchanged sample.

## A different design supplies a different warrant

Project STAR supplies a transfer with its own question:

> Among eligible kindergarten pupils in the participating Tennessee
> schools, what was the mean reading-score difference under assignment
> to a small rather than a regular class?

The outcome is *readk*, an end-of-kindergarten reading scale score.
The variable *stark* records the class-type condition. Interpreting its
small-versus-regular comparison as an assignment comparison rests on
the documented experimental protocol, not on the column name or an
*lm()* call. Random assignment occurred within participating schools
[@finn1990; @mosteller1995]. The regular-plus-aide condition is outside
this declared two-arm comparison. Class type is not a measured number
of pupils, and this analysis does not estimate a headcount dose response.

The package field does not independently verify each pupil's original
randomized assignment. An assignment-effect reading additionally requires
linking that field to the assignment record and checking switches or
implementation deviations. This teaching reanalysis does not supply that
pupil-level verification.

The data are distributed as *AER::STAR*, accompanying *Applied
Econometrics with R* and following the version used by Stock and Watson
[@kleiber2008; @stock2007]. They are loaded from the package rather
than redistributed here. Read the package's variable-format list:
the chosen outcome is reading, not an invented combined reading/math
score. The teaching analysis is retrospective, not preregistered.

The scientific target is the pupil-weighted mean assignment effect
for the eligible kindergarten cohort in participating schools. A
complete-record common school-fixed-effect coefficient is an
estimator requiring additional links to that target. Randomization
does not, by itself, guarantee those links after exclusions or under
heterogeneous effects.

## Separate eligibility from missingness

Some distributed pupil records belong to later entrants and have no
kindergarten condition recorded. Their structural absence is distinct
from missing reading or covariates among eligible kindergarten pupils.
Only variables required for this analysis enter its complete-record
rule; *mathk* must not exclude an otherwise usable reading record.

```{=latex}
\Needspace{30\baselineskip}
```

```{r}
#| label: ch8c-prepare
#| echo: true
data("STAR", package = "AER")
star_vars <- c("stark", "readk", "lunchk", "gender",
               "ethnicity", "experiencek", "schoolidk")
star_k_recorded <- !is.na(STAR$stark)
star_later_recorded <- rowSums(!is.na(
  STAR[, c("star1", "star2", "star3")]
))
star_two_arm <- star_k_recorded &
  STAR$stark %in% c("regular", "small")
star_complete <- complete.cases(STAR[, star_vars])
star_df <- droplevels(STAR[star_two_arm & star_complete, star_vars])
star_df$stark <- relevel(star_df$stark, ref = "regular")
star_observed <- droplevels(STAR[
  star_two_arm & !is.na(STAR$readk),
  c("stark", "readk", "schoolidk")
])
star_observed$stark <- relevel(star_observed$stark, ref = "regular")
c(observed_reading = nrow(star_observed),
  complete_covariates = nrow(star_df))
```

\Needspace{16\baselineskip}

```{r}
#| label: tbl-ch8c-sample
#| echo: false
#| tbl-cap: "STAR eligibility and reading-analysis completeness."
knitr::kable(data.frame(
  Stage = c("Distributed records", "Kindergarten condition recorded",
    "Eligible small or regular", "Eligible, reading observed",
    "Complete primary records", "Primary regular", "Primary small",
    "Primary schools"),
  n = c(nrow(STAR), sum(star_k_recorded), sum(star_two_arm),
    nrow(star_observed), nrow(star_df),
    sum(star_df$stark == "regular"), sum(star_df$stark == "small"),
    nlevels(star_df$schoolidk))
), row.names = FALSE)
```

Of 11,598 distributed records, 5,273 have no kindergarten condition;
all have later-grade participation in this file. They are structural late
entrants, not 5,273 failed kindergarten reading measurements. The remaining
6,325 have a recorded kindergarten arm. Across all three arms, 536 lack
reading, leaving 5,789 observed-reading records. These are counts in the
distributed file, not a replacement for every original trial enrollment
figure.

Among the 4,094 eligible small/regular records, 349 lack reading. The 3,745
observed-reading records contain 2,006 regular-arm and 1,739 small-arm pupils;
the corresponding missing-reading counts are 188 and 161. Similar missing
percentages do not establish ignorable missingness: the absent outcomes
could differ systematically within either arm.

Do not discard records for variables unused by a calculation. Across all
5,789 observed-reading records, 38 lack at least one of lunch status, gender,
ethnicity, or teacher experience. Within the focal two arms, ten records
lack a required adjustment covariate. The primary common-covariate sample
therefore has 3,735 pupils: 2,001 regular and 1,734 small, in 79 schools.
The observed-reading sensitivity retains all 3,745 focal records; no further
observed-reading records lack school ID.

\Needspace{16\baselineskip}

```{r}
#| label: ch8c-missingness-checks
#| echo: true
star_covariates <- c("lunchk", "gender", "ethnicity", "experiencek")
star_reading_recorded <- star_k_recorded & !is.na(STAR$readk)
c(missing_covariates = sum(star_reading_recorded &
    !complete.cases(STAR[, star_covariates])),
  regular_missing_reading = sum(STAR$stark == "regular" &
    is.na(STAR$readk), na.rm = TRUE),
  small_missing_reading = sum(STAR$stark == "small" &
    is.na(STAR$readk), na.rm = TRUE))
```

Complete-record inference requires an assumption that selection does
not distort the relevant within-school arm comparison, given the
declared covariates and target. The observed-reading sensitivity
relaxes the additional covariate-completeness restriction; it does
not recover the reading scores of the other 349 eligible pupils.

For a full-cohort causal contrast, observation must be connected to potential
reading scores. One possible assumption is that, within arm and school,
observation does not depend on those scores, with adequate observation in
both arms and appropriate school weighting. Observed outcomes alone cannot
verify that assumption. Selecting observed outcomes can otherwise destroy
the comparability created by randomization.

## The binary slope recovers the observed arm means

Let $D_i=1$ for the recorded small arm and 0 for regular. In the model
$\widehat Y_i=\hat\alpha+\hat\delta D_i$, minimizing RSS fits each arm's
observed mean separately. Thus

$$
\hat\alpha=\bar Y_{D=0},\qquad
\hat\delta=\bar Y_{D=1}-\bar Y_{D=0}.
$$

\Needspace{25\baselineskip}

```{r}
#| label: ch8c-binary-means
#| echo: true
star_observed_unadjusted <- lm(readk ~ stark, data = star_observed)
star_arm_means <- tapply(star_observed$readk, star_observed$stark, mean)
star_observed_difference <- unname(
  star_arm_means["small"] - star_arm_means["regular"]
)
c(regular_mean = unname(star_arm_means["regular"]),
  small_mean = unname(star_arm_means["small"]),
  small_minus_regular = star_observed_difference,
  regression_difference = unname(
    coef(star_observed_unadjusted)["starksmall"]
  ))
```

On these 3,745 records, $440.547-434.732=5.815$ reading points. The equality
is algebraic and does not require random assignment. Under independent
common-variance rows, the ordinary slope test would equal the pooled
two-sample $t$ test; that row-level reference does not address this design's
within-school dependence.

The raw arm means weight schools by each arm's observed pupil counts, and
those weights can differ by arm. Common assignment probabilities and suitable
observation patterns, or explicit standardization to target school weights,
are needed to link the pooled difference to the pupil-weighted cohort target.
Clustering the SE does not adjust the coefficient for school blocks, missing
outcomes, or unequal target weights.

## Compare two specifications on identical primary rows

The unadjusted model reproduces the small-minus-regular mean
difference on the primary records. The adjusted model adds free-lunch
status, gender, ethnicity, teacher experience, and school indicators.
School indicators respect the assignment blocks; the other recorded
covariates are the declared adjustment specification, not a substitute
for randomization.

\Needspace{24\baselineskip}

```{r}
#| label: ch8c-models
#| echo: true
star_unadjusted <- lm(readk ~ stark, data = star_df)
star_primary <- lm(
  readk ~ stark + lunchk + gender + ethnicity +
    experiencek + schoolidk,
  data = star_df
)
star_sensitivity <- lm(
  readk ~ stark + schoolidk, data = star_observed
)
c(unadjusted = nobs(star_unadjusted),
  adjusted = nobs(star_primary),
  observed_reading = nobs(star_sensitivity))
```

Adjustment can improve precision when added variables are prognostic.
It is not necessary to remove baseline confounding created by a
successfully implemented random assignment. It also does not
automatically preserve every causal target: variable timing, selection,
and the fitted comparison still need scrutiny.

On the common 3,735 primary rows, the unadjusted difference is 5.895465 and
the adjusted coefficient is 6.657510. The earlier 5.815 difference uses
3,745 observed-reading rows; it must not serve as the supposedly same-row
unadjusted baseline. The observed-reading school-adjusted sensitivity is
another specified calculation, not the raw observed mean difference.

The common *starksmall* coefficient is a coefficient of the stated
additive projection. If effects vary across schools or profiles, its
implicit weights need not equal the pupil weights defining the cohort
target. A common effect or a target-matching weighting argument is
therefore an additional requirement. Similar unadjusted and adjusted
coefficients are consistent with successful randomization, not proof
that implementation, missingness, or weighting problems are absent.

## Cluster uncertainty follows the design

Pupils are outcome units; schools are the approximately independent
units for this covariance calculation and the randomization blocks.
Schools themselves were not assigned to the two arms. Pupils sharing a
classroom may also be dependent; school clustering allows within-school
residual association, including that within classrooms, under an
independent- or weakly-dependent-school large-cluster approximation.

For school $g$, form its score $s_g=X_g'e_g$. The CR1/HC1 covariance used
here combines residual contributions **within school before taking their
outer product**:

$$
\widehat V_{\mathrm{CR1}}
=c_G(X'X)^{-1}\left(\sum_{g=1}^G s_gs_g'\right)(X'X)^{-1},
\qquad c_G=\frac{G}{G-1}\frac{n-1}{n-p}.
$$

Here $p$ counts all fitted coefficient columns, including the intercept;
for the binary-only model $p=2$. The within-school cross-products distinguish
this from treating every pupil as independent. The implementation below
uses school-clustered HC1 covariance, the finite-cluster correction, and a
$t_{G-1}$ critical value, computed with the clustered covariance estimator of
the *sandwich* package [@zeileis2020]. These are stated approximations, not
exact randomization inference.

```{=latex}
\Needspace{23\baselineskip}
```

```{r}
#| label: ch8c-cluster-results
#| echo: true
star_V <- sandwich::vcovCL(
  star_primary, cluster = star_df$schoolidk,
  type = "HC1", cadjust = TRUE
)
star_b <- unname(coef(star_primary)["starksmall"])
star_se <- sqrt(star_V["starksmall", "starksmall"])
star_G <- nlevels(star_df$schoolidk)
star_critical <- qt(.975, star_G - 1)
c(estimate = star_b, SE = star_se)
star_b + c(-1, 1) * star_critical * star_se
```

This example uses the primary adjusted model and the school IDs from its
3,735 records. Apply the same calculation to the other fitted models using
the school IDs from each model's own rows. The table collects their results;
changing the model or its records is distinct from changing the covariance
calculation.

```{r}
#| label: ch8c-format-results
#| echo: false
star_cluster_result <- function(fit, school) {
  V <- sandwich::vcovCL(fit, cluster = school,
                        type = "HC1", cadjust = TRUE)
  b <- unname(coef(fit)["starksmall"])
  se <- sqrt(V["starksmall", "starksmall"])
  G <- nlevels(droplevels(school))
  critical <- qt(.975, G - 1)
  data.frame(n = nobs(fit), G = G, Estimate = b, SE = se,
             Lower = b - critical * se,
             Upper = b + critical * se)
}
star_results <- rbind(
  star_cluster_result(star_observed_unadjusted,
                      star_observed$schoolidk),
  star_cluster_result(star_unadjusted, star_df$schoolidk),
  star_cluster_result(star_primary, star_df$schoolidk),
  star_cluster_result(star_sensitivity, star_observed$schoolidk)
)
```

\Needspace{11\baselineskip}

```{r}
#| label: tbl-ch8c-estimates
#| echo: false
#| tbl-cap: "Class-type comparisons and clustered intervals."
ch8c_display <- data.frame(
  Model = c("Observed: unadjusted", "Primary: unadjusted",
            "Primary: adjusted", "Observed: school-adjusted"),
  n = star_results$n, Estimate = star_results$Estimate,
  SE = star_results$SE,
  Interval = sprintf("[%.2f, %.2f]",
                      star_results$Lower, star_results$Upper)
)
knitr::kable(ch8c_display, digits = 3, row.names = FALSE)
```

For the unadjusted 3,745-record comparison, the 79-school covariance gives
SE 1.847617. With $t_{.975,78}\approx1.991$, its approximate interval is
$[2.137,9.493]$ around 5.815. This preserves the binary-mean comparison while
changing its uncertainty calculation; it does not repair missing outcomes
or school weighting.

The primary adjusted estimate is 6.658 reading points, with approximate
interval $[3.278,10.037]$. The observed-reading sensitivity is 6.627,
with interval $[3.084,10.170]$. The similarity is informative about
this limited sensitivity; it is not evidence that all missingness is
ignorable.

These intervals rely on independent schools and a cluster-asymptotic
approximation with 79 clusters. They are neither exact randomization
intervals nor results from a multilevel model. Row-level HC3 would
address unequal variance but not within-school dependence. A residual
plot cannot choose the design's independence unit. Nor does a count of
79 schools certify the approximation: information balance, dependence,
and the influence of particular clusters also matter.

## Read the bounded answer, then evaluate its support

A chapter-authored report can state:

> On 3,735 complete two-arm kindergarten records, the common adjusted
> small-versus-regular class-type coefficient was 6.658 reading
> points, with approximate school-clustered interval [3.278, 10.037].
> Its causal mean-effect interpretation relies on documented
> within-school randomization, a verified link from recorded class type to
> original assignment, suitable complete-record selection,
> independent-school uncertainty, and a common-effect or
> target-matching weighting argument. It concerns assigned class type,
> one kindergarten reading outcome, and participating Tennessee
> schools, not realized headcount or a portable policy dose response.

The code and sample audit carry **STAT-E** content; the design and
covariance paragraph offer **STAT-J** reasons; the estimates and
interval calculation supply **STAT-I** content; and the bounded
statistical answer supplies **STAT-C** content. **SUB-E** situates the
pupils, test, cohort, and school context. **SUB-J** augments and
contextualizes the justification through the assignment record and
the selection/weighting qualifications. **SUB-I** interprets the
comparison as assignment-related only under that support. **SUB-C**
bounds the substantive answer. These are linked logical functions
within one argument; neither layer is a separate argument.

The chapter-authored report states the selection and weighting
conditions; stating them does not establish that they hold. The
randomization protocol is located in related study publications,
not demonstrated by the coefficient table. Reading first locates
these traces and their documentary statuses. A separate evaluation
asks whether they adequately support the target and stated conclusion.
If a necessary link cannot be assessed, report that limitation rather
than inventing it or silently replacing the target with a different one.

Unlike Prestige, this example has an experimental assignment warrant.
Neither a stable coefficient nor better prediction would have created
that warrant. Under adjustment, a focal coefficient moves through the
added predictors' relations with the focal predictor and outcome,
conditional on terms already included. Within-school randomization
makes baseline relations with assignment small on average; prognostic
relations with reading may remain. This explains why useful
adjustment can retain an assignment target without treating every
model change as a new scientific question.

The primary design publications and subsequent analyses document
implementation and interpretation issues [@finn1990; @mosteller1995;
@krueger1999; @hanushek1999; @nye2000; @chetty2011]. The present result
does not settle transport to nonparticipating schools, another period,
another outcome, or another education system. Those limits belong in
the answer, not only in a list after an unrestricted causal claim.

## Apply the randomized-design transfer

Reproduce the eligibility, observed-reading, complete-record, arm,
and school counts. Explain why *mathk* does not enter the exclusion rule.
Recover the binary slope from the two observed arm means and reproduce its
school-clustered interval. Fit the two primary models on identical rows,
then the observed-reading school-adjusted sensitivity. Keep their samples
and coefficients distinct. Identify the documented reasons for school
blocking and clustered uncertainty, and distinguish the fitted coefficient
from the pupil-weighted cohort target. Read the report's STAT/SUB functions
and trace statuses before separately evaluating its selection, weighting,
causal, outcome, and transport boundaries.

# Summary and Reading Bridge {#sec-ch8-summary}

Multiple regression adds conditional comparisons, not automatic causal
control. Subtracting fitted means shows what a coefficient compares;
residualization connects it to covariance. Residual, influence, support, and
precision checks have different consequences for a bounded answer. A fitted
profile and a standardized fitted quantity are distinct didactic extensions of
the conditional model; held-out predictive performance answers a separately
owed question. Changing rows, conditioning, coding, or uncertainty
must be identified by what actually changes, not treated as interchangeable
operations.

The occupation study owes three answers. Its marginal association is a
finite-record description. Its conditional interval is a working-model
inference with limited stability and unresolved magnitude. Its held-out
errors address a separate learning-rule performance question. Shared records
make the questions coordinated; they do not make their justifications
identical. Read each answer before evaluating quality and then checking
the coherence of the synthesis.

The STAR transfer changes the warrant, not the need for careful alignment.
Documented within-school assignment can support an assignment-effect reading,
while missing outcomes, recorded assignment, school weighting, common-row
adjustment, and cluster-based uncertainty still limit the answer. The 3,745-row
observed mean difference and the 3,735-row adjusted coefficient are useful
because their samples and functions remain explicit, not because their similar
values prove every link.

For further study:

- The [OpenIntro multiple-predictor tutorial](https://openintro-ims.netlify.app/model-mlr)
  offers concrete equations and indicator interpretations. We use that
  equation-first teaching device without adopting automatic stepwise
  selection as this chapter's route.
- The [ISLR regression lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch3-linreg-lab.html)
  connects fitting, coefficients, checks, and predictions.
- The [ISLR cross-validation lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch5-resample-lab.html)
  develops the train/predict/check sequence. Our fixed rules, MAE, and
  training-size qualifications are specified for this teaching example.
- The [carData documentation](https://cran.r-project.org/web/packages/carData/carData.pdf)
  supplies the Prestige variable meanings and provenance. The coordinated
  study and R0--R4 report are chapter-authored, not claims attributed to
  that documentation.

Chapter 1a offers an optional extension of argument coordination. The next
regression question is different: might the education comparison itself
vary with another predictor, rather than remain the same additive slope?
That is the reason to introduce interaction.

# Exercises {.unnumbered}

Use the supplied numbers; rounding within 0.02 thousand dollars is
acceptable unless a different precision is requested. These prompts do not
require Chapter 1a or the appendices.

\Needspace{18\baselineskip}

<!-- prompt-id: CH08-E01 -->
## Exercise 1: Keep the three answers distinct (Foundational) {.unnumbered #ex-ch08-01}

A report promises: (a) a description of education and income across 98
occupations, (b) an education comparison at the same prestige, women, and
type, and (c) a comparison of two prediction rules on held-out records.

1. Explain why these are three question--answer pairs rather than three
   statistical strands serving one integrated answer.
2. For each, name its estimand in plain language and one kind of output
   that could contribute to its answer.
3. Does sharing the same 98 records establish the same justification for
   all three? Give one concrete reason.

<!-- prompt-id: CH08-E02 -->
## Exercise 2: Name what changes (Foundational) {.unnumbered #ex-ch08-02}

Classify each change as primarily changing the records, the comparison,
the coefficient coordinates, or the uncertainty calculation. State whether
it necessarily creates a new scientific question.

1. Refit education-only regression after excluding four missing-type records.
2. Add prestige, women, and type to the education-only association equation.
3. Change the reference category from blue-collar to white-collar.
4. Keep the fitted equation but replace classical SEs with HC3 SEs.

Finally, explain why appropriately specified adjustment in a randomized
study need not change its declared average assignment-effect estimand.

<!-- prompt-id: CH08-E03 -->
## Exercise 3: Interpret and calculate a fitted profile (Foundational) {.unnumbered #ex-ch08-03}

Use this rounded hypothetical equation, with income in thousand dollars:

$$
\widehat y=1+0.20e+0.10p-0.04w+0.30D_{\mathrm{prof}}
           +0.15D_{\mathrm{wc}}.
$$

1. Calculate the fitted mean for a white-collar profile with $e=12$,
   $p=40$, and $w=50$.
2. Calculate its modeled difference from a profile with $e=10$ and the
   same other values. Interpret the result in dollars.
3. What must be checked before calling this a comparison supported by the
   observed data?
4. Why is neither calculation an estimate of the effect of giving a worker
   two more years of education?

\Needspace{18\baselineskip}

<!-- prompt-id: CH08-E04 -->
## Exercise 4: Read held-out errors (Foundational) {.unnumbered #ex-ch08-04}

For four held-out occupations, observed incomes are $(5,8,10,13)$ thousand
dollars. The education-only predictions are $(4,10,7,12)$; full-rule
predictions are $(6,8,9,11)$. Each prediction was fitted without the
corresponding outcome.

1. Calculate the four absolute errors and MAE for each rule.
2. Calculate the average paired difference, full minus education-only.
3. Does the better average rule have to be better for every occupation?
4. Explain why this comparison is neither a coefficient interval nor
   validation of a model subsequently fitted to all available records.

<!-- prompt-id: CH08-E05 -->
## Exercise 5: Locate a reason before judging it (Integrative) {.unnumbered #ex-ch08-05}

An accessible report states:

> We retained the same 98 occupations for both equations. This makes the
> observations independent. The full equation has higher training
> $R^2$, so we conclude it predicts new occupations more accurately.

The report provides import code and fitted-model output, but no validation
split, held-out predictions, or study-specific evidence about dependence.

1. Identify the reasons offered, the inference, and the claim.
2. Record the documentary status of the independence reason and the
   held-out prediction evidence separately. Do not rewrite “stated” as
   “not located” merely because a reason is wrong.
3. Apply Justification adequacy, Inferential adequacy, and Claim calibration.
4. Give a whole-pair evaluation conclusion for the prediction answer and
   name the additional work it would require.

<!-- prompt-id: CH08-E06 -->
## Exercise 6: Evaluate the study-level synthesis (Integrative) {.unnumbered #ex-ch08-06}

Use the occupation study's reported results: marginal slope .883;
fitted conditional estimate $b_E=.116$ of $\beta_E$, with classical interval
[-.430, .663];
and held-out MAEs 2.403 for education-only versus 1.541 for the full rule.

A summary says:

> Education does not matter after adjustment, but the model proves that
> increasing workers' education raises income and predicts earnings well
> in other countries.

1. Read the claims actually offered before evaluating them.
2. Identify which separate answer obligation is being replaced or exceeded
   by each clause.
3. Apply Cross-question coherence and write a corrected three-sentence
   synthesis that retains the three reported answers and their limits.
4. If the authors now want a causal effect of education, is that a repair
   of their existing uncertainty calculation or a new/revised question-level
   cycle? Explain.

# Appendix A: Adjustment, Coding, and Group-Comparison Mathematics {.unnumbered #app-ch8-a}

This appendix develops the mathematics behind @sec-ch8-comparisons.
Appendix B retains technical reproduction and sensitivity;
@sec-ch8-star changes the design context.
The occupation calculations use the same 98-row object *d* and the
models *m0*, *m1*, and *m2* defined in the main text. They do not introduce
another sample. The type codes remain *bc*, *prof*, and *wc*.

## Why the marginal slope includes another association {.unnumbered}

Suppose a population process is

$$
Y=\beta_0+\beta_X X+\beta_Z Z+\varepsilon,
\qquad E(\varepsilon\mid X,Z)=0,
$$

with finite variances and $\operatorname{Var}(X)>0$. The marginal
least-squares slope is a covariance divided by a variance. Substituting
the outcome equation gives

$$
\begin{aligned}
\frac{\operatorname{Cov}(X,Y)}{\operatorname{Var}(X)}
&=\frac{\beta_X\operatorname{Var}(X)
+\beta_Z\operatorname{Cov}(X,Z)
+\operatorname{Cov}(X,\varepsilon)}{\operatorname{Var}(X)}\\
&=\beta_X+\beta_Z
\frac{\operatorname{Cov}(X,Z)}{\operatorname{Var}(X)}.
\end{aligned}
$$ {#eq-ch8a-omission}

The zero-error-covariance condition removes the final term. The
discrepancy therefore involves both the association between $X$ and $Z$
and $Z$'s coefficient in the outcome model. Neither ingredient alone
determines its size. If the target is the conditional coefficient,
omission creates bias relative to that target. If the target is the
marginal slope, that marginal quantity is not a biased version of an
unstated conditional target. Neither interpretation supplies a causal
warrant for observational variables.

Changing a specification does not universally change a scientific
estimand. Different estimators can target the same declared quantity,
as the randomized adjustment example in @sec-ch8-star illustrates.
Releveling a factor merely changes coordinates. By contrast, the
marginal and conditional coefficients in this example name different
comparisons.

## A known process and repeated samples {.unnumbered}

Let independent standard normal variables $X$ and $U$ generate
$Z=.65X+\sqrt{1-.65^2}U$. Then both predictors have variance one and
their correlation is .65. Independently generate a normal error with
SD 6, and set

$$
Y=50+2X+5Z+\varepsilon.
$$ {#eq-ch8a-sim-process}

The marginal slope is $2+5(.65)=5.25$; the conditional $X$ slope is 2.
The following single sample has 250 observations. Its measured strategy
and prior-attainment labels are fictional teaching interpretations.

```{=latex}
\Needspace{24\baselineskip}
```

```{r}
#| label: ch8a-one-sample
#| echo: true
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
set.seed(82108)
ch8a_n <- 250
ch8a_x <- rnorm(ch8a_n)
ch8a_z <- .65 * ch8a_x + sqrt(1 - .65^2) * rnorm(ch8a_n)
ch8a_y <- 50 + 2 * ch8a_x + 5 * ch8a_z +
  rnorm(ch8a_n, sd = 6)
ch8a_sim <- data.frame(
  strategy = ch8a_x, prior = ch8a_z, score = ch8a_y
)
ch8a_marginal <- lm(score ~ strategy, data = ch8a_sim)
ch8a_adjusted <- lm(score ~ strategy + prior, data = ch8a_sim)
c(marginal = unname(coef(ch8a_marginal)["strategy"]),
  conditional = unname(coef(ch8a_adjusted)["strategy"]))
```

```{r}
#| label: fig-ch8a-predictors
#| fig-pos: H
#| echo: false
#| fig-cap: "Joint predictor support in one generated sample."
#| fig-alt: >
#|   A diagonal cloud shows positively related strategy and prior
#|   attainment. The central region has more nearby comparisons
#|   than the sparse corners.
#| fig-width: 6.1
#| fig-height: 3.5
ggplot2::ggplot(ch8a_sim, ggplot2::aes(strategy, prior)) +
  ggplot2::geom_point(shape = 1, alpha = .65) +
  ggplot2::geom_smooth(method = "lm", se = FALSE) +
  ggplot2::labs(x = "Strategy (standardized)",
                y = "Prior attainment (standardized)")
```

```{r}
#| label: fig-ch8a-outcome
#| fig-pos: H
#| echo: false
#| fig-cap: "A marginal line does not hold prior attainment fixed."
#| fig-alt: >
#|   Score increases with strategy. Higher prior-attainment values,
#|   represented by darker points, tend to accompany higher scores
#|   and strategy. The black line is the marginal fitted relation.
#| fig-width: 6.1
#| fig-height: 3.5
ggplot2::ggplot(
  ch8a_sim, ggplot2::aes(strategy, score, colour = prior)
) +
  ggplot2::geom_point(alpha = .8) +
  ggplot2::geom_smooth(
    ggplot2::aes(group = 1), method = "lm", se = FALSE,
    colour = "black"
  ) +
  ggplot2::scale_colour_gradient(low = "grey75", high = "black") +
  ggplot2::labs(x = "Strategy (standardized)", y = "Score",
                colour = "Prior attainment")
```

The first plot displays the support needed for a conditional
comparison. The second shows that prior attainment varies along the
marginal relation. A line calculated outside the observed predictor
cloud is not made well supported by the existence of its equation.

Now use a different, explicitly stated sample size: 150 observations in
each of 1,000 repetitions. This retains the earlier repeated-sample
exercise, rather than silently treating it as repetitions of the
250-observation draw.

```{=latex}
\Needspace{16\baselineskip}
```

```{r}
#| label: ch8a-repetition
#| echo: true
set.seed(82109)
ch8a_one_rep <- function() {
  x <- rnorm(150)
  z <- .65 * x + sqrt(1 - .65^2) * rnorm(150)
  y <- 50 + 2 * x + 5 * z + rnorm(150, sd = 6)
  c(Marginal = unname(coef(lm(y ~ x))["x"]),
    Conditional = unname(coef(lm(y ~ x + z))["x"]))
}
ch8a_slopes <- t(replicate(1000, ch8a_one_rep()))
```

```{r}
#| label: tbl-ch8a-repeated
#| echo: false
#| tbl-cap: "Targets and results across 1,000 generated samples."
knitr::kable(data.frame(
  Model = colnames(ch8a_slopes), Target = c(5.25, 2),
  Mean = colMeans(ch8a_slopes),
  SD = apply(ch8a_slopes, 2, sd), row.names = NULL
), digits = 3)
```

```{r}
#| label: fig-ch8a-repeated
#| echo: false
#| fig-pos: H
#| fig-cap: "Sampling distributions around two different targets."
#| fig-alt: >
#|   Separate panels show marginal slopes centered near 5.25 and
#|   conditional slopes centered near 2. Dashed lines mark their
#|   respective population targets.
#| fig-width: 6.1
#| fig-height: 3.8
ch8a_long <- data.frame(
  Slope = c(ch8a_slopes[, 1], ch8a_slopes[, 2]),
  Model = rep(c("Marginal", "Conditional"), each = 1000)
)
ch8a_targets <- data.frame(
  Model = c("Marginal", "Conditional"), Target = c(5.25, 2)
)
ggplot2::ggplot(ch8a_long, ggplot2::aes(Slope)) +
  ggplot2::geom_density(fill = "grey80") +
  ggplot2::geom_vline(
    data = ch8a_targets, ggplot2::aes(xintercept = Target),
    linetype = 2
  ) +
  ggplot2::facet_wrap(~Model, scales = "free_x") +
  ggplot2::labs(x = "Estimated strategy slope", y = "Density")
```

Both procedures center near their own targets. The conditional
estimates are more variable here: separating correlated predictors
uses less residual predictor variation. This is a result of the
specified process, not a general promise that adjustment always lowers
precision. Prognostic covariates and their relations to the focal
predictor both matter. Simulation demonstrates consequences of its
assumptions; it cannot certify the observational occupation study.

## Indicators recover fitted group means {.unnumbered}

Start with a model containing only the three-level type factor. With
blue-collar as reference,

$$
\widehat y_g=\hat\alpha+
\hat\delta_{\mathrm{prof}}D_{\mathrm{prof}}+
\hat\delta_{\mathrm{wc}}D_{\mathrm{wc}}.
$$

The fitted means are $\hat\mu_{\mathrm{bc}}=\hat\alpha$,
$\hat\mu_{\mathrm{prof}}=\hat\alpha+\hat\delta_{\mathrm{prof}}$,
and $\hat\mu_{\mathrm{wc}}=\hat\alpha+\hat\delta_{\mathrm{wc}}$.
OLS minimizes squared deviations separately within each group, so these
equal the observed group means. The hats distinguish fitted summaries
from hypothetical population means.

```{r}
#| label: ch8a-group-means
#| echo: true
stopifnot(nrow(d) == 98,
          identical(levels(d$type), c("bc", "prof", "wc")))
ch8a_groups <- data.frame(
  type = factor(c("bc", "prof", "wc"), levels = levels(d$type))
)
model.matrix(~type, data = ch8a_groups)
ch8a_group_fit <- lm(income_k ~ type, data = d)
cbind(ch8a_groups,
      observed = as.numeric(tapply(d$income_k, d$type, mean)),
      fitted = predict(ch8a_group_fit, ch8a_groups))
```

This group-only description is not the adjusted type comparison in
*m2*. In *m2*, indicators compare fitted values at the same education,
prestige, and percentage women. The factor is one term represented by
two columns; it is neither an ordered numeric score nor one selected
reference-category coefficient.

Releveling changes the intercept and indicator coefficients, not the
fitted values, residuals, or a fixed substantive contrast:

\Needspace{14\baselineskip}

```{r}
#| label: ch8a-relevel
#| echo: true
ch8a_relevelled <- d
ch8a_relevelled$type <- relevel(d$type, ref = "wc")
ch8a_ref_fit <- lm(
  income_k ~ education + prestige + women + type,
  data = ch8a_relevelled
)
max(abs(fitted(m2) - fitted(ch8a_ref_fit)))
```

The result is floating-point noise. No new question has appeared merely
because the coordinates changed. Centering continuous predictors also
changes the intercept's reference profile without changing fitted
values. It cannot create cross-type support.

## A contrast needs coefficient covariance {.unnumbered}

For a declared linear contrast $\theta=c'\beta$, its estimate and
estimated variance are

$$
\hat\theta=c'\hat\beta,\qquad
\widehat{\operatorname{Var}}(\hat\theta)=c'\widehat Vc.
$$

The SE is the square root of that variance. Consider professional minus
white-collar in *m2*, at the same three continuous predictors. Relative
to the coefficient order intercept, education, prestige, women,
professional, white-collar, use $c=(0,0,0,0,1,-1)'$.

```{r}
#| label: ch8a-contrast
#| echo: true
ch8a_c <- c(0, 0, 0, 0, 1, -1)
ch8a_V <- vcov(m2)
ch8a_est <- sum(ch8a_c * coef(m2))
ch8a_var <- drop(ch8a_c %*% ch8a_V %*% ch8a_c)
ch8a_se <- sqrt(ch8a_var)
c(estimate = ch8a_est, variance = ch8a_var, SE = ch8a_se)
```

Numerically, the estimate is $0.324193-0.235331=0.088863$ thousand
dollars. Its variance includes the covariance:

$$
2.140641+0.969780-2(0.972018)=1.166385.
$$

Thus $SE=\sqrt{1.166385}=1.079993$, and the classical $t_{92}$
interval is approximately $[-2.056,2.234]$. Adding the two SEs, or adding
variances while discarding covariance, gives a different and incorrect
calculation. This interval is a working-model benchmark, not national
sampling uncertainty. Nor does it overcome the limited observed
cross-type support for the hypothetical same-covariate comparison.

## Classical ANOVA and explicit block tests {.unnumbered}

For the group-only model, the null is equality of all three means,
equivalently both type coefficients zero. The reduced model contains
only an intercept. The full model has three coefficient columns.
The classical statistic is

$$
F=\frac{(RSS_{\mathrm{reduced}}-RSS_{\mathrm{full}})/2}
{RSS_{\mathrm{full}}/(98-3)}.
$$ {#eq-ch8a-group-f}

This is exactly the classical one-way ANOVA statistic: the same
fitted group means, within-group RSS, two restrictions, denominator
df 95, and null reference under independent normal common-variance
errors. The equality is algebraic; it does not depend on randomized
group membership.

```{r}
#| label: ch8a-anova-equivalence
#| echo: true
ch8a_null <- lm(income_k ~ 1, data = d)
ch8a_nested <- anova(ch8a_null, ch8a_group_fit)
ch8a_aov <- aov(income_k ~ type, data = d)
c(max_fitted_difference = max(abs(
    fitted(ch8a_group_fit) - fitted(ch8a_aov))),
  regression_F = ch8a_nested$F[2],
  anova_F = summary(ch8a_aov)[[1]][["F value"]][1])
```

An adjusted type test has different full-model dimensions. It compares
*m2* with a model containing education, prestige, and women, giving
two restrictions but denominator df $98-6=92$. Adding women and type
jointly to *m1* instead tests three restrictions.

```{r}
#| label: ch8a-adjusted-blocks
#| echo: true
ch8a_no_type <- lm(
  income_k ~ education + prestige + women, data = d
)
ch8a_women_type <- anova(m1, m2)
ch8a_type <- anova(ch8a_no_type, m2)
```

\Needspace{8\baselineskip}

```{r}
#| label: tbl-ch8a-blocks
#| echo: false
#| tbl-cap: "Two declared adjusted block comparisons."
knitr::kable(data.frame(
  Added = c("Women and type", "Type after women"),
  df = c(3, 2), F = c(ch8a_women_type$F[2], ch8a_type$F[2]),
  p = c(format.pval(ch8a_women_type[["Pr(>F)"]][2],
                   digits = 3, eps = .001),
        format.pval(ch8a_type[["Pr(>F)"]][2], digits = 3))
), digits = 3, row.names = FALSE)
```

The first comparison gives $F(3,92)=11.509$; the second gives
$F(2,92)=0.032$, $p=.969$. Joint evidence for the first block does not
establish that every added variable contributes. The second result is
not proof that occupation types are meaningless.

Identical rows and the same outcome are necessary for this RSS
comparison, but not sufficient: the models must be nested, their
dimensions correct, and the reference-distribution conditions stated.
Calling *anova(m2)* alone instead gives sequential sums of squares in
formula order. Those are not order-free tests of every term. Welch and
robust covariance tests are different procedures, not interchangeable
versions of the classical formula.

## Matrix geometry and general restrictions {.unnumbered}

Let $X$ have $n$ rows and $p$ coefficient columns, including the
intercept. Full column rank makes $X'X$ invertible. Differentiating
$(y-X\beta)'(y-X\beta)$ gives the normal equations
$X'(y-X\hat\beta)=0$, hence

$$
\hat\beta=(X'X)^{-1}X'y,\qquad
\hat y=Hy,\qquad H=X(X'X)^{-1}X'.
$$

The fitted vector is a projection onto the space spanned by the
columns of $X$. Residuals are orthogonal to those columns. Changing a
full-rank coding basis preserves that space and its fitted vector;
adding a genuinely new column can enlarge it. A numerical solution is
not itself a stochastic model or a substantive explanation.

```{=latex}
\Needspace{19\baselineskip}
```

```{r}
#| label: ch8a-matrix
#| echo: true
ch8a_X <- model.matrix(m2)
ch8a_beta <- solve(crossprod(ch8a_X),
                  crossprod(ch8a_X, d$income_k))
c(columns = ncol(ch8a_X), rank = qr(ch8a_X)$rank,
  max_coefficient_difference = max(abs(ch8a_beta - coef(m2))),
  max_normal_equation_error = max(abs(
    crossprod(ch8a_X, residuals(m2)))))
```

R's *lm()* uses a numerically preferable QR calculation rather than
requiring users to form an inverse. The displayed matrix equation
explains the estimator, not a recommendation to replace its numerical
implementation [@fox2016; @fox2019].

Under the classical variance model,
$\widehat V=s^2(X'X)^{-1}$ with $s^2=RSS/(n-p)$. For $q$ independent
linear restrictions $R\beta=r$, the classical statistic is

$$
F=\frac{(R\hat\beta-r)'
[R(X'X)^{-1}R']^{-1}(R\hat\beta-r)}{q s^2}.
$$

It has the $F_{q,n-p}$ reference under the stated normal linear model.
Choosing rows of $R$ that select the two type coefficients recovers
the adjusted type-block test. A robust replacement for the covariance
requires its own approximate reference, not a claim of the same exact
finite-sample result.

Another view of a partial slope removes the fitted linear relations
with the other predictors from both income and education. If the
remaining vectors are $r_Y$ and $r_E$, then
$b_E=r_E'r_Y/(r_E'r_E)$.

\Needspace{17\baselineskip}

```{r}
#| label: ch8a-residualization
#| echo: true
ch8a_re <- residuals(lm(
  education ~ prestige + women + type, data = d
))
ch8a_ry <- residuals(lm(
  income_k ~ prestige + women + type, data = d
))
c(partial_slope = drop(
    crossprod(ch8a_re, ch8a_ry) / crossprod(ch8a_re)),
  partial_correlation = cor(ch8a_re, ch8a_ry))
```

The slope reproduces 0.116434; the partial correlation is 0.044063.
This is linear residualization, not actual matching or randomization.
A component-plus-residual plot in @sec-ch8-check is a different display:
it adds a fitted component back to the outcome residual.

## Extended practice: coding and known processes (Integrative) {.unnumbered}

1. Recover fitted *m2* values for the three types at education 11,
   prestige 50, and women 30. Treat this as algebra, then evaluate
   whether all three combinations have observed support. Relevel
   type to white-collar and verify a fixed fitted contrast is unchanged.
2. Reproduce the repeated-sample process at $n=150$. Change exactly
   one of the correlation's sign, its value to zero, or the omitted
   variable's coefficient to zero. Derive both targets, run at least
   1,000 repetitions with a recorded seed, and compare bias and SD
   relative to each target. An optional harder version uses separated
   predictor clusters and explicitly recalculates their covariance.

# Appendix B: Technical Reproduction and Sensitivity {.unnumbered #app-ch8-b}

The numbered route now carries the ordinary support, diagnostic,
standardization, and targeted-sensitivity lessons. This appendix retains
full computational reproduction and mathematical detail. It uses *d*, *m0*,
*m1*, and *m2* without changing their source rows. Classical intervals are
conditional-model benchmarks; technical refinements do not create missing
design, measurement, independence, or transport warrants.

## Reproduce validation and inspect resplit sensitivity {.unnumbered}

The following complete function implements the main comparison. Both
rules use the same training rows and the same held-out records. It
checks factor support and rank before prediction; it does not use
held-out outcomes to select variables or transform predictors.

\Needspace{30\baselineskip}

```{r}
#| label: ch8b-cv-function
#| echo: true
cv_compare <- function(data, fold) {
  stopifnot(length(fold) == nrow(data), !anyNA(fold),
            identical(levels(data$type), c("bc", "prof", "wc")))
  vars <- c("income_k", "education", "prestige", "women", "type")
  stopifnot(all(complete.cases(data[, vars])))
  pred <- matrix(NA_real_, nrow(data), 2,
                 dimnames = list(NULL, c("Education-only", "Full")))
  for (k in sort(unique(fold))) {
    train <- data[fold != k, ]
    test <- data[fold == k, ]
    stopifnot(all(table(train$type) > 0))
    simple <- lm(income_k ~ education, data = train)
    full <- lm(income_k ~ education + prestige + women + type,
               data = train)
    stopifnot(simple$rank == 2, full$rank == 6,
              nobs(simple) == nrow(train),
              nobs(full) == nrow(train))
    pred[fold == k, 1] <- predict(simple, newdata = test)
    pred[fold == k, 2] <- predict(full, newdata = test)
  }
  stopifnot(all(is.finite(pred)))
  errors <- abs(pred - data$income_k)
  mae <- colMeans(errors)
  list(predictions = pred, absolute_errors = errors, MAE = mae,
       difference = unname(mae[2] - mae[1]))
}
```

With the function defined, recreate the stated folds and collect the errors.

\Needspace{17\baselineskip}

```{r}
#| label: ch8b-cv-run
#| echo: true
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
set.seed(80806)
ch8b_fold <- sample(rep(seq_len(10), length.out = nrow(d)))
ch8b_primary <- cv_compare(d, ch8b_fold)
ch8b_primary$MAE
ch8b_primary$difference
```

The primary MAEs reproduce 2.403130 and 1.541238, with full minus
education-only difference $-0.861892$ thousand dollars. The function
averages over cases. An unweighted average of ten fold means would
give the two nine-case folds too much weight. The case-weighted target
mixes training sizes 88 and 89 with weights $80/98$ and $18/98$.

The fixed unit conversion *income/1000* is not a learned preprocessing
step. Any future learned transformation, imputation, or variable
selection would have to occur inside each training fold. The complete
record restriction instead defines this exercise's eligibility; it
does not make missingness harmless for a wider population.

The predeclared sensitivity uses 200 new random fold allocations,
without modifying either rule after looking at their errors:

```{r}
#| label: ch8b-repeat-cv
#| echo: true
set.seed(80807)
ch8b_resplits <- t(replicate(200, {
  f <- sample(rep(seq_len(10), length.out = nrow(d)))
  ans <- cv_compare(d, f)
  c(Education = ans$MAE[1], Full = ans$MAE[2],
    Difference = ans$difference)
}))
```

\Needspace{10\baselineskip}

```{r}
#| label: tbl-ch8b-resplits
#| echo: false
#| tbl-cap: "Fold-allocation sensitivity, not confidence limits."
knitr::kable(data.frame(
  Quantity = c("Education-only MAE", "Full MAE", "Full minus simple"),
  Mean = colMeans(ch8b_resplits),
  Minimum = apply(ch8b_resplits, 2, min),
  Maximum = apply(ch8b_resplits, 2, max), row.names = NULL
), digits = 3)
```

The mean difference is $-0.831149$, with range $-0.913974$ to
$-0.740997$; all 200 allocations favor the full rule. This describes
resplit sensitivity on the same data. The overlapping training sets,
reused outcomes, and repeated predictions are not independent new
samples. Neither that range nor a standard error computed by treating
fold means as independent is a sampling interval for future risk.

The internal result concerns learning rules trained at the stated
sizes, under the same-source assumptions. It is not an independent
test of the final all-98 fitted model. It cannot validate modern
occupations, another country, individual workers, or unseen types.
Nor does it establish which added predictor is responsible: prestige,
women, and type were added together. This remains retrospective
teaching reanalysis, not a historical preregistration claim.

\Needspace{12\baselineskip}

## Cook's distance and full deletion accounting {.unnumbered}

The main route interprets residual, leverage, and influence views and shows
the education-slope sensitivity. This section supplies the corresponding
Cook's distance formula and complete coefficient accounting.

With $p$ coefficient columns including the intercept, leverage is
$h_{ii}=x_i'(X'X)^{-1}x_i$. Cook's distance can be calculated as

$$
D_i=\frac{e_i^2}{p s^2}\frac{h_{ii}}{(1-h_{ii})^2}
=\frac{(\hat\beta-\hat\beta_{(-i)})'X'X
(\hat\beta-\hat\beta_{(-i)})}{p s^2}.
$$

The second expression measures a coefficient change using the design
matrix's metric, not an unweighted sum of squared coefficients with
incompatible units.

```{=latex}
\Needspace{12\baselineskip}
```

```{r}
#| label: tbl-ch8b-influence
#| echo: false
#| tbl-cap: "The five largest Cook's distances in the full fit."
knitr::kable(ch8_diag[ch8_top,
  c("occupation", "residual", "leverage", "cooks")],
  digits = 3, row.names = FALSE)
```

The record for general managers has the largest Cook's distance, approximately 0.447.
The main route explains why this observation is inspected and retained.

```{r}
#| label: tbl-ch8b-deletion-coefficients
#| echo: false
#| tbl-cap: "Coefficient sensitivity to the most influential record."
knitr::kable(data.frame(
  Term = names(coef(m2)), Full = coef(m2),
  Omit_one = coef(ch8_without_top),
  Difference = coef(ch8_without_top) - coef(m2), row.names = NULL
), digits = 3, row.names = FALSE)
```

The type coordinates also move. The table records a diagnostic perturbation
of the fit, not six new research questions or evidence that the reduced
sample is preferable.

## HC3 changes covariance, not the records or target {.unnumbered}

For independent observations, the HC3 covariance estimator is

$$
\widehat V_{\mathrm{HC3}}=(X'X)^{-1}
\left\{\sum_i x_ix_i'\frac{e_i^2}{(1-h_{ii})^2}\right\}
(X'X)^{-1}.
$$

The outer factors are the familiar design component. The middle
component allows different residual scales and adjusts for leverage
[@long2000; @zeileis2004]. It leaves the OLS coefficients unchanged.

```{r}
#| label: ch8b-hc3
#| echo: true
ch8b_Vhc3 <- sandwich::vcovHC(m2, type = "HC3")
ch8b_X <- model.matrix(m2)
ch8b_inverse <- solve(crossprod(ch8b_X))
ch8b_adjusted_e <- residuals(m2) / (1 - hatvalues(m2))
ch8b_manual <- ch8b_inverse %*%
  crossprod(ch8b_X * ch8b_adjusted_e) %*% ch8b_inverse
max(abs(ch8b_manual - ch8b_Vhc3))
```

```{r}
#| label: tbl-ch8b-hc3
#| echo: false
#| tbl-cap: "Coefficient estimates and two covariance summaries."
knitr::kable(data.frame(
  Term = names(coef(m2)), Estimate = coef(m2),
  Classical_SE = sqrt(diag(vcov(m2))),
  HC3_SE = sqrt(diag(ch8b_Vhc3)), row.names = NULL
), digits = 3, row.names = FALSE)
```

Education's HC3 SE is 0.382333 rather than 0.275226. Using $t_{92}$ as
an approximate reference gives $[-0.642912,0.875779]$. This is not an
exact finite-sample HC3 coverage result. Women's HC3 SE is smaller
than its classical counterpart: robust covariance need not widen
every interval.

For a maintained linear conditional mean, HC3 addresses unequal
variance under independent observations. For a random-design
best-linear-projection target, a suitable sandwich covariance can
support large-sample projection inference without making the mean
exactly linear. Neither interpretation repairs measurement error,
dependence, selection, causal identification, or absent joint support.

## VIF describes coefficient precision {.unnumbered}

For a single predictor column, classical conditional variance can be
written

$$
\operatorname{Var}(\hat\beta_j\mid X)=
\frac{\sigma^2}{\sum_i(X_{ij}-\bar X_j)^2}
\frac{1}{1-R_j^2}.
$$

Here $R_j^2$ comes from regressing that predictor on the remaining
columns, including the intercept. The variance inflation factor is
$VIF_j=1/(1-R_j^2)$; its square root is an SE multiplier, holding the
residual variance and focal predictor variation fixed [@kutner2005].

```{r}
#| label: ch8b-vif
#| echo: true
ch8b_aux <- lm(education ~ prestige + women + type, data = d)
ch8b_r2 <- summary(ch8b_aux)$r.squared
c(auxiliary_R2 = ch8b_r2, VIF = 1 / (1 - ch8b_r2),
  SE_multiplier = sqrt(1 / (1 - ch8b_r2)))
car::vif(m2)
```

Education gives $R_j^2=0.876495$, VIF 8.096833, and SE multiplier
2.845494. The predictor correlation between education and prestige
is 0.866480, but VIF uses all the other columns, not just that pair.

The two-column type term has a generalized VIF. Its raw GVIF is
7.700131; the dimension-adjusted value
$GVIF^{1/(2\,df)}=1.665806$ is on a comparable scale to the square
root of a one-column VIF [@fox2019]. Do not compare a raw two-column
GVIF with an ordinary one-column VIF as if their dimensions matched.

Exact collinearity creates rank deficiency. Near collinearity leaves
coefficients estimable but potentially imprecise. Under the maintained
zero-conditional-mean model it does not itself bias OLS. A cutoff of
5 or 10 neither establishes all assumptions nor justifies removing a
variable required by the question. Measurement problems also need
their own evidence: the familiar simple attenuation story requires
classical error assumptions and is not a universal rule for every
coefficient in a multiple-predictor model.

## Changed outcome scales and in-sample comparison criteria {.unnumbered}

The core already examines a prestige-quadratic sensitivity while preserving
the income outcome. A log-income model changes the outcome scale and therefore
requires an explicitly changed question.

```{r}
#| label: ch8b-outcome-scale
#| echo: true
ch8b_log <- lm(
  log(income) ~ education + prestige + women + type, data = d
)
100 * (exp(coef(ch8b_log)[c("education", "prestige", "women")]) - 1)
```

The log-income model asks about additive differences in expected log
income, or multiplicative differences in geometric means under that
model. A one-unit contrast gives $100(\exp(\hat\beta)-1)$ percent,
approximately $100\hat\beta$ for small coefficients. It is not
automatically an arithmetic-mean-income percentage effect. A changed
outcome requires a named changed question, not a claim that the data
have merely been cleaned.

```{r}
#| label: tbl-ch8b-log-sensitivity
#| echo: false
#| tbl-cap: "Log-model contrasts on a geometric-mean percent scale."
ch8b_terms <- c("education", "prestige", "women")
ch8b_log_ci <- confint(ch8b_log)[ch8b_terms, ]
knitr::kable(data.frame(
  Term = ch8b_terms, Log_coefficient = coef(ch8b_log)[ch8b_terms],
  Percent = 100 * (exp(coef(ch8b_log)[ch8b_terms]) - 1),
  Lower = 100 * (exp(ch8b_log_ci[, 1]) - 1),
  Upper = 100 * (exp(ch8b_log_ci[, 2]) - 1), row.names = NULL
), digits = 3, row.names = FALSE)
```

These intervals transform classical log-scale coefficient intervals.
Their interpretation remains conditional on the log-scale working
model; transformation does not supply a causal or population warrant.

For comparable Gaussian models on the same observations and outcome,

$$
AIC=-2\log\hat L+2k,\qquad
BIC=-2\log\hat L+k\log n.
$$

Here $k$ includes regression coefficients **and** estimated residual
variance. Smaller values are preferred under each criterion, but
neither is the observed held-out MAE [@akaike1974; @schwarz1978].

\Needspace{10\baselineskip}

```{r}
#| label: tbl-ch8b-criteria
#| echo: false
#| tbl-cap: "In-sample summaries for comparable income models."
ch8b_criteria <- do.call(rbind, lapply(
  list(m0 = m0, m1 = m1, m2 = m2), function(fit) {
    data.frame(n = nobs(fit), k = attr(logLik(fit), "df"),
      R2 = summary(fit)$r.squared,
      Adjusted_R2 = summary(fit)$adj.r.squared,
      AIC = AIC(fit), BIC = BIC(fit))
  }
))
knitr::kable(ch8b_criteria, digits = 3)
```

The raw AIC/BIC values for income and log income must not be compared
as if they used the same response density. A scale change changes the
likelihood's reference measure; a valid common-scale comparison would
need the corresponding transformation adjustment. The simpler lesson
here is to preserve the outcome and question in each comparison.

No residual graph establishes independence. The occupations share
one historical labour market, and metadata do not establish a
probability sample of a wider population. For students within schools
or repeated observations, the design may require dependence-aware
uncertainty. HC3 alone is not such a repair.

The practical response is consequential rather than pass/fail: retain
the named comparison, report its form and influence sensitivities,
distinguish classical from robust uncertainty, and bound interpretation
by measurement, support, and source. Causal adjustment-set rules about
confounders, mediators, and colliders apply to a declared causal target;
they are not universal requirements for association or prediction.

## Extended practice: reproduction and critique (Integrative) {.unnumbered}

1. Compute the white-collar conditional profile (11, 45, 50) and the
   empirical-profile standardized value, with appropriately labelled
   intervals. Name each target and assess support. Propose different
   weights and explain which part of the target changes.
2. Rebuild the 102-to-98 audit and three-model sequence from a clean
   session. Preserve names, units, type levels, and source row order.
   Add one supported display and a bounded occupation-level report.
3. A school-level analysis reports a slope 0.40, classical SE 0.18,
   and HC3 SE 0.31. Its residual smooth curves, spread increases,
   deleting one influential school changes the slope to -0.05, and
   sectors have little prior-score overlap. Assess the claim:
   "The causal effect is robust because VIFs are below 10; deleting
   the unusual school will make the model reliable." Match each
   finding to its consequence and propose a transparent sensitivity.
4. Choose another documented public dataset with a continuous outcome,
   two continuous predictors, and a factor with at least three levels.
   Declare an additive question and analytic sample before fitting.
   Document source, licence, units, missingness, coding, support,
   conditional interpretation, a block comparison, checks, and one
   sensitivity. Read the reported reasons and answer, then separately
   evaluate them. Do not require every changed estimator to imply a
   changed scientific estimand. Record versions and stochastic seeds.

# References {.unnumbered}

::: {#refs}
:::
