---
title: "Reading Group Comparisons: Categorical Outcomes and Mean Outcomes"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-cap: true
    tbl-cap: true
    fig-align: center
    include-in-header:
      - file: ../supplement/pdf-graphics-compat.tex
      - text: |
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "90%"
)
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(dev = "cairo_pdf")
}

options(digits = 4, width = 82, scipen = 4)

library(ggplot2)
library(carData)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "bottom"
    )
)

locate_data <- function(filename) {
  candidates <- c(
    file.path("supplement", filename),
    file.path("..", "supplement", filename)
  )
  found <- candidates[file.exists(candidates)]
  if (length(found) == 0L) stop("Cannot locate ", filename)
  normalizePath(found[[1]], winslash = "/")
}

fmt_p <- function(p) {
  if (p < .001) "< .001" else sub("^0", "", sprintf("%.3f", p))
}

fmt_num <- function(x, digits = 2) {
  formatC(x, format = "f", digits = digits)
}

uci <- read.csv(
  locate_data("student_performance.csv"),
  stringsAsFactors = FALSE
)
uci$higher_education <- factor(
  uci$higher_education,
  levels = c("no", "yes"),
  labels = c("No", "Yes")
)
uci$pass <- factor(uci$pass, levels = c("No", "Yes"))

uci_tab <- with(uci, table(higher_education, pass))
uci_prop <- prop.table(uci_tab, margin = 1)
uci_chi <- chisq.test(uci_tab, correct = FALSE)
uci_p_yes <- uci_prop["Yes", "Yes"]
uci_p_no <- uci_prop["No", "Yes"]
uci_n_yes <- sum(uci_tab["Yes", ])
uci_n_no <- sum(uci_tab["No", ])
uci_rd <- uci_p_yes - uci_p_no
uci_rd_se <- sqrt(
  uci_p_yes * (1 - uci_p_yes) / uci_n_yes +
    uci_p_no * (1 - uci_p_no) / uci_n_no
)
uci_rd_ci <- uci_rd + c(-1, 1) * qnorm(.975) * uci_rd_se
uci_rr <- uci_p_yes / uci_p_no
uci_or <- (uci_tab["Yes", "Yes"] / uci_tab["Yes", "No"]) /
  (uci_tab["No", "Yes"] / uci_tab["No", "No"])
uci_v <- sqrt(unname(uci_chi$statistic) / sum(uci_tab))

# The source-coded result remains primary.  This deletion analysis is a
# diagnostic for the unexplained cluster of final-grade zeros, not a recoding.
uci_zero <- uci$final_grade == 0
uci_zero_n <- sum(uci_zero)
uci_zero_all_absent_zero <- all(uci$absences[uci_zero] == 0)
uci_zero_prior_positive <- sum(uci$grade_period_2[uci_zero] > 0)
uci_nozero <- uci[!uci_zero, ]
uci_nozero_tab <- with(uci_nozero, table(higher_education, pass))
uci_nozero_prop <- prop.table(uci_nozero_tab, margin = 1)
uci_nozero_p_yes <- uci_nozero_prop["Yes", "Yes"]
uci_nozero_p_no <- uci_nozero_prop["No", "Yes"]
uci_nozero_n_yes <- sum(uci_nozero_tab["Yes", ])
uci_nozero_n_no <- sum(uci_nozero_tab["No", ])
uci_nozero_rd <- uci_nozero_p_yes - uci_nozero_p_no
uci_nozero_se <- sqrt(
  uci_nozero_p_yes * (1 - uci_nozero_p_yes) / uci_nozero_n_yes +
    uci_nozero_p_no * (1 - uci_nozero_p_no) / uci_nozero_n_no
)
uci_nozero_ci <- uci_nozero_rd + c(-1, 1) * qnorm(.975) * uci_nozero_se
uci_nozero_chi <- chisq.test(uci_nozero_tab, correct = FALSE)
uci_nozero_fisher <- fisher.test(uci_nozero_tab)
stopifnot(
  uci_zero_n == 38L,
  uci_zero_all_absent_zero,
  uci_zero_prior_positive == 25L
)

data("Baumann")
reading <- Baumann
reading$group <- factor(
  reading$group,
  levels = c("Basal", "DRTA", "Strat"),
  labels = c("DRA", "DRTA", "Think-aloud")
)

reading_groups <- levels(reading$group)
reading_summary <- do.call(
  rbind,
  lapply(reading_groups, function(g) {
    z <- reading$post.test.1[reading$group == g]
    data.frame(
      Group = g,
      Assigned = c(22L, 23L, 23L)[match(g, reading_groups)],
      Observed = length(z),
      Mean = mean(z),
      SD = sd(z),
      Minimum = min(z),
      Maximum = max(z),
      stringsAsFactors = FALSE
    )
  })
)

reading_aov <- aov(post.test.1 ~ group, data = reading)
reading_anova <- anova(reading_aov)
reading_welch <- oneway.test(
  post.test.1 ~ group,
  data = reading,
  var.equal = FALSE
)
reading_tukey <- TukeyHSD(reading_aov)$group

reading_means <- setNames(reading_summary$Mean, reading_summary$Group)
reading_ns <- setNames(reading_summary$Observed, reading_summary$Group)
reading_mse <- reading_anova["Residuals", "Mean Sq"]
reading_df <- reading_anova["Residuals", "Df"]
reading_contrast_names <- c("DRTA minus DRA", "Think-aloud minus DRA")
reading_contrast_est <- c(
  reading_means["DRTA"] - reading_means["DRA"],
  reading_means["Think-aloud"] - reading_means["DRA"]
)
reading_contrast_se <- c(
  sqrt(reading_mse * (1 / reading_ns["DRTA"] + 1 / reading_ns["DRA"])),
  sqrt(reading_mse *
    (1 / reading_ns["Think-aloud"] + 1 / reading_ns["DRA"]))
)
reading_bonf_critical <- qt(1 - .05 / (2 * 2), reading_df)
reading_contrast_low <- reading_contrast_est -
  reading_bonf_critical * reading_contrast_se
reading_contrast_high <- reading_contrast_est +
  reading_bonf_critical * reading_contrast_se
reading_contrast_t <- reading_contrast_est / reading_contrast_se
reading_contrast_p <- 2 * pt(-abs(reading_contrast_t), reading_df)
reading_ss_between <- reading_anova["group", "Sum Sq"]
reading_ss_total <- sum(reading_anova[, "Sum Sq"])
reading_eta2 <- reading_ss_between / reading_ss_total
reading_omega2 <- (
  reading_ss_between -
    reading_anova["group", "Df"] * reading_mse
) / (reading_ss_total + reading_mse)

# Bounded completed-data stress test.  The two unavailable scores can each
# take an integer value on the documented 0--16 outcome scale.
reading_completion_grid <- expand.grid(
  drta_missing = 0:16,
  think_missing = 0:16
)
reading_completion_results <- do.call(
  rbind,
  lapply(seq_len(nrow(reading_completion_grid)), function(i) {
    x_d <- reading_completion_grid$drta_missing[i]
    x_t <- reading_completion_grid$think_missing[i]
    completed <- rbind(
      reading[c("group", "post.test.1")],
      data.frame(group = factor("DRTA", levels = reading_groups),
                 post.test.1 = x_d),
      data.frame(group = factor("Think-aloud", levels = reading_groups),
                 post.test.1 = x_t)
    )
    fit <- aov(post.test.1 ~ group, data = completed)
    fit_anova <- anova(fit)
    mse <- fit_anova["Residuals", "Mean Sq"]
    df <- fit_anova["Residuals", "Df"]
    means <- tapply(completed$post.test.1, completed$group, mean)
    ns <- table(completed$group)
    estimates <- c(means["DRTA"] - means["DRA"],
                   means["Think-aloud"] - means["DRA"])
    ses <- c(
      sqrt(mse * (1 / ns["DRTA"] + 1 / ns["DRA"])),
      sqrt(mse * (1 / ns["Think-aloud"] + 1 / ns["DRA"]))
    )
    critical <- qt(1 - .05 / (2 * 2), df)
    data.frame(
      drta_estimate = estimates[1],
      think_estimate = estimates[2],
      drta_low = estimates[1] - critical * ses[1],
      drta_high = estimates[1] + critical * ses[1],
      think_low = estimates[2] - critical * ses[2],
      think_high = estimates[2] + critical * ses[2]
    )
  })
)
reading_drta_completed_bounds <- range(reading_completion_results$drta_estimate)
reading_think_completed_bounds <- range(reading_completion_results$think_estimate)
reading_drta_min_low <- min(reading_completion_results$drta_low)
reading_think_low_range <- range(reading_completion_results$think_low)
reading_think_high_range <- range(reading_completion_results$think_high)

draw_reading_route <- function() {
  old <- par(mar = rep(0.15, 4), xpd = NA)
  on.exit(par(old))
  plot.new()
  plot.window(xlim = c(0, 18), ylim = c(0, 6))
  x <- c(.3, 3.9, 7.5, 11.1, 14.7)
  labels <- c(
    "Comparison\nquestion",
    "Outcome form\nand design",
    "Counts or raw\ngrouped values",
    "Estimate and\nuncertainty",
    "Design-bounded\nanswer"
  )
  fills <- c("#DCEAF3", "#E3EDDD", "#F7EDC9", "#F3DDC9", "#E8DCEB")
  for (i in seq_along(x)) {
    rect(
      x[i], 2.0, x[i] + 3.0, 4.3,
      col = fills[i], border = "#29434E", lwd = 1.2
    )
    text(
      x[i] + 1.5, 3.15, labels[i],
      cex = .78, font = 2, col = "#26363D"
    )
    if (i < length(x)) {
      arrows(
        x[i] + 3.05, 3.15, x[i + 1] - .08, 3.15,
        length = .07, lwd = 1.2, col = "#50656E"
      )
    }
  }
  text(
    9, 5.15,
    "Read the support in question order",
    font = 2, cex = .95, col = "#244A63"
  )
  text(
    9, 1.15,
    "One estimand may have several outputs without becoming several strands",
    cex = .73, col = "#6C513B"
  )
}

draw_strand_map <- function() {
  old <- par(mar = rep(.15, 4), xpd = NA)
  on.exit(par(old))
  plot.new()
  plot.window(xlim = c(0, 16), ylim = c(0, 12))

  box <- function(x, y, w, h, label, fill, border = "#314955",
                  cex = .78) {
    rect(x, y, x + w, y + h, col = fill, border = border, lwd = 1.1)
    text(x + w / 2, y + h / 2, label, cex = cex, font = 2,
         col = "#24343B")
  }
  arrow <- function(x1, y1, x2, y2, col = "#536B76", lty = 1) {
    arrows(x1, y1, x2, y2, length = .065, lwd = 1.05,
           col = col, lty = lty)
  }

  text(.35, 11.45, "STATISTICAL LAYER", adj = 0, font = 2,
       cex = .78, col = "#244A63")
  box(.4, 8.65, 2.5, 1.65, "Shared\nSTAT-E", "#DCEAF3")
  box(3.35, 8.65, 2.5, 1.65, "Shared\nSTAT-J", "#E3EDDD")
  box(6.55, 9.45, 2.55, 1.55, "STAT-I1\nDRTA-DRA", "#F7EDC9")
  box(9.55, 9.45, 2.55, 1.55, "STAT-C1\npositive", "#F3DDC9")
  box(6.55, 7.15, 2.55, 1.55, "STAT-I2\nThink-DRA", "#F7EDC9")
  box(9.55, 7.15, 2.55, 1.55, "STAT-C2\ninconclusive", "#F3DDC9")
  arrow(2.9, 9.48, 3.35, 9.48)
  arrow(5.85, 9.48, 6.55, 10.22)
  arrow(5.85, 9.48, 6.55, 7.92)
  arrow(9.1, 10.22, 9.55, 10.22)
  arrow(9.1, 7.92, 9.55, 7.92)

  text(.35, 5.55, "SUBSTANTIVE LAYER", adj = 0, font = 2,
       cex = .78, col = "#6C513B")
  box(.4, 2.65, 2.5, 1.65, "SUB-E\nsetting", "#E8DCEB")
  box(3.35, 2.65, 2.5, 1.65, "SUB-J\ncontext", "#E8DCEB")
  box(6.55, 2.65, 2.55, 1.65, "SUB-I\nretain both", "#E8DCEB")
  box(9.55, 2.65, 2.55, 1.65, "SUB-C\none bounded\nanswer", "#E8DCEB")
  arrow(2.9, 3.48, 3.35, 3.48, "#735D73")
  arrow(5.85, 3.48, 6.55, 3.48, "#735D73")
  arrow(9.1, 3.48, 9.55, 3.48, "#735D73")

  # Routed convergence arrows stop at the substantive component boundaries.
  arrow(1.65, 8.65, 1.65, 4.3, "#A36D34", 2)
  arrow(4.6, 8.65, 4.6, 4.3, "#A36D34", 2)
  segments(7.82, 9.45, 7.82, 8.95, col = "#A36D34", lty = 2)
  segments(7.82, 8.95, 6.15, 8.95, col = "#A36D34", lty = 2)
  segments(6.15, 8.95, 6.15, 5.25, col = "#A36D34", lty = 2)
  segments(7.82, 7.15, 7.82, 6.55, col = "#A36D34", lty = 2)
  segments(7.82, 6.55, 6.15, 6.55, col = "#A36D34", lty = 2)
  arrow(6.15, 5.25, 7.82, 4.3, "#A36D34", 2)
  segments(10.82, 9.45, 10.82, 8.95, col = "#A36D34", lty = 2)
  segments(10.82, 8.95, 12.75, 8.95, col = "#A36D34", lty = 2)
  segments(10.82, 7.15, 10.82, 6.55, col = "#A36D34", lty = 2)
  segments(10.82, 6.55, 12.75, 6.55, col = "#A36D34", lty = 2)
  segments(12.75, 8.95, 12.75, 5.15, col = "#A36D34", lty = 2)
  arrow(12.75, 5.15, 10.82, 4.3, "#A36D34", 2)

  text(14.1, 9.85, "One question", cex = .75, font = 2,
       col = "#244A63")
  text(14.1, 8.95, "Two estimands", cex = .72, col = "#244A63")
  text(14.1, 8.05, "Two strands", cex = .72, col = "#244A63")
  text(14.1, 3.85, "One integrated", cex = .72, font = 2,
       col = "#6C513B")
  text(14.1, 3.05, "answer", cex = .72, font = 2,
       col = "#6C513B")
  text(8.0, 1.25,
       "The omnibus ANOVA is contextual output, not a third strand",
       cex = .68, col = "#6C513B")
}

draw_workflow <- function() {
  old <- par(mar = rep(0.2, 4), xpd = NA)
  on.exit(par(old))
  plot.new()
  plot.window(xlim = c(0, 14), ylim = c(0, 8))
  rect(0.15, 4.15, 13.85, 7.60, col = "#EEF5F8", border = "#9BB6C4")
  rect(0.15, 0.25, 13.85, 3.70, col = "#F1F6EF", border = "#A7BFA2")
  text(
    0.38, 7.30,
    "PHASE 1 - UNDERSTAND AND PLAN | iterative Stages 1-3",
    adj = 0, font = 2, cex = .66, col = "#244A63"
  )
  text(
    0.38, 3.40,
    "PHASE 2 - PRODUCE AND COMPLETE | iterative Stages 4-6",
    adj = 0, font = 2, cex = .64, col = "#466B43"
  )
  x <- c(.48, 5.05, 9.62)
  width <- 3.88
  labels <- c(
    "Review problem\nand literature",
    "Frame question and\ncandidate answer",
    "Align design, measurement,\nestimand, evidence,\nand inferential plan",
    "Produce or acquire\nand document evidence",
    "Estimate, check,\nand analyse",
    "Contextualize, interpret,\nand complete bounded answer"
  )
  stage_box <- function(x0, y0, number, label) {
    rect(
      x0, y0, x0 + width, y0 + 1.62,
      col = "white", border = "#344A55", lwd = 1
    )
    text(
      x0 + .18, y0 + 1.34, paste0("STAGE ", number),
      adj = 0, font = 2, cex = .72, col = "#203744"
    )
    text(x0 + width / 2, y0 + .66, label, cex = .65, col = "#29353B")
  }
  for (i in 1:3) stage_box(x[i], 4.65, i, labels[i])
  bottom <- c(6, 5, 4)
  for (i in 1:3) stage_box(x[i], .75, bottom[i], labels[bottom[i]])
  for (i in 1:2) {
    arrows(
      x[i] + width + .02, 5.46, x[i + 1] - .02, 5.46,
      length = .07, col = "#4F6875"
    )
  }
  for (i in 3:2) {
    arrows(
      x[i] - .02, 1.56, x[i - 1] + width + .02, 1.56,
      length = .07, col = "#557052"
    )
  }
  arrows(
    x[3] + width / 2, 4.63, x[3] + width / 2, 2.39,
    length = .07, col = "#6B6B6B"
  )
  arrows(
    x[3] + width / 2, 6.58, x[1] + width / 2, 6.58,
    length = .07, lty = 2, col = "#7A4B2A"
  )
  text(
    7, 6.84, "revise and realign within Phase 1",
    cex = .52, col = "#7A4B2A"
  )
  arrows(
    x[1] + width / 2, .64, x[3] + width / 2, .64,
    length = .07, lty = 2, col = "#7A4B2A"
  )
  text(
    7, .40, "verify, reanalyse, and recalibrate within Phase 2",
    cex = .51, col = "#7A4B2A"
  )
}
```

{{< pagebreak >}}

::: {.callout-tip title="Chapter box"}
**Central question:** How can a reader understand and compare group results
when the outcome is categorical or quantitative?

**Focal concepts:** Denominators, conditional proportions, grouped
distributions, estimands, omnibus questions, single and multiple statistical
strands, simultaneous uncertainty, and design-bounded claims.

**Main evidence:** One compact observational comparison from the UCI student
record and one transparently hypothetical teaching extension based on the
documented Baumann reading-study record. The first pair has one estimand and
one statistical strand. The second has one integrated question and two
subordinate contrast estimands.

**Scope boundary:** The main route covers ordinary two-way tables and one-way
group-mean comparisons, including the basic ANOVA model, sums of squares,
degrees of freedom, and $F$ calculation. Sparse tables, extended effect scales,
full ANOVA derivations, robust alternatives, blocking, and software
implementation are optional depth.

**Use hierarchy:** Reading published comparisons is the primary use. Quality
evaluation is the distinct secondary use after reading. The six-stage
construction workflow is the auxiliary use for conducting research.
:::

Chapter 1 supplies the shared argument-based vocabulary; Chapter 1a is not
required. The calculations build on the means, variances, standard errors,
intervals, and tests developed in Chapters 2--3. Brief reminders are included,
but introductory statistics and basic algebra are useful preparation. The
public route requires no programming or statistical software.

# Begin with the Comparison Question

Consider two statements that could appear in a report:

> Students intending higher education were more likely to pass.

> The instructional methods differed significantly.

Both may sound informative, yet neither is a complete answer. The first omits
the denominator, numerical difference, source population, and observational
boundary. The second omits the outcome, the groups, the comparison that
matters, and the size and uncertainty of the difference.

A reader can first interpret either statement by building a **comparison record**;
subsequent quality evaluation may show that its wording needs repair:

1. **Unit and analytic sample:** What does one observation represent, and
   which observations entered the result?
2. **Group or assigned condition:** Which groups are being compared, and were
   they observed or assigned?
3. **Outcome:** Is it a named event or a numerical scale with units?
4. **Estimand:** Is the quantity sought an event-probability
   difference, an equality pattern, a mean difference, or another declared
   contrast?
5. **Design-supported scope:** Which people, settings, processes, or
   assignments can the answer describe?

| Outcome form | First evidence to inspect | Typical estimand | First effect estimate |
|:-------------|:--------------------------|:---------------|:----------------------|
| Categorical | Counts and conditional proportions | Difference in event probabilities or a whole-table association | Proportion difference for a named event and direction |
| Quantitative | Raw grouped observations and group summaries | Difference or contrast among group means | Mean difference or prespecified linear contrast |

: *A reading-first orientation to group comparisons.* {#tbl-comparison-orientation tbl-colwidths="[18,28,29,25]"}

This table is an orientation, not a mechanical procedure selector. The design
and the intended claim still determine whether a repeated-sample or causal
interpretation is justified.

An **estimand** is the precisely specified quantity the study seeks to learn;
an **estimate** is the numerical result obtained from the observed evidence for
that estimand. Several tables, intervals, tests, or sensitivity checks may
describe one estimand without creating several arguments. Several estimands
become **statistical strands** only when all are subordinate to one integrated
answer and none is owed a separate question-facing conclusion. If a report
interprets a contrast separately, that conclusion belongs to its own
question--answer pair and argument.

```{r}
#| label: fig-comparison-route
#| fig-cap: "A reading route from the comparison question to a design-bounded answer."
#| fig-alt: "Five linked boxes move from comparison question to outcome form and design, counts or raw grouped values, estimate and uncertainty, and finally a design-bounded answer."
#| fig-width: 6.5
#| fig-height: 2.4
#| out-width: "100%"

draw_reading_route()
```

The remainder of the main route follows @fig-comparison-route twice. The UCI
record supplies a categorical observational comparison. The Baumann-inspired
teaching contract supplies a quantitative randomized comparison. They share a reading discipline,
but they do not share evidence or one combined argument.

\needspace{9\baselineskip}

# Read a Categorical Group Comparison {#sec-categorical-comparison}

## Start with the denominator

The first question is:

> Among students recorded in two Portuguese secondary schools, how did the
> observed passing proportion differ between students who did and did not
> report an intention to pursue higher education?

The UCI Student Performance record contains 395 students from two schools
[@cortez2008data]. One row represents one recorded student. Higher-education
intention was observed, not assigned. Passing is defined here as a final grade
of at least 10 on the source scale.

The source documentation defines final grade on a 0--20 scale and marks the
file as having no missing values. It does not, however, explain an unusual
joint pattern: 38 records have final grade 0 and absences 0, although 25 of
those records have a positive second-period grade. A zero may be a genuine
lowest grade, but the accessible record does not establish whether every zero
has the same meaning. The primary analysis therefore keeps the published
numeric coding exactly as recorded and labels the unresolved pattern; a later
deletion analysis is only a diagnostic sensitivity check.

The question asks for the proportion passing **within each intention group**.
The group totals are therefore the denominators:

$$
\widehat p_g =
\frac{\text{number passing in group }g}
     {\text{number recorded in group }g}.
$$

```{r}
#| label: tbl-uci-counts
#| tbl-cap: "Pass status and within-group passing proportions from grades as recorded in the UCI record"

uci_counts_display <- data.frame(
  "Higher-education intention" = rownames(uci_tab),
  "Did not pass" = as.integer(uci_tab[, "No"]),
  "Passed" = as.integer(uci_tab[, "Yes"]),
  "Group total" = as.integer(rowSums(uci_tab)),
  "Passing proportion" = sprintf("%.3f", uci_prop[, "Yes"]),
  check.names = FALSE
)

knitr::kable(
  uci_counts_display,
  booktabs = TRUE,
  row.names = FALSE,
  align = c("l", "r", "r", "r", "r")
)
```

The group that reported an intention to pursue higher education was much
larger: 375 students, compared with 20 who did not. Counts alone therefore
mislead. Within the two groups, the passing proportions were .688 and .350.

Column percentages would answer a different question: among students who
passed or did not pass, what proportion reported higher-education intention?
Neither row nor column percentages are universally correct. The research
question chooses the denominator.

```{r}
#| label: fig-uci-proportions
#| fig-cap: "Within-group passing proportions derived from grades as recorded in the UCI record."
#| fig-alt: "Two bars show that 35.0 percent of students without a reported higher-education intention passed, compared with 68.8 percent of students with that intention. Group sizes are 20 and 375."
#| fig-width: 6.6
#| fig-height: 3.8

uci_plot <- data.frame(
  Intention = factor(c("No", "Yes"), levels = c("No", "Yes")),
  Proportion = c(uci_p_no, uci_p_yes),
  n = c(uci_n_no, uci_n_yes)
)

ggplot(uci_plot, aes(Intention, Proportion, fill = Intention)) +
  geom_col(width = .58, show.legend = FALSE) +
  geom_text(
    aes(label = sprintf("%.1f%%\n(n = %d)", 100 * Proportion, n)),
    vjust = -0.2, size = 3.5
  ) +
  scale_y_continuous(
    limits = c(0, .82),
    breaks = seq(0, .8, .2),
    labels = function(x) paste0(round(100 * x), "%"),
    expand = expansion(mult = c(0, .02))
  ) +
  scale_fill_manual(values = c("No" = "#C98E5A", "Yes" = "#5A8FA8")) +
  labs(
    x = "Reported intention to pursue higher education",
    y = "Passing proportion"
  )
```

For the declared event and direction, the primary descriptive effect is

$$
\widehat\Delta
=
\widehat p_{\text{yes}}-\widehat p_{\text{no}}
=.688-.350=.338.
$$

Thus, among these records, the passing proportion was 33.8 percentage points
higher in the intention-yes group. This is already a complete **descriptive**
answer for the observed table. No p-value is needed to know that realized
difference.

Other scales answer other questions. **Odds** are $p/(1-p)$: passing
probability .75 gives odds $.75/.25=3$, or three passes per nonpass. A
probability ratio compares probabilities; an **odds ratio** compares odds,
not probabilities. Appendix A develops these scales. The proportion
difference remains primary because percentage points directly answer our
question.

## Read evidence against an independence pattern

Inference adds a question that description alone does not answer:

> Would a discrepancy of this size be unusual under a declared
> no-association model or assignment process?

For a two-way table, **independence** means that the modeled distribution of
pass status is the same across intention groups. Under that restriction, the
expected count in row $r$ and column $c$ is

$$
E_{rc}=
\frac{(\text{row total})(\text{column total})}{n}.
$$

For example, the no-intention group contains 20 students and the table contains
265 passes. The expected no-intention/pass count under independence is
$(20)(265)/395=13.42$, compared with 7 observed passes. The Pearson statistic
collects the cell discrepancies:

$$
X^2=\sum\frac{(O-E)^2}{E}.
$$

Here $X^2$ names the calculated Pearson statistic; $\chi^2$ names its
chi-square reference distribution. Reports often use $\chi^2$ for the
reported statistic too; these are notation conventions, not different tests.

Here $O$ and $E$ are an observed and expected cell count, and $\sum$ means
add the contributions across all cells. For the no-intention/pass cell, the
contribution is $(7-13.4177)^2/13.4177=3.0696$. The other three contributions
are 6.2573, 0.1637, and 0.3337, giving $X^2=9.8243$ after rounding. The
reference degrees of freedom are $(R-1)(C-1)$ for $R$ rows and $C$ columns:
here $(2-1)(2-1)=1$. Once both margins are fixed, specifying one cell fixes
the remaining cells; they cannot vary independently.

It is an omnibus statistic for the whole independence restriction. It is not
an effect size, a causal estimate, or a locator that by itself explains which
cells produced the pattern.

## Make the model-based uncertainty checkable

The recorded-table question has already been answered descriptively. For an
explicitly conditional illustration of inference, consider comparable
records from the same stipulated process. Write $\pi_g=P(Y=1\mid G=g)$ for
the passing probability within intention group $g$, where $Y=1$ means pass.
The model estimand is $\Delta=\pi_{\mathrm{yes}}-\pi_{\mathrm{no}}$;
its estimate is the observed difference $\widehat\Delta=.338$. That model
quantity differs from the as-recorded 395-record description. The model
calculation is contextual output here, not uncertainty about the already
known table or an independently established conclusion about other schools.

Under independent binomial sampling within the two groups, estimate the
standard error by adding the variances of their estimated proportions:

$$
\widehat{SE}(\widehat\Delta)=
\sqrt{\frac{\widehat p_1(1-\widehat p_1)}{n_1}
      +\frac{\widehat p_0(1-\widehat p_0)}{n_0}}
=\sqrt{\frac{.688(.312)}{375}+\frac{.350(.650)}{20}}
=.1093.
$$

Subscripts 1 and 0 denote the yes and no groups. An illustrative Wald interval
is $.338\pm1.96(.1093)=[.124,.552]$. The small 20-student group makes this
normal approximation fragile; a score-based interval is generally preferable
for a substantive analysis. The calculation is retained here to make the
standard-error mechanism visible, not as a recommendation to use Wald
intervals routinely.

The no-association test asks a different numerical question: if both groups
shared one passing probability, how surprising would the discrepancy be?
Its pooled estimate is $\widehat p=265/395=.6709$, so its null standard error
is $\sqrt{\widehat p(1-\widehat p)(1/375+1/20)}=.1078$. The resulting
$z=.338/.1078=3.1344$ has $z^2=9.8243$, the uncorrected Pearson statistic
for this $2\times2$ table. The pooled null SE is not the unpooled estimation
SE used for the displayed interval. These are outputs about the same directed
model difference under different restrictions, not new statistical strands.

```{r}
#| label: tbl-uci-results
#| tbl-cap: "Categorical estimate, interval, and omnibus evidence derived from grades as recorded"

uci_results_display <- data.frame(
  "Result" = c(
    "Passing proportion, intention yes",
    "Passing proportion, intention no",
    "Difference, yes minus no",
    "Large-sample 95% CI for difference",
    "Pearson chi-square",
    "Degrees of freedom",
    "p value"
  ),
  "Value" = c(
    sprintf("%.3f", uci_p_yes),
    sprintf("%.3f", uci_p_no),
    sprintf("%.3f", uci_rd),
    sprintf("[%.3f, %.3f]", uci_rd_ci[1], uci_rd_ci[2]),
    sprintf("%.2f", unname(uci_chi$statistic)),
    sprintf("%d", unname(uci_chi$parameter)),
    fmt_p(uci_chi$p.value)
  ),
  check.names = FALSE
)

knitr::kable(
  uci_results_display,
  booktabs = TRUE,
  row.names = FALSE,
  align = c("l", "r")
)
```

@tbl-uci-results keeps four tasks separate:

- The **proportion difference** gives direction and magnitude on the selected
  event scale.
- The **interval** displays values compatible with a hypothetical
  record-generating model. The displayed Wald interval is illustrative and
  should be treated cautiously because the no-intention group has only 20
  students.
- The **chi-square result**, chi-square with 1 degree of freedom = 9.82,
  $p=.002$, assesses the whole
  independence restriction under its reference conditions.
- The **question wording** determines whether the mode is descriptive,
  associational, predictive, or causal. The **design and justification**
  determine which claim is warranted.

All four expected counts exceed 5 in this table, but that fact is not a
universal license to trust the reference distribution. The analysis also
requires a defensible account of dependence, selection, clustering,
measurement, and the stochastic process to which the p-value refers. Students
share schools and perhaps classrooms, and the source is not a probability
sample of a broad population.

### Check the unexplained zero-grade cluster

If the 38 final-grade-zero records are omitted solely as a diagnostic, 357
records remain. The passing proportions become .752 in the intention-yes group
($258/343$) and .500 in the intention-no group ($7/14$), a difference of .252.
The illustrative Wald interval becomes [-.014, .518]. Pearson's
$X^2(1)=4.47$, $p=.034$, but its smallest expected count is only 3.61;
Fisher's two-sided exact test gives $p=.055$. Thus, even this simple coding
decision materially changes the numerical evidence.

This is not permission to delete the zeros. Doing so changes the analytic set
and the finite-record quantity being described, and it requires a defensible
record-definition or missingness rule that is not located in the source
documentation. The comparison based on all 395 rows remains the primary
source-coded result; the diagnostic shows why its meaning and process-level
extension require caution.

\Needspace{16\baselineskip}
::: {.callout-important title="Bounded categorical answer"}
Among the 395 recorded students, the observed passing proportions were 68.8%
for those reporting an intention to pursue higher education and 35.0% for
those not reporting it, a difference of 33.8 percentage points. Under an
illustrative independent-record model for comparable records from the same
process, a large-sample 95% interval was 12.4 to 55.2 percentage points and the
independence statistic was chi-square with 1 degree of freedom = 9.82,
$p=.002$. A diagnostic omission of the 38 unexplained final-grade-zero records
reduced the difference to 25.2 percentage points and made the inferential
conclusion sensitive to the procedure (Pearson $p=.034$; Fisher $p=.055$).
The all-row comparison is
exploratory, local, and observational. It does not show that intending higher
education caused passing or support transport to students outside the
documented setting and selection process, and the accessible materials do not
settle the meaning of the zero-grade cluster.
:::

# Read a Mean Comparison with One Integrated Answer {#sec-mean-comparison}

## Define the integrated comparison question

The second example deliberately separates a **source record** from a
**chapter-created teaching contract**. Baumann and colleagues assigned 68
fourth-grade students in one rural Midwestern school: 22 to directed reading
activity (DRA), 23 to directed reading-thinking activity (DRTA), and 23 to
think-aloud instruction [@baumann1992, p. 147]. One student in each innovative
condition had an incomplete outcome because of illness or transfer. The public
*Baumann* record in the *carData* package therefore contains 66 observed
outcomes, 22 per group [@carData2022; @fox2019].

The source study recorded two pretests and three post-tests. This chapter
selects the 0--16 error-detection outcome and uses an **unadjusted one-way
model**, rather than reproducing the source's pretest-adjusted analysis.
The uncertainty calculations are model-based, not finite-sample
randomization intervals. The package cannot recover the two unobserved
scores. These are limits of this teaching reanalysis, not details supplied
by an ANOVA result.

The chapter does **not** claim that
the source investigators prespecified the question, contrasts, or decision
rule below. Instead, imagine a future replication in this setting whose
teaching contract declares, before its outcomes are inspected:

> Does a two-format instructional package meet its integrated success
> criterion: both DRTA assignment and think-aloud assignment improve mean
> error-detection score relative to DRA assignment?

For this hypothetical contract, *improve* means a positive mean difference.
There are two subordinate estimands:

1. mean error-detection score under DRTA assignment minus mean score under DRA
   assignment; and
2. mean score under think-aloud assignment minus mean score under DRA
   assignment.

For this hypothetical replication, the intended means refer to outcomes
under each assignment for comparable eligible pupils in the stipulated
setting. The inferential calculations use an independent-pupil outcome model,
not a claim that this school represents all fourth-graders. All assignees
belong in that intended comparison; if only complete outcomes are analysed,
linking observed means to those intended means requires an explicit
missingness justification. The chapter reports that limitation rather than
quietly redefining the estimands as effects among only the observed pupils.

The declared rule requires the lower endpoint of **each** two-sided 95%
Bonferroni simultaneous interval for those two comparisons to exceed zero.
Zero is a directional benchmark, not a threshold for educational importance.
The criterion is prospective only inside this explicitly hypothetical
replication. Applying it to an already available source record is a post hoc,
secondary teaching reanalysis.

The two estimands count as statistical strands because both are needed for the
one package-level answer and neither receives its own recommendation. If a
report instead concludes separately about DRTA or think-aloud, it owes a
separate question--answer pair and argument for that conclusion.

## See grouped evidence and read ANOVA

The assignment record and observed outcomes come before a test.

```{r}
#| label: tbl-baumann-record
#| tbl-cap: "Source assignment record and observed error-detection summaries"

baumann_display <- reading_summary
baumann_display$Mean <- sprintf("%.2f", baumann_display$Mean)
baumann_display$SD <- sprintf("%.2f", baumann_display$SD)
baumann_display$Range <- paste0(
  baumann_display$Minimum, "--", baumann_display$Maximum
)
baumann_display$Minimum <- NULL
baumann_display$Maximum <- NULL

knitr::kable(
  baumann_display,
  booktabs = TRUE,
  row.names = FALSE,
  align = c("l", "r", "r", "r", "r", "r")
)
```

The assigned counts and observed counts are not interchangeable. The 66
complete records support direct description of observed outcomes. Extending
that analysis to an all-assigned estimand requires a missingness justification;
ANOVA cannot supply one.

```{r}
#| label: fig-baumann-distributions
#| fig-pos: H
#| fig-cap: "Observed error-detection scores by assigned instructional condition."
#| fig-alt: "All 66 observed scores are plotted by assigned condition. DRA has the lowest mean, DRTA the highest, and think-aloud lies between them. The distributions overlap, and think-aloud has the widest spread."
#| fig-width: 7.2
#| fig-height: 4.1

ggplot(reading, aes(group, post.test.1, color = group)) +
  geom_boxplot(
    width = .44, outlier.shape = NA, alpha = .18,
    show.legend = FALSE
  ) +
  geom_point(
    position = position_jitter(width = .08, height = 0, seed = 404),
    alpha = .72, size = 1.7, show.legend = FALSE
  ) +
  stat_summary(
    fun = mean, geom = "point", shape = 18, size = 3.5,
    color = "#1F2E35"
  ) +
  scale_color_manual(values = c("#9B6A45", "#356E8D", "#6E7F48")) +
  labs(x = "Assigned condition", y = "Error-detection score (0--16)")
```

Before seeing an ANOVA table, @fig-baumann-distributions and
@tbl-baumann-record establish four facts:

- the group centers differ;
- the observations overlap;
- within-group variation is substantial; and
- the think-aloud group is more dispersed than the other two groups.

The graph is evidence about the observed distributions. It does not prove a
model assumption or a causal claim.

### Read the omnibus ANOVA question

One-way ANOVA represents each observed score as

$$
\text{observed score}
=
\text{fitted group mean}
+
\text{within-group residual}.
$$

The corresponding **group-mean model** is $Y_{ij}=\mu_j+\varepsilon_{ij}$:
$Y_{ij}$ is pupil $i$'s score in group $j$, $\mu_j$ is that condition's model
mean, and $\varepsilon_{ij}$ is the pupil's deviation from it. Assume mean-zero
errors, a common variance $\sigma^2$, and independent pupils for this working
model. The fitted mean is the observed group mean $\bar Y_j$; a residual is
$Y_{ij}-\bar Y_j$. A fitted residual is not the same as the unknown model
error $\varepsilon_{ij}$.

### From deviations to sums of squares

The **grand mean** $\bar Y$ is the mean of all $n=66$ observed scores; it is
8.0758. Each deviation from it splits into two pieces:

$$
Y_{ij}-\bar Y=(\bar Y_j-\bar Y)+(Y_{ij}-\bar Y_j).
$$

For a DRA pupil scoring 5, this is
$5-8.0758=(6.6818-8.0758)+(5-6.6818)=-1.3940-1.6818$.
The first piece locates the group mean; the second locates the pupil within
that group. Square and add across pupils and groups to obtain

$$
\underbrace{\sum_j\sum_i(Y_{ij}-\bar Y)^2}_{SS_T}
=\underbrace{\sum_j n_j(\bar Y_j-\bar Y)^2}_{SS_B}
+\underbrace{\sum_j\sum_i(Y_{ij}-\bar Y_j)^2}_{SS_W}.
$$

The double summation means sum over pupils within each group and then over
groups. $SS_T$, $SS_B$, and $SS_W$ are total, between-group, and within-group
sums of squares. The extra cross-products vanish because residuals within
each group sum to zero. For example, DRA contributes
$22(6.681818-8.075758)^2=42.7475$ to $SS_B$. Across all three groups,
$SS_B=108.1212$ and $SS_W=640.5000$, so $SS_T=748.6212$. These numbers
measure variation in squared score units, not mean differences.

```{r}
#| label: fig-f-mechanism
#| fig-cap: "Fitted group means and within-group residuals in the Baumann data."
#| fig-alt: "Observed scores appear as points. Horizontal lines mark the three group means, and thin vertical segments connect each score to its fitted group mean. ANOVA compares separation among fitted means with the residual spread around them."
#| fig-width: 7.2
#| fig-height: 4.1

mechanism <- reading
mechanism$x <- as.numeric(mechanism$group)
mechanism$offset <- ave(
  mechanism$post.test.1,
  mechanism$group,
  FUN = function(z) seq(-.16, .16, length.out = length(z))
)
mechanism$xj <- mechanism$x + mechanism$offset
mechanism$fitted <- reading_means[as.character(mechanism$group)]

ggplot(mechanism, aes(xj, post.test.1, color = group)) +
  geom_segment(
    aes(x = xj, xend = xj, y = fitted, yend = post.test.1),
    linewidth = .35, alpha = .35
  ) +
  geom_point(size = 1.45, alpha = .74) +
  geom_segment(
    data = data.frame(
      group = reading_groups,
      x = 1:3,
      mean = as.numeric(reading_means[reading_groups])
    ),
    aes(
      x = x - .28, xend = x + .28,
      y = mean, yend = mean, color = group
    ),
    inherit.aes = FALSE, linewidth = 1.3
  ) +
  scale_x_continuous(breaks = 1:3, labels = reading_groups) +
  scale_color_manual(values = c("#9B6A45", "#356E8D", "#6E7F48")) +
  guides(color = "none") +
  labs(x = "Assigned condition", y = "Error-detection score (0--16)")
```

### From mean squares to the omnibus test

Divide each sum of squares by its degrees of freedom to obtain a **mean
square**. There are $J=3$ fitted group means, so between-group degrees of
freedom are $J-1=2$ and residual degrees of freedom are $n-J=63$. The three
within-group zero-sum constraints leave $66-3=63$ free residual pieces.

$$
\begin{aligned}
MS_B&=\frac{108.1212}{2}=54.0606,\qquad
MS_W=\frac{640.5000}{63}=10.1667,\\
F&=\frac{MS_B}{MS_W}=\frac{54.0606}{10.1667}=5.3174.
\end{aligned}
$$

$MS_W$, also called the mean square error or MSE, estimates the common error
variance. Under equal model means, both mean squares estimate that variance,
so an $F$ ratio much larger than 1 indicates an unusually large between-group
pattern relative to within-group noise. The omnibus null is

$$
H_0:\mu_{\mathrm{DRA}}
=\mu_{\mathrm{DRTA}}
=\mu_{\mathrm{Think\text{-}aloud}}.
$$

It is one joint question. The numerator degrees of freedom describe the number
of independent group restrictions; the denominator degrees of freedom describe
the residual information used to estimate within-group variation.

Under independent Normal errors with common variance, the null reference is
exactly an $F$ distribution with $J-1$ and $n-J$ degrees of freedom,
conditional on the group assignments. Without those error conditions it is
an approximation whose suitability needs justification. Random assignment
supports the assignment comparison; it does not prove Normal errors, equal
variances, or the exact finite-sample $F$ reference.

```{r}
#| label: tbl-baumann-anova
#| tbl-cap: "One-way ANOVA for observed error-detection outcomes"

baumann_results <- data.frame(
  "Source" = c("Between groups", "Within groups", "Total"),
  "df" = c(2, 63, 65),
  "Sum of squares" = c(
    reading_anova["group", "Sum Sq"],
    reading_anova["Residuals", "Sum Sq"],
    sum(reading_anova[, "Sum Sq"])
  ),
  "Mean square" = c(
    reading_anova["group", "Mean Sq"],
    reading_anova["Residuals", "Mean Sq"],
    NA
  ),
  "F" = c(reading_anova["group", "F value"], NA, NA),
  "p" = c(reading_anova["group", "Pr(>F)"], NA, NA),
  check.names = FALSE
)
baumann_results[, 3:6] <- lapply(
  baumann_results[, 3:6],
  function(x) ifelse(is.na(x), "", sprintf("%.3f", x))
)

knitr::kable(
  baumann_results,
  booktabs = TRUE,
  row.names = FALSE,
  align = c("l", "r", "r", "r", "r", "r")
)
```

The omnibus result, $F(2,63)=5.32$, $p=.007$, indicates that the
equal-means restriction is difficult to reconcile with the declared pooled
model. It does **not** identify which comparison matters, show that every pair
differs, or give the size of a meaningful comparison.

Model conditions are questions about how the evidence was produced:

- Were units independent under the assignment and classroom setting?
- Did the score have the same meaning in every condition?
- Is a common within-group variance a useful working model?
- Are unusual observations or distributional shapes influential?
- Does the treatment of missing outcomes align with the all-assigned estimand?

A plot can reveal a possible problem. It cannot certify an assumption.
For these observed outcomes, Welch's unequal-variance sensitivity result is
$F(2,41.13)=6.99$, $p=.002$. It relaxes the common-variance condition;
it does not repair missingness, dependence, outcome selection, or transport.
Appendix B gives a small-data reconstruction and comparison-family detail.

## Read two statistical strands and one integrated answer

The omnibus result is useful context, but it is not a third strand and cannot
answer the package question. It can reject equal means when only one innovative
condition differs from DRA. The integrated rule instead requires adequate
evidence for **both** declared contrasts.

| Result family | Role in this teaching case | Argument status |
|:--------------|:---------------------------|:----------------|
| Omnibus ANOVA | Assesses the broader equal-means restriction | Contextual output; neither a strand nor the package answer |
| DRTA minus DRA | Supplies a complete-record estimate and simultaneous model interval relevant to estimand 1 | Statistical strand 1 |
| Think-aloud minus DRA | Supplies a complete-record estimate and simultaneous model interval relevant to estimand 2 | Statistical strand 2 |
| Other pairwise comparisons | Answer separately declared exploratory questions | Not strands of this integrated pair |

: *Omnibus, statistical-strand, and separate-pair roles.* {#tbl-result-roles tbl-colwidths="[22,44,34]"}

### Start with differences, then generalize to a contrast

The next calculations use the 66 complete records. They estimate differences
among the observed group means. Connecting those values to the two all-assigned
estimands declared above requires the missingness bridge that the source record
cannot provide by itself.

The group means are 6.681818 (DRA), 9.772727 (DRTA), and 7.772727
(think-aloud). The two required estimates are therefore

$$
\widehat L_1=9.772727-6.681818=3.090909,\qquad
\widehat L_2=7.772727-6.681818=1.090909.
$$

In the fixed order DRA, DRTA, think-aloud, these differences use weights
$(-1,1,0)$ and $(-1,0,1)$. For the first, multiply the DRA mean by $-1$, the
DRTA mean by 1, and the think-aloud mean by 0, then add. Each weight set sums
to zero, so it compares means rather than their overall level.

A **linear contrast** generalizes this operation. With $c_j$ denoting the
chosen weight on group $j$,

$$
L=\sum_{j=1}^{J}c_j\mu_j,\qquad \sum_{j=1}^{J}c_j=0,
\qquad \widehat L=\sum_{j=1}^{J}c_j\bar Y_j.
$$

$L$ is the estimand; $\widehat L$ is its estimate. Adding the same constant
to every group mean leaves the contrast unchanged because the weights sum to
zero. Changing the weights changes the comparison and may change the
question--answer pair. In particular, averaging the two innovative conditions
against DRA uses $(-1,1/2,1/2)$; it cannot answer whether **both** comparisons
meet the declared rule.

### Calculate uncertainty for the declared two-comparison family

Under the independent common-variance model, the estimated contrast SE is

$$
\widehat{SE}(\widehat L)
=\sqrt{MS_W\sum_{j=1}^{J}\frac{c_j^2}{n_j}}.
$$

The squared weights appear because the variance of a weighted mean is
multiplied by the square of its weight. For either declared contrast,
$n_j=22$ and only the two nonzero weights contribute:
$\widehat{SE}=\sqrt{10.166667(1/22+1/22)}=0.961375$ points. This is the
uncertainty of a mean difference, not the SD of individual scores.

Bonferroni calibration protects a declared finite family of comparisons. For
$m=2$ two-sided intervals with family confidence at least 95%, allocate
$.05/2=.025$ noncoverage to each interval, split equally as
$.025/2=.0125$ in each tail. Thus $1-.0125=.9875$ lies to the left of the
upper critical point.
The resulting critical value is
$t^*=t_{1-.05/(2m),63}=t_{.9875,63}=2.296237$ (use **2.296** on a
calculator). The subscript .9875 identifies the proportion to the left of that
point in the $t$ distribution; 63 is its degrees of freedom. Thus

$$
\widehat L_k\pm t^*\widehat{SE}(\widehat L_k),\qquad k=1,2.
$$

The margin is $2.296237(.961375)=2.207545$ points. Strand 1 gives
$3.090909\pm2.207545=[0.883364,5.298454]$; strand 2 gives
$1.090909\pm2.207545=[-1.116636,3.298454]$. Round only at the end. The
family statement concerns repeated-use coverage of **both** estimands, not a
95% probability for either fixed estimand after observing this pair of
intervals. It also does not repair missingness or justify the outcome model.

Ordinary pointwise 95% intervals use a smaller critical value and do not
provide this same simultaneous guarantee. Bonferroni does not require the
two contrasts to be independent; their common DRA mean in fact makes them
dependent. Other family-calibration procedures are possible. This teaching
contract deliberately specifies Bonferroni; it is not a universal requirement
for every claim requiring two positive differences. This family covers only
the **two declared contrasts on the selected error-detection outcome**; it
does not adjust across every outcome or analysis in the source study.
Appendix B compares
alternative families.

```{r}
#| label: tbl-baumann-strands
#| tbl-cap: "Two complete-record DRA-referenced statistical strands under the hypothetical teaching contract"

strand_display <- data.frame(
  "Strand" = c("1: DRTA minus DRA", "2: Think-aloud minus DRA"),
  "Estimate (complete records)" = sprintf("%.2f", reading_contrast_est),
  "95% Bonferroni interval" = sprintf(
    "[%.2f, %.2f]", reading_contrast_low, reading_contrast_high
  ),
  "Relation to zero" = c(
    "Lower bound is above zero",
    "Interval includes zero"
  ),
  "Conclusion" = c("Positive", "Inconclusive"),
  check.names = FALSE
)

knitr::kable(
  strand_display,
  booktabs = TRUE,
  row.names = FALSE,
  align = c("l", "r", "r", "l", "l")
)
```

The DRTA-minus-DRA estimate is 3.09 points, with simultaneous interval [0.88,
5.30]. Its lower bound exceeds zero. The think-aloud-minus-DRA estimate is
1.09 points, with interval [-1.12, 3.30]. That interval remains compatible
with both no improvement and a positive improvement. It is therefore
**inconclusive** for the declared directional criterion. An interval including
zero is not proof of equality.

### Stress-test the two unavailable outcomes

The outcome scale runs from 0 to 16. Without pretending to know the two
unavailable scores, let the missing DRTA score and the missing think-aloud
score each take every integer value on that scale. The DRA group has all 22
scores, with sum 147. The observed DRTA and think-aloud sums are 215 and 171.
The resulting **completed-data difference-in-means estimates** must therefore
fall within

$$
\frac{215+x_D}{23}-\frac{147}{22}\in[2.666,3.362],
\qquad
\frac{171+x_T}{23}-\frac{147}{22}\in[.753,1.449],
$$

where $x_D,x_T\in\{0,1,\ldots,16\}$. These are bounds on what the arithmetic
difference in completed group means could be, not identification bounds for
the causal estimands.

As a second, explicitly model-dependent check, refit the same pooled one-way
model and both Bonferroni intervals for all $17^2=289$ possible completions.
The DRTA-minus-DRA lower endpoint remains above zero; its smallest value is
.271. The think-aloud-minus-DRA interval includes zero in every completion:
its largest lower endpoint is -.805 and its smallest upper endpoint is 2.996.
Consequently, the chapter's conjunctive statistical criterion remains unmet
for every bounded completion under this particular analysis rule.

This is a **bounded completed-data decision stress test**, not a formal
missing-data analysis. It demonstrates that the teaching decision is stable
across these scale-bounded completions, but it does not establish why outcomes
were unavailable, identify all-assigned causal effects, repair outcome
selection, or justify transport beyond the setting.

```{r}
#| label: fig-statistical-strands-ch4
#| fig-pos: H
#| fig-cap: "Two statistical strands supply one integrated teaching answer."
#| fig-alt: "Shared statistical Evidence and Justification feed two separate Inference-to-Claim lanes: DRTA minus DRA is positive and think-aloud minus DRA is inconclusive. Both lanes converge on one substantive chain and one bounded integrated answer. A note says the omnibus ANOVA is not a third strand."
#| fig-width: 6.5
#| fig-height: 4.4
#| out-width: "100%"

draw_strand_map()
```

@fig-statistical-strands-ch4 keeps three counting rules visible:

- several outputs for one estimand remain in one statistical strand;
- two estimands form strands here only because both are jointly required for
  one integrated answer and neither is owed a separate conclusion; and
- if either contrast is interpreted or recommended separately, it becomes a
  separate question--answer pair and argument.

The two-format statistical success criterion is **not met** in the
complete-record analysis and remains unmet in all 289 scale-bounded
completions under the same model and decision rule,
so the integrated answer is **inconclusive**. This does not show that the
package failed or that think-aloud has no effect. It says that the observed
evidence does not meet the conjunctive rule. The statistically detectable omnibus result and a
positive average of the two innovative conditions cannot replace the missing
strand-level support.

\Needspace{12\baselineskip}
::: {.callout-important title="Bounded integrated answer"}
For the 66 complete records in this secondary teaching reanalysis, DRTA minus
DRA was 3.09 error-detection points (two-sided 95% Bonferroni simultaneous
interval [0.88, 5.30]), whereas think-aloud minus DRA was 1.09 points
([-1.12, 3.30]). The first comparison met the hypothetical positive-difference
rule and the second was inconclusive. Consequently, the chapter-created
two-format statistical success criterion was not met; the integrated
substantive answer remains inconclusive. The same statistical criterion stayed
unmet when each unavailable outcome was allowed to take every integer score
from 0 through 16 under the same pooled model. That bounded completed-data
stress test does not identify the all-assigned effects or replace a missingness
analysis. This answer is post hoc
relative to the source study even though the criterion is prospective inside
the imagined replication. It is conditional on the pooled model and bounded
completion rule, with missingness unresolved; it is limited to the selected
outcome and documented setting, and is not a claim that either method failed
or that effects transport broadly.
:::

# Read, Evaluate, and Construct the Arguments

## Let the design bound the substantive answer

The two examples used different procedures, but the larger difference lies in
their designs.

| Feature | UCI categorical pair | Baumann quantitative pair |
|:--------|:---------------------|:---------------------------|
| Group variable | Observed intention | Randomly assigned instructional condition |
| Outcome | Recorded pass status | Error-detection score |
| Argument granularity | One estimand and one statistical strand | One hypothetical integrated question with two subordinate contrast estimands |
| Immediate estimate | One proportion difference | Two retained DRA-referenced mean differences |
| Strongest design-based language | Local association | Assignment comparison under stated conditions |
| Main reservations | Selection, confounding, shared setting, transport | Post hoc application of the teaching criterion, missing outcomes, selected outcome, model-based uncertainty, transport |

: *How design bounds the two group-comparison claims.* {#tbl-design-claim tbl-colwidths="[20,40,40]"}

A small p-value would not turn the UCI comparison into an experiment. A more
complex model would not create random sampling. Likewise, random assignment in
the Baumann study supports a comparison of assignment policies in its setting,
but it does not by itself solve missing outcomes, define educational
importance, or justify transport.

## A short regression bridge

The same group means, contrasts, and omnibus restriction can also be
represented using indicator-coded regression. That representation is developed
with multiple regression, after simple-regression coefficients have been
introduced. Re-encoding the same comparison does not strengthen its design.

Adding another factor creates adjustment, weighting, and possible interaction
questions developed in the later regression chapters. Appendix B's blocking discussion stays
narrower: it shows how blocking can refine
one declared group-comparison estimand without introducing coordinated
questions.

Chapter 1 treats one question--answer pair as **one argument with linked
statistical and substantive layers**. Within either layer,

$$
E+J\longrightarrow I\longrightarrow C.
$$

Reading is the primary use. It asks what content performs each component, where
that content can be inspected, and what answer the authors or teaching contract
actually report. It does not yet decide whether the support is adequate.

Four documentary statuses keep location separate from quality: **stated in the
paper**, **located in related materials**, **implicit but traceable**, and
**not located**. A contradiction across sources is a separate conflict flag.
The chapter-created questions and analyses below are labeled as such rather
than misdescribed as documentary traces of the source investigators' choices.

## Read the UCI single-strand argument

| Component | Content identified during reading | Trace and documentary status |
|:----------|:----------------------------------|:-----------------------------|
| **STAT-E** | 395 records; intention and pass coding; group counts and denominators; unexplained cluster of 38 final-grade zeros | UCI record and data dictionary: **located in related materials**; a shared meaning for all zeros is **not located** |
| **STAT-J** | Within-group denominators; declared independent-record approximation for the interval and chi-square reference; all-row coding kept primary and zero-row omission labeled diagnostic | Formula and assumptions are **stated in this chapter**; source-population and zero-grade processes are **not located** |
| **STAT-I** | Recorded-table difference .338; separate model illustration gives 95% interval [.124, .552] and $X^2(1)=9.82$, $p=.002$; deletion diagnostic gives difference .252 with method-sensitive p-values | Chapter-created analysis, reproducible from the related data |
| **STAT-C** | A known 33.8-percentage-point difference in the all-row table; model uncertainty is conditional context, and the diagnostic exposes sensitivity to record definition | **Stated in this chapter** |
| **SUB-E** | Students recorded in two Portuguese secondary schools; observed intention; passing from final grade | UCI documentation: **located in related materials** |
| **SUB-J** | Observational grouping cannot by itself exclude confounding or establish transport; documentation does not settle the meaning of the zero-grade cluster | Design relation is **implicit but traceable** from the record; broader sampling and zero-grade warrants are **not located** |
| **SUB-I** | Interpret .338 as a local association, not an effect of intending | **Stated in this chapter** |
| **SUB-C** | The bounded categorical answer in @sec-categorical-comparison | **Stated in this chapter** |

: *Reading record for the UCI single-strand argument.* {#tbl-uci-argument tbl-colwidths="[14,48,38]"}

The primary recorded-table estimand is followed through one statistical strand.
The count table and proportional plot document it; the model interval and
chi-square output show what an explicitly conditional inferential extension
would add. They do not silently turn a known finite-record description into a
population estimate. A report making a separate process-level conclusion
would owe a corresponding question--answer pair and its justification.

## Read the hypothetical two-strand argument

The source record supplies study context and outcomes. The integrated success
criterion is supplied only by the chapter's hypothetical teaching contract.
No row below attributes that criterion, its contrasts, or its Bonferroni rule
to Baumann and colleagues.

| Statistical component | Shared core | Strand 1: DRTA minus DRA | Strand 2: Think-aloud minus DRA |
|:----------------------|:------------|:-------------------------|:--------------------------------|
| **STAT-E** | Assignment record; 68 assigned; 66 observed scores; one incomplete outcome in each innovative arm | DRA and DRTA observations | DRA and think-aloud observations |
| **STAT-J** | Assignment rationale; common score; independence and pooled-variance working model; two-comparison Bonferroni family; explicit missingness limitation; bounded completion rule | Complete-record estimator and uncertainty aligned to contrast 1 | Complete-record estimator and uncertainty aligned to contrast 2 |
| **STAT-I** | Omnibus result retained as context; bounded completed-data stress test across 289 completions; the criterion is prospective only inside the hypothetical contract, while this application to existing data is post hoc | Complete-record estimate 3.09 points and interval [0.88, 5.30]; lower bound stays positive across bounded completions | Complete-record estimate 1.09 points and interval [-1.12, 3.30]; interval includes zero in every bounded completion |
| **STAT-C** | The joint criterion remains unmet under the complete-record model and every scale-bounded completion; this is not identification of the all-assigned effects | Positive for the directional criterion under the stated calculations | Inconclusive for the directional criterion under the stated calculations |

: *Statistical-layer reading record for the two-strand teaching argument.* {#tbl-baumann-stat-argument tbl-colwidths="[15,29,28,28]"}

The shared source facts are **stated in the paper** or **located in related
materials**. The exact package values are located in the public *carData*
record. The teaching question, two estimands, Bonferroni family, and integrated
rule are **chapter-created analytic content**, not statuses assigned to a
source decision. A historical trace showing that Baumann and colleagues used
this exact contract is **not located**; no conflict is claimed.

| Substantive component | Content identified during reading |
|:----------------------|:----------------------------------|
| **SUB-E** | Fourth-grade students in one rural Midwestern school; three instructional assignments; error detection as one 0--16 outcome |
| **SUB-J** | The randomized comparison contextualizes assignment effects in the documented setting; the completion stress test does not supply the missingness mechanism, implementation, educational importance, or wider relevance |
| **SUB-I** | Interpret both strand results together: one positive comparison cannot erase the other comparison's inconclusive uncertainty; stability of the rule across bounded completions is narrower than identification of effects |
| **SUB-C** | The statistical package criterion is not met; the integrated substantive answer is inconclusive and bounded to the selected outcome, pooled model, bounded completion exercise, documented setting, and post hoc application of the hypothetical contract |

: *Substantive-layer reading record for the integrated teaching argument.* {#tbl-baumann-sub-argument tbl-colwidths="[18,82]"}

SUB-E situates the statistical evidence. SUB-J **augments and contextualizes
the statistical justification**. SUB-I interprets the two statistical results
without averaging one away. SUB-C carries their uncertainty and conditions
into one bounded answer. The substantive layer cannot automatically strengthen
weak statistical support.

## Secondary use: evaluate quality

Quality evaluation begins only after the arguments have been read. Apply all
six Chapter 1 questions without adding a score.

| Evaluation question | UCI single-strand pair | Hypothetical two-strand pair |
|:----------------------------|:-----------------------|:-----------------------------|
| **Question alignment** | **Adequate for this purpose:** event, denominator, estimand, estimate, and local answer match. | **Adequate for this purpose:** both contrasts and the conjunctive rule address the one integrated question. |
| **Evidence integrity** | **Needs revision** for broader use: selection, possible within-school dependence, and the zero-grade pattern are not fully documented. | **Needs revision:** two assigned outcomes are unavailable and only one selected outcome is analysed. |
| **Justification adequacy** | **Insufficient information to judge** a broad repeated-record process or a common meaning for the zero-grade cluster; the literal all-row description itself remains reproducible. | **Needs revision:** pooled variance and observed-to-assigned missingness support are limited; bounded completion is not a missingness mechanism. |
| **Inferential adequacy** | **Adequate for the stated, limited purpose:** magnitude, interval, omnibus test, and coding diagnostic are distinguished; substantive reasoning remains local and associational. | **Adequate for the stated, limited purpose:** the complete-record model calibrates the family and the bounded completion exercise stress-tests its decision; neither identifies the all-assigned effects. |
| **Transition fidelity** | **Adequate for this purpose:** the substantive interpretation retains magnitude, model status, and observational scope. | **Adequate for this purpose:** both intervals, the inconclusive strand, the post hoc application to existing data, and missingness survive the transition. |
| **Claim calibration** | **Adequate for this purpose:** the answer is local and associational. | **Adequate for this purpose:** the answer says the criterion is not established, not that the package failed or that either method is ineffective. |

: *All six quality-evaluation questions applied after reading.* {#tbl-comparison-evaluation tbl-colwidths="[19,39,39]"}

For the bounded answers as written in this chapter, the whole-pair evaluation
conclusion is **supported as stated** for each pair. That conclusion does not
mean every underlying study feature is strong. It means each reported answer
already preserves the limitations exposed by the per-question judgments. In
particular, an inconclusive answer can be supported as stated. Failure to meet
a hypothetical success rule is not automatically the evaluation conclusion
**not supported**.

Authors and analysts have the first responsibility to document and evaluate
their own reasoning. Readers, collaborators, supervisors, reviewers, and
editors can then apply the same questions independently.

## Auxiliary use: construct the argument

The six-stage workflow supports conducting research. Here it is applied to how
a **future hypothetical replication** could construct the integrated argument,
not to infer an undocumented history for the 1992 study.

| Phase and stage | Application to the hypothetical teaching contract | Status when applied to the existing record |
|:----------------|:--------------------------------------------------|:-------------------------------------------|
| **Phase 1, Stage 1:** Review problem, setting, and literature | Study the instructional context, outcome meaning, prior evidence, and feasibility of collecting complete follow-up | Source rationale is documented; the replication decision is chapter-created |
| **Phase 1, Stage 2:** Frame question and candidate answer | Declare one integrated question, both subordinate estimands, and an affirmative answer only if both criteria are met | Prospective only inside the hypothetical contract; post hoc relative to the available data |
| **Phase 1, Stage 3:** Align design, measurement, estimand, evidence plan, and inference | Align three-arm assignment, common 0--16 measure, complete-outcome procedures, zero benchmark, two-sided 95% Bonferroni intervals, and conjunctive rule | The chapter can demonstrate the analysis but cannot create historical prospectiveness |
| **Phase 2, Stage 4:** Produce or acquire and document evidence | Assign, deliver, measure, retain flow records, and document deviations | Existing source materials supply 68 assigned and 66 observed cases |
| **Phase 2, Stage 5:** Estimate, check, and analyse | Inspect raw groups; estimate both complete-record contrasts; compute simultaneous intervals; keep the omnibus test in its contextual role; stress-test all 289 scale-bounded completions | Reproduced in @sec-mean-comparison |
| **Phase 2, Stage 6:** Contextualize, interpret, and complete the bounded answer | Retain both strands, missingness, analytic status, model conditions, completion-test limits, setting, and the inconclusive integrated result | Completed in the bounded answer |

: *Six-stage construction application for a hypothetical replication.* {#tbl-baumann-workflow tbl-colwidths="[28,45,27]"}

```{r}
#| label: fig-baumann-workflow
#| fig-cap: "The six-stage construction workflow for the hypothetical replication."
#| fig-alt: "Stages 1 through 3 run left to right in Phase 1. Stage 4 is directly below Stage 3, and Stages 4 through 6 run right to left in Phase 2. Dashed arrows permit iteration within each phase, but no arrow returns from Phase 2 to Phase 1."
#| fig-width: 6.6
#| fig-height: 3.7
#| out-width: "100%"

draw_workflow()
```

Iteration stays within a phase. Rechecking a plot, correcting a table, or
adding a same-estimand sensitivity analysis can revise Phase 2. Changing an
estimand, outcome, design, or joint rule after seeing the evidence changes the
Phase 1 anchor and starts an additional or revised cycle. Defensible material
from the earlier cycle can be reused with its provenance and changed status
visible; reuse cannot make a retrospective decision prospective.

# Summary and Reading Bridge

## A compact reading route

When reading a published group comparison:

1. Identify the unit, analytic sample, groups, outcome, estimand, and design.
2. For a categorical outcome, inspect counts and choose the denominator before
   reading percentages.
3. For a quantitative outcome, inspect raw grouped observations before reading
   ANOVA output.
4. State an effect estimate in interpretable units.
5. Use omnibus evidence only for its joint question.
6. Count statistical strands by jointly required estimands, not by output rows,
   coefficients, or tests.
7. If a result receives its own question-facing conclusion, give it a separate
   question--answer pair rather than hiding it inside another argument.
8. Carry uncertainty, analytic status, missingness, and design into the answer.
9. Read one argument with linked layers for each question--answer pair.

Reading is the primary use. Quality evaluation is the distinct secondary use
after reading; the construction workflow is the auxiliary use for conducting
research.

## Compact glossary

**Analytic sample.** The observations that actually enter a reported analysis.

**Conditional proportion.** A proportion whose denominator is restricted to a
declared group.

**Estimand.** The precisely specified quantity a study seeks to learn. An
estimate is the numerical result obtained from observed evidence for it.

**Expected count.** A cell count implied by a stated null pattern, such as
independence, given the observed margins.

**Omnibus question.** A joint question about a complete pattern, such as
whether all declared group means are equal.

**Statistical strand.** An estimand-specific statistical chain inside one
argument, used when sibling estimands are jointly required for one integrated
answer and none is owed a separate conclusion.

**Familywise calibration.** An uncertainty or testing procedure that accounts
for a declared family of comparisons.

**Design-bounded claim.** An answer whose wording retains the support and
limitations of sampling, assignment, measurement, analysis, and context.

**Inconclusive answer.** An answer stating that materially different
possibilities remain compatible with the evidence under a relevant benchmark
and the declared conditions. It is not proof of no difference.

## Reading bridge

These public sources extend the chapter without becoming prerequisites:

- [Introduction to Modern Statistics: Exploring categorical
  data](https://openintro-ims.netlify.app/explore-categorical) develops counts,
  proportions, and displays.
- [Introduction to Modern Statistics: Inference for two-way
  tables](https://openintro-ims.netlify.app/inference-tables) develops expected
  counts and chi-square reasoning.
- [Introduction to Modern Statistics: Inference for many
  means](https://openintro-ims.netlify.app/inference-many-means) develops the
  omnibus ANOVA question and follow-up comparisons.
- [NIST: Assessing a response with
  contrasts](https://www.itl.nist.gov/div898/handbook/prc/section4/prc436.htm)
  connects group means to declared contrasts; its [Bonferroni
  guide](https://www.itl.nist.gov/div898/handbook/prc/section4/prc473.htm)
  explains simultaneous intervals for a finite comparison family.
- [FDA guidance on multiple
  endpoints](https://www.fda.gov/regulatory-information/search-fda-guidance-documents/multiple-endpoints-clinical-trials)
  supplies a limited analogy for a conclusion that requires every declared
  component; it is not imported here as a social-science regulatory rule.
- [UCI Student Performance
  record](https://archive.ics.uci.edu/dataset/320/student+performance) documents
  the observational source used here.
- [The carData package](https://cran.r-project.org/package=carData) documents
  the public Baumann record. The original report supplies the design context
  [@baumann1992].

\Needspace{8\baselineskip}
\phantomsection

# Exercises {.unnumbered}

The exercises contain no answers and require no software. Use only
**Foundational** and **Integrative** as levels. Keep the question, denominator,
estimate, uncertainty, and design boundary visible in every response.

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH04-E01 -->
## Exercise 1: Choose the denominator (Foundational) {.unnumbered #ex-ch04-01}

A study records the following outcomes:

| Participation format | Completed | Did not complete | Total |
|:---------------------|----------:|-----------------:|------:|
| Online | 84 | 36 | 120 |
| In person | 63 | 17 | 80 |
| **Total** | **147** | **53** | **200** |

The question is: "Among participants using each format, how did completion
differ?"

1. Calculate the completion proportion within each format.
2. Calculate the proportion using the online format within each outcome
   category: completed and did not complete.
3. Identify which pair of proportions answers the stated question and explain
   why the other pair answers a different question.
4. State what the four raw counts contribute that percentages alone conceal.

<!-- END EX04-01 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH04-E02 -->
## Exercise 2: Name the categorical effect (Foundational) {.unnumbered #ex-ch04-02}

Use the participation-format table in Exercise 1.

1. Calculate the Online-minus-In-person difference in completion proportions.
2. Interpret its sign, direction, event, and unit in one sentence.
3. Explain how the numerical value and wording would change if the event were
   changed from completion to noncompletion.
4. State one reason a proportion difference is easier to interpret here than
   an odds ratio.

<!-- END EX04-02 PROMPT -->

{{< pagebreak >}}

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH04-E03 -->
## Exercise 3: Check expected counts and chi-square output (Foundational) {.unnumbered #ex-ch04-03}

For the participation-format table, the expected counts under independence are:

| Participation format | Completed | Did not complete |
|:---------------------|----------:|-----------------:|
| Online | 88.2 | 31.8 |
| In person | 58.8 | 21.2 |

Pearson's result is $\chi^2(1)=1.89$, $p=.170$.

1. Use the observed table in Exercise 1 to reproduce the Online/Completed
   expected count, its contribution $(O-E)^2/E$, and the table degrees of
   freedom. Explain what an expected count represents.
2. Identify the cells in which the observed count exceeds the expected count.
3. State the omnibus null in words.
4. Write one conclusion that distinguishes the observed proportion difference,
   evidence against independence, association magnitude, and causation.
5. Explain why $p=.170$ is not proof that format and completion are
   independent.

<!-- END EX04-03 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH04-E04 -->
## Exercise 4: Connect grouped observations to ANOVA (Foundational) {.unnumbered #ex-ch04-04}

Three groups have the following observed scores:

- Group A: 4, 5, 6, 6, 7, 8, 9, 9
- Group B: 5, 6, 7, 7, 8, 9, 10, 10
- Group C: 1, 4, 6, 8, 10, 12, 14, 17

Before considering an ANOVA result:

1. Compare the centers, spreads, overlap, and ranges.
2. Identify which group makes a common-variance model most questionable and
   explain why.
3. Explain what an individual-observation plot would reveal that a bar chart
   of means would conceal.
4. State why these observations cannot by themselves determine whether a group
   difference is causal or generalizable.
5. A one-way calculation gives $SS_B=20.333333$ and $SS_W=245.000000$.
   With three groups and 24 scores, calculate the between-group and
   within-group degrees of freedom, both mean squares, and $F$. Explain
   what the ratio compares; do not infer a specific pairwise difference from it.

<!-- END EX04-04 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH04-E05 -->
## Exercise 5: Distinguish outputs, strands, and pairs (Integrative) {.unnumbered #ex-ch04-05}

Classify each situation as **several outputs in one strand**, **two statistical
strands in one argument**, or **two question--answer pairs and arguments**.
Explain the answer without counting table rows or hypotheses mechanically.

1. A report gives a mean difference, standard error, confidence interval,
   p-value, and sensitivity analysis for the same estimand.
2. One prespecified package-success question requires two declared estimands
   to meet their criteria; neither receives a separate conclusion.
3. A report makes one recommendation about Method A versus control and a
   separate recommendation about Method B versus control.
4. An omnibus ANOVA and a contrast interval are printed for one estimand-
   focused question. Is the omnibus result automatically another strand?

<!-- END EX04-05 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH04-E06 -->
## Exercise 6: Construct and read a two-strand result (Integrative) {.unnumbered #ex-ch04-06}

A hypothetical teaching contract requires the lower endpoint of each two-sided
95% Bonferroni simultaneous interval to be above zero. Its two comparisons are
Method B minus reference and Method C minus reference. Use these observed
summaries:

| Group | Mean score | Observed $n$ |
|:------|-----------:|-------------:|
| Reference | 6.681818 | 22 |
| Method B | 9.772727 | 22 |
| Method C | 7.772727 | 22 |

Assume the independent, common-variance Normal model for this calculation.
The pooled within-group mean square is $MS_W=10.166667$ with 63 degrees of
freedom. For this family of two contrasts, the supplied critical value is
$t^*=2.296$; no software or quantile lookup is needed. In group order
reference, B, C, the weights are $(-1,1,0)$ and $(-1,0,1)$.

1. Calculate both contrast estimates and use
   $SE=\sqrt{MS_W(1/n_{\mathrm{method}}+1/n_{\mathrm{reference}})}$
   to calculate their common standard error.
2. Form each interval as estimate $\pm2.296(SE)$, round endpoints to two
   decimal places, and state each strand's result in ordinary language.
3. Does the integrated statistical criterion succeed? Distinguish that
   decision from the substantive conclusion, using the declared zero benchmark.
4. Explain why neither the omnibus ANOVA nor the average of the two estimates
   can replace either required comparison.
5. Write one sentence distinguishing **inconclusive** from evidence that
   Method C failed or had exactly no effect. Name one design or missingness
   condition that the calculation alone cannot establish.

<!-- END EX04-06 PROMPT -->

{{< pagebreak >}}

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH04-E07 -->
## Exercise 7: Repair two group-comparison claims (Integrative) {.unnumbered #ex-ch04-07}

Repair each claim using the evidence in this chapter.

> Higher-education intention caused students to pass because the chi-square
> test was significant.

> Both innovative instructional methods improve reading achievement for all
> fourth-graders.

For each claim:

1. identify at least three words or omissions that exceed the evidence;
2. retain the relevant magnitude and uncertainty;
3. retain the analytic sample, design, hypothetical status, and missingness
   where relevant; and
4. write one bounded replacement answer.

<!-- END EX04-07 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH04-E08 -->
## Exercise 8: Read, evaluate, then trace construction (Integrative) {.unnumbered #ex-ch04-08}

Use the hypothetical two-strand teaching argument in this chapter.

1. Identify shared STAT-E and STAT-J content, each strand's STAT-I and STAT-C,
   and the one SUB-E/J/I/C chain.
2. Give a documentary status for the source assignment counts, the package
   scores, and the chapter-created integrated criterion. Explain why the last
   item must not be attributed to the source investigators.
3. Apply all six quality-evaluation questions. Keep documentary status,
   per-question judgments, and the whole-pair evaluation conclusion distinct.
4. Classify correcting a mislabeled plot as a Phase 2 revision. Then explain
   why changing an estimand or the joint rule after seeing results requires an
   additional or revised cycle.
5. Identify evidence that could be reused in that cycle and explain why reuse
   cannot make a retrospective choice prospective.

<!-- END EX04-08 PROMPT -->

\Needspace{8\baselineskip}
\phantomsection

# Appendix A: Further Categorical Comparisons {.unnumbered}

\Needspace{6\baselineskip}
\phantomsection

## Difference, ratio, and odds scales {.unnumbered}

For event probabilities $p_1$ and $p_0$:

$$
\text{difference}=p_1-p_0,\qquad
\text{ratio}=\frac{p_1}{p_0},\qquad
\text{odds ratio}=
\frac{p_1/(1-p_1)}{p_0/(1-p_0)}.
$$

```{r}
#| label: tbl-uci-scales
#| tbl-cap: "Three descriptive scales for passing in the UCI record"

uci_scale_display <- data.frame(
  "Scale" = c(
    "Proportion difference",
    "Proportion ratio",
    "Odds ratio"
  ),
  "Estimate" = c(
    sprintf("%.3f", uci_rd),
    sprintf("%.2f", uci_rr),
    sprintf("%.2f", uci_or)
  ),
  "Reading" = c(
    "Percentage-point separation",
    "Ratio of passing probabilities",
    "Ratio of passing odds"
  ),
  check.names = FALSE
)

knitr::kable(
  uci_scale_display,
  booktabs = TRUE,
  row.names = FALSE,
  align = c("l", "r", "l")
)
```

The scales are not interchangeable. Passing is common in the intention-yes
group, so the odds ratio lies farther from 1 than the probability ratio.
Reversing the group order takes the reciprocal of a ratio and changes the sign
of a difference. Reversing the event changes the substantive story and the
sign of the difference. A report must name the event and direction.

\Needspace{6\baselineskip}
\phantomsection

## Larger and sparse categorical tables {.unnumbered}

For an $R\times C$ table, Pearson's statistic remains an omnibus comparison
of observed and expected counts. Standardized residuals help locate cells that
contribute to the pattern, while Cramer's

$$
V=\sqrt{\frac{X^2}{n\min(R-1,C-1)}}
$$

gives a unitless whole-table description. Neither supplies causal evidence or
a universal practical-importance threshold.

Sparse tables need more care than a single expected-count slogan. In the
artificial table

| Group | Event yes | Event no |
|:------|----------:|---------:|
| A | 1 | 9 |
| B | 7 | 3 |

each expected count is 4 or 6. The usual chi-square reference can be poor.
Fisher's exact test gives an exact conditional probability calculation for
fixed margins. It is not exact about sampling, measurement, independence,
causation, or truth.

\Needspace{8\baselineskip}
\phantomsection

# Appendix B: Further Mean Comparisons {.unnumbered}

\Needspace{6\baselineskip}
\phantomsection

## A complete small-data decomposition {.unnumbered}

The following six artificial scores isolate the arithmetic without adding
another study narrative: group A has 1, 3; B has 3, 5; and C has 5, 7. Their
means are 2, 4, and 6, and the grand mean is 4. Thus

$$
SS_B=2(2-4)^2+2(4-4)^2+2(6-4)^2=16,
\qquad SS_W=6(1^2)=6.
$$

Every score is one point above or below its own group mean. Directly around
the grand mean, $SS_T=9+1+1+1+1+9=22=16+6$. The mean squares are
$MS_B=16/(3-1)=8$ and $MS_W=6/(6-3)=2$, giving $F=8/2=4$.
This illustrates the calculation; six artificial scores do not establish
that the Normal common-variance model is suitable for a real study.

The algebra behind the identity is equally short. Expanding the square of
$(\bar Y_j-\bar Y)+(Y_{ij}-\bar Y_j)$ adds a cross-term. Within group $j$,
that term sums to
$2(\bar Y_j-\bar Y)\sum_i(Y_{ij}-\bar Y_j)=0$ because deviations from a
group's own mean sum to zero. What remains is exactly $SS_B+SS_W$.

\Needspace{6\baselineskip}
\phantomsection

## Simultaneous comparison families {.unnumbered}

The main text's Bonferroni intervals use

$$
\widehat\Delta_k
\mathbin{\pm}
t_{1-.05/(2g),63}\,SE(\widehat\Delta_k),
\qquad g=2.
$$

The family contains exactly the two DRA-referenced contrasts. Bonferroni is
transparent and valid for this declared finite family, but it is not uniquely
required. A Dunnett procedure can use the shared-reference dependence more
efficiently; Tukey instead calibrates all pairwise comparisons. The family must
follow the question rather than a software default.

The two contrasts share the DRA mean. Under independent groups with common
variance their covariance is

$$
\operatorname{Cov}(\bar Y_2-\bar Y_1,\bar Y_3-\bar Y_1)
=\operatorname{Var}(\bar Y_1)=\frac{\sigma^2}{n_1}.
$$

Terms involving means from different groups have zero covariance; the shared
reference contributes positively. Bonferroni remains valid without
independence between contrasts because the probability that either interval
misses cannot exceed the sum of their individual miss probabilities.
This guarantee still requires each component interval's model justification.

```{r}
#| label: tbl-baumann-bonferroni
#| tbl-cap: "Bonferroni simultaneous intervals for the two declared teaching contrasts"

bonf_display <- data.frame(
  "Comparison" = reading_contrast_names,
  "Estimate" = sprintf("%.2f", reading_contrast_est),
  "SE" = sprintf("%.2f", reading_contrast_se),
  "Lower" = sprintf("%.2f", reading_contrast_low),
  "Upper" = sprintf("%.2f", reading_contrast_high),
  check.names = FALSE
)

knitr::kable(
  bonf_display,
  booktabs = TRUE,
  row.names = FALSE,
  align = c("l", "r", "r", "r", "r")
)
```

```{r}
#| label: tbl-baumann-tukey
#| tbl-cap: "Tukey 95% familywise intervals for observed Baumann outcomes"

tukey_display <- data.frame(
  "Comparison" = rownames(reading_tukey),
  "Difference" = sprintf("%.2f", reading_tukey[, "diff"]),
  "Lower" = sprintf("%.2f", reading_tukey[, "lwr"]),
  "Upper" = sprintf("%.2f", reading_tukey[, "upr"]),
  "Adjusted p" = sprintf("%.3f", reading_tukey[, "p adj"]),
  check.names = FALSE
)

knitr::kable(
  tukey_display,
  booktabs = TRUE,
  row.names = FALSE,
  align = c("l", "r", "r", "r", "r")
)
```

The realized ANOVA summaries are
$\eta^2=SS_B/SS_T=.144$ and
$\omega^2=[SS_B-(J-1)MS_W]/[SS_T+MS_W]=.116$.
They describe how much observed score
variation the fitted grouping accounts for under these formulas. They do not
replace an outcome-scale contrast, define practical importance, or say what
proportion of variation was caused by instruction.

\Needspace{6\baselineskip}
\phantomsection

## Blocking within one comparison question {.unnumbered}

A blocking variable can improve precision when it records a feature related to
the outcome and the assignment process respects the blocks. It does not create
a second research question automatically. For one declared contrast estimand,
an unadjusted group comparison and a block-adjusted comparison are alternative
analyses of the **same** estimand and therefore remain in one statistical
strand.

Readers should ask:

- What was the unit of assignment, and what defined a block?
- Was assignment carried out within blocks or was the variable merely added
  later to a model?
- Does adjustment preserve the original estimand or change its weighting?
- Is uncertainty aligned with the actual assignment and dependence structure?

Blocking may strengthen precision or reveal sensitivity; the label itself is
not design evidence. Two-way factorial questions, interactions, marginal
weighting, and coordinated question sequences belong in Chapters 8--9.

\Needspace{8\baselineskip}
\phantomsection

# References {.unnumbered}

\begingroup\footnotesize
::: {#refs}
:::
\endgroup
