---
title: "Group Comparisons and Statistical Modelling with R"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: true
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - ../koma_heading_tags.tex
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \widowpenalty=10000
          \clubpenalty=10000
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\linespread{0.95}\selectfont\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{10\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AtBeginDocument{\renewcommand{\boldsymbol}[1]{\symbf{#1}}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---


```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                     collapse = TRUE, comment = "#>")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 10, width = 65, show.signif.stars = FALSE)
```

{{< pagebreak >}}

::: {.callout-note title="Chapter overview"}
Comparing groups asks one modelling question of several means at once.
This chapter reads two invented group comparisons as statistical models
fitted in R. One framework runs
through it: fit a stated model to the records, read what its parameters
say, compare it with a simpler named model on the same records, and check
the procedure by repetition.

Section 1 writes the group-mean model and reads the one-way F test as one
mean against three; Section 2 asks particular comparisons inside stated
families; Section 3 reads a two-way table as a sequence of named models
and reads a declared interaction contrast. Section 4 turns
to counts, a binomial model and their effect scales; Section 5 names two
fitting criteria, least squares and likelihood, and shows how numerical
optimization finds their best values; Section 6 repeats
procedures under a stated generating model and plans power and precision.
By the end you can connect each R result to its model, comparison,
uncertainty and the conditions that support its interpretation.
:::

Studies rarely stop at one mean. The invented background-sound
experiment compares mean reading comprehension under three conditions,
silence, instrumental music and music with lyrics, and its second
invented study asks how the difference between spaced and massed
practice changes across three kinds of feedback. These are the two
teaching studies introduced in Chapter 4. Here
each answer is also a statement about models: which model describes
the records, what its parameters say, and what a simpler model leaves
unexplained. Both studies are invented teaching inputs, so each analysis
shows what a procedure reports, not what background sound or feedback does.

## A Model for Group Means {#ch06-group-models}

### The question and the records

Does mean reading comprehension differ across three background-sound
conditions? Consider an invented experiment.
Each of thirty first-year university students read the same passage for
20 minutes and then answered a 20-item reading-comprehension quiz, scored
from 0 to 20. Ten students were randomly assigned to each sound condition:
Silence, Instrumental music or Music with lyrics. Each student supplies
one quiz score.

These values were constructed to keep the calculations transparent. They
are teaching inputs, not empirical findings about music and reading. The
student is both the unit of random assignment and the unit of
measurement: the ten scores in a condition come from ten different
students, not from repeated measurements of one student. Silence is the
natural reference condition, so it comes first.

The data frame `sound_data` holds these thirty records, with the columns
`sound` and `score`. Its sound conditions are ordered Silence,
Instrumental music and Music with lyrics. The following calls count the
students and summarize their scores within each condition.

```{r}
#| label: ch06-sound-setup
#| include: false
sound_data <- data.frame(
  sound = factor(rep(c("Silence", "Instrumental music",
    "Music with lyrics"), each = 10),
    levels = c("Silence", "Instrumental music",
      "Music with lyrics")),
  score = c(8, 9, 9, 11, 12, 13, 13, 14, 15, 16,
    6, 8, 8, 9, 9, 11, 11, 12, 12, 14,
    4, 5, 6, 7, 8, 8, 9, 10, 11, 12))
```

```{r}
#| label: ch06-sound-summary
table(sound_data$sound)
with(sound_data, tapply(score, sound, mean))
with(sound_data, tapply(score, sound, sd))
```

Each condition holds ten students. The condition means are 12, 10 and 8
correct answers, and the grand mean of all thirty scores is 10. The SDs,
2.7080, 2.4037 and 2.5820, are similar without being identical. Before
asking whether a difference is large relative to its uncertainty, we can
see each difference on the quiz's own scale: Instrumental music lies
2 points and Music with lyrics 4 points below Silence.

@fig-ch06-sound-groups shows every student beside the three condition
means. Each point is one student's score, moved a small random distance
sideways so that equal scores in a condition stay visible, and each
diamond marks a condition mean. Two kinds of variation appear together.
The diamonds step down by 2 points from Silence to Instrumental music and
by another 2 points to Music with lyrics, while the ten scores in each
condition spread across 8 points around their own diamond. Whether a
2-point step is large depends on that spread within conditions, and the
model below turns this visual comparison into numbers.


```{r}
#| label: ch06-show-score-graph
#| eval: false
library(ggplot2)
group_means <- aggregate(score ~ sound, sound_data, mean)
ggplot(sound_data, aes(sound, score)) +
  geom_jitter(width = .15, height = 0) +
  geom_point(data = group_means, shape = 18, size = 5)
```

```{r}
#| label: fig-ch06-sound-groups
#| echo: false
#| fig-width: 6.4
#| fig-height: 3.5
#| fig-cap: "Individual comprehension scores and their three condition means."
#| fig-alt: "Ten scores in each of Silence, Instrumental music and Music with lyrics, spread slightly sideways so that tied scores stay visible, lie around means 12, 10 and 8. Larger diamonds mark the means; the three spreads are similar."
library(ggplot2)
set.seed(12345)
group_means <- aggregate(score ~ sound, sound_data, mean)
p <- ggplot(sound_data, aes(sound, score))
p + geom_jitter(width = 0.15, height = 0, size = 2) +
  geom_point(data = group_means, shape = 18, size = 5,
    colour = "black") +
  labs(x = "Background sound",
    y = "Comprehension score (out of 20)") +
  theme_minimal(base_size = 11)
```

The points and mean diamonds represent different quantities: individual
observations and a summary of each condition. Section 1.4 examines how
such choices change what a graph makes visible.

### The group-mean model

The one-way model for student $i$ in sound
condition $j$ is

$$
Y_{ij}=\mu_j+\varepsilon_{ij},\qquad
\varepsilon_{ij}\sim N(0,\sigma^2),
$$

with independent errors. Its four parameters are the three condition
means $\mu_S$, $\mu_I$ and $\mu_L$, and the common error SD $\sigma$.
Each mean describes one sound condition's process; each error is a
student's departure from that mean. These are working model conditions,
which the invented data do not establish.

Least squares chooses each fitted mean to minimize the sum of squared
residuals within its condition, giving the condition's sample mean;
Section 5.1 examines that fitting criterion. Thus
$\widehat y_{ij}=\bar y_j$ and $e_{ij}=y_{ij}-\bar y_j$. For Silence
the fitted values are all 12 and the residuals are
$-4,-3,-3,-1,0,1,1,2,3,4$: the vertical distances from the mean diamond
in @fig-ch06-sound-groups.

A factor with three levels can also enter a linear model as two indicator
variables. Let $D_I=1$ for Instrumental music and 0 otherwise, and let
$D_L=1$ for Music with lyrics and 0 otherwise. Silence has both indicators
0. The fitted equation is

$$
\widehat y=12-2D_I-4D_L.
$$

Substituting the indicators gives 12 for Silence, $12-2=10$ for
Instrumental music and $12-4=8$ for Music with lyrics. The intercept is
the reference mean; each other coefficient is a difference from that mean,
negative here because both music conditions have lower means than Silence.
The factor order in the original setup makes Silence the reference level.

```{r}
#| label: ch06-indicator-fit
indicator_fit <- lm(score ~ sound, data = sound_data)
coef(indicator_fit)
```

```{r}
#| label: ch06-show-fit-residual
#| eval: false
fitted(indicator_fit)
round(resid(indicator_fit), 4)
```


`lm()` and `aov()` describe the same set of fitted means, so they give the
same residuals and the same ANOVA table. The formula names the outcome and
the explanatory variable; the study's design supplies the independence
assumption. A different reference level changes which comparisons the
coefficients name, but not the fitted means, the residuals or the table:
with Instrumental music as the reference, the coefficients become 10, 2
and $-2$. This indicator representation prepares the named models of
Sections 1.3 and 3.2.

### The decomposition and the F test

Fitting one mean leaves the total
sum of squares, while fitting separate group means leaves the within-group
sum of squares. On the same thirty scores these are 258 and 178; their
reduction, 80, is the between-group sum of squares. In the figure these
are the steps between the diamonds and the spread of the points around
their own diamond. `aov()` fits the same group-mean model as `lm()`, and
`anova()` prints its decomposition.

\Needspace{8\baselineskip}

```{r}
#| label: ch06-oneway-fit
one_fit <- aov(score ~ sound, data = sound_data)
anova(one_fit)
```

The two numerator degrees of freedom count the additional mean parameters
beyond one common mean. The 27 residual degrees of freedom are the thirty
observations less three fitted means. Dividing a sum of squares by its
degrees of freedom gives a mean square; the residual mean square,
$178/27=6.5926$, estimates the common error variance.

| Source | SS | df | MS | F | p |
|:--|--:|--:|--:|--:|--:|
| Sound condition | 80 | 2 | 40.0000 | 6.0674 | .0067 |
| Within groups | 178 | 27 | 6.5926 | | |
| Total | 258 | 29 | | | |

: One-way ANOVA on the thirty background-sound scores. {#tbl-ch06-oneway}

Read the result as a comparison of two named models on
the same thirty students. The null $H_0:\mu_S=\mu_I=\mu_L$ says that
the **one-mean model**, $Y_{ij}=\mu+\varepsilon_{ij}$, describes the
scores. It fits the grand mean 10 to every student and leaves the total,
$SS_T=258$ on 29 df, as its residual sum of squares. The group-mean model
leaves $SS_W=178$ on 27 df. The F statistic divides the drop in residual
sum of squares per added parameter by the larger model's residual mean
square:

$$
F=\frac{(258-178)/(29-27)}{178/27}=\frac{40}{6.5926}=6.0674.
$$

```{r}
#| label: ch06-show-one-mean
#| eval: false
common_fit <- lm(score ~ 1, data = sound_data)
anova(common_fit, indicator_fit)
```


The numerator is $MS_B$ and the denominator $MS_W$. Under independent Normal errors with a
common variance, the exact reference is $F_{2,27}$. The p-value .0067
is the probability of an F statistic at least this large under the
one-mean null model and its conditions. It provides evidence against
that model; Section 3.2 uses the same comparison principle for two
models of the feedback study. Rejection establishes an omnibus difference,
not a difference between every pair or a real effect of music on reading
from these invented observations.

The score differences remain 2 and 4 points below Silence. Alongside them,
$\eta^2=SS_B/SS_T=80/258=.3101$ is the share of the one-mean model's
residual sum of squares that the three group means remove, an in-sample
variation summary. It is neither a causal probability nor a replacement
for an interval. Subtracting the fitted-means contribution expected from
noise gives the bias-adjusted omega-squared, .2525.


```{r}
#| label: ch06-show-variation
#| eval: false
80 / 258
(80 - 2*(178/27)) / (258 + 178/27)
```

### Plotting: what does each mark represent? {#ch06-plotting}

A graph starts with a question about the records. In the sound study,
we want to see individual scores, their within-condition spread and the
location of each condition. The `ggplot2` package describes a graph by
its data, a **mapping** of variables to visual positions or colour, and
**layers** of marks. Mapping `sound` to the horizontal position and
`score` to the vertical position makes the comparison concrete. Changing
the marks can reveal or conceal features without changing the data.

```{r}
#| label: ch06-plotting-demo
#| eval: false
ggplot(sound_data, aes(sound, score)) + geom_point()
ggplot(sound_data, aes(sound, score)) +
  geom_jitter(width = .15, height = 0)
```

The first call places each student exactly at the recorded condition and
score. Tied scores fall on top of each other: 30 students occupy only 23
distinct point locations. The second separates the points slightly from
side to side. Its zero vertical displacement keeps every score at its
recorded height. @fig-ch06-plot-choices compares the two views on the same
scale. The code shows the essential mapping; labels and panel arrangement
make the printed comparison easier to read.

```{r}
#| label: fig-ch06-plot-choices
#| echo: false
#| fig-width: 6.4
#| fig-height: 3.4
#| fig-cap: "The same thirty scores, with and without horizontal displacement of tied points."
#| fig-alt: "Plain points show 23 distinct locations for 30 students. Horizontal jitter separates tied observations but leaves their vertical scores unchanged. Both panels retain the same three conditions and score scale."
plot_data <- rbind(transform(sound_data, view = "Plain points"),
  transform(sound_data, view = "Horizontal jitter"))
plot_data$view <- factor(plot_data$view,
  levels = c("Plain points", "Horizontal jitter"))
set.seed(12345)
ggplot(plot_data, aes(sound, score)) +
  geom_point(data = subset(plot_data, view == "Plain points"), size = 2) +
  geom_jitter(data = subset(plot_data, view == "Horizontal jitter"),
    width = .15, height = 0, size = 2) +
  facet_wrap(~ view) +
  scale_x_discrete(labels = c("Silence", "Instrumental\nmusic", "Music with\nlyrics")) +
  labs(x = "Background sound", y = "Comprehension score") +
  theme_minimal(base_size = 11)
```

**Predict, then explain.** If only the horizontal displacement changes,
will the fitted means, residuals or F test change? They remain identical:
the graphical displacement is separate from the recorded scores used by
the model. Conversely, moving scores vertically in the actual data would
change the analysis. A useful graph makes this distinction clear.

The mean diamonds in @fig-ch06-sound-groups add a second kind of mark:
each represents a calculated group mean, rather than another student.
The dots show spread and overlap; the diamonds make the comparison of
means easier to see. Later graphs use lines to connect cell means or
candidate criterion values, and histograms to summarize repeated studies.
In each case, begin by asking what one mark represents and what comparison
the axes make possible.

## Contrasts and Comparison Families {#ch06-contrast-families}

### Contrasts and their standard error

Suppose the two questions compare each music condition with Silence:
Instrumental music versus Silence, and Music with lyrics versus Silence.
In the order Silence, Instrumental music, Music with lyrics, their
coefficient vectors are $(-1,1,0)$ and $(-1,0,1)$. Each vector sums to
zero. A quantity

$$
C=\sum_jc_j\mu_j,\qquad \sum_jc_j=0,
$$

is a contrast of the process means. Its estimate replaces process means with
the observed group means:
$\widehat C=\sum_jc_j\bar y_j$. Thus our estimates are $-2$ and $-4$
points: both music conditions lie below Silence. Calling a comparison a
contrast specifies this mathematical form; it does not say whether the
comparison was planned before seeing the results or selected afterward.
That is additional information about the analysis.

For independent groups with a common error variance, the estimated
variance of a weighted sum adds each mean's variance with its squared
weight. Using the common estimate $MS_W$ gives

$$
SE(\widehat C)=\sqrt{MS_W\sum_j\frac{c_j^2}{n_j}}.
$$

Each comparison uses two of the means, so both have estimated SE
$\sqrt{(178/27)(1/10+1/10)}=1.1483$ points. The third condition still
contributes to the common error estimate and its 27 residual degrees of
freedom. This matters when comparing these results with an analysis
restricted to only two conditions.

```{r}
#| label: ch06-show-contrast
#| eval: false
sum(c(-1, 1, 0) * c(12, 10, 8))
sqrt((178/27) * sum(c(-1, 1, 0)^2 / 10))
```


### A declared family: Bonferroni

An individual 95% interval uses a .05 error probability for that one
comparison. The declared family here is *each music condition against
Silence*. To cover both of its comparisons with at least 95% familywise
confidence, Bonferroni allocates .025 to each two-sided comparison. Each
tail then receives $.05/(2\times2)=.0125$. The calibration uses the $t$
reference distribution with the model's 27 residual degrees of freedom
[@student1908], giving the critical value

$$
t_{1-.05/(2\times2),27}=2.3734.
$$

The estimates and SEs stay the same; the larger multiplier widens each
interval. For the Instrumental-minus-Silence comparison, the calculation is

\Needspace{8\baselineskip}

```{r}
#| label: ch06-bonferroni
#| include: false
mse <- deviance(one_fit) / df.residual(one_fit)
means <- with(sound_data, tapply(score, sound, mean))
estimate <- c(Instrumental = unname(means[2] - means[1]),
  Lyrics = unname(means[3] - means[1]))
se <- sqrt(mse * (1/10 + 1/10))
critical <- qt(1 - .05/(2*2), df.residual(one_fit))
raw_p <- 2 * pt(-abs(estimate/se), df.residual(one_fit))
cbind(estimate, SE = se, lower = estimate-critical*se,
  upper = estimate+critical*se,
  adjusted_p = p.adjust(raw_p, method = "bonferroni"))
```


```{r}
#| label: ch06-show-bonferroni
#| eval: false
se <- sqrt((178/27) * (1/10 + 1/10))
critical <- qt(1 - .05/4, 27)
-2 + c(-1, 1) * critical * se
```

| Comparison | Estimate | SE | Simultaneous interval | Adjusted p |
|:--|--:|--:|:--|--:|
| Instrumental music minus Silence | $-2$ | 1.1483 | [$-4.7253$, 0.7253] | .1859 |
| Music with lyrics minus Silence | $-4$ | 1.1483 | [$-6.7253$, $-1.2747$] | .0034 |

: Bonferroni results for the declared family of two comparisons. {#tbl-ch06-bonferroni tbl-colwidths="[34,13,11,28,14]"}

Multiplying each raw p-value by two and capping at one gives the
corresponding Bonferroni-adjusted p-value. Music with lyrics minus Silence
is distinguishable from zero with this familywise calibration: its whole
interval lies below zero. The Instrumental-music interval includes zero
although the omnibus ANOVA rejected equality. It allows a difference from
4.7253 points below Silence to 0.7253 points above it. The two results are
compatible, because the omnibus test and each contrast address different
statements.

The family calibration has a purpose and a price. The familywise error
rate is the probability of at least one
false rejection among the comparisons in a family. If both null hypotheses
were true, two independent comparisons, each calibrated at .05, would give
a familywise error probability of $1-.95^2=.0975$. These two comparisons share the Silence mean, and
dividing .05 across the family bounds the rate at .05 under any
dependence between them, regardless of which null hypotheses are true. The guarantee is paid for in width, which
Section 2.3 compares, and in power, which Section 6.2 quantifies. Declare
the family from the question before seeing the results; enlarging it
afterwards spends power on questions nobody asked.

### All pairs and every contrast: Tukey and Scheffé

If the question includes every pair, there are three comparisons. Tukey's
procedure uses the joint distribution of pairwise mean differences under the
common-variance Normal-error model. `TukeyHSD()` supplies those simultaneous
intervals and their adjusted p-values [@tukey1953].

\Needspace{8\baselineskip}

```{r}
#| label: ch06-tukey
TukeyHSD(one_fit, "sound", conf.level = .95)
```

The output labels each subtraction direction. Music with lyrics minus
Silence is $-4$, with interval [$-6.847$, $-1.153$] and adjusted p
.0047. Instrumental music minus Silence and Music with lyrics minus
Instrumental music both equal $-2$, with interval [$-4.847$, 0.847] and
adjusted p .2085. @tbl-ch06-calibrations sets one comparison, Instrumental
music minus Silence, under four calibrations. The fourth, Scheffé, protects every contrast
of the three means.

\Needspace{8\baselineskip}


```{r}
#| label: ch06-show-multipliers
#| eval: false
c(qt(.975, 27), qt(1 - .05/4, 27))
c(qtukey(.95, 3, 27)/sqrt(2), sqrt(2*qf(.95, 2, 27)))
```

| Calibration | Family protected | Multiplier | 95% interval |
|:--|:--|--:|:--|
| Pointwise t | This comparison alone | 2.0518 | [$-4.3561$, 0.3561] |
| Bonferroni | The two comparisons with Silence | 2.3734 | [$-4.7253$, 0.7253] |
| Tukey | All three pairs | 2.4794 | [$-4.8470$, 0.8470] |
| Scheffé | Every contrast of the three means | 2.5900 | [$-4.9740$, 0.9740] |

: Instrumental music minus Silence, $-2$ points with SE 1.1483, under four calibrations. {#tbl-ch06-calibrations tbl-colwidths="[17,39,16,28]"}

The estimate, $-2$, and its SE, 1.1483, are the same in every row; only
the multiplier changes. For the four calibrations in this table,
protecting a larger family uses a larger multiplier at familywise level
.05, so the intervals are wider. Here the pointwise multiplier is $t_{.975;27}$, the Tukey
multiplier is $q_{.95;3,27}/\sqrt2$, and the Scheffé multiplier is
$\sqrt{2F_{.95;2,27}}$. Scheffé covers the widest family: it also
protects non-pairwise questions such as the average of the two music
conditions minus Silence, an estimate of $-3$ points. An interval should
therefore be read together with the comparison
and the family it represents.

The all-pairs family also covers a comparison chosen after seeing the
means. Suppose Music with lyrics minus Silence were reported because it is
the largest observed difference. That choice looked at all three pairs, so
its calibration would come from the all-pairs family, [$-6.8470$,
$-1.1530$], rather than from the declared two-comparison family,
[$-6.7253$, $-1.2747$]. An omnibus rejection does not by itself protect
any comparison selected after examining the means. Conversely, a
separately specified contrast can have its own valid uncertainty procedure
without requiring omnibus rejection as an automatic preliminary gate. The
relevant information is the question, selection process and calibration,
rather than a universal sequence of significance decisions.

### Conditions and Welch's one-way ANOVA

The common-variance model assumes that within-group variability estimates
one shared error variance. In a real study, design information and observed
patterns help assess that working model. A small normality or variance test
cannot certify it, and ten students per condition give limited information
about the error distribution. In these invented data the three SDs, 2.7080,
2.4037 and 2.5820, are similar by construction, and every condition's
residuals lie between $-4$ and 4. In @fig-ch06-sound-groups each residual
is a point's vertical distance from its diamond, and the three conditions
spread alike.

Centering each score on its own fitted mean makes the spread comparison
more direct. In @fig-ch06-residuals, the three vertical strips correspond
to fitted means 8, 10 and 12. A residual above zero is a score above its
condition mean. The strips have similar spread here; in empirical data,
a much wider strip would challenge a common error variance. Only three
fitted values are available, so the plot is a comparison of group spreads,
not a detailed investigation of curvature across a continuous predictor.
The deliberately constructed regularity illustrates what to inspect; it
cannot establish the model's conditions or independence of students.

```{r}
#| label: ch06-show-residual-graph
#| eval: false
residual_data <- data.frame(fitted = fitted(one_fit),
  residual = resid(one_fit))
ggplot(residual_data, aes(fitted, residual)) + geom_point() +
  geom_hline(yintercept = 0)
```

```{r}
#| label: fig-ch06-residuals
#| echo: false
#| fig-width: 6.1
#| fig-height: 3.0
#| fig-cap: "Centering scores on their fitted condition means exposes the residual spreads."
#| fig-alt: "Residuals form three vertical strips at fitted means 8, 10 and 12. Each spans minus 4 to 4; tied residuals overlap. A horizontal zero line marks the fitted mean in each condition."
residual_data <- data.frame(fitted = fitted(one_fit),
  residual = resid(one_fit))
ggplot(residual_data, aes(fitted, residual)) + geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_x_continuous(breaks = c(8, 10, 12)) +
  labs(x = "Fitted condition mean", y = "Residual (score points)") +
  theme_minimal(base_size = 11)
```

For independent groups whose error variances may differ, Welch's one-way
procedure provides an alternative [@welch1947]. It is the several-group
counterpart of the unequal-variance two-group t test. R selects it by default in `oneway.test()` [@rcore2025]:

\Needspace{8\baselineskip}

```{r}
#| label: ch06-welch-oneway
oneway.test(score ~ sound, data = sound_data)
```

It gives $F(2, 17.9561)=5.5144$ and p .0136, beside the pooled
$F(2,27)=6.0674$ and p .0067. Welch weights each condition mean by
$n_j/s_j^2$, the inverse of its own estimated sampling variance, and
refers the statistic to fewer denominator degrees of freedom because each
condition now estimates its own variance. With similar observed SDs the
two procedures reach the same conclusion at .05, while their p-values
differ. Each p-value belongs to its own variance model and small-sample
calibration, so the two are results to explain through those assumptions
rather than votes on which numerical conclusion to prefer. With small
groups or clearly unequal spreads, the model choice can matter
appreciably.

For the same two groups and common-variance denominator, the pooled t
test (`var.equal = TRUE`) and ANOVA give $F=t^2$ and the same two-sided
p-value; neither R's default Welch test nor our
three-group contrast, which also uses the third condition's residuals,
is that matched calculation.

## Two-Way ANOVA and Nested Models {#ch06-factorial-comparisons}

### Cells and margins

A second invented study has 24 students studying the same
fractions material. Each is individually assigned to one combination of
three feedback conditions and two practice schedules, four students per
cell, and takes a common 20-item quiz one week after the final practice.
A cell is one feedback/practice combination. The question is how the
Spaced-minus-Massed comparison changes across feedback conditions.

| Feedback | Massed scores | Spaced scores |
|:--|:--|:--|
| Rereading | 4, 5, 7, 8 | 4, 5, 7, 8 |
| Answer feedback | 6, 7, 9, 10 | 10, 11, 13, 14 |
| Explanatory feedback | 8, 9, 11, 12 | 16, 17, 19, 20 |

: Twenty-four invented fractions-quiz scores. {#tbl-ch06-feedback-data tbl-colwidths="[35,32,33]"}

These values were constructed to make the calculations transparent. They are
teaching inputs, not empirical evidence that one feedback condition works
better. Each score comes from a different student; the four scores within a cell
are not repeated measurements of one student. Repeated patterns in their
deviations were deliberately chosen and cannot establish assumptions about a
population.

The data frame `feedback_data` holds these scores, their feedback and
practice conditions, and a student identifier.

```{r}
#| label: ch06-feedback-setup
#| include: false
feedback_data <- data.frame(
  student = 1:24,
  feedback = factor(rep(c("Rereading", "Answer feedback",
    "Explanatory feedback"), each = 8),
    levels = c("Rereading", "Answer feedback",
      "Explanatory feedback")),
  practice = factor(rep(rep(c("Massed", "Spaced"),
    each = 4), 3), levels = c("Massed", "Spaced")),
  score = c(4,5,7,8, 4,5,7,8, 6,7,9,10, 10,11,13,14,
    8,9,11,12, 16,17,19,20))
```

Computing cell counts and means before fitting any model makes the six
cells and their denominators visible.

\Needspace{15\baselineskip}

```{r}
#| label: ch06-cell-summaries
with(feedback_data, table(feedback, practice))
cell_mean <- with(feedback_data,
  tapply(score, list(feedback, practice), mean))
cell_mean
```

\Needspace{12\baselineskip}


```{r}
#| label: ch06-show-margins
#| eval: false
rowMeans(cell_mean)
colMeans(cell_mean)
```

@tbl-ch06-cell-means adds the margins and the three practice changes to
these cell means.

| Feedback | Massed | Spaced | Equally weighted margin | Spaced minus Massed |
|:--|--:|--:|--:|--:|
| Rereading | 6 | 6 | 6 | 0 |
| Answer feedback | 8 | 12 | 10 | 4 |
| Explanatory feedback | 10 | 18 | 14 | 8 |
| Practice margin | 8 | 12 | 10 | 4 |

: Cell means and equally weighted margins of the balanced design. {#tbl-ch06-cell-means tbl-colwidths="[31,12,12,23,22]"}

Each cell has four students, so averaging the two practice cells equally gives
the feedback group's overall mean. Similarly, each practice margin averages
equally across the three feedback conditions. In an unbalanced design, a
margin must specify its weights: the observed mix of students and an equally
weighted target can answer different questions.
In this balanced design each main effect is the average of its simple
effects: the 4-point practice difference is $(0+4+8)/3$. The additive
model below imposes that one difference on every feedback condition.

@fig-ch06-feedback-interaction draws the six cell means with one line for
each feedback condition. Each line joins two separate cell means; it does
not trace individual students across repeated measurements.


```{r}
#| label: ch06-show-interaction
#| eval: false
cell_plot <- aggregate(score ~ feedback + practice,
  feedback_data, mean)
ggplot(cell_plot, aes(practice, score, colour = feedback,
  group = feedback)) + geom_line() + geom_point()
```

```{r}
#| label: fig-ch06-feedback-interaction
#| echo: false
#| fig-pos: "!ht"
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: "The practice difference depends on feedback."
#| fig-alt: "Rereading joins 6 to 6, Answer feedback joins 8 to 12, and Explanatory feedback joins 10 to 18 from Massed to Spaced. The lines have different slopes."
cell_plot <- aggregate(score ~ feedback + practice,
  feedback_data, mean)
ggplot(cell_plot, aes(practice, score, colour = feedback,
  group = feedback)) + geom_line() + geom_point(size = 2) +
  labs(x = "Practice", y = "Mean number correct",
    colour = "Feedback") +
  theme_minimal(base_size = 11)
```

The observed changes are 0, 4 and 8 points, so the three lines have
different slopes. This describes interaction in the observed means.
Inference about a process interaction additionally requires a model and a
measure of residual variation.

### Named models and the nested-model F

The additive model is a restricted version of the full model

$$
Y_{ijk}=\mu+\alpha_j+\beta_k+(\alpha\beta)_{jk}+\varepsilon_{ijk}.
$$

Here $j$ indexes feedback, $k$ practice and $i$ students within a cell.
With sum-to-zero identifying constraints, $\mu$ is the grand mean,
$\alpha_j$ and $\beta_k$ describe marginal departures from it, and
$(\alpha\beta)_{jk}$ the remaining cell departure. These constraints
prevent the effects from shifting freely while preserving the same means.
One grand mean, two feedback directions, one practice direction and two
interaction directions describe six fitted cell means, leaving
$24-6=18$ residual df. The independent Normal errors have common variance.

The **additive model**, `score ~ feedback + practice`, gives every
feedback condition the same Spaced-minus-Massed difference. Its fitted
coefficients are 4 for Rereading under Massed practice, 4 and 8 for Answer
and Explanatory feedback minus Rereading, and 4 for Spaced minus Massed,
the average of Section 3.1's three changes. The **full model**,
`score ~ feedback * practice`, fits the six cell means and lets that
difference vary. The formula expands to
`feedback + practice + feedback:practice`.

Fit both models to the same 24 students. They leave residual sums of
squares 124 on 20 df and 60 on 18 df, respectively. Their comparison is

$$
F=\frac{(124-60)/(20-18)}{60/18}=9.6,
$$

with 2 and 18 df and $p=.0015$. The denominator comes from the larger
model, which estimates error variance without imposing the common
practice difference being tested. This is the interaction test and the
same comparison principle as Section 1.3's one mean against three means.

\Needspace{8\baselineskip}

```{r}
#| label: ch06-nested-f
additive_fit <- lm(score ~ feedback + practice, data = feedback_data)
full_fit <- lm(score ~ feedback * practice, data = feedback_data)
coef(additive_fit)
anova(additive_fit, full_fit)
```

Keep the response and the records explicitly the same: default missing-value
removal can otherwise change the comparison.

The full factorial table can also be read as a sequence of named models:
one mean (`score ~ 1`, RSS 476 on 23 df), feedback only
(`score ~ feedback`, 220 on 21), additive (124 on 20) and full (60 on 18).
Each row's sum of squares is the reduction from one fit to the next, and
every F row uses the full model's residual mean square, $60/18=3.3333$.

\Needspace{8\baselineskip}

```{r}
#| label: ch06-factorial-fit
two_fit <- aov(score ~ feedback * practice, data = feedback_data)
anova(two_fit)
```

\Needspace{12\baselineskip}

| Source | Models defining the SS reduction | SS | df | MS | F | p |
|:--|:--|--:|--:|--:|--:|--:|
| Feedback | One mean, then feedback only | 256 | 2 | 128.0000 | 38.4 | < .0001 |
| Practice | Feedback only, then additive | 96 | 1 | 96.0000 | 28.8 | < .0001 |
| Interaction | Additive, then full | 64 | 2 | 32.0000 | 9.6 | .0015 |
| Residual | Full model | 60 | 18 | 3.3333 | | |
| Total | One-mean model | 476 | 23 | | | |

: Full balanced factorial decomposition, read as a sequence of named models. {#tbl-ch06-factorial tbl-colwidths="[15,33,9,6,13,8,16]"}

The first two rows use successive models for their numerators and the
full model for their common error denominator; they are not standalone
tests of the two models named in those rows. A sequence stopping at the
additive model would divide by $124/20=6.2$ and give 20.6452 rather than
38.4 for the feedback row. The interaction row is the additive-versus-full
test. The three questions are equal feedback margins averaged equally
over practice, equal practice margins averaged equally over feedback,
and a common practice difference across feedback conditions.

The balanced layout makes these source components orthogonal, so changing
the order of the factors leaves the table unchanged. A single-model
`anova()` table produces sequential sums of squares, however, and this
convenience does not extend automatically to unequal cells.

Suppose the two Explanatory-Spaced students scoring 19 and 20 were
unavailable. Counts become 4, 4, 4, 4, 4 and 2, leaving 16 residual df.
The equally weighted Spaced margin is $(6+12+16.5)/3=11.5$, whereas the
ten remaining Spaced students average 10.5: two different questions about
weights. @tbl-ch06-unbalanced fits the full
model to these 22 students with each formula order.

\Needspace{10\baselineskip}


```{r}
#| label: ch06-show-orders
#| eval: false
dat <- subset(feedback_data, student < 23)
anova(lm(score ~ feedback * practice, data = dat))
anova(lm(score ~ practice * feedback, data = dat))
```

| Term | Formula `feedback * practice` | Formula `practice * feedback` |
|:--|--:|--:|
| Feedback | 139.7576 | 157.7500 |
| Practice | 52.0833 | 34.0909 |
| Interaction | 36.2500 | 36.2500 |
| Residual | 50.5000 | 50.5000 |

: Sequential sums of squares for the 22 remaining students, by the order of the formula. {#tbl-ch06-unbalanced tbl-colwidths="[28,36,36]"}

With feedback first, the sequence is one mean, feedback only, additive,
then full; with practice first it is one mean, practice only, additive,
then full. Thus the feedback row measures feedback against one mean
(139.7576) in one order and feedback added to practice (157.7500) in the
other. The final step stays additive against full. Its RSS values are
86.75 on 18 df and 50.50 on 16 df, giving $F(2,16)=5.7426$, $p=.0132$,
in either order. One sequential table does not give order-free tests of
main effects with unequal cells; naming the two models does.

### A declared interaction contrast and the price of selection

The study plan declared one interaction comparison before any
scores were seen: whether spacing helps more under Answer feedback than
under Rereading. Its difference of differences is

$$
\widehat{DD}=(12-8)-(6-6)=4.
$$

The coefficients are $(1,-1,-1,1)$ in cell order Rereading-Massed,
Rereading-Spaced, Answer-Massed, Answer-Spaced. Four different sets of
students supply the cell means. Under independent errors with common
variance, their estimated variances add:

$$
SE(\widehat{DD})=\sqrt{MS_E(1/4+1/4+1/4+1/4)}=1.8257.
$$

Declared as the one comparison in advance, it is a family of one, and its
pointwise 95% interval, $[0.1643,7.8357]$, is the warranted statement.
Had the same contrast been picked after inspecting the plot, the interval
would need to cover every difference of differences the plot could have
suggested, the two-dimensional interaction space. Section 2.3's Scheffé
method uses $\sqrt{2F_{.95;2,18}}=2.6663$ for that family. The same
estimate and SE would then give $[-0.868,8.868]$.

\Needspace{13\baselineskip}

```{r}
#| label: ch06-selected-interaction
#| include: false
dd <- (cell_mean["Answer feedback", "Spaced"] -
  cell_mean["Answer feedback", "Massed"]) -
  (cell_mean["Rereading", "Spaced"] -
  cell_mean["Rereading", "Massed"])
mse_two <- deviance(two_fit) / df.residual(two_fit)
se_dd <- sqrt(mse_two * 4/4)
c(estimate = dd, SE = se_dd)
dd + c(-1, 1) * qt(.975, 18) * se_dd
dd + c(-1, 1) * sqrt(2 * qf(.95, 2, 18)) * se_dd
```

```{r}
#| label: ch06-show-dd
#| eval: false
dd <- (12 - 8) - (6 - 6)
se_dd <- sqrt((60/18) * (1/4 + 1/4 + 1/4 + 1/4))
dd + c(-1, 1) * qt(.975, 18) * se_dd
dd + c(-1, 1) * sqrt(2*qf(.95, 2, 18)) * se_dd
```

The counterfactual Scheffé interval includes zero although the interaction
F test rejects. The F test asks whether both interaction directions are
zero jointly; the interval covers one direction with protection over all
of them. Other directions may carry substantial evidence, so neither
result cancels the other. @tbl-ch06-dd-quantities sets the declared DD
beside Answer feedback's simple practice effect, which happens to share
its estimate, 4.

\Needspace{12\baselineskip}


```{r}
#| label: ch06-show-simple-effect
#| eval: false
se_simple <- sqrt(2 * (60/18) / 4)
4 + c(-1, 1) * qt(.975, 18) * se_simple
```

| Quantity | Estimate | SE | 95% interval | Calibration |
|:--|--:|--:|:--|:--|
| Answer's simple practice effect | 4 | 1.2910 | [1.2877, 6.7123] | Pointwise |
| Answer-minus-Rereading DD, declared in advance | 4 | 1.8257 | [0.1643, 7.8357] | Pointwise (family of one) |
| The same DD, had it been picked after the plot | 4 | 1.8257 | [$-0.8680$, 8.8680] | Scheffé (interaction family) |

: Three quantities from the interaction section; the Scheffé row covers the two-dimensional interaction family. {#tbl-ch06-dd-quantities tbl-colwidths="[36,10,8,22,24]"}

The two DD intervals differ only in their multiplier; the comparison's
history decides which applies. Answer's simple effect compares two cell
means, so its SE, $\sqrt{2MS_E/4}$, is smaller. It contains an overall
practice component as well as an interaction component and therefore lies
outside the interaction-only family; that family's Scheffé guarantee does
not transfer to it. Choosing the narrower interval after seeing its
conclusion does not keep a family's guarantee. Name the quantity, its
history, family and interval together.

Report magnitude first in score points: the practice differences are
0, 4 and 8, and the declared DD is 4 with SE 1.8257. Variation shares
may follow with named denominators. The interaction sum of squares over
the total is $64/476=.1345$; over the interaction and residual sums of
squares it is $64/124=.5161$, the interaction's partial eta-squared.
Both are in-sample summaries of an authored
design and neither replaces the pattern of practice differences.

## Proportions and Counts {#ch06-counts}

A completion count needs a different response model: independent trials
with one common completion probability give a binomial model. Its one
parameter is that probability, fitted by the sample proportion, which
Section 5.2 also obtains by likelihood.

### One proportion

A score model is not the appropriate starting point for every outcome.
Consider the invented completion table in
@tbl-ch06-completion-counts. It contains aggregate counts for 100 records,
classified by learning format and completion. These records have no
linkage to either score dataset above. The counts themselves establish
neither random allocation nor representative sampling.

\Needspace{9\baselineskip}

| Format | Completed | Not completed | Total |
|:--|--:|--:|--:|
| Online | 32 | 18 | 50 |
| In person | 18 | 32 | 50 |
| Total | 50 | 50 | 100 |

: Authored completion counts for two learning formats. {#tbl-ch06-completion-counts}

For the Online group, the event is completion and the denominator is its
50 records. Under an independent Bernoulli model with a common completion
probability $p$, the count of completions has a binomial distribution. The
estimated probability is $\widehat p=32/50=.64$. If records were clustered
within dependent classes or came from repeated observations of the same
people, this ordinary independent-unit uncertainty model would require
modification.

Suppose the benchmark is $p_0=.5$. The score test standardizes the difference
using the variance under that null:

$$
z=\frac{\widehat p-p_0}{\sqrt{p_0(1-p_0)/50}}
=\frac{.64-.5}{\sqrt{.5(.5)/50}}=1.9799.
$$

Squaring gives $X^2=z^2=3.92$, compared with a chi-squared distribution
on one degree of freedom. The procedure uses a large-sample reference;
the expected numbers of successes and failures under this null are both 25.

\Needspace{8\baselineskip}

```{r}
#| label: ch06-one-proportion
one_prop <- prop.test(32, 50, p = .5, correct = FALSE)
one_prop$estimate
one_prop$statistic
one_prop$p.value
one_prop$conf.int
```

The two-sided p-value is .0477. The estimated completion probability is
.64, with 95% Wilson score interval [.5014, .7586]. `correct = FALSE`
explicitly selects the uncorrected score procedure; accepting R's default
continuity correction would specify a different calculation. The call and
the reported result should identify which procedure was used.

The score interval contains the candidate probabilities that their own
two-sided score tests would retain [@wilson1927]. Each candidate supplies
the variance for its null calculation. This inversion explains why the
interval is linked to the score test and is not simply
$\widehat p\pm1.96\sqrt{\widehat p(1-\widehat p)/n}$, the Wald interval,
and why its center lies slightly closer to .5 than .64 does. The p-value assesses the
stipulated benchmark; it is not the probability that $p=.5$ is true or a
measure of the completion rate's practical importance.

### The two-by-two table

Now ask whether completion probability differs by format. This uses both
rows and a different null: format and completion are independent in the
working population model, equivalently equal completion probabilities in
this two-group setting. Constructing labelled counts prevents an accidental
change in the event or group direction.

\Needspace{12\baselineskip}

```{r}
#| label: ch06-two-way-counts
#| include: false
counts <- matrix(c(32, 18, 18, 32), nrow = 2,
  byrow = TRUE,
  dimnames = list(Format = c("Online", "In person"),
    Result = c("Completed", "Not completed")))
prop.table(counts, margin = 1)
count_test <- chisq.test(counts, correct = FALSE)
count_test$expected
count_test$statistic
count_test$p.value
count_test$residuals
```

The matrix `counts` records the four cells in the displayed row and
column order. Its row proportions and the independence test are

```{r}
#| label: ch06-show-counts
prop.table(counts, margin = 1)
chisq.test(counts, correct = FALSE)
```

`margin = 1` divides each row by its own total. Completion rates are .64
for Online and .36 for In person, a difference of .28, or 28 percentage
points, in the Online-minus-In-person direction. Under independence the
fitted expected count in row $j$, column $k$ is

$$
E_{jk}=\frac{(\text{row }j\text{ total})
(\text{column }k\text{ total})}{N}.
$$

All four expected counts here equal $50\times50/100=25$. Pearson's
statistic accumulates the squared discrepancies relative to those expected
counts:

$$
X^2=\sum_{j,k}\frac{(O_{jk}-E_{jk})^2}{E_{jk}}
=4\left(\frac{7^2}{25}\right)=7.84.
$$

```{r}
#| label: ch06-show-pearson
#| eval: false
expected <- 50 * 50 / 100
sum((c(32, 18, 18, 32) - expected)^2 / expected)
```


For a $J$-by-$K$ table the usual independence reference has
$(J-1)(K-1)$ degrees of freedom. Here $df=1$ and p .0051. The observed
Pearson statistic $X^2$ is a number calculated from the table; $\chi^2_1$
names its approximate reference distribution under the null model.

The Pearson residuals are $(O-E)/\sqrt E$: 1.4, $-1.4$, $-1.4$, 1.4
in row order. Their signs locate the pattern: more Online completions and
fewer Online non-completions than the independence fit predicts, with the
opposite departures for In person. Squaring and adding these residuals
recovers $X^2$. They are related through the table margins, so their signs
do not constitute four independent cell tests.

The approximation concerns expected counts under the fitted null model,
along with independent observational units and a sampling model appropriate
to the table. Looking only at observed cell counts misses part of this
check. Sparse expected cells may require an alternative reference or a more
careful model. In the present example all expected counts are 25, supporting
the ordinary chi-squared approximation under the stated unit model.

The one-proportion test asked whether Online completion probability equals
a fixed .5 benchmark. The complete-table test asks whether the two format
probabilities are equal; it estimates their common probability under the
null from the full table. Their different p-values answer different
questions. Neither test establishes that changing format would cause
completion to change without relevant design information.


### Effect scales and alternative procedures {#ch06-categorical-reference}

A chi-square test says whether the table departs from no association, not
how large the association is or on which scale. @tbl-ch06-effect-scales
sets four common scales side by side for the completion table, with
$p_1$ the Online and $p_0$ the In-person completion proportion, and
gives the question that each one answers [@agresti2018].

\Needspace{17\baselineskip}

| Scale | Definition | Value | What it answers |
|:--|:--|--:|:--|
| Difference in proportions | $p_1-p_0$ | .2800 | How many percentage points more often the event occurs in one group: an absolute gap |
| Risk ratio | $p_1/p_0$ | 1.7778 | How many times as probable the event is in one group: a relative gap, which changes when the other outcome is called the event |
| Odds ratio | $\dfrac{p_1/(1-p_1)}{p_0/(1-p_0)}$ | 3.1605 | How many times as large the odds of the event are; it stays the same when rows and columns exchange roles, so a design that samples on the outcome can still estimate it |
| Cramér's V | $\sqrt{\dfrac{X^2}{N\min(J-1,K-1)}}$ | .2800 | How strong the association is, from 0 to 1, for a table of any size; it carries no direction and names no cell comparison |

: Four effect scales for the completion table, Online versus In person. {#tbl-ch06-effect-scales tbl-colwidths="[15,25,9,51]"}

```{r}
#| label: ch06-show-effect-scales
#| eval: false
p1 <- 32/50; p0 <- 18/50
c(difference = p1-p0, risk_ratio = p1/p0,
  odds_ratio = (p1/(1-p1))/(p0/(1-p0)), V = sqrt(7.84/100))
```


The same counts give a gap of 28 percentage points, a risk ratio of
1.7778 and an odds ratio of 3.1605. The odds ratio lies farther from 1
than the risk ratio because completion is common: Online's completion
odds are $32/18=1.7778$ and In person's are $18/32=.5625$, so an odds
ratio is not read as a risk ratio. Calling non-completion the event
changes the risk ratio to $.36/.64=.5625$ and the odds ratio to its
reciprocal, .3164, so the event and the direction are named beside every
scale. The balanced margins make Cramér's V [@cramer1946] and the
absolute difference in proportions coincide here; with other margins
they generally differ.

Other procedures change one part of the analysis while keeping the same
table, and each answers its own question. The two-proportion interface
gives a large-sample interval for the difference in proportions, a
different quantity from the one-probability Wilson interval of
Section 4.1. For the Online-minus-In-person difference, .28, the
uncorrected two-proportion procedure gives the approximate pointwise 95%
interval [.0918, .4682], under independent binomial sampling within the
two groups. It describes uncertainty about that difference, rather than
about Online's probability alone. A continuity correction changes the calibration of the
Pearson statistic. Fisher's exact test [@fisher1935] conditions on the
table margins and uses an exact reference, which matters when expected
counts are small. These procedures answer different uncertainty questions for the same
counts. Keep the event, comparison and reference model explicit when
choosing among them.


## Numerical Optimization: Least Squares and Likelihood {#ch06-likelihood}

The chapter has used two fitting criteria: least squares chose the means
of Sections 1 and 3, and likelihood chooses the binomial probability of
Section 4. With independent Normal errors of common variance the two
criteria choose the same fitted means. A count calls naturally for its
binomial likelihood. **Numerical optimization** searches over candidate
parameter values to find the minimum or maximum of a stated objective.
We begin with one parameter whose best value can also be checked exactly.

### Least squares for one center

The fitted means in Section 1 were chosen by a rule: minimize the sum
of squared deviations from the observations. We can inspect this rule
with the ten Silence scores, 8, 9, 9, 11, 12, 13, 13, 14, 15 and 16.
These observations stay fixed. What changes is the proposed center,
denoted by $c$.

An **objective function** assigns a numerical criterion to each candidate
parameter value for fixed data. For a center fitted by least squares,
the objective is

$$
Q(c)=\sum_{i=1}^{10}(y_i-c)^2
     =(8-c)^2+(9-c)^2+\cdots+(16-c)^2.
$$

At $c=12$, the deviations are the Silence residuals of Section 1, and
their squares sum to 66 squared score points. At either 11 or 13, they
sum to 76. The sample mean minimizes this criterion; here it is 12 points. A numerical search lets us reproduce the
minimizing value without writing it into the fitting instruction.

\Needspace{6\baselineskip}

```{r}
#| label: ch06-scalar-squared-loss
silence <- subset(sound_data, sound == "Silence")$score
loss <- function(center) sum((silence - center)^2)
mean_fit <- optimize(loss, interval = range(silence))
c(center = mean_fit$minimum, squared_loss = mean_fit$objective)
```

The function `loss()` evaluates the squared-loss criterion at a proposed
center. `optimize()` searches over $[8,16]$, which contains the sample
mean. This smooth bowl has one minimum to locate. Changing the candidate
center changes the criterion value while keeping all ten scores fixed.

The two returned quantities have different meanings. `minimum` is
the **parameter location**, 12 points; `objective` is the function
value there, 66 squared points. The latter is a sum of squared
deviations, Silence's contribution to $SS_W$ in Section 1, rather than an
SE or an estimate of the mean. The search interval
also belongs to the fitting instruction: excluding the best center
would restrict which answer the search could return. Numerical stopping
tolerance concerns the accuracy of that search; statistical uncertainty
concerns what an estimate would do across the repetitions justified by
the study.

@fig-ch06-loss-search draws this objective for many candidate centers.
The horizontal axis gives a proposed center in score points, and the
vertical axis gives its squared loss. Each point on the curve answers a
fitting question for the same ten observations. The bottom of the bowl
is the best center according to this criterion.


```{r}
#| label: ch06-show-loss-curve
#| eval: false
curve(sapply(x, loss), from = 8, to = 16,
  xlab = "Candidate center", ylab = "Squared loss")
```

```{r}
#| label: fig-ch06-loss-search
#| echo: false
#| fig-pos: "!ht"
#| fig-width: 6.4
#| fig-height: 3.2
#| fig-cap: "The search interval changes the permitted answer, while the data and squared-loss curve stay fixed."
#| fig-alt: "The same bowl-shaped squared-loss curve appears in two panels. Searching centers from 8 to 16 gives the minimum at 12 with loss 66. Restricting the search to 13 through 16 excludes that point and gives an endpoint minimum at 13 with loss 76. Gray shows the excluded part of the same curve."
silence <- c(8, 9, 9, 11, 12, 13, 13, 14, 15, 16)
loss <- function(center) sum((silence - center)^2)
centers <- seq(8, 16, length.out = 161)
intervals <- c("Search: 8 to 16", "Search: 13 to 16")
curve <- data.frame(center = rep(centers, 2),
  loss = rep(sapply(centers, loss), 2),
  interval = factor(rep(intervals, each = length(centers)), levels = intervals))
allowed <- subset(curve, interval == intervals[1] | center >= 13)
best <- data.frame(center = c(12, 13), loss = c(66, 76),
  interval = factor(intervals, levels = intervals), label = c("12, 66", "13, 76"))
ggplot(curve, aes(center, loss)) + geom_line(color = "grey75") +
  geom_line(data = allowed, linewidth = .8) +
  geom_point(data = best, size = 2) +
  geom_text(data = best, aes(label = label), nudge_y = 25, vjust = 0, size = 3.5) +
  facet_wrap(~ interval) +
  labs(x = "Candidate center (score points)", y = "Squared loss (squared points)") +
  theme_minimal(base_size = 11)
```

**Predict, then explain.** Suppose the search is restricted to $[13,16]$.
Which center can it return, and have the scores changed? The right panel
retains the same curve but permits only centers at least 13. Loss rises
through that interval, so the exact constrained minimum is its left
endpoint, 13, with loss 76. The unrestricted mean 12 is excluded.
`optimize(loss, interval = c(13, 16))` returns about 13.0001 with loss
76.0015: this numerical search approaches the endpoint from inside.
Evaluating `loss(13)` checks the endpoint directly and gives 76. A
successful search therefore needs a sensible parameter range as well as
a correctly defined objective.

A tighter numerical tolerance can bring the returned center closer to
the permitted minimum. It cannot make ten students into a larger sample
or make their mean more representative of a population. These
least-squares identities are arithmetic; an uncertainty statement for
the fitted mean adds a model or design and its conditions.

### Likelihood for a binomial count

For the completion example, the observations are 32 completions in 50
Online trials. Let $p$ denote a candidate probability of completion.
Under the stipulated model of independent trials with a common
probability, the number completed follows a binomial distribution.
The model assigns probability

$$
P_p(X=32)={50\choose32}p^{32}(1-p)^{18}.
$$

In a probability calculation, we choose a value of $p$ and consider
which counts might occur. For fitting, we have already observed the
count: 32 stays fixed, as does the total of 50. We now evaluate the
same expression at different candidate values of $p$. Viewed in this
direction, it is the **likelihood function**,

$$
L(p)={50\choose32}p^{32}(1-p)^{18},\qquad 0\le p\le1.
$$

The binomial coefficient counts the ways 32 completions can occur
among 50 trials. It is constant as $p$ varies. In this binomial model, likelihood is the
probability of the observed data under each candidate model parameter;
it is not a probability distribution over that parameter. In
particular, a high likelihood at a candidate $p$ does not give the
probability that $p$ is true.

**Maximum likelihood estimation** chooses the parameter value that
makes the observed data most likely under the declared model. The
logarithm is increasing, so maximizing the log-likelihood gives the
same parameter value. For $0<p<1$,

$$
\ell(p)=\log {50\choose32}+32\log p+18\log(1-p).
$$

Logs turn a product of powers into a sum and make calculation easier.
The binomial coefficient affects the reported function value, although
it cannot change the maximizing $p$. Here the interior maximum can also
be found analytically. Differentiating with respect to the candidate
probability gives

$$
\ell'(p)=\frac{32}{p}-\frac{18}{1-p}
         =\frac{32-50p}{p(1-p)}.
$$

Setting this slope to zero gives $32-50p=0$, and therefore
$\widehat p=32/50=.64$. The denominator is positive inside $(0,1)$.
The slope is positive below .64 and negative above .64: the curve rises
to this point and then falls. This verifies a maximum rather than
merely finding a flat point.

To use the default minimization operation in `optimize()`, supply the
**negative** log-likelihood. A smaller negative log-likelihood means a
larger log-likelihood, so its minimum locates the same fitted probability.

\Needspace{11\baselineskip}

```{r}
#| label: ch06-scalar-binomial-mle
negative_loglik <- function(p) {
  -dbinom(32, size = 50, prob = p, log = TRUE)
}
likelihood_fit <- optimize(negative_loglik, interval = c(0, 1))
c(numerical = likelihood_fit$minimum, analytic = 32/50,
  negative_loglik = likelihood_fit$objective)
negative_loglik(.5)
```

`dbinom()` evaluates the binomial probability for the fixed count 32.
With `log = TRUE`, it computes the log probability directly. The
numerical and analytic estimates both round to .64. The small
numerical difference from exactly .64 reflects the search tolerance.
At the fitted probability the negative log-likelihood is 2.1465,
compared with 4.1330 at $p=.5$. The smaller value is the better fit
according to this criterion. The parameter estimate .64 is a
completion probability; the criterion value 2.1465 is not.


```{r}
#| label: ch06-show-likelihood-curve
#| eval: false
curve(dbinom(32, 50, x, log = TRUE), from = .05, to = .95,
  xlab = "Candidate probability", ylab = "Log-likelihood")
```

```{r}
#| label: fig-ch06-likelihood
#| echo: false
#| fig-pos: "!ht"
#| fig-width: 6.1
#| fig-height: 3.4
#| fig-cap: "The binomial log-likelihood is largest at the observed proportion."
#| fig-alt: "For 32 successes in 50 fixed trials, the log-likelihood curve peaks at candidate probability .64. A dashed line at .5 intersects the curve at a lower value."
likelihood_grid <- data.frame(p = seq(.05, .95, length.out = 200))
likelihood_grid$loglik <- dbinom(32, size = 50,
  prob = likelihood_grid$p, log = TRUE)
ggplot(likelihood_grid, aes(p, loglik)) + geom_line() +
  geom_vline(xintercept = .64) +
  geom_vline(xintercept = .5, linetype = 2) +
  labs(x = "Candidate completion probability", y = "Log-likelihood") +
  theme_minimal(base_size = 11)
```

The horizontal axis contains candidate parameter values, and the
vertical axis gives the log-likelihood of the same observed count at
each candidate. Thus the curve displays a fitting criterion, not a
sampling distribution or a distribution of probabilities over $p$.
The peak identifies the fitted value. The score test and interval in
Section 4 answer additional questions about a proposed null and the
uncertainty of estimation under their stated calibration.

Finally, check the boundaries of the parameter space. With both
completions and non-completions observed, $p=0$ and $p=1$ each give
likelihood zero, so the maximum is interior. If all 50 trials instead
had zero completions, the maximum would be at $p=0$; if all completed,
it would be at $p=1$. An interior numerical search can approach such
an endpoint without returning it exactly. The endpoint values must
therefore be examined explicitly in those cases.

## Simulation and Planning {#ch06-simulation}

The preceding analyses started with observed data and produced estimates,
intervals and tests. We can also begin with a declared generating model,
create data from it, and examine what an analysis does over repeated studies.
Here the questions concern an omnibus ANOVA procedure and the
sample size for a specified comparison.

### One generated study and its repetition

Imagine three independent groups whose outcomes follow Normal distributions,
each with standard deviation $\sqrt{10/3}=1.8257$ points. First set all three
population means to 8, so that the equal-means null hypothesis is true;
later the unequal means 6, 8 and 10 describe an alternative. These values
borrow the scale of the feedback study in Section 3, whose Massed cells
have means 6, 8 and 10 with four students each and whose factorial fit has
residual mean square $10/3$. They define a hypothetical world in which we
know the truth; the constructed scores do not establish that this
generating model describes real students.

Begin with four observations per group and analyse one generated study.

\Needspace{13\baselineskip}

```{r}
#| label: ch06-one-generated-anova
set.seed(60630)
n_per_group <- 4
sim_group <- factor(rep(1:3, each = n_per_group))
sim_score <- rnorm(3*n_per_group,
  mean = rep(c(8, 8, 8), each = n_per_group), sd = sqrt(10/3))
anova(lm(sim_score ~ sim_group))
```

The twelve outcomes are drawn from populations with the stipulated
common mean and SD.
The seed fixes the illustrated draw. Fitting one mean per group and
comparing it with a common mean gives 2 group degrees of freedom and
$12-3=9$ error degrees of freedom. For these generated data,
$SS_B=7.5977$ and $SS_W=32.9851$, giving $F=1.0365$ and $p=.3935$.
Applying the rule $p<.05$ therefore does not reject the equal-means null
in this study. Although the three generating means are identical, the
three sample means vary, so the between-group sum of squares is
positive. The estimated error variance, $MS_W=3.665$, also varies around
the generating variance $10/3$.

```{r}
#| label: fig-ch06-repeated-studies
#| echo: false
#| fig-pos: "H"
#| fig-width: 6.4
#| fig-height: 2.8
#| fig-cap: "One study contributes one p-value to the repeated results. The shaded first bin contains p-values below .05; the panels use different count scales."
#| fig-alt: "Histograms show 2,000 p-values from four students per group under equal means and unequal means. The first bin contains 112 null rejections and 1,309 alternative rejections. A vertical line at .3935 marks the first generated null study's p-value."
one_anova_p <- function(n, means, sigma) {
  group <- factor(rep(1:3, each = n))
  score <- rnorm(3*n, mean = rep(means, each = n), sd = sigma)
  anova(lm(score ~ group))[1, "Pr(>F)"]
}
set.seed(60630)
B <- 2000
results <- data.frame(n = c(4, 12), type1 = NA_real_, power = NA_real_)
for (j in 1:nrow(results)) {
  n <- results$n[j]
  null_p <- replicate(B, one_anova_p(n, c(8, 8, 8), sqrt(10/3)))
  alternative_p <- replicate(B, one_anova_p(n, c(6, 8, 10), sqrt(10/3)))
  results$type1[j] <- mean(null_p < .05)
  results$power[j] <- mean(alternative_p < .05)
  if (j == 1) first_p <- data.frame(p = c(null_p, alternative_p),
    model = factor(rep(c("Equal means: 8, 8, 8", "Unequal means: 6, 8, 10"),
      each = B), levels = c("Equal means: 8, 8, 8", "Unequal means: 6, 8, 10")))
}
wide_null <- replicate(B, one_anova_p(4, c(8, 8, 8), 2*sqrt(10/3)))
wide_alt <- replicate(B, one_anova_p(4, c(6, 8, 10), 2*sqrt(10/3)))
wide_rates <- c(type1 = mean(wide_null < .05), power = mean(wide_alt < .05))
stopifnot(identical(results$type1, c(.056, .05)),
  identical(results$power, c(.6545, .999)),
  isTRUE(all.equal(unname(wide_rates), c(.0525, .2165))))
first_p$decision <- factor(first_p$p < .05, levels = c(FALSE, TRUE),
  labels = c("p at least .05", "p below .05"))
one_study <- data.frame(p = .3935,
  model = factor("Equal means: 8, 8, 8", levels = levels(first_p$model)))
ggplot(first_p, aes(p, fill = decision)) +
  geom_histogram(binwidth = .05, boundary = 0, color = "white") +
  geom_vline(data = one_study, aes(xintercept = p), linetype = 2) +
  geom_text(data = one_study, aes(x = p, y = Inf, label = "One study: p = .3935"),
    inherit.aes = FALSE, hjust = -.05, vjust = 1.5, size = 3) +
  facet_wrap(~ model, scales = "free_y") +
  scale_fill_manual(values = c("grey80", "#356481"), name = NULL) +
  labs(x = "p-value from one generated study", y = "Number of studies") +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")
```

The result table `first_p` records one p-value and generating-model label
per simulated study. Its graphical comparison is

```{r}
#| label: ch06-show-repeated-graph
#| eval: false
ggplot(first_p, aes(p)) +
  geom_histogram(binwidth = .05, boundary = 0) +
  facet_wrap(~ model, scales = "free_y")
```

One generated study cannot describe a procedure's long-run behaviour.
Repeating the same calculation many times gives a collection of p-values
to compare with .05. In @fig-ch06-repeated-studies, the dashed line marks
the .3935 from the study we just followed. That study contributes one
p-value to the left-hand distribution; each additional study supplies
another. The shaded first bin counts rejections. With equal generating
means, 112 of the 2,000 studies reject, giving .056. With means 6, 8 and
10, 1,309 reject, giving .6545. The latter fraction estimates power for
this alternative. The vertical scales differ, so compare the fractions
of studies in the first bin, not the bars' physical heights.

@tbl-ch06-simulation reports 2,000 studies for each of four conditions. The null conditions use means $(8,8,8)$; the
alternative conditions use $(6,8,10)$. Each condition retains independent
Normal observations and the same common SD; seed 60630 fixes the
illustrated simulation sequence.

\Needspace{9\baselineskip}

| Students per group | Null rejection rate | MCSE | Power at means 6/8/10 | MCSE |
|--:|--:|--:|--:|--:|
| 4 | .056 | 0.0051 | .6545 | 0.0106 |
| 12 | .050 | 0.0049 | .9990 | 0.0007 |

: Operating characteristics in the declared generating models, $B=2000$ studies per condition. {#tbl-ch06-simulation}

The null rejection rate estimates the Type I error probability: how
often this .05 omnibus procedure rejects when all three population means
are equal. Both estimated rates are close to .05. The power column
estimates the probability of rejecting that same null under the specified
unequal means. It rises from .6545 with four students per group to .9990 with
12. This is power against the particular means 6, 8 and 10, with the stated
spread and independence; a smaller separation would give different power.

Each rate is itself an estimate based on finitely many repetitions. If
$\widehat q$ is a simulated rejection proportion from $B$ independent studies,
its approximate Monte Carlo standard error is

$$
MCSE(\widehat q)=\sqrt{\frac{\widehat q(1-\widehat q)}{B}}.
$$

```{r}
#| label: ch06-show-mcse
#| eval: false
1309 / 2000
sqrt(.6545 * (1 - .6545) / 2000)
```


For the estimated power .6545, it is $\sqrt{.6545(1-.6545)/2000}=0.0106$,
and the table's MCSE columns apply the same formula to each rate. This
value describes the numerical uncertainty from using 2,000 simulations:
a fresh set of 2,000 simulated studies can give a somewhat different
power estimate. The MCSE is distinct from a student outcome's SD or the SE
of a group mean or contrast within one study. A rate very close to one,
such as .999, also deserves care: only two of the 2,000 studies failed
to reject, so its tiny plug-in MCSE rests on very few non-rejections.

Increasing study size $n$ changes the information within a study,
its error degrees of freedom and its ability to detect the specified
difference. Increasing the number of repetitions $B$ makes the simulated
rate more precise; it leaves the study being simulated unchanged.
Simulation describes a procedure under the declared model; agreement with
its expected behaviour does not establish that the model fits a
particular real study.

**Change the spread, keeping the comparison.** Before running another
set of studies, predict what happens if the common SD doubles from
1.8257 to 3.6515 points, while the population means, four students per
group, independence, Normality and .05 test rule stay the same. Under the
null, both the between-group and within-group variation rise together;
the F reference remains appropriate. Under the alternative, the mean
differences are smaller relative to the noise, so we expect less power.
Continuing the same simulation sequence gives a null rejection rate of
.0525, with MCSE 0.005, and power .2165, with MCSE 0.0092. Compare the
latter with .6545 under the original SD. More repetitions would estimate
this lower power more precisely; they would not restore the information
lost through greater within-group variation. The change preserves the
equal-variance condition. It compares two hypothetical population
spreads rather than establishing which one describes real students.

### Planning power and precision

The omnibus simulation asks whether some population group means differ.
A future study may instead be planned to estimate or detect one particular
difference. Consider a future balanced one-factor study of the three
feedback conditions, with a declared **two-comparison Bonferroni family**:
Answer feedback versus Rereading, and Explanatory feedback versus
Rereading. The family has the same form as Section 2's two comparisons
with Silence. Keep it, with family significance level .05, even when
calculating power for just one of its comparisons.

Suppose the comparison of interest has a population difference of
$\delta=2$ points. Plan equal group sizes $n$, independent Normal outcomes
and common SD $\sigma=\sqrt{10/3}$, the square root of the factorial fit's
residual mean square in Section 3. The planning SE for a difference of two
group means is $\sigma\sqrt{2/n}$. All three groups contribute to the
pooled error variance, so its degrees of freedom are $3(n-1)$, even though
the contrast compares only two of the means.

Bonferroni assigns two-sided significance level $.05/2=.025$ to each
comparison. Dividing that amount between its two tails puts .0125 in each
tail. The rejection cutoffs are therefore $-c$ and $c$, where

$$
c=t_{1-.05/(2\times2),\,3(n-1)}=t_{.9875,\,3(n-1)}.
$$

Under the zero-difference null, the test statistic has the central
$t$ distribution already used for the comparison. Under a specified
nonzero difference, its distribution moves towards that difference.
The **noncentral $t$ distribution** describes this statistic when its
numerator has a nonzero mean and its denominator still estimates the
common spread. Its noncentrality parameter expresses the population
difference in planning SE units:

$$
\lambda=\frac{\delta}{\sigma\sqrt{2/n}}.
$$

For this alternative, power is the area beyond the same two rejection
cutoffs: $P(T<-c)+P(T>c)$ for a noncentral $t$ variable with $3(n-1)$
degrees of freedom and noncentrality $\lambda$. These are probabilities
over future studies, rather than a p-value from observed data. Both tails
belong to the two-sided procedure, even when the planning difference is
positive.

\Needspace{10\baselineskip}

```{r}
#| label: ch06-contrast-power
#| include: false
contrast_power <- function(n, delta = 2, sigma = sqrt(10/3)) {
  df <- 3*(n-1)
  critical <- qt(1 - .05/(2*2), df)
  ncp <- delta / (sigma * sqrt(2/n))
  pt(-critical, df, ncp = ncp) +
    pt(critical, df, ncp = ncp, lower.tail = FALSE)
}
round(c(n16 = contrast_power(16), n17 = contrast_power(17)), 4)
```

```{r}
#| label: ch06-show-power
#| eval: false
df <- 3 * (17 - 1)
critical <- qt(1 - .05/4, df)
ncp <- 2 / sqrt((10/3) * 2/17)
pt(-critical, df, ncp) + pt(critical, df, ncp, lower.tail = FALSE)
```


At $n=17$, the planning SE is 0.6262, the error
degrees of freedom are 48, the critical value is 2.3139, and
$\lambda=3.1937$. Summing the two tail areas gives power
.8073. Checking successive integer sizes gives 17 students per group,
51 in total, as the first size reaching .8 power; 16 gives .7793.
Tested alone at two-sided .05, the same comparison would need only 14
per group (power .8068), against 17 inside the family: that is the power
price of the family guarantee.

This is the power of **one specified comparison within the declared
family**. It is neither the probability that both comparisons reject nor
the power of the omnibus F test. For example, with 12 students per group
the specified two-point comparison has power .6328, whereas the omnibus
simulation in @tbl-ch06-simulation gives estimated power .9990 against
the complete pattern $(6,8,10)$. The hypotheses and rejection rules differ,
so these numbers answer different questions.

A precision target concerns the interval's width rather than its
probability of excluding zero. Suppose the aim is a simultaneous interval
whose planned half-width is no more than one point. The family and
degrees of freedom stay the same, giving

$$
\text{planned half-width}
=t_{.9875,\,3(n-1)}\,\sigma\sqrt{\frac{2}{n}}.
$$

```{r}
#| label: ch06-show-width
#| eval: false
n <- c(34, 35)
qt(1 - .05/4, 3*(n-1)) * sqrt((10/3) * 2/n)
```


This calculation substitutes the planning SD $\sigma$ for the error SD
that the future study will estimate. The planned half-width is 1.0078
points at 34 per group and 0.9929 at 35, so 35 per group, or 105 students
in total, is the first size meeting this target. The realised interval
uses the observed $\sqrt{MS_W}$ and therefore has a random width.
Choosing 35 does not guarantee that its half-width will be at most one in
every study. @fig-ch06-planning shows why a power target and an
interval-width target can select quite different sample sizes.

```{r}
#| label: fig-ch06-planning
#| echo: false
#| fig-width: 6.4
#| fig-height: 2.7
#| fig-cap: "Power and planned interval width impose different sample-size targets."
#| fig-alt: "For the same two-comparison three-group procedure, power rises above .8 at 17 students per group. Planned simultaneous interval half-width falls below one point at 35 students per group."
n_grid <- 4:50
powers <- sapply(n_grid, contrast_power)
half_width <- qt(1-.05/(2*2), 3*(n_grid-1)) *
  sqrt(2*(10/3)/n_grid)
panel <- c("Power", "Planned half-width (points)")
planning <- data.frame(n = rep(n_grid, 2),
  value = c(powers, half_width),
  target = factor(rep(panel, each = length(n_grid)), levels = panel))
targets <- data.frame(target = factor(panel, levels = panel),
  value = c(.8, 1), n = c(17, 35))
ggplot(planning, aes(n, value)) + geom_line() +
  geom_hline(aes(yintercept = value), data = targets, linetype = 2) +
  geom_vline(aes(xintercept = n), data = targets, linetype = 2) +
  facet_wrap(~ target, scales = "free_y") +
  labs(x = "Students per group", y = NULL) +
  theme_minimal(base_size = 11)
```

The table `planning` contains the proposed group size `n`, each calculated
`value` and its `target` label. The same plotting structure gives

```{r}
#| label: ch06-show-planning
#| eval: false
ggplot(planning, aes(n, value)) + geom_line() +
  facet_wrap(~ target, scales = "free_y")
```

Finally, distinguish the desired number of available outcomes from the
number recruited. If outcomes are expected to be available for 90% of
recruits, the 17-per-group power target needs
$\lceil17/.9\rceil=19$ recruits per group, 57 in total, with an expected
17.1 available outcomes per group under that retention assumption. It is
an average allowance, not a guarantee of 17 outcomes in every group or a
correction for bias from systematically missing outcomes.

A two-group power calculation
declares a different study, with different error degrees of freedom, and
cannot silently replace this three-group calculation. Choose a planning
difference from the substantive question and relevant prior evidence.
Selecting statistically
significant results can favour exaggerated estimates; an observed
difference from such a selection is not automatically a suitable planning
alternative.

```{r}
#| label: ch06-anchor-check
#| include: false
local({
  near <- function(x, expected, tolerance = 5e-5) {
    stopifnot(length(x) == length(expected),
      all(abs(as.numeric(x) - expected) < tolerance))
  }
  one <- lm(score ~ 1, data = sound_data)
  group <- lm(score ~ sound, data = sound_data)
  near(c(deviance(one), df.residual(one), deviance(group),
    df.residual(group), anova(one, group)$F[2]), c(258,29,178,27,6.0674))
  se <- sqrt((178/27)*2/10)
  mult <- c(qt(.975,27), qt(1-.05/4,27),
    qtukey(.95,3,27)/sqrt(2), sqrt(2*qf(.95,2,27)))
  near(mult, c(2.0518,2.3734,2.4794,2.5900))
  near(-2-mult*se, c(-4.3561,-4.7253,-4.8470,-4.9740))
  near(-2+mult*se, c(.3561,.7253,.8470,.9740))
  fits <- list(lm(score~1,feedback_data), lm(score~feedback,feedback_data),
    lm(score~feedback+practice,feedback_data),
    lm(score~feedback*practice,feedback_data))
  near(sapply(fits,deviance),c(476,220,124,60))
  near(sapply(fits,df.residual),c(23,21,20,18))
  near(anova(fits[[3]],fits[[4]])$F[2],9.6)
  fewer <- subset(feedback_data,
    !(feedback=="Explanatory feedback" & score>=19))
  near(anova(lm(score~feedback*practice,fewer))$`Sum Sq`,
    c(139.7576,52.0833,36.25,50.5))
  near(anova(lm(score~practice*feedback,fewer))$`Sum Sq`,
    c(34.0909,157.75,36.25,50.5))
  near(anova(lm(score~feedback+practice,fewer),
    lm(score~feedback*practice,fewer))$F[2],5.7426)
  near(4+c(-1,1)*qt(.975,18)*sqrt(10/3),c(.1643,7.8357))
  near(4+c(-1,1)*sqrt(2*qf(.95,2,18))*sqrt(10/3),c(-.8680,8.8680))
  alone <- function(n) {
    df <- 3*(n-1); critical <- qt(.975,df); ncp <- 2/sqrt((10/3)*2/n)
    pt(-critical,df,ncp=ncp)+pt(critical,df,ncp=ncp,lower.tail=FALSE)
  }
  near(alone(14),.8068)
  stopifnot(alone(13)<.8, contrast_power(16)<.8, contrast_power(17)>.8)
  near(prop.test(c(32,18),c(50,50),correct=FALSE)$conf.int,
    c(.0918,.4682))
  near(-dbinom(32,50,.64,log=TRUE),2.1465)
})
```

## Reading Bridge {#ch06-reading-bridge}

Each section used one framework, which later chapters carry further:

- **A model for the records**, with parameters to read: the group means
  and indicators (Section 1), two-way models (Section 3), binomial model
  (Section 4.1) and declared generating model (Section 6.1).
- **A fitting criterion**: least squares for means (Sections 1.2 and 5.1)
  and likelihood for a probability (Section 5.2).
- **A comparison of named models** on the same records: one mean against
  three (Section 1.3), additive and full models and the sequence behind
  a two-way table (Section 3.2), and each simulated omnibus test
  (Section 6.1).
- **A comparison family and an effect scale**: pointwise, Bonferroni,
  Tukey and Scheffé intervals (Sections 2 and 3.3), and four categorical
  scales (Section 4.3).
- **Behaviour under repetition**: rejection rates, power and planned
  width under stated generating models (Section 6).

When reading a study, identify the question, analysed records, model,
estimate and uncertainty procedure before judging their adequacy. Keep a
comparison's direction, history and family beside its interval; a table's
event and effect scale beside its estimate; and generating assumptions
and the specified procedure beside any simulation or planning result.

[]{#ch06-appendix-b}Chapter 7 carries the framework from group means to a
line: squared loss for an intercept and slope, the normal equations and
projection, and uncertainty for slope and fitted mean. Chapter 8 carries
it to several predictors, named controls and the overall F as a comparison
of models. Chapter 9 generalizes indicators to factors with several levels,
gives the fuller nested-model F with block tests, and develops analysis of
covariance in Section 2, using Section 3.2's comparison for parallel slopes.
Chapter 11 develops binary-regression likelihood, exact boundary estimates,
multiple-parameter fitting, starting values and convergence checks.

\Needspace{10\baselineskip}

### Key terms and notation {.unnumbered}

The following concepts connect the chapter's models, fitting criteria
and comparisons.

| Term and common alternatives | Meaning |
|:--|:--|
| Model formula (R formula) | Notation such as `score ~ group` naming an outcome and the variables that explain it for a fitting function. It does not state whether observations are independent or paired; that comes from the design. |
| Categorical effect scales (difference in proportions; risk ratio; odds ratio; Cramér's V) | Four ways to state the size of an association in a table, each answering its own question: the difference in proportions gives an absolute gap, the risk ratio a relative gap, the odds ratio a ratio of odds that stays the same when rows and columns exchange roles, and Cramér's V a unit-free strength, from 0 to 1, for a table of any size. |
| Objective function | A numerical criterion evaluated at candidate parameter values for fixed data; fitting seeks a stated minimum or maximum. |
| Likelihood and log-likelihood | The probability or density of the observed data under a declared model, viewed as a function of its parameters; the log-likelihood is its logarithm. |
| Maximum likelihood estimation (MLE) | Choosing parameter values that maximize the likelihood, equivalently the log-likelihood, over the stated parameter space. |

: Selected concepts introduced or developed here. {#tbl-ch06-terms tbl-colwidths="[30,70]"}

The notation follows the local question. $C=\sum_jc_j\mu_j$ is a contrast,
whereas $L(\theta)$ is a likelihood function of a parameter. Similarly, Type II error probability
$\beta$ differs from a model coefficient such as $\beta_x$.
Observed Pearson $X^2$ differs from the reference distribution $\chi^2$,
and $\Delta p=p_1-p_0$ is a difference in proportions.
Section 6.1 distinguishes study size $n$ from repetitions $B$, and an
estimator's SE from finite-simulation MCSE.

### Further reading {.unnumbered}

For the linear-model basis of ANOVA, contrasts and nested-model
comparisons, see @kutner2005. @tukey1953 develops simultaneous comparison
ideas, and @wilson1927 provides the score-interval construction for a
proportion. @agresti2018 develops the effect scales for two-way tables.
@cohen1988 discusses power and the dependence of a planning result on its
chosen effect and procedure, and @cetinkaya2024 develops the
simulation-based treatment of operating characteristics used in Section 6.
R's documentation gives the arguments and returned components for `aov`,
`lm`, `TukeyHSD`, `oneway.test`, `prop.test`, `chisq.test` and `optimize`
[@rcore2025]. The plotting chapters of @wickham2023 develop the data,
mapping and layer vocabulary behind the chapter's figures.

## Exercises {.unnumbered}

The chapter's printed output, tables and figures supply the values each
exercise reads, with a calculator for the few calculations that the
reasoning needs. State the records, working conditions and comparison
direction in each interpretation.

<!-- prompt-id: CH06-E01 -->

### Exercise 1: Read the one-way fit from its output {.unnumbered #ex-ch06-01}

For the thirty background-sound scores, `anova(one_fit)` reports sums of
squares 80 and 178 on 2 and 27 df, $F=6.0674$ and $p=.0067$, and
`coef(indicator_fit)` returns 12, $-2$ and $-4$ with Silence as the
reference level.

1. Say what each sum of squares measures in these scores, and why
   Instrumental music, whose mean equals the grand mean of 10, adds
   nothing to the between-group sum of squares although its ten scores
   vary.
2. Explain what the within-group df, 27, count, and why $F$ divides by
   the residual mean square, 6.5926, rather than by the variance of all
   thirty scores, $258/29=8.8966$.
3. State the equal-means null, and interpret $p=.0067$ under the working
   model, including what the result leaves unanswered about particular
   pairs of conditions.
4. Predict the three coefficients if Instrumental music were made the
   reference level, and say which parts of the ANOVA output would change.

<!-- END EX06-01 PROMPT -->

<!-- prompt-id: CH06-E02 -->

### Exercise 2: Keep the comparison and its family together {.unnumbered #ex-ch06-02}

Instrumental music minus Silence is $-2$ points with SE 1.1483 on 27 df.
Its 95% interval is $[-4.3561, 0.3561]$ when the comparison stands alone,
$[-4.7253, 0.7253]$ inside the declared family of the two comparisons
with Silence (Bonferroni) and $[-4.8470, 0.8470]$ inside the family of
all three pairs (Tukey).

1. The estimate and SE are the same in all three intervals. Explain why
   the limits differ, naming the family that each interval protects.
2. Explain why the SE uses 27 residual df, although the comparison
   involves only the twenty students in two conditions.
3. An analyst reports Music with lyrics minus Silence, $-4$ points,
   because it is the largest of the three observed differences. Say which
   family its interval should come from, and why.
4. The omnibus F rejected equal means, yet all three intervals above
   include zero. Explain why these results are compatible, and why an
   interval that includes zero does not show that the two means are
   equal.

<!-- END EX06-02 PROMPT -->

<!-- prompt-id: CH06-E03 -->

### Exercise 3: Read interaction as a change in comparison {.unnumbered #ex-ch06-03}

In the 24-student feedback-by-practice study, the Massed and Spaced cell
means are 6 and 6 for Rereading, 8 and 12 for Answer feedback and 10 and
18 for Explanatory feedback, and the full two-way model gives the
interaction test $F(2,18)=9.6$, $p=.0015$.

1. Give the three Spaced-minus-Massed changes, and say what the
   nonparallel lines of the interaction plot show and what they cannot
   show without a model.
2. The practice margins are 8 and 12. Explain why the 4-point practice
   main effect needs the interaction beside it when a reader asks how
   spaced practice works.
3. The study plan declared one difference of differences in advance:
   Answer feedback's change minus Rereading's change, 4 points with SE
   1.8257 and pointwise 95% interval $[0.1643, 7.8357]$. Had the same
   contrast been picked after inspecting the plot, its Scheffé interval
   over the two-dimensional interaction family, $[-0.8680, 8.8680]$,
   would apply. Explain how that interval can include zero while the
   interaction F test rejects, and why the declared history warrants the
   pointwise interval.
4. Answer feedback's simple practice effect is also 4 points, with SE
   1.291. Explain why its SE is smaller and why the interaction family's
   guarantee does not cover it.

<!-- END EX06-03 PROMPT -->

<!-- prompt-id: CH06-E04 -->

### Exercise 4: Name the two models behind an interaction test {.unnumbered #ex-ch06-04}

The interaction test for the 24 feedback/practice scores compares two
models fitted to the same students: the additive model, with residual sum
of squares 124 on 20 df, and the full factorial model, with 60 on 18 df.

1. Name the two models and say what the larger one adds.
2. The comparison gives $F=9.6$ on 2 and 18 df. Explain why its
   denominator comes from the full model rather than the additive model.
3. After the two Explanatory-Spaced students scoring 19 and 20 are
   removed, the sequential sums of squares for feedback and practice
   change with the order of the formula, yet the comparison of the
   additive and full models gives $F(2,16)=5.7426$ in either order.
   Explain why.
4. Say what a nonsignificant result of this comparison would and would
   not show about additivity.

<!-- END EX06-04 PROMPT -->

<!-- prompt-id: CH06-E05 -->

### Exercise 5: Preserve the event, denominator and expected counts {.unnumbered #ex-ch06-05}

In the invented table, Online has 32 completions and 18 non-completions;
In person has 18 completions and 32 non-completions. For Online, the
uncorrected score test of $p=.5$ gives $X^2=3.92$ and $p=.0477$, with the
95% Wilson interval $[.5014, .7586]$. For the complete table, Pearson's
test gives $X^2=7.84$ on 1 df and $p=.0051$.

1. State the null and the denominator of each test, and explain why the
   two p-values answer different questions.
2. Calculate the expected count of Online completions under
   independence, and explain why the chi-squared approximation is checked
   against expected rather than observed counts.
3. The difference in proportions is .28, the risk ratio 1.7778 and the
   odds ratio 3.1605. Explain why the odds ratio lies farther from 1 than
   the risk ratio here. Cramér's V also equals .28; explain why their
   equality here does not make the measures interchangeable in general.
4. Say what design information a claim that the online format causes
   completion would additionally need.

<!-- END EX06-05 PROMPT -->

<!-- prompt-id: CH06-E06 -->

### Exercise 6: Distinguish study size, repetitions and planning targets {.unnumbered #ex-ch06-06}

In the chapter's simulation, each condition repeats 2,000 generated
studies of three Normal groups with common SD $\sqrt{10/3}$ and applies
the .05 omnibus ANOVA. With equal means $(8,8,8)$ the rejection rates are
.056 at four students per group and .050 at twelve; with means 6, 8 and
10 the estimated power is .6545 at four and .9990 at twelve.

1. Say what the rejection rates under equal means estimate, and explain
   why the power rises from four to twelve students per group while those
   rates stay near .05.
2. The Monte Carlo standard error (MCSE) of the power estimate .6545
   with $B=2000$ is 0.0106. Say what it measures, and why it differs
   from the SE of a contrast within one simulated study.
3. Predict what quadrupling $B$ to 8,000 would do to this MCSE and to the
   power that the simulation estimates.
4. For the declared two-comparison family, 17 students per group first
   reach power .8 for a 2-point comparison, and 35 per group first reach
   a planned simultaneous half-width of at most one point. Say what each
   target controls, and why a realised interval can have a different
   width from its planning value.

<!-- END EX06-06 PROMPT -->

<!-- prompt-id: CH06-E07 -->

### Exercise 7: Distinguish an estimate from its fitting criterion {.unnumbered #ex-ch06-07}

Keep 32 successes in 50 trials fixed. Minimizing the negative binomial
log-likelihood with `optimize()` over $[0,1]$ returns `minimum` .64 and
`objective` 2.1465; at $p=.5$ the negative log-likelihood is 4.1330.

1. Say which returned value is the parameter estimate and which is the
   criterion value, and why 2.1465 is neither a probability that $p=.64$
   nor an SE.
2. The derivative of the log-likelihood is
   $\ell'(p)=(32-50p)/\{p(1-p)\}$. Use its sign to explain why the
   maximum lies at $32/50$.
3. Before calculating, predict how the fitted probability and the fall in
   log-likelihood from $p=.64$ to $p=.5$ would change with 64 successes in
   100 trials. Then compare your prediction with the supplied falls,
   1.9864 for 32 of 50 and 3.9729 for 64 of 100.
4. Explain why the likelihood curve is not a probability distribution
   over $p$, and what an interior numerical search would return if all
   50 trials had been successes.

<!-- END EX06-07 PROMPT -->

<!-- prompt-id: CH06-E08 -->

### Exercise 8: Write a qualified result paragraph {.unnumbered #ex-ch06-08}

Choose the one-way comparisons, the declared interaction contrast or the
two-group count comparison from this chapter. Write a short paragraph giving the
question, data and denominator, estimate, interval and its family or
pointwise status, working conditions and scope. Identify one tempting claim
that the numerical result does not establish, and explain the missing support.

<!-- END EX06-08 PROMPT -->

## References {.unnumbered}
