---
title: "Interaction and Moderation"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - ../koma_heading_tags.tex
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{16\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AtBeginDocument{\renewcommand{\boldsymbol}[1]{\symbf{#1}}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
RNGkind("Mersenne-Twister", "Inversion", "Rejection")

required_packages <- c(
  "ggplot2", "dplyr", "carData", "knitr", "emmeans"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install the required package(s) before rendering: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(ggplot2)
library(dplyr)
library(carData)
library(emmeans)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 7, width = 68)
theme_set(theme_minimal(base_size = 12))

format_p <- function(p) {
  ifelse(p < .0001, "< .0001", format(round(p, 4), trim = TRUE))
}

hc3_vcov <- function(fit) {
  X <- model.matrix(fit)
  e <- resid(fit)
  h <- hatvalues(fit)
  bread <- solve(crossprod(X))
  meat <- crossprod(X, X * (e / (1 - h))^2)
  bread %*% meat %*% bread
}

# Finite-sample sensitivity calculation: the HC3 Wald statistic is divided by
# the number of restrictions and compared with an F reference distribution.
# This reference calibration is approximate, not an exact robust F test.
joint_wald_f <- function(fit, terms, vcov_mat = hc3_vcov(fit)) {
  b <- coef(fit)
  missing_terms <- setdiff(terms, names(b))
  if (length(missing_terms)) {
    stop("Coefficient(s) not found: ", paste(missing_terms, collapse = ", "))
  }
  R <- matrix(0, nrow = length(terms), ncol = length(b),
              dimnames = list(terms, names(b)))
  R[cbind(seq_along(terms), match(terms, names(b)))] <- 1
  delta <- R %*% b
  wald <- drop(t(delta) %*% solve(R %*% vcov_mat %*% t(R)) %*% delta)
  df1 <- nrow(R)
  df2 <- df.residual(fit)
  F_value <- wald / df1
  data.frame(F = F_value, numerator_df = df1, denominator_df = df2,
             p = pf(F_value, df1, df2, lower.tail = FALSE))
}

set.seed(909L)

n_sim <- 320L
base <- pmin(
  pmax(rnorm(n_sim, mean = 70, sd = 10), 42),
  96
)
hours <- runif(n_sim, min = 0, max = 8)
baseline_c <- base - 70

score <-
  60 +
  2.5 * hours +
  0.40 * baseline_c +
  0.12 * hours * baseline_c +
  rnorm(n_sim, mean = 0, sd = 5)

sim <- data.frame(
  score,
  hours,
  base,
  baseline_c
)

sim_audit <- data.frame(
  item = c(
    "Rows", "Complete rows", "Practice minimum", "Practice maximum",
    "Baseline minimum", "Baseline maximum"
  ),
  value = c(
    nrow(sim),
    sum(complete.cases(sim)),
    min(sim$hours),
    max(sim$hours),
    min(sim$base),
    max(sim$base)
  )
)

knitr::kable(sim_audit, digits = 2, row.names = FALSE)

m0 <- lm(
  score ~ hours + baseline_c,
  data = sim
)

m1 <- lm(
  score ~ hours * baseline_c,
  data = sim
)

sim_coef <- coef(summary(m1))
sim_ci <- confint(m1)
sim_truth <- c(60, 2.5, 0.40, 0.12)

sim_recovery <- data.frame(
  term = c(
    "Intercept", "Practice", "Centered baseline",
    "Practice x centered baseline"
  ),
  truth = sim_truth,
  estimate = coef(m1),
  SE = sim_coef[, "Std. Error"],
  lower = sim_ci[, 1],
  upper = sim_ci[, 2],
  row.names = NULL
)

knitr::kable(
  sim_recovery,
  digits = 3,
  col.names = c(
    "Term", "Truth", "Estimate", "SE", "95% lower", "95% upper"
  ),
  row.names = FALSE
)

conditional_practice_slope <- function(fit, baseline_c_value) {
  b <- coef(fit)
  V <- vcov(fit)
  L <- setNames(rep(0, length(b)), names(b))
  L["hours"] <- 1
  L["hours:baseline_c"] <- baseline_c_value

  estimate <- sum(L * b)
  SE <- sqrt(drop(t(L) %*% V %*% L))
  tcrit <- qt(0.975, df.residual(fit))

  data.frame(
    base = baseline_c_value + 70,
    estimate = estimate,
    SE = SE,
    lower = estimate - tcrit * SE,
    upper = estimate + tcrit * SE
  )
}

sim_slope_table <- bind_rows(lapply(
  c(-10, 0, 10),
  conditional_practice_slope,
  fit = m1
))

knitr::kable(
  sim_slope_table,
  digits = 3,
  col.names = c(
    "Baseline score", "Practice slope", "SE", "95% lower", "95% upper"
  ),
  row.names = FALSE
)

sim_trends <- emtrends(
  m1,
  specs = ~ baseline_c,
  var = "hours",
  at = list(baseline_c = c(-10, 0, 10))
)

sim_trends_df <- as.data.frame(confint(sim_trends)) |>
  transmute(
    base = baseline_c + 70,
    practice_slope = hours.trend,
    SE = SE,
    lower = lower.CL,
    upper = upper.CL
  )

knitr::kable(
  sim_trends_df,
  digits = 3,
  col.names = c(
    "Baseline score", "Practice slope", "SE", "95% lower", "95% upper"
  ),
  row.names = FALSE
)

# Both models must use the same rows; the named check is displayed below.
sim_row_check <- c(
  rows_additive = nobs(m0),
  rows_interaction = nobs(m1)
)

cat(sprintf(
  "Same-row check: additive model = %d; interaction model = %d.\n",
  sim_row_check["rows_additive"],
  sim_row_check["rows_interaction"]
))

sim_block_test <- anova(m0, m1)
sim_block_display <- data.frame(
  comparison = "Add practice x baseline product",
  numerator_df = sim_block_test$Df[2],
  denominator_df = sim_block_test$Res.Df[2],
  extra_SS = sim_block_test$`Sum of Sq`[2],
  F = sim_block_test$F[2],
  p = format_p(sim_block_test$`Pr(>F)`[2])
)

knitr::kable(
  sim_block_display,
  digits = 2,
  col.names = c(
    "Nested comparison", "Numerator df", "Denominator df",
    "Extra SS", "$F$", "$p$"
  ),
  caption = "Nested comparison of the additive and product-term simulation models.",
  escape = FALSE,
  row.names = FALSE
)

fit_sim_raw <- lm(
  score ~ hours * base,
  data = sim
)

centering_check <- data.frame(
  quantity = c(
    "Practice slope at coded zero",
    "Baseline lower-order slope",
    "Product coefficient",
    "Maximum fitted-value difference"
  ),
  raw_score_model = c(
    coef(fit_sim_raw)["hours"],
    coef(fit_sim_raw)["base"],
    coef(fit_sim_raw)["hours:base"],
    max(abs(fitted(fit_sim_raw) - fitted(m1)))
  ),
  centered_model = c(
    coef(m1)["hours"],
    coef(m1)["baseline_c"],
    coef(m1)["hours:baseline_c"],
    0
  )
)

knitr::kable(
  centering_check,
  digits = 6,
  col.names = c("Quantity", "Raw-score model", "Centered model"),
  row.names = FALSE
)

sim_new <- data.frame(
  hours = 4,
  baseline_c = c(-10, 0, 10)
)

sim_mean_interval <- predict(
  m1,
  newdata = sim_new,
  interval = "confidence",
  level = 0.95
)
sim_case_interval <- predict(
  m1,
  newdata = sim_new,
  interval = "prediction",
  level = 0.95
)

sim_prediction_table <- data.frame(
  base = sim_new$baseline_c + 70,
  hours = sim_new$hours,
  fitted_mean = sim_mean_interval[, "fit"],
  mean_lower = sim_mean_interval[, "lwr"],
  mean_upper = sim_mean_interval[, "upr"],
  case_lower = sim_case_interval[, "lwr"],
  case_upper = sim_case_interval[, "upr"]
)

knitr::kable(
  sim_prediction_table,
  digits = 2,
  col.names = c(
    "Baseline", "Practice", "Fitted mean", "Mean lower", "Mean upper",
    "Case lower", "Case upper"
  ),
  row.names = FALSE
)

b_jn <- coef(m1)
V_jn <- vcov(m1)
t_jn <- qt(0.975, df.residual(m1))

x_name <- "hours"
xw_name <- "hours:baseline_c"

b_x <- unname(b_jn[x_name])
b_xw <- unname(b_jn[xw_name])
v_x <- V_jn[x_name, x_name]
v_xw <- V_jn[xw_name, xw_name]
cov_x_xw <- V_jn[x_name, xw_name]

jn_A <- b_xw^2 - t_jn^2 * v_xw
jn_B <- 2 * b_x * b_xw - 2 * t_jn^2 * cov_x_xw
jn_C <- b_x^2 - t_jn^2 * v_x

jn_roots_centered <- polyroot(c(jn_C, jn_B, jn_A))
jn_roots_centered <- sort(Re(
  jn_roots_centered[abs(Im(jn_roots_centered)) < 1e-8]
))
jn_boundaries <- jn_roots_centered + 70

jn_results <- data.frame(
  boundary = c("Lower", "Upper"),
  base = jn_boundaries,
  slope_at_boundary = b_x + jn_roots_centered * b_xw,
  within_observed_range =
    jn_boundaries >= min(sim$base) &
      jn_boundaries <= max(sim$base),
  students_beyond = c(
    sum(sim$base < jn_boundaries[1]),
    sum(sim$base > jn_boundaries[2])
  )
)

knitr::kable(
  jn_results,
  digits = 2,
  col.names = c(
    "Boundary", "Baseline score", "Slope at boundary",
    "Within observed range", "Students beyond it"
  ),
  caption = paste0(
    "Johnson-Neyman boundaries for the fitted practice slope, the ",
    "slope at each boundary, and the number of simulated students ",
    "beyond it."
  ),
  row.names = FALSE
)

d0 <- carData::Salaries

dat <- d0 |>
  select(salary, yrs.service, rank) |>
  filter(complete.cases(salary, yrs.service, rank)) |>
  mutate(
    rank = droplevels(rank),
    salary_k = salary / 1000,
    service_c = yrs.service - 5
  )

salary_audit <- data.frame(
  item = c(
    "Package rows",
    "Rows missing salary",
    "Rows missing years of service",
    "Rows missing rank",
    "Complete analysis rows",
    "Rows excluded by complete-case rule"
  ),
  value = c(
    nrow(d0),
    sum(is.na(d0$salary)),
    sum(is.na(d0$yrs.service)),
    sum(is.na(d0$rank)),
    nrow(dat),
    nrow(d0) - nrow(dat)
  )
)

knitr::kable(salary_audit, row.names = FALSE)

salary_support <- dat |>
  group_by(rank) |>
  summarise(
    n = n(),
    minimum = min(yrs.service),
    first_quartile = quantile(yrs.service, 0.25),
    median = median(yrs.service),
    third_quartile = quantile(yrs.service, 0.75),
    maximum = max(yrs.service),
    mean_salary = mean(salary_k),
    .groups = "drop"
  )

common_service_range <- c(
  max(salary_support$minimum),
  min(salary_support$maximum)
)

knitr::kable(
  salary_support,
  digits = 1,
  col.names = c(
    "Rank", "$n$", "Minimum", "First quartile", "Median",
    "Third quartile", "Maximum", "Mean salary (\\$1000)"
  ),
  escape = FALSE,
  row.names = FALSE
)

s0 <- lm(
  salary_k ~ service_c + rank,
  data = dat
)

s1 <- lm(
  salary_k ~ service_c * rank,
  data = dat
)

salary_coef <- coef(summary(s1))
salary_ci <- confint(s1)

salary_coef_table <- data.frame(
  term = c(
    "Assistant mean at 5 years",
    "Assistant slope",
    "Associate - Assistant at 5 years",
    "Professor - Assistant at 5 years",
    "Associate - Assistant slope",
    "Professor - Assistant slope"
  ),
  estimate = coef(s1),
  SE = salary_coef[, "Std. Error"],
  t = salary_coef[, "t value"],
  p = format_p(salary_coef[, "Pr(>|t|)"]),
  lower = salary_ci[, 1],
  upper = salary_ci[, 2],
  row.names = NULL
)

knitr::kable(
  salary_coef_table,
  digits = 2,
  col.names = c(
    "Conditional term", "Estimate", "SE", "$t$", "$p$",
    "95% lower", "95% upper"
  ),
  escape = FALSE,
  row.names = FALSE
)

# Both models must use the same rows; the fixed sample makes the
# counts easy to confirm by displaying them.
salary_row_check <- c(
  rows_additive = nobs(s0),
  rows_interaction = nobs(s1)
)

cat(sprintf(
  "Same-row check: additive model = %d; interaction model = %d.\n",
  salary_row_check["rows_additive"],
  salary_row_check["rows_interaction"]
))

salary_block <- anova(
  s0,
  s1
)
salary_block_p <- salary_block[["Pr(>F)"]][2]

salary_block_hc3 <- joint_wald_f(
  s1,
  c("service_c:rankAssocProf", "service_c:rankProf")
)

salary_block_display <- bind_rows(
  data.frame(
    procedure = "Classical nested-model F",
    numerator_df = salary_block$Df[2],
    denominator_df = salary_block$Res.Df[2],
    F = salary_block$F[2],
    p = format_p(salary_block_p)
  ),
  data.frame(
    procedure = "Approximate HC3 joint Wald F",
    numerator_df = salary_block_hc3$numerator_df,
    denominator_df = salary_block_hc3$denominator_df,
    F = salary_block_hc3$F,
    p = format_p(salary_block_hc3$p)
  )
)

knitr::kable(
  salary_block_display,
  digits = 2,
  col.names = c(
    "Variance procedure", "Numerator df", "Denominator df", "$F$", "$p$"
  ),
  caption = "Classical and approximate HC3 tests of the two-column salary product block.",
  escape = FALSE,
  row.names = FALSE
)

salary_trends <- emtrends(
  s1,
  specs = ~ rank,
  var = "service_c"
)

salary_trends_df <- as.data.frame(confint(salary_trends)) |>
  transmute(
    rank = rank,
    slope_k_per_year = service_c.trend,
    SE = SE,
    lower = lower.CL,
    upper = upper.CL
  )

knitr::kable(
  salary_trends_df,
  digits = 2,
  col.names = c(
    "Rank", "Slope (\\$1000/yr)", "SE", "95% lower", "95% upper"
  ),
  escape = FALSE,
  row.names = FALSE
)

salary_trend_pair_tests <- as.data.frame(
  pairs(salary_trends, adjust = "tukey")
)
salary_trend_pair_intervals <- as.data.frame(
  confint(pairs(salary_trends, adjust = "tukey"))
)

salary_trend_pairs <- salary_trend_pair_intervals |>
  transmute(
    contrast = contrast,
    slope_difference = estimate,
    SE = SE,
    lower = lower.CL,
    upper = upper.CL,
    t = salary_trend_pair_tests$t.ratio,
    p_adjusted = format_p(salary_trend_pair_tests$p.value)
  )

knitr::kable(
  salary_trend_pairs,
  digits = 2,
  col.names = c(
    "Slope contrast", "Difference (\\$1000/yr)", "SE",
    "95% lower", "95% upper", "$t$", "Tukey-adjusted $p$"
  ),
  escape = FALSE,
  row.names = FALSE
)

salary_means_5 <- emmeans(
  s1,
  specs = ~ rank,
  at = list(service_c = 0)
)

salary_means_5_df <- as.data.frame(confint(salary_means_5)) |>
  transmute(
    rank = rank,
    fitted_mean_k = emmean,
    SE = SE,
    lower = lower.CL,
    upper = upper.CL
  )

knitr::kable(
  salary_means_5_df,
  digits = 1,
  col.names = c(
    "Rank", "Fitted mean (\\$1000)", "SE", "95% lower", "95% upper"
  ),
  escape = FALSE,
  row.names = FALSE
)

salary_new_5 <- data.frame(
  service_c = 0,
  rank = factor(
    levels(dat$rank),
    levels = levels(dat$rank)
  )
)

salary_ci_5 <- predict(
  s1,
  newdata = salary_new_5,
  interval = "confidence"
)
salary_pi_5 <- predict(
  s1,
  newdata = salary_new_5,
  interval = "prediction"
)

salary_interval_table <- data.frame(
  rank = salary_new_5$rank,
  fitted_mean = salary_ci_5[, "fit"],
  mean_lower = salary_ci_5[, "lwr"],
  mean_upper = salary_ci_5[, "upr"],
  case_lower = salary_pi_5[, "lwr"],
  case_upper = salary_pi_5[, "upr"]
)

knitr::kable(
  salary_interval_table,
  digits = 1,
  col.names = c(
    "Rank", "Fitted mean", "Mean lower", "Mean upper",
    "Case lower", "Case upper"
  ),
  row.names = FALSE
)

salary_prof_ref <- dat |>
  mutate(rank = relevel(rank, ref = "Prof"))

fit_salary_prof_ref <- lm(
  salary_k ~ service_c * rank,
  data = salary_prof_ref
)

reference_check <- data.frame(
  quantity = c(
    "Maximum fitted-value difference",
    "Residual standard deviation",
    "Model R-squared"
  ),
  AsstProf_reference = c(
    0,
    summary(s1)$sigma,
    summary(s1)$r.squared
  ),
  Prof_reference = c(
    max(abs(
      fitted(s1) -
        fitted(fit_salary_prof_ref)
    )),
    summary(fit_salary_prof_ref)$sigma,
    summary(fit_salary_prof_ref)$r.squared
  )
)

knitr::kable(
  reference_check,
  digits = 8,
  col.names = c(
    "Quantity", "Assistant reference", "Professor reference"
  ),
  row.names = FALSE
)

probe_years <- c(2, 3, 5, 6)
b <- coef(s1)
V <- vcov(s1)
tcrit <- qt(0.975, df.residual(s1))

contrast_estimate <- numeric(4)
contrast_se <- numeric(4)
for (j in 1:4) {
  s <- probe_years[j] - 5
  L <- setNames(rep(0, length(b)), names(b))
  L["rankProf"] <- 1
  L["service_c:rankProf"] <- s
  contrast_estimate[j] <- sum(L * b)
  contrast_se[j] <- sqrt(drop(t(L) %*% V %*% L))
}

salary_rank_contrasts <- data.frame(
  yrs.service = probe_years,
  Prof_minus_Asst_k = contrast_estimate,
  SE = contrast_se,
  lower = contrast_estimate - tcrit * contrast_se,
  upper = contrast_estimate + tcrit * contrast_se
)

knitr::kable(
  salary_rank_contrasts,
  digits = 1,
  col.names = c(
    "Years of service", "Prof - Assistant (\\$1000)", "SE",
    "95% lower", "95% upper"
  ),
  escape = FALSE,
  row.names = FALSE
)

moore <- carData::Moore |>
  mutate(
    fcategory = factor(fcategory, levels = c("low", "medium", "high")),
    partner.status = factor(partner.status, levels = c("low", "high"))
  )

moore_cell_means <- moore |>
  group_by(fcategory, partner.status) |>
  summarise(
    n = n(),
    mean_conformity = mean(conformity),
    SD = sd(conformity),
    .groups = "drop"
  )

knitr::kable(
  moore_cell_means,
  digits = 2,
  col.names = c(
    "Authoritarianism", "Partner status", "$n$", "Mean conformity", "SD"
  ),
  escape = FALSE,
  row.names = FALSE
)

fit_moore_additive <- lm(
  conformity ~ partner.status + fcategory,
  data = moore
)
fit_moore_interaction <- lm(
  conformity ~ partner.status * fcategory,
  data = moore
)

moore_block <- anova(fit_moore_additive, fit_moore_interaction)
moore_block_hc3 <- joint_wald_f(
  fit_moore_interaction,
  c(
    "partner.statushigh:fcategorymedium",
    "partner.statushigh:fcategoryhigh"
  )
)
moore_block_display <- bind_rows(
  data.frame(
    procedure = "Classical nested-model F",
    numerator_df = moore_block$Df[2],
    denominator_df = moore_block$Res.Df[2],
    F = moore_block$F[2],
    p = format_p(moore_block$`Pr(>F)`[2])
  ),
  data.frame(
    procedure = "Approximate HC3 joint Wald F",
    numerator_df = moore_block_hc3$numerator_df,
    denominator_df = moore_block_hc3$denominator_df,
    F = moore_block_hc3$F,
    p = format_p(moore_block_hc3$p)
  )
)

knitr::kable(
  moore_block_display,
  digits = 2,
  col.names = c(
    "Variance procedure", "Numerator df", "Denominator df", "$F$", "$p$"
  ),
  escape = FALSE,
  row.names = FALSE
)

moore_means_matrix <- with(
  moore,
  tapply(conformity, list(fcategory, partner.status), mean)
)

moore_did <-
  (moore_means_matrix["low", "high"] -
    moore_means_matrix["low", "low"]) -
  (moore_means_matrix["high", "high"] -
    moore_means_matrix["high", "low"])

data.frame(
  contrast =
    "(Status difference at low F) - (Status difference at high F)",
  estimate = moore_did
) |>
  knitr::kable(digits = 2, row.names = FALSE)

salary_vcov_hc3 <- hc3_vcov(s1)

salary_trends_hc3 <- emtrends(
  s1,
  specs = ~ rank,
  var = "service_c",
  vcov. = salary_vcov_hc3
)

salary_trends_classical_df <- as.data.frame(
  confint(salary_trends)
)
salary_trends_hc3_df <- as.data.frame(
  confint(salary_trends_hc3)
)

salary_hc3_comparison <- data.frame(
  rank = salary_trends_classical_df$rank,
  slope = salary_trends_classical_df$service_c.trend,
  classical_SE = salary_trends_classical_df$SE,
  HC3_SE = salary_trends_hc3_df$SE,
  HC3_lower = salary_trends_hc3_df$lower.CL,
  HC3_upper = salary_trends_hc3_df$upper.CL
)

knitr::kable(
  salary_hc3_comparison,
  digits = 2,
  col.names = c(
    "Rank", "Slope", "Classical SE", "HC3 SE",
    "Approx. HC3 95% lower", "Approx. HC3 95% upper"
  ),
  row.names = FALSE
)

salary_cook <- cooks.distance(s1)
salary_influence_index <- which.max(salary_cook)
salary_influential_case <- dat[salary_influence_index, ]

fit_salary_without_max <- update(
  s1,
  data = dat[-salary_influence_index, ]
)

salary_trends_without_max <- as.data.frame(emtrends(
  fit_salary_without_max,
  specs = ~ rank,
  var = "service_c"
))

influence_comparison <- data.frame(
  rank = salary_trends_df$rank,
  original_slope = salary_trends_df$slope_k_per_year,
  slope_without_max =
    salary_trends_without_max$service_c.trend,
  change = salary_trends_without_max$service_c.trend -
    salary_trends_df$slope_k_per_year
)

knitr::kable(
  influence_comparison,
  digits = 3,
  col.names = c(
    "Rank", "Original slope", "Slope without case", "Change"
  ),
  row.names = FALSE
)

influential_case_record <- data.frame(
  row = salary_influence_index,
  rank = salary_influential_case$rank,
  yrs.service = salary_influential_case$yrs.service,
  salary_k = salary_influential_case$salary_k,
  Cooks_D = max(salary_cook)
)

fit_salary_common_curve <- lm(
  salary_k ~ service_c * rank + I(service_c^2),
  data = dat
)

salary_curve_test <- anova(
  s1,
  fit_salary_common_curve
)

curve_display <- data.frame(
  comparison = "Add common quadratic service term",
  numerator_df = salary_curve_test$Df[2],
  denominator_df = salary_curve_test$Res.Df[2],
  F = salary_curve_test$F[2],
  p = format_p(salary_curve_test$`Pr(>F)`[2])
)

knitr::kable(
  curve_display,
  digits = 2,
  col.names = c(
    "Sensitivity comparison", "Numerator df", "Denominator df",
    "$F$", "$p$"
  ),
  escape = FALSE,
  row.names = FALSE
)

salary_coef_names <- names(coef(s1))
salary_L <- matrix(
  0,
  nrow = 5,
  ncol = length(salary_coef_names),
  dimnames = list(
    c(
      "AssocProf - AsstProf slope difference",
      "Prof - AsstProf slope difference",
      "AsstProf service slope",
      "AssocProf service slope",
      "Prof service slope"
    ),
    salary_coef_names
  )
)

salary_L[1, "service_c:rankAssocProf"] <- 1
salary_L[2, "service_c:rankProf"] <- 1
salary_L[3, "service_c"] <- 1
salary_L[4, c("service_c", "service_c:rankAssocProf")] <- 1
salary_L[5, c("service_c", "service_c:rankProf")] <- 1

salary_target_estimate <- drop(salary_L %*% coef(s1))
salary_target_classical_se <- sqrt(diag(
  salary_L %*% vcov(s1) %*% t(salary_L)
))
salary_target_hc3_se <- sqrt(diag(
  salary_L %*% salary_vcov_hc3 %*% t(salary_L)
))

set.seed(90909)
bootstrap_repetitions <- 2000
salary_bootstrap_draws <- matrix(
  NA_real_,
  nrow = nrow(salary_L),
  ncol = bootstrap_repetitions,
  dimnames = list(rownames(salary_L), NULL)
)

for (bootstrap_index in seq_len(bootstrap_repetitions)) {
  bootstrap_rows <- sample.int(nrow(dat), replace = TRUE)
  bootstrap_fit <- lm(
    salary_k ~ service_c * rank,
    data = dat[bootstrap_rows, , drop = FALSE]
  )
  bootstrap_coef <- coef(bootstrap_fit)
  if (all(salary_coef_names %in% names(bootstrap_coef)) &&
      !anyNA(bootstrap_coef[salary_coef_names])) {
    salary_bootstrap_draws[, bootstrap_index] <-
      drop(salary_L %*% bootstrap_coef[salary_coef_names])
  }
}

salary_bootstrap_summary <- data.frame(
  target = rownames(salary_L),
  estimate = salary_target_estimate,
  classical_SE = salary_target_classical_se,
  HC3_SE = salary_target_hc3_se,
  bootstrap_SE = apply(salary_bootstrap_draws, 1, sd, na.rm = TRUE),
  bootstrap_lower = apply(
    salary_bootstrap_draws, 1, quantile,
    probs = 0.025, na.rm = TRUE, names = FALSE
  ),
  bootstrap_upper = apply(
    salary_bootstrap_draws, 1, quantile,
    probs = 0.975, na.rm = TRUE, names = FALSE
  ),
  successful_resamples = rowSums(is.finite(salary_bootstrap_draws)),
  row.names = NULL
)

knitr::kable(
  salary_bootstrap_summary,
  digits = c(0, 2, 2, 2, 2, 2, 2, 0),
  col.names = c(
    "Target (thousand dollars per service year)", "Estimate",
    "Classical SE", "HC3 SE", "Bootstrap SE",
    "Bootstrap 2.5%", "Bootstrap 97.5%", "Successful resamples"
  ),
  caption = paste0(
    "Model-based, HC3, and case-bootstrap uncertainty for the two ",
    "rank slope differences and the three rank-specific service ",
    "slopes."
  ),
  row.names = FALSE
)

bootstrap_skewness <- apply(salary_bootstrap_draws, 1, function(draws) {
  draws <- draws[is.finite(draws)]
  mean((draws - mean(draws))^3) / sd(draws)^3
})

interval_comparison <- data.frame(
  target = rownames(salary_L),
  skewness = bootstrap_skewness,
  percentile_lower = salary_bootstrap_summary$bootstrap_lower,
  percentile_upper = salary_bootstrap_summary$bootstrap_upper,
  normal_lower = salary_target_estimate -
    1.96 * salary_bootstrap_summary$bootstrap_SE,
  normal_upper = salary_target_estimate +
    1.96 * salary_bootstrap_summary$bootstrap_SE,
  row.names = NULL
)

knitr::kable(
  interval_comparison,
  digits = 2,
  col.names = c(
    "Target", "Bootstrap skewness", "Percentile 2.5%", "Percentile 97.5%",
    "Normal 2.5%", "Normal 97.5%"
  ),
  caption = "Percentile and normal-approximation intervals from the same 2,000 resamples, with the skewness of each resampling distribution.",
  row.names = FALSE
)
stopifnot(nrow(dat)==397, df.residual(s1)==391, nrow(moore)==45, all(salary_bootstrap_summary$successful_resamples==2000))


# Paired simple-slope draws: each column comes from one complete-row refit.
set.seed(90910)
practice_coef_boot <- replicate(2000, coef(lm(score ~ hours * baseline_c,
  data = sim[sample.int(nrow(sim), replace = TRUE), ])))
practice_L <- cbind(0, 1, 0, c(-10, 0, 10))
practice_draws <- practice_L %*% practice_coef_boot
practice_difference <- practice_draws[3, ] - practice_draws[1, ]
practice_boot_summary <- data.frame(base = c(60,70,80),
  estimate = drop(practice_L %*% coef(m1)),
  SE = apply(practice_draws,1,sd),
  lower = apply(practice_draws,1,quantile,.025),
  upper = apply(practice_draws,1,quantile,.975))
practice_boot_long <- data.frame(
  base = factor(rep(c(60,70,80), each=2000)),
  slope = as.vector(t(practice_draws)))
# Controlled change in predictor coverage, with identical rows and errors.
precision_original <- sim
precision_error <- with(sim, score-(60+2.5*hours+.4*baseline_c+.12*hours*baseline_c))
precision_narrow <- sim
precision_narrow$baseline_c <- .3*sim$baseline_c
precision_narrow$base <- 70+precision_narrow$baseline_c
precision_narrow$score <- with(precision_narrow,
  60+2.5*hours+.4*baseline_c+.12*hours*baseline_c)+precision_error
precision_fit <- lm(score ~ hours * baseline_c, precision_narrow)
precision_intervals <- data.frame(design=c("Original coverage","Compressed coverage"),
  estimate=c(coef(m1)[4],coef(precision_fit)[4]),
  lower=c(confint(m1)[4,1],confint(precision_fit)[4,1]),
  upper=c(confint(m1)[4,2],confint(precision_fit)[4,2]))
stopifnot(all(is.finite(practice_coef_boot)),
  abs(sd(practice_difference)-.2737093)<1e-6,
  abs(coef(precision_fit)[4]-.10295382)<1e-7)

```

{{< pagebreak >}}

::: {.callout-note title="Chapter overview"}
An additive straight-line model gives each predictor one partial slope.
This chapter asks when a comparison should instead vary across another
predictor's values or groups. A product term makes that possibility
explicit. We follow its meaning from generated practice-and-preparation
records to faculty salaries and conformity responses: interpret the
coefficients at their anchors, estimate conditional comparisons, show
where the records support them, and evaluate their uncertainty. By the end,
you can distinguish a varying association from a causal explanation and
write an informative report even when the slopes are poorly distinguished.
:::

The same equation can pose different research questions. An extra hour
of practice may have different associations with achievement at different
preparation levels. A year's service may have different salary associations
across academic ranks. We begin with the comparison that an additive model
holds fixed, then examine what changes when it is allowed to vary.

## A Comparison That Depends on Another Predictor {#ch10-question}


### The restriction the additive model imposes

Suppose a researcher asks whether students who spend more time in
deliberate practice tend to achieve higher posttest scores. An additive
regression can allow students with stronger prior preparation to have
higher scores while requiring the practice-score relationship to be the
same at every preparation level. That requirement may be the very claim
the researcher needs to examine. Perhaps an additional hour of practice
is associated with a small score difference for students with lower
prior preparation and a larger difference for students with higher
prior preparation. Or perhaps the apparent difference is only sampling
noise.

The central question is therefore conditional:

> How does the expected outcome difference associated with one unit of a
> focal predictor vary across meaningful values or groups of another
> predictor?

An **interaction** is the model term that permits such variation. On the
modeled scale, interaction means departure from additivity. Parallel
lines are additive; nonparallel lines are interactive. The lines need
not cross inside the observed data, and a statistically detectable
product term does not by itself identify a causal mechanism.


```{r}
#| label: fig-ch10-additive-interactive
#| fig-pos: H
#| echo: false
#| fig-cap: "The same practice model with and without its product term. Arrows compare two with four practice hours at baseline 60 and 80: both differences are 5 under additivity, versus 2.6 and 7.4 under interaction. These are conditional model means."
#| fig-alt: "Two panels show expected score against practice. Additive lines at baseline 60 and 80 have equal two-hour differences of 5. Interactive lines have two-hour differences of 2.6 and 7.4, shown by arrows. The lines do not cross."
#| fig-width: 6.3
#| fig-height: 3.4
opening_grid <- expand.grid(hours=seq(0,8,length.out=81),
  base=c(60,80), pattern=c("Additive","Interactive"))
opening_grid$pattern <- factor(opening_grid$pattern, c("Additive","Interactive"))
opening_grid$predicted <- with(opening_grid, 60+2.5*hours+.4*(base-70)+
  ifelse(pattern=="Interactive",.12*hours*(base-70),0))
opening_grid$probe <- factor(opening_grid$base, c(60,80), c("Baseline 60","Baseline 80"))
opening_arrows <- subset(opening_grid,hours==2)
opening_arrows$end <- with(opening_arrows,predicted+2*(2.5+
  ifelse(pattern=="Interactive",.12*(base-70),0)))
opening_arrows$label <- paste0("+",round(opening_arrows$end-opening_arrows$predicted,1))
ggplot(opening_grid,aes(hours,predicted,linetype=probe))+
  geom_line(linewidth=.8)+facet_wrap(~pattern)+
  geom_segment(data=opening_arrows,aes(x=4,xend=4,yend=end),
    arrow=grid::arrow(length=grid::unit(.08,"inches"),ends="both"),linetype=1)+
  geom_segment(data=opening_arrows,aes(x=2,xend=4,yend=predicted),linetype=3)+
  geom_text(data=opening_arrows,aes(x=4.65,y=(predicted+end)/2,label=label),
    inherit.aes=FALSE,size=3.2)+
  labs(x="Practice hours",y="Expected posttest score",linetype=NULL)
```


Read the arrows as conditional comparisons. In the interactive panel,
two additional practice hours correspond to 2.6 score points at baseline
60 and 7.4 at baseline 80. Their difference is 4.8 points. The product
term will express this changing comparison in one equation. The lines
are model means at two named baseline values: they are neither individual
students' trajectories nor separate regressions fitted after splitting
the students into groups. Section 2.4 declares this model and generates
one sample from it.

This chapter first uses a simulation in which the conditional
relationship is known. It then studies a real observational example:
nine-month salary, years of service, and academic rank for the faculty
of one U.S. college in 2008-09, from the package-distributed
`carData::Salaries` data. The two examples play different roles. The
simulation verifies algebra and repeated-sample logic against known
truth. The salary analysis forces the harder decisions about units,
severely unequal support, factor coding, uncertainty, diagnostics, and
bounded reporting — and it shows what an honest interaction report looks
like when the data do not clearly distinguish rank-specific slopes from
one common slope.


### The same idea for groups and continuous predictors

Chapter 4's factorial model allowed the practice-schedule difference to
vary across feedback groups: its interaction test gave $F(2,18)=9.6$,
$p=.0015$, and its declared difference of differences was 4 with SE 1.8257.
Chapter 6 fitted that model in R. Here a continuous predictor replaces one
factor, so a surface replaces a finite cell table and a moving slope
replaces a single difference of differences. The ANOVA foundations remain
the same; what changes is the form of the conditional comparison.

### Interaction is defined on a stated scale

Additivity is always defined on a scale. Lines that are nonparallel for
salary in dollars remain nonparallel after converting dollars to
thousands of dollars, although every outcome coefficient is divided by
1,000. A
nonlinear transformation such as a logarithm can change whether a
pattern appears additive because it changes the model itself, not merely
the label on the axis.

The same caution matters in later generalized linear models: interaction
on a linear-predictor scale need not look the same on a response scale.
This chapter stays with continuous outcomes and the identity-link linear
model. Logistic interaction, multilevel interaction, and three-way
interaction are outside its worked scope.

::: {.callout-warning title="Moderation is not mediation or mechanism"}
A moderator names where a focal relationship is allowed to vary. That
reading describes when an association differs, rather than how an effect
is transmitted. An intervening variable and a causal mechanism need
additional design and identification evidence. Chapter 14 develops those
requirements.
:::

## The Product Term and Its Coefficients {#ch10-product}


### The product coefficient is a difference in slopes

Begin with two continuous predictors, $X$ and $W$, and a continuous
outcome, $Y$. An additive conditional-mean model is

$$
E(Y_i\mid X_i,W_i)=\beta_0+\beta_X X_i+\beta_W W_i.
$$

Holding $W$ fixed, the expected difference associated with a one-unit
difference in $X$ is always $\beta_X$. Holding $X$ fixed, the expected
difference associated with a one-unit difference in $W$ is always
$\beta_W$. The model permits different levels but not different slopes.

The product-term model is

$$
E(Y_i\mid X_i,W_i)
=\beta_0+\beta_X X_i+\beta_W W_i+\beta_{XW}X_iW_i.
$$ {#eq-continuous-interaction}

Regrouping the terms that involve $X$ gives

$$
E(Y_i\mid X_i,W_i)
=\left(\beta_0+\beta_W W_i\right)
 + \left(\beta_X+\beta_{XW}W_i\right)X_i.
$$

The conditional slope of $X$ at $W=w$ is therefore

$$
\frac{\partial E(Y\mid X,W)}{\partial X}
=\beta_X+\beta_{XW}w.
$$ {#eq-conditional-slope-x}

If $\beta_{XW}=0$, the slope does not vary with $W$ and the model is
additive. If $\beta_{XW}\ne0$, a one-unit difference in $W$ changes the
conditional slope of $X$ by $\beta_{XW}$. This is what the product
coefficient measures. It is a **difference in slopes**, not the slope of
$X$ at every value of $W$.

The same model can be regrouped around $W$:

$$
\frac{\partial E(Y\mid X,W)}{\partial W}
=\beta_W+\beta_{XW}x.
$$ {#eq-conditional-slope-w}

The algebra is symmetric: each predictor's conditional slope depends on
the other. Substantive language need not be symmetric. Researchers often
call $W$ a **moderator** when theory makes $X$ the focal predictor and
$W$ the variable across which the $X$ relationship is expected to vary;
the conditional slope $\beta_X+\beta_{XW}w$ is then called a **simple
slope** or conditional effect at $w$ [@aiken1991; @hayes2022]. That label
comes from the research question, not from a privileged side of the
multiplication sign.

For two values $x_a$ and $x_b$ at the same moderator value $w$, the
conditional mean contrast is

$$
E(Y\mid x_b,w)-E(Y\mid x_a,w)
=(x_b-x_a)\left(\beta_X+\beta_{XW}w\right).
$$

That sentence contains several safeguards:

- it identifies the outcome scale;
- it identifies the focal predictor difference;
- it holds the other predictor at a stated value;
- it describes a conditional mean, not every case; and
- it remains associational unless the research design supports a causal
  interpretation.


### Hierarchy: the lower-order terms stay

The formula

```r
y ~ x * w
```

expands to `x + w + x:w`. The lower-order terms belong with their
product term. This **hierarchy principle** has both mathematical and
interpretive reasons [@cohen2003; @kutner2005].

First, deleting $X$ while retaining $XW$ constrains the slope of $X$ to
zero when $W=0$. Deleting $W$ imposes the analogous constraint at
$X=0$. Such constraints are rarely part of the research question.
Second, shifting either predictor changes which lower-order coefficients
represent slopes and differences. A hierarchical model remains the same
fitted surface after a change of origin; a product-only model generally
does not. Third, factor product terms are often represented by several
columns, so retaining the full lower-order structure keeps the block's
meaning coherent.

Hierarchy does not mean every possible product must be added. The
interaction should be motivated by the research question, graph, design,
or a prespecified model comparison. It means that once a product is in
the model, its component terms remain.


### Centering changes the question at zero

Let $W_c=W-c$. Substituting $W=W_c+c$ into
@eq-continuous-interaction gives

$$
E(Y\mid X,W_c)
=(\beta_0+c\beta_W)
 +(\beta_X+c\beta_{XW})X
 +\beta_WW_c+\beta_{XW}XW_c.
$$

Centering $W$ at $c$ therefore changes the intercept and the
lower-order coefficient of $X$. The new $X$ coefficient is the
conditional slope at the original value $W=c$. It does not change the
interaction coefficient, fitted values, residuals, or the corresponding
full-versus-reduced product-block test. Centering is a
**reparameterization**: it puts zero at a value that makes a coefficient
worth interpreting [@aiken1991].

Centering is not a remedy for confounding, selection bias, measurement
error, omitted nonlinearity, lack of overlap, or a poorly specified
research question. It can change correlations among constructed columns
and may make numerical output easier to read, but it does not add
information or make unsupported comparisons valid.

Scaling is another reparameterization. If $X_{10}=X/10$, then a one-unit
difference in $X_{10}$ is a ten-unit difference in $X$. Coefficients and
standard errors change units together; fitted values and substantive
conclusions should agree after correct translation.


### A known-truth example: what one sample recovers

Consider a deliberately simplified educational setting. Each row is one
student. The focal predictor is weekly deliberate practice in hours,
$X$. The second predictor is a baseline assessment score, $W$. The
outcome is a posttest score, $Y$. The simulation asks an associational
model question under known conditions: how does the practice-score slope
vary with baseline score?

The data-generating process centers baseline score at 70:

$$
\begin{aligned}
Y_i={}&60+2.5X_i+0.4(W_i-70)\\
 &+0.12X_i(W_i-70)+\varepsilon_i,\\
\varepsilon_i&\sim\mathcal N(0,5^2).
\end{aligned}
$$ {#eq-simulation-dgp}

The coefficients are fixed by design. At a baseline score of 70, the
true practice slope is 2.5 score points per hour. Ten baseline points
below that anchor, it is
$2.5+0.12(-10)=1.3$ points per hour. Ten points above, it is
$2.5+0.12(10)=3.7$ points per hour. The positive product coefficient
does not say that practice changes baseline preparation, or vice versa;
it says that the conditional mean surface is nonadditive.

Practice hours and baseline scores are generated independently here so
that the interaction mechanics are transparent. Real educational data
would demand attention to attendance, access to support, prior courses,
measurement reliability, clustering within classrooms or schools, and
the process by which practice time was chosen. None of those design
complications is solved by fitting a product term.


The explicit seed makes this one sample reproducible. It does not make
the sample representative of a real population. The observed baseline
range is 42.8747 to
96, and practice lies between
0.0141 and
7.9802 hours. Conditional estimates
outside this rectangle would be extrapolations even though the generating
equation is known to us.

The working object `sim` holds `score`, `hours`, `base` and
`baseline_c = base - 70`. Generating the baseline, practice and independent
error in that order with seed 909 gives the declared sample.

```{r}
#| eval: false
#| echo: true
set.seed(909)
base <- pmin(pmax(rnorm(320, 70, 10), 42), 96)
hours <- runif(320, 0, 8)
baseline_c <- base - 70
score <- 60 + 2.5*hours + .4*baseline_c +
  .12*hours*baseline_c + rnorm(320, 0, 5)
sim <- data.frame(score, hours, base, baseline_c)
m1 <- lm(score ~ hours * baseline_c, data = sim)
```

The R formula uses `*` to include both lower-order terms and their
product. The table below reports each estimate with its standard error
(SE), the estimated standard deviation of that estimate across
hypothetical repeated samples, and a 95% confidence interval.


| Term | Target | Estimate | SE | 95% lower | 95% upper |
|:--|--:|--:|--:|--:|--:|
| Intercept | 60.00 | 59.6123 | 0.5550 | 58.5204 | 60.7043 |
| Practice | 2.50 | 2.6972 | 0.1244 | 2.4524 | 2.9421 |
| Centered baseline | 0.40 | 0.3822 | 0.0625 | 0.2593 | 0.5051 |
| Practice by centered baseline | 0.12 | 0.1149 | 0.0139 | 0.0875 | 0.1422 |

: One generated sample against its declared targets (seed 909, 320 rows). {#tbl-ch10-recovery tbl-colwidths="[34,14,14,12,13,13]"}


One noisy sample does not reproduce every truth exactly. The fitted
practice slope at baseline 70 is
2.6972 points per
hour, compared with the true value 2.5. The fitted interaction is
0.1149
points per hour per baseline-score point, compared with the true value
0.12. Its interval is
[0.0875,
0.1422]. The interval
quantifies repeated-sample uncertainty under the fitted model; it does
not quantify every possible source of design or measurement error.

The fitted lower-order coefficient for centered baseline,
0.3822, is the baseline
score slope when practice is zero. It is not a baseline-score slope
averaged over all practice values. In an interaction model, every
lower-order coefficient is conditional on the other component being at
its coded zero or reference level.


### The product block against the additive restriction

The reduced model is written first because it states the restriction
being evaluated: one common practice slope and one common baseline
slope. The full model releases the common-slope restriction by adding
the product.


The nested comparison gives

$$
F=\frac{(RSS_R-RSS_F)/q}{RSS_F/(n-p)}
=\frac{(9108.7071-7488.7866)/1}{7488.7866/316}=68.3548,
\qquad p<.0001.
$$

Here $q=1$ added coefficient and $p=4$ total coefficients. The product
coefficient's $t=8.2677$ has the same squared statistic from full precision.
Both models use the same 320 outcomes.

Here the product block has one degree of freedom because both predictors
are continuous. The comparison asks whether allowing the practice slope
to vary linearly with baseline score reduces residual variation beyond
the additive model. It does not test whether the entire model is true,
whether the relationship is causal, or whether the interaction is large
enough to matter in a particular decision.

In the one-degree-of-freedom case, the block $F$ statistic equals the
squared $t$ statistic for the product coefficient. That identity does
not generalize to a factor with three levels: a continuous-by-three-level
factor interaction requires two product columns and a joint two-degree-
of-freedom test.


```{r}
#| eval: false
#| echo: true
m0 <- lm(score ~ hours + baseline_c, data = sim)
anova(m0, m1)
```


### Recentering leaves the surface alone

The centered model uses baseline 70 as zero. A raw-score model uses the
unobserved and impossible baseline score zero as its anchor.


| Quantity | Raw-score model | Centered model |
|:--|--:|--:|
| Coefficient of practice at the coded zero | $-5.3448$ | 2.6972 |
| Lower-order baseline coefficient | 0.3822 | 0.3822 |
| Product coefficient | 0.1149 | 0.1149 |
| Product block $F$ | 68.3548 | 68.3548 |
| Maximum fitted-value difference | below $10^{-12}$ | below $10^{-12}$ |

: What centering changes and what it leaves alone. {#tbl-ch10-centering tbl-colwidths="[46,27,27]"}


The raw model's practice coefficient is the fitted practice slope at a
baseline score of zero. Its unattractive value is not evidence that the
model failed; it answers an unsupported question. After centering, the
practice coefficient answers the supported question at baseline 70.
The product coefficient and every fitted value agree to numerical
precision.

Centering can change the p-values of lower-order coefficients because it
changes the hypotheses they test. It should not change the one-degree-
of-freedom product test in this corresponding model. A lower-order
p-value becoming smaller after centering is not repaired confounding;
it is a different conditional comparison.


```{r}
#| eval: false
#| echo: true
raw <- lm(score ~ hours * base, data = sim)
max(abs(fitted(raw) - fitted(m1)))
```


## Probe the Conditional Comparison {#ch10-probe}


### A conditional slope needs the covariance

At a centered baseline value $w$, the fitted practice slope is

$$
\widehat\beta_X(w)=\widehat\beta_X+w\widehat\beta_{XW}.
$$

Because the two estimated coefficients are correlated, its variance is
not obtained by adding their variances alone:

$$
\begin{aligned}
\operatorname{Var}\{\widehat\beta_X(w)\}
={}&\operatorname{Var}(\widehat\beta_X)
+w^2\operatorname{Var}(\widehat\beta_{XW})\\
&+2w\operatorname{Cov}
(\widehat\beta_X,\widehat\beta_{XW}).
\end{aligned}
$$ {#eq-simple-slope-variance}

The conditional-slope table applies this covariance formula to three
meaningful baseline scores. Each interval uses $t_{.975,316}=1.9675$.


| Baseline score | Target slope | Fitted slope | SE | 95% lower | 95% upper |
|:--|--:|--:|--:|--:|--:|
| 60 | 1.3 | 1.5484 | 0.1780 | 1.1982 | 1.8986 |
| 70 | 2.5 | 2.6972 | 0.1244 | 2.4524 | 2.9421 |
| 80 | 3.7 | 3.8461 | 0.1947 | 3.4630 | 4.2291 |

: Conditional practice slopes at three supported baseline scores, with their declared targets. {#tbl-ch10-slopes tbl-colwidths="[20,16,16,12,18,18]"}


The estimated practice association rises from
1.5484 points per hour at baseline 60
to 3.8461 points per hour at baseline
80. These are three evaluations of one fitted surface, not three
separate regressions and not evidence that students belong to three
latent types.

The `emtrends()` function in the `emmeans` package calculates these
conditional trends and their covariance-based intervals directly.


Agreement with the manual table is a computational check. The software
does not choose defensible probe values. That responsibility remains
with the analyst.

For the slope at centered baseline $w=10$, applying the variance formula
to the full-precision covariance matrix gives variance 0.0379 and
SE 0.1947. Keep the entries at their stored precision until the final
result is reported: early rounding of small covariance terms can alter
the reconstructed SE.

Omitting covariance gives 0.1865. Adding the focal SE to ten times the
product SE gives 0.2634. Neither computes the uncertainty of the fitted
linear combination. A short contrast-vector calculation includes every
needed covariance automatically:

```{r}
#| eval: false
#| echo: true
b <- coef(m1); V <- vcov(m1)
cvec <- c(0, 1, 0, 10)  # intercept, hours, baseline_c, product
sum(cvec * b)
sqrt(drop(cvec %*% V %*% cvec))
```


```{r}
#| eval: false
#| echo: true
emmeans::emtrends(m1, ~ baseline_c, var = "hours",
  at = list(baseline_c = c(-10, 0, 10)))
```


A **simple slope** fixes the moderator and reads the focal comparison
from the whole fitted model. For a two-hour comparison, multiply the
slope and both endpoints by two: the fitted differences are 3.0969 at
baseline 60 and 7.6923 at baseline 80. A different practice interval
changes that multiplier, not the fitted relationship.

Now ask the interaction question directly: how much greater is the
practice slope at 80 than at 60? Because both slopes use the same model,
the focal coefficient cancels:

$$
\widehat\beta_X(10)-\widehat\beta_X(-10)
=20\widehat\beta_{XW}=2.2977,
\qquad SE=20\,SE(\widehat\beta_{XW})=0.2779.
$$

The 95% interval is [1.7509, 2.8445]. This compares **two slopes**, whereas
each row of @tbl-ch10-slopes compares **one slope with zero**. Their
uncertainty calculations answer different questions. In general,
$\operatorname{Var}(\widehat s_2-\widehat s_1)=
\operatorname{Var}(\widehat s_2)+\operatorname{Var}(\widehat s_1)
-2\operatorname{Cov}(\widehat s_2,\widehat s_1)$.
The two slope estimates share coefficients; subtracting interval
endpoints or adding their SEs ignores that relationship.

```{r}
#| eval: false
#| echo: true
20 * coef(m1)["hours:baseline_c"]
20 * confint(m1)["hours:baseline_c", ]
```

A large simple slope can coexist with zero interaction: both preparation
levels might have the same strong practice association. Conversely, a
changing slope can pass through zero at one baseline. Neither a test
against zero at a chosen probe nor two separate significance labels
replace a direct slope contrast [@gelman2006].

### Choosing probe values and showing where the data are

The **pick-a-point** approach — also called the analysis of simple
slopes — evaluates conditional slopes at a small set of substantively
chosen moderator values [@aiken1991; @hayes2022]. There is no rule that
these values must be the mean and one standard deviation above and below
it. Those values may be
uninterpretable, outside the data, or different across studies. Better
choices come from the measurement scale and the research question:
policy thresholds, clinically or educationally meaningful cutoffs,
prespecified percentiles, or values needed for an actual decision.

The 60, 70, and 80 probes work here because baseline score has an
interpretable unit and all three values are well supported. A complete
report also displays the moderator distribution rather than presenting
the probes as if observations were equally dense everywhere.


### The conditional-slope curve is the primary display

Three probe values are readings of one
continuous object: the fitted conditional slope
$\widehat\beta_X(w)=\widehat\beta_X+w\widehat\beta_{XW}$ with its pointwise
band, drawn over the moderator values the data actually support. That
curve, not a boundary number, is the primary display for a continuous
moderator [@bauer2005; @hayes2022]. The marked zero-crossing boundaries are explained in Section 3.4.
Their position relative to the observed support matters as much as
their numerical value in @fig-ch10-slope-curve.


```{r}
#| label: fig-ch10-slope-curve
#| echo: false
#| fig-cap: Fitted practice slope across the supported baseline-score range with a 95% pointwise band. Dashed vertical lines mark the two Johnson-Neyman boundaries, the shaded panel marks baseline scores below the smallest generated value, and the rug shows the observed baseline distribution.
#| fig-alt: The fitted practice slope rises almost linearly from slightly below zero at the lowest observed baseline score to nearly six score points per hour at the highest. Its confidence band leaves zero at the upper Johnson-Neyman boundary near baseline 51, while the lower boundary near baseline 39 sits inside the shaded unsupported region where no data exist.
#| fig-pos: "!ht"
#| fig-width: 6.3
#| fig-height: 3.4

slope_grid_c <- seq(
  min(sim$baseline_c),
  max(sim$baseline_c),
  length.out = 200
)

sim_slope_curve <- data.frame(
  base = slope_grid_c + 70,
  estimate = b_x + slope_grid_c * b_xw,
  SE = sqrt(v_x + slope_grid_c^2 * v_xw + 2 * slope_grid_c * cov_x_xw)
)
sim_slope_curve$lower <-
  sim_slope_curve$estimate - t_jn * sim_slope_curve$SE
sim_slope_curve$upper <-
  sim_slope_curve$estimate + t_jn * sim_slope_curve$SE

ggplot(sim_slope_curve, aes(x = base, y = estimate)) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = min(sim$base),
    ymin = -Inf,
    ymax = Inf,
    alpha = 0.10
  ) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.20) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, color = "grey40") +
  geom_vline(
    xintercept = jn_boundaries,
    linetype = "22",
    linewidth = 0.7
  ) +
  geom_rug(
    data = sim,
    aes(x = base),
    sides = "b",
    alpha = 0.25,
    inherit.aes = FALSE
  ) +
  labs(
    x = "Baseline score",
    y = "Fitted practice slope (score points per hour)"
  )
```


Read left to right, the curve tells one continuous story rather than
three separate group findings. At the lowest supported baseline scores
the fitted slope is slightly negative — about
$-0.4191$ score points per hour — and
its band covers zero. The band leaves zero at the upper boundary,
51.284, and the slope then rises steadily to
about 5.6843 score points per hour
at the highest observed baseline score. The three pick-a-point probes are
three readings of this same curve. The band widens toward both ends
because a conditional slope evaluated far from the centering value
borrows heavily from the product coefficient, whose own uncertainty is
then multiplied by a large $w$.

Marking the boundaries on the curve also makes the support limitation visible
rather than hidden in a table: the lower boundary sits inside the shaded
panel, where no student was generated, so nothing in these data speaks to
it.


### Levels, cases and the limit of the algebra

A slope compares changes; a fitted mean describes a level. At four
practice hours and baseline 70, the fitted mean is 70.4013 with mean-response
interval [69.8644, 70.9382]. The new-case interval [60.8082, 79.9944] is
wider because it also includes individual error variance. At baselines 60
and 80, the fitted means are 61.9836 and 78.8189. Precision about a mean
does not imply precise prediction of each student.

A **Johnson-Neyman** calculation searches the moderator continuum for
values at which a pointwise confidence interval for a conditional slope
touches zero, and it labels the values on one side of those boundaries a
**region of significance** [@johnson1936; @bauer2005; @hayes2022]. It can
complement pick-a-point probes when the moderator is genuinely continuous
and no small set of theory-based values captures the question. It
replaces neither the conditional-slope curve, the magnitude estimates,
nor the support audit.

For this fitted curve the pointwise interval excludes zero below baseline
39.039 and above 51.284; between these values it covers zero. Appendix A
shows how the boundary equation and its sign determine those regions.
The figure lets us check what those numbers mean before interpreting them.

This example reveals a support
limitation that a pair of printed roots would conceal. The lower boundary,
39.039, lies below the smallest generated
baseline score, 42.8747, so
0 of the 320 simulated
students fall in the “significantly negative” region: it exists in the
algebra and nowhere in the data. The upper boundary,
51.284, does lie inside the observed range, and
313 students
(97.8125 percent
of the sample) sit above it. The supported reading is therefore
one-sided: across almost the entire baseline range these data cover, the
practice slope is positive and separated from zero, and only a thin
low-baseline sliver of
7 students falls in the
interval-covers-zero region.

Any boundary is descriptive of that fitted model and its pointwise
procedure. Scanning many moderator values raises multiplicity questions;
a pointwise band is not a simultaneous band. A root outside the observed
range has no empirical support, and a root inside a sparse tail may still
be unstable. The phrase “region of significance” can also distract from
magnitude and uncertainty. Pick-a-point answers questions at chosen
values; Johnson-Neyman locates model-based zero-crossing boundaries.
Neither method discovers a uniquely correct moderator cutoff, protects
against specification search, or supplies simultaneous error control
[@preacher2006].


A **simultaneous band** answers a wider question: does one band cover the
entire conditional-slope curve? Under the correctly specified linear model
with independent Normal errors of constant variance, a Scheffé band for
the two coefficients determining that curve replaces 1.9675 by
$\sqrt{2F_{.95;2,316}}=2.4594$ in the estimate-plus-or-minus-SE calculation.
It is wider because its coverage applies jointly to every linear
combination of those two coefficients, hence every conditional slope.
The two counts the dimensions of this slope family, not all four
coefficients in the fitted mean model. Showing only the supported range
makes the display useful; it does not validate extrapolation. Section 6.4
compares this band with pointwise intervals from two calculations.


## Continuous by Categorical: Rank-Specific Slopes {#ch10-salary}


### The question, the records and the analysis sample

The sustained empirical question is:

> Among the faculty of this college in 2008-09, how does the linear
> association between years of service and mean nine-month salary differ
> across the assistant, associate, and full professor ranks?

Each row is one faculty member. The outcome is nine-month salary,
rescaled to thousands of dollars. The focal continuous predictor is
years of service at the institution. Rank is a three-level categorical
predictor. The target is a rank-conditional descriptive association and
conditional mean prediction within observed support. It is not a return
to loyalty, a promotion effect, or a claim about faculty salaries in
general.

::: {.callout-note title="Data and software source"}

The `Salaries` data are called directly from the
[`carData` package](https://cran.r-project.org/package=carData)
([package DOI](https://doi.org/10.32614/CRAN.package.carData)), which
documents them as the 2008-09 nine-month salaries of the assistant,
associate, and full professors at one U.S. college, collected as part of
the administration's ongoing monitoring of salary differences between
male and female faculty [@carData2022; @fox2019]. This chapter uses only
the package-distributed data and does not copy a local data file. One college
in one year is not a probability sample of academia, so all inferential
language below is bounded to these faculty under the working model.

Conditional trends are computed with the official
[`emmeans` package](https://rvlenth.github.io/emmeans/), whose
`emtrends()` function evaluates fitted slopes from a model reference
grid. The chapter also verifies key calculations from the fitted
coefficient vector and covariance matrix.
:::

The package dataset has 397 rows. The analysis rule is fixed before
model fitting: retain a row only when salary, years of service, and rank
are all observed, and use exactly the same rows for the additive and
interaction models. Completeness is verified rather than assumed.


The fixed analytic sample contains 397 faculty members:
67 assistant professors,
64 associate professors, and
266 full professors. Here the
complete-case rule excludes nothing, but stating and checking it is
still part of a prespecified analysis: the audit is what shows the rule
to be harmless rather than an untested assumption.


```{r}
#| eval: false
#| echo: true
d0 <- carData::Salaries
dat <- d0[complete.cases(d0[c("salary", "yrs.service", "rank")]), ]
dat$rank <- droplevels(dat$rank)
dat$salary_k <- dat$salary/1000
dat$service_c <- dat$yrs.service - 5
```


### Support first: where each rank actually lives

An interaction graph can tempt analysts to compare fitted lines where a
rank has few or no observations. Support must therefore be examined
within ranks, not only in the pooled sample.


| Rank | $n$ | Minimum | First quartile | Median | Third quartile | Maximum | Mean salary | SD |
|:--|--:|--:|--:|--:|--:|--:|--:|--:|
| Assistant | 67 | 0 | 1 | 3 | 3 | 6 | 80.7760 | 8.1741 |
| Associate | 64 | 1 | 7 | 8 | 11 | 53 | 93.8764 | 13.8317 |
| Full | 266 | 0 | 15 | 21 | 30 | 60 | 126.7721 | 27.7187 |

: Years-of-service support and salary summaries by rank (thousand dollars). {#tbl-ch10-support tbl-colwidths="[14,8,11,15,10,16,11,9,6]"}


The support asymmetry here is extreme, and it is the most important fact
in this analysis. Assistant professors span only
0 to
6 years of
service, while associate and full professors extend past fifty years.
The range common to all three ranks is only
1 to 6 years.
Whatever the model later reports about the assistant-professor slope is
estimated from that six-year window; the model can say nothing about
assistant professors with long service because none appear in these data.
The dataset cannot establish whether such faculty existed at the college
outside the recorded observations.


```{r}
#| label: fig-ch10-support
#| echo: false
#| fig-cap: Rank-specific years-of-service support. Thick segments show interquartile ranges, thin segments show observed ranges, and points mark medians.
#| fig-alt: Assistant-professor service years span only zero to six, while associate and full professors extend past fifty years. Only the narrow interval from one to six years is common to all ranks.
#| fig-pos: "!ht"
#| fig-width: 6.3
#| fig-height: 3.4

ggplot(salary_support, aes(y = rank)) +
  geom_segment(
    aes(x = minimum, xend = maximum, yend = rank),
    linewidth = 0.7,
    color = "grey55"
  ) +
  geom_segment(
    aes(
      x = first_quartile,
      xend = third_quartile,
      yend = rank
    ),
    linewidth = 4,
    color = "grey35"
  ) +
  geom_point(aes(x = median), size = 2.5) +
  annotate(
    "rect",
    xmin = common_service_range[1],
    xmax = common_service_range[2],
    ymin = -Inf,
    ymax = Inf,
    alpha = 0.08
  ) +
  labs(
    x = "Years of service",
    y = NULL
  )
```


```{r}
#| eval: false
#| echo: true
tapply(dat$yrs.service, dat$rank, range)
```


### The centered model and its two product columns

The model centers years of service at 5:

$$
S_c=S-5.
$$

This anchor lies in the observed range of all three ranks — it is inside
even the narrow assistant-professor window. It is chosen so that rank
lower-order coefficients represent salary differences at a supported
common service value. It is not chosen to force a desired p-value or to
make the ranks comparable in every other respect.

With assistant professor as the reference level, the model is

$$
\begin{aligned}
E(Y\mid S_c,R)={}&\beta_0+\beta_SS_c
 +\beta_A I(R=\text{AssocProf})\\
&+\beta_P I(R=\text{Prof})
 +\beta_{SA}S_c I(R=\text{AssocProf})\\
&+\beta_{SP}S_c I(R=\text{Prof}).
\end{aligned}
$$ {#eq-salary-interaction}

The three lines have the following interpretations:

$$
\begin{array}{ll}
\text{Assistant:}&\beta_0+\beta_SS_c,\\
\text{Associate:}&(\beta_0+\beta_A)+(\beta_S+\beta_{SA})S_c,\\
\text{Professor:}&(\beta_0+\beta_P)+(\beta_S+\beta_{SP})S_c.
\end{array}
$$

The two product coefficients are slope differences from the assistant
rank. They are not the associate and full-professor slopes themselves.


| Conditional term | Estimate | SE | 95% lower | 95% upper |
|:--|--:|--:|--:|--:|
| Assistant mean salary at 5 years | 82.8231 | 5.8693 | 71.2837 | 94.3625 |
| Assistant slope | 0.7793 | 1.9452 | $-3.0450$ | 4.6036 |
| Associate $-$ assistant at 5 years | 13.7977 | 6.8835 | 0.2644 | 27.3310 |
| Full $-$ assistant at 5 years | 46.0748 | 6.4445 | 33.4046 | 58.7449 |
| Associate $-$ assistant slope | $-1.1740$ | 1.9674 | $-5.0420$ | 2.6940 |
| Full $-$ assistant slope | $-0.8986$ | 1.9492 | $-4.7309$ | 2.9336 |

: Coefficients of the service-by-rank interaction model, read at their own anchors. {#tbl-ch10-salary-fit tbl-colwidths="[38,14,12,18,18]"}


At five years of service, the fitted mean salary for assistant
professors is
82.8231 thousand
dollars. The fitted associate-minus-assistant difference at that service
value is
13.7977 thousand
dollars, with an interval from
0.2644 to
27.331. The fitted
professor-minus-assistant difference is
46.0748 thousand dollars,
with an interval from 33.4046 to
58.7449.

Those are conditional comparisons at one service value. They are not
overall rank differences and are not comparisons at each rank's typical
service. The product terms further show that the associate slope differs
from the assistant slope by an estimated
$-1.174$
thousand dollars per year, while the professor slope differs by
$-0.8986$ — both
with wide intervals, dominated by the six-year assistant support window.


```{r}
#| eval: false
#| echo: true
s1 <- lm(salary_k ~ service_c * rank, data = dat)
coef(s1)
```


### The two-degree-of-freedom product block

Because rank has three levels, its interaction with one continuous
predictor contributes two columns. The appropriate nested comparison
tests them jointly, and this omnibus block question comes before any
examination of individual slopes or their pairwise differences.


| Variance procedure | Numerator df | Denominator df | $F$ | $p$ |
|:--|--:|--:|--:|--:|
| Classical nested-model $F$ | 2 | 391 | 0.4860 | .6155 |
| Approximate HC3 joint Wald $F$ | 2 | 391 | 1.3138 | .2700 |

: Classical and approximate robust tests of the two-column service-by-rank product block. {#tbl-ch10-blocks tbl-colwidths="[40,16,18,13,13]"}


The block comparison gives
$F(2, 391)=
0.486$,
$p=.6155$. Under the linear-model working
assumptions, the data provide no clear evidence against one common
service slope for the three ranks. That is a statement about what these
data can distinguish, not a demonstration that the three slopes are
equal — the later interval estimates show how wide a range of slope
differences remains compatible. The test also does not say whether the
lines are exactly linear or whether rank causally changes the salary
return to service.

An approximate joint Wald sensitivity check using the HC3 covariance matrix — the
estimator named in the chapter box, which keeps the ordinary least
squares (OLS) coefficients that minimize the residual sum of squares but
rebuilds their covariance from leverage-adjusted squared residuals
instead of one pooled variance [@long2000] — gives
$F(2,
391)=
1.3138$,
$p=.27$. Dividing the robust Wald statistic by
the number of restrictions and comparing it with an $F$ distribution using
the residual degrees of freedom is a transparent finite-sample approximation,
not an exact robust $F$ reference result. The robust and classical block
procedures therefore support the same bounded conclusion here, although
the p-values differ because the procedures use different variance models,
not because the fitted slopes changed.

The order of `anova(reduced, full)` matters for readable output and for
the narrative. Both models must use the same response and exactly the
same rows. If the full model silently loses more observations because a
new variable is missing, the models are not nested on a common dataset
and the comparison no longer answers the intended block question. Here
the analysis sample was fixed once, before fitting, so the printed row
counts agree by construction.


```{r}
#| eval: false
#| echo: true
s0 <- lm(salary_k ~ service_c + rank, data = dat)
anova(s0, s1)
```


### Rank-specific slopes and their comparisons

With the joint block assessed, the rank slopes are more direct than the
reference-coded product coefficients. `emtrends()` obtains the fitted
service trend within each rank.


| Rank | Slope (thousand dollars per year) | SE | 95% lower | 95% upper |
|:--|--:|--:|--:|--:|
| Assistant | 0.7793 | 1.9452 | $-3.0450$ | 4.6036 |
| Associate | $-0.3947$ | 0.2949 | $-0.9744$ | 0.1850 |
| Full | $-0.1193$ | 0.1253 | $-0.3656$ | 0.1270 |

: Rank-specific fitted service slopes with covariance-based standard errors. {#tbl-ch10-rank-slopes tbl-colwidths="[20,34,14,16,16]"}


Within the observed rank-specific ranges, one additional year of service
is associated with an estimated mean salary difference of about
0.7793
thousand dollars for assistant professors,
$-0.3947$
for associate professors, and
$-0.1193$
for full professors. Every classical interval includes zero, and the
assistant-professor interval is several times wider than the others —
its slope is estimated from a six-year window. The rank, unit, support and associational reading belong together. The
slightly negative continuing-rank estimates are compatible with cohort
and market effects that these cross-sectional data cannot separate from
service itself.

Pairwise slope differences address a different family of questions. We
use a Tukey-style adjustment for the three pairwise comparisons; the
prespecified omnibus block test reported above remains the primary check
of whether slopes vary somewhere across ranks.

\Needspace{8\baselineskip}


The adjusted $p$-values are 0.8220, 0.8895 and 0.6661 for associate-minus-assistant,
full-minus-assistant and full-minus-associate slopes. The multiplier is
2.3527 on 391 degrees of freedom. This calibration is approximate for
unequal slope-estimation variances, rather than the exact balanced
one-way Tukey result. No pairwise slope contrast is distinguishable from zero, in agreement
with the omnibus block. This coherent presentation — an omnibus check,
then interval estimates for every slope and contrast — is more
informative than announcing that one reference-coded product term is
small and another is not.


```{r}
#| eval: false
#| echo: true
emmeans::emtrends(s1, ~ rank, var = "service_c")
```


### Coordinates, common-value contrasts and the picture

To verify that claim, refit the salary model with full professor as the
reference rank. Coefficient labels and lower-order interpretations
change, but the fitted surface does not.


The maximum fitted-value difference is numerical noise. Reference coding
changes which line is described directly by $\beta_0$ and $\beta_S$ and
which slope differences appear as product coefficients. It does not
change the three lines, residuals, predictions, or the interaction-block
test.

The same principle applies to centering years of service. Had we
centered at 3 instead of 5 years, rank lower-order coefficients would
compare fitted means at 3 years. The slopes and slope differences would
remain the same. An analyst should choose the anchor before looking for
a more favorable coefficient table.

| Question | Quantity | Salary example |
|:--|:--|:--|
| How does salary vary with service within a rank? | Conditional slope | $-0.1193$ thousand dollars per year for full professors |
| How do ranks differ at one common service value? | Conditional mean contrast | Full $-$ assistant 46.0748 at 5 years |
| What mean salary does the model fit for a rank and service value? | Conditional mean | Associate 96.6208 at 5 years |
| What range is plausible for one new case under the model? | New-case interval | [49.6127, 143.6288] for one associate at 5 years |
| Is one common slope adequate across the ranks? | Product-block test | $F(2,391)=0.486$ |

: Four conditional questions and one block question, each with its own quantity. {#tbl-ch10-question-map tbl-colwidths="[38,22,40]"}

At two, three, five and six service years, the fitted full-minus-assistant
salary contrasts are 48.7706, 47.8720, 46.0748 and 45.1762 thousand dollars,
with SEs 4.2152, 4.2532, 6.4445 and 8.0405. These are linked readings of
one fitted surface, all inside the common service window. Their uncertainty
grows toward the edge of the short assistant range.

A single grid from the pooled minimum to pooled maximum would draw the
assistant-professor line across fifty unsupported years. The plotted prediction
grid uses the minimum and maximum service within each rank.


```{r}
#| label: fig-ch10-salary-fitted
#| echo: false
#| fig-cap: Rank-specific fitted mean salary with 95% mean-response bands. Each line and band stop at that rank's observed service limits.
#| fig-alt: Three nearly flat rank-specific salary lines sit at clearly different levels. The assistant-professor line and band cover only zero to six years of service, while the other lines extend past fifty years. No line is extended beyond its rank's observed range.
#| fig-pos: "!ht"
#| fig-width: 6.3
#| fig-height: 3.4

salary_prediction_grid <- bind_rows(lapply(
  levels(dat$rank),
  function(level) {
    rank_data <- dat[dat$rank == level, ]
    data.frame(
      yrs.service = seq(
        min(rank_data$yrs.service),
        max(rank_data$yrs.service),
        length.out = 100
      ),
      rank = factor(level, levels = levels(dat$rank))
    )
  }
)) |>
  mutate(service_c = yrs.service - 5)

salary_prediction_values <- predict(
  s1,
  newdata = salary_prediction_grid,
  interval = "confidence"
)
salary_prediction_grid <- cbind(
  salary_prediction_grid,
  fit = salary_prediction_values[, "fit"],
  lower = salary_prediction_values[, "lwr"],
  upper = salary_prediction_values[, "upr"]
)

ggplot(
  dat,
  aes(
    x = yrs.service,
    y = salary_k,
    linetype = rank,
    shape = rank
  )
) +
  geom_ribbon(
    data = salary_prediction_grid,
    aes(
      x = yrs.service,
      ymin = lower,
      ymax = upper,
      fill = rank,
      group = rank
    ),
    alpha = 0.16,
    color = NA,
    inherit.aes = FALSE
  ) +
  geom_point(alpha = 0.42) +
  geom_line(
    data = salary_prediction_grid,
    aes(y = fit, group = rank),
    linewidth = 0.9
  ) +
  scale_fill_grey(start = 0.82, end = 0.45) +
  labs(
    x = "Years of service",
    y = "Nine-month salary (thousand dollars)",
    linetype = "Rank",
    shape = "Rank",
    fill = "Rank"
  )
```


The visual and the trend table agree: rank differences in salary level
dominate, the fitted within-rank service trends are close to flat, and
the assistant-professor line is both short and imprecisely estimated.
The graph also keeps a limitation visible that a coefficient table can
hide: the ranks occupy drastically different parts of the service range.
The model estimates each within-rank slope from its own support; a
cross-rank comparison at a particular service value is most credible
inside the narrow common support of
1 to 6 years.


```{r}
#| eval: false
#| echo: true
predict(s1, newdata = data.frame(service_c = 0,
  rank = factor("AssocProf", levels = levels(dat$rank))),
  interval = "confidence")
```


## Factor by Factor: Cell Means Revisited {#ch10-cells}


### Two factors: the same nonadditive question

When both predictors are factors, an interaction is nonadditivity among
cell means. The `carData::Moore` data provide a compact
illustration from social psychology: 45 participants in a conformity
experiment [@moore1971], with the number of conforming
responses as the outcome, the manipulated status of a partner (low or
high) as one factor, and the participant's measured authoritarianism
category (low, medium, or high F-scale score) as a three-level factor.
The package documentation describes partner status as manipulated but
does not document random assignment or the full experimental protocol;
the example therefore treats it as an experimental-condition contrast
rather than assuming causal identification from the dataset description
alone.
This example is separate from the salary analysis and is not needed for
its conclusions. Under the classical common-variance linear model, its
product block is statistically detectable; the conclusion is sensitive
to the variance model and outcome scale, as shown below. The cell
summary reports, for each of the six cells, its size, its mean number of
conforming responses, and its standard deviation (SD), the typical
distance of a participant's score from that cell's mean.


| Authoritarianism | Partner status | $n$ | Mean conformity | SD |
|:--|:--|--:|--:|--:|
| Low | Low | 10 | 8.9000 | 2.6437 |
| Low | High | 5 | 17.4000 | 4.5056 |
| Medium | Low | 4 | 7.2500 | 3.9476 |
| Medium | High | 11 | 14.2727 | 3.9520 |
| High | Low | 8 | 12.6250 | 7.3473 |
| High | High | 7 | 11.8571 | 3.9340 |

: Conformity cell counts, means and standard deviations. {#tbl-ch10-moore-cells tbl-colwidths="[24,20,10,24,22]"}


An additive factor model requires the partner-status difference to be
the same at every authoritarianism level. The product model allows that
difference to vary. Its factor levels are ordered low, medium, high for
authoritarianism and low, high for partner status.


```{r}
#| label: fig-ch10-moore-cells
#| echo: false
#| fig-cap: Moore conformity cell means by partner status and authoritarianism category. The status difference is large at low and medium authoritarianism and essentially absent at high authoritarianism, a crossing nonadditive profile.
#| fig-alt: Mean conformity is much higher with a high-status partner among low- and medium-authoritarianism participants, but the two partner-status lines meet at the high-authoritarianism category.
#| fig-pos: "!ht"
#| fig-width: 6.3
#| fig-height: 3.4

ggplot(
  moore_cell_means,
  aes(
    x = fcategory,
    y = mean_conformity,
    group = partner.status,
    linetype = partner.status,
    shape = partner.status
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.8) +
  labs(
    x = "Authoritarianism category",
    y = "Mean conforming responses",
    linetype = "Partner status",
    shape = "Partner status"
  )
```


### The block, simple differences and interaction contrast

The block has two degrees of freedom because a three-level
authoritarianism factor contributes two indicator columns and therefore
two product columns. No middle category is discarded to manufacture a
2-by-2 table. The classical test gives
$F(2, 39)=
4.1846$,
$p=.0226$, whereas the approximate HC3
joint Wald test gives $F(2,
39)=
2.8294$,
$p=.0712$. The cells are small and unbalanced,
their spreads differ, and conformity is a bounded count. Ordinary linear
model inference is therefore a working approximation here, not a
scale-robust verdict. The cell-mean pattern remains pedagogically useful;
the strength of inferential evidence must be reported as
variance-model-sensitive.

For any two authoritarianism categories, a **statistical
difference-in-differences (DiD)** — the difference between two group
differences — is an interaction contrast. Comparing the low and high
categories gives

$$
\begin{aligned}
\widehat C_{\mathrm{DD}}={}&(\bar y_{\text{low},H}-\bar y_{\text{low},L})\\
&-(\bar y_{\text{high},H}-\bar y_{\text{high},L}),
\end{aligned}
$$

where $H$ and $L$ denote high- and low-status partners.


The contrast is 9.2679 conforming responses: the
partner-status difference is about
8.5
responses among low-authoritarianism participants and about
$-0.7679$
among high-authoritarianism participants. Algebraically it is a
product-term contrast. Partner status is described as manipulated, but
the package documentation alone does not establish random assignment,
implementation fidelity, or the absence of post-assignment complications.
The status comparison should therefore be described as an
experimental-condition contrast unless the original study documentation
supplies those warrants. Authoritarianism is measured rather than
assigned, so any variation of the status contrast across F-scale
categories also depends on observational group comparability. In a policy
or before-after study, a causal difference-in-differences
interpretation would additionally require a defensible treatment and
timing design, a counterfactual trend argument such as parallel trends,
absence of relevant anticipatory or spillover behavior, and uncertainty
appropriate to the assignment and dependence structure. Those
design-based methods are outside this chapter.

### What the difference of differences does not carry

::: {.callout-warning title="Statistical interaction is not a causal design"}

A product term describes nonadditivity in a model. Causal interpretation
comes from design and assumptions outside the coefficient algebra. The
same product term can appear in an observational cross-section, a
randomized factorial experiment, or a longitudinal policy comparison;
those settings do not inherit the same causal meaning.
:::


```{r}
#| eval: false
#| echo: true
moore <- carData::Moore
with(moore, tapply(conformity, list(fcategory, partner.status), mean))
c1 <- lm(conformity ~ partner.status * fcategory, data = moore)
```

For low versus high authoritarianism, the interaction contrast is

$$
\widehat C_{\mathrm{DD}}
=(\bar y_{\mathrm{low},H}-\bar y_{\mathrm{low},L})
-(\bar y_{\mathrm{high},H}-\bar y_{\mathrm{high},L})
=\mathbf c{}^{\mathsf T}\widehat{\boldsymbol\beta}=9.2679,
\qquad SE=\sqrt{\mathbf c{}^{\mathsf T}\widehat{\mathbf V}\mathbf c}=3.4507.
$$

The classical pointwise interval is [2.2883, 16.2475]. If this comparison
was selected after inspecting the cell pattern, the comparison family
needs a simultaneous calibration as in Chapter 4; an attractive contrast
is not retroactively prespecified.

## Check, Qualify and Resample {#ch10-checks}


### Where the model is likely to strain

The salary interaction model assumes an adequately linear conditional
mean within each rank, independent information from the modeled rows,
and sufficient joint service-by-rank support. Classical standard errors
and prediction intervals additionally use a common residual variance;
small-sample exact $t$ and $F$ results use a conditional Normal-error
model. Leverage and influence matter because product columns can make
sparse predictor combinations especially consequential. Residual plots
do not certify these assumptions; they show where parts of the
approximation may be incomplete [@fox2016; @gelman2020].


```{r}
#| label: fig-ch10-salary-residuals
#| echo: false
#| fig-cap: Residuals versus fitted values for the salary interaction model. Points are marked by rank and the smooth summarizes remaining mean pattern.
#| fig-alt: Residuals scatter around zero across fitted salary, with visibly larger spread at the higher fitted values that belong mostly to full professors.
#| fig-pos: "!ht"
#| fig-width: 6.3
#| fig-height: 3.4

salary_diag <- data.frame(
  fitted = fitted(s1),
  residual = resid(s1),
  standardized = rstandard(s1),
  rank = dat$rank
)

ggplot(salary_diag, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, color = "grey55") +
  geom_point(aes(shape = rank), alpha = 0.60) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    se = FALSE,
    color = "black",
    linewidth = 0.8
  ) +
  labs(
    x = "Fitted salary (thousand dollars)",
    y = "Residual salary (thousand dollars)",
    shape = "Rank"
  )
```


The residual-versus-fitted plot should be read with the raw plot and
support display. Ranks occupy different fitted ranges, and residual
spread is visibly larger at professor-level fitted values, so the
common-variance working assumption deserves the HC3 sensitivity check
below rather than silent acceptance.


```{r}
#| label: fig-ch10-salary-qq
#| echo: false
#| fig-cap: Normal quantile plot of standardized residuals from the salary interaction model. Tail departures matter most for small-sample interval calibration and unusual cases.
#| fig-alt: Standardized residuals track a reference line through most of the distribution with heavier departures in the upper tail.
#| fig-pos: "!ht"
#| fig-width: 6.3
#| fig-height: 3.4

qq_values <- qqnorm(
  salary_diag$standardized,
  plot.it = FALSE
)
qq_df <- data.frame(
  theoretical = qq_values$x,
  standardized = qq_values$y
)

ggplot(qq_df, aes(x = theoretical, y = standardized)) +
  geom_abline(intercept = 0, slope = 1, color = "grey55") +
  geom_point(alpha = 0.65) +
  labs(
    x = "Theoretical Normal quantile",
    y = "Standardized residual quantile"
  )
```


The Q-Q plot is an approximation check, not a requirement that the
salaries themselves be Normally distributed. The linear-model assumption
concerns the conditional errors used for exact small-sample reference
results.


```{r}
#| eval: false
#| echo: true
tapply(residuals(s1), dat$rank, sd)
plot(fitted(s1), residuals(s1))
abline(h = 0, lty = 2)
```


### The same slopes under a different covariance

The HC3 standard errors introduced with the block test replace the
common-variance coefficient covariance matrix with a leverage-adjusted,
residual-based estimate while retaining the OLS coefficients and the
fitted mean model [@long2000; @zeileis2004]. Standard errors, approximate
intervals, and approximate joint Wald tests derived from that matrix can
change. Applied to the rank-specific trends, they are a sensitivity analysis
for uncertainty, not a repair for a wrong conditional mean and not a license
to ignore dependence.

\Needspace{8\baselineskip}


| Rank | Slope | Classical SE | HC3 SE | Approximate HC3 lower | Approximate HC3 upper |
|:--|--:|--:|--:|--:|--:|
| Assistant | 0.7793 | 1.9452 | 0.8686 | $-0.9284$ | 2.4870 |
| Associate | $-0.3947$ | 0.2949 | 0.1679 | $-0.7248$ | $-0.0645$ |
| Full | $-0.1193$ | 0.1253 | 0.1884 | $-0.4898$ | 0.2512 |

: Rank-specific slopes under the classical and HC3 coefficient covariances. {#tbl-ch10-hc3 tbl-colwidths="[16,12,18,12,21,21]"}


Here the two variance treatments disagree in an instructive way. The
HC3 standard errors are smaller for the two junior ranks — whose
residual spread is below the pooled level — and larger for full
professors. The approximate HC3 sensitivity interval for the
associate-professor slope excludes zero
while the classical interval does not. A borderline conclusion that
moves when the variance model changes should be reported as sensitive,
not resolved by quietly choosing whichever calculation looks better.
There is a second trap here. Under HC3 the associate slope interval
excludes zero and the full-professor interval covers zero. That does not
show that the two slopes differ. Their directly estimated
associate-minus-full contrast is $-0.2754$, with approximate HC3 SE
0.2524 and interval [$-0.7716$, 0.2209] using the same residual-df
reference ($p=.2759$). The interval leaves substantial uncertainty about
the direction of the slope difference. Comparing two labels discards the
very covariance that the contrast needs; its nonrejection also does not
establish equal slopes.

Neither calculation repairs the fitted mean, the design, or the
dependence structure.


```{r}
#| eval: false
#| echo: true
V <- sandwich::vcovHC(s1, type = "HC3")
cvec <- c(0, 1, 0, 0, 1, 0)  # associate slope
sqrt(drop(cvec %*% V %*% cvec))
```


### Influence and a competing explanation

The most influential row is a full professor with
60 years of service and a salary of
192.253 thousand dollars — a
long-service, high-leverage point at the far edge of the professor
range. Omitting it moves the professor slope slightly and leaves the
other slopes essentially unchanged; the largest Cook's distance is
small, and the broad pattern of imprecise, overlapping slopes remains.
This leave-one-out fit is a sensitivity description. Deletion would
require a documented recording error, an explicit target-population
rule, or another substantive reason independent of Cook's distance.

If the true relationship is curved and ranks occupy different ranges,
separate straight lines can appear to have different slopes even when
the same curved function applies. A compact sensitivity model adds one
common quadratic service term while retaining the full lower-order
interaction structure.


This one-degree-of-freedom check provides little evidence that a shared
quadratic term materially improves the fitted mean. It does not prove
that every rank relationship is exactly linear. Rank-specific curvature
would require additional terms and, for assistant professors,
substantially more support than a six-year window can offer. A
sensitivity model should clarify a plausible failure mode, not become an
unrestricted search for significance.


```{r}
#| eval: false
#| echo: true
which.max(cooks.distance(s1))
anova(s1, lm(salary_k ~ service_c * rank + I(service_c^2), dat))
```


### Bootstrap the conditional comparison

Resampling follows the question. For the generated students, that
question is the practice slope at baseline 60, 70 or 80, or the difference
between two such slopes. One complete-row resample supplies one new fit,
and that fit supplies **all** these answers. Resampling each data column
separately would destroy the observed association. Resampling the two
slopes separately would destroy their shared estimation uncertainty.

Treating the generated student rows as independent and identically
distributed, draw 320 complete rows with replacement and refit the same
hierarchical model. Repeat 2,000 times with seed 90910. This approximates
repeating the sampling and fitting process for the stated slope
estimands; it does not generate new independent evidence [@efron1993;
@freedman1981]. The three distributions in @fig-ch10-bootstrap-slopes
contain estimates, not individual students' scores. Their centers and
widths differ because the three probes ask different conditional
questions of the same records.

| Baseline | Original slope | Bootstrap SE | Percentile lower | Percentile upper |
|:--|--:|--:|--:|--:|
| 60 | 1.5484 | 0.1764 | 1.1866 | 1.8790 |
| 70 | 2.6972 | 0.1262 | 2.4532 | 2.9328 |
| 80 | 3.8461 | 0.1954 | 3.4523 | 4.2187 |

: The same three simple slopes followed through 2,000 complete-row resamples. {#tbl-ch10-practice-bootstrap tbl-colwidths="[12,20,20,24,24]"}

```{r}
#| label: fig-ch10-bootstrap-slopes
#| fig-pos: H
#| echo: false
#| fig-cap: "Bootstrap distributions of the three conditional practice slopes. Solid lines mark original estimates; dashed lines mark the 2.5th and 97.5th percentiles. Each panel uses the same 2,000 row resamples, with a shared horizontal scale for comparing their spreads."
#| fig-alt: "Three histograms for baseline 60, 70 and 80 center near slopes 1.55, 2.70 and 3.85. The middle probe has the smallest spread. Solid estimate lines and dashed percentile limits distinguish centers from uncertainty."
#| fig-width: 6.3
#| fig-height: 3.1
ggplot(practice_boot_long,aes(slope))+
  geom_histogram(binwidth=.06,fill="grey55",color=NA)+
  geom_vline(data=practice_boot_summary,aes(xintercept=estimate),linewidth=.7)+
  geom_vline(data=practice_boot_summary,aes(xintercept=lower),linetype=2)+
  geom_vline(data=practice_boot_summary,aes(xintercept=upper),linetype=2)+
  facet_wrap(~base,labeller=label_both)+
  labs(x="Simple slope (score points per hour)",y="Resamples")

```

For every resample, subtract its slope at 60 from **that same resample's**
slope at 80. The resulting difference has bootstrap SE 0.2737 and
percentile interval [1.7404, 2.8368], close to the classical interval
[1.7509, 2.8445]. If `draws` holds one slope per row and one resample per
column, the calculation is short:

```{r}
#| eval: false
#| echo: true
diff <- draws[3, ] - draws[1, ]
sd(diff)
quantile(diff, c(.025, .975))
```

The agreement here is expected from a deliberately simple, well-behaved
generating model. It is not a general promise that bootstrap intervals
are wider, more conservative or more accurate. A percentile interval
reads quantiles of a resampling distribution; it is not a statement that
this realized interval contains the true slope with 95% posterior
probability. More resamples make those quantiles less sensitive to Monte
Carlo variation. They do not make the study's 320 observations more
informative.

@fig-ch10-calibrations answers another question: what changes when the
coverage statement concerns a whole curve? All panels use the same
original estimate. The first two show **pointwise** intervals: one
classical, one percentile bootstrap. The third uses Section 3.4's
classical simultaneous multiplier and is visibly wider. Its additional
protection concerns the family of slopes, under the classical model
assumptions; it is not a correction for a wrong model or weak design.
Taking percentiles at many grid points does not turn the second panel
into a simultaneous band.

```{r}
#| label: fig-ch10-calibrations
#| fig-pos: H
#| echo: false
#| fig-cap: "One slope curve, three coverage statements. Classical and bootstrap pointwise bands concern a fixed moderator value; the wider classical Scheffé band simultaneously covers the slope family under its stated linear-model assumptions. All displays stop at observed support."
#| fig-alt: "Three panels show the same rising conditional-slope curve. Classical and bootstrap pointwise bands are similar. The classical simultaneous band is wider, especially at the ends. Rugs show sparse support in the tails."
#| fig-width: 6.3
#| fig-height: 3.3
w <- seq(min(sim$base),max(sim$base),length.out=151)
L <- cbind(0,1,0,w-70)
est <- drop(L%*%coef(m1))
se <- sqrt(rowSums((L%*%vcov(m1))*L))
q <- qt(.975,df.residual(m1)); k <- sqrt(2*qf(.95,2,df.residual(m1)))
bounds <- t(apply(L%*%practice_coef_boot,1,quantile,c(.025,.975)))
cal <- rbind(data.frame(base=w,estimate=est,lower=est-q*se,upper=est+q*se,
    band="Classical pointwise"),
  data.frame(base=w,estimate=est,lower=bounds[,1],upper=bounds[,2],
    band="Bootstrap pointwise"),
  data.frame(base=w,estimate=est,lower=est-k*se,upper=est+k*se,
    band="Classical simultaneous"))
cal$band <- factor(cal$band,c("Classical pointwise","Bootstrap pointwise","Classical simultaneous"))
ggplot(cal,aes(base,estimate))+
  geom_ribbon(aes(ymin=lower,ymax=upper),fill="grey75")+
  geom_line(linewidth=.7)+geom_hline(yintercept=0,color="grey45")+
  geom_rug(data=sim,aes(x=base),inherit.aes=FALSE,alpha=.2)+
  facet_wrap(~band,ncol=3)+labs(x="Baseline score",y="Fitted practice slope")

```

Return now to the salary example, where unequal spread made the choice
of uncertainty calculation consequential.

The associate slope is the quantity whose reading changed under HC3.
A case bootstrap asks what another sampling approximation says about it:
draw whole faculty records with replacement and refit the same model.
This preserves the empirical association of rank, service and spread,
while treating rows as independent and exchangeable. With seed 90909,
all 2,000 resamples here contain all three ranks. The bootstrap machinery
is the case-resampling scheme of Chapter 9 [@efron1993; @freedman1981].

\Needspace{9\baselineskip}

| Rank slope | Estimate | Classical SE | HC3 SE | Bootstrap SE | Percentile 2.5% | Percentile 97.5% |
|:--|--:|--:|--:|--:|--:|--:|
| Assistant | 0.7793 | 1.9452 | 0.8686 | 0.8074 | $-0.7172$ | 2.3841 |
| Associate | $-0.3947$ | 0.2949 | 0.1679 | 0.1736 | $-0.8066$ | $-0.1436$ |
| Full | $-0.1193$ | 0.1253 | 0.1884 | 0.1872 | $-0.4928$ | 0.2522 |

: Three uncertainty calculations for the same three conditional slopes (seed 90909, $B=2000$). {#tbl-ch10-bootstrap tbl-colwidths="[18,13,15,12,15,13,14]"}

The associate bootstrap SE 0.1736 is close to HC3's 0.1679 rather than
the classical 0.2949. Its percentile interval also excludes zero. The
resampling skewness is $-0.8463$, so a symmetric normal approximation,
[$-0.7350$, $-0.0544$], conveys a somewhat different shape than the
percentile interval [$-0.8066$, $-0.1436$]. The interval method belongs in
the report. Neither interval is an independent replication of the study.

If `draws` holds one bootstrap associate-slope estimate per element,
these two operations summarize that resampling distribution:

```{r}
#| eval: false
#| echo: true
sd(draws)
quantile(draws, c(.025, .975))
```

Row resampling cannot represent unknown departmental dependence, create
long-serving assistant professors, repair measurement problems, or identify
a promotion effect. A design with clusters, strata or weights requires
a resampling scheme that follows them. Agreement between HC3 and the
case bootstrap is a sensitivity result about these same observations.

## From Analysis to a Defensible Claim {#ch10-report}


### The pattern before the verdict

The interaction block answers whether one common slope is adequate under
the fitted linear model. A defensible report does not stop there. It
describes the conditional pattern, gives interval estimates in original
units, shows where the comparisons are supported, and states what the
design does not identify.

For the salary analysis, the main pieces are:

- 397 faculty members at one college in 2008-09, with
  rank-specific service ranges shown explicitly;
- a hierarchical model with salary in thousands of dollars, service
  centered at 5 years, and a two-degree-of-freedom product block;
- an omnibus interaction comparison of
  $F(2, 391)=
  0.486$,
  $p=.6155$;
- estimated within-rank slopes with confidence intervals;
- fitted means or contrasts at a supported common service value;
- residual, unequal-variance, influence, curvature, and support checks;
- three uncertainty procedures for the same conditional slopes, including
  the one slope whose approximate sensitivity interval excludes zero under
  HC3 and case resampling but not under the classical common-variance model;
  and
- an observational interpretation bounded to the represented data.

::: {.callout-tip title="A bounded report"}

Among the 397 faculty members of this college in
2008-09, rank differences in mean nine-month salary were large, but the
data provided no clear evidence that the linear service-salary
association differed across ranks
($F(2, 391)=
0.486$,
$p=.6155$). Within each rank's observed range,
the estimated slopes were
0.7793 thousand dollars per
service year for assistant professors
(95% confidence interval [$-3.045$,
4.6036]),
$-0.3947$ for associate
professors ([$-0.9744$,
0.185]), and
$-0.1193$ for full professors
([$-0.3656$,
0.127]). Every classical interval includes zero
and the intervals overlap; the assistant-professor slope rests on a
service range of only
0 to
6 years. The
associate-professor slope was the one quantity sensitive to the variance
procedure: its approximate sensitivity interval excluded zero under both an HC3 variance model
([$-0.7248$,
$-0.0645$]) and a 2,000-resample
case bootstrap
([$-0.8066$,
$-0.1436$]) but not under
the classical common-variance model, and its resampling distribution was
skewed, so the interval type matters as well. The result is a descriptive
conditional association for these faculty; it does not estimate a return
to service, identify promotion or cohort mechanisms, or support
extrapolation beyond observed rank-service combinations.
:::

This paragraph reports a null block result without converting it into a
claim that the slopes are equal, and it reports the variance-model
sensitivity instead of quietly resolving it. Reporting what the data
cannot distinguish is as much a finding as reporting what they can.


### A reusable template and the statements it blocks

A reusable interaction report can be organized in six sentences.

1. **Target and unit:** “Among [analytic units], we examined whether the
   [outcome-unit] association between [focal predictor] and [outcome]
   varied across [moderator values or groups].”
2. **Model and support:** “We fit a hierarchical [model scale] model on
   [fixed sample], with probes restricted to [observed joint support].”
3. **Block evidence:** “Adding the [named] product block changed model fit
   by [degrees of freedom, statistic, interval or p-value].”
4. **Conditional estimates:** “At [prespecified supported values], the
   estimated slopes or contrasts were [estimates and intervals in units].”
5. **Checks:** “Residual, variance, influence, curvature, and alternative-
   coding checks suggested [specific consequences].”
6. **Boundary:** “The result supports [bounded associational or predictive
   conclusion] but does not establish [causality, mechanism,
   generalization, or unsupported extrapolation].”

Not every paper must use these exact sentences. Every report should make
their information recoverable.

| Attractive statement | Why it fails | Defensible replacement |
|:--|:--|:--|
| "The interaction proves moderation." | A product term is model-based nonadditivity; substantive moderation needs a theory-defined role and a design. | "The fitted conditional slope varied across the named values or groups." |
| "There is no effect of $X$." | The lower-order coefficient is the slope only at the coded zero or reference value. | "At $W=0$ under this coding the estimated slope of $X$ was ..." |
| "The lines interact only if they cross." | Interaction means nonparallelism on the modelled scale. | "The estimated slopes differed over the displayed support." |
| "Centering fixed the collinearity and the confounding." | Centering changes the origin, not the information, the omitted variables, or the design. | "Centering made the zero-value comparison interpretable." |
| "The product $p$-value shows where the effect exists." | One block test does not describe conditional magnitudes across the moderator. | "The slope curve with its band and support shows where the comparison is estimated and how precisely." |
| "One group's slope is significant and another's is not, so they differ." | Two significance verdicts are not a test of their difference. | "The slope contrast was ... with interval ..." |
| "A large product $p$-value shows the slopes are equal." | Nonrejection is not evidence of no difference. | "A slope difference between ... and ... remains compatible with these records; claims are restricted to observed joint support." |

: Common interaction overclaims and their defensible replacements. {#tbl-ch10-overclaims tbl-colwidths="[28,38,34]"}

The sustained salary example is exactly this situation, so it deserves
its own statement of principle. A large product p-value is not proof
that conditional slopes are exactly equal. It says the data and model do
not distinguish the product block from zero with the chosen precision
and reference procedure. A bounded null report gives the estimated
difference and its interval, then asks whether values of practical
importance remain compatible with the data.

For example, suppose an estimated slope difference is 0.08 outcome units
per predictor unit with a 95% interval from $-0.12$ to 0.28. “No
interaction” hides the fact that modest negative and positive differences
remain compatible. The relevant next question is substantive: would any
part of that interval change the decision or theory?

Why can a study estimate a large average association while leaving
moderation imprecise? A product coefficient needs information about how
comparisons vary across **joint predictor combinations**. Consider the
same 320 generated students. Keep practice hours and each student's
original generated error fixed, compress the centered baseline scores
to 30% of their original deviations, and generate scores again from the
same equation, including the true product coefficient 0.12. This changes
the design's predictor coverage, not the true interaction or error SD.

\Needspace{5\baselineskip}

**Predict before reading the intervals:** would retaining 320 students
and the same error realization preserve the precision of the interaction
estimate? Which part of the information has changed?

```{r}
#| label: fig-ch10-precision
#| fig-pos: H
#| echo: false
#| fig-cap: "The same sample size and true interaction, with different predictor coverage. Left: compression narrows baseline variation while preserving all practice values. Right: the product-coefficient interval widens around the unchanged generating coefficient 0.12 (dashed line)."
#| fig-alt: "The left panel overlays the original broad baseline-by-practice cloud and a compressed cloud concentrated near baseline 70. The right panel shows the original product interval about 0.09 to 0.14 and the much wider compressed interval about 0.01 to 0.19; both refer to a true product coefficient of 0.12."
#| fig-width: 6.3
#| fig-height: 3.3
old <- par(mfrow=c(1,2),mar=c(4.2,4.2,1.2,.6))
plot(sim$base,sim$hours,pch=1,col="grey65",xlab="Baseline score",ylab="Practice hours")
points(precision_narrow$base,precision_narrow$hours,pch=16,cex=.55)
legend("topright",c("Original","Compressed"),pch=c(1,16),col=c("grey65","black"),bty="n",cex=.8)
z <- precision_intervals
plot(z$estimate,2:1,xlim=c(0,.22),ylim=c(.5,2.5),yaxt="n",pch=16,
  xlab="Product coefficient",ylab="")
axis(2,at=2:1,labels=c("Original","Compressed"),las=1,cex.axis=.8)
segments(z$lower,2:1,z$upper,2:1,lwd=2)
abline(v=.12,lty=2,col="grey45")
par(old)
```

Baseline support contracts from 42.8747--96 to 61.8624--77.8000. The product
estimate changes from 0.1149 (SE 0.0139, [0.0875, 0.1422]) to 0.103
(SE 0.0463, [0.0118, 0.1941]). Both samples arise from a true coefficient
of 0.12, yet the compressed design leaves much more uncertainty about its
size. The same row count is not the same information. In this controlled
comparison the SE is multiplied by $1/0.3$; that numerical relation uses
the shared predictor values and errors, not a universal rule for every
restricted dataset.

Sparse tails, imbalanced groups and noisy measurements can similarly
weaken an interaction comparison [@gelman2020]. A single simulated
sample does not estimate statistical power. It shows what a design
calculation should investigate: which effect sizes and intervals could
this joint predictor distribution resolve? More bootstrap repetitions
cannot replace missing predictor variation. Probing baseline 60 in the
compressed sample would also be extrapolation, although it was supported
in the original sample.

### The research record for this analysis

| Element | Closing answer for the faculty-salary analysis |
|:--|:--|
| Question | Whether the service-salary association varies across academic ranks, rather than assuming one common slope or searching over possible products |
| Records and design | One faculty member per row, one U.S. college, one academic year; rank and service were observed, not assigned, and no broader sampling scheme is documented |
| Measurement | Nine-month salary and years of service are recorded administrative variables; they do not measure productivity, market conditions, or the mechanisms behind a salary structure |
| Estimand and support | Rank-specific fitted service slopes and their differences over each rank's own support; cross-rank contrasts are credible only inside the 1-to-6 year common window |
| Model and estimation | One hierarchical OLS model with two product columns, service centered at a supported value, and every reported quantity built as a linear combination of the coefficients |
| Uncertainty and checks | The additive restriction compared by a two-degree-of-freedom block test; classical, HC3 and case-bootstrap uncertainty for the same slopes; residual form, influence, curvature, support and scale examined and reported |
| Bounded claim | These records do not clearly distinguish three rank-specific slopes from one common linear association; the weak interaction evidence is a bounded association, not a moderator mechanism |

: Closing research record for the faculty-salary interaction analysis. {#tbl-ch10-record tbl-colwidths="[24,76]"}

The record includes the package datasets, complete-case rule, scales,
model formulas, anchors, probes and resampling unit. The generated example
uses seed 909 with 320 rows. The practice and salary bootstraps use
seeds 90910 and 90909 respectively, each with 2,000 whole-row resamples. Reproducibility recovers the numerical calculation;
the design record establishes which interpretation it can carry.

## Reading Bridge

A product term releases a common-slope restriction. Lower-order
coefficients are comparisons at stated anchors; centering and releveling
move the coordinates without changing the fitted surface. Conditional
slopes, group contrasts and differences of differences all use coefficient
covariance for their uncertainty. The block test examines additivity;
curves, intervals and support show what the comparison means.

The salary records illustrated an honest weak-interaction result and one
slope sensitive to the variance assumption. Chapter 11 changes the outcome
to binary and the mean scale to the logit. The conditional question
continues, but interaction on the linear-predictor and probability scales
will become different statements about the same fit.

### Key terms {.unnumbered}

| Term and common alternatives | Meaning or essential distinction |
|:--|:--|
| Product term (interaction term; multiplicative term) | A column formed by multiplying two predictors, so that each one's fitted comparison depends on the other's value. Its coefficient is a difference in slopes, not a slope, and nonadditivity is defined on the modelled scale: a nonlinear change of outcome scale can create or remove it. |
| Conditional slope (simple slope; conditional effect) | The fitted difference in outcome per one-unit difference in a focal predictor evaluated at a stated value or group of the other predictor. It is a linear combination of coefficients, so its SE uses their covariance, and it describes a conditional mean, not an individual case. |
| Moderator (moderating variable; moderation) | The variable across whose values or groups a named focal relationship is allowed to vary. The product-term algebra is symmetric in the two predictors, so the label comes from the research question; a moderation reading describes when a relationship differs, not how an effect is transmitted. |
| Hierarchy principle (marginality; well-formulated model) | The rule that a model containing a product also contains both component terms. Dropping a component forces the other's slope to zero at its coded zero and makes the fitted surface depend on the origin; hierarchy does not require that every possible product be added. |
| Centering a predictor (anchoring; reparameterization) | Replacing a predictor by its difference from a chosen value, so that coefficients anchored at zero become comparisons at that value. It changes the intercept and the lower-order coefficients and leaves the product coefficient, the fitted values, the residuals and the product-block test unchanged; it repairs no design, selection, or measurement problem. |
| Probe value (pick-a-point value) | A stated value of the moderator at which a conditional slope or contrast is evaluated, chosen from the measurement scale, a threshold, or a decision need rather than by an automatic mean-and-SD rule, and restricted to values the records support. |
| Johnson--Neyman boundary (region of significance) | A moderator value at which a pointwise interval for a conditional slope touches zero, found as a root of a quadratic whose leading coefficient decides which side is the region. A boundary outside the observed support describes the algebra and not the data, and a pointwise band is not a simultaneous band. |
| Interaction contrast (difference in differences within a model) | A linear combination of coefficients comparing two conditional comparisons, such as one group's difference minus another group's difference. Under reference coding it equals a product coefficient up to sign; its SE uses the coefficient covariance, and its causal reading depends on design, not on the algebra. |

: Concepts introduced here. {#tbl-ch10-terms tbl-colwidths="[30,70]"}

### Further reading {.unnumbered}

Aiken and West [@aiken1991] and Cohen and colleagues [@cohen2003]
develop centering and conditional comparisons. Bauer and Curran
[@bauer2005] and Preacher and colleagues [@preacher2006] explain probing
and significance boundaries. Fox and Weisberg [@fox2019] supply the data
and regression software route. The [emmeans interaction vignette](https://rvlenth.github.io/emmeans/articles/interactions.html)
connects conditional comparisons to implementation.

{{< pagebreak >}}

## Exercises {.unnumbered}

<!-- prompt-id: CH10-E01 -->

### Exercise 1: Read the graph before the table (Foundational) {.unnumbered}

A study
plots a continuous outcome against weekly practice hours for two program
groups. Over the observed range of 0 to 6 hours, the fitted Group A line
rises from 50 to 62 and the fitted Group B line rises from 58 to 76. The
confidence bands overlap at some practice values and the lines do not
cross. (1) Calculate each fitted slope in outcome units per hour and
their difference. (2) Give the fitted group gap at 0 hours and at 6
hours, and the difference between those two gaps. (3) Describe the
pattern as additive or nonadditive without using a $p$-value, and say
why crossing is not required. (4) A reader concludes from the
overlapping bands that the two slopes are the same. State what would
have to be estimated instead.

<!-- END exercise-id: CH10-E01 -->

<!-- prompt-id: CH10-E02 -->

### Exercise 2: A conditional slope needs the covariance (Foundational) {.unnumbered}

In a fitted product-term model the focal coefficient is 2.6972 and the
product coefficient is 0.1149. At centered moderator $w=10$, the full
covariance calculation gives a combined slope variance of 0.0379; the
residual degrees of freedom are 316 and $t_{.975,316}=1.9675$.
Here $w=10$ means ten units above the anchor. (1) Write the conditional
slope and its three-term variance formula, retaining the covariance;
calculate the slope at $w=10$. (2) Use the supplied combined variance
to calculate its SE and 95% interval. (3) A student omits the covariance
term and reports 0.1865; another adds the focal standard error to ten
times the product standard error and reports 0.2634. Say what each
number is and why neither is the standard error required here. (4)
To compare the slopes at $w=-10$ and $w=10$, which quantity should be estimated directly? Explain why subtracting the endpoints of their two separate intervals does not supply an interval for that difference.

<!-- END exercise-id: CH10-E02 -->

<!-- prompt-id: CH10-E03 -->

### Exercise 3: Translate a centered model (Foundational) {.unnumbered}

A model
fitted with the raw moderator has intercept 32.8571, focal coefficient
$-5.3448$, moderator coefficient 0.3822 and product coefficient 0.1149;
the raw focal coefficient has $t=-5.5131$. The moderator is a test score
and the analyst now centers it at 70. (1) Give the focal coefficient and
the intercept of the centered model. (2) State which of the following
change: the product coefficient, the fitted values, the residuals, the
product-block $F=68.3548$. (3) Explain what the raw focal coefficient's
small $p$-value is a test of, and why that does not make the raw
parameterization preferable. (4) A colleague says centering removed a
confounder. State what centering did and did not do.

<!-- END exercise-id: CH10-E03 -->

<!-- prompt-id: CH10-E04 -->

### Exercise 4: A two-column product block (Foundational) {.unnumbered}

On the same
397 records, a model with a common service slope for three academic
ranks has residual sum of squares 219014.23 on 393 degrees of freedom,
and the model adding two service-by-rank product columns has 218471.17
on 391. The 5% critical value of $F(2,391)$ is 3.0188. (1) Calculate
the block $F$ statistic. (2) State the decision at 5% and what it does
and does not say about the three rank-specific slopes. (3) Explain why
the result would be unchanged if the full professor rather than the
assistant rank were the reference, while the two individual product
$t$ tests would not be. (4) Name one quantity a reader needs before
accepting "the slopes are the same" as a summary.

<!-- END exercise-id: CH10-E04 -->

<!-- prompt-id: CH10-E05 -->

### Exercise 5: A difference of differences from cell means (Integrative) {.unnumbered}

In a conformity experiment the mean number of conforming responses is
8.9 with a low-status partner and 17.4 with a high-status partner
among low-authoritarianism participants; 7.2500 and 14.2727 among
medium-authoritarianism participants; and 12.6250 and 11.8571 among
high-authoritarianism participants. (1) Give the partner-status
difference in each authoritarianism category. (2) Give the difference of
differences comparing the low and high categories. (3) Its standard
error is 3.4507 on 39 degrees of freedom, with $t_{.975,39}=2.0227$;
give the 95% interval and the $t$ statistic. (4) The cell counts are
10, 5, 4, 11, 8 and 7 and the cell standard deviations range from 2.6437
to 7.3473. State one consequence for the reference distribution used in
(3), and one thing the contrast does not establish about partner status.

<!-- END exercise-id: CH10-E05 -->

<!-- prompt-id: CH10-E06 -->

### Exercise 6: Report a sensitive conditional slope (Integrative) {.unnumbered}

A
conditional slope is estimated as $-0.3947$ thousand dollars per year.
Three standard errors are available for it: 0.2949 from the classical
common-variance model, 0.1679 from an HC3 covariance, and 0.1736 from a
2,000-resample case bootstrap whose percentile interval is
[$-0.8066$, $-0.1436$]. Use $t_{.975,391}=1.966$ for the first two.
Residual standard deviations in the three groups are 8.0906, 13.2448 and
27.6842 against a pooled 23.6379. (1) Give the classical and HC3
intervals. (2) Say which intervals exclude zero and explain, using the
residual standard deviations, why the calculations disagree. (3) Write
one sentence reporting this slope for a reader, without choosing a
winner. (4) A student joins pointwise bootstrap intervals into a curve and claims 95% simultaneous coverage. Explain the error and name one study problem that resampling does not repair.

<!-- END exercise-id: CH10-E06 -->


{{< pagebreak >}}

## Appendix A: Conditional quantities as linear combinations {.unnumbered}

### One formula for four questions {.unnumbered}

A conditional slope, conditional mean, group contrast and interaction
contrast can all be written $\mathbf c{}^{\mathsf T}\boldsymbol\beta$. The vector $\mathbf c$ names the
question; the estimated variance is $\mathbf c{}^{\mathsf T}\widehat{\mathbf V}\mathbf c$. For coefficient order
$(\beta_0,\beta_X,\beta_W,\beta_{XW})$:

| Quantity | Contrast vector $\mathbf c{}^{\mathsf T}$ |
|:--|:--|
| Practice slope at $w$ | $(0,1,0,w)$ |
| Mean at $(x,w)$ | $(1,x,w,xw)$ |
| Mean difference $x_b-x_a$ at $w$ | $(0,x_b-x_a,0,w(x_b-x_a))$ |
| Slope difference at $w_b$ and $w_a$ | $(0,0,0,w_b-w_a)$ |

: Different questions use the same coefficient covariance. {#tbl-ch10-linear tbl-colwidths="[50,50]"}

At $w=10$ the first row gives 3.8461 with SE 0.1947 from full precision.
The faculty associate slope uses $\mathbf c{}^{\mathsf T}=(0,1,0,0,1,0)$ and gives $-0.3947$
with classical SE 0.2949. The uncertainty belongs to the whole comparison,
not a sum of separate coefficient SEs.

### Centering translates the coefficients {.unnumbered}

For $W_c=W-c$,

$$
\boldsymbol\beta^{\mathrm{new}}=\mathbf T\boldsymbol\beta,
\qquad T=\begin{pmatrix}1&0&c&0\\0&1&0&c\\0&0&1&0\\0&0&0&1\end{pmatrix},
\qquad \mathbf V^{\mathrm{new}}=\mathbf T\mathbf V\mathbf T{}^{\mathsf T}.
$$

The last row fixes the product coefficient. The first two translate the
anchor. With $c=70$, the full-precision raw estimates give intercept
59.6123 and practice coefficient 2.6972. A translated contrast uses the
translated covariance; its estimate and SE then agree with the original.

### The boundary quadratic {.unnumbered}

Writing $a=\widehat\beta_X$, $b=\widehat\beta_{XW}$,
$v_a=\widehat{\mathbf V}_{XX}$, $v_b=\widehat{\mathbf V}_{XW,XW}$ and
$v_{ab}=\widehat{\mathbf V}_{X,XW}$, the squared endpoint condition becomes

$$
Aw^2+Bw+C=0,\quad A=b^2-t^2v_b,\quad
B=2ab-2t^2v_{ab},\quad C=a^2-t^2v_a.
$$

For distinct real roots, the band excludes zero outside them if $A>0$
and inside if $A<0$. With $A=0$ the condition is linear; with no real
roots the sign is constant and must be checked at a supported value.
The generated fit has $A=0.0125$ and baseline roots 39.039 and 51.284.
Substitution into the quadratic gives numerical zero, and the pointwise
interval touches zero there. The lower root remains outside the data.

## Appendix B: What changes beyond this chapter {.unnumbered}

A three-way product $XWZ$ allows a two-way pattern to vary with a third
predictor. Hierarchy keeps all two-way products and all component terms.
Its interpretation requires conditional comparisons across three dimensions;
Chapter 4's three-way table gives the categorical reading.

With a nonlinear link, additivity on the linear-predictor scale and on
the response scale are different properties. Even without a product, a
logistic model can give probability differences that vary with another
predictor. Always name the scale.

A conditional indirect effect concerns mediation. Its causal interpretation
needs defensible temporal ordering, adequate functional form and assumptions
about common causes of the intervening variable and outcome. An interval
around an unidentified causal quantity leaves that problem in place.

Department, classroom or school dependence changes the appropriate
uncertainty calculation. Adding a product changes a conditional mean;
it does not make records independent.

**Extended practice.** (1) Refit the faculty model with log salary.
Explain how the comparison changes from dollars to proportional changes,
and why raw residual sums of squares or likelihood criteria on different
outcome scales cannot be compared without accounting for the transformation.
(2) Resample faculty within rank rather than across all records. Explain
which sampling story fixes the rank counts and which lets them vary.
Compare the associate slope intervals without treating the two schemes
as independent substantive evidence.

## References {.unnumbered}

::: {#refs}
:::
