---
title: "Basic Inferential Statistics with R"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-cap: true
    tbl-cap: true
    fig-align: center
    include-in-header:
      - file: ../supplement/pdf-graphics-compat.tex
      - text: |
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
if (knitr::is_latex_output()) knitr::opts_chunk$set(
    dev = "cairo_pdf")
options(digits = 4, width = 72)
library(ggplot2)
theme_set(theme_minimal(base_size = 11))
```


{{< pagebreak >}}

::: {.callout-tip title="Chapter box"}
**Central question:** How can we reproduce a basic estimate, confidence
interval or statistical test in R and explain its estimand, assumptions
and bounded interpretation?

**Focal concepts:** Sampling and working models; reference distributions;
SEs, intervals and tests for means, proportions and associations; paired
and independent comparisons; one-way ANOVA and focused comparisons.

**Main evidence:** Paired measurements for ten patients in `sleep` continue
Chapter 5.
`mtcars` and `ToothGrowth` provide separate basic-method examples, and a
short declared Normal simulation makes sampling variability visible.

**Scope:** Chapters 3-4 supply inferential foundations and Chapter 5 supplies
descriptive calculations and simple R. Chapter 1 supplies the reading
approach; Chapter 1a remains optional. Statistics organizes the route.
Chapter 7 develops regression; optional appendices retain technical depth.
:::

**Core route.** Sections 1--3 carry the paired `sleep` example from its
recorded differences to one inferential answer. Sections 4--5 then show how the
same question--estimand--procedure discipline changes for other basic methods.
Section 6 uses repeated sampling to check what an inferential procedure does,
and Section 7 completes the bounded answer and bridges to regression. This
route is complete without the appendices; they provide optional mathematical,
computational, and design extensions.

# From a Descriptive Result to an Inferential Question {#sec-ch6-question}

Chapter 5 described ten people's drug-2-minus-drug-1 differences in extra
sleep: a mean of 1.58 hours and an SD of 1.23 hours. The descriptive arithmetic
concerns those recorded people. An interval concerns an additional estimand under
a stated sampling or working model. A descriptive population question can
therefore require inference; these are analytical roles, not competing
question-mode labels.

## Recall the paired observations

The [sleep documentation](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/sleep.html)
defines extra sleep relative to control under two drugs. `ID` identifies
the person, `group` the condition and `extra` the hours. The conditions
are not baseline and follow-up. Recreate the ten aligned pairs locally:

```{r}
#| label: ch6-retain
#| echo: true
sleep_wide <- tidyr::pivot_wider(sleep, id_cols = ID,
  names_from = group, values_from = extra, names_prefix = "drug_")
complete_pair <- complete.cases(sleep_wide[c("drug_1", "drug_2")])
sleep_wide <- sleep_wide[complete_pair, ]
difference <- sleep_wide$drug_2 - sleep_wide$drug_1
c(n = length(difference), mean = mean(difference), SD = sd(difference))
```

All ten pairs are complete. Their mean is exactly calculable from the
records. The working inferential question is: **What do these differences
estimate about a process mean, if they behave as independent observations
from the stated process?** The file does not itself establish a sampling,
treatment-order or carryover protocol. Keep that provenance separate from
the model supporting a reference distribution.

## A declaration comes before a default

A formula such as `score ~ group` names an outcome and grouping variable.
It does not state whether observations are independent or paired. Begin
with the question and dependence structure, then choose the calculation.

| Question | Relevant quantity and calculation | Condition to establish |
|:--|:--|:--|
| Mean paired difference | Mean of within-person differences; paired or one-sample analysis | Correct pairs; independence between people for ordinary inference |
| Mean difference between separate groups | Difference of means; Welch comparison | Independent observations within and across groups |
| Fraction meeting a criterion | Proportion and suitable interval | Stated denominator and appropriate repeated-trial model |
| Association between categories | Counts, association measure and reference test | Independent units; adequate reference conditions |
| Differences across several groups | Group means, omnibus comparison or contrast | Stated design, variance and multiplicity conditions |

: Questions guide the calculation. {#tbl-ch6-selection}

This is an orientation, not an automatic selection algorithm. Sections 4
and 5 work through the additional procedures. Functions return structured
objects; their printed summaries are only one view. We will extract the
quantities relevant to a question after understanding the calculation.

# Probability and Reference Distributions {#sec-ch6-reference}

A probability calculation describes a stipulated model. It supplies tail
areas and critical values for an inferential procedure when that model is
appropriate. Use direct R calls and read their statistical question before
the function name. Sections 4-5 apply these references to additional records.

\Needspace{12\baselineskip}

## Four prefixes, four probability questions

| Prefix | Returned quantity | Example |
|:--|:--|:--|
| `d` | Discrete probability mass or continuous density height | `dbinom()`, `dnorm()`, `dt()` |
| `p` | Cumulative probability | `pbinom()`, `pnorm()`, `pt()` |
| `q` | Quantile at a cumulative probability | `qbinom()`, `qnorm()`, `qt()` |
| `r` | Random draws from the specified model | `rbinom()`, `rnorm()`, `rt()` |

For $X\sim\operatorname{Binomial}(20,.30)$, the model describes 20
independent trials with a common success probability .30. These are not
20 observations discovered in either additional dataset.

```{r}
#| label: ch6app-a-distributions
#| echo: true
mass_at_4 <- dbinom(4, size = 20, prob = .30)
cumulative_at_4 <- pbinom(4, size = 20, prob = .30)
percentile_90 <- qbinom(.90, size = 20, prob = .30)
c(mass = mass_at_4, cumulative = cumulative_at_4,
  percentile = percentile_90)
dnorm(0)
pnorm(1) - pnorm(-1)
qnorm(.975)
```

`dbinom(4, 20, .30)` is $P(X=4)$. For a continuous normal variable,
`dnorm(0)` is a density height, not $P(X=0)$; the latter is zero.
An interval probability is an area, here $\Phi(1)-\Phi(-1)\approx.6827$.
A continuous density can exceed one without assigning a probability above
one. The binomial quantile is the smallest integer whose cumulative
probability reaches the requested level; discrete quantiles need not attain
that probability exactly.

## Normal, t, chi-square, and F references

For an independent normal sample, replacing known $\sigma$ by its sample
estimate in $(\bar Y-\mu)/(\sigma/\sqrt n)$ changes the exact reference
from standard normal to $t_{n-1}$. The heavier t tails reflect this estimated
scale. This is not a universal cutoff saying that every sample above some
size is normal or independent.

```{r}
#| label: ch6app-a-reference-quantiles
#| echo: true
c(normal = qnorm(.975), t_5 = qt(.975, 5),
  t_9 = qt(.975, 9), t_30 = qt(.975, 30))
qchisq(.95, df = 4)
qf(.95, df1 = 2, df2 = 40)
pchisq(10, df = 4, lower.tail = FALSE)
pf(4, df1 = 2, df2 = 40, lower.tail = FALSE)
```

A chi-square variable is a sum of squared independent standard normals.
If $U\sim\chi^2_a$ and $V\sim\chi^2_b$ independently, then
$(U/a)/(V/b)\sim F_{a,b}$. This explains the two degrees of freedom of an
F reference. Pearson count discrepancies can have an approximate chi-square
reference; normal linear-model mean-square comparisons can have an exact F
reference. The same distribution name does not establish the same conditions
for every statistic using it.

```{r}
#| label: fig-ch6app-a-normal-t
#| echo: false
#| fig-cap: "Standard normal and t densities. The heavier t tails produce larger finite-df critical values."
#| fig-alt: "The t curve with five degrees of freedom is lower near zero and higher in the tails than the standard normal curve."
ref_x <- seq(-4, 4, length.out = 801)
ref <- rbind(
  data.frame(x = ref_x, density = dnorm(ref_x),
    Model = "Normal"),
  data.frame(x = ref_x, density = dt(ref_x, 5),
    Model = "t: 5 df")
)
ggplot2::ggplot(ref, ggplot2::aes(x, density,
    linetype = Model)) +
  ggplot2::geom_line(linewidth = .8) +
  ggplot2::labs(x = "Reference statistic", y = "Density",
    linetype = NULL)
```

# Estimate a Mean Difference and Its Uncertainty {#sec-ch6-uncertainty}

## Introduce a different question when uncertainty is needed

The ten-record mean is already known. For a **working-model question**,
suppose the differences behave as independent draws from a stable process
with mean $\mu_D$ and variance $\sigma_D^2$. What do these observations
estimate about that process mean?

This changes the estimand from the average of named records to a process
parameter. The short historical data description does not itself establish
random sampling, treatment order or absence of carryover. Keep those design
questions separate from the following conditional calculation.

For independent normally distributed differences, the one-sample $t$
interval is exact under the model. For other independent distributions it
is an approximation whose adequacy depends on shape and sample size.
Here $n=10$, so normality is a consequential assumption. Pairing accounts
for dependence within a person; independence concerns differences between
people. It is the differences, not both marginal condition distributions,
to which the normal model applies.

## Execute the arithmetic before reading output

The SD of differences is

$$
s_D=\sqrt{\frac{\sum_i(d_i-\bar d)^2}{n-1}}=1.2300.
$$

Under the independent-differences model,

$$
\widehat{SE}(\bar D)=\frac{s_D}{\sqrt n}
=\frac{1.2300}{\sqrt{10}}=0.3890\text{ hours}.
$$

The 95% interval uses $t_{.975,9}=2.2622$:

$$
\bar d\pm t_{.975,9}\frac{s_D}{\sqrt n}
=1.5800\pm2.2622(0.38896)
=[0.7001,\ 2.4599]\text{ hours}.
$$

```{r}
#| label: ch6-interval-arithmetic
#| echo: true
n <- length(difference)
estimate <- mean(difference)
s <- sd(difference)
se <- s / sqrt(n)
critical <- qt(.975, df = n - 1)
interval <- estimate + c(-1, 1) * critical * se
interval
```

The endpoint calculation uses unrounded values. For $H_0:\mu_D=0$,

$$
t=\frac{\bar d-0}{s_D/\sqrt n}
=\frac{1.58}{0.38896}=4.0621,\qquad df=9.
$$

The two-sided $p$ value is about .00283 under the stated null model. It is
the probability of a statistic at least as extreme under that model,
not the probability that the null is true. The interval concerns a mean
difference, not a range containing 95% of individual differences.

```{r}
#| label: ch6-paired-t
#| echo: true
result <- t.test(sleep_wide$drug_2, sleep_wide$drug_1,
                 paired = TRUE, mu = 0, conf.level = .95)
result$estimate
result$stderr
result$conf.int
result$statistic
result$parameter
result$p.value
```

R subtracts the second vector from the first: drug 2 minus drug 1 here.
`paired = TRUE` uses within-person differences, but does not match IDs
for you. The preceding pivot provided that alignment.

The one-sample call below gives the same estimate, interval and test:

```{r}
#| label: ch6-equivalence
#| echo: true
difference_result <- t.test(difference, mu = 0, conf.level = .95)
difference_result$estimate
difference_result$conf.int
```

This is a useful computational identity: a paired $t$ procedure is a
one-sample procedure on differences. Agreement verifies the calculation,
while its model and design still need their own justification.

## Direction, confidence level and test sidedness

Reversing the subtraction reverses the estimate and both interval endpoints;
it does not change the two-sided p-value. A higher confidence level uses a
larger critical value and gives a wider interval on the same data. The
alternative hypothesis must be specified to match the question before
examining the observed sign:

```{r}
#| label: ch6-sidedness
#| echo: true
t.test(difference, mu = 0, alternative = "greater")$p.value
t.test(difference, mu = 0, conf.level = .99)$conf.int
```

For a positive t statistic and a symmetric reference, the upper-tail
p-value is half the two-sided value. The lower-tail value is its complement.
Halving is inappropriate for an alternative opposite to the observed sign.
Nonrejection is not proof of equality; significance does not specify a
practical threshold or establish a causal explanation.

# Basic Procedures for Means and a Proportion {#sec-ch6-procedures}

## Built-in records and explicit questions

The later examples use data already installed with R. They are separate
illustrations, each with its own row unit and question. No supporting
script or download is needed.

The [official `mtcars` documentation](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html)
describes 32 automobile models from 1973–74. `mpg` is miles per US gallon,
`wt` is weight in thousands of pounds, and `am` codes automatic (0) or
manual (1) transmission. These observational records are not a probability
sample of all cars or a randomized transmission comparison.

```{r}
#| label: ch6-method-data
#| echo: true
cars <- mtcars
cars$transmission <- factor(cars$am, levels = c(0, 1),
                            labels = c("Automatic", "Manual"))
cars$mpg20 <- factor(cars$mpg >= 20,
                     levels = c(FALSE, TRUE),
                     labels = c("Below 20", "At least 20"))
mpg <- cars$mpg[!is.na(cars$mpg)]
head(cars[c("mpg", "wt", "transmission", "mpg20")])
c(observed = length(mpg), missing = sum(is.na(cars$mpg)))
```

The threshold 20 is a declared teaching definition, not a universal fuel
standard. All 32 cars have the variables used here. Keeping an explicit
missing-value check teaches how to retain the same denominator when another
dataset is incomplete.

The mean and proportions describe these records directly. An interval or
reference test adds a working process model. For example, a model of
independent car outcomes is an additional assumption; running a function
does not establish how the cars entered the file.

## Known-sigma z and one-sample t

For independent normal outcomes with process mean $\mu$ and externally
known SD $\sigma$, $SE(\bar Y)=\sigma/\sqrt n$ and
$z=(\bar y-\mu_0)/SE$. Stipulate $\sigma=5$ mpg and $\mu_0=20$ mpg
only for this hypothetical calculation:

```{r}
#| label: ch6-known-sd
#| echo: true
n_mpg <- length(mpg)
estimate_mpg <- mean(mpg)
z_se <- 5 / sqrt(n_mpg)
z <- (estimate_mpg - 20) / z_se
z_ci <- estimate_mpg + c(-1, 1) * qnorm(.975) * z_se
c(mean = estimate_mpg, SE = z_se, z = z)
z_ci
```

The recorded mean is 20.0906 mpg. The stipulated model gives SE .8839 and
interval [18.3582, 21.8230] mpg. A known SD must come from an appropriate
source; the observed file does not make 5 a known population SD.

When SD is estimated, replace $\sigma$ by the sample SD $s$ and use the
$t_{n-1}$ reference under the independent-normal model [@student1908].

```{r}
#| label: ch6-one-t
#| echo: true
mpg_se <- sd(mpg) / sqrt(n_mpg)
mpg_ci <- estimate_mpg + c(-1, 1) * qt(.975, n_mpg - 1) * mpg_se
one_t <- t.test(mpg, mu = 20)
c(mean = estimate_mpg, SD = sd(mpg), SE = mpg_se)
mpg_ci
one_t$statistic
one_t$p.value
```

Here $s=6.0269$, $SE=1.0654$, $t=.0851$, $df=31$, and the 95% interval
is [17.9177, 22.2636] mpg. The two-sided $p$ value is .9328. This interval
contains plausible process means under the model; it is not a range for
individual cars. Its inclusion of 20 does not prove equality to 20.

Exact $t$ calibration requires normal independent outcomes. Approximate
use beyond that model depends on distribution shape, finite variance,
influential values and the dependence structure, not a universal sample-size
cutoff. The observed mean itself needs no such model to describe these cars.

## Two independent means: Welch's comparison

For independent groups, define $\Delta=\mu_A-\mu_M$, here automatic minus
manual mean mpg. The estimate and its SE are

$$
\widehat\Delta=\bar y_A-\bar y_M,\qquad
SE=\sqrt{s_A^2/n_A+s_M^2/n_M}.
$$

Writing $a=s_A^2/n_A$ and $b=s_M^2/n_M$, the Welch degrees of freedom are

$$
\nu=\frac{(a+b)^2}{a^2/(n_A-1)+b^2/(n_M-1)}.
$$

```{r}
#| label: ch6-welch
#| echo: true
car_rows <- complete.cases(cars[c("mpg", "transmission")])
car_groups <- cars[car_rows, ]
auto <- car_groups$mpg[car_groups$transmission == "Automatic"]
manual <- car_groups$mpg[car_groups$transmission == "Manual"]
a <- var(auto) / length(auto)
b <- var(manual) / length(manual)
welch_se <- sqrt(a + b)
welch_df <- (a + b)^2 /
  (a^2 / (length(auto) - 1) + b^2 / (length(manual) - 1))
welch <- t.test(auto, manual)
c(n_auto = length(auto), n_manual = length(manual),
  auto_mean = mean(auto), manual_mean = mean(manual),
  difference = mean(auto) - mean(manual), SE = welch_se, df = welch_df)
welch$conf.int
welch$p.value
```

There are 19 automatic and 13 manual cars, with means 17.1474 and 24.3923
mpg. Their difference is $-7.2449$ mpg, SE 1.9232, with approximate
$df=18.3323$ and interval [$-11.2802,-3.2097$]. The two-sided $p$ value
is .0014. The sign records the comparison's direction.

Welch's reference is approximate even for unequal-variance normal groups
[@welch1947]. It is the default two-sample `t.test()` calculation.
`var.equal = TRUE` instead requests a pooled-variance procedure with
$n_A+n_M-2$ degrees of freedom. Neither option resolves dependence or
confounding: these car models also differ in weight, engine and other
features. The observed mpg gap does not isolate the effect of transmission.

The paired calculation in Section 3 instead uses within-person differences.
Select paired or independent analysis from the design, not from which
procedure gives a smaller p-value. Positive within-person covariance changes
the variance of a difference, as Chapter 5 demonstrated.

## One proportion: score test and Wilson interval

For an independent-Bernoulli working process with probability $p$, the
estimate is $\widehat p=s/n$. The score statistic for null $p_0$ is

$$
z=\frac{\widehat p-p_0}{\sqrt{p_0(1-p_0)/n}}.
$$

Inverting $|z(p_0)|\le z_{.975}$ gives the Wilson interval [@wilson1927]:

$$
\frac{\widehat p+z^2/(2n)}{1+z^2/n}
\ \pm\ \frac{z}{1+z^2/n}
\sqrt{\frac{\widehat p(1-\widehat p)}n+\frac{z^2}{4n^2}}.
$$

```{r}
#| label: ch6-proportion
#| echo: true
successes <- sum(cars$mpg >= 20, na.rm = TRUE)
trials <- sum(!is.na(cars$mpg))
phat <- successes / trials
p0 <- .5
score_z <- (phat - p0) / sqrt(p0 * (1 - p0) / trials)
zcrit <- qnorm(.975)
wilson_center <- (phat + zcrit^2 / (2 * trials)) /
  (1 + zcrit^2 / trials)
wilson_half <- zcrit / (1 + zcrit^2 / trials) *
  sqrt(phat * (1 - phat) / trials + zcrit^2 / (4 * trials^2))
wilson_ci <- wilson_center + c(-1, 1) * wilson_half
score <- prop.test(successes, trials, p = p0, correct = FALSE)
c(successes = successes, trials = trials, proportion = phat, z = score_z)
wilson_ci
score$statistic
score$p.value
```

Fourteen of 32 recorded cars meet the definition: $\widehat p=.4375$.
The score statistic is $z=-.7071$; with continuity correction disabled,
R reports $X^2=z^2=.5$, one df and $p=.4795$. The Wilson interval is
[.2817, .6067]. The default continuity correction changes the statistic
and interval, so declare the convention when reproducing a result.

This approximate reference requires an appropriate independent-trial model
and adequate expected successes and failures. The Bernoulli model and the
.50 reference are additional teaching assumptions. They do not turn the
32 recorded models into a random sample of all cars.

# Association and Several Group Means {#sec-ch6-comparisons}

Chapter 5 supplied descriptive tables, correlations and group summaries.
Here a reference model adds uncertainty to an explicitly stated question.

## Correlation inference uses a different process target

Recreate the complete weight-mpg pairs introduced descriptively in Chapter 5:

```{r}
#| label: ch6-correlation-data
#| echo: true
car_pairs <- cars[complete.cases(cars[c("wt", "mpg")]), ]
r <- cor(car_pairs$wt, car_pairs$mpg)
r
```

The recorded $r=-.8677$ describes these 32 pairs. For an additional process
correlation $\rho$, the zero-correlation statistic is

$$
t=r\sqrt{\frac{n-2}{1-r^2}},\qquad df=n-2.
$$

Its null reference is exact for independent bivariate-normal pairs. The
approximate Fisher interval transforms $z_F=\operatorname{atanh}(r)$,
uses $SE(z_F)=1/\sqrt{n-3}$ and transforms back:

$$
CI_\rho=\tanh\!\left(z_F\pm\frac{1.96}{\sqrt{n-3}}\right).
$$

```{r}
#| label: ch6-correlation-test
#| echo: true
n_r <- nrow(car_pairs)
fisher_ci <- tanh(atanh(r) + c(-1, 1) * qnorm(.975) / sqrt(n_r - 3))
cor_test <- cor.test(car_pairs$wt, car_pairs$mpg)
fisher_ci
cor_test$statistic
cor_test$p.value
```

The interval is [$-.9338,-.7441$], $t=-9.5590$ with 30 df, and
$p\approx1.294\times10^{-10}$. These precise calculations do not verify
bivariate normality, independent sampling or representation of all cars.
When computing a whole covariance matrix, use a common row set: different
pairwise missing-value subsets need not form a coherent joint covariance
structure.

## Categorical association: counts, residuals, and magnitude

Does the fraction with at least 20 mpg differ between automatic and manual
cars in this file? Fix the observed rows before counting.

```{r}
#| label: ch6-categorical
#| echo: true
categorical <- cars[complete.cases(cars[c("transmission", "mpg20")]), ]
counts <- table(categorical$transmission, categorical$mpg20)
counts
rowSums(counts)
prop.table(counts, margin = 1)
chi_result <- chisq.test(counts, correct = FALSE)
chi_result$expected
chi_result$residuals
chi_result$stdres
chi_result$statistic
chi_result$p.value
cramers_v <- sqrt(as.numeric(chi_result$statistic) / sum(counts))
cramers_v
```

Four of 19 automatic cars and ten of 13 manual cars meet the threshold:
.2105 versus .7692. Under independence, the expected first-row/below-20
count is $19(18)/32=10.6875$, compared with 15 observed.

For cell counts $O_{jk}$, the fitted independence count, Pearson statistic,
degrees of freedom and Cramér's magnitude measure are

$$
\begin{aligned}
E_{jk}&=\frac{O_{j+}O_{+k}}N, &
X^2&=\sum_{jk}\frac{(O_{jk}-E_{jk})^2}{E_{jk}},\\[2pt]
df&=(J-1)(K-1), &
V&=\sqrt{\frac{X^2}{N\min(J-1,K-1)}}.
\end{aligned}
$$

Here $X^2=9.7906$, one df, $p=.0018$ and $V=.5531$ [@cramer1946].
Pearson residuals are $(O-E)/\sqrt E$. Standardized residuals additionally
account for fitting the margins; they locate departures but are not a set
of independent confirmatory tests. Row proportions retain the denominators
needed to interpret the association.

The chi-square approximation needs independent units and adequate expected
counts; the smallest expected cell here is 5.6875. The default two-by-two
continuity correction is disabled explicitly. Fisher's alternative uses a
conditional distribution given the margins under its null [@fisher1935]:

```{r}
#| label: ch6-fisher
#| echo: true
fisher <- fisher.test(counts)
fisher$p.value
```

The Fisher $p$ value is .0033. Conditioning must fit the sampling or
assignment question; Fisher's test is not a universal repair for dependence,
selection or measurement problems. Neither result establishes that changing
a car's transmission would cause it to cross the threshold.

## Raw group means and one-way ANOVA

The [official `ToothGrowth` documentation](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html)
describes measurements from 60 guinea pigs. Each animal received one of
three vitamin-C doses, .5, 1 or 2 mg/day, delivered as orange juice (`OJ`)
or ascorbic acid (`VC`). `len` records odontoblast length; the help page does
not specify a length unit, so we report the recorded length scale. This is
not a human outcome or a measured before/after change. The short help page
does not itself document a randomization protocol.

```{r}
#| label: ch6-growth-data
#| echo: true
growth <- ToothGrowth
growth$dose <- factor(growth$dose, levels = c(.5, 1, 2))
growth$supp <- factor(growth$supp, levels = c("OJ", "VC"))
table(growth$dose, growth$supp)
dose_n <- table(growth$dose)
dose_mean <- tapply(growth$len, growth$dose, mean)
dose_sd <- tapply(growth$len, growth$dose, sd)
data.frame(dose = names(dose_n), n = as.integer(dose_n),
           mean = dose_mean, SD = dose_sd)
```

There are 20 animals per dose and ten per dose–supplement cell. The three
raw dose means are 10.605, 19.735 and 26.100. Each averages equally across
the two delivery methods. Treating `dose` as a factor fits separate means;
it does not impose a linear dose–response relation.

```{r}
#| label: fig-ch6-dose
#| echo: false
#| fig-pos: H
#| fig-cap: "Individual lengths and distributions for the three dose groups."
#| fig-alt: "Boxplots and individual observations show progressively higher recorded lengths at doses 0.5, 1 and 2 mg/day."
ggplot(growth, aes(dose, len)) + geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitter(width = .08, height = 0, seed = 606)) +
  labs(x = "Vitamin-C dose (mg/day)", y = "Recorded odontoblast length")
```

For $k$ groups and $N$ records,

$$
SS_B=\sum_j n_j(\bar y_j-\bar y)^2,\quad
SS_E=\sum_j\sum_i(y_{ij}-\bar y_j)^2,\quad
F=\frac{SS_B/(k-1)}{SS_E/(N-k)}.
$$

The omnibus null is equality of the three process means. Its exact normal
reference requires independent errors, normal outcomes within groups and
one common error variance. It does not require a pooled histogram to be
normal.

```{r}
#| label: ch6-dose-anova
#| echo: true
dose_aov <- aov(len ~ dose, data = growth)
summary(dose_aov)
dose_lm <- lm(len ~ dose, data = growth)
SSB <- sum(dose_n * (dose_mean - mean(growth$len))^2)
SSE <- sum(resid(dose_lm)^2)
MSE <- SSE / df.residual(dose_lm)
c(SSB = SSB, SSE = SSE, MSE = MSE)
anova(dose_lm)
oneway.test(len ~ dose, data = growth)
```

Here $F=(2426.4343/2)/(1025.775/57)=67.4157$, with $p<.001$ under
that model. Welch's one-way extension gives $F=68.4010$ with approximate
denominator df 37.7433; it relaxes common variance through an approximate
reference. Neither omnibus result selects the scientifically important
contrast or validates the treatment protocol. Omitted delivery-method
variation is one reason to examine the adjusted extension in Appendix C.

## Magnitude, planned contrasts, and multiple comparisons

The one-way sample variation measures are

$$
\eta^2=\frac{SS_B}{SS_T},\qquad
\widehat\omega^2=\frac{SS_B-(k-1)MSE}{SS_T+MSE},\qquad
SS_T=SS_B+SS_E.
$$

The omega estimate can be negative through sampling variation; report any
zero-truncation convention explicitly. Neither measure is automatically a
causal proportion of an outcome or a substantive importance threshold.

```{r}
#| label: ch6-magnitude
#| echo: true
SST <- SSB + SSE
eta <- SSB / SST
omega <- (SSB - 2 * MSE) / (SST + MSE)
c(eta_squared = eta, omega_squared = omega)
```

Here $\eta^2=.7029$ and $\widehat\omega^2=.6888$.

Declare a focused dose contrast before examining its results:
the 2 mg/day mean minus the equally weighted average of the .5 and 1 mg/day
means. Its weights are $\mathbf c=(-.5,-.5,1)$.

$$
\widehat L=\sum_j c_j\bar y_j,\quad
SE(\widehat L)=\sqrt{MSE\sum_jc_j^2/n_j},\quad
CI=\widehat L\pm t_{.975,N-k}SE(\widehat L).
$$

```{r}
#| label: ch6-contrast
#| echo: true
weights <- c(-.5, -.5, 1)
contrast <- sum(weights * dose_mean)
contrast_se <- sqrt(MSE * sum(weights^2 / dose_n))
contrast_ci <- contrast + c(-1, 1) *
  qt(.975, df.residual(dose_lm)) * contrast_se
c(contrast = contrast, SE = contrast_se)
contrast_ci
TukeyHSD(dose_aov, "dose")
```

The contrast is $26.100-(10.605+19.735)/2=10.930$, SE 1.1618, with
interval [8.6036, 13.2564] on the recorded length scale. Equal group sizes
make the chosen average equal the pooled mean of the two lower-dose groups
here; those targets would differ if their sizes were unequal.

Tukey's output answers a different family: all three pairwise differences.
Their estimates are 9.130, 15.495 and 6.365; the simultaneous intervals are
[5.9018, 12.3582], [12.2668, 18.7232] and [3.1368, 9.5932]. Exact familywise
coverage applies in the balanced normal common-variance case; `TukeyHSD()`
uses an unequal-size adjustment otherwise. It does not account for searching
across other outcomes or models. A numerical printed adjusted $p$ value of
zero is rounding, not an event with literally zero probability.

Keep the declared contrast and exploratory pairwise family distinct. Their
positive findings concern these model-based comparisons; practical and
causal interpretation also need the substantive measurement and design record.

# Check Inference Through Repeated Sampling {#sec-ch6-repetition}

## Check decisions, not only whether the function ran

Inspect the IDs, measurement order and variables used to identify complete
pairs. A wrong pairing changes the calculation. An unsupported independence
assumption changes whether the SE has its proposed interpretation. A missing
measurement restricts which people the description includes. Repeating a
command cannot answer all three questions.

For `sleep`, the table establishes ten complete recorded pairs, and the
arithmetic establishes their mean difference. The independent normal model
supplies a conditional reference for uncertainty. The computation does not
verify an unrecorded sampling or treatment protocol.

## Simulate a known process

A simulation lets us choose the process and inspect its consequences.
Generate independent normal differences with **known mean 1.5 hours** and
**SD 1.2 hours**. Each generated study has $n=10$; repeat it $B=2000$ times.
These are teaching constants, not estimated features or a reconstruction of
the historical `sleep` study. The true mean is known in this experiment,
so we can ask whether each generated interval contains it.

First follow one generated study: draw ten values, estimate its mean and
SE, and form one interval. Check whether the declared mean 1.5 is inside.

```{r}
#| label: ch6-one-generated-study
#| echo: true
set.seed(60610)
one_sample <- rnorm(10, mean = 1.5, sd = 1.2)
one_mean <- mean(one_sample)
one_se <- sd(one_sample) / sqrt(10)
one_interval <- one_mean + c(-1, 1) * qt(.975, 9) * one_se
one_interval
1.5 >= one_interval[1] & 1.5 <= one_interval[2]
```

The loop below repeats those same steps for fresh samples. Reset the seed
before the full run so that its results do not depend on whether the
one-study illustration was run first.

```{r}
#| label: ch6-simulation
#| echo: true
set.seed(60610)
mu <- 1.5
sigma <- 1.2
n_sim <- 10
B <- 2000
sample_means <- numeric(B)
covered <- logical(B)
critical_sim <- qt(.975, df = n_sim - 1)
for (b in seq_len(B)) {
  sample_d <- rnorm(n_sim, mean = mu, sd = sigma)
  sample_means[b] <- mean(sample_d)
  sample_se <- sd(sample_d) / sqrt(n_sim)
  margin <- critical_sim * sample_se
  covered[b] <- abs(sample_means[b] - mu) <= margin
}
c(mean_of_means = mean(sample_means), empirical_SE = sd(sample_means),
  theoretical_SE = sigma / sqrt(n_sim), coverage = mean(covered))
```

`numeric(B)` and `logical(B)` reserve one result per repetition. The loop
draws a fresh sample, estimates its mean and SE, constructs its interval,
then records whether it covers the known mean. A logical `TRUE` counts as
one in `mean(covered)`. Repeating a calculation on one unchanged sample
would not produce a sampling distribution.

```{r}
#| label: fig-ch6-repetition
#| fig-cap: "Sample means from the declared independent normal-differences process."
#| fig-alt: "Simulated means cluster around the known value 1.5 hours and vary much less than the individual differences."
#| fig-width: 6.2
#| fig-height: 3.4
ggplot(data.frame(sample_means), aes(sample_means)) +
  geom_histogram(bins = 25, fill = "grey80", colour = "white") +
  geom_vline(xintercept = mu, linetype = 2) +
  labs(x = "Mean difference in a generated study (hours)",
       y = "Repetitions")
```

The distribution of 2,000 means approximates the **sampling distribution**
under the chosen process. Its theoretical SE is $1.2/\sqrt{10}=.3795$
hours. Normality makes the usual $t$ interval exact for this generated
process. The observed coverage is `r sprintf("%.4f", mean(covered))`;
finite simulations need not reproduce .9500 exactly.

Increasing $n$ from 10 to 40 halves the theoretical SE to
$1.2/\sqrt{40}=.1897$ hours. Increasing $B$ leaves each study's information
unchanged; it improves the numerical approximation from the simulation.

\Needspace{16\baselineskip}

## Separate three kinds of variability

| Quantity | What varies? | Illustrative scale |
|:--|:--|:--|
| Individual-difference SD | Observations within a generated study | 1.2 hours |
| SE of the mean | Means across generated studies of size 10 | $1.2/\sqrt{10}=.3795$ hours |
| Monte Carlo SE of coverage | Approximation from a finite number of simulations | About .0049 in probability units at $B=2000$ |

: Study variation and simulation approximation are different. {#tbl-ch6-variability}

If true coverage is $p=.95$, its Monte Carlo SE is

$$
MCSE=\sqrt{\frac{p(1-p)}B}
=\sqrt{\frac{.95(.05)}{2000}}=.0049.
$$

For an unknown simulated probability, substitute the observed coverage.
Increasing $B$ from 2,000 to 8,000 halves this numerical SE, not the study
mean's SE. The average simulated mean also has Monte Carlo uncertainty:
about $.3795/\sqrt{2000}=.0085$ hours. A seed makes results repeatable;
it does not make generating assumptions true for an actual study.
Appendix A retains exact probabilities, skewed processes and further
sample-size experiments.

# Read the Answer and Prepare for Regression {#sec-ch6-reading}

## A compact computational record

Read the calculation and its stated reasoning before evaluating a stronger
claim. A concise report is:

> The ten complete person pairs had a mean recorded drug-2-minus-drug-1
> difference of 1.58 hours of extra sleep, with SD 1.23 hours. Under a
> separately stated independent normal-differences model, the paired $t$
> procedure gave SE .389 hours and a 95% interval [.700, 2.460] hours for
> a process mean. The calculation does not itself establish the study's
> sampling, treatment-order or carryover conditions.

The inclusion code identifies the records, the dictionary explains the
variables, and the returned object supplies the numerical quantities.
The model declaration offers a reason for using the $t$ reference; locating
that offered reason is separate from deciding whether it is adequate for
the study. No full component map is needed for this computational lesson.

**Separate quality check.** “Drug 2 causes everyone to sleep 1.58 hours
longer” changes an observed mean into a universal causal effect. It fails
**Question alignment** and **Claim calibration**. The recorded mean remains
checkable, but a population or causal answer requires a corresponding
question and additional design support. Positive interval endpoints alone
cannot supply that support.

## A short bridge from a mean to a fitted line

For 2, 4 and 6, squared deviations total 8 around their mean of 4 and
11 around either 3 or 5. The mean minimizes the sum of squared deviations.
With group A values 2 and 4 and group B values 6 and 10, the group means
are 3 and 8. A fitted expression $3+5D$, where $D=0$ for A and 1 for B,
reproduces those means. The coefficient 5 is their observed difference.

These identities connect simple descriptive calculations to regression.
They do not require Normal observations to hold; uncertainty for a model
coefficient adds assumptions. Chapter 7 develops least squares, slope
uncertainty and diagnostics, and Chapter 8 develops indicators, adjustment
and coordinated prediction questions. Appendix B retains optional criterion
and optimization details; Appendix C retains adjusted and factorial examples.

## Summary and Reading Bridge {#sec-ch6-summary}

Name the estimand, relevant observations and dependence structure, then
calculate an estimate, SE, interval or test under its stated reference.
Read a returned number in that role. The same mean can be a fixed-record
description or an estimate of a process parameter under an additional model.

The core methods are direct R calculations of familiar statistics. Their
assumptions and interpretation are part of the analytical result. A sampling
simulation illustrates a chosen process; it does not validate that process
for historical records. Increasing sample size and increasing simulation
repetitions improve different quantities.

The [R t-test reference](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/t.test.html)
and local help such as `?prop.test`, `?cor.test` and `?aov` document the
returned procedures. Chapter 7 extends the descriptive/inferential distinction
to regression and Chapter 8 to multiple questions and model comparisons.


\Needspace{8\baselineskip}
\phantomsection

# Exercises {.unnumbered}

Use the supplied summaries and reference values. Keep intermediate values
unrounded; report SEs and intervals to three decimals and probabilities
to four unless the question specifies otherwise.

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH06-E01 -->

## Exercise 1: Preserve the pairs for inference (Foundational) {.unnumbered #ex-ch06-01}

Four students' baseline scores are 40, 50, 60 and 70. Their corresponding
follow-up scores are 46, 54, missing and 75. Calculate the complete-pair
mean change and its sample SD. Under an independent normal-differences
working model, calculate its estimated SE. Explain which observations
the model concerns and why treating the measurements as independent groups
would be a different calculation.

<!-- END EX06-01 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH06-E02 -->

## Exercise 2: Reproduce an interval (Foundational) {.unnumbered #ex-ch06-02}

Twenty-five independent paired differences have mean 3.0 and SD 5.0.
Under the normal-difference working model, use $t_{.975,24}=2.064$.
Calculate the estimated SE, the 95% interval and the standardized statistic
for a zero process mean. Explain what the interval concerns and why it is
not a range for individual differences.

<!-- END EX06-02 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH06-E03 -->

## Exercise 3: Match the procedure and output to the question (Foundational) {.unnumbered #ex-ch06-03}

Study A measures the same 20 people twice. Study B compares two independent
groups of 20 and 25 people without an equal-variance assumption. Name the
appropriate mean-comparison procedures. A paired analysis returns estimate
2.4, SE 0.8, 95% interval [0.75,4.05] and $t=3.0$. Explain each quantity.
Name two additional pieces of information needed to assess a claim that
the result is practically important and causal.

<!-- END EX06-03 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH06-E04 -->

## Exercise 4: More students or more simulations? (Foundational) {.unnumbered #ex-ch06-04}

A simulation uses independent observations with SD 12, sample size 36
and 1,000 repetitions. Calculate the theoretical SE of a sample mean.
If coverage is .95, calculate the Monte Carlo SE of the estimated coverage.
What changes if sample size becomes 144? What changes if only the number
of repetitions becomes 4,000? Report probabilities to four decimal places.

<!-- END EX06-04 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH06-E05 -->

## Exercise 5: Read a categorical comparison (Integrative) {.unnumbered #ex-ch06-05}

Among 20 people in group A, 12 answer Yes and 8 No; among 20 in group B,
6 answer Yes and 14 No. Calculate the A-minus-B percentage-point difference
and the four expected counts under independence. An uncorrected Pearson
test reports $X^2=3.6364$, $df=1$ and $p=.0565$. State the 5% decision
and evaluate the statement "the groups have been proved identical."

<!-- END EX06-05 PROMPT -->

\Needspace{6\baselineskip}
\phantomsection

<!-- prompt-id: CH06-E06 -->

## Exercise 6: Separate an omnibus result from specific comparisons (Integrative) {.unnumbered #ex-ch06-06}

A three-group ANOVA under an independent Normal common-variance model
reports $F(2,27)=4.50$ and $p=.0206$. Two prespecified comparisons are
B minus A and C minus A. Their simultaneous 95% intervals are [.2,3.8]
and [-.1,6.1]. State what the omnibus result establishes at 5%. Does the
evidence meet a rule requiring both comparisons to be positive? Explain
why the second interval does not prove a zero difference.

<!-- END EX06-06 PROMPT -->

\Needspace{8\baselineskip}
\phantomsection

# Appendix A: Simulation Extensions {.unnumbered #sec-ch6-app-a}

These optional examples deepen the distinction between study uncertainty and
Monte Carlo approximation. Section 6 already supplies the chapter's essential
simulation lesson.

\Needspace{6\baselineskip}
\phantomsection

## Exact probabilities and Monte Carlo error {.unnumbered}

Simulation approximates $p=P(X\le4)$ under the binomial model. Let
$I_b=1$ when draw $b$ meets that event and 0 otherwise. The Monte Carlo
estimate is $\widehat p_{MC}=B^{-1}\sum_b I_b$ with estimated numerical SE
$\sqrt{\widehat p_{MC}(1-\widehat p_{MC})/B}$.

```{r}
#| label: ch6app-a-binomial-simulation
#| echo: true
set.seed(6062)
B_binomial <- 20000
binomial <- rbinom(B_binomial, size = 20, prob = .30)
exact <- pbinom(4, size = 20, prob = .30)
mc <- mean(binomial <= 4)
mcse <- sqrt(mc * (1 - mc) / B_binomial)
```

```{=latex}
\Needspace{9\baselineskip}
```

```{r}
#| label: ch6app-a-binomial-simulation-display-1
#| echo: false
knitr::kable(data.frame(
  Method = c("Exact model probability", "Monte Carlo approximation"),
  Probability = c(exact, mc),
  MC_SE = c(NA_real_, mcse)
), digits = 5,
  caption = paste(
    "A simulated probability and its numerical uncertainty."
  )
)
```


The exact probability is `r sprintf("%.5f", exact)`; the Monte Carlo
estimate is `r sprintf("%.5f", mc)`, with MC SE
`r sprintf("%.5f", mcse)`. At $B=20{,}000$, the substitution is
$\sqrt{\widehat p_{MC}(1-\widehat p_{MC})/20{,}000}$.
Quadrupling $B$ approximately halves this numerical SE. It does not improve
the study's sample size, measurement, assignment, or the truth of the
binomial assumptions.

\Needspace{6\baselineskip}
\phantomsection

## Helpers, `replicate()`, and sample-size comparisons {.unnumbered}

A function can expose the sample size, repetition count, and generating rule
as separate arguments. `replicate()` evaluates its expression repeatedly;
the generator is called anew on every repetition, rather than taking the
mean of one sample repeatedly.

```{r}
#| label: ch6app-a-sample-means
#| echo: true
simulate_means <- function(sample_size, repetitions,
    generator) {
  replicate(repetitions, mean(generator(sample_size)))
}
set.seed(6063)
sizes <- c(5, 20, 80)
sampling <- do.call(rbind, lapply(sizes,
    function(n_now) {
  data.frame(
    n = n_now,
    sample_mean = simulate_means(
      n_now, 5000, function(n) rlnorm(n, meanlog = 0, sdlog = .8)
    )
  )
}))
population_mean <- exp(.8^2 / 2)
population_sd <- sqrt((exp(.8^2) - 1) * exp(.8^2))
sampling_summary <- do.call(rbind, lapply(sizes,
    function(n_now) {
  estimates <- sampling$sample_mean[
    sampling$n == n_now]
  data.frame(n = n_now, Mean_estimate = mean(estimates),
             Empirical_SD = sd(estimates),
             Theoretical_SD = population_sd / sqrt(n_now),
             MC_SE_of_mean_estimate = sd(estimates) / sqrt(
    length(estimates)))
}))
```

```{r}
#| label: ch6app-a-sample-means-display-1
#| echo: false
knitr::kable(sampling_summary, digits = 4,
  col.names = c("n", "Mean estimate", "Empirical SD", "Theoretical SD", "MC SE"),
               caption = paste(
    "Observation variability, estimator variability, and",
                 "Monte Carlo error."
               )
             )
```


The programmed lognormal population has mean $e^{.8^2/2}$ and variance
$(e^{.8^2}-1)e^{.8^2}$. Its observation SD stays the same in all three
conditions. The SD of its sample mean is $\sigma/\sqrt n$, so moving from
$n=5$ to 20 halves it. The last column instead measures numerical precision
in the average of 5,000 simulated estimates. Increasing $B$ improves that
Monte Carlo approximation without making any individual simulated study
more informative.

```{r}
#| label: fig-ch6app-a-sampling
#| echo: false
#| fig-cap: "Sample means from the same right-skewed lognormal process at three sample sizes."
#| fig-alt: "Histograms at sample sizes five, twenty, and eighty become progressively narrower and more symmetric, although the generating observation distribution is unchanged."
ggplot2::ggplot(sampling, ggplot2::aes(sample_mean)) +
  ggplot2::geom_histogram(bins = 35, fill = "grey55",
    colour = "white") +
  ggplot2::facet_wrap(~n, ncol = 1, labeller = ggplot2::label_both) +
  ggplot2::labs(x = "Sample mean", y = "Repetitions")
```

\clearpage

\Needspace{8\baselineskip}
\phantomsection

# Appendix B: Estimation, Optimization, and Likelihood {.unnumbered #sec-ch6-app-b}

\Needspace{6\baselineskip}
\phantomsection

## A criterion defines the estimate {.unnumbered}

For values $y_i$, squared loss at candidate center $m$ is
$L(m)=\sum_i(y_i-m)^2$. Expanding around the mean gives

$$
L(m)=\sum_i(y_i-\bar y)^2+n(m-\bar y)^2.
$$

The cross term vanishes because $\sum_i(y_i-\bar y)=0$. Thus $\bar y$ is
the unique minimizer when $n>0$. Equivalently, $L'(m)=2n(m-\bar y)$ and
$L''(m)=2n>0$. This identity does not require normality or sampling.

```{r}
#| label: ch6-loss-optimizer
#| echo: true
mpg <- mtcars$mpg
squared_loss <- function(m, y) sum((y - m)^2)
mean_fit <- optimize(squared_loss, interval = range(mpg), y = mpg)
c(numerical = mean_fit$minimum, arithmetic = mean(mpg),
  minimum_loss = mean_fit$objective)
```

The numerical minimum is 20.0906 mpg, matching the arithmetic mean; loss
is 1126.0472 squared mpg. Moving one mpg away raises it by $32(1)^2=32$.
The search interval must contain the relevant minimum. An optimizer cannot
find a value excluded by its constraints.

```{r}
#| label: fig-ch6-loss
#| echo: false
#| fig-pos: H
#| fig-cap: "Squared loss is minimized at the recorded mean mpg."
#| fig-alt: "A convex squared-loss curve has its minimum at mean mpg 20.09."
loss_grid <- data.frame(m = seq(10, 35, length.out = 100))
loss_grid$loss <- sapply(loss_grid$m, squared_loss, y = mpg)
ggplot(loss_grid, aes(m, loss)) + geom_line() +
  geom_vline(xintercept = mean(mpg), linetype = 2) +
  labs(x = "Candidate mean (mpg)", y = "Sum of squared deviations")
```

Chapter 7 extends squared loss to a fitted line. The criterion defines
what the estimate optimizes; it does not establish representation or causality.

\Needspace{6\baselineskip}
\phantomsection

## A binomial likelihood and boundary solutions {.unnumbered}

For $s$ successes in $n$ independent Bernoulli trials with probability $p$,

$$
L(p)={n\choose s}p^s(1-p)^{n-s},\qquad
\ell(p)=\text{constant}+s\log p+(n-s)\log(1-p).
$$

For $0<s<n$, setting $s/p-(n-s)/(1-p)=0$ gives $\widehat p=s/n$.
Use 14 of 32 cars reaching 20 mpg as an illustration of this separately
stipulated likelihood. Counting those outcomes does not verify independent
Bernoulli sampling.

```{r}
#| label: ch6-binomial-likelihood
#| echo: true
successes <- sum(mtcars$mpg >= 20)
trials <- nrow(mtcars)
binomial_nll <- function(p, successes, trials) {
  -dbinom(successes, size = trials, prob = p, log = TRUE)
}
prop_fit <- optimize(binomial_nll, interval = c(0, 1),
                     successes = successes, trials = trials)
c(numerical = prop_fit$minimum, proportion = successes / trials)
```

Both values are .4375. The combinatorial term is included by `dbinom()`
but does not affect which $p$ maximizes the likelihood. If $s=0$ or $s=n$,
the maximum is at 0 or 1. Check endpoints explicitly in such cases; an
interior-only search may omit the correct boundary solution. `log = TRUE`
avoids taking the log of a probability that may already have underflowed.

\Needspace{6\baselineskip}
\phantomsection

## Multiple parameters and convergence checks {.unnumbered}

Under a hypothetical independent normal model,

$$
-\ell(\mu,\sigma)=\frac n2\log(2\pi)+n\log\sigma+
\frac{\sum_i(y_i-\mu)^2}{2\sigma^2}.
$$

Parameterizing $\sigma=e^\eta$ lets an unconstrained optimizer search in
$\eta$ while keeping scale positive. This illustrates likelihood fitting;
it is not a claim that the recorded car outcomes follow that exact model.

```{r}
#| label: ch6-normal-likelihood
#| echo: true
normal_nll <- function(par, y) {
  mu <- par[1]
  sigma <- exp(par[2])
  -sum(dnorm(y, mean = mu, sd = sigma, log = TRUE))
}
normal_fit <- optim(c(mean(mpg) - 1, log(sd(mpg))),
                    normal_nll, y = mpg, method = "BFGS", hessian = TRUE)
second_fit <- optim(c(mean(mpg) + 1, log(sd(mpg) / 2)),
                    normal_nll, y = mpg, method = "BFGS")
c(numerical_mean = normal_fit$par[1],
  numerical_scale = exp(normal_fit$par[2]),
  arithmetic_mean = mean(mpg),
  closed_form_scale = sqrt(mean((mpg - mean(mpg))^2)))
normal_fit$convergence
normal_fit$value - second_fit$value
eigen(normal_fit$hessian)$values
```

The mean is approximately 20.0906 and normal MLE scale 5.9320. The latter
divides squared deviations by $n$, whereas `sd()` divides by $n-1$.
The fits return convergence code zero and nearly identical objective values.
A zero code reports an algorithm's stopping criterion. Compare starting
values, the returned objective and an analytic answer where available.
Positive Hessian eigenvalues support local curvature, not a universal proof
of global identification or scientific model adequacy.

\Needspace{6\baselineskip}
\phantomsection

## Extended practice: an estimand and its criterion (Integrative) {.unnumbered}

Plot squared loss for a quantitative variable and compare its minimum with
the arithmetic mean. Deliberately narrow the search interval to exclude the
mean, then explain why the numerical result changes. For binomial counts,
compare an interior case with zero successes and all successes, checking the
boundary likelihoods. Distinguish the scientific question, chosen criterion
and numerical search.

\Needspace{8\baselineskip}
\phantomsection

# Appendix C: Adjusted and Factorial Model Extensions {.unnumbered #sec-ch6-app-c}

These optional extensions develop the group comparisons in Section 5.
They are useful when adjustment or effect variation belongs to the question;
they are not prerequisites for the chapter's main paired-analysis route.

\Needspace{6\baselineskip}
\phantomsection

## Indicator regression reproduces the same fitted group means {.unnumbered}

With .5 mg/day as reference, the intercept is its mean; the other two
coefficients are the 1-minus-.5 and 2-minus-.5 differences. Adding each
coefficient to the intercept reconstructs its group mean.

```{r}
#| label: ch6-indicator-equivalence
#| echo: true
coef(dose_lm)
dose_levels <- levels(growth$dose)
new_doses <- data.frame(dose = factor(dose_levels, levels = dose_levels))
predict(dose_lm, new_doses)
```

The coefficients are 10.605, 9.130 and 15.495, and the fitted group means
are 10.605, 19.735 and 26.100. The intercept-plus-indicators model and
one-way ANOVA fit the same space, so fitted values, residuals and the
classical joint $F$ agree. This is a computational identity; it adds no
new design evidence. Chapter 8 develops the regression algebra further.


\Needspace{6\baselineskip}
\phantomsection

## Adjustment and balanced comparisons {.unnumbered}

In `ToothGrowth`, supplement is a second recorded treatment dimension,
not a school block or a before/after occasion. An additive model assumes
a common dose comparison within OJ and VC. A marginal dose mean instead
averages over the supplements observed at that dose. Here the balanced
cell counts give equal supplement weights at every dose. Consequently the
raw and additive-model contrasts have the same point estimate; their SEs
use different residual variances. This equality need not hold with unequal
supplement compositions.

```{r}
#| label: ch6-adjusted-model
#| echo: true
growth <- ToothGrowth
growth$dose <- factor(growth$dose, levels = c(.5, 1, 2))
growth$supp <- factor(growth$supp, levels = c("OJ", "VC"))
supp_fit <- lm(len ~ supp, data = growth)
add_fit <- lm(len ~ supp + dose, data = growth)
coef(add_fit)
anova(supp_fit, add_fit)
```

The additive intercept is 12.455. The VC-minus-OJ coefficient is $-3.700$;
dose coefficients are 9.130 and 15.495 relative to .5 mg/day. The dose
block comparison gives $F(2,56)=82.8109$, $p<.001$. This is a joint test
of dose terms, not the test of the single high-dose contrast.

For the 2 mg/day mean minus the equally weighted average of .5 and 1,
the coefficient weights are $(0,0,-.5,1)$ in the displayed order.

```{r}
#| label: ch6-adjusted-contrast
#| echo: true
L <- c(0, 0, -.5, 1)
adjusted <- sum(L * coef(add_fit))
adjusted_se <- sqrt(drop(t(L) %*% vcov(add_fit) %*% L))
adjusted_ci <- adjusted + c(-1, 1) *
  qt(.975, df.residual(add_fit)) * adjusted_se
c(contrast = adjusted, SE = adjusted_se)
adjusted_ci
```

The adjusted contrast is 10.9300, SE 1.0482, interval [8.8301, 13.0299].
Compare the raw one-way SE 1.1618 and interval [8.6036, 13.2564]. The
smaller residual variance after including supplement changes the precision
under the additive model; the data have not gained more animals. With exact
additivity, a common within-supplement dose contrast also equals a contrast
standardized to any common supplement weights. Whether a common contrast
is useful depends on how the comparison varies by supplement.

\Needspace{6\baselineskip}
\phantomsection

## Six cells and explicit model comparisons {.unnumbered}

Two supplements and three doses form six observed cells, with ten animals
per cell. Inspect the counts and means before fitting an interaction:

```{r}
#| label: ch6-factorial-cells
#| echo: true
cell_n <- with(growth, table(supp, dose))
cell_mean <- with(growth, tapply(len, list(supp, dose), mean))
cell_n
cell_mean
```

The OJ means are 13.23, 22.70 and 26.06; the VC means are 7.98, 16.77 and
26.14. The high-minus-low dose difference is 12.83 under OJ and 18.16 under
VC. Interaction concerns such variation in comparisons, not simply different
cell means.

```{r}
#| label: fig-ch6-factorial
#| echo: false
#| fig-pos: H
#| fig-cap: "Mean recorded lengths for the six dose–supplement combinations."
#| fig-alt: "OJ and VC profiles rise with dose but converge at 2 mg/day, showing nonparallel dose comparisons."
ggplot(growth, aes(dose, len, group = supp, linetype = supp, shape = supp)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  labs(x = "Dose (mg/day)", y = "Mean recorded length",
       linetype = "Supplement", shape = "Supplement")
```

The additive model has four coefficients: intercept, one supplement and
two dose terms. The full product model has six cell means, adding
$(2-1)(3-1)=2$ interaction parameters and leaving $60-6=54$ residual df.

For nested models, let $SS_{\text{added}}=RSS_R-RSS_F$ and let $q$ be the
number of extra parameters. The ordinary nested-model statistic is

$$
F=\frac{SS_{\text{added}}/q}{RSS_F/df_F}.
$$

\Needspace{22\baselineskip}

```{r}
#| label: ch6-factorial-test
#| echo: true
full_fit <- lm(len ~ supp * dose, data = growth)
anova(add_fit, full_fit)
rss_add <- sum(resid(add_fit)^2)
rss_full <- sum(resid(full_fit)^2)
df_full <- df.residual(full_fit)
interaction_F <- ((rss_add - rss_full) / 2) / (rss_full / df_full)
interaction_p <- pf(interaction_F, 2, df_full, lower.tail = FALSE)
c(F = interaction_F, p = interaction_p)
```

The extra SS is 108.319, full-model residual SS 712.106, and
$F=(108.319/2)/(712.106/54)=4.1070$, $p=.0219$. Under the working model,
this is evidence against the additive mean restriction. Interpret the dose
profiles before deciding whether a single average comparison is adequate.
The exact reference uses independent normal, common-variance errors and the
specified nested mean spaces [@kutner2005].

One can also report several components against a deliberately common
full-model residual denominator. Name that choice explicitly:

```{r}
#| label: ch6-factorial-components
#| echo: true
dose_only <- lm(len ~ dose, data = growth)
component_ss <- c(
  sum(resid(dose_only)^2) - rss_add,
  sum(resid(supp_fit)^2) - rss_add,
  rss_add - rss_full
)
component_df <- c(1, 2, 2)
component_F <- (component_ss / component_df) / (rss_full / df_full)
data.frame(component = c("Supplement after dose", "Dose after supplement",
                          "Interaction"),
           df = component_df, SS = component_ss, F = component_F,
           p = pf(component_F, component_df, df_full, lower.tail = FALSE))
```

The component $F$ values are 15.5720, 92.0000 and 4.1070. The first two
use a denominator from the full product model; they therefore differ from
ordinary comparisons ending at the additive model. Only the last is the
same nested interaction test just shown. In this balanced design, supplement
and dose terms are orthogonal. With unequal cells, sequential sums of squares
can depend on term order, so name the compared spaces instead of treating any
row labelled a main effect as interchangeable.

None of these model identities supplies missing assignment documentation or
extends animal results to people. Statistical fit, precision, substantive
importance and causal interpretation remain separate questions.

\Needspace{8\baselineskip}
\phantomsection

# Appendix D: Extended Practice and Synthesis {.unnumbered #sec-ch6-app-d}

These optional tasks combine procedure selection, simulation, and bounded
multi-procedure interpretation.

\Needspace{6\baselineskip}
\phantomsection

## Extended practice: simulation questions (Foundational) {.unnumbered}

Write distinct mass/density, cumulative-probability, quantile, and simulation
questions for a binomial and a normal distribution. Approximate one exactly
computable probability at two values of $B$ and report both MC SEs. Then
compare means at two values of $n$ while holding the generator fixed. Explain
separately what varies among observations, among study estimates, and because
the simulation used finitely many repetitions. Do not describe this
lognormal example as the source of the built-in observed records.

\Needspace{6\baselineskip}
\phantomsection

## Extended practice: match procedure to target (Integrative) {.unnumbered}

For a known-SD mean, unknown-SD mean, independent mean difference, paired
difference, and one proportion, name the estimand and repeated world before
the function. Reproduce estimate, SE, interval, and null standardization;
state whether the reference is exact or approximate and which conditions
make that statement true. Do not describe a stipulated SD as empirically known.

For two quantitative variables, fix a complete-pair sample, plot it, and
reproduce covariance, Pearson r, and the Fisher interval. For a two-way
categorical table, show observed and expected counts, denominators,
standardized residuals, chi-square, and Cramer's V. Explain when Fisher's
conditional test addresses the same intended null and what neither test
can say about design or causation.

\Needspace{6\baselineskip}
\phantomsection

## A bounded multi-procedure reading and extended practice (Integrative) {.unnumbered}

An extended report might state: “In the 60 recorded guinea pigs, dose means
were 10.605, 19.735 and 26.100 on the recorded length scale. The declared
high-dose-minus-average-lower-dose contrast was 10.930, with a common-variance
working-model interval [8.604, 13.256]. An additive supplement-adjusted model
gave the same estimate and interval [8.830, 13.030]. The dose profiles differed
by supplement; a nested interaction comparison gave $F(2,54)=4.107$.”

Read each quantity with its question, denominator and working model. Equal
raw and adjusted estimates here follow from balance, not a universal rule.
The interaction result invites interpretation of conditional dose comparisons;
it does not replace that interpretation with a significance label. The dataset
help page does not supply a threshold of substantive importance or a complete
randomization record.

For practice, reproduce the raw means, one-way ANOVA, magnitude measures,
declared contrast and Tukey family. Explain their different questions.
Verify that indicator regression returns the same fitted means. Then reproduce
the six cell means, additive contrast and nested interaction statistic. Compare
the residual denominators used for the different reported $F$ values.

Complete a short computational record using one actual built-in dataset or
explicitly labelled supplied summary. State the question, row unit, variables,
calculation, result, assumptions and bounded interpretation. Include one useful
display or compact table. Read what the calculation establishes before judging
one consequential limitation. Finish by explaining what a later regression
model would add to the chosen question.

\Needspace{8\baselineskip}
\phantomsection

# References {.unnumbered}

::: {#refs}
:::
