---
title: "Multiple Regression II: Categorical Predictors, Model Comparison, Nonlinearity, and Robust Uncertainty"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{16\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AddToHook{cmd/subsubsection/before}{\Needspace{6\baselineskip}\phantomsection}
          \AtBeginDocument{\renewcommand{\boldsymbol}[1]{\symbf{#1}}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.pos = "!ht")
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")
options(digits = 7, width = 68)
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 11))
```

\clearpage

::: {.callout-note title="Chapter overview"}
Chapters 7 and 8 fitted lines and additive models and read them as
conditional comparisons. Along the way they used tools they could only
name: a factor coded as 0/1 columns, a contrast between two categories, a
joint test of several added coefficients, an information criterion, a
squared predictor, a logged outcome, a covariance estimate that allows
unequal spread, and a covariance that respects pupils sitting in schools.
This chapter develops each of them once, on the same 98 occupation
records, on a small randomized reading study with a pretest, and then on
Project STAR, where a randomized design supplies a different warrant. By
the end of the chapter you will be able to code and read a factor with
several levels, estimate a contrast with the right uncertainty, read an
analysis of covariance as a regression on a randomized pretest--posttest
study, compare nested models and read AIC and BIC, change a predictor's
form or an outcome's scale and say what question changed, choose between
classical, HC3, cluster-robust, and bootstrap uncertainty for a stated
reason, and say what none of these calculations repairs.
:::

A regression report rarely stops at one additive equation. It compares
groups that are categories, adjusts a comparison for a baseline score,
asks whether a set of added variables matters jointly, ranks candidate
models, lets a relationship curve or a scale change, and reports
uncertainty calculated in more than one way. Each of these steps has a
small mathematics of its own and a temptation of its own: reading a
coefficient as "the effect of type", adding two standard errors,
comparing criteria across different outcomes, or treating a robust
standard error as a repair for a design problem.

We keep the occupation records of Chapter 8, with their 98 complete rows,
income in thousand 1971 Canadian dollars, and the additive model of
education, prestige, percentage of women, and type, so that every tool is
seen on numbers already read. Two further studies join them where the
occupations cannot teach the point: a randomized reading study with a
pretest, for the analysis of covariance, and Project STAR, where the
uncertainty calculation has to follow the design that produced the data.
Keeping the records fixed makes one distinction visible throughout:
changing a model, changing a question, and changing an uncertainty
calculation are three different acts.

The R code shown is the one call that produces each new quantity; the
Computer Lab carries the preparation, the checks and the variations. The
type levels remain *bc*, *prof*, and *wc*, the full additive formula
remains `income_k ~ education + prestige + women + type`, and every
adjusted type comparison names education, prestige, and the percentage of
women as its conditioning set. Where Chapter 8's appendices already
worked a tool through on these records (the group means recovered by
indicators, the overall $F$, the HC3 covariance, and the Normal
likelihood behind AIC and BIC), this chapter points back to that
working and adds the new use.

\clearpage

## Predictors That Are Categories {#sec-ch9-factors}

### One indicator, then several

Chapter 7, Section 3.3, fitted a single 0/1 predictor to 24 invented
feedback-by-practice scores. The fitted line passed through the two group
means, its slope was the difference between them, 4 with SE 1.6967 and
$t=2.3575$ on 22 df, and that test was the equal-variance two-sample test
written as a regression. Nothing about that arrangement is special to two
groups.

A factor with $J$ levels enters a regression as $J-1$ indicator columns.
One level is omitted and becomes the **reference**: the intercept is its
fitted mean, and each indicator coefficient is a difference from it. With
three occupation types and blue-collar as the reference,

$$
\widehat y_g=\widehat\alpha+\widehat\delta_{\mathrm{prof}}D_{\mathrm{prof}}
+\widehat\delta_{\mathrm{wc}}D_{\mathrm{wc}},\qquad
\widehat\mu_{\mathrm{bc}}=\widehat\alpha,\quad
\widehat\mu_{\mathrm{prof}}=\widehat\alpha+\widehat\delta_{\mathrm{prof}},\quad
\widehat\mu_{\mathrm{wc}}=\widehat\alpha+\widehat\delta_{\mathrm{wc}}.
$$

Two indicators, not three, carry three means. A third column would equal
the intercept column minus the other two, so the design matrix would lose
full rank and no coefficient would be separately estimable. Which level
is omitted is therefore a choice of coordinates, not a choice of model.

The occupation records and models are those of Chapter 8, re-created here
so that every later calculation has a stated origin: `m0`, `m1`, and `m2`
are the education-only, the education-and-prestige, and the full additive
models.

```{r}
#| label: ch9-records
#| echo: true
prestige <- carData::Prestige
prestige$income_k <- prestige$income / 1000
d <- prestige[!is.na(prestige$type), ]
d$type <- factor(d$type, levels = c("bc", "prof", "wc"))
m0 <- lm(income_k ~ education, data = d)
m1 <- lm(income_k ~ education + prestige, data = d)
m2 <- lm(income_k ~ education + prestige + women + type, data = d)
```

The 98 retained records hold 44 blue-collar, 31 professional, and 23
white-collar occupations, and the full model fits six coefficients, leaving
92 residual degrees of freedom. Complete-case selection is how these rows
were chosen; it is not random sampling from a labour market
[@carData2022; @fox2019].

### Indicators recover fitted group means

Fit income on type alone and read the coefficients against
@fig-ch9-indicators. A blue-collar occupation carries 0 in both indicator
columns, a professional one a 1 in the first, and a white-collar one a 1
in the second, so each type's fitted value is the intercept plus that
type's coefficient.

```{r}
#| label: ch9-group-means
#| echo: true
group_fit <- lm(income_k ~ type, data = d)
coef(group_fit)
```

```{r}
#| label: fig-ch9-indicators
#| fig-width: 6.4
#| fig-height: 3.1
#| fig-cap: "Income of the 98 occupations by type. Bars mark the group means; the dashed line is the blue-collar mean, the intercept, and the arrows are the two indicator coefficients, 5.1853 and −0.3218."
#| fig-alt: "Jittered open circles of income in thousand dollars for 44 blue-collar, 31 professional and 23 white-collar occupations, with a thick bar at each group mean, about 5.37, 10.56 and 5.05. A dashed horizontal line at the blue-collar mean runs across the plot. An upward arrow from the dashed line to the professional mean is labelled 5.1853 higher and a short downward arrow to the white-collar mean is labelled 0.3218 lower. Professional incomes spread from about 4.6 to 25.9 thousand dollars; the other two types stay below 9."
type_means <- data.frame(x = 1:3,
  mean = as.numeric(tapply(d$income_k, d$type, mean)))
ggplot2::ggplot(d, ggplot2::aes(as.numeric(type), income_k)) +
  ggplot2::geom_hline(yintercept = type_means$mean[1], linetype = 2,
    colour = "grey45") +
  ggplot2::geom_point(shape = 1, colour = "grey55",
    position = ggplot2::position_jitter(width = .12, height = 0,
      seed = 12345)) +
  ggplot2::geom_segment(data = type_means, ggplot2::aes(x = x - .28,
    xend = x + .28, y = mean, yend = mean), linewidth = 1.1) +
  ggplot2::annotate("segment", x = c(2.4, 3.4), xend = c(2.4, 3.4),
    y = type_means$mean[1], yend = type_means$mean[2:3],
    arrow = ggplot2::arrow(length = ggplot2::unit(.15, "cm"))) +
  ggplot2::annotate("text", x = c(2.46, 3.46), y = c(9.9, 6.9),
    label = c("5.1853 higher", "0.3218 lower"), hjust = 0, size = 3.4) +
  ggplot2::scale_x_continuous(breaks = 1:3, limits = c(.6, 3.9),
    labels = c("Blue-collar (44)", "Professional (31)",
      "White-collar (23)")) +
  ggplot2::labs(x = NULL, y = "Income (thousand dollars)") +
  ggplot2::theme(panel.grid.major.x = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank())
```

The coefficients are 5.3741, 5.1853, and $-0.3218$. The intercept is the
blue-collar mean, the dashed line in the figure, and each coefficient is
an arrow from that line to another type's mean: professional occupations
average 5.1853 thousand dollars more, white-collar occupations 0.3218
less. The fitted means, 5.3741, 10.5595, and 5.0523, are exactly the
observed means, for the reason Chapter 8, Appendix A, gave: with one
free parameter per group, least squares returns each group's own mean. A
group-only regression is therefore a description of three means with no
adjustment in it at all.

The figure also shows what three means leave out. Professional incomes
spread from about 4.6 to 25.9 thousand dollars, while the other two types
stay below 9; Section 5 returns to that difference in spread.

Keep this description apart from the adjusted comparison of the full
model. In `m2` the professional and white-collar coefficients are 0.3242
and 0.2353: they compare fitted incomes at the *same* education,
prestige, and percentage of women, not the observed incomes of
differently composed groups. The raw professional advantage of 5.1853
thousand dollars and the adjusted 0.3242 answer different questions, and
neither is "the effect of occupational type".

### Releveling and centering change coordinates, not the fit

Releveling makes a different level the reference. Refit the group-only
model with white-collar as the reference:

```{r}
#| label: ch9-relevel
#| echo: true
group_fit_wc <- lm(income_k ~ relevel(type, ref = "wc"), data = d)
coef(group_fit_wc)
```

The coefficients are now 5.0523, 0.3218, and 5.5071: the intercept is the
white-collar mean, and the two coefficients measure the blue-collar and
professional means from it. The fitted values and residuals are
unchanged, as Chapter 8, Appendix A, stated; the same three means are
reported in a different coordinate system. The full model behaves the
same way. Releveling `m2` to
white-collar gives intercept 0.6559, blue-collar $-0.2353$, and
professional 0.0889, while the education, prestige, and women
coefficients stay exactly as they were.

Centering a continuous predictor is the analogous operation: subtracting
a constant moves the profile the intercept refers to, and leaves every
fitted value, residual, and fixed substantive contrast alone. Two
consequences are worth keeping. A coefficient's size means nothing
without its reference; "the professional coefficient" is short for
"professional minus whichever level is currently omitted". And neither
releveling nor centering creates information. Chapter 8 documented how
thin the joint support across types is at common education and prestige
values; changing coordinates adds no observed comparison.

### The one-way analysis of variance of Chapter 4 as a regression

Chapter 4 tested equality of several group means with a one-way analysis
of variance. In the group-only regression that null says that both
indicator coefficients are zero, so the reduced model is the
intercept-only model, whose residual sum of squares is the total sum of
squares about the grand mean. Comparing the two models gives

$$
F=\frac{(RSS_{\mathrm{reduced}}-RSS_{\mathrm{full}})/2}
{RSS_{\mathrm{full}}/(98-3)}
=\frac{(1734.0792-1138.123)/2}{1138.123/95}=24.8725.
$$

The numerator is the reduction in residual sum of squares that the two
indicators buy, per indicator; the denominator is the group-only model's
residual mean square on its 95 degrees of freedom.

```{r}
#| label: ch9-anova-identity
#| echo: true
anova(lm(income_k ~ 1, data = d), group_fit)
```

`aov()` returns the same statistic from the same fitted means, the same
residual sum of squares, and the same degrees of freedom. The equality is
algebraic. It holds whether or not group membership was randomized, so it
supplies no causal warrant of its own; Section 7 shows what a randomized
design adds. The Chapter 4 cautions travel with the identity. An omnibus
$F$ says that the three means are not all equal under the stated
reference distribution; it does not say which differ, and any comparison
chosen after looking at this result still needs the family calibration of
Chapter 4. The next section gives the calculation that a single declared
comparison requires.

## Contrasts and Adjusted Group Comparisons {#sec-ch9-contrasts}

### A contrast needs coefficient covariance

A **contrast** is a declared linear combination of coefficients,
$C=\mathbf c^{\mathsf T}\boldsymbol\beta$, named before the comparison is
read. Its estimate is the same combination of the fitted coefficients,
and its variance is the corresponding quadratic form in the coefficient
covariance matrix $\widehat{\mathbf V}$:

$$
\widehat C=\mathbf c^{\mathsf T}\widehat{\boldsymbol\beta},\qquad
\widehat{\operatorname{Var}}(\widehat C)=\mathbf c^{\mathsf T}\widehat{\mathbf V}\mathbf c,\qquad
SE(\widehat C)=\sqrt{\mathbf c^{\mathsf T}\widehat{\mathbf V}\mathbf c}.
$$

For a difference of two coefficients the quadratic form reduces to
$\operatorname{Var}(b_1)+\operatorname{Var}(b_2)-2\operatorname{Cov}(b_1,b_2)$.
Take professional minus white-collar in `m2`, whose coefficients are
ordered intercept, education, prestige, women, professional,
white-collar, so that $\mathbf c=(0,0,0,0,1,-1)^{\mathsf T}$.

```{r}
#| label: ch9-contrast
#| echo: true
c_vec <- c(0, 0, 0, 0, 1, -1)
V <- vcov(m2)
c(estimate = sum(c_vec * coef(m2)),
  SE = sqrt(drop(c_vec %*% V %*% c_vec)))
```

The estimate is $0.3242-0.2353=0.0889$ thousand dollars. Its variance
assembles the two coefficient variances and their covariance,

$$
2.1406+0.9698-2(0.972)=1.1664,
$$

so the SE is 1.08 and, with $t_{.975,92}=1.9861$, the 95% interval is
$[-2.0561,\ 2.2338]$.

The covariance matters because the two type coefficients move together.
Both are measured from the same blue-collar reference at the same values
of the continuous predictors, and their estimated correlation is .6746: a
sample that overstates one tends to overstate the other, so their
difference varies less than either movement alone suggests. Dropping the
covariance would give an SE of 1.7636, and adding the two standard
errors, $1.4631+0.9848=2.4479$, gives a number that is not the standard
error of anything. Only the quadratic form is the variance of the
declared comparison.

The vector $\mathbf c$ is the question written down. One entry set to 1
and the rest to 0 returns an ordinary coefficient with its usual SE.
Entries on the type coefficients that sum to zero, with the intercept
entry 0, make a contrast in the Chapter 4 sense, and such a contrast names
the same comparison whichever level is the reference; that is why Section
1.3 found the same 0.0889 as the professional coefficient of the releveled
fit. Declaring the comparison in advance matters as much as computing it
correctly: a contrast chosen because it looked large in the fitted
coefficients is a selection from a family, and it needs Chapter 4's
calibration for that family.

Read the interval for what it is: a working-model benchmark for
professional against white-collar occupations at the same education,
prestige, and percentage of women. It covers zero, so these records leave
the comparison unresolved, and its width does not overcome the thin
cross-type support that Chapter 8 documented. An interval can be
arithmetically correct and still describe a comparison that the records
barely observe.

### Adjusted block tests

A contrast tests one declared combination. A **block test** asks whether
a set of coefficients is jointly zero; it is the nested-model comparison
of Section 1.4 with a different reduced model. Two blocks are worth
testing here, and they are different questions.

```{r}
#| label: ch9-blocks
#| echo: true
#| results: hide
no_type <- lm(income_k ~ education + prestige + women, data = d)
anova(m1, m2)
anova(no_type, m2)
```

Adding the percentage of women and the two type indicators to `m1` tests
three restrictions against the full model's 92 residual degrees of
freedom; dropping type from the full model tests two restrictions against
the same denominator.

| Added | df | $F$ | $p$ |
|:--|--:|--:|--:|
| Women and type | 3 | 11.5088 | $<.0001$ |
| Type after women | 2 | 0.0319 | .9686 |

: Two declared adjusted block comparisons. {#tbl-ch09-blocks tbl-colwidths="[40,15,20,25]"}

The first block is strongly supported: $F(3,92)=11.5088$ against a 5%
critical value of 2.7036. The second is not: $F(2,92)=0.0319$ with
$p=.9686$ against a critical value of 3.0954. Neither result says what it
is often read as saying. Joint evidence for the first block does not
establish that each of the three added columns contributes; the
percentage of women could be carrying the whole block. And the second
result is not proof that occupational type is irrelevant. It says that,
once education, prestige, and the percentage of women are in the
equation, the two type indicators add no detectable reduction in residual
sum of squares *on these 98 records*, where the types barely overlap on
the other predictors.

Two further distinctions matter. `anova(m2)` on a single model gives
**sequential** sums of squares in formula order: each line tests a term
given only the terms above it, so the answers depend on the order in which
the formula was written, and only the last line coincides with the
corresponding block test. And Welch-type and robust-covariance tests are
different procedures with different reference distributions; Section 5
gives the robust version of this type block.

### The analysis of covariance as regression {#ch09-ancova}

A randomized study with a baseline measurement asks the group question in
a sharper form. Baumann and colleagues randomly assigned 68 fourth-grade
pupils in one school to three reading-instruction conditions
[@baumann1992]: DRA, a directed reading activity from the basal
programme; DRTA, a directed reading-thinking activity; and think-aloud
strategy instruction. Before and after the lessons, each pupil took an
error-detection test of comprehension monitoring. The `carData::Baumann`
distribution holds the 66 pupils with a recorded first posttest, 22 per
group. The analysis here uses the first pretest and the first posttest,
labels the groups DRA, DRTA, and Think-aloud, and takes DRA as the
reference; the original investigators analysed further outcomes and both
pretests.

The **analysis of covariance** adds the pretest to the group comparison
as one more predictor with a single slope:

$$
Y_{ij}=\mu_j+\beta(X_{ij}-\bar X)+\varepsilon_{ij},\qquad
\bar Y_j^{\,\mathrm{adj}}=\bar Y_j-\widehat\beta(\bar X_j-\bar X).
$$

Here $Y_{ij}$ and $X_{ij}$ are the posttest and pretest of pupil $i$ in
group $j$. The group terms $\mu_j$ shift the line up or down, and the
pretest enters once, so every group's line has the same slope $\beta$. An
**adjusted mean** is a group's fitted mean at a stated pretest value. At
the grand pretest mean $\bar X$ it is the group's observed posttest mean
$\bar Y_j$ moved along the common slope by the group's pretest departure
from $\bar X$, which is the second equation.

```{r}
#| label: ch9-baumann
#| echo: true
reading <- carData::Baumann
reading$group <- factor(reading$group,
  levels = c("Basal", "DRTA", "Strat"),
  labels = c("DRA", "DRTA", "Think-aloud"))
ancova <- lm(post.test.1 ~ group + pretest.1, data = reading)
coef(ancova)
predict(ancova, interval = "confidence",
  newdata = data.frame(group = levels(reading$group),
    pretest.1 = mean(reading$pretest.1)))
```

```{r}
#| label: fig-ch9-ancova
#| fig-width: 6.4
#| fig-height: 3.9
#| fig-cap: "Posttest against pretest for the 66 Baumann pupils. The parallel lines are the common-slope fit over the common pretest support 6--14; each filled symbol marks a group's observed pretest and posttest means, and its arrow follows the group's line to the grand pretest mean 9.7879, where the open symbol marks the adjusted mean."
#| fig-alt: "Grey points for DRA (circles), DRTA (triangles) and Think-aloud (squares) pupils, pretest from 4 to 16 and posttest from 1 to 15, slightly jittered. Three parallel fitted lines with slope about 0.69 run from pretest 6 to 14: DRTA highest, Think-aloud in the middle and DRA lowest. A vertical line marks the grand pretest mean 9.79, where open symbols mark the adjusted means. The DRA mean, a filled circle at pretest 10.5 and posttest 6.68, moves down its line to 6.19; the Think-aloud mean, at 9.14 and 7.77, moves up to 8.22; the DRTA mean, at 9.73 and 9.77, barely moves, to 9.81, so its filled and open triangles almost coincide."
grand_pretest <- mean(reading$pretest.1)
line_grid <- expand.grid(pretest.1 = c(6, 14),
  group = levels(reading$group))
line_grid$fitted <- predict(ancova, newdata = line_grid)
shift <- data.frame(group = levels(reading$group),
  x = as.numeric(tapply(reading$pretest.1, reading$group, mean)),
  y = as.numeric(tapply(reading$post.test.1, reading$group, mean)))
shift$adjusted <- predict(ancova, newdata = data.frame(
  group = shift$group, pretest.1 = grand_pretest))
ggplot2::ggplot(reading, ggplot2::aes(pretest.1, post.test.1,
    shape = group, linetype = group)) +
  ggplot2::geom_point(colour = "grey60", size = 1.6,
    position = ggplot2::position_jitter(width = .15, height = .15,
      seed = 12345)) +
  ggplot2::geom_line(data = line_grid, ggplot2::aes(y = fitted),
    linewidth = .7) +
  ggplot2::geom_vline(xintercept = grand_pretest, colour = "grey40",
    linewidth = .4) +
  ggplot2::geom_segment(data = shift, ggplot2::aes(x = x, y = y,
    xend = grand_pretest, yend = adjusted), inherit.aes = FALSE,
    linewidth = .9,
    arrow = ggplot2::arrow(length = ggplot2::unit(.22, "cm"))) +
  ggplot2::geom_point(data = shift, ggplot2::aes(x, y), size = 2.8) +
  ggplot2::geom_point(data = shift, ggplot2::aes(grand_pretest, adjusted),
    shape = c(1, 2, 0), size = 3.8, stroke = .9, show.legend = FALSE) +
  ggplot2::scale_shape_manual(values = c(16, 17, 15)) +
  ggplot2::scale_linetype_manual(values = c(1, 2, 4)) +
  ggplot2::scale_x_continuous(breaks = seq(4, 16, 2)) +
  ggplot2::labs(x = "Pretest score", y = "Posttest score", shape = NULL,
    linetype = NULL) +
  ggplot2::theme(legend.position = "bottom",
    panel.grid.minor = ggplot2::element_blank())
```

Read @fig-ch9-ancova from each filled symbol, which marks a group's
observed pretest and posttest means, to the open symbol on the vertical
line, which marks its adjusted mean. The DRA group happened to start
highest, 0.7121 points above the grand pretest mean of 9.7879. Following
its line to the grand mean lowers its posttest mean by
$0.6932\times0.7121$, from 6.6818 to 6.1882. Think-aloud started 0.6515
points below the grand mean and rises from 7.7727 to 8.2243, while DRTA,
almost exactly at the grand mean, barely moves, from 9.7727 to 9.8147.
The fit gives the common slope 0.6932 (SE 0.1015), the differences from
DRA of 3.6266 (SE 0.7362) for DRTA and 2.0362 (SE 0.7450) for
Think-aloud, and a residual mean square of 5.894 on 62 degrees of
freedom. Tested together after the pretest, the two group coefficients
give $F(2,62)=12.1689$, $p<.0001$, and the printed adjusted means carry
the 95% intervals $[5.1435,\ 7.2329]$, $[8.7800,\ 10.8495]$, and
$[7.1813,\ 9.2674]$.

Three conditions carry that reading. The adjusted means need a **common
reference**, here the grand pretest mean. That reference must lie inside
every group's observed pretest range, the **common support**: the groups'
pretests run from 4 to 16, 6 to 16, and 4 to 14, so the lines are drawn
over 6 to 14, and an adjusted comparison read outside that band would rest
on extrapolation. And the comparison needs a stated **averaging
population**: for this additive linear model, predicting every group at
the grand pretest mean gives the same values as predicting every group
over the pupils' common pretest distribution and averaging. Because the
lines are parallel, the adjusted differences, 3.6266 and 2.0362, are the
same at any pretest value inside the common support; the adjusted means
themselves are not, which is why the reference value is always stated.

### Posttest only, gain score, and covariance adjustment as one regression {#ch09-ancova-pretest}

Three analyses are common for a randomized pretest--posttest study:
compare the posttest means, compare the gains (posttest minus pretest), or
adjust for the pretest as above. They are one regression that differs only
in the pretest coefficient $\gamma$:

$$
Y_i-\gamma X_i=\alpha+\delta_{\mathrm{DRTA}}D_{\mathrm{DRTA},i}
+\delta_{\mathrm{TA}}D_{\mathrm{TA},i}+\varepsilon_i,\qquad
\gamma=0,\ 1,\ \text{or estimated}.
$$

With $\gamma=0$ the model compares posttest means, with $\gamma=1$ it
compares gains, and with $\gamma$ estimated it is the analysis of
covariance. In R, `offset()` enters the pretest with its coefficient fixed
at 1:

```{r}
#| label: ch9-three-analyses
#| echo: true
post_only <- lm(post.test.1 ~ group, data = reading)
gain <- lm(post.test.1 ~ group + offset(pretest.1), data = reading)
```

| Analysis | Pretest coefficient | DRTA minus DRA (SE) | Think-aloud minus DRA (SE) | Residual SD (df) |
|:--|:--|:--|:--|:--|
| Posttest only | 0 | 3.0909 (0.9614) | 1.0909 (0.9614) | 3.1885 (63) |
| Gain score | 1 | 3.8636 (0.7779) | 2.4545 (0.7779) | 2.5799 (63) |
| Analysis of covariance | 0.6932, estimated | 3.6266 (0.7362) | 2.0362 (0.7450) | 2.4278 (62) |

: Three analyses of the Baumann records as one regression. {#tbl-ch09-three-analyses tbl-colwidths="[24,18,20,20,18]"}

Under random assignment the expected pretest difference between groups is
zero, so all three analyses estimate the same average difference in
posttest scores. Strictly, random assignment covers the 68 pupils
assigned, while these analyses use the 66 with a recorded first
posttest; the argument carries over to the 66 if, for example, the two
posttests went missing for reasons unrelated to what those pupils would
have scored, a complete-record condition that the design does not
guarantee and the records cannot check. The three analyses differ in precision and in how
this sample's chance pretest imbalance enters. Here the pretest means
are 10.5000, 9.7273, and 9.1364, so DRA started with a head start of 0.7727 points over DRTA,
about a quarter of the pretest SD of 3.0205 across all 66 pupils. The
size of that gap is what matters. A significance test of it would add
nothing: random assignment already makes any baseline gap a chance gap,
so such a test would examine a null hypothesis known to be true.

The gap accounts exactly for the spread of the estimates in
@tbl-ch09-three-analyses. Each DRTA estimate is the posttest-only
difference plus $\gamma$ times the DRA group's pretest head start,
$3.0909+\gamma(10.5-9.7273)$, which gives 3.0909, 3.8636, and 3.6266
for the three values of $\gamma$ (3.6266 from full precision; 3.6265
from the rounded inputs).

The last column shows why the estimated coefficient is usually preferred.
The pretest predicts the posttest with a slope near 0.7, so fixing its
coefficient at 0 leaves the pretest's predictive information in the error
(residual SD 3.1885), and fixing it at 1 over-corrects (2.5799).
Estimating it leaves the smallest residual SD, 2.4278, and the smallest
standard errors. The choice among the three belongs in the analysis plan,
made before the outcomes are seen; picking whichever estimate looks best
after fitting all three turns three defensible analyses into one selected
result.

### What adjustment cannot supply {#ch09-ancova-limits}

A common slope is a condition, not a finding. The check adds the
group-by-pretest product block, which lets each group have its own slope,
and compares the two named models on the same records:

```{r}
#| label: ch9-parallel
#| echo: true
varying_slopes <- lm(post.test.1 ~ group * pretest.1, data = reading)
anova(ancova, varying_slopes)
```

The product block gives $F(2,60)=1.5054$, $p=.2302$, with group slopes of
0.4663 (DRA), 0.6868 (DRTA), and 0.8768 (Think-aloud). These 66 records do
not distinguish the two models, which is not the same as showing that the
slopes are equal. When slopes do differ, an adjusted difference depends on
the pretest value at which it is read; Chapter 10 develops that
varying-slope model as an interaction.

What adjustment supplies depends on how the groups were formed. In groups
formed by random assignment, the pretest only sharpens a comparison that
the design already warrants. In observational groups, a fitted comparison
at a common covariate value cannot balance what was not measured; a
covariate measured with error under-adjusts, leaving part of the
baseline difference in place (Chapter 14, Section 5.1); and a covariate
recorded after the groups formed can remove part of the effect being
estimated.

**Lord's paradox** makes the point with intact groups that differ at
baseline [@lord1967]: the gain-score analysis asks whether the groups
changed by different amounts, the covariance analysis asks whether pupils
with the same pretest ended differently, and the two can reach opposite
conclusions from the same records, partly through the regression toward
the mean of Chapter 7, Section 7. Neither analysis says which question is
the causal one; the design does, and Chapter 14 returns to what a design
supplies that adjustment cannot.

## Compare Nested Models {#sec-ch9-comparison}

### The extra-sum-of-squares statistic

Sections 1.4 and 2.2 used one statistic. Let the reduced model be the
full model with $q$ coefficients set to zero, both fitted to the same
records and the same outcome. Then

$$
F=\frac{(RSS_{R}-RSS_{F})/q}{RSS_{F}/(n-p)}.
$$

The numerator is the extra sum of squares that the added block buys, per
coefficient it spends. The denominator is the fuller model's residual
mean square, the same $s^2$ that appears in every classical standard
error. Under the Normal linear model with independent common-variance
errors, the ratio has an $F$ reference with $q$ and $n-p$ degrees of
freedom.

```{r}
#| label: fig-ch9-rss
#| fig-width: 6.4
#| fig-height: 2.6
#| fig-cap: "Residual sum of squares as columns are added to the occupation model on the same 98 records; each label gives the drop that the added column or block buys."
#| fig-alt: "Horizontal bars of residual sum of squares for five nested models. The intercept-only bar is labelled RSS 1734.0792; adding education drops it by 571.5324, adding prestige by 294.8931, adding the percentage of women by 236.3266, and adding the two type columns by only 0.4375, so the last two bars look the same length."
rss_path <- data.frame(model = c("Intercept only", "+ education",
    "+ prestige", "+ women", "+ type (2 columns)"),
  rss = sapply(list(lm(income_k ~ 1, data = d), m0, m1, no_type, m2),
    deviance))
rss_path$label <- paste(c("RSS", rep("drop", nrow(rss_path) - 1)),
  format(round(c(rss_path$rss[1], -diff(rss_path$rss)), 4), digits = 15,
    scientific = FALSE, trim = TRUE))
rss_path$model <- factor(rss_path$model, levels = rev(rss_path$model))
ggplot2::ggplot(rss_path, ggplot2::aes(rss, model)) +
  ggplot2::geom_col(width = .6, fill = "grey75") +
  ggplot2::geom_text(ggplot2::aes(label = label), hjust = -.1,
    size = 3.3) +
  ggplot2::scale_x_continuous(limits = c(0, 2150), expand = c(0, 0)) +
  ggplot2::labs(x = "Residual sum of squares", y = NULL)
```

@fig-ch9-rss shows that bargain for the occupation models. Education buys
571.5324 of the total 1734.0792, and prestige a further 294.8931. The
percentage of women then buys 236.3266 for one coefficient, while the two
type columns buy only 0.4375 for two. Set against the full model's
residual mean square of 6.8575, the type block's 0.4375 spread over two
coefficients gives $F(2,92)=0.0319$, and the women-and-type block's 236.764
over three gives 11.5088, the entries of @tbl-ch09-blocks. Section 1.4's
test was the same comparison with the intercept-only model as the reduced
model.

The subtraction means something only when the two models use the same
rows and the same outcome, and when the reduced model is obtainable from
the fuller one by restrictions; Chapter 8's complete-case selection is why
every comparison in this chapter runs on the same 98 records. The
**overall** regression test is the special case whose reduced model is
the intercept alone, so that all $p-1$ non-intercept coefficients are set
to zero. Chapter 8, Appendix A, gave its formula and read it from
`summary()`: the full model gives $F(5,92)=32.1747$, and the
education-only model on the same 98 records gives $F(1,96)=47.1956$.
Placed beside the general statistic, the two values are easy to misread.
The larger one is not evidence that `m0` is the better model, because the
two statistics test different restrictions in different models, each
against the intercept alone. Neither verifies the mean function, the
error model, or the independence of the rows, and neither says how a
fitted rule would perform on records it has not seen, which is a separate
question with a separate calculation (Chapter 13).

### Restrictions in matrix form

The same statistic has a compact matrix form, worth one pass because it
makes clear what "the same records" is doing in every comparison above.
Let $\mathbf X$ hold $n$ rows and $p$ coefficient columns including the
intercept, with full column rank. Minimizing
$(\mathbf y-\mathbf X\boldsymbol\beta)^{\mathsf T}(\mathbf y-\mathbf X\boldsymbol\beta)$
gives the normal equations
$\mathbf X^{\mathsf T}(\mathbf y-\mathbf X\widehat{\boldsymbol\beta})=\mathbf 0$ and
hence
$\widehat{\boldsymbol\beta}=(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}\mathbf y$,
$\widehat{\mathbf y}=\mathbf H\mathbf y$ with
$\mathbf H=\mathbf X(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}$, and
$\widehat{\mathbf V}=s^2(\mathbf X^{\mathsf T}\mathbf X)^{-1}$. For $q$ independent
linear restrictions $\mathbf R\boldsymbol\beta=\mathbf r$,

$$
F=\frac{(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)^{\mathsf T}\,
[\mathbf R(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf R^{\mathsf T}]^{-1}\,
(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)}{q\,s^{2}}\sim F_{q,\,n-p}.
$$

For `m2`, $\mathbf X$ has six columns and rank 6: the two type indicators
do not reproduce the intercept column, so every coefficient is separately
estimable. Choosing the two rows of $\mathbf R$ that select the type
coefficients, with $\mathbf r=\mathbf 0$, returns 0.0319, the adjusted
type block test of Section 2.2, without fitting a second model.

Two practical points follow. `lm()` does not form
$(\mathbf X^{\mathsf T}\mathbf X)^{-1}$; it uses a QR decomposition, which
is numerically better behaved, so the displayed equation explains what the
estimator *is*, not how R computes it [@fox2016; @fox2019]. And because
$\widehat{\mathbf y}=\mathbf H\mathbf y$ is a projection onto the column
space of $\mathbf X$, any full-rank recoding of a factor spans the same
space and leaves the fitted vector alone, which is Section 1.3's
releveling result stated geometrically. Appendix A carries the
derivation.

### A local Normal likelihood and information criteria

Nested $F$ tests compare two models when one contains the other.
Information criteria rank models that need not be nested, using a
different quantity: the maximized likelihood, penalized for the number of
fitted parameters.

Chapter 8, Appendix B, built that likelihood for these occupation models
by carrying Chapter 6's binomial likelihood, for 32 successes in 50
trials, over to outcomes treated as independent Normal draws around the
regression. Three of its results carry the reading here. Maximizing the
Normal likelihood over the coefficients reproduces least squares. Its
scale estimate divides the residual sum of squares by $n$,
$\sqrt{RSS/n}=2.5373$, where the residual SD reported by `lm()` divides
by the residual degrees of freedom, $\sqrt{RSS/(n-p)}=2.6187$. And AIC
adds $2k$ to $-2\log\widehat L$ while BIC adds $k\log n$, where $k$
counts the fitted coefficients **and** the residual variance, so `m2` has
$k=7$, not 6. One call returns the maximized log-likelihood and both
criteria for a fitted model:

```{r}
#| label: ch9-likelihood
#| echo: true
c(logLik = logLik(m2), AIC = AIC(m2), BIC = BIC(m2))
```

For `m2` it returns $-230.3020$, 474.6040, and 492.6988. Chapter 8,
Appendix B, set the same criteria for the three nested models side by
side: AIC 526.5052, 499.8328, and 474.6040 for `m0`, `m1`, and `m2`, and
BIC 534.2601, 510.1727, and 492.6988. Smaller is preferred under either
criterion, and both prefer `m2` here.

What this chapter adds is how to use them as tools for comparing models.
BIC charges the same seven parameters $7\log 98=32.0948$ instead of AIC's
14, because at $n=98$ the factor $\log n=4.585$ exceeds 2. The two
criteria can therefore disagree, and a disagreement is information about
how strongly the extra parameters are being charged, not a tie to be
broken by preference. The conditions for reading them are strict and easy
to breach. The candidates must be fitted to the same records, with the
same outcome on the same scale, under the same likelihood convention;
Section 4.2 shows what happens when the scale changes. Above all, neither
criterion is a held-out error. Both are calculated on the records that
fitted the model, and both are comparative rather than absolute: a
smaller AIC says one candidate is preferred to another, not that either
describes the records well [@akaike1974; @schwarz1978]. Chapter 11
develops likelihood fully, for models where least squares is no longer
available.

## Change the Form of a Predictor or the Scale of the Outcome {#sec-ch9-forms}

### A polynomial term

Chapter 8, Section 4.3, refitted the full model with a squared prestige
term as a disclosed sensitivity check on functional form. Here that refit
is the worked example. A squared term is an ordinary extra column,
`I(prestige^2)`, and the model is still linear in its coefficients. What
changes is the reading:

$$
E(Y\mid x,\ldots)=\beta_0+\beta_1x+\beta_2x^2+\cdots,\qquad
\frac{\partial E(Y)}{\partial x}=\beta_1+2\beta_2x,\qquad
x^{*}=-\frac{\beta_1}{2\beta_2}.
$$

The predictor no longer has a single slope. It has a slope that depends on
its own value, $\beta_1+2\beta_2x$, and a turning point $x^{*}$ where that
slope is zero.

```{=latex}
\Needspace{18\baselineskip}
```

```{r}
#| label: ch9-quadratic
#| echo: true
m2q <- lm(income_k ~ education + prestige + I(prestige^2) + women + type,
  data = d)
anova(m2, m2q)
```

```{r}
#| label: fig-ch9-quadratic
#| fig-width: 6.4
#| fig-height: 3.4
#| fig-cap: "Component-plus-residual display for prestige in the model with a squared prestige term. The solid curve is the fitted quadratic component and the dashed line the linear component of the additive model; the labels give the curve's slope at prestige 30, 50 and 70."
#| fig-alt: "Open grey circles for the 98 occupations, prestige from 17.3 to 87.2 on the horizontal axis and component plus residual from about −6 to 18 on the vertical axis. A solid curve rises slowly at low prestige and more steeply at high prestige; a dashed straight line crosses it twice. Black dots on the curve at prestige 30, 50 and 70 carry the labels slope 0.0697, slope 0.1665 and slope 0.2633."
b_q <- coef(m2q)
component <- function(x) b_q["prestige"] * x + b_q["I(prestige^2)"] * x^2
centre <- mean(component(d$prestige))
partial_data <- data.frame(prestige = d$prestige,
  partial = component(d$prestige) - centre + residuals(m2q))
curve_grid <- data.frame(prestige = seq(17.3, 87.2, length.out = 200))
curve_grid$quadratic <- component(curve_grid$prestige) - centre
curve_grid$linear <- coef(m2)["prestige"] *
  (curve_grid$prestige - mean(d$prestige))
marks <- data.frame(prestige = c(30, 50, 70))
marks$partial <- component(marks$prestige) - centre
marks$label <- paste("slope", format(round(
  b_q["prestige"] + 2 * b_q["I(prestige^2)"] * marks$prestige, 4),
  digits = 15, scientific = FALSE, trim = TRUE))
ggplot2::ggplot(partial_data, ggplot2::aes(prestige, partial)) +
  ggplot2::geom_point(shape = 1, colour = "grey65") +
  ggplot2::geom_line(data = curve_grid, ggplot2::aes(y = linear),
    linetype = 2) +
  ggplot2::geom_line(data = curve_grid, ggplot2::aes(y = quadratic),
    linewidth = .8) +
  ggplot2::geom_point(data = marks, size = 2.6) +
  ggplot2::geom_label(data = marks, ggplot2::aes(y = partial + 3.2,
    label = label), size = 3.2, linewidth = 0) +
  ggplot2::labs(x = "Prestige", y = "Prestige component plus residual")
```

The quadratic coefficient is 0.0024 with SE 0.0009 and $t=2.5954$; the
one-restriction block test gives $F(1,91)=6.736$ with $p=.011$; and the
criteria move from 474.6040 and 492.6988 to 469.6058 and 490.2855. Adding
the term also changes the education coefficient from 0.1164 to $-0.0298$,
a reminder that coefficients in an additive model are defined relative to
the other columns present.

@fig-ch9-quadratic, the component-plus-residual display of Chapter 8 with
the fitted curve added, shows what the squared term claims. With linear
coefficient $-0.0754$ and quadratic coefficient 0.0024, the curve's slope
is 0.0697 thousand dollars per prestige point at prestige 30, 0.1665 at
50, and 0.2633 at 70: income rises with prestige throughout the records,
and rises faster at the top, where the straight line of the additive
model falls below several high-prestige occupations. The turning point,
15.5906, lies below the smallest observed prestige of 17.3, so the
descending branch of the parabola is entirely outside the records.
Reporting a "negative prestige coefficient" from this model would misread
the coordinate $-0.0754$, which is the slope at prestige zero, a value no
occupation has.

Centering repairs that coordinate problem without changing the curve,
because it is a change of coordinates. Prestige and its square correlate
.9849; centering prestige at its mean, 47.3276, before squaring lowers
that correlation to .3216. The quadratic coefficient and its SE stay
0.0024 and 0.0009, and every fitted value stays the same. The linear
coefficient becomes 0.1536 with SE 0.0347: it now estimates the slope at
the mean prestige, a different quantity from $-0.0754$, the slope at
prestige zero, whose SE is about 2.6 times as large. The smaller SE
belongs to a slope read in the middle of the records, not to a sharper
estimate of the old one. Keep the status of the exercise straight:
this is a disclosed sensitivity check on functional form, declared before
the refit, and its result is that a modest curvature is detectable on
these records. It is not a search across candidate terms for the smallest
$p$-value, and a $p$-value read after such a search would not mean what it
appears to mean.

### A log-transformed outcome

Changing a predictor's form keeps the question. Changing the outcome's
scale changes the question, and that difference is the point of this
subsection.

$$
\log Y_i=\beta_0+\sum_j\beta_jx_{ij}+\varepsilon_i,\qquad
100\,(e^{\beta_j}-1)\ \text{percent per unit of } x_j .
$$

A coefficient in a log-income model is an additive difference in expected
log income. Exponentiating turns it into a multiplicative difference in
geometric mean income under that model, and $100(e^{\widehat\beta}-1)$
expresses that multiplier as a percentage. The geometric mean is not the
arithmetic mean, so this is not automatically a statement about average
income in dollars.

```{r}
#| label: ch9-log-outcome
#| echo: true
m_log <- lm(log(income) ~ education + prestige + women + type, data = d)
100 * (exp(coef(m_log)[c("education", "prestige", "women")]) - 1)
```

| Term | Log coefficient | Percent | Lower | Upper |
|:--|--:|--:|--:|--:|
| education | 0.0076 | 0.7661 | $-4.2041$ | 5.9941 |
| prestige | 0.0207 | 2.0965 | 1.4351 | 2.7622 |
| women | $-0.0085$ | $-0.8430$ | $-1.0189$ | $-0.6667$ |

: Log-model contrasts on a geometric-mean percent scale. {#tbl-ch09-log tbl-colwidths="[22,22,18,19,19]"}

One more prestige point is associated with about 2.0965 percent higher
geometric-mean income at the same education, percentage of women, and
type, with interval $[1.4351,\ 2.7622]$ percent; one more percentage point
of women with about $-0.843$ percent. The two type coefficients, $-0.0137$
and 0.0708 on the log scale, are $-1.3584$ and 7.3341 percent. The
intervals in @tbl-ch09-log are the classical log-scale intervals with both
endpoints transformed, which is legitimate because the transformation is
monotone; they inherit the log-scale working model and supply no new
warrant.

The criteria will not travel across this change. This model's AIC is
8.091 against `m2`'s 474.604, and the difference says nothing at all:
the two likelihoods are densities of different response variables, so
their values are not on a common scale. The discipline to keep is simple:
fix the outcome and the question inside any comparison, and treat a scale
change as a new question rather than as data cleaning [@fox2019].

### Coefficient scales: standardized coefficients and unique contributions

Two summaries are often asked for once several predictors are in an
equation, and both are more limited than their names suggest:

$$
\beta_j^{*}=\widehat\beta_j\frac{s_{x_j}}{s_y},\qquad
\Delta R^2_j=R^2_{\mathrm{full}}-R^2_{\mathrm{without}\ j}.
$$

A **standardized coefficient**, or beta weight, rescales a partial slope
into outcome-SD units per predictor SD. A **drop-one $R^2$** reports how
much of the explained variation disappears when one predictor is removed
from the full model. The beta weights are 0.0757, 0.5685, and $-0.3971$
for education, prestige, and the percentage of women, and education's
drop-one $R^2$ is .0007: removing it from the full model costs essentially
nothing in fit, the same message the education coefficient's wide interval
carried in Chapter 8.

Both quantities are rescalings of an in-sample fit, and neither measures
importance. A standardized coefficient depends on the predictor's observed
spread in these particular records, so a restricted range shrinks it for
reasons that have nothing to do with the relationship. It also differs
from the standardization over named profiles that Chapter 8 used, which
averages fitted values over a declared distribution rather than dividing
by an SD. The drop-one change describes only the *unique* contribution:
because predictor information overlaps, the separate drop-one values need
not sum to the joint $R^2$, and a predictor that duplicates another can
show a near-zero value while mattering. Neither number establishes a
causal ranking or predicts held-out performance.

## Robust Uncertainty Under Unequal Spread {#sec-ch9-hc3}

### The HC3 covariance

Every classical standard error so far assumed one common error variance
for all 98 occupations. @fig-ch9-spread shows why that assumption is
doubtful here.

```{r}
#| label: fig-ch9-spread
#| fig-width: 6.4
#| fig-height: 3
#| fig-cap: "Residuals of the full occupation model by type, with each type's residual SD."
#| fig-alt: "Jittered open circles of residuals in thousand dollars, one column per type, around a horizontal line at zero. Blue-collar and white-collar residuals stay between about −2.4 and 2.6, with SDs 1.0437 and 1.0243; professional residuals run from about −7.8 to 14.2, with SD 4.3242."
spread <- data.frame(type = as.numeric(d$type), residual = residuals(m2))
type_sd <- data.frame(type = 1:3, label = paste("SD", format(round(
  tapply(spread$residual, d$type, sd), 4), digits = 15, scientific = FALSE,
  trim = TRUE)))
ggplot2::ggplot(spread, ggplot2::aes(type, residual)) +
  ggplot2::geom_hline(yintercept = 0, colour = "grey45") +
  ggplot2::geom_point(shape = 1, position = ggplot2::position_jitter(
    width = .12, height = 0, seed = 12345)) +
  ggplot2::geom_text(data = type_sd, ggplot2::aes(type, 15.5,
    label = label), size = 3.3) +
  ggplot2::scale_x_continuous(breaks = 1:3,
    labels = c("Blue-collar", "Professional", "White-collar")) +
  ggplot2::labs(x = NULL, y = "Residual (thousand dollars)") +
  ggplot2::theme(panel.grid.major.x = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank())
```

The residuals of professional occupations spread about four times as
widely as the others: their SD is 4.3242 thousand dollars against 1.0437
for blue-collar and 1.0243 for white-collar occupations. A single $s^2$
averages these very different spreads, so it overstates the uncertainty
contributed by some rows and understates it for others.

The remedy is already in hand. Chapter 7, Section 6.1, first used a
robust standard error on the simple regression line, where the education
slope's SE moved from 0.1270 to 0.1491. Chapter 8 then applied the same
HC3 estimator to the full model: its Section 3.3 computed the education
interval with `sandwich::vcovHC()`, and its Appendix B displayed the
estimator, which keeps the design component outside and replaces the
single $s^2$ inside by each row's own leverage-adjusted squared
residual, and set classical and HC3 standard errors side by side for all
six coefficients [@white1980; @mackinnon1985; @long2000; @zeileis2004].
Appendix A of this chapter shows where the estimator comes from.

What the figure adds is a reason for the pattern in that table. The
professional coefficient's SE grows by about 40%, from 1.4631 to 2.0928,
as the wide professional spread predicts, while the white-collar SE grows
by only about 7%, from 0.9848 to 1.0548. The education SE grows from
0.2752 to 0.3823, and its 95% interval on the robust covariance is
$[-0.6429,\ 0.8758]$: wider than the classical one, with the same
reading of an imprecisely estimated comparison. Every coefficient, fitted
value, and residual stays exactly what it was, because HC3 replaces a
covariance estimate, not a fit.

The percentage of women is the instructive case: its SE *falls* from
0.0097 to 0.0079. A robust covariance is not a penalty applied to every
coefficient.
It is a different estimate of the same target, and where the classical
common-variance assumption happened to overstate a coefficient's
variability, the robust version is smaller. Reporting only the
coefficients whose intervals widened, or switching estimators after seeing
which choice gives the preferred conclusion, turns a stated method into a
selection.

### A block test with a robust covariance

The restriction statistic of Section 3.2 needs only a coefficient vector
and a covariance matrix. Substituting the HC3 covariance
$\widehat{\mathbf V}_{\mathrm{HC3}}$ for $\widehat{\mathbf V}$ gives an
approximate Wald version of the same block test. For the type block it
gives 0.0345, with $p=.9661$, against
the classical 0.0319 and .9686: here the substitution changes nothing of
substance. The reason is that the type coefficients sit close to the
restriction, each within a quarter of its classical SE of zero, so both
statistics are near zero; the SEs would have to shrink about tenfold
before either statistic reached the .05 critical value, 3.0954. A change
of covariance matters most when a block statistic lies near its critical
value.

The reference distribution is where the two differ in kind. The classical
statistic has an exact $F_{2,92}$ distribution under the stated Normal
linear model. The robust version has only an approximate reference,
justified as the sample grows, and $F$ with the same degrees of freedom is
a convention rather than a derivation. Quoting the robust $p$-value to
several digits, or treating it as the same object as the classical one
because both are labelled $F$, overstates what the calculation delivers.

### What HC3 does not repair

HC3 has one job. Under a maintained linear conditional mean and
independent observations, it estimates the coefficient covariance without
assuming a common error variance.

It does not repair curvature: if the mean function is wrong, the
coefficients estimate a different quantity (Appendix A names it), and a
robust covariance describes that quantity's variability more honestly
without making it the quantity of interest. It does not repair dependence:
rows that are not independent break the estimator's own derivation, and
Section 7 shows a case where a robust row-level SE and a naive one agree
closely while both understate the uncertainty by nearly half. It does not
repair selection or missing outcomes, which change what the records
represent, and it does not supply causal identification, which comes from
design or from an argument about the conditioning set. Error in measuring
a predictor also pulls its slope toward zero, by roughly the predictor's
reliability when it is the only predictor, and no covariance estimator
repairs it [@cohen2003]; independent mean-zero error in the outcome, by
contrast, widens the residual spread and the standard errors but leaves
the slope's target unchanged. Chapter 14, Section 5.1, develops
attenuation.
Nor does HC3 create joint support: the type contrast's robust SE of 1.1827
is still the uncertainty of a comparison the records barely observe.

The practical response is to say which estimator was used and why, and to
keep the answer's limits attached to it. Chapter 8's advice holds here:
report the named comparison, its form and influence sensitivities, and the
covariance calculation, rather than treating any one of them as a
pass/fail certificate.

## Resample the Records {#sec-ch9-bootstrap}

### Why resample, and which scheme

Every uncertainty statement so far came from a formula: a coefficient
covariance matrix, classical or robust, propagated through a linear
combination. Resampling offers a different approximation. Refit the model
to many resamples of the records, and read the spread of the refitted
estimates directly [@efron1979; @efron1993; @davison1997]. Where the
formula and the resampling distribution disagree, the disagreement is
itself informative.

The choice of what to resample is an assumption, not a default.

- **Case resampling** draws whole records with replacement and refits. It
  inherits whatever unequal spread the rows contain, but it requires the
  resampling unit to match the sampling unit, with units independent and
  exchangeable.
- **Residual resampling** holds the design fixed, resamples rescaled
  residuals, and adds them back to the fitted values. It assumes
  identically distributed errors, the assumption HC3 exists to relax.
- **The wild bootstrap** also holds the design fixed but multiplies each
  rescaled residual by an independent mean-zero, unit-variance draw, so
  every record keeps its own error scale.

This chapter uses case resampling, for two stated reasons. The occupation
predictors were not designed: no one assigned education or prestige to an
occupation, so a joint-sampling view of the rows is the natural one. And
the residual spread differs by type, as @fig-ch9-spread showed, so the
identically distributed errors behind residual resampling are not
available [@freedman1981]. Appendix B compares the other two schemes on
the same targets.

### Case resampling for the full model

Resample the 98 rows with replacement, refit `m2`, and keep the education
slope. The seed and the number of resamples $B$ belong in the record, so
both appear in the code; $B$ counts bootstrap resamples here, as it
counted simulation repetitions in Chapter 5.

```{=latex}
\Needspace{14\baselineskip}
```

```{r}
#| label: ch9-bootstrap
#| echo: true
set.seed(90909)
B <- 2000
boot_education <- replicate(B, {
  rows <- sample.int(nrow(d), replace = TRUE)
  coef(lm(income_k ~ education + prestige + women + type,
    data = d[rows, ]))["education"]
})
c(SE = sd(boot_education), quantile(boot_education, c(.025, .975)))
```

```{r}
#| label: fig-ch9-bootstrap
#| fig-width: 6.4
#| fig-height: 3
#| fig-cap: "The 2,000 case-resampled education slopes (seed 90909). Solid lines mark the percentile interval; the dashed curve is the Normal distribution with the classical SE 0.2752."
#| fig-alt: "Histogram of refitted education slopes from about −1.2 to 1.1, centred near 0.1 and slightly longer on the left, with solid vertical lines at −0.683 and 0.744. A dashed Normal curve centred at 0.1164 is narrower and taller than the histogram."
normal_curve <- data.frame(x = seq(-1.5, 1.5, length.out = 300))
normal_curve$density <- dnorm(normal_curve$x, coef(m2)["education"],
  sqrt(vcov(m2)["education", "education"]))
ggplot2::ggplot(data.frame(slope = boot_education),
    ggplot2::aes(slope)) +
  ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = 40, fill = "grey80", colour = "white") +
  ggplot2::geom_line(data = normal_curve, ggplot2::aes(x, density),
    linetype = 2) +
  ggplot2::geom_vline(xintercept = quantile(boot_education,
    c(.025, .975)), linewidth = .7) +
  ggplot2::labs(x = "Education slope in a resample (thousand dollars per year)",
    y = "Density")
```

@fig-ch9-bootstrap is the resampling distribution itself. The 2,000
refitted slopes spread more widely than the Normal curve built from the
classical SE: their SD, the bootstrap SE, is 0.3699, and their middle 95%,
the percentile interval, runs from $-0.683$ to 0.744. The same resamples
give the other two targets of @tbl-ch09-bootstrap, the
professional-minus-white-collar contrast and the professional coefficient.
A resample could omit an entire occupational type, and its refit would
then lose a type coefficient; here all 2,000 resamples contained all
three types, so all 2,000 refits are usable, and that count is part of
the result rather than an assumption.

| Target | Estimate | Classical SE | HC3 SE | Bootstrap SE | Percentile 2.5% | Percentile 97.5% |
|:--|--:|--:|--:|--:|--:|--:|
| Education slope | 0.1164 | 0.2752 | 0.3823 | 0.3699 | $-0.6830$ | 0.7440 |
| prof $-$ wc | 0.0889 | 1.0800 | 1.1827 | 1.1187 | $-1.9613$ | 2.4313 |
| prof coefficient | 0.3242 | 1.4631 | 2.0928 | 1.9926 | $-3.1416$ | 4.5443 |

: Model-based, HC3, and case-bootstrap uncertainty for three targets in the full occupation model (seed 90909, $B=2000$). {#tbl-ch09-bootstrap tbl-colwidths="[22,12,14,12,14,13,13]"}

Read the table one row at a time. The education slope's bootstrap SE of
0.3699 sits between the classical 0.2752 and the HC3 0.3823: three
calculations, one conclusion that this comparison is imprecise. The type
contrast's bootstrap SE of 1.1187 exceeds the classical 1.0800, with
percentile interval $[-1.9613,\ 2.4313]$. The professional coefficient
shows the largest movement, from a classical 1.4631 to 1.9926, close to
the HC3 2.0928; with only 31 professional occupations spreading as widely
as @fig-ch9-spread shows, a resample's composition matters, and the
formula that assumes a common error variance is the most optimistic of the
three.

Agreement between HC3 and the bootstrap is not independent confirmation.
Both are sensitivity procedures applied to the same 98 records, and both
inherit the same limitations of those records; agreement means the two
approximations behave similarly here, not that the answer has been
replicated. Monte Carlo variation also means that the last decimal of a
bootstrap summary carries no weight: a different seed would move it, which
is exactly why the seed and $B=2000$ are reported.

### Which interval, and what resampling does not repair

Two intervals come directly from the resampling distribution. The
**percentile interval** takes its 2.5th and 97.5th percentiles, as in
@fig-ch9-bootstrap. The **normal-approximation interval** takes the
original estimate plus and minus 1.96 bootstrap standard errors. The two
agree when the resampling distribution is roughly symmetric; here the
three distributions have skewness $-0.3891$, 0.2635, and 0.3785, read by
their sign as in Chapter 2, Section 2.1: the education slopes have the
slightly longer left tail visible in the figure, the other two a slightly
longer right tail. The asymmetry is mild enough that the two kinds of
endpoint agree closely. For a strongly skewed or biased resampling
distribution, **bootstrap-$t$** and **BCa** intervals are the higher-order
alternatives; Appendix B defines both in words.

What resampling cannot do matters more than which interval to choose.
Drawing occupations with replacement treats them as independent and
exchangeable draws from one population, and there are stated reasons to
doubt both. The occupations share one historical labour market, where
common conditions can move related occupations' incomes together; the
98 rows were chosen by complete records rather than sampled at random
from a population of occupations; and their measurements are aggregates
over workers. Resampling does not repair the thin cross-type support, correct the measurement problems Chapter 8
documented, or convert an observational comparison into a causal one. And
if a design contained clusters, strata, or weights, the resampling unit
would have to follow the design rather than the rows, which is the subject
of the next section.

## Dependence in the Design: Project STAR {#sec-ch9-star}

```{=latex}
\hypertarget{sec-ch9-star}{}
```

```{r}
#| label: ch9-star-prepare
data("STAR", package = "AER")
star_vars <- c("stark", "readk", "lunchk", "gender", "ethnicity",
  "experiencek", "schoolidk")
star_two_arm <- !is.na(STAR$stark) & STAR$stark %in% c("regular", "small")
star_df <- droplevels(STAR[star_two_arm &
  complete.cases(STAR[, star_vars]), star_vars])
star_df$stark <- relevel(star_df$stark, ref = "regular")
star_observed <- droplevels(STAR[star_two_arm & !is.na(STAR$readk),
  c("stark", "readk", "schoolidk")])
star_observed$stark <- relevel(star_observed$stark, ref = "regular")
```

Every calculation so far has used observational records or one small
experiment, and every qualification has been about what those records
cannot support. This section changes the design. Project STAR randomized
kindergarten pupils to class types within their schools, which supplies a
warrant no amount of adjustment can manufacture, and a dependence
structure that the uncertainty calculation must follow.

### A different design supplies a different warrant

The question is specific:

> Among eligible kindergarten pupils in the participating Tennessee
> schools, what was the mean reading-score difference under assignment to
> a small rather than a regular class?

The outcome is `readk`, an end-of-kindergarten reading scale score. The
field `stark` records the kindergarten class-type condition. Reading its
small-versus-regular comparison as an assignment comparison rests on the
documented experimental protocol, not on the column name or on the `lm()`
call: random assignment occurred **within** participating schools
[@finn1990; @mosteller1995]. The regular-with-aide arm is outside this
declared two-arm comparison and is excluded by design, not dropped for
convenience.

Three limits belong with the question from the start. Class **type** is
not a measured headcount, so this analysis estimates no dose response in
pupils per room. The distributed field does not independently verify each
pupil's original randomized assignment; an assignment-effect reading
additionally requires linking the recorded type to the assignment record
and checking switches and implementation deviations, which this teaching
reanalysis does not supply. And the scientific target is the
pupil-weighted mean assignment effect for the eligible cohort in
participating schools, while a common coefficient from a complete-record
model is an estimator that needs further links, a common effect across
schools or an explicit weighting argument, to reach that target.

The data are distributed as `AER::STAR`, accompanying *Applied
Econometrics with R* and following the version used by Stock and Watson
[@kleiber2008; @stock2007]. They are loaded from the package rather than
redistributed here, and the analysis is retrospective rather than
preregistered.

### Separate eligibility from missingness

Two different reasons for a record's absence from an analysis are easy to
merge and must be kept apart. Some distributed rows belong to pupils who
entered the study after kindergarten: they have no kindergarten condition
at all, and their absence is **structural ineligibility**. Other rows
belong to eligible kindergarten pupils whose reading score or covariates
were not recorded: that is **missingness**, and it can distort a
comparison. Only the variables a calculation actually uses may enter its
complete-record rule; in particular, a missing mathematics score (`mathk`)
must not exclude an otherwise usable reading record.

\Needspace{12\baselineskip}

| Stage | $n$ |
|:--|--:|
| Distributed records | 11,598 |
| Kindergarten condition recorded | 6,325 |
| Eligible small or regular | 4,094 |
| Eligible, reading observed | 3,745 |
| Complete primary records | 3,735 |
| Primary regular | 2,001 |
| Primary small | 1,734 |
| Primary schools | 79 |

: STAR eligibility and reading-analysis completeness. {#tbl-ch09-star-sample tbl-colwidths="[60,40]"}

Of 11,598 distributed records, 5,273 have no kindergarten condition; they
are later entrants, not 5,273 failed measurements. The remaining 6,325
have a recorded kindergarten arm, and across all three arms 536 lack a
reading score. Restricting to the two declared arms leaves 4,094 eligible
records, of which 349 lack reading, 188 in the regular arm and 161 in the
small arm, leaving 3,745 with an observed reading score, 2,006 regular and
1,739 small.

Similar missing percentages in the two arms are reassuring but prove
nothing: the absent scores could differ systematically within either arm,
and no observed value can reveal that. Complete-record inference here
requires an assumption that, given the declared covariates and within arm
and school, observation does not depend on the reading score that would
have been recorded. Stating the assumption is not verifying it. Among the
observed-reading records across all arms, 38 lack at least one of lunch
status, gender, ethnicity, and teacher experience; within the two focal
arms, ten do. The primary common-covariate sample therefore has 3,735
pupils, 2,001 regular and 1,734 small, in 79 schools, and the
observed-reading sample keeps all 3,745 focal records as a declared
sensitivity, not as an alternative headline.

### The binary slope recovers the observed arm means

Section 1.1's identity applies unchanged. With $D_i=1$ for the small arm
and 0 for regular,

$$
\widehat Y_i=\widehat\alpha+\widehat\delta D_i,\qquad
\widehat\alpha=\bar Y_{D=0},\qquad
\widehat\delta=\bar Y_{D=1}-\bar Y_{D=0}=440.5474-434.7323=5.8151.
$$

On the 3,745 observed-reading records the regression slope reproduces the
arm difference of 5.8151 reading points exactly, as Chapter 7's binary
predictor did on 24 pupils. The identity is algebraic and requires no
randomization; what randomization adds is the reason to read the
difference as an assignment comparison.

Two cautions attach to the number itself. Under independent
common-variance rows, the slope's $t$ test equals the pooled two-sample
test, here $t=5.6039$; that row-level reference ignores the fact that
pupils sit in schools, which Section 7.5 addresses. And raw arm means
weight each school by its own arm-specific pupil count, so the pooled
difference equals the pupil-weighted cohort target only under common
assignment probabilities and comparable observation patterns, or after
explicit standardization to target school weights.

### Compare two specifications on identical primary rows

The adjusted model adds free-lunch status, gender, ethnicity, teacher
experience, and school indicators. School indicators respect the
assignment blocks; the remaining covariates are the declared adjustment
specification, not a substitute for randomization.

```{r}
#| label: ch9-star-models
#| echo: true
star_unadjusted <- lm(readk ~ stark, data = star_df)
star_primary <- lm(readk ~ stark + lunchk + gender + ethnicity +
  experiencek + schoolidk, data = star_df)
```

On the common 3,735 primary rows, the unadjusted difference is 5.8955 and
the adjusted coefficient is 6.6575. Adjustment can improve precision when
the added variables are prognostic for reading, and school indicators
match the blocks in which assignment happened. It is not needed to remove
baseline confounding that a successfully implemented randomization
already prevents, and it does not automatically preserve a causal target:
timing, selection, and the fitted comparison still need scrutiny. This is
Section 2.5's point about randomized groups, on a larger design.

The earlier 5.8151 comes from 3,745 rows and must not serve as the
same-row unadjusted baseline for 6.6575. Comparing 5.8151 with 6.6575
would confound two changes, a change of records and a change of
specification, and the whole discipline of this chapter is to keep those
apart. The same-row pair to compare is 5.8955 against 6.6575. Similar
unadjusted and adjusted coefficients are consistent with successful
randomization; they are not proof that implementation, missingness, or
weighting problems are absent.

### Cluster uncertainty follows the design

Pupils are the outcome units. Schools are the randomization blocks and the
approximately independent units for the uncertainty calculation; the
schools themselves were not assigned to arms. Chapter 3, Section 2.3,
showed that the uncertainty of a school-blocked comparison follows its
within-school comparisons, and @fig-ch9-star-schools shows those
comparisons for STAR.

```{r}
#| label: fig-ch9-star-schools
#| fig-width: 6.4
#| fig-height: 3
#| fig-cap: "Small-minus-regular mean reading difference in each of the 78 schools that observed both class types, ordered by size. The dashed line is their unweighted mean, 6.7094."
#| fig-alt: "Dot plot of 78 school differences, one dot per school, rising from about −32 to 59 reading points in an S-shaped column; 25 dots lie left of the solid zero line, and a dashed line marks the mean, about 6.71."
school_means <- with(star_observed,
  tapply(readk, list(schoolidk, stark), mean))
school_diff <- sort(school_means[, "small"] - school_means[, "regular"])
ggplot2::ggplot(data.frame(difference = school_diff,
    school = seq_along(school_diff)), ggplot2::aes(difference, school)) +
  ggplot2::geom_vline(xintercept = 0, colour = "grey45") +
  ggplot2::geom_vline(xintercept = mean(school_diff), linetype = 2) +
  ggplot2::geom_point(size = 1.4) +
  ggplot2::labs(x = "Small minus regular mean reading score in the school",
    y = "Schools, ordered") +
  ggplot2::theme(axis.text.y = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank())
```

In the 78 schools that observed pupils in both arms, the small-class
advantage ranges from about $-32$ to 59 reading points, and it is negative
in 25 schools. Pupils in one school share whatever made their school's
difference large or small: its teachers, its intake, its classrooms. The
3,745 pupils therefore do not supply 3,745 independent pieces of
information about the arm difference; the schools do. A rough check makes
the scale visible. An average of 78 independent differences with SD
15.4192 has SE $15.4192/\sqrt{78}=1.7459$, the size of the clustered
standard errors below and far from the row-level 1.0377.

For school $g$, form its score $\mathbf s_g=\mathbf X_g^{\mathsf T}\mathbf e_g$
by summing the residual contributions **within** the school before taking
an outer product. The CR1/HC1 cluster-robust covariance is then

$$
\widehat{\mathbf V}_{\mathrm{CR1}}=c_G(\mathbf X^{\mathsf T}\mathbf X)^{-1}
\Big(\sum_{g=1}^{G}\mathbf s_g\mathbf s_g^{\mathsf T}\Big)(\mathbf X^{\mathsf T}\mathbf X)^{-1},
\qquad \mathbf s_g=\mathbf X_g^{\mathsf T}\mathbf e_g,\qquad
c_G=\frac{G}{G-1}\frac{n-1}{n-p}.
$$

The within-school cross-products are exactly what distinguishes this from
treating every pupil as independent, and $c_G$ is the finite-cluster
correction. With $G=79$ schools, intervals use a $t_{78}$ critical value,
1.9908 [@zeileis2020].

```{=latex}
\Needspace{10\baselineskip}
```

```{r}
#| label: ch9-star-cluster
#| echo: true
star_V <- sandwich::vcovCL(star_primary, cluster = star_df$schoolidk,
  type = "HC1", cadjust = TRUE)
sqrt(star_V["starksmall", "starksmall"])
```

\Needspace{12\baselineskip}

| Model | $n$ | Estimate | SE | Interval |
|:--|--:|--:|--:|:--|
| Observed: unadjusted | 3,745 | 5.8151 | 1.8476 | [2.1368, 9.4935] |
| Primary: unadjusted | 3,735 | 5.8955 | 1.8491 | [2.2143, 9.5767] |
| Primary: adjusted | 3,735 | 6.6575 | 1.6974 | [3.2782, 10.0368] |
| Observed: school-adjusted | 3,745 | 6.6273 | 1.7797 | [3.0842, 10.1703] |

: Class-type comparisons and school-clustered intervals. {#tbl-ch09-star-estimates tbl-colwidths="[34,12,16,14,24]"}

@tbl-ch09-star-estimates holds four distinct calculations, each with its
own records and its own specification; each row's interval uses the
school identifiers from that model's own rows. Changing the model or its
records is a different act from changing the covariance calculation, and
the table keeps them separable.

Set the clustered SE beside Section 5's robust one. For the same
3,745-record unadjusted comparison, treating pupils as independent rows
gives SE 1.0377; HC3 gives 1.0416, barely different; school clustering
gives 1.8476, nearly twice as large. Unequal spread across pupils was not
this design's problem. Dependence within schools was, and no row-level
robust covariance addresses it. The adjusted primary model shows the same
pattern: its row-level SE is 0.9161 and its HC3 SE 0.9509, against the
clustered 1.6974.

These are cluster-asymptotic approximations with 79 clusters, not exact
randomization inference and not a multilevel model. A residual plot
cannot choose the independence unit; the design does. And a count of 79
clusters does not certify the approximation: the balance of information
across clusters and the influence of particular schools also matter.

### Read the bounded answer, then evaluate its support

A chapter-authored report can now state:

> On 3,735 complete two-arm kindergarten records, the common adjusted
> small-versus-regular class-type coefficient was 6.6575 reading points,
> with approximate school-clustered interval [3.2782, 10.0368]. Its causal
> mean-effect interpretation relies on documented within-school
> randomization, a verified link from recorded class type to original
> assignment, suitable complete-record selection, independent-school
> uncertainty, and a common-effect or target-matching weighting argument.
> It concerns assigned class type, one kindergarten reading outcome, and
> participating Tennessee schools, not realized headcount or a portable
> policy dose response.

Read the report's functions before judging it, applying Chapter 1's
argument components as Chapters 7 and 8 did. The sample audit and the
model code carry **STAT-E** content; the design and covariance reasoning
supplies **STAT-J**; the estimate and its clustered interval supply
**STAT-I**; and the first sentence is the bounded **STAT-C** answer.
**SUB-E** situates the pupils, the test, the cohort, and the school
context; **SUB-J** adds the assignment record together with the selection
and weighting qualifications; **SUB-I** reads the coefficient as
assignment-related only under that support; and **SUB-C** bounds the
substantive claim. These are linked functions within one argument, not two
arguments.

Then separate what is *stated* from what is *demonstrated*. The report
states the selection and weighting conditions; stating them does not
establish that they hold. The randomization protocol is documented in the
study's own publications, not shown by this coefficient table. The link
from the recorded class-type field to the original assignment is asserted
here and not verified. Where a necessary link cannot be assessed, the
honest move is to report the limitation rather than to invent it or to
quietly substitute a different target.

The design literature documents implementation and interpretation issues
that a single coefficient cannot settle [@krueger1999; @hanushek1999;
@nye2000; @chetty2011]. Nor does this result transport by itself to
nonparticipating schools, another period, another outcome, or another
education system. Those limits belong inside the answer, not in a list
appended after an unrestricted claim, and a crossed significance boundary
is not what makes a result [@gelman2006; @gelman2020]. Chapter 14 returns
to dependence as a design fact, reuses this calculation as its example,
and asks what a design supplies that adjustment cannot.

## Reading Bridge {#sec-ch9-summary}

A factor enters a regression as indicators, and its fitted means,
contrasts, and omnibus test are the analysis of variance written as a
model; a contrast's uncertainty needs the covariance between coefficients,
not the sum of two standard errors. The analysis of covariance is the same
regression with a pretest added: the posttest-only, gain-score, and
adjusted analyses differ only in the pretest coefficient, and adjustment
sharpens a randomized comparison without supplying what randomization
supplies. Nested models compare on the same records and the same outcome
by an extra-sum-of-squares $F$, and information criteria rank them by
penalized in-sample fit, never by held-out error. A squared predictor
gives a comparison that depends on the predictor's value; a logged outcome
changes the question to a geometric-mean scale, and its criteria no longer
speak to the original one. HC3, the case bootstrap, and a clustered
covariance each change the uncertainty calculation for a stated reason and
leave the fitted coefficients alone; none repairs curvature, missing
outcomes, weighting, or the absence of a design. Project STAR showed the
uncertainty following the design: schools, not pupils, were the
approximately independent units, and the difference was nearly a factor of
two.

Chapter 10 asks a different question of the same equations. Every
comparison in this chapter was the same at every value of the other
predictors, which is what "additive" means, and Section 2.5's common-slope
check was the first test of it. Chapter 10 lets a comparison itself vary
with another predictor, and develops how to read, probe, and report such
an interaction.

\Needspace{14\baselineskip}

### Key terms {.unnumbered}

Use this short reference to revisit the concepts introduced here. The
Part One Concept Reference gives the broader reference and the principal
home of each concept; indicator coding, partial slopes, coefficient
covariance, and HC3 are explained in Chapters 7 and 8, and regression
toward the mean in Chapter 7.

| Term and common alternatives | Meaning |
|:--|:--|
| ANCOVA (analysis of covariance) | A regression comparing groups while including a quantitative covariate with one common slope; its adjusted group means are read at a stated covariate value within common support. The posttest-only, gain-score and ANCOVA analyses are one regression whose pretest coefficient is 0, 1 or estimated; adjustment cannot supply what random assignment supplies. |
| Homogeneity of regression slopes (parallel slopes) | ANCOVA's condition that the outcome--covariate slope is the same in every group, checked by testing the group-by-covariate product block; when the slopes differ, the adjusted difference depends on the covariate value at which it is read. |
| Nested-model $F$ test (extra sum of squares; block test) | Compares a reduced model with a fuller one fitted to the same records and outcome: the RSS reduction per added coefficient, divided by the fuller model's residual mean square, referred to $F$ with the number of restrictions and the fuller model's residual df. It tests the added block jointly, not each coefficient separately. |
| Information criteria (AIC; BIC) | Penalized in-sample fit: $-2\log\widehat L$ plus $2k$ (AIC) or $k\log n$ (BIC), with $k$ counting the coefficients and the residual variance. Smaller is preferred among models fitted to the same records and outcome scale; neither is held-out error. |
| Polynomial term (quadratic term; curvature) | A power of a predictor added as a further column so that the fitted mean can curve. The predictor's comparison then depends on its value; centering before squaring reduces collinearity without changing the fitted curve. |
| Log-transformed outcome (geometric-mean scale) | Fitting the logarithm of the outcome, so that a coefficient is an additive difference in expected log outcome and $100(e^{\beta}-1)$ percent is a multiplicative difference in geometric means. A changed outcome scale is a changed question, and its criteria are not comparable with the original scale's. |
| Standardized coefficient (beta weight) | A partial slope rescaled by the predictor and outcome SDs, in outcome-SD units per predictor SD, with the other predictors held fixed. It differs from standardization over named profiles and does not measure causal importance. |
| Bootstrap (case resampling; percentile interval) | Refitting the model to many resamples of the records drawn with replacement and using the spread of the refitted estimates as an SE, or their middle 95% as a percentile interval. It treats the resampled units as independent and exchangeable; it does not repair dependence, thin support, measurement error, or an observational design. |
| Cluster-robust standard error (school-clustered covariance) | A coefficient covariance that sums residual contributions within clusters, such as schools, before combining them, allowing dependence within clusters under an approximation that treats clusters as independent. It changes uncertainty, not the coefficient, and does not repair missing outcomes or weighting. |

: Selected concepts for the regression toolbox. {#tbl-ch09-key-terms tbl-colwidths="[28,72]"}

### Further reading {.unnumbered}

- The [OpenIntro multiple-predictor and model-selection tutorial](https://openintro-ims.netlify.app/model-mlr)
  gives concrete indicator interpretations and criterion-based reading. We
  use its equation-first device without adopting automatic stepwise
  selection.
- The [ISLR regression lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch3-linreg-lab.html)
  and the [ISLR resampling lab](https://hastie.su.domains/ISLR2/Labs/Rmarkdown_Notebooks/Ch5-resample-lab.html)
  connect fitting, factors, criteria, and resampling in one workflow
  [@james2021].
- Fox [@fox2016] and Kutner and colleagues [@kutner2005] develop the
  linear-model algebra, the restriction statistic, and the diagnostics
  behind this chapter's matrix section.
- Efron and Tibshirani [@efron1993] and Davison and Hinkley [@davison1997]
  are the standard bootstrap references, including the bootstrap-$t$ and
  BCa intervals named in Section 6.3.
- The [sandwich documentation](https://cran.r-project.org/web/packages/sandwich/sandwich.pdf)
  specifies the HC and CR covariance calls used here [@zeileis2020].

## Exercises {.unnumbered}

Use the supplied numbers; rounding within 0.02 is acceptable unless a
different precision is requested. These prompts do not require the
appendices.

\Needspace{12\baselineskip}

<!-- prompt-id: CH09-E01 -->
### Exercise 1: Read indicator coefficients (Foundational) {.unnumbered #ex-ch09-01}

A group-only regression of income (thousand dollars) on occupation type
has observed group means 5.3741 (blue-collar, 44 records), 10.5595
(professional, 31), and 5.0523 (white-collar, 23). (1) With blue-collar as
reference, give the intercept and the two indicator coefficients. (2) Give
the intercept and both coefficients with white-collar as reference. (3)
Explain why the fitted values are the same under both codings. (4) The
full additive model's professional coefficient is 0.3242. Explain why it
differs from the professional coefficient in (1) and why neither is "the
effect of occupational type".

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E02 -->
### Exercise 2: A contrast needs covariance (Foundational) {.unnumbered #ex-ch09-02}

In the full additive model the professional and white-collar coefficients
are 0.3242 and 0.2353, with estimated variances 2.1406 and 0.9698 and
covariance 0.9720, on 92 residual df. (1) The
professional-minus-white-collar contrast is 0.0889. Calculate its
variance, and explain why the covariance enters with a minus sign and what
a positive covariance does to the result. (2) The contrast's SE is 1.08
and its 95% interval is $[-2.0561,\ 2.2338]$. A student instead adds the
two coefficient SEs (1.4631 and 0.9848). State what that number is and why
it is wrong. (3) Explain what the interval does and does not say about the
two types.

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E03 -->
### Exercise 3: A nested comparison from residual sums of squares (Foundational) {.unnumbered #ex-ch09-03}

On the same 98 records, the model with education and prestige has RSS
867.6537 on 95 df, and the full model adding the percentage of women and
the two type indicators has RSS 630.8897 on 92 df. The 5% critical value
of $F(3,92)$ is 2.7036. (1) Calculate the $F$ statistic for the added
block. (2) State the decision at 5% and what it does and does not say
about the three added columns. (3) Explain why the sequential sums of
squares from a single `anova()` call on the full model would not answer
this question.

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E04 -->
### Exercise 4: What information criteria compare (Foundational) {.unnumbered #ex-ch09-04}

The full additive model has maximized log-likelihood $-230.302$ on 98
records, with AIC 474.6040 and BIC 492.6988; the model with education and
prestige only has AIC 499.8328 and BIC 510.1727. (1) The criteria count
seven fitted parameters for the full model. Say what the seven are, and
why BIC charges the full model more for them than AIC does
($\log 98=4.585$). (2) Which model do the criteria favour, and on what
condition is that comparison meaningful? (3) A log-income version of the
full model has AIC 8.091. Explain why this number cannot be compared with
474.604.

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E05 -->
### Exercise 5: Percent and curvature (Integrative) {.unnumbered #ex-ch09-05}

Two changed models of the occupation records raise different questions.
(1) In a log-income model the prestige coefficient is 0.0207. Give the
percent difference in geometric-mean income per prestige point and say
what "geometric mean" adds to the reading. (2) In an income model with a
squared prestige term, the fitted slope is 0.0697 thousand dollars per
prestige point at prestige 30, 0.1665 at 50 and 0.2633 at 70, and the
turning point is at prestige 15.5906; prestige ranges from 17.3 to 87.2.
Explain why the model has no single prestige slope, and what the position
of the turning point tells you. (3) State in one sentence what the curve
claims about income and prestige over the observed range, and one reason
that claim is a sensitivity check rather than a discovery.

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E06 -->
### Exercise 6: Choose the uncertainty calculation (Integrative) {.unnumbered #ex-ch09-06}

In Project STAR the small-minus-regular kindergarten reading difference on
3,745 observed records is 5.8151 points. Three standard errors are
available: 1.0377 treating pupils as independent, 1.0416 with HC3, and
1.8476 clustering pupils within their 79 schools. Treating pupils as
independent gives the 95% interval $[3.7807,\ 7.8496]$. (1) Using
$t_{.975,78}=1.9908$, give the interval under the clustered calculation
and compare its width with the first. (2) Say which interval a report
should use and why, naming the unit at which assignment happened, and
explain why HC3 barely changes the first standard error. (3) Name two
problems that none of the three calculations repairs.

\Needspace{10\baselineskip}

<!-- prompt-id: CH09-E07 -->
### Exercise 7: One randomized study, three analyses (Integrative) {.unnumbered #ex-ch09-07}

In the Baumann records, 66 pupils in three randomized instruction groups
have a pretest and a posttest. With DRA as reference, the DRTA difference
is 3.0909 (SE 0.9614) from the posttest alone, 3.8636 (SE 0.7779) from the
gain score, and 3.6266 (SE 0.7362) from the analysis of covariance, whose
pretest coefficient is 0.6932; the DRA and DRTA pretest means are 10.5000
and 9.7273. (1) Write the three analyses as one regression and give the
pretest coefficient each uses. (2) Explain why all three estimate the same
average difference under random assignment, and show how the pretest gap
accounts for the differences between the three estimates. (3) Say which
analysis you would report, and when that choice has to be made. (4)
Suppose the groups had been intact classes that differed at the pretest.
Explain what changes, naming the paradox that can arise.

## Appendix A: Matrix Geometry, Restrictions, and Sandwich Constructions {.unnumbered #app-ch9-a}

This appendix derives the results that Sections 3.2, 5.2, and 7.5 used,
including the HC3 estimator that Chapter 8, Appendix B, stated, and
names the target least squares estimates when a linear mean is only an
approximation. It is reference material: the exercises do not require
it, and the main route is complete without it. The Computer Lab carries
the numerical reproductions of each construction.

### Projection and the normal equations {.unnumbered}

Let $\mathbf X$ be $n\times p$ with full column rank, so
$\mathbf X^{\mathsf T}\mathbf X$ is invertible. The least-squares
criterion is
$S(\boldsymbol\beta)=(\mathbf y-\mathbf X\boldsymbol\beta)^{\mathsf T}(\mathbf y-\mathbf X\boldsymbol\beta)$.
Expanding gives
$S=\mathbf y^{\mathsf T}\mathbf y-2\boldsymbol\beta^{\mathsf T}\mathbf X^{\mathsf T}\mathbf y
+\boldsymbol\beta^{\mathsf T}\mathbf X^{\mathsf T}\mathbf X\boldsymbol\beta$,
and differentiating with respect to $\boldsymbol\beta$ yields the normal
equations

$$
\mathbf X^{\mathsf T}\mathbf X\widehat{\boldsymbol\beta}=\mathbf X^{\mathsf T}\mathbf y,
\qquad\text{equivalently}\qquad
\mathbf X^{\mathsf T}(\mathbf y-\mathbf X\widehat{\boldsymbol\beta})=\mathbf 0 .
$$

The second form is the geometry: the residual vector is orthogonal to
every column of $\mathbf X$. Hence
$\widehat{\mathbf y}=\mathbf X\widehat{\boldsymbol\beta}=\mathbf H\mathbf y$
with $\mathbf H=\mathbf X(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}$,
and $\mathbf H$ is symmetric and idempotent, the orthogonal projection
onto the column space of $\mathbf X$. Its trace is $p$, which is why the
leverages $h_{ii}$ sum to the number of fitted coefficients; for `m2` they
sum to 6.

Two consequences used earlier follow immediately. Any full-rank recoding
$\mathbf X\mathbf A$, with $\mathbf A$ invertible, spans the same column
space, so $\mathbf H$ and $\widehat{\mathbf y}$ are unchanged: releveling a
factor or centering a predictor cannot alter a single fitted value. Adding
a genuinely new column enlarges the space and can change every fitted
value. And under the classical error model,
$\widehat{\mathbf V}=s^2(\mathbf X^{\mathsf T}\mathbf X)^{-1}$ with
$s^2=RSS/(n-p)$, which is where every classical standard error in this
chapter came from. Solving the normal equations directly for `m2`
reproduces `lm()`'s coefficients to about $10^{-13}$.

### General linear restrictions {.unnumbered}

Let $\mathbf R$ be $q\times p$ of rank $q$ and consider
$\mathbf R\boldsymbol\beta=\mathbf r$. Since $\widehat{\boldsymbol\beta}$
has covariance $\sigma^2(\mathbf X^{\mathsf T}\mathbf X)^{-1}$ under the
classical model, the vector $\mathbf R\widehat{\boldsymbol\beta}-\mathbf r$
has covariance
$\sigma^2\mathbf R(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf R^{\mathsf T}$.
Standardizing by that covariance and dividing by $q$ gives the statistic of
Section 3.2, which has an exact $F_{q,\,n-p}$ reference under Normal
errors, because its numerator and denominator quadratic forms are
independent chi-squares.

It is the same number the extra-sum-of-squares formula produces.
Minimizing $S(\boldsymbol\beta)$ subject to
$\mathbf R\boldsymbol\beta=\mathbf r$ with a Lagrange multiplier gives the
restricted estimator
$\widetilde{\boldsymbol\beta}=\widehat{\boldsymbol\beta}
-(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf R^{\mathsf T}
[\mathbf R(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf R^{\mathsf T}]^{-1}
(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)$, a correction applied
only in the directions the restriction touches. Substituting it into the
criterion shows that the increase in residual sum of squares is

$$
RSS_R-RSS_F=(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r)^{\mathsf T}
[\mathbf R(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf R^{\mathsf T}]^{-1}
(\mathbf R\widehat{\boldsymbol\beta}-\mathbf r),
$$

which is exactly the numerator of the restriction statistic. Fitting the
reduced model and differencing, or evaluating the quadratic form in the
unrestricted fit, are two routes to one number; for the two type
coefficients of `m2` both return 0.0319. Replacing
$s^2(\mathbf X^{\mathsf T}\mathbf X)^{-1}$ by a robust covariance in the
middle of the expression gives the Wald statistic of Section 5.2. The
algebra survives the substitution; the exact reference distribution does
not.

### The HC3 sandwich {.unnumbered}

Chapter 8, Appendix B, stated the HC3 estimator; this subsection shows
where it comes from. Write the coefficient error as
$\widehat{\boldsymbol\beta}-\boldsymbol\beta
=(\mathbf X^{\mathsf T}\mathbf X)^{-1}\mathbf X^{\mathsf T}\boldsymbol\varepsilon$.
Under independent rows with possibly different variances $\sigma_i^2$, its
exact covariance is

$$
(\mathbf X^{\mathsf T}\mathbf X)^{-1}
\Big(\sum_i\sigma_i^2\,\mathbf x_i\mathbf x_i^{\mathsf T}\Big)
(\mathbf X^{\mathsf T}\mathbf X)^{-1},
$$

a *sandwich* with the design component as bread and the error-scale
component as filling. The classical estimator substitutes a common $s^2$,
which collapses the filling to $s^2\mathbf X^{\mathsf T}\mathbf X$. The HC
family substitutes each row's own squared residual instead. Because
$E(e_i^2)=\sigma_i^2(1-h_{ii})$ under the classical model, the raw residual
understates the error at high-leverage rows; HC3 corrects by dividing by
$(1-h_{ii})^2$, which is deliberately slightly conservative and performs
well in small samples [@mackinnon1985; @long2000]. Built directly from
these pieces, the estimator reproduces `sandwich::vcovHC(m2, type = "HC3")`
to about $10^{-12}$.

### The CR1 cluster sandwich {.unnumbered}

Cluster robustness changes the filling, not the bread. Partition the rows
into $G$ clusters and let $\mathbf X_g$ and $\mathbf e_g$ hold cluster
$g$'s rows and residuals. The cluster score
$\mathbf s_g=\mathbf X_g^{\mathsf T}\mathbf e_g$ sums the within-cluster
contributions *before* the outer product is taken, so the estimator of
Section 7.5 retains the within-cluster residual correlations that the
independent-row filling discards. If every cluster held one row, the
filling would reduce to the HC0 form, which is why a row-level robust SE
is the special case that assumes away the dependence.

The asymptotics are in $G$, not in $n$: 3,745 pupils in 79 schools supply
79 approximately independent pieces of information for this covariance,
which is why the clustered SE of 1.8476 is so much larger than the
row-level 1.0377 and why a $t_{G-1}$ reference is used. For the
observed-reading model the finite-cluster factor is $c_G=1.0131$, and
building the estimator from the school scores reproduces
`sandwich::vcovCL(..., type = "HC1", cadjust = TRUE)` to about $10^{-13}$.

### What least squares estimates when the mean is not linear {.unnumbered #app-ch09-projection-target}

A linear regression is often fitted where the conditional mean of the
outcome is not exactly linear in the predictors, as Section 4.1's curved
prestige component shows. Least squares then still estimates a defined
target: the **best linear projection**, the coefficients that minimize
expected squared error over the population the records represent,

$$
\begin{gathered}
(\beta_0^*,\beta_1^*)=\arg\min_{a,b}E\{(Y-a-bX)^2\},\qquad
\beta_1^*=\frac{\operatorname{Cov}(X,Y)}{\operatorname{Var}(X)}, \\
\boldsymbol\beta^*=\{E(\mathbf x\mathbf x^{\mathsf T})\}^{-1}E(\mathbf x Y).
\end{gathered}
$$

With one predictor the target slope is the population covariance over the
predictor's variance, the population version of Chapter 7's
covariance-to-slope identity; with several, it is the population version
of the normal equations. The target is well defined, but it is an
average: the projection slope of income on prestige in the additive model
averages the curve's steeper and flatter stretches, weighted by where the
occupations' prestige values fall. Its value therefore depends on the
distribution of the predictors, and a sample with a different prestige
mix would have a different target even with the same curve. It is a
declared alternative target, not a rescue for a misspecified mean. HC3
standard errors remain appropriate for it under independent sampling,
because the curvature the linear fit misses behaves like one more source
of unequal spread [@fox2016].

## Appendix B: Bootstrap Schemes and Intervals {.unnumbered #app-ch9-b}

Section 6 used case resampling and said why. @tbl-ch09-schemes sets it
beside the other two schemes on the same model, the same three targets,
the same seed, and $B=2000$.

| Scheme | What is resampled | Assumption carried | Education slope | prof $-$ wc | prof coefficient |
|:--|:--|:--|--:|--:|--:|
| Case | Whole records | Records independent and exchangeable | 0.3699 | 1.1187 | 1.9926 |
| Residual | Rescaled residuals, moved across records; design fixed | One error distribution for every record | 0.2820 | 1.1081 | 1.4861 |
| Wild | Each record's own rescaled residual, sign flipped at random; design fixed | Each record keeps its own error scale | 0.3673 | 1.1151 | 1.9767 |

: Three bootstrap schemes for the full occupation model, with each scheme's bootstrap SE for three targets (seed 90909, $B=2000$). {#tbl-ch09-schemes tbl-colwidths="[11,27,24,13,12,13]"}

Both design-fixed schemes rebuild outcomes from the fitted values.
Residual resampling draws leverage-rescaled residuals $e_i/\sqrt{1-h_{ii}}$
with replacement and adds them to the fitted values, which treats the
rescaled residuals as exchangeable across rows. The wild bootstrap keeps
each row's own rescaled residual and multiplies it by $\pm1$ with equal
probability, so each row retains its own error scale.

Read the table by column. For the education slope the residual scheme is
the outlier, 0.2820, close to the classical 0.2752, which is what one
should expect, since both assume one common error distribution. The wild
bootstrap, which does not, lands at 0.3673, near the case result and near
HC3's 0.3823. For the professional coefficient the pattern repeats more
strongly: 1.9926, 1.4861, and 1.9767. When the schemes disagree, the
disagreement is about the error model, not about the data, and it is a
reason to say which scheme was used.

Four interval definitions are worth distinguishing in words.

- The **percentile interval** takes the 2.5th and 97.5th percentiles of
  the resampling distribution of the estimate.
- The **normal-approximation interval** takes the original estimate plus
  and minus 1.96 bootstrap standard errors. It uses the resampling
  distribution only for its spread, so it is insensitive to skewness.
- The **bootstrap-$t$ interval** resamples a studentized quantity, the
  resampled estimate minus the original estimate, divided by the
  resample's own standard error, and inverts the empirical quantiles of
  that ratio. It needs a standard error inside every resample, and it is
  more accurate when the studentized quantity is closer to pivotal than
  the estimate itself.
- The **BCa interval** adjusts the percentile endpoints for median bias
  and for an acceleration constant that captures how the standard error
  changes with the parameter. It is the usual recommendation when the
  resampling distribution is visibly skewed [@efron1993; @davison1997].

For the three targets of Section 6 the skewness values $-0.3891$, 0.2635,
and 0.3785 are mild, so percentile and normal-approximation endpoints agree
closely and the higher-order corrections would change little. That
agreement is a property of this example, not a general licence.

## References {.unnumbered}

::: {#refs}
:::
