---
title: "Extending the Argument-Based Approach to Quantitative Studies"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-cap: true
    tbl-cap: true
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(dev = "cairo_pdf")
}

draw_workflow <- function() {
  old <- par(mar = rep(0.2, 4), xpd = NA)
  on.exit(par(old))
  plot.new()
  plot.window(xlim = c(0, 14), ylim = c(0, 8))

  rect(0.15, 4.15, 13.85, 7.60, col = "#EEF5F8", border = "#9BB6C4")
  rect(0.15, 0.25, 13.85, 3.70, col = "#F1F6EF", border = "#A7BFA2")
  text(0.38, 7.30, "PHASE 1 - UNDERSTAND AND PLAN | iterate within Stages 1-3",
       adj = 0, font = 2, cex = 0.69, col = "#244A63")
  text(0.38, 3.40, "PHASE 2 - PRODUCE AND COMPLETE | iterate within Stages 4-7",
       adj = 0, font = 2, cex = 0.65, col = "#466B43")

  top_x <- c(0.48, 5.565, 10.65)
  labels <- c(
    "Literature\nand problem",
    "Questions, answers,\nand status",
    "Align design, measures,\nestimands, evidence,\nand inference",
    "Produce or acquire\nand document evidence",
    "Estimate, check,\nand analyse",
    "Contextualize, justify,\ninterpret, and complete\nbounded answers",
    "State outcomes\nand synthesize"
  )

  stage_box <- function(x0, y0, number, label) {
    rect(x0, y0, x0 + 2.72, y0 + 1.62, col = "white",
         border = "#344A55", lwd = 1)
    text(x0 + 0.18, y0 + 1.34, paste0("STAGE ", number), adj = 0,
         font = 2, cex = 0.60, col = "#203744")
    text(x0 + 1.36, y0 + 0.66, label, cex = 0.66, col = "#29353B")
  }

  bottom_x <- c(10.65, 7.26, 3.87, 0.48)
  for (i in 1:3) stage_box(top_x[i], 4.65, i, labels[i])
  for (i in 1:4) stage_box(bottom_x[i], 0.75, i + 3, labels[i + 3])
  for (i in 1:2) {
    arrows(top_x[i] + 2.74, 5.46, top_x[i + 1] - 0.02, 5.46,
           length = 0.07, col = "#4F6875")
  }
  for (i in 1:3) {
    arrows(bottom_x[i] - 0.02, 1.56,
           bottom_x[i + 1] + 2.74, 1.56,
           length = 0.07, col = "#557052")
  }
  arrows(top_x[3] + 1.36, 4.63, top_x[3] + 1.36, 2.39,
         length = 0.07, col = "#6B6B6B")
  arrows(11.35, 6.58, 1.20, 6.58, length = 0.07, lty = 2,
         col = "#7A4B2A")
  text(6.25, 6.84, "revise and realign before evidence production or acquisition",
        cex = 0.65, col = "#7A4B2A")
  arrows(3.20, 0.52, 12.00, 0.52, length = 0.07, lty = 2,
         col = "#7A4B2A")
  text(7.60, 0.33, "reanalyse, reinterpret, and recalibrate within Phase 2",
        cex = 0.65, col = "#7A4B2A")
}

draw_statistical_strands <- function() {
  old <- par(mar = rep(0.15, 4), xpd = NA)
  on.exit(par(old))
  plot.new()
  plot.window(xlim = c(0, 20), ylim = c(0, 13.8))

  ink <- "#203744"
  e_fill <- "#DCEAF3"
  j_fill <- "#DDEEDB"
  i_fill <- "#F8EDC7"
  c_fill <- "#F5D9C8"
  link <- "#9A5A35"

  rect(0.25, 0.72, 12.55, 13.35, col = "#F5F9FB",
       border = "#9BB6C4")
  rect(13.05, 0.72, 19.75, 13.35, col = "#F7FAF6",
       border = "#A7BFA2")
  text(0.55, 12.98, "STATISTICAL LAYER", adj = 0, font = 2,
       cex = 0.78, col = "#244A63")
  text(0.55, 12.52, "shared E/J core; two estimand-specific strand columns",
       adj = 0, cex = 0.62, col = "#56656C")
  text(13.38, 12.98, "SUBSTANTIVE LAYER", adj = 0, font = 2,
       cex = 0.78, col = "#466B43")
  text(13.38, 12.52, "one integrated answer", adj = 0,
       cex = 0.62, col = "#56656C")

  component_box <- function(cx, y0, y1, width, code, label, fill,
                            code_cex = 0.72, label_cex = 0.61) {
    rect(cx - width / 2, y0, cx + width / 2, y1,
         col = fill, border = ink, lwd = 1.05)
    text(cx, y1 - 0.30, code, font = 2, cex = code_cex, col = ink)
    text(cx, mean(c(y0, y1)) - 0.17, label, cex = label_cex,
         col = "#29353B")
  }

  stat_cx <- 6.30
  stat_w <- 10.60
  strand_cx <- c(3.60, 9.00)
  strand_w <- 5.00
  sub_cx <- 16.35
  sub_w <- 5.20
  rows <- c(E = 10.85, J = 8.60, I = 5.80, C = 2.85)

  component_box(stat_cx, 10.20, 11.50, stat_w, "STAT-E CORE",
                "Shared assignments, cohort,\nand timing records", e_fill)
  component_box(stat_cx, 7.95, 9.25, stat_w, "STAT-J CORE",
                "Shared randomized-design warrant", j_fill)
  text(stat_cx, 9.70, "+", font = 2, cex = 1.05, col = ink)

  component_box(strand_cx[1], 5.05, 6.55, strand_w,
                "STRAND 1  |  STAT-I1",
                "Score estimate, interval,\nchecks, and status", i_fill,
                code_cex = 0.68, label_cex = 0.59)
  component_box(strand_cx[2], 5.05, 6.55, strand_w,
                "STRAND 2  |  STAT-I2",
                "Completion estimate,\nuncertainty, and checks", i_fill,
                code_cex = 0.68, label_cex = 0.59)
  component_box(strand_cx[1], 2.10, 3.60, strand_w, "STAT-C1",
                "Score input to joint criterion\nwith stated limits", c_fill,
                label_cex = 0.59)
  component_box(strand_cx[2], 2.10, 3.60, strand_w, "STAT-C2",
                "Completion input to joint criterion\nwith stated limits", c_fill,
                label_cex = 0.59)

  # The shared core feeds both strands; each strand then moves I -> C.
  arrows(strand_cx[1], 7.90, strand_cx[1], 6.60,
         length = 0.06, col = ink)
  arrows(strand_cx[2], 7.90, strand_cx[2], 6.60,
         length = 0.06, col = ink)
  arrows(strand_cx[1], 5.00, strand_cx[1], 3.65,
         length = 0.06, col = ink)
  arrows(strand_cx[2], 5.00, strand_cx[2], 3.65,
         length = 0.06, col = ink)

  sub_specs <- list(
    E = c(10.20, 11.50, "SUB-E", "Situated students, outcomes,\nand setting"),
    J = c(7.95, 9.25, "SUB-J", "Meaning, implementation,\nand scope"),
    I = c(5.05, 6.55, "SUB-I", "Integrates both\nstrand results"),
    C = c(2.10, 3.60, "SUB-C", "Reports whether the joint\ncriterion was met")
  )
  sub_fills <- c(E = e_fill, J = j_fill, I = i_fill, C = c_fill)
  for (nm in names(sub_specs)) {
    spec <- sub_specs[[nm]]
    component_box(sub_cx, as.numeric(spec[1]), as.numeric(spec[2]), sub_w,
                  spec[3], spec[4], sub_fills[[nm]], label_cex = 0.59)
  }
  text(sub_cx, 9.70, "+", font = 2, cex = 1.05, col = ink)
  arrows(sub_cx, 7.90, sub_cx, 6.60, length = 0.06, col = ink)
  arrows(sub_cx, 5.00, sub_cx, 3.65, length = 0.06, col = ink)

  # Ochre outlines make convergence explicit without running connectors
  # through either strand. Each paired row maps to one like SUB component.
  rect(0.92, 4.90, 11.68, 6.70, border = link, lwd = 1.05, lty = 2)
  rect(0.92, 1.95, 11.68, 3.75, border = link, lwd = 1.05, lty = 2)
  arrows(11.72, rows["E"], 13.70, rows["E"], length = 0.07,
         col = link, lwd = 1.10)
  arrows(11.72, rows["J"], 13.70, rows["J"], length = 0.07,
         col = link, lwd = 1.10)
  arrows(11.72, rows["I"], 13.70, rows["I"], length = 0.07,
         col = link, lwd = 1.10)
  arrows(11.72, rows["C"], 13.70, rows["C"], length = 0.07,
         col = link, lwd = 1.10)

  text(10, 0.28,
       "Each strand adds specific E/J  |  Ochre groups like components  |  One declared question, not an omnibus across questions",
       cex = 0.58, col = "#56656C")
}
```

{{< pagebreak >}}

::: {.callout-tip title="Chapter box"}
**Central question:** How can a reader extend the argument-based approach from
a simple study to a relatively complex quantitative study with coordinated
questions whose estimands, evidence, models, and interpretations interact?

**Focal components:** Every question--answer pair retains its own argument with
linked statistical and substantive layers. To support that reading task, the
chapter distinguishes multiple question-level arguments from multiple
statistical strands inside one argument. It then extends the basic six-stage
construction workflow to seven stages for
coordinated questions, develops seven quality-evaluation questions, and
shows how regression can instantiate question-specific estimands.

**Main evidence:** An upgraded hypothetical special-practice study with 320
randomized participants, three distinct questions, imperfect attendance,
missing delayed outcomes, section structure, adjustment, and interaction; and a
compact transfer application to the Deslauriers et al. published study and its
supporting materials.

**Scope:** Reading quantitative studies remains the approach's **primary use**.
Quality evaluation is its **secondary use**: it follows and depends on reading
but is not part of reading's definition. The construction workflow is its
**auxiliary use** for researchers conducting a study. *How to Read a
Quantitative Study as an Argument* is
sufficient for a simple study with one primary question, or with multiple
independent questions that can be addressed separately. This advanced tutorial
is optional and is not a prerequisite for Chapters 2--14. It addresses
relatively complex studies with coordinated questions. Its construction
workflow and quality-evaluation questions retain those ranked uses. The
workflow covers much, but not all, of the research process; the evaluation questions
do not produce a mechanical score; and regression
does not supply measurement meaning, design support, substantive importance,
or generalizability by itself. All learner-directed activities follow Section
9 in the unnumbered Exercises section.
:::

## Coordinating Question-Specific Arguments in More Complex Studies

### What remains unchanged

The core rule remains:

> Each question--answer pair has one argument with linked layers.

The basic tutorial can therefore be applied more than once when a study has
several independent questions that can be read separately. The
advanced treatment becomes useful when questions are coordinated through
shared estimands, evidence, models, priorities, or synthesis, so that decisions
and interpretations for one pair can affect another.

Within each statistical or substantive layer, Evidence and Justification
support Inference, which supports a bounded Claim:

$$
E + J \longrightarrow I \longrightarrow C.
$$

The statistical layer addresses what the recorded evidence supports about a
defined estimand. The substantive layer situates, augments and contextualizes,
interprets, and bounds that result as an answer about the people, practices, settings, or
phenomena named in the question.

Complexity does not create permission to merge questions that are owed
separately reportable answers. If a study separately asks about an average assignment effect,
variation in that effect, and the relationship between attendance and
performance, it owes three separately reportable question-facing answers. The
three question--answer pairs therefore have three arguments. They may share the
same participants, design records, outcome, and fitted model, but shared support
does not make their obligations identical.

### What becomes more demanding

Four extensions matter in this chapter.

First, **several questions require coordination**. A primary question may be
prespecified, a secondary question may ask about effect variation, and an
exploratory question may arise after evidence production has begun. Such a new
question starts its own explicitly exploratory argument cycle; it is not added
retrospectively to a planned cycle. Priorities and analytic status must remain
visible. Study-level synthesis should integrate question-specific answers
without turning them into a new unsupported conclusion.

Second, **the evidence-generating process becomes less tidy**. Participants may
enter through nonprobability recruitment, attend different amounts, miss an
outcome, or be nested in sections. These features affect what counts as
evidence, which estimand can be estimated, and how uncertainty should be calculated.

Third, **the three uses remain ranked and distinct**. Reading the arguments is
the primary use. Quality evaluation is the secondary use because it judges what
reading first identifies. The construction workflow is the auxiliary use for
researchers building the support. Stages and evaluation questions are not
additional argument layers.

Fourth, **regression can express several estimands in one mathematical language**.
A group indicator can represent a randomized comparison; baseline variables can
improve precision; an interaction can express effect variation. The equation
does not supply design, measurement meaning, or substantive importance by
itself.

### Shared material, question-specific obligations

Suppose two questions use the same delayed score. The score records are shared
STAT-E. Yet the average-effect question and the effect-variation question use
those records differently. The latter additionally requires adequate variation
and overlap in baseline preparation, an interaction specification, and enough
precision to distinguish meaningful patterns.

Likewise, a common randomization record supports the assignment-effect question
directly. It does not support a causal effect of attendance, because attendance
occurs after assignment and is not itself randomized. The role and adequacy of
support are always question-specific.

The same discipline applies to conclusions. “Assignment improved performance
on average” does not imply “the improvement was larger for lower-preparation
students.” An inconclusive interaction is not evidence that effects are equal.
A positive average effect and uncertain variation can coexist without
contradiction.

### One question can contain several statistical strands

A different structure arises when several statistical estimands are jointly
needed for one integrated substantive answer. Each estimand then has a
**statistical strand** within the pair's one statistical layer:

$$
STAT\text{-}E_k + STAT\text{-}J_k \longrightarrow STAT\text{-}I_k
\longrightarrow STAT\text{-}C_k.
$$

The strands feed one substantive layer and one question-facing answer. An estimand
counts as a strand when it is subordinate to one prespecified integrated answer
obligation and is not itself owed or interpreted as a standalone substantive
conclusion. Separately owed conclusions form separate pairs. Independent
answerability is a useful diagnostic, but the number of hypotheses,
coefficients, tests, estimands, tables, or pooled estimates does not decide.
A hypothesis is a proposition about one or more estimands, not itself an
estimand. Neither hypothesis count nor estimand count mechanically determines
the number of pairs or strands.
Author wording and analytic status are cues to the intended obligations, not
permission to merge conclusions that the study actually owes separately.
The display may compress shared support as clarity permits; it cannot compress
separately owed conclusions into strands.

For a compact illustration, imagine that the upgraded study had asked one
integrated decision question: **Did assignment meet the prespecified success
criterion of a positive delayed-performance effect and no completion-rate
reduction greater than five percentage points?** One
strand addresses the score effect; the other addresses the completion-rate
difference and its prespecified non-reduction margin. Nonsignificance or nearly
equal observed rates alone would not establish non-reduction. The strands share
an assignment-and-design core but retain estimand-specific evidence,
justification, inferences, and statistical claims. SUB-I integrates the two
results, and SUB-C says whether the single success criterion was met while
preserving both sets of conditions and limitations. The two statistical claims
must still be reported transparently, but here they are inputs to that one
decision rather than standalone substantive conclusions. If the study instead
owed a separate question-facing conclusion for either outcome, it would have
separate pairs. The formulation is a granularity illustration, not a fourth
worked question below.

```{r}
#| label: fig-statistical-strands
#| fig-cap: "One question-level argument with two statistical strands and one substantive layer."
#| fig-alt: "A statistical panel has shared Evidence and Justification across its width. Beneath them, a score strand and a completion strand form two columns, each moving from its own Inference to its own bounded statistical Claim. Ochre outlines group the two Inference boxes and the two Claim boxes; short horizontal arrows map each grouped component, as well as the shared Evidence and Justification, to one corresponding component in a single substantive layer. A note says this is one declared integrated question, not an omnibus argument across separate questions."
#| fig-width: 6.4
#| fig-height: 4.7
#| fig-pos: H

draw_statistical_strands()
```

Parallel strands need not be statistically independent. Their E and J content
may have a shared core and estimand-specific additions, as the figure shows; a
common documentary trace can serve both without becoming either component.
Several estimators, checks, or sensitivity analyses for the same estimand also do
not create additional strands. Chapters 3 and 4 develop the formal estimand and
comparison distinctions.

## The Upgraded Special-Practice Study

### The study record

The university now implements the study in four first-year course sections.
Each section contains 80 consenting students. Within every section, 40 students
are randomly assigned to Special Practice and 40 to Usual Practice, giving
320 randomized participants in total.

As in the basic tutorial, *Special Practice* is a neutral name for the
hypothetical format rather than a conclusion about its quality.

Before assignment, students complete a 0--40 baseline-preparation assessment.
Both formats contain four scheduled sessions matched in duration and topic
coverage. Attendance is recorded initially as implementation information, not
as the focus of a research question. A 0--40 delayed assessment is scheduled
two weeks after the final session.

Unlike the basic example, not every participant attends most sessions or
completes the delayed outcome.

```{=latex}
\Needspace{0.30\textheight}
```

| Feature | Special Practice | Usual Practice |
|:--------|--------------------:|---------------:|
| Assigned students | 160 | 160 |
| Baseline score, M (SD) | 20.3 (5.6) | 19.9 (5.8) |
| Attended at least three sessions, n (%) | 118 (73.8%) | 126 (78.8%) |
| Delayed outcome observed, n (%) | 150 (93.8%) | 150 (93.8%) |
| Observed delayed score, M (SD) | 27.2 (5.4) | 24.1 (4.6) |

: *Descriptive record for the upgraded special-practice study.* {#tbl-upgraded-record tbl-colwidths="[52,24,24]"}

The two observed-outcome means differ by 3.1 points. That difference describes
the 300 observed outcomes. It is not automatically the primary answer for all
320 randomized students because outcome missingness and the planned estimand
must still be addressed.

### Three question--answer pairs

The study has three explicitly separated questions.

**Primary, prespecified assignment question**

> Among all 320 randomized participants, what was the average effect of
> assignment to Special rather than Usual Practice on the delayed score?

The estimand is the average assignment effect for all randomized participants.
The analysis uses the prespecified multiple-imputation and standardization
procedure described in @sec-advanced-missingness. It adjusts for baseline and
section while retaining all 320 participants in the population named by the
estimand. A
two-point difference was identified prospectively as a potentially important
benchmark.

The estimated assignment effect is 2.9 points, 95% interval [1.9, 4.0]. The
report gives a positive average assignment-effect answer under its stated
conditions. Because the interval includes values just below and above two
points, the reported answer does not claim that the effect exceeded the
importance benchmark.

**Secondary, prespecified effect-variation question**

> How did the Special-minus-Usual assignment effect vary with baseline
> preparation?

The fitted interaction estimates that the assignment effect changes by -0.8
points for each ten-point higher baseline score, 95% interval [-2.7, 1.1].
The compatible values include a meaningfully decreasing pattern, little or no
variation, and an increasing pattern. This analysis uses the same imputed
datasets and the same 320-person target cohort as the primary analysis. The
reported answer therefore does not resolve the direction or substantive size
of effect variation.

**Exploratory attendance question**

After evidence production had begun, the researchers formulated a new
attendance question. It therefore began a separate, explicitly exploratory
argument cycle rather than being inserted into the prospectively coordinated
cycles for Pairs 1 and 2. The new cycle reused attendance, assignment,
baseline, section, and outcome records, but its question and analytical plan
were retrospective to those already produced records.

> After accounting for assignment, baseline preparation, and section, how was
> the number of attended sessions associated with delayed score?

The adjusted association is 0.3 points per additional session, 95% interval
[-0.4, 0.9] among the 300 students with observed delayed outcomes. Modest
negative and positive observed-case associations both remain compatible with
the result, so it does not resolve direction. More importantly,
attendance was not randomized, and this exploratory analysis does not address
outcome missingness beyond restricting the analysis to observed cases. Even a
precise positive association would not automatically estimate the causal effect
of attending one more session.

### A question-to-argument map

| Pair | Estimand and status | Result | Reported answer and stated bounds |
|:-----|:------------------|:-------|:---------------|
| **1. Assignment effect** | Average effect of assignment among all 320 participants; primary and prespecified | 2.9 points, 95% interval [1.9, 4.0] | Positive average assignment effect under the missingness, implementation, and model conditions; exceeding two points remains uncertain |
| **2. Effect variation** | Change in assignment effect per ten-point higher baseline score among all 320 participants; secondary and prespecified | -0.8 points, 95% interval [-2.7, 1.1] | The result does not resolve the direction or substantive size of effect variation |
| **3. Attendance association** | Conditional association per additional attended session among 300 observed-outcome cases; exploratory | 0.3 points, 95% interval [-0.4, 0.9] | Modest negative and positive observed-case associations remain compatible; the analysis does not establish a causal effect of attendance |

: *Three question--answer pairs and their distinct arguments.* {#tbl-question-argument-map tbl-colwidths="[18,32,23,27]"}

The three arguments share recruitment, assignment, baseline, attendance, and
outcome records. They do not share every justification. Random assignment
supports Pair 1 and the assignment component of Pair 2. It does not randomize
baseline preparation or attendance. Pair 2 needs interaction-specific support.
Pair 3 needs an associational interpretation and explicit protection against
causal wording.

They are three pairs because the study owes a separately reportable,
question-facing conclusion for each one, not simply because the estimands or
analytic statuses differ. Independent answerability helps diagnose that
structure, while the authors' labels help locate the obligations; neither is a
mechanical counting rule.

### The primary argument across both layers

The primary question can be mapped without repeating the full eight-component
table from the basic tutorial.

**STAT-E** contains records for all 320 assignments, baseline scores, sections,
attendance, outcome availability, delayed scores, implementation, and scoring.
The 20 missing outcomes remain part of the evidence record even though their
scores are unavailable.

**STAT-J** records the study's offered reasons connecting the within-section
randomization and all-randomized estimand to baseline and section adjustment,
the prespecified multiple-imputation procedure, standardization,
interval-combination method, and planned checks. Its force
depends on implementation, absence of material interference, measurement
comparability, and a plausible connection between observed information and
missing outcomes.

**STAT-I** brings together the 2.9-point standardized estimate, 95% interval
[1.9, 4.0], the completed missingness stress test reported in
@sec-advanced-missingness, model checks, confirmatory status, and the two-point
benchmark. **STAT-C** states a positive average assignment contrast for the
320-person estimand under those conditions while retaining uncertainty about
whether the contrast exceeds two points.

Across layers, **SUB-E** situates the records as students in four sections,
assigned to two matched-time practice formats and assessed on a common delayed
course measure. **SUB-J** augments and contextualizes the statistical
justification with the intended educational meaning of the formats and
assessment, implementation evidence, plausible alternatives, the importance
benchmark, and limits on population transport.

**SUB-I** interprets the 2.9-point estimate and interval on the 0--40 delayed
assessment scale. It considers whether that magnitude is important without
declaring the benchmark crossed. **SUB-C** records the study's participant- and
implementation-bounded causal answer; Section 6 evaluates whether it is
supported.

This component map also shows why the observed-case difference of 3.1 and the
primary estimate of 2.9 are not competing answers. They describe different
analytical objects. One is a raw comparison among observed outcomes; the other
addresses the all-randomized estimand through the prespecified adjusted and
standardized procedure.

### Study-level synthesis

The report gives a study-level synthesis that integrates the three answers
without erasing their differences:

> Assignment to Special Practice improved delayed-assessment performance on
> average in this implementation. The study does not establish that the average
> improvement exceeded two points or that the assignment effect varied with
> baseline preparation. The observed-case attendance association was
> inconclusive, and the analysis does not establish that attending an additional
> session caused better performance.

The synthesis does not invent a fourth quantitative result. If the authors
wished to claim that Special Practice works specifically by increasing
attendance, that mechanism claim would require another question and argument.

## Extending Evidence, Measurement, Estimands, and Missingness

### Evidence is produced, not merely available

Evidence is not a neutral table handed to a statistical method. It records who
or what was observed, how observations were produced, which comparison the
design created, and which intended cases are absent.

For the upgraded study, STAT-E includes:

- eligibility, consent, section, and assignment records for all 320 students;
- baseline and delayed assessment scores and their scoring rules;
- session schedules, delivery, and attendance;
- the identities and timing of the 20 missing delayed outcomes; and
- processing and inclusion decisions.

The means and intervals are derived results in STAT-I, not the empirical
evidence itself. This narrower use of *Evidence* keeps the inferential move
visible even though everyday writing often calls results “evidence.”

### Population, unit, case, and level

The **target population or target set** contains the entities, settings, or
occasions to which an intended answer refers. The **sampling frame** is the
reachable set from which participants can arise. The **analytic records** are
the cases and observations used by a stated procedure. These sets need not
coincide.

The direct target cohort for Pair 1 is the 320 consenting participants. The
study does not use probability sampling from all first-year students. Random
assignment can support a causal comparison for these participants while wider
generalization remains uncertain.

The **unit** is the entity about which the question is asked. Here the student
is the unit of assignment and outcome measurement. Section is a randomization
block and a contextual grouping. Repeated records for one student remain
observations on one unit; they are not independent students.

Level matters because a relationship among students does not automatically
describe sections or institutions. Four sections also provide little direct
evidence about variation across the larger population of possible sections.
Including section indicators respects the blocked comparison but does not
justify a general claim about all course contexts.

### Estimands keep questions distinct

Recall the Chapter 1 contrast: the **estimand** is what the study seeks to
learn, whereas the **estimate** is the numerical result obtained from observed
evidence. The advanced point is that an estimand can answer a question directly
or contribute as one strand of an integrated answer. Defining estimands before
choosing estimators helps connect substantive questions, design and assumptions,
and empirical evidence [@lundberg2021].

Pair 1 has an estimand of the average assignment effect for all 320
participants. Pair 2 has an estimand describing variation in that assignment
effect with baseline preparation. Pair 3 has an estimand of an observed
conditional association with attendance. The same outcome
and regression family neither make these one estimand nor decide the number of
question--answer pairs.

Variable roles also come from questions. Baseline preparation is a precision
variable in Pair 1, a moderator in Pair 2, and an adjustment variable in Pair
3. Attendance is an implementation record for Pair 1 and the focal explanatory
variable in Pair 3. Its role cannot be inferred merely from its column position
in a dataset.

| Pair | Evidence and estimand obligation | Main vulnerability | Consequence for the claim |
|:-----|:-------------------------------|:-------------------|:--------------------------|
| **Assignment effect** | Preserve all 320 randomized participants in the target set and use baseline, section, implementation, and outcome records appropriately | Twenty delayed outcomes are missing; participation is from one university setting | The answer remains conditional on missingness handling and limited to this implementation and represented participants |
| **Effect variation** | Retain the randomized contrast while estimating how it changes over the supported baseline range | Interaction form, limited precision, and sparse values near the baseline extremes | Do not claim a subgroup pattern from the negative point estimate |
| **Attendance association** | Define attendance consistently and preserve its post-assignment timing | Motivation, difficulty, health, and other factors can affect both attendance and score | Report association only; do not inherit the causal support of assignment |

: *Question-specific evidence and estimand obligations.* {#tbl-evidence-target-obligations tbl-colwidths="[18,34,24,24]"}

### Observable measures still require interpretation

Both assessments are directly recorded, but neither is theoretically
transparent. Their meaning depends on content coverage, administration,
scoring, reliability, and evidence supporting the intended interpretation
[@aeraApaNcme2014; @kane2013].

The delayed score is evidence about performance on a defined course assessment.
It may support a bounded interpretation of delayed learning over the assessed
content, but it does not measure every kind of knowledge, transfer, motivation,
or long-term retention. Baseline preparation likewise represents performance
on a particular instrument, not a complete amount of preparation possessed by
each student.

Measurement comparability matters across assignment groups and baseline
levels. If one format teaches test-specific cues unavailable in the other, the
observed score difference may still be real while its interpretation as broader
learning becomes weaker. If the baseline scale behaves differently near its
ends, the interaction may partly reflect measurement rather than genuine
effect variation.

### Missing outcomes are part of the evidence-generating process {#sec-advanced-missingness}

Ten students in each assigned group lack the delayed outcome. Equal counts do
not make missingness harmless. The reasons could differ by baseline
preparation, section, attendance, assignment, or an unobserved outcome.

The primary estimand remains the average assignment effect for all 320
randomized participants. An analysis of only the 300 observed outcomes estimates
an observed-case contrast; it does not by itself change the declared estimand.
Without conditions connecting observed cases to all randomized cases, however,
that analysis does not identify the declared estimand.

The prespecified missing-outcome procedure uses assignment, section, baseline
preparation, the assignment-by-baseline relationship, attendance, and available
administrative information. It is designed to preserve the information needed
for both the average-effect and effect-variation questions while keeping all
320 randomized participants represented in those estimands. Attendance helps inform the
missing outcomes; it neither defines the assignment estimand nor becomes a
randomized exposure. Later chapters develop the imputation, model-fitting,
standardization, and uncertainty-combination mechanics.

This procedure assumes that, after accounting for the recorded predictors, the
unobserved delayed score does not add further information about whether it is
observed. That condition cannot be proved from observed data. Sensitivity
analysis should vary plausible departures and carry consequential changes into
STAT-C and SUB-C.

One completed stress test subtracts four points from each imputed missing
Special Practice outcome while leaving the Usual Practice imputations
unchanged. The standardized estimate becomes 2.6 points, 95% interval
[1.5, 3.7]. This deliberately unfavorable departure retains a positive
direction while leaving the exact magnitude uncertain and whether the effect
exceeds the two-point benchmark unresolved.
The next question is a tipping-point or claim-change question: how different
would plausible unobserved outcomes or assumptions need to be before the
bounded answer changed direction, crossed the two-point benchmark, or otherwise
changed materially? A departure that leaves materially important benefit and
harm both compatible would require an inconclusive directional answer. That
judgment depends on the question and meaningful benchmarks, not merely on
whether an interval crosses zero. Sensitivity is not a competition to find a
preferred result; it describes what the claim can withstand.

### Documentary trace and status

An argument component names a **logical function**; study-specific content
fills that function, and documentary traces show where the content can be
inspected. The components are not filing categories for documents. A shared
protocol can trace several functions without establishing that every reported
action was performed correctly or that every offered reason is sound.

A reader uses Chapter 1's four **documentary statuses**--**stated in the
paper**, **located in related materials**, **implicit but traceable**, and **not
located**--for the most direct available trace of relevant component content.

A contradiction across sources is a separate flag, not a fifth status. These
statuses describe what can be inspected; they do not evaluate its quality. A
protocol can establish that an estimand was prospectively stated without proving
that it was well chosen, and a complete script can establish computational
traceability without establishing measurement quality. Conversely, **not
located** does not prove that an underlying action never occurred.

Consider a deliberately mistaken alternative report, **not the upgraded
study's actual account**: “Because practice-format assignment was randomized,
we interpret the attendance association causally.” Reading records that
offered reason in **STAT-J**, with **stated in the paper** as its documentary
status. It remains an identifiable justification because that is the function
the authors give it, even though the reason is mistaken. Evaluation separately
finds that assignment to a format did not randomize attendance. Sharing the
assignment record therefore does not supply the same justification for these
different inferences.

Subject expertise helps readers recognize an offered function and make a
conventional implicit connection explicit when it is already traceable in the
record. Recognizing that connection is distinct from deciding whether it is
correct or sufficient. Expertise cannot supply an undocumented study-specific
premise, such as assuming that attendance itself was randomized.

A reader-added premise may be named during evaluation only as a proposed
repair. It does not count as support in the argument currently being read. The
word **proof** is therefore usually too strong: readers locate component
content through documentary traces and then evaluate adequacy separately.

## Research Design and the Claims It Can Support

### Representation and comparison are different

Research design makes an observed comparison informative. It supplies
distinguishable support for representation, comparability, timing,
measurement, and implementation. A study can be strong in one dimension and
weak in another [@shadish2002].

**Random sampling** primarily concerns who enters from a population and can
support population generalization when selection probabilities and
participation are handled appropriately. **Random assignment** concerns which
condition participants receive and can support causal comparison within the
study. One does not imply the other.

The upgraded study uses nonprobability participation and random assignment.
Its design can support an assignment-effect answer for the 320 participants,
but extension to other students, courses, institutions, or implementations
needs additional evidence.

### Design families and intended claims

An **observational study** observes conditions that arise without researcher
assignment. It can directly support description, association, and prediction.
Causal interpretation needs additional design-and-assumption support--often
called **identification support** because it explains why the intended causal
comparison can be learned from the observed comparison. Regression adjustment
alone does not create it.

A **randomized experiment** uses chance to assign conditions. Randomization
supports a causal comparison of assigned conditions when implementation,
outcome observation, interference, and analysis are handled appropriately.

A **quasi-experiment** uses a threshold, timing change, policy rule, or
comparison structure to approximate an experiment without random assignment.
Its causal support comes from design-specific conditions rather than the word
*quasi* or the presence of regression.

The strongest claim mode follows the question and design, not the complexity of
the model. Description, association, prediction, and causal comparison can all
use simple or elaborate analyses.

### Blocking, dependence, and interference

Assignment in the upgraded study occurs separately within four sections, so
section is a randomization block rather than the unit of assignment. The
analysis and uncertainty procedure should respect the within-section
comparison. A generic cluster correction based on only four sections is not an
automatic substitute; later chapters develop the relevant methods.

Students may interact within sections. **Interference** occurs when one
student's assigned format affects another student's outcome. Separate sessions
and monitoring may make material interference less plausible, but peer
discussion outside sessions could remain. The estimand should name the
assignment strategy actually implemented; the claim should retain any
unresolved spillover reservation.

### Assignment is not attendance

Randomization most directly supports the effect of **assignment**. Attendance
occurs after assignment and can be influenced by motivation, scheduling,
difficulty, health, or early reactions to the format. Comparing students who
attend different numbers of sessions generally loses randomization.

The attendance association is still useful descriptively. It can identify an
implementation pattern, motivate a future attendance intervention, or support
prediction. It cannot be promoted to the effect of receiving an additional
session without stronger assumptions and design.

### Analytic status and multiplicity

The primary and secondary questions were specified before outcomes were
examined; the attendance question is exploratory. These statuses affect how the
results should be calibrated but do not predetermine whether their answers are
positive or inconclusive.

Several outcomes, subgroups, interactions, and model variants create multiple
opportunities to find an interesting result. Multiplicity can be addressed
through priority, prespecification, adjusted uncertainty where appropriate,
transparent reporting, and independent confirmation. It cannot be solved by
reporting only the model that produced the most favourable estimate.

### Inferential conditions belong to particular arguments

Conditions should be linked to the inferential move they support rather than
recited as one universal checklist.

**Estimand and specification conditions** ask whether the population, contrast,
scale, functional form, and variable roles match the question. Pair 1 requires
an all-randomized average assignment estimand; Pair 2 additionally requires a
defensible scale and form for effect variation.

**Evidence-production conditions** concern recruitment, assignment, timing,
measurement, implementation, and processing. Within-section randomization
supports the assigned comparison, but only if the assignment record is
preserved and the outcome has a comparable meaning across formats.

**Uncertainty conditions** identify the variation represented by the interval.
The procedure should respect blocked assignment and student-level units. An
interval can be numerically narrow while referring to the wrong estimand or
ignoring a dependency created by design.

**Selection and missingness conditions** connect the observed records to the
declared population. The primary argument needs an explicit reason why its
missing-outcome procedure can address the all-randomized estimand and a
sensitivity analysis for plausible departures.

**Analytic-conduct conditions** concern priority, prospectiveness, model
changes, multiplicity, and reuse of the same evidence. A secondary
prespecified interaction and an exploratory attendance analysis should not be
reported as if they had the same evidential status as the primary contrast.

**Checking and sensitivity conditions** ask whether the main answer survives
plausible alternatives. Diagnostics may reveal poor functional form or
influential observations, but passing a small set of checks does not certify a
model as true. The consequence of a material failure is a changed inference or
claim, not merely an additional appendix figure.

| Design feature | Support supplied | Remaining limitation |
|:---------------|:-----------------|:----------------------|
| **Within-section random assignment** | Average causal comparison of assigned formats for participants | Implementation, missing outcomes, interference, and transport |
| **Nonprobability participation** | Direct evidence about consenting participants | Wider population representation requires additional support |
| **Baseline measurement** | Precision adjustment and a moderator for a secondary question | It does not identify a causal effect of preparation |
| **Attendance recording** | Implementation description and an exploratory association | Attendance is post-assignment and not randomized |
| **Common delayed assessment** | Comparable observed outcome under stated scoring | Score meaning remains bounded to the assessment and conditions |
| **Prospective priorities** | Distinguishes primary, secondary, and exploratory analyses | Prespecification does not guarantee adequate design or inference |

: *Design support and remaining claim limits in the upgraded study.* {#tbl-design-claim-limits tbl-colwidths="[28,35,37]"}

Statistical analysis can express these design-supported estimands and improve
precision. It cannot manufacture probability sampling, randomization of
attendance, absence of interference, comparable measurement, or prospective
status.

## Extending the Construction Workflow

### A seven-stage extension

The argument components describe what logically supports one answer. A
**construction-workflow stage** describes an activity used to construct or
revise that support. Components and stages therefore do different work.

The seven stages below extend Chapter 1's six-stage workflow for studies with
coordinated questions. The diagram compresses a coordinated set of
question-level cycles; it does not create one omnibus argument cycle for the
study. The extension covers much of the question-to-answer research process,
but, as Chapter 1 notes, it is not a complete research-lifecycle model.

Basic Stages 1--3 remain advanced Stages 1--3, with prospective
cross-question alignment added to Stage 3. Basic Stages 4--6 remain advanced
Stages 4--6. New Stage 7 gives every question an explicit outcome and
synthesizes across those outcomes without merging their arguments.

| Basic six-stage workflow | Advanced seven-stage extension |
|:-------------------------|:--------------------------------|
| **1. Review the problem, setting, and relevant literature** | **1. Review literature and the problem** for the coordinated question set |
| **2. Frame the question and, where useful, a candidate answer** | **2. Frame questions and candidate answers**, priorities, and analytic status |
| **3. Align design, measurement, estimand, evidence-production or acquisition plan, and inferential plan** | **3. Plan and align prospectively** across questions: design, measurement, estimands, evidence production or acquisition, stopping, inference, checking, uncertainty, and sensitivity |
| **4. Produce or acquire, and document, empirical and contextual evidence** | **4. Produce or acquire and document** shared and question-specific empirical and contextual evidence |
| **5. Estimate, check, and analyse** | **5. Estimate or predict, check, conduct sensitivity analysis, and reason statistically** for every question-specific estimand |
| **6. Contextualize, interpret, and complete the bounded answer** | **6. Contextualize, justify, interpret, and complete** each bounded substantive answer |
| **New for coordinated questions** | **7. Give every question an explicit outcome and synthesize** across questions |

: *How the basic six-stage workflow extends to seven stages for coordinated questions.* {#tbl-basic-advanced-workflow tbl-colwidths="[43,57]"}

The workflow adapts the iterative character of statistical enquiry emphasized
by PPDAC and statistics-education guidance [@wild1999; @asa2016gaise].
Iteration is confined to the phase currently under construction. Phase 1
Stages 1--3 can iterate before a one-way handoff to Phase 2. Phase 2 Stages
4--7 can then iterate while evidence is produced, analysed, interpreted, and
synthesized. Preregistration or another time-stamped plan can strengthen the
record of prospective decisions, but it does not guarantee that the question,
design, or analysis is adequate [@nosek2018].

| Phase and stage | Main activity | Application to the upgraded study |
|:----------------|:--------------|:----------------------------------|
| **Phase 1 -- 1. Review literature and the problem** | Develop prospective statistical and substantive justification | Examine relevant learning theory and prior evidence about attempted problem solving, feedback, delayed performance, assessment meaning, implementation, and likely moderators |
| **Phase 1 -- 2. Frame questions and candidate answers** | Specify every question--answer obligation and its priority. Declare separate questions, priorities, and analytic statuses when conclusions will be reported separately; declare one question with jointly necessary estimands only when one integrated conclusion is owed. | In the initial coordinated cycles, declare the primary assignment question, secondary baseline-variation question, two-point benchmark, and candidate positive average effect |
| **Phase 1 -- 3. Plan and align prospectively across questions** | Specify and align evidence production or acquisition, any stopping rule, estimands, design, measurement, inference, uncertainty, checking, missingness handling, sensitivity, and planned synthesis across pairs | Align the planned Pair 1 and Pair 2 cycles: randomize within section, stop recruitment after 320 participants are randomized without outcome-dependent stopping, measure baseline, define the delayed outcome, retain all participants in the population named by the primary estimand, and plan interaction and sensitivity analyses |
| **Phase 2 -- 4. Produce or acquire and document evidence** | Produce STAT-E or obtain and designate question-specific analysis records; retain the contextual records needed for SUB-E | Record consent, assignment, delivery, attendance, baseline, outcomes, missingness reasons, and section context |
| **Phase 2 -- 5. Estimate, check, and analyse** | Develop STAT-I and a provisional STAT-C for each pair | Estimate and check the assignment contrast and interaction; in the later exploratory attendance cycle, analyse the reused records only after separately framing its associational estimand |
| **Phase 2 -- 6. Contextualize, justify, interpret, and complete each bounded substantive answer** | Complete the substantive reasoning and bounded answer for every pair | Complete the planned average-effect and effect-variation answers and the separate noncausal attendance answer with their distinct estimands, status, uncertainty, and limitations |
| **Phase 2 -- 7. Give explicit outcomes and synthesize** | Record an explicit outcome for every question, then integrate those outcomes without erasing distinctions | Bring the completed cycles together: record a positive average assignment effect, inconclusive effect variation, and an inconclusive noncausal attendance association before writing the bounded study-level synthesis |

: *The seven-stage extension applied to the upgraded special-practice study.* {#tbl-workflow-application tbl-colwidths="[25,31,44]"}

For each question, Stages 1--6 constitute its own argument-construction cycle.
The planned Pair 1 and Pair 2 cycles share activities and are aligned in Stage
3. Once Stage 6 has completed their bounded answers, Stage 7 records their
outcomes and coordinates the study-level synthesis. The attendance question,
which arose after Phase 2 had begun, first requires a separate exploratory
cycle before its answer may enter that synthesis. Stage 7 does not create a
fourth argument or complete the three arguments a second time.

The Phase 1 anchor for each cycle includes its question and mode, design,
measurement, estimand, evidence-production or acquisition plan, stopping rule,
and inferential plan. A prospectively documented candidate answer may be
contradicted by the results without starting a new cycle. Stage 4 begins when
evidence production starts or,
for existing data, when the question-specific analysis records or extract are
obtained or designated. First access to relevant outcomes or unblinding is a
separate timing landmark: it may precede, coincide with, or follow that handoff.
Prospectiveness is always a relation between a decision and the evidence
relevant to that decision.

@fig-research-workflow emphasizes iteration within each phase and a one-way
handoff from Phase 1 to Phase 2.

```{r}
#| label: fig-research-workflow
#| fig-cap: "A compressed seven-stage view of coordinated question-level cycles, with iteration confined to each phase."
#| fig-alt: "Stages 1 through 3 run from left to right across Phase 1, with a dashed return arrow from Stage 3 to Stage 1. A solid downward arrow gives the one-way handoff from Stage 3 to Stage 4. Stages 4 through 7 then run from right to left across Phase 2. A dashed arrow from Stage 7 to Stage 4 reopens Phase 2 work only in affected question-level cycles. No arrow returns from Stage 7 or any other Phase 2 stage to Phase 1."
#| fig-width: 7.4
#| fig-height: 4.25
#| fig-pos: H

draw_workflow()
```

The dashed Stage 7-to-Stage 4 arrow means that synthesis may reveal a need to
recheck evidence, analysis, or interpretation within Phase 2 of one or more
affected question-level cycles. It does not send one merged study cycle back
to Stage 4.

::: {.callout-note title="Argument, cycle, and execution"}
An **argument** is the logical support for one answer obligation. A **workflow
cycle** is the documented sequence used to construct that argument. An
**execution or data attempt** is one operational production or acquisition of
evidence within the cycle.

After an operational failure, evidence production or acquisition may be
re-executed within the same cycle only when the Phase 1 anchor--including its
acquisition plan and stopping rule--is unchanged, the reason is independent of
the observed answer, and every attempt remains documented. Otherwise, the
change requires an explicit outcome for the existing cycle and a new or revised
cycle. Prospective status belongs to particular decisions relative to
particular evidence; it is not inherited automatically.
:::

### Disciplined iteration

Iteration is disciplined when it stays within the current phase, responds to a
documented problem, and changes analytic status or claims transparently.

Suppose a pilot reveals that reminder messages differ across formats and could
affect attendance. Within Phase 1, before the focal study, the team can return
from Stage 3 to Stages 1 and 2, clarify what belongs to each format, standardize
reminders, and realign the coordinated question set within Stage 3.

Suppose instead that one section experiences a scheduling disruption after
implementation begins. Phase 2 has started, so the team cannot erase the
disruption or treat the old plan as if implementation had not occurred. It
should document the disruption in STAT-E, examine a prespecified or clearly
labelled sensitivity analysis in STAT-I, and narrow the claim if the result
depends on that section. These actions iterate within Stages 4--7.

If Phase 2 reveals that a Phase 1 question, design, measurement, estimand,
evidence-production or acquisition plan, stopping rule, or inferential plan
must change, the same workflow cycle does not return to Phase 1. The team gives
that cycle an explicit outcome and begins a new or revised argument cycle.
Records, code, decisions, and other materials from the earlier cycle may be
reused with their provenance and documentary status intact, but they do not
automatically retain prospective status in the new cycle.

Applied to the upgraded study, the planned delayed-score pair keeps its own
bounded answer. If the team comes to prefer a newly introduced outcome after
evidence production, that outcome requires a separate exploratory pair and
cycle rather than becoming a replacement primary result. If the planned delayed
assessment cannot answer the original question, the current attempt is **not answerable as planned**;
a redesigned study begins a new or revised cycle while
reusing suitable earlier material transparently.

A within-phase return is not a licence to search until a desired answer
appears. An apparently post hoc change can retain prospective status only when
dated records show that it preceded the evidence relevant to that change. A
question or plan chosen after inspecting existing relevant records is not
prospective to those records; a plan fixed before genuinely new evidence may be
prospective for that evidence while remaining explicitly prior-informed. The
workflow makes within-phase revision and new-cycle replacement visible so both
can be evaluated.

### One stage can contribute to several components

Stage 4 produces assignment and outcome records for STAT-E, but it also
produces delivery and setting information needed for SUB-E. Stage 3 develops
parts of STAT-J through the randomization and analysis plan while developing
parts of SUB-J through measurement and implementation choices. Stage 6 does
not manufacture substantive justification after seeing the result; it checks
how the planned support fits what happened, interprets each result, and
completes each bounded answer. Stage 7 then brings the completed
question-level cycles together, gives every question an explicit outcome, and
synthesizes across them without forming an omnibus argument.

This many-to-many relationship is why workflow cannot replace the argument
components. The stages describe work; the components describe logical roles in
the support for a particular answer.

### What readers can and cannot infer from the documented workflow

A reader can usually trace the logical support more confidently than the
historical workflow. A dated protocol can support the claim that a question,
estimand, or analysis was prospective. Versioned code and deviation logs can show
when changes occurred. A Methods section written after analysis cannot by
itself establish timing.

A documented-workflow trace should therefore use restrained timing
descriptions:

- **prospectively documented** when a dated record precedes the relevant
  evidence;
- **implemented as planned** when study records support correspondence between
  plan and conduct;
- **documented deviation** when the change and its timing are reported;
- **retrospectively described** when the paper gives the decision but not its
  timing; and
- **timing not located** when the accessible record does not establish when the
  decision occurred.

These are workflow-timing descriptions, not the four documentary statuses and
not quality-evaluation judgments. They prevent a polished narrative from being mistaken for a
chronological record. They also prevent an absent protocol from being treated
as proof of poor research. The reader states what can be established from the
materials; quality evaluation may then conclude **insufficient information to
judge** for a
particular quality question when the missing timing matters.

## Extending Quality Evaluation for Coordinated Questions

### Responsibility and use

As Chapter 1 explains, authors have the first responsibility to evaluate and
document their arguments, while collaborators, supervisors, reviewers,
editors, and readers independently evaluate them. Researchers can use the
questions while planning, conducting, and reporting a study; readers and
reviewers usually work from the accessible paper and related materials.

A judgment under one quality-evaluation question may be **adequate for this
purpose**, **needs revision**, **insufficient
information to judge**, or **not applicable**. The output is a reasoned profile,
not a summed score. One
decisive weakness can require a claim to be withdrawn even when other features
are strong.

### The six basic questions plus Cross-question coherence

Chapter 1 introduced six quality-evaluation questions. Coordinated questions
require one addition. The seven questions can be remembered in four groups
without renaming them:

- **Alignment:** Question alignment.
- **Support:** Evidence integrity, Justification adequacy, and Inferential
  adequacy.
- **Translation:** Transition fidelity and Claim calibration.
- **Synthesis:** Cross-question coherence--do the question-specific arguments
  remain coherent with one another, and is their study-level synthesis
  warranted?

When one argument contains statistical strands, **Inferential adequacy** asks
whether the strands jointly support the intended statistical inference;
this includes their analytic status, multiplicity, dependence, and uncertainty,
because parallel strands need not be statistically independent.
The same question also evaluates the reasoning that integrates their results
within SUB-I; it is not restricted to statistical procedure.
**Transition fidelity** asks whether SUB-I preserves each strand's result and
uncertainty; and **Claim calibration** asks whether SUB-C retains the conditions
and limitations of every contributing STAT-C. Relations among separate
question--answer pairs belong instead to **Cross-question coherence**.

Suppose a draft conclusion for the upgraded study states:

> Special Practice produced an educationally important improvement,
> especially for less-prepared students, and greater attendance increased the
> benefit.

The conclusion combines one supported result with three unsupported extensions:
importance, effect variation, and attendance causality or mechanism.
The evaluation profile identifies where the problems begin and what they require
the writer to change.

| Group and quality-evaluation question | Beginner question | Judgment and required response |
|:--------------------------------------|:---------------|:-------------------------------|
| **Alignment -- 1. Question alignment** | Do question, estimand, design, method, and intended answer agree? | **Needs revision.** The primary assignment question is aligned; the attendance sentence changes from association to causality and must be revised |
| **Support -- 2. Evidence integrity** | What constitutes evidence, how was it produced, and what do missingness, dependence, and processing imply? | **Insufficient information to judge.** Records are largely adequate, but the missingness process is not fully documented, so support for inference to the all-randomized estimand remains conditional on the missingness analysis |
| **Support -- 3. Justification adequacy** | Do design and statistical conditions support the statistical move, and do measurement and context support the substantive move? | **Needs revision.** Randomization supports assignment, not attendance; assessment meaning and one-setting scope remain bounded |
| **Support -- 4. Inferential adequacy** | Do statistical procedure, uncertainty, checks, sensitivity, multiplicity, and substantive reasoning support the inferences within both layers? | **Needs revision.** The primary interval supports a positive effect but does not establish an effect above two points; the compatible interaction and attendance patterns do not resolve their directions |
| **Translation -- 5. Transition fidelity** | Are magnitude, uncertainty, status, and reservations retained across layers? | **Needs revision.** The draft loses uncertainty and turns secondary point estimates into definite substantive patterns |
| **Translation -- 6. Claim calibration** | Are claims no stronger than evidence, design, uncertainty, alternatives, and scope permit? | **Needs revision.** Replace “educationally important” with uncertainty about whether the effect exceeds the benchmark; remove subgroup and attendance causal claims |
| **Synthesis -- 7. Cross-question coherence** | Are distinct estimands and outcomes preserved, and is synthesis transparent? | **Needs revision.** Keep the positive primary answer beside two inconclusive answers rather than making all three positive |

: *Seven quality-evaluation questions applied to an overstrong synthesis.* {#tbl-quality-evaluation tbl-colwidths="[24,35,41]"}

The repaired synthesis is:

> Assignment to Special Practice improved mean delayed-assessment
> performance for these participants under the stated conditions. Whether the
> effect exceeded two points or varied with baseline preparation remains
> uncertain. Among observed-outcome cases, the attendance association was
> inconclusive; the exploratory analysis does not identify whether attending
> more sessions caused better performance.

Keep four outputs distinct. Reading records each **documentary status** and the
authors' **reported answer**. Each quality question produces a judgment such as
adequate for this purpose, needs revision, insufficient information to judge, or not
applicable.
Evaluation then states the **supported bounded answer** and one conclusion for
each question--answer pair. The study-level synthesis is evaluated separately
under Cross-question coherence.

For the three correctly bounded pairs in the question map, the supported
answers match what was reported, so each is **supported as stated**. The causal
attendance clause in the overstrong synthesis creates a different implied pair;
because the accessible design provides no defensible causal attendance answer,
that implied pair is **not supported**. The synthesis itself **requires
revision** under Cross-question coherence; it does not receive one of the four
pair-level conclusions. **Not assessable from the accessible record** remains
available when the record is insufficient to decide a pair-level conclusion.

No total score is needed. The evaluation changes the answer directly without
turning documentary presence into a quality judgment.

### Locate the originating problem, then trace its consequences

Do not count one weakness several times merely because it affects several parts
of an argument. Locate the originating problem under the question whose primary
object contains it, then trace its consequences. Here the attendance mechanism
first fails Question alignment because it changes an associational estimand into
a causal one, and it fails Justification adequacy because attendance was not
randomized. The unsupported move then propagates to Transition fidelity, Claim
calibration, and Cross-question coherence. Those later questions show what the
final wording and synthesis must repair.

Cross-question coherence is normally not applicable to a genuinely
one-question study. If a conclusion introduces an unasked subgroup, mechanism,
or prediction claim, however, the study now contains another implied
question--answer pair that must be separated and evaluated. In a coordinated
study, Cross-question coherence tracks priorities, shared evidence,
multiplicity, tensions, and synthesis across those distinct pairs.

## How Regression-Based Methods Develop and Instantiate the Approach

Here **instantiate** means *put selected parts of the approach into practice*.
Regression can express a declared comparison, help estimate its estimand,
quantify uncertainty, organize checking, and support a bounded statistical
claim. It does not define the research question or estimand and does not replace
the full argument.

### From a two-group comparison to adjustment

Let $Y$ be delayed score and $Z$ indicate Special Practice assignment.
A simple two-group regression is:

$$
E(Y\mid Z)=\beta_0+\beta_1 Z.
$$

Here $\beta_0$ represents the Usual Practice mean and $\beta_1$ represents
the Special-minus-Usual mean difference for the analysed records. This is
the same comparison that can be calculated directly from two group means.
Regression provides a common language; it does not create a new question.

For the primary analysis, baseline preparation $X$ and section indicators
can improve precision and respect blocked assignment:

$$
E(Y\mid Z,X,S)=
\beta_0+\beta_1 Z+\beta_2(X-20)+\gamma_S.
$$

The reported primary result is not simply whichever coefficient appears in the
output. Regression standardization averages fitted assignment comparisons over
the target group. In this additive linear model it reproduces the constant
assignment coefficient; interactions or nonlinear models can make the averaged
comparison differ, and neither is the crude 3.1-point observed-case difference.

### An interaction represents a different question

The secondary model adds an assignment-by-baseline product:

$$
E(Y\mid Z,X,S)=
\beta_0+\beta_1 Z+\beta_2(X-20)+
\beta_3 Z\{(X-20)/10\}+\gamma_S.
$$

Now $\beta_3$ represents how the assignment contrast changes per ten-point
higher baseline score under the specified scale and functional form. It does
not estimate the causal effect of baseline preparation. It also does not prove
that effect variation is absent when its interval includes zero.

Interaction is scale- and specification-dependent. A pattern on a mean-score
scale can differ from a pattern on a probability or odds scale. Interpretation
should stay within the supported baseline range and retain uncertainty.

### Attendance uses the same language with different support

A regression for Pair 3 includes attendance alongside assignment, baseline, and
section to estimate a conditional attendance association among the 300
observed-outcome cases. Attendance occurs after assignment. Conditioning on it
can introduce post-assignment selection bias and can also change the
interpretation of the assignment coefficient [@elwert2014]. Randomization
supports the unconditioned or appropriately standardized assignment comparison;
it does not automatically support either coefficient in an
attendance-conditioned model.

The regression is therefore used only for an exploratory association argument.
It does not estimate a causal effect of attendance. Claim mode comes from the
question and design, not the appearance of regression.

| Regression move | Question-level role | Result in the upgraded study | What remains outside regression |
|:----------------|:--------------------|:-----------------------------|:--------------------------------|
| **Two-group indicator** | Expresses the assigned-format comparison | Observed-outcome difference of 3.1 points | Missing-outcome estimand and wider scope |
| **Baseline and section adjustment** | Improves precision and respects blocked assignment while estimating the average estimand through standardization | Primary effect 2.9, 95% interval [1.9, 4.0] | Measurement meaning, missingness plausibility, implementation, and transport |
| **Assignment-by-baseline interaction** | Expresses effect variation over the baseline scale | -0.8 per ten points, 95% interval [-2.7, 1.1] | Adequacy of the moderator measure, functional form, and substantive meaning |
| **Post-assignment attendance association** | Expresses an observed-case conditional association | 0.3 per session, 95% interval [-0.4, 0.9] | Causal effect of attendance, missing outcomes, and unmeasured reasons for participation |

: *How regression instantiates distinct parts of three arguments.* {#tbl-regression-instantiation tbl-colwidths="[22,30,25,23]"}

### Where regression enters the statistical layer

Regression implements different parts of the statistical reasoning, but it
does not create the study's empirical evidence:

- **STAT-E is the input.** The outcome, assignment, baseline, section,
  attendance, missingness, and processing records must exist before the model is
  fitted. The analysis must keep exclusions and transformations visible rather
  than treating model-ready data as the original evidence.
- In **STAT-J**, the model and estimator help connect a declared estimand to the
  records through choices about functional form, adjustment, dependence,
  missingness, uncertainty, and checking when the associated warrant is stated
  or implicit but traceable. A fitted model does not show that those choices or
  their conditions are adequate.
- In **STAT-I**, fitting, checking, and sensitivity analysis produce estimates,
  intervals, diagnostics, and comparisons that must be reasoned about for the
  declared estimand.
- In **STAT-C**, the study states its statistical conclusion about magnitude,
  direction, uncertainty, and conditions; evaluation later judges its
  calibration.

The substantive layer still has to situate the evidence, augment and
contextualize the statistical justification with reasons about measurement and
design relevance, interpret magnitude, consider alternatives, and state the
substantive answer with its bounds. Quality evaluation is a separate task.

Regression cannot supply by itself:

- population representation;
- measurement validity;
- causal identification;
- a plausible missingness condition;
- implementation fidelity;
- practical or theoretical importance;
- mechanism; or
- transport to another setting.

These are not afterthoughts called “limitations.” They help determine whether
the coefficient answers the question.

### Checking, sensitivity, and claim revision

Regression makes several kinds of checking possible, but their role must remain
bounded.

A residual display can reveal tension with a linear mean pattern or constant
spread. An influence analysis can show that an estimate depends heavily on a
few observations. Comparing adjusted and unadjusted estimates can reveal how
the chosen specification changes the estimate or precision, or whether the
fitted procedure addresses a quantity different from the declared estimand.
None of these checks proves that the model is true or that unmeasured problems
are absent.

For Pair 1, useful checks include whether the baseline relationship is
reasonably represented over the observed range, whether section-specific
contrasts point in compatible directions, and whether the standardized result
depends strongly on the missingness procedure. A serious failure should enter
STAT-I and change STAT-C.

For Pair 2, plotting estimated contrasts over supported baseline values is more
informative than announcing one interaction coefficient. Sparse data near the
scale ends and alternative nonlinear forms can make an apparent subgroup
pattern unstable. Because the interval includes materially decreasing,
negligible, and increasing patterns, the reported answer remains inconclusive;
width alone does not determine that conclusion. Exploratory alternatives should
not be used to search for a preferred threshold.

For Pair 3, no residual check can remove unmeasured reasons for attendance.
Sensitivity can show how large an omitted association would need to be to
change the estimate, but the result remains observational. Checking improves
calibration; it does not upgrade design.

This is why model checking belongs within inference rather than in a detached
technical appendix. If a check matters, its consequence should be visible in
the statistical claim and carried through the substantive layer.

### Continuity with later chapters

An intercept-only model represents a mean. A group indicator represents a mean
difference. Continuous variables express slopes or curves. Additional variables
define conditional or standardized comparisons, and product terms express
variation in comparisons. Later chapters develop these methods in detail.

The central discipline does not change: begin with a question, declare the
estimand, preserve the design, make uncertainty and checking visible, interpret
through the substantive layer, and end with a bounded answer, which may be
inconclusive.

## Transferring the Approach to a Published Study

### The study and its materials

Deslauriers and colleagues compared active and passive instruction in two
large-enrolment introductory physics courses [@deslauriers2019]. The
[open-access paper](https://doi.org/10.1073/pnas.1821936116) and
[supporting appendix](https://www.pnas.org/doi/suppl/10.1073/pnas.1821936116/suppl_file/pnas.1821936116.sapp.pdf)
provide a compact opportunity to transfer the approach from a designed
hypothetical record to actual research reporting.

The intervention occurred during two consecutive class meetings. Students were
randomly assigned to two groups. In the first meeting, one group received
active instruction on static equilibrium while the other received a passive
lecture. In the second meeting, the instructional conditions were reversed for
the fluids topic. The instructional condition therefore crossed over within two
fixed group--instructor--room combinations. The study used both random
assignment to group and a crossover of instructional condition.

The groups received identical content and handouts. In the active condition,
students attempted problems in small groups before receiving complete
solutions. In the passive condition, the instructor presented the same
material and solutions. Of 157 enrolled students, 149 participated; eight opted
out or did not complete the study.

After each class, students completed:

- a 12-item test of learning (TOL); and
- four feeling-of-learning (FOL) items, analysed using both a designated primary
  item and a principal-component-weighted composite of all four.

The tests were written from learning objectives by an author who did not see
the course materials, and instructors did not see the tests. These design
details contribute to measurement and comparison justification.

### Two focal question--answer pairs

The paper contains at least two focal pairs:

1. What effect did active rather than passive instruction have on measured
   learning?
2. What effect did active rather than passive instruction have on students'
   feeling of learning?

The adjusted standardized regression results were:

- active instruction was associated with a **0.46 standard-deviation higher**
  TOL score than passive instruction; and
- active instruction was associated with a **0.56 standard-deviation lower**
  FOL score than passive instruction.

For each outcome, the reported test used zero as the null target value for the
adjusted active-minus-passive standardized contrast and had
$p<.001$. These tests belong in STAT-I beside their respective 0.46- and
-0.56-SD estimates. As Chapter 1's numerical-result safeguard emphasizes, a
$p$-value alone supplies neither direction, magnitude, nor precision; an omnibus test might
not even supply direction. Chapter 3 develops the calculation and fuller
interpretation of tests, intervals, and inconclusive results. The point
estimates changed little when student covariates were removed or student fixed
effects were added. The paper also considered peer effects and provided study
materials in its supporting appendix.

The two results do not form one answer. They answer different questions with
different outcomes and measurement meanings. Their study-level synthesis is
that active instruction improved performance on the immediate tests while
students felt that they learned less.

### Tracing the supporting record

| Pair and component | Documentary trace | Documentary status | Component content located |
|:-------------------|:------------------|:-------------------|:--------------------------|
| **TOL -- STAT-E** | Methods, assignment description, crossover table, participation information, 12-item tests | Stated in the paper; test material located in related materials | Group assignment, crossover, participation, and immediate test records; adjusted-model analytic N after missing pretests is not located in the compact result table |
| **TOL -- STAT-J** | Design description, model specification, robustness descriptions, supplement | Stated in the paper; additional checks located in related materials | Randomized crossover warrant, adjustment strategy, and alternative specifications |
| **TOL -- STAT-I** | Regression Table 3 and robustness descriptions | Stated in the paper | Standardized 0.46 active-minus-passive estimate, test of a zero adjusted difference with *p* < .001, and stability across selected specifications |
| **TOL -- STAT-C** | Results text, Table 3, and Figure 2 | Stated in the paper | Reported positive adjusted comparison for immediate test performance |
| **TOL -- SUB-E/J/I/C** | Learning objectives, course-material-blinded test preparation, matched content, Discussion | Stated in the paper; located in related materials (supplement) | Immediate topic-specific performance interpretation, instructional context, and reported scope |
| **FOL -- STAT-E** | Four survey items, response scale, assignment and crossover records | Stated in the paper; item detail located in related materials | Immediate self-reported perceptions after each class |
| **FOL -- STAT-J** | Composite description, model specification, and designated-item analysis | Stated in the paper | Composite warrant, adjustment strategy, and alternative item analysis |
| **FOL -- STAT-I** | Regression Table 3 and designated-item analysis | Stated in the paper | Standardized -0.56 active-minus-passive estimate, test of a zero adjusted difference with *p* < .001, and consistency with the designated item |
| **FOL -- STAT-C** | Results text, Table 3, and Figure 2 | Stated in the paper | Reported negative adjusted comparison for immediate feeling of learning |
| **FOL -- SUB-E/J/I/C** | Survey wording, discussion of fluency and metacognition, interviews | Stated in the paper; interview detail located in related materials | Immediate perceived-learning interpretation and proposed mechanism context |
| **Cross-question synthesis** | Abstract, Results, Discussion, figures | Stated in the paper | Higher measured performance and lower perceived learning are reported as distinct outcomes |
| **Transport and mechanism context** | Discussion, interviews, semester intervention, supporting literature | Stated in the paper; further material located in related materials | Proposed cognitive-effort explanation and discussion of possible extension beyond the two class meetings |

: *Documentary traceability for two arguments in a published randomized study.* {#tbl-published-trace tbl-colwidths="[21,27,21,31]"}

### What is strongly supported

Random assignment, the instructional-condition crossover within fixed
group--instructor--room combinations, matched content, and common timing provide
substantial support for comparing active with passive instruction in the
implemented classes. The different directions for TOL and FOL are not a
contradiction because the outcomes answer different questions.

The crossover also creates obligations. Each student experienced two ordered
class meetings on different topics, so readers should examine topic, period,
instructional order, and possible carryover. Groups, instructors, and rooms
remained fixed rather than being separately crossed. The randomized group
assignment, instructional-condition crossover, and relevant indicators reduce
some concerns, but they do not make those design conditions disappear.

```{=latex}
\Needspace{0.10\textheight}
```

The paper's use of student background measures, instructor and topic indicators,
alternative specifications, and peer-effect discussion adds statistical
justification. Course-material-blinded test preparation by a coauthor and the
supplied materials make parts of the measurement support inspectable.

### Where the answer should stop

The direct study population consists of students in two Harvard introductory
physics courses. The intervention lasted two class meetings, and the outcomes
were measured immediately. The TOL therefore supports a bounded claim about
short-term performance on topic-specific tests, not durable learning across
disciplines. FOL is a self-reported perception measure, not learning itself.

Under the crossover's topic, period, order, and carryover conditions, the
randomized comparisons support effects of assigned instructional mode on TOL
and FOL. They do not by themselves establish that increased cognitive effort
caused the difference, that novice metacognition is the unique mechanism, or
that the same effects will occur in every course. The interviews, literature,
and later semester intervention provide relevant evidence, but those mechanism
and transport claims require their own arguments.

A calibrated synthesis is:

> In two introductory physics courses, active rather than passive instruction
> increased immediate topic-specific test performance while reducing students'
> immediate reported feeling of learning under the implemented crossover
> conditions. The study does not by itself establish long-term learning,
> a unique psychological mechanism, or effects in other populations and
> disciplines.

This transfer example illustrates why the accessible record must be read by
logical function rather than paper section. It also shows why one paper can
contain several related but distinct arguments. For the bounded TOL and FOL
answers stated above, the supported answers match the reported answers, so the
pair-level conclusions are **supported as stated**. Stronger claims about a
unique mechanism, durable learning, or all disciplines are **not supported** as
answers to those additional implied questions by the two focal randomized
comparisons alone.

{{< pagebreak >}}

## Summary and Reading Bridge

### What the advanced unit adds

A relatively complex study may coordinate several questions without collapsing
them into one argument. Each question--answer pair keeps its own estimand,
support, analytic status, reported answer with its bounds, and question-level
construction cycle. When several estimands are jointly necessary for one
integrated answer, they can instead form statistical strands inside that one
argument. The seven-stage extension adds prospective cross-question alignment to
Stage 3 and a new Stage 7 that brings completed cycles together for explicit
question outcomes and study-level synthesis. The seventh
quality-evaluation question checks the coherence of that synthesis. Regression
can instantiate selected comparisons and
inferential moves, but the complete reading must still trace design,
measurement, context, uncertainty, and scope.

### Further extension for latent-variable questions

Observable scores already require measurement reasoning. When a question
contains a latent variable or construct, the statistical part of its one
question-level argument may be elaborated into linked measurement and
structural layers rather than retained as one undifferentiated statistical
layer.

A **measurement layer** addresses what observed indicators support about a
latent variable, score, or classification. A **structural layer** addresses
relationships among latent variables or between latent and observed variables.
Each layer can contain Evidence, Justification, Inference, and bounded Claim.

Both layers elaborate the statistical layer and then connect to the substantive
layer. This remains one argument for one question--answer pair, not separate
measurement and structural arguments. Parallel component labels do not imply
that every cross-layer operation is identical. The full architecture will be
developed in the book's latent-variable part.

\Needspace{0.55\textheight}

### Intellectual and practical reading bridge

The approach is a tutorial synthesis:

- Toulmin emphasizes evidence, warrants, qualifications, and reservations
  [@toulmin2003].
- LeMire offers a narrower statistical argumentation precedent
  [@lemire2010].
- Kane and the testing standards organize support for score interpretations
  and uses [@kane2013; @aeraApaNcme2014].
- Estimand-first reasoning connects theoretical questions to estimands,
  identification, and evidence [@lundberg2021].
- PPDAC and GAISE emphasize iterative statistical enquiry, context, real data,
  and conceptual understanding [@wild1999; @asa2016gaise].
- Design principles explain why representation, assignment, measurement,
  implementation, and transport bound claims [@shadish2002].
- Preregistration and reporting standards can strengthen prospective
  transparency and traceability without guaranteeing validity
  [@nosek2018; @appelbaum2018].

The approach is not offered as a complete new theory of inference. Its
contribution is pedagogical integration: one argument for every
question--answer pair, linked layers, a two-phase construction workflow with
within-phase iteration, and quality-evaluation questions that lead directly to
calibrated answers.

{{< pagebreak >}}

\Needspace{8\baselineskip}\phantomsection

## Exercises {.unnumbered}

\Needspace{6\baselineskip}\phantomsection

<!-- prompt-id: CH01A-E01 -->
### Exercise 1: Separate the questions (Foundational) {.unnumbered #ex-ch01a-01}

A study asks whether a mentoring program:

1. changes average end-of-year score;
2. works differently for students with different baseline scores;
3. improves attendance; and
4. predicts which students will leave the course.

For each question, state:

- its intended claim mode;
- a plausible estimand;
- whether random assignment to mentoring directly supports the answer; and
- why it needs a distinct argument.

<!-- END EX01-01 PROMPT -->

\Needspace{6\baselineskip}\phantomsection

<!-- prompt-id: CH01A-E02 -->
### Exercise 2: Identify shared and question-specific support (Foundational) {.unnumbered #ex-ch01a-02}

For the upgraded special-practice study, classify the following as shared
across all three arguments, shared across only Pairs 1 and 2, or primarily
question-specific:

- participant consent records;
- within-section randomization;
- baseline score;
- session attendance;
- delayed outcome;
- missing-outcome condition;
- assignment-by-baseline interaction;
- attendance-confounding justification; and
- two-point importance benchmark.

Explain why “shared evidence” does not mean “identical adequacy.”

Then consider the alternative integrated decision question, “Did assignment
meet the prespecified success criterion of a positive delayed-performance
effect and no completion-rate reduction greater than five percentage points?”
Explain why its score and completion estimands can be two statistical strands of
one argument, and what would have to change for them to become two separately
owed question--answer pairs.

<!-- END EX01-02 PROMPT -->

\Needspace{6\baselineskip}\phantomsection

<!-- prompt-id: CH01A-E03 -->
### Exercise 3: Diagnose an estimand--analysis mismatch (Foundational) {.unnumbered #ex-ch01a-03}

An analyst reports the observed-case difference of 3.1 points as the average
effect for all 320 randomized students. Explain:

1. how the observed-case analysis and available evidence fail to match the
   declared all-randomized estimand;
2. which component or components are affected;
3. what condition or method is needed to connect 300 observed outcomes to the
   320-person estimand; and
4. how the final claim should change if that connection is not defensible.

<!-- END EX01-03 PROMPT -->

\Needspace{6\baselineskip}\phantomsection

<!-- prompt-id: CH01A-E04 -->
### Exercise 4: Map the construction workflow (Integrative) {.unnumbered #ex-ch01a-04}

Use the seven-stage workflow: Stages 1--3 form Phase 1, and Stages 4--7 form
Phase 2. Assign each event to one or more workflow stages:

- reviewing evidence about delayed assessment;
- declaring the two-point benchmark;
- fixing the 320-participant evidence-production plan and stopping rule;
- discovering a scheduling disruption;
- discovering during Phase 2 that the delayed assessment cannot answer the
  declared primary question;
- formulating the exploratory attendance question after evidence production
  had begun;
- documenting session attendance;
- revising a missingness sensitivity analysis;
- interpreting the interaction interval;
- integrating the three answer outcomes.

Identify (a) one justified return within the current phase, (b) one Phase 2
discovery that requires closing the current cycle and beginning a new or
revised argument cycle, and (c) one new question whose timing requires a
separate exploratory cycle.

<!-- END EX01-04 PROMPT -->

\Needspace{6\baselineskip}\phantomsection

<!-- prompt-id: CH01A-E05 -->
### Exercise 5: Apply the seven quality-evaluation questions (Integrative) {.unnumbered #ex-ch01a-05}

Evaluate this conclusion:

> Special Practice is especially effective for less-prepared students and
> works because it makes students attend more sessions.

Give a judgment for all seven quality-evaluation questions. Use **adequate for this purpose**,
**needs revision**, **insufficient information to judge**, or **not
applicable**. Then write a
replacement synthesis that preserves all three reported pair outcomes,
including the inconclusive observed-case attendance association. State a
supported bounded answer and pair-level evaluation conclusion for each original
pair. Finally, identify the implied causal-attendance pair, give its pair-level
conclusion, and say whether the study synthesis is coherent as written or
requires revision.

<!-- END EX01-05 PROMPT -->

\Needspace{6\baselineskip}\phantomsection

<!-- prompt-id: CH01A-E06 -->
### Exercise 6: Assignment versus attendance (Integrative) {.unnumbered #ex-ch01a-06}

Draw two causal stories in words:

1. assignment affects attendance, which affects score; and
2. motivation affects both attendance and score.

Explain why an association between attendance and score can arise in both
stories and why random assignment to the practice format does not randomize
attendance. Propose a future design that would answer a causal attendance
question more directly.

<!-- END EX01-06 PROMPT -->

\Needspace{6\baselineskip}\phantomsection

<!-- prompt-id: CH01A-E07 -->
### Exercise 7: Decode three regression estimands (Integrative) {.unnumbered #ex-ch01a-07}

For each model below, identify the question, estimand, key justification, and
strongest possible claim:

1. delayed score regressed only on assigned format;
2. delayed score regressed on assigned format, baseline score, and section; and
3. delayed score regressed on assigned format, baseline score, section, and an
   assignment-by-baseline interaction.

State one condition under which the assigned-format coefficient would not
equal the marginal average assignment effect without standardization.

<!-- END EX01-07 PROMPT -->

\Needspace{6\baselineskip}\phantomsection

<!-- prompt-id: CH01A-E08 -->
### Exercise 8: Published-paper traceability (Integrative) {.unnumbered #ex-ch01a-08}

Using the Deslauriers et al. paper and supporting appendix:

1. locate the two focal questions;
2. map one STAT component and one SUB component for each outcome;
3. identify where the randomized crossover design is documented;
4. compare the paper's direct numerical support for TOL and FOL;
5. identify one mechanism claim that needs a separate argument; and
6. rewrite one broad generalization as a bounded answer.

For each located item, use **stated in the paper**, **located in related
materials**, or **implicit but traceable**; use **not located** when appropriate
and flag contradictions separately. Record documentary status separately from
quality: an item can be stated in the paper yet inadequately justified.

<!-- END EX01-08 PROMPT -->

\Needspace{6\baselineskip}\phantomsection

<!-- prompt-id: CH01A-E09 -->
### Exercise 9: Independent multi-question audit (Integrative) {.unnumbered #ex-ch01a-09}

Choose a quantitative paper with at least two coordinated research questions
whose estimands, evidence, models, priorities, or interpretations are connected,
and at least one related material such as a protocol, supplement, code archive,
or data dictionary. Produce:

1. a question-to-argument map;
2. one complete eight-component map;
3. a shared-versus-question-specific support analysis;
4. a documented-workflow trace, showing within-phase returns and marking the
   start of a new cycle whenever Phase 2 work led to revision of a Phase 1
   decision;
5. a seven-question quality-evaluation profile;
6. a study-level synthesis; and
7. a list of component content that is **not located**, followed separately by
   any whole-pair conclusion of **not assessable from the accessible record**.

Do not infer that a decision was prospective solely because it appears in the
Methods section.

<!-- END EX01-09 PROMPT -->

\Needspace{6\baselineskip}\phantomsection

## References {.unnumbered}

::: {#refs}
:::
