---
title: "Part One: Foundations, Regression, and Generalized Linear Models"
subtitle: "An Introduction to the Chapters and How They Fit Together"
author: "Jinsong Chen"

lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    fontsize: 11pt
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - ../koma_heading_tags.tex
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{12\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{12\baselineskip}\phantomsection}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: intro-setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center",
  dev = if (knitr::is_latex_output()) "cairo_pdf" else "png"
)

draw_part_one_route <- function() {
  old <- par(mar = rep(.1, 4), xaxs = "i", yaxs = "i")
  on.exit(par(old), add = TRUE)
  plot.new()
  plot.window(xlim = c(0, 4), ylim = c(0, 1))

  labels <- c(
    "Chapters 1-2\nQuestions, evidence,\nrecords, and variation",
    "Chapters 3-4\nEstimation, uncertainty,\nand group comparisons",
    "Chapters 5-6\nThe same methods\ncarried out in R",
    "Chapters 7-9\nRegression, from one\npredictor to the toolbox"
  )
  fills <- c("#F3F5F6", "#E8EEF1", "#DCE7EB", "#D0E0E5")

  for (i in 0:3) {
    x0 <- i + .03
    x1 <- i + .90
    rect(x0, .18, x1, .82, col = fills[i + 1], border = "#405768",
         lwd = 1.2)
    text((x0 + x1) / 2, .50, labels[i + 1], cex = .72,
         col = "#243746", font = 2)
    if (i < 3) {
      arrows(x1 + .015, .50, i + 1.025, .50, length = .07,
             lwd = 1.2, col = "#405768")
    }
  }
}
```

## What Part One Offers

*Quantitative Methods for Social and Behavioral Research* develops
the statistical understanding needed to read empirical studies, interpret
their results, and conduct quantitative research. Part One is where the
book begins: fourteen numbered chapters that start with what a study asks
and how it produces its evidence, and end with regression models for a
range of outcomes. It is written for postgraduate students and applied researchers.
A mean, an interval, or a regression coefficient becomes useful when a
reader understands what it describes, how it was obtained, and which
question it helps answer.

Some introductory familiarity with statistics and algebra is useful, and
little else is assumed. Each chapter explains the notation it uses and
develops its calculations in worked steps, so the algebra arrives with the
idea that needs it. Calculus and proof-based mathematical statistics are
not prerequisites. Chapters 1--4 can be read without running R or SPSS:
Chapter 1 introduces the research journey, and Chapters 2--4 teach
statistics directly, through explanations, figures, and worked
calculations. Computation joins the statistics in Chapters 5 and 6.

Results are rounded to four decimals from full-precision computation,
with trailing zeros dropped (3.1 rather than 3.1000), unless a source's own
precision applies.

## How the Chapters Build on Each Other

Part One contains fourteen chapters. The first nine establish a route from
questions and records to regression. Chapter 1,
*The Quantitative Research Journey*, orients the reader in questions,
evidence, design, and interpretation. Chapter 2, *Data, Distributions, and
Variation*, develops records, distributions, and variation across repeated
samples. Chapter 3, *Estimates, Uncertainty, and Statistical Tests*,
develops estimation, intervals, and tests. Chapter 4, *Group Comparisons
with Analysis of Variance*, develops comparisons among several groups.
Chapter 5, *Description and Basic Inference with R*, and Chapter 6,
*Group Comparisons and Statistical Modelling with R*, make those
foundations executable. Chapters 7--9 then develop regression from one
predictor to the regression toolbox: Chapter 7, *Simple Linear
Regression*; Chapter 8, *Multiple Regression I: Conditioning, Checks, and
Coordinated Questions*; and Chapter 9, *Multiple Regression II:
Categorical Predictors, Model Comparison, Nonlinearity, and Robust
Uncertainty*.

Chapters 10--14 extend the regression framework to conditional relationships,
noncontinuous outcomes and prediction, then return to the design of a study
and the limits of its claims. Interaction changes how a relationship varies
across conditions; generalized linear models change the outcome model;
validation asks how a fitted rule performs on new cases; and causal design
asks which comparisons can support an intervention claim.

```{r}
#| label: fig-part-one-route
#| fig-width: 7
#| fig-height: 1.75
#| out-width: "100%"
#| fig-cap: "The foundations of Part One, from questions and records to inference, computation, and regression in Chapters 1–9."
#| fig-alt: "Four boxes connected from left to right: Chapters 1 to 2, questions, evidence, records, and variation; Chapters 3 to 4, estimation, uncertainty, and group comparisons; Chapters 5 to 6, the same methods carried out in R; Chapters 7 to 9, regression from one predictor to the regression toolbox."

draw_part_one_route()
```

@tbl-part-one-sequence states what each chapter is responsible for and what
it builds on.

| Chapter | What it is responsible for | What it builds on |
|:--|:--|:--|
| **1. The Quantitative Research Journey** | Turning a concern into a question with a stated target quantity; how a study produces its evidence; how comparison, timing, sampling, and assignment support descriptive, associational, predictive, and causal answers; reading a reported result as a bounded claim | The starting point of the part |
| **2. Data, Distributions, and Variation** | Reading records, identifiers, codes, and denominators; sampling schemes; distributions, their skew and tail weight, centre, spread, and groups; relationships between two variables; how a summary varies across repeated samples | Reads the dataset behind Chapter 1's account of evidence and applies its design reasoning to an interpretation |
| **3. Estimates, Uncertainty, and Statistical Tests** | The target quantity, the rule that estimates it, and the value obtained; standard errors for a mean and a difference, including the pooled standard error, and the conditions behind them; confidence intervals and confidence level; hypothesis tests, p-values, and significance; the one-sample, paired, equal-variance, and Welch t tests; the chi-square test of association and tables of three or more categorical variables; magnitude, error rates, power, and choices made after seeing the results | Turns Chapter 2's sampling distributions and standard error into inference from one observed sample |
| **4. Group Comparisons with Analysis of Variance** | Group-mean models and their comparison by the F test; one-way and two-way analysis of variance; sums of squares, degrees of freedom, and mean squares; contrasts and comparison families; interaction and simple effects, extended to a worked three-way table and the structure of four-way designs | Applies Chapter 2's description of group variation and Chapter 3's inference to comparisons among several groups |
| **5. Description and Basic Inference with R** | Preparing records in base R with explicit missing values; distributions, graphics and probability calculations; relationships, covariance and correlation matrices, and one-control partial correlation; simulated sampling variation, bias and standard error; mean intervals and their repeated-study coverage, z and t tests, and selected correlation inference; a short reproducible report | The computational route through Chapter 2's order of distributions, relationships and sampling, followed by Chapter 3's mean inference |
| **6. Group Comparisons and Statistical Modelling with R** | Group-mean and factorial models in R, with the F test comparing named models; graphical comparisons with ggplot2; contrasts and comparison families, including a declared contrast and the protection needed if it had been selected after inspection; proportions and counts, with their effect scales; numerical optimization through squared loss and a scalar likelihood; simulation, precision and power | Implements Chapter 4's model comparisons and extends Chapter 3's categorical and planning lessons on Chapter 5's computational foundation |
| **7. Simple Linear Regression** | The model for a conditional mean and its error variation, first through a known process and then occupation records; least-squares estimates and slope inference; repeated-study uncertainty; mean-response and new-case intervals; diagnostics; regression toward the mean, its paradox and its causal misreading, with Galton's heights and repeated exams | Extends Chapter 4's group-mean model to a line, using Chapters 5--6's covariance, inference, simulation and objective-function work |
| **8. Multiple Regression I: Conditioning, Checks, and Coordinated Questions** | The multiple-regression model and partial slopes through a known process; estimation and inference on the same occupation records, with the overall F test; checks, coefficient covariance, fitted profiles and standardization; a separate held-out prediction question; three coordinated answers in one study | Extends Chapter 7's model and line with Chapter 5's partial correlation and Chapter 4's model comparisons, implemented in R in Chapter 6 |
| **9. Multiple Regression II: Categorical Predictors, Model Comparison, Nonlinearity, and Robust Uncertainty** | Changes to the mean part: indicators and contrasts, covariance adjustment, nested and non-nested model comparisons, and changed forms and scales; changes to uncertainty: HC3, resampling, and a randomized study with school clusters | Extends Chapter 8's model, connecting each method to the part it changes and the conditions it needs |
| **10. Interaction and Moderation** | Product terms, conditional slopes, comparisons within joint support, and bootstrap uncertainty for conditional relationships | Extends Chapters 8 and 9's regression models, contrasts and resampling to relationships that vary with another variable |
| **11. Binary Logistic Regression** | The Bernoulli likelihood, the logit link, probability and odds scales, fitting, comparison and checks for a binary outcome | Changes Chapter 8's outcome model while retaining its conditional-question discipline and Chapter 6's likelihood foundation |
| **12. Extending the GLM: Nominal, Ordinal, and Count Outcomes** | Outcome classification, generalized and cumulative logits, counts and rates, overdispersion, and model checks | Extends Chapter 11's response-distribution, link and linear-predictor framework |
| **13. Prediction, Cross-Validation, and Regularization** | Held-out error, leakage, penalized fitting, classification performance and changes in the deployment population | Develops Chapter 8's separate prediction question with the regression and classification models of Chapters 9--12 |
| **14. Regression Pitfalls and Causal Design** | Adjustment structures, reversals, selection, measurement, dependence and the design support for a bounded claim | Re-examines the preceding models through Chapter 1's questions, evidence and interpretation, including what adjustment alone cannot supply |

: What each chapter is responsible for. {#tbl-part-one-sequence tbl-colwidths="[24,46,30]"}

Each idea has one principal home, and later chapters use it rather than
teach it again. Chapter 1 previews uncertainty while reading a study;
Chapter 2 develops sampling distributions and the standard error through
repeated samples; Chapter 3 develops their inferential use and states the
conditions they require. Chapter 2 explains the descriptive ingredients of
group variation; Chapter 4 develops the full analysis-of-variance
decomposition and its inferential use, comparing one common mean with
group means and an additive model with a model that includes interaction. Where a later chapter needs an
earlier idea, it recalls it briefly and then puts it to work in the new
setting.

## How to Read a Chapter and Use the Concept Reference

The numbered chapters share an arrangement. A **Chapter overview** box opens the
chapter with the route it follows and what a reader will be able to do
afterwards. Numbered sections develop the material in order. A closing
**Reading Bridge** brings the principal ideas together and connects them
to the next chapter, and a selective **Key terms** table gives a compact
reference to the concepts explained principally there. Unnumbered material
follows the numbered sections: exercises for working with the ideas,
appendices holding fuller derivations and specialized extensions, and the
references. The arrangement keeps the main explanation distinct from fuller derivations
and specialized extensions.

There are two sensible routes through the chapters. The first is
the chapter sequence. The second pairs statistics with computation:
Chapters 2 and 3 with Chapter 5, then Chapters 3 and 4 with Chapter 6.
Either way, follow a method from its data and question through its
calculation to its interpretation; a result separated from the question it
answers is difficult to use.

The [**Part One Concept Reference**](https://jinsong-chen.github.io/QMSBR/part-one/part_one_concept_reference.html) is published beside this introduction.
It gathers selected concepts from Chapters 1--14 into nine tables, arranged
by concept group and then by term, with each term's meaning and the place
where the book explains it principally. A further table lists the
procedures of Chapters 3 and 4 with what each estimates and the
conditions it relies on, and a closing table records the notation those
chapters use. It is a place to look something up while
reading rather than a substitute for reading, because first-use
explanations and worked examples stay in the chapters themselves.

Readers who would like to run R without installing anything can use the
[public browser practice pages](https://jinsong-chen.github.io/QMSBR/practice/) that accompany Chapters 3 and 4.

## What Comes Later

Three argument chapters extend the argument-based approach. Argument
Chapter A, *How to Read a Quantitative Study as an Argument*, develops
the reading, evaluation, and construction of one question-level argument.
Argument Chapter B, *Extending the Argument-Based Approach to Quantitative
Studies*, develops arguments with several statistical strands and studies
with coordinated questions. Argument Chapter C concerns arguments involving
latent variables or an additional measurement layer, drawing on Part Two.
The argument chapters are optional, and no numbered chapter depends on
them.

Argument Chapter C is reserved for that further development.

Part Two of the book turns to latent-variable modeling, where the variable
of primary interest is one that no column of the data set holds. Its twelve
chapters proceed from factor analysis and measurement models to structural
systems, causal identification, indirect and conditional relationships,
group comparisons, incomplete data, change and study planning.
