---
title: "Part One: Foundations, Regression, and Generalized Linear Models"
subtitle: "An Introduction to the Chapters and How They Fit Together"
author: "Jinsong Chen"

lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    fontsize: 11pt
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
          \ifdefined\pdftrailerid\pdftrailerid{}\fi
          \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
          \ifdefined\pdfvariable
            \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
          \fi
          \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
          \addtokomafont{sectionentry}{\small}
          \let\chapteroldtableofcontents\tableofcontents
          \renewcommand{\tableofcontents}{\begingroup\small\chapteroldtableofcontents\endgroup}
          \usepackage{caption}
          \captionsetup[table]{
            position=above,
            format=plain,
            justification=raggedright,
            singlelinecheck=false,
            labelsep=newline,
            labelfont=bf,
            textfont=it
          }
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{float}
          \usepackage{needspace}
          \AddToHook{cmd/section/before}{\Needspace{12\baselineskip}\phantomsection}
          \AddToHook{cmd/subsection/before}{\Needspace{12\baselineskip}\phantomsection}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: intro-setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center",
  dev = if (knitr::is_latex_output()) "cairo_pdf" else "png"
)

draw_part_one_route <- function() {
  old <- par(mar = rep(.1, 4), xaxs = "i", yaxs = "i")
  on.exit(par(old), add = TRUE)
  plot.new()
  plot.window(xlim = c(0, 4), ylim = c(0, 1))

  labels <- c(
    "Chapters 1-2\nQuestions, evidence,\nrecords, and variation",
    "Chapters 3-4\nEstimation, uncertainty,\nand group comparisons",
    "Chapters 5-6\nThe same methods\ncarried out in R",
    "Chapters 7-9\nRegression, from one\npredictor to the toolbox"
  )
  fills <- c("#F3F5F6", "#E8EEF1", "#DCE7EB", "#D0E0E5")

  for (i in 0:3) {
    x0 <- i + .04
    x1 <- i + .86
    rect(x0, .18, x1, .82, col = fills[i + 1], border = "#405768",
         lwd = 1.2)
    text((x0 + x1) / 2, .50, labels[i + 1], cex = .62,
         col = "#243746", font = 2)
    if (i < 3) {
      arrows(x1 + .015, .50, i + 1.025, .50, length = .07,
             lwd = 1.2, col = "#405768")
    }
  }
}
```

## What Part One Offers

*Quantitative Methods for Social and Behavioral Research* develops
the statistical understanding needed to read empirical studies, interpret
their results, and conduct quantitative research. Part One is where the
book begins: fourteen numbered chapters that start with what a study asks
and how it produces its evidence, and end with regression models for a
range of outcomes. It is written for postgraduate students and applied researchers.
A mean, an interval, or a regression coefficient becomes useful when a
reader understands what it describes, how it was obtained, and which
question it helps answer.

Some introductory familiarity with statistics and algebra is useful, and
little else is assumed. Each chapter explains the notation it uses and
develops its calculations in worked steps, so the algebra arrives with the
idea that needs it. Calculus and proof-based mathematical statistics are
not prerequisites. Chapters 1--4 can be read without running R or SPSS:
Chapter 1 introduces the research journey, and Chapters 2--4 teach
statistics directly, through explanations, figures, and worked
calculations. Computation joins the statistics in Chapters 5 and 6.

Results are reported to four decimals from full-precision computation
unless a source's own precision applies.

## How the Chapters Build on Each Other

Nine chapters are available, and they form one route. Chapter 1,
*The Quantitative Research Journey*, orients the reader in questions,
evidence, design, and interpretation. Chapter 2, *Data, Distributions, and
Variation*, develops records, distributions, and variation across repeated
samples. Chapter 3, *Estimates, Uncertainty, and Statistical Tests*,
develops estimation, intervals, and tests. Chapter 4, *Group Comparisons
with Analysis of Variance*, develops comparisons among several groups.
Chapter 5, *Description and Basic Inference with R*, and Chapter 6,
*Group Comparisons and Statistical Modelling with R*, make those
foundations executable. Chapters 7--9 then develop regression from one
predictor to the regression toolbox: Chapter 7, *Simple Linear
Regression*; Chapter 8, *Multiple Regression I: Conditioning, Checks, and
Coordinated Questions*; and Chapter 9, *Multiple Regression II:
Categorical Predictors, Model Comparison, Nonlinearity, and Robust
Uncertainty*.

```{r}
#| label: fig-part-one-route
#| fig-width: 7
#| fig-height: 1.75
#| out-width: "100%"
#| fig-cap: "The route through the available chapters, from questions and records to inference, computation, and regression."
#| fig-alt: "Four boxes connected from left to right: Chapters 1 to 2, questions, evidence, records, and variation; Chapters 3 to 4, estimation, uncertainty, and group comparisons; Chapters 5 to 6, the same methods carried out in R; Chapters 7 to 9, regression from one predictor to the regression toolbox."

draw_part_one_route()
```

@tbl-part-one-sequence states what each chapter is responsible for and what
it builds on.

| Chapter | What it is responsible for | What it builds on |
|:--|:--|:--|
| **1. The Quantitative Research Journey** | Turning a concern into a question with a stated target quantity; how a study produces its evidence; how comparison, timing, sampling, and assignment support descriptive, associational, predictive, and causal answers; reading a reported result as a bounded claim | The starting point of the part |
| **2. Data, Distributions, and Variation** | Reading records, identifiers, codes, and denominators; sampling schemes; distributions, centre, spread, and groups; relationships between two variables; how a summary varies across repeated samples | Reads the dataset behind Chapter 1's account of evidence and applies its design reasoning to an interpretation |
| **3. Estimates, Uncertainty, and Statistical Tests** | The target quantity, the rule that estimates it, and the value obtained; standard errors and the conditions behind them; confidence intervals and tests; magnitude, error rates, power, and several comparisons weighed together | Turns Chapter 2's sampling distributions and standard error into inference from one observed sample |
| **4. Group Comparisons with Analysis of Variance** | One-way and two-way analysis of variance; sums of squares, degrees of freedom, mean squares, and the F test; contrasts and comparison families; interaction and simple effects | Applies Chapter 2's description of group variation and Chapter 3's inference to comparisons among several groups |
| **5. Description and Basic Inference with R** | Preparing records in base R with explicit missing values; description and graphics; probability calculations; simulated sampling variation, bias, and standard error; mean intervals and z and t tests; covariance and correlation matrices and selected correlation inference; a short reproducible report | The computational route through Chapter 2 and the mean inference of Chapter 3 |
| **6. Group Comparisons and Statistical Modelling with R** | The one-way and factorial analyses in R, with contrasts and comparison families; a small additional plotting vocabulary; proportions and counts; a bounded analysis of covariance; simulation, precision and power, and one scalar likelihood | Extends Chapter 4 and Chapter 3's categorical and planning lessons on Chapter 5's computational foundation |
| **7. Simple Linear Regression** | A line fitted by least squares; slope, intercept, fitted values, and residuals; the matched correlation, slope, and F tests; an interval for a mean response and an interval for a new case; form, spread, influence, and supported predictor range | Uses the covariance, correlation, and interval reasoning of Chapters 2, 3, and 5 |
| **8. Multiple Regression I: Conditioning, Checks, and Coordinated Questions** | Several predictors and what conditioning changes; coefficient covariance and the quantities built from it; a fixed profile, standardization, and held-out prediction as different questions; several separately owed answers in one study | Extends Chapter 7's line with Chapter 5's partial correlation and Chapter 6's adjusted comparison |
| **9. Multiple Regression II: Categorical Predictors, Model Comparison, Nonlinearity, and Robust Uncertainty** | Factors, indicator coding, and declared contrasts; nested model comparison and information criteria; polynomial terms and a changed outcome scale; heteroskedasticity-robust, cluster-robust, and bootstrap uncertainty; a randomized design whose pupils sit in schools | Collects the regression toolbox on Chapters 7--8 and supplies what the later chapters need |

: What each available chapter is responsible for. {#tbl-part-one-sequence tbl-colwidths="[24,46,30]"}

Each idea has one principal home, and later chapters use it rather than
teach it again. Chapter 1 previews uncertainty while reading a study;
Chapter 2 develops sampling distributions and the standard error through
repeated samples; Chapter 3 develops their inferential use and states the
conditions they require. Chapter 2 explains the descriptive ingredients of
group variation; Chapter 4 develops the full analysis-of-variance
decomposition and its inferential use. Where a later chapter needs an
earlier idea, it recalls it briefly and then puts it to work in the new
setting.

## How to Read a Chapter and Use the Concept Reference

Chapters 1--9 share an arrangement. A **Chapter overview** box opens the
chapter with the route it follows and what a reader will be able to do
afterwards. Numbered sections develop the material in order. A closing
**Reading Bridge** brings the principal ideas together and connects them
to the next chapter, and a selective **Key terms** table gives a compact
reference to the concepts explained principally there. Unnumbered material
follows the numbered sections: exercises for working with the ideas,
appendices holding fuller derivations and specialized extensions, and the
references. Later chapters may use a different arrangement as the
collection grows.

There are two sensible routes through the available chapters. The first is
the chapter sequence. The second pairs statistics with computation:
Chapters 2 and 3 with Chapter 5, then Chapters 3 and 4 with Chapter 6.
Either way, follow a method from its data and question through its
calculation to its interpretation; a result separated from the question it
answers is difficult to use.

The [**Part One Concept Reference**](https://jinsong-chen.github.io/QMSBR/part-one/part_one_concept_reference.html) is published beside this introduction.
It gathers selected concepts from Chapters 1--9 into six tables, arranged
by concept group and then by term, with each term's meaning and the place
where the book explains it principally; a closing table records the
notation those chapters use. It is a place to look something up while
reading rather than a substitute for reading, because first-use
explanations and worked examples stay in the chapters themselves.

Readers who would like to run R without installing anything can use the
[public browser practice pages](https://jinsong-chen.github.io/QMSBR/practice/) that accompany Chapters 3 and 4.

## What Comes Later

Part One continues with five further chapters. Chapter 10, *Interaction
and Moderation*, develops product terms, conditional slopes, and probing
within joint support. Chapter 11, *Binary Logistic Regression*, develops
the Bernoulli likelihood, the logit link, and the probability, odds, and
log-odds scales. Chapter 12, *Extending the GLM: Nominal, Ordinal, and
Count Outcomes*, varies only the response distribution and the link for
nominal, ordered, and count or rate outcomes. Chapter 13, *Prediction,
Cross-Validation, and Regularization*, develops prediction as a separate
aim, with held-out error and penalized fitting. Chapter 14, *Regression
Pitfalls and Causal Design*, examines what can undermine a
regression-based claim: reversals and Simpson's paradox, regression to the
mean, selection, measurement, dependence, and what a study's design
supplies that adjustment cannot.

Two supplementary units extend Chapter 1's argument-based reading: *How to
Read a Quantitative Study as an Argument* and *Extending the
Argument-Based Approach to Quantitative Studies*. Both are optional, and
no numbered chapter depends on either.

Chapters 10--14 and the two supplementary units are not yet available.

Part Two of the book turns to latent-variable modeling, where the variable
of primary interest is one that no column of the data set holds.
