---
title: "Specifying and Estimating Confirmatory Factor Models"
author: "Jinsong Chen"
bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
execute:
  echo: false
  warning: false
  message: false
  cache: false
format:
  pdf:
    output-file: chapter_03.pdf
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    include-in-header:
      - file: ../supplement/pdf-graphics-compat.tex
      - text: |
          \setlength{\parindent}{0pt}
          \usepackage{float}
          \usepackage{needspace}
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{caption}
          \captionsetup[table]{position=above,format=plain,justification=raggedright,singlelinecheck=false,labelsep=newline,labelfont=bf,textfont=it}
  html:
    toc: true
    toc-depth: 2
    number-sections: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(dev=if(knitr::is_latex_output()) "cairo_pdf" else "png",
  dpi=160,fig.pos="H",fig.align="center")
options(digits=3,width=80)

if (!requireNamespace("lavaan", quietly = TRUE))
  stop("Required package: lavaan")

# ---- SLE population and in-memory samples ---------------------------------
# The population is declared here and every sample is simulated from a fixed
# seed, so this chapter reads no data file and sources no script. These are the
# declarations and seeds that produced the distributed sle_continuous.csv.

sle_population <- function() {
  items <- c(paste0("P", 1:4), paste0("S", 1:4), paste0("B", 1:4))
  factors <- c("Planning", "Strategy", "Belonging")
  Lambda_z <- matrix(0, 12, 3, dimnames = list(items, factors))
  Lambda_z[1:4, 1] <- c(.78, .74, .70, .55)
  Lambda_z[5:8, 2] <- c(.76, .72, .68, .55)
  Lambda_z[9:12, 3] <- c(.80, .74, .67, .50)
  Lambda_z[4, 2] <- .32
  Lambda_z[8, 3] <- .30
  Phi <- matrix(c(1,.42,.30, .42,1,.35, .30,.35,1), 3, 3,
                dimnames = list(factors, factors))
  common <- Lambda_z %*% Phi %*% t(Lambda_z)
  Theta_z <- diag(1 - diag(common))
  dimnames(Theta_z) <- list(items, items)
  # Shared peer context not fully represented by the broad factors.
  Theta_z[9, 10] <- Theta_z[10, 9] <- .12
  scale <- c(8,10,12,9, 10,8,11,9, 10,12,8,11)
  D <- diag(scale)
  nu <- setNames(c(50,52,48,51, 49,50,52,48, 51,49,50,52), items)
  Lambda <- D %*% Lambda_z
  Theta <- D %*% Theta_z %*% D
  dimnames(Lambda) <- dimnames(Lambda_z)
  dimnames(Theta) <- dimnames(Theta_z)
  stopifnot(min(eigen(Phi, symmetric=TRUE)$values) > 0,
            min(eigen(Theta, symmetric=TRUE)$values) > 0)
  list(version="SLE-continuous-v1", items=items, factors=factors,
       Lambda_z=Lambda_z, Phi=Phi, Theta_z=Theta_z, scale=scale,
       nu=nu, alpha=setNames(rep(0,3),factors), Lambda=Lambda, Theta=Theta)
}

sle_manifest <- function()
  data.frame(sample = LETTERS[1:3],
    role = c("development", "confirmation", "revalidation"),
    n = 360L, population = "SLE-continuous-v1", seed = 20260921:20260923,
    rng = "Mersenne-Twister/Inversion/Rejection", stringsAsFactors = FALSE)

sle_simulate <- function(role) {
  manifest <- sle_manifest()
  row <- manifest[manifest$role == role, , drop = FALSE]
  stopifnot(nrow(row) == 1L)
  pop <- sle_population()
  RNGkind("Mersenne-Twister", "Inversion", "Rejection")
  set.seed(row$seed)
  n <- row$n
  eta <- matrix(rnorm(n * 3), n, 3) %*% chol(pop$Phi)
  epsilon <- matrix(rnorm(n * 12), n, 12) %*% chol(pop$Theta)
  Y <- sweep(eta %*% t(pop$Lambda) + epsilon, 2, pop$nu, "+")
  colnames(Y) <- pop$items
  x <- as.data.frame(Y)
  stopifnot(identical(dim(x), c(360L, 12L)), !anyNA(x))
  list(x = x, ids = sprintf("%s%03d", row$sample, seq_len(n)), manifest = row)
}

# ---- the candidate frozen at the end of the exploratory chapter ------------
# Declared in full rather than loaded from a saved object.
sle_candidate <- function() {
  primary <- paste("Planning =~ p1*P1 + p2*P2 + P3 + P4",
    "Strategy =~ S1 + S2 + S3 + S4 + P4",
    "Belonging =~ B1 + B2 + B3 + B4 + S4", sep = "\n")
  items <- sle_population()$items
  list(version = "SLE-candidate-v1", frozen_on = "2026-09-04",
    active_items = items, factors = c("Planning", "Strategy", "Belonging"),
    syntax = primary,
    rivals = list(
      M0_simple = paste("Planning =~ p1*P1 + p2*P2 + P3 + P4",
        "Strategy =~ S1 + S2 + S3 + S4",
        "Belonging =~ B1 + B2 + B3 + B4", sep = "\n"),
      M1_cross = primary,
      M2_orthogonal = paste(primary, "Planning ~~ 0*Strategy + 0*Belonging",
        "Strategy ~~ 0*Belonging", sep = "\n"),
      M3_equal = paste(primary, "p1 == p2", sep = "\n"),
      M4_one = paste("General =~", paste(items, collapse = " + "))),
    free_parameters = c(M0_simple = 27, M1_cross = 29, M2_orthogonal = 26,
      M3_equal = 28, M4_one = 24),
    df = c(M0_simple = 51, M1_cross = 49, M2_orthogonal = 52, M3_equal = 50,
      M4_one = 54),
    settings = list(estimator = "ML", likelihood = "normal", std.lv = TRUE,
      meanstructure = FALSE),
    secondary_loadings = c("P4 on Strategy", "S4 on Belonging"),
    rationale = "Item content recorded before generation plus stable Sample A secondary relations; no .30 cutoff, no item exclusions, no residual covariance freed.",
    history = "Sample A used for development only; B and C were not analysed before this specification was fixed.")
}

# ---- render helpers --------------------------------------------------------

pt2_table <- function(x, digits=3, caption=NULL, row.names=FALSE) {
  x <- as.data.frame(x,check.names=FALSE)
  integer_columns <- c("df","Df","Factors","Moments","Free","npar","Dimension",
    "Effective_parameters","Candidate_df","Delta_df","Missing")
  for(nm in names(x)) if(is.numeric(x[[nm]])) {
    d <- if(nm %in% integer_columns) 0 else digits
    x[[nm]] <- formatC(x[[nm]],digits=d,format="f")
  }
  names(x) <- gsub("_"," ",names(x),fixed=TRUE)
  out <- knitr::kable(x,caption=caption,row.names=row.names,booktabs=TRUE,longtable=FALSE)
  if(knitr::is_latex_output()) knitr::asis_output(paste0(
    "\\Needspace{",min(nrow(x)+7,25),"\\baselineskip}\n\n",
    paste(out,collapse="\n"),"\n\n")) else out
}

pt2_course_map <- function(labels) {
  n <- length(labels); par(mar=c(0,0,0,0))
  plot.new(); plot.window(xlim=c(0,n),ylim=c(0,1))
  for(i in seq_len(n)) {
    rect(i-.95,.25,i-.15,.75,col="#edf2f5",border="#405768")
    text(i-.55,.5,labels[i],cex=.8)
    if(i<n) arrows(i-.13,.5,i+.03,.5,length=.08)
  }
}

pt2_num <- function(x, digits = 3) formatC(x, digits = digits, format = "f")
pt2_p <- function(x) ifelse(x < .001, "< .001", paste0("= ", pt2_num(x)))

pt2_heatmap <- function(M,title="",limit=max(abs(M)),digits=NULL,cex=.7) {
  nr <- nrow(M); nc <- ncol(M)
  par(mar=c(4,4,.9,1))
  palette <- grDevices::colorRampPalette(c("#8c510a","#f7f7f7","#2166ac"))(101)
  image(1:nc,1:nr,t(M[nr:1,,drop=FALSE]),zlim=c(-limit,limit),
        col=palette,axes=FALSE,xlab="",ylab="",main=title)
  axis(1,at=1:nc,labels=colnames(M),las=2,cex.axis=cex)
  axis(2,at=1:nr,labels=rev(rownames(M)),las=2,cex.axis=cex)
  if (!is.null(digits)) for (i in 1:nr) for (j in 1:nc) if(is.finite(M[i,j]))
    text(j,nr+1-i,formatC(M[i,j],digits=digits,format="f"),
         cex=cex*.8,col=if(abs(M[i,j])>limit*.6) "white" else "black")
  box()
}

# ---- neutral six-indicator microcase and the donor teaching record ---------

pt3_microcase <- function(cross=0) {
  Lambda <- matrix(c(1,.8,1.2,0,0,0,0,0,cross,1,1.1,.9),6,2)
  dimnames(Lambda) <- list(paste0("Y",1:6),c("F1","F2"))
  Phi <- matrix(c(4,1.2,1.2,9),2,2,dimnames=list(colnames(Lambda),colnames(Lambda)))
  Theta <- diag(c(2,3,4,3,4,5)); dimnames(Theta) <- list(rownames(Lambda),rownames(Lambda))
  nu <- c(10,12,11,15,14,13)
  Sigma <- Lambda%*%Phi%*%t(Lambda)+Theta
  list(Lambda=Lambda,Phi=Phi,Theta=Theta,nu=nu,Sigma=Sigma)
}

# Simulated in memory from its declared population and seed; the distributed
# CSV holds the same 500 records.
pt3_donor <- function() {
  pop <- pt3_microcase(cross = .25)
  RNGkind("Mersenne-Twister", "Inversion", "Rejection"); set.seed(20260930L)
  n <- 500L
  eta <- matrix(rnorm(n * 2), n, 2) %*% chol(pop$Phi)
  epsilon <- matrix(rnorm(n * 6), n, 6) %*% chol(pop$Theta)
  x <- sweep(eta %*% t(pop$Lambda) + epsilon, 2, pop$nu, "+")
  colnames(x) <- rownames(pop$Lambda)
  list(x = as.data.frame(x), population = pop, n = n, seed = 20260930L,
    id = "D-continuous-v1",
    role = "Separate synthetic continuous donor microcase; never Sample A, B or C.",
    models = list(
      D1 = "F1 =~ Y1 + Y2 + Y3\nF2 =~ Y4 + Y5 + Y6",
      D2 = "F1 =~ Y1 + Y2\nF2 =~ Y3 + Y4 + Y5 + Y6",
      D3 = "F1 =~ Y1 + Y2 + Y3\nF2 =~ Y4 + Y5 + Y6 + Y3",
      D4 = "F1 =~ Y1 + Y2 + Y3\nF2 =~ Y4 + Y5 + Y6 + Y3\nF1 ~~ 0*F2"),
    restrictions = "All unspecified loadings and residual covariances fixed at zero; residual variances free; D4 factor covariance fixed at zero; no equalities.",
    analytic_status = "D1-D4 specified before microcase generation; numerical teaching comparisons only.")
}

# ---- fit and audit ---------------------------------------------------------

pt3_fit <- function(syntax,x,std.lv=TRUE,meanstructure=FALSE) {
  warns <- character()
  fit <- withCallingHandlers(lavaan::cfa(syntax,data=x,estimator="ML",
    likelihood="normal",std.lv=std.lv,meanstructure=meanstructure,
    missing="listwise",se="standard",test="standard",information="expected"),
    warning=function(w) { warns <<- c(warns,conditionMessage(w)); invokeRestart("muffleWarning") })
  list(fit=fit,warnings=warns)
}

pt3_audit <- function(record,items=NULL) {
  fit <- record$fit; opt <- lavaan::lavInspect(fit,"options")
  est <- lavaan::lavInspect(fit,"est"); Sigma <- lavaan::fitted(fit)$cov
  if(is.null(items)) items <- lavaan::lavNames(fit,"ov")
  pt <- lavaan::parTable(fit)
  # Effective free coordinates: raw free slots less the rank of equality rows.
  eq <- pt[pt$op=="==",,drop=FALSE]
  nfree <- length(unique(pt$free[pt$free>0]))-nrow(eq)
  u <- length(items)*(length(items)+1)/2+if(opt$meanstructure) length(items) else 0
  Delta <- lavaan::lavInspect(fit,"delta")
  out <- c(Converged=isTRUE(lavaan::lavInspect(fit,"converged")),
    No_warnings=length(record$warnings)==0L,
    Positive_residual_variances=all(diag(est$theta)>0),
    Positive_factor_covariance=min(eigen(est$psi,symmetric=TRUE)$values)>1e-8,
    Positive_implied_covariance=min(eigen(Sigma,symmetric=TRUE)$values)>1e-8,
    ML=opt$estimator=="ML",Normal=opt$likelihood=="normal",
    Expected_information=all(opt$information=="expected"),
    Standard_errors=opt$se=="standard",Complete=opt$missing=="listwise",
    Item_order=identical(lavaan::lavNames(fit,"ov"),items),
    Count=abs(u-nfree-lavaan::fitMeasures(fit,"df"))<1e-8)
  list(checks=out,nfree=nfree,moments=u,df=u-nfree,
    delta_rank=qr(Delta,tol=1e-7)$rank,delta_columns=ncol(Delta),
    max_gradient=max(abs(lavaan::lavInspect(fit,"gradient"))),
    min_information_eigen=min(eigen(lavaan::lavInspect(fit,"information"),symmetric=TRUE)$values),
    min_residual=min(diag(est$theta)),min_phi_eigen=min(eigen(est$psi,symmetric=TRUE)$values),
    options=opt,parameters=lavaan::parameterEstimates(fit,standardized=TRUE),
    matrices=est,Sigma=Sigma,vcov=lavaan::vcov(fit))
}

pt3_build <- function() {
  candidate <- sle_candidate()
  input <- sle_simulate("confirmation")
  primary <- pt3_fit(candidate$syntax, input$x)
  audit <- pt3_audit(primary, candidate$active_items)
  stopifnot(all(audit$checks), audit$nfree == 29, audit$df == 49,
    audit$delta_rank == audit$delta_columns)
  marker <- pt3_fit(candidate$syntax, input$x, std.lv = FALSE)
  means <- pt3_fit(candidate$syntax, input$x, meanstructure = TRUE)
  simple_means <- pt3_fit(candidate$rivals$M0_simple, input$x, meanstructure = TRUE)
  stopifnot(all(pt3_audit(marker)$checks), all(pt3_audit(means)$checks),
    all(pt3_audit(simple_means)$checks),
    max(abs(lavaan::fitted(marker$fit)$cov - audit$Sigma)) < 1e-3,
    max(abs(lavaan::fitted(means$fit)$cov - audit$Sigma)) < 1e-3,
    max(abs(lavaan::fitted(means$fit)$mean - colMeans(input$x))) < 1e-5)
  fit_measure_names <- c("chisq", "df", "pvalue", "rmsea", "rmsea.ci.lower",
    "rmsea.ci.upper", "cfi", "tli", "srmr", "logl", "aic", "bic",
    "baseline.chisq", "baseline.df")
  measures <- lavaan::fitMeasures(primary$fit, fit_measure_names)
  stopifnot(all(is.finite(measures)))
  pop <- pt3_microcase()
  stopifnot(abs(pop$Sigma[2, 3] - .8 * 1.2 * 4) < 1e-12,
    abs(pop$Sigma[2, 5] - .8 * 1.1 * 1.2) < 1e-12)
  D <- diag(1 / sqrt(diag(pop$Phi)))
  scaled <- list(Lambda = pop$Lambda %*% solve(D), Phi = D %*% pop$Phi %*% D)
  stopifnot(max(abs(scaled$Lambda %*% scaled$Phi %*% t(scaled$Lambda) +
    pop$Theta - pop$Sigma)) < 1e-12)
  registry <- pt3_donor()
  donor <- pt3_fit(registry$models$D1, registry$x)
  donor_marker <- pt3_fit(registry$models$D1, registry$x, std.lv = FALSE)
  stopifnot(all(pt3_audit(donor)$checks), all(pt3_audit(donor_marker)$checks),
    max(abs(lavaan::fitted(donor$fit)$cov -
      lavaan::fitted(donor_marker$fit)$cov)) < 1e-4)
  list(version = "SLE-CFA-handoff-v1", candidate = candidate,
    input = input, fit = primary$fit, warnings = primary$warnings, audit = audit,
    marker = marker, means = means, simple_means = simple_means,
    fit_measure_names = fit_measure_names, measures = measures,
    S_N = unclass(lavaan::lavInspect(primary$fit, "sampstat")$cov),
    parameter_table = lavaan::parTable(primary$fit),
    wald = lavaan::lavTestWald(primary$fit, "p1 == p2"),
    microcase = pop, scaled = scaled, donor = donor, donor_marker = donor_marker,
    donor_registry = registry,
    metadata = list(R = as.character(getRversion()),
      lavaan = as.character(packageVersion("lavaan")),
      sample = "B", role = "confirmation",
      status = "Estimation only; this chapter does not assess model adequacy.",
      sample_C = "Not read in this chapter."))
}

pt3_estimation_plot <- function() {
  par(mar=c(1,1,1,1));plot.new();plot.window(xlim=c(0,10),ylim=c(0,5))
  t <- seq(1,8,length.out=150); y <- 1+.08*(t-4)^2
  lines(t,y,lwd=2,col="#405768")
  points(5,4,pch=16);text(5,4.4,"Observed moments S",cex=.9)
  points(4.5,1.02,pch=17,col="#9c5600");text(6.6,.45,"Selected implied moments",cex=.85)
  arrows(5,3.8,4.5,1.25,length=.09,lty=2)
  text(2,2.6,"Restricted model\nset of moments",cex=.85)
  text(7.7,3,"ML discrepancy;\nnot Euclidean distance",cex=.8)
}

# ---- shared Part Two path-diagram grammar ----------------------------------
# Rectangles are recorded variables, circles are variables introduced by the
# model, single-headed arrows are loadings, double-headed arrows covariances.
# Drawn with base graphics: no external renderer, script or figure directory.

pt2_ink        <- "#243746"
pt2_box_fill   <- "#fff1c7"
pt2_circ_fill  <- "#dcecf5"
pt2_path_col   <- "#2c5d77"
pt2_cross_col  <- "#94a3ad"
pt2_cov_col    <- "#8c4f2b"
pt2_resid_col  <- "#596b75"

# Render a label that may be plain text (".70", "0") or a plotmath
# expression string ("lambda[11]", "eta[1]").
.pt2_label <- function(x) {
  if (is.expression(x) || is.language(x)) return(x)
  if (!nzchar(x)) return("")
  if (grepl("^[-.0-9]", x)) return(x)
  # Plotmath where the label parses (eta[1], phi[12]); literal text otherwise,
  # so that factor names such as "School support" render as written.
  out <- tryCatch(parse(text = x), error = function(e) NULL)
  if (is.null(out)) x else out
}

.pt2_arrow <- function(x0, y0, x1, y1, dashed = FALSE, both = FALSE,
                       col = pt2_path_col, lwd = 1.35) {
  arrows(x0, y0, x1, y1, length = .07, angle = 20,
         code = if (both) 3 else 2,
         lty = if (dashed) 2 else 1, lwd = lwd, col = col)
}

.pt2_box <- function(x, y, label, hw = .74, hh = .30) {
  rect(x - hw, y - hh, x + hw, y + hh,
       col = pt2_box_fill, border = pt2_ink, lwd = 1.35)
  text(x, y, .pt2_label(label), cex = .92)
}

.pt2_circle <- function(x, y, label, r = .55) {
  symbols(x, y, circles = r, inches = FALSE, add = TRUE,
          bg = pt2_circ_fill, fg = pt2_ink, lwd = 1.35)
  text(x, y, .pt2_label(label), cex = .95)
}

# Draw a label at a fraction along a path, nudged perpendicular to it.
.pt2_path_label <- function(x0, y0, x1, y1, label, at = .58, nudge = .15,
                            col = pt2_ink, cex = .70) {
  if (is.null(label) || !nzchar(label)) return(invisible(NULL))
  text(x0 + at * (x1 - x0), y0 + at * (y1 - y0) + nudge,
       .pt2_label(label), cex = cex, col = col)
}

# A double-headed covariance arc drawn to the left of two circles.
.pt2_cov_arc <- function(x, y1, y2, label, r = .55, bulge = .85) {
  tt <- seq(0, 1, length.out = 60)
  ym <- (y1 + y2) / 2
  half <- abs(y2 - y1) / 2 - r * .55
  ys <- ym + half * cos(pi * tt) * sign(y1 - y2)
  xs <- x - bulge * sin(pi * tt)
  lines(xs, ys, col = pt2_cov_col, lwd = 1.4)
  .pt2_arrow(xs[2], ys[2], xs[1], ys[1], col = pt2_cov_col, lwd = 1.4)
  n <- length(xs)
  .pt2_arrow(xs[n - 1], ys[n - 1], xs[n], ys[n], col = pt2_cov_col, lwd = 1.4)
  if (!is.null(label) && nzchar(label))
    text(x - bulge - .14, ym, .pt2_label(label), adj = 1, cex = .82)
  invisible(NULL)
}

.pt2_canvas <- function(xlim, ylim) {
  op <- par(mar = rep(.15, 4), xaxs = "i", yaxs = "i")
  plot.new()
  plot.window(xlim = xlim, ylim = ylim)
  op
}

.pt2_indicator_y <- function(p, top = 7.6, bottom = .9) {
  if (p == 1) return(mean(c(top, bottom)))
  seq(top, bottom, length.out = p)
}

# ---------------------------------------------------------------------------
# First-order model with m correlated (or orthogonal) factors.
#
#   assign  integer vector of length p giving each indicator's primary factor,
#           or a p x m logical matrix for a fully general pattern.
#   cross   optional p x m logical matrix of additional (dashed) loadings.
# ---------------------------------------------------------------------------
pt2_fa_diagram <- function(assign, m = NULL, cross = NULL,
                           loading_labels = NULL, factor_labels = NULL,
                           indicator_labels = NULL, factor_cov_labels = NULL,
                           show_residuals = TRUE, x_factor = 2.2,
                           x_indicator = 7.0) {
  if (is.matrix(assign)) {
    primary <- assign
    p <- nrow(primary); m <- ncol(primary)
  } else {
    p <- length(assign); if (is.null(m)) m <- max(assign)
    primary <- matrix(FALSE, p, m)
    primary[cbind(seq_len(p), assign)] <- TRUE
  }
  if (is.null(cross)) cross <- matrix(FALSE, p, m)
  if (is.null(factor_labels))
    factor_labels <- paste0("eta[", seq_len(m), "]")
  if (is.null(indicator_labels))
    indicator_labels <- paste0("Y[", seq_len(p), "]")

  yi <- .pt2_indicator_y(p)
  hh <- if (p > 1) min(.30, .40 * abs(yi[1] - yi[2])) else .30
  yf <- if (m == 1) mean(range(yi)) else
    seq(max(yi) - .9, min(yi) + .9, length.out = m)
  x_resid <- x_indicator + 3.0
  op <- .pt2_canvas(c(0, x_resid + 1.0), c(.3, 8.2))
  on.exit(par(op))

  # factor covariances: adjacent pairs, arcs on the left
  if (m > 1 && !is.null(factor_cov_labels)) {
    if (length(factor_cov_labels) == 1L && m > 2)
      factor_cov_labels <- rep(factor_cov_labels, m - 1)
    for (k in seq_len(m - 1))
      .pt2_cov_arc(x_factor - .55, yf[k], yf[k + 1],
                   factor_cov_labels[k],
                   bulge = .70 + .30 * (k %% 2))
  }

  for (j in seq_len(p)) for (k in seq_len(m)) {
    is_cross <- isTRUE(cross[j, k])
    if (!isTRUE(primary[j, k]) && !is_cross) next
    .pt2_arrow(x_factor + .55, yf[k], x_indicator - .74, yi[j],
               dashed = is_cross,
               col = if (is_cross) pt2_cross_col else pt2_path_col,
               lwd = if (is_cross) 1.1 else 1.35)
    if (!is.null(loading_labels))
      .pt2_path_label(x_factor + .55, yf[k], x_indicator - .74, yi[j],
                      loading_labels[j, k], nudge = if (is_cross) -.16 else .16,
                      col = if (is_cross) pt2_cross_col else pt2_ink)
  }

  for (j in seq_len(p)) {
    .pt2_box(x_indicator, yi[j], indicator_labels[j], hh = hh)
    if (show_residuals) {
      .pt2_arrow(x_resid - .52, yi[j], x_indicator + .74, yi[j],
                 col = pt2_resid_col, lwd = 1.15)
      text(x_resid, yi[j], .pt2_label(paste0("epsilon[", j, "]")), cex = .85)
    }
  }
  for (k in seq_len(m)) .pt2_circle(x_factor, yf[k], factor_labels[k])
  invisible(NULL)
}

# ---------------------------------------------------------------------------
# Higher-order model: one second-order factor above m first-order factors.
# ---------------------------------------------------------------------------
pt2_higher_order_diagram <- function(assign, m = NULL, gamma_labels = NULL,
                                     loading_labels = NULL,
                                     factor_labels = NULL,
                                     indicator_labels = NULL,
                                     higher_label = "xi",
                                     show_disturbances = TRUE) {
  p <- length(assign); if (is.null(m)) m <- max(assign)
  if (is.null(factor_labels)) factor_labels <- paste0("eta[", seq_len(m), "]")
  if (is.null(indicator_labels)) indicator_labels <- paste0("Y[", seq_len(p), "]")
  x_high <- 1.0; x_factor <- 4.0; x_indicator <- 8.0; x_resid <- 10.8

  yi <- .pt2_indicator_y(p)
  yf <- seq(max(yi) - .9, min(yi) + .9, length.out = m)
  yh <- mean(range(yf))
  op <- .pt2_canvas(c(0, x_resid + 1.0), c(.3, 8.2))
  on.exit(par(op))

  for (k in seq_len(m)) {
    .pt2_arrow(x_high + .55, yh, x_factor - .55, yf[k])
    if (!is.null(gamma_labels))
      .pt2_path_label(x_high + .55, yh, x_factor - .55, yf[k],
                      gamma_labels[k], at = .55, nudge = .17)
    if (show_disturbances) {
      .pt2_arrow(x_factor, yf[k] + 1.05, x_factor, yf[k] + .55,
                 col = pt2_resid_col, lwd = 1.15)
      text(x_factor, yf[k] + 1.24,
           .pt2_label(paste0("zeta[", k, "]")), cex = .82)
    }
  }
  for (j in seq_len(p)) {
    k <- assign[j]
    .pt2_arrow(x_factor + .55, yf[k], x_indicator - .74, yi[j])
    if (!is.null(loading_labels))
      .pt2_path_label(x_factor + .55, yf[k], x_indicator - .74, yi[j],
                      loading_labels[j], nudge = .16)
    .pt2_box(x_indicator, yi[j], indicator_labels[j])
    .pt2_arrow(x_resid - .52, yi[j], x_indicator + .74, yi[j],
               col = pt2_resid_col, lwd = 1.15)
    text(x_resid, yi[j], .pt2_label(paste0("epsilon[", j, "]")), cex = .85)
  }
  for (k in seq_len(m)) .pt2_circle(x_factor, yf[k], factor_labels[k])
  .pt2_circle(x_high, yh, higher_label)
  invisible(NULL)
}

# ---------------------------------------------------------------------------
# Bifactor model: one general factor on every indicator, plus orthogonal
# group factors on their own blocks.
# ---------------------------------------------------------------------------
pt2_bifactor_diagram <- function(assign, m = NULL, general_label = "eta[G]",
                                 group_labels = NULL, indicator_labels = NULL,
                                 general_loadings = NULL,
                                 group_loadings = NULL) {
  p <- length(assign); if (is.null(m)) m <- max(assign)
  if (is.null(group_labels))
    group_labels <- paste0("eta[", seq_len(m), "]")
  if (is.null(indicator_labels)) indicator_labels <- paste0("Y[", seq_len(p), "]")
  x_gen <- 1.0; x_group <- 4.2; x_indicator <- 8.0; x_resid <- 10.8

  yi <- .pt2_indicator_y(p)
  yg <- seq(max(yi) - .8, min(yi) + .8, length.out = m)
  ygen <- mean(range(yi))
  op <- .pt2_canvas(c(0, x_resid + 1.0), c(.3, 8.2))
  on.exit(par(op))

  # general factor reaches every indicator; drawn first and lighter so the
  # group structure stays readable
  for (j in seq_len(p)) {
    .pt2_arrow(x_gen + .55, ygen, x_indicator - .74, yi[j],
               col = pt2_cross_col, lwd = 1.0)
    if (!is.null(general_loadings))
      .pt2_path_label(x_gen + .55, ygen, x_indicator - .74, yi[j],
                      general_loadings[j], at = .30, nudge = .15,
                      col = pt2_cross_col, cex = .64)
  }
  for (j in seq_len(p)) {
    k <- assign[j]
    .pt2_arrow(x_group + .55, yg[k], x_indicator - .74, yi[j])
    if (!is.null(group_loadings))
      .pt2_path_label(x_group + .55, yg[k], x_indicator - .74, yi[j],
                      group_loadings[j], nudge = .16)
    .pt2_box(x_indicator, yi[j], indicator_labels[j])
    .pt2_arrow(x_resid - .52, yi[j], x_indicator + .74, yi[j],
               col = pt2_resid_col, lwd = 1.15)
    text(x_resid, yi[j], .pt2_label(paste0("epsilon[", j, "]")), cex = .85)
  }
  for (k in seq_len(m)) .pt2_circle(x_group, yg[k], group_labels[k])
  .pt2_circle(x_gen, ygen, general_label)
  text(x_gen, ygen - 1.15, "all factors orthogonal", cex = .68,
       col = pt2_resid_col)
  invisible(NULL)
}

# ---------------------------------------------------------------------------
# Named-group front end used by the confirmatory chapters. `groups` is a named
# list mapping each factor to its indicators; `cross` and `residual` are
# optional data frames of additional free relations.
# ---------------------------------------------------------------------------
pt2_groups_diagram <- function(groups, correlated = TRUE, marker = FALSE,
                               cross = NULL, residual = NULL,
                               factor_cov_labels = NULL) {
  stopifnot(length(groups) >= 1L, !is.null(names(groups)))
  items <- unlist(groups, use.names = FALSE)
  assign <- rep(seq_along(groups), lengths(groups))
  p <- length(items); m <- length(groups)
  x_factor <- 2.2; x_indicator <- 7.0; x_resid <- x_indicator + 3.0
  yi <- .pt2_indicator_y(p)
  hh <- if (p > 1) min(.30, .40 * abs(yi[1] - yi[2])) else .30
  yf <- if (m == 1) mean(range(yi)) else
    vapply(seq_len(m), function(k) mean(yi[assign == k]), numeric(1))
  op <- .pt2_canvas(c(0, x_resid + 1.3), c(.3, 8.2))
  on.exit(par(op))

  if (correlated && m > 1) {
    if (is.null(factor_cov_labels)) factor_cov_labels <- rep("", m - 1)
    if (length(factor_cov_labels) == 1L) factor_cov_labels <- rep(factor_cov_labels, m - 1)
    for (k in seq_len(m - 1))
      .pt2_cov_arc(x_factor - .55, yf[k], yf[k + 1], factor_cov_labels[k],
                   bulge = .70 + .30 * (k %% 2))
  }
  for (j in seq_len(p)) {
    k <- assign[j]
    .pt2_arrow(x_factor + .55, yf[k], x_indicator - .74, yi[j])
    if (marker && j == which(assign == k)[1])
      .pt2_path_label(x_factor + .55, yf[k], x_indicator - .74, yi[j], "1",
                      at = .42, nudge = .17)
  }
  if (!is.null(cross) && nrow(cross)) for (i in seq_len(nrow(cross))) {
    k <- match(cross$factor[i], names(groups)); j <- match(cross$item[i], items)
    .pt2_arrow(x_factor + .55, yf[k], x_indicator - .74, yi[j],
               dashed = TRUE, col = pt2_cov_col, lwd = 1.3)
  }
  for (j in seq_len(p)) {
    .pt2_box(x_indicator, yi[j], items[j], hh = hh)
    .pt2_arrow(x_resid - .52, yi[j], x_indicator + .74, yi[j],
               col = pt2_resid_col, lwd = 1.15)
    text(x_resid, yi[j], .pt2_label(paste0("epsilon[", j, "]")), cex = .82)
  }
  if (!is.null(residual) && nrow(residual)) for (i in seq_len(nrow(residual))) {
    a <- match(residual$item1[i], items); b <- match(residual$item2[i], items)
    tt <- seq(0, 1, length.out = 40)
    ym <- (yi[a] + yi[b]) / 2; half <- abs(yi[b] - yi[a]) / 2 - .18
    ys <- ym + half * cos(pi * tt) * sign(yi[a] - yi[b])
    xs <- x_resid + .40 + .55 * sin(pi * tt)
    lines(xs, ys, col = pt2_cov_col, lwd = 1.4, lty = 3)
    .pt2_arrow(xs[2], ys[2], xs[1], ys[1], col = pt2_cov_col, lwd = 1.4)
    n <- length(xs)
    .pt2_arrow(xs[n-1], ys[n-1], xs[n], ys[n], col = pt2_cov_col, lwd = 1.4)
  }
  for (k in seq_len(m)) .pt2_circle(x_factor, yf[k], names(groups)[k])
  invisible(NULL)
}

# Two small panels contrasting a correlated first-order model with an
# equivalent second-order representation.
pt2_equivalence_pair <- function() {
  op <- par(mfrow = c(1, 2)); on.exit(par(op))
  pt2_groups_diagram(list(F1 = c("Y1", "Y2", "Y3"), F2 = c("Y4", "Y5", "Y6")),
                     correlated = TRUE, factor_cov_labels = "phi[12]")
  title("Correlated first-order", line = -1.4, cex.main = .95)
  pt2_higher_order_diagram(rep(1:2, each = 3),
                           factor_labels = c("F1", "F2"),
                           gamma_labels = c("1", "1"), higher_label = "G")
  title("Second-order, loadings fixed", line = -1.4, cex.main = .95)
  invisible(NULL)
}

h <- pt3_build()
pe <- h$audit$parameters
pop <- h$microcase
L <- pop$Lambda; Phi <- pop$Phi; Theta <- pop$Theta
```

{{< pagebreak >}}

::: {.callout-tip title="Chapter box"}
**Five themes.**

*A restriction is a claim.* Confirmatory analysis differs from exploratory
analysis in what it fixes, not in the equation it uses. Every zero is a
statement about the moments that could turn out to be false.

*Fixed, free and constrained are three different statuses.* A diagram shows
which relations exist; it does not show which are estimated, which are held
at a value, and which are tied to another.

*A latent variable needs a metric before it has coefficients.* Marker
scaling and fixed-factor scaling give the same implied covariance and
different printed numbers.

*Counting is necessary and not sufficient.* Moments minus free coordinates
gives degrees of freedom; identification asks the separate question of
whether the moments determine the parameters.

*Estimation minimizes a stated discrepancy.* Convergence, admissibility and
a well-conditioned solution are read before any coefficient is interpreted.

**Prerequisites.** The covariance decomposition, factor scale and
orientation, and pattern coefficients from the preceding chapters, together
with the candidate model frozen at the end of the exploratory analysis.

**Records.** A neutral six-indicator model makes the algebra visible; a
four-factor topology tests diagram-to-syntax transfer; a separate synthetic
microcase supports compact comparisons. The twelve-indicator candidate is
estimated on the confirmation sample, untouched by the development analysis.

**Scope.** We specify, identify, estimate and audit. Convergence and
admissibility are settled here; overall model adequacy is not assessed, and
no fit index is interpreted.
:::

# From exploratory structure to restricted CFA

## The change is in the restrictions

The common-factor equation has not changed:

$$
\mathbf y_i=\boldsymbol\nu+\boldsymbol\Lambda\boldsymbol\eta_i
 +\boldsymbol\epsilon_i,
\qquad
\boldsymbol\Sigma(\boldsymbol\vartheta)
=\boldsymbol\Lambda\boldsymbol\Phi\boldsymbol\Lambda^{\mathsf T}
 +\boldsymbol\Theta.
$$

What changes in confirmatory factor analysis is how the parameters are
restricted before estimation. A CFA might assign each indicator to one
factor, allow a few specified secondary loadings, constrain selected
loadings to equality, or specify a local residual covariance. The model
then has testable implications beyond the number of factors alone.

Here $\boldsymbol\vartheta$ denotes the vector of free parameters;
$\boldsymbol\Theta$ denotes the residual covariance matrix. They are
not two notations for the same object. For $p$ indicators and $m$ factors,
$\boldsymbol\Lambda$ is $p\times m$, $\boldsymbol\Phi$ is $m\times m$,
and $\boldsymbol\Theta$ is $p\times p$. As throughout Part Two,
$\boldsymbol\Phi=\operatorname{Var}(\boldsymbol\eta)$.

```{=latex}
\Needspace{14\baselineskip}
```

| Question | EFA in Chapter 2 | CFA in this chapter |
|:--|:--|:--|
| Which loadings are estimated? | All, before choosing an orientation | Only the declared free relations |
| How is orientation handled? | Rotation and subsequent labels | Loading restrictions and factor metrics |
| What determines the candidate? | Development evidence plus content | A frozen specification, with named rivals |
| What data are being used? | Sample A for development | Sample B for estimation of that candidate |
| What has not been established? | Unique substantive interpretation | Adequacy, validity, or causality |

“Confirmatory” describes restrictions and analytic history, not a
guaranteed favorable conclusion. A CFA estimated after repeated same-data
changes is still mathematically a CFA, but the changed specification is
data-developed. Conversely, a prespecified cross-loading does not make a
model nonconfirmatory merely because it departs from independent-clusters
simple structure.

## Missing arrows have statistical consequences

An absent loading arrow generally means a coefficient fixed to zero.
It does not mean that the indicator and factor are uncorrelated. Under
correlated factors, an indicator can correlate with a factor on which
it has no direct loading. Chapter 2's pattern--structure distinction
therefore remains essential.

An absent residual covariance means that two residuals are uncorrelated
under the model. That is a moment restriction. To assert conditional
independence of indicators given the factors, we additionally need an
appropriate conditional distribution or factorization; in a jointly
normal linear model, diagonal residual covariance supplies that stronger
property. Zero covariance alone does not generally imply independence.

A factor-to-indicator arrow specifies a loading relation. Interpreting
the arrow causally requires design and substantive assumptions beyond
the covariance equation. Good fit cannot supply temporal precedence or
exclude every alternative explanation.

```{r}
#| label: fig-cfa-sequence
#| fig-width: 7
#| fig-height: 1.6
#| fig-cap: "Specification, recoverability, and numerical estimation are distinct tasks; evaluating the fitted record is a fourth."
#| fig-alt: "Four linked boxes show restrictions, identification, estimation, and evaluation."
pt2_course_map(c("Restrictions","Identification","Estimation","Evaluation"))
```

# Write one model four ways

## A verbal specification

Begin with six continuous indicators, $Y_1,\ldots,Y_6$, and two
correlated factors. The first three indicators load only on $\eta_1$;
the last three load only on $\eta_2$. All six residual variances are
free and all residual covariances are zero. Factors and residuals are
uncorrelated. Use $Y_1$ and $Y_4$ as markers, fixing their loadings to
one. The factor variances and their covariance remain free.

The labels are intentionally neutral. This compact model is adapted from
the donor's two-factor/six-indicator teaching structure, not extracted
from the SLE items. It lets us follow each restriction without introducing
a second substantive measurement story.

## Six scalar equations

With zero latent intercepts and zero-mean residuals,

$$
\begin{aligned}
Y_{i1}&=\nu_1+\eta_{i1}+\epsilon_{i1},&
Y_{i4}&=\nu_4+\eta_{i2}+\epsilon_{i4},\\
Y_{i2}&=\nu_2+\lambda_{21}\eta_{i1}+\epsilon_{i2},&
Y_{i5}&=\nu_5+\lambda_{52}\eta_{i2}+\epsilon_{i5},\\
Y_{i3}&=\nu_3+\lambda_{31}\eta_{i1}+\epsilon_{i3},&
Y_{i6}&=\nu_6+\lambda_{62}\eta_{i2}+\epsilon_{i6}.
\end{aligned}
$$

The subscript $i$ denotes a person; the first subscript of
$\lambda_{ja}$ identifies the indicator, and the second identifies
the factor. Writing all six equations is useful once: it prevents a
compact diagram from hiding a cross-loading or a mistaken marker.
Later, matrix notation avoids repeating the same structure.

## Three parameter matrices

The same model is

$$
\boldsymbol\Lambda=
\begin{bmatrix}
1&0\\\lambda_{21}&0\\\lambda_{31}&0\\
0&1\\0&\lambda_{52}\\0&\lambda_{62}
\end{bmatrix},\qquad
\boldsymbol\Phi=\begin{bmatrix}\phi_{11}&\phi_{12}\\
\phi_{12}&\phi_{22}\end{bmatrix},\qquad
\boldsymbol\Theta=\operatorname{diag}(\theta_{11},\ldots,\theta_{66}).
$$

There are twelve loading slots, not twelve freely estimated loadings:
four are free, two are fixed at one for scaling, and six are fixed at
zero. Symmetry means $\phi_{12}$ and $\phi_{21}$ are the same covariance,
not two independent parameters. Likewise, a freely estimated residual
covariance would occupy two symmetric matrix entries but add only one
free coordinate.

The implied mean vector is
$\boldsymbol\mu=\boldsymbol\nu+\boldsymbol\Lambda\boldsymbol\alpha$.
Here $\boldsymbol\alpha=\mathbf0$, so
$\boldsymbol\mu=\boldsymbol\nu$. The latent intercept equals the
latent mean because this CFA has no structural paths or latent
predictors. This convention will need an explicit distinction between
intercepts and marginal means once structural regressions enter.

## A diagram and executable syntax

```{r}
#| label: fig-cfa-six
#| fig-width: 7
#| fig-height: 4.5
#| fig-cap: "Marker-scaled six-indicator CFA. Residual variances are free; residual covariances and unshown loadings are zero. Intercepts are omitted from this covariance diagram."
#| fig-alt: "F1 predicts Y1, Y2, and Y3; F2 predicts Y4, Y5, and Y6. The Y1 and Y4 loadings equal one. Factor variances and their covariance are free."
pt2_groups_diagram(
  list(F1=paste0("Y",1:3),F2=paste0("Y",4:6)),marker=TRUE,
  factor_cov_labels="phi[12]")
```

In R, the `lavaan` package represents factor loadings with `=~` and
variances or covariances with `~~`. The following is the marker-scaled
specification:

```r
model6 <- '
  F1 =~ 1*Y1 + Y2 + Y3
  F2 =~ 1*Y4 + Y5 + Y6
  F1 ~~ F2
'
```

The `cfa()` convenience function supplies free indicator residual
variances and fixes unmentioned residual covariances to zero in this
model. Defaults must still be audited; a short syntax string is not a
complete parameter ledger. The [official CFA tutorial](https://lavaan.ugent.be/tutorial/cfa.html)
is a useful orientation to the specify--fit--extract workflow.

| Representation | What to check |
|:--|:--|
| Words | Indicator assignments, residual assumptions, metric, and means |
| Scalar equations | Every indicator has exactly the stated loading terms |
| Matrices | All zero, fixed, and shared entries are explicit |
| Diagram and syntax | Every drawn or coded relation matches the same ledger |

If two representations disagree, no amount of numerical convergence can
resolve which model the researcher intended. Fix the specification before
interpreting output.

## Transfer the grammar to four factors

A larger diagram should be read with the same ledger rather than as a new
kind of model. As a published-reading exercise, reconstruct the topology of
Figure 3 in @chou2021: four correlated factors---self-efficacy, school
support, privacy concerns, and technostress---each measured by four
indicators. The reconstruction below records the displayed assignments and
factor relations; it does not reproduce the paper's estimated coefficients
or imply that every software default used in the original analysis is known.

```{r}
#| label: fig-cfa-four
#| fig-width: 7
#| fig-height: 5.4
#| fig-cap: "Four-factor, sixteen-indicator CFA reconstructed as a diagram-reading transfer. All six factor correlations are free; fixed-factor scaling is used for the count."
#| fig-alt: "Four correlated factors called Self-efficacy, School support, Privacy concerns, and Technostress each predict four separate indicators. Each indicator has a residual."
pt2_groups_diagram(
  list(`Self-efficacy`=paste0("E",1:4),`School support`=paste0("S",1:4),
    `Privacy concerns`=paste0("P",1:4),Technostress=paste0("T",1:4)),
  correlated=TRUE,marker=FALSE)
```

Under fixed-factor scaling, a direct `lavaan` translation is

```r
model4 <- '
  SelfEfficacy =~ E1 + E2 + E3 + E4
  SchoolSupport =~ S1 + S2 + S3 + S4
  Privacy =~ P1 + P2 + P3 + P4
  Technostress =~ T1 + T2 + T3 + T4
'
fit4 <- lavaan::cfa(model4, data = dat, std.lv = TRUE)
```

The omitted loading arrows are fixed zero, the four latent variances set the
factor metrics, and the six pairwise factor covariances are free. With 16
continuous indicators, the covariance ledger is therefore

$$
u=\frac{16(17)}2=136,\qquad
t=16\text{ loadings}+16\text{ residual variances}
  +\frac{4(3)}2\text{ factor correlations}=38,
\qquad df=98.
$$

This transfer matters because visual density can conceal a counting error.
Four double-headed arrows would omit two factor correlations; counting 16
drawn loading slots as 16 *free* loadings would be wrong under marker
scaling but correct under the fixed-factor convention used here. Always
state the metric before reading freedom from a diagram.

# Free, fixed, and constrained parameters

## Different restrictions do different jobs

A free parameter is estimated from the data under the other restrictions.
A fixed parameter is assigned a known constant. An equality constraint
requires two or more parameter entries to share a value; it does not
necessarily assign that value in advance. A bound restricts the allowable
range. A defined parameter is a computed function of fitted parameters
and does not add a new free coordinate.

```{=latex}
\Needspace{18\baselineskip}
```

| Entry | Status | Statistical role | Reason to state |
|:--|:--|:--|:--|
| $\lambda_{11}=1$ | Fixed | Factor metric | Marker choice |
| $\lambda_{21}$ | Free | Target loading | Indicator assignment |
| $\lambda_{22}=0$ | Fixed | Excluded secondary loading | Content/design restriction |
| $\lambda_{21}=\lambda_{31}$ | Equality | Shared loading value | Equality on a stated scale |
| $\theta_{23}=0$ | Fixed | Residual relation excluded | Local measurement assumption |
| $\phi_{12}$ | Free | Factor association | Correlation permitted |
| $\theta_{jj}>0$ | Admissibility condition | Positive residual variance | Statistical parameter space |
| $\lambda_{21}-\lambda_{31}$ | Defined | Loading difference | Derived estimand, not extra freedom |

Fixing one loading to one selects a unit. Fixing a secondary loading to
zero excludes a relation. They are both constants in software, but only
the latter ordinarily adds a substantive covariance restriction when
the former replaces another valid scale normalization.

“Unconstrained” is also relative. A free loading is estimated within a
model whose other loadings, residual structure, and factor variances
may be strongly restricted. Its estimate is not model-free evidence
about that indicator.

## Change one loading restriction

Allow $Y_3$ to load on factor 2 as well:

$$
Y_{i3}=\nu_3+\lambda_{31}\eta_{i1}
+\lambda_{32}\eta_{i2}+\epsilon_{i3}.
$$

One formerly zero matrix entry becomes free. The new coefficient affects
the variance of $Y_3$ and its covariances with all indicators connected
to either factor. It does not merely repair the covariance of one chosen
pair. The coefficient's interpretation is conditional on factor 1, just
as an oblique EFA pattern coefficient was conditional on the other factors.

```{r}
#| label: fig-cfa-cross
#| fig-width: 7
#| fig-height: 4.3
#| fig-cap: "A single additional loading changes several implied moments. The dashed path is lambda32."
#| fig-alt: "The six-indicator model adds one dashed arrow from F2 to Y3; all other relations match the marker-scaled base model."
pt2_groups_diagram(
  list(F1=paste0("Y",1:3),F2=paste0("Y",4:6)),marker=TRUE,
  cross=data.frame(factor="F2",item="Y3",stringsAsFactors=FALSE),
  factor_cov_labels="phi[12]")
```

## Change one residual relation instead

Alternatively, retain the original loading matrix but free
$\theta_{23}=\operatorname{Cov}(\epsilon_2,\epsilon_3)$.
Then only the $(2,3)$ and $(3,2)$ entries of the residual matrix change.
This can represent item-specific shared wording, a common task, or
another source of association outside the two modeled factors.
Its meaning must be justified; “the software suggested it” is not a
measurement explanation.

```{r}
#| label: fig-cfa-residual
#| fig-width: 7
#| fig-height: 4.3
#| fig-cap: "A residual covariance is a different model change from a cross-loading. The dotted relation connects e2 and e3."
#| fig-alt: "The base six-indicator model adds a covariance between residuals e2 and e3 without adding a factor-to-indicator loading."
pt2_groups_diagram(
  list(F1=paste0("Y",1:3),F2=paste0("Y",4:6)),marker=TRUE,
  residual=data.frame(item1="Y2",item2="Y3",stringsAsFactors=FALSE),
  factor_cov_labels="phi[12]")
```

Both changes add one parameter, but they generally produce different
sets of implied covariance matrices. Equal df therefore does not imply
equivalent models. When two modifications improve the same observed
pair, their consequences for the other pairs help distinguish them.
Local diagnosis depends on that difference.

## Equality requires a common metric

The restriction $\lambda_{21}=\lambda_{31}$ states equal raw changes
in $Y_2$ and $Y_3$ for a one-unit change in their common factor. If
$Y_3$ is rescaled from points to tens of points, the equivalent restriction
must be transformed. Repeating the old equality on rescaled data tests
a different hypothesis.

Equality of loadings is not equality of residual variances, intercepts,
or item distributions. Equal loadings with unequal residual variances
allow different precision. Stronger parallel-measurement conditions
require additional restrictions. A short-form scale cannot be justified
simply by retaining the largest estimated loadings; content coverage
and intended score use also change when items are removed.

# Give latent variables a metric

## Marker scaling and fixed-factor scaling

Latent variables have no observed unit until we assign one. Under marker
scaling, $\lambda_{11}=1$ means that a one-unit change in $\eta_1$
contributes one unit to $Y_1$ through its loading relation. It does not
set the factor variance to one or make the marker error-free.

Under fixed-factor scaling, set $\phi_{11}=\phi_{22}=1$ and estimate
all six target loadings. A factor unit is then one model-standard-deviation
unit in this single-group CFA. The indicators retain their raw units.
In `lavaan`, `std.lv=TRUE` requests this scale choice. It is not a command
to standardize every observed column.

| Quantity | Marker scaling | Fixed-factor scaling |
|:--|:--|:--|
| Marker loadings | Fixed at one | Free |
| Factor variances | Free | Fixed at one |
| Free target loadings | 4 | 6 |
| Free factor variance/covariance entries | 3 | 1 |
| Free residual variances | 6 | 6 |
| Total free covariance coordinates | 13 | 13 |

One must also choose factor signs. With unit factor variances, changing
the sign of a factor and every loading on it leaves the covariance
unchanged. A declared positive orientation for central indicators makes
reported signs interpretable; a negative factor does not represent a
different fit merely because its column points the other way.

## Derive the reparameterization

For an invertible diagonal $m\times m$ matrix $\mathbf D$, define

$$
\boldsymbol\eta_i^*=\mathbf D\boldsymbol\eta_i,
\qquad
\boldsymbol\Lambda^*=\boldsymbol\Lambda\mathbf D^{-1},
\qquad
\boldsymbol\Phi^*=\mathbf D\boldsymbol\Phi\mathbf D^{\mathsf T}.
$$

Then

$$
\boldsymbol\Lambda^*\boldsymbol\Phi^*
(\boldsymbol\Lambda^*)^{\mathsf T}
=\boldsymbol\Lambda\boldsymbol\Phi\boldsymbol\Lambda^{\mathsf T}.
$$

For positive factor variances, choosing
$\mathbf D=\operatorname{diag}(\phi_{11}^{-1/2},\phi_{22}^{-1/2})$
makes the transformed factor variances one. The residual matrix is
unchanged. If latent means are modeled, transform
$\boldsymbol\alpha^*=\mathbf D\boldsymbol\alpha$ as well.

Use the following controlled values for hand calculation:

$$
\boldsymbol\Lambda=
\begin{bmatrix}1&0\\.8&0\\1.2&0\\0&1\\0&1.1\\0&.9\end{bmatrix},
\quad
\boldsymbol\Phi=\begin{bmatrix}4&1.2\\1.2&9\end{bmatrix},
\quad
\boldsymbol\Theta=\operatorname{diag}(2,3,4,3,4,5).
$$

The transformation divides factor 1 by 2 and factor 2 by 3. Accordingly,
the transformed loading columns multiply by 2 and 3. The covariance
1.2 becomes a correlation $1.2/(2\times3)=.20$. The transformed
primary loading of $Y_2$ is 1.6, but its implied variance has not changed.

```{r}
#| label: tbl-cfa-scale-values
#| tbl-cap: "Controlled values under two metrics; these are stipulated parameters, not sample estimates."
pt2_table(data.frame(Item=rownames(L),Marker_loading=c(L[1:3,1],L[4:6,2]),
  Fixed_factor_loading=c(h$scaled$Lambda[1:3,1],h$scaled$Lambda[4:6,2]),
  Residual_variance=diag(Theta),Implied_variance=diag(pop$Sigma)))
```

This is a reparameterization of the same covariance model, not evidence
that marker and fixed-factor methods estimate different constructs.
Equivalence can fail if a constraint is not transformed with the metric,
if a required marker loading is zero, or if a factor variance is on an
inadmissible boundary. Multigroup means and nonlinear models need
additional care. A convenient single-group result should not be
generalized without checking those conditions.

## Changing observed units is a different operation

Latent rescaling leaves the observed variables untouched. Changing an
indicator from raw points to tens of points changes the numerical record
itself. Let

$$
\mathbf y_i^*=\mathbf b+\boldsymbol\Delta\mathbf y_i,
$$

where $\boldsymbol\Delta$ is a nonsingular diagonal matrix of observed-unit
conversion factors. The matching model is

$$
\begin{aligned}
\boldsymbol\nu^*&=\mathbf b+\boldsymbol\Delta\boldsymbol\nu,
&\boldsymbol\Lambda^*&=\boldsymbol\Delta\boldsymbol\Lambda,\\
\boldsymbol\Theta^*&=\boldsymbol\Delta\boldsymbol\Theta
\boldsymbol\Delta^{\mathsf T},
&\boldsymbol\Sigma^*&=\boldsymbol\Delta\boldsymbol\Sigma
\boldsymbol\Delta^{\mathsf T}.
\end{aligned}
$$ {#eq-cfa-observed-rescaling}

Thus multiplying one indicator by 10 multiplies its intercept and loading by
10, its variance and residual variance by 100, and its covariance with an
unchanged indicator by 10. A fixed shift changes its intercept and mean but
not its covariance. Nothing is unidentified in this operation: the same
observed information is expressed in different units, and every affected
model quantity must change coherently.

Let $\mathbf D_y=\operatorname{diag}(\boldsymbol\Sigma)$ contain the positive
observed variances. The corresponding correlation model is

$$
\begin{aligned}
\mathbf R
&=\mathbf D_y^{-1/2}\boldsymbol\Sigma\mathbf D_y^{-1/2}\\
&=\widetilde{\boldsymbol\Lambda}\boldsymbol\Phi
\widetilde{\boldsymbol\Lambda}^{\mathsf T}
+\widetilde{\boldsymbol\Theta},
\end{aligned}
$$ {#eq-cfa-correlation-reparameterization}

where $\widetilde{\boldsymbol\Lambda}=\mathbf D_y^{-1/2}
\boldsymbol\Lambda$ and $\widetilde{\boldsymbol\Theta}=
\mathbf D_y^{-1/2}\boldsymbol\Theta\mathbf D_y^{-1/2}$. This standardizes
the indicators, not the factors. With continuous ML and a consistently
transformed model, observed-unit rescaling does not change fit. Repeating an
untransformed numerical equality constraint after changing units, however,
tests a different model.

## Three kinds of coefficient output

For a loading $\lambda_{ja}$, latent-standardized and fully standardized
versions are, respectively,

$$
\lambda_{ja}^{(\mathrm{lv})}
=\lambda_{ja}\sqrt{\phi_{aa}},
\qquad
\lambda_{ja}^{(\mathrm{all})}
=\lambda_{ja}\frac{\sqrt{\phi_{aa}}}{\sqrt{\sigma_{jj}}}.
$$

The first expresses a factor change in standard-deviation units while
retaining the indicator's raw unit. The second standardizes both. In
our fixed-factor CFA, raw and latent-standardized loadings coincide,
but fully standardized loadings generally differ.

Standardizing output after fitting does not impose equal indicator
variances during estimation. Moreover, the uncertainty of a standardized
coefficient depends on the estimated quantities used in the
transformation. Dividing a raw standard error by a displayed rounded
standard deviation is not generally the correct standard error for the
fully standardized coefficient.

A loading greater than one in raw units is unremarkable. Even a
standardized pattern coefficient needs to be interpreted in its
multivariate context; admissibility concerns the full covariance
decomposition. For a single-loading item with uncorrelated residuals,
the squared fully standardized loading is its communality. For a
cross-loading item, use the full quadratic form, not a single square.

# Trace restrictions into implied moments

## Within-factor and between-factor covariance

From the controlled marker-scaled model,

$$
\begin{aligned}
\operatorname{Var}(Y_2)&=\lambda_{21}^2\phi_{11}+\theta_{22},\\
\operatorname{Cov}(Y_2,Y_3)&=\lambda_{21}\lambda_{31}\phi_{11},\\
\operatorname{Cov}(Y_2,Y_5)&=\lambda_{21}\lambda_{52}\phi_{12},\\
\operatorname{Cov}(Y_5,Y_6)&=\lambda_{52}\lambda_{62}\phi_{22}.
\end{aligned}
$$

The numerical values are $5.56$, $3.84$, $1.056$, and $8.91$.
Each expression has a reason. $Y_2$'s variance contains common and
residual variance. $Y_2$ and $Y_3$ share factor 1. $Y_2$ and $Y_5$
are connected through the factor covariance. There is no residual
cross-term because the base residual matrix is diagonal.

The diagonal and off-diagonal equations do different work during
estimation. The covariances inform relations among loadings and factor
covariances; the variances also constrain residual variances. An
implausibly large common part may force a residual variance toward zero
or below it. A Heywood case is therefore a symptom of the whole
decomposition, not a harmless cosmetic warning.

```{r}
#| label: tbl-cfa-six-sigma
#| tbl-cap: "Full implied covariance matrix from the controlled six-indicator model."
pt2_table(data.frame(Indicator=rownames(pop$Sigma),pop$Sigma,check.names=FALSE))
```

```r
Sigma <- Lambda %*% Phi %*% t(Lambda) + Theta
Sigma[2, 3]
Sigma[2, 5]
cov2cor(Sigma)
```

The final line changes covariance units to correlations; it does not
re-estimate the model. In a fitted model, compare the implied covariance
to the sample covariance using matching divisors and variable order.

## What the cross-loading adds

Now add $\lambda_{32}$ to the third indicator's row. Its variance is

$$
\sigma_{33}=\lambda_{31}^2\phi_{11}
+\lambda_{32}^2\phi_{22}
+2\lambda_{31}\lambda_{32}\phi_{12}+\theta_{33}.
$$

The covariances with $Y_2$ and $Y_5$ become

$$
\begin{aligned}
\sigma_{23}&=\lambda_{21}
(\lambda_{31}\phi_{11}+\lambda_{32}\phi_{12}),\\
\sigma_{35}&=\lambda_{52}
(\lambda_{31}\phi_{12}+\lambda_{32}\phi_{22}).
\end{aligned}
$$

If $\lambda_{32}=.25$ while the other controlled parameters stay fixed,
$\sigma_{23}$ increases by $.8\times.25\times1.2=.24$,
whereas $\sigma_{35}$ increases by $1.1\times.25\times9=2.475$.
The unequal changes follow from the pathways and factor variances,
not a software accident.

The independently generated continuous donor microcase used for
estimation has precisely this additional .25 population loading,
$n=500$, and a separately recorded seed. Its file is
`chapter_03_donor_continuous.csv` beside this chapter.
It is not an archived ordinal data set and is not Sample B. The base
six-indicator fit is useful for illustrating parameterization even
when its adequacy remains a separate question.

## What a residual covariance adds

Instead, in the base model, adding $\theta_{23}=.50$ changes
$\sigma_{23}$ from 3.84 to 4.34 and leaves all other unique covariance
entries unchanged when the remaining parameters are held fixed.
This contrasts sharply with the cross-loading's multi-pair changes.

After refitting, other estimates will generally change too. The
controlled comparison holds them fixed only to expose the direct
algebraic consequences of the specification. Do not confuse that
hand calculation with the total change after an optimizer re-estimates
every free parameter.

## Means are a separate set of moments

In covariance-only analysis there are $p(p+1)/2$ observed moments.
Adding a mean structure adds $p$ more:

$$
u_{\mathrm{cov}}=\frac{p(p+1)}2,
\qquad
u_{\mathrm{mean+cov}}=\frac{p(p+1)}2+p.
$$

If all indicator intercepts are free and latent intercepts are fixed
to zero, the extra $p$ mean moments are matched by $p$ extra parameters.
The covariance restrictions and df are unchanged. This is different
from testing structured mean differences or constraining intercepts
across groups, topics that will become important later.

Freeing every indicator intercept and every latent intercept at once
does not identify a latent origin. For any vector $\mathbf c$,

$$
\boldsymbol\alpha^*=\boldsymbol\alpha+\mathbf c,
\qquad
\boldsymbol\nu^*=\boldsymbol\nu-\boldsymbol\Lambda\mathbf c
$$

produce the same observed means. Location restrictions are needed in
addition to scale restrictions. Fixing factor variances to one does not
remove this location ambiguity.

# Can the parameters be recovered?

## Identification is about the parameter-to-moment mapping

After accounting for scale and sign conventions, global identification
requires that

$$
\boldsymbol\Sigma(\boldsymbol\vartheta_1)
=\boldsymbol\Sigma(\boldsymbol\vartheta_2)
\quad\Longrightarrow\quad
\boldsymbol\vartheta_1=\boldsymbol\vartheta_2
$$

for admissible parameter vectors in the declared model. Local
identification requires uniqueness in a neighborhood. Generic
identification permits exceptional parameter sets, such as special
zero loadings, where recoverability fails. For models with means,
replace the covariance mapping by the joint mean-and-covariance mapping.

Identification is not the existence of an exact solution to every
sample covariance matrix. An identified overrestricted model usually
cannot reproduce arbitrary sample moments exactly; estimation finds
the closest admissible representation under a stated discrepancy.
Conversely, multiple parameter vectors can reproduce the same moments
even if an optimizer prints only one of them.

## Count moments and free coordinates

Counting begins by choosing one coherent moment representation. For $p$
continuous indicators, a covariance matrix supplies $p(p+1)/2$ distinct
variances and covariances. A correlation matrix supplies only $p(p-1)/2$
freely varying off-diagonal entries because its $p$ diagonal entries equal
one. Adding an unrestricted observed mean vector contributes another $p$
moments.

For six indicators this gives

| Moment representation | Distinct information |
|:--|--:|
| Covariances, including variances | $6(7)/2=21$ |
| Correlations, with unit diagonals fixed | $6(5)/2=15$ |
| Covariances plus observed means | $21+6=27$ |

In covariance form, the fixed-factor two-factor simple-structure model has
six loadings, one factor correlation, and six residual variances: $t=13$ and
$df=21-13=8$. In correlation form, the six unit-diagonal equations determine
the six residual variances, leaving seven free coordinates and the same
$df=15-7=8$. Subtracting all 13 covariance-model parameters from 15
correlations would count standardization twice.

For the marker-scaled six-indicator base model,

$$
u=\frac{6(7)}2=21,\qquad
t=4+3+6=13,\qquad df=21-13=8.
$$

Adding one cross-loading yields $t=14$, $df=7$. Adding one residual
covariance instead gives the same count but a different model. Adding
both gives $t=15$, $df=6$. A single independent equality constraint
on two otherwise free loadings reduces the free-coordinate count by
one. Merely giving a free parameter a label does not change the count.

| Six-indicator specification | Loading coordinates | Factor covariance coordinates | Residual coordinates | Candidate df |
|:--|--:|--:|--:|--:|
| Marker-scaled base | 4 | 3 | 6 | 8 |
| Fixed-factor base | 6 | 1 | 6 | 8 |
| Base plus one cross-loading | 5 | 3 | 6 | 7 |
| Base plus one residual covariance | 4 | 3 | 7 | 7 |
| Base with one loading equality | 3 | 3 | 6 | 9 |

The count $df=u-t$ is necessary bookkeeping, not sufficient proof.
The term “observed moments” is preferable to “number of observations”
here: 21 unique covariance entries are not 21 people. Increasing the
sample size improves precision under suitable conditions but does not
create new types of covariance moments or repair structural
nonidentification.

The same bookkeeping clarifies why a displayed unrestricted EFA loading
matrix contains more entries than the model has distinguishable coordinates.
With six indicators and two unit-variance orthogonal factors, there are 12
raw loading entries and six residual variances. A two-factor orthogonal
rotation has $2(2-1)/2=1$ continuous degree of freedom and changes the loading
display without changing its reproduced covariance. The generic dimension is

$$
t_{\mathrm{EFA}}=12+6-\frac{2(2-1)}2=17,
\qquad df=21-17=4.
$$ {#eq-cfa-efa-dimension}

This subtraction counts distinct covariance-model coordinates at regular,
full-rank parameter points. It does not declare any substantively preferred
loading to be known, and it does not replace an identification analysis.

## A recoverable block and a counterexample

For a single marker-scaled factor with three indicators and diagonal
residuals,

$$
\sigma_{12}=\lambda_2\phi,
\quad \sigma_{13}=\lambda_3\phi,
\quad \sigma_{23}=\lambda_2\lambda_3\phi.
$$

When the relevant covariances are nonzero, these imply

$$
\phi=\frac{\sigma_{12}\sigma_{13}}{\sigma_{23}},
\quad \lambda_2=\frac{\sigma_{23}}{\sigma_{13}},
\quad \lambda_3=\frac{\sigma_{23}}{\sigma_{12}}.
$$

The diagonal equations then determine residual variances. These
expressions explain generic recoverability for this simple block,
subject to admissibility. If a denominator vanishes, the reasoning
must be revisited; a generic result does not cover every exceptional
configuration.

```{r}
#| label: fig-cfa-three-indicator
#| fig-width: 5.6
#| fig-height: 3.7
#| fig-cap: "A marker-scaled three-indicator block. The three off-diagonal covariances can recover the factor variance and two non-marker loadings at regular, nonzero parameter points."
#| fig-alt: "One factor predicts three indicators, the first loading is fixed to one, and each indicator has a residual."
pt2_groups_diagram(list(F=paste0("Y",1:3)),correlated=FALSE,marker=TRUE)
```

Now consider two *orthogonal*, isolated two-indicator factors, with
unit factor variances. Each block has two free loadings and two residual
variances but only three covariance moments. Across both blocks there
are ten available covariance entries and eight free parameters, giving
a superficially favorable $df=2$. Yet the cross-block covariances are
all fixed to zero and provide no information to separate the two
loadings within either block. For one block,

$$
\sigma_{12}=\lambda_1\lambda_2,
\quad \theta_{11}=\sigma_{11}-\lambda_1^2,
\quad \theta_{22}=\sigma_{22}-\lambda_2^2.
$$

A range of choices for $\lambda_1$ can yield the same moments by
adjusting $\lambda_2$ and both residual variances, while retaining
positive variances. The whole-model count did not reveal the local
deficiency. Additional indicators or justified restrictions can help;
adding an arbitrary constraint solely to make software run changes the
measurement model and needs a rationale.

The diagram pair makes the missing information visible. In the isolated
model, all four cross-block covariances are fixed to zero, so the positive
whole-model df do not help identify either two-indicator block. Allowing a
nonzero factor covariance connects the blocks: cross-block moments then
contain products of loadings from both factors. The connected model is
generically identified under regular nonzero loadings and
$\phi_{12}\ne0$; the argument degenerates when that connection is zero.

```{r}
#| label: fig-cfa-identification-pair
#| layout-ncol: 2
#| fig-subcap:
#|   - "Isolated orthogonal blocks"
#|   - "Connected correlated blocks"
#| fig-cap: "Two-indicator blocks can have the same whole-model count but different recoverability because the covariance edge changes which moments carry loading information."
#| fig-alt: "The left panel shows two separate two-indicator factors. The right panel adds a covariance between the two factors."
op <- par(mfrow=c(1,2)); on.exit(par(op))
pt2_groups_diagram(list(F1=c("Y1","Y2"),F2=c("Y3","Y4")),correlated=FALSE)
title("Isolated blocks",line=-1.4,cex.main=.95)
pt2_groups_diagram(list(F1=c("Y1","Y2"),F2=c("Y3","Y4")),correlated=TRUE,
  factor_cov_labels="phi[12]")
title("Connected by a factor covariance",line=-1.4,cex.main=.95)
```

## Structural, weak, and numerical problems

Structural nonidentification is built into the parameter mapping.
Weak identification describes a formally recoverable model whose
parameters are poorly distinguished in a particular configuration.
Small loadings, nearly proportional loading columns, or factor
correlations close to one can make very different parameter vectors
produce nearly indistinguishable moments. Numerical difficulty is
another issue: unsuitable starts or badly scaled variables can impede
an optimizer even when a well-behaved optimum exists.

```{=latex}
\Needspace{18\baselineskip}
```

| Audit | Warning | What it warrants |
|:--|:--|:--|
| Moment count | More free coordinates than moments | Reconsider specification |
| Metric and sign | No normalization or inconsistent orientation | Specify units and signs |
| Local connections | Isolated insufficient indicator block | Examine block equations |
| Loading rank | Redundant factor patterns | Examine distinct common directions |
| Factor covariance | Near singularity | Investigate weak separation |
| Residual variances | Zero or negative estimates | Inspect admissibility and misspecification |
| Information/SEs | Singular matrix or unstable large SEs | Check recoverability and numerical behavior |
| Starts and gradients | Different optima or nonzero gradient | Resolve optimization before interpretation |

A numerical derivative matrix provides a local diagnostic. If
$\mathbf s(\boldsymbol\vartheta)$ stacks the distinct implied moments,
its Jacobian is
$\boldsymbol\Delta=\partial\mathbf s/\partial\boldsymbol\vartheta^{\mathsf T}$.
Full column rank at a regular point supports local identification;
it does not prove global uniqueness or uniform stability throughout
the parameter space. Students need not derive this matrix to understand
what the diagnostic adds beyond counting.

The SLE primary fit has numerical Jacobian rank
`r h$audit$delta_rank` for `r h$audit$delta_columns` free coordinates.
That is a useful local check of the actual fitted candidate. The
three blocks also have several pure primary indicators, so the two
retained cross-loadings do not leave every factor's orientation open
as in unrestricted EFA.

## Nested, nonnested, and equivalent models

A restricted model is nested in another if every covariance matrix it
can generate belongs to the larger model's covariance set. Show the
restrictions that produce that inclusion; a parameter-count comparison
alone is not enough. The base six-indicator model is nested in its
cross-loading extension by setting $\lambda_{32}=0$.

Two models can have the same df and different covariance sets, as with
an alternative indicator assignment. They can also have different-looking
parameters but identical covariance sets, as with the regular marker
and fixed-factor reparameterizations. Observational equivalence limits
what covariance data can distinguish. It does not establish that
higher-order, bifactor, and correlated-residual models are generally
equivalent; each such claim requires an exact mapping and conditions.

One deliberately constrained example shows what an exact mapping looks like.
For two positively correlated first-order factors, fix both second-order
loadings to one, set $\operatorname{Var}(G)=\phi_{12}$, and set the two
first-order disturbance variances to $\phi_{11}-\phi_{12}$ and
$\phi_{22}-\phi_{12}$. When these variances are nonnegative, the second-order
parameterization reproduces exactly the same $2\times2$ factor covariance
matrix. Without the two extra scale/equality restrictions, two first-order
factors do not by themselves identify a second-order factor. The
substantively more useful cases have three or more first-order factors.

```{r}
#| label: fig-cfa-equivalent-pair
#| fig-width: 7
#| fig-height: 4.2
#| fig-cap: "A correlated two-factor model and one constrained second-order reparameterization. The equivalence requires the stated fixed loadings and admissible disturbance variances."
#| fig-alt: "The left model connects two first-order factors with a covariance. The right model replaces that covariance with one general factor loading one-to-one on each first-order factor."
pt2_equivalence_pair()
```

These distinctions prepare comparison. They do not yet say which
candidate is adequate or which restriction the data oppose.

# What estimation does

## Minimize a stated discrepancy

Let $\mathcal P$ be the admissible parameter space and $\mathbf S$
the sample covariance under a declared divisor. Covariance estimation
can be written

$$
\widehat{\boldsymbol\vartheta}
=\arg\min_{\boldsymbol\vartheta\in\mathcal P}
F\{\mathbf S,\boldsymbol\Sigma(\boldsymbol\vartheta)\}.
$$

For normal-theory ML, a conventional covariance discrepancy is

$$
F_{\mathrm{ML}}
=\log|\boldsymbol\Sigma(\boldsymbol\vartheta)|
+\operatorname{tr}\{\mathbf S\boldsymbol\Sigma(\boldsymbol\vartheta)^{-1}\}
-\log|\mathbf S|-p.
$$

The determinant summarizes multivariate dispersion; the trace term
compares the sample covariance to the model's inverse-covariance
weighting. The last two terms depend only on the input matrix and
dimension. They normalize the discrepancy to zero when
$\boldsymbol\Sigma=\mathbf S$, because
$\operatorname{tr}(\mathbf S\mathbf S^{-1})=p$.
The dimension term is the number of observed variables, not the
number of indicators plus factors.

At a proper optimum, numerical changes in the free coordinates cannot
materially lower the objective under the specified constraints.
That is an optimization statement. It does not imply small absolute
discrepancy, plausible parameter meaning, or a unique substantive model.

```{r}
#| label: fig-cfa-estimation
#| fig-width: 6.6
#| fig-height: 3.2
#| fig-cap: "Conceptual estimation geometry. The curve represents a restricted set of moments, not the literal dimension or shape of this CFA."
#| fig-alt: "Observed moments lie away from a curve of model-implied moments; a dashed arrow points to a selected model point, with a warning that ML is not Euclidean distance."
pt3_estimation_plot()
```

## Normality belongs to the likelihood contract

The factor covariance equation follows from linearity, finite second
moments, and the stated zero covariances. It does not require normality.
Conventional normal-theory ML standard errors and tests additionally
use sampling and distributional regularity. Independently distributed
normal factors and normal residuals give a jointly normal observed
vector in this linear model. Marginally normal histograms alone do not
establish multivariate normality or independent sampling.

Sample B consists of 360 complete, independently simulated continuous
records. There is no silent missing-case removal, imputation, weighting,
clustering adjustment, or ordinal recoding. The raw indicators retain
the frozen process-index units. The fitted settings are explicit:

```{=latex}
\Needspace{18\baselineskip}
```

```r
dat <- sle_simulate("confirmation")$x
items <- c(paste0("P", 1:4), paste0("S", 1:4), paste0("B", 1:4))
model <- '
  Planning =~ p1*P1 + p2*P2 + P3 + P4
  Strategy =~ S1 + S2 + S3 + S4 + P4
  Belonging =~ B1 + B2 + B3 + B4 + S4
'
fit <- lavaan::cfa(model, data = dat[items],
                   estimator = "ML", likelihood = "normal",
                   std.lv = TRUE, meanstructure = FALSE,
                   missing = "listwise", se = "standard",
                   test = "standard", information = "expected")
```

The confirmation sample is simulated from the declared population and seed,
exactly as the exploratory chapter simulates its development sample; that
chapter's [data appendix](chapter_02.qmd#sec-sle-data) gives the codebook
and sample manifest. The explicit syntax above matches the
frozen candidate; no saved-model download is needed to run this example.
The canonical script separately checks the manifest's continuous-data declaration,
item names and order, sample role, completeness, size, and hashes before
the call. Numerically coded columns alone are not sufficient evidence
of a continuous measurement scale.

Categorical indicators are covered elsewhere. Categorical group labels
and dummy predictors remain available for later multigroup and MIMIC
models. Later SEM chapters address robust ML inference for nonnormality
and FIML for missing continuous data under suitable assumptions; neither
is a categorical-measurement switch.

## Record the covariance divisor

The normal-likelihood convention uses

$$
\mathbf S_N=\frac1N\sum_{i=1}^N
(\mathbf y_i-\bar{\mathbf y})(\mathbf y_i-\bar{\mathbf y})^{\mathsf T}
=\frac{N-1}{N}\mathbf S_{N-1}.
$$

R's `cov()` uses $N-1$; `lavaan` performs the normal-likelihood
rescaling internally. Do not pre-rescale raw observations to compensate.
When working from a covariance matrix rather than raw data, disclose
the supplied divisor and package rescaling options. The [estimator
documentation](https://lavaan.ugent.be/tutorial/est.html) distinguishes
the normal and Wishart conventions explicitly.

For the normal convention, the model statistic uses $N$ times the
covariance discrepancy. The Wishart convention uses an $N-1$ covariance
and multiplier. Chapter 2's corrected exploratory statistic is a
further distinct finite-sample convention. This chapter records the
normal family without interpreting its model test.

## A small mean-structure comparison

The frozen M1-cross has 78 covariance moments, 29 free covariance
coordinates, and 49 df. Adding twelve free indicator intercepts while
keeping latent intercepts zero gives 90 moments, 41 free coordinates,
and the same 49 df. For the stricter M0-simple model the corresponding
counts are 78/27/51 and 90/39/51.

```{r}
#| label: tbl-cfa-mean-counts
#| tbl-cap: "Counts verified against actual fitted parameter tables. The mean extension does not constrain the indicator means."
audits <- list(pt3_audit(list(fit=h$fit,warnings=h$warnings)),
  pt3_audit(h$means),pt3_audit(h$simple_means))
pt2_table(data.frame(Model=c("M1 covariance","M1 free intercepts","M0 free intercepts"),
  Moments=sapply(audits,function(a)a$moments),
  Free=sapply(audits,function(a)a$nfree),Df=sapply(audits,function(a)a$df)),digits=0)
```

The largest difference between the M1 covariance-only and free-intercept
implied covariance matrices is
`r format(max(abs(lavaan::fitted(h$means$fit)$cov-h$audit$Sigma)),scientific=TRUE,digits=2)`.
The estimated free intercepts reproduce the sample means. This verifies
the intended separation of covariance restrictions from an unrestricted
mean extension. It does not justify freely estimating latent origins
without additional constraints.

# Read an estimated model before reading its fit

## Convergence and admissibility first

The optimizer reports convergence for the inherited M1-cross model,
with no captured fitting warnings. The maximum absolute gradient is
`r format(h$audit$max_gradient,scientific=TRUE,digits=2)`.
All residual variances are positive, the factor covariance and implied
covariance matrices are positive definite, and standard errors are
available. The smallest information-matrix eigenvalue is positive,
although its magnitude depends on parameter scaling and is not a
universal quality index.

```{r}
#| label: tbl-cfa-audit
#| tbl-cap: "Selected estimation gates for the actual Sample B primary fit. Passing these gates does not establish adequate fit."
gate <- h$audit$checks[1:5]
pt2_table(data.frame(Check=gsub("_"," ",names(gate)),Passed=unname(gate)))
```

If convergence fails, first inspect the stated model, data, scale, and
starting values. If convergence succeeds but a variance is negative,
the solution is improper for this model. Do not proceed directly to
fit-index interpretation or replace the negative value by zero in a
report. If an admissible fit has large standard errors or near-perfect
factor correlations, examine weak separation and sensitivity. Different
warnings imply different investigations.

These checks are reproducible from `lavInspect()` and the parameter
table. They are diagnostics at the fitted point; they do not substitute
for the earlier structural identification reasoning.

## Read loadings in their stated units

```{r}
#| label: tbl-cfa-loadings
#| tbl-cap: "Sample B loadings for the frozen M1-cross. Intervals are 95 percent normal-approximation intervals for raw coefficients. Std. all is a point estimate on the fully standardized metric."
tab <- pe[pe$op=="=~",c("lhs","rhs","est","se","ci.lower","ci.upper","std.all")]
names(tab) <- c("Factor","Item","Raw","SE","Lower","Upper","Std.all")
pt2_table(tab)
```

For P1, a one-standard-deviation increase in Planning corresponds to
an estimated `r pt2_num(pe$est[1])` process-index-point change through
the specified loading relation. Its standard error is
`r pt2_num(pe$se[1])`, and its 95% interval runs from
`r pt2_num(pe$ci.lower[1])` to `r pt2_num(pe$ci.upper[1])`.
The fully standardized loading, `r pt2_num(pe$std.all[1])`, expresses
the same relation relative to P1's model-implied standard deviation.
Neither magnitude is a causal effect established by this analysis.

P4 and S4 each have two rows because two loadings were frozen as free
before Sample B was inspected. In this restricted CFA, the secondary
coefficients are estimated jointly with all other free parameters.
They need not equal the rotated EFA coefficients from another sample,
another matrix metric, and a less restricted loading model.

The signs and intervals support reading these as positive conditional
relations within the specified model. They do not license removing
other restrictions without evaluation. A precise coefficient can be
estimated inside a model whose broader covariance representation is
inadequate.

## Standard errors, intervals, and tests

For a regular interior parameter with null value $\vartheta_0$,

$$
z=\frac{\widehat\vartheta-\vartheta_0}
{\operatorname{SE}(\widehat\vartheta)},
\qquad
\widehat\vartheta\ \pm\ 1.96\operatorname{SE}(\widehat\vartheta)
$$

give a conventional asymptotic test and 95% interval under the stated
ML assumptions. The interval concerns repeated-sample coverage of a
population parameter under that model. It is not a range containing
95% of individual factor scores or item responses.

Normal approximations deserve particular caution for boundary variances,
strongly asymmetric sampling distributions, weak identification, and
misspecified distributional assumptions. A fixed parameter has no
sampling standard error as a freely estimated quantity: its printed
value of one or zero is a specification, not a highly precise discovery.
Software may print zero SE and omit its $z$ test for that reason.

Statistical significance also differs from measurement importance.
A small loading can be precisely estimated in a large sample; a
content-essential loading may be uncertain in a small sample. Deleting
items solely because $p>.05$ changes the model and the content represented.
That decision needs more than a significance column.

## Factor relations and residual variance

```{r}
#| label: tbl-cfa-factor-relations
#| tbl-cap: "Estimated factor correlations under fixed-factor scaling, with raw-metric intervals. Here the raw factor covariance is itself a correlation."
fp <- pe[pe$op=="~~" & pe$lhs!=pe$rhs,c("lhs","rhs","est","se","ci.lower","ci.upper")]
names(fp) <- c("Factor1","Factor2","Correlation","SE","Lower","Upper")
pt2_table(fp)
```

All three correlations are positive and distinct from one in the
reported point solution. A correlation is not the percentage of one
factor caused by another. It also does not measure the reliability
of either factor. Later structural models place restrictions on these
associations; measurement estimation alone does not select their
direction.

For a single-loading item, the standardized residual variance is
$\theta_{jj}/\sigma_{jj}$ and the common proportion is its complement.
For a cross-loading item,

$$
h_j^2=\frac{\boldsymbol\lambda_j^{\mathsf T}
\boldsymbol\Phi\boldsymbol\lambda_j}{\sigma_{jj}}.
$$

The numerator includes cross-products involving factor correlations.
It is not generally the sum of two squared standardized coefficients.
Residual variance can include specific stable influences and random
measurement error; the factor model does not automatically separate
those components.

## A planned equality test

M3-equal was named in Chapter 2: it constrains the raw Planning loadings
of P1 and P2 to equality. Let $\mathbf R\boldsymbol\vartheta=\mathbf r$
represent $q$ independent linear restrictions. The Wald statistic is

$$
W=(\mathbf R\widehat{\boldsymbol\vartheta}-\mathbf r)^{\mathsf T}
\left[\mathbf R\widehat{\operatorname{Var}}
(\widehat{\boldsymbol\vartheta})\mathbf R^{\mathsf T}\right]^{-1}
(\mathbf R\widehat{\boldsymbol\vartheta}-\mathbf r).
$$

Under suitable regularity, its reference is $\chi_q^2$. For one
loading difference, the estimated variance is

$$
\operatorname{Var}(\widehat\lambda_{\mathrm{P1}}
-\widehat\lambda_{\mathrm{P2}})
=\operatorname{Var}(\widehat\lambda_{\mathrm{P1}})
+\operatorname{Var}(\widehat\lambda_{\mathrm{P2}})
-2\operatorname{Cov}(\widehat\lambda_{\mathrm{P1}},
\widehat\lambda_{\mathrm{P2}}).
$$

Ignoring the covariance between estimates can give the wrong test.
The syntax labels `p1` and `p2` identify the coefficients:

```r
lavaan::lavTestWald(fit, constraints = "p1 == p2")
```

Here $W=$ `r pt2_num(h$wald$stat)` on one df, with
$p$ `r pt2_p(h$wald$p.value)`. Under the declared model and
normal approximation, the data oppose this particular raw-loading
equality. This is not a test that M1 is globally adequate, nor a
test of equality of fully standardized loadings. The Wald perspective is
one of three views of the same restriction; the restricted-model and
likelihood-ratio views belong with model comparison.

# Build the evaluation handoff record

## Carry the exact candidate, not an assumed baseline

The inherited model has twelve indicators, three correlated factors,
and two additional loadings. Its fixed-factor covariance ledger is

$$
u=78,\qquad
t=14\text{ loadings}+12\text{ residual variances}
+3\text{ factor correlations}=29,\qquad df=49.
$$

The simpler 27-parameter, 51-df model remains M0-simple, a named rival.
It is not silently substituted for the inherited model. The all-item
one-factor rival, the orthogonal restriction, and the raw-loading
equality retain their Chapter 2 definitions. Models keep their M0--M4
identifiers across the three chapters; the separate donor models use
D1--D4, and a later data-developed revision will use an R identifier.

```{r}
#| label: fig-cfa-handoff
#| fig-width: 7
#| fig-height: 5.3
#| fig-cap: "The exact M1-cross structure carried into evaluation, not a redrawn simple-structure substitute."
#| fig-alt: "Three correlated SLE factors have twelve primary indicator relations and the two frozen additional loadings, P4 on Strategy and S4 on Belonging."
pt2_groups_diagram(
  list(Planning=paste0("P",1:4),Strategy=paste0("S",1:4),
    Belonging=paste0("B",1:4)),correlated=TRUE,marker=FALSE,
  cross=data.frame(factor=c("Strategy","Belonging"),item=c("P4","S4"),
    stringsAsFactors=FALSE))
```

## Save what another analyst needs

```{=latex}
\Needspace{22\baselineskip}
```

| Record component | Saved content |
|:--|:--|
| Evidence identity | Sample B role, 360 rows, item order, IDs and hashes |
| Measurement specification | Frozen syntax, item codebook, factors and all rivals |
| Parameter conventions | Unit factor variances, zero latent location, covariance-only model |
| Estimation | ML, normal likelihood, expected-information standard errors |
| Data handling | Complete continuous raw observations; no weights or clustering |
| Numerical status | Convergence, warnings, gradients, admissibility, local rank |
| Parameter output | Full free/fixed ledger, estimates, SEs, intervals, standardizations |
| Fitted moments | Observed $\mathbf S_N$ and exact model-implied covariance |
| Evaluation fields | Conventional test, df, RMSEA/interval, CFI/TLI, SRMR, log likelihood, AIC/BIC |
| Reproduction | R and package versions, source hashes, saved fitted object |

The record is rebuilt from the declarations in this chapter each time it
renders. Its baseline is the fitted `lavaan` object itself, not a
reconstruction from rounded coefficients. A clean-session
reproduction checks that the candidate, coefficients, and settings
match before this chapter renders.

The numerical marker/fixed-factor comparison also succeeds for both
records. For the SLE fit, the largest absolute difference in implied
covariance is
`r format(max(abs(lavaan::fitted(h$marker$fit)$cov-h$audit$Sigma)),scientific=TRUE,digits=2)`;
for the six-indicator donor fit it is
`r format(max(abs(lavaan::fitted(h$donor$fit)$cov-lavaan::fitted(h$donor_marker$fit)$cov)),scientific=TRUE,digits=2)`.
The small numerical tolerances reflect separate optimization, not a
theoretical discrepancy between valid scale choices.

Saving evaluation fields is not interpreting them. They are checked for
availability and matching conventions here so that evaluation can begin
with an intact record. The revalidation sample has not been analysed in
building it.

# Summary and reading bridge

Five equations organize the chapter: the scalar measurement equation,
the covariance decomposition, the scale transformation, the
parameter-to-moment identification mapping, and the discrepancy
minimization. Together they distinguish a proposed structure from a
recoverable parameterization and a successful numerical estimate.

A disciplined specification audit asks whether words, equations,
matrices, diagram, and syntax describe the same model. A parameter audit
then separates free entries, scale normalizations, substantive zeros,
equalities, and derived quantities. The count of distinct moments minus
free coordinates is essential but does not replace local-block or
identification reasoning. Mean models need location conventions as well
as factor scales.

The Sample B candidate passes the stated estimation and admissibility
checks. Its parameters can be read with explicit units and conventional
uncertainty. That is the stopping point: **an identified, converged,
admissible estimate is necessary for ordinary interpretation, but it is
not evidence by itself of adequate model fit.**

The technical route follows the CFA specification tradition developed
in @joreskog1969, @bollen1989, and @brown2006. For software details,
the [parameter-syntax tutorial](https://lavaan.ugent.be/tutorial/syntax2.html)
and [output-extraction guide](https://lavaan.ugent.be/tutorial/inspect.html)
help connect restrictions and fitted-object components. What follows
examines the exact-fit test, fit profile, local residuals, planned
comparisons, and the conditions under which a revision would need new
evidence.

# Exercises {.unnumbered}

1. **Foundational -- represent a model.** Write the six-indicator base
   model in equations, matrices, and syntax. Identify its four free
   non-marker loadings, two marker constraints, six target zeros, and
   residual assumptions. Explain why no drawn arrow is causal proof.
2. **Foundational -- trace a changed relation.** Using the controlled
   parameter values in Section 4, compute $\sigma_{16}$ and
   $\sigma_{36}$. Add $\lambda_{32}=.25$ and recompute the affected
   moments. Contrast adding $\theta_{23}=.50$ instead.
3. **Foundational -- transform the metric.** For marker-scale factor
   variances 9 and 4, covariance 1.8, and target loadings
   $(1,.7,1.1)$ and $(1,.8,1.2)$, obtain the fixed-factor loading matrix
   and factor correlation. Verify a within-factor and a between-factor
   covariance before and after transformation.
4. **Foundational -- count carefully.** Audit the six-indicator base,
   its one-cross-loading extension, and its loading-equality restriction.
   Reconstruct the four-factor, sixteen-indicator count
   $136-(16+16+6)=98$. Then derive $78/(27+k)/(51-k)$ for twelve
   indicators and $k$ extra free cross-loadings under the stated
   simple-structure assumptions.
5. **Foundational -- add means.** Reproduce 90/39/51 for the M0 free-
   intercept extension and 90/41/49 for M1. Explain why freeing all
   latent and indicator intercepts together creates a location ambiguity.
6. **Foundational -- count is not identification.** Construct two
   distinct admissible loading/residual-variance parameter vectors for
   a two-indicator unit-variance factor with
   $\sigma_{11}=\sigma_{22}=1$ and $\sigma_{12}=.4$. Explain how
   an isolated copy of this block can remain unidentified inside a
   positive-df whole model.
7. **Foundational -- read uncertainty.** From the Sample B loading
   table, interpret P3's raw coefficient, SE, interval, and fully
   standardized coefficient. Explain why its raw coefficient above
   one is not an improper solution and why the printed raw SE is not
   the fully standardized SE.
8. **Integrative -- audit the contract.** An analyst feeds R's `cov()`
   matrix into another program, leaves its divisor unspecified, fixes
   marker loadings to one, and compares its test directly with this
   chapter's raw-data fit. List the settings and invariances that must
   be checked before interpreting a numerical difference.
9. **Integrative -- test a restriction.** Explain the P1/P2 raw-loading
   equality, write its $\mathbf R$ row, and identify the covariance
   term required for the variance of the estimated difference. Why
   does this test not decide global fit or full parallel measurement?
10. **Integrative -- preserve the handoff.** Prepare a one-page
    specification-and-estimation record for M1-cross using the saved
    item order, parameter count, data role, options, and admissibility
    checks. List the outputs an evaluation would need, but do not give a
    model-adequacy verdict. Explain why the revalidation sample remains
    unopened at this stage.

# Appendix: Structured means and parcels {.unnumbered}

## Latent location is not identified by naming a factor {.unnumbered}

The mean equation $\boldsymbol\mu=\boldsymbol\nu+
\boldsymbol\Lambda\boldsymbol\alpha$ contains both indicator
intercepts and latent intercepts. In a single group, zero latent
intercepts with free indicator intercepts are a convenient reference
convention. In multigroup work, some intercept restrictions can anchor
latent mean differences, but those restrictions carry measurement
content. A numerical origin does not arise from labelling one group
“control” or one factor “ability.” Later invariance chapters develop
the required scale and location links.

The covariance-only model is unaffected by translating every observed
column by a constant. A structured-mean model need not be. Similarly,
rescaling a factor requires transforming its loadings, covariance, and
latent intercept; rescaling an observed variable requires transforming
its intercept and residual variance as well. An equality imposed after
only some of these transformations can silently change the hypothesis.

## Parceling changes the observed system {.unnumbered}

Let a vector of parcel scores be $\mathbf z=\mathbf A\mathbf y$,
where rows of $\mathbf A$ specify item averages or sums. Then

$$
E(\mathbf z)=\mathbf A\boldsymbol\mu,
\qquad
\operatorname{Var}(\mathbf z)
=\mathbf A\boldsymbol\Sigma\mathbf A^{\mathsf T}.
$$

If $\mathbf A$ reduces dimension, several different item-level
covariance patterns can yield the same parcel covariance. Opposing
secondary loadings or local dependencies may partly cancel within
averages. A well-fitting parcel model therefore does not establish
that the item-level restrictions would fit, and its AIC/BIC is not
directly comparable to an item-level model with different outcomes.

Parceling can be useful under a justified measurement design, but it
is not a generic normality repair or a route around identification
and dimensionality questions. Evaluate item structure and preserve
the meaning of the resulting score before changing the observed
system. No parcels are used in the SLE course record.

# References {.unnumbered}

::: {#refs}
:::
