---
title: "Evaluating and Comparing Confirmatory Factor Models"
author: "Jinsong Chen"
bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
execute:
  echo: false
  warning: false
  message: false
  cache: false
format:
  pdf:
    output-file: chapter_04.pdf
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    include-in-header:
      - file: ../supplement/pdf-graphics-compat.tex
      - text: |
          \setlength{\parindent}{0pt}
          \usepackage{float}
          \usepackage{needspace}
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{caption}
          \captionsetup[table]{position=above,format=plain,justification=raggedright,singlelinecheck=false,labelsep=newline,labelfont=bf,textfont=it}
  html:
    toc: true
    toc-depth: 2
    number-sections: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(dev=if(knitr::is_latex_output()) "cairo_pdf" else "png",
  dpi=160,fig.pos="H",fig.align="center")
options(digits=3,width=80)
source("chapter_02.R")
source("chapter_03.R")
source("chapter_04.R")
b <- pt4_analyze_B()
v <- pt4_revalidate(b)
primary <- b$profiles$M1_cross
fm <- primary$measures
W <- matrix(0,12,4)
W[1:4,1] <- 1/4; W[5:8,2] <- 1/4; W[9:12,3] <- 1/4; W[,4] <- 1/12
colnames(W) <- c("Planning mean","Strategy mean","Belonging mean","All-item mean")
omega <- apply(W,2,function(w)pt4_omega(primary,w))
score <- pt4_scores(primary,b$h$input$x)
added_B <- subset(v$revised_B$parameters,lhs=="Planning" & op=="=~" & rhs=="B4")
added_C <- subset(v$revised_C$parameters,lhs=="Planning" & op=="=~" & rhs=="B4")
extensions <- pt4_analyze_extensions(b)
population_audit <- pt4_population_audit(b)

# Higher-order and bifactor diagrams are QMD-local adaptations of the donor's
# DiagrammeR/Graphviz grammar. Only temporary PDF graphics are produced.
pt4_graphviz_file <- function(dot,prefix="fa4-") {
  needed <- c("DiagrammeR","DiagrammeRsvg","rsvg","digest")
  missing <- needed[!vapply(needed,requireNamespace,logical(1),quietly=TRUE)]
  if(length(missing)) stop("Diagram packages missing: ",paste(missing,collapse=", "))
  svg <- DiagrammeRsvg::export_svg(DiagrammeR::grViz(dot))
  # Deterministic name inside the figure directory so the artifact is captured
  # by _freeze and restored on a clean CI checkout. A random tempfile() name
  # can never be committed, which silently yields a figure-less deployed page.
  dir <- file.path("chapter_04_files","figure-html")
  dir.create(dir,recursive=TRUE,showWarnings=FALSE)
  stem <- paste0(prefix,substr(digest::digest(dot,algo="sha256"),1,12))
  if(knitr::is_latex_output()) {
    path <- file.path(dir,paste0(stem,".pdf"))
    rsvg::rsvg_pdf(charToRaw(svg),file=path)
  } else {
    path <- file.path(dir,paste0(stem,".png"))
    rsvg::rsvg_png(charToRaw(svg),file=path,width=1600)
  }
  path
}

pt4_sle_extension_file <- function(type=c("higher","bifactor")) {
  type <- match.arg(type)
  domains <- c("Planning","Strategy","Belonging")
  prefixes <- c("P","S","B"); x <- c(0,4.2,8.4)
  nodes <- edges <- character()
  for(g in seq_along(domains)) {
    fid <- if(type=="higher") paste0("f",g) else paste0("s",g)
    flabel <- if(type=="higher") domains[g] else paste0(domains[g]," specific")
    nodes <- c(nodes,sprintf('%s [label="%s", shape=ellipse, width=1.45, height=.55, pos="%.2f,7.0!", fillcolor="#DCEAF2"];',fid,flabel,x[g]))
    if(type=="higher") {
      did <- paste0("d",g)
      nodes <- c(nodes,sprintf('%s [label="d%s", shape=circle, width=.38, height=.38, fixedsize=true, pos="%.2f,7.0!", fillcolor="#F2F2F2", fontsize=9];',did,g,x[g]+1.45))
      edges <- c(edges,sprintf('%s -> %s [arrowsize=.55,color="#666666"];',did,fid))
    }
    for(j in 1:4) {
      item <- paste0(prefixes[g],j); eid <- paste0("e",item); y <- 5.35-(j-1)*1.08
      nodes <- c(nodes,
        sprintf('%s [label="%s", shape=box, width=.72, height=.38, pos="%.2f,%.2f!", fillcolor="#F7F1DF"];',item,item,x[g],y),
        sprintf('%s [label="e%s", shape=circle, width=.35, height=.35, fixedsize=true, pos="%.2f,%.2f!", fillcolor="#F2F2F2", fontsize=9];',eid,item,x[g]+1.15,y))
      edges <- c(edges,sprintf('%s -> %s;',fid,item),
        sprintf('%s -> %s [arrowsize=.5,color="#666666"];',eid,item))
    }
  }
  if(type=="higher") {
    nodes <- c(nodes,'g [label="General", shape=ellipse, width=1.25, height=.58, pos="4.20,8.75!", fillcolor="#D9EAD3"];')
    edges <- c(edges,'g -> f1;','g -> f2;','g -> f3;',
      'f2 -> P4 [style=dashed,color="#A45719",penwidth=1.5];',
      'f3 -> S4 [style=dashed,color="#A45719",penwidth=1.5];')
  } else {
    nodes <- c(nodes,'g [label="General", shape=ellipse, width=1.25, height=.58, pos="-2.20,3.70!", fillcolor="#D9EAD3"];')
    edges <- c(edges,vapply(c(paste0("P",1:4),paste0("S",1:4),paste0("B",1:4)),
      function(item)sprintf('g -> %s [color="#6B5A8E",penwidth=1.15];',item),character(1)))
  }
  dot <- paste0('digraph sle { graph [layout=neato, overlap=false, outputorder=edgesfirst, splines=curved, bgcolor="transparent", pad=.2]; ',
    'node [fontname="Arial", fontsize=10, style="filled", color="#405768", penwidth=1.05]; ',
    'edge [fontname="Arial", fontsize=9, color="#405768", arrowsize=.68]; ',
    paste(nodes,collapse=" ")," ",paste(edges,collapse=" ")," }")
  pt4_graphviz_file(dot,paste0(type,"-"))
}
```

{{< pagebreak >}}

::: {.callout-tip title="Chapter box"}
**Central question:** How should a fitted CFA be evaluated, compared
with planned rivals, and revised without confusing development with
independent support; what do higher-order and bifactor representations
add, and what can fit not decide between them?

**Prerequisites:** The covariance equation, restrictions, identification,
ML estimation, parameter metrics, and uncertainty from Chapters 1--3.

**Record:** The exact twelve-indicator SLE fit from Chapter 3, plus a
separate six-indicator continuous microcase for compact comparisons.

**Destination:** A reproducible evaluation, an explicit decision about
a tentative revision, higher-order and bifactor extensions, a retrospective
known-population audit, and bounded interpretations of reliability and
scores. Fit does not establish construct reality, invariance, or causality.
:::

# From an estimated model to an evaluation record

## Begin with what is already fixed

Chapter 3 supplied a converged, admissible estimate of M1-cross on
Sample B. The candidate was frozen using Sample A: twelve continuous
indicators, three correlated factors, P4 additionally loading on Strategy,
and S4 additionally loading on Belonging. All other cross-loadings and
all residual covariances are zero. Factor variances equal one, and
the primary model is covariance-only. It has 29 free parameters and
49 df.

This chapter reads the saved fitted object directly. It does not
reconstruct the baseline from rounded estimates, change the item order,
or substitute a simple-structure model. The inherited object includes
the data role, codebook, parameter table, software settings, warnings,
and sample covariance under the normal-likelihood convention.

```r
handoff <- readRDS(
  "chapter_03_chapter_04_handoff.rds")
fit <- handoff$fit
lavaan::fitMeasures(fit, handoff$fit_measure_names)
```

The code is short because the previous chapter preserved the record.
It is not a license to skip checking it. The render verifies the
candidate and Sample B hashes, package version, item order, and
estimation gates before using any fit statistics.

## Separate five questions

| Question | Evidence | What a favorable answer does not establish |
|:--|:--|:--|
| Did the optimizer finish? | Convergence, gradients, warnings | Admissibility or adequacy |
| Is the solution admissible? | Variances, covariance matrices, local rank | Small model discrepancy |
| How large is global discrepancy? | Model test and fit profile | Absence of important local problems |
| Where are discrepancies located? | Residual patterns and parameter diagnostics | A unique explanation |
| Is interpretation defensible? | Content, estimates, uncertainty, analytic history | Validity for every score use |

These questions are related but not interchangeable. An optimizer can
converge to a badly fitting model. A globally favorable fit profile can
coexist with a questionable item or weakly identified parameter. A
well-measured factor can still be used in an inappropriate substantive
analysis.

The conventional output order often begins with the test statistic,
but our reading order starts with the model and estimation contract.
Fit measures belong to a specified analysis, not to the data set in
isolation.

## Keep the evidence roles visible

Sample A developed the loading candidate. Sample B estimates and
evaluates that frozen candidate and its planned rivals. A new relation
suggested by B is exploratory in B, even if only one path is changed.
Sample C can assess that relation independently only after the changed
specification is frozen. All three samples have 360 cases from the same
documented synthetic continuous population; no new seed is selected to
obtain a preferred teaching outcome.

The six-indicator donor microcase is separate. Its 500 newly generated
continuous records illustrate four model specifications and transparent
likelihood comparisons. They are not a recoded version of the donor's
ordinal observations and cannot borrow the SLE samples' development
or confirmation status.

# What estimation is fitting: one continuous ML family

## Reconstruct the discrepancy before interpreting it

Write the implied covariance as

$$
\boldsymbol\Sigma(\boldsymbol\vartheta)
=\boldsymbol\Lambda\boldsymbol\Phi\boldsymbol\Lambda^{\mathsf T}
+\boldsymbol\Theta.
$$

For $p$ continuous outcomes and the declared sample covariance,

$$
F_{\mathrm{ML}}=
\log|\boldsymbol\Sigma(\boldsymbol\vartheta)|
+\operatorname{tr}\{\mathbf S\boldsymbol\Sigma(\boldsymbol\vartheta)^{-1}\}
-\log|\mathbf S|-p.
$$

The first two terms depend on the model. The remaining terms make the
discrepancy zero when the implied covariance equals the observed one.
The fitted value is the minimum over the specified parameter space,
not necessarily a small number by any substantive standard.

In our normal-likelihood analysis, $\mathbf S=\mathbf S_N$ uses
divisor $N$, and the model statistic is

$$
T_M=N\widehat F_{\mathrm{ML}}.
$$

For M1-cross, the reconstructed discrepancy is
`r pt2_num(primary$Fml,6)`; multiplying by 360 gives
`r pt2_num(fm["chisq"])`. The calculation agrees with the stored
software statistic. This check catches mismatched divisors, a wrongly
ordered matrix, or an incorrect dimension term before those mistakes
propagate into fit-index calculations.

```r
S <- lavaan::lavInspect(fit, "sampstat")$cov
Sigma <- lavaan::fitted(fit)$cov
p <- ncol(S)
Fml <- determinant(Sigma, logarithm = TRUE)$modulus +
       sum(diag(solve(Sigma, S))) -
       determinant(S, logarithm = TRUE)$modulus - p
as.numeric(nrow(handoff$input$x) * Fml)
```

## Do not mix normal, Wishart, and robust conventions

R's ordinary sample covariance uses divisor $N-1$. With raw-data
normal ML, `lavaan` handles conversion to divisor $N$ internally.
Wishart ML uses an $N-1$ covariance and test multiplier. The
[official estimator documentation](https://lavaan.ugent.be/tutorial/est.html)
explains this distinction. A small difference between test statistics
can be a convention difference rather than a scientific disagreement.

Changing an observed unit can leave ML discrepancy unchanged when
the entire model and all restrictions are transformed consistently.
Leaving a raw-loading equality unchanged after changing an item's
unit is not such a transformation. Scale invariance is conditional
on comparing equivalent model specifications.

The present calculations use conventional ML throughout. Later robust
ML procedures adjust standard errors and tests for particular forms
of nonnormality; one cannot subtract two displayed scaled statistics
and assume an ordinary difference test. FIML later handles missing
continuous observations under stated missingness and model assumptions.
Neither topic supplies a categorical-indicator measurement model,
which is outside this series' present scope.

## Large-sample properties are conditional

Consistency concerns increasing sample size under identification and
other appropriate conditions, not merely increasing the number of
indicators. Asymptotic normality of estimates and chi-square references
are approximations, not guarantees for every finite sample or boundary
configuration. More indicators can add information, but also add
parameters, content heterogeneity, and possible misspecification.

Different starts need not always converge to the same optimum. Negative
residual variances, singular factor covariance matrices, or very large
standard errors call for investigation. Here all planned SLE and
donor-model fits pass the recorded convergence and admissibility gates.
That makes ordinary evaluation interpretable; it does not predetermine
the evaluation's conclusion.

# Read global fit as a profile

## The exact-fit null

The covariance-only null is that there exists an admissible parameter
vector such that

$$
H_0:\boldsymbol\Sigma_{\mathrm{population}}
=\boldsymbol\Sigma(\boldsymbol\vartheta).
$$

Under regularity and the conventional ML assumptions,
$T_M\ \dot\sim\ \chi^2_{df_M}$ when that null holds. The dot
marks an asymptotic approximation. The df refer to distinct covariance
restrictions after identification and parameter accounting, not
$N$ minus the number of regression coefficients.

For M1-cross, $T_M=$ `r pt2_num(fm["chisq"])` on 49 df, with
$p=$ `r pt2_num(fm["pvalue"])`. We fail to reject exact fit under
this test. We do not “accept the true model.” Sampling variability,
power, alternative covariance-equivalent models, and departures not
well detected by the test remain relevant.

Large samples can detect small fixed discrepancies. That is not a
reason to declare the test untrustworthy solely because $N$ is large;
it is a reason to separate evidence against *exact* fit from the
magnitude and consequences of the discrepancy. Conversely, a small
sample can fail to reject a consequential departure.

## RMSEA: discrepancy per degree of freedom

For the present $N$ convention,

$$
\widehat{\mathrm{RMSEA}}
=\sqrt{\max\left\{\frac{T_M-df_M}{N\,df_M},0\right\}}.
$$

Under the commonly taught $N-1$ convention, replace $N$ by $N-1$.
State which calculation is being used. The numerator estimates a
noncentrality contribution; it is truncated at zero when $T_M\leq df_M$.
Thus RMSEA equal to zero does not require zero residuals or $T_M=0$.

The interval is obtained by inverting a noncentral chi-square reference
and transforming its noncentrality limits onto the RMSEA scale. A
90% interval is conventional in this literature, whereas the parameter
intervals in Chapter 3 were 95%. These different levels should be
labelled, not silently mixed.

For M1-cross, RMSEA is `r pt2_num(fm["rmsea"])`, with 90% interval
[`r pt2_num(fm["rmsea.ci.lower"])`,
`r pt2_num(fm["rmsea.ci.upper"])`]. The zero estimate arises because
the test statistic is slightly below its df. The nonzero upper limit
expresses uncertainty about population approximation error.

Low-df models can have unstable RMSEA estimates and broad intervals.
A point estimate without its interval can conceal that uncertainty.
Nor does an interval imply equal plausibility of every value inside
it. Explicit close-fit or not-close-fit hypotheses require stated
population thresholds and their own tests; a threshold should not
be inferred retroactively from whichever endpoint looks favorable.
The interval and power perspective is developed by @maccallum1996.

## CFI and TLI need a baseline

For continuous covariance analysis, the usual independence baseline
frees the observed variances and fixes covariances to zero. With twelve
indicators, it has 66 covariance restrictions. It is not a competing
substantive theory of the three factors; it supplies a reference for
incremental improvement.

One conventional CFI expression is

$$
\widehat{\mathrm{CFI}}
=1-\frac{\max(T_M-df_M,0)}
{\max(T_M-df_M,T_B-df_B,0)}.
$$

If the denominator is zero, this expression requires special handling;
it is not an ordinary ratio with a meaningful zero denominator. In
our record the baseline statistic is
`r pt2_num(fm["baseline.chisq"])` on 66 df, so that issue does not
arise. A high CFI means the model improves substantially relative to
this baseline under the stated correction, not that the construct
interpretation has been established.

For positive baseline discrepancy, TLI can be written

$$
\widehat{\mathrm{TLI}}
=\frac{T_B/df_B-T_M/df_M}{T_B/df_B-1}.
$$

Unlike a correlation coefficient, this formula need not remain between
zero and one. A value slightly above one is possible when $T_M<df_M$;
it is not “better than perfect measurement.” We retain the software's
uncapped conventional value rather than altering it to look more
familiar.

Baseline dependence explains why a model with substantial local
discrepancy can still have a high incremental index. The baseline
may be much worse. An index's favorable appearance must be interpreted
relative to the question it answers.

## Compare the planned profiles

```{r}
#| label: tbl-cfa-planned-fit
#| tbl-cap: "Sample B planned models, all using identical observations, indicators, and conventional normal ML. M1-cross is the inherited target."
pt2_table(pt4_fit_table(b$profiles))
```

```{r}
#| label: fig-cfa-rmsea
#| fig-width: 6.8
#| fig-height: 3.4
#| fig-cap: "RMSEA point estimates and 90 percent intervals for the planned models. No traffic-light cutoffs are imposed."
#| fig-alt: "Horizontal intervals compare five planned SLE models. M1 and M3 have low estimates, M0 and M2 have larger estimates, and the one-factor model has the largest."
pt4_interval_plot(b$profiles)
```

The profiles do not reduce to unanimous votes. M2-orthogonal has CFI
about .955 but SRMR about .139: appreciable standardized residual
discrepancy remains despite a sizable improvement over independence.
M3-equal has a favorable global profile even though the targeted
loading-equality test is opposed. Global approximation and a focused
parameter restriction are different questions.

M1-cross has the most favorable overall balance among these
prespecified alternatives. M0-simple leaves more discrepancy after
forcing the two secondary loadings to zero. M4-one does not provide
a satisfactory broad covariance representation relative to these
multi-factor alternatives. These are judgments about this synthetic
sample and candidate set, not universal thresholds for instrument
validation.

# Locate local discrepancy

## Three residual scales

The raw covariance residual is

$$
e_{jk}=s_{jk}-\widehat\sigma_{jk}.
$$

Its unit is the product of the two indicators' units. A residual of
10 has a different meaning for indicators with SDs 2 and 3 than for
indicators with SDs 10 and 12. Two useful rescalings are

$$
e_{jk}^{(s)}=\frac{s_{jk}-\widehat\sigma_{jk}}
{\sqrt{s_{jj}s_{kk}}},
\qquad
e_{jk}^{(r)}=r_{jk}-\widehat r_{jk}.
$$

The first divides both observed and fitted covariance by *observed*
SDs. The second separately converts each covariance matrix to a
correlation matrix. They coincide off the diagonal when the observed
and fitted variances match; otherwise they can differ. Neither is a
residual divided by its sampling standard error. A standardized residual
in the variance-scaling sense is not automatically a $z$ test.

For this covariance-only ML model with free residual variances, fitted
variances closely reproduce observed variances, so the two off-diagonal
rescalings are numerically very close. The script nevertheless calculates
them separately and checks the actual relationship rather than assuming
it in every future model.

## SRMR is a root mean square, not a mean absolute residual

For the covariance-based convention verified here,

$$
\mathrm{SRMR}=\sqrt{\frac{2}{p(p+1)}
\sum_{j=1}^p\sum_{k=1}^j\left(e_{jk}^{(s)}\right)^2}.
$$

The average includes the diagonal and one triangle. Squaring gives
large discrepancies extra influence, and the square root returns the
scale to standardized covariance units. This differs from the mean
absolute residual and from Chapter 2's off-diagonal-only RMS.

The direct calculation gives `r pt2_num(primary$manual_srmr)`,
matching the package SRMR. A small summary does not imply every cell
is small: many near-zero cells can dilute a few larger ones. Mean
structures, multiple groups, and alternative estimators require their
own checked conventions.

## Read patterns in the twelve-item system

```{r}
#| label: fig-cfa-local-residuals
#| fig-width: 6.8
#| fig-height: 5.6
#| fig-cap: "M1-cross standardized covariance residuals in Sample B, lower triangle including the diagonal. Numeric cells retain the sign of observed minus implied covariance."
#| fig-alt: "A twelve-by-twelve lower-triangle residual display shows several negative residuals involving B4 and smaller positive and negative discrepancies elsewhere."
Eplot <- primary$Es; Eplot[upper.tri(Eplot)] <- NA
pt2_heatmap(Eplot,"Sample B: standardized residuals",limit=.10,digits=2,cex=.70)
```

```{r}
#| label: tbl-cfa-local-pairs
#| tbl-cap: "Eight largest absolute standardized covariance residuals. Raw and correlation residuals are shown separately."
pt2_table(head(primary$pairs,8))
```

For B4 and S3, the observed covariance is
`r pt2_num(primary$S["B4","S3"])` and the implied covariance is
`r pt2_num(primary$Sigma["B4","S3"])`. Their raw residual is
`r pt2_num(primary$E["B4","S3"])`; scaled by the observed SDs it
is `r pt2_num(primary$Es["B4","S3"])`. The model overpredicts
this pair's covariance. Several other negative cross-domain residuals
also involve B4, so it is reasonable to inspect that item rather than
treat the largest cell as an isolated software signal.

B4 concerns connection to the institution as a whole, whereas B1--B3
refer more directly to peers, participation, and classroom voice.
That distinction suggests questions about content breadth and local
heterogeneity. It does not uniquely imply that B4 should load negatively
on Planning. A cross-loading, an omitted narrower factor, a response
process difference, and sampling fluctuation can leave overlapping
residual patterns.

Before modifying anything, check coding, distributions, data dependence,
and whether the proposed relation changes the intended measurement
meaning. A globally favorable fit does not prohibit local investigation,
but it also does not require a modification merely to make every
residual smaller.

# Compare planned rivals and test constraints

## Establish nesting by exact restrictions

For compatible conventional ML fits, if restricted model $R$ is
regularly nested in unrestricted model $U$,

$$
\Delta T=T_R-T_U,
\qquad \Delta df=df_R-df_U,
\qquad \Delta T\ \dot\sim\ \chi^2_{\Delta df}
$$

under the null restrictions and regularity conditions. The null is a
population parameter restriction, not that the two observed fit
statistics are literally equal. The larger model cannot have a worse
optimized likelihood if both optimizations have found the corresponding
optima. That monotonicity alone does not establish the chi-square
reference.

For M0 versus M1, set the P4-on-Strategy and S4-on-Belonging loadings
to zero. For M2 versus M1, set all three factor correlations to zero.
For M3 versus M1, impose the one raw-loading equality. These are
different hypotheses with different df changes.

```r
lavaan::lavTestLRT(fit_M1, fit_M0, method = "standard")
```

The compatible fits must use identical indicators, cases, data scale,
mean structure, likelihood convention, and sampling assumptions.
A smaller free-parameter count is not proof of nesting. The one-factor
model's relation to a three-factor model passes through degenerate
latent covariance configurations, so we do not assign an ordinary
chi-square difference test to M4 merely by subtracting df.

```{r}
#| label: tbl-cfa-planned-lr
#| tbl-cap: "Eligible planned restrictions tested against M1 in Sample B. Values reproduce lavTestLRT under conventional ML."
lr <- b$lr[1:3,,drop=FALSE]
pt2_table(data.frame(Comparison=rownames(lr),Delta_T=lr[,1],Delta_df=lr[,2],
  p=ifelse(lr[,3]<.001,"<.001",pt2_num(lr[,3])),row.names=NULL))
```

The two zero cross-loading restrictions cost
`r pt2_num(b$lr["M0_vs_M1","Delta_T"])` chi-square units on two
df. The orthogonality restrictions cost
`r pt2_num(b$lr["M2_vs_M1","Delta_T"])` on three df. Under this
family, both are strongly opposed. The raw-loading equality costs
`r pt2_num(b$lr["M3_vs_M1","Delta_T"])` on one df, with
$p=$ `r pt2_num(b$lr["M3_vs_M1","p"])`. Its relatively favorable
global fit profile did not make that particular restriction innocuous.

Failure to reject a restriction would not establish equivalence.
It would leave the data insufficient to reject it at the chosen level
under that test. If practical equivalence is the question, specify a
meaningful discrepancy region and an appropriate procedure rather
than translate nonsignificance into equality.

## The donor's four-model comparison, rebuilt continuously

The separate donor registry specifies the following models before
the continuous microcase is generated. All use fixed-factor scaling,
free residual variances, diagonal residual covariance, and conventional
normal ML.

```{=latex}
\Needspace{12\baselineskip}
```

| Model | F1 indicators | F2 indicators | Factor relation | Free / df |
|:--|:--|:--|:--|:--|
| D1 | Y1, Y2, Y3 | Y4, Y5, Y6 | Correlated | 13 / 8 |
| D2 | Y1, Y2 | Y3, Y4, Y5, Y6 | Correlated | 13 / 8 |
| D3 | Y1, Y2, Y3 | Y3, Y4, Y5, Y6 | Correlated | 14 / 7 |
| D4 | Y1, Y2, Y3 | Y3, Y4, Y5, Y6 | Orthogonal | 13 / 8 |

```{r}
#| label: fig-cfa-donor-nesting
#| fig-width: 7
#| fig-height: 3.1
#| fig-cap: "Three different one-parameter restrictions of D3. D1 and D2 are not nested in one another merely because both have eight df."
#| fig-alt: "D3 branches to D1 by fixing lambda32 to zero, to D2 by fixing lambda31 to zero, and to D4 by fixing the factor covariance to zero."
pt4_donor_nesting()
```

```{r}
#| label: tbl-cfa-donor-fit
#| tbl-cap: "Recomputed fit profiles for the independent continuous donor microcase, N = 500. These are not the archived donor's numerical results."
pt2_table(pt4_fit_table(b$donor))
```

D1 has CFI about .974 but RMSEA about .093, with a 90% upper limit
above .12. This is another reason not to label models by one index.
D3's additional loading changes the covariance representation and
substantially reduces discrepancy. The D1--D3 difference is
`r pt2_num(b$donor_lr["D1_vs_D3","Delta_T"])` on one df.
Its result is computed for this new continuous microcase; it is not
inherited from an older example with different data.

The two-indicator F1 block in D2 is not isolated: it is linked through
a freely estimated nonzero factor covariance to a factor with several
indicators. That distinguishes it from Chapter 3's orthogonal isolated-
block counterexample. The actual numerical fits pass the local and
admissibility checks. A generic “two indicators always fail” rule
would miss the role of the rest of the model.

## Score, Wald, and LR: three views of one restriction

In D2, $\lambda_{31}=0$: Y3 has no F1 loading. D3 frees it.
The score test uses the restricted fit and asks whether the likelihood
has a slope suggesting that release. A modification index is the
corresponding local one-parameter score statistic. The Wald test uses
the unrestricted estimate and its uncertainty. The LR test fits both
and compares their optimized likelihoods.

```{r}
#| label: tbl-cfa-test-triangle
#| tbl-cap: "Three tests of the same D2-to-D3 loading restriction. All use one df, but their finite-sample numerical values differ."
pt2_table(b$triangle)
```

The numbers are not interchangeable. Here the omitted loading is large
enough that the local score approximation and unrestricted Wald
approximation differ appreciably from the actual LR change. Their
asymptotic relationship under regular local conditions does not
require exact numerical agreement in every finite example.

For equality constraints, use an appropriate score-test routine such
as `lavTestScore()` rather than assuming the ordinary zero-parameter
MI table tests every labelled equality. The chapter's analysis does
not request MIs for the equality-constrained models, avoiding that
misinterpretation.

## AIC and BIC compare eligible likelihoods

For log likelihood $\ell(\widehat{\boldsymbol\vartheta})$ and the
software's effective parameter count $q$,

$$
\mathrm{AIC}=-2\ell+2q,
\qquad
\mathrm{BIC}=-2\ell+q\log N.
$$

Smaller values indicate a relative preference within the eligible
set under each criterion's tradeoff. Neither is a model-test $p$ value,
an absolute adequacy index, or a guarantee that the preferred candidate
is scientifically meaningful. AIC and BIC can disagree because they
penalize complexity differently.

```{r}
#| label: tbl-cfa-information
#| tbl-cap: "Likelihood and information criteria for planned Sample B models. Compare rows within this table, not against another sample or the donor microcase."
ic <- do.call(rbind,lapply(b$profiles,function(p)p$measures[c("logl","npar","aic","bic")]))
pt2_table(data.frame(Model=rownames(ic),ic,row.names=NULL),digits=2)
```

For M1, the package count is 29. Substituting its log likelihood and
$N=360$ reproduces both criteria. The same observations and outcomes
make these rows comparable; the one-factor model can enter this
relative likelihood comparison without being assigned a regular
nested LR reference.

D1 and D2 provide a compact nonnested comparison with the same six
outcomes and 500 cases. Their AICs are
`r pt2_num(b$donor$D1$measures["aic"],2)` and
`r pt2_num(b$donor$D2$measures["aic"],2)`, respectively. D1 has
the smaller value, but D3's profile still matters in the larger
candidate set. Being preferable to a poor alternative does not
establish adequate fit.

Do not compare AIC/BIC across A, B, and C, across item subsets or
parcels, or across incompatible likelihood definitions. Do not carry
an EFA statistic with the same printed name into this table without
checking its exact definition. Eligibility precedes arithmetic.

## Loading equality is not full parallelism

M3 tests a raw-loading equality for P1 and P2. As a separate teaching
sensitivity, V1 constrains their *residual variances* equal while
leaving their loadings free. V1 is not retroactively listed as an
A-frozen rival; it illustrates a different measurement restriction.
Its LR difference from M1 is
`r pt2_num(b$lr["V1_vs_M1","Delta_T"])` on one df, with
$p=$ `r pt2_num(b$lr["V1_vs_M1","p"])`.

These restrictions concern different aspects of the model and are
scale-dependent. Equal loadings plus equal residual variances are
still not a complete statement about parallel measurements when
intercepts and means matter. Equal intercepts and the appropriate
common true-score structure must also be considered. Nonsignificant
tests would not establish exact parallelism, and a loading equality
for just one pair would not establish it for an entire instrument.

# Extensions of the measurement model

## Ask what accounts for the factor correlations

M1-cross represents Planning, Strategy, and Belonging as three correlated
first-order factors. That representation estimates their associations
without claiming what produces them. A higher-order model asks a more
restrictive question: can one additional latent variable reproduce the
covariance among those first-order factors?

Let $\boldsymbol\Lambda_l$ contain the item loadings on the first-order
factors. A one-factor higher-order model writes

$$
\boldsymbol\eta_l=\boldsymbol\gamma G+\boldsymbol\delta,
\qquad
\boldsymbol\Phi_l
=\boldsymbol\gamma\boldsymbol\gamma^{\mathsf T}
 +\boldsymbol\Psi_h,
$$ {#eq-cfa-higher-phi}

where $G$ has unit variance for convenience and
$\boldsymbol\Psi_h$ is diagonal in this version. Substitution into the
measurement covariance gives

$$
\boldsymbol\Sigma
=\boldsymbol\Lambda_l
 (\boldsymbol\gamma\boldsymbol\gamma^{\mathsf T}
  +\boldsymbol\Psi_h)
 \boldsymbol\Lambda_l^{\mathsf T}
 +\boldsymbol\Theta.
$$ {#eq-cfa-higher-sigma}

The components of $\boldsymbol\delta$ are first-order *disturbances*: the
parts of Planning, Strategy, and Belonging not represented by $G$. They are
not indicator residuals. The diagram retains M1-cross's two frozen secondary
item loadings; changing those while adding $G$ would confound two model
changes.

```{r}
#| label: fig-cfa-higher-order
#| fig-width: 7
#| fig-height: 6.1
#| fig-cap: "Higher-order representation of the inherited SLE measurement model. General accounts for first-order factor covariances; first-order disturbances remain, and the P4-on-Strategy and S4-on-Belonging loadings are retained."
#| fig-alt: "One General factor predicts Planning, Strategy, and Belonging. Those three first-order factors predict their four indicators, with two dashed secondary loadings retained. Each first-order factor has a disturbance and each item has a residual."
knitr::include_graphics(pt4_sle_extension_file("higher"))
```

An arrow from $G$ to a first-order factor represents a hypothesized common
source of covariance. Cross-sectional covariance data do not establish that
causal direction. Nor does a favorable fit show that $G$ has the intended
substantive meaning.

## Three correlations exactly determine three higher-order loadings

With three standardized first-order factors and positive nonzero
correlations,

$$
\phi_{12}=\gamma_1\gamma_2,
\qquad \phi_{13}=\gamma_1\gamma_3,
\qquad \phi_{23}=\gamma_2\gamma_3.
$$

Multiplying the first two equations and dividing by the third yields

$$
\gamma_1^2=\frac{\phi_{12}\phi_{13}}{\phi_{23}},\qquad
\gamma_2^2=\frac{\phi_{12}\phi_{23}}{\phi_{13}},\qquad
\gamma_3^2=\frac{\phi_{13}\phi_{23}}{\phi_{12}}.
$$ {#eq-cfa-higher-inversion}

Signs are chosen consistently with the factor orientation. Each ratio must
be nonnegative and finite, and the standardized disturbance variances
$1-\gamma_j^2$ must be nonnegative. These are substantive parameter-space
conditions, not algebraic decoration.

For the fitted Sample B M1-cross factor correlations, the calculation is:

```{r}
#| label: tbl-cfa-higher-loadings
#| tbl-cap: "Closed-form higher-order coordinates from the fitted Sample B first-order factor correlations. These are standardized coordinates, not the SLE-generating population values."
phi <- extensions$Phi
gam <- data.frame(Factor=names(extensions$gamma),
  Gamma=unname(extensions$gamma),
  Disturbance=unname(extensions$disturbances))
pt2_table(gam)
```

Thus $\widehat\gamma_1=$ `r pt2_num(extensions$gamma["Planning"])`,
$\widehat\gamma_2=$ `r pt2_num(extensions$gamma["Strategy"])`, and
$\widehat\gamma_3=$ `r pt2_num(extensions$gamma["Belonging"])`.
For example, the ratio for $\widehat\gamma_1^2$ is
`r pt2_num(phi[1,2])` times `r pt2_num(phi[1,3])`, divided by
`r pt2_num(phi[2,3])`, which equals
`r pt2_num(extensions$gamma[1]^2)`.

The raw higher-order loadings printed by `lavaan` are on a different
first-order disturbance metric when `std.lv = TRUE`; their standardized
values are the coordinates that reproduce the fitted factor-correlation
matrix above. Mixing the raw and standardized columns would break the
closed-form check.

## With three first-order factors, fit cannot choose between the forms

The correlated model estimates $m(m-1)/2$ first-order factor correlations.
The one-factor higher-order form estimates $m$ higher-order loadings. Their
difference is

$$
\frac{m(m-1)}2-m=\frac{m(m-3)}2.
$$ {#eq-cfa-higher-count}

For $m=3$, the difference is zero. At a regular admissible solution, the
three correlations and three higher-order loadings are alternative
coordinates for the same information. The fitted records confirm this
equivalence rather than treating matching fit as accidental.

```{r}
#| label: tbl-cfa-extension-fit
#| tbl-cap: "Sample B fit profiles for M1-cross, its equivalent higher-order representation, and the nonnested orthogonal bifactor rival. The extension fits are didactic post-confirmation analyses, not Sample-A-frozen rivals."
pt2_table(pt4_fit_table(list(M1_cross=primary,
  Higher_order=extensions$profiles$HO_SLE,
  Bifactor=extensions$profiles$BIF_SLE)))
```

M1-cross and the higher-order form have 29 free covariance parameters,
49 df, the same fitted covariance matrix to numerical tolerance, and the
same likelihood, AIC, and BIC. Therefore these data cannot decide which
story is preferable. Interpretation must come from the intended construct
representation and from evidence beyond the reproduced covariance.

For $m=4$, six factor correlations are replaced by four higher-order
loadings, so the higher-order model imposes two restrictions. The
four-factor model introduced in Chapter 3 therefore supplies a testable
case; three first-order factors do not. More first-order factors do not make
the model automatically correct---they merely create restrictions that can
be evaluated.

## A bifactor model allocates item covariance directly

A standard bifactor model takes a different route. Every item loads directly
on one general factor and on one assigned domain-specific factor:

$$
Y_{ij}=\nu_j+\lambda_{jg}G_i+\lambda_{js}S_{is}+\epsilon_{ij}.
$$

In the version fitted here, $G$, Planning-specific, Strategy-specific, and
Belonging-specific are mutually orthogonal, residual covariances are zero,
and each item has exactly one specific loading. A specific factor represents
common variance among its assigned items after the general factor is held
constant; it is not an item residual.

```{r}
#| label: fig-cfa-bifactor
#| fig-width: 7
#| fig-height: 6.1
#| fig-cap: "Orthogonal SLE bifactor model. General and the three domain-specific factors load directly on items; all four factors are mutually orthogonal and unshown loadings are zero."
#| fig-alt: "A General factor loads on all twelve items. Planning-specific, Strategy-specific, and Belonging-specific each load on their four assigned items. No factor covariance arrows are present, and each item has a residual."
knitr::include_graphics(pt4_sle_extension_file("bifactor"))
```

The corresponding core syntax is explicit about orthogonality:

```r
General =~ P1 + P2 + P3 + P4 + S1 + S2 + S3 + S4 +
           B1 + B2 + B3 + B4
Planning_s  =~ P1 + P2 + P3 + P4
Strategy_s  =~ S1 + S2 + S3 + S4
Belonging_s =~ B1 + B2 + B3 + B4
General ~~ 0*Planning_s + 0*Strategy_s + 0*Belonging_s
Planning_s ~~ 0*Strategy_s + 0*Belonging_s
Strategy_s ~~ 0*Belonging_s
```

Fixed-factor scaling leaves 24 loadings and 12 item residual variances,
so $t=36$ and $df=78-36=42$. Relative to M1-cross, the bifactor adds a
general loading for every item but also removes the three factor
correlations and replaces the two M1 secondary loadings with exclusive
domain-specific assignments. It is therefore a distinct, nonnested rival,
not M1 plus twelve harmless arrows. An ordinary likelihood-ratio difference
test between the two is not licensed.

```{r}
#| label: tbl-cfa-bifactor-loadings
#| tbl-cap: "Fully standardized bifactor loadings in Sample B. Specific-factor coefficients describe covariance remaining after the orthogonal general factor; they need not all be large or positive."
bl <- extensions$bifactor_loadings
general <- bl[bl$Factor=="General",c("Item","Std_all")]
names(general)[2] <- "General"
specific <- bl[bl$Factor!="General",c("Factor","Item","Std_all","p")]
names(specific)[3:4] <- c("Specific","p_specific")
bwide <- merge(general,specific,by="Item",sort=FALSE)
bwide <- bwide[match(c(paste0("P",1:4),paste0("S",1:4),paste0("B",1:4)),bwide$Item),]
pt2_table(bwide)
```

The bifactor fit converges, has a full-rank local derivative matrix, and has
positive estimated residual variances. Yet the Strategy-specific loading for
S4 is `r pt2_num(subset(bl,Factor=="Strategy_s" & Item=="S4")$Std_all)`,
with $p=$ `r pt2_num(subset(bl,Factor=="Strategy_s" & Item=="S4")$p)`.
That weak negative coefficient is not a numerical failure, but it strains a
simple claim that every item carries a clear positive domain-specific signal.
The bifactor also has higher AIC and BIC than M1-cross. Greater flexibility
and a familiar name do not guarantee a more useful representation.

## Omega-hierarchical answers a score-specific question

For declared score weights $\mathbf w$, the share of observed score variance
attributable specifically to the bifactor general factor is

$$
\omega_h(\mathbf w)=
\frac{(\mathbf w^{\mathsf T}\boldsymbol\lambda_g)^2\phi_{gg}}
     {\mathbf w^{\mathsf T}\boldsymbol\Sigma\mathbf w}.
$$ {#eq-cfa-omega-h}

For the unit-weighted sum of all 12 items, the fitted bifactor gives
$\omega_h=$ `r pt2_num(extensions$omega_h)` and total model-based omega
$=$ `r pt2_num(extensions$omega_total)`. Thus about
`r pt2_num(extensions$reliable_general_share)` of the score's modeled
reliable variance is attributed to the general factor. The remainder is
modeled reliable specific-factor variance, not automatically error.

These coefficients are conditional on this model, these weights, and this
sample. A substantial $\omega_h$ does not prove essential unidimensionality,
validate the score interpretation, or erase the weak specific-loading
pattern. It answers a narrower variance-decomposition question.

# Diagnose, revise, and revalidate

## MI and EPC are conditional local diagnostics

A modification index approximates the improvement in the model
statistic if one fixed parameter were released and the model re-estimated.
The expected parameter change, EPC, approximates the parameter's change
under that local release. The standardized EPC expresses a corresponding
rescaled magnitude, with the exact scale depending on the parameter type.

Read them together with the residual pattern, parameter meaning, and
analytic history:

$$
(\mathrm{MI},\mathrm{EPC},\mathrm{standardized\ EPC},
\mathrm{residual\ pattern},\mathrm{meaning},\mathrm{status}).
$$

An MI is not the exact LR change, a posterior probability of a missing
path, or a guarantee that the proposed parameter is substantively real.
Searching many MIs and reporting only the largest ordinary $p$ value
does not retain the interpretation of a single prespecified test.
The [official modification-index guide](https://lavaan.ugent.be/tutorial/modindices.html)
also distinguishes the different standardized EPC columns.

```{r}
#| label: tbl-cfa-mi
#| tbl-cap: "Largest one-parameter diagnostics for M1 in Sample B. No cutoff is used as an automatic release rule."
mi <- head(primary$MI,6)[,c("lhs","op","rhs","mi","epc","sepc.all")]
names(mi) <- c("Left","Relation","Right","MI","EPC","Std.EPC")
pt2_table(mi)
```

The top suggestion is a negative B4 loading on Planning, with MI
`r pt2_num(primary$MI$mi[1])` and EPC
`r pt2_num(primary$MI$epc[1])`. It aligns with some negative
cross-domain residuals involving B4. But institutional connection
wording does not offer a compelling Planning mechanism. The numerical
clue is therefore weaker than a justified measurement revision.

By contrast, the D2 microcase's missing F1 loading on Y3 was already
one of the registry's specified alternatives. A planned rival and a
newly discovered relation may be represented by the same software
operation, yet have different analytic status.

## A deliberately tentative one-path sensitivity

For teaching, we retain one tentative model, **R1-B4**, which frees
Planning's loading on B4 while leaving all other M1 restrictions
unchanged. It has 30 free parameters and 48 df. This is a transparent
sensitivity demonstration, not a recommendation to modify a favorable
model whenever an MI is available.

The revision's syntax, rationale limitations, parent model, and handoff
hash were frozen before Sample C was read. Only this one diagnostic
revision is fitted. We do not chase additional MIs, remove items,
select a new seed, or invent an explanation after seeing C.

In Sample B, the actual decrease in the model statistic is
`r pt2_num(v$lr_B["Delta_T"])` on one df, close to but not
identical to the MI. The estimated added loading is
`r pt2_num(added_B$est)`, with 95% interval
[`r pt2_num(added_B$ci.lower)`, `r pt2_num(added_B$ci.upper)`].
Its conventional same-sample LR $p$ value is
`r pt2_num(v$lr_B["p"])`. Because this path was selected from
the diagnostics, that value is descriptive of the refit, not an
unadjusted confirmatory discovery.

## The frozen retest does not support the added path

```{r}
#| label: tbl-cfa-revalidation
#| tbl-cap: "One frozen revision across its development and revalidation records. Compare models within B or within C; do not compare likelihood values across samples."
pt2_table(pt4_fit_table(list(B_M1=primary,B_R1=v$revised_B,
  C_M1=v$primary_C,C_R1=v$revised_C)))
```

```{r}
#| label: fig-cfa-revision
#| fig-width: 6.8
#| fig-height: 2.7
#| fig-cap: "The added B4-on-Planning loading in the sample that suggested it and in the untouched retest. Intervals are 95 percent conventional parameter intervals."
#| fig-alt: "The added loading is negative in B with an interval below zero, but near zero in C with an interval spanning both signs."
pt4_revision_plot(v)
```

In Sample C the added coefficient is
`r pt2_num(added_C$est)`, with interval
[`r pt2_num(added_C$ci.lower)`, `r pt2_num(added_C$ci.upper)`].
The LR improvement is only `r pt2_num(v$lr_C["Delta_T"])` on
one df, with $p=$ `r pt2_num(v$lr_C["p"])`.
The candidate relation does not receive independent support in this
retest. We retain M1-cross as the course's working measurement model
and do not promote R1 to the final specification.

This is not proof that the population coefficient is exactly zero.
It is a bounded decision based on weak content justification, a
data-selected improvement in B, and no corresponding evidence in C.
The failed sensitivity remains in the report rather than disappearing
from the analytic history.

## What revalidation does and does not fix

A held-out sample protects a specified retest from using the same
sampling fluctuations that suggested the change. It does not repair
poor content coverage, turn simulation into external validity, or
justify unlimited searches in the held-out record. Once C is used to
choose further modifications, it becomes development evidence for
those decisions.

Revalidation here means re-estimating and evaluating a frozen model
structure in independent observations from the same synthetic
population. It does not mean holding every B estimate fixed as a
predictive parameter value. That would be a different, more stringent
out-of-sample prediction question requiring its own objective and
evaluation protocol.

```{=latex}
\Needspace{22\baselineskip}
```

| Stage | Decision | Evidence role |
|:--|:--|:--|
| A | Retain three factors and two secondary loadings | Development |
| Before B | Freeze M0--M4 and M1 target | Specification fixed |
| B, Chapter 3 | Estimate M1 and audit its solution | Confirmation-record estimation |
| B, Chapter 4 | Evaluate planned models; inspect local diagnostics | Planned evaluation plus labelled exploration |
| Before C | Freeze tentative R1-B4 without further changes | Retest specification fixed |
| C | Evaluate M1 and R1; do not search again | Independent synthetic revalidation |
| Report | Retain M1; disclose unsupported R1 | Bounded conclusion |

# Audit the frozen model against the known population

## Reveal the generating feature only after the B/C sequence

The SLE records are synthetic, so the generating covariance is available to
the course author. It was deliberately withheld while the frozen B/C
sequence was conducted. We now use it retrospectively to ask what the
working M1-cross representation misses. This audit cannot retroactively make
a diagnostic planned, and it does not authorize another model search.

The generating residual covariance between B1 and B2 is
`r pt2_num(population_audit$omitted_standardized_residual)` on the
standardized item scale (`r pt2_num(population_audit$omitted_raw_residual)`
on the raw covariance scale). M1-cross fixes that relation to zero. A simple
four-indicator tetrad makes the restriction visible. In the generating
correlation matrix,

$$
r_{12}r_{34}-r_{13}r_{24}
=.712(.335)-.536(.370)=.0402,
$$

whereas the fitted one-factor Belonging block imposes zero, apart from
numerical tolerance. The nonzero population tetrad identifies a specific
covariance restriction that global indices summarize but cannot name.

Now that the generator is revealed, its first-order correlations also give a
clean higher-order calculation:

$$
\gamma_1^2=\frac{.42(.30)}{.35}=.36,
\qquad
\boldsymbol\gamma=(.60,.70,.50)^{\mathsf T},
\qquad
\operatorname{diag}(\boldsymbol\Psi_h)=(.64,.51,.75).
$$

These are **generating-population coordinates**, not the Sample B estimates
in Section 6. Because three factor correlations can be reparameterized this
way, the calculation does not show that a higher-order causal process
generated the data; the documented SLE generator simply used a correlated
factor covariance matrix.

## Population fitting separates displacement from sampling fluctuation

Fit the unchanged M1-cross model directly to the exact generating covariance
matrix. The optimizer then finds the population-ML, or pseudo-true,
coordinates under that misspecified model. No random sample is involved in
the following loading displacement:

```{r}
#| label: tbl-cfa-population-displacement
#| tbl-cap: "Belonging loadings in the generating model and at the population-ML optimum under M1-cross. Displacement is deterministic under these two specifications; it is not inferred from one realized sample."
pt2_table(population_audit$displacement)
```

The two indicators sharing the omitted residual relation have higher
pseudo-true primary loadings; the other two have lower loadings. This is a
systematic consequence of fitting the stated restricted model to the stated
population covariance. By contrast, comparing two estimates in Sample B can
show sensitivity to a specification choice but cannot by itself establish
estimator bias.

## Population discrepancy is not the sample RMSEA formula

For population covariance $\boldsymbol\Sigma_0$ and its pseudo-true fitted
covariance $\boldsymbol\Sigma^*$, define

$$
F_0=log|\boldsymbol\Sigma^*|
+\operatorname{tr}(\boldsymbol\Sigma_0
 \boldsymbol\Sigma^{*-1})
-\log|\boldsymbol\Sigma_0|-p.
$$

The audit gives $F_0=$ `r format(population_audit$population["F0"],digits=7)`.
Population RMSEA is therefore

$$
\operatorname{RMSEA}_0=\sqrt{F_0/df}
=\sqrt{`r format(population_audit$population["F0"],digits=7)`/49}
=`r pt2_num(population_audit$population["RMSEA"])`.
$$

```{r}
#| label: tbl-cfa-population-fit
#| tbl-cap: "Population discrepancy measures versus sample-index formulas applied to the same population covariance with N = 360. The latter are not observed-sample evidence."
pa <- population_audit$population
sf <- population_audit$sample_formula
pt2_table(data.frame(Quantity=c("Model statistic","df","RMSEA","CFI","SRMR"),
  Population_definition=c(NA,population_audit$df,pa["RMSEA"],pa["CFI"],pa["SRMR"]),
  Sample_formula_at_N360=c(sf["chisq"],sf["df"],sf["rmsea"],sf["cfi"],sf["srmr"])))
```

Applying the ordinary finite-sample formulas at $N=360$ to population input
produces RMSEA .000 and CFI 1.000 because their degree-of-freedom corrections
truncate the small discrepancy. Those numbers do **not** make the model true:
the population discrepancy, the tetrad, and the loading displacement remain
nonzero. Even exact knowledge of the population covariance does not turn a
global index into a certificate of specification.

# Interpret parameters, reliability, and score boundaries

## Parameters still depend on the model

Chapter 3's parameter intervals remain conditional on M1's restrictions
and ML assumptions. The favorable fit profile makes ordinary
interpretation more defensible than it would be under severe
misspecification, but it does not remove that conditionality.
Residual inspection also shows that individual relations are not
reproduced perfectly.

B4's relatively weak primary loading can reflect a broader or different
content emphasis. Removing it would alter the domain represented by
Belonging. It would also change the observed outcome set and invalidate
a direct AIC/BIC comparison with the twelve-item fit. “Higher loading
after deletion” is not a complete scale-development rationale.

Factor correlations are estimated associations, not reliability
coefficients or directional effects. A correlation close to one can
raise a question about empirical separation, but a universal cutoff
cannot replace checking uncertainty, item content, and alternative
model structures. The present three correlations are moderate positive
relations in the fixed-factor metric.

## Define a score before calculating reliability

For a continuous composite

$$
c_i=\mathbf w^{\mathsf T}\mathbf y_i,
$$

its model-implied variance is
$\mathbf w^{\mathsf T}\boldsymbol\Sigma\mathbf w$.
The proportion attributed to all modeled common factors is

$$
\omega(\mathbf w)=
\frac{\mathbf w^{\mathsf T}\boldsymbol\Lambda
\boldsymbol\Phi\boldsymbol\Lambda^{\mathsf T}\mathbf w}
{\mathbf w^{\mathsf T}\boldsymbol\Sigma\mathbf w}.
$$

The numerator is the variance of the model's common part for that
weighting rule. The denominator also includes
$\mathbf w^{\mathsf T}\boldsymbol\Theta\mathbf w$.
This expression retains cross-loadings, factor correlations, and any
modeled residual covariances. A shortcut based only on selected
loadings may omit essential terms.

For the course example, define four scores on the *raw process-index
scale*: the mean of P1--P4, the mean of S1--S4, the mean of B1--B4,
and the mean of all twelve indicators. Weights are .25 on the relevant
four items and zero elsewhere for a subscale, or $1/12$ on all items
for the total. A standardized-item mean would be a different weighting
rule and requires a separately defined calculation.

```{r}
#| label: tbl-cfa-reliability
#| tbl-cap: "Model-conditional common-variance proportions for explicitly defined raw-item composites under M1 in Sample B. These are point estimates, not validation thresholds."
pt2_table(data.frame(Composite=names(omega),Common_variance_proportion=unname(omega)))
```

The all-item mean has a larger common-variance proportion than the
individual subscale means, but that does not make it unidimensional.
Its numerator combines common variance from three correlated factors.
Similarly, the Planning and Strategy means include their retained
secondary-loading contributions. Their values are not proportions
attributable exclusively to the intended named factor.

Reliability concerns a specified score under a specified model and
population. It does not establish dimensionality, content coverage,
fairness, predictive usefulness, or invariance across groups. The
reported estimates also have sampling uncertainty. A bootstrap interval
would require a declared resampling design, model-refitting checks,
and an explicit treatment of failed or improper solutions; it would
not account automatically for every earlier model-selection decision.

## A small calculation exposes the assumptions

For a unit-weighted sum of three indicators loading only on a
unit-variance factor, with loadings $.8,.7,.6$ and uncorrelated residual
variances $.36,.51,.64$,

$$
\omega=\frac{(.8+.7+.6)^2}
{(.8+.7+.6)^2+.36+.51+.64}
=\frac{4.41}{5.92}\approx .745.
$$

If the first two residuals instead have covariance .10, the denominator
increases by $.20$, because both symmetric covariance entries enter
the sum variance. The common numerator is unchanged. Ignoring that
residual relation would overstate the common-variance proportion.

The signs also matter. Replacing loadings by absolute values silently
changes the implied score unless the indicator coding and weights
are correspondingly changed. Reliability is not improved by hiding
opposing contributions inside an absolute-value formula.

AVE can summarize the average communality for a simple,
fully-standardized single-factor item set. It is not interchangeable
with reliability of a weighted composite, and the SLE cross-loading
subscales do not satisfy that simple interpretation without qualification.
Coefficient $H$ requires its own optimal-weighting conditions and is
left to the optional note.

## Factor scores are predictions, not observed latent values

A factor score uses a model-dependent weighting matrix:

$$
\widehat{\boldsymbol\eta}_i
=\mathbf A(\mathbf y_i-\widehat{\boldsymbol\mu})
$$

under the zero latent-location convention. The regression score
targets a conditional mean under a normal model and typically shrinks
uncertain individuals toward the latent mean. Bartlett's score uses
a different weighting criterion, emphasizing residual-variance
weighting. Both use estimated parameters and can differ for the same
person under the same fitted covariance model.

```{r}
#| label: fig-cfa-scores
#| fig-width: 6.7
#| fig-height: 3.4
#| fig-cap: "Two ways of scoring Planning in the same synthetic Sample B. Agreement is not identity and neither axis is an observed true factor value."
#| fig-alt: "A scatterplot compares regression and Bartlett Planning scores, with an identity reference line; points are closely related but do not lie exactly on that line."
par(mar=c(4,4,1,1))
plot(score$regression[,1],score$bartlett[,1],pch=16,cex=.45,col="#40576888",
  xlab="Regression factor score",ylab="Bartlett factor score")
abline(0,1,lty=2,col="#9c5600")
```

Under the fitted normal model with parameters treated as known, the
conditional SDs of the three latent factors given the indicators are
approximately `r paste(pt2_num(sqrt(diag(score$posterior))),collapse=", ")`.
The factors have unit marginal variance, so these are not zero
uncertainties. They also omit parameter-estimation and model-selection
uncertainty. A numerical score is a prediction, not the elimination
of the measurement problem.

Exporting estimated factor scores into an ordinary regression and
treating them as perfectly measured covariates can change the target
and distort inference. For many substantive questions, a joint latent
SEM is the appropriate next representation, provided its measurement
and structural assumptions are justified. Scores can still be useful
for clearly stated descriptive purposes; their method, scale, and
uncertainty should accompany them.

# Report a reproducible CFA evaluation

## Report in the order of dependence

A report should state the sample and data role, indicators and coding,
model restrictions, factor metric, estimator and likelihood convention,
mean structure, and missing-data/design handling before presenting
fit. Follow with convergence and admissibility, global and local
evidence, planned comparisons, any data-developed changes, parameter
interpretation, and the limits of the conclusion.

A possible compact report for this record is:

> We evaluated a previously frozen twelve-indicator, three-factor model
> in a simulated confirmation sample of 360 complete continuous records.
> The model retained two specified secondary loadings, free factor
> correlations, unit factor variances, and diagonal residual covariance.
> Conventional normal ML was used without a mean structure. The fit
> converged without warnings and was admissible. M1-cross had
> $\chi^2(49)=$ `r pt2_num(fm["chisq"],2)`,
> $p=$ `r pt2_num(fm["pvalue"])`, RMSEA =
> `r pt2_num(fm["rmsea"])` (90% CI
> [`r pt2_num(fm["rmsea.ci.lower"])`,
> `r pt2_num(fm["rmsea.ci.upper"])`]), CFI =
> `r pt2_num(fm["cfi"])`, TLI = `r pt2_num(fm["tli"])`,
> and SRMR = `r pt2_num(fm["srmr"])`.
> Planned restrictions removing the secondary loadings or forcing
> orthogonality were opposed by compatible ML difference tests.
> Local residuals involving B4 motivated one tentative sensitivity
> model. Its added loading did not receive support in an independent
> frozen retest of 360 simulated records, so the primary model was
> retained. Reliability estimates refer to declared raw-item composites
> and do not establish unidimensionality or general validity.

The report includes a favorable conclusion about a working covariance
representation without claiming the factors have been directly
observed or the measurement theory uniquely confirmed. It also retains
the unsuccessful sensitivity analysis, a material part of what was done.

## Preserve reproducibility without overwhelming the main text

Archive the exact syntax, data manifest and codebook, parameter tables,
fitted objects, comparison eligibility checks, software versions, and
frozen revision record. The main text need not print every diagnostic
row, but its selected rows must be traceable and must not conceal an
improper fit or unsuccessful comparison.

The canonical analysis reconstructs the ML statistic, RMSEA and its
interval, CFI, TLI, SRMR, AIC, BIC, and eligible LR differences from
the same fitted family. Reproduction is an arithmetic and provenance
check; it does not replace substantive review. Public chapter renders
have no dependency on private teaching files or archived donor data.

# Bridge from CFA to structural equation modeling

## Add structural restrictions without discarding measurement

The measurement equation remains
$\mathbf y=\boldsymbol\nu+\boldsymbol\Lambda\boldsymbol\eta
+\boldsymbol\epsilon$.
A later structural model can write

$$
\boldsymbol\eta=\boldsymbol\alpha+\mathbf B\boldsymbol\eta
+\boldsymbol\Gamma\boldsymbol\xi+\boldsymbol\zeta,
$$

where $\mathbf B$ contains latent structural paths,
$\boldsymbol\Gamma$ links exogenous predictors, and
$\boldsymbol\zeta$ is a structural disturbance. If
$\mathbf I-\mathbf B$ is invertible,

$$
E(\boldsymbol\eta)=
(\mathbf I-\mathbf B)^{-1}
\{\boldsymbol\alpha+\boldsymbol\Gamma E(\boldsymbol\xi)\}.
$$

Thus $\boldsymbol\alpha$ remains the latent *intercept*, not generally
the marginal latent mean. In the earlier CFA, $\mathbf B=\mathbf0$
and there were no predictor terms, so the two coincided. No new
intercept symbol is required when structural paths arrive.

Let $\boldsymbol\Psi=\operatorname{Var}(\boldsymbol\zeta)$.
If predictors and disturbances are uncorrelated, the total latent
covariance is

$$
\boldsymbol\Phi=
(\mathbf I-\mathbf B)^{-1}
\{\boldsymbol\Gamma\operatorname{Var}(\boldsymbol\xi)
\boldsymbol\Gamma^{\mathsf T}+\boldsymbol\Psi\}
(\mathbf I-\mathbf B)^{-\mathsf T}.
$$

$\boldsymbol\Phi$ still means total latent covariance;
$\boldsymbol\Psi$ now means structural disturbance covariance;
$\boldsymbol\Theta$ remains indicator residual covariance. Keeping
these roles distinct prevents a notation change from hiding a change
in the statistical model.

## What the next chapters will add

Path analysis first makes structural restrictions visible with
observed variables. General SEM then combines those restrictions with
measurement models and develops direct and indirect effects. Invariance
examines whether measurement parameters can be linked across groups
or occasions. SEM-based latent growth uses structured loadings and
means to represent trajectories. Multilevel growth belongs to Part
Three.

The familiar checks do not disappear: identification, scaling,
estimation, global and local evaluation, and analytic history remain
necessary. A favorable CFA is not automatic permission to interpret
every later path causally. Nor does adding structural arrows repair a
poor measurement model. The first four chapters establish the
measurement foundation on which those later restrictions will depend.

# Summary and reading bridge

CFA evaluation is not a vote among fit-index cutoffs. The model test
addresses exact covariance restrictions; RMSEA describes approximation
per df with uncertainty; CFI and TLI use a baseline; SRMR summarizes
standardized covariance residuals. Local patterns and parameter
meaning remain necessary after those summaries are read.

Planned comparisons require exact nesting and compatible likelihoods.
Information criteria require comparable observations and outcomes.
Score, Wald, and LR tests take different views of a restriction and
need not agree numerically. Modification indices are local
approximations whose interpretation changes when selected from a
search.

The SLE record supports retaining M1-cross as a working measurement
model within the synthetic course example. A tentative B4 cross-loading
improved its development record but failed to receive support in the
untouched retest. Reliability calculations remain conditional on the
model and weighting rule; factor scores remain uncertain predictions.

The fitted three-factor higher-order form is an admissible
reparameterization of M1-cross, so identical fit cannot choose between
their interpretations. The orthogonal bifactor model is a nonnested rival:
it fits reasonably but less well by AIC/BIC here and contains a weak negative
specific loading. Omega-hierarchical describes the declared all-item score's
general-factor variance; it does not certify unidimensionality. Finally, the
known-population audit reveals a nonzero omitted B1--B2 residual relation,
loading displacement, and population discrepancy even though finite-sample
fit formulas applied at $N=360$ look perfect.

For further study, return to @bollen1989 and @brown2006 for the
measurement-model framework, and @maccallum1996 for the connection
between fit uncertainty and power. Read software tutorials alongside
their fitted-object definitions, not as authority for universal
cutoffs. The next topic is how structural restrictions change the
model-implied latent means and covariances while retaining the
measurement system.

# Exercises {.unnumbered}

1. **Foundational -- reconstruct ML.** For
   $\mathbf S=\begin{bmatrix}1&.5\\.5&1\end{bmatrix}$ and
   $\boldsymbol\Sigma=\begin{bmatrix}1&.3\\.3&1\end{bmatrix}$,
   calculate $F_{\mathrm{ML}}$ and $NF_{\mathrm{ML}}$ for $N=200$.
   Explain why a chi-square reference still requires the model's df
   and regularity conditions, not only this arithmetic.
2. **Foundational -- RMSEA and uncertainty.** Compute RMSEA under the
   $N$ convention for $T=60$, $df=40$, $N=400$, and for $T=35$ with
   the same df and $N$. Explain why the second result does not imply
   zero residuals. What does a wide 90% interval add to interpretation?
3. **Foundational -- baseline dependence.** With $T_M=60$, $df_M=40$,
   compare CFI when $(T_B,df_B)=(500,60)$ versus $(100,60)$.
   Explain why the same target-model statistic can yield different
   incremental fit and why neither result establishes construct truth.
4. **Foundational -- distinguish residual scales.** Let observed
   variances be 4 and 9, fitted variances 5 and 8, observed covariance
   2, and fitted covariance 1.5. Calculate the raw residual, the
   observed-SD-standardized covariance residual, and the correlation
   residual. Explain why they are not all the same number.
5. **Foundational -- establish nesting.** Show the exact restrictions
   producing D1, D2, and D4 from D3. Use the printed statistics to
   calculate D4 versus D3's LR difference and df. Explain why D1 and
   D2 cannot be compared by the same ordinary subtraction test.
6. **Foundational -- likelihood eligibility.** Recompute M1's AIC and
   BIC from its log likelihood and free-parameter count. Explain why
   comparison with M4 is eligible here but comparison with the donor
   microcase, Sample C, or a ten-item short form is not.
7. **Foundational -- define the composite.** For loadings $.7,.6,.5$
   on a unit-variance factor and residual variances $.51,.64,.75$,
   compute the common-variance proportion of a unit-weighted sum.
   Recompute it with residual covariance .08 between the first two
   indicators. State every assumption needed for the calculation.
8. **Integrative -- interpret disagreement.** Compare M2's CFI and
   SRMR, and M3's global profile and equality test. Explain the
   distinct questions rather than selecting one statistic to override
   the others.
9. **Integrative -- evaluate the tentative revision.** Combine B4's
   wording, residual pattern, MI/EPC, B refit, and C retest. Explain
   the decision to retain M1 without claiming that C proves the added
   loading is exactly zero. What would be lost by omitting R1 from
   the report?
10. **Integrative -- compare three tests.** Use D2/D3 to explain where
    the score, Wald, and LR statistics obtain their information. Why
    do they differ numerically, and why should an equality-constrained
    model not be diagnosed solely with ordinary zero-parameter MIs?
11. **Integrative -- scores are not measurements without error.**
    Compare a raw Planning mean with a regression factor score.
    Describe their weights, model dependence, scale, and uncertainty.
    Explain the risk of treating exported factor scores as error-free
    predictors in a later regression.
12. **Integrative -- write the report and next step.** Draft an
    evaluation paragraph that includes the primary model, sample roles,
    estimation family, global/local evidence, planned comparisons,
    unsupported revision, and reliability scope. Then explain how
    adding $\mathbf B$ changes the meanings of latent intercepts,
    latent means, total latent covariance, and disturbance covariance.
13. **Integrative -- invert three factor correlations.** Starting from
    the fitted Sample B correlations in Section 6, recover the three
    standardized higher-order loadings and disturbance variances. State
    the sign, denominator, and admissibility conditions. Explain why the
    higher-order form has the same df and fit as M1-cross, and why a
    four-first-order-factor version would impose two restrictions.
14. **Integrative -- compare a bifactor rival.** Count the 36 free
    covariance parameters and 42 df in the displayed bifactor model.
    Explain why it is not nested in M1-cross, interpret the weak
    Strategy-specific S4 loading, and compare their fit profiles without
    claiming that a general factor has been proved.
15. **Integrative -- read the population audit.** Use the displayed
    tetrad and loading-displacement table to identify the omitted
    restriction. Explain why the population RMSEA is nonzero while the
    finite-$N$ sample formulas applied to the population covariance return
    RMSEA .000 and CFI 1.000, and why neither result is a specification
    certificate.

# Appendix: Interval and score calculations {.unnumbered}

## Inverting the noncentral chi-square reference {.unnumbered}

Let $G(t;d,\delta)$ be the noncentral chi-square CDF with df $d$
and noncentrality $\delta$. For a two-sided 90% interval, solve

$$
G(T_M;df_M,\delta_L)=.95,
\qquad
G(T_M;df_M,\delta_U)=.05,
$$

with a lower boundary of zero when an interior solution does not
exist. Transform the limits using
$\sqrt{\delta/(Ndf_M)}$ under this chapter's convention. The
shared numerical helper brackets the roots and reproduces the package
intervals for every displayed model. The numerical inversion is
optional; interpreting why a point estimate needs uncertainty is core.

## Regression and Bartlett weights {.unnumbered}

With a jointly normal model, zero latent mean, and fitted parameters
temporarily treated as known, the regression weights are

$$
\mathbf A_R=\boldsymbol\Phi\boldsymbol\Lambda^{\mathsf T}
\boldsymbol\Sigma^{-1},
$$

and the conditional covariance is

$$
\operatorname{Var}(\boldsymbol\eta\mid\mathbf y)
=\boldsymbol\Phi-\boldsymbol\Phi\boldsymbol\Lambda^{\mathsf T}
\boldsymbol\Sigma^{-1}\boldsymbol\Lambda\boldsymbol\Phi.
$$

For nonsingular residual covariance and a full-column-rank loading
matrix, Bartlett weights are

$$
\mathbf A_B=
(\boldsymbol\Lambda^{\mathsf T}\boldsymbol\Theta^{-1}
\boldsymbol\Lambda)^{-1}
\boldsymbol\Lambda^{\mathsf T}\boldsymbol\Theta^{-1}.
$$

These formulas clarify why scoring depends on more than selecting the
largest loadings. With nonzero latent means, add the appropriate
latent mean and center indicators at their implied means. Parameter
uncertainty and model uncertainty require additional treatment; the
conditional covariance above does not include them.

For a factor $a$, model-based regression-score determinacy is the
correlation between the latent factor and its regression prediction.
Under the stated model its square is
$1-\operatorname{Var}(\eta_a\mid\mathbf y)/\phi_{aa}$.
High determinacy is not proof of a correct measurement model, and
it does not eliminate prediction error for individuals.

## Coefficient H under a restrictive one-factor model {.unnumbered}

For a single unit-variance factor with diagonal positive residual
covariance, let
$a=\boldsymbol\lambda^{\mathsf T}\boldsymbol\Theta^{-1}
\boldsymbol\lambda$. The maximum common-variance proportion among
linear composites is

$$
H=\frac{a}{1+a}.
$$

With fully standardized simple indicators,
$a=\sum_j\lambda_j^2/(1-\lambda_j^2)$.
This is an optimal-weighting result under specific assumptions,
not reliability of every unit-weighted score and not a general
formula for a multifactor cross-loading subscale. It should not
be attached to the SLE factors by ignoring their other loadings.

# References {.unnumbered}

::: {#refs}
:::
