---
title: "From Latent-Variable Modeling to Factor Analysis"
author: "Jinsong Chen"

bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    output-file: chapter_01.pdf
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-cap: true
    tbl-cap: true
    fig-align: center
    include-in-header:
      text: |
        \setlength{\parindent}{0pt}
        \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
        \ifdefined\pdftrailerid\pdftrailerid{}\fi
        \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
        \ifdefined\pdfvariable
          \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
        \fi
        \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
        \addtokomafont{sectionentry}{\scriptsize}
        \let\chapteroldtableofcontents\tableofcontents
        \renewcommand{\tableofcontents}{\begingroup\scriptsize
          \setlength{\parskip}{0pt}\chapteroldtableofcontents\endgroup}
        \usepackage{caption}
        \captionsetup[table]{
          position=above, format=plain, justification=raggedright,
          singlelinecheck=false, labelsep=newline, labelfont=bf, textfont=it
        }
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{float}
        \usepackage{needspace}
        \usepackage{etoolbox}
        \AtBeginEnvironment{CSLReferences}{\small\setlength{\parskip}{1pt}}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center",
  dev = if (knitr::is_latex_output()) "cairo_pdf" else "png"
)
options(digits = 3, width = 72, scipen = 4)

# This source is deliberately self-contained. The constructor below checks the
# known-value records used in the prose, and the drawing function produces all
# five path diagrams with base R.
chapter_examples <- local({
  implied_cov <- function(Lambda, Phi, Theta) {
    Lambda %*% Phi %*% t(Lambda) + Theta
  }

  Lambda_one <- matrix(.70, nrow = 6, ncol = 1)
  Phi_one <- matrix(1, 1, 1)
  Theta_one <- diag(.51, 6)
  R_one <- implied_cov(Lambda_one, Phi_one, Theta_one)

  Lambda_two <- rbind(
    c(.80, 0), c(.80, 0), c(.80, 0),
    c(0, .60), c(0, .60), c(0, .60)
  )
  Phi_two <- diag(2)
  Theta_two <- diag(c(rep(.36, 3), rep(.64, 3)))
  R_two <- implied_cov(Lambda_two, Phi_two, Theta_two)

  Q <- matrix(c(1, 1, -1, 1), 2, 2) / sqrt(2)
  Lambda_rotated <- Lambda_two %*% Q
  R_rotated <- implied_cov(Lambda_rotated, Phi_two, Theta_two)

  stopifnot(
    max(abs(diag(R_one) - 1)) < 1e-12,
    max(abs(R_one[row(R_one) != col(R_one)] - .49)) < 1e-12,
    max(abs(R_two[1:3, 1:3][row(R_two[1:3, 1:3]) !=
                              col(R_two[1:3, 1:3])] - .64)) < 1e-12,
    max(abs(R_two[4:6, 4:6][row(R_two[4:6, 4:6]) !=
                              col(R_two[4:6, 4:6])] - .36)) < 1e-12,
    max(abs(R_two[1:3, 4:6])) < 1e-12,
    max(abs(R_rotated - R_two)) < 1e-12,
    max(abs(colSums(Lambda_two^2) - c(1.92, 1.08))) < 1e-12,
    max(abs(colSums(Lambda_rotated^2) - c(1.50, 1.50))) < 1e-12
  )

  list(
    Lambda_one = Lambda_one, Phi_one = Phi_one, Theta_one = Theta_one,
    R_one = R_one, Lambda_two = Lambda_two, Phi_two = Phi_two,
    Theta_two = Theta_two, R_two = R_two, Q = Q,
    Lambda_rotated = Lambda_rotated
  )
})

# Shared Part Two path-diagram grammar. Rectangles are recorded
# variables, circles are variables introduced by the model, single-headed
# arrows are loadings and double-headed arrows are covariances. Drawn with
# base graphics so the chapter needs no external script or renderer.

pt2_ink        <- "#243746"
pt2_box_fill   <- "#fff1c7"
pt2_circ_fill  <- "#dcecf5"
pt2_path_col   <- "#2c5d77"
pt2_cross_col  <- "#94a3ad"
pt2_cov_col    <- "#8c4f2b"
pt2_resid_col  <- "#596b75"

# Render a label that may be plain text (".70", "0") or a plotmath
# expression string ("lambda[11]", "eta[1]").
.pt2_label <- function(x) {
  if (is.expression(x) || is.language(x)) return(x)
  if (!nzchar(x)) return("")
  if (grepl("^[-.0-9]", x)) return(x)
  # Plotmath where the label parses (eta[1], phi[12]); literal text otherwise,
  # so that factor names such as "School support" render as written.
  out <- tryCatch(parse(text = x), error = function(e) NULL)
  if (is.null(out)) x else out
}

.pt2_arrow <- function(x0, y0, x1, y1, dashed = FALSE, both = FALSE,
                       col = pt2_path_col, lwd = 1.35) {
  arrows(x0, y0, x1, y1, length = .07, angle = 20,
         code = if (both) 3 else 2,
         lty = if (dashed) 2 else 1, lwd = lwd, col = col)
}

.pt2_box <- function(x, y, label, hw = .74, hh = .30) {
  rect(x - hw, y - hh, x + hw, y + hh,
       col = pt2_box_fill, border = pt2_ink, lwd = 1.35)
  text(x, y, .pt2_label(label), cex = .92)
}

.pt2_circle <- function(x, y, label, r = .55) {
  symbols(x, y, circles = r, inches = FALSE, add = TRUE,
          bg = pt2_circ_fill, fg = pt2_ink, lwd = 1.35)
  text(x, y, .pt2_label(label), cex = .95)
}

# Draw a label at a fraction along a path, nudged perpendicular to it.
.pt2_path_label <- function(x0, y0, x1, y1, label, at = .58, nudge = .15,
                            col = pt2_ink, cex = .70) {
  if (is.null(label) || !nzchar(label)) return(invisible(NULL))
  text(x0 + at * (x1 - x0), y0 + at * (y1 - y0) + nudge,
       .pt2_label(label), cex = cex, col = col)
}

# A double-headed covariance arc drawn to the left of two circles.
.pt2_cov_arc <- function(x, y1, y2, label, r = .55, bulge = .85) {
  tt <- seq(0, 1, length.out = 60)
  ym <- (y1 + y2) / 2
  half <- abs(y2 - y1) / 2 - r * .55
  ys <- ym + half * cos(pi * tt) * sign(y1 - y2)
  xs <- x - bulge * sin(pi * tt)
  lines(xs, ys, col = pt2_cov_col, lwd = 1.4)
  .pt2_arrow(xs[2], ys[2], xs[1], ys[1], col = pt2_cov_col, lwd = 1.4)
  n <- length(xs)
  .pt2_arrow(xs[n - 1], ys[n - 1], xs[n], ys[n], col = pt2_cov_col, lwd = 1.4)
  if (!is.null(label) && nzchar(label))
    text(x - bulge - .14, ym, .pt2_label(label), adj = 1, cex = .82)
  invisible(NULL)
}

.pt2_canvas <- function(xlim, ylim) {
  op <- par(mar = rep(.15, 4), xaxs = "i", yaxs = "i")
  plot.new()
  plot.window(xlim = xlim, ylim = ylim)
  op
}

.pt2_indicator_y <- function(p, top = 7.6, bottom = .9) {
  if (p == 1) return(mean(c(top, bottom)))
  seq(top, bottom, length.out = p)
}

# ---------------------------------------------------------------------------
# First-order model with m correlated (or orthogonal) factors.
#
#   assign  integer vector of length p giving each indicator's primary factor,
#           or a p x m logical matrix for a fully general pattern.
#   cross   optional p x m logical matrix of additional (dashed) loadings.
# ---------------------------------------------------------------------------
pt2_fa_diagram <- function(assign, m = NULL, cross = NULL,
                           loading_labels = NULL, factor_labels = NULL,
                           indicator_labels = NULL, factor_cov_labels = NULL,
                           show_residuals = TRUE, x_factor = 2.2,
                           x_indicator = 7.0) {
  if (is.matrix(assign)) {
    primary <- assign
    p <- nrow(primary); m <- ncol(primary)
  } else {
    p <- length(assign); if (is.null(m)) m <- max(assign)
    primary <- matrix(FALSE, p, m)
    primary[cbind(seq_len(p), assign)] <- TRUE
  }
  if (is.null(cross)) cross <- matrix(FALSE, p, m)
  if (is.null(factor_labels))
    factor_labels <- paste0("eta[", seq_len(m), "]")
  if (is.null(indicator_labels))
    indicator_labels <- paste0("Y[", seq_len(p), "]")

  yi <- .pt2_indicator_y(p)
  yf <- if (m == 1) mean(range(yi)) else
    seq(max(yi) - .9, min(yi) + .9, length.out = m)
  x_resid <- x_indicator + 3.0
  op <- .pt2_canvas(c(0, x_resid + 1.0), c(.3, 8.2))
  on.exit(par(op))

  # factor covariances: adjacent pairs, arcs on the left
  if (m > 1 && !is.null(factor_cov_labels)) {
    if (length(factor_cov_labels) == 1L && m > 2)
      factor_cov_labels <- rep(factor_cov_labels, m - 1)
    for (k in seq_len(m - 1))
      .pt2_cov_arc(x_factor - .55, yf[k], yf[k + 1],
                   factor_cov_labels[k],
                   bulge = .70 + .30 * (k %% 2))
  }

  for (j in seq_len(p)) for (k in seq_len(m)) {
    is_cross <- isTRUE(cross[j, k])
    if (!isTRUE(primary[j, k]) && !is_cross) next
    .pt2_arrow(x_factor + .55, yf[k], x_indicator - .74, yi[j],
               dashed = is_cross,
               col = if (is_cross) pt2_cross_col else pt2_path_col,
               lwd = if (is_cross) 1.1 else 1.35)
    if (!is.null(loading_labels))
      .pt2_path_label(x_factor + .55, yf[k], x_indicator - .74, yi[j],
                      loading_labels[j, k], nudge = if (is_cross) -.16 else .16,
                      col = if (is_cross) pt2_cross_col else pt2_ink)
  }

  for (j in seq_len(p)) {
    .pt2_box(x_indicator, yi[j], indicator_labels[j])
    if (show_residuals) {
      .pt2_arrow(x_resid - .52, yi[j], x_indicator + .74, yi[j],
                 col = pt2_resid_col, lwd = 1.15)
      text(x_resid, yi[j], .pt2_label(paste0("epsilon[", j, "]")), cex = .85)
    }
  }
  for (k in seq_len(m)) .pt2_circle(x_factor, yf[k], factor_labels[k])
  invisible(NULL)
}

# ---------------------------------------------------------------------------
# Higher-order model: one second-order factor above m first-order factors.
# ---------------------------------------------------------------------------
pt2_higher_order_diagram <- function(assign, m = NULL, gamma_labels = NULL,
                                     loading_labels = NULL,
                                     factor_labels = NULL,
                                     indicator_labels = NULL,
                                     higher_label = "xi",
                                     show_disturbances = TRUE) {
  p <- length(assign); if (is.null(m)) m <- max(assign)
  if (is.null(factor_labels)) factor_labels <- paste0("eta[", seq_len(m), "]")
  if (is.null(indicator_labels)) indicator_labels <- paste0("Y[", seq_len(p), "]")
  x_high <- 1.0; x_factor <- 4.0; x_indicator <- 8.0; x_resid <- 10.8

  yi <- .pt2_indicator_y(p)
  yf <- seq(max(yi) - .9, min(yi) + .9, length.out = m)
  yh <- mean(range(yf))
  op <- .pt2_canvas(c(0, x_resid + 1.0), c(.3, 8.2))
  on.exit(par(op))

  for (k in seq_len(m)) {
    .pt2_arrow(x_high + .55, yh, x_factor - .55, yf[k])
    if (!is.null(gamma_labels))
      .pt2_path_label(x_high + .55, yh, x_factor - .55, yf[k],
                      gamma_labels[k], at = .55, nudge = .17)
    if (show_disturbances) {
      .pt2_arrow(x_factor, yf[k] + 1.05, x_factor, yf[k] + .55,
                 col = pt2_resid_col, lwd = 1.15)
      text(x_factor, yf[k] + 1.24,
           .pt2_label(paste0("zeta[", k, "]")), cex = .82)
    }
  }
  for (j in seq_len(p)) {
    k <- assign[j]
    .pt2_arrow(x_factor + .55, yf[k], x_indicator - .74, yi[j])
    if (!is.null(loading_labels))
      .pt2_path_label(x_factor + .55, yf[k], x_indicator - .74, yi[j],
                      loading_labels[j], nudge = .16)
    .pt2_box(x_indicator, yi[j], indicator_labels[j])
    .pt2_arrow(x_resid - .52, yi[j], x_indicator + .74, yi[j],
               col = pt2_resid_col, lwd = 1.15)
    text(x_resid, yi[j], .pt2_label(paste0("epsilon[", j, "]")), cex = .85)
  }
  for (k in seq_len(m)) .pt2_circle(x_factor, yf[k], factor_labels[k])
  .pt2_circle(x_high, yh, higher_label)
  invisible(NULL)
}

# ---------------------------------------------------------------------------
# Bifactor model: one general factor on every indicator, plus orthogonal
# group factors on their own blocks.
# ---------------------------------------------------------------------------
pt2_bifactor_diagram <- function(assign, m = NULL, general_label = "eta[G]",
                                 group_labels = NULL, indicator_labels = NULL,
                                 general_loadings = NULL,
                                 group_loadings = NULL) {
  p <- length(assign); if (is.null(m)) m <- max(assign)
  if (is.null(group_labels))
    group_labels <- paste0("eta[", seq_len(m), "]")
  if (is.null(indicator_labels)) indicator_labels <- paste0("Y[", seq_len(p), "]")
  x_gen <- 1.0; x_group <- 4.2; x_indicator <- 8.0; x_resid <- 10.8

  yi <- .pt2_indicator_y(p)
  yg <- seq(max(yi) - .8, min(yi) + .8, length.out = m)
  ygen <- mean(range(yi))
  op <- .pt2_canvas(c(0, x_resid + 1.0), c(.3, 8.2))
  on.exit(par(op))

  # general factor reaches every indicator; drawn first and lighter so the
  # group structure stays readable
  for (j in seq_len(p)) {
    .pt2_arrow(x_gen + .55, ygen, x_indicator - .74, yi[j],
               col = pt2_cross_col, lwd = 1.0)
    if (!is.null(general_loadings))
      .pt2_path_label(x_gen + .55, ygen, x_indicator - .74, yi[j],
                      general_loadings[j], at = .30, nudge = .15,
                      col = pt2_cross_col, cex = .64)
  }
  for (j in seq_len(p)) {
    k <- assign[j]
    .pt2_arrow(x_group + .55, yg[k], x_indicator - .74, yi[j])
    if (!is.null(group_loadings))
      .pt2_path_label(x_group + .55, yg[k], x_indicator - .74, yi[j],
                      group_loadings[j], nudge = .16)
    .pt2_box(x_indicator, yi[j], indicator_labels[j])
    .pt2_arrow(x_resid - .52, yi[j], x_indicator + .74, yi[j],
               col = pt2_resid_col, lwd = 1.15)
    text(x_resid, yi[j], .pt2_label(paste0("epsilon[", j, "]")), cex = .85)
  }
  for (k in seq_len(m)) .pt2_circle(x_group, yg[k], group_labels[k])
  .pt2_circle(x_gen, ygen, general_label)
  text(x_gen, ygen - 1.15, "all factors orthogonal", cex = .68,
       col = pt2_resid_col)
  invisible(NULL)
}

```

{{< pagebreak >}}

::: {.callout-tip title="Chapter box"}
**Five themes.**

*A latent variable is introduced, not found.* It is an unobserved random
variable in a model, distinct from the construct that names it, from a
composite of the items, and from any score computed for a person.

*One model, four representations.* The common-factor model can be written
as scalar equations, as matrices, as a path diagram, and as a sentence.
The four must agree, and moving between them is how a specification is
checked.

*Latent coordinates carry no units.* A factor has no inherited scale, sign,
or location. Conventions supply them, and several coordinate systems
describe the same observed covariance equally well.

*Covariance splits into two parts.* The model implies
$\boldsymbol{\Sigma}=\boldsymbol{\Lambda}\boldsymbol{\Phi}
\boldsymbol{\Lambda}^{\mathsf T}+\boldsymbol{\Theta}$: a common part
carried by the factors and a residual part. Communality and uniqueness are
conditional on the declared model, and a residual is not the same thing as
measurement error.

*Restrictions make a model refutable.* A model with fewer free parameters
than distinct moments asserts something about those moments that could turn
out false. Exact reproduction shows compatibility; it does not establish
that the factors are the attributes they are named for.

**Prerequisites.** Variance, covariance, correlation, random variables, and
linear regression notation. Part One Chapter 1 is an optional reference for
the collection's general reading method; no argument framework vocabulary is
required for the core chapter.

**Scope.** We formulate, translate and interrogate a common-factor model
using exact teaching records. No data are estimated and no learner-facing
software is used. Indicators are treated as continuous throughout Part Two;
methods for ordered-categorical indicators belong elsewhere and receive one
short scope note.
:::

# Latent variables and what they model {#sec-lvm}

## Observed responses, an unobserved account {#sec-what-is-latent}

**Guiding question.** When several indicators covary, when is a latent
factor a useful statistical account, and when may it be interpreted as the
construct named in a study?

**Provisional answer.** A common-factor model can represent their covariance
through unobserved factors and residuals. Compatibility with that covariance
supports the statistical representation. Interpreting a factor as a construct
additionally requires a defensible account of the indicator domain, response
process, population, and intended use. Naming a factor, obtaining a simple
pattern, or even reproducing covariance exactly is not enough.

Behavioral and social research often concerns attributes that are not
recorded directly. A dataset may contain responses to several planning
items, but no column contains planning ability itself. A test records
item scores, not error-free proficiency. Symptoms may be observed while
the syndrome proposed to organize them remains a theoretical and
statistical construction.

A latent-variable model does not turn the proposed attribute into an
observation. Instead, it introduces one or more unobserved random
variables and states how they are related to the variables that were
recorded. The useful question is therefore not simply whether a factor
can be named. It is whether the proposed model has observable
consequences, and what conclusions those consequences support.

Answering that pair takes both halves: a covariance representation that can
be checked, and a conclusion that stays within what such a check can
establish. The first task is to distinguish the quantities that enter the
representation.

@tbl-quantities keeps four kinds of quantity apart.

| Quantity | Status | Example |
|:--|:--|:--|
| **Observed variable** | Recorded for each unit | An item response $Y_j$ |
| **Latent random variable** | Introduced by the model; not recorded | A common factor $\eta_k$ |
| **Unknown fixed parameter** | A population quantity to be estimated | A loading $\lambda_{jk}$ |
| **Residual random variable** | The part of an indicator the declared factors do not represent | $\epsilon_j$ |

: *Four kinds of quantity in a factor model.* {#tbl-quantities}

A factor is not merely a parameter; it is a random variable in the
model, with a distribution over units. A loading is not a hidden score
attached to a person; it is a parameter describing how the model relates
an indicator to a factor. A residual is unobserved too, but that does
not make it a substantive construct: it collects whatever the declared
common factors leave out.

@fig-lvm shows the smallest interesting case. Three observed indicators
are drawn as rectangles, one unobserved variable as a circle, and the
single-headed arrows run from the unobserved variable to the observed
ones.

```{r}
#| label: fig-lvm
#| fig-cap: "A latent variable with three indicators. Rectangles are recorded variables, the circle is a variable introduced by the model, and single-headed arrows denote loading relations."
#| fig-alt: "One latent factor eta points to observed indicators Y1 through Y3; each indicator also receives a residual epsilon arrow."
#| fig-width: 7
#| fig-height: 3.2
#| out-width: "58%"

pt2_fa_diagram(
  rep(1L, 3),
  loading_labels = matrix(c("lambda[1]", "lambda[2]", "lambda[3]"), 3, 1)
)
```

For three indicators, the picture summarizes equations of the form

$$
Y_{ij} = \nu_j + \lambda_j\eta_i + \epsilon_{ij},
\qquad j = 1,2,3 .
$$

The indicator $Y_{ij}$ is observed. The factor $\eta_i$ and residual
$\epsilon_{ij}$ are random variables introduced by the model. The
intercept $\nu_j$ and loading $\lambda_j$ are fixed but unknown
population parameters.

The arrows deserve care. They state a *direction of representation*: the
model writes each indicator as a function of the factor plus a residual,
not the reverse. That is a modeling choice with real consequences, and
it is the choice this book makes throughout. It is not, by itself,
evidence that the factor causes the responses. A diagram is a compact
statement of a model; establishing that a latent attribute produces
observed behaviour requires theory about the response process and a
design capable of supporting the claim. Model fit does not supply either.

Two inferential links are easy to conflate. First, relations among
predictors, exposures and outcomes require a design capable of supporting
causal interpretation; sampling units from a target population bears mainly
on population generalization. Second, relating a proposed construct to a set
of indicators requires defensible coverage of the construct's content or
behavioural domain. This second task is often called **domain sampling**,
although indicator selection is commonly systematic and theory-guided rather
than a literal probability sample.

Domain coverage supports a broad construct interpretation, but it is not
sufficient by itself: the response process, dimensionality, population and
intended use still matter. Without such a warrant, a factor may remain useful
as a statistical device - for example, as a method factor - without being
interpreted as the construct itself. Likewise, drawing an arrow from a factor
to an indicator states the model's reflective direction; neither domain
coverage nor the arrow alone establishes a causal effect.

Why introduce something that is not observed? Because the model has
consequences for quantities that are. Under assumptions developed below,
two indicators connected to the same factor have the common covariance
contribution

$$
\lambda_j\operatorname{Var}(\eta)\lambda_k .
$$

When their residual covariance is zero, this is their entire model-implied
covariance. @sec-implied derives the complete result rather than asking
the reader to accept this preview.

Only after the model and its terms are clear will we ask it to reproduce
particular numerical patterns. The two six-indicator matrices in
@sec-records are constructed, known-value teaching records. They are not
sample correlations or evidence that particular constructs exist.

## Classical test theory and the common-factor model {#sec-ctt}

Readers arriving from measurement courses will know classical test
theory, which decomposes an observed score into a true score and an
error:

$$
X_{ir} = T_i + E_{ir},
\qquad
T_i = E_r\!\left(X_{ir} \mid i\right),
\qquad
E_r\!\left(E_{ir} \mid i\right) = 0 .
$$

Here $X_{ir}$ is the score of person $i$ on a hypothetical repetition
$r$ of a specified measurement procedure, $T_i$ is its expected value
over those repetitions, and $E_{ir}$ is the deviation. The expectation
holds the person fixed and averages over measurement conditions; it does
not average over people. The true score is therefore a property of a
person *and a procedure*, not automatically a pure measure of the
intended attribute: an influence that is stable across repetitions stays
inside $T_i$ whether or not it belongs to the construct
[@livingston2018].

The two frameworks overlap, but neither is simply a relabeling of the
other. Under additional assumptions, a classical true-score
decomposition can be represented as a one-factor model by identifying
the common factor with true-score variation and fixing the relevant
loadings at one. A common-factor model need not impose that restriction:
it can let indicators relate to one or more factors with different
strengths and can examine the resulting covariance restrictions. The
connection is useful precisely because its assumptions must be stated;
classical test theory does not in general assert that every set of item
responses follows a one-factor model.

That is a difference in scope, not a ranking. Classical test theory
answers questions about a score and its replication conditions with very
few assumptions, and it remains the right tool for many of them.
Common-factor models, item-response models and the neighbouring
families add structure to the joint distribution of responses, and buy
item-level and multidimensional statements at the price of stronger
assumptions and larger samples. A residual in a factor model is also not
automatically classical error: stable item-specific content sits in the
residual without being error under the declared repetition procedure.
Reliability connects the two decompositions and needs conditions from
both; it is not derivable from a loading pattern alone.

## A map of latent-variable models {#sec-map}

@tbl-map crosses the form of the latent variable with the form of the
observed indicators to give a serviceable orientation.

| Latent variable | Observed indicators | Common model families |
|:--|:--|:--|
| Continuous | Continuous | Common-factor models, confirmatory factor analysis, many structural equation models |
| Continuous | Binary or ordered | Item-response models, categorical factor models |
| Categorical | Continuous | Latent-profile and finite-mixture models |
| Categorical | Categorical | Latent-class models, cognitive diagnosis models |

: *A qualified map of latent-variable model families.* {#tbl-map}

The map is an orientation, not a set of sealed boxes. Factor models can
be fitted to ordered indicators through underlying-response
formulations; item-response theory and categorical factor analysis share
deep mathematical connections while using different parameterizations
and traditions; mixture models combine continuous and categorical latent
structure. Structural equation modeling is broader still, and can
contain a factor measurement portion, regressions among variables, or
both.

This chapter takes the first row: continuous factors and continuous
indicators. Part Two as a whole stays there. That narrowing is what
allows the algebra to be seen plainly before estimation choices and
software output accumulate. Responses are treated as continuous throughout
Part Two.

## Exploratory and confirmatory orientations {#sec-orientations}

The same common-factor equation can be used with very different amounts
declared in advance.

An **exploratory** analysis ordinarily lets every indicator relate to
every retained factor and uses rotation or equivalent constraints to
choose interpretable coordinates. It suits situations where a construct
exists conceptually but its dimensionality is unsettled. A
**confirmatory** analysis fixes or constrains loadings and residual
relations according to a specification stated before the focal analysis,
and asks what follows.

The words describe orientations, not two unrelated models. Both use the
same covariance decomposition. Theory can inform an exploratory
analysis, and a confirmatory model can be revised after data are
inspected. What separates them is the *restrictions* and the *history of
the decisions* — not the name of the procedure. Practice usually falls
somewhere along the continuum rather than at either end. @sec-approaches
returns to the distinction once the shared model is in hand.

# The common-factor model {#sec-model}

## One indicator at a time {#sec-scalar}

Write $y_{ij}$ for the response of unit $i$ on indicator $j$. The
common-factor model expresses it as

$$
y_{ij} = \nu_j + \boldsymbol{\lambda}_j^{\mathsf T}\boldsymbol{\eta}_i
         + \epsilon_{ij},
$$ {#eq-scalar}

where $i = 1,\ldots,n$ indexes units, $j = 1,\ldots,p$ indexes
indicators, $\nu_j$ is an intercept, $\boldsymbol{\lambda}_j$ is an
$m \times 1$ vector of loadings, $\boldsymbol{\eta}_i$ is the
$m \times 1$ vector of common factors for unit $i$, and $\epsilon_{ij}$
is the residual. With two factors this reads

$$
y_{ij} = \nu_j + \lambda_{j1}\eta_{i1} + \lambda_{j2}\eta_{i2}
         + \epsilon_{ij} .
$$

The resemblance to multiple regression is useful and has one decisive
limit: the predictors $\eta_{i1}$ and $\eta_{i2}$ are not columns in the
dataset. Their scale and orientation must be supplied by model
constraints, and the parameters are learned from the joint pattern among
all $p$ indicators rather than from a single outcome at a time.

## From equations to matrices {#sec-matrix}

Collecting the $p$ indicators for unit $i$ into $\mathbf{y}_i$,

$$
\mathbf{y}_i = \boldsymbol{\nu} + \boldsymbol{\Lambda}\boldsymbol{\eta}_i
               + \boldsymbol{\epsilon}_i ,
$$ {#eq-matrix}

with $\mathbf{y}_i$, $\boldsymbol{\nu}$ and $\boldsymbol{\epsilon}_i$ in
$\mathbb{R}^{p}$, $\boldsymbol{\eta}_i$ in $\mathbb{R}^{m}$, and
$\boldsymbol{\Lambda}$ a $p \times m$ loading matrix. The multiplication
conforms because a $p \times m$ matrix times an $m \times 1$ vector
returns a $p \times 1$ vector: one common-factor contribution for each
indicator. Row $j$ of $\boldsymbol{\Lambda}$ is
$\boldsymbol{\lambda}_j^{\mathsf T}$.

The model's moment assumptions are

$$
\begin{aligned}
E(\boldsymbol{\eta}_i) &= \boldsymbol{\alpha}, &
\operatorname{Var}(\boldsymbol{\eta}_i) &= \boldsymbol{\Phi}, \\
E(\boldsymbol{\epsilon}_i) &= \mathbf{0}, &
\operatorname{Var}(\boldsymbol{\epsilon}_i) &= \boldsymbol{\Theta}, \\
\operatorname{Cov}(\boldsymbol{\eta}_i, \boldsymbol{\epsilon}_i)
  &= \mathbf{0}. & &
\end{aligned}
$$ {#eq-assumptions}

$\boldsymbol{\Phi}$ is the $m \times m$ matrix of factor variances and
covariances; $\boldsymbol{\Theta}$ is the $p \times p$ matrix of
residual variances and any residual covariances the model permits. Both
are symmetric and positive semidefinite. The zero matrix in the last
line is $m \times p$.

The notation $\boldsymbol{\alpha}$ is chosen deliberately. It is the
latent intercept vector. In this chapter there are no regressions among
latent variables, so $\boldsymbol{\alpha}=E(\boldsymbol{\eta}_i)$. Once
a structural coefficient matrix $\mathbf B$ is introduced later, the
marginal mean becomes
$E(\boldsymbol{\eta}_i)=(\mathbf I-\mathbf B)^{-1}
\boldsymbol{\alpha}$. We will therefore call $\boldsymbol{\alpha}$ the
factor mean only in the present measurement-only setting.

Each assumption does a specific job, and it is worth naming them
separately rather than accepting them as a block. The factor–residual
condition is what will let the covariance decomposition split cleanly
into two parts in @sec-implied. The mean conditions are what let the
model say anything about $E(\mathbf{y}_i)$. Note also what has *not*
been assumed: zero covariance between factors and residuals is a moment
restriction, not a statement that they are statistically independent,
and $\boldsymbol{\Theta}$ has not been required to be diagonal.

## What a residual contains {#sec-residual}

It is tempting to read every $\epsilon_{ij}$ as measurement error. That
is too narrow. Residual variation can include random response
fluctuation, but also stable item-specific content, wording or method
effects, occasion effects, an omitted common factor, or a
misspecification of the model. A residual *covariance* $\theta_{jk}$
represents a relation between two indicators that the declared common
factors do not account for. Whether such a relation is substantively
defensible is a modeling question, not something a large diagnostic
statistic settles on its own.

A diagonal $\boldsymbol{\Theta}$ has a precise but limited meaning, and
three claims are worth keeping apart. Under the moment assumptions of
@eq-assumptions, a diagonal $\boldsymbol{\Theta}$ says that distinct
residual components have zero **unconditional** covariance. If we add
the conditional moment assumptions
$E(\boldsymbol{\epsilon}_i \mid \boldsymbol{\eta}_i) = \mathbf{0}$ and
$\operatorname{Var}(\boldsymbol{\epsilon}_i \mid \boldsymbol{\eta}_i)
= \boldsymbol{\Theta}$, it also gives zero **conditional** residual
covariance. Full **local independence** — the indicators being
conditionally independent given the factors — is stronger still, and
needs a distributional statement or an explicit factorization of the
conditional response distribution. Zero covariance is not independence:
two variables can have zero correlation and remain functionally related.
None of the three claims says that the observed items are unrelated;
that is the whole point of the model.

# One model, four representations {#sec-four-ways}

We now fix a single system and keep it in view for the rest of the
chapter: six observed indicators and two common factors. Allowing every
indicator to relate to both factors gives six equations,

$$
\begin{aligned}
y_{i1} &= \nu_1 + \lambda_{11}\eta_{i1} + \lambda_{12}\eta_{i2} + \epsilon_{i1},\\
y_{i2} &= \nu_2 + \lambda_{21}\eta_{i1} + \lambda_{22}\eta_{i2} + \epsilon_{i2},\\
y_{i3} &= \nu_3 + \lambda_{31}\eta_{i1} + \lambda_{32}\eta_{i2} + \epsilon_{i3},\\
y_{i4} &= \nu_4 + \lambda_{41}\eta_{i1} + \lambda_{42}\eta_{i2} + \epsilon_{i4},\\
y_{i5} &= \nu_5 + \lambda_{51}\eta_{i1} + \lambda_{52}\eta_{i2} + \epsilon_{i5},\\
y_{i6} &= \nu_6 + \lambda_{61}\eta_{i1} + \lambda_{62}\eta_{i2} + \epsilon_{i6},
\end{aligned}
$$ {#eq-six}

which collect into the loading matrix

$$
\boldsymbol{\Lambda} =
\begin{bmatrix}
\lambda_{11} & \lambda_{12}\\
\lambda_{21} & \lambda_{22}\\
\lambda_{31} & \lambda_{32}\\
\lambda_{41} & \lambda_{42}\\
\lambda_{51} & \lambda_{52}\\
\lambda_{61} & \lambda_{62}
\end{bmatrix},
\qquad
\boldsymbol{\Phi} =
\begin{bmatrix}
\phi_{11} & \phi_{12}\\
\phi_{12} & \phi_{22}
\end{bmatrix} .
$$ {#eq-general-loading}

Rows of $\boldsymbol{\Lambda}$ belong to indicators and columns to
factors, so $\lambda_{42}$ relates $Y_4$ to the second factor. Reading
the subscripts in that order is the quickest way to catch a path diagram
that disagrees with its matrix. The residual covariance matrix is
symmetric and $6 \times 6$; a diagonal working version is
$\boldsymbol{\Theta} = \operatorname{diag}(\theta_{11},\ldots,\theta_{66})$.

@fig-general draws the same system. Every indicator receives an arrow
from both factors, and the double-headed arrow between the factors is
the covariance $\phi_{12}$.

```{r}
#| label: fig-general
#| fig-cap: "The two-factor, six-indicator system with every loading present. Single-headed arrows denote loading relations; the double-headed arrow denotes the factor covariance."
#| fig-alt: "Both correlated factors point to all six indicators; cross-loading paths are dashed, and each indicator has a residual."
#| fig-width: 7
#| fig-height: 6.2
#| out-width: "88%"

cross_all <- matrix(TRUE, 6, 2)
cross_all[cbind(seq_len(6), rep(1:2, each = 3))] <- FALSE
pt2_fa_diagram(
  rep(1:2, each = 3), cross = cross_all,
  factor_cov_labels = "phi[12]"
)
```

## Restricting the pattern {#sec-simple}

Setting the cross-loadings to zero, so the first three indicators relate
only to $\eta_1$ and the last three only to $\eta_2$, gives

$$
\boldsymbol{\Lambda}_{\text{simple}} =
\begin{bmatrix}
\lambda_{11} & 0\\
\lambda_{21} & 0\\
\lambda_{31} & 0\\
0 & \lambda_{42}\\
0 & \lambda_{52}\\
0 & \lambda_{62}
\end{bmatrix} .
$$ {#eq-simple-loading}

@fig-simple redraws the same system after the cross-loadings are fixed at
zero.

```{r}
#| label: fig-simple
#| fig-cap: "The same system restricted to a simple structure. Each indicator has one loading and one residual; the six cross-loadings are fixed at zero."
#| fig-alt: "Eta 1 points only to Y1 through Y3 and eta 2 only to Y4 through Y6; the factors covary and every indicator has a residual."
#| fig-width: 7
#| fig-height: 6.2
#| out-width: "88%"

labels_simple <- matrix("", 6, 2)
labels_simple[cbind(seq_len(6), c(1, 1, 1, 2, 2, 2))] <-
  paste0("lambda[", seq_len(6), c(1, 1, 1, 2, 2, 2), "]")
pt2_fa_diagram(
  rep(1:2, each = 3), loading_labels = labels_simple,
  factor_cov_labels = "phi[12]"
)
```

Each zero is a restriction, not a drawing convenience. It says that,
according to this model, the second factor makes no direct contribution
to $Y_1$. Restrictions of this kind are the substance of a confirmatory
specification, and they are what make a model capable of being wrong: a
model in which everything is free forbids nothing.

The four representations must agree, and it is worth checking that they
do. For the restricted model: the path from $\eta_1$ to $Y_4$ is absent
in the diagram; $\lambda_{41} = 0$ in the equations; row 4, column 1 of
$\boldsymbol{\Lambda}$ is zero; and any common covariance between $Y_1$
and $Y_4$ must therefore travel through $\phi_{12}$, unless a residual
covariance is also permitted. The fourth representation is the sentence:
*six indicators are represented by two possibly correlated factors; each
indicator has one loading and one residual; cross-loadings and residual
covariances are fixed at zero.* That sentence should be recoverable from
any of the other three without guessing.

## Reading a parameter family {#sec-families}

Counting arrows in a diagram is not a reliable way to know what a model
contains, because the same picture can represent different models. What
matters is the status of each element. @tbl-families turns that status into
four sets of questions that should be answered for every specification.

| Family | Matrix | Questions to settle |
|:--|:--|:--|
| Intercepts | $\boldsymbol{\nu}$ | Are means modeled, or is the analysis covariance-only? |
| Loadings | $\boldsymbol{\Lambda}$ | Which entries are fixed at zero, fixed for scale, constrained equal, or free? |
| Factor variances and covariances | $\boldsymbol{\Phi}$ | Are variances fixed to set a scale? Is the covariance free? |
| Residual variances and covariances | $\boldsymbol{\Theta}$ | Is the matrix diagonal? Are any residual covariances justified and free? |

: *Parameter families and what must be declared about each.* {#tbl-families}

A loading can also be fixed at a value other than zero or one. That
possibility looks like a technicality here and becomes central later:
growth models fix a whole column of $\boldsymbol{\Lambda}$ to values
such as $0, 1, 2, 3$ so that the corresponding factor expresses change
per unit of time. The point is that fixed, free and constrained are three
different things, and a diagram alone does not distinguish them.

\Needspace{0.22\textheight}

# Coordinates without units {#sec-coordinates}

**Question.** If a factor is never observed, what determines its units, sign,
and zero?

**Short answer.** The observed moments do not determine them uniquely.
Explicit constraints choose a coordinate system, while invariant combinations
preserve the same observable implications.

An observed variable arrives with a scale: if height is recorded in
centimetres, a one-unit change means something definite. An unobserved
factor arrives with nothing. This section works through the
consequences, which are less an inconvenience than a permanent feature
of latent-variable models.

## The factor scale is indeterminate {#sec-scale}

Take a one-factor model and any nonzero constant $a$. Define

$$
\eta_i^{*} = a\,\eta_i, \qquad
\alpha^{*} = a\,\alpha, \qquad
\lambda_j^{*} = \lambda_j / a, \qquad
\phi^{*} = a^{2}\phi .
$$ {#eq-rescale}

The factor's contribution to an indicator's mean is unchanged, since
$\lambda_j^{*}\alpha^{*} = (\lambda_j/a)(a\alpha) = \lambda_j\alpha$, and
its contribution to a covariance is unchanged too:

$$
\lambda_j^{*}\phi^{*}\lambda_k^{*}
= \frac{\lambda_j}{a}\,\bigl(a^{2}\phi\bigr)\,\frac{\lambda_k}{a}
= \lambda_j\phi\lambda_k .
$$

Concretely, with $\lambda_j = .80$ and $\phi = 1$, setting $a = 2$ gives
a loading of $.40$ and a factor variance of $4$; the common contribution
to the indicator's variance is $.40^{2}(4) = .64$ either way. Everything
else — $\nu_j$, $\epsilon_{ij}$, $\theta_{jk}$ — is untouched, because
the transformation acts only on the factor's coordinates. The observed
data cannot choose between these descriptions, because they imply
exactly the same observed moments.

@tbl-scaling contrasts two conventions in general use. Both simply pick a
coordinate system.

| Convention | What is fixed | What is estimated |
|:--|:--|:--|
| **Marker variable** | One loading per factor, usually at 1 | The factor variance and the remaining loadings |
| **Fixed factor** | The factor variance, usually at 1 | All permitted loadings |

: *Two ways to give a factor a scale.* {#tbl-scaling}

Neither is superior. Unstandardized loadings and factor variances differ
between them; the implied covariance among the indicators does not.

With several factors, the coordinate freedom is wider than changing one
scale at a time. A nonsingular transformation can mix or rotate the
factor axes while corresponding changes to the loading and
factor-covariance matrices leave the observed covariance unchanged. The
new loading coordinates can look quite different even though they
represent the same common covariance. @sec-argument works one orthogonal
example numerically; the general transformation belongs with rotation,
where the freedom is used deliberately for interpretation.

## Sign and location {#sec-sign-location}

Setting $a = -1$ in @eq-rescale reverses the sign of every factor value
and every loading while leaving each product, and therefore the implied
covariance, unchanged. A factor with loadings $(.80, .70, .60)$ and one
with $(-.80, -.70, -.60)$ describe the same observed correlations. The
direction is fixed by convention and by what the items say, not by the
data; a negative sign is not by itself a problem, and a sign flip
between two equivalent solutions is not evidence that anything changed.

Location behaves the same way once means are in the model. For any fixed
$m \times 1$ vector $\mathbf{c}$,

$$
\boldsymbol{\eta}_i^{*} = \boldsymbol{\eta}_i + \mathbf{c},
\qquad
\boldsymbol{\alpha}^{*} = \boldsymbol{\alpha} + \mathbf{c},
\qquad
\boldsymbol{\nu}^{*} = \boldsymbol{\nu} - \boldsymbol{\Lambda}\mathbf{c},
$$ {#eq-location}

leaves the implied mean untouched:
$\boldsymbol{\nu}^{*} + \boldsymbol{\Lambda}\boldsymbol{\alpha}^{*}
= \boldsymbol{\nu} + \boldsymbol{\Lambda}\boldsymbol{\alpha}$. So the
observed means alone cannot determine both unrestricted factor means and
unrestricted intercepts. In a single-group model the usual convention
fixes the factor means at zero and estimates the intercepts. Comparing
latent means across groups or occasions requires additional equality
restrictions and a reference group; that is the subject of measurement
invariance, later in Part Two, and it is worth noticing now that the
question is one of *identification*, not of finding a better estimator.

## Changing the units of the indicators {#sec-units}

A different operation is to rescale the observed variables themselves —
recording a test in percentages rather than raw points, say. This also
produces a model that fits equally well, but for an entirely different
reason: the loadings, residual variances and implied covariances all
change in step with the new units, coherently and predictably. Nothing
is indeterminate here; the information is the same information in a
different metric.

The distinction matters because the two are easy to conflate. Choosing a
factor's scale is choosing coordinates for something that has none.
Changing an indicator's units is re-expressing something that already
had them. The same distinction governs the relationship between analysing
a covariance matrix and analysing a correlation matrix.

# The model-implied covariance structure {#sec-implied}

The common-factor model earns its keep because it implies a specific
covariance structure among the observed indicators. That structure, not
the raw data, is what most factor-analytic estimation actually works
with — which is why factor analysis and structural equation modeling are
together known as **covariance structure analysis**. We derive the
structure rather than quote it.

For indicators $j$ and $k$, subtract the means and use @eq-scalar:

$$
\begin{aligned}
y_{ij} - E(y_{ij}) &=
  \boldsymbol{\lambda}_j^{\mathsf T}
  (\boldsymbol{\eta}_i - \boldsymbol{\alpha}) + \epsilon_{ij}, \\
y_{ik} - E(y_{ik}) &=
  \boldsymbol{\lambda}_k^{\mathsf T}
  (\boldsymbol{\eta}_i - \boldsymbol{\alpha}) + \epsilon_{ik} .
\end{aligned}
$$

Expanding the covariance shows all four contributions:

$$
\begin{aligned}
\operatorname{Cov}(y_{ij},y_{ik})
={}&
\boldsymbol{\lambda}_j^{\mathsf T}\boldsymbol{\Phi}
  \boldsymbol{\lambda}_k \\
&+\operatorname{Cov}\!\left[
  \boldsymbol{\lambda}_j^{\mathsf T}
  (\boldsymbol{\eta}_i-\boldsymbol{\alpha}),\epsilon_{ik}\right] \\
&+\operatorname{Cov}\!\left[
  \epsilon_{ij},
  \boldsymbol{\lambda}_k^{\mathsf T}
  (\boldsymbol{\eta}_i-\boldsymbol{\alpha})\right]
  +\theta_{jk}.
\end{aligned}
$$

The first term is the common-factor contribution, the middle two are
factor–residual cross terms, and the last is the residual covariance.
The factor–residual assumption in @eq-assumptions sets both middle terms
to zero, leaving

$$
\boxed{\;
\operatorname{Cov}(y_{ij}, y_{ik})
= \boldsymbol{\lambda}_j^{\mathsf T}\boldsymbol{\Phi}
  \boldsymbol{\lambda}_k + \theta_{jk} \; }
$$ {#eq-cov}

and, for $j = k$,

$$
\operatorname{Var}(y_{ij})
= \boldsymbol{\lambda}_j^{\mathsf T}\boldsymbol{\Phi}
  \boldsymbol{\lambda}_j + \theta_{jj} .
$$ {#eq-var}

Stacking over all pairs gives the matrix form:

$$
\boxed{\;
\boldsymbol{\Sigma}(\boldsymbol{\vartheta})
= \boldsymbol{\Lambda}\boldsymbol{\Phi}\boldsymbol{\Lambda}^{\mathsf T}
  + \boldsymbol{\Theta} \; }
$$ {#eq-implied}

where $\boldsymbol{\vartheta}$ collects the free parameters and
$\theta_{jk}$ remains an entry of $\boldsymbol{\Theta}$. Every observed
covariance is thus split into a **common** part, carried by the factors,
and a **residual** part. If $\theta_{jk}=0$, a strong covariance requires
the two indicators to receive a sufficiently strong joint contribution
from shared factors. If $\theta_{jk}\ne0$, residual association can also
contribute; the decomposition prevents that association from being
silently attributed to the factors.

This identity needs finite second moments and the stated moment
restrictions. It does not need normality. Distributional assumptions
enter separately, through particular estimators, standard errors and test
statistics; that boundary is worth keeping explicit.

The same assumptions give a second useful matrix:

$$
\operatorname{Cov}(\mathbf y_i,\boldsymbol{\eta}_i)
=\boldsymbol{\Lambda}\boldsymbol{\Phi}.
$$

$\boldsymbol{\Lambda}$ is the **pattern matrix**: its entries are the
coefficients in the equations for the indicators.
$\boldsymbol{\Lambda}\boldsymbol{\Phi}$ is the **structure matrix**: its
entries are indicator–factor covariances. They are indicator–factor
correlations only when both sides have been standardized. When factors
are orthogonal and have unit variance, pattern and structure coincide;
when factors correlate, they need not. The two matrices have to be read
side by side whenever an oblique solution is interpreted.

## Communality and uniqueness {#sec-communality}

The common part of an indicator's variance,

$$
h_j^{2}
= \boldsymbol{\lambda}_j^{\mathsf T}\boldsymbol{\Phi}
  \boldsymbol{\lambda}_j ,
$$ {#eq-communality}

is its **communality under the declared model**. The qualifier matters:
changing the number of factors, the loading pattern, or the factor
covariances changes $h_j^{2}$. For a standardized indicator whose
variance decomposes into just these two parts,
$1 = h_j^{2} + \theta_{jj}$, so the residual variance $\theta_{jj}$ —
often called the **uniqueness** — is $1 - h_j^{2}$. The larger the share
the factors account for, the smaller the remainder.

If a standardized indicator has one nonzero loading on a unit-variance
factor, this reduces to $h_j^{2} = \lambda_j^{2}$. That shortcut is used
constantly and is not the general definition. With two correlated
factors the communality expands to

$$
h_j^{2} = \lambda_{j1}^{2}\phi_{11}
        + 2\lambda_{j1}\lambda_{j2}\phi_{12}
        + \lambda_{j2}^{2}\phi_{22},
$$ {#eq-communality-two}

and the middle term is why squared loadings cannot simply be added when
factors correlate. In an admissible standardized solution the
communality lies between zero and one; an estimated $\theta_{jj} < 0$ is
an improper, or **Heywood**, solution and signals a problem to
investigate rather than a discovery that an indicator has more than
100% common variance.

Uniqueness means variance the declared common factors do not represent.
It does not mean that all of that variance is random error, for the
reasons given in @sec-residual.

## A moment structure, not only a covariance structure {#sec-moments}

The same assumptions also imply a mean structure:

$$
E(\mathbf{y}_i)
= \boldsymbol{\mu}(\boldsymbol{\vartheta})
= \boldsymbol{\nu} + \boldsymbol{\Lambda}\boldsymbol{\alpha} .
$$ {#eq-mean}

Most of Part Two analyses covariances alone, holding the mean structure
aside, and that is a legitimate and common choice. It is worth being
clear that it is a *choice*. Means are part of the model even when they
are set aside, and two of Part Two's later destinations depend on them
entirely: measurement invariance compares intercepts and factor means
across groups, and latent growth models place a whole trajectory in
$\boldsymbol{\alpha}$ while fixing $\boldsymbol{\Lambda}$ to a
time basis.

A brief look at the second of those makes the point concrete. Suppose
four measurements are taken on the same people at equally spaced
occasions, with

$$
\boldsymbol{\Lambda} =
\begin{bmatrix}
1 & 0\\ 1 & 1\\ 1 & 2\\ 1 & 3
\end{bmatrix},
\qquad
\boldsymbol{\alpha} = (\alpha_1, \alpha_2)^{\mathsf T} .
$$

Every loading is fixed. The column of ones makes the first factor
contribute equally at every occasion, so $\alpha_1$ is an average
starting level; the column $0, 1, 2, 3$ is a time basis, so $\alpha_2$
is an average change per occasion. With intercepts fixed at zero, the
implied means $\boldsymbol{\nu} + \boldsymbol{\Lambda}\boldsymbol{\alpha}$
trace the average trajectory. The variances and covariances in
$\boldsymbol{\Phi}$ then describe how much individuals differ in where
they start and how fast they change, which is usually the substantive
point. We are not fitting this model here, and its full assumptions come
much later. It is shown because it uses only @eq-matrix and
@eq-mean, with the loadings fixed to substantively chosen values rather
than estimated — a possibility @tbl-families raised and this makes real.

Finally, a note on what a restriction buys. Six indicators have
$6(7)/2 = 21$ distinct variances and covariances. A model with fewer
free parameters than that makes claims about the 21 numbers which could
turn out to be false — which is exactly what makes it worth testing.
Counting moments against free parameters is the systematic form of this
observation.

# Reading the records {#sec-records}

We can now put numbers into the representation. Both records in this
section are constructed: the parameters came first and the correlations
were computed from them. That is deliberate. Because the values are
exact, any disagreement is an arithmetic error rather than sampling
variation. They are ideal for learning the algebra and useless as
evidence about any real population.

## One common factor {#sec-scenario-one}

The first record, shown in @tbl-record-one, has the same $.49$ correlation for
every distinct pair.

\Needspace{0.30\textheight}

| | $Y_1$ | $Y_2$ | $Y_3$ | $Y_4$ | $Y_5$ | $Y_6$ |
|:--|--:|--:|--:|--:|--:|--:|
| $Y_1$ | 1.00 | | | | | |
| $Y_2$ | .49 | 1.00 | | | | |
| $Y_3$ | .49 | .49 | 1.00 | | | |
| $Y_4$ | .49 | .49 | .49 | 1.00 | | |
| $Y_5$ | .49 | .49 | .49 | .49 | 1.00 | |
| $Y_6$ | .49 | .49 | .49 | .49 | .49 | 1.00 |

: *A constructed one-factor correlation record.* {#tbl-record-one}

The repeated value suggests a single common contribution shared evenly
across all six indicators. Let us check that representation rather than
infer it from appearance alone.

Let all six standardized indicators relate to a single unit-variance
factor with loading $.70$:

$$
\boldsymbol{\Lambda}_1 = (.70,\,.70,\,.70,\,.70,\,.70,\,.70)^{\mathsf T},
\qquad
\boldsymbol{\Phi}_1 = [1],
\qquad
\boldsymbol{\Theta}_1 = .51\,\mathbf{I}_6 .
$$

@fig-onefactor draws this parameterization before the entrywise check.

```{r}
#| label: fig-onefactor
#| fig-cap: "The one-factor model that reproduces the first record. Every loading is .70."
#| fig-alt: "One factor points to six indicators, with every loading labelled point 70 and a separate residual for every indicator."
#| fig-width: 6.5
#| fig-height: 6.3
#| out-width: "64%"

pt2_fa_diagram(
  rep(1L, 6),
  loading_labels = matrix(rep(".70", 6), 6, 1)
)
```

Apply @eq-cov and @eq-var. Each diagonal entry is
$.70^{2} + .51 = .49 + .51 = 1$, as a correlation matrix requires. Each
off-diagonal entry is $.70 \times 1 \times .70 = .49$. That reproduces
@tbl-record-one exactly. Every communality is $.70^{2} = .49$ and every
uniqueness is $1 - .70^{2} = .51$.

The column sum of squared loadings is $6(.70^{2}) = 2.94$. This quantity
is often called the factor's eigenvalue, and the label is safe only
under specific conditions. Here it is the sum of squares of a loading
column — an **SS loading**. It happens to equal the single nonzero
eigenvalue of the common part
$\boldsymbol{\Lambda}_1\boldsymbol{\Lambda}_1^{\mathsf T}$, but it is
*not* an eigenvalue of the correlation matrix in @tbl-record-one, whose
eigenvalues are $3.45$ and five values of $.51$. Adding $.51$ to the
diagonal shifts every eigenvalue by $.51$. Wherever eigenvalues do real
work in deciding how many factors to retain, the matrix being decomposed
has to be named; the habit starts here.

## Two orthogonal factors {#sec-scenario-two}

The second record in @tbl-record-two has two blocks: correlations are $.64$
within the first three indicators, $.36$ within the last three, and zero
across blocks.

| | $Y_1$ | $Y_2$ | $Y_3$ | $Y_4$ | $Y_5$ | $Y_6$ |
|:--|--:|--:|--:|--:|--:|--:|
| $Y_1$ | 1.00 | | | | | |
| $Y_2$ | .64 | 1.00 | | | | |
| $Y_3$ | .64 | .64 | 1.00 | | | |
| $Y_4$ | .00 | .00 | .00 | 1.00 | | |
| $Y_5$ | .00 | .00 | .00 | .36 | 1.00 | |
| $Y_6$ | .00 | .00 | .00 | .36 | .36 | 1.00 |

: *A constructed two-factor correlation record.* {#tbl-record-two}

For the second record, let the first three indicators relate to $\eta_1$
with loading $.80$ and the last three to $\eta_2$ with loading $.60$,
with the factors uncorrelated:

$$
\boldsymbol{\Lambda}_2 =
\begin{bmatrix}
.80 & 0\\ .80 & 0\\ .80 & 0\\
0 & .60\\ 0 & .60\\ 0 & .60
\end{bmatrix},
\qquad
\boldsymbol{\Phi}_2 = \mathbf{I}_2,
\qquad
\boldsymbol{\Theta}_2 = \operatorname{diag}(.36,.36,.36,.64,.64,.64).
$$

@fig-orthogonal draws the corresponding two-factor parameterization.

```{r}
#| label: fig-orthogonal
#| fig-cap: "The orthogonal two-factor model that reproduces the second record."
#| fig-alt: "Eta 1 loads point 80 on Y1 through Y3, eta 2 loads point 60 on Y4 through Y6, and their covariance is zero."
#| fig-width: 7
#| fig-height: 6.2
#| out-width: "88%"

labels_orthogonal <- matrix("", 6, 2)
labels_orthogonal[cbind(seq_len(6), c(1, 1, 1, 2, 2, 2))] <-
  c(rep(".80", 3), rep(".60", 3))
pt2_fa_diagram(
  rep(1:2, each = 3), loading_labels = labels_orthogonal,
  factor_cov_labels = "0"
)
```

Within the first block,
$\operatorname{Cov}(Y_1, Y_2) = .80 \times 1 \times .80 = .64$. Within
the second, $.60 \times 1 \times .60 = .36$. Across the blocks,

$$
\operatorname{Cov}(Y_1, Y_4) =
\begin{bmatrix} .80 & 0 \end{bmatrix}
\begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix}
\begin{bmatrix} 0 \\ .60 \end{bmatrix} = 0 ,
$$

which reproduces @tbl-record-two. The first three communalities are
$.64$ with uniquenesses $.36$; the last three are $.36$ with
uniquenesses $.64$. The SS loadings are $3(.80^{2}) = 1.92$ and
$3(.60^{2}) = 1.08$.

The exact zeros between blocks follow from three restrictions acting
together: the simple-structure loading pattern, the orthogonality of the
factors, and the absence of cross-block residual covariances. They are
not a general property of two-factor models.

## Why one factor will not do {#sec-tetrad}

The first record yielded to a single factor; the second did not. It is
worth seeing that this is a fact about the numbers, not an impression
from the layout.

Under a one-factor model with uncorrelated residuals, any covariance is
$\sigma_{jk} = \lambda_j \phi \lambda_k$. Take any four indicators and
form two products of two covariances each:

$$
\sigma_{12}\sigma_{45} - \sigma_{14}\sigma_{25}
= \lambda_1\lambda_2\lambda_4\lambda_5\phi^{2}
- \lambda_1\lambda_4\lambda_2\lambda_5\phi^{2}
= 0 .
$$ {#eq-tetrad}

The two products contain the same four loadings and must cancel. This
**tetrad** restriction is a testable consequence: a single factor
predicts it for every set of four indicators, whatever the loadings are.

In @tbl-record-one the restriction holds, as it must. In
@tbl-record-two, taking $Y_1, Y_2, Y_4, Y_5$,

$$
\sigma_{12}\sigma_{45} - \sigma_{14}\sigma_{25}
= (.64)(.36) - (0)(0) = .2304 \ne 0 ,
$$

so no one-factor model with uncorrelated residuals can reproduce that
record exactly, whatever values its loadings take. The result concerns
exact reproduction of this known-value matrix; approximate fit to a
sample covariance matrix is a different question. This is the first
genuine model comparison in the book: a restriction with observable
consequences, and a record that violates it. Degrees of freedom are a
systematic count of restrictions of exactly this kind.

## Letting the factors correlate {#sec-correlated}

Orthogonality was an assumption, not a finding. Replacing
$\boldsymbol{\Phi}_2 = \mathbf{I}_2$ with

$$
\boldsymbol{\Phi}_{2\rho} =
\begin{bmatrix} 1 & \rho \\ \rho & 1 \end{bmatrix},
\qquad -1 \leq \rho \leq 1 ,
$$

with strict $|\rho|<1$ when a nonsingular factor covariance matrix is
required, and keeping the residuals diagonal gives a cross-block
covariance of

$$
\operatorname{Cov}(Y_1, Y_4) = .80\,\rho\,(.60) = .48\rho .
$$ {#eq-cross}

At $\rho = .40$ this is $.192$, and every one of the nine cross-block
pairs becomes nonzero. The calculation shows exactly what the
orthogonality restriction had been doing, and it also sets a ceiling:
because the common part of a covariance is bounded by the product of the
two communality square roots, no factor correlation can push
$\operatorname{Cov}(Y_1, Y_4)$ beyond $.48$ under this loading pattern.
If two such indicators correlated at $.60$, no two-factor model of this
shape could account for it.

Real sample covariance matrices do not usually repeat these clean values
or contain exact structural zeros. Their deviations combine population
discrepancy with sampling variation, so arithmetic reconstruction gives
way to estimation, uncertainty, and model evaluation. No empirical result
is smuggled
into the present known-value demonstration.

## Neighbouring objects {#sec-objects}

**Question.** Is a factor the construct, the item total, a principal
component, or the score estimated for a respondent?

**Short answer.** None of these objects is interchangeable with another.
@tbl-objects separates them before we compare factor-score estimates and
principal components more closely.

| Object | What it is | Recorded for unit $i$? |
|:--|:--|:--|
| **Construct** | A concept in the research question, such as planning; its meaning comes from theory, item content, response processes, population and intended use | Not made observable by naming a factor |
| **Common factor** $\eta_{ik}$ | An unobserved random variable in the model | No |
| **Factor-score estimate** $\hat{\eta}_{ik}$ | A prediction from observed responses, a fitted model and a scoring rule | Computed, but not the factor itself |
| **Principal component** | A weighted composite of observed variables chosen to summarize variance | Computed from the data |
| **Prespecified composite** | A declared rule such as a sum or mean of items | Computed from the data |

: *Five objects that are easy to conflate.* {#tbl-objects}

A factor-score estimate depends on the fitted parameters, the items
included, their coding and the scoring rule; different rules give
different numbers from the same fitted model. Because the factor itself
is never observed, no scoring rule turns an estimate into an error-free
measurement of it.

Principal component analysis deserves a sentence of its own, because it
is often reached for in the same situations. A principal component is a
weighted composite of the observed variables, with weights chosen to
capture as much observed variance as possible. The common-factor model
proposes something different in kind: a decomposition of the covariance
into a common part and a residual part, as in @eq-implied. The two can
produce similar-looking numbers and answer different questions.
Substituting one for the other changes what is being claimed; neither is
generally superior.

## Exploratory, confirmatory and hybrid uses {#sec-approaches}

With the model in hand, the orientations of @sec-orientations can be
stated more precisely. @tbl-approaches summarizes what each orientation
declares in advance and what it asks the analysis to choose.

| Approach | What is declared in advance | What the analysis chooses |
|:--|:--|:--|
| **Exploratory** | The items, the range of factor counts considered, and the estimation and rotation choices | The number of factors, and the loading pattern within an equivalent family of coordinates |
| **Confirmatory** | The number of factors, the loading pattern, factor relations and residual structure | Parameter values, and whether the declared restrictions are compatible with the evidence |
| **Hybrid or partial** | Some relations prespecified, others left free | Depends on the exact parameterization, which must be stated |

: *What each orientation fixes and what it leaves open.* {#tbl-approaches}

Exploratory analysis is not mechanical discovery: the items entered, the
association matrix, the number of factors considered, the extraction
criterion, the rotation family and the interpretation are all decisions.
Confirmatory analysis is not automatic confirmation: it evaluates the
consequences of a specification and can expose strain, equivalent
alternatives, or evidence too weak to distinguish anything.

The two are often combined, and there is a real hazard in doing so
carelessly. A pattern discovered in one body of evidence and then
"confirmed" on the same evidence has not been confirmed; the second
analysis is a restatement of the first. The remedy is not to avoid
combining them but to keep the *analytic status* of each decision
visible: what was specified in advance, what was chosen after looking,
and what evidence has been used already. A structure chosen using one
body of evidence receives a genuinely new confirmatory test only from
evidence not used to choose it - for example, a prespecified split or a
new sample. If evidence is reused, the resulting analysis should be
described as exploratory or revised rather than independently confirmed.

## The exploratory-to-confirmatory workflow {#sec-workflow}

The two orientations are most often used in sequence, and the sequence has a
structural requirement that is easy to state and easy to violate. An
exploratory analysis chooses a factor count, a loading pattern and a set of
interpretations *by consulting the data*. A confirmatory analysis asks
whether a stated pattern is compatible with the data. If both consult the
same records, the second analysis cannot fail in any way the first did not
already rule out: its restrictions were chosen because they fit those very
numbers. Nothing has been confirmed except the analyst's arithmetic.

The remedy is to give the two stages different evidence. In practice this
means splitting the available records into at least two parts before the
exploratory work begins - one part to develop on, one held back - or
collecting a second sample. The split must be declared in advance and made
without reference to the outcome; a split chosen after seeing which half
behaves better is not a split at all.

Splitting is not free. Two halves of $n$ each carry less information than one
sample of $2n$, and factor recovery degrades in small samples with modest
loadings, so a study too small to split may be too small for the sequence
[@fabrigar1999]. The honest alternatives are to collect more data, to report
the analysis as exploratory and leave confirmation to later work, or to
declare the model in advance and skip the exploratory stage entirely. What is
not an alternative is to run both stages on one sample and describe the
second as confirmation.

Passing a model from one stage to the other requires more than a factor
count. @tbl-handoff lists what has to be fixed in writing before the held-back
records are opened, because each of these is a decision the exploratory stage
made and the confirmatory stage must not remake.

| What is transferred | Why it must be fixed in advance |
|:--|:--|
| The item set and its order | Dropping a poorly behaved item after seeing the second sample makes that sample exploratory too |
| The number of factors | The count is the exploratory stage's principal finding, and the claim under test |
| Which loadings are free and which are fixed at zero | The zeros are the restrictions being tested; adding one back is a new model |
| Factor variances, covariances and the scaling convention | These determine the metric in which every coefficient is read |
| The residual structure | A residual covariance freed after inspecting fit is a revision, not a confirmation |
| Named rival models | Comparisons decided in advance are interpretable; comparisons invented afterwards are selected on the same evidence |

: *What an exploratory-to-confirmatory handoff has to declare.* {#tbl-handoff}

A revision discovered while examining the held-back sample is a legitimate and
often necessary thing to do. It simply changes the status of that sample: it
has now been used to develop, and the revised model needs evidence of its own.
Analyses can move down this ladder freely and never move back up.

A third option refuses the sequence. Exploratory structural equation modeling
estimates a measurement model with rotational freedom retained inside a larger
structural model, rather than forcing every cross-loading to zero
[@asparouhovMuthen2009; @marsh2014]. It trades the sharpness of exact zero
restrictions for a specification that does not pretend to a precision the
evidence never supplied - a trade whose merits are still actively argued.

## A note on response scales {#sec-response}

**Question.** What changes when indicators are ordered categories rather than
genuinely continuous measurements?

**Short answer.** Treating them as continuous becomes an explicit
approximation; an underlying-response model introduces thresholds and changes
estimation and interpretation. Part Two keeps the continuous treatment so
that categorical-indicator methods can be taught elsewhere.

The indicators here are treated as continuous, and Part Two keeps that
treatment throughout. Many questionnaire items are recorded on ordered
categories, and treating such responses as continuous is a modeling
decision with a literature behind it — generally defensible with enough
categories and reasonably symmetric distributions, and less so with few
categories or strong floor and ceiling effects. The alternative is to
model an underlying continuous response together with thresholds, which
changes the estimator and the interpretation of the parameters. That
route belongs to a different course. What matters here is that the
choice be made deliberately and stated, not made by default.

# Six commitments {#sec-commitments}

The material above is enough to state the working commitments that govern the
rest of Part Two. They are not slogans: each one is a consequence of something
this chapter has already established, and each rules out a specific move that
is easy to make by accident.

**1. The model agrees across representations.** @sec-four-ways set out four
ways of writing one model - scalar equations, matrices, a diagram, and a
sentence - and required them to agree. This is worth treating as a routine
check rather than a stylistic preference. A path missing from a diagram but
free in a matrix is not a drawing slip; it is two different models, and the
one that gets estimated is whichever the software was given. When syntax and
output enter later, they join the same obligation: everything that describes
the model must describe the *same* restrictions.

**2. Exploration and confirmation are historical roles.** @sec-workflow argued
that what separates them is when a decision was made and on what evidence, not
which procedure was run. Nothing in a fitted object records this. A model
specified in advance and the identical model arrived at by inspecting the data
produce the same coefficients, the same fit, and the same printout; the
difference lives only in the analytic history, which is why that history has to
be written down rather than reconstructed afterwards.

**3. A fitted solution is read in stages.** The stages are ordered by
dependence, not by preference. Whether the parameters are recoverable at all is
prior to estimating them; whether an estimator converged is prior to reading
what it produced; whether the solution is admissible - no negative variance, no
correlation outside its bounds - is prior to interpreting a coefficient; and
all of that is prior to asking how well the model reproduces the observed
moments. A summary computed from an inadmissible solution is arithmetic
performed on an impossibility. @sec-coordinates and @sec-implied supply the
first two stages; the later ones need data this chapter does not use.

**4. Comparison requires eligibility.** @sec-tetrad made the first genuine
comparison in this book: a restriction with an observable consequence, and a
record that violates it. That is what makes comparison possible - a difference
in what two models forbid. Two models that forbid nothing different cannot be
distinguished by any amount of evidence, and @sec-argument shows the sharpest
version of this, where an orthogonal rotation leaves every implied moment
unchanged. Having fewer parameters is not the same as being nested, and
appearing in adjacent rows of a table is not a test.

**5. Computation reproduces rather than justifies.** This chapter's own code
checks its arithmetic; it settles nothing else. Code can verify that a
constructed matrix really does follow from the parameters claimed to generate
it, and it can preserve a record another analyst can re-run. It cannot supply a
design that was never conducted, an item set that covers the domain, or a
reason to interpret a factor as an attribute. A calculation that runs without
error has demonstrated that it runs without error.

**6. Claims remain bounded.** Reproducing a covariance record establishes
compatibility between that record and the model, and stops there. It does not
by itself establish how many dimensions an attribute has, that the factors are
the constructs they are named for, that a score is valid or comparable across
groups, or that any arrow represents a causal effect. Each of those requires
its own evidence, and none of it is contained in the fit.

The sixth commitment is the one the rest of the chapter has been building
toward, and the optional section that follows works it through on a case where
the arithmetic is exact and the conclusion still fails.

# Applying the argument-based approach to latent variables (Optional) {#sec-argument}

This optional section returns to the Chapter Box using the argument
vocabulary of Part One Chapters 1 and 1a. It may be skipped without loss.

## Can covariance select unique factor coordinates?

Sections 2 to 6 established the statistical account. Here we ask whether
covariance selects unique latent coordinates.

> **Operational question.** Do the original and a rotated two-factor
> description imply the same six-indicator covariance, and which parts
> of their interpretation survive the coordinate change?

Using the constructed record in @tbl-record-two, begin with
$\boldsymbol{\Lambda}_2$ and $\boldsymbol{\Phi}_2=\mathbf I_2$ from
@sec-scenario-two and define the orthogonal matrix

$$
\mathbf Q=\frac{1}{\sqrt{2}}
\begin{bmatrix}
1&-1\\
1& 1
\end{bmatrix},
\qquad
\boldsymbol{\eta}_i^*=\mathbf Q^{\mathsf T}\boldsymbol{\eta}_i,
\qquad
\boldsymbol{\Lambda}_2^*=\boldsymbol{\Lambda}_2\mathbf Q .
$$

Because $\mathbf Q^{\mathsf T}\mathbf Q=\mathbf I_2$, the rotated factor
covariance remains $\boldsymbol{\Phi}_2^*=\mathbf I_2$. The loading
matrix becomes

$$
\boldsymbol{\Lambda}_2^*=
\begin{bmatrix}
.80/\sqrt{2}&-.80/\sqrt{2}\\
.80/\sqrt{2}&-.80/\sqrt{2}\\
.80/\sqrt{2}&-.80/\sqrt{2}\\
.60/\sqrt{2}& .60/\sqrt{2}\\
.60/\sqrt{2}& .60/\sqrt{2}\\
.60/\sqrt{2}& .60/\sqrt{2}
\end{bmatrix}.
$$

The simple zeros have disappeared, yet the implied covariance has not.
For example, the common covariance of $Y_1$ and $Y_4$ is now

$$
\frac{.80}{\sqrt{2}}\frac{.60}{\sqrt{2}}
+
\frac{-.80}{\sqrt{2}}\frac{.60}{\sqrt{2}}
=.24-.24=0.
$$

Every other entry is preserved by the same matrix identity,
$\boldsymbol{\Lambda}_2^*\boldsymbol{\Phi}_2^*
\boldsymbol{\Lambda}_2^{*\mathsf T}
=\boldsymbol{\Lambda}_2\boldsymbol{\Phi}_2
\boldsymbol{\Lambda}_2^{\mathsf T}$. Indicator communalities therefore
stay at $.64$ for $Y_1$ to $Y_3$ and $.36$ for $Y_4$ to $Y_6$. What
changes is the allocation to the two columns, as @tbl-coordinate-ss shows:

| Coordinates | SS loading 1 | SS loading 2 | Total common variance |
|:--|--:|--:|--:|
| Original | 1.92 | 1.08 | 3.00 |
| Rotated | 1.50 | 1.50 | 3.00 |

: *Factor-specific allocations change while their total is invariant.* {#tbl-coordinate-ss}

## Does exact reproduction confirm two constructs?

Suppose a results section uses the original simple axes and states:

> "Because the two-factor model reproduces the six-indicator record exactly,
> the analysis confirms that Planning and Strategy are two distinct
> psychological constructs."

The arithmetic in the first clause is correct. The conclusion is not. The
same covariance record admits the rotated description, whose columns no
longer preserve the proposed Planning and Strategy partition. The four
argument roles make the break visible.

**Evidence.** The evidence is the declared correlation record together
with the original and rotated numerical representations. Its provenance
matters: the record was constructed from parameters, not estimated from
respondents. Exact arithmetic is therefore available, but sampling and
generalization claims are not.

**Justification.** The common-factor covariance equation
@eq-implied, the unchanged residual covariance matrix, and the
orthogonality of $\mathbf Q$ connect the two loading matrices to the same
observable covariance. The choice of axes belongs here as a coordinate
convention; it is not an observed finding.

**Inference.** Substitution gives the same implied covariance,
communalities, and total common variance of $3.00$. The displayed
cancellation $.24-.24=0$ verifies a cross-block entry, while the SS
loading calculation shows that factor-specific allocations change from
$1.92$ and $1.08$ to $1.50$ and $1.50$.

**Claim.** The quoted claim moves from exact reproduction to distinct
psychological constructs. That transition requires a substantive warrant
about indicator content and response processes which the constructed record
does not provide.

## Audit the claim

@tbl-coordinate-verdicts applies the six evaluation questions to the quoted
claim. The mixed verdicts matter: correct arithmetic at one stage does not
repair an unsupported transition at a later one.

\Needspace{0.72\textheight}

| Evaluation question | Verdict | Reason in this record |
|:--|:--|:--|
| Question alignment | Satisfied | The sentence addresses whether the proposed factors may be interpreted as distinct constructs. |
| Evidence integrity | Satisfied within the declared record | All numbers come from the two stated parameterizations of one constructed matrix. |
| Justification adequacy | Partly satisfied | Orthogonality and @eq-implied license covariance reproduction and equivalence, not the construct labels. |
| Inferential adequacy | Not satisfied | The equivalent rotation prevents the record from selecting the original pair of axes uniquely. |
| Transition fidelity | Not satisfied | The sentence moves from a statistical representation to psychological entities without a domain or response-process warrant. |
| Claim calibration | **Not satisfied** | Exact reproduction cannot by itself confirm that Planning and Strategy are distinct psychological constructs. |

: *Six applied evaluation verdicts for an overclaim based on the coordinate comparison.* {#tbl-coordinate-verdicts}

## Repair the claim

A calibrated replacement is:

> "The record is compatible with the displayed two-factor parameterization,
> but the equivalent rotated description shows that covariance reproduction
> does not uniquely identify Planning and Strategy axes. Interpreting those
> axes as distinct constructs requires theory, defensible domain coverage,
> response-process evidence, and relevant empirical checks."

**Answer returned to the opening pair.** The two descriptions imply the
same indicator covariance, communalities, and total common variance of
$3.00$. Their loading coordinates and factor-specific SS allocations
differ. Thus the covariance representation survives the coordinate
change, whereas the convenience and factor-specific allocation of the
original simple axes do not. The rotated description is outside the
original fixed-zero CFA class: equivalence in the broader factor
representation does not imply that an identified restricted CFA lacks a
unique fitted solution.

As an optional transfer exercise, rewrite the quoted overclaim as four
sentences - one each for evidence, justification, inference and claim - and
identify every phrase you changed to keep covariance compatibility separate
from construct interpretation.

# Summary and reading bridge {#sec-summary}

@fig-chapter-route retraces the chapter's route from observed covariance to a
bounded interpretation.

```{r}
#| label: fig-chapter-route
#| fig-cap: "The chapter's route from observed covariance to a calibrated conclusion."
#| fig-alt: "A six-step vertical sequence runs from observed indicators covary, through proposing factors and residuals, translating the model, supplying latent coordinates, and deriving implied moments and restrictions, to calibrating the substantive claim."
#| fig-width: 7
#| fig-height: 6.5
#| out-width: "78%"

route_labels <- c(
  "Observed indicators\ncovary",
  "Propose factors\nand residuals",
  "Translate the model:\nequations, matrix,\ndiagram and prose",
  "Supply latent\ncoordinates",
  "Derive implied moments\nand test restrictions",
  "Calibrate the\nsubstantive claim"
)
route_y <- seq(.91, .09, length.out = length(route_labels))

op <- par(mar = c(.3, .3, .3, .3), xpd = NA)
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 1))
for (k in seq_along(route_labels)) {
  rect(.14, route_y[k] - .052, .86, route_y[k] + .052,
       col = if (k %in% c(1, 6)) "grey88" else "white",
       border = "black", lwd = 1.1)
  text(.5, route_y[k], route_labels[k], cex = .90)
  if (k < length(route_labels)) {
    arrows(.5, route_y[k] - .057, .5, route_y[k + 1] + .060,
           length = .08, lwd = 1.1)
  }
}
par(op)
```

The model is

$$
\mathbf{y}_i = \boldsymbol{\nu} + \boldsymbol{\Lambda}\boldsymbol{\eta}_i
               + \boldsymbol{\epsilon}_i ,
\qquad
\boldsymbol{\Sigma}
= \boldsymbol{\Lambda}\boldsymbol{\Phi}\boldsymbol{\Lambda}^{\mathsf T}
  + \boldsymbol{\Theta} ,
\qquad
\boldsymbol{\mu}
= \boldsymbol{\nu} + \boldsymbol{\Lambda}\boldsymbol{\alpha} .
$$

Its loadings, factor covariances and residual covariances jointly
determine the observed moments. Its latent coordinates have no inherited
scale, sign or location, and several coordinate systems describe the
same observed covariance. Communality and uniqueness are conditional on
the declared model. A diagonal residual covariance matrix does not by
itself establish local independence.

What this chapter could not do is the point of the next one. Every
numerical record here was exact by construction. Given a sample
correlation matrix that no simple model reproduces exactly, how many
factors should be retained, how should the coordinates be chosen, and
how much residual discrepancy is tolerable? Chapter 2 takes those
questions up as an exploratory analysis, with the retention decision —
the one this chapter supplied by construction — treated as the
consequential judgment it is.

For fuller treatments of latent-variable representation and
covariance-structure modeling, see @bartholomewKnott1999, @mulaik2010,
@joreskog1969, @bollen1989 and @brown2006. This chapter uses their
shared vocabulary selectively and leaves estimation and model evaluation
to later chapters.

## Questions this chapter leaves open {#sec-open}

The following are not exercises with reference answers. Each names a
disagreement that is still live among competent methodologists, and each
bears directly on how the algebra above should be read. They suit seminar
discussion or a short written position; the citations are starting points.

1. **Is a latent variable discovered or introduced?** @borsboom2003
   distinguish treating it as a real attribute that causes the responses, as
   a summary construction, and as shorthand for the measurement operations.
   Which does this chapter presuppose - and does the algebra require any of
   them?
2. **Which way should the arrows point?** For constructs such as
   socioeconomic status, indicators arguably *produce* the construct rather
   than express it, in which case internal consistency stops being a virtue
   and the common-factor model no longer applies [@bollenLennox1991;
   @edwardsBagozzi2000]. Name a construct in your field where the direction
   is genuinely arguable, and say what could settle it.
3. **Are construct, factor, composite and score kept apart in practice?**
   Using @tbl-objects, find one paper in your area that treats two of them
   as interchangeable, and say what follows.
4. **Is forcing cross-loadings to zero honest or convenient?** Exploratory
   structural equation modeling retains rotational freedom inside a larger
   model [@asparouhovMuthen2009; @marsh2014]; critics reply that a
   specification forbidding nothing is harder to refute. Which risk concerns
   you more?
5. **When is splitting the sample the wrong call?** @sec-workflow required
   separate evidence for development and confirmation, but splitting halves
   the information available to each stage and factor analyses are often run
   on samples already too small [@fabrigar1999]. Where would you draw the
   line?
6. **On what basis is a coordinate system chosen?** @sec-argument showed the
   data cannot choose. Is "interpretability" a scientific criterion, a
   convention, or a theoretical commitment in disguise?

# Exercises {.unnumbered .unlisted}

Show your assumptions and arithmetic, and end every numerical answer
with a sentence of interpretation. None of these requires software.

## Foundational 1 - Classify the quantities {.unnumbered .unlisted}

For each item below, say whether it is an observed variable, a latent
random variable, an unknown fixed parameter, or a residual random
variable. Then explain, in one or two sentences, what distinction is
lost if every unobserved quantity is simply called "latent".

1. A respondent's answer to Item 4.
2. The loading relating Item 4 to the second factor.
3. The value of the second factor for that respondent.
4. The part of Item 4 not represented by the declared factors.
5. The population covariance between the two factors.

## Foundational 2 - Place the model {.unnumbered .unlisted}

1. Using @tbl-map, say which family fits each of: a five-category
   attitude scale analysed with a continuous latent trait; a set of
   continuous school-performance measures thought to identify unobserved
   groups of schools; six continuous items thought to reflect two
   continuous attributes.
2. Under what additional interpretation can a classical true-score
   decomposition be represented as a one-factor model? State the
   equal-loading restriction, then explain why the two frameworks are
   not generally interchangeable.
3. Give one question about a six-item scale that an exploratory
   orientation suits, and one that a confirmatory orientation suits.
   Explain what makes each appropriate.

## Foundational 3 - Translate between representations {.unnumbered .unlisted}

Consider a two-factor model in which $Y_1, Y_2, Y_3$ relate only to
$\eta_1$, $Y_4, Y_5, Y_6$ relate only to $\eta_2$, the factors may
correlate, and all residual covariances are zero.

1. Write the six scalar equations.
2. Write $\boldsymbol{\Lambda}$, $\boldsymbol{\Phi}$ and
   $\boldsymbol{\Theta}$, marking which elements are fixed and which are
   free.
3. Sketch the path diagram.
4. Describe the model in one paragraph, then check that every statement
   in your paragraph is visible in all three of the other
   representations.

## Foundational 4 - Follow the independence ladder {.unnumbered .unlisted}

Let $U\sim N(0,1)$, $\epsilon_1=U$, and $\epsilon_2=U^2-1$. Use $E(U^2)=1$,
$E(U^3)=0$, and $E(U^4)=3$.

1. Show that both residuals have mean zero.
2. Compute $\operatorname{Var}(\epsilon_1)$,
   $\operatorname{Var}(\epsilon_2)$, and
   $\operatorname{Cov}(\epsilon_1,\epsilon_2)$.
3. Show that the residuals are nevertheless dependent, first from their
   deterministic relation and then by calculating
   $\operatorname{Cov}(\epsilon_1^2,\epsilon_2)$.
4. Explain what this example shows about a diagonal $\boldsymbol\Theta$,
   residual independence, and local independence conditional on the factors.

## Foundational 5 - Reconstruct and perturb {.unnumbered .unlisted}

1. For the model of @sec-scenario-one, recompute one diagonal and one
   off-diagonal entry of the implied correlation matrix, the communality
   and uniqueness of any indicator, and the SS loading. State which
   quantity in your answer is *not* an eigenvalue of @tbl-record-one,
   and why.
2. For the model of @sec-scenario-two, recompute one within-block entry
   from each block and one cross-block entry.
3. Now suppose $Y_1$ also relates to $\eta_2$ with a loading of $.25$,
   with the factors still uncorrelated. Which entries of the implied
   matrix change, and which do not? Recompute
   $\operatorname{Cov}(Y_1, Y_4)$.
4. Instead of the cross-loading, suppose a residual covariance
   $\theta_{14} = .05$ is permitted. Which entries change now? Explain
   why these two modifications are not interchangeable.

## Foundational 6 - Coordinates and conventions {.unnumbered .unlisted}

Begin with a standardized one-factor indicator having $\lambda = .80$,
$\phi = 1$ and $\theta = .36$.

1. Rescale the factor by $a = 2$. Give the new loading and factor
   variance, and verify that the implied indicator variance is
   unchanged.
2. Set $a = -1$. What changes, and what does not? Would you be able to
   tell the two solutions apart from the correlations alone?
3. State the difference between fixing a factor's scale and changing the
   units in which an indicator is recorded. Give an example of each.
4. In a single-group model the factor mean is usually fixed at zero.
   Using @eq-location, explain why some such convention is unavoidable.

## Integrative 7 - Test an exact restriction, then bound the claim {.unnumbered .unlisted}

A colleague reports that six items "clearly measure a single underlying
trait", offering @tbl-record-two as support and noting that all six
items are positively worded.

1. Compute the tetrad $\sigma_{13}\sigma_{46} - \sigma_{16}\sigma_{34}$
   for @tbl-record-two. What does a one-factor model require of it?
2. State, with your calculation, whether a one-factor model with
   uncorrelated residuals can reproduce this constructed record exactly.
   Why is that narrower than deciding whether six real items measure one
   trait?
3. Suppose a sample covariance matrix yields the corresponding tetrad
   $.0402$ rather than the model-imposed value zero. Explain why a
   nonzero sample tetrad is not by itself a population-model refutation.
   What sampling information or fitted-model assessment would you need?
4. Suppose a two-factor model does reproduce a record exactly. Write two
   sentences stating what that establishes, and two stating what it does
   not.

::: {#refs}
:::
