---
title: "Exploratory Factor Analysis and Model Development"
author: "Jinsong Chen"
bibliography: ../references.bib
csl: ../apa.csl
link-citations: true
lang: en
execute:
  echo: false
  warning: false
  message: false
  cache: false
format:
  pdf:
    output-file: chapter_02.pdf
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      - text: |
          % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
          % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
          % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
          % https://github.com/latex3/graphics-def/pull/48
          % https://github.com/latex3/graphics-def/issues/47
          % The upstream PR was unmerged when this local backport was made.
          % Original driver copyright: LaTeX3 Project and its listed contributors.
          % This adaptation is distributed under LPPL 1.3c or later:
          % https://www.latex-project.org/lppl.txt
          %
          % Preserve the imported PDF's transparency group. Put the driver's additional
          % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
          % keys. Do not replace this with PDF post-processing or disabling transparency.
          % Exact-token patches deliberately fail on an unexpected/new driver: review
          % whether the upstream fix is then installed before removing this backport.
          \usepackage{etoolbox}
          \makeatletter
          \newcommand{\QMSBRgraphicsPatchError}[1]{%
            \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
              {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
          \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
          \patchcmd{\Gread@@pdftex}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
            {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
             \ifGread@transgroup
               \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
               \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
             \fi}
            {}{\QMSBRgraphicsPatchError{3}}
          \patchcmd{\Gread@@pdftex}
            {\useimageresource\the\lastsavedimageresourceindex}
            {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
            {}{\QMSBRgraphicsPatchError{4}}
          \makeatother
          \setlength{\parindent}{0pt}
          \usepackage{float}
          \usepackage{needspace}
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{caption}
          \captionsetup[table]{position=above,format=plain,justification=raggedright,singlelinecheck=false,labelsep=newline,labelfont=bf,textfont=it}
  html:
    toc: true
    toc-depth: 2
    number-sections: true
editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(dev=if(knitr::is_latex_output()) "cairo_pdf" else "png",
  dpi=160,fig.pos="H",fig.align="center")
options(digits=3,width=80)

# This chapter is self-contained. The SLE population is declared below and the
# development records are simulated in memory from a fixed seed, so the QMD
# needs no external script. The released sle_continuous.csv is read only by the
# provenance check below, which confirms that the regenerated records match the
# file readers download; the analysis itself runs on the regenerated values.
for (p in c("psych","GPArotation"))
  if (!requireNamespace(p,quietly=TRUE)) stop("Required package: ",p)

# ---- SLE codebook, manifest, population -----------------------------------

sle_codebook_lines <- function() c(
  "item,domain,indicator_definition,wording,unit,coding,missing_code,response_process",
  "P1,Planning,Planning-routine consistency,I establish a study plan before beginning a task.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous index; not a single ordered-category response",
  "P2,Planning,Priority-setting clarity,I identify which study tasks deserve attention first.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous index; not a single ordered-category response",
  "P3,Planning,Progress-monitoring regularity,I compare my progress with the plan I made.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous index; not a single ordered-category response",
  "P4,Planning,Adaptive planning,I revise my plan when a study method is not working.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous index combining planning and method adjustment",
  "S1,Strategy,Strategy-adjustment flexibility,I change my learning strategy when understanding is weak.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous index; not a single ordered-category response",
  "S2,Strategy,Method-selection deliberateness,I select a learning method suited to the task.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous index; not a single ordered-category response",
  "S3,Strategy,Reflective revision,I revise my explanation after checking my understanding.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous index; not a single ordered-category response",
  "S4,Strategy,Socially supported strategy change,I use discussion with others to improve my learning method.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous index combining learning strategy and social participation",
  "B1,Belonging,Peer acceptance,I feel accepted by other students.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous experience index; shared peer context with B2",
  "B2,Belonging,Peer participation,I feel able to join other students in learning activities.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous experience index; shared peer context with B1",
  "B3,Belonging,Classroom voice,I feel able to express my views in class.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous experience index; not a single ordered-category response",
  "B4,Belonging,Institutional connection,I feel connected to the university as a whole.,process-index points,higher means more of the stated attribute,NA,hypothetical continuous experience index; broad content rather than one setting"
)

sle_codebook <- function()
  read.csv(text=paste(sle_codebook_lines(),collapse="\n"),stringsAsFactors=FALSE)

sle_manifest <- function()
  data.frame(sample=LETTERS[1:3],
    role=c("development","confirmation","revalidation"),
    n=360L,population="SLE-continuous-v1",seed=20260921:20260923,
    rng="Mersenne-Twister/Inversion/Rejection",stringsAsFactors=FALSE)

sle_population <- function() {
  items <- c(paste0("P", 1:4), paste0("S", 1:4), paste0("B", 1:4))
  factors <- c("Planning", "Strategy", "Belonging")
  Lambda_z <- matrix(0, 12, 3, dimnames = list(items, factors))
  Lambda_z[1:4, 1] <- c(.78, .74, .70, .55)
  Lambda_z[5:8, 2] <- c(.76, .72, .68, .55)
  Lambda_z[9:12, 3] <- c(.80, .74, .67, .50)
  Lambda_z[4, 2] <- .32
  Lambda_z[8, 3] <- .30
  Phi <- matrix(c(1,.42,.30, .42,1,.35, .30,.35,1), 3, 3,
                dimnames = list(factors, factors))
  common <- Lambda_z %*% Phi %*% t(Lambda_z)
  Theta_z <- diag(1 - diag(common))
  dimnames(Theta_z) <- list(items, items)
  # Shared peer context not fully represented by the broad factors.
  Theta_z[9, 10] <- Theta_z[10, 9] <- .12
  scale <- c(8,10,12,9, 10,8,11,9, 10,12,8,11)
  D <- diag(scale)
  nu <- setNames(c(50,52,48,51, 49,50,52,48, 51,49,50,52), items)
  Lambda <- D %*% Lambda_z
  Theta <- D %*% Theta_z %*% D
  dimnames(Lambda) <- dimnames(Lambda_z)
  dimnames(Theta) <- dimnames(Theta_z)
  stopifnot(min(eigen(Phi, symmetric=TRUE)$values) > 0,
            min(eigen(Theta, symmetric=TRUE)$values) > 0)
  list(version="SLE-continuous-v1", items=items, factors=factors,
       Lambda_z=Lambda_z, Phi=Phi, Theta_z=Theta_z, scale=scale,
       nu=nu, alpha=setNames(rep(0,3),factors), Lambda=Lambda, Theta=Theta)
}

# Simulate one declared sample in memory. Deterministic given the seed.
sle_simulate <- function(role="development") {
  manifest <- sle_manifest()
  row <- manifest[manifest$role==role,,drop=FALSE]
  stopifnot(nrow(row)==1L)
  pop <- sle_population()
  RNGkind("Mersenne-Twister","Inversion","Rejection")
  set.seed(row$seed)
  n <- row$n
  eta <- matrix(rnorm(n*3),n,3) %*% chol(pop$Phi)
  epsilon <- matrix(rnorm(n*12),n,12) %*% chol(pop$Theta)
  Y <- sweep(eta %*% t(pop$Lambda)+epsilon,2,pop$nu,"+")
  colnames(Y) <- pop$items
  x <- as.data.frame(Y)
  stopifnot(identical(dim(x),c(360L,12L)),!anyNA(x),all(is.finite(as.matrix(x))))
  list(x=x,ids=sprintf("%s%03d",row$sample,seq_len(n)),manifest=row,
    manifest_all=manifest)
}

# Provenance check on the whole released file: every sample, every case
# identifier and all twelve indicators must match what the source regenerates.
# The released CSV is part of the chapter, so a missing file is an error.
sle_verify_release <- function(path = "sle_continuous.csv") {
  if (!file.exists(path))
    stop("Released file ", path, " not found; it is part of this chapter.")
  released <- read.csv(path, stringsAsFactors = FALSE)
  manifest <- sle_manifest()
  items <- sle_population()$items
  stopifnot(identical(names(released), c("sample", "id", items)),
            nrow(released) == sum(manifest$n),
            identical(sort(unique(released$sample)), sort(manifest$sample)))
  for (role in manifest$role) {
    row <- manifest[manifest$role == role, , drop = FALSE]
    got <- sle_simulate(role)
    ref <- released[released$sample == row$sample, , drop = FALSE]
    stopifnot(nrow(ref) == row$n,
              identical(ref$id, got$ids),
              max(abs(as.matrix(ref[items]) - as.matrix(got$x))) < 1e-8)
  }
  TRUE
}

# Only the development sample is analyzed here. Samples B and C are
# regenerated by the provenance check above and then discarded.
efa_read_development <- function() {
  input <- sle_simulate("development")
  sle_verify_release()
  list(x=input$x,codebook=sle_codebook(),manifest=input$manifest,
    manifest_all=input$manifest_all)
}

# ---- exploratory analysis --------------------------------------------------

efa_pa_roots <- function(R) {
  stopifnot(min(eigen(R,symmetric=TRUE,only.values=TRUE)$values)>1e-10)
  reduced <- R
  diag(reduced) <- 1 - 1/diag(solve(R))
  cbind(component=eigen(R,symmetric=TRUE,only.values=TRUE)$values,
        factor=eigen(reduced,symmetric=TRUE,only.values=TRUE)$values)
}

efa_permutation_pa <- function(x, reps=100L, seed=12345L) {
  RNGkind("Mersenne-Twister","Inversion","Rejection")
  set.seed(seed)
  observed <- efa_pa_roots(cor(x))
  null <- array(NA_real_, c(ncol(x),2,reps))
  for (b in seq_len(reps)) {
    permuted <- vapply(x, function(v) sample(v, replace=FALSE), numeric(nrow(x)))
    # Do not discard failures or smooth a matrix: stop the analysis instead.
    null[,,b] <- efa_pa_roots(cor(permuted))
  }
  reference <- apply(null,c(1,2),quantile,probs=.95,type=7)
  # Retain the leading uninterrupted run, not later isolated crossings.
  leading <- function(ok) if (all(ok)) length(ok) else which(!ok)[1]-1L
  counts <- vapply(1:2,function(j) leading(observed[,j]>reference[,j]),integer(1))
  list(observed=observed,reference=reference,null=null,reps=reps,seed=seed,
       counts=setNames(counts,c("component","factor")),failures=0L)
}

efa_extract <- function(R,n,m,fm="ml") {
  warnings <- character()
  fit <- withCallingHandlers(psych::fa(R,nfactors=m,n.obs=n,fm=fm,rotate="none",
    SMC=TRUE,min.err=1e-8,max.iter=1000,smooth=FALSE,scores="none",warnings=TRUE),
    warning=function(w) {warnings <<- c(warnings,conditionMessage(w)); invokeRestart("muffleWarning")})
  list(fit=fit,warnings=unique(warnings))
}

efa_align <- function(P,Phi,reference=NULL) {
  if (ncol(P)!=3L) return(list(P=P,Phi=Phi))
  perms <- rbind(c(1,2,3),c(1,3,2),c(2,1,3),c(2,3,1),c(3,1,2),c(3,2,1))
  if (is.null(reference)) {
    anchors <- list(1:3,5:7,9:11)
    strength <- t(vapply(anchors,function(i) colMeans(abs(P[i,,drop=FALSE])),numeric(3)))
    score <- apply(perms,1,function(p) sum(strength[cbind(1:3,p)]))
  } else {
    congruence <- t(reference) %*% P /
      sqrt(outer(colSums(reference^2),colSums(P^2)))
    score <- apply(perms,1,function(p) sum(abs(congruence[cbind(1:3,p)])))
  }
  order <- perms[which.max(score),]
  P <- P[,order,drop=FALSE]; Phi <- Phi[order,order,drop=FALSE]
  signs <- if (is.null(reference)) sign(c(sum(P[1:3,1]),sum(P[5:7,2]),sum(P[9:11,3]))) else
    sign(colSums(P*reference))
  signs[signs==0] <- 1
  P <- sweep(P,2,signs,"*"); Phi <- diag(signs)%*%Phi%*%diag(signs)
  colnames(P) <- c("Planning","Strategy","Belonging")
  dimnames(Phi) <- list(colnames(P),colnames(P))
  list(P=P,Phi=Phi)
}

efa_rotate <- function(A,method="oblimin",seed=12345L,reference=NULL) {
  set.seed(seed)
  if (ncol(A)==1L) return(list(P=A,Phi=matrix(1),converged=TRUE))
  rot <- switch(method,
    oblimin=GPArotation::oblimin(A,gam=0,normalize=FALSE,eps=1e-7,maxit=10000,randomStarts=20),
    geomin=GPArotation::geominQ(A,delta=.01,normalize=FALSE,eps=1e-7,maxit=10000,randomStarts=20),
    varimax=GPArotation::Varimax(A,normalize=FALSE,eps=1e-7,maxit=10000,randomStarts=20))
  Phi <- if (is.null(rot$Phi)) diag(ncol(A)) else rot$Phi
  aligned <- efa_align(unclass(rot$loadings),Phi,reference)
  c(aligned,list(converged=isTRUE(rot$convergence),rotation=rot))
}

efa_discrepancy <- function(S,Sigma) {
  p <- nrow(S)
  as.numeric(determinant(Sigma,logarithm=TRUE)$modulus + sum(diag(solve(Sigma,S))) -
    determinant(S,logarithm=TRUE)$modulus - p)
}

run_efa_analysis <- function() {
  for (p in c("psych","GPArotation"))
    if (!requireNamespace(p,quietly=TRUE)) stop("Required package: ",p)
  input <- efa_read_development()
  x <- input$x; n <- nrow(x); p <- ncol(x); R <- cor(x); S <- cov(x)
  pa <- efa_permutation_pa(x)
  fits <- lapply(1:5,function(m) efa_extract(R,n,m))
  # psych does not expose its ML optimizer convergence flag. Independently
  # check the same likelihood optimum using stats::factanal (five starts).
  set.seed(12345L)
  optimizer_checks <- lapply(1:5,function(m)
    stats::factanal(covmat=R,n.obs=n,factors=m,rotation="none",control=list(nstart=5)))
  rotations <- lapply(1:5,function(m) efa_rotate(unclass(fits[[m]]$fit$loadings),seed=12345L))
  primary <- rotations[[3]]
  A <- unclass(fits[[3]]$fit$loadings)
  varimax <- efa_rotate(A,"varimax",reference=primary$P)
  geomin <- efa_rotate(A,"geomin",seed=12345L,reference=primary$P)
  minres <- efa_extract(R,n,3,"minres")
  minres_rot <- efa_rotate(unclass(minres$fit$loadings),seed=12345L,reference=primary$P)
  describe_fit <- function(f,m,rot) {
    common <- tcrossprod(unclass(f$loadings))
    Sigma <- common + diag(f$uniquenesses)
    residual <- R-Sigma
    Fml <- efa_discrepancy(R,Sigma)
    df <- ((p-m)^2-p-m)/2
    multiplier <- n-1-(2*p+5)/6-2*m/3
    data.frame(Factors=m,df=df,F_ML=Fml,Chi_square=multiplier*Fml,
      p_value=pchisq(multiplier*Fml,df,lower.tail=FALSE),
      RMS_offdiag=sqrt(mean(residual[lower.tri(residual)]^2)),
      Max_residual=max(abs(residual[lower.tri(residual)])),
      Min_uniqueness=min(f$uniquenesses),
      Optimizer_check=isTRUE(optimizer_checks[[m]]$converged),
      Objective_gap=abs(Fml-unname(optimizer_checks[[m]]$criteria["objective"])),Rotated=rot$converged,
      Boundary=min(f$uniquenesses)<=.0051)
  }
  candidates <- do.call(rbind,lapply(1:5,function(m) describe_fit(fits[[m]]$fit,m,rotations[[m]])))
  common <- primary$P %*% primary$Phi %*% t(primary$P)
  h2 <- diag(common); u2 <- fits[[3]]$fit$uniquenesses
  reproduced <- common+diag(u2); residual <- R-reproduced
  index <- which(lower.tri(residual),arr.ind=TRUE)
  residual_pairs <- data.frame(Item1=rownames(R)[index[,1]],Item2=colnames(R)[index[,2]],
    Observed=R[index],Reproduced=reproduced[index],Residual=residual[index])
  residual_pairs <- residual_pairs[order(abs(residual_pairs$Residual),decreasing=TRUE),]
  inverse <- solve(R); partial <- -cov2cor(inverse); diag(partial) <- 0
  kmo <- sum(R[lower.tri(R)]^2)/(sum(R[lower.tri(R)]^2)+sum(partial[lower.tri(partial)]^2))
  bartlett <- -(n-1-(2*p+5)/6)*as.numeric(determinant(R,logarithm=TRUE)$modulus)
  audit <- data.frame(Item=names(x),Mean=vapply(x,mean,numeric(1)),SD=vapply(x,sd,numeric(1)),
    Min=vapply(x,min,numeric(1)),Max=vapply(x,max,numeric(1)),
    Skew=vapply(x,function(z) mean((z-mean(z))^3)/mean((z-mean(z))^2)^1.5,numeric(1)),
    Missing=vapply(x,function(z) sum(is.na(z)),integer(1)))
  sensitivity <- do.call(rbind,lapply(list(ML_oblimin=primary,ML_geomin=geomin,Minres_oblimin=minres_rot),
    function(z) data.frame(Max_pattern_difference=max(abs(z$P-primary$P)),
      Max_Phi_difference=max(abs(z$Phi-primary$Phi)),
      Min_congruence=min(colSums(z$P*primary$P)/sqrt(colSums(z$P^2)*colSums(primary$P^2))))))
  rownames(sensitivity) <- c("ML oblimin","ML geomin","Minres oblimin")
  # A single donor-aligned PAF update, separate from ML and the PA target.
  reduced <- R; diag(reduced) <- 1-1/diag(solve(R))
  ev <- eigen(reduced,symmetric=TRUE)
  paf1 <- ev$vectors[,1:3]%*%diag(sqrt(ev$values[1:3]))
  result <- list(input=input,n=n,p=p,R=R,S=S,pa=pa,fits=fits,rotations=rotations,
    primary=primary,varimax=varimax,geomin=geomin,minres=minres,minres_rot=minres_rot,
    candidates=candidates,optimizer_checks=optimizer_checks,common=common,h2=h2,u2=u2,reproduced=reproduced,residual=residual,
    structure=primary$P%*%primary$Phi,residual_pairs=residual_pairs,kmo=kmo,bartlett=bartlett,
    audit=audit,sensitivity=sensitivity,paf_update=data.frame(Item=names(x),
      SMC=diag(reduced),First_update=rowSums(paf1^2)),
    versions=vapply(c("psych","GPArotation"),function(p) as.character(packageVersion(p)),character(1)))
  result
}

verify_efa_results <- function(z) {
  checks <- list()
  add <- function(name,value) checks[[name]] <<- isTRUE(value)
  add("A only",identical(z$input$manifest$sample,"A"))
  add("360 by 12",identical(dim(z$input$x),c(360L,12L)))
  add("PA finite",all(is.finite(z$pa$null)))
  add("PA no failures",z$pa$failures==0L)
  add("moment counts",all(z$candidates$df==c(54,43,33,24,16)))
  add("independent optimizer convergence",all(z$candidates$Optimizer_check))
  add("independent ML objective",max(z$candidates$Objective_gap)<1e-6)
  add("rotation convergence",all(z$candidates$Rotated))
  add("primary admissible",min(z$u2)>0 && min(eigen(z$primary$Phi,symmetric=TRUE)$values)>0)
  add("structure",max(abs(z$structure-z$primary$P%*%z$primary$Phi))<1e-10)
  add("communalities",max(abs(z$h2-diag(z$common)))<1e-10)
  add("diagonal reconstruction",max(abs(z$h2+z$u2-1))<1e-6)
  add("residual identity",max(abs(z$R-z$reproduced-z$residual))<1e-12)
  A <- unclass(z$fits[[3]]$fit$loadings)
  for (name in c("primary","varimax","geomin")) {
    rot <- z[[name]]
    add(paste("rotation invariance",name),max(abs(rot$P%*%rot$Phi%*%t(rot$P)-tcrossprod(A)))<1e-6)
  }
  for (m in 1:5)
    add(paste("ML statistic",m),abs(z$candidates$Chi_square[m]-z$fits[[m]]$fit$STATISTIC)<1e-5)
  add("KMO",abs(z$kmo-psych::KMO(z$R)$MSA)<1e-10)
  add("Bartlett",abs(z$bartlett-psych::cortest.bartlett(z$R,z$n)$chisq)<1e-8)
  values <- unlist(checks)
  if(!all(values)) stop("Failed: ",paste(names(values)[!values],collapse=", "))
  values
}

# ---- render helpers --------------------------------------------------------

pt2_table <- function(x, digits=3, caption=NULL, row.names=FALSE) {
  x <- as.data.frame(x,check.names=FALSE)
  integer_columns <- c("df","Df","Factors","Moments","Free","npar","Dimension",
    "Effective_parameters","Candidate_df","Delta_df","Missing")
  for(nm in names(x)) if(is.numeric(x[[nm]])) {
    d <- if(nm %in% integer_columns) 0 else digits
    x[[nm]] <- formatC(x[[nm]],digits=d,format="f")
  }
  names(x) <- gsub("_"," ",names(x),fixed=TRUE)
  out <- knitr::kable(x,caption=caption,row.names=row.names,booktabs=TRUE,longtable=FALSE)
  if(knitr::is_latex_output()) knitr::asis_output(paste0(
    "\\Needspace{",min(nrow(x)+7,25),"\\baselineskip}\n\n",
    paste(out,collapse="\n"),"\n\n")) else out
}

pt2_sle_appendix_table <- function(x,widths) {
  stopifnot(ncol(x)==length(widths))
  if(!knitr::is_latex_output()) return(knitr::kable(x,row.names=FALSE))
  out <- knitr::kable(x,format="latex",booktabs=TRUE,longtable=TRUE,
    align=rep("l",ncol(x)),row.names=FALSE,linesep="")
  old <- paste0("\\begin{longtable}{",paste(rep("l",ncol(x)),collapse=""),"}")
  columns <- paste0("p{",widths,"\\linewidth}",collapse="")
  out <- sub(old,paste0("\\begin{longtable}{@{}",columns,"@{}}"),out,fixed=TRUE)
  # repeat the header row on every continuation page
  out <- sub("\\midrule","\\midrule\n\\endhead",out,fixed=TRUE)
  knitr::asis_output(paste0("\\begingroup\\small\n",out,"\n\\endgroup\n"))
}

pt2_num <- function(x,digits=3) formatC(x,digits=digits,format="f")
pt2_p <- function(x) ifelse(x<.001,"< .001",paste0("= ",pt2_num(x)))

pt2_heatmap <- function(M,title="",limit=max(abs(M)),digits=NULL,cex=.7) {
  nr <- nrow(M); nc <- ncol(M)
  par(mar=c(4,4,.9,1))
  palette <- grDevices::colorRampPalette(c("#8c510a","#f7f7f7","#2166ac"))(101)
  image(1:nc,1:nr,t(M[nr:1,,drop=FALSE]),zlim=c(-limit,limit),
        col=palette,axes=FALSE,xlab="",ylab="",main=title)
  axis(1,at=1:nc,labels=colnames(M),las=2,cex.axis=cex)
  axis(2,at=1:nr,labels=rev(rownames(M)),las=2,cex.axis=cex)
  if (!is.null(digits)) for (i in 1:nr) for (j in 1:nc) if(is.finite(M[i,j]))
    text(j,nr+1-i,formatC(M[i,j],digits=digits,format="f"),
         cex=cex*.8,col=if(abs(M[i,j])>limit*.6) "white" else "black")
  box()
}

pt2_pa_plot <- function(pa) {
  par(mfrow=c(1,2),mar=c(4,4,2,1))
  for (j in 1:2) {
    ylim <- range(c(pa$observed[,j],pa$reference[,j]))
    plot(1:nrow(pa$observed),pa$observed[,j],type="b",pch=16,ylim=ylim,
      xlab="Ordered dimension",ylab="Eigenvalue",
      main=c("Full Pearson matrix","SMC-reduced matrix")[j],cex.main=.95)
    lines(1:nrow(pa$observed),pa$reference[,j],lty=2,lwd=2,col="#9c5600")
    abline(h=0,col="gray80")
    legend("topright",c("Observed","Permutation 95th percentile"),
      lty=c(1,2),pch=c(16,NA),col=c("black","#9c5600"),bty="n",cex=.65)
  }
}

pt2_neighbor_plot <- function(z) {
  par(mfrow=c(1,3))
  for (m in 2:4) {
    P <- z$rotations[[m]]$P
    colnames(P) <- paste0("F",1:m)
    pt2_heatmap(P,paste(m,"ML factors"),limit=1,digits=2,cex=.8)
  }
}

pt2_rotation_plot <- function() {
  A <- rbind(c(.8,0),c(.7,.1),c(0,.6),c(.1,.7))
  Q <- matrix(c(1,1,-1,1),2,2)/sqrt(2)
  T <- matrix(c(1,0,.5,1),2,2)
  sets <- list(A,A%*%Q,A%*%solve(T))
  par(mfrow=c(1,3),mar=c(4,3,2,1))
  for (k in 1:3) {
    plot(sets[[k]],xlim=c(-.3,1.1),ylim=c(-.8,1),pch=16,
      xlab="Loading coordinate 1",ylab="Loading coordinate 2",
      main=c("Original","Orthogonal Q","Inverse-paired T")[k],cex.main=.9)
    abline(h=0,v=0,col="gray70")
    text(sets[[k]],labels=paste0("Y",1:4),pos=3,cex=.7)
  }
}

pt2_course_map <- function(labels) {
  n <- length(labels); par(mar=c(0,0,0,0))
  plot.new(); plot.window(xlim=c(0,n),ylim=c(0,1))
  for(i in seq_len(n)) {
    rect(i-.95,.25,i-.15,.75,col="#edf2f5",border="#405768")
    text(i-.55,.5,labels[i],cex=.8)
    if(i<n) arrows(i-.13,.5,i+.03,.5,length=.08)
  }
}

# ---- shared Part Two path-diagram grammar ----------------------------------
# Rectangles are recorded variables, circles are variables introduced by the
# model, single-headed arrows are loadings, double-headed arrows covariances.

pt2_ink        <- "#243746"
pt2_box_fill   <- "#fff1c7"
pt2_circ_fill  <- "#dcecf5"
pt2_path_col   <- "#2c5d77"
pt2_cross_col  <- "#94a3ad"
pt2_cov_col    <- "#8c4f2b"
pt2_resid_col  <- "#596b75"

# Render a label that may be plain text (".70", "0") or a plotmath
# expression string ("lambda[11]", "eta[1]").
pt2_greek <- c("alpha","beta","gamma","delta","epsilon","zeta","eta","theta",
  "iota","kappa","lambda","mu","nu","xi","pi","rho","sigma","tau","upsilon",
  "phi","chi","psi","omega","Gamma","Delta","Theta","Lambda","Xi","Pi",
  "Sigma","Phi","Psi","Omega")

# Render as plotmath only where the label is meant as notation: a subscripted
# token such as eta[1] or Planning[s], or a bare Greek name. Everything else is
# literal, so a factor named "Self-efficacy" is not parsed as a subtraction.
.pt2_label <- function(x) {
  if (is.expression(x) || is.language(x)) return(x)
  if (!nzchar(x)) return("")
  if (grepl("^[-.0-9]", x)) return(x)
  notation <- grepl("^[A-Za-z]+\\[[A-Za-z0-9]+\\]$", x) || x %in% pt2_greek
  if (!notation) return(x)
  out <- tryCatch(parse(text = x), error = function(e) NULL)
  if (is.null(out)) x else out
}

# Half-width of a factor node. Long substantive names get an ellipse rather
# than spilling out of a fixed circle, following the reference decks.
.pt2_frx <- function(label, r = .55) {
  n <- nchar(gsub("\\[|\\]", "", as.character(label)))
  max(r, .105 * n + .22)
}

.pt2_arrow <- function(x0, y0, x1, y1, dashed = FALSE, both = FALSE,
                       col = pt2_path_col, lwd = 1.35) {
  arrows(x0, y0, x1, y1, length = .07, angle = 20,
         code = if (both) 3 else 2,
         lty = if (dashed) 2 else 1, lwd = lwd, col = col)
}

.pt2_box <- function(x, y, label, hw = .74, hh = .30) {
  rect(x - hw, y - hh, x + hw, y + hh,
       col = pt2_box_fill, border = pt2_ink, lwd = 1.35)
  text(x, y, .pt2_label(label), cex = .92)
}

.pt2_circle <- function(x, y, label, r = .55) {
  rx <- .pt2_frx(label, r)
  tt <- seq(0, 2 * pi, length.out = 96)
  polygon(x + rx * cos(tt), y + r * sin(tt), col = pt2_circ_fill,
          border = pt2_ink, lwd = 1.35)
  text(x, y, .pt2_label(label), cex = .92)
}

# Draw a label at a fraction along a path, nudged perpendicular to it.
.pt2_path_label <- function(x0, y0, x1, y1, label, at = .58, nudge = .15,
                            col = pt2_ink, cex = .70) {
  if (is.null(label) || !nzchar(label)) return(invisible(NULL))
  text(x0 + at * (x1 - x0), y0 + at * (y1 - y0) + nudge,
       .pt2_label(label), cex = cex, col = col)
}

# A double-headed covariance arc drawn to the left of two circles.
.pt2_cov_arc <- function(x, y1, y2, label, r = .55, bulge = .85) {
  tt <- seq(0, 1, length.out = 60)
  ym <- (y1 + y2) / 2
  half <- abs(y2 - y1) / 2 - r * .55
  ys <- ym + half * cos(pi * tt) * sign(y1 - y2)
  xs <- x - bulge * sin(pi * tt)
  lines(xs, ys, col = pt2_cov_col, lwd = 1.4)
  .pt2_arrow(xs[2], ys[2], xs[1], ys[1], col = pt2_cov_col, lwd = 1.4)
  n <- length(xs)
  .pt2_arrow(xs[n - 1], ys[n - 1], xs[n], ys[n], col = pt2_cov_col, lwd = 1.4)
  if (!is.null(label) && nzchar(label))
    text(x - bulge - .14, ym, .pt2_label(label), adj = 1, cex = .82)
  invisible(NULL)
}

# Draw a double-headed arc for every distinct factor pair. Wider separations
# bow further out so that the arcs nest instead of overlapping, matching the
# reference decks' one-sided layout for three or more factors.
.pt2_cov_pairs <- function(x, yf, labels = NULL) {
  m <- length(yf)
  if (m < 2) return(invisible(NULL))
  pairs <- which(upper.tri(matrix(0, m, m)), arr.ind = TRUE)
  span <- pairs[, 2] - pairs[, 1]
  ord <- order(span, pairs[, 1])
  pairs <- pairs[ord, , drop = FALSE]; span <- span[ord]
  labels <- if (is.null(labels)) rep("", nrow(pairs)) else
    rep_len(labels, nrow(pairs))
  bulge <- if (m == 2) .80 else .50 + 1.10 * (span - 1) / (m - 2)
  for (i in seq_len(nrow(pairs)))
    .pt2_cov_arc(x, yf[pairs[i, 1]], yf[pairs[i, 2]], labels[i],
                 bulge = bulge[i])
  invisible(NULL)
}

.pt2_canvas <- function(xlim, ylim) {
  op <- par(mar = rep(.15, 4), xaxs = "i", yaxs = "i")
  plot.new()
  plot.window(xlim = xlim, ylim = ylim)
  op
}

.pt2_indicator_y <- function(p, top = 7.6, bottom = .9) {
  if (p == 1) return(mean(c(top, bottom)))
  seq(top, bottom, length.out = p)
}

# ---------------------------------------------------------------------------
# First-order model with m correlated (or orthogonal) factors.
#
#   assign  integer vector of length p giving each indicator's primary factor,
#           or a p x m logical matrix for a fully general pattern.
#   cross   optional p x m logical matrix of additional (dashed) loadings.
# ---------------------------------------------------------------------------
pt2_fa_diagram <- function(assign, m = NULL, cross = NULL,
                           loading_labels = NULL, factor_labels = NULL,
                           indicator_labels = NULL, factor_cov_labels = NULL,
                           show_residuals = TRUE, x_factor = 2.2,
                           x_indicator = 7.0) {
  if (is.matrix(assign)) {
    primary <- assign
    p <- nrow(primary); m <- ncol(primary)
  } else {
    p <- length(assign); if (is.null(m)) m <- max(assign)
    primary <- matrix(FALSE, p, m)
    primary[cbind(seq_len(p), assign)] <- TRUE
  }
  if (is.null(cross)) cross <- matrix(FALSE, p, m)
  if (is.null(factor_labels))
    factor_labels <- paste0("eta[", seq_len(m), "]")
  if (is.null(indicator_labels))
    indicator_labels <- paste0("Y[", seq_len(p), "]")

  yi <- .pt2_indicator_y(p)
  yf <- if (m == 1) mean(range(yi)) else
    seq(max(yi) - .9, min(yi) + .9, length.out = m)
  x_resid <- x_indicator + 3.0
  op <- .pt2_canvas(c(-.7, x_resid + 1.0), c(.3, 8.2))
  on.exit(par(op))

  # every factor pair is drawn, nested on the left
  if (m > 1 && !is.null(factor_cov_labels))
    .pt2_cov_pairs(x_factor - .55, yf, factor_cov_labels)

  for (j in seq_len(p)) for (k in seq_len(m)) {
    is_cross <- isTRUE(cross[j, k])
    if (!isTRUE(primary[j, k]) && !is_cross) next
    .pt2_arrow(x_factor + .pt2_frx(factor_labels[k]), yf[k],
               x_indicator - .74, yi[j], dashed = is_cross,
               col = if (is_cross) pt2_cross_col else pt2_path_col,
               lwd = if (is_cross) 1.1 else 1.35)
    if (!is.null(loading_labels))
      .pt2_path_label(x_factor + .pt2_frx(factor_labels[k]), yf[k],
                      x_indicator - .74, yi[j],
                      loading_labels[j, k], nudge = if (is_cross) -.16 else .16,
                      col = if (is_cross) pt2_cross_col else pt2_ink)
  }

  for (j in seq_len(p)) {
    .pt2_box(x_indicator, yi[j], indicator_labels[j])
    if (show_residuals) {
      .pt2_arrow(x_resid - .52, yi[j], x_indicator + .74, yi[j],
                 col = pt2_resid_col, lwd = 1.15)
      text(x_resid, yi[j], .pt2_label(paste0("epsilon[", j, "]")), cex = .85)
    }
  }
  for (k in seq_len(m)) .pt2_circle(x_factor, yf[k], factor_labels[k])
  invisible(NULL)
}

# ---------------------------------------------------------------------------
# Higher-order model: one second-order factor above m first-order factors.
# ---------------------------------------------------------------------------
pt2_higher_order_diagram <- function(assign, m = NULL, gamma_labels = NULL,
                                     loading_labels = NULL,
                                     factor_labels = NULL,
                                     indicator_labels = NULL,
                                     higher_label = "xi", cross = NULL,
                                     show_disturbances = TRUE) {
  p <- length(assign); if (is.null(m)) m <- max(assign)
  if (is.null(factor_labels)) factor_labels <- paste0("eta[", seq_len(m), "]")
  if (is.null(indicator_labels)) indicator_labels <- paste0("Y[", seq_len(p), "]")
  x_high <- 1.0; x_factor <- 4.0; x_indicator <- 8.0; x_resid <- 10.8

  yi <- .pt2_indicator_y(p)
  hh <- if (p > 1) min(.30, .40 * abs(yi[1] - yi[2])) else .30
  yf <- seq(max(yi) - .9, min(yi) + .9, length.out = m)
  yh <- mean(range(yf))
  # leave room for the widest node on each side so nothing is clipped
  left <- min(0, x_high - .pt2_frx(higher_label) - .35)
  op <- .pt2_canvas(c(left, x_resid + .9), c(.3, 8.2))
  on.exit(par(op))

  for (k in seq_len(m)) {
    .pt2_arrow(x_high + .pt2_frx(higher_label), yh,
               x_factor - .pt2_frx(factor_labels[k]), yf[k])
    if (!is.null(gamma_labels))
      .pt2_path_label(x_high + .pt2_frx(higher_label), yh,
                      x_factor - .pt2_frx(factor_labels[k]), yf[k],
                      gamma_labels[k], at = .55, nudge = .17)
    if (show_disturbances) {
      .pt2_arrow(x_factor, yf[k] + 1.05, x_factor, yf[k] + .55,
                 col = pt2_resid_col, lwd = 1.15)
      text(x_factor, yf[k] + 1.24,
           .pt2_label(paste0("zeta[", k, "]")), cex = .82)
    }
  }
  if (!is.null(cross) && nrow(cross)) for (i in seq_len(nrow(cross))) {
    k <- match(cross$factor[i], factor_labels)
    j <- match(cross$item[i], indicator_labels)
    .pt2_arrow(x_factor + .pt2_frx(factor_labels[k]), yf[k],
               x_indicator - .74, yi[j], dashed = TRUE,
               col = pt2_cov_col, lwd = 1.3)
  }
  for (j in seq_len(p)) {
    k <- assign[j]
    .pt2_arrow(x_factor + .pt2_frx(factor_labels[k]), yf[k],
               x_indicator - .74, yi[j])
    if (!is.null(loading_labels))
      .pt2_path_label(x_factor + .pt2_frx(factor_labels[k]), yf[k],
                      x_indicator - .74, yi[j],
                      loading_labels[j], nudge = .16)
    .pt2_box(x_indicator, yi[j], indicator_labels[j], hh = hh)
    .pt2_arrow(x_resid - .52, yi[j], x_indicator + .74, yi[j],
               col = pt2_resid_col, lwd = 1.15)
    text(x_resid, yi[j], .pt2_label(paste0("epsilon[", j, "]")), cex = .85)
  }
  for (k in seq_len(m)) .pt2_circle(x_factor, yf[k], factor_labels[k])
  .pt2_circle(x_high, yh, higher_label)
  invisible(NULL)
}

# ---------------------------------------------------------------------------
# Bifactor model: one general factor on every indicator, plus orthogonal
# group factors on their own blocks.
# ---------------------------------------------------------------------------
pt2_bifactor_diagram <- function(assign, m = NULL, general_label = "eta[G]",
                                 group_labels = NULL, indicator_labels = NULL,
                                 general_loadings = NULL,
                                 group_loadings = NULL) {
  p <- length(assign); if (is.null(m)) m <- max(assign)
  if (is.null(group_labels))
    group_labels <- paste0("eta[", seq_len(m), "]")
  if (is.null(indicator_labels)) indicator_labels <- paste0("Y[", seq_len(p), "]")
  x_gen <- 1.0; x_group <- 4.2; x_indicator <- 8.0; x_resid <- 10.8

  yi <- .pt2_indicator_y(p)
  hh <- if (p > 1) min(.30, .40 * abs(yi[1] - yi[2])) else .30
  yg <- seq(max(yi) - .8, min(yi) + .8, length.out = m)
  ygen <- mean(range(yi))
  left <- min(0, x_gen - .pt2_frx(general_label) - .35)
  op <- .pt2_canvas(c(left, x_resid + .9), c(.3, 8.2))
  on.exit(par(op))

  # general factor reaches every indicator; drawn first and lighter so the
  # group structure stays readable
  for (j in seq_len(p)) {
    .pt2_arrow(x_gen + .pt2_frx(general_label), ygen,
               x_indicator - .74, yi[j], col = pt2_cross_col, lwd = 1.0)
    if (!is.null(general_loadings))
      .pt2_path_label(x_gen + .pt2_frx(general_label), ygen,
                      x_indicator - .74, yi[j],
                      general_loadings[j], at = .30, nudge = .15,
                      col = pt2_cross_col, cex = .64)
  }
  for (j in seq_len(p)) {
    k <- assign[j]
    .pt2_arrow(x_group + .pt2_frx(group_labels[k]), yg[k],
               x_indicator - .74, yi[j])
    if (!is.null(group_loadings))
      .pt2_path_label(x_group + .pt2_frx(group_labels[k]), yg[k],
                      x_indicator - .74, yi[j],
                      group_loadings[j], nudge = .16)
    .pt2_box(x_indicator, yi[j], indicator_labels[j], hh = hh)
    .pt2_arrow(x_resid - .52, yi[j], x_indicator + .74, yi[j],
               col = pt2_resid_col, lwd = 1.15)
    text(x_resid, yi[j], .pt2_label(paste0("epsilon[", j, "]")), cex = .85)
  }
  for (k in seq_len(m)) .pt2_circle(x_group, yg[k], group_labels[k])
  .pt2_circle(x_gen, ygen, general_label)
  text(x_gen, ygen - 1.15, "all factors orthogonal", cex = .68,
       col = pt2_resid_col)
  invisible(NULL)
}

# Survey of retention rules, computed on the same development record.
# Each rule states a reference; the rule is the comparison, not the number.
efa_retention_survey <- function(z) {
  R <- z$R; n <- z$n; p <- z$p
  observed <- z$pa$observed
  mean_ref <- apply(z$pa$null, c(1,2), mean)
  leading <- function(obs, ref) {
    ok <- obs > ref
    if (all(ok)) length(ok) else which(!ok)[1] - 1L
  }
  # Empirical Kaiser criterion: a sample-adjusted reference, floored at one.
  ekc_ref <- numeric(p)
  for (j in seq_len(p)) {
    explained <- if (j == 1L) 0 else sum(observed[seq_len(j-1), 1])
    ekc_ref[j] <- max(((1+sqrt(p/n))^2)*(p-explained)/(p-j+1), 1)
  }
  # Sequential chi-square model tests on the same ML fits already computed.
  smt_p <- z$candidates$p_value
  smt <- if (any(smt_p > .05)) which(smt_p > .05)[1] else NA_integer_
  survey <- data.frame(
    Rule = c("Kaiser-Guttman", "Empirical Kaiser criterion",
             "PA, component roots, mean", "PA, component roots, 95th",
             "PA, reduced roots, mean", "PA, reduced roots, 95th",
             "Sequential chi-square tests"),
    Reference = c("Root exceeds 1",
                  "Root exceeds a sample-adjusted bound, floored at 1",
                  "Root exceeds the mean null root",
                  "Root exceeds the 95th-percentile null root",
                  "Reduced root exceeds the mean null reduced root",
                  "Reduced root exceeds the 95th-percentile null reduced root",
                  "First non-significant exact-fit test"),
    Retained = c(sum(observed[,1] > 1), leading(observed[,1], ekc_ref),
                 leading(observed[,1], mean_ref[,1]),
                 leading(observed[,1], z$pa$reference[,1]),
                 leading(observed[,2], mean_ref[,2]),
                 leading(observed[,2], z$pa$reference[,2]),
                 smt),
    stringsAsFactors = FALSE)
  survey$Retained <- as.character(survey$Retained)
  list(survey=survey, ekc_ref=ekc_ref, mean_ref=mean_ref,
       mp_bound=(1+sqrt(p/n))^2)
}

z <- run_efa_analysis()
checks <- verify_efa_results(z)
P <- z$primary$P
Phi <- z$primary$Phi
retention <- efa_retention_survey(z)
```

{{< pagebreak >}}

::: {.callout-tip title="Chapter box"}
**Five themes.**

*The direction of work reverses.* The preceding chapter built a covariance
matrix from declared parameters. Here a sample correlation matrix comes
first, and no simple model reproduces it exactly.

*Every diagnostic compares against a reference.* Extraction, retention and
rotation each state what the data are being held up against. Naming the
reference is what makes a result interpretable, and what makes two
procedures with the same name different.

*Axes are chosen, not found.* Rotation changes the loading display
without changing the fitted covariance, so pattern coefficients, structure
correlations and factor correlations are three different quantities that
must be read together.

*Discrepancy has a location.* A residual correlation matrix says where a
retained solution strains; an average residual cannot.

*A development sample cannot confirm what it was used to develop.* The
analysis produces a candidate whose restrictions are stated explicitly, and
the evidence for it has to come from records not yet inspected.

**Prerequisites.** Sample covariance and correlation, elementary matrix
algebra, and the common-factor model, communalities, uniquenesses and
scale and rotational indeterminacies developed in the preceding chapter. Basic R helps
reproduction; the printed matrices, equations and output support the
reasoning without running it.

**Record.** Sample A contains 360 complete, simulated continuous records on
twelve Student Learning Experiences indicators. No observations concern
actual students. Samples B and C are not analyzed here.

**Scope.** We develop a candidate measurement model on Sample A and stop
there. Neither confirmation sample is analyzed or inspected for
model-development decisions, no restricted model is estimated,
and no fit index is interpreted as evidence for a construct. Indicators are
treated as continuous throughout Part Two. This is model development, not
instrument validation or individual scoring.
:::

# Define the exploratory measurement problem {#sec-efa-problem}

Chapter 1 began with known loadings and factor covariances and worked
forward to an implied covariance matrix. Here the direction of work
changes. We observe a sample of indicator values, calculate their
associations, and ask which common-factor representation is defensible.
The observed correlations fluctuate from sample to sample. Their pattern
does not arrive with a factor count, a preferred rotation, or substantive
names attached. Those have to be developed and documented.

Exploratory factor analysis, or **EFA**, leaves most loading restrictions
open while seeking a comparatively low-dimensional representation of
shared covariation. It is not a procedure with no assumptions. Selecting
the indicators, defining the population, choosing an association matrix,
restricting residual covariance, considering a range of dimensions, and
choosing rotation criteria all shape what the analysis can reveal. A
carefully designed exploratory study therefore has substantial work to
do before an extraction routine runs.

The distinction between a provisional representation and a confirmed
restriction is important from the outset. Suppose an EFA suggests that
an indicator of adaptive planning also relates to strategy adjustment.
That can be a useful discovery. It is not evidence that a cross-loading
specified after seeing these observations was predicted before them.
The later CFA can evaluate the resulting restriction on unused data,
but changing the name of the fitting function does not change the
history of the decision.

## The common-factor model is not restriction-free

For person or observational unit $i$, retain Chapter 1's notation:

$$
\mathbf y_i=\boldsymbol\nu+\boldsymbol\Lambda\boldsymbol\eta_i
 +\boldsymbol\epsilon_i,\qquad
\boldsymbol\Sigma=\boldsymbol\Lambda\boldsymbol\Phi
 \boldsymbol\Lambda^{\mathsf T}+\boldsymbol\Theta.
$$ {#eq-efa-model}

There are $p$ indicators and $m$ factors. Thus $\boldsymbol\Lambda$ is
$p\times m$, $\boldsymbol\Phi$ is $m\times m$, and
$\boldsymbol\Theta$ and $\boldsymbol\Sigma$ are $p\times p$.
The factor-residual covariance is zero. Classical EFA initially makes
$\boldsymbol\Theta$ diagonal: remaining indicator-specific variation
does not covary across indicators. That restriction can be wrong even
when most loadings are freely estimated.

The loading pattern is initially open, but its dimension is not. A
two-factor and a three-factor EFA are different covariance models.
Scale restrictions are also necessary. One convenient extraction
representation uses unit-variance, uncorrelated factors; a subsequent
oblique transformation expresses the same common covariance using
correlated axes. Orthogonality at this computational stage does
not establish independence of substantively named attributes.

The latent intercept $\boldsymbol\alpha$ equals the factor mean in
this model without structural paths or separate predictors. It remains
an intercept symbol throughout Part Two. Likewise $\boldsymbol\Phi$
denotes total factor covariance, not residual covariance and not the
later structural-disturbance matrix. These distinctions prevent the
same letter from silently changing meaning between EFA, CFA, and SEM.

## A twelve-indicator content record

The sustained example concerns **Student Learning Experiences (SLE)**.
The proposed domains are Planning, Strategy, and Belonging. These names
organize the starting content questions; they do not determine how many
factors Sample A will support. All twelve variables are explicitly
simulated continuous process indices. Their units are index points,
not response categories, percentages, or counts. The statements below
clarify intended content, but they are not presented as an administered
and validated questionnaire.

```{r}
#| label: tbl-efa-items
#| tbl-cap: "Proposed content before factor analysis. All indices are positively directed."
pt2_table(z$input$codebook[c("item","domain","wording")])
```

P4 concerns changing a plan when a learning method fails. S4 concerns
using discussion with others to improve a learning method. Their content
already suggests possible overlap between domains. A cross-loading
would not automatically make either indicator defective. Conversely,
clear wording does not guarantee that its covariance pattern will
match the intended domain.

The twelve columns have different raw dispersions. All are coded so
that larger values mean more of the stated attribute, and none is
reverse-scored. The simulation imposes neither missing observations
nor response limits. Its unbounded normal records are a deliberate
teaching design, not a claim that actual student experiences are
normally distributed or measured without limits.

Download [sle_continuous.csv](sle_continuous.csv), the only SLE data file
needed for the examples. The [SLE data appendix](#sec-sle-data) contains
the complete codebook and sample manifest. Its `sample` column identifies
three independent 360-case samples, not three groups to pool. Select A
before inspecting indicators or calculating statistics in this chapter.
Population parameters are documented in Appendix A, separate from the
learner-facing analysis: knowing the recipe is not a substitute for
reasoning from Sample A. No empirical
summary, correlation, factor solution, or fit statistic from B or C
has been used to make the decisions reported here.

## Check the indicators before interpreting correlations

Data examining is part of defining the statistical input. An unintended
reverse code can change signs. A nearly constant indicator can make a
correlation unstable. Duplicate records alter the apparent amount of
information. Pairwise deletion can produce different effective samples
for different correlations, and sometimes a matrix that is not positive
definite. None of these problems is repaired merely by rotating a
loading matrix until it looks recognizable.

```{r}
#| label: tbl-efa-audit
#| tbl-cap: "Sample A audit, in raw process-index points; skewness is the third standardized central moment."
pt2_table(z$audit,digits=2)
```

There are `r z$n` distinct rows, no missing values, and positive
variance in every indicator. The minimum eigenvalue of the Pearson
matrix is `r pt2_num(min(eigen(z$R,symmetric=TRUE)$values))`, so its
inverse exists without matrix repair. This is a useful computational
check, not evidence that the factor model is true. Normal-theory ML
also needs an appropriate sampling and distributional interpretation.
The simulation supplies independent normal factor/residual draws; a
real study would have to justify or investigate those assumptions.

```{r}
#| label: fig-efa-distributions
#| fig-width: 7
#| fig-height: 3.2
#| fig-cap: "Raw distributions in Sample A. Different dispersions motivate an explicit scaling choice."
#| fig-alt: "Boxplots of twelve continuous SLE indices, grouped by Planning, Strategy, and Belonging; all have nonzero dispersion."
par(mar=c(3,4,1,1))
boxplot(z$input$x,col=rep(c("#dce8ef","#eee5d3","#e5e5e5"),each=4),
        ylab="Process-index points",outline=TRUE)
```

In an actual application, an unusual observation deserves examination
of provenance and measurement, not automatic removal because it makes
normality less convenient. A miscoded value, a genuine extreme case,
and a member of another target population call for different actions.
Any exclusion changes the estimand's population or the analyzed record
and must be described. Here no item or case is removed. That keeps the
same twelve-variable target available for later confirmatory comparisons.

# Choose the matrix and extraction target {#sec-efa-target}

## Covariance and correlation answer different scaling questions

Let $\mathbf S$ denote the ordinary sample covariance matrix computed
with denominator $n-1$, and let
$\mathbf D_S=\operatorname{diag}(s_{11},\ldots,s_{pp})$.
The Pearson correlation matrix is

$$
\mathbf R=\mathbf D_S^{-1/2}\mathbf S\mathbf D_S^{-1/2}.
$$ {#eq-efa-standardization}

This operation divides each covariance by the two sample standard
deviations. It does not remove dependencies, correct measurement error,
or make a nonnormal distribution normal. It changes the units in which
the association structure is represented. Covariance-scale analysis
preserves raw variances; correlation-scale analysis gives each observed
variable a unit diagonal.

Our primary analysis uses $\mathbf R$. The rationale is that the
process-index dispersions are not intended to set the relative weight
of the content domains. Raw $\mathbf S$, means, and standard deviations
are retained, however. Later CFA fits raw observations, and an explicit
scale transformation connects those fits with this standardized EFA.
No variable is replaced by an ordinal score or a new construct along
the way.

The observations are synthetic samples generated from a prespecified
continuous latent-variable population, and the exact released records are
distributed as `sle_continuous.csv`. Throughout the fragments below, `X` is
Sample A's 360-by-12 indicator matrix and `R` is its correlation matrix:

```r
sle <- read.csv("sle_continuous.csv")

items <- c(paste0("P", 1:4),
           paste0("S", 1:4),
           paste0("B", 1:4))

X <- sle[sle$sample == "A", items]
R <- cor(X)
```
 For
these complete observations, standardizing the covariance matrix gives the
same Pearson matrix as direct calculation. That identity checks the
transformation, not the suitability of Pearson correlation for every dataset.
Numeric codes for categories would require a separate
measurement treatment; that topic belongs to other courses or later
series. Categorical group labels and coded predictors remain possible
in later multigroup or MIMIC work. This brief boundary introduces no
categorical-estimator track here.

```{r}
#| label: fig-efa-correlation
#| fig-pos: "H"
#| fig-width: 6.5
#| fig-height: 4.2
#| fig-cap: "Sample A Pearson correlations, with a fixed scale from -1 to 1. The ordering follows the content map, not a fitted cluster algorithm."
#| fig-alt: "Twelve-by-twelve Pearson matrix with generally stronger within-domain correlations and positive relations between domains."
pt2_heatmap(z$R,limit=1,digits=2,cex=.8)
```

The broad blocks are suggestive, not conclusive. A correlation between
a Planning and a Strategy indicator might arise from correlated
factors, a cross-loading, residual dependence, or some combination.
The same visible association can therefore be represented by different
restrictions. The task is not to give each colored block a name and
stop; it is to evaluate a coherent factor representation of the matrix.

## PCA: a decomposition of total observed variance

For a symmetric positive-definite correlation matrix,

$$
\mathbf R=\mathbf V\mathbf D\mathbf V^{\mathsf T},\qquad
\mathbf V^{\mathsf T}\mathbf V=\mathbf I,
$$ {#eq-efa-spectral}

where the columns of $\mathbf V$ are orthonormal eigenvectors and
$\mathbf D=\operatorname{diag}(d_1,\ldots,d_p)$ contains ordered
eigenvalues. A rank-$m$ PCA reconstruction uses

$$
\mathbf A_m=\mathbf V_m\mathbf D_m^{1/2},\qquad
\widehat{\mathbf R}_{\mathrm{PCA},m}=\mathbf A_m\mathbf A_m^{\mathsf T}.
$$ {#eq-efa-pca}

The full decomposition is exact; a truncated one generally is not.
PCA chooses directions of high observed variance. Component scores
are weighted functions of observed variables, not unobserved random
variables whose common contribution is separated from item residuals.
Calling $\mathbf A_m$ a loading matrix in software does not change
that distinction.

The first three Sample A component eigenvalues sum to
`r pt2_num(sum(z$pa$observed[1:3,1]))`, or
`r pt2_num(100*sum(z$pa$observed[1:3,1])/12,1)`% of the twelve units
of standardized observed variance. This is a PCA summary. It is not
the common variance attributed to three latent factors by the ML model.
The two summaries can differ because their objectives differ, even
when their broad dimensional patterns look similar.

PCA may be appropriate when compact prediction, compression, or a
descriptive orientation is the objective. The problem is not
using PCA; it is interpreting its output as if a common-factor
measurement model had been estimated. A method should be judged
against the task it actually solves.

## Principal-axis factoring: estimate the diagonal contribution

Principal-axis factoring, or PAF, works with a reduced correlation
matrix. At iteration $t$, replace the unit diagonal by estimated
communalities:

$$
\mathbf R_c^{(t)}=\mathbf R-
 \operatorname{diag}(1-\widehat{\mathbf h}^{2(t)}).
$$ {#eq-efa-paf}

Decompose this matrix, use its leading positive eigenvalues to form
a provisional loading matrix, and update the communalities from its
row sums of squared loadings in an orthogonal representation. Repeat
until a declared rule indicates sufficiently small changes. The
initial diagonal, handling of nonpositive roots, and stopping rule
are implementation choices, not properties of factor analysis itself.

One common initialization is the squared multiple correlation of
each variable with the remaining variables. This quantity is the **squared
multiple correlation**, written SMC throughout: for indicator $j$ it is the
$R^2$ from regressing $Y_j$ on all eleven other indicators, so it says how much
of that indicator the rest of the set can already account for. For nonsingular
$\mathbf R$,

$$
\widehat h_{j,0}^{2}=1-
 \frac{1}{(\mathbf R^{-1})_{jj}}.
$$ {#eq-efa-smc}

These are regression summaries of the observed matrix, not known
common variances. They can underestimate some true communalities and
can be unstable when predictors are highly redundant. They provide
a starting point, not an answer to the factor-count question.

```{r}
#| label: tbl-efa-paf-update
#| tbl-cap: "A single three-factor PAF-style spectral update from an SMC diagonal. This is an illustration, not the primary ML fit or a converged PAF solution."
pt2_table(z$paf_update)
```

For P1, the SMC start is `r pt2_num(z$paf_update$SMC[1])` and the
first updated communality is `r pt2_num(z$paf_update$First_update[1])`.
Repeating that update would change the reduced diagonal again. This
explains why a PAF root is not automatically the same as a root of
the original unit-diagonal correlation matrix.

## Minimum residual and maximum likelihood extraction

Minimum-residual extraction seeks small discrepancies between observed
and reproduced off-diagonal correlations. A representative objective is

$$
F_{\mathrm{minres}}=
 \sum_{j<k}\left[r_{jk}-(\boldsymbol\Lambda\boldsymbol\Phi
 \boldsymbol\Lambda^{\mathsf T})_{jk}\right]^2.
$$ {#eq-efa-minres}

The diagonal residual variances then complete the observed variances.
This criterion gives the same squared-error weight to each displayed
off-diagonal discrepancy. Software algorithms can differ in numerical
details; the objective, convergence criteria, and resulting covariance
still need to be checked. PAF and minres are both common-factor
approaches, but they are not names for the same algorithm.

The primary method here is **maximum-likelihood common-factor
extraction**. For an observed covariance matrix $\mathbf S$ and a
positive-definite model-implied covariance
$\boldsymbol\Sigma(\boldsymbol\vartheta)$, the normal-theory
discrepancy can be written

$$
F_{\mathrm{ML}}(\boldsymbol\vartheta)=
 \log|\boldsymbol\Sigma(\boldsymbol\vartheta)|
 +\operatorname{tr}\!\left[\mathbf S\boldsymbol\Sigma(\boldsymbol\vartheta)^{-1}\right]
 -\log|\mathbf S|-p.
$$ {#eq-efa-ml}

Here $\boldsymbol\vartheta$ collects free parameters; it is not the
residual matrix $\boldsymbol\Theta$. The determinant terms measure
covariance volume, while the trace term measures discrepancy relative
to the model's covariance geometry. The two sample-only terms set
the discrepancy to zero for exact reproduction. Minimization chooses
parameter values under the model's restrictions, rather than choosing
a covariance matrix independently of them.

The same expression applies to our correlation-scale analysis with
$\mathbf R$ in place of $\mathbf S$. Under freely rescalable loadings
and residual variances, transforming observed units has a corresponding
parameter transformation. This does not mean every arbitrary
constraint is invariant to units. Equal unstandardized loadings,
for example, must be re-expressed when the indicators change scale.

Normal ML supplies a distributional framework for standard errors and
test statistics when its assumptions and regularity conditions hold.
Independent observations, a suitable population sampling interpretation,
identification, and admissible covariance matrices matter. Good-looking
loadings cannot supply any of these by themselves. Numerical convergence
only says that the optimizer met a stopping rule; it does not show that
the fitted representation is adequate or that the solution is uniquely
identified.

| Method | Immediate target | What the output does not establish |
|:--|:--|:--|
| PCA | Low-rank reconstruction of total observed variance | A latent measurement decomposition |
| PAF | Iterative reduced-matrix factor representation | A universal diagonal start or stopping tolerance |
| Minres | Small off-diagonal squared residuals | Normal-likelihood inference without additional work |
| ML factor analysis | Best normal likelihood within the factor model | Substantive truth, validity, or acceptable fit merely from convergence |

The maintained sequence is therefore Pearson correlations, ML
extraction, and oblimin rotation. Minres is a bounded sensitivity
comparison on the same matrix. The later CFA uses the same continuous
ML family; it is not a transition from one response model to another.

# Determine a plausible factor count {#sec-efa-retention}

## Underfactoring and overfactoring have different costs

Too few factors force distinctions into an inadequate space. Indicators
of an omitted dimension can acquire misleading loadings on the retained
dimensions, factor correlations can shift, and structured residuals can
remain. Too many factors can divide one broad pattern into weak or
unstable directions, make residual variances approach a boundary, and
encourage names for sample-specific features. Adding a dimension is not
free just because numerical reconstruction improves.

Retention is therefore a question about the representation to carry
forward, not a race to maximize variance accounted for. We consider
one through five factors before examining the results. Three is a
content-based expectation, not a required outcome of the analysis.
The full range is retained in the record even when neighboring models
give inconvenient results.

## Specify the matrix whose eigenvalues are being read

A spectral eigenvalue satisfies

$$
\mathbf R\mathbf v_k=d_k\mathbf v_k.
$$ {#eq-efa-eigenvalue}

Its meaning depends on the matrix. Eigenvalues of the full Pearson
matrix include all standardized observed variance and sum to $p$.
Eigenvalues of an SMC-reduced matrix use a different diagonal and need
not all be positive. An estimated common covariance has yet another
spectrum. A printed column headed "eigenvalue" must therefore be
interpreted alongside the software's construction of that column.

The post-rotation column statistic

$$
SS_k=\sum_{j=1}^p\lambda_{jk}^2
$$ {#eq-efa-ss}

is an **SS loading**, not generally a spectral eigenvalue. In suitable
unrotated orthogonal decompositions these quantities coincide; after
rotation the column allocation can change while common covariance
stays fixed. With correlated factors, pattern-column SS values do not
provide a unique additive partition of common variance. Chapter 1's
rotated two-factor example already demonstrated the simpler orthogonal
version of this distinction.

## Scree and the eigenvalue-greater-than-one heuristic

The familiar Kaiser--Guttman heuristic retains roots above one in a
unit-diagonal correlation analysis. It compares a component's variance
with one observed standardized variable's variance. Sampling variation
alone can create roots above one, so this is neither a necessary
condition for a useful common factor nor a reliable stand-alone count.
Applying the same threshold to a reduced matrix changes the comparison
and should not be treated as an interchangeable implementation.

A scree plot instead asks where large decreases give way to a flatter
sequence. It can reveal a distinct elbow, multiple plausible elbows,
or no clear break. The method involves judgment: a visually appealing
elbow does not supply a sampling reference. Full- and reduced-matrix
plots need not agree, even when they do in the present example.

In Sample A, the first three full-matrix roots are
`r paste(pt2_num(z$pa$observed[1:3,1]),collapse=", ")`, followed by
`r pt2_num(z$pa$observed[4,1])`. That drop makes three a plausible
candidate. It becomes more informative when compared with what equally
sized data containing no cross-variable dependence would produce.

## Parallel analysis compares against a matched null

The leading eigenvalues reported just above are comfortably greater than one. That is not by itself evidence of anything, because
*random* data behave the same way. Correlate twelve unrelated variables in a
sample of 360 and the largest eigenvalue will still exceed one, simply because
sampling noise produces some correlations by chance and the leading eigenvalue
collects whatever structure happens to be there. The question a retention rule
has to answer is not "is this eigenvalue large?" but "is it larger than chance
alone would produce, in a dataset of exactly this shape?"

Parallel analysis answers that question directly [@horn1965;
@bujaEyuboglu1992]. It builds datasets that have no common factors but match
the real one in the two respects that govern the size of a chance eigenvalue -
the sample size and the empirical marginal distribution of each variable -
computes their eigenvalues, and uses those as the yardstick. A dimension is retained while the observed eigenvalue stands
above what the no-factor datasets produced at the same rank.

**Building the null.** The trick is to destroy the relationships between
indicators while leaving everything else alone. Independently shuffling each
column does exactly that. After shuffling, item P1 still has its own 360
values, its own mean, spread and skew - but which respondent's P1 sits beside
which respondent's S3 is now arbitrary, so any correlation between them is
accidental. Repeating this many times gives a distribution of eigenvalues under
"same items, same sample size, no common structure."

**Choosing the threshold.** We compare each observed eigenvalue with the 95th
percentile of its null distribution, not the mean:

$$
d_k^{\mathrm{observed}}>
 Q_{.95}\{d_{k,1}^{\mathrm{null}},\ldots,d_{k,B}^{\mathrm{null}}\}.
$$ {#eq-efa-pa}

The difference matters. The mean is a central summary of the null
distribution, so an eigenvalue only a little above it clears the bar; the 95th
percentile asks the eigenvalue to exceed all but the most extreme five per cent
of chance results. The percentile reference is the more stringent of the two
and generally retains no more dimensions than the mean, though which performs
better depends on the population, the sample size and the strength of the
factors. They are two different decision rules, not two labels for one.

**Counting.** We retain the *leading consecutive run* of dimensions that clear
their references, and stop at the first that does not. Without that rule, an
isolated crossing among the small trailing eigenvalues - where observed and
null curves nearly coincide and either may edge ahead - would be counted as a
further meaningful dimension.

@tbl-efa-pa-settings records the choices that would change the answer if made
differently; anyone rerunning the analysis needs all four.

| Choice | Value used here | Why it matters |
|:--|:--|:--|
| Null construction | Independent within-column permutation, without replacement | Preserves each item's own distribution while removing between-item structure |
| Matrix compared | Pearson correlations, computed the same way for observed and null | Comparing an observed correlation matrix with a null covariance matrix is not a like-for-like test |
| Reference | 95th percentile | The mean retains more dimensions; the choice is a decision rule, not a detail |
| Replications and seed | 100, seed 12345 | Enough for a stable retained count in this example; the seed makes that count reproducible |

: *The parallel-analysis settings that affect the retained count.* {#tbl-efa-pa-settings}

A short note on implementation, because a function name can hide the rule it
applies. The installed `psych` version's `fa.parallel(sim = FALSE)` resamples
each column **with replacement**, which is a bootstrap rather than the
permutation described above. Both are defensible null constructions, but they
are not the same one, so this chapter uses an explicit loop:

```r
set.seed(12345)
null_roots <- replicate(100, {
  X0 <- vapply(X, function(v) sample(v, replace = FALSE), numeric(nrow(X)))
  eigen(cor(X0), symmetric = TRUE, only.values = TRUE)$values
})
reference_95 <- apply(null_roots, 1, quantile, probs = .95, type = 7)
```

Read from the inside out: `sample(v, replace = FALSE)` shuffles one column;
`vapply` does that to all twelve, producing one null dataset; `eigen(cor(X0))`
returns its twelve eigenvalues; `replicate` repeats the whole thing 100 times,
giving a 12-by-100 matrix of null eigenvalues; and the last line takes the 95th
percentile at each rank. A hundred replications is ample here - going to a
thousand moves the reference values by about $.01$ and changes no retained
count.

**Two targets, two panels.** @fig-efa-pa runs this comparison twice, and the
difference between its panels is the diagonal of the matrix being decomposed.
The left panel uses the full Pearson matrix, whose diagonal is all ones: its
eigenvalues describe *total* standardized variance, which is what a principal
component analysis decomposes. The right panel replaces that diagonal with SMCs
before taking eigenvalues, so it describes only the variance each indicator
shares with the others - closer to what a common-factor model tries to explain.
Reduced-matrix eigenvalues are smaller, and the later ones can be negative,
which is why the two panels have different vertical scales and why an observed
curve must only ever be read against the reference drawn beside it. The same
diagonal rule is applied to every permuted matrix as well, so each panel is an
internally consistent comparison.

```{r}
#| label: fig-efa-pa
#| fig-width: 7.2
#| fig-height: 3.6
#| fig-cap: "Scree and parallel-analysis evidence. Each observed sequence is compared only with its matching permutation reference."
#| fig-alt: "Two panels show full Pearson and SMC-reduced eigenvalues against their separate permutation 95th percentiles. The first three observed roots exceed their references in each panel."
pt2_pa_plot(z$pa)
```

```{r}
#| label: tbl-efa-pa
#| tbl-cap: "Leading observed and null-reference eigenvalues; full and reduced targets are not pooled."
pa_table <- data.frame(Dimension=1:5,Component=z$pa$observed[1:5,1],
  Component_95=z$pa$reference[1:5,1],Factor=z$pa$observed[1:5,2],Factor_95=z$pa$reference[1:5,2])
pt2_table(pa_table)
```

Both comparisons retain a leading run of three dimensions. For the
third component root, `r pt2_num(z$pa$observed[3,1])` exceeds
`r pt2_num(z$pa$reference[3,1])`; for the fourth,
`r pt2_num(z$pa$observed[4,1])` does not exceed
`r pt2_num(z$pa$reference[4,1])`. The reduced-matrix comparison
reaches the same count by a different set of numerical comparisons.
Agreement makes the candidate less dependent on this particular
diagonal choice, but does not turn the procedure into a proof that
three substantive attributes exist.

PA addresses a matched null reference, not every plausible alternative
population. Factor correlations, weak factors, minor dimensions,
sample size, and the chosen root algorithm can change its performance.
It should be read alongside content, neighboring solutions, and
residuals.

## A wider family of retention rules {#sec-efa-family}

Retention is the decision an exploratory analysis most often gets wrong, and
the familiar heuristics are among the weaker available options.
@auerswaldMoshagen2019 compared the main extraction criteria across realistic
conditions - correlated factors, unequal loadings, minor factors, non-normal
responses, varying sample sizes - and found the eigenvalue-greater-than-one
rule and the unaided scree plot performing poorly enough that neither should
carry a retention decision alone. Several better-behaved alternatives exist,
and they are worth knowing not as a longer checklist but because each states a
different reference, and knowing the reference is what lets a disagreement be
interpreted.

**Parallel analysis is a family, not a method.** Two choices generate four
distinct procedures. The roots may come from the full Pearson matrix, as in a
component analysis, or from a reduced matrix with communality estimates on the
diagonal, closer to the common-factor model. The null reference may be the
mean of the simulated roots or their 95th percentile. Across
@auerswaldMoshagen2019's conditions the component-based versions generally
outperformed the reduced-matrix versions, and the mean reference tends to
retain more dimensions than the percentile reference. All four share one
structural weakness: the reference data are uncorrelated by construction, so
after a strong first factor has absorbed much of the observed covariance, the
comparison for the second and later roots is no longer like for like, and a
genuine second dimension can be missed.

**The empirical Kaiser criterion** repairs the Kaiser--Guttman rule rather
than discarding it [@braekenVanAssen2017]. Under the null model the sample
eigenvalues of uncorrelated variables are described asymptotically by a
Marchenko--Pastur law
whose upper bound is $(1+\sqrt{p/n})^2$; this is the reference for the first
root. Later roots are corrected for the variance already accounted for, and
the reference is never allowed below one:

$$
\text{EKC}_j=\max\!\left\{(1+\sqrt{p/n})^{2}\,
\frac{p-\sum_{i<j} d_i}{p-j+1},\; 1\right\}.
$$ {#eq-efa-ekc}

With $p=12$ and $n=360$ the first reference is
`r pt2_num(retention$mp_bound)`, appreciably above one: a leading root of
$1.1$ would satisfy Kaiser--Guttman and fail the EKC. From the second root
onward the correction term falls below one in this record, so the floor
applies and the remaining references equal one. The criterion is therefore
stricter than Kaiser--Guttman at the first root and identical afterwards -
here, on this data, which is exactly the sort of detail that a printed count
conceals.

**Sequential model tests** use the exact-fit statistic rather than an
eigenvalue. Fit a one-factor model; if its test is significant, fit two, and
continue until a test is not significant. The logic is clean, but it carries a
caveat worth stating: once the number of factors exceeds the true number, the
regularity conditions behind the $\chi^2$ reference fail, so the sequence is
only trustworthy while it is still under-fitting. An earlier test that was
significant by chance therefore propagates into further over-extraction rather
than being corrected.

**Two further methods** appear in the comparison literature, though neither is
computed here. The Hull method chooses the count at which the balance
between fit and number of parameters stops improving, borrowing the convex-hull
idea from model selection [@lorenzoSeva2011]. Comparison data generates
reference datasets that *reproduce* the observed correlation matrix under a
known factorial structure of $1,2,3,\ldots$ factors, rather than under
independence, and retains the smallest count whose reference data resemble the
observed data [@ruscioRoche2012]. Both address the same weakness in parallel
analysis: an uncorrelated null is not the relevant comparison once one factor
is granted.

@tbl-efa-family applies seven of these rules to the present record. The four
parallel-analysis rows share the same permutation replications, so they differ
only in the target matrix and the reference statistic; Kaiser--Guttman, the
empirical Kaiser criterion and the sequential tests use no permutations at all,
and are included because they answer the same question from different
premises.

```{r}
#| label: tbl-efa-family
#| tbl-cap: "Seven retention rules on the same development record. What separates them is the reference each uses, not the data."
pt2_table(retention$survey)
```

Every rule returns three. That agreement is a fact about this record, not a
general property of the methods, and it is what a well-behaved simulated
population with three clearly separated factors is expected to produce. Real
records disagree often, which is why @auerswaldMoshagen2019 recommend
**combination rules** rather than any single criterion: when sequential model
tests agree with either the 95th-percentile component parallel analysis or the
empirical Kaiser criterion, the pair identified the correct count with high
consistency across their conditions. Sequential tests perform well with highly
correlated factors but over-extract in the presence of minor factors and under
non-normality; the other two are robust and do not over-extract, which is why
they pair well.

The value of a combination rule is not that it counts votes. It is that
disagreement is informative: when these criteria diverge, the dimensions are
genuinely hard to detect in the available evidence, and the appropriate
response is usually more data or a narrower claim rather than a casting vote.
A majority vote among heuristics cannot replace explaining what each
diagnostic is comparing.

# Estimate neighboring dimensional solutions

## Count effective parameters, not every printed loading

For $p$ indicators, the symmetric covariance matrix supplies
$u=p(p+1)/2$ distinct second moments. An exploratory model with $m$
orthonormal factors appears to contain $pm$ loadings and $p$ uniquenesses.
However, orthogonal rotations leave its implied covariance unchanged.
The $m(m-1)/2$ rotational degrees of freedom do not add independently
recoverable covariance parameters. The effective count is therefore

$$
t_m=pm+p-\frac{m(m-1)}2,
\qquad
df_m=\frac{p(p+1)}2-t_m
=\frac{(p-m)^2-p-m}{2}.
$$

This is a dimension count for the unrestricted common-factor covariance
model at regular points. An oblique rotation does not add factor
correlations to this count: it re-expresses the same fitted covariance.
Counting every oblique loading, every factor correlation, and every
uniqueness as independently identified would count the same freedom
more than once.

```{r}
#| label: tbl-efa-counts
#| tbl-cap: "Covariance-moment accounting for 12 indicators."
m <- 1:5
pt2_table(data.frame(Factors=m,Moments=78,
  Effective_parameters=12*m+12-m*(m-1)/2,
  Candidate_df=z$candidates$df),digits=0)
```

The nonnegative-df condition also yields the **Ledermann bound**, the largest
factor count the moments can generically support:

$$
m\leq\frac{2p+1-\sqrt{8p+1}}2 ,
\qquad\text{equivalently}\qquad
(p-m)^{2}\ \geq\ p+m .
$$ {#eq-ledermann}

The two forms state the same condition, and the second shows what it is doing:
the supply of distinct moments has to keep pace with the demand for
parameters. For $p=12$ the bound is about $7.58$, so at most seven factors,
and the accounting above passes through zero exactly there - three degrees of
freedom at seven factors, $-2$ at eight. The bound is not a recommendation to
extract seven, a count of positive eigenvalues, or a proof of identification.

## What the count leaves undetermined {#sec-efa-identification}

Chapter 1 separated counting from **identification**: counting can rule a model
out, but a nonnegative $df$ never establishes that the moments determine the
parameters. Exploratory analysis is the clearest illustration, because an
unrestricted factor model is *by construction* unidentified. There is really
one problem here rather than three - the model is invariant under a family of
transformations of the factors - but that family has a continuous part and two
discrete parts, and they show up differently in practice.

**Rotational indeterminacy** is the substantial one. For any nonsingular
$m\times m$ matrix $\mathbf T$,

$$
\boldsymbol\Lambda^{*}=\boldsymbol\Lambda\mathbf T^{-1},
\qquad
\boldsymbol\Phi^{*}=\mathbf T\boldsymbol\Phi\mathbf T^{\mathsf T}
$$ {#eq-efa-rotation}

leaves $\boldsymbol\Lambda\boldsymbol\Phi\boldsymbol\Lambda^{\mathsf T}$
unchanged, so a continuous family of loading matrices reproduces the observed
covariance identically. Restricting $\mathbf T$ to be orthogonal leaves a
family of dimension $m(m-1)/2$ - precisely the quantity subtracted in the
parameter count above. The count has therefore already paid for this freedom;
it has not removed it.

**Sign reversal** and **column permutation** are the discrete members of the
same family: both are special cases of $\mathbf T$, one a diagonal matrix of
$\pm 1$ and the other a permutation matrix. Reversing the sign of an entire loading column together with its
factor changes nothing observable, so $2^{m}$ sign patterns are equivalent;
and the factors carry no intrinsic order, so any of the $m!$ permutations of
the columns describes the same model. Neither costs a degree of freedom,
because neither is a continuous direction, which is why $df$ is silent about
them. They are still real. Two runs of the same routine on the same data can
return one solution with a column negated or two columns exchanged, and
comparing the two naively would suggest a difference that does not exist. It
is why the sensitivity comparisons later in this chapter align column order
and sign before anything is compared.

What an exploratory analysis identifies is therefore the *common space* the
factors span - not the particular axes that get printed. At regular,
generically identified parameter values the communalities and residual
variances are also determined, since they are invariant under the whole
transformation family; exceptional configurations, such as a boundary solution
or a rank-deficient loading matrix, are not covered by that statement. Rotation does not resolve the
indeterminacy; it selects one member of the family on stated grounds, which is
exactly why the criterion has to be declared rather than accepted as a
default. A confirmatory model takes the other route, fixing enough loadings in
advance that no nontrivial $\mathbf T$ preserves the pattern - at which point
identification stops being automatic and has to be argued.

## Fit one through five; inspect two, three, and four closely

We estimate normal-theory ML common-factor models for one through five
factors. Each fit uses the same Pearson matrix, $n=360$, SMC starts,
and the same uniqueness-bound convention. The primary extraction is
unrotated; rotation is a subsequent choice of axes. The printed code
shows the essential calls; the chapter source carries the convergence checks
and the full settings.

```r
fit3 <- psych::fa(R, nfactors = 3, n.obs = nrow(X),
                  fm = "ml", rotate = "none", SMC = TRUE,
                  smooth = FALSE, scores = "none")
A <- unclass(fit3$loadings)
```

Two details about this call repay attention, because both are easy to
misread. `psych::fa()` accepts `min.err` and `max.iter`, and it is tempting to
present them as convergence controls for every extraction method. Under
`fm = "ml"` in the installed version they do not govern the optimizer: the
same solution is returned whether they are set tightly or loosely. They
control the iterative principal-axis path instead. Second, the smallest
uniqueness this implementation will return under ML is bounded below at
$.005$, which is why the four- and five-factor solutions in the table below
report a minimum uniqueness at exactly that value rather than at zero. Neither
fact is a defect; both are reasons to read a function's documentation for the
method actually being used rather than for the function as a whole.

The implemented analysis also checks the ML objective using
`stats::factanal()` with five starting values. Agreement of independently
computed objectives is a useful numerical check, not proof that the
scientific model is correct. A numerical algorithm can converge perfectly
to an inadequate model.

```{r}
#| label: tbl-efa-neighbors
#| tbl-cap: "Neighboring ML solutions. RMS uses distinct off-diagonal correlation residuals."
pt2_table(z$candidates[,c("Factors","df","F_ML","RMS_offdiag",
  "Max_residual","Min_uniqueness","Boundary")])
```

The two-factor solution leaves a largest absolute residual correlation
of `r pt2_num(z$candidates$Max_residual[2])`, compared with
`r pt2_num(z$candidates$Max_residual[3])` for three factors. That is a
substantial change in the covariance reconstruction, not merely an
additional column of interpretable words. Adding a fourth factor reduces
the discrepancy further, but its smallest uniqueness is approximately
.005, at the numerical lower bound. The fifth-factor solution does the
same. A bound prevents an optimizer from reporting a negative uniqueness;
it does not make that boundary estimate substantively credible.

An interior uniqueness of .20 and a bound-constrained uniqueness of .005
are qualitatively different evidence. The latter can signal too many
factors, a highly redundant pair, sampling fluctuation, or a misspecified
common-factor model. Inspect its indicator and neighboring pattern before
assigning a new construct name. Do not silently omit the bound from the
report or describe it as evidence of nearly perfect measurement.

```{r}
#| label: fig-efa-neighbors
#| fig-width: 7.2
#| fig-height: 5.3
#| fig-cap: "Oblimin patterns for neighboring dimensions. Column signs and order are arbitrary across different factor counts; compare item groupings, not column numbers."
#| fig-alt: "Three heatmaps display all loadings in the two-, three-, and four-factor solutions, allowing comparison of merged and split item groups."
pt2_neighbor_plot(z)
```

Read this display in two directions. Down a column, ask which items
define a coherent factor. Across a row, ask whether an item has a stable
primary relation or is distributed across factors. A smaller model must
represent more covariance using fewer common directions; it may merge
domains or leave patterned residuals. A larger model can separate narrow
item-specific covariance that does not justify an additional broad
construct. In neither case is a short label an adequate diagnostic.

## An exact-fit test is one piece of evidence

For reference, `psych` reports a corrected ML statistic of the form

$$
T_m=\left[n-1-\frac{2p+5}{6}-\frac{2m}{3}\right]
\widehat F_{\mathrm{ML}}.
$$

Under the exact factor model and suitable regularity conditions, its
reference is approximately $\chi^2_{df_m}$. For the three-factor fit,
$T_3=$ `r pt2_num(z$candidates$Chi_square[3])` with 33 df, giving
$p=$ `r pt2_num(z$candidates$p_value[3])`. This does not establish that
the factor interpretation is true; it says that this discrepancy is not
unusual under the exact-model reference used by the test. Power and
misspecification size matter. A nonsignificant test cannot establish
equivalence or rule out consequential departures.

The fourth- and fifth-factor solutions reach a variance boundary, where
ordinary interior-point chi-square arguments need additional care. Their
large conventional $p$ values should not be used to endorse extra factors.
Nor should one subtract any two displayed factor tests and automatically
declare a regular nested likelihood-ratio test: changes of factor count
involve singular or boundary configurations under the smaller model.

Chapter 3 uses raw-data ML with an explicit normal-likelihood convention.
Its test multiplier is not this small-sample EFA correction. Comparability
of estimation families does not mean every package reports an identical
finite-sample test statistic.

# Resolve the orientation by rotation

## What rotation changes, and what it preserves

An unrotated solution is convenient for estimation but often inconvenient
for interpretation. Suppose the extracted common covariance is
$\mathbf A\mathbf A^{\mathsf T}$. For an orthogonal matrix $\mathbf Q$,
with $\mathbf Q^{\mathsf T}\mathbf Q=\mathbf I$,

$$
\mathbf P=\mathbf A\mathbf Q
\quad\Longrightarrow\quad
\mathbf P\mathbf P^{\mathsf T}
=\mathbf A\mathbf A^{\mathsf T}.
$$

The rows of $\mathbf P$ locate the indicators in a rotated axis
system. The reproduced covariance, communalities, uniquenesses, and
extraction discrepancy are unchanged. The individual loading coefficients
change because they refer to different axes.

The general nonsingular transformation requires a paired change. Let
$\boldsymbol\eta^*=\mathbf T\boldsymbol\eta$, where the original
factor covariance is $\mathbf I$. Then

$$
\mathbf P=\mathbf A\mathbf T^{-1},\qquad
\boldsymbol\Phi^*=\mathbf T\mathbf T^{\mathsf T},
$$

and

$$
\begin{aligned}
\mathbf P\boldsymbol\Phi^*\mathbf P^{\mathsf T}
&=\mathbf A\mathbf T^{-1}\mathbf T\mathbf T^{\mathsf T}
  (\mathbf T^{-1})^{\mathsf T}\mathbf A^{\mathsf T}\\
&=\mathbf A\mathbf A^{\mathsf T}.
\end{aligned}
$$

If the transformed factor variances are subsequently rescaled to one,
the loading columns must be rescaled correspondingly. Multiplying the
loadings by an arbitrary matrix while leaving $\boldsymbol\Phi$
unchanged is not an equivalent oblique rotation. This inverse-paired
transformation is the same principle introduced in Chapter 1.

```{r}
#| label: fig-efa-rotation
#| fig-width: 7.2
#| fig-height: 3.3
#| fig-cap: "A change of axes in a two-dimensional illustration. These are illustrative loadings, not a second empirical data set."
#| fig-alt: "Three panels show the same four indicators under original, orthogonal, and oblique factor coordinates; oblique coordinates require a matching factor covariance."
pt2_rotation_plot()
```

## Orthogonal and oblique simple structure

Simple structure favors interpretable differentiation: relatively few
substantial relations per item and identifiable groups of indicators.
It is an orientation criterion, not a new criterion of fit to the observed
covariance matrix. Rotation does not improve $F_{\mathrm{ML}}$.

Varimax is an orthogonal criterion. In an unnormalized version it
maximizes, across columns, the dispersion of squared loadings:

$$
V(\mathbf P)=\sum_{a=1}^m
\left\{\frac1p\sum_{j=1}^p p_{ja}^4
-\left(\frac1p\sum_{j=1}^p p_{ja}^2\right)^2\right\}.
$$

A column with some larger and some smaller squared loadings has a higher
criterion than a column whose squared loadings are all similar. The
orthogonality constraint remains in force; a clean-looking varimax
pattern does not establish that the substantive attributes are unrelated.

Oblique criteria permit factor correlations while seeking a simpler
pattern. Here the primary rotation is direct oblimin with $\gamma=0$
(quartimin), no Kaiser normalization, convergence tolerance $10^{-7}$,
up to 10,000 iterations, and 20 random starts. At $\gamma=0$, the
criterion discourages a row from having large squared coefficients in
several columns. It does not set the factor correlations to zero.

```r
set.seed(12345)
rot3 <- GPArotation::oblimin(A, gam = 0, normalize = FALSE,
                            eps = 1e-7, maxit = 10000,
                            randomStarts = 20)
pattern <- unclass(rot3$loadings)
factor_correlations <- rot3$Phi
```

Kaiser normalization, when requested, divides each row by its
communality-based length before optimizing and restores its scale
afterward. That changes the criterion's relative weighting of indicators.
It is not the same operation as standardizing the original observed
variables. Because implementation details can affect the chosen
orientation, the rotation name alone is not a sufficient specification.

Multiple starts reduce dependence on a single starting orientation but
do not prove that a global optimum was found. Signs and column order are
also arbitrary. Our reported solution aligns columns with the first
three content anchors in each domain and makes their summed primary
loadings positive. This is a labeling convention applied after fitting;
it is not an extra model restriction.

## Why the oblique solution is primary here

Planning, strategy adjustment, and belonging need not be independent.
The oblimin solution estimates the following factor correlations:

```{r}
#| label: tbl-efa-phi
#| tbl-cap: "Factor correlation matrix in the aligned three-factor oblimin solution."
pt2_table(data.frame(Factor=rownames(Phi),Phi,check.names=FALSE))
```

These are estimated relations between the chosen common-factor
axes. They are not causal paths, observed scale correlations, or
proof that the same orientation would be uniquely recovered in another
sample. Varimax provides a useful comparison precisely because it
represents the same common covariance with orthogonal axes.
The maximum absolute difference in reproduced common covariance between
the reported oblimin and varimax solutions is
`r format(max(abs(z$common-z$varimax$P%*%z$varimax$Phi%*%t(z$varimax$P))),scientific=TRUE,digits=2)`.
The near-zero value verifies invariance; it does not make their loading
tables interchangeable.

# Read an oblique factor system

## Pattern coefficients and structure correlations

With standardized indicators and unit-variance factors, the pattern
matrix $\mathbf P$ contains the coefficients in

$$
\mathbf z_i=\mathbf P\boldsymbol\eta_i+\boldsymbol\epsilon_i.
$$

Each row is a simultaneous linear decomposition. A coefficient describes
the relation of that factor to the indicator conditional on the other
factors in this representation. The structure matrix instead gives the
zero-order indicator--factor correlations:

$$
\mathbf C=\operatorname{Cov}(\mathbf z,\boldsymbol\eta)
=\mathbf P\boldsymbol\Phi.
$$

Since both sides have unit marginal variances, those covariances are
correlations. In an orthogonal solution $\boldsymbol\Phi=\mathbf I$,
so pattern and structure coincide. In an oblique solution they generally
do not.

```{r}
#| label: tbl-efa-pattern
#| tbl-cap: "All pattern coefficients. Small entries remain visible; none were suppressed at .30."
pt2_table(data.frame(Item=rownames(P),P,check.names=FALSE))
```

```{r}
#| label: tbl-efa-structure
#| tbl-cap: "Structure correlations for the same orientation."
pt2_table(data.frame(Item=rownames(P),z$structure,check.names=FALSE))
```

For P4, the Strategy structure correlation is

$$
c_{\mathrm{P4,S}}
=p_{\mathrm{P4,P}}\phi_{\mathrm{PS}}
 +p_{\mathrm{P4,S}}+p_{\mathrm{P4,B}}\phi_{\mathrm{BS}}.
$$

Substituting the unrounded estimates gives
`r pt2_num(z$structure["P4","Strategy"])`. The direct Strategy
pattern coefficient is only `r pt2_num(P["P4","Strategy"])`.
Much of the zero-order relation reflects P4's Planning loading together
with the Planning--Strategy association. Reading the structure matrix as
if every entry were a separate direct loading would therefore exaggerate
cross-loading complexity.

Conversely, dismissing every secondary pattern coefficient below .30
would conceal the two content-relevant relations under consideration.
P4 concerns revising a plan when a method fails; S4 concerns using others
to improve a learning strategy. Their secondary coefficients,
`r pt2_num(P["P4","Strategy"])` and
`r pt2_num(P["S4","Belonging"])`, are evidence to evaluate alongside
content, orientation sensitivity, and later independent estimation.
They are not established nonzero population coefficients merely because
we can give them plausible names.

## Communality is a quadratic form

For indicator $j$, let $\mathbf p_j^{\mathsf T}$ be row $j$ of the
pattern matrix. Its common variance is

$$
h_j^2=\mathbf p_j^{\mathsf T}\boldsymbol\Phi\mathbf p_j,
\qquad
u_j^2=1-h_j^2.
$$

For three standardized factors this expands to

$$
h_j^2=p_{j1}^2+p_{j2}^2+p_{j3}^2
 +2p_{j1}p_{j2}\phi_{12}
 +2p_{j1}p_{j3}\phi_{13}
 +2p_{j2}p_{j3}\phi_{23}.
$$

The cross-products are part of the common variance. Summing squared
pattern coefficients alone drops them. For P4, the row sum of squared
pattern coefficients is `r pt2_num(sum(P["P4",]^2))`, whereas the
communality is `r pt2_num(z$h2["P4"])`. Neither number should be
labeled the variance uniquely explained by Planning.

```{r}
#| label: tbl-efa-communalities
#| tbl-cap: "Common and unique variance in the standardized metric. Row sums of squared pattern coefficients are included only to expose the oblique distinction."
pt2_table(data.frame(Item=rownames(P),Pattern_SS=rowSums(P^2),
  Communality=z$h2,Uniqueness=z$u2))
```

The column sums of squared loadings reported by software are useful
descriptions of an orientation, but correlated factors do not provide a
unique additive allocation of common variance to separate constructs.
One cannot turn overlapping factor relations into mutually exclusive
percentages simply by dividing each column sum by $p$.

A low communality can mean that an indicator is weakly represented by
this common-factor system. It does not by itself show poor wording or
justify removal. An indicator may intentionally cover content absent
from the other items. Removing it can make the covariance model cleaner
while narrowing the construct. A high communality can likewise arise
from redundancy rather than desirable breadth.

## Reconstruct a pair, then the whole matrix

Under diagonal uniqueness, the reproduced correlation of distinct
indicators $j$ and $k$ is

$$
\widehat r_{jk}=\widehat{\mathbf p}_j^{\mathsf T}
\widehat{\boldsymbol\Phi}\widehat{\mathbf p}_k.
$$

All primary and secondary coefficients participate. For P4 and S4,
the observed correlation is `r pt2_num(z$R["P4","S4"])` and the
reproduced correlation is `r pt2_num(z$reproduced["P4","S4"])`.
The residual, defined as observed minus reproduced, is
`r pt2_num(z$residual["P4","S4"])`. Its sign is interpretable only
after the subtraction convention is stated.

```r
common <- pattern %*% factor_correlations %*% t(pattern)
communality <- diag(common)
reproduced <- common + diag(fit3$uniquenesses)
residual <- R - reproduced
```

This short calculation is more informative than memorizing the names of
several output tables. It links loadings, factor correlations,
communalities, uniquenesses, and residuals to one covariance equation.
When a table cannot be reconstructed from that equation, first check
the standardization and rotation conventions rather than inventing a
substantive explanation.

## What is uncertain about this interpretation?

Sampling affects extraction, factor count, orientation, and the apparent
stability of item assignments. A coefficient reported to three decimals
is not thereby precise to three decimals. The comparisons below examine
specified analytic alternatives on one sample; they do not provide a
sampling confidence interval. Bootstrap analysis would additionally
require re-estimation, alignment, and a decision about whether the factor
count is held fixed. Unaligned bootstrap columns cannot sensibly be
averaged.

Factor scores are another distinct object. EFA estimates a population
covariance representation; individual latent values are not observed
merely because loadings have been estimated. Alternative scoring rules
can produce different scores under the same factor model. Chapter 4
develops this issue after measurement-model evaluation.

# Check reconstruction and analytic sensitivity

## Examine local residuals, not only an average

Define the correlation residual matrix

$$
\mathbf E=\mathbf R-\widehat{\mathbf R},
\qquad
\mathrm{RMS}_{\mathrm{off}}
=\sqrt{\frac{2}{p(p-1)}\sum_{j<k}e_{jk}^2}.
$$

The off-diagonal average deliberately excludes variances. It is not
automatically identical to every software statistic called RMSR or SRMR.
Its definition here is transparent and reproducible. The largest
absolute residual supplies different information: one problematic pair
can be concealed by averaging many small residuals.

```{r}
#| label: fig-efa-residual
#| fig-width: 6.3
#| fig-height: 5.5
#| fig-cap: "Observed minus reproduced correlations for the three-factor ML solution. The scale is centered on zero."
#| fig-alt: "A labeled residual heatmap shows small positive and negative discrepancies across the twelve indicators, with values printed in every cell."
pt2_heatmap(z$residual,"Three-factor correlation residuals",limit=.05,digits=2,cex=.64)
```

```{r}
#| label: tbl-efa-residual-pairs
#| tbl-cap: "Eight largest absolute off-diagonal residuals, with observed and reproduced values."
pt2_table(head(z$residual_pairs,8),row.names=FALSE)
```

Ask whether residuals concentrate within a domain, between two domains,
or around a particular item. Those patterns suggest different questions
about omitted common dimensions, secondary loadings, and local residual
relations. A residual is evidence of unreproduced covariance, not a label
for its cause. An unrestricted exploratory loading pattern can also
absorb a local relation into its common part. Small EFA residuals therefore
do not certify that every later simple-structure CFA restriction is safe.

## KMO and Bartlett answer preliminary questions

Let $a_{jk}$ be the partial correlation between indicators $j$ and $k$
controlling the remaining indicators. A common overall KMO statistic is

$$
\mathrm{KMO}=
\frac{\sum_{j<k}r_{jk}^2}
{\sum_{j<k}r_{jk}^2+\sum_{j<k}a_{jk}^2}.
$$

The sample value is `r pt2_num(z$kmo)`. It is larger when the squared
partial correlations are small relative to the squared zero-order
correlations. That is a useful compact description, but it neither
selects the number of factors nor establishes the meaning of any factor.
It is also not a universal minimum-sample-size calculation.

Bartlett's sphericity test addresses $H_0:\mathbf R_{\mathrm{population}}
=\mathbf I$. A familiar normal-theory approximation is

$$
T_B=-\left[n-1-\frac{2p+5}{6}\right]\log|\mathbf R|,
\qquad df_B=\frac{p(p-1)}2.
$$

Here $T_B=$ `r pt2_num(z$bartlett,1)` on 66 df. Rejecting an identity
correlation matrix says that the indicators are not mutually uncorrelated
in the population under the test assumptions. It does not show that a
particular low-dimensional factor model fits. A matrix with complicated
local dependencies can readily reject sphericity.

Use these summaries to characterize the input, not as entrance tickets
that replace inspecting the correlation matrix, positive definiteness,
indicator content, and the fitted model.

## Change one analytic choice at a time

Two bounded sensitivity analyses use the same observations, variables,
and Pearson matrix. First, keep ML extraction and use oblique geomin
rotation with $\delta=.01$, no normalization, and 20 starts. Second,
keep three factors and oblimin but replace ML extraction by minres.
The first changes the orientation criterion; the second changes how
the covariance discrepancy is weighted.

Signs and factor order are aligned before comparing coefficients.
For loading columns $\mathbf a$ and $\mathbf b$, Tucker's congruence
coefficient is

$$
c(\mathbf a,\mathbf b)=
\frac{\mathbf a^{\mathsf T}\mathbf b}
{\sqrt{\mathbf a^{\mathsf T}\mathbf a}
 \sqrt{\mathbf b^{\mathsf T}\mathbf b}}.
$$

It compares the directions of two coefficient vectors, not the
correlation between participant scores. Near-one congruence can coexist
with an important difference for one item, so also inspect the largest
coefficient difference and the relevant rows.

```{r}
#| label: tbl-efa-sensitivity
#| tbl-cap: "Aligned sensitivity comparisons against primary ML oblimin. These are same-sample analytic checks, not independent replications."
pt2_table(data.frame(Analysis=rownames(z$sensitivity),z$sensitivity,
  row.names=NULL,check.names=FALSE))
```

The largest pattern change under geomin is
`r pt2_num(z$sensitivity[2,"Max_pattern_difference"])`; under minres
it is `r pt2_num(z$sensitivity[3,"Max_pattern_difference"])`.
The principal item groupings and the two content-relevant secondary
relations persist. This supports carrying a specific candidate forward
without claiming that all reasonable analyses have been exhausted.
Changing the item set, sample, distributional assumptions, or factor
count would be a different sensitivity question.

The defensible conclusion is deliberately narrower than “the scale is
validated.” Three correlated dimensions summarize Sample A better than
two, the main orientation is interpretable and stable across these
limited checks, and the larger fitted solutions encounter boundaries.
These observations justify a candidate to test elsewhere.

# Summary and reading bridge {#sec-summary-bridge}

EFA is a sequence of linked decisions. The input matrix establishes the
observed metric. Extraction states which covariance structure is being
approximated and how discrepancies are weighted. Retention selects a
plausible dimension using matched reference comparisons, content, and
neighboring solutions. Rotation chooses interpretable axes
without changing the fitted covariance. Interpretation then requires
pattern, structure, factor correlations, and the quadratic-form
communality to be read together.

The Sample A record supports a three-factor candidate, not a final
measurement verdict. Both declared parallel analyses retain three
dimensions; two factors leave appreciably larger residuals; four and
five reach uniqueness bounds. The primary oblique pattern and limited
same-sample sensitivity checks support three content-defined domains -
Planning, Strategy and Belonging - together with two secondary loadings that
are retained on content grounds rather than by a numerical cutoff: P4 on
Strategy and S4 on Belonging. All twelve indicators are kept, in the codebook
order, and every other cross-loading and every residual covariance would be
fixed at zero in a restricted version of this model. That specification, and
the stricter simple-structure alternative that fixes both secondary loadings
to zero, are what a confirmatory analysis on unused data would have to test.

Naming those alternatives now is part of the transition, because a comparison
decided in advance is interpretable and one invented after seeing the
confirmation data is not. Five models are therefore fixed here, before any
confirmation outcomes or statistics are inspected or used:

| Label | Specification | Free covariance parameters | Candidate df |
|:--|:--|--:|--:|
| **M1-cross** | The primary candidate: three correlated factors, P4 also on Strategy, S4 also on Belonging | 29 | 49 |
| **M0-simple** | Both secondary loadings fixed at zero | 27 | 51 |
| **M2-orthogonal** | All three factor correlations fixed at zero | 26 | 52 |
| **M3-equal** | The raw P1 and P2 Planning loadings constrained equal | 28 | 50 |
| **M4-one** | A single common factor for all twelve indicators | 24 | 54 |

: *The five models fixed before any confirmation outcome is inspected or used.* {#tbl-efa-rivals}

All five use the same twelve indicators in codebook order, fixed-factor
scaling, and no residual covariances. M0, M2 and M3 are restrictions of
M1-cross; M4-one is a broader dimensional rival whose relation to a correlated
three-factor model is not an ordinary nested comparison, for reasons taken up
where models are compared.

A compact report could read:

> We analyzed Pearson correlations for 12 complete continuous indicators
> in a simulated development sample of 360 cases. Matched 95th-percentile
> permutation references for full and SMC-reduced eigenvalues both
> retained three dimensions. Neighboring ML models and residual inspection
> supported examining a three-factor representation; four- and five-factor
> fits reached uniqueness bounds. Direct oblimin with stated settings
> produced interpretable correlated domains. Geomin and minres checks
> gave closely aligned patterns. A 12-indicator CFA candidate with two
> content-motivated secondary loadings was fixed for independent-sample
> estimation; this development analysis is not itself confirmation.

For mathematical background, connect the common-factor representation
and rotational indeterminacy to @mulaik2010 and @bartholomewKnott1999.
Read @horn1965 for the parallel-analysis idea and @bujaEyuboglu1992 for
the permutation perspective and qualifications. The historical oblique
rotation literature includes @jennrichSampson1966. Software documentation
is indispensable for implementation details, especially distinctions
between simulation, resampling, and permutation.

Chapter 3 now changes the question: **given the fixed zero, free, and
equality restrictions, can a CFA be specified, identified, and estimated
coherently?** An exploratory orientation becomes a confirmatory model
only when its restrictions and analytic history are made explicit.

# Exercises {.unnumbered}

The following exercises are answer-free. Use the printed tables where
possible; code is optional unless a prompt explicitly asks for it.

1. **Identify the target.** An analyst extracts principal
   components from $\mathbf R$ and calls the discarded component variance
   “measurement error.” Explain the missing common-factor assumptions.
   For $\mathbf R=\begin{bmatrix}1&.6\\.6&1\end{bmatrix}$, obtain its
   eigenvalues and the variance represented by the first component.
2. **Update a diagonal.** For the three-variable matrix
   $\mathbf R=\begin{bmatrix}1&.5&.4\\.5&1&.3\\.4&.3&1\end{bmatrix}$,
   calculate the SMCs using $1-1/(\mathbf R^{-1})_{jj}$. Form the reduced
   matrix and describe one principal-axis update. Explain why a negative
   reduced-matrix eigenvalue is not automatically a negative variance.
3. **Examine the count.** Derive the effective parameter
   counts and candidate df for $p=8$, $m=1,2,3$. Explain the rotational
   subtraction and why a nonnegative count does not prove identification.
4. **Match the null.** At ranks 1--4, observed roots are
   $3.8,1.5,.9,.2$, and matching permutation 95th percentiles are
   $.5,.3,.2,.1$. Determine the leading crossing count. Explain why this
   table cannot be interpreted without knowing which matrix generated
   the roots, and why a full-matrix reference cannot replace a reduced one.
5. **Read an oblique row.** For pattern row $(.6,.3)$ and
   factor correlation .4, compute the two structure correlations,
   communality, and uniqueness of a standardized indicator. Explain why
   the sum of squared pattern coefficients gives a different result.
6. **Compare dimensions.** Using the one-to-five-factor
   table and neighboring-pattern figure, recommend a dimension for further
   study. Discuss the uniqueness boundary and explain why neither the
   smallest discrepancy nor the largest conventional $p$ value is a
   sufficient decision rule.
7. **Evaluate an item.** Contrast P4 and B4 using wording,
   pattern, structure, and communalities. Propose what additional evidence
   would be needed before deleting either. Do not use a loading cutoff
   as the complete rationale.
8. **State the transition.** Write the free/fixed relation
    ledger for two restricted models: the three-factor candidate that keeps the
    P4 and S4 secondary loadings free, and the stricter version that fixes both
    at zero. Give the free-parameter count and candidate df for each against
    the 78 distinct moments of twelve indicators. Then identify which evidence
    could and could not be used to choose between them, and explain what
    changes if a model is revised after inspecting a second sample.

```{=latex}
\Needspace{10\baselineskip}
```

# Appendix A: SLE data dictionary and sample manifest {#sec-sle-data .unnumbered}

```{=latex}
\Needspace{6\baselineskip}
```

The downloadable file [sle_continuous.csv](sle_continuous.csv) contains 1,080
simulated records in 14 columns: two identifiers and twelve continuous
indicators. The three samples are combined for distribution only, and no
observations have been pooled, rounded, standardized, reassigned or
regenerated for analysis.

## Fields and common coding {.unnumbered}

```{r}
#| label: tbl-sle-fields
#| tbl-cap: "CSV fields and coding rules. Neither sample nor id is a factor indicator."
pt2_sle_appendix_table(data.frame(
  Field=c("sample","id","P1-P4","S1-S4","B1-B4","All twelve indicators",
    "Direction","Missing values","Response format"),
  Meaning=c("A = development; B = confirmation; C = revalidation.",
    "Unique synthetic record ID; the first letter agrees with sample.",
    "Proposed Planning domain.","Proposed Strategy domain.",
    "Proposed Belonging domain.","Continuous, unbounded process-index points.",
    "Higher values mean more of the stated attribute; no reverse coding.",
    "NA is the missing-value code; the supplied records have no missing observations.",
    "Hypothetical continuous indices, not single ordered-category responses, percentages, or counts.")
),c(.22,.69))
```

## Item codebook {.unnumbered}

The wording gives illustrative content, not evidence that a questionnaire
was administered or validated. Proposed domains precede EFA and do not
guarantee the recovered factor structure.

```{r}
#| label: tbl-sle-codebook
#| tbl-cap: "Complete SLE item definitions and illustrative wording; the common coding rules above apply to every row."
book <- z$input$codebook
pt2_sle_appendix_table(data.frame(Item=book$item,Domain=book$domain,
  Definition=book$indicator_definition,Wording=book$wording),
  c(.06,.13,.24,.43))
```

The Planning and Strategy indicators are hypothetical continuous process
indices; Belonging indicators are hypothetical continuous experience
indices. The following qualifications make overlapping content explicit.
They do not disclose an estimated loading or authorize a model change.

```{r}
#| label: tbl-sle-process
#| tbl-cap: "Indicator-specific response-process qualifications."
pt2_sle_appendix_table(data.frame(
  Item=c("P4","S4","B1 and B2","B4"),
  Qualification=c("Combines planning and method adjustment.",
    "Combines learning strategy and social participation.",
    "Both refer to a shared peer context.",
    "Broad institutional content rather than one setting.")
),c(.16,.75))
```

```{=latex}
\Needspace{14\baselineskip}
```

## Population parameters {.unnumbered}

The three samples are drawn from the population in @tbl-sle-population, using
the seeds in @tbl-sle-manifest below. Standardized loadings and factor
correlations are given; residual variances are whatever makes each
standardized indicator variance one, apart from the single residual covariance
shown. Indicator means and scale factors convert the standardized system to
the recorded units.

This table documents the *generating population* and the role of each sample.
It is not a complete recipe for reproducing the released numbers, which would
additionally require the random-number convention, the order in which the
factor and residual draws are made, and the factorization used to impose the
covariance structure. The exact observations analysed in Chapters 2 to 4 are
the ones distributed in `sle_continuous.csv`; the chapter source regenerates
them during rendering and checks them against that file.

```{r}
#| label: tbl-sle-population
#| tbl-cap: "The generating population. Loadings are standardized; a blank means the loading is fixed at zero."
pop <- sle_population()
Lz <- pop$Lambda_z
blank <- function(v) ifelse(abs(v) < 1e-12, "", pt2_num(v, 2))
pt2_sle_appendix_table(data.frame(
  Item = rownames(Lz),
  Planning = blank(Lz[,1]), Strategy = blank(Lz[,2]),
  Belonging = blank(Lz[,3]),
  Intercept = pt2_num(pop$nu, 0), Scale = pt2_num(pop$scale, 0),
  check.names = FALSE), c(.12,.16,.16,.18,.19,.19))
```

Factor correlations are
$\phi_{12}=`r pt2_num(pop$Phi[1,2],2)`$,
$\phi_{13}=`r pt2_num(pop$Phi[1,3],2)`$ and
$\phi_{23}=`r pt2_num(pop$Phi[2,3],2)`$, with all three factor variances
fixed at one. One residual covariance is nonzero,
$\theta_{\mathrm{B1,B2}}=`r pt2_num(pop$Theta_z[9,10],2)`$ on the
standardized scale, representing shared peer context that the three broad
factors do not capture. Every other residual covariance is zero.

```{=latex}
\Needspace{12\baselineskip}
```

## Sample manifest {.unnumbered}

```{r}
#| label: tbl-sle-manifest
#| tbl-cap: "Three independent samples in one file. The first-use rules concern analytical evidence, not physical file accessibility."
manifest <- z$input$manifest_all
pt2_sle_appendix_table(data.frame(
  Sample=paste(manifest$sample,manifest$role,sep=": "),
  N=as.character(manifest$n),
  IDs=c("A001-A360","B001-B360","C001-C360"),
  Seed=as.character(manifest$seed),
  `Permitted first use`=c("Chapter 2: EFA development.",
    "Chapter 3: estimate the candidate frozen after A; Chapter 4: evaluate it.",
    "Chapter 4: retest the revision frozen after B."),
  check.names=FALSE),c(.18,.05,.17,.15,.29))
```

Each row belongs to one sample only. The 360-case sample sizes are
pedagogical choices, not general adequacy recommendations. Importing the
combined file does not authorize earlier use of B or C. Select the eligible
sample before descriptive summaries, missing-data decisions, rescaling,
correlations, extraction, estimation, or diagnostics. Keep `sample` and
`id` out of the indicator matrix.

The records are simulated from the declared population with the seeds in
@tbl-sle-manifest. The source generates Sample A internally and, when the
released CSV is present, verifies agreement across all three samples, their
case identifiers and all twelve indicators. Readers reproduce the released
analysis from `sle_continuous.csv`. The generation specification was declared
before any sample was inspected, and no seed was chosen for its fit results.

# Appendix B: Spectral and rotation details {.unnumbered}

## Why a truncated eigendecomposition is a component solution {.unnumbered}

For a symmetric positive-semidefinite matrix,
$\mathbf R=\mathbf V\mathbf D\mathbf V^{\mathsf T}$ with orthonormal
eigenvectors. Retaining the first $m$ eigenpairs gives
$\mathbf R_m=\mathbf V_m\mathbf D_m\mathbf V_m^{\mathsf T}$.
Its trace is the sum of the retained eigenvalues. Among unrestricted
rank-$m$ approximations, this truncation minimizes squared Frobenius
error. The omitted diagonal is not required to be a diagonal residual
covariance from a latent-variable measurement model. That is the key
reason a convenient low-rank approximation and common-factor estimation
answer different questions.

Replacing the diagonal by communality estimates changes the eigenproblem.
The reduced matrix need not be positive semidefinite: its diagonal is
no longer the full variance of an observed random vector. In a
principal-axis iteration, the retained positive roots define a current
common part whose diagonal supplies the next communality estimates.
Convergence of that iteration and its sensitivity to starts are numerical
questions separate from whether the resulting factors make sense.

## Other retention summaries are not interchangeable votes {.unnumbered}

The minimum average partial (MAP) idea progressively removes component
directions and examines the remaining partial-correlation structure.
It is not the same reference comparison as parallel analysis. Very
simple structure (VSS) compares approximations obtained by simplifying
loading patterns to a small number of entries per row; it consequently
depends on the kind of simplicity one is prepared to impose. Sequential
model tests target exact covariance restrictions under their reference
conditions. Information criteria trade likelihood against parameter
count but also require a carefully specified likelihood and admissible
candidate family.

These methods can disagree because they target different features. A
useful extension would state each target, its tuning choices, and the
implications of disagreement, rather than assemble a majority-vote
factor count. Newer eigenvalue corrections likewise require their own
derivation and calibration; they should not be introduced merely by
renaming the eigenvalue-greater-than-one rule.

## Two explicit oblique criteria {.unnumbered}

Ignoring a positive multiplicative constant, the quartimin objective is

$$
Q_{\mathrm{quartimin}}(\mathbf P)
=\sum_{j=1}^p\sum_{a<b}p_{ja}^2p_{jb}^2.
$$

Large coefficients in two columns of the same row increase this
objective. A geomin objective takes the form

$$
Q_{\mathrm{geomin}}(\mathbf P)
=\sum_{j=1}^p
\left[\prod_{a=1}^m(p_{ja}^2+\delta)\right]^{1/m},
\qquad\delta>0.
$$

The small positive $\delta$ regularizes behavior near zero. These
criteria may have several local optima. Convergence checks, multiple
starts, and sensitivity comparisons are therefore part of a complete
rotation report. Neither objective supplies a probability that a factor
name is correct, and neither changes the fitted common covariance when
implemented as an equivalent rotation.

## Factor-score indeterminacy in one equation {.unnumbered}

If factors and indicators are jointly normal, a model-based conditional
mean score in the standardized metric is

$$
E(\boldsymbol\eta\mid\mathbf z)
=\boldsymbol\Phi\mathbf P^{\mathsf T}
\boldsymbol\Sigma_z^{-1}\mathbf z.
$$

The conditional variance generally remains positive. Thus knowing the
model parameters does not identify each person's latent value without
error. Other scoring criteria lead to other point estimates. This
appendix supplies the conceptual warning; Chapter 4 develops scoring
and its uncertainty after the CFA model has been evaluated.

# References {.unnumbered}

::: {#refs}
:::
