---
title: "Part Two: Factor Analysis and Structural Equation Modeling"
author: "Jinsong Chen"

lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    output-file: part_two_introduction.pdf
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-cap: true
    tbl-cap: true
    fig-align: center
    include-in-header:
      text: |
        \setlength{\parindent}{0pt}
        \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
        \ifdefined\pdftrailerid\pdftrailerid{}\fi
        \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
        \ifdefined\pdfvariable
          \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
        \fi
        \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
        \addtokomafont{sectionentry}{\small}
        \let\chapteroldtableofcontents\tableofcontents
        \renewcommand{\tableofcontents}{\begingroup\small
          \setlength{\parskip}{0pt}\chapteroldtableofcontents\endgroup}
        \usepackage{caption}
        \captionsetup[table]{
          position=above,
          format=plain,
          justification=raggedright,
          singlelinecheck=false,
          labelsep=newline,
          labelfont=bf,
          textfont=it
        }
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{float}
        \usepackage{needspace}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: intro-setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center",
  dev = if (knitr::is_latex_output()) "cairo_pdf" else "png"
)

draw_part_two_route <- function() {
  old <- par(mar = rep(.1, 4), xaxs = "i", yaxs = "i")
  on.exit(par(old), add = TRUE)
  plot.new()
  plot.window(xlim = c(0, 4), ylim = c(0, 1))

  labels <- c(
    "Chapter 1\nGeneral and conceptual\nintroduction",
    "Chapter 2\nExploratory approach\nconcepts and practice",
    "Chapter 3\nFoundational CFA\nspecification and\nestimation",
    "Chapter 4\nAdvanced CFA\nevaluation, comparison,\nand extensions"
  )
  fills <- c("#F3F5F6", "#E8EEF1", "#DCE7EB", "#D0E0E5")

  for (i in 0:3) {
    x0 <- i + .04
    x1 <- i + .86
    rect(x0, .18, x1, .82, col = fills[i + 1], border = "#405768",
         lwd = 1.2)
    text((x0 + x1) / 2, .50, labels[i + 1], cex = .62,
         col = "#243746", font = 2)
    if (i < 3) {
      arrows(x1 + .015, .50, i + 1.025, .50, length = .07,
             lwd = 1.2, col = "#405768")
    }
  }
}
```

## What Part Two Is

Part Two of *Quantitative Methods for Social and Behavioral Research with R*
introduces latent-variable modeling through factor analysis and then prepares
the transition to structural equation modeling. It is written for
postgraduate students and applied researchers who need to understand what a
latent-variable model asserts, how its parameters are estimated, and how far
its results can support a substantive interpretation.

The first four chapters form one measurement-model sequence. Chapter 1
introduces the common-factor model and its representations. Chapter 2 develops
an exploratory factor model. Chapter 3 turns the resulting proposal into a
restricted confirmatory model and estimates it. Chapter 4 evaluates that fit,
compares planned rivals, studies a frozen revision, fits higher-order and
bifactor representations, and closes with a retrospective known-population
audit. The sequence is cumulative: later chapters assume the model, notation,
and analytic history established earlier.

This is not a catalogue of software commands or fit-index cutoffs. Equations,
matrices, diagrams, prose, and code are treated as different representations
of the same model. Software makes the calculations reproducible; it does not
choose the research question, justify a construct label, or turn an
exploratory decision into an independently confirmed result.

**What is assumed.** Readers should understand variance, covariance,
correlation, sampling uncertainty, and elementary regression notation at the
level developed in Part One. Basic matrix multiplication is helpful, but the
dimensions and operations needed for each model are explained where they are
used. Calculus and proof-based mathematical statistics are not prerequisites.

**How to use this map.** This introduction explains why Chapters 1--4 appear
in their present order, how their evidence roles differ, and what they hand to
the later structural-modeling sequence. It is an orientation document, not a
substitute for any chapter.

## The Measurement Problem at a Glance

Several observed indicators may vary together even though the attribute used
to explain their association is not recorded directly. A common-factor model
represents that proposal as

$$
\mathbf y_i
=\boldsymbol\nu+\boldsymbol\Lambda\boldsymbol\eta_i
+\boldsymbol\epsilon_i,
\qquad
\begin{aligned}
E(\mathbf y_i)
  &=\boldsymbol\nu+\boldsymbol\Lambda\boldsymbol\alpha,\\
\boldsymbol\Sigma(\boldsymbol\vartheta)
  &=\boldsymbol\Lambda\boldsymbol\Phi
    \boldsymbol\Lambda^{\mathsf T}+\boldsymbol\Theta.
\end{aligned}
$$

Here $\mathbf y_i$ contains observed indicators,
$\boldsymbol\eta_i$ contains unobserved factors,
$\boldsymbol\Lambda$ contains loadings, and $\boldsymbol\epsilon_i$ contains
what the declared factors leave unexplained. The vectors $\boldsymbol\nu$ and
$\boldsymbol\alpha$ are indicator and latent intercepts; with no structural
regressions in this measurement-only model,
$\boldsymbol\alpha=E(\boldsymbol\eta_i)$. Under
$E(\boldsymbol\epsilon_i)=\mathbf 0$ and
$\operatorname{Cov}(\boldsymbol\eta_i,\boldsymbol\epsilon_i)=\mathbf 0$,
$\boldsymbol\Phi=\operatorname{Var}(\boldsymbol\eta_i)$ and
$\boldsymbol\Theta=\operatorname{Var}(\boldsymbol\epsilon_i)$ produce the
displayed moments; $\boldsymbol\vartheta$ collects the model's free
parameters. The covariance equation divides observed variation into a common
part and a residual part. It also gives the model observable consequences:
proposed loadings, factor relations, and residual restrictions imply a
particular covariance matrix.

The equation does not make a factor identical to the construct used to name
it. A factor is an unobserved random variable within a statistical model. A
construct is a substantive concept whose interpretation also depends on the
indicator domain, response process, population, design, and intended use.
Likewise, a factor-to-indicator arrow states the direction of representation;
by itself it does not establish that the factor causally produces a response.
Good covariance fit can support a statistical representation without proving
that a latent entity exists exactly as named.

Part One's question--answer discipline remains in the background: a method
should answer a declared question, and the conclusion should not outrun its
evidence and justification. Part Two does not repeat the complete
argument-based framework. Chapter 1 contains one optional late application to
the latent-variable context; Chapters 2--4 concentrate on the statistical
work needed to develop, estimate, and evaluate measurement models.

## How Chapters 1--4 Progress

The four chapters make four connected moves:

1. **Represent:** state what factors and residuals mean and derive the moments
   implied by a common-factor model.
2. **Explore and develop:** use a development record to investigate
   dimensionality, orientation, interpretation, and sensitivity, then freeze
   a candidate.
3. **Specify and estimate:** express that candidate as explicit restrictions,
   establish recoverability, fit it to unused data, and audit the solution.
4. **Evaluate, compare, and extend:** assess global and local discrepancy,
   compare eligible rivals, study revisions honestly, and examine advanced
   measurement representations.

@fig-part-two-route makes the architecture explicit. Each stage changes the
question and evidence role while retaining the same common-factor foundation.

```{r}
#| label: fig-part-two-route
#| fig-width: 7
#| fig-height: 1.75
#| out-width: "100%"
#| fig-cap: "The four-chapter architecture moves from representation to exploratory development, foundational CFA, and advanced CFA."
#| fig-alt: "Four boxes connected from left to right: Chapter 1 general and conceptual introduction; Chapter 2 exploratory concepts and practice; Chapter 3 foundational CFA specification and estimation; Chapter 4 advanced CFA evaluation, comparison, and extension."

draw_part_two_route()
```

@tbl-part-two-sequence summarizes the responsibility and handoff of each
chapter.

| Chapter | Main responsibility | What it hands forward |
|:--|:--|:--|
| **1. From Latent-Variable Modeling to Factor Analysis** | Distinguish observed variables, factors, constructs, composites, and scores; translate the common-factor model among scalar equations, matrices, diagrams, and prose; derive implied means and covariances; explain residuals and latent-coordinate conventions | A technically explicit common-factor representation, known-value teaching records, and the questions that exploratory analysis must answer |
| **2. Exploratory Factor Analysis and Model Development** | Audit indicators and their correlations; choose an extraction target; investigate factor count; rotate and interpret neighboring solutions; examine residuals, sensitivity, and stability | A documented three-factor candidate, named alternatives, and unresolved questions frozen before unused confirmation data are inspected |
| **3. Specifying and Estimating Confirmatory Factor Models** | Translate the candidate into fixed, free, and constrained parameters; set factor metrics; count moments and coordinates; examine identification with small-model diagrams; translate a published four-factor model into syntax and a parameter ledger; estimate by continuous ML; read convergence, admissibility, coefficients, and uncertainty | A reproducible fitted object and parameter ledger whose overall adequacy has deliberately not yet been decided |
| **4. Evaluating and Comparing Confirmatory Factor Models** | Read global fit as a profile; locate local discrepancy; compare compatible planned rivals; interpret modification diagnostics; freeze and revalidate a revision; fit and compare higher-order and bifactor representations; conduct a retrospective known-population audit; examine reliability and score boundaries | A calibrated measurement-model conclusion and a bridge from relations among factors to later structural equation models |

: *The four-chapter measurement-model sequence.* {#tbl-part-two-sequence tbl-colwidths="[27,38,35]"}

The stopping points are deliberate. Chapter 2 does not relabel a developed
pattern as confirmation. Chapter 3 does not treat convergence as adequate fit.
Chapter 4 does not treat favorable fit as construct validation. Each chapter
completes one responsibility before transferring a clearly defined record to
the next.

The diagrams also form a deliberate progression. Chapter 3 reads the same
visual grammar across a two-factor algebra model, the three-factor SLE model,
and a published four-factor transfer, then uses small paired diagrams to make
identification and equivalence visible. Chapter 4 carries that grammar into
the three-factor higher-order and bifactor alternatives. The diagrams, syntax,
and parameter counts are generated from matching structural ledgers rather
than treated as independent illustrations.

## One Model Family, Several Responsibilities

The common-factor covariance equation remains the mathematical backbone of
all four chapters. What changes is the analytical responsibility assigned to
it. Chapter 1 works forward from known parameters to implied moments. Chapter
2 works from sample associations toward a provisional low-dimensional
representation. Chapter 3 asks whether a prespecified parameterization can be
recovered and estimated. Chapter 4 asks how much discrepancy remains, where it
is located, and which comparisons are warranted.

Four mathematical threads connect the sequence:

- **Common and residual variation.** Scalar covariance expansions become the
  matrix decomposition
  $\boldsymbol\Lambda\boldsymbol\Phi\boldsymbol\Lambda^{\mathsf T}
  +\boldsymbol\Theta$, then become fitted residuals and local diagnostics.
- **Coordinates and restrictions.** Scale, sign, and rotational freedom lead
  to EFA rotation, CFA scaling constraints, identification checks, and the
  distinction between equivalent and testably different models.
- **Information and estimation.** Sample means and covariances become targets
  for a restricted model; the number of distinct moments, free coordinates,
  and candidate degrees of freedom is counted consistently before numerical
  output is interpreted.
- **Evaluation and use.** Parameter uncertainty, global fit, local residuals,
  planned comparisons, reliability, and scores answer different questions.
  None is a universal pass--fail certificate for the model or construct.

The statistical level is intentionally higher than an introductory survey.
Core derivations remain in the chapters and are explained in words rather
than removed. The lecture decks stage the same ideas across shorter teaching
steps; exercises require calculations and interpretations; and the R labs
make selected analyses executable without turning the chapter into a command
reference.

Chapter 1's main route requires no student software. Chapter 2 introduces the
`psych` and `GPArotation` workflow for EFA. Chapters 3--4 use `lavaan` for CFA
estimation and evaluation. In every case the printed chapter supplies enough
information to understand the model and the conclusion even when the reader
does not run the code.

## Evidence Roles Across the Sequence

Repeated use of one topic does not mean that every record has the same
inferential role. Part Two makes the analysis history visible so that
development, first evaluation, and revalidation are not collapsed into one
favorable story. @tbl-part-two-evidence gives the principal roles.

| Record | First permitted use | What it cannot establish by itself |
|:--|:--|:--|
| **Known-value six-indicator records** | Chapter 1 reconstructs covariance from declared parameters | Sampling behavior, empirical factor count, or the existence of named constructs |
| **SLE Sample A, 360 cases** | Chapter 2 develops and freezes the exploratory candidate | Independent confirmation of choices made while examining A |
| **SLE Sample B, 360 cases** | Chapter 3 estimates the frozen candidate; Chapter 4 evaluates it and planned rivals | Independent support for a new relation first suggested by B |
| **SLE Sample C, 360 cases** | Chapter 4 evaluates one revision frozen after B | General validity, invariance, or permission to search repeatedly in C |
| **Known SLE generating moments** | Chapter 4 reveals them retrospectively, after the B/C analysis history is complete | Contemporaneous justification for earlier choices, or finite-sample bias inferred from one realization |
| **Separate donor microcases** | Compact demonstrations of specified calculations and model comparisons | The development or confirmation status of the SLE sequence |

: *Records keep distinct analytical roles even when they illustrate the same model family.* {#tbl-part-two-evidence tbl-colwidths="[27,36,37]"}

The Student Learning Experiences (SLE) records are synthetic continuous data,
not observations from actual students. One file, `sle_continuous.csv`, stores
1,080 rows across three independent samples of 360 cases and twelve indicators:
four proposed Planning indicators, four Strategy indicators, and four
Belonging indicators. A `sample` column identifies A, B, or C, and the Chapter
2 appendix supplies the codebook and manifest. The samples are selected for
their declared roles, never pooled to manufacture a larger analysis. A record
that becomes exploratory through revision does not recover confirmatory
status merely because the next command calls the model CFA.

Two numerical roles are kept separate in Chapter 4. Its extension fits use
the factor correlations estimated from Sample B. Only afterward does the
retrospective audit reveal the generating correlation triplet
$(.42,.30,.35)$ and its exact higher-order coordinates
$(.60,.70,.50)$. Those known values sharpen the teaching comparison; they are
not Sample-B estimates and were not evidence used to develop or revise the
candidate.

This separation is also a teaching device. It lets readers see that a model
can be technically coherent, reproducible, and even apparently well fitting
while still owing a narrower conclusion than its name might suggest.

## Boundaries and What Follows

The initial Part Two sequence treats indicators as continuous and keeps one
coherent continuous-ML lineage. Ordered categorical indicators, thresholds,
polychoric correlations, and their estimator-specific testing conventions are
signposted but taught elsewhere. This boundary does not exclude categorical
grouping variables or coded predictors: later multigroup, invariance, and
structural models may use them while retaining continuous indicators.

The four-chapter architecture focuses on reflective common-factor measurement
models. It does not make a factor score an error-free observation, equate fit
with validity, or establish causality from a path diagram. Within that
architecture, Chapter 4 treats higher-order and bifactor models as rival
measurement representations, not automatic remedies for a weak core model.
Specialized topics such as item-response theory,
mixture models, multilevel factor analysis, and a full treatment of
categorical indicators belong to other courses or later series.

After Chapter 4, Part Two turns from relations between factors and indicators
to relations among variables in a structural system. Observed-variable path
analysis prepares directed systems; general structural equation modeling
combines structural relations with measurement models; measurement invariance
supports comparisons across groups or occasions; and latent growth models
represent both average trajectories and variation in change. Multilevel
growth belongs to Part Three.

The notation is chosen to survive that transition. In particular, a latent
intercept is not generally the same as the marginal latent mean once
structural regressions are introduced. The later chapters extend the model
rather than quietly changing what its symbols mean.

## Routes and Commitments

**A taught course** should follow Chapters 1--4 in order. The lecture for each
chapter covers most of its conceptual and mathematical route; the exercise
companion supplies worked reference answers; and the R lab offers in-class
practice. In Chapters 3--4 these companions carry the same diagram-reading,
extension, and audit distinctions as the public chapters while keeping R-lab
datasets in their separate practice role. Homework extends the same
responsibilities. Part Two quizzes are not yet part of the teaching package.

**A reader working alone** should begin with Chapter 1, then keep Chapters
2--4 in sequence so that the exploratory candidate, confirmatory restrictions,
fitted object, and evaluation history remain intelligible. The labs may be run
alongside the chapters but are not prerequisites for the mathematical route.

**A reader consulting one method** should read the chapter's opening question,
record and destination before jumping to an output table. A fit index, loading,
modification index, or reliability estimate has meaning only within the model,
metric, evidence role, and intended use that produced it.

Across these routes, six commitments remain constant:

1. **The model agrees across representations.** Prose, equations, matrices,
   diagrams, syntax, and output must describe the same restrictions.
2. **Exploration and confirmation are historical roles.** They depend on when
   a decision was made and which evidence informed it, not on a function name.
3. **A fitted solution is read in stages.** Identification, numerical
   convergence, admissibility, uncertainty, global fit, and local fit are
   connected but not interchangeable checks.
4. **Comparison requires eligibility.** Nested tests and information criteria
   apply only when their assumptions and data-likelihood bases make the
   comparison meaningful.
5. **Computation reproduces rather than justifies.** Code can verify arithmetic
   and preserve an analysis record; it cannot supply missing design,
   measurement, or substantive warrants.
6. **Claims remain bounded.** Covariance compatibility does not by itself prove
   dimensionality, construct reality, validity, invariance, score fitness, or
   causal direction.

Each numbered chapter closes with a **Summary and reading bridge**, followed
by unnumbered exercises. Appendices may add derivations or software details,
but the principal question and its calibrated answer remain understandable
from the numbered route.
