---
title: "Part Two: Factor Analysis and Structural Equation Modeling"
author: "Jinsong Chen"

lang: en
tbl-cap-location: top
fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false
  cache: false

format:
  pdf:
    output-file: part_two_introduction.pdf
    pdf-engine: lualatex
    pdf-standard: ua-2
    colorlinks: false
    latex-min-runs: 3
    number-sections: true
    toc: true
    toc-depth: 2
    fig-align: center
    include-in-header:
      text: |
        % QMSBR local compatibility backport for luatex.def 2025/09/29 v1.2f.
        % Adapted from latex3/graphics-def PR 48 (Joseph Wright), commit
        % edda2fc20708ead2b7c18ed4e4f66cae4d7b1ee7:
        % https://github.com/latex3/graphics-def/pull/48
        % https://github.com/latex3/graphics-def/issues/47
        % The upstream PR was unmerged when this local backport was made.
        % Original driver copyright: LaTeX3 Project and its listed contributors.
        % This adaptation is distributed under LPPL 1.3c or later:
        % https://www.latex-project.org/lppl.txt
        %
        % Preserve the imported PDF's transparency group. Put the driver's additional
        % group on a separate outer Form XObject, avoiding conflicting duplicate /Group
        % keys. Do not replace this with PDF post-processing or disabling transparency.
        % Exact-token patches deliberately fail on an unexpected/new driver: review
        % whether the upstream fix is then installed before removing this backport.
        \usepackage{etoolbox}
        \makeatletter
        \newcommand{\QMSBRgraphicsPatchError}[1]{%
          \PackageError{qmsbr-graphics-compat}{Graphics compatibility patch #1 failed}%
            {Expected LuaTeX graphics driver 2025/09/29 v1.2f. Review the upstream fix and local header before rendering.}}
        \patchcmd{\Gread@@pdftex}{\ifGread@transgroup 1\fi}{}{}{\QMSBRgraphicsPatchError{1}}
        \patchcmd{\Gread@@pdftex}{\ifGread@transgroup/Group<</S/Transparency/K false/I false>>\fi}{}{}{\QMSBRgraphicsPatchError{2}}
        \patchcmd{\Gread@@pdftex}
          {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}}
          {\setbox\@tempboxa=\hbox{\useimageresource\lastsavedimageresourceindex}%
           \ifGread@transgroup
             \saveboxresource attr{/Group<</S/Transparency/K false/I false>>}\@tempboxa
             \setbox\@tempboxa=\hbox{\useboxresource\lastsavedboxresourceindex}%
           \fi}
          {}{\QMSBRgraphicsPatchError{3}}
        \patchcmd{\Gread@@pdftex}
          {\useimageresource\the\lastsavedimageresourceindex}
          {\ifGread@transgroup\useboxresource\the\lastsavedboxresourceindex\else\useimageresource\the\lastsavedimageresourceindex\fi}
          {}{\QMSBRgraphicsPatchError{4}}
        \makeatother
        \setlength{\parindent}{0pt}
        \ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
        \ifdefined\pdftrailerid\pdftrailerid{}\fi
        \ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1\fi
        \ifdefined\pdfvariable
          \pdfvariable suppressoptionalinfo \numexpr1+2+4+8+16+32+64+128+256+512\relax
        \fi
        \AtBeginDocument{\hypersetup{pdfcreator={},pdfproducer={},pdfcreationdate={},pdfmoddate={}}}
        \addtokomafont{sectionentry}{\small}
        \let\chapteroldtableofcontents\tableofcontents
        \renewcommand{\tableofcontents}{\begingroup\small
          \setlength{\parskip}{0pt}\chapteroldtableofcontents\endgroup}
        \usepackage{caption}
        \captionsetup[table]{
          position=above,
          format=plain,
          justification=raggedright,
          singlelinecheck=false,
          labelsep=newline,
          labelfont=bf,
          textfont=it
        }
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{float}
        \usepackage{needspace}
        \AddToHook{cmd/section/before}{\Needspace{12\baselineskip}\phantomsection}
        \AddToHook{cmd/subsection/before}{\Needspace{12\baselineskip}\phantomsection}
  html:
    df-print: kable
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
  docx:
    df-print: kable
    toc: true
    toc-depth: 3
    number-sections: false

editor_options:
  chunk_output_type: console

editor:
  markdown:
    wrap: 72
---

```{r}
#| label: intro-setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center",
  dev = if (knitr::is_latex_output()) "cairo_pdf" else "png"
)

draw_part_two_route <- function() {
  old <- par(mar = rep(.1, 4), xaxs = "i", yaxs = "i")
  on.exit(par(old), add = TRUE)
  plot.new()
  plot.window(xlim = c(0, 4), ylim = c(0, 1))

  labels <- c(
    "Chapter 1\nGeneral and conceptual\nintroduction",
    "Chapter 2\nExploratory approach\nconcepts and practice",
    "Chapter 3\nFoundational CFA\nspecification and\nestimation",
    "Chapter 4\nAdvanced CFA\nevaluation, comparison,\nand extensions"
  )
  fills <- c("#F3F5F6", "#E8EEF1", "#DCE7EB", "#D0E0E5")

  for (i in 0:3) {
    x0 <- i + .04
    x1 <- i + .86
    rect(x0, .18, x1, .82, col = fills[i + 1], border = "#405768",
         lwd = 1.2)
    text((x0 + x1) / 2, .50, labels[i + 1], cex = .62,
         col = "#243746", font = 2)
    if (i < 3) {
      arrows(x1 + .015, .50, i + 1.025, .50, length = .07,
             lwd = 1.2, col = "#405768")
    }
  }
}
```

## What Part Two Is and Who It Is For

Part Two of *Quantitative Methods for Social and Behavioral Research*
is about latent-variable modeling. The methods of Part One worked with
variables that were recorded: a test score, a count, a group label. Here the
variable of primary interest is one that no column of the data set holds.
Twelve questionnaire items are recorded; the planning, the learning
strategies and the sense of belonging those items are meant to register are
not. Part Two asks what follows from taking that situation seriously, and
answers through factor analysis: how observed responses represent latent
variables, and how a measurement model is developed, specified, estimated,
evaluated and extended.

It is written for postgraduate students and applied researchers who have read
Part One or who already command the statistics it develops: variable form,
variance, covariance and correlation; estimates, sampling uncertainty,
intervals and tests, with the limits of a dichotomous decision; regression
equations, residuals and conditional comparisons; nested model comparison and
penalized in-sample criteria; held-out evidence; and reproducible work in R.
Part One's argument-based reading of a study is useful background rather than
a prerequisite.

Four chapters form the measurement sequence, and this introduction is about
them. Chapter 1, *From Latent-Variable Modeling to Factor Analysis*, is the
general and conceptual presentation. Chapter 2, *Exploratory Factor Analysis
and Model Development*, is the exploratory approach, conceptual and
practical. Chapter 3, *Specifying and Estimating Confirmatory Factor Models*,
is the foundational confirmatory approach, covering concepts, specification
and estimation. Chapter 4, *Evaluating and Comparing Confirmatory Factor
Models*, is the advanced confirmatory approach, covering evaluation,
comparison and the higher-order and bifactor extensions. The chapters after
them carry the structural half of the subject, from relations between factors
and indicators to relations among variables in a structural system; the last
section here says what they cover and what they cannot confer.

Part Two asks for more mathematics than Part One did. Readers manipulate
equations, count the moments a covariance matrix supplies against the
parameters a model spends, and work with small vectors and matrices: a matrix
product, a transpose, a symmetric matrix, a diagonal matrix. No previous
course in factor algebra or matrix calculus is assumed; each new operation
appears first in a scalar or two-variable case, substantive and graphical
meaning comes before notation, and every symbol, index and dimension is
defined before it is used. Calculus and proof-based mathematical statistics
are not prerequisites, and the longer derivations, identification theorems
and asymptotic arguments are placed so that a reader can pass over them
without losing a definition, an equation or an assumption the main route
needs.

The situation the four chapters share can be put in one sentence and one
equation. Several recorded responses vary together, and the attribute used to
explain their association is not recorded at all; the common-factor model
writes each recorded response as a common part, shared with the other
responses through a small number of unobserved variables, plus a residual
part that the declared common variables do not represent. For $p$ indicators
and $m$ factors, with $\mathbf y_i$ the $p\times1$ vector of responses
recorded for unit $i$,

$$
\mathbf y_i=\boldsymbol\nu+\boldsymbol\Lambda\boldsymbol\eta_i
  +\boldsymbol\epsilon_i,
\qquad
\begin{aligned}
\boldsymbol\mu(\boldsymbol\vartheta)
  &=\boldsymbol\nu+\boldsymbol\Lambda\boldsymbol\alpha,\\
\boldsymbol\Sigma(\boldsymbol\vartheta)
  &=\boldsymbol\Lambda\boldsymbol\Phi
    \boldsymbol\Lambda^{\mathsf T}+\boldsymbol\Theta,
\end{aligned}
$$

with $E(\boldsymbol\epsilon_i)=\mathbf 0$ and
$\operatorname{Cov}(\boldsymbol\eta_i,\boldsymbol\epsilon_i)=\mathbf 0$. Here
$\boldsymbol\nu$ holds the $p$ indicator intercepts, $\boldsymbol\eta_i$ the
$m$ factors for unit $i$, $\boldsymbol\Lambda$ the $p\times m$ loadings that
relate the two, and $\boldsymbol\epsilon_i$ the $p$ residuals;
$\boldsymbol\alpha$ holds the $m$ latent intercepts, $\boldsymbol\Phi$ is the
$m\times m$ factor covariance matrix, $\boldsymbol\Theta$ the $p\times p$
residual covariance matrix, and $\boldsymbol\vartheta$ the complete vector of
free parameters. What the model produces is the implied pair: a mean
structure $\boldsymbol\mu(\boldsymbol\vartheta)$ and a covariance structure
$\boldsymbol\Sigma(\boldsymbol\vartheta)$. An analysis working from
covariances alone sets the mean structure aside or profiles it out under
stated assumptions, which is a choice rather than an absence. One notation
serves all four chapters, and it is chosen to survive the move to structural
models, so that later chapters extend the model rather than quietly changing
what its symbols mean.

The covariance equation is what makes the proposal answerable. Declared
loadings, factor relations and residual restrictions imply one particular
covariance matrix, and that matrix can be held against the one a sample
supplies. The comparison has force because a covariance matrix for $p$
indicators holds only $p(p+1)/2$ distinct entries: a model that spends fewer
parameters than that makes claims which can fail. That counting is a
recurring operation in Part Two, and it is kept separate from the further
question of whether those moments determine the parameters at all.

None of this makes a factor identical to the construct that names it. A
**construct** is a substantive idea belonging to the research question, and
its meaning comes from theory, item content, response processes, the
population studied and the use intended for the results, none of which a
covariance matrix supplies. A **factor** is a random variable inside a
statistical model, and an observed or estimated **score** is neither; the
three are never treated as synonyms. The arrow a path diagram draws from a
factor to an indicator states the direction of representation the model
adopts, writing each response as a function of the factors plus a residual.
It is not a finding that the factor produces the response, and it is the
wrong direction for a construct whose indicators produce it.

The scope is narrower than the vocabulary might suggest. Part Two models
continuous indicators and continuous outcomes; ordered categorical
indicators, thresholds, polychoric correlations and ordered-response factor
models are signposted where a reader would meet them and are taught
elsewhere, and numeric category codes are not automatically continuous
measurements. The boundary concerns the indicators rather than every variable
in a study: categorical grouping variables and dummy-coded predictors stay in
scope, and the later multigroup and MIMIC work uses them alongside continuous
indicators. The family modelled here is the reflective common-factor model with
its higher-order and bifactor relatives, estimated by normal-theory maximum
likelihood across exploratory and confirmatory work; principal components
analysis answers a different question and is not a common-factor estimator.
What the family cannot do is as settled as what it can. No factor analysis
establishes that a latent entity exists, that a reflective direction is
causal, that a scale is valid, or that a structural claim is warranted;
higher-order and bifactor models are rival representations of an adequate
core model rather than repairs for a weak one; and reliability follows a
defined score, model, population and use, so it is not validity and does not
mend a weak measurement representation.

## How the Chapters Build on One Another

The four chapters make four connected moves on one covariance equation:
**represent** a common-factor model and derive the moments it implies;
**explore** a development record to settle dimensionality, orientation and
interpretation; **specify and estimate** the resulting proposal as explicit
restrictions on evidence that played no part in producing it; and
**evaluate, compare and extend** the fitted model. The equation is the same
throughout. What changes at each stage is the responsibility it carries and
the evidence it must answer to. @fig-part-two-route sets out that
architecture; no chapter redraws it.

```{r}
#| label: fig-part-two-route
#| fig-width: 7
#| fig-height: 1.75
#| out-width: "100%"
#| fig-cap: "The four-chapter architecture moves from representation to exploratory development, foundational CFA, and advanced CFA."
#| fig-alt: "Four boxes connected from left to right: Chapter 1 general and conceptual introduction; Chapter 2 exploratory approach, concepts and practice; Chapter 3 foundational CFA specification and estimation; Chapter 4 advanced CFA evaluation, comparison, and extensions."

draw_part_two_route()
```

@tbl-part-two-sequence states what each chapter is responsible for and what
it hands to the chapter after it.

| Chapter | Main responsibility | What it hands forward |
|:--|:--|:--|
| **1. From Latent-Variable Modeling to Factor Analysis** | Separate an observed variable, a latent random variable, a fixed parameter and a residual, and a factor from the construct that names it; write one two-factor, six-indicator system as scalar equations, matrices, a path diagram and a sentence and check the four against one another; say what a scaling, sign, location or coordinate convention supplies and what it cannot decide; derive $\boldsymbol\Sigma=\boldsymbol\Lambda\boldsymbol\Phi\boldsymbol\Lambda^{\mathsf T}+\boldsymbol\Theta$; and say what exact reproduction does not establish | The vocabulary the three later chapters use without redefinition, and the known-value six-indicator records on which the algebra is checked before any sample is read |
| **2. Exploratory Factor Analysis and Model Development** | Audit indicators, units, coding and missingness before reading a correlation matrix; state which matrix an extraction targets; compare observed eigenvalues with a matched permutation reference and read a structured retention profile across seven rules; count effective parameters and candidate degrees of freedom; transform loadings and factor covariance together and say what rotation preserves; and read pattern coefficients, structure correlations, factor correlations and communalities together, locating discrepancy in a residual matrix rather than an average | A frozen candidate specification with its named rivals, its retained cross-loadings, the retention evidence behind it and an explicit list of unresolved alternatives, settled before any confirmation evidence is inspected |
| **3. Specifying and Estimating Confirmatory Factor Models** | State one restricted model as words, equations, parameter matrices and a diagram and check that they agree; separate free, fixed, equality-constrained, bounded and defined entries; set a latent metric by marker or fixed-factor scaling and transform between them; trace a cross-loading and a residual covariance into the moments each changes; count moments, free parameters and degrees of freedom and say why the count is not identification; and check convergence, admissibility, gradients and local rank before reading any coefficient | An evaluation record built in memory and never written to disk: the exact specification and parameter ledger, the active items in order, the moment and degrees-of-freedom audit, the estimator and scaling settings, the convergence and admissibility record, the estimates with their uncertainty, and the rivals declared in advance |
| **4. Evaluating and Comparing Confirmatory Factor Models** | Reconstruct the discrepancy the estimator minimized under a named divisor convention; read the exact-fit test, RMSEA with its interval, CFI, TLI and SRMR as answers to different questions; separate the three residual scales; establish exact nesting, run a compatible difference test and say when AIC and BIC are eligible; specify, count, fit and compare the higher-order and bifactor extensions; read modification indices and expected parameter change as leads that change analytic status; and freeze one revision and evaluate it on untouched records | A calibrated measurement conclusion whose analytic history and unresolved strain stay intact, and the structural question the later chapters take up: which latent variables are endogenous, which directional restrictions enter, and how measurement uncertainty is carried forward |

: *The four-chapter measurement sequence.* {#tbl-part-two-sequence tbl-colwidths="[23,47,30]"}

Using one model family repeatedly does not give every record the same
inferential role. Part Two keeps the analysis history visible so that
development, first evaluation and independent revalidation are not collapsed
into one favorable account. Every decision is developmental, planned in
advance, a planned sensitivity, or post hoc, always in relation to a named
body of evidence, and that status follows the timing and provenance of the
decision rather than the name of the function that fitted it: a model
developed after inspecting a record is exploratory with respect to that
record whatever routine estimates it, and a record made exploratory by
revision does not recover confirmatory status because the next command is
named after confirmation. The samples are never pooled to manufacture a
larger analysis, and availability is not permission to use a record early.
@tbl-part-two-evidence gives the principal roles.

| Record | First permitted use | What it cannot establish by itself |
|:--|:--|:--|
| **Known-value six-indicator records** | Chapter 1 reconstructs an implied covariance matrix entrywise from parameters declared in advance | Sampling behavior, an empirical factor count, a unique rotation, a best model, or the existence of the constructs the columns are named after |
| **SLE Sample A, 360 cases, development** | Chapter 2 audits the indicators, develops the exploratory solution, and freezes the candidate with its named rivals | Independent confirmation of any choice made while Sample A was being inspected |
| **SLE Sample B, 360 cases, confirmation** | Chapter 3 estimates the frozen candidate; Chapter 4 evaluates it, diagnoses it, and compares the rivals declared in advance | Independent support for a relation first noticed in Sample B, which stays exploratory until fresh evidence tests it |
| **SLE Sample C, 360 cases, revalidation** | Chapter 4 evaluates the one revision frozen after Sample B | General validity, measurement invariance, or permission to search Sample C repeatedly for a revision that survives |
| **Known SLE generating moments** | Chapter 4 discloses them only after the Sample B and Sample C sequence is complete, then computes the population optimum and the population discrepancies directly | Contemporaneous justification for an earlier choice, or a finite-sample conclusion from one realization beyond what the audit separates |
| **Separate donor microcases** | Compact demonstrations of a specified calculation or model comparison, declared from their own population and seed | The development or confirmation status of the SLE sequence, which they never join |

: *Records keep distinct analytical roles even when they illustrate the same model family.* {#tbl-part-two-evidence tbl-colwidths="[24,38,38]"}

The Student Learning Experiences (SLE) records are a synthetic course case:
continuous simulated data, not observations from actual students and not a
validated instrument. Three independent samples of 360 complete cases carry
twelve indicators -- four Planning, four Strategy and four Belonging -- drawn
from one declared population and distributed together in
`sle_continuous.csv` beside the chapters. Each analysis uses only the sample
its role entitles it to, Chapter 1 uses none of them, and the Chapter 2
appendix supplies the codebook, the sample manifest and the generating
population, so that every reported value can be reproduced. The retrospective
audit that closes Chapter 4 is possible only because that population is
known, and it is disclosed only after the Sample B and Sample C sequence has
run its course: first the evidence a researcher could actually have had, then
the generating structure the candidate omitted. It adds no further sample, no
new candidate, no successful revision and no after-the-fact justification for
a choice made earlier; its value is that a reader watches a model stay
coherent, reproducible and apparently well fitting while still owing a
narrower conclusion than its name suggests -- which no real study can show.

The stopping points are as deliberate as the moves. Chapter 2 does not
relabel a developed pattern as confirmation. Chapter 3 does not treat
convergence or admissibility as adequate fit, and says so in its own section
title: it reads an estimated model *before its fit*. Chapter 4 does not treat
favorable fit as construct validation. Each chapter completes one
responsibility and transfers a defined record, rather than carrying an
unexamined conclusion forward under a new name.

## How to Read a Chapter and Use the Concept Reference

Each chapter opens in the same way. A **Chapter overview** box summarizes the
whole chapter in ordinary words: the route, and what a reader will be able to
do at the end. The prose that follows it begins from a real measurement
problem or concern in an applied study and states the question the chapter
answers. Each chapter closes with a numbered **Summary and reading bridge**
that gathers what has been settled and states the questions the next chapter
owes. A short **Key terms** table follows where a chapter carries one, then
exercises, then any appendices, which hold derivations, data documentation
and extensions the main route does not require. A term is explained where it
is first used, and the chapters carry the statistics themselves: a model
appears as equations, as parameter matrices, as a path diagram and as an
explanation, and every reported number is computed for the reader and read in
prose and in named tables. Results are reported to four decimals from
full-precision computation unless a source's own precision applies.

The software is named where the methods need it rather than taught as a
course of its own. Chapter 1 asks for none: its records are constructed, its
results exact, and its conclusions readable without a computer. `psych` and
`GPArotation` enter with the exploratory work of Chapter 2, and `lavaan`
enters with the confirmatory fits of Chapter 3. A chapter names the software
it relies on and states the consequential defaults in words -- scale setting,
factor covariances, mean structures, missing-data handling, standard errors
and the test-statistic family -- so that a reader who never opens R can still
follow every model, every count and every conclusion from the printed
evidence.

Part One's argument vocabulary is reused rather than redefined, and Part Two
does not re-teach the argument-based approach as a recurring apparatus. One
optional section of Chapter 1 shows how evidence, justification,
interpretation and claim boundaries change when a statistical variable is
latent, applies the six quality-evaluation questions under Part One's names,
records at least one failed verdict and repairs the claim; a reader may skip
it without losing a definition, an equation, an exercise or a prerequisite
for the chapters that follow. Chapters 2 to 4
express prespecification, uncertainty, limitations and claim boundaries in the
ordinary technical vocabulary of exploratory and confirmatory factor analysis,
and *analytic status* is the one Part Two term that carries the development,
confirmation and replication rules set out above.

The [*Part Two Concept Reference*](https://jinsong-chen.github.io/QMSBR/part-two/part_two_concept_reference.html) publishes beside these chapters and holds the
single vocabulary the four of them share. Its four concept tables gather
related terms into broad groups and give each term's meaning or essential
distinction, and a closing table records the notation; each row names the
chapter and section where the term is principally explained, so a term met in
one chapter can be traced to the place that teaches it. It is written for
lookup rather than for reading through: consult it when a term or a symbol is
unfamiliar, and return to the chapter for the explanation, the worked example
and the reason behind the distinction.

## What Comes Later

The chapters after Chapter 4 continue Part Two from relations between factors
and indicators to relations among variables in a structural system. Their
subject is observed-variable path analysis; general structural equation
modeling; measurement invariance and multigroup measurement; MIMIC models;
latent regression and indirect effects; latent growth in a structural
equation form; and structural model comparison, ordered by their research
questions rather than by a software menu. Multilevel modeling, including
multilevel growth, belongs to Part Three -- a division of the curriculum, not
a claim that structural and multilevel parameterizations stop being
equivalent when observations are unbalanced or incomplete. Item-response
theory, mixture models, multilevel factor analysis and a full treatment of
categorical indicators lie outside Part Two altogether.

The measurement sequence ends where a reader can read and evaluate a
measurement representation, and structural work begins where a new question
requires relations among variables or among factors. An acceptable
measurement model is necessary evidence for many latent-variable structural
analyses, but a measurement portion plus a structural portion confers no
substantive warrant: structural equation modeling does not upgrade an
associational design to a causal one, does not erase measurement uncertainty,
and does not make an indirect effect causal by calling it mediation. Before
latent means or structural parameters are compared across groups, settings or
occasions, the relevant level of measurement comparability is addressed with
the loading, intercept and other restrictions it requires; and a growth model
addresses both mean trajectories and the variation and covariation in
trajectories.

Chapter 4 therefore leaves four questions to that continuation: which latent
variables are endogenous; which directional restrictions enter; whether the
measurement representation is stable for the populations, groups, occasions
and comparisons a structural question requires; and how measurement
uncertainty is carried forward rather than replaced by exported scores. The
bridge out of Chapter 4 is written with those questions in view, so that a
latent intercept is not mistaken for the marginal latent mean once structural
regressions enter.
